### 2023-09-24 tmp/gm_addition_to_cv.Rmd
GM addition to cv vignette

---
title: "Cross-validation of regression models: mixed model examples"
author: "John Fox and Georges Monette"
date: "`r Sys.Date()`"
package: cv
output: 
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["cv.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Cross-validation of regression models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.align = "center",
  fig.height = 6,
  fig.width = 7,
  fig.path = "fig/",
  dev = "png",
  comment = "#>" #,
  # eval = nzchar(Sys.getenv("REBUILD_VIGNETTES"))
)

# save some typing
knitr::set_alias(w = "fig.width",
                 h = "fig.height",
                 cap = "fig.cap")

# colorize text: use inline as `r colorize(text, color)`
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}

.opts <- options(digits = 5)
```

```{r packages, include=FALSE}
library(magrittr)
library(bibtex)
library(kableExtra)
library(latex2exp)
library(cv)
library(car)
library(lattice)
library(latticeExtra)
layer <- latticeExtra::layer  # to mask ggplot2::layer
library(lme4)
library(nlme)
knitr::opts_chunk$set(comment='    ')
```
The following artificial data set exemplifies a difference between cluster-based
and individual-based cross-validation in hierarchical models.

It also shows how, in contrast with ordinary least-squares
models, the variance at different levels of a hierarchical
model need not necessarily become smaller as the model becomes larger.

In this example, the simplest random intercept model results in a between-cluster variance
that is much smaller than that of a model that also includes $x$ as a predictor.
Consequently, different methods of assessing model fit may result in
different best models, underscoring the importance of assessing hierarchical models
at different levels.

Adjusting the parameters generating this data can create an example 
that illustrates problems of non-convergence with hierarchical models.
Fitted models with relatively small between-cluster variance, even simple
random intercept models, are prone to non-convergence due to the 
nature of optimizing algorithms used in `lme` and `lmer`.

In some contexts, non-convergence,
can be viewed as a diagnostic of a poor model. However, with a simple 
random intercept model where it may indicate a small between-cluster
variance, non-convergence is an annoyance since it would be statistically correct to
report that the data are consistent with a zero between-cluster population variance,
i.e. that observed between-cluster variance in the data can be explained by 
the between-individual variability within clusters.  This leads to perfectly
valid statistical inferences. Although the calculation of accurate p-values for hypotheses concerning
random effects may be complicated technically
by the fact that the estimated variance is at the boundary of the parameter space,
the estimated variance-covariance matrix of fixed-effect parameters will generally
be non-singular and may be used for valid Wald tests of linear hypotheses 
of fixed-effect parameters.

Using cross-validation based on fitting random subsets compounds the problem
since subsets on which there is non-convergence cause the cross-validation to
fail unless non-converging subsets are omitted, which may bias results since non-convergence
is likely to be related to the estimated between-cluster variances.

By varying the parameter for between-cluster variance, we can use the approach
below to generate data sets with varying propensities to produce
non-convergence in order to explore the consequences of non-convergence and to 
compare the performance of different optimizers. **to do**

At present, the results below are based on a data set that does not
produce non-convergence and are limited to comparisons obtained
by varying three factors:

- Three models: random intercept only, adding x as a predictor, and also adding the group means of x as a predictor,
- The method used for fitting: `lmer` in `lme4` or `lme` in `nlme`,
- Cross-validation resampling individual cases or resampling clusters.

Note that since the independence of clusters is a reasonable assumption
while the independence of individual observations is not, 
cluster-based cross-validation has a stronger theoretical justification.

We start by generating and plotting the data.

```{r data}
{
  # Generate data set

  # Parameters

  set.seed(123)
  Nb <- 10       # number of groups
  Nw <- 5        # number of individuals within groups
  Bb <- 0        # between group regression coefficient on group mean
  SDre <- 2.0    # between group SD of random level relative to group mean of X

  SDwithin <- 0.5    # within group SD

  Bw <- 1       # within group effect of X

  dd <- data.frame(
    group = factor(rep(1:Nb, each = Nw)),
    x = 20 + seq(1, Nb*Nw)
  ) %>% 
    within(
      {
        xm  <- ave(x, group, FUN = mean) # withing group mean
        y <- 10 +
          Bb * xm +                  # contextual effect
          Bw * (x - xm) +            # within-group effect
          SDre * rnorm(Nb)[group] +  # random level by group
          SDwithin * rnorm(Nb*Nw)    # random error within groups
      }
    )
}
```

```{r plot1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Hierarchical data set with 10 clusters"
pl <- xyplot(y ~ x, dd, groups = group,
             ylim = c(0,20),
             par.settings = list(superpose.symbol = list(pch = 16, cex = 1))) + 
    latticeExtra::layer(panel.ellipse(...,center.cex =0))
pl
pl <- update(pl, par.settings = list(superpose.symbol = list(pch = 1, cex = .7)))
```

The Y variable could represent scores on a programming aptitude test of 10 students
of varying ages, each tested
annually during a 5-year training program.  The X variable is the age
at which each test occurred.

We next obtain six fits with three models and two fitting functions.
```{r}
# random intercept models

fit0 <- lmer(y ~ 1 + (1|group), dd)
fit0l <- lme(y ~ 1, dd, random = ~ 1|group)

# effect of x and random intercept

fit1 <- lmer(y ~ x + (1|group), dd)
fit1l <- lme(y ~ x, dd, random = ~ 1|group)

# effect of x, mean of x and random intercept

fit2 <- lmer(y ~ x + xm + (1|group), dd)
fit2l <- lme(y ~ x + xm, dd, random = ~1|group)

modlist <- list(fit0, fit0l, fit1, fit1l, fit2, fit2l)
names(modlist) <- sapply(modlist, formula)
```
We can compare estimated fixed-effect coefficients.

```{r, message=FALSE,warning=FALSE}
tbl <- do.call(compareCoefs, modlist) 
tbl[] <- round(tbl,4)
tbl[is.na(tbl)] <- ''
kbl(tbl, align = 'r', caption = "Comparison of estimated fixed effects")
```

With this data set, the two functions `lme` and `lmer` converged for all models and produced consistent results. 
The differences between these two functions, and differences that depend on the optimizer used
within each function, would manifest themselves if the data were generated with a smaller
between-cluster variance.

We visualize predicted values using population predictions with the fixed effects model only 
and using BLUPs for random effects.  Since `lme` and `lmer` produced almost identical results, the
figures below are produced only with the `lmer` results.

```{r}
dd$fit_mod0__lmer.fe <- predict(fit0, re.form = ~ 0)
dd$fit_mod0__lmer.re <- predict(fit0)
dd$fit_mod0__lme.fe <- predict(fit0l, level = 0)
dd$fit_mod0__lme.re <- predict(fit0l, level = 1)

dd$fit_mod1__lmer.fe <- predict(fit1, re.form = ~ 0)
dd$fit_mod1__lmer.re <- predict(fit1)
dd$fit_mod1__lme.fe <- predict(fit1l, level = 0)
dd$fit_mod1__lme.re <- predict(fit1l, level = 1)

dd$fit_mod2__lmer.fe <- predict(fit2, re.form = ~ 0)
dd$fit_mod2__lmer.re <- predict(fit2)
dd$fit_mod2__lme.fe <- predict(fit2l, level = 0)
dd$fit_mod2__lme.re <- predict(fit2l, level = 1)

dl <- stats::reshape(dd, direction = 'long', sep = '.', timevar = 'effect', varying = grep('\\.', names(dd)))
dl$id <- 1:nrow(dl)
dl <- stats::reshape(dl, direction = 'long', sep = '__', timevar = 'method',  varying = grep('__', names(dl)))
dl$id <- 1:nrow(dl)
dl <- stats::reshape(dl, direction = 'long', sep = '_', timevar = 'modelcode',  varying = grep('_', names(dl)))

dl$model <- factor(c('~ 1', '~ 1 + x', '~ 1 + x + xm')[match(dl$modelcode, c('mod0','mod1','mod2'))])

```

```{r plot-fits-mod0}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values using lmer with random intercept model"

(pl +
  xyplot(fit ~ x, subset(dl, modelcode == 'mod0' & effect == 'fe' & method == 'lmer'), groups = group, type = 'l', lwd =2) + 
  xyplot(fit ~ x, subset(dl, modelcode == 'mod0' & effect == 're' & method == 'lmer'), groups = group, type = 'l', lwd =2, lty=3) 
) %>% update(
  key=list(
    space = 'right', 
    text = list(c('fixed effects only','fixed and random')),
    lines = list(lty=c(1,3))))

```
```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values using lmer with random intercept model and x as a predictor"
(pl +
  xyplot(fit ~ x, subset(dl, modelcode == 'mod1' & effect == 'fe' & method == 'lmer'), groups = group, type = 'l', lwd =2) + 
  xyplot(fit ~ x, subset(dl, modelcode == 'mod1' & effect == 're' & method == 'lmer'), groups = group, type = 'l', lwd =2, lty=3) 
) %>% update(
  ylim = c(-20,40),
  key=list(
    space = 'right', 
    text = list(c('fixed effects only','fixed and random')),
    lines = list(lty=c(1,3))))
```
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fitted values using lmer with random intercept model and x and the group mean of x as predictors"
(pl +
  xyplot(fit ~ x, subset(dl, modelcode == 'mod2' & effect == 'fe' & method == 'lmer'), groups = group, type = 'l', lwd =2) + 
  xyplot(fit ~ x, subset(dl, modelcode == 'mod2' & effect == 're' & method == 'lmer'), groups = group, type = 'l', lwd =2, lty=3) 
) %>% update(
  ylim = c(0,20),
  key=list(
    space = 'right', 
    text = list(c('fixed effects only','fixed and random')),
    lines = list(lty=c(1,3))))
```

### Cross-validation of models

```{r cross-validation}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Cross-validation using lmer with random intercept model and x and the group mean of x as predictors"

cvs_cases <- lapply(modlist, cv) %>% lapply(rbind) %>% do.call(rbind,.)
cvs_clusters <- lapply(modlist, cv, cluster = 'group') %>% lapply(rbind) %>% do.call(rbind,.)

dout <- as.data.frame( rbind( cvs_cases, cvs_clusters))
# nrow(dout)
dout$cv_method <- rep(c('cases','clusters'), each = nrow(dout)/2)
dout$model <- rep(names(modlist), 2)
# dout$model <- factor(rep(c('~ 1','~ x','~ x + xm'), nrow(dout)/3))
dout$method <- rep(c('lmer','lme'), nrow(dout)/2)
dout$CV <- as.numeric(as.character(dout$`CV crit`))
dout$model <- factor(sub(' \\+ .1 .*$','', dout$model))

dout <- dout[order(dout$model),]
head(dout)
xyplot(CV ~ model , subset(dout, method == 'lmer'), groups = cv_method,
       par.settings = list(superpose.line = list(lty=c(2,3),lwd = 3)),
       type = 'l',
       auto.key = T,
       scales = list(x = list(alternating = F), y = list(log=T)))

```

### 2023-10-05 R/cv.modList.R
graph range rather than SD of reps

@@ -147,7 +147,7 @@ plot.cvModList <- function(x, y,
   if (missing(ylab)){
     ylab <- if (inherits(x[[1L]], "cvList")){
       paste("Cross-Validated", x[[1L]][[1L]]$criterion,
-            "(Averaged  +/- SD)")
+            "(Average and Range)")
     } else {
       paste("Cross-Validated", x[[1L]]$criterion)
     }
@@ -179,10 +179,13 @@ plot.cvModList <- function(x, y,
     on.exit(par(save.mai))
   }
   if (inherits(x[[1L]], "cvList")){
-    sd <- paste("SD", y)
+    # sd <- paste("SD", y)
+    ynm <- paste(y, "range")
     sumry <- lapply(x, summarizeReps)
-    min.y <- sapply(sumry, function(x) x[[y]] - x[[sd]])
-    max.y <- sapply(sumry, function(x) x[[y]] + x[[sd]])
+    # min.y <- sapply(sumry, function(x) x[[y]] - x[[sd]])
+    # max.y <- sapply(sumry, function(x) x[[y]] + x[[sd]])
+    min.y <- sapply(sumry, function(x) x[[ynm]][1])
+    max.y <- sapply(sumry, function(x) x[[ynm]][2])
     plot(c(1L, length(x)), c(min(min.y), max(max.y)),
          xlab=xlab, ylab=ylab,
          main=main, axes=FALSE, type="n")

### 2023-10-09 scratch/gm_addition_to_cv.Rmd
edits to gm_addition_to_cv.Rmd

--- 2023-09-24 tmp/gm_addition_to_cv.Rmd
@@
 
-Adjusting the parameters generating this data can create an example 
+Adjusting the parameters generating this data can be used to create examples
 that illustrates problems of non-convergence with hierarchical models.
-Fitted models with relatively small between-cluster variance, even simple
+Fitted models with relatively small between-cluster variance, even those using simple
 random intercept models, are prone to non-convergence due to the
@@
 random intercept model where it may indicate a small between-cluster
-variance, non-convergence is an annoyance since it would be statistically correct to
+variance, non-convergence is an annoyance because it would be statistically correct to
 report that the data are consistent with a zero between-cluster population variance,
 i.e. that observed between-cluster variance in the data can be explained by
-the between-individual variability within clusters.  This leads to perfectly
+the between-individual variability within clusters.  This leads to
 valid statistical inferences. Although the calculation of accurate p-values for hypotheses concerning
-random effects may be complicated technically
+random effects may be complicated
 by the fact that the estimated variance is at the boundary of the parameter space,
@@
 Using cross-validation based on fitting random subsets compounds the problem
-since subsets on which there is non-convergence cause the cross-validation to
+since subsets for which there is non-convergence cause the cross-validation to
 fail unless non-converging subsets are omitted, which may bias results since non-convergence
@@
 
-By varying the parameter for between-cluster variance, we can use the approach
-below to generate data sets with varying propensities to produce
-non-convergence in order to explore the consequences of non-convergence and to 
-compare the performance of different optimizers. **to do**
+Varying the parameter for between-cluster variance (`SDre` in the code below),
+will generate data sets with varying propensities to produce
+non-convergence. These data sets can be used to explore the consequences of non-convergence and to
+compare the performance of different optimizers.
 
@@
 
-Note that since the independence of clusters is a reasonable assumption
-while the independence of individual observations is not, 
-cluster-based cross-validation has a stronger theoretical justification.
+Since the independence of clusters may be a reasonable assumption, as is the
+the independence of individual observations from the same cluster *conditional*
+on clusters, each form of resampling addresses the
+suitability of models
+to answer different questions.
+Results based on individual resampling generalize to new samples from the same
+clusters. Resampling clusters generalizes to new samples of clusters and
+cases.
 
@@
   Nw <- 5        # number of individuals within groups
-  Bb <- 0        # between group regression coefficient on group mean
-  SDre <- 2.0    # between group SD of random level relative to group mean of X
+  Bb <- 0        # between-group regression coefficient on group mean
+  SDre <- 2.0    # between-group SD of random level relative to group mean of X
 
@@
       {
-        xm  <- ave(x, group, FUN = mean) # withing group mean
+        xm  <- ave(x, group, FUN = mean) # within-group mean
         y <- 10 +
@@
 ```{r}
-# random intercept models
+# random intercept only models
 
@@
 modlist <- list(fit0, fit0l, fit1, fit1l, fit2, fit2l)
-names(modlist) <- sapply(modlist, formula)
+names(modlist) <- c('lmer:~1','lme:~1','lmer:~x','lme:~x','lmer:~x+xm','lme:~x+xm')
+modlist <- do.call(models, modlist)
 ```
@@
 tbl <- do.call(compareCoefs, modlist)
+colnames(tbl) <- names(modlist)
 tbl[] <- round(tbl,4)
@@
 
+We carry out cross-validation resampling individuals within clusters, based on
+fixed and predicted random effects, and resampling clusters, based on
+fixed effects only.
+In order to reduce between-model variability in comparisons of models,
+the `cv` method applied to the list of models created by the `models` function,
+performs cross-validation with the same training and validation sets
+for each model in order to reduce the between-model variability in the cross-validation
+criterion that stems from the random variation in training set selection.
+
 ```{r cross-validation}
@@
 #| fig.height = 6,
-#| fig.cap = "Cross-validation using lmer with random intercept model and x and the group mean of x as predictors"
+#| fig.cap = "Cross-validation using lmer and lme comparing random intercept models with varying fixed effects and with case- and cluster-based resampling."
 
-cvs_cases <- lapply(modlist, cv) %>% lapply(rbind) %>% do.call(rbind,.)
-cvs_clusters <- lapply(modlist, cv, cluster = 'group') %>% lapply(rbind) %>% do.call(rbind,.)
+cvs_cases <- cv(modlist, data = dd)
+cvs_clusters <- cv(modlist, data = dd, cluster = 'group')
 
-dout <- as.data.frame( rbind( cvs_cases, cvs_clusters))
-# nrow(dout)
-dout$cv_method <- rep(c('cases','clusters'), each = nrow(dout)/2)
-dout$model <- rep(names(modlist), 2)
-# dout$model <- factor(rep(c('~ 1','~ x','~ x + xm'), nrow(dout)/3))
-dout$method <- rep(c('lmer','lme'), nrow(dout)/2)
-dout$CV <- as.numeric(as.character(dout$`CV crit`))
-dout$model <- factor(sub(' \\+ .1 .*$','', dout$model))
+cvs_cases <- cvs_cases %>% lapply(rbind) %>% do.call(rbind,.) %>% as.data.frame
+cvs_clusters <- cvs_clusters %>% lapply(rbind) %>% do.call(rbind,.) %>% as.data.frame
+
+cvs_cases$modlist <- names(modlist)
+cvs_clusters$modlist <- names(modlist)
+cvs_cases$cv_method <- 'cases'
+cvs_clusters$cv_method <- 'clusters'
+
+dout <-
+  within(
+    as.data.frame( rbind( cvs_cases, cvs_clusters)),
+    {
+      model <- factor(sub('lme.*:','',modlist))
+      method <- factor(sub(':.*$','',modlist))
+      CV <- as.numeric(as.character(`CV crit`))
+      cv_method <- factor(cv_method)
+    }
+)
 
 dout <- dout[order(dout$model),]
-head(dout)
-xyplot(CV ~ model , subset(dout, method == 'lmer'), groups = cv_method,
-       par.settings = list(superpose.line = list(lty=c(2,3),lwd = 3)),
-       type = 'l',
+
+xyplot(CV ~ model , dout, groups = cv_method:method,
+       par.settings =
+         list(superpose.line = list(lty=c(2,3),lwd = 3),superpose.symbol=list(pch=21:24, fill='transparent', cex =1)),
+       type = 'b',
        auto.key = T,
-       scales = list(x = list(alternating = F), y = list(log=T)))
+       scales = list(y = list(log=T)))
 
 ```
+
+With this example, `lme` and `lmer` give similar results, but resampling cases versus resampling clusters provides
+very different comparisons of models.  The model with X alone, without the contextual mean of X, is assessed as
+fitting very poorly sampling clusters and, relatively, much better resampling individual cases.  This illustrates the
+need to perform cross-validation at different levels.
+
+B
+B
+B
+B
+B
+B
+B
+B
+B
+B
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A
+A

### 2023-10-10 scratch/gm_addition_to_cv-edited.Rmd
add edited version of mixed-models vignette section

--- 2023-10-09 scratch/gm_addition_to_cv.Rmd
@@
 ---
-title: "Cross-validation of regression models: mixed model examples"
+title: "Cross-validation of regression models: mixed model examples (edited)"
 author: "John Fox and Georges Monette"
@@
 .opts <- options(digits = 5)
+
+knitr::opts_chunk$set(comment="    ")
 ```
@@
 ```{r packages, include=FALSE}
-library(magrittr)
-library(bibtex)
-library(kableExtra)
-library(latex2exp)
 library(cv)
@@
 library(latticeExtra)
-layer <- latticeExtra::layer  # to mask ggplot2::layer
 library(lme4)
-library(nlme)
-knitr::opts_chunk$set(comment='    ')
 ```
@@
 ```{r data}
-{
 # Generate data set
 
-  # Parameters
-
-  set.seed(123)
+# Parameters:
+set.seed(123) # set.seed(9693)
 Nb <- 10     # number of groups
@@
 Bb <- 0      # between-group regression coefficient on group mean
-  SDre <- 2.0    # between-group SD of random level relative to group mean of X
+SDre <- 2.0  # between-group SD of random level relative to group mean of x
+SDwithin <- 0.5  # within group SD
+Bw <- 1  # within group effect of x
 
-  SDwithin <- 0.5    # within group SD
-
-  Bw <- 1       # within group effect of X
-
-  dd <- data.frame(
-    group = factor(rep(1:Nb, each = Nw)),
+Data <- data.frame(
+  group = factor(rep(1:Nb, each=Nw)),
   x = 20 + seq(1, Nb*Nw)
-  ) %>%
+) |>
   within(
@@
   )
-}
 ```
@@
 #| fig.cap = "Hierarchical data set with 10 clusters"
-pl <- xyplot(y ~ x, dd, groups = group,
-             ylim = c(0,20),
-             par.settings = list(superpose.symbol = list(pch = 16, cex = 1))) +
-    latticeExtra::layer(panel.ellipse(...,center.cex =0))
-pl
-pl <- update(pl, par.settings = list(superpose.symbol = list(pch = 1, cex = .7)))
+plot <- xyplot(y ~ x, data=Data, group=group,
+               ylim=c(0, 20),
+               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
+    layer(panel.ellipse(..., center.cex=0))
+plot
 ```
@@
 ```{r}
-# random intercept only models
+# random intercept only model:
+fit0 <- lmer(y ~ 1 + (1|group), Data)
 
-fit0 <- lmer(y ~ 1 + (1|group), dd)
-fit0l <- lme(y ~ 1, dd, random = ~ 1|group)
+# effect of x and random intercept:
+fit1 <- lmer(y ~ x + (1|group), Data)
 
-# effect of x and random intercept
-
-fit1 <- lmer(y ~ x + (1|group), dd)
-fit1l <- lme(y ~ x, dd, random = ~ 1|group)
-
-# effect of x, mean of x and random intercept
-
-fit2 <- lmer(y ~ x + xm + (1|group), dd)
-fit2l <- lme(y ~ x + xm, dd, random = ~1|group)
-
-modlist <- list(fit0, fit0l, fit1, fit1l, fit2, fit2l)
-names(modlist) <- c('lmer:~1','lme:~1','lmer:~x','lme:~x','lmer:~x+xm','lme:~x+xm')
-modlist <- do.call(models, modlist)
+# effect of x, mean of x, and random intercept:
+fit2 <- lmer(y ~ x + xm + (1|group), Data)
 ```
@@
 ```{r, message=FALSE,warning=FALSE}
-tbl <- do.call(compareCoefs, modlist)
-colnames(tbl) <- names(modlist)
-tbl[] <- round(tbl,4)
-tbl[is.na(tbl)] <- ''
-kbl(tbl, align = 'r', caption = "Comparison of estimated fixed effects")
+compareCoefs(fit0,  fit1, fit2)
 ```
@@
 ```{r}
-dd$fit_mod0__lmer.fe <- predict(fit0, re.form = ~ 0)
-dd$fit_mod0__lmer.re <- predict(fit0)
-dd$fit_mod0__lme.fe <- predict(fit0l, level = 0)
-dd$fit_mod0__lme.re <- predict(fit0l, level = 1)
+Data <- within(Data, {
+  fit_mod0.fe <- predict(fit0, re.form = ~ 0)
+  fit_mod0.re <- predict(fit0)
+  fit_mod1.fe <- predict(fit1, re.form = ~ 0)
+  fit_mod1.re <- predict(fit1)
+  fit_mod2.fe <- predict(fit2, re.form = ~ 0)
+  fit_mod2.re <- predict(fit2)
+})
 
-dd$fit_mod1__lmer.fe <- predict(fit1, re.form = ~ 0)
-dd$fit_mod1__lmer.re <- predict(fit1)
-dd$fit_mod1__lme.fe <- predict(fit1l, level = 0)
-dd$fit_mod1__lme.re <- predict(fit1l, level = 1)
+Data_long <- reshape(Data, direction = "long", sep = ".", 
+              timevar = "effect", varying = grep("\\.", names(Data)))
+Data_long$id <- 1:nrow(Data_long)
+Data_long <- reshape(Data_long, direction = "long", sep = "_", 
+              timevar = "modelcode",  varying = grep("_", names(Data_long)))
 
-dd$fit_mod2__lmer.fe <- predict(fit2, re.form = ~ 0)
-dd$fit_mod2__lmer.re <- predict(fit2)
-dd$fit_mod2__lme.fe <- predict(fit2l, level = 0)
-dd$fit_mod2__lme.re <- predict(fit2l, level = 1)
-
-dl <- stats::reshape(dd, direction = 'long', sep = '.', timevar = 'effect', varying = grep('\\.', names(dd)))
-dl$id <- 1:nrow(dl)
-dl <- stats::reshape(dl, direction = 'long', sep = '__', timevar = 'method',  varying = grep('__', names(dl)))
-dl$id <- 1:nrow(dl)
-dl <- stats::reshape(dl, direction = 'long', sep = '_', timevar = 'modelcode',  varying = grep('_', names(dl)))
-
-dl$model <- factor(c('~ 1', '~ 1 + x', '~ 1 + x + xm')[match(dl$modelcode, c('mod0','mod1','mod2'))])
-
+Data_long$model <- factor(c("~ 1", "~ 1 + x", "~ 1 + x + xm")
+                   [match(Data_long$modelcode, c("mod0","mod1","mod2"))])
 ```
@@
 #| fig.height = 6,
-#| fig.cap = "Fitted values using lmer with random intercept model"
-
-(pl +
-  xyplot(fit ~ x, subset(dl, modelcode == 'mod0' & effect == 'fe' & method == 'lmer'), groups = group, type = 'l', lwd =2) +
-  xyplot(fit ~ x, subset(dl, modelcode == 'mod0' & effect == 're' & method == 'lmer'), groups = group, type = 'l', lwd =2, lty=3)
-) %>% update(
+#| fig.cap = "Fitted values with random intercept model"
+(plot +
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & 
+                         effect == "fe"),
+         groups=group, type="l", lwd=2) +
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & 
+                         effect == "re"),
+         groups=group, type="l", lwd=2, lty=3)
+) |> update(
   key=list(
-    space = 'right',
-    text = list(c('fixed effects only','fixed and random')),
-    lines = list(lty=c(1,3))))
-
+    space="right",
+    text=list(c("fixed effects only","fixed and random")),
+    lines=list(lty=c(1, 3))))
 ```
@@
 #| fig.height = 6,
-#| fig.cap = "Fitted values using lmer with random intercept model and x as a predictor"
-(pl +
-  xyplot(fit ~ x, subset(dl, modelcode == 'mod1' & effect == 'fe' & method == 'lmer'), groups = group, type = 'l', lwd =2) +
-  xyplot(fit ~ x, subset(dl, modelcode == 'mod1' & effect == 're' & method == 'lmer'), groups = group, type = 'l', lwd =2, lty=3)
-) %>% update(
-  ylim = c(-20,40),
+#| fig.cap = "Fitted values with random intercept model and x as a predictor"
+(plot +
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & 
+                         effect == "fe"), 
+         groups=group, type="l", lwd=2) +
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & 
+                         effect == "re"), 
+         groups=group, type="l", lwd=2, lty=3)
+) |> update(
+  ylim=c(-20, 40),
   key=list(
-    space = 'right',
-    text = list(c('fixed effects only','fixed and random')),
-    lines = list(lty=c(1,3))))
+    space="right",
+    text=list(c("fixed effects only","fixed and random")),
+    lines=list(lty=c(1, 3))))
 ```
@@
 #| fig.height = 6,
-#| fig.cap = "Fitted values using lmer with random intercept model and x and the group mean of x as predictors"
-(pl +
-  xyplot(fit ~ x, subset(dl, modelcode == 'mod2' & effect == 'fe' & method == 'lmer'), groups = group, type = 'l', lwd =2) +
-  xyplot(fit ~ x, subset(dl, modelcode == 'mod2' & effect == 're' & method == 'lmer'), groups = group, type = 'l', lwd =2, lty=3)
-) %>% update(
-  ylim = c(0,20),
+#| fig.cap = "Fitted values with random intercept model and x and the group mean of x as predictors"
+(plot +
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" &
+
+                           effect == "fe"), 
+         groups=group, type="l", lwd=2) +
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & 
+                         effect == "re"), 
+         groups=group, type="l", lwd=2, lty=3)
+) |> update(
+  ylim = c(0, 20),
   key=list(
-    space = 'right',
-    text = list(c('fixed effects only','fixed and random')),
-    lines = list(lty=c(1,3))))
+    space="right",
+    text=list(c("fixed effects only","fixed and random")),
+    lines=list(lty=c(1, 3))))
 ```
@@
 #| fig.height = 6,
-#| fig.cap = "Cross-validation using lmer and lme comparing random intercept models with varying fixed effects and with case- and cluster-based resampling."
+#| fig.cap = "Cross-validation comparing random intercept models with varying fixed effects and with case- and cluster-based resampling."
+modlist <- models("~ 1"=fit0, "~ x"=fit1, "~ x + xm"=fit2)
+cvs_cases <- cv(modlist, data=Data, seed=9693)
+cvs_clusters <- cv(modlist, data=Data, cluster="group", seed=4576)
 
-cvs_cases <- cv(modlist, data = dd)
-cvs_clusters <- cv(modlist, data = dd, cluster = 'group')
-
-cvs_cases <- cvs_cases %>% lapply(rbind) %>% do.call(rbind,.) %>% as.data.frame
-cvs_clusters <- cvs_clusters %>% lapply(rbind) %>% do.call(rbind,.) %>% as.data.frame
+cvs_cases <- cvs_cases |> lapply(rbind) |> 
+  do.call(rbind, args = _) |> as.data.frame()
+cvs_clusters <- cvs_clusters |> lapply(rbind) |> 
+  do.call(rbind, args = _) |> as.data.frame()
 
@@
 cvs_clusters$modlist <- names(modlist)
-cvs_cases$cv_method <- 'cases'
-cvs_clusters$cv_method <- 'clusters'
+cvs_cases$cv_method <- "cases"
+cvs_clusters$cv_method <- "clusters"
 
-dout <-
-  within(
-    as.data.frame( rbind( cvs_cases, cvs_clusters)),
+dout <- within(
+  as.data.frame(rbind(cvs_cases, cvs_clusters)),
   {
-      model <- factor(sub('lme.*:','',modlist))
-      method <- factor(sub(':.*$','',modlist))
+    model <- factor(modlist)
     CV <- as.numeric(as.character(`CV crit`))
-      cv_method <- factor(cv_method)
   }
@@
 
-dout <- dout[order(dout$model),]
+dout <- dout[order(dout$model), ]
 
-xyplot(CV ~ model , dout, groups = cv_method:method,
+xyplot(CV ~ model, data=dout, groups=cv_method,
        par.settings =
-         list(superpose.line = list(lty=c(2,3),lwd = 3),superpose.symbol=list(pch=21:24, fill='transparent', cex =1)),
-       type = 'b',
-       auto.key = T,
-       scales = list(y = list(log=T)))
-
+         list(superpose.line=list(lty=c(2, 3), lwd=3), 
+              superpose.symbol=list(pch=15:16, 
+                                    cex=1.5)),
+       type="b",
+       auto.key=TRUE,
+       scales=list(y=list(log=TRUE)))
 ```
@@
 need to perform cross-validation at different levels.
-
-B
-B
-B
-B
-B
-B
-B
-B
-B
-B
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A
-A

### 2023-10-11 scratch/gm_addition_to_cv-edited.Rmd
add errors/warnings/messages for bad reps and seed args to cv() methods

@@
 cvs_cases <- cv(modlist, data=Data, seed=9693)
-cvs_clusters <- cv(modlist, data=Data, cluster="group", seed=4576)
+cvs_clusters <- cv(modlist, data=Data, cluster="group")
 

### 2023-10-11 scratch/gm_addition_to_cv-edited.Rmd
increase n of clusters to 100 in mixed models addendum to vignette

@@
 # Parameters:
-set.seed(123) # set.seed(9693)
-Nb <- 10     # number of groups
+set.seed(9693) # set.seed(123) 
+Nb <- 100     # number of groups
 Nw <- 5      # number of individuals within groups
@@
 #| fig.cap = "Hierarchical data set with 10 clusters"
-plot <- xyplot(y ~ x, data=Data, group=group,
+plot <- xyplot(y ~ x, data=Data[ 1:50,], group=group,
                ylim=c(0, 20),
@@
 
-Data_long <- reshape(Data, direction = "long", sep = ".", 
-              timevar = "effect", varying = grep("\\.", names(Data)))
+Data_long <- reshape(Data[1:50, ], direction = "long", sep = ".", 
+              timevar = "effect", varying = grep("\\.", names(Data.10)))
 Data_long$id <- 1:nrow(Data_long)
@@
 cvs_cases <- cv(modlist, data=Data, seed=9693)
-cvs_clusters <- cv(modlist, data=Data, cluster="group")
+cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
 

### 2023-10-13 scratch/gm_addition_to_cv-edited_v2.Rmd
further code edits for mixed-models vignette section

--- 2023-10-11 scratch/gm_addition_to_cv-edited.Rmd
@@
 ---
-title: "Cross-validation of regression models: mixed model examples (edited)"
+title: "Cross-validation of regression models: mixed model examples (edited v2)"
 author: "John Fox and Georges Monette"
@@
 # Parameters:
-set.seed(123) # set.seed(9693)
-Nb <- 10     # number of groups
+set.seed(9693) 
+Nb <- 100    # number of groups
 Nw <- 5      # number of individuals within groups
@@
 Bw <- 1     # within group effect of x
+Ay <- 10    # intercept for response
+Ax <- 20    # starting level of x
+Nx <- Nw*10 # number of distinct x values
 
@@
   group = factor(rep(1:Nb, each=Nw)),
-  x = 20 + seq(1, Nb*Nw)
+  x = Ax + rep(1:Nx, length.out = Nw*Nb)
 ) |>
@@
       xm  <- ave(x, group, FUN = mean) # within-group mean
-      y <- 10 +
+      y <- Ay +
         Bb * xm +                  # contextual effect
@@
 #| fig.height = 6,
-#| fig.cap = "Hierarchical data set with 10 clusters"
-plot <- xyplot(y ~ x, data=Data, group=group,
-               ylim=c(0, 20),
+#| fig.cap = "Hierarchical data set, showing first 10 of 100 clusters."
+plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
+               ylim=c(4, 16),
                par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
@@
 ```{r}
-# random intercept only model:
-fit0 <- lmer(y ~ 1 + (1|group), Data)
+# random intercept only:
+fit0 <- lmer(y ~ 1 + (1 | group), Data)
 
 # effect of x and random intercept:
-fit1 <- lmer(y ~ x + (1|group), Data)
+fit1 <- lmer(y ~ x + (1 | group), Data)
 
 # effect of x, mean of x, and random intercept:
-fit2 <- lmer(y ~ x + xm + (1|group), Data)
+fit2 <- lmer(y ~ x + xm + (1 | group), Data)
+        # equivalent to y ~ I(x - xm) + xm + (1 | group)
+
+# model generating the data (where Bb = 0)
+fit3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
 ```
@@
 ```{r, message=FALSE,warning=FALSE}
-compareCoefs(fit0,  fit1, fit2)
+compareCoefs(fit0,  fit1, fit2, fit3)
 ```
@@
   fit_mod2.re <- predict(fit2)
+  fit_mod3.fe <- predict(fit3, re.form = ~ 0)
+  fit_mod3.re <- predict(fit3)
 })
 
-Data_long <- reshape(Data, direction = "long", sep = ".", 
-              timevar = "effect", varying = grep("\\.", names(Data)))
+Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
+              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
 Data_long$id <- 1:nrow(Data_long)
@@
 
-Data_long$model <- factor(c("~ 1", "~ 1 + x", "~ 1 + x + xm")
-                   [match(Data_long$modelcode, c("mod0","mod1","mod2"))])
+Data_long$model <- factor(
+  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
+  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
+)
 ```
@@
 #| fig.height = 6,
-#| fig.cap = "Fitted values with random intercept model"
+#| fig.cap = "Fitted values from random intercept model."
 (plot +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & 
-                         effect == "fe"),
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
          groups=group, type="l", lwd=2) +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & 
-                         effect == "re"),
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
          groups=group, type="l", lwd=2, lty=3)
 ) |> update(
+  main="Model: y ~ 1 + (1 | group)",
   key=list(
-    space="right",
+    corner=c(0.05, 0.05),
     text=list(c("fixed effects only","fixed and random")),
@@
 #| fig.height = 6,
-#| fig.cap = "Fitted values with random intercept model and x as a predictor"
+#| fig.cap = "Fitted values from model with random intercept and x as a predictor."
 (plot +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & 
-                         effect == "fe"), 
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
          groups=group, type="l", lwd=2) +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & 
-                         effect == "re"), 
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
          groups=group, type="l", lwd=2, lty=3)
 ) |> update(
-  ylim=c(-20, 40),
+  main="Model: y ~ 1 + x + (1 | group)",
+  ylim=c(-15, 35),
   key=list(
-    space="right",
+    corner=c(0.95, 0.05),
     text=list(c("fixed effects only","fixed and random")),
@@
 #| fig.height = 6,
-#| fig.cap = "Fitted values with random intercept model and x and the group mean of x as predictors"
+#| fig.cap = "Fitted values from model with random intercept, x, and the group mean of x as predictors."
 (plot +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" &
-
-                           effect == "fe"), 
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
          groups=group, type="l", lwd=2) +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & 
-                         effect == "re"), 
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
          groups=group, type="l", lwd=2, lty=3)
 ) |> update(
-  ylim = c(0, 20),
+  main="Model: y ~ 1 + x + xm + (1 | group)",
+  ylim=c(4, 16),
   key=list(
-    space="right",
+    corner=c(0.05, 0.05),
+    text=list(c("fixed effects only","fixed and random")),
+    lines=list(lty=c(1, 3))))
+```
+
+```{r plot-fits-mod3}
+#| out.width = "100%",
+#| fig.height = 6,
+#| fig.cap = "Fitted values from the estimated model generating the data."
+(plot +
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
+         groups=group, type="l", lwd=2) +
+  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
+         groups=group, type="l", lwd=2, lty=3)
+) |> update(
+  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
+  ylim=c(4, 16),
+  key=list(
+    corner=c(0.05, 0.05),
     text=list(c("fixed effects only","fixed and random")),
@@
 
-```{r cross-validation}
+```{r cross-validation-clusters}
 #| out.width = "100%",
 #| fig.height = 6,
-#| fig.cap = "Cross-validation comparing random intercept models with varying fixed effects and with case- and cluster-based resampling."
-modlist <- models("~ 1"=fit0, "~ x"=fit1, "~ x + xm"=fit2)
+#| fig.cap = "Cross-validation comparing random intercept models with varying fixed effects and with cluster-based resampling."
+modlist <- models("~ 1"=fit0, "~ 1 + x"=fit1, 
+                  "~ 1 + x + xm"=fit2, "~ 1 + I(x - xm)"=fit3)
+cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
+plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")
+```
+```{r cross-validation-cases}
+#| out.width = "100%",
+#| fig.height = 6,
+#| fig.cap = "Cross-validation comparing random intercept models with varying fixed effects and with case-based resampling."
 cvs_cases <- cv(modlist, data=Data, seed=9693)
-cvs_clusters <- cv(modlist, data=Data, cluster="group")
-
-cvs_cases <- cvs_cases |> lapply(rbind) |> 
-  do.call(rbind, args = _) |> as.data.frame()
-cvs_clusters <- cvs_clusters |> lapply(rbind) |> 
-  do.call(rbind, args = _) |> as.data.frame()
-
-cvs_cases$modlist <- names(modlist)
-cvs_clusters$modlist <- names(modlist)
-cvs_cases$cv_method <- "cases"
-cvs_clusters$cv_method <- "clusters"
-
-dout <- within(
-  as.data.frame(rbind(cvs_cases, cvs_clusters)),
-  {
-    model <- factor(modlist)
-    CV <- as.numeric(as.character(`CV crit`))
-  }
-)
-
-dout <- dout[order(dout$model), ]
-
-xyplot(CV ~ model, data=dout, groups=cv_method,
-       par.settings =
-         list(superpose.line=list(lty=c(2, 3), lwd=3), 
-              superpose.symbol=list(pch=15:16, 
-                                    cex=1.5)),
-       type="b",
-       auto.key=TRUE,
-       scales=list(y=list(log=TRUE)))
+plot(cvs_cases, main="Model Comparison, Case-Based CV")
 ```

### 2023-10-13 scratch/gm_addition_to_cv-edited_v2.Rmd
tweak code for mixed-model vignette addendum

@@
         Bw * (x - xm) +            # within-group effect
-        SDre * rnorm(Nb)[group] +  # random level by group
-        SDwithin * rnorm(Nb*Nw)    # random error within groups
+        rnorm(Nb, sd=SDre)[group] +  # random level by group
+        rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
     }
@@
 
+# prepare for plotting:
 Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
@@
               timevar = "modelcode",  varying = grep("_", names(Data_long)))
-
 Data_long$model <- factor(
@@
 ```
-
 ```{r plot-fits-mod3}

### 2023-10-15 scratch/gm_addition_to_cv-edited_v2.Rmd
Further edits to mixed model portion of vignette

@@
 ```
-The following artificial data set exemplifies a difference between cluster-based
-and individual-based cross-validation in hierarchical models.
+The following artificial data set exemplifies striking aspects of cross-validation
+with hierarchical models. Model comparisons using cluster-based versus individual-based 
+cross-validation may not agree on a best model. Furthermore, commonly-used measures
+of fit, such as least-squares, do not necessarily become smaller as 
+models becomes larger.
 
-It also shows how, in contrast with ordinary least-squares
-models, the variance at different levels of a hierarchical
-model need not necessarily become smaller as the model becomes larger.
+Varying the parameters of the model, particularly the between-subject variance
+relative to the within-subject variance, can also reveal strong differences in the
+behaviour of different algorithms used to fit mixed models. These
+issues are not pursued in any depth in this vignette.
 
-In this example, the simplest random intercept model results in a between-cluster variance
-that is much smaller than that of a model that also includes $x$ as a predictor.
-Consequently, different methods of assessing model fit may result in
-different best models, underscoring the importance of assessing hierarchical models
-at different levels.
+This example exhibits both phenomena described above. Consider a researcher studying the improvement
+in a skill, yodelling for example, among students enrolled in a four-year
+yodelling program. The plan is to measure each student's skill level
+at the beginning of the program and every year thereafter until
+the end of the program, resulting in five annual measurements for each student.
 
-Adjusting the parameters generating this data can be used to create examples
-that illustrates problems of non-convergence with hierarchical models.
+It turns out that yodelling appeals to students of all ages.
+Students enrolling in the program range in age from 20 to 70.
+Moreover, participants' untrained yodelling skill is similar at all ages, as is
+their rate of progress with training. All students complete the
+five-year program.
+
+The researcher, with more expertise in yodelling than modelling,
+decides to model the response $Y$, yodelling skill, against
+age, $X$, reasoning that students get older during their stay in the program,
+and age can serve as a proxy for elapsed time.
+The researcher knows that a mixed model should be used to account for clustering
+due to the expected similarity of measurements taken from each student.
+
+The figures below show the pattern of responses expected from a representative
+subset of individuals.
+
+The between-student effect of age is zero but the within-student effect is 1.
+Due to the large variation in ages between students, a least-squares model
+fitting age as a predictor would produce an estimated slope that is close to
+zero, since the estimated slope is heavily weighted toward the between-student effect.
+Consider a mixed model using the same fixed-effects model with the addition of a random intercept.
+
+Depending on the estimated variance parameters of the model, a mixed model will apply 
+varying degrees of
+shrinkage to the random intercept BLUPs [insert reference to ARA&GLM] that
+correspond to the variation in the heights of the parallel fitted lines for each student, measured
+at a common value of age.
+
+In the example below, the model applies little shrinkage, allowing a large variability in the heights
+of fitted lines, which approach closely the observed values for each student.
+Consequently, the model's fit is similar to that of
+a fixed-effects model with an effect of age together with a categorical predictor for 
+individual students. Thus, the model fits individual observations well, resulting in a
+favourable assessment using individual-based cross-validation. In contrast, the very large
+variability in the BLUPs results in large residuals for
+cluster-based cross-validation, producing an
+unfavourable assessment compared with the smaller model using only a random intercept.
+
+By comparison, if the mixed model had applied considerable shrinkage, neither the
+between-cluster nor the between-individual cross-validations would have shown much 
+improvement over the random-intercept only model.  In our experience, the degree of
+shrinkage does not vary smoothly as parameters are changed but tends to be
+"all or nothing." Its behaviour is affected considerably by the choice of the
+algorithm used to fit the model.
+
+Finally, a model that also includes a compositional variable consisting of the mean age
+of each student, results in the random intercept BLUPs representing variation in the students' fitted
+lines measured at the mean age of each student. Thus, with this example, both 
+the cluster-based and the individual-based fits are relatively good.
+
+To serve as a comparison, we also include cross-validation using the model known to 
+have generated the data, namely a model with a between-subject age effect equal to zero by
+modelling the response $Y$ with a fixed effects model using an age centered within
+each student, i.e. the age at assessment less the mean age of each student's 
+assessments, plus a random term for each student.
+
+A closely related question is that of non-convergence of mixed-model
+algorithms. Adjusting the parameters generating this data can be used to create examples
+illustrating problems of non-convergence with hierarchical models.
 Fitted models with relatively small between-cluster variance, even those using simple
@@
 produce non-convergence and are limited to comparisons obtained
-by varying three factors:
+by varying two factors:
 
-- Three models: random intercept only, adding x as a predictor, and also adding the group means of x as a predictor,
-- The method used for fitting: `lmer` in `lme4` or `lme` in `nlme`,
+- Four models: random intercept only, adding x as a predictor, adding the 
+  group means of x (xm) as a predictor, and
+  a fourth model with a random intercept and x centered within each student.
 - Cross-validation resampling individual cases or resampling clusters.
@@
 
-With this data set, the two functions `lme` and `lmer` converged for all models and produced consistent results.
-The differences between these two functions, and differences that depend on the optimizer used
-within each function, would manifest themselves if the data were generated with a smaller
-between-cluster variance.
-
 We visualize predicted values using population predictions with the fixed effects model only
-and using BLUPs for random effects.  Since `lme` and `lmer` produced almost identical results, the
-figures below are produced only with the `lmer` results.
-
+and using BLUPs for random effects. 
 ```{r}
@@
 performs cross-validation with the same training and validation sets
-for each model in order to reduce the between-model variability in the cross-validation
-criterion that stems from the random variation in training set selection.
+for each model.
 
@@
 
-With this example, `lme` and `lmer` give similar results, but resampling cases versus resampling clusters provides
-very different comparisons of models.  The model with X alone, without the contextual mean of X, is assessed as
-fitting very poorly sampling clusters and, relatively, much better resampling individual cases.  This illustrates the
-need to perform cross-validation at different levels.
+In summary, the model with $X$ alone, without the compositional mean of $X$, is assessed as
+fitting very poorly sampling clusters but, relatively, much better resampling individual cases.
+The model that includes both $X$ and its compositional mean produces better results using both
+criteria: resampling individual cases and resampling cluster, illustrating the need to
+assess models at different levels of resampling.

### 2023-10-17 scratch/gm_addition_to_cv-edited_v3.Rmd
version 3 of mixed-model vignette section

---
title: "Cross-validation of regression models: mixed model examples (edited v2)"
author: "John Fox and Georges Monette"
date: "`r Sys.Date()`"
package: cv
output:
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["cv.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Cross-validation of regression models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.align = "center",
  fig.height = 6,
  fig.width = 7,
  fig.path = "fig/",
  dev = "png",
  comment = "#>" #,
  # eval = nzchar(Sys.getenv("REBUILD_VIGNETTES"))
)

# save some typing
knitr::set_alias(w = "fig.width",
                 h = "fig.height",
                 cap = "fig.cap")

# colorize text: use inline as `r colorize(text, color)`
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}

.opts <- options(digits = 5)

knitr::opts_chunk$set(comment="    ")
```

```{r packages, include=FALSE}
library(cv)
library(car)
library(lattice)
library(latticeExtra)
library(lme4)
```
In this section, we introduce an artificial data set that exemplifies  aspects of cross-validation particular to hierarchical models. We show that model comparisons using cluster-based versus case-based cross-validation may not agree on a best model. Furthermore, commonly used measures of fit, such as mean-squared error computed for the full data set, do not necessarily become smaller as models become larger, even when the models are nested.

<!-- Varying the parameters of the model, particularly the between-subject variance -->
<!-- relative to the within-subject variance, can also reveal strong differences in the -->
<!-- behaviour of different algorithms used to fit mixed models. These -->
<!-- issues are not pursued in any depth in this vignette. -->

<!-- This example exhibits both phenomena described above.  -->

<!-- The figures below show the pattern of responses expected from a representative -->
<!-- subset of individuals.     -->

<!-- Consider a mixed model using the same fixed-effects model with the addition of a random intercept. -->

<!-- Finally, a model that also includes a compositional variable consisting of the mean age -->
<!-- of each student, results in the random intercept BLUPs representing variation in the students' fitted -->
<!-- lines measured at the mean age of each student. Thus, with this example, both  -->
<!-- the cluster-based and the individual-based fits are relatively good. -->

<!-- To serve as a comparison, we also include cross-validation using the model known to  -->
<!-- have generated the data, namely a model with a between-subject age effect equal to zero by -->
<!-- modelling the response $Y$ with a fixed effects model using an age centered within -->
<!-- each student, i.e. the age at assessment less the mean age of each student's  -->
<!-- assessments, plus a random term for each student. -->

<!-- A closely related question is that of non-convergence of mixed-model -->
<!-- algorithms. Adjusting the parameters generating this data can be used to create examples -->
<!-- illustrating problems of non-convergence with hierarchical models. -->
<!-- Fitted models with relatively small between-cluster variance, even those using simple -->
<!-- random intercept models, are prone to non-convergence due to the -->
<!-- nature of optimizing algorithms used in `lme` and `lmer`. -->

<!-- In some contexts, non-convergence, -->
<!-- can be viewed as a diagnostic of a poor model. However, with a simple -->
<!-- random intercept model where it may indicate a small between-cluster -->
<!-- variance, non-convergence is an annoyance because it would be statistically correct to -->
<!-- report that the data are consistent with a zero between-cluster population variance, -->
<!-- i.e. that observed between-cluster variance in the data can be explained by -->
<!-- the between-individual variability within clusters.  This leads to -->
<!-- valid statistical inferences. Although the calculation of accurate p-values for hypotheses concerning -->
<!-- random effects may be complicated -->
<!-- by the fact that the estimated variance is at the boundary of the parameter space, -->
<!-- the estimated variance-covariance matrix of fixed-effect parameters will generally -->
<!-- be non-singular and may be used for valid Wald tests of linear hypotheses -->
<!-- of fixed-effect parameters. -->

<!-- Using cross-validation based on fitting random subsets compounds the problem -->
<!-- since subsets for which there is non-convergence cause the cross-validation to -->
<!-- fail unless non-converging subsets are omitted, which may bias results since non-convergence -->
<!-- is likely to be related to the estimated between-cluster variances. -->

<!-- Varying the parameter for between-cluster variance (`SDre` in the code below), -->
<!-- will generate data sets with varying propensities to produce -->
<!-- non-convergence. These data sets can be used to explore the consequences of non-convergence and to -->
<!-- compare the performance of different optimizers. -->

<!-- At present, the results below are based on a data set that does not -->
<!-- produce non-convergence and are limited to comparisons obtained -->
<!-- by varying two factors: -->

<!-- - Four models: random intercept only, adding x as a predictor, adding the  -->
<!--   group means of x (xm) as a predictor, and -->
<!--   a fourth model with a random intercept and x centered within each student. -->
<!-- - Cross-validation resampling individual cases or resampling clusters. -->

<!-- Since the independence of clusters may be a reasonable assumption, as is the -->
<!-- the independence of individual observations from the same cluster *conditional* -->
<!-- on clusters, each form of resampling addresses the -->
<!-- suitability of models -->
<!-- to answer different questions. -->
<!-- Results based on individual resampling generalize to new samples from the same -->
<!-- clusters. Resampling clusters generalizes to new samples of clusters and -->
<!-- cases. -->

Consider a researcher studying improvement in a skill, yodelling for example, among students enrolled in a four-year yodelling program. The plan is to measure each student's skill level at the beginning of the program and every year thereafter until the end of the program, resulting in five annual measurements for each student. It turns out that yodelling appeals to students of all ages, and students enrolling in the program range in age from 20 to 70. Moreover, participants' untrained yodelling skill is similar at all ages, as is their rate of progress with training. All students complete the five-year program.

The researcher, who has more expertise in yodelling than modelling, decides to model the response $y$, yodelling skill, as a function of age, $x$, reasoning that students get older during their stay in the program, and (spuriously) that age can serve as a proxy for elapsed time. The researcher knows that a mixed model should be used to account for clustering due to the expected similarity of measurements taken from each student.

We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^vary-parameters]

[^vary-parameters]: We invite the interested reader to experiment with varying the parameters of our example.

```{r data}
# Generate data set

# Parameters:
set.seed(9693) 
Nb <- 100    # number of groups
Nw <- 5      # number of individuals within groups
Bb <- 0      # between-group regression coefficient on group mean
SDre <- 2.0  # between-group SD of random level relative to group mean of x
SDwithin <- 0.5  # within group SD
Bw <- 1     # within group effect of x
Ay <- 10    # intercept for response
Ax <- 20    # starting level of x
Nx <- Nw*10 # number of distinct x values

Data <- data.frame(
  group = factor(rep(1:Nb, each=Nw)),
  x = Ax + rep(1:Nx, length.out = Nw*Nb)
) |>
  within(
    {
      xm  <- ave(x, group, FUN = mean) # within-group mean
      y <- Ay +
        Bb * xm +                  # contextual effect
        Bw * (x - xm) +            # within-group effect
        rnorm(Nb, sd=SDre)[group] +  # random level by group
        rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
    }
  )
```

Here is scatterplot of the data for a representative group of 10 (without loss of generality, the first 10) of 100 students, showing the 95% concentration ellipse for each cluster:[^lattice]

[^lattice]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.

```{r plot1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Hierarchical data set, showing the first 10 of 100 students."
library("lattice")
library("latticeExtra")
plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
               ylim=c(4, 16),
               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
    layer(panel.ellipse(..., center.cex=0))
plot
```

The between-student effect of age is 0 but the within-student effect is 1. Due to the large variation in ages between students, the least-squares regression of yodelling skill on age produces an estimated slope close to 0 (though with a small $p$-value), because the  slope is heavily weighted toward the between-student effect:
```{r}
summary(lm(y ~ x, data=Data))
```

The initial mixed-effects model that we fit to the data is a simple random-intercepts model:
```{r}
# random intercept only:
fit0 <- lmer(y ~ 1 + (1 | group), Data)
summary(fit0)
```
We will shortly consider three other, more complex, mixed models; because of data-management considerations, it is convenient to fit them now, but we defer discussion of these models:
```{r}
# effect of x and random intercept:
fit1 <- lmer(y ~ x + (1 | group), Data)

# effect of x, mean of x, and random intercept:
fit2 <- lmer(y ~ x + xm + (1 | group), Data)
        # equivalent to y ~ I(x - xm) + xm + (1 | group)

# model generating the data (where Bb = 0)
fit3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
```

We proceed to obtain predictions from the random-intercept model (`fit0`) and the other models (`fit1`, `fit2`, and `fit3`) based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students):
```{r}
Data <- within(Data, {
  fit_mod0.fe <- predict(fit0, re.form = ~ 0) # fixed effects only
  fit_mod0.re <- predict(fit0) # fixed and random effects (BLUPs)
  fit_mod1.fe <- predict(fit1, re.form = ~ 0)
  fit_mod1.re <- predict(fit1)
  fit_mod2.fe <- predict(fit2, re.form = ~ 0)
  fit_mod2.re <- predict(fit2)
  fit_mod3.fe <- predict(fit3, re.form = ~ 0)
  fit_mod3.re <- predict(fit3)
})
```

We then prepare the data for plotting:
```{r}
Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
Data_long$id <- 1:nrow(Data_long)
Data_long <- reshape(Data_long, direction = "long", sep = "_", 
              timevar = "modelcode",  varying = grep("_", names(Data_long)))
Data_long$model <- factor(
  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
)
```

Predictions based on the random-intercept model `fit0` for the first 10 students are shown in the following graph:
```{r plot-fits-mod0}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictions from the random intercept model."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + (1 | group)",
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean. Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not terribly different.

Our next model, `fit1` includes a fixed intercept and fixed effect of `x` along with a random intercept:
```{r}
summary(fit1)
```
Predictions from this model appear in the following graph:
```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictions from the model with random intercepts and $x$ as a fixed-effect predictor."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + (1 | group)",
  ylim=c(-15, 35),
  key=list(
    corner=c(0.95, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model `fit0`. We therefore anticipate (and show later in this section) that case-based cross-validation will prefer `mod1` to `mod0` but that cluster-based cross-validation will prefer `mod0` to `mod1`.

Our third model, `fit2`, includes the contextual effect of $x$---that is, the cluster mean `xm`---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:
```{r}
summary(fit2)
```
This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | group)`, which is the model that generated the data once the coefficient of the contextual predictor `xm` is set to 0.

Predictions from model `fit2` appear in the following graph:
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictors from the model with random intercepts, $x$, and the group (student) mean of $x$ as predictors."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + xm + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```
Depending on the estimated variance parameters of the model, a mixed model like this will apply varying degrees of shrinkage to the random-intercept BLUPs [insert reference to ARA&GLM] that correspond to variation in the height of the parallel fitted line for each student. In our contrived data, the model applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student. Consequently, the model's fit is similar to that of a fixed-effects model with an effect of age together with a categorical predictor for individual students. The model therefore fits individual observations well, and we anticipate a favorable assessment using individual-based cross-validation. In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for model `fit2` compared to the smaller model `fit0` which includes only fixed and random intercepts.

Had the mixed model applied considerable shrinkage, then neither cluster-based nor case-based cross-validation would show much improvement over the random-intercept-only model.  In our experience, the degree of shrinkage does not vary smoothly as parameters are changed but tends to be "all or nothing," and near the tipping point, the behavior of estimates can be  affected considerably by the choice of algorithm used to fit the model.

<!-- We can compare estimated fixed-effect coefficients. -->

<!-- ```{r, message=FALSE,warning=FALSE} -->
<!-- compareCoefs(fit0,  fit1, fit2, fit3) -->
<!-- ``` -->

<!-- We visualize predicted values using population predictions with the fixed effects model only -->
<!-- and using BLUPs for random effects.  -->

Finally, `fit3` directly estimates the model used to generate the data. As mentioned, it is a constrained version of `fit2`, with the coefficient of `xm` set to 0, and with `x` expressed as a deviation from the cluster mean `xm`:
```{r}
summary(fit3)
```

The predictions from `fit3` are therefore similar to those from `fit2`:
```{r plot-fits-mod3}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictions from the estimated model generating the data."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

We next carry out case-based cross-validation, which as we have explained is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only. In order to reduce between-model variability in comparisons of models, we apply `cv()` to the list of models created by the `models()` function, performing cross-validation with the same folds for each model:

```{r cross-validation-clusters}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept models with varying fixed effects."
modlist <- models("~ 1"=fit0, "~ 1 + x"=fit1, 
                  "~ 1 + x + xm"=fit2, "~ 1 + I(x - xm)"=fit3)
cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")
```
```{r cross-validation-cases}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "10-fold case-based cross-validation comparing random intercept models with varying fixed effects."
cvs_cases <- cv(modlist, data=Data, seed=9693)
plot(cvs_cases, main="Model Comparison, Case-Based CV")
```

In summary, the model with $x$ alone, without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. The model that includes both $x$ and its contextual mean produces better results using both cluster-based and case-based CV. The data-generating model, which includes the fixed effect of `x - xm` in place of separate terms in `x` and `xm`, isn't distinguishable from the model including `x` and `xm` separately, even though the latter has an unnecessary parameter (recall that the population coefficient of `xm` is 0 when `x` is expressed as deviations from the contextual mean). These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the need to assess mixed-effect models at different hierarchical levels.

### 2023-10-17 for-CRAN/tests/testthat/test-cvSelect.R
add for-CRAN directory for CRAN submission with prebuilt html vignettes

@@ -1,0 +1,75 @@
+# test that different algorithms produce the same results
+#  using selectStepAIC() for a lm
+
+data(Auto, package="ISLR2")
+m <- lm(mpg ~ . - name - origin, data=Auto)
+
+test_that("cvSelect naive vs Woodbury lm", {
+  expect_equal(cvSelect(selectStepAIC, Auto, k=5, seed=123,
+                        model=m, method="naive"),
+               cvSelect(selectStepAIC, Auto, k=5, seed=123,
+                        model=m, method="Woodbury")
+  )
+})
+
+# test that parallel computations work correctly using selectStepAIC()
+#  for a lm
+
+test_that("cvSelect parallel lm", {
+  expect_equal(cvSelect(selectStepAIC, Auto, k=5, seed=123,
+                        model=m),
+               cvSelect(selectStepAIC, Auto, k=5, seed=123,
+                        model=m, ncores=2)
+               )
+})
+
+# test that different algorithms produce the same results
+#  using selectStepAIC() for a Gaussian glm
+
+m.glm <- lm(mpg ~ . - name - origin, data=Auto)
+
+test_that("cvSelect exact vs Woodbury glm", {
+  expect_equal(cvSelect(selectStepAIC, Auto, k=5, seed=123,
+                        model=m.glm, method="exact"),
+               cvSelect(selectStepAIC, Auto, k=5, seed=123,
+                        model=m.glm, method="Woodbury")
+  )
+})
+
+
+# longer tests
+
+test_that("cvSelect exact vs hatvalues loo glm", {
+  skip_on_cran()
+  expect_equal(cvSelect(selectStepAIC, Auto, k="loo",
+                        model=m.glm, method="exact"),
+               cvSelect(selectStepAIC, Auto, k="loo",
+                        model=m.glm, method="hatvalues")
+  )
+})
+
+test_that("cvSelect Woodbury vs hatvalues loo glm", {
+  skip_on_cran()
+  expect_equal(cvSelect(selectStepAIC, Auto, k="loo",
+                        model=m.glm, method="Woodbury"),
+               cvSelect(selectStepAIC, Auto, k="loo",
+                        model=m.glm, method="hatvalues")
+  )
+})
+
+# test that parallel computations work correctly using selectStepAIC()
+#  for a glm
+
+data("Caravan", package="ISLR2")
[…]

### 2023-10-18 scratch/gm_addition_to_cv-edited_v3.Rmd
copy mixed-model contrived example to cv.Rmd vignette

@@
 ---
-title: "Cross-validation of regression models: mixed model examples (edited v2)"
+title: "Cross-validation of regression models: mixed model examples (edited v3)"
 author: "John Fox and Georges Monette"
@@
 ```
-In this section, we introduce an artificial data set that exemplifies  aspects of cross-validation particular to hierarchical models.
-We show that model comparisons using cluster-based versus case-based cross-validation may not agree on a best model.
-Furthermore, commonly used measures of fit, such as mean-squared error computed for the full data set, do not necessarily become smaller as models become larger, even when the models are nested.
 
-<!-- Varying the parameters of the model, particularly the between-subject variance -->
-<!-- relative to the within-subject variance, can also reveal strong differences in the -->
-<!-- behaviour of different algorithms used to fit mixed models. These -->
-<!-- issues are not pursued in any depth in this vignette. -->
+### Example: Contrived Hierarchical Data
 
-<!-- This example exhibits both phenomena described above.  -->
+We introduce an artificial data set that exemplifies aspects of cross-validation particular to hierarchical models.
+Using this data set, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model.
+Furthermore, commonly used measures of fit, such as mean-squared error, do not necessarily become smaller as models become larger, even when the models are nested, and even when the measure of fit is computed for the whole data set.
 
-<!-- The figures below show the pattern of responses expected from a representative -->
-<!-- subset of individuals.     -->
-
-<!-- Consider a mixed model using the same fixed-effects model with the addition of a random intercept. -->
-
-<!-- Finally, a model that also includes a compositional variable consisting of the mean age -->
-<!-- of each student, results in the random intercept BLUPs representing variation in the students' fitted -->
-<!-- lines measured at the mean age of each student. Thus, with this example, both  -->
-<!-- the cluster-based and the individual-based fits are relatively good. -->
-
-<!-- To serve as a comparison, we also include cross-validation using the model known to  -->
-<!-- have generated the data, namely a model with a between-subject age effect equal to zero by -->
-<!-- modelling the response $Y$ with a fixed effects model using an age centered within -->
-<!-- each student, i.e. the age at assessment less the mean age of each student's  -->
-<!-- assessments, plus a random term for each student. -->
-
-<!-- A closely related question is that of non-convergence of mixed-model -->
-<!-- algorithms. Adjusting the parameters generating this data can be used to create examples -->
-<!-- illustrating problems of non-convergence with hierarchical models. -->
-<!-- Fitted models with relatively small between-cluster variance, even those using simple -->
-<!-- random intercept models, are prone to non-convergence due to the -->
-<!-- nature of optimizing algorithms used in `lme` and `lmer`. -->
-
-<!-- In some contexts, non-convergence, -->
-<!-- can be viewed as a diagnostic of a poor model. However, with a simple -->
-<!-- random intercept model where it may indicate a small between-cluster -->
-<!-- variance, non-convergence is an annoyance because it would be statistically correct to -->
-<!-- report that the data are consistent with a zero between-cluster population variance, -->
-<!-- i.e. that observed between-cluster variance in the data can be explained by -->
-<!-- the between-individual variability within clusters.  This leads to -->
-<!-- valid statistical inferences. Although the calculation of accurate p-values for hypotheses concerning -->
-<!-- random effects may be complicated -->
-<!-- by the fact that the estimated variance is at the boundary of the parameter space, -->
-<!-- the estimated variance-covariance matrix of fixed-effect parameters will generally -->
-<!-- be non-singular and may be used for valid Wald tests of linear hypotheses -->
-<!-- of fixed-effect parameters. -->
-
-<!-- Using cross-validation based on fitting random subsets compounds the problem -->
-<!-- since subsets for which there is non-convergence cause the cross-validation to -->
-<!-- fail unless non-converging subsets are omitted, which may bias results since non-convergence -->
-<!-- is likely to be related to the estimated between-cluster variances. -->
-
-<!-- Varying the parameter for between-cluster variance (`SDre` in the code below), -->
-<!-- will generate data sets with varying propensities to produce -->
-<!-- non-convergence. These data sets can be used to explore the consequences of non-convergence and to -->
-<!-- compare the performance of different optimizers. -->
-
-<!-- At present, the results below are based on a data set that does not -->
-<!-- produce non-convergence and are limited to comparisons obtained -->
-<!-- by varying two factors: -->
-
-<!-- - Four models: random intercept only, adding x as a predictor, adding the  -->
-<!--   group means of x (xm) as a predictor, and -->
-<!--   a fourth model with a random intercept and x centered within each student. -->
-<!-- - Cross-validation resampling individual cases or resampling clusters. -->
-
-<!-- Since the independence of clusters may be a reasonable assumption, as is the -->
-<!-- the independence of individual observations from the same cluster *conditional* -->
-<!-- on clusters, each form of resampling addresses the -->
-<!-- suitability of models -->
-<!-- to answer different questions. -->
-<!-- Results based on individual resampling generalize to new samples from the same -->
-<!-- clusters. Resampling clusters generalizes to new samples of clusters and -->
-<!-- cases. -->
-
-Consider a researcher studying improvement in a skill, yodelling for example, among students enrolled in a four-year yodelling program.
+Consider a researcher studying improvement in a skill, yodelling, for example, among students enrolled in a four-year yodelling program.
 The plan is to measure each student's skill level at the beginning of the program and every year thereafter until the end of the program, resulting in five annual measurements for each student.
@@
 
-The researcher, who has more expertise in yodelling than modelling, decides to model the response $y$, yodelling skill, as a function of age, $x$, reasoning that students get older during their stay in the program, and (spuriously) that age can serve as a proxy for elapsed time.
+The researcher, who has more expertise in yodelling than in modelling, decides to model the response, $y$, yodelling skill, as a function of age, $x$, reasoning that students get older during their stay in the program, and (incorrectly) that age can serve as a proxy for elapsed time.
 The researcher knows that a mixed model should be used to account for clustering due to the expected similarity of measurements taken from each student.
 
-We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^vary-parameters]
+We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^1]
 
-[^vary-parameters]: We invite the interested reader to experiment with varying the parameters of our example.
+[^1]: We invite the interested reader to experiment with varying the parameters of our example.
 
 ```{r data}
-# Generate data set
-
 # Parameters:
@@
 
-Here is scatterplot of the data for a representative group of 10 (without loss of generality, the first 10) of 100 students, showing the 95% concentration ellipse for each cluster:[^lattice]
+Here is scatterplot of the data for a representative group of 10 (without loss of generality, the first 10) of 100 students, showing the 95% concentration ellipse for each cluster:[^2]
 
-[^lattice]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.
+[^2]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.
 
@@
     layer(panel.ellipse(..., center.cex=0))
-plot
+plot # display graph
 ```
@@
 The between-student effect of age is 0 but the within-student effect is 1.
-Due to the large variation in ages between students, the least-squares regression of yodelling skill on age produces an estimated slope close to 0 (though with a small $p$-value), because the  slope is heavily weighted toward the between-student effect:
+Due to the large variation in ages between students, the least-squares regression of yodelling skill on age (for the 500 observations among all 100 students) produces an estimated slope close to 0 (though with a small $p$-value), because the slope is heavily weighted toward the between-student effect:
+
 ```{r}
@@
 The initial mixed-effects model that we fit to the data is a simple random-intercepts model:
+
 ```{r}
 # random intercept only:
-fit0 <- lmer(y ~ 1 + (1 | group), Data)
-summary(fit0)
+mod.0 <- lmer(y ~ 1 + (1 | group), Data)
+summary(mod.0)
 ```
+
 We will shortly consider three other, more complex, mixed models; because of data-management considerations, it is convenient to fit them now, but we defer discussion of these models:
+
 ```{r}
 # effect of x and random intercept:
-fit1 <- lmer(y ~ x + (1 | group), Data)
+mod.1 <- lmer(y ~ x + (1 | group), Data)
 
-# effect of x, mean of x, and random intercept:
-fit2 <- lmer(y ~ x + xm + (1 | group), Data)
+# effect of x, contextual (student) mean of x, and random intercept:
+mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
         # equivalent to y ~ I(x - xm) + xm + (1 | group)
@@
 # model generating the data (where Bb = 0)
-fit3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
+mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
 ```
 
-We proceed to obtain predictions from the random-intercept model (`fit0`) and the other models (`fit1`, `fit2`, and `fit3`) based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students):
+We proceed to obtain predictions from the random-intercept model (`mod.0`) and the other models (`mod.1`, `mod.2`, and `mod.3`) based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students):
+
 ```{r}
 Data <- within(Data, {
-  fit_mod0.fe <- predict(fit0, re.form = ~ 0) # fixed effects only
-  fit_mod0.re <- predict(fit0) # fixed and random effects (BLUPs)
-  fit_mod1.fe <- predict(fit1, re.form = ~ 0)
-  fit_mod1.re <- predict(fit1)
-  fit_mod2.fe <- predict(fit2, re.form = ~ 0)
-  fit_mod2.re <- predict(fit2)
-  fit_mod3.fe <- predict(fit3, re.form = ~ 0)
-  fit_mod3.re <- predict(fit3)
+  fit_mod0.fe <- predict(mod.0, re.form = ~ 0) # fixed effects only
+  fit_mod0.re <- predict(mod.0) # fixed and random effects (BLUPs)
+  fit_mod1.fe <- predict(mod.1, re.form = ~ 0)
+  fit_mod1.re <- predict(mod.1)
+  fit_mod2.fe <- predict(mod.2, re.form = ~ 0)
+  fit_mod2.re <- predict(mod.2)
+  fit_mod3.fe <- predict(mod.3, re.form = ~ 0)
+  fit_mod3.re <- predict(mod.3)
 })
@@
 We then prepare the data for plotting:
+
 ```{r}
@@
 
-Predictions based on the random-intercept model `fit0` for the first 10 students are shown in the following graph:
+Predictions based on the random-intercept model `mod.0` for the first 10 students are shown in the following graph:
+
 ```{r plot-fits-mod0}
@@
 ```
+
 The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean.
-Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not terribly different.
+Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.
 
-Our next model, `fit1` includes a fixed intercept and fixed effect of `x` along with a random intercept:
+Our next model, `mod.1`, includes a fixed intercept and fixed effect of `x` along with a random intercept:
+
 ```{r}
-summary(fit1)
+summary(mod.1)
 ```
+
 Predictions from this model appear in the following graph:
+
 ```{r plot-fits-mod1}
@@
 ```
-The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model `fit0`.
-We therefore anticipate (and show later in this section) that case-based cross-validation will prefer `mod1` to `mod0` but that cluster-based cross-validation will prefer `mod0` to `mod1`.
 
-Our third model, `fit2`, includes the contextual effect of $x$---that is, the cluster mean `xm`---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:
+The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, `mod.0`.
+We therefore anticipate (and show later in this section) that case-based cross-validation will prefer `mod1` to `mod0`, but that cluster-based cross-validation will prefer `mod0` to `mod1`.
+
+Our third model, `mod.2`, includes the contextual effect of $x$---that is, the cluster mean `xm`---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:
+
 ```{r}
-summary(fit2)
+summary(mod.2)
 ```
-This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | group)`, which is the model that generated the data once the coefficient of the contextual predictor `xm` is set to 0.
 
-Predictions from model `fit2` appear in the following graph:
+This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | group)`, which is the model that generated the data once the coefficient of the contextual predictor `xm` is set to 0 (as it is in `mod.3`, discussed below).
+
+Predictions from model `mod.2` appear in the following graph:
 ```{r plot-fits-mod2}
@@
 ```
-Depending on the estimated variance parameters of the model, a mixed model like this will apply varying degrees of shrinkage to the random-intercept BLUPs [insert reference to ARA&GLM] that correspond to variation in the height of the parallel fitted line for each student.
-In our contrived data, the model applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student.
-Consequently, the model's fit is similar to that of a fixed-effects model with an effect of age together with a categorical predictor for individual students.
-The model therefore fits individual observations well, and we anticipate a favorable assessment using individual-based cross-validation.
-In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for model `fit2` compared to the smaller model `fit0` which includes only fixed and random intercepts.
+
+Depending on the estimated variance parameters of the model, a mixed model like `mod.2` will apply varying degrees of shrinkage to the random-intercept BLUPs that correspond to variation in the heights of the parallel fitted lines for the individual students.
+In our contrived data, the `mod.2` applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student.
+The fit of the mixed model `mod.2` is consequently similar to that of a fixed-effects model with age and a categorical predictor for individual students (i.e., treating students as a factor, and not shown here).
+
+The  mixed model `mod.2` therefore fits individual observations well, and we anticipate a favorable assessment using individual-based cross-validation.
+In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for model `mod.2` compared to the smaller model `mod.0`, which includes only fixed and random intercepts.
 
@@
 
-<!-- We can compare estimated fixed-effect coefficients. -->
+Finally, `mod.3` directly estimates the model used to generate the data.
+As mentioned, it is a constrained version of `mod.2`, with the coefficient of `xm` set to 0, and with `x` expressed as a deviation from the cluster mean `xm`:
 
-<!-- ```{r, message=FALSE,warning=FALSE} -->
-<!-- compareCoefs(fit0,  fit1, fit2, fit3) -->
-<!-- ``` -->
-
-<!-- We visualize predicted values using population predictions with the fixed effects model only -->
-<!-- and using BLUPs for random effects.  -->
-
-Finally, `fit3` directly estimates the model used to generate the data.
-As mentioned, it is a constrained version of `fit2`, with the coefficient of `xm` set to 0, and with `x` expressed as a deviation from the cluster mean `xm`:
 ```{r}
-summary(fit3)
+summary(mod.3)
 ```
 
-The predictions from `fit3` are therefore similar to those from `fit2`:
+The predictions from `mod.3` are therefore similar to those from `mod.2`:
+
 ```{r plot-fits-mod3}
@@
 
-We next carry out case-based cross-validation, which as we have explained is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.
-In order to reduce between-model variability in comparisons of models, we apply `cv()` to the list of models created by the `models()` function, performing cross-validation with the same folds for each model:
+We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.
+In order to reduce between-model random variability in comparisons of models, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model:
 
@@
 #| fig.cap = "10-fold cluster-based cross-validation comparing random intercept models with varying fixed effects."
-modlist <- models("~ 1"=fit0, "~ 1 + x"=fit1, 
-                  "~ 1 + x + xm"=fit2, "~ 1 + I(x - xm)"=fit3)
+modlist <- models("~ 1"=mod.0, "~ 1 + x"=mod.1, 
+                  "~ 1 + x + xm"=mod.2, "~ 1 + I(x - xm)"=mod.3)
 cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
@@
 ```
+
 ```{r cross-validation-cases}
@@
 
-In summary, the model with $x$ alone, without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
-The model that includes both $x$ and its contextual mean produces better results using both cluster-based and case-based CV.
-The data-generating model, which includes the fixed effect of `x - xm` in place of separate terms in `x` and `xm`, isn't distinguishable from the model including `x` and `xm` separately, even though the latter has an unnecessary parameter (recall that the population coefficient of `xm` is 0 when `x` is expressed as deviations from the contextual mean).
-These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the need to assess mixed-effect models at different hierarchical levels.
+In summary, model `mod.1`, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
+Model `mod.2`, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV.
+The data-generating model, `mod.3`, which includes the fixed effect of `x - xm` in place of separate terms in `x` and `xm`, isn't distinguishable from model `mod.2`, which incldues `x` and `xm` separately, even though `mod.2` has an unnecessary parameter (recall that the population coefficient of `xm` is 0 when `x` is expressed as deviations from the contextual mean).
+These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

### 2023-10-18 vignettes/cv.R

@@ -9,7 +9,6 @@ knitr::opts_chunk$set(
   fig.path = "fig/",
   dev = "png",
   comment = "#>" #,
-  # eval = nzchar(Sys.getenv("REBUILD_VIGNETTES"))
 )
 
 # save some typing
@@ -157,6 +156,159 @@ data("MathAchSchool", package="nlme")
 dim(MathAchSchool)
 head(MathAchSchool, 2)
 tail(MathAchSchool, 2)
+
+## ----data---------------------------------------------------------------------
+# Parameters:
+set.seed(9693) 
+Nb <- 100     # number of groups
+Nw <- 5       # number of individuals within groups
+Bb <- 0       # between-group regression coefficient on group mean
+SDre <- 2.0   # between-group SD of random level relative to group mean of x
+SDwithin <- 0.5  # within group SD
+Bw <- 1          # within group effect of x
+Ay <- 10         # intercept for response
+Ax <- 20         # starting level of x
+Nx <- Nw*10      # number of distinct x values
+
+Data <- data.frame(
+  group = factor(rep(1:Nb, each=Nw)),
+  x = Ax + rep(1:Nx, length.out = Nw*Nb)
+) |>
+  within(
+    {
+      xm  <- ave(x, group, FUN = mean) # within-group mean
+      y <- Ay +
+        Bb * xm +                    # contextual effect
+        Bw * (x - xm) +              # within-group effect
+        rnorm(Nb, sd=SDre)[group] +  # random level by group
+        rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
+    }
+  )
+
+## ----plot1--------------------------------------------------------------------
+library("lattice")
+library("latticeExtra")
+plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
+               ylim=c(4, 16),
+               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
+    layer(panel.ellipse(..., center.cex=0))
+plot # display graph
+
+## -----------------------------------------------------------------------------
+summary(lm(y ~ x, data=Data))
+
+## -----------------------------------------------------------------------------
+# random intercept only:
+mod.0 <- lmer(y ~ 1 + (1 | group), Data)
+summary(mod.0)
+
+## -----------------------------------------------------------------------------
[…]

### 2023-10-29 vignettes/cv-extend.R
reset options() at end of vignettes

@@ -242,3 +242,6 @@ selectSubsets(model=m.swiss, indices=seq(5, 45, by=10))
 ## ----best-models-by-folds-----------------------------------------------------
 compareFolds(cv.swiss)
 
+## ----coda, include = FALSE----------------------------------------------------
+options(.opts)
+

### 2023-10-29 vignettes/cv.R

@@ -448,3 +448,6 @@ system.time(m.mroz.sel.cv.p <- cvSelect(selectStepAIC, Mroz,
                           ncores=2))
 all.equal(m.mroz.sel.cv, m.mroz.sel.cv.p)
 
+## ----coda, include = FALSE----------------------------------------------------
+options(.opts)
+

### 2023-10-29 scratch/notes_john/simple-var-selection.Rmd
restore scratch dir with notes_john subdir; update tarball under scratch

Now suppose that we use a simple-minded variable-selection procedure to pick the $p' = 5$ predictors that are most highly correlated with $y$ and use them in a least-squares regression:
```{r pick-best-5}
r <- as.vector(cor(X, y))
names(r) <- 1:100
(best5 <- sort(r, decreasing=TRUE)[1:5])
m <- lm(y ~ X[, names(best5)])
summary(m)
mse(y, fitted(m))
```
The $R^2$ for the selected model is small, but the omnibus $F$-test that all of the 5 $\beta$s are 0 is (spuriously) "statistically significant," as are the regression coefficients for 2 of the 5 selected $x$s. These results are, of course, nonsense. Notice that the mean-squared error of prediction for the model is $0.962$, which is smaller than the true error variance $\sigma^2 = 1$. The estimated error variance for the least-squares regression is also too small, $\widehat{\sigma}^2 = 0.984^2 = 0.968$.

Now let us cross-validate the MSE for the selected model:
```{r cv-best-5}
cv(m, seed=4354)
```
Cross-validation consequently also over-states the predictive power of the model because, to reiterate, the model was selected on the same data used in the CV process.

How then can one proceed correctly to cross-validate a model-selection procedure? The model-selection procedure should be applied with each fold removed, and the resulting selected model applied to predict the response for the cases in the omitted fold.

For our simple variable-selection procedure this process can be automated:
```{r cvSelect}
selectBest <- function(data, n.best, y, x){
  # data: data frame containing all of the data
  # n.best: number of xs to select
  # y: name or index of the response
  # x: vector of names or indices of candidate predictors
  y <- data[ , y]
  X <- as.matrix(data[ , x])
  colnames(X) <- 1:ncol(X)
  r <- as.vector(cor(X, y))
  names(r) <- 1:ncol(X)
  (best <- sort(r, decreasing=TRUE)[1:n.best])
  best <- names(best)
  X <- X[, best]
  data <- data.frame(y, X)
  lm(y ~ X, data=data)
  }

D <- data.frame(y, X)
selectBest(data=D, n.best=5, y=1, x=2:101)
```
Next, we write a function to apply `selectBest()` with a fold removed:
```{r cvSelectBest}
cvSelectBest <- function(data, indices, n.best, y, x){
  # data: data frame containing all of the data
  # indices: row indices in data for the current fold
  #          if omitted, the function returns the MSE
  #          for the full data set
  # n.best: number of xs to select
  # y: name or index of the response
  # x: vector of names or indices of candidate predictors
  if (missing(indices)) {
    model.i <- selectBest(data, n.best=n.best, y=y, x=x)
    fit.o.i <- fitted(model.i)
    return(mse(data[, y], fit.o.i))
  }
  data.minus.fold <- data[-indices, ]
  model.i <- selectBest(data.minus.fold, n.best, y, x)
  X <- cbind(1, as.matrix(data[, names(coef(model.i))[-1]]))
  fit.o.i <- as.vector(X %*% coef(model.i))
  fit.i <- fit.o.i[indices]
  c(mse(data[indices, y], fit.i), mse(data[, y], fit.o.i))
}
```
For example:
```{r try-cvSelectBest}
cvSelectBest(data=D, n.best=5, y=1, x=2:101) # whole data set
cvSelectBest(data=D, indices=1:100,
             n.best=5, y=1, x=2:101) # omit first 100 cases
```

We wrote `cvSelectBest()` so that it's useable  with the `cvSelect()` function in the **cv** package. `cvSelect()` applies a model-selection procedure, such as our `cvSelectBest()`, in $k$-fold or LOO CV.

`cvSelectBest()` has two required arguments: `procedure`, a model-selection function whose first two arguments should be `data`, a full data frame, and `indices`, the case indices for the current fold. The user can also supply `k` and `seed` arguments to `cvSelectBest()`; as for `cv()`, `k` defaults to `10`, and if `seed` isn't supplied, a random-number generator seed is randomly selected and saved. Any other arguments to `cvSelect()` are passed through to the model-selection function.

So, we can cross-validate our simple model-selection procedure applied to the randomly generated data by
```{r cv-select-best-5}
cvSelect(cvSelectBest, data=D, seed=321,
         n.best=5, y=1, x=2:101)
```
Properly cross-validating model selection for the example produces a realistic estimate of the poor predictive power of the selected model, with the estimated MSE of prediction very slightly larger than the marginal variance of $y$, that is, $\sigma^2 = 1$.

### 2023-10-29 scratch/notes_john/cvSelect-tests.R

@@ -1,0 +1,31 @@
+library(cv)
+library(MASS)
+library(ISLR)
+
+m <- lm(mpg ~ . - name - origin, data=Auto)
+
+stepAIC(m, trace=FALSE)
+stepAIC(m, k=log(nrow(Auto)), trace=FALSE) # BIC
+
+stepAIC(lm(mpg ~ . - name - origin, data=Auto[1:50, ]),
+           trace=FALSE)
+stepAIC(lm(mpg ~ . - name - origin, data=Auto[1:50, ]),
+        k=log(nrow(Auto)), trace=FALSE) # BIC
+
+selectStepAIC(Auto, 1:50, model=m)
+selectStepAIC(Auto, 51:100, model=m)
+selectStepAIC(Auto, model=m)
+
+selectStepAIC(Auto, 1:50, model=m, k.=log(nrow(Auto))) # BIC
+
+cvSelect(selectStepAIC, Auto, k=5, seed=123, model=m)
+cvSelect(selectStepAIC, Auto, k=5, seed=321, model=m)
+cvSelect(selectStepAIC, Auto, k=5, seed=123, model=m,
+         k.=log(nrow(Auto))) # BIC
+
+cvSelect(selectStepAIC, Auto, k=5, seed=321, model=m, reps=5)
+
+
+system.time(print(cvSelect(selectStepAIC, Auto, seed=123, model=m)))
+system.time(print(cvSelect(selectStepAIC, Auto, seed=123, model=m,
+                           ncores=5))) # slower

### 2023-10-29 scratch/notes_john/hex-sticker.R

@@ -1,0 +1,33 @@
+# library(plotrix)
+
+hexagon <- function (x, y, unitcell = 1, col = NA, border = "black", lwd=1)
+{
+  polygon(c(x, x, x + unitcell/2, x + unitcell, x + unitcell,
+            x + unitcell/2), c(y + unitcell * 0.125, y + unitcell *
+                                 0.875, y + unitcell * 1.125, y + unitcell * 0.875, y +
+                                 unitcell * 0.125, y - unitcell * 0.125), col = col, border = border,
+          lwd=lwd)
+}
+
+
+row1 <- seq(0, 0.8, length=5)*5
+row2 <- rep(0.2, 5)*5
+row3 <- seq(0.8, 0, length=5)*5
+heights <- rbind(row1, row2, row3)
+
+
+png("~/temp/hex.png", width=500, height=500)
+
+par(mar=rep(c(2.5, 0, 3.5, 2.5)))
+
+barplot(heights, xlim=c(-2.5, 7.5), ylim=c(-2.5, 7.5),
+        col=c("gray", "blue", "gray"),
+        axes=FALSE)
+
+hexagon(-0.65, -1.1, 7.5, lwd=5, border="blue",
+        col=scales::alpha("blue", 0.10))
+
+text(3, 6, "cv", cex=4, col="blue")
+text(3.1, -0.6, "cross-validation", col="blue", cex=2)
+
+dev.off()

### 2023-10-29 scratch/notes_john/run_all_cv_tests.R

@@ -1,0 +1,5 @@
+run_all_cv_tests <- function(run=TRUE){
+  if (run) Sys.setenv(RUN_ALL_CV_TESTS = "true")
+  else Sys.setenv(RUN_ALL_CV_TESTS = "")
+}
+

### 2023-10-29 scratch/notes_john/test-cv.modList.R

@@ -1,0 +1,49 @@
+library(car)
+library(cv)
+
+m1 <- lm(prestige ~ income + education, data=Duncan)
+m2 <- lm(prestige ~ income + education + type, data=Duncan)
+m3 <- glm(prestige ~ income, data=Duncan)
+m4 <- lm(prestige ~ income + education, data=Prestige)
+m5 <- lm(prestige ~ income + education + type, data=Prestige)
+m6 <- lm(women ~ income + education, data=Prestige)
+
+
+lst1 <- models(m1=m1, m2=m2)
+lst2 <- models(m1, m2)
+lst3 <- models(m1, m3)
+lst4 <- models(m1, m4)
+lst5 <- models(m4, m5)
+lst6 <- models(m4, m6)
+
+cv(models(m1=m1, m2=m2), data=Duncan)
+
+cv(models(m1=m1, m2=m2), data=Duncan, k="loo")
+
+
+# ---- application to Auto data
+
+data("Auto", package="ISLR2")
+for (p in 1:10){
+  assign(paste0("m.", p),
+         lm(mpg ~ poly(horsepower, p), data=Auto))
+}
+
+cv.auto.10 <- cv(models(m.1, m.2, m.3, m.4, m.5,
+                     m.6, m.7, m.8, m.9, m.10),
+              data=Auto, seed=2120)
+cv.auto.10[[2]]
+cv.auto.loo <- cv(models(m.1, m.2, m.3, m.4, m.5,
+                        m.6, m.7, m.8, m.9, m.10),
+                 data=Auto, k="loo")
+cv.auto.loo[[2]]
+cv.mse.10 <- sapply(cv.auto.10, function(x) x[["adj CV crit"]])
+cv.mse.loo <- sapply(cv.auto.loo, function(x) x[["CV crit"]])
+plot(c(1, 10), range(cv.mse.10, cv.mse.loo), type="n",
+     xlab="Degree of polynomial, p",
+     ylab="Cross-Validated MSE")
+lines(1:10, cv.mse.10, lwd=2, lty=1, col=2, pch=16, type="b")
+lines(1:10, cv.mse.loo, lwd=2, lty=2, col=3, pch=17, type="b")
+legend("topright", inset=0.02,
+       legend=c("10-Fold CV", "LOO CV"),
+       lwd=2, lty=2:1, col=3:2, pch=17:16)

### 2023-10-30 scratch/notes_georges/gm_addition_to_cv-edited.Rmd
move __notes_georges to scratch

--- 2023-10-11 scratch/gm_addition_to_cv-edited.Rmd
@@
 plot <- xyplot(y ~ x, data=Data[ 1:50,], group=group,
-               ylim=c(0, 20),
+               ylim=c(4, 16),
                par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
@@
 Data_long <- reshape(Data[1:50, ], direction = "long", sep = ".", 
-              timevar = "effect", varying = grep("\\.", names(Data.10)))
+              timevar = "effect", varying = grep("\\.", names(Data[1:50, ])))
 Data_long$id <- 1:nrow(Data_long)
@@
 ) |> update(
-  ylim=c(-20, 40),
+  ylim=c(4, 16),
   key=list(
@@
 ) |> update(
-  ylim = c(0, 20),
+  ylim=c(4, 16),
   key=list(

### 2023-10-30 scratch/notes_georges/gm_addition_to_cv-edited_v2_tmp.Rmd
move __notes_georges to scratch

--- 2023-10-30 scratch/notes_georges/gm_addition_to_cv-edited.Rmd
@@
 # Parameters:
-set.seed(9693) # set.seed(123) 
+set.seed(9693) 
 Nb <- 100     # number of groups
@@
   group = factor(rep(1:Nb, each=Nw)),
-  x = 20 + seq(1, Nb*Nw)
+  x = 20 + rep(seq(1, 50), length.out = Nw*Nb)
 ) |>
@@
 ) |> update(
-  ylim=c(4, 16),
+  ylim=c(-15, 35),
   key=list(
@@
        type="b",
-       auto.key=TRUE,
+       auto.key=list(reverse.rows=TRUE),
        scales=list(y=list(log=TRUE)))

### 2023-10-30 scratch/notes_georges/test.Rmd
move __notes_georges to scratch

--- 2023-10-11 scratch/gm_addition_to_cv-edited.Rmd
@@
 # Parameters:
-set.seed(123) # set.seed(9693)
+set.seed(9665) # set.seed(123) # set.seed(9693)
 Nb <- 10     # number of groups
@@
 #| fig.cap = "Cross-validation comparing random intercept models with varying fixed effects and with case- and cluster-based resampling."
+
 modlist <- models("~ 1"=fit0, "~ x"=fit1, "~ x + xm"=fit2)
 cvs_cases <- cv(modlist, data=Data, seed=9693)
+cvs_clusters <- cv(modlist, data=Data, cluster="group", seed=4576)
+
+{
+
+  # play
+  cv0 <- replicate(10,cv(fit0))
+  cv1 <- replicate(10,cv(fit1))
+  cv2 <- replicate(10,cv(fit2))
+
+  cv0
+  cv1
+  cv2
+
+  cv12 <- cv(fit1)
+  cv13 <- cv(fit1)
+  cv14 <- cv(fit1)
+  cv01 <- cv(fit0)
+  cv12 <- cv(fit0)
+  cv13 <- cv(fit0)
+  cv14 <- cv(fit0)
+  # 
+  # 
+  cv11
+  cv12
+  cv13
+  cv14
+
+  cvs_cases <- cv(modlist, data=Data)
   cvs_clusters <- cv(modlist, data=Data, cluster="group")
@@
          scales=list(y=list(log=TRUE)))
+}
 ```

### 2023-10-30 scratch/notes_georges/for-CRAN/R/cost-functions.R

@@ -1,0 +1,44 @@
+#' Cost Functions for Fitted Regression Models
+#'
+#' Compute cost functions (cross-validation criteria) for fitted
+#' regression models.
+#'
+#' @aliases costFunctions
+#'
+#' @param y response
+#' @param yhat fitted value
+#'
+#' @details
+#' Two cost functions (cross-validation criteria) are provided: (1)
+#' \code{mse()} returns the mean-squared error of prediction for
+#' a numeric response variable \code{y} and predictions \code{yhat}.
+#' (2) \code{BayesRule()} and \code{BayesRule2()} report the proportion
+#' of correct predictions for a dichotomous response variable \code{y}, assumed
+#' coded \code{0} and \code{1}. The \code{yhat} values are
+#' predicted probabilities and are rounded to 0 or 1. The distinction
+#' between \code{BayesRule()} and \code{BayesRule2()} is that the former
+#' checks that the \code{y} values are all either \code{0} or \code{1}
+#' and that the \code{yhat} values are all between 0 and 1, while
+#' the latter doesn't and is therefore faster.
+#'
+#' @describeIn cost-functions Mean-square error
+#' @export
+mse <- function(y, yhat){
+  mean((y - yhat)^2)
+}
+
+#' @describeIn cost-functions Bayes Rule for a binary response
+#' @export
+BayesRule <- function(y, yhat){
+  if (!all(y %in% c(0, 1))) stop("response values not all 0 or 1")
+  if (any(yhat < 0) || any(yhat > 1)) stop("fitted values outside of interval [0, 1]")
+  yhat <- round(yhat)
+  mean(y != yhat) # proportion in error
+}
+
+#' @describeIn cost-functions Bayes rule for a binary response (without bounds checking)
+#' @export
+BayesRule2 <- function(y, yhat){
+  yhat <- round(yhat)
+  mean(y != yhat) # proportion in error
+}

### 2023-10-30 scratch/notes_georges/for-CRAN/R/cv.modList.R

@@ -1,0 +1,218 @@
+#' Cross-Validate Several Models Fit to the Same Data
+#'
+#' A \code{\link{cv}()} method for an object of class  \code{"modlist"},
+#' created by the \code{models()} function. This \code{cv()} method simplifies
+#' the process of cross-validating several models on the same set of folds.
+#' \code{models()} performs some
+#' "sanity" checks, warning if the models are of different classes, and
+#' reporting an error if they are fit to apparently different data sets or
+#' different response variables.
+#' @param model a list of regression model objects,
+#' created by \code{models()}.
+#' @param data (required) the data set to which the models were fit.
+#' @param criterion the CV criterion (cost) function, defaults to
+#' \code{\link{mse}}.
+#' @param k the number of CV folds; may be omitted, in which case the value
+#' will depend on the default for the \code{cv()} method invoked for the
+#' individual models.
+#' @param reps number of replications of CV for each model.
+#' @param seed (optional) seed for R's pseudo-random-number generator,
+#' to be used to create the same set of CV folds for all of the models;
+#' if omitted, a seed will be randomly generated and saved.
+#' @param quietly If \code{TRUE} (the default), simple messages (for example about the
+#' value to which the random-number generator seed is set), but not warnings or
+#' errors, are suppressed.
+#' @param ... for \code{cv()}, additional arguments to be passed to the \code{cv()} method
+#' applied to each model. For \code{models()}, two or more competing models fit to the
+#' the same data; the several models may be named. For the \code{print()}
+#' method, arguments to be passed to the \code{print()} method for
+#' the individual model cross-validations. For the \code{plot()},
+#' method, arguments to be passed to the base \code{\link[base]{plot}()}
+#' function.
+#' @param x an object of class \code{"cvModList"} to be printed or plotted.
+#' @param y the name of the element in each \code{"cv"} object to be
+#' plotted; defaults to \code{"adj CV crit"}, if it exists, or to
+#' \code{"CV crit"}.
+#' @param xlab label for the x-axis (defaults to blank).
[…]

### 2023-11-01 R/cv.R
version -> 1.0.2; suppress (more) spurious warnings in cv.modList(); tweak main vignette

@@ -139,7 +139,7 @@ cv.default <- function(model, data=insight::get_data(model),
     message("R RNG seed set to ", seed)
   } else {
     if (reps > 1) stop("reps should not be > 1 for n-fold CV")
-    if (!missing(seed) && !is.null(seed)) warning("seed ignored for n-fold CV")
+    if (!missing(seed) && !is.null(seed)) message("Note: seed ignored for n-fold CV")
     seed <- NULL
   }
   nk <-  n %/% k # number of cases in each fold
@@ -286,7 +286,7 @@ cv.lm <- function(model, data=insight::get_data(model), criterion=mse, k=10,
   }
   if (k == n){
     if (reps > 1) stop("reps should not be > 1 for n-fold CV")
-    if (!missing(seed) && !is.null(seed)) warning("seed ignored for n-fold CV")
+    if (!missing(seed) && !is.null(seed)) message("Note: seed ignored for n-fold CV")
     seed <- NULL
   }
 
@@ -410,7 +410,7 @@ cv.glm <- function(model, data=insight::get_data(model), criterion=mse, k=10,
   }
   if (k == n){
     if (reps > 1) stop("reps should not be > 1 for n-fold CV")
-    if (!missing(seed) && !is.null(seed)) warning("seed ignored for n-fold CV")
+    if (!missing(seed) && !is.null(seed)) message("Note: seed ignored for n-fold CV")
     seed <- NULL
   }
   method <- match.arg(method)

### 2023-11-01 R/cvSelect.R

@@ -83,7 +83,7 @@ cvSelect <- function(procedure, data, k=10, reps=1,
     if (reps > 1) stop("reps should not be > 1 for n-fold CV")
     if (k == n){
       if (reps > 1) stop("reps should not be > 1 for n-fold CV")
-      if (!missing(seed) && !is.null(seed)) warning("seed ignored for n-fold CV")
+      if (!missing(seed) && !is.null(seed)) message("Note: seed ignored for n-fold CV")
       seed <- NULL
     }
     seed <- NULL

### 2023-11-16 R/cv-mixed-models.R
fix parallel computation bug in cvMixed()

@@ -12,6 +12,7 @@
 #' It can be used to extend \code{cv()} to other classes of mixed-effects models.
 #'
 #' @param model a mixed-effects model object for which a \code{cv()} method is available.
+#' @param fun the mixed-modeling function employed.
 #' @param data data frame to which the model was fit (not usually necessary)
 #' @param criterion cross-validation ("cost" or lack-of-fit) criterion function of form \code{f(y, yhat)}
 #'        where \code{y} is the observed values of the response and
@@ -70,9 +71,17 @@
 #' cv(fm2) # LOO CV of cases
 #' cv(fm2, clusterVariables="Subject", k=5, seed=321) # 5-fold CV of clusters
 #'
+#' library("glmmTMB")
+#' # from ?glmmTMB
+#' (m1 <- glmmTMB(count ~ mined + (1|site),
+#'                zi=~mined,
+#'                family=poisson, data=Salamanders))
+#' cv(m1, seed=97816, k=5) # 5-fold CV of clusters
+#' cv(m1, seed=34506, k=5) # 5-fold CV of cases
 #' @describeIn cvMixed not to be called directly
 #' @export
 cvMixed <- function(model,
+                    fun,
                     data=insight::get_data(model),
                     criterion=mse,
                     k,
@@ -84,18 +93,41 @@ cvMixed <- function(model,
                     predict.cases.args=list(object=model, newdata=data),
                     ...){
 
-  f.clusters <- function(i){
+  # f.clusters <- function(i){
+  #   indices.i <- indices[starts[i]:ends[i]]
+  #   index <- selectClusters(clusters[- indices.i, , drop=FALSE], data=data)
+  #   predict.clusters.args$object <- update(model, data=data[index, ], ...)
+  #   fit.all.i <- do.call(predict, predict.clusters.args)
+  #   fit.i <- fit.all.i[!index]
+  #   c(criterion(y[!index], fit.i), criterion(y, fit.all.i))
+  # }
+
+  f.clusters <- function(i, predict.clusters.args, predict.cases.args, ...){
     indices.i <- indices[starts[i]:ends[i]]
     index <- selectClusters(clusters[- indices.i, , drop=FALSE], data=data)
-    predict.clusters.args$object <- update(model, data=data[index, ], ...)
+    update.args <- list(...)
+    update.args$object <- model
[…]

### 2023-11-16 tests/testthat/test-cvMixed.R

@@ -1,0 +1,78 @@
+# test parallel computations for mixed models
+
+data("sleepstudy", package="lme4")
+fm1 <- lme4::lmer(Reaction ~ Days + (Days | Subject), sleepstudy)
+
+test_that("parallel computations lmer loo clusters", {
+  expect_equal(cv(fm1, clusterVariables="Subject"),
+               cv(fm1, clusterVariables="Subject", ncores=2))
+})
+
+test_that("parallel computations lmer k-fold clusters", {
+  expect_equal(cv(fm1, clusterVariables="Subject", k=5, seed=123),
+               cv(fm1, clusterVariables="Subject", k=5, seed=123,
+                  ncores=2))
+})
+
+test_that("parallel computations lmer k-fold cases", {
+  expect_equal(suppressWarnings(cv(fm1, k="loo")),
+               cv(fm1, k="loo", ncores=2))
+})
+
+test_that("parallel computations lmer k-fold cases", {
+  expect_equal(cv(fm1, k=5, seed=321),
+               cv(fm1, k=5, ncores=2, seed=321))
+})
+
+data("Orthodont", package="nlme")
+fm2 <- nlme::lme(distance ~ age + Sex, data = Orthodont,
+           random = ~ 1 | Subject)
+
+test_that("parallel computations lme loo clusters", {
+  expect_equal(cv(fm2, clusterVariables="Subject"),
+               cv(fm2, clusterVariables="Subject", ncores=2))
+})
+
+test_that("parallel computations lme k-fold clusters", {
+  expect_equal(cv(fm2, clusterVariables="Subject", k=5, seed=123),
+               cv(fm2, clusterVariables="Subject", k=5, seed=123,
+                  ncores=2))
+})
+
+test_that("parallel computations lme k-fold cases", {
+  expect_equal(cv(fm2, k=5, seed=321),
+               cv(fm2, k=5, seed=321, ncores=2))
+})
+
+test_that("parallel computations lme LOO cases", {
+  expect_equal(cv(fm2, k="loo"),
+               cv(fm2, k="loo", ncores=2))
+})
+
+data("Salamanders", package="glmmTMB")
+m1 <- glmmTMB::glmmTMB(count ~ mined + (1|site),
+                       zi=~mined,
+                       family=poisson, data=Salamanders)
+
+test_that("parallel computations glmmTMB k-fold cases", {
+  expect_equal(cv(m1, seed=123, k=5),
[…]

### 2023-11-16 vignettes/cv-extend.R

@@ -117,6 +117,7 @@ cv.glmmPQL <- function(model, data = model$data, criterion = mse,
                      k, reps = 1, seed, ncores = 1, clusterVariables, ...){
   cvMixed(
     model,
+    fun=MASS::glmmPQL,
     data=data,
     criterion=criterion,
     k=k,

### 2023-11-20 R/cv-mixed-models.R
add Vehtari CV FAQ reference; mixed-model methods no longer "experimental"

@@ -6,8 +6,7 @@
 #' fit by the \code{\link[nlme]{lme}()} function in the \pkg{nlme}
 #' package; and for models of class \code{"glmmTMB"} fit by the
 #' \code{\link[glmmTMB]{glmmTMB}()} function in the \pkg{glmmTMB} package.
-#' The implementations here should be regarded
-#' as experimental. The \code{cvMixed()} function is meant to be called by
+#' The \code{cvMixed()} function is meant to be called by
 #' \code{cv()} methods for mixed-effect models and not directly by the user.
 #' It can be used to extend \code{cv()} to other classes of mixed-effects models.
 #'

### 2023-11-27 R/Pigs.R
version -> 1.1.0; add Pigs data; add crossed-re example to cv vignette; update vignettes

@@ -1,0 +1,28 @@
+#' @name Pigs
+#' @docType data
+#'
+#' @title Body Weights of 48 Pigs in 9 Successive Weeks
+#'
+#' @description This data set appears in Table 3.1 of Diggle, Liang,
+#' and Zeger (1994).
+#'
+#' @usage data("Pigs", package = "cv")
+#'
+#' @format A data frame with 432 rows and 3 columns.
+#' \describe{
+#'   \item{id}{Pig id number, 1--48.}
+#'   \item{week}{Week number, 1--9.}
+#'   \item{weight}{Weight in kg.}
+#'   }
+#'
+#' @source P. J. Diggle, K.-Y. Liang, and S. L. Zeger,
+#' \emph{Analysis of Longitudinal Data} (Oxford, 1994).
+#'
+#' @examples
+#' library("lme4")
+#' m.p <- lmer(weight ~ week + (1 | id) + (1 | week),
+#'             data=Pigs, REML=FALSE,
+#'             control=lmerControl(optimizer="bobyqa"))
+#' summary(m.p)
+#' cv(m.p, clusterVariables=c("id", "week"), k=10, seed=8469)
+"Pigs"

### 2023-11-27 R/cost-functions.R

@@ -13,7 +13,9 @@
 #' @details
 #' Cost functions (cross-validation criteria) are meant to measure lack-of-fit. Three cost functions are provided:
 #' 1. \code{mse()} returns the mean-squared error of prediction for
-#' a numeric response variable \code{y} and predictions \code{yhat}.
+#' a numeric response variable \code{y} and predictions \code{yhat};
+#' \code{rmse()} returns the root-mean-squared error and is just the
+#' square-root of \code{mse()}.
 #' 2. \code{medAbsErr()} returns the median absolute error of predictor for a numeric
 #' response \code{y} and predictions \code{yhat}.
 #' 3. \code{BayesRule()} and \code{BayesRule2()} report the proportion
@@ -33,6 +35,12 @@
 #' @export
 mse <- function(y, yhat){
   mean((y - yhat)^2)
+}
+
+#' @describeIn cost-functions Root-mean-square error
+#' @export
+rmse <- function(y, yhat){
+  sqrt(mean((y - yhat)^2))
 }
 
 #' @describeIn cost-functions Median absolute error

### 2023-11-27 R/cv-mixed-models.R

@@ -29,7 +29,7 @@
 #' @param ncores number of cores to use for parallel computations
 #'        (default is \code{1}, i.e., computations aren't done in parallel)
 #' @param clusterVariables a character vector of names of the variables
-#' defining clusters for a mixed model with nested random effects;
+#' defining clusters for a mixed model with nested or crossed random effects;
 #' if missing, cross-validation is performed for individual cases rather than
 #' for clusters
 #' @param predict.clusters.args a list of arguments to be used to predict
@@ -51,7 +51,6 @@
 #' individual observations. If the former, predictions are based only on fixed
 #' effects; if the latter, predictions include the random effects (i.e., are the
 #' best linear unbiased predictors or "BLUPS").
-#' Only mixed models with fully nested random effects are supported.
 #'
 #' @seealso \code{\link{cv}}, \code{\link[lme4]{lmer}}, \code{\link[lme4]{glmer}},
 #' \code{\link[nlme]{lme}}, \code{\link[glmmTMB]{glmmTMB}}

### 2023-11-27 vignettes/cv.R

@@ -313,6 +313,37 @@ plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")
 cvs_cases <- cv(modlist, data=Data, seed=9693)
 plot(cvs_cases, main="Model Comparison, Case-Based CV")
 
+## ----pigs---------------------------------------------------------------------
+head(Pigs, 9)
+head(xtabs(~ id + week, data=Pigs), 3)
+tail(xtabs(~ id + week, data=Pigs), 3)
+
+## ----pigs-graph---------------------------------------------------------------
+plot(weight ~ week, data=Pigs, type="n")
+for (i in unique(Pigs$id)){
+  with(Pigs, lines(x=1:9, y=Pigs[id == i, "weight"],
+                   col="gray"))
+}
+abline(lm(weight ~ week, data=Pigs), col="blue", lwd=2)
+lines(with(Pigs, loess.smooth(week, weight, span=0.5)),
+      col="magenta", lty=2, lwd=2)
+
+## ----pigs-lmer----------------------------------------------------------------
+m.p <- lmer(weight ~ week + (1 | id) + (1 | week),
+            data=Pigs, REML=FALSE, # i.e., ML
+            control=lmerControl(optimizer="bobyqa"))
+summary(m.p)
+
+## ----pigs-cv------------------------------------------------------------------
+cv(m.p, clusterVariables="id")
+
+cv(m.p, clusterVariables="week")
+
+cv(m.p, clusterVariables=c("id", "week"), k=10, seed=8469)
+
+## ----pigs-cv-cases------------------------------------------------------------
+cv(m.p, k=10, seed=8469)
+
 ## ----mroz-reps----------------------------------------------------------------
 cv(m.mroz, criterion=BayesRule, seed=248, reps=5, 
    method="Woodbury")

### 2023-12-04 R/cv-mixed-models.R
update selectTransAndStepAIC(), vignettes, etc.

@@ -158,7 +158,13 @@ cvMixed <- function(model,
   ends <- cumsum(folds) # end of each fold
   starts <- c(1, ends + 1)[-(k + 1)] # start of each fold
   indices <- if (n > k) sample(n, n)  else 1:n # permute clusters/cases
-  yhat <- numeric(n)
+  yhat <- if (is.factor(y)){
+    factor(rep(NA, n), levels=levels(y))
+  } else if (is.character(y)) {
+    character(n)
+  } else {
+    numeric(n)
+  }
 
   if (ncores > 1L){
     cl <- makeCluster(ncores)

### 2023-12-04 R/cv.R

@@ -164,7 +164,13 @@ cv.default <- function(model, data=insight::get_data(model),
   ends <- cumsum(folds) # end of each fold
   starts <- c(1, ends + 1)[-(k + 1)] # start of each fold
   indices <- if (n > k) sample(n, n)  else 1:n # permute cases
-  yhat <- numeric(n)
+  yhat <- if (is.factor(y)){
+    factor(rep(NA, n), levels=levels(y))
+  } else if (is.character(y)) {
+    character(n)
+  } else {
+    numeric(n)
+  }
   if (ncores > 1){
     dots <- list(...)
     cl <- makeCluster(ncores)

### 2023-12-04 R/cvSelect.R

@@ -10,6 +10,11 @@
 #'
 #' @param procedure a model-selection procedure function (see Details).
 #' @param data full data frame for model selection.
+#' @param y.expression normally the response variable is found from the
+#' \code{model} argument; but if, for a particular selection procedure, the
+#' \code{model} argument is absent, or if the response can't be inferred from the
+#' model, the response can be specified by an expression, such as \code{expression(log(income))},
+#' to be evaluated within the data set provided by the \code{data} argument.
 #' @param k perform k-fold cross-validation (default is 10); \code{k}
 #' may be a number or \code{"loo"} or \code{"n"} for n-fold (leave-one-out)
 #' cross-validation.
@@ -44,10 +49,12 @@
 #'   \item{other arguments}{to be passed via \code{...}
 #'   from \code{cvSelect()}.}
 #' }
-#' \code{procedure()} should return a two-element vector with the result
-#' of applying a cross-validation criterion to the cases in
-#' the current fold for the model deleting that fold, and to
-#' all of the cases, again for the model deleting the current fold.
+#' \code{procedure()} should return a list with the following
+#' named elements: \code{fit.i}, the vector of predicted values for the cases in
+#' the current fold computed from the model omitting these cases;
+#' \code{crit.all.i}, the CV criterion computed for all of the cases using
+#' the model omitting the current fold; and (optionally) \code{coefficients},
+#' parameter estimates from the model computed omitting the current fold.
 #'
 #' When the \code{indices} argument is missing, \code{procedure()} returns the cross-validation criterion for all of the cases based on
 #' the model fit to all of the cases.
@@ -104,7 +111,13 @@ cvSelect <- function(procedure, data, criterion=mse,
   ends <- cumsum(folds) # end of each fold
   starts <- c(1, ends + 1)[-(k + 1)] # start of each fold
   indices <- if (n > k) sample(n, n)  else 1:n # permute cases
-  yhat <- numeric(n)
+  yhat <- if (is.factor(y)){
+    factor(rep(NA, n), levels=levels(y))
+  } else if (is.character(y)) {
+    character(n)
+  } else {
+    numeric(n)
+  }
   crit.all.i <- numeric(k)
 
   if (ncores > 1){
@@ -151,9 +164,11 @@ cvSelect <- function(procedure, data, criterion=mse,
   if (reps == 1) {
     return(result)
   } else {
+    if (missing(y.expression)) y.expression <- NULL
[…]

### 2023-12-05 R/cv-mixed-models.R
change default cv criterion to rmse()

@@ -16,7 +16,7 @@
 #' @param data data frame to which the model was fit (not usually necessary)
 #' @param criterion cross-validation ("cost" or lack-of-fit) criterion function of form \code{f(y, yhat)}
 #'        where \code{y} is the observed values of the response and
-#'        \code{yhat} the predicted values; the default is \code{\link{mse}}
+#'        \code{yhat} the predicted values; the default is \code{\link{rmse}}
 #'        (the mean-squared error)
 #' @param k perform k-fold cross-validation; \code{k}
 #' may be a number or \code{"loo"} or \code{"n"} for n-fold (leave-one-out)
@@ -82,7 +82,7 @@
 cvMixed <- function(model,
                     package,
                     data=insight::get_data(model),
-                    criterion=mse,
+                    criterion=rmse,
                     k,
                     reps=1,
                     seed,
@@ -229,7 +229,7 @@ cvMixed <- function(model,
 
 #' @describeIn cvMixed \code{cv()} method
 #' @export
-cv.merMod <- function(model, data = insight::get_data(model), criterion = mse,
+cv.merMod <- function(model, data = insight::get_data(model), criterion = rmse,
                       k, reps = 1, seed, ncores = 1, clusterVariables, ...){
   cvMixed(
     model,
@@ -255,7 +255,7 @@ cv.merMod <- function(model, data = insight::get_data(model), criterion = mse,
 
 #' @describeIn cvMixed \code{cv()} method
 #' @export
-cv.lme <- function(model, data = insight::get_data(model), criterion = mse,
+cv.lme <- function(model, data = insight::get_data(model), criterion = rmse,
                    k, reps = 1, seed, ncores = 1, clusterVariables, ...){
   cvMixed(
     model,
@@ -278,7 +278,7 @@ cv.lme <- function(model, data = insight::get_data(model), criterion = mse,
 
 #' @describeIn cvMixed \code{cv()} method
 #' @export
-cv.glmmTMB <- function(model, data = insight::get_data(model), criterion = mse,
+cv.glmmTMB <- function(model, data = insight::get_data(model), criterion = rmse,
                        k, reps = 1, seed, ncores = 1, clusterVariables, ...){
   cvMixed(
     model,

### 2023-12-05 R/cvSelect.R

@@ -72,7 +72,7 @@
 #' \code{\link[car]{powerTransform}}
 #'
 #' @export
-cvSelect <- function(procedure, data, criterion=mse,
+cvSelect <- function(procedure, data, criterion=rmse,
                      model, y.expression,
                      k=10, reps=1,
                      save.coef = k <= 10,
@@ -197,7 +197,7 @@ cvSelect <- function(procedure, data, criterion=mse,
 #'          AIC=FALSE, k=5, reps=3) # via BIC
 #' @export
 selectStepAIC <- function(data, indices,
-                          model, criterion=mse, AIC=TRUE,
+                          model, criterion=rmse, AIC=TRUE,
                           save.coef=TRUE, ...){
   y <- getResponse(model)
   if (missing(indices)) {
@@ -302,7 +302,7 @@ yjPowerInverse <- function(y, lambda) {
 #' cv(m.pres, seed=123)
 #' @export
 selectTrans <- function(data, indices, save.coef=TRUE, model,
-                        criterion=mse, predictors, response,
+                        criterion=rmse, predictors, response,
                         family=c("bcPower", "bcnPower", "yjPower", "basicPower"),
                         family.y=c("bcPower", "bcnPower", "yjPower", "basicPower"),
                         rounded=TRUE,
@@ -426,7 +426,7 @@ selectTransStepAIC <- function(data,
                                indices,
                                save.coef = TRUE,
                                model,
-                               criterion = mse,
+                               criterion = rmse,
                                predictors,
                                response,
                                family = c("bcPower", "bcnPower", "yjPower", "basicPower"),

### 2023-12-06 R/cost-functions.R
small edits to cv vignette and docs

@@ -28,8 +28,11 @@
 #' the latter doesn't and is therefore faster.
 #'
 #' @returns In general, cost functions should return a single numeric
-#' value measuring lack-of-fit. `mse()` returns the mean-squared error; `BayesRule()` and
-#' `BayesRule2()` return the proportion of misclassified cases; etc.
+#' value measuring lack-of-fit. `mse()` returns the mean-squared error;
+#' `rmse()` returns the root-mean-squared error;
+#' `medAbsErr()` returns the median absolute error;
+#' and BayesRule()` and
+#' `BayesRule2()` return the proportion of misclassified cases.
 #' @examples
 #' data("Duncan", package="carData")
 #' m.lm <- lm(prestige ~ income + education, data=Duncan)

### 2023-12-06 R/cv-mixed-models.R

@@ -17,7 +17,7 @@
 #' @param criterion cross-validation ("cost" or lack-of-fit) criterion function of form \code{f(y, yhat)}
 #'        where \code{y} is the observed values of the response and
 #'        \code{yhat} the predicted values; the default is \code{\link{rmse}}
-#'        (the mean-squared error)
+#'        (the root-mean-squared error)
 #' @param k perform k-fold cross-validation; \code{k}
 #' may be a number or \code{"loo"} or \code{"n"} for n-fold (leave-one-out)
 #' cross-validation; the default is \code{10} if cross-validating individual

### 2023-12-19 R/cost-functions.R
make mse() the default; differentiate casewise-averaged criteria; compute CIs for bias-adj criteria

@@ -46,7 +46,7 @@
 #' @export
 mse <- function(y, yhat){
   result <- mean((y - yhat)^2)
-  attr(result, "linear") <- TRUE
+  attr(result, "casewise average") <- TRUE
   result
 }
 
@@ -70,7 +70,7 @@ BayesRule <- function(y, yhat){
   if (any(yhat < 0) || any(yhat > 1)) stop("fitted values outside of interval [0, 1]")
   yhat <- round(yhat)
   result <- mean(y != yhat) # proportion in error
-  attr(result, "linear") <- TRUE
+  attr(result, "casewise average") <- TRUE
   result
 }
 
@@ -79,6 +79,6 @@ BayesRule <- function(y, yhat){
 BayesRule2 <- function(y, yhat){
   yhat <- round(yhat)
   result <- mean(y != yhat) # proportion in error
-  attr(result, "linear") <- TRUE
+  attr(result, "casewise average") <- TRUE
   result
 }

### 2023-12-19 R/cv-mixed-models.R

@@ -16,13 +16,17 @@
 #' @param data data frame to which the model was fit (not usually necessary)
 #' @param criterion cross-validation ("cost" or lack-of-fit) criterion function of form \code{f(y, yhat)}
 #'        where \code{y} is the observed values of the response and
-#'        \code{yhat} the predicted values; the default is \code{\link{rmse}}
-#'        (the root-mean-squared error)
+#'        \code{yhat} the predicted values; the default is \code{\link{mse}}
+#'        (the mean-squared error)
 #' @param k perform k-fold cross-validation; \code{k}
 #' may be a number or \code{"loo"} or \code{"n"} for n-fold (leave-one-out)
 #' cross-validation; the default is \code{10} if cross-validating individual
 #' cases and \code{"loo"} if cross-validating clusters.
-#' @param reps number of times to replicate k-fold CV (default is \code{1})
+#' @param reps number of times to replicate k-fold CV (default is \code{1}),
+#' @param confint if \code{TRUE} (the default if the number of cases is 400
+#' or greater), compute a confidence interval for the bias-corrected CV
+#' criterion, if the criterion is the average of casewise components.
+#' @param level confidence level (default \code{0.95}).
 #' @param seed for R's random number generator; optional, if not
 #' supplied a random seed will be selected and saved; not needed
 #' for n-fold cross-validation
@@ -75,16 +79,17 @@
 #' (m1 <- glmmTMB(count ~ mined + (1|site),
 #'                zi=~mined,
 #'                family=poisson, data=Salamanders))
-#' cv(m1, seed=97816, k=5) # 5-fold CV of clusters
+#' cv(m1, seed=97816, k=5, clusterVariables="site") # 5-fold CV of clusters
 #' cv(m1, seed=34506, k=5) # 5-fold CV of cases
 #' @describeIn cvMixed not to be called directly
 #' @export
 cvMixed <- function(model,
                     package,
                     data=insight::get_data(model),
-                    criterion=rmse,
+                    criterion=mse,
                     k,
                     reps=1,
+                    confint, level=0.95,
                     seed,
                     ncores=1,
                     clusterVariables,
@@ -184,7 +189,7 @@ cvMixed <- function(model,
     }
   }
   cv <- criterion(y, yhat)
-  args <-
+ # args <-
[…]

### 2023-12-19 R/cv.modList.R

@@ -11,7 +11,7 @@
 #' created by \code{models()}.
 #' @param data (required) the data set to which the models were fit.
 #' @param criterion the CV criterion ("cost" or lack-of-fit) function, defaults to
-#' \code{\link{rmse}}.
+#' \code{\link{mse}}.
 #' @param k the number of CV folds; may be omitted, in which case the value
 #' will depend on the default for the \code{cv()} method invoked for the
 #' individual models.
@@ -91,7 +91,7 @@ models <- function(...){
 
 #' @describeIn models \code{cv()} method for \code{"modList"} objects
 #' @exportS3Method
-cv.modList <- function(model, data, criterion=rmse, k, reps=1, seed, quietly=TRUE, ...){
+cv.modList <- function(model, data, criterion=mse, k, reps=1, seed, quietly=TRUE, ...){
   n.models <- length(model)
   if (missing(seed)) seed <- sample(1e6, 1L)
   result <- vector(n.models, mode="list")

### 2023-12-19 R/cvSelect.R

@@ -18,6 +18,10 @@
 #' @param k perform k-fold cross-validation (default is 10); \code{k}
 #' may be a number or \code{"loo"} or \code{"n"} for n-fold (leave-one-out)
 #' cross-validation.
+#' @param confint if \code{TRUE} (the default if the number of cases is 400
+#' or greater), compute a confidence interval for the bias-corrected CV
+#' criterion, if the criterion is the average of casewise components.
+#' @param level confidence level (default \code{0.95}).
 #' @param save.coef save the coefficients from the selected models? (default is \code{TRUE} if
 #' \code{k} is 10 or smaller, \code{FALSE} otherwise)
 #' @param reps number of times to replicate k-fold CV (default is \code{1})
@@ -72,9 +76,10 @@
 #' \code{\link[car]{powerTransform}}
 #'
 #' @export
-cvSelect <- function(procedure, data, criterion=rmse,
+cvSelect <- function(procedure, data, criterion=mse,
                      model, y.expression,
-                     k=10, reps=1,
+                     k=10, confint = n >= 400, level=0.95,
+                     reps=1,
                      save.coef = k <= 10,
                      seed, ncores=1, ...){
   n <- nrow(data)
@@ -156,8 +161,24 @@ cvSelect <- function(procedure, data, criterion=rmse,
   }
   cv <- criterion(y, yhat)
   cv.full <- procedure(data, model=model, criterion=criterion, ...)
-  adj.cv <- cv + cv.full - weighted.mean(crit.all.i, folds)
+
+  casewise.average <- attr(cv, "casewise average")
+  if (!is.null(casewise.average) && casewise.average) {
+    loss <- getLossFn(criterion) # casewise loss function
+    adj.cv <- cv + cv.full - weighted.mean(crit.all.i, folds)
+    se.cv <- sd(loss(y, yhat))/sqrt(n)
+    halfwidth <- qnorm(1 - (1 - level)/2)*se.cv
+    ci <- if (confint) c(lower = adj.cv - halfwidth, upper = adj.cv + halfwidth,
+                         level=round(level*100)) else NULL
+  } else {
+    adj.cv <- NULL
+    ci <- NULL
+  }
+
+  # adj.cv <- cv + cv.full - weighted.mean(crit.all.i, folds)
+
   result <- list("CV crit" = cv, "adj CV crit" = adj.cv, "full crit" = cv.full,
+                 "confint" = ci,
[…]

### 2023-12-19 vignettes/cv.R

@@ -50,23 +50,23 @@ for (p in 1:5){
 legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=2,
        title="Degree", inset=0.02)
 
-## ----mpg-horsepower-RMSE-se---------------------------------------------------
-library("cv") # for rmse() and other functions
+## ----mpg-horsepower-MSE-se----------------------------------------------------
+library("cv") # for mse() and other functions
 
-se <- rmse <- numeric(10)
+var <- mse <- numeric(10)
 for (p in 1:10){
   m <- lm(mpg ~ poly(horsepower, p), data=Auto)
-  rmse[p] <- rmse(Auto$mpg, fitted(m))
-  se[p] <- summary(m)$sigma
+  mse[p] <- mse(Auto$mpg, fitted(m))
+  var[p] <- summary(m)$sigma^2
 }
 
-plot(c(1, 10), range(rmse, se), type="n",
+plot(c(1, 10), range(mse, var), type="n",
      xlab="Degree of polynomial, p",
-     ylab="Estimated Error")
-lines(1:10, rmse, lwd=2, lty=1, col=2, pch=16, type="b")
-lines(1:10, se, lwd=2, lty=2, col=3, pch=17, type="b")
+     ylab="Estimated Squared Error")
+lines(1:10, mse, lwd=2, lty=1, col=2, pch=16, type="b")
+lines(1:10, var, lwd=2, lty=2, col=3, pch=17, type="b")
 legend("topright", inset=0.02,
-       legend=c(expression(hat(sigma)), "RMSE"),
+       legend=c(expression(hat(sigma)^2), "MSE"),
        lwd=2, lty=2:1, col=3:2, pch=17:16)
 
 ## ----cv-lm-1------------------------------------------------------------------
@@ -105,13 +105,13 @@ cv.auto.loo <- cv(models(m.1, m.2, m.3, m.4, m.5,
 cv.auto.loo[1:2] # linear and quadratic models
 
 ## ----polynomial-regression-CV-graph-------------------------------------------
-cv.rmse.10 <- sapply(cv.auto.10, function(x) x[["adj CV crit"]])
-cv.rmse.loo <- sapply(cv.auto.loo, function(x) x[["CV crit"]])
-plot(c(1, 10), range(cv.rmse.10, cv.rmse.loo), type="n",
+cv.mse.10 <- sapply(cv.auto.10, function(x) x[["adj CV crit"]])
[…]

### 2023-12-19 scratch/notes_john/plot.cvModLiist-tests.R
modify plot.cvModList() to plot CIs

@@ -1,0 +1,51 @@
+data(Auto, package="ISLR2")
+
+for (p in 1:10){
+  assign(paste0("m.", p),
+         lm(mpg ~ poly(horsepower, p), data=Auto))
+}
+
+cv.auto.10.noci <- cv(models(m.1, m.2, m.3, m.4, m.5,
+                        m.6, m.7, m.8, m.9, m.10),
+                 data=Auto, seed=2120)
+
+plot(cv.auto.10.noci)
+
+cv.auto.10 <- cv(models(m.1, m.2, m.3, m.4, m.5,
+                        m.6, m.7, m.8, m.9, m.10),
+                 data=Auto, seed=2120, confint=TRUE,
+                 level=.50)
+
+plot(cv.auto.10)
+
+cv.auto.10.someci <- cv.auto.10
+cv.auto.10.someci[[2]] <- cv.auto.10.noci[[2]]
+cv.auto.10.someci[[7]] <- cv.auto.10.noci[[7]]
+plot(cv.auto.10.someci)
+
+cv.auto.10.reps <- cv(models(m.1, m.2, m.3, m.4, m.5,
+                             m.6, m.7, m.8, m.9, m.10),
+                      data=Auto, seed=2120, reps=10,
+                      confint=TRUE, level=0.50)
+
+plot(cv.auto.10.reps)
+
+data(Duncan, package="carData")
+m1 <- lm(prestige ~ income + education, data=Duncan)
+m2 <- lm(prestige ~ income + education + type, data=Duncan)
+m3 <- lm(prestige ~ (income + education)*type, data=Duncan)
+(cv.models <- cv(models(m1=m1, m2=m2, m3=m3),
+                  data=Duncan, seed=7949, reps=5))
+plot(cv.models)
+
+(cv.models.ci <- cv(models(m1=m1, m2=m2, m3=m3),
+                 data=Duncan, seed=5962, confint=TRUE,
+                 level=0.50))
+plot(cv.models.ci) # nb: n too small for accurate CIs
+
+
+cv.auto.loo <- cv(models(m.1, m.2, m.3, m.4, m.5,
+                         m.6, m.7, m.8, m.9, m.10),
+                  data=Auto, k="loo")
+cv.auto.loo
+plot(cv.auto.loo)

### 2023-12-20 R/cv.R
update extending cv vignette; improve getLossFn()

@@ -634,11 +634,14 @@ getLossFn <- function(criterion){
   which <- grepl("result <- mean\\(", expressions)
   if (!(any(which))) stop(deparse(substitute(criterion)),
                           " does not set 'result <- mean(. . .)'")
-  expression <- sub(")$" , "",
-                    sub("result <- mean\\(", "result <- ",
-                        expressions[which][sum(which)])
-  )
+  expression <- sub("result <- mean\\(", "result <- ",
+                    expressions[which][sum(which)])
+  close.parens <- gregexpr(")", expression)[[1]]
+  expression <- substr(expression, 1,
+                       close.parens[length(close.parens)] - 1)
   expressions[which] <- expression
+  which <- grepl("attr\\(", expressions)
+  expressions <- expressions[!which]
   eval(parse(text=paste0("function(y, yhat) ",
                          paste(expressions, collapse="\n"),
                          "}")))

### 2023-12-21 R/cv.R
implement alternative approach to getLossFn()

@@ -199,9 +199,9 @@ cv.default <- function(model, data=insight::get_data(model),
   }
   cv <- criterion(y, yhat)
   cv.full <- criterion(y, predict(model, type=type, ...))
-  casewise.average <- attr(cv, "casewise average")
-  if (!is.null(casewise.average) && casewise.average) {
-    loss <- getLossFn(criterion) # casewise loss function
+  casewise.loss <- attr(cv, "casewise loss")
+  if (!is.null(casewise.loss)) {
+    loss <- getLossFn(cv) # casewise loss function
     adj.cv <- cv + cv.full -
       weighted.mean(sapply(result, function(x) x$crit.all.i), folds)
     se.cv <- sd(loss(y, yhat))/sqrt(n)
@@ -410,9 +410,9 @@ cv.lm <- function(model, data=insight::get_data(model),
   }
   cv <- criterion(y, yhat)
   cv.full <- criterion(y, fitted(model))
-  casewise.average <- attr(cv, "casewise average")
-  if (!is.null(casewise.average) && casewise.average) {
-    loss <- getLossFn(criterion) # casewise loss function
+  casewise.loss <- attr(cv, "casewise loss")
+  if (!is.null(casewise.loss)) {
+    loss <- getLossFn(cv) # casewise loss function
     adj.cv <- cv + cv.full -
       weighted.mean(sapply(result, function(x) x$crit.all.i), folds)
     se.cv <- sd(loss(y, yhat))/sqrt(n)
@@ -562,9 +562,9 @@ cv.glm <- function(model, data=insight::get_data(model),
     }
     cv <- criterion(y, yhat)
     cv.full <- criterion(y, fitted(model))
-    casewise.average <- attr(cv, "casewise average")
-    if (!is.null(casewise.average) && casewise.average) {
-      loss <- getLossFn(criterion) # casewise loss function
+    casewise.loss <- attr(cv, "casewise loss")
+    if (!is.null(casewise.loss)) {
+      loss <- getLossFn(cv) # casewise loss function
       adj.cv <- cv + cv.full -
         weighted.mean(sapply(result, function(x) x$crit.all.i), folds)
       se.cv <- sd(loss(y, yhat))/sqrt(n)
@@ -629,19 +629,8 @@ summarizeReps <- function(x){
        "adj CV crit range" = adjCVcritRange)
 }
 
-getLossFn <- function(criterion){
-  expressions <- as.vector(as.character(functionBody(criterion)))
-  which <- grepl("result <- mean\\(", expressions)
[…]

### 2023-12-21 R/cvSelect.R

@@ -162,9 +162,9 @@ cvSelect <- function(procedure, data, criterion=mse,
   cv <- criterion(y, yhat)
   cv.full <- procedure(data, model=model, criterion=criterion, ...)
 
-  casewise.average <- attr(cv, "casewise average")
-  if (!is.null(casewise.average) && casewise.average) {
-    loss <- getLossFn(criterion) # casewise loss function
+  casewise.loss <- attr(cv, "casewise loss")
+  if (!is.null(casewise.loss)) {
+    loss <- getLossFn(cv) # casewise loss function
     adj.cv <- cv + cv.full - weighted.mean(crit.all.i, folds)
     se.cv <- sd(loss(y, yhat))/sqrt(n)
     halfwidth <- qnorm(1 - (1 - level)/2)*se.cv

### 2023-12-21 R/cv.R
update docs

@@ -33,9 +33,10 @@
 #' and may be recognized or ignored by \code{predict()} methods for other model classes.
 #' @param ... to match generic; passed to \code{predict()} for the default method.
 #'
-#' @returns The \code{cv()} methods return an object of class \code{"cv"}, with the averaged CV criterion
-#' (\code{"CV crit"}), the adjusted average CV criterion (\code{"adj CV crit"}),
+#' @returns The \code{cv()} methods return an object of class \code{"cv"}, with the CV criterion
+#' (\code{"CV crit"}), the bias-adjusted CV criterion (\code{"adj CV crit"}),
 #' the criterion for the model applied to the full data (\code{"full crit"}),
+#' the confidence interval and level for the bias-adjusted CV criterion (\code{"confint"}),
 #' the number of folds (\code{"k"}), and the seed for R's random-number
 #' generator (\code{"seed"}). Some methods may return a
 #' subset of these components and may add additional information.
@@ -631,6 +632,7 @@ summarizeReps <- function(x){
 
 getLossFn <- function(cv){
   eval(parse(text=paste0("function(y, yhat) {",
-                         attr(cv, "casewise loss"),
+                         paste(attr(cv, "casewise loss"), collapse="\n"),
                          "}")))
 }
+

### 2023-12-21 scratch/notes_john/getLossFn.R

@@ -1,5 +1,11 @@
 getLossFn <- function(cv){
   eval(parse(text=paste0("function(y, yhat) {",
-                         attr(cv, "casewise loss"),
+                         paste(attr(cv, "casewise loss"), collapse="\n"),
                          "}")))
+}
+
+mse <- function(y, yhat){
+  result <- mean((y - yhat)^2)
+  attr(result, "casewise loss") <- c("result <- (y - yhat)^2", "result")
+  result
 }

### 2023-12-21 vignettes/cv-extend.R

@@ -47,7 +47,7 @@ cv(m.mroz, criterion=AUCcomp, seed=3639)
 ## -----------------------------------------------------------------------------
 mse
 
-cv:::getLossFn(mse)
+cv:::getLossFn(mse(rnorm(100), rnorm(100)))
 
 ## ----BEPS-data----------------------------------------------------------------
 data("BEPS", package="carData")

### 2023-12-30 scratch/notes_georges/AUC_not_casewise-edited-1.Rmd
edit AUC appendix to vignette

---
title: "Appendix: Why the complement of AUC isn't a casewise CV criterion"
output: html_document
---

Consider calculating AUC for folds in which a validation set contains $n_v$ observations. To calculate AUC in the validation set, we need the vector of prediction criteria, $\widehat{\mathbf{y}} = (\widehat{y}_1, ..., \widehat{y}_{n_v})^T$, and the vector of observed responses in the validation set, $\mathbf{y} = (y_1, \ldots, y_{n_v})^T$ with $y_i \in \{0,1\}, \; i = 1, \ldots, n_v$.

To construct the ROC curve, only the ordering of the values in 
$\mathbf{\widehat{y}}$ is relevant. Thus, assuming that there are no ties, and reordering observations if necessary, we can set 
$\mathbf{\widehat{y}} = (1, 2, \ldots, n_v)^T$.

If the AUC can be expressed as the casewise mean or sum of a function $\mathrm{cv}(\widehat{y}_i,y_i)$, where $\mathrm{cv}: \{1,2,...,n_v\}\times\{0,1\} \rightarrow [0,1]$, then
\begin{equation}
\label{eq:cw}
\tag{1}
\sum_{i=1}^{n_v} \mathrm{cv}(\widehat{y}_i,y_i) = \mathrm{AUC}(\mathbf{\widehat{y}},\mathbf{y})
\end{equation}
must hold for all $2^{n_v}$ possible values of $\mathbf{y} = (y_1,...,y_{n_v})^T$. If all $y\mathrm{s}$ have the same value, either 1 or 0, then the definition of AUC is ambiguous. AUC could be considered undefined, or it could be set to 0 if all $y$s are 0 and to 1 if all $y$s are 1. If AUC is considered to be undefined in these cases, we have $2^{n_v} - 2$ admissible values for $\mathbf{y}$.

Thus, equation (\ref{eq:cw}) produces either $2^{n_v}$ or $2^{n_v}-2$ constraints. Although there are only $2n$ possible values for the $\mathrm{cv(\cdot)}$ function, equation (\ref{eq:cw}) could, nevertheless, have consistent solutions.  We therefore need to determine whether there is a value of $n_v$ for which (\ref{eq:cw}) has no consistent solution for all admissible values of $\mathbf{y}$. In that eventuality, we will have shown that AUC cannot, in general, be expressed through a casewise sum.

If $n_v=3$, we show below that (\ref{eq:cw}) has no consistent solution if we include all possibilities for $\mathbf{y}$, but does if we exclude cases where all $y$s have the same value. If $n_v=4$, we show that there no consistent solutions in either case.

The following R function computes AUC from $\mathbf{\widehat{y}}$ and $\mathbf{y}$:

```{r}
AUC <- function(y, yhat = seq_along(y)) {
  # AUC:
  # y is a vector of 0s and 1s
  # yhat is a predictor of y, only used for its order
  #    i.e. the result is invariant under monotonic transformations of yhat
  #    If there are ties in yhat, this function splits the ties
  s <- sum(y)
  n_v <- length(y)
  if (s == 0) return(0)
  if (s == n_v) return(1)
  a <- sum(outer(yhat[y == 0], yhat[y == 1], `<`)) + 
    sum(outer(yhat[y == 0], yhat[y == 1], `==`))/2
  a/(s*(n_v - s))
}
```

We then define a function to generate all possible $\mathbf{y}$s of length $n_v$ as rows of the matrix $\mathbf{Y}$:

```{r}
Ymat <- function(n_v, exclude_identical = FALSE) {
  stopifnot(n_v > 0 && round(n_v) == n_v)    # n_v must be a positive integer
  ret <- sapply(0:(2^n_v - 1),
                function(x) as.integer(intToBits(x)) )[1:n_v, ]
  ret <- if (is.matrix(ret)) t(ret) else matrix(ret)
  colnames(ret) <- paste0("y", 1:ncol(ret))
  if (exclude_identical) ret[-c(1, nrow(ret)), ] else ret
}
```

For $n_v=3$,

```{r}
Ymat(3)
```

If we exclude $\mathbf{y}$s with identical values, then

```{r}
Ymat(3, TRUE)
```

Here is $\mathbf{Y}$ with corresponding values of AUC:

```{r}
cbind(Ymat(3), AUC = apply(Ymat(3), 1, AUC))
```

The values of $\mathrm{cv}(\widehat{y}_i, y_i)$ that express AUC as a sum of casewise values are solutions of equation (\ref{eq:cw}), which can be written as solutions of the following linear equation:
\begin{equation}
\label{eq:lin}
\tag{2}
(\mathbf{U} -\mathbf{Y}) \mathbf{c}_0 + \mathbf{Y} \mathbf{c}_1
=
[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]
\begin{bmatrix}
\mathbf{c}_0 \\ \mathbf{c}_1
\end{bmatrix}
= \mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})
\end{equation}
where $\mathbf{U}$ is a matrix of 1s conformable with $\mathbf{Y}$; $\mathbf{c}_0 = [\mathrm{cv}(1,0), c(2,0), ..., \mathrm{cv}(n_v,0)]^T$; $\mathbf{c}_1 = [\mathrm{cv}(1,1), c(2,1), ..., \mathrm{cv}(n_v,1)]^T$; $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]$ and
$\begin{bmatrix}\begin{aligned}
\mathbf{c}_0 \\ \mathbf{c}_1
\end{aligned}
\end{bmatrix}$ 
are partitioned matrices; and $\mathbf{\widehat{Y}}$ is a matrix each of whose rows consists of the integers 1 to $n_v$.

We can test whether equation (\ref{eq:lin}) has a solution for any given $n_v$ by considering whether the residuals of the associated linear model are all 0, using the "design matrix" $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]$ to predict the outcome
$\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})$.

```{r}
resids <- function(n_v, exclude_identical = FALSE, 
                   tol = sqrt(.Machine$double.eps)) {
  Y <- Ymat(n_v, exclude_identical = exclude_identical)
  AUC <- apply(Y, 1, AUC)
  X <- cbind(1-Y, Y)
  opts <- options(warn = -1)
  on.exit(options(opts))
  fit <- lsfit(X, AUC, intercept = FALSE)
  ret <- max(abs(residuals(fit)))
  if(ret < tol){
    ret <- 0
    solution <- coef(fit)
    names(solution) <- paste0("c(", c(1:n_v, 1:n_v), ",", 
                              rep(0:1, each = n_v), ")")
    attr(ret, "solution") <- zapsmall(solution)
  }
  ret
}
```

The case $n_v=3$, excluding identical $y$s, has a solution:

```{r}
resids(3, exclude_identical = TRUE)
```

But, if identical $y$s are included, the equation is not consistent:

```{r}
resids(3, exclude_identical = FALSE)
```

For $n_v=4$, there are no solutions in either case:

```{r}
resids(4, exclude_identical = TRUE)
resids(4, exclude_identical = FALSE)
```

Consequently, the widely employed AUC measure of fit for binary regression cannot in general be used for a casewise cross-validation criterion.

### 2023-12-30 scratch/notes_georges/AUC_not_casewise-edited-1.Rmd
tweak AUC appendix

@@
 
-The following R function computes AUC from $\mathbf{\widehat{y}}$ and $\mathbf{y}$:
+The following R function computes AUC from $\mathbf{\widehat{y}}$ and $\mathbf{y}$, accommodating the cases where $\mathbf{y}$ is all 0s or all 1s:
 
@@
 AUC <- function(y, yhat = seq_along(y)) {
-  # AUC:
-  # y is a vector of 0s and 1s
-  # yhat is a predictor of y, only used for its order
-  #    i.e. the result is invariant under monotonic transformations of yhat
-  #    If there are ties in yhat, this function splits the ties
   s <- sum(y)
-  n_v <- length(y)
   if (s == 0) return(0)
-  if (s == n_v) return(1)
-  a <- sum(outer(yhat[y == 0], yhat[y == 1], `<`)) + 
-    sum(outer(yhat[y == 0], yhat[y == 1], `==`))/2
-  a/(s*(n_v - s))
+  if (s == length(y)) return(1)
+  Metrics::auc(y, yhat)
 }
@@
 
-We can test whether equation (\ref{eq:lin}) has a solution for any given $n_v$ by considering whether the residuals of the associated linear model are all 0, using the "design matrix" $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]$ to predict the outcome
-$\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})$.
+We can test whether equation (\ref{eq:lin}) has a solution for any given $n_v$ by trying to solve it as a least-squares problem, considering whether the residuals of the associated linear model are all 0, using the "design matrix" $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]$ to predict the "outcome"
+$\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})$:
 

### 2023-12-31 scratch/notes_georges/AUC_not_casewise-edited-1.Rmd
show dimensions of matrices in AUC appendix

@@
 Consider calculating AUC for folds in which a validation set contains $n_v$ observations.
-To calculate AUC in the validation set, we need the vector of prediction criteria, $\widehat{\mathbf{y}} = (\widehat{y}_1, ..., \widehat{y}_{n_v})^T$, and the vector of observed responses in the validation set, $\mathbf{y} = (y_1, \ldots, y_{n_v})^T$ with $y_i \in \{0,1\}, \; i = 1, \ldots, n_v$.
+To calculate AUC in the validation set, we need the vector of prediction criteria, $\widehat{\mathbf{y}}_{(n_v \times 1)} = (\widehat{y}_1, ..., \widehat{y}_{n_v})^T$, and the vector of observed responses in the validation set, $\mathbf{y}_{(n_v \times 1)} = (y_1, \ldots, y_{n_v})^T$ with $y_i \in \{0,1\}, \; i = 1, \ldots, n_v$.
 
@@
 
-We then define a function to generate all possible $\mathbf{y}$s of length $n_v$ as rows of the matrix $\mathbf{Y}$:
+We then define a function to generate all possible $\mathbf{y}$s of length $n_v$ as rows of the matrix $\mathbf{Y}_{(2^{n_v} \times n_v)}$:
 
@@
 \end{equation}
-where $\mathbf{U}$ is a matrix of 1s conformable with $\mathbf{Y}$; $\mathbf{c}_0 = [\mathrm{cv}(1,0), c(2,0), ..., \mathrm{cv}(n_v,0)]^T$; $\mathbf{c}_1 = [\mathrm{cv}(1,1), c(2,1), ..., \mathrm{cv}(n_v,1)]^T$; $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]$ and
+where $\mathbf{U}_{(2^{n_v} \times n_v)}$ is a matrix of 1s conformable with $\mathbf{Y}$; $\mathbf{c}_0 = [\mathrm{cv}(1,0), c(2,0), ..., \mathrm{cv}(n_v,0)]^T$; $\mathbf{c}_1 = [\mathrm{cv}(1,1), c(2,1), ..., \mathrm{cv}(n_v,1)]^T$; $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]_{(2^{n_v} \times 2n_v)}$ and
 $\begin{bmatrix}\begin{aligned}
@@
 \end{aligned}
-\end{bmatrix}$ 
-are partitioned matrices; and $\mathbf{\widehat{Y}}$ is a matrix each of whose rows consists of the integers 1 to $n_v$.
+\end{bmatrix}_{(2n_v \times 1)}$ 
+are partitioned matrices; and $\mathbf{\widehat{Y}}_{(2^{n_v} \times n_v)}$ is a matrix each of whose rows consists of the integers 1 to $n_v$.
 
 We can test whether equation (\ref{eq:lin}) has a solution for any given $n_v$ by trying to solve it as a least-squares problem, considering whether the residuals of the associated linear model are all 0, using the "design matrix" $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]$ to predict the "outcome"
-$\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})$:
+$\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})_{(2n_v \times 1)}$:
 

### 2023-12-31 scratch/notes_georges/AUC_not_casewise-edited-1.Rmd
fix typo in AUC appendix

@@
 We can test whether equation (\ref{eq:lin}) has a solution for any given $n_v$ by trying to solve it as a least-squares problem, considering whether the residuals of the associated linear model are all 0, using the "design matrix" $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]$ to predict the "outcome"
-$\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})_{(2n_v \times 1)}$:
+$\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})_{(2^{n_v} \times 1)}$:
 

### 2023-12-31 scratch/notes_georges/AUC_not_casewise-edited-1.Rmd
more tweaks to AUC appendix

@@
 ```{r}
-Ymat(3, TRUE)
+Ymat(3, exclude_identical = TRUE)
 ```
@@
 
-The values of $\mathrm{cv}(\widehat{y}_i, y_i)$ that express AUC as a sum of casewise values are solutions of equation (\ref{eq:cw}), which can be written as solutions of the following linear equation:
+The values of $\mathrm{cv}(\widehat{y}_i, y_i)$ that express AUC as a sum of casewise values are solutions of equation (\ref{eq:cw}), which can be written as solutions of the following system of $2^{n_v}$ linear simultaneous equations in $2n_v$ unknowns:
 \begin{equation}

### 2023-12-31 R/cv.R
add AUC appendix to cv-extend vignette

@@ -108,7 +108,7 @@ cv <- function(model, data, criterion, k, reps=1, seed, ...){
 }
 
 #' @describeIn cv \code{default} method
-#' @importFrom stats coef family fitted lm.wfit model.frame
+#' @importFrom stats coef family fitted lm.wfit lsfit model.frame
 #' model.matrix model.response predict qnorm
 #' update weighted.mean weights
 #' residuals hatvalues printCoefmat sd

### 2023-12-31 vignettes/cv-extend.R

@@ -251,6 +251,63 @@ selectSubsets(model=m.swiss, indices=seq(5, 45, by=10))
 ## ----best-models-by-folds-----------------------------------------------------
 compareFolds(cv.swiss)
 
+## -----------------------------------------------------------------------------
+AUC <- function(y, yhat = seq_along(y)) {
+  s <- sum(y)
+  if (s == 0) return(0)
+  if (s == length(y)) return(1)
+  Metrics::auc(y, yhat)
+}
+
+## -----------------------------------------------------------------------------
+Ymat <- function(n_v, exclude_identical = FALSE) {
+  stopifnot(n_v > 0 && round(n_v) == n_v)    # n_v must be a positive integer
+  ret <- sapply(0:(2^n_v - 1),
+                function(x) as.integer(intToBits(x)) )[1:n_v, ]
+  ret <- if (is.matrix(ret)) t(ret) else matrix(ret)
+  colnames(ret) <- paste0("y", 1:ncol(ret))
+  if (exclude_identical) ret[-c(1, nrow(ret)), ] else ret
+}
+
+## -----------------------------------------------------------------------------
+Ymat(3)
+
+## -----------------------------------------------------------------------------
+Ymat(3, exclude_identical = TRUE)
+
+## -----------------------------------------------------------------------------
+cbind(Ymat(3), AUC = apply(Ymat(3), 1, AUC))
+
+## -----------------------------------------------------------------------------
+resids <- function(n_v, exclude_identical = FALSE, 
+                   tol = sqrt(.Machine$double.eps)) {
+  Y <- Ymat(n_v, exclude_identical = exclude_identical)
+  AUC <- apply(Y, 1, AUC)
+  X <- cbind(1-Y, Y)
+  opts <- options(warn = -1)
+  on.exit(options(opts))
+  fit <- lsfit(X, AUC, intercept = FALSE)
+  ret <- max(abs(residuals(fit)))
+  if(ret < tol){
+    ret <- 0
+    solution <- coef(fit)
+    names(solution) <- paste0("c(", c(1:n_v, 1:n_v), ",", 
+                              rep(0:1, each = n_v), ")")
+    attr(ret, "solution") <- zapsmall(solution)
+  }
+  ret
+}
+
+## -----------------------------------------------------------------------------
+resids(3, exclude_identical = TRUE)
+
+## -----------------------------------------------------------------------------
+resids(3, exclude_identical = FALSE)
+
[…]

### 2024-01-01 R/cv.modList.R
change getResponse() to GetResponse() to avoid name clash with nlme

@@ -76,9 +76,9 @@ models <- function(...){
   if (!all(n[1L] == n[-1L])) {
     stop("models are fit to data sets of differing numbers of cases")
   }
-  response <- getResponse(models[[1L]])
+  response <- GetResponse(models[[1L]])
   for (i in 2L:length(models)){
-    if (!isTRUE(all.equal(response, getResponse(models[[i]]),
+    if (!isTRUE(all.equal(response, GetResponse(models[[i]]),
                           check.attributes=FALSE))){
       stop("models are not all fit to the same response variable")
     }

### 2024-01-01 R/getResponse.R

@@ -22,15 +22,15 @@
 #'
 #' @examples
 #'     fit <- lm(mpg ~ gear, mtcars)
-#'     getResponse(fit)
+#'     GetResponse(fit)
 #' @export
-getResponse <- function(model, ...){
-  UseMethod("getResponse")
+GetResponse <- function(model, ...){
+  UseMethod("GetResponse")
 }
 
-#' @describeIn getResponse \code{default} method
+#' @describeIn GetResponse \code{default} method
 #' @export
-getResponse.default <- function(model, ...){
+GetResponse.default <- function(model, ...){
   y <- if (!isS4(model)) model$y else insight::get_response(model)
   if (is.null(y)) y <- model.response(model.frame(model))
   if (!is.vector(y)) stop("non-vector response")
@@ -38,9 +38,9 @@ getResponse.default <- function(model, ...){
   y
 }
 
-#' @describeIn getResponse \code{merMod} method
+#' @describeIn GetResponse \code{merMod} method
 #' @export
-getResponse.merMod <- function(model, ...){
+GetResponse.merMod <- function(model, ...){
   y <- insight::get_response(model)
   if (is.factor(y)) {
     levels <- levels(y)
@@ -58,10 +58,10 @@ getResponse.merMod <- function(model, ...){
   y
 }
 
-#' @describeIn getResponse \code{merMod} method
+#' @describeIn GetResponse \code{merMod} method
 #' @export
-getResponse.lme <- function(model, ...) insight::get_response(model)
+GetResponse.lme <- function(model, ...) insight::get_response(model)
 
-#' @describeIn getResponse \code{glmmTMB} method
+#' @describeIn GetResponse \code{glmmTMB} method
 #' @export
-getResponse.glmmTMB <- getResponse.merMod
+GetResponse.glmmTMB <- GetResponse.merMod

### 2024-01-01 vignettes/cv-extend.R

@@ -87,12 +87,12 @@ xtabs(~ vote, data=BEPS)/nrow(BEPS)
 ## ----BEPS-test-default, error=TRUE--------------------------------------------
 cv(m.beps, seed=3465, criterion=BayesRuleMulti)
 
-## ----getResponse.multinom-----------------------------------------------------
-getResponse.multinom <- function(model, ...) {
+## ----GetResponse.multinom-----------------------------------------------------
+GetResponse.multinom <- function(model, ...) {
   insight::get_response(model)
 }
 
-head(getResponse(m.beps))
+head(GetResponse(m.beps))
 
 ## ----BEPS-test-default-2, error=TRUE------------------------------------------
 cv(m.beps, seed=3465, criterion=BayesRuleMulti)
@@ -110,13 +110,13 @@ cv(m.beps, seed=3465)
 ## ----cv.lme-------------------------------------------------------------------
 cv:::cv.lme
 
-## ----getResponse.glmmPQL------------------------------------------------------
-getResponse.glmmPQL <- function(model, ...){
+## ----GetResponse.glmmPQL------------------------------------------------------
+GetResponse.glmmPQL <- function(model, ...){
   f <- formula(model)
   f[[3]] <- 1 # regression constant only on RHS
   model <- suppressWarnings(glm(f, data=model$data, family=model$family,
                                 control=list(maxit=1)))
-  cv::getResponse(model)
+  cv::GetResponse(model)
 }
 
 ## ----cv.glmmPQL---------------------------------------------------------------
@@ -195,7 +195,7 @@ selectSubsets <- function(data=insight::get_data(model),
   if (inherits(model, "lm", which=TRUE) != 1)
     stop("selectSubsets is appropriate only for 'lm' models")
   
-  y <- getResponse(model)
+  y <- GetResponse(model)
   formula <- formula(model)
   X <- model.matrix(model)
 

### 2024-01-01 scratch/notes_john/JSS-article/JSS-article.Rmd
add initial "draft" of JSS article to scratch

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0000-0000-0000
    affiliation: 'York University'
    # To add another line, use \AND at the end of the previous one as above
    address: |
      | Toronto, Ontario, Canada
title:
  formatted: "\\pkg{cv}: An R Package for Cross-Validation of Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validation of Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validation"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: ["cv.bib"]
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
```

# Cross-validation

Cross-validation (CV) is an essentially simple and intuitively reasonable approach to estimating the predictive accuracy of regression models. CV is developed in many standard sources on regression modeling and "machine learning"---we particularly recommend @JamesEtAl:2021 [Secs. 5.1, 5.3]---and so we will describe the method only briefly here before taking up computational issues and some examples. See @ArlotCelisse:2010 for a wide-ranging, if technical, survey of cross-validation and related methods that emphasizes the statistical properties of CV.

Validating research by replication on independently collected data is a common scientific norm. Emulating this process in a single study by data-division is less common: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data. Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, particularly in smaller samples, the results can vary substantially based on the random division of the data: See @Harrell:2015 [Sec. 5.3] for this and other remarks about data-division and cross-validation.

Cross-validation speaks to both of these issues. In CV, the data are randomly divided as equally as possible into several, say $k$, parts, called "folds." The statistical model is fit $k$ times, leaving each fold out in turn. Each fitted model is then used to predict the response variable for the cases in the omitted fold. A CV criterion or "cost" measure, such as the mean-squared error ("MSE") of prediction, is then computed using these predicted values. In the extreme $k = n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."

Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of prediction error. The $n$ regression models are highly statistical dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has relatively large variance. In contrast, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) are somewhat biased but have smaller variance. It is also possible to correct $k$-fold CV for bias (see below).

# Examples

## Polynomial regression for the Auto data

The data for this example are drawn from the **ISLR2** package for R, associated with @JamesEtAl:2021. The presentation here is close (though not identical) to that in the original source [@JamesEtAl:2021 Secs. 5.1, 5.3], and it demonstrates the use of the `cv()` function in the **cv** package.[^boot]

[^boot]: @JamesEtAl:2021 use the `cv.glm()` function in the **boot** package [@CantyRipley2022; @DavisonHinkley:1997]. Despite its name, `cv.glm()` is an independent function and not a method of a `cv()` generic function.

The `Auto` dataset contains information about 392 cars:

```{r Auto}
data("Auto", package="ISLR2")
head(Auto)
dim(Auto)
```
With the exception of `origin` (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see `help("Auto", package="ISLR2")`.

We'll focus here on the relationship of `mpg` (miles per gallon) to `horsepower`, as displayed in the following scatterplot:

```{r mpg-horsepower-scatterplot}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "`mpg` vs `horsepower` for the `Auto` data"
plot(mpg ~ horsepower, data=Auto)
```
The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in the following figure:

[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is probably not the best choice here. Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader).

```{r mpg-horsepower-scatterplot-polynomials}
#| out.width = "100%",
#| fig.height = 5,
#| fig.cap = "`mpg` vs `horsepower` for the `Auto` data"
plot(mpg ~ horsepower, data=Auto)
horsepower <- with(Auto, 
                   seq(min(horsepower), max(horsepower), 
                       length=1000))
for (p in 1:5){
  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
  lines(horsepower, mpg, col=p + 1, lty=p, lwd=2)
}
legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=2,
       title="Degree", inset=0.02)
```
The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).

The following graph shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$. The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.

```{r mpg-horsepower-MSE-se}
#| out.width = "100%",
#| fig.height = 5,
#| fig.cap = "Estimated squared error as a function of polynomial degree, $p$"
library("cv") # for mse() and other functions

var <- mse <- numeric(10)
for (p in 1:10){
  m <- lm(mpg ~ poly(horsepower, p), data=Auto)
  mse[p] <- mse(Auto$mpg, fitted(m))
  var[p] <- summary(m)$sigma^2
}

plot(c(1, 10), range(mse, var), type="n",
     xlab="Degree of polynomial, p",
     ylab="Estimated Squared Error")
lines(1:10, mse, lwd=2, lty=1, col=2, pch=16, type="b")
lines(1:10, var, lwd=2, lty=2, col=3, pch=17, type="b")
legend("topright", inset=0.02,
       legend=c(expression(hat(sigma)^2), "MSE"),
       lwd=2, lty=2:1, col=3:2, pch=17:16)
```
The code for this graph uses the `mse()` function from the **cv** package to compute the MSE for each fit.

### Using cv()

The generic `cv()` function has an `"lm"` method, which by default performs $k = 10$-fold CV:
```{r cv-lm-1}
m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
summary(m.auto)

cv(m.auto)
```
The `"lm"` method by default uses `mse()` as the CV criterion and the Woodbury matrix identity to update the regression with each fold deleted without having literally to refit the model. Computational details are discussed in the final section of this vignette. The function reports the CV estimate of MSE, a biased-adjusted estimate of the MSE (the bias adjustment is explained in the final section), and the MSE is also computed for the original, full-sample regression. Because the division of the data into 10 folds is random, `cv()` explicitly (randomly) generates and saves a seed for R's pseudo-random number generator, to make the results replicable. The user can also specify the seed directly via the `seed` argument to `cv()`.

To perform LOO CV, we can set the `k` argument to `cv()` to the number of cases in the data, here `k=392`, or, more conveniently, to `k="loo"` or `k="n"`:

```{r cv.lm-2`}
cv(m.auto, k="loo")
```
For LOO CV of a linear model, `cv()` by default uses the hatvalues from the model fit to the full data for the LOO updates, and reports only the CV estimate of MSE. Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted. All three methods produce exact results for a linear model (within the precision of floating-point computations):
```{r cv.lm-3}
cv(m.auto, k="loo", method="naive")

cv(m.auto, k="loo", method="Woodbury")
```
The `"naive"` and `"Woodbury"` methods also return the bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an issue for LOO CV.

This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed. In this comparison, we include the `cv.glm()` function from the **boot** package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM. We use the `microbenchmark()` function from the package of the same name for the timings [@Mersmann:2023]:
```{r cv.lm.timings, cache=TRUE}
m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
boot::cv.glm(Auto, m.auto.glm)$delta

microbenchmark::microbenchmark(
  hatvalues = cv(m.auto, k="loo"),
  Woodbury = cv(m.auto, k="loo", method="Woodbury"),
  naive = cv(m.auto, k="loo", method="naive"),
  cv.glm = boot::cv.glm(Auto, m.auto.glm),
  times=10
)
```
On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.[^microbenchmark]

[^microbenchmark]: Out of impatience, we asked `microbenchmark()` to execute each command only 10 times rather than the default 100. With the exception of the last columns, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable. Because of the small number of repetitions (i.e., 10), the `"hatvalues"` and `"Woodbury"` methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default `times=100` repetitions.

### Comparing competing models

The `cv()` function also has a method that can be applied to a list of regression models for the same data, composed using the `models()` function. For $k$-fold CV, the same folds are used for the competing models, which reduces random error in their comparison. This result can also be obtained by specifying a common seed for R's random-number generator while applying `cv()` separately to each model, but employing a list of models is more convenient for both $k$-fold and LOO CV (where there is no random component to the composition of the $n$ folds).

We illustrate with the polynomial regression models of varying degree for the `Auto` data (discussed previously), beginning by fitting and saving the 10 models:
```{r polyomial-models}
for (p in 1:10){
  assign(paste0("m.", p),
         lm(mpg ~ poly(horsepower, p), data=Auto))
}
objects(pattern="m\\.[0-9]")
summary(m.2) # for example, the quadratic fit
```
We then apply `cv()` to the list of 10 models (the `data` argument is required):
```{r polynomial-regression-CV}
# 10-fold CV
cv.auto.10 <- cv(models(m.1, m.2, m.3, m.4, m.5,
                     m.6, m.7, m.8, m.9, m.10),
              data=Auto, seed=2120)
cv.auto.10[1:2] # for the linear and quadratic models

# LOO CV
cv.auto.loo <- cv(models(m.1, m.2, m.3, m.4, m.5,
                        m.6, m.7, m.8, m.9, m.10),
                 data=Auto, k="loo")
cv.auto.loo[1:2] # linear and quadratic models
```
Because we didn't supply names for the models in the calls to the `models()` function, the names `model.1`, `model.2`, etc., are generated by the function.

Finally, we extract and graph the adjusted MSEs for $10$-fold CV and the MSEs for LOO CV:
```{r polynomial-regression-CV-graph}
#| out.width = "100%",
#| fig.height = 5,
#| fig.cap = "Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$"
cv.mse.10 <- sapply(cv.auto.10, function(x) x[["adj CV crit"]])
cv.mse.loo <- sapply(cv.auto.loo, function(x) x[["CV crit"]])
plot(c(1, 10), range(cv.mse.10, cv.mse.loo), type="n",
     xlab="Degree of polynomial, p",
     ylab="Cross-Validated MSE")
lines(1:10, cv.mse.10, lwd=2, lty=1, col=2, pch=16, type="b")
lines(1:10, cv.mse.loo, lwd=2, lty=2, col=3, pch=17, type="b")
legend("topright", inset=0.02,
       legend=c("10-Fold CV", "LOO CV"),
       lwd=2, lty=2:1, col=3:2, pch=17:16)
```

Alternatively, we can use the `plot()` method for `"cvModList"` objects to compare the models, though with separate graphs for 10-fold and LOO CV:

```{r, polynomial-regression-CV-graph-2, fig.show="hold"}
#| out.width = "45%",
#| fig.height = 6,
#| fig.cap = "Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$"
plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
plot(cv.auto.loo, main="Polynomial Regressions, LOO CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
```

In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model, reported above (except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting).

## Logistic regression for the Mroz data

The `Mroz` data set from the **carData** package [associated with @FoxWeisberg:2019] has been used by several authors to illustrate binary logistic regression; see, in particular @FoxWeisberg:2019. The data were originally drawn from the U.S. Panel Study of Income Dynamics and pertain to married women. Here are a few cases in the data set:
```{r Mroz-data}
data("Mroz", package="carData")
head(Mroz, 3)
tail(Mroz, 3)
```
The response variable in the logistic regression is `lfp`, labor-force participation, a factor coded `"yes"` or `"no"`. The remaining variables are predictors:

* `k5`, number of children 5 years old of younger in the woman's household;
* `k618`, number of children between 6 and 18 years old;
* `age`, in years;
* `wc`, wife's college attendance, `"yes"` or `"no"`;
* `hc`, husband's college attendance;
* `lwg`, the woman's log wage rate if she is employed, or her *imputed* wage rate, if she is not [a variable that @FoxWeisberg:2019 show is problematically defined]; and
* `inc`, family income, in $1000s, exclusive of wife's income.

We use the `glm()` function to fit a binary logistic regression to the `Mroz` data:
```{r Mroz-logistic-regresion}
m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
summary(m.mroz)

BayesRule(ifelse(Mroz$lfp == "yes", 1, 0), 
          fitted(m.mroz, type="response"))
```
In addition to the usually summary output for a GLM, we show the result of applying the `BayesRule()` function from the **cv** package to predictions derived from the fitted model. Bayes rule, which predicts a "success" in a binary regression model when the fitted probability of success [i.e., $\phi = \Pr(y = 1)$] is $\widehat{\phi} \ge .5$ and a "failure" if $\widehat{\phi} < .5$.[^BayesRule] The first argument to `BayesRule()` is the binary {0, 1} response, and the second argument is the predicted probability of success. `BayesRule()` returns the proportion of predictions that are *in error*, as appropriate for a "cost" function.

[^BayesRule]: `BayesRule()` does some error checking; `BayesRule2()` is similar, but omits the error checking, and so can be faster for large problems.

The value returned by `BayesRule()` is associated with an "attribute" named `"casewise loss"` and set to `"y != round(yhat)"`, signifying that the Bayes rule CV criterion is computed as the mean of casewise values, here 0 if the prediction for a case matches the observed value and 1 if it does not (signifying a prediction error). The `mse()` function for numeric responses is also calculated as a casewise average. Some other criteria, such as the median absolute error, computed by the `medAbsErr()` function in the **cv** package, aren't averages of casewise components. The distinction is important because, to our knowledge, the statistical theory of cross-validation, for example, in @DavisonHinkley:1997, @BatesHastieTibshirani:2023, and @ArlotCelisse:2010, is developed for CV criteria like MSE that are means of casewise components. As a consequence, we limit computation of bias adjustment and confidence intervals (see below) to criteria that are casewise averages.

In this example, the fitted logistic regression incorrectly predicts 31% of the responses; we expect this estimate to be optimistic given that the model is used to "predict" the data to which it is fit.

The `"glm"` method for `cv()` is largely similar to the `"lm"` method, although the default algorithm, selected explicitly by `method="exact"`, refits the model with each fold removed (and is thus equivalent to `method="naive"` for `"lm"` models). For generalized linear models, `method="Woodbury"` or (for LOO CV) `method="hatvalues"` provide approximate results (see the last section of the vignette for details):

```{r cv-Mroz-10-fold}
cv(m.mroz, criterion=BayesRule, seed=248)

cv(m.mroz, criterion=BayesRule, seed=248, method="Woodbury")
```
To ensure that the two methods use the same 10 folds, we specify the seed for R's random-number generator explicitly; here, and as is common in our experience, the `"exact"` and `"Woodbury"` algorithms produce nearly identical results. The CV estimates of prediction error are slightly higher than the estimate based on all of the cases.

The printed output includes a 95% confidence interval for the bias-adjusted Bayes rule CV criterion. @BatesHastieTibshirani:2023 show that these confidence intervals are unreliable for models fit to small samples, and by default `cv()` computes them only when the sample size is 400 or larger and when the CV criterion employed is an average of casewise components, as is the case for Bayes rule. See the final section of the vignette for details of the computation of confidence intervals for bias-adjusted CV criteria.

Here are results of applying LOO CV to the Mroz model, using both the exact and the approximate methods:
```{r cv-Mroz-LOO}
cv(m.mroz, k="loo", criterion=BayesRule)

cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury")

cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues")
```
To the number of decimal digits shown, the three methods produce identical results for this example.

As for linear models, we report some timings for the various `cv()` methods of computation in LOO CV as well as for the `cv.glm()` function from the **boot** package (which, recall, refits the model with each case removed, and thus is comparable to `cv()` with `method="exact"`):
```{r glm.timings, cache=TRUE}
microbenchmark::microbenchmark(
  hatvalues=cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues"),
  Woodbury=cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury"),
  exact=cv(m.mroz, k="loo", criterion=BayesRule),
  cv.glm=boot::cv.glm(Mroz, m.mroz,
               cost=BayesRule),
  times=10)
```
There is a substantial time penalty associated with exact computations.

# Cross-validating mixed-effects models

The fundamental analogy for cross-validation is to the collection of new data. That is, predicting the response in each fold from the model fit to data in the other folds is like using the model fit to all of the data to predict the response for new cases from the values of the predictors for those new cases. As we explained, the application of this idea to independently sampled cases is straightforward---simply partition the data into random folds of equal size and leave each fold out in turn, or, in the case of LOO CV, simply omit each case in turn.

In contrast, mixed-effects models are fit to *dependent* data, in which cases as clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases repeated observations on the individuals over time.[^crossed-effects]

[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects. For example, consider students within classes within schools. In primary schools, students typically are in a single class, and so classes are nested within schools. In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed. The `lmer()` function in the **lme4** package is capable of modeling both nested and crossed random effects, and the `cv()` methods for mixed models in the **cv** package pertain to both nested and crossed random effects. We present an example of the latter later in the vignette.

We can think of two approaches to applying cross-validation to clustered data:[^cv-faq]

[^cv-faq]: We subsequently discovered that @Vehtari:2023 [Section 8] makes similar points.

1. Treat CV as analogous to predicting the response for one or more cases in a *newly observed cluster*. In this instance, the folds comprise one or more whole clusters; we refit the model with all of the cases in clusters in the current fold removed; and then we predict the response for the cases in clusters in the current fold. These predictions are based only on fixed effects because the random effects for the omitted clusters are presumably unknown, as they would be for data on cases in newly observed clusters.

2. Treat CV as analogous to predicting the response for a newly observed case in an *existing cluster*. In this instance, the folds comprise one or more individual cases, and the predictions can use both the fixed and random effects.

## Example: The High-School and Beyond data

Following their use by @RaudenbushBryk:2002, data from the 1982 *High School and Beyond* (HSB) survey have become a staple of the literature on mixed-effects models. The HSB data are used by @FoxWeisberg:2019 [Sec. 7.2.2] to illustrate the application of linear mixed models to hierarchical data, and we'll closely follow their example here.

The HSB data are included in the `MathAchieve` and `MathAchSchool` data sets in the **nlme** package  [@PinheiroBates:2000]. `MathAchieve` includes individual-level data on 7185  students in 160 high schools, and `MathAchSchool` includes school-level data:
```{r HSB-data}
data("MathAchieve", package="nlme")
dim(MathAchieve)
head(MathAchieve, 3)
tail(MathAchieve, 3)

data("MathAchSchool", package="nlme")
dim(MathAchSchool)
head(MathAchSchool, 2)
tail(MathAchSchool, 2)
```
The first few students are in school number 1224 and the last few in school 9586.

We'll use only the `School`, `SES` (students' socioeconomic status), and `MathAch` (their score on a standardized math-achievement test) variables in the `MathAchieve` data set, and `Sector` (`"Catholic"` or `"Public"`) in the `MathAchSchool` data set.

Some data-management is required before fitting a mixed-effects model to the HSB data, for which we use the **dplyr** package [@WickhamEtAl:2023]:
```{r HSB-data-management, cache=TRUE}
library("dplyr")
MathAchieve %>% group_by(School) %>%
  summarize(mean.ses = mean(SES)) -> Temp
Temp <- merge(MathAchSchool, Temp, by="School")
HSB <- merge(Temp[, c("School", "Sector", "mean.ses")],
             MathAchieve[, c("School", "SES", "MathAch")], by="School")
names(HSB) <- tolower(names(HSB))

HSB$cses <- with(HSB, ses - mean.ses)
```
In the process, we created two new school-level variables: `meanses`, which is the average SES for students in each school; and `cses`, which is school-average SES centered at its mean. For details, see @FoxWeisberg:2019 [Sec. 7.2.2].

Still following Fox and Weisberg, we proceed to use the `lmer()` function in the **lme4** package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
```{r HSB-lmer, cache=TRUE}
library("lme4")
hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
                   + (cses | school), data=HSB)
summary(hsb.lmer, correlation=FALSE)
```

We can then cross-validate at the cluster (i.e., school) level,
```{r HSB-lmer-CV-cluster, cache=TRUE}
cv(hsb.lmer, k=10, clusterVariables="school", seed=5240)
```
or at the case (i.e., student) level,
```{r HSB-lmer-CV-case, cache=TRUE}
cv(hsb.lmer, seed=1575)
```
For cluster-level CV, the `clusterVariables` argument tells `cv()` how the clusters are defined. Were there more than one clustering variable, say classes within schools, these would be provided as a character vector of variable names: `clusterVariables = c("school", "class")`. For cluster-level CV, the default is `k = "loo"`, that is, leave one cluster out at a time; we instead specify `k = 10` folds of clusters, each fold therefore comprising $160/10 = 16$ schools.

If the `clusterVariables` argument is omitted, then case-level CV is employed, with `k = 10` folds as the default, here each with $7185/10 \approx 719$ students. Notice that one of the 10 models refit with a fold removed failed to converge. Convergence problems are common in mixed-effects modeling. The apparent issue here is that an estimated variance component is close to or equal to 0, which is at a boundary of the parameter space. That shouldn't disqualify the fitted model for the kind of prediction required for cross-validation.

There is also a `cv()` method for linear mixed models fit by the `lme()` function in the **nlme** package, and the arguments for `cv()` in this case are the same as for a model fit by `lmer()` or `glmer()`. We illustrate with the mixed model fit to the HSB data:
```{r hsb-lme, cache=TRUE}
library(nlme)
hsb.lme <- lme(mathach ~ mean.ses*cses + sector*cses,
                 random = ~ cses | school, data=HSB,
               control=list(opt="optim"))
summary(hsb.lme)

cv(hsb.lme, k=10, clusterVariables="school", seed=5240)

cv(hsb.lme, seed=1575)
```
We used the same random-number generator seeds as in the previous example cross-validating the model fit by `lmer()`, and so the same folds are employed in both cases.[^optimizer] The estimated covariance components and fixed effects in the summary output differ slightly between the `lmer()` and `lme()` solutions, although both functions seek to maximize the REML criterion. This is, of course, to be expected when different algorithms are used for numerical optimization.  To the precision reported, the cluster-level CV results for the `lmer()` and `lme()` models are identical, while the case-level CV results are very similar but not identical.

[^optimizer]: The observant reader will notice that we set the argument `control=list(opt="optim")` in the call to `lme()`, changing the optimizer employed from the default `"nlminb"`. We did this because with the default optimizer, `lme()` encountered the same convergence issue as `lmer()`, but rather than issuing a warning, `lme()` failed, reporting an error. As it turns out, setting the optimizer to `"optim"` avoids this problem.

## Example: Contrived hierarchical data

We introduce an artificial data set that exemplifies aspects of cross-validation particular to hierarchical models. Using this data set, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model. Furthermore, commonly used measures of fit, such as mean-squared error, do not necessarily become smaller as models become larger, even when the models are nested, and even when the measure of fit is computed for the whole data set.

Consider a researcher studying improvement in a skill, yodeling, for example, among students enrolled in a four-year yodeling program. The plan is to measure each student's skill level at the beginning of the program and every year thereafter until the end of the program, resulting in five annual measurements for each student. It turns out that yodeling appeals to students of all ages, and students enrolling in the program range in age from 20 to 70. Moreover, participants' untrained yodeling skill is similar at all ages, as is their rate of progress with training. All students complete the four-year program.

The researcher, who has more expertise in yodeling than in modeling, decides to model the response, $y$, yodeling skill, as a function of age, $x$, reasoning that students get older during their stay in the program, and (incorrectly) that age can serve as a proxy for elapsed time. The researcher knows that a mixed model should be used to account for clustering due to the expected similarity of measurements taken from each student.

We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^1]

[^1]: We invite the interested reader to experiment with varying the parameters of our example.

```{r data}
# Parameters:
set.seed(9693) 
Nb <- 100     # number of groups
Nw <- 5       # number of individuals within groups
Bb <- 0       # between-group regression coefficient on group mean
SDre <- 2.0   # between-group SD of random level relative to group mean of x
SDwithin <- 0.5  # within group SD
Bw <- 1          # within group effect of x
Ay <- 10         # intercept for response
Ax <- 20         # starting level of x
Nx <- Nw*10      # number of distinct x values

Data <- data.frame(
  group = factor(rep(1:Nb, each=Nw)),
  x = Ax + rep(1:Nx, length.out = Nw*Nb)
) |>
  within(
    {
      xm  <- ave(x, group, FUN = mean) # within-group mean
      y <- Ay +
        Bb * xm +                    # contextual effect
        Bw * (x - xm) +              # within-group effect
        rnorm(Nb, sd=SDre)[group] +  # random level by group
        rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
    }
  )
```

Here is a scatterplot of the data for a representative group of 10 (without loss of generality, the first 10) of 100 students, showing the 95% concentration ellipse for each cluster:[^2]

[^2]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.

```{r plot1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Hierarchical data set, showing the first 10 of 100 students."
library("lattice")
library("latticeExtra")
plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
               ylim=c(4, 16),
               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
    layer(panel.ellipse(..., center.cex=0))
plot # display graph
```

The between-student effect of age is 0 but the within-student effect is 1. Due to the large variation in ages between students, the least-squares regression of yodeling skill on age (for the 500 observations among all 100 students) produces an estimated slope close to 0 (though with a small $p$-value), because the slope is heavily weighted toward the between-student effect:

```{r}
summary(lm(y ~ x, data=Data))
```

The initial mixed-effects model that we fit to the data is a simple random-intercepts model:
```{r include=FALSE, echo=FALSE}
library(lme4) # necessary for some reason to knit vignette in RStudio, harmless otherwise
```
```{r}
# random intercept only:
mod.0 <- lmer(y ~ 1 + (1 | group), Data)
summary(mod.0)
```

We will shortly consider three other, more complex, mixed models; because of data-management considerations, it is convenient to fit them now, but we defer discussion of these models:

```{r}
# effect of x and random intercept:
mod.1 <- lmer(y ~ x + (1 | group), Data)

# effect of x, contextual (student) mean of x, and random intercept:
mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
        # equivalent to y ~ I(x - xm) + xm + (1 | group)

# model generating the data (where Bb = 0)
mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
```

We proceed to obtain predictions from the random-intercept model (`mod.0`) and the other models (`mod.1`, `mod.2`, and `mod.3`) based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students):

```{r}
Data <- within(Data, {
  fit_mod0.fe <- predict(mod.0, re.form = ~ 0) # fixed effects only
  fit_mod0.re <- predict(mod.0) # fixed and random effects (BLUPs)
  fit_mod1.fe <- predict(mod.1, re.form = ~ 0)
  fit_mod1.re <- predict(mod.1)
  fit_mod2.fe <- predict(mod.2, re.form = ~ 0)
  fit_mod2.re <- predict(mod.2)
  fit_mod3.fe <- predict(mod.3, re.form = ~ 0)
  fit_mod3.re <- predict(mod.3)
})
```

We then prepare the data for plotting:

```{r}
Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
Data_long$id <- 1:nrow(Data_long)
Data_long <- reshape(Data_long, direction = "long", sep = "_", 
              timevar = "modelcode",  varying = grep("_", names(Data_long)))
Data_long$model <- factor(
  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
)
```

Predictions based on the random-intercept model `mod.0` for the first 10 students are shown in the following graph:

```{r plot-fits-mod0}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictions from the random intercept model."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + (1 | group)",
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean. Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.

Our next model, `mod.1`, includes a fixed intercept and fixed effect of `x` along with a random intercept:

```{r}
summary(mod.1)
```

Predictions from this model appear in the following graph:

```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictions from the model with random intercepts and $x$ as a fixed-effect predictor."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + (1 | group)",
  ylim=c(-15, 35),
  key=list(
    corner=c(0.95, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, `mod.0`. We therefore anticipate (and show later in this section) that case-based cross-validation will prefer `mod1` to `mod0`, but that cluster-based cross-validation will prefer `mod0` to `mod1`.

Our third model, `mod.2`, includes the contextual effect of $x$---that is, the cluster mean `xm`---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:

```{r}
summary(mod.2)
```

This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | group)`, which is the model that generated the data once the coefficient of the contextual predictor `xm` is set to 0 (as it is in `mod.3`, discussed below).

Predictions from model `mod.2` appear in the following graph:
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictors from the model with random intercepts, $x$, and the group (student) mean of $x$ as predictors."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + xm + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

Depending on the estimated variance parameters of the model, a mixed model like `mod.2` will apply varying degrees of shrinkage to the random-intercept BLUPs that correspond to variation in the heights of the parallel fitted lines for the individual students. In our contrived data, the `mod.2` applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student. The fit of the mixed model `mod.2` is consequently similar to that of a fixed-effects model with age and a categorical predictor for individual students (i.e., treating students as a factor, and not shown here).

The  mixed model `mod.2` therefore fits individual observations well, and we anticipate a favorable assessment using individual-based cross-validation. In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for model `mod.2` compared to the smaller model `mod.0`, which includes only fixed and random intercepts.

Had the mixed model applied considerable shrinkage, then neither cluster-based nor case-based cross-validation would show much improvement over the random-intercept-only model. In our experience, the degree of shrinkage does not vary smoothly as parameters are changed but tends to be "all or nothing," and near the tipping point, the behavior of estimates can be affected considerably by the choice of algorithm used to fit the model.

Finally, `mod.3` directly estimates the model used to generate the data. As mentioned, it is a constrained version of `mod.2`, with the coefficient of `xm` set to 0, and with `x` expressed as a deviation from the cluster mean `xm`:

```{r}
summary(mod.3)
```

The predictions from `mod.3` are therefore similar to those from `mod.2`:

```{r plot-fits-mod3}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictions from the estimated model generating the data."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
  ylim=c(4, 16),
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only. In order to reduce between-model random variability in comparisons of models, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model:

```{r cross-validation-clusters}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept models with varying fixed effects. The error bars show the 95% confidence interval around the CV estimate of the MSE for each model."
modlist <- models("~ 1"=mod.0, "~ 1 + x"=mod.1, 
                  "~ 1 + x + xm"=mod.2, "~ 1 + I(x - xm)"=mod.3)
cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")
```

```{r cross-validation-cases}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "10-fold case-based cross-validation comparing random intercept models with varying fixed effects."
cvs_cases <- cv(modlist, data=Data, seed=9693)
plot(cvs_cases, main="Model Comparison, Case-Based CV")
```

In summary, model `mod.1`, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. Model `mod.2`, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV. The data-generating model, `mod.3`, which includes the fixed effect of `x - xm` in place of separate terms in `x` and `xm`, isn't distinguishable from model `mod.2`, which includes `x` and `xm` separately, even though `mod.2` has an unnecessary parameter (recall that the population coefficient of `xm` is 0 when `x` is expressed as deviations from the contextual mean). These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

## Example: Crossed random effects

Crossed random effects arise when the structure of the data aren't strictly hierarchical. Nevertheless, crossed and nested random effects can be handled in much the same manner, by refitting the mixed-effects model to the data with a fold of clusters or cases removed and using the refitted model to predict the response in the removed fold.

We'll illustrate with data on pig growth, introduced by @DiggleLiangZeger:1994 [Table 3.1]. The data are in the `Pigs` data frame in the **cv** package:
```{r pigs}
head(Pigs, 9)
head(xtabs(~ id + week, data=Pigs), 3)
tail(xtabs(~ id + week, data=Pigs), 3)
```
Each of 48 pigs is observed weekly over a period of 9 weeks, with the weight of the pig recorded in kg. The data are in "long" format, as is appropriate for use with the `lmer()` function in the **lme4** package. The data are very regular, with no missing cases.

The following graph, showing the growth trajectories of the pigs, is similar to Figure 3.1 in @DiggleLiangZeger:1994; we add an overall least-squares line and a loess smooth, which are nearly indistinguishable:
```{r pigs-graph}
#| out.width = "60%",
#| fig.height = 6,
#| fig.cap = "Growth trajectories for 48 pigs, with overall least-squares line (sold blue) and loess line (broken magenta)."
plot(weight ~ week, data=Pigs, type="n")
for (i in unique(Pigs$id)){
  with(Pigs, lines(x=1:9, y=Pigs[id == i, "weight"],
                   col="gray"))
}
abline(lm(weight ~ week, data=Pigs), col="blue", lwd=2)
lines(with(Pigs, loess.smooth(week, weight, span=0.5)),
      col="magenta", lty=2, lwd=2)
```
The individual "growth curves" and the overall trend are generally linear, with some tendency for variability of pig weight to increase over weeks (a feature of the data that we ignore in the mixed model that we fit to the data below).

The **Stata** mixed-effects models manual proposes a model with crossed random effects for the `Pigs` data [@Stata:2023 page 37]:

> [S]uppose that we wish to fit
$$
\mathrm{weight}_{ij} = \beta_0 + \beta_1 \mathrm{week}_{ij} + u_i + v_j + \varepsilon_{ij} 
$$
for the $i = 1, \ldots, 9$ weeks and $j = 1, \dots, 48$ pigs and
$$
u_i \sim N(0, \sigma^2_u); v_j \sim N(0, \sigma^2_v ); \varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon)
$$
all independently. That is, we assume an overall population-average growth curve $\beta_0 + \beta_1 \mathrm{week}$ and a random pig-specific shift. In other words, the effect due to week, $u_i$, is systematic to that week and  common to all pigs. The rationale behind [this model] could be that, assuming that the pigs were measured contemporaneously, we might be concerned that week-specific random factors such as weather and feeding patterns had significant systematic effects on all pigs.

Although we might prefer an alternative model,[^pig-alternative] we think that this is a reasonable specification.

[^pig-alternative]: These are repeated-measures data, which would be more conventionally modeled with autocorrelated errors within pigs. The `lme()` function in the **nlme** package, for example, is capable of fitting a mixed-model of this form.

The **Stata** manual fits the mixed model by maximum likelihood (rather than REML), and we duplicate the results reported there using `lmer()`:
```{r pigs-lmer}
m.p <- lmer(weight ~ week + (1 | id) + (1 | week),
            data=Pigs, REML=FALSE, # i.e., ML
            control=lmerControl(optimizer="bobyqa"))
summary(m.p)
```
We opt for the non-default `"bobyqa"` optimizer because it provides more numerically stable results for subsequent cross-validation in this example.

We can then cross-validate the model by omitting folds composed of pigs, folds composed of weeks, or folds composed of pig-weeks (which in the `Pigs` data set correspond to individual cases, using only the fixed effects):
```{r pigs-cv}
cv(m.p, clusterVariables="id")

cv(m.p, clusterVariables="week")

cv(m.p, clusterVariables=c("id", "week"), k=10, seed=8469)
```
We can also cross-validate the individual cases taking account of the random effects (employing the same 10 folds):
```{r pigs-cv-cases}
cv(m.p, k=10, seed=8469)
```
Because these predictions are based on BLUPs, they are more accurate than the predictions based only on fixed effects.[^crossed-vs-nested] As well, the difference between the MSE computed for the model fit to the full data and the CV estimates of the MSE is greater here than for cluster-based predictions.

[^crossed-vs-nested]: Even though there is only one observation per combination of pigs and weeks, we can use the BLUP for the omitted case because of the crossed structure of the random effects; that is each pig-week has a pig random effect and a week random effect. Although it probably isn't sensible, we can imagine a mixed model for the pig data that employs nested random effects, which would be specified by `lmer(weight ~ week + (1 | id/week), data=Pigs)`---that is, a random intercept that varies by combinations of `id` (pig) and `week`. This model can't be fit, however: With only one case per combination of `id` and `week`, the nested random-effect variance is indistinguishable from the case-level variance.

# Replicating cross-validation

Assuming that the number of cases $n$ is a multiple of the number of folds $k$---a slightly simplifying assumption---the number of possible partitions of cases into folds is $\frac{n!}{[(n/k)!]^k}$, a number that grows very large very quickly. For example, for $n = 10$ and $k = 5$, so that the folds are each of size $n/k = 2$, there are $113,400$ possible partitions; for $n=100$ and $k=5$, where $n/k = 20$, still a small problem, the number of possible partitions is truly astronomical, $1.09\times 10^{66}$.

Because the partition into folds that's employed is selected randomly, the resulting CV criterion estimates are subject to sampling error. (An exception is LOO cross-validation, which is not at all random.) To get a sense of the magnitude of the sampling error, we can repeat the CV procedure with different randomly selected partitions into folds. All of the CV functions in the **cv** package are capable of repeated cross-validation, with the number of repetitions controlled by the `reps` argument, which defaults to `1`.

Here, for example, is 10-fold CV for the Mroz logistic regression, repeated 5 times:

```{r mroz-reps}
cv(m.mroz, criterion=BayesRule, seed=248, reps=5, 
   method="Woodbury")
```
When `reps` > `1`, the result returned by `cv()` is an object of class `"cvList"`---literally a list of `"cv"` objects. The results are reported for each repetition and then averaged across repetitions, with the standard deviations of the CV criterion and the biased-adjusted CV criterion given in parentheses. In this example, there is therefore little variation across repetitions, increasing our confidence in the reliability of the results.

Notice that the seed that's set in the `cv()` command pertains to the first repetition and the seeds for the remaining repetitions are then selected pseudo-randomly.[^replicates] Setting the first seed, however, makes the entire process easily replicable, and the seed for each repetition is stored in the corresponding element of the `"cvList"` object (which isn't, however, saved in the example).

[^replicates]: Because of the manner in which the computation is performed, the order of the replicates in the `"cvList"` object returned by `cv()` isn't the same as the order in which the replicates are computed. Each element of the result, however, is a `"cv"` object with the correct random-number seed saved, and so this technical detail can be safely ignored. The individual `"cv"` objects are printed in the order in which they are stored rather than the order in which they are computed.

It's also possible to replicate CV when comparing competing models via the `cv()` method for `"modList"` objects. Recall our comparison of polynomial regressions of varying degree fit to the `Auto` data; we performed 10-fold CV for each of 10 models. Here, we replicate that process 5 times for each model and graph the results:
```{r model-comparison-with-reps}
#| out.width = "100%",
#| fig.height = 5,
#| fig.cap = " Replicated cross-validated 10-fold CV as a function of polynomial degree, $p$"
cv.auto.reps <- cv(models(m.1, m.2, m.3, m.4, m.5,
                        m.6, m.7, m.8, m.9, m.10),
                 data=Auto, seed=8004, reps=5)
plot(cv.auto.reps)
```
The graph shows both the average CV criterion and its range for each of the competing models.

# Cross-validating model selection

## A preliminary example

As @HastieTibshiraniFriedman:2009 [Sec. 7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain, if the whole data are used to select or fine-tune a statistical model, subsequent cross-validation of the model is intrinsically misleading, because the model is selected to fit the whole data, including the part of the data that remains when each fold is removed.

The following example is similar in spirit to one employed by @HastieTibshiraniFriedman:2009. Suppose that we randomly generate $n = 1000$ independent observations for a response variable variable $y \sim N(\mu = 10, \sigma^2 = 0)$, and independently sample $1000$ observations for $p = 100$ "predictors," $x_1, \ldots, x_{100}$, each from $x_j \sim N(0, 1)$. The response has nothing to do with the predictors and so the population linear-regression model $y_i = \alpha + \beta_1 x_{i1} + \cdots + \beta_{100} x_{i,100} + \varepsilon_i$ has $\alpha = 10$ and all $\beta_j = 0$.

```{r generate-selection-data}
set.seed(24361) # for reproducibility
D <- data.frame(
  y = rnorm(1000, mean=10),
  X = matrix(rnorm(1000*100), 1000, 100)
)
head(D[, 1:6])
```
Least-squares provides accurate estimates of the regression constant $\alpha = 10$ and the error variance $\sigma^2 = 1$ for the "null model" including only the regression constant; moreover, the omnibus $F$-test of the correct null hypothesis that all of the $\beta$s are 0 for the "full model" with all 100 $x$s is associated with a large $p$-value:
```{r omnibus-F}
m.full <- lm(y ~ ., data=D)
m.null <- lm(y ~ 1, data=D)
anova(m.null, m.full)

summary(m.null)
```
Next, using the `stepAIC()` function in the **MASS** package [@VenablesRipley:2002], let us perform a forward stepwise regression to select a "best" model, starting with the null model, and using AIC as the model-selection criterion (see the help page for `stepAIC()` for details):[^selection-order]

[^selection-order]: It's generally advantageous to start with the largest model, here the one with 100 predictors, and proceed by backward elimination. In this demonstration, however, where all of the $\beta$s are really 0, the selected model will be small, and so we proceed by forward selection from the null model to save computing time.

```{r forward-selection}
library("MASS")  # for stepAIC()
m.select <- stepAIC(m.null,
                    direction="forward", trace=FALSE,
                    scope=list(lower=~1, upper=formula(m.full)))
summary(m.select)
mse(D$y, fitted(m.select))
```
The resulting model has 15 predictors, a very modest $R^2 = .044$, but a small $p$-value for its omnibus $F$-test (which, of course, is entirely spurious because the same data were used to select and test the model). The MSE for the selected model is smaller than the true error variance $\sigma^2 = 1$, as is the estimated error variance for the selected model, $\widehat{\sigma}^2 = 0.973^2 = 0.947$.

If we cross-validate the selected model, we also obtain an optimistic estimate of its predictive power (although the confidence interval for the bias-adjusted MSE includes 1):
```{r cv-selectedModel}
cv(m.select, seed=2529)
```

The `cvSelect()` function in the **cv** package allows us to cross-validate the whole model-selection procedure. The first argument to `cvSelect()` is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion. The `selectStepAIC()` function, also in **cv** and based on `stepAIC()`, is suitable for use with `cvSelect()`:
```{r cvSelect-artificial-data, cache=TRUE}
cv.select <- cvSelect(selectStepAIC, data=D, seed=3791,
                      model=m.null, direction="forward",
                      scope=list(lower=~1, 
                                 upper=formula(m.full)))
cv.select
```

The other arguments to `cvSelect()` are:

* `data`, the data set to which the model is fit;
* `seed`, an optional seed for R's pseudo-random-number generator; as for `cv()`, if the seed isn't supplied by the user, a seed is randomly selected and saved;
* additional arguments required by the model-selection function, here the starting `model` argument, the `direction` of model selection, and the `scope` of models considered (from the model with only a regression constant to the model with all 100 predictors).

By default, `cvSelect()` performs 10-fold CV, and produces an estimate of MSE for the model-selection procedure even *larger* than the true error variance, $\sigma^2 = 1$.

Also by default, when the number of folds is 10 or fewer, `cvSelect()` saves the coefficients of the selected models. In this example, the `compareFolds()` function reveals that the variables retained by the model-selection process in the several folds are quite different:
```{r compare-selected-models}
compareFolds(cv.select)
```
## Mroz's logistic regression revisited

For a contrasting example we apply model selection to Mroz's logistic regression for married women's labor-force participation. First, recall the logistic regression model that we fit to the `Mroz` data:
```{r recall-Mroz-regression}
summary(m.mroz)
```

Applying stepwise model selection Mroz's logistic regression, using BIC as the model-selection criterion (via the argument `k=log(nrow(Mroz))` to `stepAIC()`) selects 5 of the 7 original predictors:
```{r mroz-selection}
m.mroz.sel <- stepAIC(m.mroz, k=log(nrow(Mroz)),
                      trace=FALSE)
summary(m.mroz.sel)
BayesRule(Mroz$lfp == "yes",
          predict(m.mroz.sel, type="response"))
```
Bayes rule applied to the selected model misclassifies 32% of the cases in the `Mroz` data.

Cross-validating the selected model produces a similar, slightly larger, estimate of misclassification, about 33%:
```{r cv-mroz-regression}
cv(m.mroz.sel, criterion=BayesRule, seed=345266)
```
Is this estimate of predictive performance optimistic?

We proceed to apply the model-selection procedure by cross-validation, producing more or less the same result:
```{r cv-mroz-selection}
m.mroz.sel.cv <- cvSelect(selectStepAIC, Mroz, 
                          seed=6681,
                          criterion=BayesRule,
                          model=m.mroz,
                          AIC=FALSE)
m.mroz.sel.cv
```
Setting `AIC=FALSE` in the call to `cvSelect()` uses the BIC rather than the AIC as the model-selection criterion. As it turns out, exactly the same predictors are selected when each of the 10 folds are omitted, and the several coefficient estimates are very similar, as we show using `compareFolds()`:
```{r compare-selected-models-mroz}
compareFolds(m.mroz.sel.cv)
```
In this example, therefore, we appear to obtain a realistic estimate of model performance directly from the selected model, because there is little added uncertainty induced by model selection.

## Cross-validating choice of transformations in regression

The **cv** package also provides a `cvSelect()` procedure, `selectTrans()`, for choosing transformations of the predictors and the response in regression.

Some background: As @Weisberg:2014 [Sec. 8.2] explains, there are technical advantages to having (numeric) predictors in linear regression analysis that are themselves linearly related. If the predictors *aren't* linearly related, then the relationships between them can often be straightened by power transformations. Transformations can be selected after graphical examination of the data, or by analytic methods. Once the relationships between the predictors are linearized, it can be advantageous similarly to transform the response variable towards normality.

Selecting transformations analytically raises the possibility of automating the process, as would be required for cross-validation. One could, in principle, apply graphical methods to select transformations for each fold, but because a data analyst couldn't forget the choices made for previous folds, the process wouldn't really be applied independently to the folds.

To illustrate, we adapt an example appearing in several places in @FoxWeisberg:2019 (for example in Chapter 3 on transforming data), using data on the prestige and other characteristics of 102 Canadian occupations circa 1970. The data are in the `Prestige` data frame in the **carData** package:
```{r Prestige-data}
data("Prestige", package="carData")
head(Prestige)
summary(Prestige)
```
The variables in the `Prestige` data set are:

* `education`: average years of education for incumbents in the occupation, from the 1971 Canadian Census.
* `income`: average dollars of annual income for the occupation, from the Census.
* `women`: percentage of occupational incumbents who were women, also from the Census.
* `prestige`: the average prestige rating of the occupation on a 0--100 "thermometer" scale, in a Canadian social survey conducted around the same time.
* `type`, type of occupation, and `census`, the Census occupational code, which are not used in our example.

The object of a regression analysis for the `Prestige` data (and their original purpose) is to predict occupational prestige from the other variables in the data set.

A scatterplot matrix (using the `scatterplotMatrix()` function in the **car** package) of the numeric variables in the data reveals that the distributions of `income` and `women` are positively skewed, and that some of the relationships among the three predictors, and between the predictors and the response (i.e., `prestige`), are nonlinear:
```{r scatterplot-matrix}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Scatterplot matrix for the `Prestige` data."
library("car")
scatterplotMatrix(~ prestige + income + education + women,
                  data=Prestige, smooth=list(spread=FALSE))
```
The `powerTransform()` function in the **car** package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964]. Several "families" of power transformations can be used, including the original Box-Cox family, simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap. 8], and @FoxWeisberg:2019 [Chap. 3] for details. Because `women` has some zero values, we use the Yeo-Johnson family:
```{r power-transform-Prestige}
trans <- powerTransform( cbind(income, education, women) ~ 1,
                         data=Prestige, family="yjPower")
summary(trans)
```
We thus have evidence of the desirability of transforming `income` (by the $1/3$ power) and `women` (by the $0.16$ power---which is close to the "0" power, i.e., the log transformation), but not `education`. Applying the "rounded" power transformations makes the predictors better-behaved:
```{r transformed-predictors}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Scatterplot matrix for the `Prestige` data with the predictors transformed."
P <- Prestige[, c("prestige", "income", "education", "women")]
(lambdas <- trans$roundlam)
names(lambdas) <- c("income", "education", "women")
for (var in c("income", "education", "women")){
  P[, var] <- yjPower(P[, var], lambda=lambdas[var])
}
summary(P)

scatterplotMatrix(~ prestige + income + education + women,
                  data=P, smooth=list(spread=FALSE))
```
Comparing the MSE for the regressions with the original and transformed predictors shows a advantage to the latter:
```{r prestige-regressions}
m.pres <- lm(prestige ~ income + education + women, data=Prestige)
m.pres.trans <- lm(prestige ~ income + education + women, data=P)
mse(Prestige$prestige, fitted(m.pres))
mse(P$prestige, fitted(m.pres.trans))
```
Similarly, component+residual plots for the two regressions, produced by the `crPlots()` function in the **car** package, suggest that the partial relationship of `prestige` to `income` is more nearly linear in the transformed data, but the transformation of `women` fails to capture what appears to be a slight quadratic partial relationship; the partial relationship of `prestige` to `education` is close to linear in both regressions:
```{r CR-plots-untransformed}
#| fig.cap = "Component+residual plots for the `Prestige` regression with the original predictors."
crPlots(m.pres)
```
```{r CR-plots-transformed}
#| fig.cap = "Component+residual plots for the `Prestige` regression with transformed predictors."
crPlots(m.pres.trans)
```

Having transformed the predictors towards multinormality, we now consider whether there's evidence for transforming the response (using `powerTransform()` for Box and Cox's original method), and we discover that there's not:
```{r transform-response}
summary(powerTransform(m.pres.trans))
```

The `selectTrans()` function in the **cv** package automates the process of selecting predictor and response transformations. The function takes a `data` set and "working" `model` as arguments, along with the candidate `predictors` and `response` for transformation, and the transformation `family` to employ. If the `predictors` argument is missing then only the response is transformed, and if the `response` argument is missing, only the supplied predictors are transformed. The default `family` for transforming the predictors is `"bcPower"`---the original Box-Cox family---as is the default `family.y` for transforming the response; here we specify `family="yjPower` because of the zeros in `women`. `selectTrans()` returns the result of applying a lack-of-fit criterion to the model after the selected transformation is applied, with the default `criterion=mse`:
```{r selectTrans}
selectTrans(data=Prestige, model=m.pres,
            predictors=c("income", "education", "women"),
            response="prestige", family="yjPower")
```
`selectTrans()` also takes an optional `indices` argument, making it suitable for doing computations on a subset of the data (i.e., a CV fold), and hence for use with `cvSelect()` (see `?selectTrans` for details):

```{r cv-select-transformations}
cvs <- cvSelect(selectTrans, data=Prestige, model=m.pres, seed=1463,
                predictors=c("income", "education", "women"),
                response="prestige",
                family="yjPower")
cvs

cv(m.pres, seed=1463) # untransformed model with same folds

compareFolds(cvs)
```
The results suggest that the predictive power of the transformed regression is reliably greater than that of the untransformed regression (though in both case, the cross-validated MSE is considerably higher than the MSE computed for the whole data). Examining the selected transformations for each fold reveals that the predictor `education` and the response `prestige` are never transformed; that the $1/3$ power is selected for `income` in all of the folds; and that the transformation selected for `women` varies narrowly across the folds between the $0$th power (i.e., log) and the $1/3$ power.

## Selecting both transformations and predictors[^Venables]

[^Venables]: The presentation in the section benefits from an email conversation with Bill Venables, who of course isn't responsible for the use to which we've put his insightful remarks.

As we mentioned, @HastieTibshiraniFriedman:2009 [Sec. 7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain that honest cross-validation has to take account of model specification and selection. Statistical modeling is at least partly a craft, and one could imagine applying that craft to successive partial data sets, each with a fold removed. The resulting procedure would be tedious, though possibly worth the effort, but it would also be difficult to realize in practice: After all, we can hardly erase our memory of statistical modeling choices between analyzing partial data sets.

Alternatively, if we're able to automate the process of model selection, then we can more realistically apply CV mechanically. That's what we did in the preceding two sections, first for predictor selection and then for selection of transformations in regression. In this section, we consider the case where we both select variable transformations and then proceed to select predictors. It's insufficient to apply these steps sequentially, first, for example, using `cvSelect()` with `selectTrans()` and then with `selectStepAIC()`; rather we should apply the whole model-selection procedure with each fold omitted. The `selectTransAndStepAIC()` function, also supplied by the **cv** package, does exactly that.

To illustrate this process, we return to the `Auto` data set:
```{r Auto-redux}
summary(Auto)
xtabs(~ year, data=Auto)
xtabs(~ origin, data=Auto)
xtabs(~ cylinders, data=Auto)
```
We previously used the `Auto` here in a preliminary example where we employed CV to inform the selection of the order of a polynomial regression of `mpg` on `horsepower`. Here, we consider more generally the problem of predicting `mpg` from the other variables in the `Auto` data. We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set:
```{r Auto-explore}
#| out.width = "100%",
#| fig.height = 7,
#| fig.cap = "Scatterplot matrix for the numeric variables in the `Auto` data"
Auto$cylinders <- factor(Auto$cylinders,
                         labels=c("3.4", "3.4", "5.6", "5.6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels=c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique=TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
                  smooth=list(spread=FALSE), data=Auto)
```
A comment before we proceed: `origin` is clearly categorical and so converting it to a factor is natural, but we could imagine treating `cylinders` and `year` as numeric predictors. There are, however, only 5 distinct values of `cylinders` (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare. and none of the cars has 7 cylinders. There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between `mpg` and `year` is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.

[^year]: Of course, making the decision to treat `year` as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process. It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation. We believe, however, that it's still useful to automate as much of the process as we can to obtain a more realistic, if still biased, estimate of the predictive power of a model.

We begin with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
```{r Auto-working-model}
#| out.width = "100%",
#| fig.height = 7,
#| fig.cap = "Component+residual plots for the working model fit to the `Auto` data"
m.auto <- lm(mpg ~ ., data = Auto)
summary(m.auto)

Anova(m.auto)

crPlots(m.auto)
```
The component+residual plots, created with the `crPlots()` function in the previously loaded **car** package, clearly reveal the inadequacy of the model.

We proceed to transform the numeric predictors towards multi-normality: 
```{r Auto-transform}
num.predictors <- c("displacement", "horsepower", "weight", "acceleration")
tr.x <- powerTransform(Auto[, num.predictors])
summary(tr.x)
```
We then apply the (rounded) transformations---all, as it turns out, logs---to the data and re-estimate the model:
```{r Auto-with-transformed-predictors}
A <- Auto
powers <- tr.x$roundlam
for (pred in num.predictors){
  A[, pred] <- bcPower(A[, pred], lambda=powers[pred])
}
head(A)

m <- update(m.auto, data=A)
```
Finally, we perform Box-Cox regression to transform the response (also obtaining a log transformation):
```{r Auto-Box-Cox}
summary(powerTransform(m))

m <- update(m, log(mpg) ~ .)
summary(m)

Anova(m)
```

The transformed numeric variables are much better-behaved:
```{r Auto-transformed-scatterplot-matrix}
#| out.width = "100%",
#| fig.height = 7,
#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the `Auto` data"
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread=FALSE), data=A)
```
And the partial relationships in the model fit to the transformed data are much more nearly linear:
```{r Auto-CR-plots-transformed}
#| out.width = "100%",
#| fig.height = 7,
#| fig.cap = "Component+residual plots for the model fit to the transformed `Auto` data"
crPlots(m)
```

Having transformed both the numeric predictors and the response, we proceed to use the `stepAIC()` function in the **MASS** package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log(n)$):
```{r}
m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
summary(m.step)

Anova(m.step)
```
The selected model includes three of the numeric predictors, `horsepower`, `weight`, and `acceleration`, along with the factors `year` and `origin`. We can calculate the MSE for this model, but we expect that the result will be optimistic because we used the whole data to help specify the model
```{r MSE-whole-selected-model}
mse(Auto$mpg, exp(fitted(m.step)))
```
This is considerably smaller than the MSE for the original working model:
```{r MSE-working-model}
mse(Auto$mpg, fitted(m.auto))
```
A perhaps subtle point is that we compute the MSE for the selected model on the original `mpg` response scale rather than the log scale, so as to make the selected model comparable to the working model. That's slightly uncomfortable given the skewed distribution of `mpg`. An alternative is to use the median absolute error instead of the mean-squared error, employing the `medAbsErr()` function from the **cv** package:
```{r Auto-median-absolute-error}
medAbsErr(Auto$mpg, exp(fitted(m.step)))
medAbsErr(Auto$mpg, fitted(m.auto))
```

Now let's use `cvSelect()` with `selectTransAndStepAIC()` to automate and cross-validate the whole model-specification process:
```{r Auto-transform-and-select}
num.predictors
cvs <- cvSelect(selectTransStepAIC, data=Auto, seed=76692, model=m.auto,
                predictors=num.predictors,
                response="mpg", AIC=FALSE, criterion=medAbsErr)
cvs

compareFolds(cvs)
```
Here, as for `selectTrans()`, the `predictors` and `response` arguments specify candidate variables for transformation,
and `AIC=FALSE` uses the BIC for model selection. The starting model, `m.auto`, is the working model fit to the `Auto` data. The CV criterion isn't bias-adjusted because median absolute error isn't a mean of casewise error components.

Some noteworthy points:

* `selectTransStepAIC()` automatically computes CV cost criteria, here the median absolute error, on the untransformed response scale.
* The estimate of the median absolute error that we obtain by cross-validating the whole model-specification process is a little larger than the median absolute error computed for the model we fit to the `Auto` data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
* When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of `compareFolds()`), we see that there is little uncertainty in choosing variable transformations (the `lam.*`s for the $x$s and `lambda` for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: `horsepower`, `weight`, and `year` are always included among the selected predictors; `acceleration` and `displacement` are each included respectively in 4 and 3 of 10 selected models; and `cylinders` and `origin` are each included in only 1 of 10 models. Recall that when we selected predictors for the full data, we obtained a model with `horsepower`, `weight`, `acceleration`, `year`, and `origin`.

# Parallel computations

The CV functions in the **cv** package are all capable of performing parallel computations by setting the `ncores` argument (specifying the number of computer cores to be used) to a number > `1` (which is the default). Parallel computation can be advantageous for large problems, reducing the execution time of the program.

To illustrate, let's time model selection in Mroz's logistic regression, repeating the computation as performed previously and then doing it in parallel using 2 cores:
```{r parallel-computation}
system.time(m.mroz.sel.cv <- cvSelect(selectStepAIC, Mroz,
                          seed=6681,
                          criterion=BayesRule,
                          model=m.mroz,
                          AIC=FALSE))

system.time(m.mroz.sel.cv.p <- cvSelect(selectStepAIC, Mroz,
                          seed=6681,
                          criterion=BayesRule,
                          model=m.mroz,
                          AIC=FALSE,
                          ncores=2))
all.equal(m.mroz.sel.cv, m.mroz.sel.cv.p)
```

In this small problem, the parallel computation is actually *slower*, because there is an overhead cost to parallelization, but we can see that it produces the same result as before.

# Extending the cv package

The **cv** package is designed to be extensible in several directions. In this vignette, we discuss three kinds of extensions, ordered by increasing general complexity: (1) adding a cross-validation cost criterion; (2) adding a model class that's not directly accommodated by the `cv()` default method or by another directly inherited method, with separate consideration of mixed-effects models; and (3) adding a new model-selection procedure suitable for use with `selectModel()`.

## Adding a cost criterion

A cost criterion suitable for use with `cv()` or `cvSelect()` should take two arguments, `y` (the observed response vector) and `yhat` (a vector of fitted or predicted response values), and return a numeric index of lack of fit. The **cv** package supplies several such criteria: `mse(y, yhat)`, which returns the mean-squared prediction error for a numeric response; `rmse(y, yhat)`, which returns the (square-)root mean-squared error; `medAbsErr(y, yhat)`, which returns the median absolute error; and `BayesRule(y, yhat)` (and its non-error-checking version, `BayesRule2(y, yhat))`, suitable for use with a binary regression model, where `y` is the binary response coded `0` for a "failure" or `1` for a "success"; where `yhat` is the predicted probability of success; and where the proportion of *incorrectly* classified cases is returned.

To illustrate using a different prediction cost criterion, we'll base a cost criterion on the area under the receiver operating characteristic ("ROC") curve for a logistic regression. The ROC curve is a graphical representation of the classification power of a binary regression model, and the area under the ROC curve ("AUC"), which varies from 0 to 1, is a common summary measure based on the ROC [see @Wikipedia-ROC:2023]. The **Metrics** package [@HamnerFrasco:2018] includes a variety of measures useful for model selection, including an `auc()` function. We convert the AUC into a cost measure by taking its complement:

```{r AUCcomp}
AUCcomp <- function(y, yhat) 1 - Metrics::auc(y, yhat)
```

We then apply `AUCcomp()` to the the Mroz logistic regression discussed in the main **cv** package vignette, which we reproduce here, using the `Mroz` data frame from the **carData** package [@FoxWeisberg:2019]:
```{r Mroz-regression}
data("Mroz", package="carData")
m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
summary(m.mroz)

AUCcomp(with(Mroz, as.numeric(lfp == "yes")), fitted(m.mroz))
```
Cross-validating this cost measure is straightforward:
```{r Mroz-CV-ROC}
library("cv")
cv(m.mroz, criterion=AUCcomp, seed=3639)
```
As expected, the cross-validated complement to the AUC is somewhat less optimistic than the criterion computed from the model fit to the whole data set.

As we explain in the vignette "Cross-validation of regression models," the `cv()` function differentiates between CV criteria that are averages of casewise components and criteria that are not. Computation of bias corrections and confidence intervals is limited to the former. We show in the appendix to this vignette that the AUC, and hence its complement, cannot be expressed as averages of casewise components.

`cv()` looks for a `"casewise loss"` attribute of the value returned by a CV criterion function. If this attribute exists, then the criterion is treated as the mean of casewise components, and `cv()` uses the unexported function `getLossFn()` to construct a function that returns the casewise components of the criterion.

We illustrate with the `mse()`:
```{r}
mse

cv:::getLossFn(mse(rnorm(100), rnorm(100)))
```
For this scheme to work, the "casewise loss" attribute must be a character string (or vector of character strings), here `"(y - yhat)^2"`, that evaluates to an expression that is a function of `y` and `yhat`, and that computes the vector of casewise components of the CV criterion.

## Adding a model class not covered by the default cv() method

### Independently sampled cases

Suppose that we want to cross-validate a multinomial logistic regression model fit by the `multinom()` function in the **nnet** package [@VenablesRipley:2002]. We borrow an example from @Fox:2016 [Sec. 14.2.1], with data from the British Election Panel Study on vote choice in the 2001 British election. Data for the example are in the `BEPS` data frame in the **carData** package:
```{r BEPS-data}
data("BEPS", package="carData")
head(BEPS)
```
The polytomous (multi-category) response variable is `vote`, a factor with levels `"Conservative"`, `"Labour"`, and `"Liberal Democrat"`. The predictors of `vote` are:

* `age`, in years;
* `econ.cond.national` and `econ.cond.household`, the respondent's ratings of the state of the economy, on 1 to 5 scales.
* `Blair`, `Hague`, and `Kennedy`, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
* `Europe`, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism."
* `political.knowledge`, knowledge of the parties' positions on European integration, with scores from 0 to 3.
* `gender`, `"female"` or `"male"`.

The model fit to the data includes an interaction between `Europe` and `political.knowledge`; the other predictors enter the model additively:
```{r BEPS-model}
library("nnet")
m.beps <- multinom(vote ~ age + gender + economic.cond.national +
                       economic.cond.household + Blair + Hague + Kennedy +
                       Europe*political.knowledge, data=BEPS)

car::Anova(m.beps)
```
Most of the predictors, including the `Europe` $\times$ `political.knowledge` interaction, are associated with very small $p$-values; the `Anova()` function is from the **car** package [@FoxWeisberg:2019].

Here's an "effect plot", using the the **effects** package [@FoxWeisberg:2019] to visualize the `Europe` $\times$ `political.knowledge` interaction in a "stacked-area" graph:
```{r BEPS-plot, fig.width=9, fig.height=5}
plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
            xlevels=list(Europe=1:11, political.knowledge=0:3),
            fixed.predictors=list(given.values=c(gendermale=0.5))),
     lines=list(col=c("blue", "red", "orange")),
     axes=list(x=list(rug=FALSE), y=list(style="stacked")))
```

To cross-validate this multinomial-logit model we need an appropriate cost criterion. None of the criteria supplied by the **cv** package---for example, neither `mse()`, which is appropriate for a numeric response, nor `BayesRule()`, which is appropriate for a binary response---will do. One possibility is to adapt Bayes rule to a polytomous response:
```{r BayesRuleMulti}
head(BEPS$vote)
yhat <- predict(m.beps, type="class")
head(yhat)

BayesRuleMulti <- function(y, yhat){
  result <- mean(y != yhat)
  attr(result, "casewise loss") <- "y != yhat"
  result
}

BayesRuleMulti(BEPS$vote, yhat)
```
The `predict()` method for `"multinom"` models called with argument `type="class"` reports the Bayes-rule prediction for each case---that is, the response category with the highest predicted probability. Our `BayesRuleMulti()` function calculates the proportion of misclassified cases. Because this value is the mean of casewise components, we attach a `"casewise loss"` attribute to the result (as explained in the preceding section).

The marginal proportions for the response categories are
```{r BEPS-response-distribution}
xtabs(~ vote, data=BEPS)/nrow(BEPS)
```
and so the marginal Bayes-rule prediction, that everyone will vote Labour, produces an error rate of $1 - 0.47213 = 0.52787$. The multinomial-logit model appears to do substantially better than that, but does its performance hold up to cross-validation?

We check first whether the default `cv()` method works "out-of-the-box" for the `"multinom"` model:
```{r BEPS-test-default, error=TRUE}
cv(m.beps, seed=3465, criterion=BayesRuleMulti)
```
The default method of `GetResponse()` (a function supplied by the **cv** package---see `?GetResponse`) fails for a `"multinom"` object. A straightforward solution is to supply a `GetResponse.multinom()` method that returns the factor response [using the `get_response()` function from the **insight** package, @LudeckeWaggonerMakowski:2019],
```{r GetResponse.multinom}
GetResponse.multinom <- function(model, ...) {
  insight::get_response(model)
}

head(GetResponse(m.beps))
```
and to try again:
```{r BEPS-test-default-2, error=TRUE}
cv(m.beps, seed=3465, criterion=BayesRuleMulti)
```
A `traceback()` (not shown) reveals that the problem is that the default method of `cv()` calls the `"multinom"` method for `predict()` with the argument `type="response"`, when the correct argument should be `type="class"`.  We therefore must write a "`multinom`" method for `cv()`, but that proves to be very simple:
```{r cv.nultinom}
cv.multinom <- function (model, data, criterion=BayesRuleMulti, k, reps,
                         seed, ...){
  NextMethod(type="class", criterion=criterion)
}
```
That is, we simply call the default `cv()` method with the `type` argument properly set. In addition to supplying the correct `type` argument, our method sets the default `criterion` for the `cv.multinom()` method to `BayesRuleMulti`.

Then:
```{r BEPS-cv}
m.beps <- update(m.beps, trace=FALSE)
cv(m.beps, seed=3465)
```
Prior to invoking `cv()`, we called `update()` with `trace=FALSE` to suppress the iteration history reported by default by `multinom()`---it would be tedious to see the iteration history for each fold. The cross-validated polytomous Bayes-rule criterion confirms that the fitted model does substantially better than the marginal Bayes-rule prediction that everyone votes for Labour.

### Mixed-effects models

Adding a `cv()` method for a mixed-model class is somewhat more complicated. We provide the `cvMixed()` function to facilitate this process, and to see how that works, consider the `"lme"` method from the **cv** package:
```{r cv.lme}
cv:::cv.lme
```
Notice that `cv.lme()` sets up a call to `cvMixed()`, which does the computational work.

Most of the arguments of `cvMixed()` are familiar:

* `model` is the mixed-model object, here of class `"lme"`.

* `package` is the name of the package in which the mixed-modeling function used to fit the model, here  `lme()`, resides---i.e., `"nlme"`; `cvMixed()` uses this argument to retrieve the package namespace.

* `data` is the data set to which the model is fit, by default extracted by the `get_data()` function in the **insight** package.

* `criterion` is the CV criterion, defaulting to the `mse()` function.

* `k` is the number of CV folds, defaulting to `"loo"` for CV by clusters and `10` for CV by cases.

* `reps` is the number of times the CV process is repeated, defaulting to `1`.

* `seed` is the seed for R's random-number generator, defaulting to a randomly selected (and saved) value.

* `ncores` is the number of cores to use for parallel computation; if `1`, the default, then the computation isn't parallelized.

* `clusterVariables` is a character vector of the names of variables defining clusters; if missing, then CV is based on cases rather than clusters.

The remaining two arguments are unfamiliar:

* `predict.clusters.args` is a named list of arguments to be passed to the `predict()` function to obtain predictions for the full data set from a model fit to a subset of the data for cluster-based CV. The first two arguments should be `object` and `newdata`. It is typically necessary to tell `cvMixed()` how to base predictions only on fixed effects; in the case of `"lme"` models, this is done by setting `level = 0`.

* Similarly, `predict.cases.args` is a named list of arguments to be passed to `predict()` for case-based CV. Setting `level = 1` includes random effects in the predictions.

Finally, any additional arguments, absorbed by `...`, are passed to `update()` when the model is refit with each fold omitted. `cvMixed()` returns an object of class `"cv"`.

Now imagine that we want to support a new class of mixed-effects models. To be concrete, we illustrate with the `glmmPQL()` function in the **MASS** package [@VenablesRipley:2002], which fits generalized-linear mixed-effects models by penalized quasi-likelihood.[^glmmPQL] Not coincidentally, the arguments of `glmmPQL()` are similar to those of `lme()` (with an additional `family` argument), because the former iteratively invokes the latter; so `cv.glmmPQL()` should resemble `cv.lme()`.

[^glmmPQL]: This example is somewhat artificial in that `glmmPQL()` has largely been superseded by computationally superior functions, such the `glmer()` function in the **lme4** package. There is, however, one situation in which `glmmPQL()` might prove useful: to specify serial dependency in case-level errors within clusters for longitudinal data, which is not currently supported by `glmer()`.

As it turns out, neither the default method for `GetResponse()` nor `insight::get_data()` work for `"glmmPQL"` objects. These objects include a `"data"` element, however, and so we can simply extract this element as the default for the `data` argument of our `cv.glmmPQL()` method.

To get the response variable is more complicated: We refit the fixed part of the model as a GLM with only the regression constant on the right-hand side, and extract the response from that; because all we need is the response variable, we limit the number of GLM iterations to 1 and suppress warning messages about non-convergence:

```{r GetResponse.glmmPQL}
GetResponse.glmmPQL <- function(model, ...){
  f <- formula(model)
  f[[3]] <- 1 # regression constant only on RHS
  model <- suppressWarnings(glm(f, data=model$data, family=model$family,
                                control=list(maxit=1)))
  cv::GetResponse(model)
}
```

Writing the `cv()` method is then straightforward:
```{r cv.glmmPQL}
cv.glmmPQL <- function(model, data = model$data, criterion = mse,
                     k, reps = 1, seed, ncores = 1, clusterVariables, ...){
  cvMixed(
    model,
    package="MASS",
    data=data,
    criterion=criterion,
    k=k,
    reps=reps,
    seed=seed,
    ncores=ncores,
    clusterVariables=clusterVariables,
    predict.clusters.args=list(object=model,
                               newdata=data,
                               level=0,
                               type="response"),
    predict.cases.args=list(object=model,
                            newdata=data,
                            level=1,
                            type="response"),
    verbose=FALSE,
    ...)
}
```
We set the argument `verbose=FALSE` to suppress `glmmPQL()`'s iteration counter when `cvMixed()` calls `update()`.

Let's apply our newly minted method to a logistic regression with a random intercept in an example that appears in `?glmmPQL`:
```{r glmmPQL-example}
library("MASS")
m.pql <- glmmPQL(y ~ trt + I(week > 2), random = ~ 1 | ID,
             family = binomial, data = bacteria)
summary(m.pql)
```
We compare this result to that obtained from `glmer()` in the **lme4** package:
```{r compare-to-lme4}
library("lme4")
m.glmer <- glmer(y ~ trt + I(week > 2) + (1 | ID),
               family = binomial, data = bacteria)
summary(m.glmer)

  # comparison of fixed effects:
car::compareCoefs(m.pql, m.glmer) 
```
The two sets of estimates are similar, but not identical

Finally, we try out our `cv.glmmPQL()` method, cross-validating both by clusters and by cases,
```{r cv-example-glmmPQL, cache=TRUE}
cv(m.pql, clusterVariables="ID", criterion=BayesRule)

cv(m.pql, data=bacteria, criterion=BayesRule, seed=1490)
```
and again compare to `glmer()`:
```{r cv-example-glmer, cache=TRUE}
cv(m.glmer, clusterVariables="ID", criterion=BayesRule)

cv(m.glmer, data=bacteria, criterion=BayesRule, seed=1490)
```

## Adding a model-selection procedure

The `selectStepAIC()` function supplied by the **cv** package, which is based on the `stepAIC()` function from the **nnet** package [@VenablesRipley:2002] for stepwise model selection, is suitable for the `procedure` argument of `cvSelect()`. The use of `selectStepAIC()` is illustrated in the principal vignette for the package.

We'll employ `selectStepAIC()` as a "template" for writing a CV model-selection procedure. To see the code for this function, type `cv::selectStepAIC` at the R command prompt, or examine the sources for the **cv** package at https://github.com/gmonette/cv (the code for `selectStepAIC()` is in https://github.com/gmonette/cv/blob/main/R/cvSelect.R).

Another approach to model selection is all-subsets regression. The `regsubsets()` function in the **leaps** package [@LumleyMiller:2020] implements an efficient algorithm for selecting the best-fitting linear least-squares regressions for subsets of predictors of all sizes, from 1 through the maximum number of candidate predictors.[^1] To illustrate the use of `regsubsets()`, we employ the `swiss` data frame supplied by the **leaps** package:

[^1]: The `regsubsets()` function computes several measures of model predictive performance, including the $R^2$ and $R^2$ adjusted for degrees of freedom, the residual sums of squares, Mallows's $C_p$, and the BIC. Several of these are suitable for comparing models with differing numbers of coefficients---we use the BIC below---but all necessarily agree when comparing models with the *same* number of coefficients.

```{r swiss}
library("leaps")
head(swiss)
nrow(swiss)
```
The data set includes the following variables, for each of 47 French-speaking Swiss provinces circa 1888:

* `Fertility`: A standardized fertility measure.
* `Agriculture`: The percentage of the male population engaged in agriculture.
* `Examination`: The percentage of draftees into the Swiss army receiving the highest grade on an examination.
* `Education`: The percentage of draftees with more than a primary-school education.
* `Catholic`: The percentage of the population who were Catholic.
* `Infant.Mortality`: The infant-mortality rate, expressed as the percentage of live births surviving less than a year.

Following @LumleyMiller:2020, we treat `Fertility` as the response and the other variables as predictors in a linear least-squares regression:
```{r swiss-lm}
m.swiss <- lm(Fertility ~ ., data=swiss)
summary(m.swiss)

cv(m.swiss, seed=8433)
```
Thus, the RMSE for the model fit to the complete data is considerably smaller than the CV estimate of the RMSE. Can we do better by selecting a subset of the predictors, taking account of the additional uncertainty induced by model selection?

First, let's apply best-subset selection to the complete data set:
```{r subset-selection}
#| fig.cap = "Selecting the best model of each size."
swiss.sub <- regsubsets(Fertility ~ ., data=swiss)
summary(swiss.sub)

(bics <- summary(swiss.sub)$bic)
which.min(bics)

car::subsets(swiss.sub, legend="topright")
```
The graph, produced by the `subsets()` function in the **car** package, shows that the model with the smallest BIC is the best model with 4 predictors, including `Agriculture`, `Education`, `Catholic`, and `Infant.Mortality`, but not `Examination`:
```{r best-model}
m.best <- update(m.swiss, . ~ . - Examination)
summary(m.best)

cv(m.best, seed=8433) # use same folds as before
```
The RMSE for the selected model is (of course) slightly higher than for the full model fit previously, but the cross-validated RMSE is a bit lower; as we explain in the main vignette, however, it isn't kosher to select and cross-validate a model on the same data.

Here's a function named `selectSubsets()`, meant to be used with `cvSelect()`, suitable for cross-validating the model-selection process:
```{r selectSubsets}
selectSubsets <- function(data=insight::get_data(model), 
                          model,
                          indices,
                          criterion=mse,
                          save.coef=TRUE, ...){

  if (inherits(model, "lm", which=TRUE) != 1)
    stop("selectSubsets is appropriate only for 'lm' models")

  y <- GetResponse(model)
  formula <- formula(model)
  X <- model.matrix(model)

  if (missing(indices)) {
    # select the best model from the full data by BIC
    sel <- leaps::regsubsets(formula, data=data, ...)
    bics <- summary(sel)$bic
    best <- coef(sel, 1:length(bics))[[which.min(bics)]]
    x.names <- names(best)
    # fit the best model; intercept is already in X, hence - 1:
    m.best <- lm(y ~ X[, x.names] - 1) 
    fit.all <- predict(m.best, newdata=data)
    return(criterion(y, fit.all)) # return the CV criterion
  }

  # select the best model omitting the i-th fold (given by indices)
  sel.i <- leaps::regsubsets(formula, data[-indices, ], ...)
  bics.i <- summary(sel.i)$bic
  best.i <- coef(sel.i, 1:length(bics.i))[[which.min(bics.i)]]
  x.names.i <- names(best.i)
  m.best.i <- lm(y[-indices] ~ X[-indices, x.names.i] - 1)
              # predict() doesn't work here:
  fit.all.i <- as.vector(X[, x.names.i] %*% coef(m.best.i))
  fit.i <- fit.all.i[indices]
  # return the fitted values for i-th fold, CV criterion for all cases, 
  #   and the regression coefficients
  list(fit.i=fit.i, # fitted values for i-th fold
       crit.all.i=criterion(y, fit.all.i), # CV crit for all cases
       coefficients = if (save.coef){ # regression coefficients
         coefs <- coef(m.best.i)

         # fix coefficient names
         names(coefs) <- sub("X\\[-indices, x.names.i\\]", "",
                             names(coefs))

         coefs
       }  else {
         NULL
       }
  )
}
```
A slightly tricky point is that because of scoping issues, `predict()` doesn't work with the model fit omitting the $i$th fold, and so the fitted values for all cases are computed directly as $\widehat{\mathbf{y}}_{-i} = \mathbf{X} \mathbf{b}_{-i}$, where $\mathbf{X}$ is the model-matrix for all of the cases, and $\mathbf{b}_{-i}$ is the vector of least-squares coefficients for the selected model with the $i$th fold omitted.

Additionally, the command `lm(y[-indices] ~ X[-indices, x.names.i] - 1)`, which is the selected model with the $i$th fold deleted, produces awkward coefficient names like `"X[-indices, x.names.i]Infant.Mortality"`. Purely for aesthetic reasons, the command `sub("X\\[-indices, x.names.i\\]", "", names(coefs))` fixes these awkward names, removing the extraneous text, `"X[-indices, x.names.i]"`.

Applying `selectSubsets()` to the full data produces the full-data cross-validated RMSE (which we obtained previously):
%%```{r test-selectSubsets}
%%selectSubsets(model=m.swiss)
%%```
Similarly, applying the function to an imaginary "fold" of 5 cases returns the RMSE for the cases in the fold, based on the model selected and fit to the cases omitting the fold; the RMSE for all of the cases, based on the same model; and the coefficients of the selected model, which includes 4 or the 5 predictors (and the intercept):
%%```{r test-selectSubsets-fold}
%%selectSubsets(model=m.swiss, indices=seq(5, 45, by=10))
%%```

Then, using `selectSubsets()` in cross-validation, we get:
```{r cvSelect-selectSubsets}
(cv.swiss <- cvSelect(selectSubsets, model=m.swiss,
                      data=swiss, seed=8433)) # use same folds
```
Cross-validation shows that model selection exacts a penalty in RMSE. Examining the models selected for the 10 folds reveals that there is some uncertainty in identifying the predictors in the "best" model, with `Agriculture` sometimes appearing and sometimes not:
```{r best-models-by-folds}
compareFolds(cv.swiss)
```

# Computational notes

## Efficient computations for linear and generalized linear models

The most straightforward way to implement cross-validation in R for statistical modeling functions that are written in the canonical manner is to use `update()` to refit the model with each fold removed. This is the approach taken in the default method for `cv()`, and it is appropriate if the cases are independently sampled. Refitting the model in this manner for each fold is generally feasible when the number of folds in modest, but can be prohibitively costly for leave-one-out cross-validation when the number of cases is large.

The `"lm"` and `"glm"` methods for `cv()` take advantage of computational efficiencies by avoiding refitting the model with each fold removed. Consider, in particular, the weighted linear model $\mathbf{y}_{n \times 1} = \mathbf{X}_{n \times p}\boldsymbol{\beta}_{p \times 1} + \boldsymbol{\varepsilon}_{n \times 1}$, where $\boldsymbol{\varepsilon} \sim \mathbf{N}_n \left(\mathbf{0}, \sigma^2 \mathbf{W}^{-1}_{n \times n}\right)$. Here, $\mathbf{y}$ is the response vector,  $\mathbf{X}$ the model matrix, and $\boldsymbol{\varepsilon}$ the error vector, each for $n$ cases, and $\boldsymbol{\beta}$ is the vector of $p$ population regression coefficients. The errors are assumed to be multivariately normally distributed with 0 means and covariance matrix $\sigma^2 \mathbf{W}^{-1}$, where $\mathbf{W} = \mathrm{diag}(w_i)$ is a diagonal matrix of inverse-variance weights. For the linear model with constant error variance, the weight matrix is taken to be $\mathbf{W} = \mathbf{I}_n$, the order-$n$ identity matrix.

The weighted-least-squares (WLS) estimator of $\boldsymbol{\beta}$ is [see, e.g., @Fox:2016, Sec. 12.2.2] [^WLS]
$$
\mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} 
  \mathbf{X}^T \mathbf{W} \mathbf{y}
$$

[^WLS]: This is a definitional formula, which assumes that the model matrix $\mathbf{X}$ is of full column rank, and which can be subject to numerical instability when $\mathbf{X}$ is ill-conditioned. `lm()` uses the singular-value decomposition of the model matrix to obtain computationally more stable results.

Fitted values are then $\widehat{\mathbf{y}} = \mathbf{X}\mathbf{b}_{\mathrm{WLS}}$.

The LOO fitted value for the $i$th case can be efficiently computed by $\widehat{y}_{-i} = y_i - e_i/(1 - h_i)$ where $h_i = \mathbf{x}^T_i \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} \mathbf{x}_i$ (the so-called "hatvalue"). Here, $\mathbf{x}^T_i$ is the $i$th row of $\mathbf{X}$, and $\mathbf{x}_i$ is the $i$th row written as a column vector. This approach can break down when one or more hatvalues are equal to 1, in which case the formula for $\widehat{y}_{-i}$ requires division by 0.

To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identify [@Wikipedia-Woodbury:2023],
$$
\left(\mathbf{A}_{m \times m} + \mathbf{U}_{m \times k} 
\mathbf{C}_{k \times k} \mathbf{V}_{k \times m} \right)^{-1} = \mathbf{A}^{-1} - \mathbf{A}^{-1}\mathbf{U} \left(\mathbf{C}^{-1} + 
\mathbf{VA}^{-1}\mathbf{U} \right)^{-1} \mathbf{VA}^{-1}
$$
where $\mathbf{A}$ is a nonsingular order-$n$ matrix. We apply this result by letting
\begin{align*}
	\mathbf{A} &= \mathbf{X}^T \mathbf{W} \mathbf{X} \\
	\mathbf{U} &= \mathbf{X}_\mathbf{j}^T \\
	\mathbf{V} &= - \mathbf{X}_\mathbf{j} \\
	\mathbf{C} &= \mathbf{W}_\mathbf{j} \\
\end{align*}
where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jm})^T$ represents the vector of indices for the cases in the $j$th fold, $j = 1, \ldots, k$. The negative sign in $\mathbf{V} = - \mathbf{X}_\mathbf{j}$ reflects the *removal*, rather than addition, of the cases in $\mathbf{j}$.

Applying the Woodbury identity isn't quite as fast as using the hatvalues, but it is generally much faster than refitting the model. A disadvantage of the Woodbury identity, however, is that it entails explicit matrix inversion and thus may be numerically unstable. The inverse of $\mathbf{A} = \mathbf{X}^T \mathbf{W} \mathbf{X}$ is available directly in the `"lm"` object, but the second term on the right-hand side of the Woodbury identity requires a matrix inversion with each fold deleted. (In contrast, the inverse of each $\mathbf{C} = \mathbf{W}_\mathbf{j}$ is straightforward because $\mathbf{W}$ is diagonal.)

The Woodbury identity also requires that the model matrix be of full rank. We impose that restriction in our code by removing redundant regressors from the model matrix for all of the cases, but that doesn't preclude rank deficiency from surfacing when a fold is removed. Rank deficiency of $\mathbf{X}$ doesn't disqualify cross-validation because all we need are fitted values under the estimated model.

`glm()` computes the maximum-likelihood estimates for a generalized linear model by iterated weighted least squares [see, e.g., @FoxWeisberg:2019, Sec. 6.12]. The last iteration is therefore just a WLS fit of the "working response" on the model matrix using "working weights." Both the working weights and the working response at convergence are available from the information in the object returned by `glm()`.

We then treat re-estimation of the model with a case or cases deleted as a WLS problem, using the hatvalues or the Woodbury matrix identity. The resulting fitted values for the deleted fold aren't exact---that is, except for the Gaussian family, the result isn't identical to what we would obtain by literally refitting the model---but in our (limited) experience, the approximation is very good, especially for LOO CV, which is when we would be most tempted to use it. Nevertheless, because these results are approximate, the default for the `"glm"` `cv()` method is to perform the exact computation, which entails refitting the model with each fold omitted.

## Computation of the bias-corrected CV criterion and confidence intervals

Let $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ represent a cross-validation cost criterion, such as mean-squared error, computed for all of the $n$ values of the response $\mathbf{y}$ based on fitted values $\widehat{\mathbf{y}}$ from the model fit to all of the data. We require that $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ is the mean of casewise components, that is, $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n\mathrm{cv}(y_i, \widehat{y}_i)$.[^contrast-function] For example, $\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$.

[^contrast-function]: @ArlotCelisse:2010 term the casewise loss, $\mathrm{cv}(y_i, \widehat{y}_i)$, the "contrast function."

We divide the $n$ cases into $k$ folds of approximately $n_j \approx n/k$ cases each, where $n = \sum n_j$. As above, let $\mathbf{j}$ denote the indices of the cases in the $j$th fold.

Now define $\mathrm{CV}_j = \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}^{(j)})$. The superscript $(j)$ on $\widehat{\mathbf{y}}^{(j)}$ represents fitted values computed  for all of the cases from the model with fold $j$ omitted. Let $\widehat{\mathbf{y}}^{(-i)}$ represent the vector of fitted values for all $n$ cases where the fitted value for the $i$th case is computed from the model fit with the fold including the $i$th case omitted (i.e., fold $j$ for which $i \in \mathbf{j}$).

Then the cross-validation criterion is just $\mathrm{CV} = \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}^{(-i)})$.
Following @DavisonHinkley:1997[pp. 293--295], the bias-adjusted cross-validation criterion is
$$
\mathrm{CV}_{\mathrm{adj}} = \mathrm{CV} + \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) - \frac{1}{n} \sum_{j=1}^{k} n_j \mathrm{CV}_j
$$

We compute the standard error of CV as 
$$
\mathrm{SE}(\mathrm{CV}) = \frac{1}{\sqrt n} \sqrt{ \frac{\sum_{i=1}^n \left[ \mathrm{cv}(y_i, \widehat{y}_i^{(-i)} ) - \mathrm{CV} \right]^2 }{n - 1} }
$$
that is, as the standard deviation of the casewise components of CV divided by the square-root of the number of cases.

We then use $\mathrm{SE}(\mathrm{CV})$ to construct a $100 \times (1 - \alpha)$% confidence interval around the *adjusted* CV estimate of error:
$$
\left[ \mathrm{CV}_{\mathrm{adj}} - z_{1 - \alpha/2}\mathrm{SE}(\mathrm{CV}), \mathrm{CV}_{\mathrm{adj}} + z_{1 - \alpha/2}\mathrm{SE}(\mathrm{CV})  \right]
$$
where $z_{1 - \alpha/2}$ is the $1 - \alpha/2$ quantile of the standard-normal distribution (e.g, $z \approx 1.96$ for a 95% confidence interval, for which $1 - \alpha/2 = .975$).

@BatesHastieTibshirani:2023 show that the coverage of this confidence interval is poor for small samples, and they suggest a much more computationally intensive procedure, called *nested cross-validation*, to compute better estimates of error and confidence intervals with better coverage for small samples. We may implement Bates et al.'s approach in a later release of the **cv** package. At present we use the confidence interval above for sufficiently large $n$, which, based on Bates et al.'s results, we take by default to be $n \ge 400$.

# Why the complement of AUC isn't a casewise CV criterion

Consider calculating AUC for folds in which a validation set contains $n_v$ observations. To calculate AUC in the validation set, we need the vector of prediction criteria, $\widehat{\mathbf{y}}_{v_{(n_v \times 1)}} = (\widehat{y}_1, ..., \widehat{y}_{n_v})^T$, and the vector of observed responses in the validation set, $\mathbf{y}_{v_{(n_v \times 1)}} = (y_1, \ldots, y_{n_v})^T$ with $y_i \in \{0,1\}, \; i = 1, \ldots, n_v$.

To construct the ROC curve, only the ordering of the values in 
$\mathbf{\widehat{y}}_v$ is relevant. Thus, assuming that there are no ties, and reordering observations if necessary, we can set 
$\mathbf{\widehat{y}}_v = (1, 2, \ldots, n_v)^T$.

If the AUC can be expressed as the casewise mean or sum of a function $\mathrm{cv}(\widehat{y}_i,y_i)$, where $\mathrm{cv}: \{1,2,...,n_v\}\times\{0,1\} \rightarrow [0,1]$, then
\begin{equation}
\label{eq:cw}
\tag{1}
\sum_{i=1}^{n_v} \mathrm{cv}(\widehat{y}_i,y_i) = \mathrm{AUC}(\mathbf{\widehat{y}}_v,\mathbf{y}_v)
\end{equation}
must hold for all $2^{n_v}$ possible values of $\mathbf{y}_v = (y_1,...,y_{n_v})^T$. If all $y\mathrm{s}$ have the same value, either 1 or 0, then the definition of AUC is ambiguous. AUC could be considered undefined, or it could be set to 0 if all $y$s are 0 and to 1 if all $y$s are 1. If AUC is considered to be undefined in these cases, we have $2^{n_v} - 2$ admissible values for $\mathbf{y}_v$.

Thus, equation (\ref{eq:cw}) produces either $2^{n_v}$ or $2^{n_v}-2$ constraints. Although there are only $2n_v$ possible values for the $\mathrm{cv(\cdot)}$ function, equation (\ref{eq:cw}) could, nevertheless, have consistent solutions.  We therefore need to determine whether there is a value of $n_v$ for which (\ref{eq:cw}) has no consistent solution for all admissible values of $\mathbf{y}_v$. In that eventuality, we will have shown that AUC cannot, in general, be expressed through a casewise sum.

If $n_v=3$, we show below that (\ref{eq:cw}) has no consistent solution if we include all possibilities for $\mathbf{y}_v$, but does if we exclude cases where all $y$s have the same value. If $n_v=4$, we show that there are no consistent solutions in either case.

The following R function computes AUC from $\mathbf{\widehat{y}}_v$ and $\mathbf{y}_v$, accommodating the cases where $\mathbf{y}_v$ is all 0s or all 1s:

```{r}
AUC <- function(y, yhat = seq_along(y)) {
  s <- sum(y)
  if (s == 0) return(0)
  if (s == length(y)) return(1)
  Metrics::auc(y, yhat)
}
```

We then define a function to generate all possible $\mathbf{y}_v$s of length $n_v$ as rows of the matrix $\mathbf{Y}_{(2^{n_v} \times n_v)}$:

```{r}
Ymat <- function(n_v, exclude_identical = FALSE) {
  stopifnot(n_v > 0 && round(n_v) == n_v)    # n_v must be a positive integer
  ret <- sapply(0:(2^n_v - 1),
                function(x) as.integer(intToBits(x)) )[1:n_v, ]
  ret <- if (is.matrix(ret)) t(ret) else matrix(ret)
  colnames(ret) <- paste0("y", 1:ncol(ret))
  if (exclude_identical) ret[-c(1, nrow(ret)), ] else ret
}
```

For $n_v=3$,

```{r}
Ymat(3)
```

If we exclude $\mathbf{y}_v$s with identical values, then

```{r}
Ymat(3, exclude_identical = TRUE)
```

Here is $\mathbf{Y}$ with corresponding values of AUC:

```{r}
cbind(Ymat(3), AUC = apply(Ymat(3), 1, AUC))
```

The values of $\mathrm{cv}(\widehat{y}_i, y_i)$ that express AUC as a sum of casewise values are solutions of equation (\ref{eq:cw}), which can be written as solutions of the following system of $2^{n_v}$ linear simultaneous equations in $2n_v$ unknowns:
\begin{equation}
\label{eq:lin}
\tag{2}
(\mathbf{U} -\mathbf{Y}) \mathbf{c}_0 + \mathbf{Y} \mathbf{c}_1
=
[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]
\begin{bmatrix}
\mathbf{c}_0 \\ \mathbf{c}_1
\end{bmatrix}
= \mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})
\end{equation}
where $\mathbf{U}_{(2^{n_v} \times n_v)}$ is a matrix of 1s conformable with $\mathbf{Y}$; $\mathbf{c}_0 = [\mathrm{cv}(1,0), c(2,0), ..., \mathrm{cv}(n_v,0)]^T$; $\mathbf{c}_1 = [\mathrm{cv}(1,1), c(2,1), ..., \mathrm{cv}(n_v,1)]^T$; $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]_{(2^{n_v} \times 2n_v)}$ and
$\begin{bmatrix}\begin{aligned}
\mathbf{c}_0 \\ \mathbf{c}_1
\end{aligned}
\end{bmatrix}_{(2n_v \times 1)}$ 
are partitioned matrices; and $\mathbf{\widehat{Y}}_{(2^{n_v} \times n_v)}$ is a matrix each of whose rows consists of the integers 1 to $n_v$.

We can test whether equation (\ref{eq:lin}) has a solution for any given $n_v$ by trying to solve it as a least-squares problem, considering whether the residuals of the associated linear model are all 0, using the "design matrix" $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]$ to predict the "outcome"
$\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})_{(2^{n_v} \times 1)}$:

```{r}
resids <- function(n_v, exclude_identical = FALSE, 
                   tol = sqrt(.Machine$double.eps)) {
  Y <- Ymat(n_v, exclude_identical = exclude_identical)
  AUC <- apply(Y, 1, AUC)
  X <- cbind(1-Y, Y)
  opts <- options(warn = -1)
  on.exit(options(opts))
  fit <- lsfit(X, AUC, intercept = FALSE)
  ret <- max(abs(residuals(fit)))
  if(ret < tol){
    ret <- 0
    solution <- coef(fit)
    names(solution) <- paste0("c(", c(1:n_v, 1:n_v), ",", 
                              rep(0:1, each = n_v), ")")
    attr(ret, "solution") <- zapsmall(solution)
  }
  ret
}
```

The case $n_v=3$, excluding identical $y$s, has a solution:

```{r}
resids(3, exclude_identical = TRUE)
```

But, if identical $y$s are included, the equation is not consistent:

```{r}
resids(3, exclude_identical = FALSE)
```

For $n_v=4$, there are no solutions in either case:

```{r}
resids(4, exclude_identical = TRUE)
resids(4, exclude_identical = FALSE)
```

Consequently, the widely employed AUC measure of fit for binary regression cannot in general be used for a casewise cross-validation criterion.

<!-- # Introduction -->

<!-- This template demonstrates some of the basic LaTeX that you need to know to create a JSS article. -->

<!-- ## Code formatting -->

<!-- In general, don't use Markdown, but use the more precise LaTeX commands instead: -->

<!-- * \proglang{Java} -->
<!-- * \pkg{plyr} -->

<!-- One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax). -->

<!-- If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example. -->

<!-- # \proglang{R} code {short-title="R code" #r-code} -->

<!-- Can be inserted in regular R markdown blocks. -->

<!-- ```{r} -->
<!-- x <- 1:10 -->
<!-- x -->
<!-- ``` -->

<!-- ## Features specific to \pkg{rticles} {short-title="Features specific to rticles"} -->

<!-- * Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409). -->
<!-- * Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format. -->

### 2024-01-01 scratch/notes_john/JSS-article/JSS-article.tex
add initial "draft" of JSS article to scratch

\documentclass[
]{jss}

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

\usepackage[utf8]{inputenc}

\author{
John Fox~\orcidlink{0000-0002-1196-8012}\\McMaster
University \And Georges Monette~\orcidlink{0000-0000-0000-0000}\\York
University
}
\title{\pkg{cv}: An R Package for Cross-Validation of Regression Models}

\Plainauthor{John Fox, Georges Monette}
\Plaintitle{cv: An R Package for Cross-Validation of Regression Models}
\Shorttitle{\pkg{cv}: Cross-Validation}

\Abstract{
The abstract of the article.
}

\Keywords{cross-validation, regression analysis, model
selection, \proglang{R}}
\Plainkeywords{cross-validation, regression analysis, model
selection, R}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    John Fox\\
    McMaster University\\
    Hamilton, Ontario, Canada\\
  E-mail: \email{jfox@mcmaster.ca}\\
  URL: \url{https://www.john-fox.ca/}\\~\\
      Georges Monette\\
    York University\\
    Toronto, Ontario, Canada\\

  }

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{amsmath}

\begin{document}

\hypertarget{cross-validation}{%
\section{Cross-validation}\label{cross-validation}}

Cross-validation (CV) is an essentially simple and intuitively
reasonable approach to estimating the predictive accuracy of regression
models. CV is developed in many standard sources on regression modeling
and ``machine learning''---we particularly recommend \citet[Secs. 5.1,
5.3]{JamesEtAl:2021}---and so we will describe the method only briefly
here before taking up computational issues and some examples. See
\citet{ArlotCelisse:2010} for a wide-ranging, if technical, survey of
cross-validation and related methods that emphasizes the statistical
properties of CV.

Validating research by replication on independently collected data is a
common scientific norm. Emulating this process in a single study by
data-division is less common: The data are randomly divided into two,
possibly equal-size, parts; the first part is used to develop and fit a
statistical model; and then the second part is used to assess the
adequacy of the model fit to the first part of the data. Data-division,
however, suffers from two problems: (1) Dividing the data decreases the
sample size and thus increases sampling error; and (2), even more
disconcertingly, particularly in smaller samples, the results can vary
substantially based on the random division of the data: See \citet[Sec.
5.3]{Harrell:2015} for this and other remarks about data-division and
cross-validation.

Cross-validation speaks to both of these issues. In CV, the data are
randomly divided as equally as possible into several, say \(k\), parts,
called ``folds.'' The statistical model is fit \(k\) times, leaving each
fold out in turn. Each fitted model is then used to predict the response
variable for the cases in the omitted fold. A CV criterion or ``cost''
measure, such as the mean-squared error (``MSE'') of prediction, is then
computed using these predicted values. In the extreme \(k = n\), the
number of cases in the data, thus omitting individual cases and
refitting the model \(n\) times---a procedure termed ``leave-one-out
(LOO) cross-validation.''

Because the \(n\) models are each fit to \(n - 1\) cases, LOO CV
produces a nearly unbiased estimate of prediction error. The \(n\)
regression models are highly statistical dependent, however, based as
they are on nearly the same data, and so the resulting estimate of
prediction error has relatively large variance. In contrast, estimated
prediction error for \(k\)-fold CV with \(k = 5\) or \(10\) (commonly
employed choices) are somewhat biased but have smaller variance. It is
also possible to correct \(k\)-fold CV for bias (see below).

\hypertarget{examples}{%
\section{Examples}\label{examples}}

\hypertarget{polynomial-regression-for-the-auto-data}{%
\subsection{Polynomial regression for the Auto
data}\label{polynomial-regression-for-the-auto-data}}

The data for this example are drawn from the \textbf{ISLR2} package for
R, associated with \citet{JamesEtAl:2021}. The presentation here is
close (though not identical) to that in the original source \citep[
Secs. 5.1, 5.3]{JamesEtAl:2021}, and it demonstrates the use of the
\texttt{cv()} function in the \textbf{cv} package.\footnote{\citet{JamesEtAl:2021}
  use the \texttt{cv.glm()} function in the \textbf{boot} package
  \citep{CantyRipley2022, DavisonHinkley:1997}. Despite its name,
  \texttt{cv.glm()} is an independent function and not a method of a
  \texttt{cv()} generic function.}

The \texttt{Auto} dataset contains information about 392 cars:

\begin{CodeChunk}
\begin{CodeInput}
R> data("Auto", package="ISLR2")
R> head(Auto)
\end{CodeInput}
\begin{CodeOutput}
  mpg cylinders displacement horsepower weight acceleration year origin
1  18         8          307        130   3504         12.0   70      1
2  15         8          350        165   3693         11.5   70      1
3  18         8          318        150   3436         11.0   70      1
4  16         8          304        150   3433         12.0   70      1
5  17         8          302        140   3449         10.5   70      1
6  15         8          429        198   4341         10.0   70      1
                       name
1 chevrolet chevelle malibu
2         buick skylark 320
3        plymouth satellite
4             amc rebel sst
5               ford torino
6          ford galaxie 500
\end{CodeOutput}
\begin{CodeInput}
R> dim(Auto)
\end{CodeInput}
\begin{CodeOutput}
[1] 392   9
\end{CodeOutput}
\end{CodeChunk}

With the exception of \texttt{origin} (which we don't use here), these
variables are largely self-explanatory, except possibly for units of
measurement: for details see \texttt{help("Auto",\ package="ISLR2")}.

We'll focus here on the relationship of \texttt{mpg} (miles per gallon)
to \texttt{horsepower}, as displayed in the following scatterplot:

\begin{CodeChunk}
\begin{CodeInput}
R> plot(mpg ~ horsepower, data=Auto)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/mpg-horsepower-scatterplot-1}

}

\caption[`mpg` vs `horsepower` for the `Auto` data]{`mpg` vs `horsepower` for the `Auto` data}\label{fig:mpg-horsepower-scatterplot}
\end{figure}
\end{CodeChunk}

The relationship between the two variables is monotone, decreasing, and
nonlinear. Following \citet{JamesEtAl:2021}, we'll consider
approximating the relationship by a polynomial regression, with the
degree of the polynomial \(p\) ranging from 1 (a linear regression) to
10.\footnote{Although it serves to illustrate the use of CV, a
  polynomial is probably not the best choice here. Consider, for example
  the scatterplot for log-transformed \texttt{mpg} and
  \texttt{horsepower}, produced by
  \texttt{plot(mpg\ \textasciitilde{}\ horsepower,\ data=Auto,\ log="xy")}
  (execution of which is left to the reader).} Polynomial fits for
\(p = 1\) to \(5\) are shown in the following figure:

\begin{CodeChunk}
\begin{CodeInput}
R> plot(mpg ~ horsepower, data=Auto)
R> horsepower <- with(Auto, 
+                    seq(min(horsepower), max(horsepower), 
+                        length=1000))
R> for (p in 1:5){
+   m <- lm(mpg ~ poly(horsepower,p), data=Auto)
+   mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
+   lines(horsepower, mpg, col=p + 1, lty=p, lwd=2)
+ }
R> legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=2,
+        title="Degree", inset=0.02)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}

}

\caption[`mpg` vs `horsepower` for the `Auto` data]{`mpg` vs `horsepower` for the `Auto` data}\label{fig:mpg-horsepower-scatterplot-polynomials}
\end{figure}
\end{CodeChunk}

The linear fit is clearly inappropriate; the fits for \(p = 2\)
(quadratic) through \(4\) are very similar; and the fit for \(p = 5\)
may over-fit the data by chasing one or two relatively high \texttt{mpg}
values at the right (but see the CV results reported below).

The following graph shows two measures of estimated (squared) error as a
function of polynomial-regression degree: The mean-squared error
(``MSE''), defined as
\(\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\), and
the usual residual variance, defined as
\(\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2\).
The former necessarily declines with \(p\) (or, more strictly, can't
increase with \(p\)), while the latter gets slightly larger for the
largest values of \(p\), with the ``best'' value, by a small margin, for
\(p = 7\).

\begin{CodeChunk}
\begin{CodeInput}
R> library("cv") # for mse() and other functions
\end{CodeInput}
\begin{CodeOutput}
Loading required package: doParallel
\end{CodeOutput}
\begin{CodeOutput}
Loading required package: foreach
\end{CodeOutput}
\begin{CodeOutput}
Loading required package: iterators
\end{CodeOutput}
\begin{CodeOutput}
Loading required package: parallel
\end{CodeOutput}
\begin{CodeInput}
R> var <- mse <- numeric(10)
R> for (p in 1:10){
+   m <- lm(mpg ~ poly(horsepower, p), data=Auto)
+   mse[p] <- mse(Auto$mpg, fitted(m))
+   var[p] <- summary(m)$sigma^2
+ }
R> 
R> plot(c(1, 10), range(mse, var), type="n",
+      xlab="Degree of polynomial, p",
+      ylab="Estimated Squared Error")
R> lines(1:10, mse, lwd=2, lty=1, col=2, pch=16, type="b")
R> lines(1:10, var, lwd=2, lty=2, col=3, pch=17, type="b")
R> legend("topright", inset=0.02,
+        legend=c(expression(hat(sigma)^2), "MSE"),
+        lwd=2, lty=2:1, col=3:2, pch=17:16)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/mpg-horsepower-MSE-se-1}

}

\caption[Estimated squared error as a function of polynomial degree, $p$]{Estimated squared error as a function of polynomial degree, $p$}\label{fig:mpg-horsepower-MSE-se}
\end{figure}
\end{CodeChunk}

The code for this graph uses the \texttt{mse()} function from the
\textbf{cv} package to compute the MSE for each fit.

\hypertarget{using-cv}{%
\subsubsection{Using cv()}\label{using-cv}}

The generic \texttt{cv()} function has an \texttt{"lm"} method, which by
default performs \(k = 10\)-fold CV:

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
R> summary(m.auto)
\end{CodeInput}
\begin{CodeOutput}

Call:
lm(formula = mpg ~ poly(horsepower, 2), data = Auto)

Residuals:
     Min       1Q   Median       3Q      Max 
-14.7135  -2.5943  -0.0859   2.2868  15.8961

Coefficients:
                      Estimate Std. Error t value Pr(>|t|)
(Intercept)            23.4459     0.2209  106.13   <2e-16
poly(horsepower, 2)1 -120.1377     4.3739  -27.47   <2e-16
poly(horsepower, 2)2   44.0895     4.3739   10.08   <2e-16

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 534539
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.22944
bias-adjusted cross-validation criterion = 19.21651
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

The \texttt{"lm"} method by default uses \texttt{mse()} as the CV
criterion and the Woodbury matrix identity to update the regression with
each fold deleted without having literally to refit the model.
Computational details are discussed in the final section of this
vignette. The function reports the CV estimate of MSE, a biased-adjusted
estimate of the MSE (the bias adjustment is explained in the final
section), and the MSE is also computed for the original, full-sample
regression. Because the division of the data into 10 folds is random,
\texttt{cv()} explicitly (randomly) generates and saves a seed for R's
pseudo-random number generator, to make the results replicable. The user
can also specify the seed directly via the \texttt{seed} argument to
\texttt{cv()}.

To perform LOO CV, we can set the \texttt{k} argument to \texttt{cv()}
to the number of cases in the data, here \texttt{k=392}, or, more
conveniently, to \texttt{k="loo"} or \texttt{k="n"}:

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.auto, k="loo")
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: hatvalues
criterion: mse
cross-validation criterion = 19.24821
\end{CodeOutput}
\end{CodeChunk}

For LOO CV of a linear model, \texttt{cv()} by default uses the
hatvalues from the model fit to the full data for the LOO updates, and
reports only the CV estimate of MSE. Alternative methods are to use the
Woodbury matrix identity or the ``naive'' approach of literally
refitting the model with each case omitted. All three methods produce
exact results for a linear model (within the precision of floating-point
computations):

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.auto, k="loo", method="naive")
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: naive
criterion: mse
cross-validation criterion = 19.24821
bias-adjusted cross-validation criterion = 19.24787
full-sample criterion = 18.98477 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, k="loo", method="Woodbury")
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.24821
bias-adjusted cross-validation criterion = 19.24787
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

The \texttt{"naive"} and \texttt{"Woodbury"} methods also return the
bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an
issue for LOO CV.

This is a small regression problem and all three computational
approaches are essentially instantaneous, but it is still of interest to
investigate their relative speed. In this comparison, we include the
\texttt{cv.glm()} function from the \textbf{boot} package, which takes
the naive approach, and for which we have to fit the linear model as an
equivalent Gaussian GLM. We use the \texttt{microbenchmark()} function
from the package of the same name for the timings \citep{Mersmann:2023}:

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta
\end{CodeInput}
\begin{CodeOutput}
[1] 19.24821 19.24787
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k="loo"),
+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
+   naive = cv(m.auto, k="loo", method="naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: microseconds
      expr        min        lq       mean     median         uq        max
 hatvalues    977.973    993.43   1143.433   1173.113   1202.981   1296.256
  Woodbury   9932.004  10178.21  11461.993  10363.201  10865.943  17762.266
     naive 215384.152 217809.02 224546.766 219116.505 221670.231 275518.278
    cv.glm 384357.616 388159.59 409612.546 390652.100 441741.913 443226.605
 neval cld
    10 a  
    10 a  
    10  b 
    10   c
\end{CodeOutput}
\end{CodeChunk}

On our computer, using the hatvalues is about an order of magnitude
faster than employing Woodbury matrix updates, and more than two orders
of magnitude faster than refitting the model.\footnote{Out of
  impatience, we asked \texttt{microbenchmark()} to execute each command
  only 10 times rather than the default 100. With the exception of the
  last columns, the output is self-explanatory. The last column shows
  which methods have average timings that are statistically
  distinguishable. Because of the small number of repetitions (i.e.,
  10), the \texttt{"hatvalues"} and \texttt{"Woodbury"} methods aren't
  distinguishable, but the difference between these methods persists
  when we perform more repetitions---we invite the reader to redo this
  computation with the default \texttt{times=100} repetitions.}

\hypertarget{comparing-competing-models}{%
\subsubsection{Comparing competing
models}\label{comparing-competing-models}}

The \texttt{cv()} function also has a method that can be applied to a
list of regression models for the same data, composed using the
\texttt{models()} function. For \(k\)-fold CV, the same folds are used
for the competing models, which reduces random error in their
comparison. This result can also be obtained by specifying a common seed
for R's random-number generator while applying \texttt{cv()} separately
to each model, but employing a list of models is more convenient for
both \(k\)-fold and LOO CV (where there is no random component to the
composition of the \(n\) folds).

We illustrate with the polynomial regression models of varying degree
for the \texttt{Auto} data (discussed previously), beginning by fitting
and saving the 10 models:

\begin{CodeChunk}
\begin{CodeInput}
R> for (p in 1:10){
+   assign(paste0("m.", p),
+          lm(mpg ~ poly(horsepower, p), data=Auto))
+ }
R> objects(pattern="m\\.[0-9]")
\end{CodeInput}
\begin{CodeOutput}
 [1] "m.1"  "m.10" "m.2"  "m.3"  "m.4"  "m.5"  "m.6"  "m.7"  "m.8"  "m.9" 
\end{CodeOutput}
\begin{CodeInput}
R> summary(m.2) # for example, the quadratic fit
\end{CodeInput}
\begin{CodeOutput}

Call:
lm(formula = mpg ~ poly(horsepower, p), data = Auto)

Residuals:
     Min       1Q   Median       3Q      Max 
-14.7135  -2.5943  -0.0859   2.2868  15.8961

Coefficients:
                      Estimate Std. Error t value Pr(>|t|)
(Intercept)            23.4459     0.2209  106.13   <2e-16
poly(horsepower, p)1 -120.1377     4.3739  -27.47   <2e-16
poly(horsepower, p)2   44.0895     4.3739   10.08   <2e-16

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\end{CodeChunk}

We then apply \texttt{cv()} to the list of 10 models (the \texttt{data}
argument is required):

\begin{CodeChunk}
\begin{CodeInput}
R> # 10-fold CV
R> cv.auto.10 <- cv(models(m.1, m.2, m.3, m.4, m.5,
+                      m.6, m.7, m.8, m.9, m.10),
+               data=Auto, seed=2120)
R> cv.auto.10[1:2] # for the linear and quadratic models
\end{CodeInput}
\begin{CodeOutput}

Model model.1:
10-Fold Cross Validation
method: Woodbury
cross-validation criterion = 24.24642
bias-adjusted cross-validation criterion = 24.23039
full-sample criterion = 23.94366

Model model.2:
10-Fold Cross Validation
method: Woodbury
cross-validation criterion = 19.34601
bias-adjusted cross-validation criterion = 19.32699
full-sample criterion = 18.98477 
\end{CodeOutput}
\begin{CodeInput}
R> # LOO CV
R> cv.auto.loo <- cv(models(m.1, m.2, m.3, m.4, m.5,
+                         m.6, m.7, m.8, m.9, m.10),
+                  data=Auto, k="loo")
R> cv.auto.loo[1:2] # linear and quadratic models
\end{CodeInput}
\begin{CodeOutput}

Model model.1:
n-Fold Cross Validation
method: hatvalues
cross-validation criterion = 24.23151
Model model.2:
n-Fold Cross Validation
method: hatvalues
cross-validation criterion = 19.24821
\end{CodeOutput}
\end{CodeChunk}

Because we didn't supply names for the models in the calls to the
\texttt{models()} function, the names \texttt{model.1},
\texttt{model.2}, etc., are generated by the function.

Finally, we extract and graph the adjusted MSEs for \(10\)-fold CV and
the MSEs for LOO CV:

\begin{CodeChunk}
\begin{CodeInput}
R> cv.mse.10 <- sapply(cv.auto.10, function(x) x[["adj CV crit"]])
R> cv.mse.loo <- sapply(cv.auto.loo, function(x) x[["CV crit"]])
R> plot(c(1, 10), range(cv.mse.10, cv.mse.loo), type="n",
+      xlab="Degree of polynomial, p",
+      ylab="Cross-Validated MSE")
R> lines(1:10, cv.mse.10, lwd=2, lty=1, col=2, pch=16, type="b")
R> lines(1:10, cv.mse.loo, lwd=2, lty=2, col=3, pch=17, type="b")
R> legend("topright", inset=0.02,
+        legend=c("10-Fold CV", "LOO CV"),
+        lwd=2, lty=2:1, col=3:2, pch=17:16)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/polynomial-regression-CV-graph-1}

}

\caption[Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$]{Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$}\label{fig:polynomial-regression-CV-graph}
\end{figure}
\end{CodeChunk}

Alternatively, we can use the \texttt{plot()} method for
\texttt{"cvModList"} objects to compare the models, though with separate
graphs for 10-fold and LOO CV:

\begin{CodeChunk}
\begin{CodeInput}
R> plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",
+      axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
R> plot(cv.auto.loo, main="Polynomial Regressions, LOO CV",
+      axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=0.45\linewidth]{JSS-article_files/figure-latex/polynomial-regression-CV-graph-2-1} \includegraphics[width=0.45\linewidth]{JSS-article_files/figure-latex/polynomial-regression-CV-graph-2-2}

}

\caption[Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$]{Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$}\label{fig:polynomial-regression-CV-graph-2}
\end{figure}
\end{CodeChunk}

In this example, 10-fold and LOO CV produce generally similar results,
and also results that are similar to those produced by the estimated
error variance \(\widehat{\sigma}^2\) for each model, reported above
(except for the highest-degree polynomials, where the CV results more
clearly suggest over-fitting).

\hypertarget{logistic-regression-for-the-mroz-data}{%
\subsection{Logistic regression for the Mroz
data}\label{logistic-regression-for-the-mroz-data}}

The \texttt{Mroz} data set from the \textbf{carData} package
\citep[associated with][]{FoxWeisberg:2019} has been used by several
authors to illustrate binary logistic regression; see, in particular
\citet{FoxWeisberg:2019}. The data were originally drawn from the U.S.
Panel Study of Income Dynamics and pertain to married women. Here are a
few cases in the data set:

\begin{CodeChunk}
\begin{CodeInput}
R> data("Mroz", package="carData")
R> head(Mroz, 3)
\end{CodeInput}
\begin{CodeOutput}
  lfp k5 k618 age wc hc       lwg   inc
1 yes  1    0  32 no no 1.2101647 10.91
2 yes  0    2  30 no no 0.3285041 19.50
3 yes  1    3  35 no no 1.5141279 12.04
\end{CodeOutput}
\begin{CodeInput}
R> tail(Mroz, 3)
\end{CodeInput}
\begin{CodeOutput}
    lfp k5 k618 age wc hc       lwg    inc
751  no  0    0  43 no no 0.8881401  9.952
752  no  0    0  60 no no 1.2249736 24.984
753  no  0    3  39 no no 0.8532125 28.363
\end{CodeOutput}
\end{CodeChunk}

The response variable in the logistic regression is \texttt{lfp},
labor-force participation, a factor coded \texttt{"yes"} or
\texttt{"no"}. The remaining variables are predictors:

\begin{itemize}
\tightlist
\item
  \texttt{k5}, number of children 5 years old of younger in the woman's
  household;
\item
  \texttt{k618}, number of children between 6 and 18 years old;
\item
  \texttt{age}, in years;
\item
  \texttt{wc}, wife's college attendance, \texttt{"yes"} or
  \texttt{"no"};
\item
  \texttt{hc}, husband's college attendance;
\item
  \texttt{lwg}, the woman's log wage rate if she is employed, or her
  \emph{imputed} wage rate, if she is not \citep[a variable that][ show
  is problematically defined]{FoxWeisberg:2019}; and
\item
  \texttt{inc}, family income, in \$1000s, exclusive of wife's income.
\end{itemize}

We use the \texttt{glm()} function to fit a binary logistic regression
to the \texttt{Mroz} data:

\begin{CodeChunk}
\begin{CodeInput}
R> m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
R> summary(m.mroz)
\end{CodeInput}
\begin{CodeOutput}

Call:
glm(formula = lfp ~ ., family = binomial, data = Mroz)

Coefficients:
             Estimate Std. Error z value Pr(>|z|)
(Intercept)  3.182140   0.644375   4.938 7.88e-07
k5          -1.462913   0.197001  -7.426 1.12e-13
k618        -0.064571   0.068001  -0.950 0.342337
age         -0.062871   0.012783  -4.918 8.73e-07
wcyes        0.807274   0.229980   3.510 0.000448
hcyes        0.111734   0.206040   0.542 0.587618
lwg          0.604693   0.150818   4.009 6.09e-05
inc         -0.034446   0.008208  -4.196 2.71e-05

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 1029.75  on 752  degrees of freedom
Residual deviance:  905.27  on 745  degrees of freedom
AIC: 921.27

Number of Fisher Scoring iterations: 4
\end{CodeOutput}
\begin{CodeInput}
R> BayesRule(ifelse(Mroz$lfp == "yes", 1, 0), 
+           fitted(m.mroz, type="response"))
\end{CodeInput}
\begin{CodeOutput}
[1] 0.3067729
attr(,"casewise loss")
[1] "y != round(yhat)"
\end{CodeOutput}
\end{CodeChunk}

In addition to the usually summary output for a GLM, we show the result
of applying the \texttt{BayesRule()} function from the \textbf{cv}
package to predictions derived from the fitted model. Bayes rule, which
predicts a ``success'' in a binary regression model when the fitted
probability of success {[}i.e., \(\phi = \Pr(y = 1)\){]} is
\(\widehat{\phi} \ge .5\) and a ``failure'' if
\(\widehat{\phi} < .5\).\footnote{\texttt{BayesRule()} does some error
  checking; \texttt{BayesRule2()} is similar, but omits the error
  checking, and so can be faster for large problems.} The first argument
to \texttt{BayesRule()} is the binary \{0, 1\} response, and the second
argument is the predicted probability of success. \texttt{BayesRule()}
returns the proportion of predictions that are \emph{in error}, as
appropriate for a ``cost'' function.

The value returned by \texttt{BayesRule()} is associated with an
``attribute'' named \texttt{"casewise\ loss"} and set to
\texttt{"y\ !=\ round(yhat)"}, signifying that the Bayes rule CV
criterion is computed as the mean of casewise values, here 0 if the
prediction for a case matches the observed value and 1 if it does not
(signifying a prediction error). The \texttt{mse()} function for numeric
responses is also calculated as a casewise average. Some other criteria,
such as the median absolute error, computed by the \texttt{medAbsErr()}
function in the \textbf{cv} package, aren't averages of casewise
components. The distinction is important because, to our knowledge, the
statistical theory of cross-validation, for example, in
\citet{DavisonHinkley:1997}, \citet{BatesHastieTibshirani:2023}, and
\citet{ArlotCelisse:2010}, is developed for CV criteria like MSE that
are means of casewise components. As a consequence, we limit computation
of bias adjustment and confidence intervals (see below) to criteria that
are casewise averages.

In this example, the fitted logistic regression incorrectly predicts
31\% of the responses; we expect this estimate to be optimistic given
that the model is used to ``predict'' the data to which it is fit.

The \texttt{"glm"} method for \texttt{cv()} is largely similar to the
\texttt{"lm"} method, although the default algorithm, selected
explicitly by \texttt{method="exact"}, refits the model with each fold
removed (and is thus equivalent to \texttt{method="naive"} for
\texttt{"lm"} models). For generalized linear models,
\texttt{method="Woodbury"} or (for LOO CV) \texttt{method="hatvalues"}
provide approximate results (see the last section of the vignette for
details):

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.mroz, criterion=BayesRule, seed=248)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 248
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: exact
criterion: BayesRule
cross-validation criterion = 0.3240372
bias-adjusted cross-validation criterion = 0.3195223
95% CI for bias-adjusted CV criterion = (0.2860721, 0.3529724)
full-sample criterion = 0.3067729 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.mroz, criterion=BayesRule, seed=248, method="Woodbury")
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 248
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3240372
bias-adjusted cross-validation criterion = 0.3192577
95% CI for bias-adjusted CV criterion = (0.2858076, 0.3527079)
full-sample criterion = 0.3067729 
\end{CodeOutput}
\end{CodeChunk}

To ensure that the two methods use the same 10 folds, we specify the
seed for R's random-number generator explicitly; here, and as is common
in our experience, the \texttt{"exact"} and \texttt{"Woodbury"}
algorithms produce nearly identical results. The CV estimates of
prediction error are slightly higher than the estimate based on all of
the cases.

The printed output includes a 95\% confidence interval for the
bias-adjusted Bayes rule CV criterion.
\citet{BatesHastieTibshirani:2023} show that these confidence intervals
are unreliable for models fit to small samples, and by default
\texttt{cv()} computes them only when the sample size is 400 or larger
and when the CV criterion employed is an average of casewise components,
as is the case for Bayes rule. See the final section of the vignette for
details of the computation of confidence intervals for bias-adjusted CV
criteria.

Here are results of applying LOO CV to the Mroz model, using both the
exact and the approximate methods:

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.mroz, k="loo", criterion=BayesRule)
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: exact
criterion: BayesRule
cross-validation criterion = 0.3200531
bias-adjusted cross-validation criterion = 0.3183001
95% CI for bias-adjusted CV criterion = (0.2849584, 0.3516418)
full-sample criterion = 0.3067729 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury")
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3200531
bias-adjusted cross-validation criterion = 0.3183001
95% CI for bias-adjusted CV criterion = (0.2849584, 0.3516418)
full-sample criterion = 0.3067729 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues")
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: hatvalues
criterion: BayesRule
cross-validation criterion = 0.3200531
\end{CodeOutput}
\end{CodeChunk}

To the number of decimal digits shown, the three methods produce
identical results for this example.

As for linear models, we report some timings for the various
\texttt{cv()} methods of computation in LOO CV as well as for the
\texttt{cv.glm()} function from the \textbf{boot} package (which,
recall, refits the model with each case removed, and thus is comparable
to \texttt{cv()} with \texttt{method="exact"}):

\begin{CodeChunk}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues=cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues"),
+   Woodbury=cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury"),
+   exact=cv(m.mroz, k="loo", criterion=BayesRule),
+   cv.glm=boot::cv.glm(Mroz, m.mroz,
+                cost=BayesRule),
+   times=10)
\end{CodeInput}
\begin{CodeOutput}
Unit: milliseconds
      expr         min          lq        mean      median          uq
 hatvalues    1.316141    1.326842    1.654063    1.407571    1.442462
  Woodbury   37.127427   38.641639   39.368192   39.356904   40.562202
     exact 1757.625925 1827.861385 1905.072191 1911.421128 1951.049944
    cv.glm 2062.693354 2122.649130 2149.761134 2140.079440 2184.788525
         max neval cld
    4.022428    10 a  
   41.060188    10 a  
 2128.307540    10  b 
 2299.544573    10   c
\end{CodeOutput}
\end{CodeChunk}

There is a substantial time penalty associated with exact computations.

\hypertarget{cross-validating-mixed-effects-models}{%
\section{Cross-validating mixed-effects
models}\label{cross-validating-mixed-effects-models}}

The fundamental analogy for cross-validation is to the collection of new
data. That is, predicting the response in each fold from the model fit
to data in the other folds is like using the model fit to all of the
data to predict the response for new cases from the values of the
predictors for those new cases. As we explained, the application of this
idea to independently sampled cases is straightforward---simply
partition the data into random folds of equal size and leave each fold
out in turn, or, in the case of LOO CV, simply omit each case in turn.

In contrast, mixed-effects models are fit to \emph{dependent} data, in
which cases as clustered, such as hierarchical data, where the clusters
comprise higher-level units (e.g., students clustered in schools), or
longitudinal data, where the clusters are individuals and the cases
repeated observations on the individuals over time.\footnote{There are,
  however, more complex situations that give rise to so-called
  \emph{crossed} (rather than \emph{nested}) random effects. For
  example, consider students within classes within schools. In primary
  schools, students typically are in a single class, and so classes are
  nested within schools. In secondary schools, however, students
  typically take several classes and students who are together in a
  particular class may not be together in other classes; consequently,
  random effects based on classes within schools are crossed. The
  \texttt{lmer()} function in the \textbf{lme4} package is capable of
  modeling both nested and crossed random effects, and the \texttt{cv()}
  methods for mixed models in the \textbf{cv} package pertain to both
  nested and crossed random effects. We present an example of the latter
  later in the vignette.}

We can think of two approaches to applying cross-validation to clustered
data:\footnote{We subsequently discovered that \citet[Section
  8]{Vehtari:2023} makes similar points.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Treat CV as analogous to predicting the response for one or more cases
  in a \emph{newly observed cluster}. In this instance, the folds
  comprise one or more whole clusters; we refit the model with all of
  the cases in clusters in the current fold removed; and then we predict
  the response for the cases in clusters in the current fold. These
  predictions are based only on fixed effects because the random effects
  for the omitted clusters are presumably unknown, as they would be for
  data on cases in newly observed clusters.
\item
  Treat CV as analogous to predicting the response for a newly observed
  case in an \emph{existing cluster}. In this instance, the folds
  comprise one or more individual cases, and the predictions can use
  both the fixed and random effects.
\end{enumerate}

\hypertarget{example-the-high-school-and-beyond-data}{%
\subsection{Example: The High-School and Beyond
data}\label{example-the-high-school-and-beyond-data}}

Following their use by \citet{RaudenbushBryk:2002}, data from the 1982
\emph{High School and Beyond} (HSB) survey have become a staple of the
literature on mixed-effects models. The HSB data are used by \citet[Sec.
7.2.2]{FoxWeisberg:2019} to illustrate the application of linear mixed
models to hierarchical data, and we'll closely follow their example
here.

The HSB data are included in the \texttt{MathAchieve} and
\texttt{MathAchSchool} data sets in the \textbf{nlme} package
\citep{PinheiroBates:2000}. \texttt{MathAchieve} includes
individual-level data on 7185 students in 160 high schools, and
\texttt{MathAchSchool} includes school-level data:

\begin{CodeChunk}
\begin{CodeInput}
R> data("MathAchieve", package="nlme")
R> dim(MathAchieve)
\end{CodeInput}
\begin{CodeOutput}
[1] 7185    6
\end{CodeOutput}
\begin{CodeInput}
R> head(MathAchieve, 3)
\end{CodeInput}
\begin{CodeOutput}
Grouped Data: MathAch ~ SES | School
  School Minority    Sex    SES MathAch MEANSES
1   1224       No Female -1.528   5.876  -0.428
2   1224       No Female -0.588  19.708  -0.428
3   1224       No   Male -0.528  20.349  -0.428
\end{CodeOutput}
\begin{CodeInput}
R> tail(MathAchieve, 3)
\end{CodeInput}
\begin{CodeOutput}
Grouped Data: MathAch ~ SES | School
     School Minority    Sex    SES MathAch MEANSES
7183   9586       No Female  1.332  19.641   0.627
7184   9586       No Female -0.008  16.241   0.627
7185   9586       No Female  0.792  22.733   0.627
\end{CodeOutput}
\begin{CodeInput}
R> data("MathAchSchool", package="nlme")
R> dim(MathAchSchool)
\end{CodeInput}
\begin{CodeOutput}
[1] 160   7
\end{CodeOutput}
\begin{CodeInput}
R> head(MathAchSchool, 2)
\end{CodeInput}
\begin{CodeOutput}
     School Size Sector PRACAD DISCLIM HIMINTY MEANSES
1224   1224  842 Public   0.35   1.597       0  -0.428
1288   1288 1855 Public   0.27   0.174       0   0.128
\end{CodeOutput}
\begin{CodeInput}
R> tail(MathAchSchool, 2)
\end{CodeInput}
\begin{CodeOutput}
     School Size   Sector PRACAD DISCLIM HIMINTY MEANSES
9550   9550 1532   Public   0.45   0.791       0   0.059
9586   9586  262 Catholic   1.00  -2.416       0   0.627
\end{CodeOutput}
\end{CodeChunk}

The first few students are in school number 1224 and the last few in
school 9586.

We'll use only the \texttt{School}, \texttt{SES} (students'
socioeconomic status), and \texttt{MathAch} (their score on a
standardized math-achievement test) variables in the
\texttt{MathAchieve} data set, and \texttt{Sector} (\texttt{"Catholic"}
or \texttt{"Public"}) in the \texttt{MathAchSchool} data set.

Some data-management is required before fitting a mixed-effects model to
the HSB data, for which we use the \textbf{dplyr} package
\citep{WickhamEtAl:2023}:

\begin{CodeChunk}
\begin{CodeInput}
R> library("dplyr")
\end{CodeInput}
\begin{CodeOutput}

Attaching package: 'dplyr'
\end{CodeOutput}
\begin{CodeOutput}
The following objects are masked from 'package:stats':

    filter, lag
\end{CodeOutput}
\begin{CodeOutput}
The following objects are masked from 'package:base':

    intersect, setdiff, setequal, union
\end{CodeOutput}
\begin{CodeInput}
R> MathAchieve %>% group_by(School) %>%
+   summarize(mean.ses = mean(SES)) -> Temp
R> Temp <- merge(MathAchSchool, Temp, by="School")
R> HSB <- merge(Temp[, c("School", "Sector", "mean.ses")],
+              MathAchieve[, c("School", "SES", "MathAch")], by="School")
R> names(HSB) <- tolower(names(HSB))
R> 
R> HSB$cses <- with(HSB, ses - mean.ses)
\end{CodeInput}
\end{CodeChunk}

In the process, we created two new school-level variables:
\texttt{meanses}, which is the average SES for students in each school;
and \texttt{cses}, which is school-average SES centered at its mean. For
details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.

Still following Fox and Weisberg, we proceed to use the \texttt{lmer()}
function in the \textbf{lme4} package \citep{BatesEtAl:2015} to fit a
mixed model for math achievement to the HSB data:

\begin{CodeChunk}
\begin{CodeInput}
R> library("lme4")
\end{CodeInput}
\begin{CodeOutput}
Loading required package: Matrix
\end{CodeOutput}
\begin{CodeInput}
R> hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
+                    + (cses | school), data=HSB)
R> summary(hsb.lmer, correlation=FALSE)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed model fit by REML ['lmerMod']
Formula: mathach ~ mean.ses * cses + sector * cses + (cses | school)
   Data: HSB

REML criterion at convergence: 46503.7

Scaled residuals: 
     Min       1Q   Median       3Q      Max 
-3.15926 -0.72319  0.01704  0.75444  2.95822

Random effects:
 Groups   Name        Variance Std.Dev. Corr
 school   (Intercept)  2.380   1.5426       
          cses         0.101   0.3179   0.39
 Residual             36.721   6.0598       
Number of obs: 7185, groups:  school, 160

Fixed effects:
                    Estimate Std. Error t value
(Intercept)          12.1279     0.1993  60.856
mean.ses              5.3329     0.3692  14.446
cses                  2.9450     0.1556  18.928
sectorCatholic        1.2266     0.3063   4.005
mean.ses:cses         1.0393     0.2989   3.477
cses:sectorCatholic  -1.6427     0.2398  -6.851
\end{CodeOutput}
\end{CodeChunk}

We can then cross-validate at the cluster (i.e., school) level,

\begin{CodeChunk}
\begin{CodeInput}
R> cv(hsb.lmer, k=10, clusterVariables="school", seed=5240)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 5240
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation based on 160 {school} clusters
cross-validation criterion = 39.15662
bias-adjusted cross-validation criterion = 39.14844
95% CI for bias-adjusted CV criterion = (38.06554, 40.23135)
full-sample criterion = 39.00599 
\end{CodeOutput}
\end{CodeChunk}

or at the case (i.e., student) level,

\begin{CodeChunk}
\begin{CodeInput}
R> cv(hsb.lmer, seed=1575)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 1575
\end{CodeOutput}
\begin{CodeOutput}
Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv, :
Model failed to converge with max|grad| = 0.00587228 (tol = 0.002, component 1)
\end{CodeOutput}
\begin{CodeOutput}
boundary (singular) fit: see help('isSingular')
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 37.44473
bias-adjusted cross-validation criterion = 37.33801
95% CI for bias-adjusted CV criterion = (36.28761, 38.38841)
full-sample criterion = 36.06767 
\end{CodeOutput}
\end{CodeChunk}

For cluster-level CV, the \texttt{clusterVariables} argument tells
\texttt{cv()} how the clusters are defined. Were there more than one
clustering variable, say classes within schools, these would be provided
as a character vector of variable names:
\texttt{clusterVariables\ =\ c("school",\ "class")}. For cluster-level
CV, the default is \texttt{k\ =\ "loo"}, that is, leave one cluster out
at a time; we instead specify \texttt{k\ =\ 10} folds of clusters, each
fold therefore comprising \(160/10 = 16\) schools.

If the \texttt{clusterVariables} argument is omitted, then case-level CV
is employed, with \texttt{k\ =\ 10} folds as the default, here each with
\(7185/10 \approx 719\) students. Notice that one of the 10 models refit
with a fold removed failed to converge. Convergence problems are common
in mixed-effects modeling. The apparent issue here is that an estimated
variance component is close to or equal to 0, which is at a boundary of
the parameter space. That shouldn't disqualify the fitted model for the
kind of prediction required for cross-validation.

There is also a \texttt{cv()} method for linear mixed models fit by the
\texttt{lme()} function in the \textbf{nlme} package, and the arguments
for \texttt{cv()} in this case are the same as for a model fit by
\texttt{lmer()} or \texttt{glmer()}. We illustrate with the mixed model
fit to the HSB data:

\begin{CodeChunk}
\begin{CodeInput}
R> library(nlme)
R> hsb.lme <- lme(mathach ~ mean.ses*cses + sector*cses,
+                  random = ~ cses | school, data=HSB,
+                control=list(opt="optim"))
R> summary(hsb.lme)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed-effects model fit by REML
  Data: HSB 
       AIC      BIC    logLik
  46524.78 46593.57 -23252.39

Random effects:
 Formula: ~cses | school
 Structure: General positive-definite, Log-Cholesky parametrization
            StdDev     Corr  
(Intercept) 1.54117685 (Intr)
cses        0.01817364 0.006 
Residual    6.06349216

Fixed effects:  mathach ~ mean.ses * cses + sector * cses 
                        Value Std.Error   DF  t-value p-value
(Intercept)         12.128207 0.1991964 7022 60.88567   0e+00
mean.ses             5.336665 0.3689784  157 14.46335   0e+00
cses                 2.942145 0.1512240 7022 19.45554   0e+00
sectorCatholic       1.224531 0.3061139  157  4.00025   1e-04
mean.ses:cses        1.044406 0.2910747 7022  3.58810   3e-04
cses:sectorCatholic -1.642148 0.2331162 7022 -7.04433   0e+00
 Correlation: 
                    (Intr) men.ss cses   sctrCt mn.ss:
mean.ses             0.256                            
cses                 0.000  0.000                     
sectorCatholic      -0.699 -0.356  0.000              
mean.ses:cses        0.000  0.000  0.295  0.000       
cses:sectorCatholic  0.000  0.000 -0.696  0.000 -0.351

Standardized Within-Group Residuals:
        Min          Q1         Med          Q3         Max 
-3.17010624 -0.72487654  0.01489162  0.75426269  2.96549829

Number of Observations: 7185
Number of Groups: 160 
\end{CodeOutput}
\begin{CodeInput}
R> cv(hsb.lme, k=10, clusterVariables="school", seed=5240)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 5240
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation based on 160 {school} clusters
cross-validation criterion = 39.1569
bias-adjusted cross-validation criterion = 39.14881
95% CI for bias-adjusted CV criterion = (38.06591, 40.23171)
full-sample criterion = 39.0062 
\end{CodeOutput}
\begin{CodeInput}
R> cv(hsb.lme, seed=1575)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 1575
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 37.44163
bias-adjusted cross-validation criterion = 37.40222
95% CI for bias-adjusted CV criterion = (36.35136, 38.45309)
full-sample criterion = 36.14707 
\end{CodeOutput}
\end{CodeChunk}

We used the same random-number generator seeds as in the previous
example cross-validating the model fit by \texttt{lmer()}, and so the
same folds are employed in both cases.\footnote{The observant reader
  will notice that we set the argument
  \texttt{control=list(opt="optim")} in the call to \texttt{lme()},
  changing the optimizer employed from the default \texttt{"nlminb"}. We
  did this because with the default optimizer, \texttt{lme()}
  encountered the same convergence issue as \texttt{lmer()}, but rather
  than issuing a warning, \texttt{lme()} failed, reporting an error. As
  it turns out, setting the optimizer to \texttt{"optim"} avoids this
  problem.} The estimated covariance components and fixed effects in the
summary output differ slightly between the \texttt{lmer()} and
\texttt{lme()} solutions, although both functions seek to maximize the
REML criterion. This is, of course, to be expected when different
algorithms are used for numerical optimization. To the precision
reported, the cluster-level CV results for the \texttt{lmer()} and
\texttt{lme()} models are identical, while the case-level CV results are
very similar but not identical.

\hypertarget{example-contrived-hierarchical-data}{%
\subsection{Example: Contrived hierarchical
data}\label{example-contrived-hierarchical-data}}

We introduce an artificial data set that exemplifies aspects of
cross-validation particular to hierarchical models. Using this data set,
we show that model comparisons employing cluster-based and those
employing case-based cross-validation may not agree on a ``best'' model.
Furthermore, commonly used measures of fit, such as mean-squared error,
do not necessarily become smaller as models become larger, even when the
models are nested, and even when the measure of fit is computed for the
whole data set.

Consider a researcher studying improvement in a skill, yodeling, for
example, among students enrolled in a four-year yodeling program. The
plan is to measure each student's skill level at the beginning of the
program and every year thereafter until the end of the program,
resulting in five annual measurements for each student. It turns out
that yodeling appeals to students of all ages, and students enrolling in
the program range in age from 20 to 70. Moreover, participants'
untrained yodeling skill is similar at all ages, as is their rate of
progress with training. All students complete the four-year program.

The researcher, who has more expertise in yodeling than in modeling,
decides to model the response, \(y\), yodeling skill, as a function of
age, \(x\), reasoning that students get older during their stay in the
program, and (incorrectly) that age can serve as a proxy for elapsed
time. The researcher knows that a mixed model should be used to account
for clustering due to the expected similarity of measurements taken from
each student.

We start by generating the data, using parameters consistent with the
description above and meant to highlight the issues that arise in
cross-validating mixed-effects models:\footnote{The
  \texttt{regsubsets()} function computes several measures of model
  predictive performance, including the \(R^2\) and \(R^2\) adjusted for
  degrees of freedom, the residual sums of squares, Mallows's \(C_p\),
  and the BIC. Several of these are suitable for comparing models with
  differing numbers of coefficients---we use the BIC below---but all
  necessarily agree when comparing models with the \emph{same} number of
  coefficients.}

\begin{CodeChunk}
\begin{CodeInput}
R> # Parameters:
R> set.seed(9693) 
R> Nb <- 100     # number of groups
R> Nw <- 5       # number of individuals within groups
R> Bb <- 0       # between-group regression coefficient on group mean
R> SDre <- 2.0   # between-group SD of random level relative to group mean of x
R> SDwithin <- 0.5  # within group SD
R> Bw <- 1          # within group effect of x
R> Ay <- 10         # intercept for response
R> Ax <- 20         # starting level of x
R> Nx <- Nw*10      # number of distinct x values
R> 
R> Data <- data.frame(
+   group = factor(rep(1:Nb, each=Nw)),
+   x = Ax + rep(1:Nx, length.out = Nw*Nb)
+ ) |>
+   within(
+     {
+       xm  <- ave(x, group, FUN = mean) # within-group mean
+       y <- Ay +
+         Bb * xm +                    # contextual effect
+         Bw * (x - xm) +              # within-group effect
+         rnorm(Nb, sd=SDre)[group] +  # random level by group
+         rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
+     }
+   )
\end{CodeInput}
\end{CodeChunk}

Here is a scatterplot of the data for a representative group of 10
(without loss of generality, the first 10) of 100 students, showing the
95\% concentration ellipse for each cluster:\footnote{We find it
  convenient to use the \textbf{lattice} \citep{Sarkar:2008} and
  \textbf{latticeExtra} \citep{SarkarAndrews:2022} packages for this and
  other graphs in this section.}

\begin{CodeChunk}
\begin{CodeInput}
R> library("lattice")
R> library("latticeExtra")
R> plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
+                ylim=c(4, 16),
+                par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
+     layer(panel.ellipse(..., center.cex=0))
R> plot # display graph
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot1-1}

}

\caption[Hierarchical data set, showing the first 10 of 100 students]{Hierarchical data set, showing the first 10 of 100 students.}\label{fig:plot1}
\end{figure}
\end{CodeChunk}

The between-student effect of age is 0 but the within-student effect is
1. Due to the large variation in ages between students, the
least-squares regression of yodeling skill on age (for the 500
observations among all 100 students) produces an estimated slope close
to 0 (though with a small \(p\)-value), because the slope is heavily
weighted toward the between-student effect:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(lm(y ~ x, data=Data))
\end{CodeInput}
\begin{CodeOutput}

Call:
lm(formula = y ~ x, data = Data)

Residuals:
    Min      1Q  Median      3Q     Max 
-5.7713 -1.6583 -0.0894  1.5520  7.6240

Coefficients:
            Estimate Std. Error t value Pr(>|t|)
(Intercept) 9.050430   0.347189  26.068  < 2e-16
x           0.020908   0.007273   2.875  0.00422

Residual standard error: 2.347 on 498 degrees of freedom
Multiple R-squared:  0.01632,   Adjusted R-squared:  0.01435 
F-statistic: 8.263 on 1 and 498 DF,  p-value: 0.004219
\end{CodeOutput}
\end{CodeChunk}

The initial mixed-effects model that we fit to the data is a simple
random-intercepts model:

\begin{CodeChunk}
\begin{CodeInput}
R> # random intercept only:
R> mod.0 <- lmer(y ~ 1 + (1 | group), Data)
R> summary(mod.0)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed model fit by REML ['lmerMod']
Formula: y ~ 1 + (1 | group)
   Data: Data

REML criterion at convergence: 2103.1

Scaled residuals: 
     Min       1Q   Median       3Q      Max 
-2.03514 -0.72645 -0.01169  0.78477  2.04377

Random effects:
 Groups   Name        Variance Std.Dev.
 group    (Intercept) 2.900    1.703   
 Residual             2.712    1.647   
Number of obs: 500, groups:  group, 100

Fixed effects:
            Estimate Std. Error t value
(Intercept)  10.0018     0.1855   53.91
\end{CodeOutput}
\end{CodeChunk}

We will shortly consider three other, more complex, mixed models;
because of data-management considerations, it is convenient to fit them
now, but we defer discussion of these models:

\begin{CodeChunk}
\begin{CodeInput}
R> # effect of x and random intercept:
R> mod.1 <- lmer(y ~ x + (1 | group), Data)
R> 
R> # effect of x, contextual (student) mean of x, and random intercept:
R> mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
R>         # equivalent to y ~ I(x - xm) + xm + (1 | group)
R> 
R> # model generating the data (where Bb = 0)
R> mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
\end{CodeInput}
\end{CodeChunk}

We proceed to obtain predictions from the random-intercept model
(\texttt{mod.0}) and the other models (\texttt{mod.1}, \texttt{mod.2},
and \texttt{mod.3}) based on fixed effects alone, as would be used for
cross-validation based on clusters (i.e., students), and for fixed and
random effects---so-called best linear unbiased predictions or
BLUPs---as would be used for cross-validation based on cases (i.e.,
occasions within students):

\begin{CodeChunk}
\begin{CodeInput}
R> Data <- within(Data, {
+   fit_mod0.fe <- predict(mod.0, re.form = ~ 0) # fixed effects only
+   fit_mod0.re <- predict(mod.0) # fixed and random effects (BLUPs)
+   fit_mod1.fe <- predict(mod.1, re.form = ~ 0)
+   fit_mod1.re <- predict(mod.1)
+   fit_mod2.fe <- predict(mod.2, re.form = ~ 0)
+   fit_mod2.re <- predict(mod.2)
+   fit_mod3.fe <- predict(mod.3, re.form = ~ 0)
+   fit_mod3.re <- predict(mod.3)
+ })
\end{CodeInput}
\end{CodeChunk}

We then prepare the data for plotting:

\begin{CodeChunk}
\begin{CodeInput}
R> Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
+               timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
R> Data_long$id <- 1:nrow(Data_long)
R> Data_long <- reshape(Data_long, direction = "long", sep = "_", 
+               timevar = "modelcode",  varying = grep("_", names(Data_long)))
R> Data_long$model <- factor(
+   c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
+   [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
+ )
\end{CodeInput}
\end{CodeChunk}

Predictions based on the random-intercept model \texttt{mod.0} for the
first 10 students are shown in the following graph:

\begin{CodeChunk}
\begin{CodeInput}
R> (plot +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
+          groups=group, type="l", lwd=2) +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
+          groups=group, type="l", lwd=2, lty=3)
+ ) |> update(
+   main="Model: y ~ 1 + (1 | group)",
+   key=list(
+     corner=c(0.05, 0.05),
+     text=list(c("fixed effects only","fixed and random")),
+     lines=list(lty=c(1, 3))))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod0-1}

}

\caption[Predictions from the random intercept model]{Predictions from the random intercept model.}\label{fig:plot-fits-mod0}
\end{figure}
\end{CodeChunk}

The fixed-effect predictions for the various individuals are
identical---the estimated fixed-effects intercept or estimated general
mean of \(y\)---while the BLUPs are the sums of the fixed-effects
intercept and the random intercepts, and are only slightly shrunken
towards the general mean. Because in our artificial data there is no
population relationship between age and skill, the fixed-effect-only
predictions and the BLUPs are not very different.

Our next model, \texttt{mod.1}, includes a fixed intercept and fixed
effect of \texttt{x} along with a random intercept:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(mod.1)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed model fit by REML ['lmerMod']
Formula: y ~ x + (1 | group)
   Data: Data

REML criterion at convergence: 1564.5

Scaled residuals: 
     Min       1Q   Median       3Q      Max 
-2.90160 -0.63501  0.01879  0.55407  2.82932

Random effects:
 Groups   Name        Variance Std.Dev.
 group    (Intercept) 192.9406 13.8903 
 Residual               0.2569  0.5068 
Number of obs: 500, groups:  group, 100

Fixed effects:
             Estimate Std. Error t value
(Intercept) -33.91892    1.56446  -21.68
x             0.96529    0.01581   61.05

Correlation of Fixed Effects:
  (Intr)
x -0.460
\end{CodeOutput}
\end{CodeChunk}

Predictions from this model appear in the following graph:

\begin{CodeChunk}
\begin{CodeInput}
R> (plot +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
+          groups=group, type="l", lwd=2) +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
+          groups=group, type="l", lwd=2, lty=3)
+ ) |> update(
+   main="Model: y ~ 1 + x + (1 | group)",
+   ylim=c(-15, 35),
+   key=list(
+     corner=c(0.95, 0.05),
+     text=list(c("fixed effects only","fixed and random")),
+     lines=list(lty=c(1, 3))))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod1-1}

}

\caption[Predictions from the model with random intercepts and $x$ as a fixed-effect predictor]{Predictions from the model with random intercepts and $x$ as a fixed-effect predictor.}\label{fig:plot-fits-mod1}
\end{figure}
\end{CodeChunk}

The BLUPs fit the observed data very closely, but predictions based on
the fixed effects alone, with a common intercept and slope for all
clusters, are very poor---indeed, much worse than the fixed-effects-only
predictions based on the simpler random-intercept model, \texttt{mod.0}.
We therefore anticipate (and show later in this section) that case-based
cross-validation will prefer \texttt{mod1} to \texttt{mod0}, but that
cluster-based cross-validation will prefer \texttt{mod0} to
\texttt{mod1}.

Our third model, \texttt{mod.2}, includes the contextual effect of
\(x\)---that is, the cluster mean \texttt{xm}---along with \(x\) and the
intercept in the fixed-effect part of the model, and a random intercept:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(mod.2)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed model fit by REML ['lmerMod']
Formula: y ~ x + xm + (1 | group)
   Data: Data

REML criterion at convergence: 1169.2

Scaled residuals: 
     Min       1Q   Median       3Q      Max 
-2.98466 -0.63750  0.00191  0.55682  2.73246

Random effects:
 Groups   Name        Variance Std.Dev.
 group    (Intercept) 3.3986   1.8435  
 Residual             0.2552   0.5052  
Number of obs: 500, groups:  group, 100

Fixed effects:
            Estimate Std. Error t value
(Intercept)  9.47866    0.61705   15.36
x            0.99147    0.01597   62.07
xm          -0.97998    0.02055  -47.68

Correlation of Fixed Effects:
   (Intr) x     
x   0.000       
xm -0.600 -0.777
\end{CodeOutput}
\end{CodeChunk}

This model is equivalent to fitting
\texttt{y\ \textasciitilde{}\ I(x\ -\ xm)\ +\ xm\ +\ (1\ \textbar{}\ group)},
which is the model that generated the data once the coefficient of the
contextual predictor \texttt{xm} is set to 0 (as it is in
\texttt{mod.3}, discussed below).

Predictions from model \texttt{mod.2} appear in the following graph:

\begin{CodeChunk}
\begin{CodeInput}
R> (plot +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
+          groups=group, type="l", lwd=2) +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
+          groups=group, type="l", lwd=2, lty=3)
+ ) |> update(
+   main="Model: y ~ 1 + x + xm + (1 | group)",
+   ylim=c(4, 16),
+   key=list(
+     corner=c(0.05, 0.05),
+     text=list(c("fixed effects only","fixed and random")),
+     lines=list(lty=c(1, 3))))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod2-1}

}

\caption[Predictors from the model with random intercepts, $x$, and the group (student) mean of $x$ as predictors]{Predictors from the model with random intercepts, $x$, and the group (student) mean of $x$ as predictors.}\label{fig:plot-fits-mod2}
\end{figure}
\end{CodeChunk}

Depending on the estimated variance parameters of the model, a mixed
model like \texttt{mod.2} will apply varying degrees of shrinkage to the
random-intercept BLUPs that correspond to variation in the heights of
the parallel fitted lines for the individual students. In our contrived
data, the \texttt{mod.2} applies little shrinkage, allowing substantial
variability in the heights of the fitted lines, which closely approach
the observed values for each student. The fit of the mixed model
\texttt{mod.2} is consequently similar to that of a fixed-effects model
with age and a categorical predictor for individual students (i.e.,
treating students as a factor, and not shown here).

The mixed model \texttt{mod.2} therefore fits individual observations
well, and we anticipate a favorable assessment using individual-based
cross-validation. In contrast, the large variability in the BLUPs
results in larger residuals for predictions based on fixed effects
alone, and so we expect that cluster-based cross-validation won't show
an advantage for model \texttt{mod.2} compared to the smaller model
\texttt{mod.0}, which includes only fixed and random intercepts.

Had the mixed model applied considerable shrinkage, then neither
cluster-based nor case-based cross-validation would show much
improvement over the random-intercept-only model. In our experience, the
degree of shrinkage does not vary smoothly as parameters are changed but
tends to be ``all or nothing,'' and near the tipping point, the behavior
of estimates can be affected considerably by the choice of algorithm
used to fit the model.

Finally, \texttt{mod.3} directly estimates the model used to generate
the data. As mentioned, it is a constrained version of \texttt{mod.2},
with the coefficient of \texttt{xm} set to 0, and with \texttt{x}
expressed as a deviation from the cluster mean \texttt{xm}:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(mod.3)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed model fit by REML ['lmerMod']
Formula: y ~ I(x - xm) + (1 | group)
   Data: Data

REML criterion at convergence: 1163.2

Scaled residuals: 
     Min       1Q   Median       3Q      Max 
-2.97703 -0.63204  0.00627  0.56032  2.72489

Random effects:
 Groups   Name        Variance Std.Dev.
 group    (Intercept) 3.3913   1.8415  
 Residual             0.2552   0.5052  
Number of obs: 500, groups:  group, 100

Fixed effects:
            Estimate Std. Error t value
(Intercept) 10.00176    0.18553   53.91
I(x - xm)    0.99147    0.01597   62.07

Correlation of Fixed Effects:
          (Intr)
I(x - xm) 0.000 
\end{CodeOutput}
\end{CodeChunk}

The predictions from \texttt{mod.3} are therefore similar to those from
\texttt{mod.2}:

\begin{CodeChunk}
\begin{CodeInput}
R> (plot +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
+          groups=group, type="l", lwd=2) +
+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
+          groups=group, type="l", lwd=2, lty=3)
+ ) |> update(
+   main="Model: y ~ 1 + I(x - xm) + (1 | group)",
+   ylim=c(4, 16),
+   key=list(
+     corner=c(0.05, 0.05),
+     text=list(c("fixed effects only","fixed and random")),
+     lines=list(lty=c(1, 3))))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod3-1}

}

\caption[Predictions from the estimated model generating the data]{Predictions from the estimated model generating the data.}\label{fig:plot-fits-mod3}
\end{figure}
\end{CodeChunk}

We next carry out case-based cross-validation, which, as we have
explained, is based on both fixed and predicted random effects (i.e.,
BLUPs), and cluster-based cross-validation, which is based on fixed
effects only. In order to reduce between-model random variability in
comparisons of models, we apply \texttt{cv()} to the list of models
created by the \texttt{models()} function (introduced previously),
performing cross-validation with the same folds for each model:

\begin{CodeChunk}
\begin{CodeInput}
R> modlist <- models("~ 1"=mod.0, "~ 1 + x"=mod.1, 
+                   "~ 1 + x + xm"=mod.2, "~ 1 + I(x - xm)"=mod.3)
R> cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
R> plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/cross-validation-clusters-1}

}

\caption[10-fold cluster-based cross-validation comparing random intercept models with varying fixed effects]{10-fold cluster-based cross-validation comparing random intercept models with varying fixed effects. The error bars show the 95\% confidence interval around the CV estimate of the MSE for each model.}\label{fig:cross-validation-clusters}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> cvs_cases <- cv(modlist, data=Data, seed=9693)
R> plot(cvs_cases, main="Model Comparison, Case-Based CV")
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/cross-validation-cases-1}

}

\caption[10-fold case-based cross-validation comparing random intercept models with varying fixed effects]{10-fold case-based cross-validation comparing random intercept models with varying fixed effects.}\label{fig:cross-validation-cases}
\end{figure}
\end{CodeChunk}

In summary, model \texttt{mod.1}, with \(x\) alone and without the
contextual mean of \(x\), is assessed as fitting very poorly by
cluster-based CV, but relatively much better by case-based CV. Model
\texttt{mod.2}, which includes both \(x\) and its contextual mean,
produces better results using both cluster-based and case-based CV. The
data-generating model, \texttt{mod.3}, which includes the fixed effect
of \texttt{x\ -\ xm} in place of separate terms in \texttt{x} and
\texttt{xm}, isn't distinguishable from model \texttt{mod.2}, which
includes \texttt{x} and \texttt{xm} separately, even though
\texttt{mod.2} has an unnecessary parameter (recall that the population
coefficient of \texttt{xm} is 0 when \texttt{x} is expressed as
deviations from the contextual mean). These conclusions are consistent
with our observations based on graphing predictions from the various
models, and they illustrate the desirability of assessing mixed-effect
models at different hierarchical levels.

\hypertarget{example-crossed-random-effects}{%
\subsection{Example: Crossed random
effects}\label{example-crossed-random-effects}}

Crossed random effects arise when the structure of the data aren't
strictly hierarchical. Nevertheless, crossed and nested random effects
can be handled in much the same manner, by refitting the mixed-effects
model to the data with a fold of clusters or cases removed and using the
refitted model to predict the response in the removed fold.

We'll illustrate with data on pig growth, introduced by \citet[Table
3.1]{DiggleLiangZeger:1994}. The data are in the \texttt{Pigs} data
frame in the \textbf{cv} package:

\begin{CodeChunk}
\begin{CodeInput}
R> head(Pigs, 9)
\end{CodeInput}
\begin{CodeOutput}
  id week weight
1  1    1   24.0
2  1    2   32.0
3  1    3   39.0
4  1    4   42.5
5  1    5   48.0
6  1    6   54.5
7  1    7   61.0
8  1    8   65.0
9  1    9   72.0
\end{CodeOutput}
\begin{CodeInput}
R> head(xtabs(~ id + week, data=Pigs), 3)
\end{CodeInput}
\begin{CodeOutput}
   week
id  1 2 3 4 5 6 7 8 9
  1 1 1 1 1 1 1 1 1 1
  2 1 1 1 1 1 1 1 1 1
  3 1 1 1 1 1 1 1 1 1
\end{CodeOutput}
\begin{CodeInput}
R> tail(xtabs(~ id + week, data=Pigs), 3)
\end{CodeInput}
\begin{CodeOutput}
    week
id   1 2 3 4 5 6 7 8 9
  46 1 1 1 1 1 1 1 1 1
  47 1 1 1 1 1 1 1 1 1
  48 1 1 1 1 1 1 1 1 1
\end{CodeOutput}
\end{CodeChunk}

Each of 48 pigs is observed weekly over a period of 9 weeks, with the
weight of the pig recorded in kg. The data are in ``long'' format, as is
appropriate for use with the \texttt{lmer()} function in the
\textbf{lme4} package. The data are very regular, with no missing cases.

The following graph, showing the growth trajectories of the pigs, is
similar to Figure 3.1 in \citet{DiggleLiangZeger:1994}; we add an
overall least-squares line and a loess smooth, which are nearly
indistinguishable:

\begin{CodeChunk}
\begin{CodeInput}
R> plot(weight ~ week, data=Pigs, type="n")
R> for (i in unique(Pigs$id)){
+   with(Pigs, lines(x=1:9, y=Pigs[id == i, "weight"],
+                    col="gray"))
+ }
R> abline(lm(weight ~ week, data=Pigs), col="blue", lwd=2)
R> lines(with(Pigs, loess.smooth(week, weight, span=0.5)),
+       col="magenta", lty=2, lwd=2)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=0.6\linewidth]{JSS-article_files/figure-latex/pigs-graph-1}

}

\caption[Growth trajectories for 48 pigs, with overall least-squares line (sold blue) and loess line (broken magenta)]{Growth trajectories for 48 pigs, with overall least-squares line (sold blue) and loess line (broken magenta).}\label{fig:pigs-graph}
\end{figure}
\end{CodeChunk}

The individual ``growth curves'' and the overall trend are generally
linear, with some tendency for variability of pig weight to increase
over weeks (a feature of the data that we ignore in the mixed model that
we fit to the data below).

The \textbf{Stata} mixed-effects models manual proposes a model with
crossed random effects for the \texttt{Pigs} data \citep[ page
37]{Stata:2023}:

\begin{quote}
{[}S{]}uppose that we wish to fit \[
\mathrm{weight}_{ij} = \beta_0 + \beta_1 \mathrm{week}_{ij} + u_i + v_j + \varepsilon_{ij} 
\] for the \(i = 1, \ldots, 9\) weeks and \(j = 1, \dots, 48\) pigs and
\[
u_i \sim N(0, \sigma^2_u); v_j \sim N(0, \sigma^2_v ); \varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon)
\] all independently. That is, we assume an overall population-average
growth curve \(\beta_0 + \beta_1 \mathrm{week}\) and a random
pig-specific shift. In other words, the effect due to week, \(u_i\), is
systematic to that week and common to all pigs. The rationale behind
{[}this model{]} could be that, assuming that the pigs were measured
contemporaneously, we might be concerned that week-specific random
factors such as weather and feeding patterns had significant systematic
effects on all pigs.
\end{quote}

Although we might prefer an alternative model,\footnote{These are
  repeated-measures data, which would be more conventionally modeled
  with autocorrelated errors within pigs. The \texttt{lme()} function in
  the \textbf{nlme} package, for example, is capable of fitting a
  mixed-model of this form.} we think that this is a reasonable
specification.

The \textbf{Stata} manual fits the mixed model by maximum likelihood
(rather than REML), and we duplicate the results reported there using
\texttt{lmer()}:

\begin{CodeChunk}
\begin{CodeInput}
R> m.p <- lmer(weight ~ week + (1 | id) + (1 | week),
+             data=Pigs, REML=FALSE, # i.e., ML
+             control=lmerControl(optimizer="bobyqa"))
R> summary(m.p)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed model fit by maximum likelihood  ['lmerMod']
Formula: weight ~ week + (1 | id) + (1 | week)
   Data: Pigs
Control: lmerControl(optimizer = "bobyqa")

     AIC      BIC   logLik deviance df.resid 
  2037.6   2058.0  -1013.8   2027.6      427

Scaled residuals: 
    Min      1Q  Median      3Q     Max 
-3.7750 -0.5418  0.0054  0.4762  3.9816

Random effects:
 Groups   Name        Variance Std.Dev.
 id       (Intercept) 14.83622 3.8518  
 week     (Intercept)  0.08499 0.2915  
 Residual              4.29733 2.0730  
Number of obs: 432, groups:  id, 48; week, 9

Fixed effects:
            Estimate Std. Error t value
(Intercept) 19.35561    0.63340   30.56
week         6.20990    0.05393  115.14

Correlation of Fixed Effects:
     (Intr)
week -0.426
\end{CodeOutput}
\end{CodeChunk}

We opt for the non-default \texttt{"bobyqa"} optimizer because it
provides more numerically stable results for subsequent cross-validation
in this example.

We can then cross-validate the model by omitting folds composed of pigs,
folds composed of weeks, or folds composed of pig-weeks (which in the
\texttt{Pigs} data set correspond to individual cases, using only the
fixed effects):

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.p, clusterVariables="id")
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation based on 48 {id} clusters
cross-validation criterion = 19.97309
bias-adjusted cross-validation criterion = 19.96496
95% CI for bias-adjusted CV criterion = (17.1249, 22.80502)
full-sample criterion = 19.20076 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.p, clusterVariables="week")
\end{CodeInput}
\begin{CodeOutput}
boundary (singular) fit: see help('isSingular')
\end{CodeOutput}
\begin{CodeOutput}
n-Fold Cross Validation based on 9 {week} clusters
cross-validation criterion = 19.31219
bias-adjusted cross-validation criterion = 19.30499
95% CI for bias-adjusted CV criterion = (16.56624, 22.04373)
full-sample criterion = 19.20076 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.p, clusterVariables=c("id", "week"), k=10, seed=8469)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 8469
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation based on 432 {id, week} clusters
cross-validation criterion = 19.2352
bias-adjusted cross-validation criterion = 19.23294
95% CI for bias-adjusted CV criterion = (16.49263, 21.97325)
full-sample criterion = 19.20076 
\end{CodeOutput}
\end{CodeChunk}

We can also cross-validate the individual cases taking account of the
random effects (employing the same 10 folds):

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.p, k=10, seed=8469)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 8469
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 5.15835
bias-adjusted cross-validation criterion = 5.072933
95% CI for bias-adjusted CV criterion = (4.122952, 6.022913)
full-sample criterion = 3.795972 
\end{CodeOutput}
\end{CodeChunk}

Because these predictions are based on BLUPs, they are more accurate
than the predictions based only on fixed effects.\footnote{Even though
  there is only one observation per combination of pigs and weeks, we
  can use the BLUP for the omitted case because of the crossed structure
  of the random effects; that is each pig-week has a pig random effect
  and a week random effect. Although it probably isn't sensible, we can
  imagine a mixed model for the pig data that employs nested random
  effects, which would be specified by
  \texttt{lmer(weight\ \textasciitilde{}\ week\ +\ (1\ \textbar{}\ id/week),\ data=Pigs)}---that
  is, a random intercept that varies by combinations of \texttt{id}
  (pig) and \texttt{week}. This model can't be fit, however: With only
  one case per combination of \texttt{id} and \texttt{week}, the nested
  random-effect variance is indistinguishable from the case-level
  variance.} As well, the difference between the MSE computed for the
model fit to the full data and the CV estimates of the MSE is greater
here than for cluster-based predictions.

\hypertarget{replicating-cross-validation}{%
\section{Replicating
cross-validation}\label{replicating-cross-validation}}

Assuming that the number of cases \(n\) is a multiple of the number of
folds \(k\)---a slightly simplifying assumption---the number of possible
partitions of cases into folds is \(\frac{n!}{[(n/k)!]^k}\), a number
that grows very large very quickly. For example, for \(n = 10\) and
\(k = 5\), so that the folds are each of size \(n/k = 2\), there are
\(113,400\) possible partitions; for \(n=100\) and \(k=5\), where
\(n/k = 20\), still a small problem, the number of possible partitions
is truly astronomical, \(1.09\times 10^{66}\).

Because the partition into folds that's employed is selected randomly,
the resulting CV criterion estimates are subject to sampling error. (An
exception is LOO cross-validation, which is not at all random.) To get a
sense of the magnitude of the sampling error, we can repeat the CV
procedure with different randomly selected partitions into folds. All of
the CV functions in the \textbf{cv} package are capable of repeated
cross-validation, with the number of repetitions controlled by the
\texttt{reps} argument, which defaults to \texttt{1}.

Here, for example, is 10-fold CV for the Mroz logistic regression,
repeated 5 times:

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.mroz, criterion=BayesRule, seed=248, reps=5, 
+    method="Woodbury")
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 248
\end{CodeOutput}
\begin{CodeOutput}
R RNG seed set to 68134
\end{CodeOutput}
\begin{CodeOutput}
R RNG seed set to 767359
\end{CodeOutput}
\begin{CodeOutput}
R RNG seed set to 556270
\end{CodeOutput}
\begin{CodeOutput}
R RNG seed set to 882966
\end{CodeOutput}
\begin{CodeOutput}

Replicate 1:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3200531
bias-adjusted cross-validation criterion = 0.3130109
95% CI for bias-adjusted CV criterion = (0.2796692, 0.3463526)
full-sample criterion = 0.3067729

Replicate 2:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3160691
bias-adjusted cross-validation criterion = 0.311697
95% CI for bias-adjusted CV criterion = (0.2784665, 0.3449274)
full-sample criterion = 0.3067729

Replicate 3:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.314741
bias-adjusted cross-validation criterion = 0.3086212
95% CI for bias-adjusted CV criterion = (0.2754285, 0.341814)
full-sample criterion = 0.3067729

Replicate 4:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3240372
bias-adjusted cross-validation criterion = 0.318069
95% CI for bias-adjusted CV criterion = (0.2846189, 0.3515192)
full-sample criterion = 0.3067729

Replicate 5:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3240372
bias-adjusted cross-validation criterion = 0.3192577
95% CI for bias-adjusted CV criterion = (0.2858076, 0.3527079)
full-sample criterion = 0.3067729

Average:
10-Fold Cross Validation
method: Woodbury
criterion: BayesRule
cross-validation criterion = 0.3198318 (0.003886968)
bias-adjusted cross-validation criterion = 0.3139445 (0.004009339)
full-sample criterion = 0.3067729 
\end{CodeOutput}
\end{CodeChunk}

When \texttt{reps} \textgreater{} \texttt{1}, the result returned by
\texttt{cv()} is an object of class \texttt{"cvList"}---literally a list
of \texttt{"cv"} objects. The results are reported for each repetition
and then averaged across repetitions, with the standard deviations of
the CV criterion and the biased-adjusted CV criterion given in
parentheses. In this example, there is therefore little variation across
repetitions, increasing our confidence in the reliability of the
results.

Notice that the seed that's set in the \texttt{cv()} command pertains to
the first repetition and the seeds for the remaining repetitions are
then selected pseudo-randomly.\footnote{Because of the manner in which
  the computation is performed, the order of the replicates in the
  \texttt{"cvList"} object returned by \texttt{cv()} isn't the same as
  the order in which the replicates are computed. Each element of the
  result, however, is a \texttt{"cv"} object with the correct
  random-number seed saved, and so this technical detail can be safely
  ignored. The individual \texttt{"cv"} objects are printed in the order
  in which they are stored rather than the order in which they are
  computed.} Setting the first seed, however, makes the entire process
easily replicable, and the seed for each repetition is stored in the
corresponding element of the \texttt{"cvList"} object (which isn't,
however, saved in the example).

It's also possible to replicate CV when comparing competing models via
the \texttt{cv()} method for \texttt{"modList"} objects. Recall our
comparison of polynomial regressions of varying degree fit to the
\texttt{Auto} data; we performed 10-fold CV for each of 10 models. Here,
we replicate that process 5 times for each model and graph the results:

\begin{CodeChunk}
\begin{CodeInput}
R> cv.auto.reps <- cv(models(m.1, m.2, m.3, m.4, m.5,
+                         m.6, m.7, m.8, m.9, m.10),
+                  data=Auto, seed=8004, reps=5)
R> plot(cv.auto.reps)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/model-comparison-with-reps-1}

}

\caption[ Replicated cross-validated 10-fold CV as a function of polynomial degree, $p$]{ Replicated cross-validated 10-fold CV as a function of polynomial degree, $p$}\label{fig:model-comparison-with-reps}
\end{figure}
\end{CodeChunk}

The graph shows both the average CV criterion and its range for each of
the competing models.

\hypertarget{cross-validating-model-selection}{%
\section{Cross-validating model
selection}\label{cross-validating-model-selection}}

\hypertarget{a-preliminary-example}{%
\subsection{A preliminary example}\label{a-preliminary-example}}

As \citet[Sec. 7.10.2: ``The Wrong and Right Way to Do
Cross-validation'']{HastieTibshiraniFriedman:2009} explain, if the whole
data are used to select or fine-tune a statistical model, subsequent
cross-validation of the model is intrinsically misleading, because the
model is selected to fit the whole data, including the part of the data
that remains when each fold is removed.

The following example is similar in spirit to one employed by
\citet{HastieTibshiraniFriedman:2009}. Suppose that we randomly generate
\(n = 1000\) independent observations for a response variable variable
\(y \sim N(\mu = 10, \sigma^2 = 0)\), and independently sample \(1000\)
observations for \(p = 100\) ``predictors,'' \(x_1, \ldots, x_{100}\),
each from \(x_j \sim N(0, 1)\). The response has nothing to do with the
predictors and so the population linear-regression model
\(y_i = \alpha + \beta_1 x_{i1} + \cdots + \beta_{100} x_{i,100} + \varepsilon_i\)
has \(\alpha = 10\) and all \(\beta_j = 0\).

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(24361) # for reproducibility
R> D <- data.frame(
+   y = rnorm(1000, mean=10),
+   X = matrix(rnorm(1000*100), 1000, 100)
+ )
R> head(D[, 1:6])
\end{CodeInput}
\begin{CodeOutput}
          y        X.1        X.2         X.3         X.4         X.5
1 10.031647 -1.2388628 -0.2648705 -0.03539048 -2.57697337  0.81104761
2  9.664989  0.1228689 -0.1774440  0.37290421 -0.93513788  0.62867324
3 10.023249 -0.9505172 -0.7348667 -1.05978180  0.88294443  0.02391808
4  8.990969  1.1357103  0.3241085  0.11036901  1.37630285 -0.42211426
5  9.071249  1.4947403  1.8753802  0.10574793  0.29213991 -0.18456833
6 11.349283 -0.1845331 -0.7803709 -1.23803778 -0.01094861  0.69103395
\end{CodeOutput}
\end{CodeChunk}

Least-squares provides accurate estimates of the regression constant
\(\alpha = 10\) and the error variance \(\sigma^2 = 1\) for the ``null
model'' including only the regression constant; moreover, the omnibus
\(F\)-test of the correct null hypothesis that all of the \(\beta\)s are
0 for the ``full model'' with all 100 \(x\)s is associated with a large
\(p\)-value:

\begin{CodeChunk}
\begin{CodeInput}
R> m.full <- lm(y ~ ., data=D)
R> m.null <- lm(y ~ 1, data=D)
R> anova(m.null, m.full)
\end{CodeInput}
\begin{CodeOutput}
Analysis of Variance Table

Model 1: y ~ 1
Model 2: y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7 + X.8 + X.9 + X.10 + 
    X.11 + X.12 + X.13 + X.14 + X.15 + X.16 + X.17 + X.18 + X.19 + 
    X.20 + X.21 + X.22 + X.23 + X.24 + X.25 + X.26 + X.27 + X.28 + 
    X.29 + X.30 + X.31 + X.32 + X.33 + X.34 + X.35 + X.36 + X.37 + 
    X.38 + X.39 + X.40 + X.41 + X.42 + X.43 + X.44 + X.45 + X.46 + 
    X.47 + X.48 + X.49 + X.50 + X.51 + X.52 + X.53 + X.54 + X.55 + 
    X.56 + X.57 + X.58 + X.59 + X.60 + X.61 + X.62 + X.63 + X.64 + 
    X.65 + X.66 + X.67 + X.68 + X.69 + X.70 + X.71 + X.72 + X.73 + 
    X.74 + X.75 + X.76 + X.77 + X.78 + X.79 + X.80 + X.81 + X.82 + 
    X.83 + X.84 + X.85 + X.86 + X.87 + X.88 + X.89 + X.90 + X.91 + 
    X.92 + X.93 + X.94 + X.95 + X.96 + X.97 + X.98 + X.99 + X.100
  Res.Df    RSS  Df Sum of Sq      F Pr(>F)
1    999 973.65                            
2    899 888.44 100    85.208 0.8622  0.825
\end{CodeOutput}
\begin{CodeInput}
R> summary(m.null)
\end{CodeInput}
\begin{CodeOutput}

Call:
lm(formula = y ~ 1, data = D)

Residuals:
    Min      1Q  Median      3Q     Max 
-3.4585 -0.6809  0.0190  0.6365  2.9346

Coefficients:
            Estimate Std. Error t value Pr(>|t|)
(Intercept)  9.93704    0.03122   318.3   <2e-16

Residual standard error: 0.9872 on 999 degrees of freedom
\end{CodeOutput}
\end{CodeChunk}

Next, using the \texttt{stepAIC()} function in the \textbf{MASS} package
\citep{VenablesRipley:2002}, let us perform a forward stepwise
regression to select a ``best'' model, starting with the null model, and
using AIC as the model-selection criterion (see the help page for
\texttt{stepAIC()} for details):\footnote{It's generally advantageous to
  start with the largest model, here the one with 100 predictors, and
  proceed by backward elimination. In this demonstration, however, where
  all of the \(\beta\)s are really 0, the selected model will be small,
  and so we proceed by forward selection from the null model to save
  computing time.}

\begin{CodeChunk}
\begin{CodeInput}
R> library("MASS")  # for stepAIC()
R> m.select <- stepAIC(m.null,
+                     direction="forward", trace=FALSE,
+                     scope=list(lower=~1, upper=formula(m.full)))
R> summary(m.select)
\end{CodeInput}
\begin{CodeOutput}

Call:
lm(formula = y ~ X.99 + X.90 + X.87 + X.40 + X.65 + X.91 + X.53 + 
    X.45 + X.31 + X.56 + X.61 + X.60 + X.46 + X.35 + X.92, data = D)

Residuals:
    Min      1Q  Median      3Q     Max 
-3.2620 -0.6446  0.0236  0.6406  3.1180

Coefficients:
            Estimate Std. Error t value Pr(>|t|)
(Intercept)  9.93716    0.03098 320.803  < 2e-16
X.99        -0.09103    0.03082  -2.953  0.00322
X.90        -0.08205    0.03135  -2.617  0.00901
X.87        -0.06942    0.03105  -2.235  0.02561
X.40        -0.04759    0.03076  -1.547  0.12211
X.65        -0.05523    0.03147  -1.755  0.07952
X.91         0.05245    0.03084   1.700  0.08937
X.53        -0.04921    0.03048  -1.615  0.10672
X.45         0.05543    0.03182   1.742  0.08183
X.31         0.04525    0.03108   1.456  0.14570
X.56         0.05433    0.03273   1.660  0.09723
X.61        -0.05085    0.03170  -1.604  0.10908
X.60        -0.05133    0.03194  -1.607  0.10832
X.46         0.05158    0.03272   1.576  0.11529
X.35         0.04696    0.03146   1.493  0.13584
X.92         0.04430    0.03100   1.429  0.15329

Residual standard error: 0.9725 on 984 degrees of freedom
Multiple R-squared:  0.04419,   Adjusted R-squared:  0.02962 
F-statistic: 3.033 on 15 and 984 DF,  p-value: 8.338e-05
\end{CodeOutput}
\begin{CodeInput}
R> mse(D$y, fitted(m.select))
\end{CodeInput}
\begin{CodeOutput}
[1] 0.9306254
attr(,"casewise loss")
[1] "(y - yhat)^2"
\end{CodeOutput}
\end{CodeChunk}

The resulting model has 15 predictors, a very modest \(R^2 = .044\), but
a small \(p\)-value for its omnibus \(F\)-test (which, of course, is
entirely spurious because the same data were used to select and test the
model). The MSE for the selected model is smaller than the true error
variance \(\sigma^2 = 1\), as is the estimated error variance for the
selected model, \(\widehat{\sigma}^2 = 0.973^2 = 0.947\).

If we cross-validate the selected model, we also obtain an optimistic
estimate of its predictive power (although the confidence interval for
the bias-adjusted MSE includes 1):

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.select, seed=2529)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 2529
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 0.9593695
bias-adjusted cross-validation criterion = 0.9578478
95% CI for bias-adjusted CV criterion = (0.8766138, 1.039082)
full-sample criterion = 0.9306254 
\end{CodeOutput}
\end{CodeChunk}

The \texttt{cvSelect()} function in the \textbf{cv} package allows us to
cross-validate the whole model-selection procedure. The first argument
to \texttt{cvSelect()} is a model-selection function capable of
refitting the model with a fold omitted and returning a CV criterion.
The \texttt{selectStepAIC()} function, also in \textbf{cv} and based on
\texttt{stepAIC()}, is suitable for use with \texttt{cvSelect()}:

\begin{CodeChunk}
\begin{CodeInput}
R> cv.select <- cvSelect(selectStepAIC, data=D, seed=3791,
+                       model=m.null, direction="forward",
+                       scope=list(lower=~1, 
+                                  upper=formula(m.full)))
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 3791
\end{CodeOutput}
\begin{CodeInput}
R> cv.select
\end{CodeInput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 1.06873
bias-adjusted cross-validation criterion = 1.061183
95% CI for bias-adjusted CV criterion = (0.9717229, 1.150642)
full-sample criterion = 0.9306254 
\end{CodeOutput}
\end{CodeChunk}

The other arguments to \texttt{cvSelect()} are:

\begin{itemize}
\tightlist
\item
  \texttt{data}, the data set to which the model is fit;
\item
  \texttt{seed}, an optional seed for R's pseudo-random-number
  generator; as for \texttt{cv()}, if the seed isn't supplied by the
  user, a seed is randomly selected and saved;
\item
  additional arguments required by the model-selection function, here
  the starting \texttt{model} argument, the \texttt{direction} of model
  selection, and the \texttt{scope} of models considered (from the model
  with only a regression constant to the model with all 100 predictors).
\end{itemize}

By default, \texttt{cvSelect()} performs 10-fold CV, and produces an
estimate of MSE for the model-selection procedure even \emph{larger}
than the true error variance, \(\sigma^2 = 1\).

Also by default, when the number of folds is 10 or fewer,
\texttt{cvSelect()} saves the coefficients of the selected models. In
this example, the \texttt{compareFolds()} function reveals that the
variables retained by the model-selection process in the several folds
are quite different:

\begin{CodeChunk}
\begin{CodeInput}
R> compareFolds(cv.select)
\end{CodeInput}
\begin{CodeOutput}
        (Intercept)    X.87    X.90    X.99    X.91    X.54    X.53    X.56
Fold 1       9.9187 -0.0615 -0.0994 -0.0942  0.0512  0.0516                
Fold 2       9.9451 -0.0745 -0.0899 -0.0614          0.0587          0.0673
Fold 3       9.9423 -0.0783 -0.0718 -0.0987  0.0601                  0.0512
Fold 4       9.9410 -0.0860 -0.0831 -0.0867  0.0570         -0.0508        
Fold 5       9.9421 -0.0659 -0.0849 -0.1004  0.0701  0.0511 -0.0487  0.0537
Fold 6       9.9633 -0.0733 -0.0874 -0.0960  0.0555  0.0629 -0.0478        
Fold 7       9.9279 -0.0618 -0.0960 -0.0838  0.0533         -0.0464        
Fold 8       9.9453 -0.0610 -0.0811 -0.0818          0.0497 -0.0612  0.0560
Fold 9       9.9173 -0.0663 -0.0894 -0.1100  0.0504  0.0524          0.0747
Fold 10      9.9449 -0.0745 -0.0906 -0.0891  0.0535  0.0482 -0.0583  0.0642
           X.40    X.45    X.65    X.68    X.92    X.15    X.26    X.46    X.60
Fold 1                  -0.0590                 -0.0456  0.0658  0.0608        
Fold 2                                   0.0607          0.0487                
Fold 3  -0.0496         -0.0664          0.0494                                
Fold 4  -0.0597  0.0579 -0.0531          0.0519 -0.0566                 -0.0519
Fold 5                           0.0587                          0.0527 -0.0603
Fold 6  -0.0596  0.0552          0.0474                                        
Fold 7           0.0572          0.0595                                        
Fold 8           0.0547 -0.0617  0.0453  0.0493 -0.0613  0.0591  0.0703 -0.0588
Fold 9  -0.0552  0.0573 -0.0635  0.0492         -0.0513  0.0484         -0.0507
Fold 10 -0.0558                          0.0529                  0.0710        
           X.61     X.8    X.28    X.29    X.31    X.35    X.70    X.89    X.17
Fold 1  -0.0490          0.0616 -0.0537                  0.0638                
Fold 2           0.0671                  0.0568                  0.0523        
Fold 3  -0.0631          0.0616                                                
Fold 4           0.0659         -0.0549          0.0527                  0.0527
Fold 5           0.0425                  0.0672  0.0613          0.0493        
Fold 6           0.0559         -0.0629  0.0498          0.0487                
Fold 7                                                           0.0611  0.0472
Fold 8  -0.0719                                          0.0586                
Fold 9                   0.0525                                                
Fold 10 -0.0580                                  0.0603                        
           X.25     X.4    X.64    X.81    X.97    X.11     X.2    X.33    X.47
Fold 1                                   0.0604          0.0575                
Fold 2   0.0478          0.0532  0.0518                                        
Fold 3                           0.0574                          0.0473        
Fold 4                   0.0628                                                
Fold 5   0.0518                                                                
Fold 6                                           0.0521                        
Fold 7           0.0550                                                        
Fold 8                                                                         
Fold 9                                   0.0556                          0.0447
Fold 10          0.0516                                                        
            X.6    X.72    X.73    X.77    X.79 X.88
Fold 1   0.0476                                     
Fold 2                   0.0514                     
Fold 3                                              
Fold 4                                  -0.0473     
Fold 5           0.0586                         0.07
Fold 6                          -0.0489             
Fold 7                                              
Fold 8                                              
Fold 9                                              
Fold 10                                             
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{mrozs-logistic-regression-revisited}{%
\subsection{Mroz's logistic regression
revisited}\label{mrozs-logistic-regression-revisited}}

For a contrasting example we apply model selection to Mroz's logistic
regression for married women's labor-force participation. First, recall
the logistic regression model that we fit to the \texttt{Mroz} data:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(m.mroz)
\end{CodeInput}
\begin{CodeOutput}

Call:
glm(formula = lfp ~ ., family = binomial, data = Mroz)

Coefficients:
             Estimate Std. Error z value Pr(>|z|)
(Intercept)  3.182140   0.644375   4.938 7.88e-07
k5          -1.462913   0.197001  -7.426 1.12e-13
k618        -0.064571   0.068001  -0.950 0.342337
age         -0.062871   0.012783  -4.918 8.73e-07
wcyes        0.807274   0.229980   3.510 0.000448
hcyes        0.111734   0.206040   0.542 0.587618
lwg          0.604693   0.150818   4.009 6.09e-05
inc         -0.034446   0.008208  -4.196 2.71e-05

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 1029.75  on 752  degrees of freedom
Residual deviance:  905.27  on 745  degrees of freedom
AIC: 921.27

Number of Fisher Scoring iterations: 4
\end{CodeOutput}
\end{CodeChunk}

Applying stepwise model selection Mroz's logistic regression, using BIC
as the model-selection criterion (via the argument
\texttt{k=log(nrow(Mroz))} to \texttt{stepAIC()}) selects 5 of the 7
original predictors:

\begin{CodeChunk}
\begin{CodeInput}
R> m.mroz.sel <- stepAIC(m.mroz, k=log(nrow(Mroz)),
+                       trace=FALSE)
R> summary(m.mroz.sel)
\end{CodeInput}
\begin{CodeOutput}

Call:
glm(formula = lfp ~ k5 + age + wc + lwg + inc, family = binomial, 
    data = Mroz)

Coefficients:
            Estimate Std. Error z value Pr(>|z|)
(Intercept)  2.90193    0.54290   5.345 9.03e-08
k5          -1.43180    0.19320  -7.411 1.25e-13
age         -0.05853    0.01142  -5.127 2.94e-07
wcyes        0.87237    0.20639   4.227 2.37e-05
lwg          0.61568    0.15014   4.101 4.12e-05
inc         -0.03367    0.00780  -4.317 1.58e-05

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 1029.75  on 752  degrees of freedom
Residual deviance:  906.46  on 747  degrees of freedom
AIC: 918.46

Number of Fisher Scoring iterations: 3
\end{CodeOutput}
\begin{CodeInput}
R> BayesRule(Mroz$lfp == "yes",
+           predict(m.mroz.sel, type="response"))
\end{CodeInput}
\begin{CodeOutput}
[1] 0.3187251
attr(,"casewise loss")
[1] "y != round(yhat)"
\end{CodeOutput}
\end{CodeChunk}

Bayes rule applied to the selected model misclassifies 32\% of the cases
in the \texttt{Mroz} data.

Cross-validating the selected model produces a similar, slightly larger,
estimate of misclassification, about 33\%:

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.mroz.sel, criterion=BayesRule, seed=345266)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 345266
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: exact
criterion: BayesRule
cross-validation criterion = 0.3306773
bias-adjusted cross-validation criterion = 0.3333245
95% CI for bias-adjusted CV criterion = (0.2996998, 0.3669493)
full-sample criterion = 0.3187251 
\end{CodeOutput}
\end{CodeChunk}

Is this estimate of predictive performance optimistic?

We proceed to apply the model-selection procedure by cross-validation,
producing more or less the same result:

\begin{CodeChunk}
\begin{CodeInput}
R> m.mroz.sel.cv <- cvSelect(selectStepAIC, Mroz, 
+                           seed=6681,
+                           criterion=BayesRule,
+                           model=m.mroz,
+                           AIC=FALSE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 6681
\end{CodeOutput}
\begin{CodeInput}
R> m.mroz.sel.cv
\end{CodeInput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 0.3306773
bias-adjusted cross-validation criterion = 0.3345203
95% CI for bias-adjusted CV criterion = (0.3008955, 0.368145)
full-sample criterion = 0.3187251 
\end{CodeOutput}
\end{CodeChunk}

Setting \texttt{AIC=FALSE} in the call to \texttt{cvSelect()} uses the
BIC rather than the AIC as the model-selection criterion. As it turns
out, exactly the same predictors are selected when each of the 10 folds
are omitted, and the several coefficient estimates are very similar, as
we show using \texttt{compareFolds()}:

\begin{CodeChunk}
\begin{CodeInput}
R> compareFolds(m.mroz.sel.cv)
\end{CodeInput}
\begin{CodeOutput}
        (Intercept)     age     inc      k5     lwg wcyes
Fold 1       2.5014 -0.0454 -0.0388 -1.3613  0.5653  0.85
Fold 2       3.0789 -0.0659 -0.0306 -1.5335  0.6923  0.79
Fold 3       3.0141 -0.0595 -0.0305 -1.3994  0.5428  0.86
Fold 4       2.7251 -0.0543 -0.0354 -1.4474  0.6298  1.09
Fold 5       2.7617 -0.0566 -0.0320 -1.4752  0.6324  0.74
Fold 6       3.0234 -0.0621 -0.0348 -1.4537  0.6618  0.94
Fold 7       2.9615 -0.0600 -0.0351 -1.4127  0.5835  0.97
Fold 8       2.9598 -0.0603 -0.0329 -1.3865  0.6210  0.69
Fold 9       3.2481 -0.0650 -0.0381 -1.4138  0.6093  0.94
Fold 10      2.7724 -0.0569 -0.0295 -1.4503  0.6347  0.85
\end{CodeOutput}
\end{CodeChunk}

In this example, therefore, we appear to obtain a realistic estimate of
model performance directly from the selected model, because there is
little added uncertainty induced by model selection.

\hypertarget{cross-validating-choice-of-transformations-in-regression}{%
\subsection{Cross-validating choice of transformations in
regression}\label{cross-validating-choice-of-transformations-in-regression}}

The \textbf{cv} package also provides a \texttt{cvSelect()} procedure,
\texttt{selectTrans()}, for choosing transformations of the predictors
and the response in regression.

Some background: As \citet[Sec. 8.2]{Weisberg:2014} explains, there are
technical advantages to having (numeric) predictors in linear regression
analysis that are themselves linearly related. If the predictors
\emph{aren't} linearly related, then the relationships between them can
often be straightened by power transformations. Transformations can be
selected after graphical examination of the data, or by analytic
methods. Once the relationships between the predictors are linearized,
it can be advantageous similarly to transform the response variable
towards normality.

Selecting transformations analytically raises the possibility of
automating the process, as would be required for cross-validation. One
could, in principle, apply graphical methods to select transformations
for each fold, but because a data analyst couldn't forget the choices
made for previous folds, the process wouldn't really be applied
independently to the folds.

To illustrate, we adapt an example appearing in several places in
\citet{FoxWeisberg:2019} (for example in Chapter 3 on transforming
data), using data on the prestige and other characteristics of 102
Canadian occupations circa 1970. The data are in the \texttt{Prestige}
data frame in the \textbf{carData} package:

\begin{CodeChunk}
\begin{CodeInput}
R> data("Prestige", package="carData")
R> head(Prestige)
\end{CodeInput}
\begin{CodeOutput}
                    education income women prestige census type
gov.administrators      13.11  12351 11.16     68.8   1113 prof
general.managers        12.26  25879  4.02     69.1   1130 prof
accountants             12.77   9271 15.70     63.4   1171 prof
purchasing.officers     11.42   8865  9.11     56.8   1175 prof
chemists                14.62   8403 11.68     73.5   2111 prof
physicists              15.64  11030  5.13     77.6   2113 prof
\end{CodeOutput}
\begin{CodeInput}
R> summary(Prestige)
\end{CodeInput}
\begin{CodeOutput}
   education          income          women           prestige    
 Min.   : 6.380   Min.   :  611   Min.   : 0.000   Min.   :14.80  
 1st Qu.: 8.445   1st Qu.: 4106   1st Qu.: 3.592   1st Qu.:35.23  
 Median :10.540   Median : 5930   Median :13.600   Median :43.60  
 Mean   :10.738   Mean   : 6798   Mean   :28.979   Mean   :46.83  
 3rd Qu.:12.648   3rd Qu.: 8187   3rd Qu.:52.203   3rd Qu.:59.27  
 Max.   :15.970   Max.   :25879   Max.   :97.510   Max.   :87.20  
     census       type   
 Min.   :1113   bc  :44  
 1st Qu.:3120   prof:31  
 Median :5135   wc  :23  
 Mean   :5402   NA's: 4  
 3rd Qu.:8312            
 Max.   :9517            
\end{CodeOutput}
\end{CodeChunk}

The variables in the \texttt{Prestige} data set are:

\begin{itemize}
\tightlist
\item
  \texttt{education}: average years of education for incumbents in the
  occupation, from the 1971 Canadian Census.
\item
  \texttt{income}: average dollars of annual income for the occupation,
  from the Census.
\item
  \texttt{women}: percentage of occupational incumbents who were women,
  also from the Census.
\item
  \texttt{prestige}: the average prestige rating of the occupation on a
  0--100 ``thermometer'' scale, in a Canadian social survey conducted
  around the same time.
\item
  \texttt{type}, type of occupation, and \texttt{census}, the Census
  occupational code, which are not used in our example.
\end{itemize}

The object of a regression analysis for the \texttt{Prestige} data (and
their original purpose) is to predict occupational prestige from the
other variables in the data set.

A scatterplot matrix (using the \texttt{scatterplotMatrix()} function in
the \textbf{car} package) of the numeric variables in the data reveals
that the distributions of \texttt{income} and \texttt{women} are
positively skewed, and that some of the relationships among the three
predictors, and between the predictors and the response (i.e.,
\texttt{prestige}), are nonlinear:

\begin{CodeChunk}
\begin{CodeInput}
R> library("car")
\end{CodeInput}
\begin{CodeOutput}
Loading required package: carData
\end{CodeOutput}
\begin{CodeInput}
R> scatterplotMatrix(~ prestige + income + education + women,
+                   data=Prestige, smooth=list(spread=FALSE))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/scatterplot-matrix-1}

}

\caption[Scatterplot matrix for the `Prestige` data]{Scatterplot matrix for the `Prestige` data.}\label{fig:scatterplot-matrix}
\end{figure}
\end{CodeChunk}

The \texttt{powerTransform()} function in the \textbf{car} package
transforms variables towards multivariate normality by a generalization
of Box and Cox's maximum-likelihood-like approach \citep{BoxCox:1964}.
Several ``families'' of power transformations can be used, including the
original Box-Cox family, simple powers (and roots), and two adaptations
of the Box-Cox family to data that may include negative values and
zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see
\citet[Chap. 8]{Weisberg:2014}, and \citet[Chap. 3]{FoxWeisberg:2019}
for details. Because \texttt{women} has some zero values, we use the
Yeo-Johnson family:

\begin{CodeChunk}
\begin{CodeInput}
R> trans <- powerTransform( cbind(income, education, women) ~ 1,
+                          data=Prestige, family="yjPower")
R> summary(trans)
\end{CodeInput}
\begin{CodeOutput}
yjPower Transformations to Multinormality 
          Est Power Rounded Pwr Wald Lwr Bnd Wald Upr Bnd
income       0.2678        0.33       0.1051       0.4304
education    0.5162        1.00      -0.2822       1.3145
women        0.1630        0.16       0.0112       0.3149

 Likelihood ratio test that all transformation parameters are equal to 0
                               LRT df      pval
LR test, lambda = (0 0 0) 15.73879  3 0.0012827
\end{CodeOutput}
\end{CodeChunk}

We thus have evidence of the desirability of transforming
\texttt{income} (by the \(1/3\) power) and \texttt{women} (by the
\(0.16\) power---which is close to the ``0'' power, i.e., the log
transformation), but not \texttt{education}. Applying the ``rounded''
power transformations makes the predictors better-behaved:

\begin{CodeChunk}
\begin{CodeInput}
R> P <- Prestige[, c("prestige", "income", "education", "women")]
R> (lambdas <- trans$roundlam)
\end{CodeInput}
\begin{CodeOutput}
   income education     women 
0.3300000 1.0000000 0.1630182 
\end{CodeOutput}
\begin{CodeInput}
R> names(lambdas) <- c("income", "education", "women")
R> for (var in c("income", "education", "women")){
+   P[, var] <- yjPower(P[, var], lambda=lambdas[var])
+ }
R> summary(P)
\end{CodeInput}
\begin{CodeOutput}
    prestige         income        education          women      
 Min.   :14.80   Min.   :22.15   Min.   : 6.380   Min.   :0.000  
 1st Qu.:35.23   1st Qu.:44.17   1st Qu.: 8.445   1st Qu.:1.731  
 Median :43.60   Median :50.26   Median :10.540   Median :3.362  
 Mean   :46.83   Mean   :50.76   Mean   :10.738   Mean   :3.502  
 3rd Qu.:59.27   3rd Qu.:56.24   3rd Qu.:12.648   3rd Qu.:5.591  
 Max.   :87.20   Max.   :83.62   Max.   :15.970   Max.   :6.830  
\end{CodeOutput}
\begin{CodeInput}
R> scatterplotMatrix(~ prestige + income + education + women,
+                   data=P, smooth=list(spread=FALSE))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/transformed-predictors-1}

}

\caption[Scatterplot matrix for the `Prestige` data with the predictors transformed]{Scatterplot matrix for the `Prestige` data with the predictors transformed.}\label{fig:transformed-predictors}
\end{figure}
\end{CodeChunk}

Comparing the MSE for the regressions with the original and transformed
predictors shows a advantage to the latter:

\begin{CodeChunk}
\begin{CodeInput}
R> m.pres <- lm(prestige ~ income + education + women, data=Prestige)
R> m.pres.trans <- lm(prestige ~ income + education + women, data=P)
R> mse(Prestige$prestige, fitted(m.pres))
\end{CodeInput}
\begin{CodeOutput}
[1] 59.15265
attr(,"casewise loss")
[1] "(y - yhat)^2"
\end{CodeOutput}
\begin{CodeInput}
R> mse(P$prestige, fitted(m.pres.trans))
\end{CodeInput}
\begin{CodeOutput}
[1] 50.60016
attr(,"casewise loss")
[1] "(y - yhat)^2"
\end{CodeOutput}
\end{CodeChunk}

Similarly, component+residual plots for the two regressions, produced by
the \texttt{crPlots()} function in the \textbf{car} package, suggest
that the partial relationship of \texttt{prestige} to \texttt{income} is
more nearly linear in the transformed data, but the transformation of
\texttt{women} fails to capture what appears to be a slight quadratic
partial relationship; the partial relationship of \texttt{prestige} to
\texttt{education} is close to linear in both regressions:

\begin{CodeChunk}
\begin{CodeInput}
R> crPlots(m.pres)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{JSS-article_files/figure-latex/CR-plots-untransformed-1}

}

\caption[Component+residual plots for the `Prestige` regression with the original predictors]{Component+residual plots for the `Prestige` regression with the original predictors.}\label{fig:CR-plots-untransformed}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> crPlots(m.pres.trans)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{JSS-article_files/figure-latex/CR-plots-transformed-1}

}

\caption[Component+residual plots for the `Prestige` regression with transformed predictors]{Component+residual plots for the `Prestige` regression with transformed predictors.}\label{fig:CR-plots-transformed}
\end{figure}
\end{CodeChunk}

Having transformed the predictors towards multinormality, we now
consider whether there's evidence for transforming the response (using
\texttt{powerTransform()} for Box and Cox's original method), and we
discover that there's not:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(powerTransform(m.pres.trans))
\end{CodeInput}
\begin{CodeOutput}
bcPower Transformation to Normality 
   Est Power Rounded Pwr Wald Lwr Bnd Wald Upr Bnd
Y1    1.0194           1       0.6773       1.3615

Likelihood ratio test that transformation parameter is equal to 0
 (log transformation)
                          LRT df       pval
LR test, lambda = (0) 32.2174  1 1.3785e-08

Likelihood ratio test that no transformation is needed
                             LRT df    pval
LR test, lambda = (1) 0.01238421  1 0.91139
\end{CodeOutput}
\end{CodeChunk}

The \texttt{selectTrans()} function in the \textbf{cv} package automates
the process of selecting predictor and response transformations. The
function takes a \texttt{data} set and ``working'' \texttt{model} as
arguments, along with the candidate \texttt{predictors} and
\texttt{response} for transformation, and the transformation
\texttt{family} to employ. If the \texttt{predictors} argument is
missing then only the response is transformed, and if the
\texttt{response} argument is missing, only the supplied predictors are
transformed. The default \texttt{family} for transforming the predictors
is \texttt{"bcPower"}---the original Box-Cox family---as is the default
\texttt{family.y} for transforming the response; here we specify
\texttt{family="yjPower} because of the zeros in \texttt{women}.
\texttt{selectTrans()} returns the result of applying a lack-of-fit
criterion to the model after the selected transformation is applied,
with the default \texttt{criterion=mse}:

\begin{CodeChunk}
\begin{CodeInput}
R> selectTrans(data=Prestige, model=m.pres,
+             predictors=c("income", "education", "women"),
+             response="prestige", family="yjPower")
\end{CodeInput}
\begin{CodeOutput}
[1] 50.60016
attr(,"casewise loss")
[1] "(y - yhat)^2"
\end{CodeOutput}
\end{CodeChunk}

\texttt{selectTrans()} also takes an optional \texttt{indices} argument,
making it suitable for doing computations on a subset of the data (i.e.,
a CV fold), and hence for use with \texttt{cvSelect()} (see
\texttt{?selectTrans} for details):

\begin{CodeChunk}
\begin{CodeInput}
R> cvs <- cvSelect(selectTrans, data=Prestige, model=m.pres, seed=1463,
+                 predictors=c("income", "education", "women"),
+                 response="prestige",
+                 family="yjPower")
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 1463
\end{CodeOutput}
\begin{CodeInput}
R> cvs
\end{CodeInput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 54.4871
bias-adjusted cross-validation criterion = 54.30824
full-sample criterion = 50.60016 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.pres, seed=1463) # untransformed model with same folds
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 1463
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 63.2926
bias-adjusted cross-validation criterion = 63.07251
full-sample criterion = 59.15265 
\end{CodeOutput}
\begin{CodeInput}
R> compareFolds(cvs)
\end{CodeInput}
\begin{CodeOutput}
        lam.education lam.income lam.women lambda
Fold 1          1.000      0.330     0.330      1
Fold 2          1.000      0.330     0.169      1
Fold 3          1.000      0.330     0.330      1
Fold 4          1.000      0.330     0.330      1
Fold 5          1.000      0.330     0.000      1
Fold 6          1.000      0.330     0.330      1
Fold 7          1.000      0.330     0.330      1
Fold 8          1.000      0.330     0.000      1
Fold 9          1.000      0.330     0.000      1
Fold 10         1.000      0.330     0.000      1
\end{CodeOutput}
\end{CodeChunk}

The results suggest that the predictive power of the transformed
regression is reliably greater than that of the untransformed regression
(though in both case, the cross-validated MSE is considerably higher
than the MSE computed for the whole data). Examining the selected
transformations for each fold reveals that the predictor
\texttt{education} and the response \texttt{prestige} are never
transformed; that the \(1/3\) power is selected for \texttt{income} in
all of the folds; and that the transformation selected for
\texttt{women} varies narrowly across the folds between the \(0\)th
power (i.e., log) and the \(1/3\) power.

\hypertarget{selecting-both-transformations-and-predictorsvenables}{%
\subsection[Selecting both transformations and
predictors]{\texorpdfstring{Selecting both transformations and
predictors\footnote{The presentation in the section benefits from an
  email conversation with Bill Venables, who of course isn't responsible
  for the use to which we've put his insightful remarks.}}{Selecting both transformations and predictors}}\label{selecting-both-transformations-and-predictorsvenables}}

As we mentioned, \citet[Sec. 7.10.2: ``The Wrong and Right Way to Do
Cross-validation'']{HastieTibshiraniFriedman:2009} explain that honest
cross-validation has to take account of model specification and
selection. Statistical modeling is at least partly a craft, and one
could imagine applying that craft to successive partial data sets, each
with a fold removed. The resulting procedure would be tedious, though
possibly worth the effort, but it would also be difficult to realize in
practice: After all, we can hardly erase our memory of statistical
modeling choices between analyzing partial data sets.

Alternatively, if we're able to automate the process of model selection,
then we can more realistically apply CV mechanically. That's what we did
in the preceding two sections, first for predictor selection and then
for selection of transformations in regression. In this section, we
consider the case where we both select variable transformations and then
proceed to select predictors. It's insufficient to apply these steps
sequentially, first, for example, using \texttt{cvSelect()} with
\texttt{selectTrans()} and then with \texttt{selectStepAIC()}; rather we
should apply the whole model-selection procedure with each fold omitted.
The \texttt{selectTransAndStepAIC()} function, also supplied by the
\textbf{cv} package, does exactly that.

To illustrate this process, we return to the \texttt{Auto} data set:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(Auto)
\end{CodeInput}
\begin{CodeOutput}
      mpg          cylinders      displacement     horsepower        weight    
 Min.   : 9.00   Min.   :3.000   Min.   : 68.0   Min.   : 46.0   Min.   :1613  
 1st Qu.:17.00   1st Qu.:4.000   1st Qu.:105.0   1st Qu.: 75.0   1st Qu.:2225  
 Median :22.75   Median :4.000   Median :151.0   Median : 93.5   Median :2804  
 Mean   :23.45   Mean   :5.472   Mean   :194.4   Mean   :104.5   Mean   :2978  
 3rd Qu.:29.00   3rd Qu.:8.000   3rd Qu.:275.8   3rd Qu.:126.0   3rd Qu.:3615  
 Max.   :46.60   Max.   :8.000   Max.   :455.0   Max.   :230.0   Max.   :5140

  acceleration        year           origin                      name    
 Min.   : 8.00   Min.   :70.00   Min.   :1.000   amc matador       :  5  
 1st Qu.:13.78   1st Qu.:73.00   1st Qu.:1.000   ford pinto        :  5  
 Median :15.50   Median :76.00   Median :1.000   toyota corolla    :  5  
 Mean   :15.54   Mean   :75.98   Mean   :1.577   amc gremlin       :  4  
 3rd Qu.:17.02   3rd Qu.:79.00   3rd Qu.:2.000   amc hornet        :  4  
 Max.   :24.80   Max.   :82.00   Max.   :3.000   chevrolet chevette:  4  
                                                 (Other)           :365  
\end{CodeOutput}
\begin{CodeInput}
R> xtabs(~ year, data=Auto)
\end{CodeInput}
\begin{CodeOutput}
year
70 71 72 73 74 75 76 77 78 79 80 81 82 
29 27 28 40 26 30 34 28 36 29 27 28 30 
\end{CodeOutput}
\begin{CodeInput}
R> xtabs(~ origin, data=Auto)
\end{CodeInput}
\begin{CodeOutput}
origin
  1   2   3 
245  68  79 
\end{CodeOutput}
\begin{CodeInput}
R> xtabs(~ cylinders, data=Auto)
\end{CodeInput}
\begin{CodeOutput}
cylinders
  3   4   5   6   8 
  4 199   3  83 103 
\end{CodeOutput}
\end{CodeChunk}

We previously used the \texttt{Auto} here in a preliminary example where
we employed CV to inform the selection of the order of a polynomial
regression of \texttt{mpg} on \texttt{horsepower}. Here, we consider
more generally the problem of predicting \texttt{mpg} from the other
variables in the \texttt{Auto} data. We begin with a bit of data
management, and then examine the pairwise relationships among the
numeric variables in the data set:

\begin{CodeChunk}
\begin{CodeInput}
R> Auto$cylinders <- factor(Auto$cylinders,
+                          labels=c("3.4", "3.4", "5.6", "5.6", "8"))
R> Auto$year <- as.factor(Auto$year)
R> Auto$origin <- factor(Auto$origin,
+                       labels=c("America", "Europe", "Japan"))
R> rownames(Auto) <- make.names(Auto$name, unique=TRUE)
R> Auto$name <- NULL
R> 
R> scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
+                   smooth=list(spread=FALSE), data=Auto)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/Auto-explore-1}

}

\caption[Scatterplot matrix for the numeric variables in the `Auto` data]{Scatterplot matrix for the numeric variables in the `Auto` data}\label{fig:Auto-explore}
\end{figure}
\end{CodeChunk}

A comment before we proceed: \texttt{origin} is clearly categorical and
so converting it to a factor is natural, but we could imagine treating
\texttt{cylinders} and \texttt{year} as numeric predictors. There are,
however, only 5 distinct values of \texttt{cylinders} (ranging from 3 to
8), but cars with 3 or 5 cylinders are rare. and none of the cars has 7
cylinders. There are similarly only 13 distinct years between 1970 and
1982 in the data, and the relationship between \texttt{mpg} and
\texttt{year} is difficult to characterize.\footnote{Of course, making
  the decision to treat \texttt{year} as a factor on this basis could be
  construed as cheating in the current context, which illustrates the
  difficulty of automating the whole model-selection process. It's
  rarely desirable, in our opinion, to forgo exploration of the data to
  ensure the purity of model validation. We believe, however, that it's
  still useful to automate as much of the process as we can to obtain a
  more realistic, if still biased, estimate of the predictive power of a
  model.} It's apparent that most these variables are positively skewed
and that many of the pairwise relationships among them are nonlinear.

We begin with a ``working model'' that specifies linear partial
relationships of the response to the numeric predictors:

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto <- lm(mpg ~ ., data = Auto)
R> summary(m.auto)
\end{CodeInput}
\begin{CodeOutput}

Call:
lm(formula = mpg ~ ., data = Auto)

Residuals:
    Min      1Q  Median      3Q     Max 
-9.0064 -1.7450 -0.0917  1.5251 10.9504

Coefficients:
               Estimate Std. Error t value Pr(>|t|)
(Intercept)  37.0341323  1.9693933  18.805  < 2e-16
cylinders5.6 -2.6029412  0.6552000  -3.973 8.54e-05
cylinders8   -0.5824578  1.1714516  -0.497 0.619335
displacement  0.0174253  0.0067340   2.588 0.010043
horsepower   -0.0413534  0.0133786  -3.091 0.002145
weight       -0.0055479  0.0006323  -8.774  < 2e-16
acceleration  0.0615272  0.0883132   0.697 0.486431
year71        0.9680584  0.8373899   1.156 0.248408
year72       -0.6014345  0.8251155  -0.729 0.466517
year73       -0.6876890  0.7402723  -0.929 0.353510
year74        1.3755758  0.8765000   1.569 0.117408
year75        0.9299288  0.8590716   1.082 0.279742
year76        1.5598929  0.8225051   1.897 0.058669
year77        2.9094161  0.8417285   3.456 0.000611
year78        3.1751976  0.7989396   3.974 8.48e-05
year79        5.0192987  0.8457587   5.935 6.76e-09
year80        9.0997634  0.8972933  10.141  < 2e-16
year81        6.6886597  0.8852181   7.556 3.28e-13
year82        8.0711248  0.8706683   9.270  < 2e-16
originEurope  2.0466642  0.5171236   3.958 9.07e-05
originJapan   2.1448874  0.5077169   4.225 3.02e-05

Residual standard error: 2.924 on 371 degrees of freedom
Multiple R-squared:  0.8668,    Adjusted R-squared:  0.8596 
F-statistic: 120.7 on 20 and 371 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> Anova(m.auto)
\end{CodeInput}
\begin{CodeOutput}
Anova Table (Type II tests)

Response: mpg
             Sum Sq  Df F value    Pr(>F)
cylinders     292.3   2 17.0915 7.935e-08
displacement   57.3   1  6.6959  0.010043
horsepower     81.7   1  9.5544  0.002145
weight        658.3   1 76.9801 < 2.2e-16
acceleration    4.2   1  0.4854  0.486431
year         3016.8  12 29.3987 < 2.2e-16
origin        190.3   2 11.1287 2.024e-05
Residuals    3172.5 371                  
\end{CodeOutput}
\begin{CodeInput}
R> crPlots(m.auto)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/Auto-working-model-1}

}

\caption[Component+residual plots for the working model fit to the `Auto` data]{Component+residual plots for the working model fit to the `Auto` data}\label{fig:Auto-working-model}
\end{figure}
\end{CodeChunk}

The component+residual plots, created with the \texttt{crPlots()}
function in the previously loaded \textbf{car} package, clearly reveal
the inadequacy of the model.

We proceed to transform the numeric predictors towards multi-normality:

\begin{CodeChunk}
\begin{CodeInput}
R> num.predictors <- c("displacement", "horsepower", "weight", "acceleration")
R> tr.x <- powerTransform(Auto[, num.predictors])
R> summary(tr.x)
\end{CodeInput}
\begin{CodeOutput}
bcPower Transformations to Multinormality 
             Est Power Rounded Pwr Wald Lwr Bnd Wald Upr Bnd
displacement   -0.0509           0      -0.2082       0.1065
horsepower     -0.1249           0      -0.2693       0.0194
weight         -0.0870           0      -0.2948       0.1208
acceleration    0.3061           0      -0.0255       0.6376

Likelihood ratio test that transformation parameters are equal to 0
 (all log transformations)
                                 LRT df    pval
LR test, lambda = (0 0 0 0) 4.872911  4 0.30059

Likelihood ratio test that no transformations are needed
                                 LRT df       pval
LR test, lambda = (1 1 1 1) 390.0777  4 < 2.22e-16
\end{CodeOutput}
\end{CodeChunk}

We then apply the (rounded) transformations---all, as it turns out,
logs---to the data and re-estimate the model:

\begin{CodeChunk}
\begin{CodeInput}
R> A <- Auto
R> powers <- tr.x$roundlam
R> for (pred in num.predictors){
+   A[, pred] <- bcPower(A[, pred], lambda=powers[pred])
+ }
R> head(A)
\end{CodeInput}
\begin{CodeOutput}
                          mpg cylinders displacement horsepower   weight
chevrolet.chevelle.malibu  18         8     5.726848   4.867534 8.161660
buick.skylark.320          15         8     5.857933   5.105945 8.214194
plymouth.satellite         18         8     5.762051   5.010635 8.142063
amc.rebel.sst              16         8     5.717028   5.010635 8.141190
ford.torino                17         8     5.710427   4.941642 8.145840
ford.galaxie.500           15         8     6.061457   5.288267 8.375860
                          acceleration year  origin
chevrolet.chevelle.malibu     2.484907   70 America
buick.skylark.320             2.442347   70 America
plymouth.satellite            2.397895   70 America
amc.rebel.sst                 2.484907   70 America
ford.torino                   2.351375   70 America
ford.galaxie.500              2.302585   70 America
\end{CodeOutput}
\begin{CodeInput}
R> m <- update(m.auto, data=A)
\end{CodeInput}
\end{CodeChunk}

Finally, we perform Box-Cox regression to transform the response (also
obtaining a log transformation):

\begin{CodeChunk}
\begin{CodeInput}
R> summary(powerTransform(m))
\end{CodeInput}
\begin{CodeOutput}
bcPower Transformation to Normality 
   Est Power Rounded Pwr Wald Lwr Bnd Wald Upr Bnd
Y1    0.0024           0      -0.1607       0.1654

Likelihood ratio test that transformation parameter is equal to 0
 (log transformation)
                               LRT df    pval
LR test, lambda = (0) 0.0008015428  1 0.97741

Likelihood ratio test that no transformation is needed
                           LRT df       pval
LR test, lambda = (1) 124.1307  1 < 2.22e-16
\end{CodeOutput}
\begin{CodeInput}
R> m <- update(m, log(mpg) ~ .)
R> summary(m)
\end{CodeInput}
\begin{CodeOutput}

Call:
lm(formula = log(mpg) ~ cylinders + displacement + horsepower + 
    weight + acceleration + year + origin, data = A)

Residuals:
     Min       1Q   Median       3Q      Max 
-0.33412 -0.05774  0.00410  0.06072  0.38081

Coefficients:
             Estimate Std. Error t value Pr(>|t|)
(Intercept)   8.89652    0.35822  24.835  < 2e-16
cylinders5.6 -0.06355    0.02574  -2.469   0.0140
cylinders8   -0.07691    0.03900  -1.972   0.0493
displacement  0.02799    0.05146   0.544   0.5868
horsepower   -0.29010    0.05631  -5.152 4.20e-07
weight       -0.54274    0.08193  -6.624 1.23e-10
acceleration -0.14214    0.05630  -2.525   0.0120
year71        0.02505    0.02891   0.866   0.3869
year72       -0.01680    0.02894  -0.580   0.5620
year73       -0.04257    0.02602  -1.636   0.1026
year74        0.04932    0.03041   1.622   0.1056
year75        0.04715    0.02959   1.594   0.1118
year76        0.07087    0.02845   2.491   0.0132
year77        0.13241    0.02927   4.523 8.21e-06
year78        0.14472    0.02777   5.211 3.13e-07
year79        0.23354    0.02921   7.994 1.67e-14
year80        0.32381    0.03170  10.216  < 2e-16
year81        0.25655    0.03094   8.291 2.10e-15
year82        0.30756    0.03036  10.131  < 2e-16
originEurope  0.04921    0.01955   2.518   0.0122
originJapan   0.04409    0.01947   2.265   0.0241

Residual standard error: 0.1043 on 371 degrees of freedom
Multiple R-squared:  0.9108,    Adjusted R-squared:  0.906 
F-statistic: 189.4 on 20 and 371 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> Anova(m)
\end{CodeInput}
\begin{CodeOutput}
Anova Table (Type II tests)

Response: log(mpg)
             Sum Sq  Df F value    Pr(>F)
cylinders    0.0663   2  3.0521   0.04845
displacement 0.0032   1  0.2959   0.58679
horsepower   0.2885   1 26.5420 4.198e-07
weight       0.4769   1 43.8805 1.229e-10
acceleration 0.0693   1  6.3745   0.01199
year         4.4521  12 34.1339 < 2.2e-16
origin       0.0807   2  3.7128   0.02532
Residuals    4.0325 371                  
\end{CodeOutput}
\end{CodeChunk}

The transformed numeric variables are much better-behaved:

\begin{CodeChunk}
\begin{CodeInput}
R> scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
+                   + acceleration, 
+                   smooth=list(spread=FALSE), data=A)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/Auto-transformed-scatterplot-matrix-1}

}

\caption[Scatterplot matrix for the transformed numeric variables in the `Auto` data]{Scatterplot matrix for the transformed numeric variables in the `Auto` data}\label{fig:Auto-transformed-scatterplot-matrix}
\end{figure}
\end{CodeChunk}

And the partial relationships in the model fit to the transformed data
are much more nearly linear:

\begin{CodeChunk}
\begin{CodeInput}
R> crPlots(m)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/Auto-CR-plots-transformed-1}

}

\caption[Component+residual plots for the model fit to the transformed `Auto` data]{Component+residual plots for the model fit to the transformed `Auto` data}\label{fig:Auto-CR-plots-transformed}
\end{figure}
\end{CodeChunk}

Having transformed both the numeric predictors and the response, we
proceed to use the \texttt{stepAIC()} function in the \textbf{MASS}
package to perform predictor selection, employing the BIC
model-selection criterion (by setting the \texttt{k} argument of
\texttt{stepAIC()} to \(\log(n)\)):

\begin{CodeChunk}
\begin{CodeInput}
R> m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
R> summary(m.step)
\end{CodeInput}
\begin{CodeOutput}

Call:
lm(formula = log(mpg) ~ horsepower + weight + acceleration + 
    year + origin, data = A)

Residuals:
     Min       1Q   Median       3Q      Max 
-0.35230 -0.05682  0.00677  0.06741  0.35861

Coefficients:
              Estimate Std. Error t value Pr(>|t|)
(Intercept)   9.434594   0.261529  36.075  < 2e-16
horsepower   -0.276254   0.056143  -4.921 1.30e-06
weight       -0.609071   0.056003 -10.876  < 2e-16
acceleration -0.131380   0.053195  -2.470  0.01397
year71        0.027984   0.028936   0.967  0.33412
year72       -0.007111   0.028446  -0.250  0.80274
year73       -0.039529   0.026014  -1.520  0.12947
year74        0.052752   0.029986   1.759  0.07936
year75        0.053199   0.029280   1.817  0.07004
year76        0.074317   0.028212   2.634  0.00878
year77        0.137931   0.028875   4.777 2.56e-06
year78        0.145876   0.027529   5.299 1.99e-07
year79        0.236036   0.029080   8.117 6.99e-15
year80        0.335274   0.031148  10.764  < 2e-16
year81        0.262872   0.030555   8.603  < 2e-16
year82        0.323391   0.029608  10.922  < 2e-16
originEurope  0.055818   0.016785   3.326  0.00097
originJapan   0.043554   0.017479   2.492  0.01314

Residual standard error: 0.1049 on 374 degrees of freedom
Multiple R-squared:  0.909, Adjusted R-squared:  0.9049 
F-statistic: 219.8 on 17 and 374 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> Anova(m.step)
\end{CodeInput}
\begin{CodeOutput}
Anova Table (Type II tests)

Response: log(mpg)
             Sum Sq  Df  F value    Pr(>F)
horsepower   0.2663   1  24.2120 1.296e-06
weight       1.3010   1 118.2805 < 2.2e-16
acceleration 0.0671   1   6.0998  0.013965
year         4.7589  12  36.0549 < 2.2e-16
origin       0.1366   2   6.2090  0.002225
Residuals    4.1137 374                   
\end{CodeOutput}
\end{CodeChunk}

The selected model includes three of the numeric predictors,
\texttt{horsepower}, \texttt{weight}, and \texttt{acceleration}, along
with the factors \texttt{year} and \texttt{origin}. We can calculate the
MSE for this model, but we expect that the result will be optimistic
because we used the whole data to help specify the model

\begin{CodeChunk}
\begin{CodeInput}
R> mse(Auto$mpg, exp(fitted(m.step)))
\end{CodeInput}
\begin{CodeOutput}
[1] 6.512144
attr(,"casewise loss")
[1] "(y - yhat)^2"
\end{CodeOutput}
\end{CodeChunk}

This is considerably smaller than the MSE for the original working
model:

\begin{CodeChunk}
\begin{CodeInput}
R> mse(Auto$mpg, fitted(m.auto))
\end{CodeInput}
\begin{CodeOutput}
[1] 8.093171
attr(,"casewise loss")
[1] "(y - yhat)^2"
\end{CodeOutput}
\end{CodeChunk}

A perhaps subtle point is that we compute the MSE for the selected model
on the original \texttt{mpg} response scale rather than the log scale,
so as to make the selected model comparable to the working model. That's
slightly uncomfortable given the skewed distribution of \texttt{mpg}. An
alternative is to use the median absolute error instead of the
mean-squared error, employing the \texttt{medAbsErr()} function from the
\textbf{cv} package:

\begin{CodeChunk}
\begin{CodeInput}
R> medAbsErr(Auto$mpg, exp(fitted(m.step)))
\end{CodeInput}
\begin{CodeOutput}
[1] 1.339604
\end{CodeOutput}
\begin{CodeInput}
R> medAbsErr(Auto$mpg, fitted(m.auto))
\end{CodeInput}
\begin{CodeOutput}
[1] 1.666121
\end{CodeOutput}
\end{CodeChunk}

Now let's use \texttt{cvSelect()} with \texttt{selectTransAndStepAIC()}
to automate and cross-validate the whole model-specification process:

\begin{CodeChunk}
\begin{CodeInput}
R> num.predictors
\end{CodeInput}
\begin{CodeOutput}
[1] "displacement" "horsepower"   "weight"       "acceleration"
\end{CodeOutput}
\begin{CodeInput}
R> cvs <- cvSelect(selectTransStepAIC, data=Auto, seed=76692, model=m.auto,
+                 predictors=num.predictors,
+                 response="mpg", AIC=FALSE, criterion=medAbsErr)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 76692
\end{CodeOutput}
\begin{CodeInput}
R> cvs
\end{CodeInput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 1.495075
full-sample criterion = 1.339604 
\end{CodeOutput}
\begin{CodeInput}
R> compareFolds(cvs)
\end{CodeInput}
\begin{CodeOutput}
        (Intercept) horsepower lam.acceleration lam.displacement lam.horsepower
Fold 1      9.71384   -0.17408          0.50000          0.00000        0.00000
Fold 2      9.21713   -0.31480          0.00000          0.00000        0.00000
Fold 3      9.61824   -0.19248          0.00000          0.00000        0.00000
Fold 4      8.69910   -0.25523          0.50000          0.00000        0.00000
Fold 5      9.14403   -0.14934          0.00000          0.00000        0.00000
Fold 6      9.63481   -0.16739          0.50000          0.00000        0.00000
Fold 7      9.98933   -0.36847          0.00000          0.00000       -0.15447
Fold 8      9.06301   -0.29721          0.00000          0.00000        0.00000
Fold 9      8.88315   -0.22684          0.00000          0.00000        0.00000
Fold 10     9.61727   -0.17086          0.00000          0.00000        0.00000
        lam.weight   lambda   weight   year71   year72   year73   year74
Fold 1     0.00000  0.00000 -0.74636  0.03764 -0.00327 -0.02477  0.05606
Fold 2     0.00000  0.00000 -0.47728  0.02173 -0.01488 -0.03770  0.04312
Fold 3     0.00000  0.00000 -0.72085  0.01128 -0.02569 -0.03872  0.05187
Fold 4     0.00000  0.00000 -0.53846  0.02153 -0.02922 -0.05181  0.04136
Fold 5     0.00000  0.00000 -0.69081  0.02531 -0.01062 -0.04625  0.05039
Fold 6     0.00000  0.00000 -0.74049  0.02456  0.00759 -0.03412  0.06266
Fold 7     0.00000  0.00000 -0.72843  0.02532 -0.01271 -0.04144  0.04568
Fold 8     0.00000  0.00000 -0.46392  0.02702 -0.02041 -0.05605  0.04437
Fold 9     0.00000  0.00000 -0.47136  0.00860 -0.03620 -0.04835  0.01906
Fold 10    0.00000  0.00000 -0.73550  0.02937 -0.00899 -0.03814  0.05408
          year75   year76   year77   year78   year79   year80   year81   year82
Fold 1   0.07080  0.07250  0.14420  0.14281  0.23266  0.35127  0.25635  0.30546
Fold 2   0.04031  0.06718  0.13094  0.14917  0.21871  0.33192  0.26196  0.30943
Fold 3   0.03837  0.06399  0.11593  0.12601  0.20499  0.32821  0.24478  0.29204
Fold 4   0.04072  0.05537  0.12292  0.14083  0.22878  0.32947  0.25140  0.27248
Fold 5   0.05596  0.07044  0.13356  0.14724  0.24675  0.33331  0.26938  0.32594
Fold 6   0.06940  0.07769  0.14211  0.14647  0.23532  0.34761  0.26737  0.33062
Fold 7   0.03614  0.07385  0.12976  0.14040  0.23976  0.33998  0.27652  0.30659
Fold 8   0.06573  0.08135  0.13158  0.13987  0.23011  0.32880  0.25886  0.30538
Fold 9   0.03018  0.05846  0.10536  0.11722  0.20665  0.31533  0.23352  0.29375
Fold 10  0.04881  0.07862  0.14101  0.14313  0.23258  0.35649  0.26214  0.32421
        acceleration displacement cylinders5.6 cylinders8 originEurope
Fold 1                                                                
Fold 2      -0.18909     -0.09197                                     
Fold 3                                                                
Fold 4      -0.03484                  -0.09080   -0.10909             
Fold 5                                                         0.06261
Fold 6                                                                
Fold 7                                                                
Fold 8      -0.17676     -0.10542                                     
Fold 9      -0.14514     -0.13452                                     
Fold 10                                                               
        originJapan
Fold 1             
Fold 2             
Fold 3             
Fold 4             
Fold 5         0.04
Fold 6             
Fold 7             
Fold 8             
Fold 9             
Fold 10            
\end{CodeOutput}
\end{CodeChunk}

Here, as for \texttt{selectTrans()}, the \texttt{predictors} and
\texttt{response} arguments specify candidate variables for
transformation, and \texttt{AIC=FALSE} uses the BIC for model selection.
The starting model, \texttt{m.auto}, is the working model fit to the
\texttt{Auto} data. The CV criterion isn't bias-adjusted because median
absolute error isn't a mean of casewise error components.

Some noteworthy points:

\begin{itemize}
\tightlist
\item
  \texttt{selectTransStepAIC()} automatically computes CV cost criteria,
  here the median absolute error, on the untransformed response scale.
\item
  The estimate of the median absolute error that we obtain by
  cross-validating the whole model-specification process is a little
  larger than the median absolute error computed for the model we fit to
  the \texttt{Auto} data separately selecting transformations of the
  predictors and the response and then selecting predictors for the
  whole data set.
\item
  When we look at the transformations and predictors selected with each
  of the 10 folds omitted (i.e., the output of \texttt{compareFolds()}),
  we see that there is little uncertainty in choosing variable
  transformations (the \texttt{lam.*}s for the \(x\)s and
  \texttt{lambda} for \(y\) in the output), but considerably more
  uncertainty in subsequently selecting predictors: \texttt{horsepower},
  \texttt{weight}, and \texttt{year} are always included among the
  selected predictors; \texttt{acceleration} and \texttt{displacement}
  are each included respectively in 4 and 3 of 10 selected models; and
  \texttt{cylinders} and \texttt{origin} are each included in only 1 of
  10 models. Recall that when we selected predictors for the full data,
  we obtained a model with \texttt{horsepower}, \texttt{weight},
  \texttt{acceleration}, \texttt{year}, and \texttt{origin}.
\end{itemize}

\hypertarget{parallel-computations}{%
\section{Parallel computations}\label{parallel-computations}}

The CV functions in the \textbf{cv} package are all capable of
performing parallel computations by setting the \texttt{ncores} argument
(specifying the number of computer cores to be used) to a number
\textgreater{} \texttt{1} (which is the default). Parallel computation
can be advantageous for large problems, reducing the execution time of
the program.

To illustrate, let's time model selection in Mroz's logistic regression,
repeating the computation as performed previously and then doing it in
parallel using 2 cores:

\begin{CodeChunk}
\begin{CodeInput}
R> system.time(m.mroz.sel.cv <- cvSelect(selectStepAIC, Mroz,
+                           seed=6681,
+                           criterion=BayesRule,
+                           model=m.mroz,
+                           AIC=FALSE))
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 6681
\end{CodeOutput}
\begin{CodeOutput}
   user  system elapsed 
  0.249   0.004   0.253 
\end{CodeOutput}
\begin{CodeInput}
R> system.time(m.mroz.sel.cv.p <- cvSelect(selectStepAIC, Mroz,
+                           seed=6681,
+                           criterion=BayesRule,
+                           model=m.mroz,
+                           AIC=FALSE,
+                           ncores=2))
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 6681
\end{CodeOutput}
\begin{CodeOutput}
   user  system elapsed 
  0.035   0.004   0.949 
\end{CodeOutput}
\begin{CodeInput}
R> all.equal(m.mroz.sel.cv, m.mroz.sel.cv.p)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

In this small problem, the parallel computation is actually
\emph{slower}, because there is an overhead cost to parallelization, but
we can see that it produces the same result as before.

\hypertarget{extending-the-cv-package}{%
\section{Extending the cv package}\label{extending-the-cv-package}}

The \textbf{cv} package is designed to be extensible in several
directions. In this vignette, we discuss three kinds of extensions,
ordered by increasing general complexity: (1) adding a cross-validation
cost criterion; (2) adding a model class that's not directly
accommodated by the \texttt{cv()} default method or by another directly
inherited method, with separate consideration of mixed-effects models;
and (3) adding a new model-selection procedure suitable for use with
\texttt{selectModel()}.

\hypertarget{adding-a-cost-criterion}{%
\subsection{Adding a cost criterion}\label{adding-a-cost-criterion}}

A cost criterion suitable for use with \texttt{cv()} or
\texttt{cvSelect()} should take two arguments, \texttt{y} (the observed
response vector) and \texttt{yhat} (a vector of fitted or predicted
response values), and return a numeric index of lack of fit. The
\textbf{cv} package supplies several such criteria:
\texttt{mse(y,\ yhat)}, which returns the mean-squared prediction error
for a numeric response; \texttt{rmse(y,\ yhat)}, which returns the
(square-)root mean-squared error; \texttt{medAbsErr(y,\ yhat)}, which
returns the median absolute error; and \texttt{BayesRule(y,\ yhat)} (and
its non-error-checking version, \texttt{BayesRule2(y,\ yhat))}, suitable
for use with a binary regression model, where \texttt{y} is the binary
response coded \texttt{0} for a ``failure'' or \texttt{1} for a
``success''; where \texttt{yhat} is the predicted probability of
success; and where the proportion of \emph{incorrectly} classified cases
is returned.

To illustrate using a different prediction cost criterion, we'll base a
cost criterion on the area under the receiver operating characteristic
(``ROC'') curve for a logistic regression. The ROC curve is a graphical
representation of the classification power of a binary regression model,
and the area under the ROC curve (``AUC''), which varies from 0 to 1, is
a common summary measure based on the ROC
\citep[see][]{Wikipedia-ROC:2023}. The \textbf{Metrics} package
\citep{HamnerFrasco:2018} includes a variety of measures useful for
model selection, including an \texttt{auc()} function. We convert the
AUC into a cost measure by taking its complement:

\begin{CodeChunk}
\begin{CodeInput}
R> AUCcomp <- function(y, yhat) 1 - Metrics::auc(y, yhat)
\end{CodeInput}
\end{CodeChunk}

We then apply \texttt{AUCcomp()} to the the Mroz logistic regression
discussed in the main \textbf{cv} package vignette, which we reproduce
here, using the \texttt{Mroz} data frame from the \textbf{carData}
package \citep{FoxWeisberg:2019}:

\begin{CodeChunk}
\begin{CodeInput}
R> data("Mroz", package="carData")
R> m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
R> summary(m.mroz)
\end{CodeInput}
\begin{CodeOutput}

Call:
glm(formula = lfp ~ ., family = binomial, data = Mroz)

Coefficients:
             Estimate Std. Error z value Pr(>|z|)
(Intercept)  3.182140   0.644375   4.938 7.88e-07
k5          -1.462913   0.197001  -7.426 1.12e-13
k618        -0.064571   0.068001  -0.950 0.342337
age         -0.062871   0.012783  -4.918 8.73e-07
wcyes        0.807274   0.229980   3.510 0.000448
hcyes        0.111734   0.206040   0.542 0.587618
lwg          0.604693   0.150818   4.009 6.09e-05
inc         -0.034446   0.008208  -4.196 2.71e-05

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 1029.75  on 752  degrees of freedom
Residual deviance:  905.27  on 745  degrees of freedom
AIC: 921.27

Number of Fisher Scoring iterations: 4
\end{CodeOutput}
\begin{CodeInput}
R> AUCcomp(with(Mroz, as.numeric(lfp == "yes")), fitted(m.mroz))
\end{CodeInput}
\begin{CodeOutput}
[1] 0.2636161
\end{CodeOutput}
\end{CodeChunk}

Cross-validating this cost measure is straightforward:

\begin{CodeChunk}
\begin{CodeInput}
R> library("cv")
R> cv(m.mroz, criterion=AUCcomp, seed=3639)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 3639
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: exact
criterion: AUCcomp
cross-validation criterion = 0.2747088
full-sample criterion = 0.2636161 
\end{CodeOutput}
\end{CodeChunk}

As expected, the cross-validated complement to the AUC is somewhat less
optimistic than the criterion computed from the model fit to the whole
data set.

As we explain in the vignette ``Cross-validation of regression models,''
the \texttt{cv()} function differentiates between CV criteria that are
averages of casewise components and criteria that are not. Computation
of bias corrections and confidence intervals is limited to the former.
We show in the appendix to this vignette that the AUC, and hence its
complement, cannot be expressed as averages of casewise components.

\texttt{cv()} looks for a \texttt{"casewise\ loss"} attribute of the
value returned by a CV criterion function. If this attribute exists,
then the criterion is treated as the mean of casewise components, and
\texttt{cv()} uses the unexported function \texttt{getLossFn()} to
construct a function that returns the casewise components of the
criterion.

We illustrate with the \texttt{mse()}:

\begin{CodeChunk}
\begin{CodeInput}
R> mse
\end{CodeInput}
\begin{CodeOutput}
 [1] 23.94366 18.98477 18.94499 18.87633 18.42697 18.24065 18.07817 18.06613
 [9] 18.02697 18.00953
\end{CodeOutput}
\begin{CodeInput}
R> cv:::getLossFn(mse(rnorm(100), rnorm(100)))
\end{CodeInput}
\begin{CodeOutput}
function (y, yhat) 
{
    (y - yhat)^2
}
<environment: 0x12bdf3978>
\end{CodeOutput}
\end{CodeChunk}

For this scheme to work, the ``casewise loss'' attribute must be a
character string (or vector of character strings), here
\texttt{"(y\ -\ yhat)\^{}2"}, that evaluates to an expression that is a
function of \texttt{y} and \texttt{yhat}, and that computes the vector
of casewise components of the CV criterion.

\hypertarget{adding-a-model-class-not-covered-by-the-default-cv-method}{%
\subsection{Adding a model class not covered by the default cv()
method}\label{adding-a-model-class-not-covered-by-the-default-cv-method}}

\hypertarget{independently-sampled-cases}{%
\subsubsection{Independently sampled
cases}\label{independently-sampled-cases}}

Suppose that we want to cross-validate a multinomial logistic regression
model fit by the \texttt{multinom()} function in the \textbf{nnet}
package \citep{VenablesRipley:2002}. We borrow an example from
\citet[Sec. 14.2.1]{Fox:2016}, with data from the British Election Panel
Study on vote choice in the 2001 British election. Data for the example
are in the \texttt{BEPS} data frame in the \textbf{carData} package:

\begin{CodeChunk}
\begin{CodeInput}
R> data("BEPS", package="carData")
R> head(BEPS)
\end{CodeInput}
\begin{CodeOutput}
              vote age economic.cond.national economic.cond.household Blair
1 Liberal Democrat  43                      3                       3     4
2           Labour  36                      4                       4     4
3           Labour  35                      4                       4     5
4           Labour  24                      4                       2     2
5           Labour  41                      2                       2     1
6           Labour  47                      3                       4     4
  Hague Kennedy Europe political.knowledge gender
1     1       4      2                   2 female
2     4       4      5                   2   male
3     2       3      3                   2   male
4     1       3      4                   0 female
5     1       4      6                   2   male
6     4       2      4                   2   male
\end{CodeOutput}
\end{CodeChunk}

The polytomous (multi-category) response variable is \texttt{vote}, a
factor with levels \texttt{"Conservative"}, \texttt{"Labour"}, and
\texttt{"Liberal\ Democrat"}. The predictors of \texttt{vote} are:

\begin{itemize}
\tightlist
\item
  \texttt{age}, in years;
\item
  \texttt{econ.cond.national} and \texttt{econ.cond.household}, the
  respondent's ratings of the state of the economy, on 1 to 5 scales.
\item
  \texttt{Blair}, \texttt{Hague}, and \texttt{Kennedy}, ratings of the
  leaders of the Labour, Conservative, and Liberal Democratic parties,
  on 1 to 5 scales.
\item
  \texttt{Europe}, an 11-point scale on attitude towards European
  integration, with high scores representing ``Euro-skepticism.''
\item
  \texttt{political.knowledge}, knowledge of the parties' positions on
  European integration, with scores from 0 to 3.
\item
  \texttt{gender}, \texttt{"female"} or \texttt{"male"}.
\end{itemize}

The model fit to the data includes an interaction between
\texttt{Europe} and \texttt{political.knowledge}; the other predictors
enter the model additively:

\begin{CodeChunk}
\begin{CodeInput}
R> library("nnet")
R> m.beps <- multinom(vote ~ age + gender + economic.cond.national +
+                        economic.cond.household + Blair + Hague + Kennedy +
+                        Europe*political.knowledge, data=BEPS)
\end{CodeInput}
\begin{CodeOutput}
# weights:  36 (22 variable)
initial  value 1675.383740 
iter  10 value 1240.047788
iter  20 value 1163.199642
iter  30 value 1116.519687
final  value 1116.519666 
converged
\end{CodeOutput}
\begin{CodeInput}
R> car::Anova(m.beps)
\end{CodeInput}
\begin{CodeOutput}
Analysis of Deviance Table (Type II tests)

Response: vote
                           LR Chisq Df Pr(>Chisq)
age                          13.872  2  0.0009721
gender                        0.453  2  0.7972568
economic.cond.national       30.604  2  2.262e-07
economic.cond.household       5.652  2  0.0592570
Blair                       135.369  2  < 2.2e-16
Hague                       166.770  2  < 2.2e-16
Kennedy                      68.878  2  1.105e-15
Europe                       78.033  2  < 2.2e-16
political.knowledge          55.568  2  8.582e-13
Europe:political.knowledge   50.804  2  9.291e-12
\end{CodeOutput}
\end{CodeChunk}

Most of the predictors, including the \texttt{Europe} \(\times\)
\texttt{political.knowledge} interaction, are associated with very small
\(p\)-values; the \texttt{Anova()} function is from the \textbf{car}
package \citep{FoxWeisberg:2019}.

Here's an ``effect plot'', using the the \textbf{effects} package
\citep{FoxWeisberg:2019} to visualize the \texttt{Europe} \(\times\)
\texttt{political.knowledge} interaction in a ``stacked-area'' graph:

\begin{CodeChunk}
\begin{CodeInput}
R> plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
+             xlevels=list(Europe=1:11, political.knowledge=0:3),
+             fixed.predictors=list(given.values=c(gendermale=0.5))),
+      lines=list(col=c("blue", "red", "orange")),
+      axes=list(x=list(rug=FALSE), y=list(style="stacked")))
\end{CodeInput}

\begin{center}\includegraphics{JSS-article_files/figure-latex/BEPS-plot-1} \end{center}

\end{CodeChunk}

To cross-validate this multinomial-logit model we need an appropriate
cost criterion. None of the criteria supplied by the \textbf{cv}
package---for example, neither \texttt{mse()}, which is appropriate for
a numeric response, nor \texttt{BayesRule()}, which is appropriate for a
binary response---will do. One possibility is to adapt Bayes rule to a
polytomous response:

\begin{CodeChunk}
\begin{CodeInput}
R> head(BEPS$vote)
\end{CodeInput}
\begin{CodeOutput}
[1] Liberal Democrat Labour           Labour           Labour          
[5] Labour           Labour          
Levels: Conservative Labour Liberal Democrat
\end{CodeOutput}
\begin{CodeInput}
R> yhat <- predict(m.beps, type="class")
R> head(yhat)
\end{CodeInput}
\begin{CodeOutput}
[1] Labour           Labour           Labour           Labour          
[5] Liberal Democrat Labour          
Levels: Conservative Labour Liberal Democrat
\end{CodeOutput}
\begin{CodeInput}
R> BayesRuleMulti <- function(y, yhat){
+   result <- mean(y != yhat)
+   attr(result, "casewise loss") <- "y != yhat"
+   result
+ }
R> 
R> BayesRuleMulti(BEPS$vote, yhat)
\end{CodeInput}
\begin{CodeOutput}
[1] 0.3186885
attr(,"casewise loss")
[1] "y != yhat"
\end{CodeOutput}
\end{CodeChunk}

The \texttt{predict()} method for \texttt{"multinom"} models called with
argument \texttt{type="class"} reports the Bayes-rule prediction for
each case---that is, the response category with the highest predicted
probability. Our \texttt{BayesRuleMulti()} function calculates the
proportion of misclassified cases. Because this value is the mean of
casewise components, we attach a \texttt{"casewise\ loss"} attribute to
the result (as explained in the preceding section).

The marginal proportions for the response categories are

\begin{CodeChunk}
\begin{CodeInput}
R> xtabs(~ vote, data=BEPS)/nrow(BEPS)
\end{CodeInput}
\begin{CodeOutput}
vote
    Conservative           Labour Liberal Democrat 
       0.3029508        0.4721311        0.2249180 
\end{CodeOutput}
\end{CodeChunk}

and so the marginal Bayes-rule prediction, that everyone will vote
Labour, produces an error rate of \(1 - 0.47213 = 0.52787\). The
multinomial-logit model appears to do substantially better than that,
but does its performance hold up to cross-validation?

We check first whether the default \texttt{cv()} method works
``out-of-the-box'' for the \texttt{"multinom"} model:

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.beps, seed=3465, criterion=BayesRuleMulti)
\end{CodeInput}
\begin{CodeOutput}
Error in GetResponse.default(model): non-vector response
\end{CodeOutput}
\end{CodeChunk}

The default method of \texttt{GetResponse()} (a function supplied by the
\textbf{cv} package---see \texttt{?GetResponse}) fails for a
\texttt{"multinom"} object. A straightforward solution is to supply a
\texttt{GetResponse.multinom()} method that returns the factor response
\citep[using the \texttt{get\_response()} function from the
\textbf{insight} package,][]{LudeckeWaggonerMakowski:2019},

\begin{CodeChunk}
\begin{CodeInput}
R> GetResponse.multinom <- function(model, ...) {
+   insight::get_response(model)
+ }
R> 
R> head(GetResponse(m.beps))
\end{CodeInput}
\begin{CodeOutput}
[1] Liberal Democrat Labour           Labour           Labour          
[5] Labour           Labour          
Levels: Conservative Labour Liberal Democrat
\end{CodeOutput}
\end{CodeChunk}

and to try again:

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.beps, seed=3465, criterion=BayesRuleMulti)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 3465
\end{CodeOutput}
\begin{CodeOutput}
# weights:  36 (22 variable)
initial  value 1507.296060 
iter  10 value 1134.575036
iter  20 value 1037.413231
iter  30 value 1007.705242
iter  30 value 1007.705235
iter  30 value 1007.705235
final  value 1007.705235 
converged
\end{CodeOutput}
\begin{CodeOutput}
Error in match.arg(type): 'arg' should be one of "class", "probs"
\end{CodeOutput}
\end{CodeChunk}

A \texttt{traceback()} (not shown) reveals that the problem is that the
default method of \texttt{cv()} calls the \texttt{"multinom"} method for
\texttt{predict()} with the argument \texttt{type="response"}, when the
correct argument should be \texttt{type="class"}. We therefore must
write a ``\texttt{multinom}'' method for \texttt{cv()}, but that proves
to be very simple:

\begin{CodeChunk}
\begin{CodeInput}
R> cv.multinom <- function (model, data, criterion=BayesRuleMulti, k, reps,
+                          seed, ...){
+   NextMethod(type="class", criterion=criterion)
+ }
\end{CodeInput}
\end{CodeChunk}

That is, we simply call the default \texttt{cv()} method with the
\texttt{type} argument properly set. In addition to supplying the
correct \texttt{type} argument, our method sets the default
\texttt{criterion} for the \texttt{cv.multinom()} method to
\texttt{BayesRuleMulti}.

Then:

\begin{CodeChunk}
\begin{CodeInput}
R> m.beps <- update(m.beps, trace=FALSE)
R> cv(m.beps, seed=3465)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 3465
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 0.3245902
bias-adjusted cross-validation criterion = 0.3236756
95% CI for bias-adjusted CV criterion = (0.300168, 0.3471831)
full-sample criterion = 0.3186885 
\end{CodeOutput}
\end{CodeChunk}

Prior to invoking \texttt{cv()}, we called \texttt{update()} with
\texttt{trace=FALSE} to suppress the iteration history reported by
default by \texttt{multinom()}---it would be tedious to see the
iteration history for each fold. The cross-validated polytomous
Bayes-rule criterion confirms that the fitted model does substantially
better than the marginal Bayes-rule prediction that everyone votes for
Labour.

\hypertarget{mixed-effects-models}{%
\subsubsection{Mixed-effects models}\label{mixed-effects-models}}

Adding a \texttt{cv()} method for a mixed-model class is somewhat more
complicated. We provide the \texttt{cvMixed()} function to facilitate
this process, and to see how that works, consider the \texttt{"lme"}
method from the \textbf{cv} package:

\begin{CodeChunk}
\begin{CodeInput}
R> cv:::cv.lme
\end{CodeInput}
\begin{CodeOutput}
function(model, data = insight::get_data(model), criterion = mse,
                   k, reps = 1, seed, ncores = 1, clusterVariables, ...){
  cvMixed(
    model,
    package="nlme",
    data=data,
    criterion=criterion,
    k=k,
    reps=reps,
    seed=seed,
    ncores=ncores,
    clusterVariables=clusterVariables,
    predict.clusters.args=list(object=model,
                               newdata=data,
                               level=0),
    predict.cases.args=list(object=model,
                            newdata=data,
                            level=1),
    ...)
}
<bytecode: 0x117501408>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

Notice that \texttt{cv.lme()} sets up a call to \texttt{cvMixed()},
which does the computational work.

Most of the arguments of \texttt{cvMixed()} are familiar:

\begin{itemize}
\item
  \texttt{model} is the mixed-model object, here of class
  \texttt{"lme"}.
\item
  \texttt{package} is the name of the package in which the
  mixed-modeling function used to fit the model, here \texttt{lme()},
  resides---i.e., \texttt{"nlme"}; \texttt{cvMixed()} uses this argument
  to retrieve the package namespace.
\item
  \texttt{data} is the data set to which the model is fit, by default
  extracted by the \texttt{get\_data()} function in the \textbf{insight}
  package.
\item
  \texttt{criterion} is the CV criterion, defaulting to the
  \texttt{mse()} function.
\item
  \texttt{k} is the number of CV folds, defaulting to \texttt{"loo"} for
  CV by clusters and \texttt{10} for CV by cases.
\item
  \texttt{reps} is the number of times the CV process is repeated,
  defaulting to \texttt{1}.
\item
  \texttt{seed} is the seed for R's random-number generator, defaulting
  to a randomly selected (and saved) value.
\item
  \texttt{ncores} is the number of cores to use for parallel
  computation; if \texttt{1}, the default, then the computation isn't
  parallelized.
\item
  \texttt{clusterVariables} is a character vector of the names of
  variables defining clusters; if missing, then CV is based on cases
  rather than clusters.
\end{itemize}

The remaining two arguments are unfamiliar:

\begin{itemize}
\item
  \texttt{predict.clusters.args} is a named list of arguments to be
  passed to the \texttt{predict()} function to obtain predictions for
  the full data set from a model fit to a subset of the data for
  cluster-based CV. The first two arguments should be \texttt{object}
  and \texttt{newdata}. It is typically necessary to tell
  \texttt{cvMixed()} how to base predictions only on fixed effects; in
  the case of \texttt{"lme"} models, this is done by setting
  \texttt{level\ =\ 0}.
\item
  Similarly, \texttt{predict.cases.args} is a named list of arguments to
  be passed to \texttt{predict()} for case-based CV. Setting
  \texttt{level\ =\ 1} includes random effects in the predictions.
\end{itemize}

Finally, any additional arguments, absorbed by \texttt{...}, are passed
to \texttt{update()} when the model is refit with each fold omitted.
\texttt{cvMixed()} returns an object of class \texttt{"cv"}.

Now imagine that we want to support a new class of mixed-effects models.
To be concrete, we illustrate with the \texttt{glmmPQL()} function in
the \textbf{MASS} package \citep{VenablesRipley:2002}, which fits
generalized-linear mixed-effects models by penalized
quasi-likelihood.\footnote{This example is somewhat artificial in that
  \texttt{glmmPQL()} has largely been superseded by computationally
  superior functions, such the \texttt{glmer()} function in the
  \textbf{lme4} package. There is, however, one situation in which
  \texttt{glmmPQL()} might prove useful: to specify serial dependency in
  case-level errors within clusters for longitudinal data, which is not
  currently supported by \texttt{glmer()}.} Not coincidentally, the
arguments of \texttt{glmmPQL()} are similar to those of \texttt{lme()}
(with an additional \texttt{family} argument), because the former
iteratively invokes the latter; so \texttt{cv.glmmPQL()} should resemble
\texttt{cv.lme()}.

As it turns out, neither the default method for \texttt{GetResponse()}
nor \texttt{insight::get\_data()} work for \texttt{"glmmPQL"} objects.
These objects include a \texttt{"data"} element, however, and so we can
simply extract this element as the default for the \texttt{data}
argument of our \texttt{cv.glmmPQL()} method.

To get the response variable is more complicated: We refit the fixed
part of the model as a GLM with only the regression constant on the
right-hand side, and extract the response from that; because all we need
is the response variable, we limit the number of GLM iterations to 1 and
suppress warning messages about non-convergence:

\begin{CodeChunk}
\begin{CodeInput}
R> GetResponse.glmmPQL <- function(model, ...){
+   f <- formula(model)
+   f[[3]] <- 1 # regression constant only on RHS
+   model <- suppressWarnings(glm(f, data=model$data, family=model$family,
+                                 control=list(maxit=1)))
+   cv::GetResponse(model)
+ }
\end{CodeInput}
\end{CodeChunk}

Writing the \texttt{cv()} method is then straightforward:

\begin{CodeChunk}
\begin{CodeInput}
R> cv.glmmPQL <- function(model, data = model$data, criterion = mse,
+                      k, reps = 1, seed, ncores = 1, clusterVariables, ...){
+   cvMixed(
+     model,
+     package="MASS",
+     data=data,
+     criterion=criterion,
+     k=k,
+     reps=reps,
+     seed=seed,
+     ncores=ncores,
+     clusterVariables=clusterVariables,
+     predict.clusters.args=list(object=model,
+                                newdata=data,
+                                level=0,
+                                type="response"),
+     predict.cases.args=list(object=model,
+                             newdata=data,
+                             level=1,
+                             type="response"),
+     verbose=FALSE,
+     ...)
+ }
\end{CodeInput}
\end{CodeChunk}

We set the argument \texttt{verbose=FALSE} to suppress
\texttt{glmmPQL()}'s iteration counter when \texttt{cvMixed()} calls
\texttt{update()}.

Let's apply our newly minted method to a logistic regression with a
random intercept in an example that appears in \texttt{?glmmPQL}:

\begin{CodeChunk}
\begin{CodeInput}
R> library("MASS")
R> m.pql <- glmmPQL(y ~ trt + I(week > 2), random = ~ 1 | ID,
+              family = binomial, data = bacteria)
\end{CodeInput}
\begin{CodeOutput}
iteration 1
\end{CodeOutput}
\begin{CodeOutput}
iteration 2
\end{CodeOutput}
\begin{CodeOutput}
iteration 3
\end{CodeOutput}
\begin{CodeOutput}
iteration 4
\end{CodeOutput}
\begin{CodeOutput}
iteration 5
\end{CodeOutput}
\begin{CodeOutput}
iteration 6
\end{CodeOutput}
\begin{CodeInput}
R> summary(m.pql)
\end{CodeInput}
\begin{CodeOutput}
Linear mixed-effects model fit by maximum likelihood
  Data: bacteria 
  AIC BIC logLik
   NA  NA     NA

Random effects:
 Formula: ~1 | ID
        (Intercept)  Residual
StdDev:    1.410637 0.7800511

Variance function:
 Structure: fixed weights
 Formula: ~invwt 
Fixed effects:  y ~ trt + I(week > 2) 
                    Value Std.Error  DF   t-value p-value
(Intercept)      3.412014 0.5185033 169  6.580506  0.0000
trtdrug         -1.247355 0.6440635  47 -1.936696  0.0588
trtdrug+        -0.754327 0.6453978  47 -1.168779  0.2484
I(week > 2)TRUE -1.607257 0.3583379 169 -4.485311  0.0000
 Correlation: 
                (Intr) trtdrg trtdr+
trtdrug         -0.598              
trtdrug+        -0.571  0.460       
I(week > 2)TRUE -0.537  0.047 -0.001

Standardized Within-Group Residuals:
       Min         Q1        Med         Q3        Max 
-5.1985361  0.1572336  0.3513075  0.4949482  1.7448845

Number of Observations: 220
Number of Groups: 50 
\end{CodeOutput}
\end{CodeChunk}

We compare this result to that obtained from \texttt{glmer()} in the
\textbf{lme4} package:

\begin{CodeChunk}
\begin{CodeInput}
R> library("lme4")
R> m.glmer <- glmer(y ~ trt + I(week > 2) + (1 | ID),
+                family = binomial, data = bacteria)
R> summary(m.glmer)
\end{CodeInput}
\begin{CodeOutput}
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: y ~ trt + I(week > 2) + (1 | ID)
   Data: bacteria

     AIC      BIC   logLik deviance df.resid 
   202.3    219.2    -96.1    192.3      215

Scaled residuals: 
    Min      1Q  Median      3Q     Max 
-4.5615  0.1359  0.3022  0.4217  1.1276

Random effects:
 Groups Name        Variance Std.Dev.
 ID     (Intercept) 1.543    1.242   
Number of obs: 220, groups:  ID, 50

Fixed effects:
                Estimate Std. Error z value Pr(>|z|)
(Intercept)       3.5479     0.6958   5.099 3.41e-07
trtdrug          -1.3667     0.6770  -2.019 0.043517
trtdrug+         -0.7826     0.6831  -1.146 0.251925
I(week > 2)TRUE  -1.5985     0.4759  -3.359 0.000783

Correlation of Fixed Effects:
            (Intr) trtdrg trtdr+
trtdrug     -0.593              
trtdrug+    -0.537  0.487       
I(wk>2)TRUE -0.656  0.126  0.064
\end{CodeOutput}
\begin{CodeInput}
R>   # comparison of fixed effects:
R> car::compareCoefs(m.pql, m.glmer) 
\end{CodeInput}
\begin{CodeOutput}
Warning in car::compareCoefs(m.pql, m.glmer): models to be compared are of
different classes
\end{CodeOutput}
\begin{CodeOutput}
Calls:
1: glmmPQL(fixed = y ~ trt + I(week > 2), random = ~1 | ID, family = 
  binomial, data = bacteria)
2: glmer(formula = y ~ trt + I(week > 2) + (1 | ID), data = bacteria, 
  family = binomial)

                Model 1 Model 2
(Intercept)       3.412   3.548
SE                0.514   0.696

trtdrug          -1.247  -1.367
SE                0.638   0.677

trtdrug+         -0.754  -0.783
SE                0.640   0.683

I(week > 2)TRUE  -1.607  -1.598
SE                0.355   0.476

\end{CodeOutput}
\end{CodeChunk}

The two sets of estimates are similar, but not identical

Finally, we try out our \texttt{cv.glmmPQL()} method, cross-validating
both by clusters and by cases,

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.pql, clusterVariables="ID", criterion=BayesRule)
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation based on 50 {ID} clusters
cross-validation criterion = 0.1954545
bias-adjusted cross-validation criterion = 0.1954545
full-sample criterion = 0.1954545 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.pql, data=bacteria, criterion=BayesRule, seed=1490)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 1490
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 0.2090909
bias-adjusted cross-validation criterion = 0.2072727
full-sample criterion = 0.1454545 
\end{CodeOutput}
\end{CodeChunk}

and again compare to \texttt{glmer()}:

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.glmer, clusterVariables="ID", criterion=BayesRule)
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation based on 50 {ID} clusters
cross-validation criterion = 0.1954545
bias-adjusted cross-validation criterion = 0.1954545
full-sample criterion = 0.1954545 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.glmer, data=bacteria, criterion=BayesRule, seed=1490)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 1490
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 0.1954545
bias-adjusted cross-validation criterion = 0.1936364
full-sample criterion = 0.15 
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{adding-a-model-selection-procedure}{%
\subsection{Adding a model-selection
procedure}\label{adding-a-model-selection-procedure}}

The \texttt{selectStepAIC()} function supplied by the \textbf{cv}
package, which is based on the \texttt{stepAIC()} function from the
\textbf{nnet} package \citep{VenablesRipley:2002} for stepwise model
selection, is suitable for the \texttt{procedure} argument of
\texttt{cvSelect()}. The use of \texttt{selectStepAIC()} is illustrated
in the principal vignette for the package.

We'll employ \texttt{selectStepAIC()} as a ``template'' for writing a CV
model-selection procedure. To see the code for this function, type
\texttt{cv::selectStepAIC} at the R command prompt, or examine the
sources for the \textbf{cv} package at
\url{https://github.com/gmonette/cv} (the code for
\texttt{selectStepAIC()} is in
\url{https://github.com/gmonette/cv/blob/main/R/cvSelect.R}).

Another approach to model selection is all-subsets regression. The
\texttt{regsubsets()} function in the \textbf{leaps} package
\citep{LumleyMiller:2020} implements an efficient algorithm for
selecting the best-fitting linear least-squares regressions for subsets
of predictors of all sizes, from 1 through the maximum number of
candidate predictors.\footnote{The \texttt{regsubsets()} function
  computes several measures of model predictive performance, including
  the \(R^2\) and \(R^2\) adjusted for degrees of freedom, the residual
  sums of squares, Mallows's \(C_p\), and the BIC. Several of these are
  suitable for comparing models with differing numbers of
  coefficients---we use the BIC below---but all necessarily agree when
  comparing models with the \emph{same} number of coefficients.} To
illustrate the use of \texttt{regsubsets()}, we employ the
\texttt{swiss} data frame supplied by the \textbf{leaps} package:

\begin{CodeChunk}
\begin{CodeInput}
R> library("leaps")
R> head(swiss)
\end{CodeInput}
\begin{CodeOutput}
             Fertility Agriculture Examination Education Catholic
Courtelary        80.2        17.0          15        12     9.96
Delemont          83.1        45.1           6         9    84.84
Franches-Mnt      92.5        39.7           5         5    93.40
Moutier           85.8        36.5          12         7    33.77
Neuveville        76.9        43.5          17        15     5.16
Porrentruy        76.1        35.3           9         7    90.57
             Infant.Mortality
Courtelary               22.2
Delemont                 22.2
Franches-Mnt             20.2
Moutier                  20.3
Neuveville               20.6
Porrentruy               26.6
\end{CodeOutput}
\begin{CodeInput}
R> nrow(swiss)
\end{CodeInput}
\begin{CodeOutput}
[1] 47
\end{CodeOutput}
\end{CodeChunk}

The data set includes the following variables, for each of 47
French-speaking Swiss provinces circa 1888:

\begin{itemize}
\tightlist
\item
  \texttt{Fertility}: A standardized fertility measure.
\item
  \texttt{Agriculture}: The percentage of the male population engaged in
  agriculture.
\item
  \texttt{Examination}: The percentage of draftees into the Swiss army
  receiving the highest grade on an examination.
\item
  \texttt{Education}: The percentage of draftees with more than a
  primary-school education.
\item
  \texttt{Catholic}: The percentage of the population who were Catholic.
\item
  \texttt{Infant.Mortality}: The infant-mortality rate, expressed as the
  percentage of live births surviving less than a year.
\end{itemize}

Following \citet{LumleyMiller:2020}, we treat \texttt{Fertility} as the
response and the other variables as predictors in a linear least-squares
regression:

\begin{CodeChunk}
\begin{CodeInput}
R> m.swiss <- lm(Fertility ~ ., data=swiss)
R> summary(m.swiss)
\end{CodeInput}
\begin{CodeOutput}

Call:
lm(formula = Fertility ~ ., data = swiss)

Residuals:
     Min       1Q   Median       3Q      Max 
-15.2743  -5.2617   0.5032   4.1198  15.3213

Coefficients:
                 Estimate Std. Error t value Pr(>|t|)
(Intercept)      66.91518   10.70604   6.250 1.91e-07
Agriculture      -0.17211    0.07030  -2.448  0.01873
Examination      -0.25801    0.25388  -1.016  0.31546
Education        -0.87094    0.18303  -4.758 2.43e-05
Catholic          0.10412    0.03526   2.953  0.00519
Infant.Mortality  1.07705    0.38172   2.822  0.00734

Residual standard error: 7.165 on 41 degrees of freedom
Multiple R-squared:  0.7067,    Adjusted R-squared:  0.671 
F-statistic: 19.76 on 5 and 41 DF,  p-value: 5.594e-10
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.swiss, seed=8433)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 8433
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 59.68315
bias-adjusted cross-validation criterion = 58.84614
full-sample criterion = 44.78815 
\end{CodeOutput}
\end{CodeChunk}

Thus, the RMSE for the model fit to the complete data is considerably
smaller than the CV estimate of the RMSE. Can we do better by selecting
a subset of the predictors, taking account of the additional uncertainty
induced by model selection?

First, let's apply best-subset selection to the complete data set:

\begin{CodeChunk}
\begin{CodeInput}
R> swiss.sub <- regsubsets(Fertility ~ ., data=swiss)
R> summary(swiss.sub)
\end{CodeInput}
\begin{CodeOutput}
Subset selection object
Call: regsubsets.formula(Fertility ~ ., data = swiss)
5 Variables  (and intercept)
                 Forced in Forced out
Agriculture          FALSE      FALSE
Examination          FALSE      FALSE
Education            FALSE      FALSE
Catholic             FALSE      FALSE
Infant.Mortality     FALSE      FALSE
1 subsets of each size up to 5
Selection Algorithm: exhaustive
         Agriculture Examination Education Catholic Infant.Mortality
1  ( 1 ) " "         " "         "*"       " "      " "             
2  ( 1 ) " "         " "         "*"       "*"      " "             
3  ( 1 ) " "         " "         "*"       "*"      "*"             
4  ( 1 ) "*"         " "         "*"       "*"      "*"             
5  ( 1 ) "*"         "*"         "*"       "*"      "*"             
\end{CodeOutput}
\begin{CodeInput}
R> (bics <- summary(swiss.sub)$bic)
\end{CodeInput}
\begin{CodeOutput}
[1] -19.60287 -28.61139 -35.65643 -37.23388 -34.55301
\end{CodeOutput}
\begin{CodeInput}
R> which.min(bics)
\end{CodeInput}
\begin{CodeOutput}
[1] 4
\end{CodeOutput}
\begin{CodeInput}
R> car::subsets(swiss.sub, legend="topright")
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{JSS-article_files/figure-latex/subset-selection-1}

}

\caption[Selecting the best model of each size]{Selecting the best model of each size.}\label{fig:subset-selection}
\end{figure}
\end{CodeChunk}

The graph, produced by the \texttt{subsets()} function in the
\textbf{car} package, shows that the model with the smallest BIC is the
best model with 4 predictors, including \texttt{Agriculture},
\texttt{Education}, \texttt{Catholic}, and \texttt{Infant.Mortality},
but not \texttt{Examination}:

\begin{CodeChunk}
\begin{CodeInput}
R> m.best <- update(m.swiss, . ~ . - Examination)
R> summary(m.best)
\end{CodeInput}
\begin{CodeOutput}

Call:
lm(formula = Fertility ~ Agriculture + Education + Catholic + 
    Infant.Mortality, data = swiss)

Residuals:
     Min       1Q   Median       3Q      Max 
-14.6765  -6.0522   0.7514   3.1664  16.1422

Coefficients:
                 Estimate Std. Error t value Pr(>|t|)
(Intercept)      62.10131    9.60489   6.466 8.49e-08
Agriculture      -0.15462    0.06819  -2.267  0.02857
Education        -0.98026    0.14814  -6.617 5.14e-08
Catholic          0.12467    0.02889   4.315 9.50e-05
Infant.Mortality  1.07844    0.38187   2.824  0.00722

Residual standard error: 7.168 on 42 degrees of freedom
Multiple R-squared:  0.6993,    Adjusted R-squared:  0.6707 
F-statistic: 24.42 on 4 and 42 DF,  p-value: 1.717e-10
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.best, seed=8433) # use same folds as before
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 8433
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 58.46664
bias-adjusted cross-validation criterion = 57.7779
full-sample criterion = 45.91637 
\end{CodeOutput}
\end{CodeChunk}

The RMSE for the selected model is (of course) slightly higher than for
the full model fit previously, but the cross-validated RMSE is a bit
lower; as we explain in the main vignette, however, it isn't kosher to
select and cross-validate a model on the same data.

Here's a function named \texttt{selectSubsets()}, meant to be used with
\texttt{cvSelect()}, suitable for cross-validating the model-selection
process:

\begin{CodeChunk}
\begin{CodeInput}
R> selectSubsets <- function(data=insight::get_data(model), 
+                           model,
+                           indices,
+                           criterion=mse,
+                           save.coef=TRUE, ...){
+   
+   if (inherits(model, "lm", which=TRUE) != 1)
+     stop("selectSubsets is appropriate only for 'lm' models")
+   
+   y <- GetResponse(model)
+   formula <- formula(model)
+   X <- model.matrix(model)
+ 
+   if (missing(indices)) {
+     # select the best model from the full data by BIC
+     sel <- leaps::regsubsets(formula, data=data, ...)
+     bics <- summary(sel)$bic
+     best <- coef(sel, 1:length(bics))[[which.min(bics)]]
+     x.names <- names(best)
+     # fit the best model; intercept is already in X, hence - 1:
+     m.best <- lm(y ~ X[, x.names] - 1) 
+     fit.all <- predict(m.best, newdata=data)
+     return(criterion(y, fit.all)) # return the CV criterion
+   }
+ 
+   # select the best model omitting the i-th fold (given by indices)
+   sel.i <- leaps::regsubsets(formula, data[-indices, ], ...)
+   bics.i <- summary(sel.i)$bic
+   best.i <- coef(sel.i, 1:length(bics.i))[[which.min(bics.i)]]
+   x.names.i <- names(best.i)
+   m.best.i <- lm(y[-indices] ~ X[-indices, x.names.i] - 1)
+               # predict() doesn't work here:
+   fit.all.i <- as.vector(X[, x.names.i] %*% coef(m.best.i))
+   fit.i <- fit.all.i[indices]
+   # return the fitted values for i-th fold, CV criterion for all cases, 
+   #   and the regression coefficients
+   list(fit.i=fit.i, # fitted values for i-th fold
+        crit.all.i=criterion(y, fit.all.i), # CV crit for all cases
+        coefficients = if (save.coef){ # regression coefficients
+          coefs <- coef(m.best.i)
+          
+          # fix coefficient names
+          names(coefs) <- sub("X\\[-indices, x.names.i\\]", "",
+                              names(coefs))
+          
+          coefs
+        }  else {
+          NULL
+        }
+   )
+ }
\end{CodeInput}
\end{CodeChunk}

A slightly tricky point is that because of scoping issues,
\texttt{predict()} doesn't work with the model fit omitting the \(i\)th
fold, and so the fitted values for all cases are computed directly as
\(\widehat{\mathbf{y}}_{-i} = \mathbf{X} \mathbf{b}_{-i}\), where
\(\mathbf{X}\) is the model-matrix for all of the cases, and
\(\mathbf{b}_{-i}\) is the vector of least-squares coefficients for the
selected model with the \(i\)th fold omitted.

Additionally, the command
\texttt{lm(y{[}-indices{]}\ \textasciitilde{}\ X{[}-indices,\ x.names.i{]}\ -\ 1)},
which is the selected model with the \(i\)th fold deleted, produces
awkward coefficient names like
\texttt{"X{[}-indices,\ x.names.i{]}Infant.Mortality"}. Purely for
aesthetic reasons, the command
\texttt{sub("X\textbackslash{}\textbackslash{}{[}-indices,\ x.names.i\textbackslash{}\textbackslash{}{]}",\ "",\ names(coefs))}
fixes these awkward names, removing the extraneous text,
\texttt{"X{[}-indices,\ x.names.i{]}"}.

Applying \texttt{selectSubsets()} to the full data produces the
full-data cross-validated RMSE (which we obtained previously):
\%\%\texttt{\{r\ test-selectSubsets\}\ \%\%selectSubsets(model=m.swiss)\ \%\%}
Similarly, applying the function to an imaginary ``fold'' of 5 cases
returns the RMSE for the cases in the fold, based on the model selected
and fit to the cases omitting the fold; the RMSE for all of the cases,
based on the same model; and the coefficients of the selected model,
which includes 4 or the 5 predictors (and the intercept):
\%\%\texttt{\{r\ test-selectSubsets-fold\}\ \%\%selectSubsets(model=m.swiss,\ indices=seq(5,\ 45,\ by=10))\ \%\%}

Then, using \texttt{selectSubsets()} in cross-validation, we get:

\begin{CodeChunk}
\begin{CodeInput}
R> (cv.swiss <- cvSelect(selectSubsets, model=m.swiss,
+                       data=swiss, seed=8433)) # use same folds
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 8433
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 65.83495
bias-adjusted cross-validation criterion = 63.64431
full-sample criterion = 45.91637 
\end{CodeOutput}
\end{CodeChunk}

Cross-validation shows that model selection exacts a penalty in RMSE.
Examining the models selected for the 10 folds reveals that there is
some uncertainty in identifying the predictors in the ``best'' model,
with \texttt{Agriculture} sometimes appearing and sometimes not:

\begin{CodeChunk}
\begin{CodeInput}
R> compareFolds(cv.swiss)
\end{CodeInput}
\begin{CodeOutput}
        (Intercept) Catholic Education Infant.Mortality Agriculture
Fold 1      59.0852   0.1397   -1.0203           1.2985       -0.17
Fold 2      67.0335   0.1367   -1.0499           0.9413       -0.20
Fold 3      55.0453   0.1221   -0.8757           1.3541       -0.15
Fold 4      62.5543   0.1236   -0.9719           1.0679       -0.16
Fold 5      50.4643   0.1057   -0.7863           1.2144            
Fold 6      68.0289   0.1195   -1.0073           0.8294       -0.17
Fold 7      66.5219   0.1357   -1.0827           0.9523       -0.19
Fold 8      46.3507   0.0776   -0.7637           1.4463            
Fold 9      62.2632   0.1230   -1.0067           1.1000       -0.17
Fold 10     52.5112   0.1005   -0.7232           1.0809            
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{computational-notes}{%
\section{Computational notes}\label{computational-notes}}

\hypertarget{efficient-computations-for-linear-and-generalized-linear-models}{%
\subsection{Efficient computations for linear and generalized linear
models}\label{efficient-computations-for-linear-and-generalized-linear-models}}

The most straightforward way to implement cross-validation in R for
statistical modeling functions that are written in the canonical manner
is to use \texttt{update()} to refit the model with each fold removed.
This is the approach taken in the default method for \texttt{cv()}, and
it is appropriate if the cases are independently sampled. Refitting the
model in this manner for each fold is generally feasible when the number
of folds in modest, but can be prohibitively costly for leave-one-out
cross-validation when the number of cases is large.

The \texttt{"lm"} and \texttt{"glm"} methods for \texttt{cv()} take
advantage of computational efficiencies by avoiding refitting the model
with each fold removed. Consider, in particular, the weighted linear
model
\(\mathbf{y}_{n \times 1} = \mathbf{X}_{n \times p}\boldsymbol{\beta}_{p \times 1} + \boldsymbol{\varepsilon}_{n \times 1}\),
where
\(\boldsymbol{\varepsilon} \sim \mathbf{N}_n \left(\mathbf{0}, \sigma^2 \mathbf{W}^{-1}_{n \times n}\right)\).
Here, \(\mathbf{y}\) is the response vector, \(\mathbf{X}\) the model
matrix, and \(\boldsymbol{\varepsilon}\) the error vector, each for
\(n\) cases, and \(\boldsymbol{\beta}\) is the vector of \(p\)
population regression coefficients. The errors are assumed to be
multivariately normally distributed with 0 means and covariance matrix
\(\sigma^2 \mathbf{W}^{-1}\), where \(\mathbf{W} = \mathrm{diag}(w_i)\)
is a diagonal matrix of inverse-variance weights. For the linear model
with constant error variance, the weight matrix is taken to be
\(\mathbf{W} = \mathbf{I}_n\), the order-\(n\) identity matrix.

The weighted-least-squares (WLS) estimator of \(\boldsymbol{\beta}\) is
\citep[see, e.g.,][Sec. 12.2.2]{Fox:2016} \footnote{This is a
  definitional formula, which assumes that the model matrix
  \(\mathbf{X}\) is of full column rank, and which can be subject to
  numerical instability when \(\mathbf{X}\) is ill-conditioned.
  \texttt{lm()} uses the singular-value decomposition of the model
  matrix to obtain computationally more stable results.} \[
\mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} 
  \mathbf{X}^T \mathbf{W} \mathbf{y}
\]

Fitted values are then
\(\widehat{\mathbf{y}} = \mathbf{X}\mathbf{b}_{\mathrm{WLS}}\).

The LOO fitted value for the \(i\)th case can be efficiently computed by
\(\widehat{y}_{-i} = y_i - e_i/(1 - h_i)\) where
\(h_i = \mathbf{x}^T_i \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} \mathbf{x}_i\)
(the so-called ``hatvalue''). Here, \(\mathbf{x}^T_i\) is the \(i\)th
row of \(\mathbf{X}\), and \(\mathbf{x}_i\) is the \(i\)th row written
as a column vector. This approach can break down when one or more
hatvalues are equal to 1, in which case the formula for
\(\widehat{y}_{-i}\) requires division by 0.

To compute cross-validated fitted values when the folds contain more
than one case, we make use of the Woodbury matrix identify
\citep{Wikipedia-Woodbury:2023}, \[
\left(\mathbf{A}_{m \times m} + \mathbf{U}_{m \times k} 
\mathbf{C}_{k \times k} \mathbf{V}_{k \times m} \right)^{-1} = \mathbf{A}^{-1} - \mathbf{A}^{-1}\mathbf{U} \left(\mathbf{C}^{-1} + 
\mathbf{VA}^{-1}\mathbf{U} \right)^{-1} \mathbf{VA}^{-1}
\] where \(\mathbf{A}\) is a nonsingular order-\(n\) matrix. We apply
this result by letting \begin{align*}
    \mathbf{A} &= \mathbf{X}^T \mathbf{W} \mathbf{X} \\
    \mathbf{U} &= \mathbf{X}_\mathbf{j}^T \\
    \mathbf{V} &= - \mathbf{X}_\mathbf{j} \\
    \mathbf{C} &= \mathbf{W}_\mathbf{j} \\
\end{align*} where the subscript
\(\mathbf{j} = (i_{j1}, \ldots, i_{jm})^T\) represents the vector of
indices for the cases in the \(j\)th fold, \(j = 1, \ldots, k\). The
negative sign in \(\mathbf{V} = - \mathbf{X}_\mathbf{j}\) reflects the
\emph{removal}, rather than addition, of the cases in \(\mathbf{j}\).

Applying the Woodbury identity isn't quite as fast as using the
hatvalues, but it is generally much faster than refitting the model. A
disadvantage of the Woodbury identity, however, is that it entails
explicit matrix inversion and thus may be numerically unstable. The
inverse of \(\mathbf{A} = \mathbf{X}^T \mathbf{W} \mathbf{X}\) is
available directly in the \texttt{"lm"} object, but the second term on
the right-hand side of the Woodbury identity requires a matrix inversion
with each fold deleted. (In contrast, the inverse of each
\(\mathbf{C} = \mathbf{W}_\mathbf{j}\) is straightforward because
\(\mathbf{W}\) is diagonal.)

The Woodbury identity also requires that the model matrix be of full
rank. We impose that restriction in our code by removing redundant
regressors from the model matrix for all of the cases, but that doesn't
preclude rank deficiency from surfacing when a fold is removed. Rank
deficiency of \(\mathbf{X}\) doesn't disqualify cross-validation because
all we need are fitted values under the estimated model.

\texttt{glm()} computes the maximum-likelihood estimates for a
generalized linear model by iterated weighted least squares \citep[see,
e.g.,][Sec. 6.12]{FoxWeisberg:2019}. The last iteration is therefore
just a WLS fit of the ``working response'' on the model matrix using
``working weights.'' Both the working weights and the working response
at convergence are available from the information in the object returned
by \texttt{glm()}.

We then treat re-estimation of the model with a case or cases deleted as
a WLS problem, using the hatvalues or the Woodbury matrix identity. The
resulting fitted values for the deleted fold aren't exact---that is,
except for the Gaussian family, the result isn't identical to what we
would obtain by literally refitting the model---but in our (limited)
experience, the approximation is very good, especially for LOO CV, which
is when we would be most tempted to use it. Nevertheless, because these
results are approximate, the default for the \texttt{"glm"}
\texttt{cv()} method is to perform the exact computation, which entails
refitting the model with each fold omitted.

\hypertarget{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}{%
\subsection{Computation of the bias-corrected CV criterion and
confidence
intervals}\label{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}}

Let \(\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})\) represent a
cross-validation cost criterion, such as mean-squared error, computed
for all of the \(n\) values of the response \(\mathbf{y}\) based on
fitted values \(\widehat{\mathbf{y}}\) from the model fit to all of the
data. We require that \(\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})\)
is the mean of casewise components, that is,
\(\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n\mathrm{cv}(y_i, \widehat{y}_i)\).\footnote{\citet{ArlotCelisse:2010}
  term the casewise loss, \(\mathrm{cv}(y_i, \widehat{y}_i)\), the
  ``contrast function.''} For example,
\(\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\).

We divide the \(n\) cases into \(k\) folds of approximately
\(n_j \approx n/k\) cases each, where \(n = \sum n_j\). As above, let
\(\mathbf{j}\) denote the indices of the cases in the \(j\)th fold.

Now define
\(\mathrm{CV}_j = \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}^{(j)})\).
The superscript \((j)\) on \(\widehat{\mathbf{y}}^{(j)}\) represents
fitted values computed for all of the cases from the model with fold
\(j\) omitted. Let \(\widehat{\mathbf{y}}^{(-i)}\) represent the vector
of fitted values for all \(n\) cases where the fitted value for the
\(i\)th case is computed from the model fit with the fold including the
\(i\)th case omitted (i.e., fold \(j\) for which \(i \in \mathbf{j}\)).

Then the cross-validation criterion is just
\(\mathrm{CV} = \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}^{(-i)})\).
Following \citet[pp.~293--295]{DavisonHinkley:1997}, the bias-adjusted
cross-validation criterion is \[
\mathrm{CV}_{\mathrm{adj}} = \mathrm{CV} + \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) - \frac{1}{n} \sum_{j=1}^{k} n_j \mathrm{CV}_j
\]

We compute the standard error of CV as \[
\mathrm{SE}(\mathrm{CV}) = \frac{1}{\sqrt n} \sqrt{ \frac{\sum_{i=1}^n \left[ \mathrm{cv}(y_i, \widehat{y}_i^{(-i)} ) - \mathrm{CV} \right]^2 }{n - 1} }
\] that is, as the standard deviation of the casewise components of CV
divided by the square-root of the number of cases.

We then use \(\mathrm{SE}(\mathrm{CV})\) to construct a
\(100 \times (1 - \alpha)\)\% confidence interval around the
\emph{adjusted} CV estimate of error: \[
\left[ \mathrm{CV}_{\mathrm{adj}} - z_{1 - \alpha/2}\mathrm{SE}(\mathrm{CV}), \mathrm{CV}_{\mathrm{adj}} + z_{1 - \alpha/2}\mathrm{SE}(\mathrm{CV})  \right]
\] where \(z_{1 - \alpha/2}\) is the \(1 - \alpha/2\) quantile of the
standard-normal distribution (e.g, \(z \approx 1.96\) for a 95\%
confidence interval, for which \(1 - \alpha/2 = .975\)).

\citet{BatesHastieTibshirani:2023} show that the coverage of this
confidence interval is poor for small samples, and they suggest a much
more computationally intensive procedure, called \emph{nested
cross-validation}, to compute better estimates of error and confidence
intervals with better coverage for small samples. We may implement Bates
et al.'s approach in a later release of the \textbf{cv} package. At
present we use the confidence interval above for sufficiently large
\(n\), which, based on Bates et al.'s results, we take by default to be
\(n \ge 400\).

\hypertarget{why-the-complement-of-auc-isnt-a-casewise-cv-criterion}{%
\section{Why the complement of AUC isn't a casewise CV
criterion}\label{why-the-complement-of-auc-isnt-a-casewise-cv-criterion}}

Consider calculating AUC for folds in which a validation set contains
\(n_v\) observations. To calculate AUC in the validation set, we need
the vector of prediction criteria,
\(\widehat{\mathbf{y}}_{v_{(n_v \times 1)}} = (\widehat{y}_1, ..., \widehat{y}_{n_v})^T\),
and the vector of observed responses in the validation set,
\(\mathbf{y}_{v_{(n_v \times 1)}} = (y_1, \ldots, y_{n_v})^T\) with
\(y_i \in \{0,1\}, \; i = 1, \ldots, n_v\).

To construct the ROC curve, only the ordering of the values in
\(\mathbf{\widehat{y}}_v\) is relevant. Thus, assuming that there are no
ties, and reordering observations if necessary, we can set
\(\mathbf{\widehat{y}}_v = (1, 2, \ldots, n_v)^T\).

If the AUC can be expressed as the casewise mean or sum of a function
\(\mathrm{cv}(\widehat{y}_i,y_i)\), where
\(\mathrm{cv}: \{1,2,...,n_v\}\times\{0,1\} \rightarrow [0,1]\), then
\begin{equation}
\label{eq:cw}
\tag{1}
\sum_{i=1}^{n_v} \mathrm{cv}(\widehat{y}_i,y_i) = \mathrm{AUC}(\mathbf{\widehat{y}}_v,\mathbf{y}_v)
\end{equation} must hold for all \(2^{n_v}\) possible values of
\(\mathbf{y}_v = (y_1,...,y_{n_v})^T\). If all \(y\mathrm{s}\) have the
same value, either 1 or 0, then the definition of AUC is ambiguous. AUC
could be considered undefined, or it could be set to 0 if all \(y\)s are
0 and to 1 if all \(y\)s are 1. If AUC is considered to be undefined in
these cases, we have \(2^{n_v} - 2\) admissible values for
\(\mathbf{y}_v\).

Thus, equation (\ref{eq:cw}) produces either \(2^{n_v}\) or
\(2^{n_v}-2\) constraints. Although there are only \(2n_v\) possible
values for the \(\mathrm{cv(\cdot)}\) function, equation (\ref{eq:cw})
could, nevertheless, have consistent solutions. We therefore need to
determine whether there is a value of \(n_v\) for which (\ref{eq:cw})
has no consistent solution for all admissible values of
\(\mathbf{y}_v\). In that eventuality, we will have shown that AUC
cannot, in general, be expressed through a casewise sum.

If \(n_v=3\), we show below that (\ref{eq:cw}) has no consistent
solution if we include all possibilities for \(\mathbf{y}_v\), but does
if we exclude cases where all \(y\)s have the same value. If \(n_v=4\),
we show that there are no consistent solutions in either case.

The following R function computes AUC from \(\mathbf{\widehat{y}}_v\)
and \(\mathbf{y}_v\), accommodating the cases where \(\mathbf{y}_v\) is
all 0s or all 1s:

\begin{CodeChunk}
\begin{CodeInput}
R> AUC <- function(y, yhat = seq_along(y)) {
+   s <- sum(y)
+   if (s == 0) return(0)
+   if (s == length(y)) return(1)
+   Metrics::auc(y, yhat)
+ }
\end{CodeInput}
\end{CodeChunk}

We then define a function to generate all possible \(\mathbf{y}_v\)s of
length \(n_v\) as rows of the matrix
\(\mathbf{Y}_{(2^{n_v} \times n_v)}\):

\begin{CodeChunk}
\begin{CodeInput}
R> Ymat <- function(n_v, exclude_identical = FALSE) {
+   stopifnot(n_v > 0 && round(n_v) == n_v)    # n_v must be a positive integer
+   ret <- sapply(0:(2^n_v - 1),
+                 function(x) as.integer(intToBits(x)) )[1:n_v, ]
+   ret <- if (is.matrix(ret)) t(ret) else matrix(ret)
+   colnames(ret) <- paste0("y", 1:ncol(ret))
+   if (exclude_identical) ret[-c(1, nrow(ret)), ] else ret
+ }
\end{CodeInput}
\end{CodeChunk}

For \(n_v=3\),

\begin{CodeChunk}
\begin{CodeInput}
R> Ymat(3)
\end{CodeInput}
\begin{CodeOutput}
     y1 y2 y3
[1,]  0  0  0
[2,]  1  0  0
[3,]  0  1  0
[4,]  1  1  0
[5,]  0  0  1
[6,]  1  0  1
[7,]  0  1  1
[8,]  1  1  1
\end{CodeOutput}
\end{CodeChunk}

If we exclude \(\mathbf{y}_v\)s with identical values, then

\begin{CodeChunk}
\begin{CodeInput}
R> Ymat(3, exclude_identical = TRUE)
\end{CodeInput}
\begin{CodeOutput}
     y1 y2 y3
[1,]  1  0  0
[2,]  0  1  0
[3,]  1  1  0
[4,]  0  0  1
[5,]  1  0  1
[6,]  0  1  1
\end{CodeOutput}
\end{CodeChunk}

Here is \(\mathbf{Y}\) with corresponding values of AUC:

\begin{CodeChunk}
\begin{CodeInput}
R> cbind(Ymat(3), AUC = apply(Ymat(3), 1, AUC))
\end{CodeInput}
\begin{CodeOutput}
     y1 y2 y3 AUC
[1,]  0  0  0 0.0
[2,]  1  0  0 0.0
[3,]  0  1  0 0.5
[4,]  1  1  0 0.0
[5,]  0  0  1 1.0
[6,]  1  0  1 0.5
[7,]  0  1  1 1.0
[8,]  1  1  1 1.0
\end{CodeOutput}
\end{CodeChunk}

The values of \(\mathrm{cv}(\widehat{y}_i, y_i)\) that express AUC as a
sum of casewise values are solutions of equation (\ref{eq:cw}), which
can be written as solutions of the following system of \(2^{n_v}\)
linear simultaneous equations in \(2n_v\) unknowns: \begin{equation}
\label{eq:lin}
\tag{2}
(\mathbf{U} -\mathbf{Y}) \mathbf{c}_0 + \mathbf{Y} \mathbf{c}_1
=
[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]
\begin{bmatrix}
\mathbf{c}_0 \\ \mathbf{c}_1
\end{bmatrix}
= \mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})
\end{equation} where \(\mathbf{U}_{(2^{n_v} \times n_v)}\) is a matrix
of 1s conformable with \(\mathbf{Y}\);
\(\mathbf{c}_0 = [\mathrm{cv}(1,0), c(2,0), ..., \mathrm{cv}(n_v,0)]^T\);
\(\mathbf{c}_1 = [\mathrm{cv}(1,1), c(2,1), ..., \mathrm{cv}(n_v,1)]^T\);
\([\mathbf{U} -\mathbf{Y}, \mathbf{Y}]_{(2^{n_v} \times 2n_v)}\) and
\(\begin{bmatrix}\begin{aligned} \mathbf{c}_0 \\ \mathbf{c}_1 \end{aligned} \end{bmatrix}_{(2n_v \times 1)}\)
are partitioned matrices; and
\(\mathbf{\widehat{Y}}_{(2^{n_v} \times n_v)}\) is a matrix each of
whose rows consists of the integers 1 to \(n_v\).

We can test whether equation (\ref{eq:lin}) has a solution for any given
\(n_v\) by trying to solve it as a least-squares problem, considering
whether the residuals of the associated linear model are all 0, using
the ``design matrix'' \([\mathbf{U} -\mathbf{Y}, \mathbf{Y}]\) to
predict the ``outcome''
\(\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})_{(2^{n_v} \times 1)}\):

\begin{CodeChunk}
\begin{CodeInput}
R> resids <- function(n_v, exclude_identical = FALSE, 
+                    tol = sqrt(.Machine$double.eps)) {
+   Y <- Ymat(n_v, exclude_identical = exclude_identical)
+   AUC <- apply(Y, 1, AUC)
+   X <- cbind(1-Y, Y)
+   opts <- options(warn = -1)
+   on.exit(options(opts))
+   fit <- lsfit(X, AUC, intercept = FALSE)
+   ret <- max(abs(residuals(fit)))
+   if(ret < tol){
+     ret <- 0
+     solution <- coef(fit)
+     names(solution) <- paste0("c(", c(1:n_v, 1:n_v), ",", 
+                               rep(0:1, each = n_v), ")")
+     attr(ret, "solution") <- zapsmall(solution)
+   }
+   ret
+ }
\end{CodeInput}
\end{CodeChunk}

The case \(n_v=3\), excluding identical \(y\)s, has a solution:

\begin{CodeChunk}
\begin{CodeInput}
R> resids(3, exclude_identical = TRUE)
\end{CodeInput}
\begin{CodeOutput}
[1] 0
attr(,"solution")
c(1,0) c(2,0) c(3,0) c(1,1) c(2,1) c(3,1) 
   1.0    0.0   -0.5    0.5    0.0    0.0 
\end{CodeOutput}
\end{CodeChunk}

But, if identical \(y\)s are included, the equation is not consistent:

\begin{CodeChunk}
\begin{CodeInput}
R> resids(3, exclude_identical = FALSE)
\end{CodeInput}
\begin{CodeOutput}
[1] 0.125
\end{CodeOutput}
\end{CodeChunk}

For \(n_v=4\), there are no solutions in either case:

\begin{CodeChunk}
\begin{CodeInput}
R> resids(4, exclude_identical = TRUE)
\end{CodeInput}
\begin{CodeOutput}
[1] 0.08333333
\end{CodeOutput}
\begin{CodeInput}
R> resids(4, exclude_identical = FALSE)
\end{CodeInput}
\begin{CodeOutput}
[1] 0.25
\end{CodeOutput}
\end{CodeChunk}

Consequently, the widely employed AUC measure of fit for binary
regression cannot in general be used for a casewise cross-validation
criterion.

\bibliography{cv.bib}

\end{document}

### 2024-01-02 scratch/notes_john/JSS-article/JSS-article.Rmd
use JSS markup \pkg() etc.

@@
 title:
-  formatted: "\\pkg{cv}: An R Package for Cross-Validation of Regression Models"
+  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validation of Regression Models"
   # If you use tex in the formatted title, also supply version without
@@
 
-The data for this example are drawn from the **ISLR2** package for R, associated with @JamesEtAl:2021.
+The data for this example are drawn from the \pkg{ISLR2} package for \proglang{R}, associated with @JamesEtAl:2021.
 The presentation here is close (though not identical) to that in the original source [@JamesEtAl:2021 Secs.
-5.1, 5.3], and it demonstrates the use of the `cv()` function in the **cv** package.[^boot]
-
-[^boot]: @JamesEtAl:2021 use the `cv.glm()` function in the **boot** package [@CantyRipley2022; @DavisonHinkley:1997].
+5.1, 5.3], and it demonstrates the use of the `cv()` function in the \pkg{cv} package.[^boot]
+
+[^boot]: @JamesEtAl:2021 use the `cv.glm()` function in the \pkg{boot} package [@CantyRipley2022; @DavisonHinkley:1997].
 Despite its name, `cv.glm()` is an independent function and not a method of a `cv()` generic function.
@@
 ```
-The code for this graph uses the `mse()` function from the **cv** package to compute the MSE for each fit.
+The code for this graph uses the `mse()` function from the \pkg{cv} package to compute the MSE for each fit.
 
@@
 ```
-The `"lm"` method by default uses `mse()` as the CV criterion and the Woodbury matrix identity to update the regression with each fold deleted without having literally to refit the model.
+The \code{"lm"} method by default uses \code{mse()} as the CV criterion and the Woodbury matrix identity to update the regression with each fold deleted without having literally to refit the model.
 Computational details are discussed in the final section of this vignette.
 The function reports the CV estimate of MSE, a biased-adjusted estimate of the MSE (the bias adjustment is explained in the final section), and the MSE is also computed for the original, full-sample regression.
-Because the division of the data into 10 folds is random, `cv()` explicitly (randomly) generates and saves a seed for R's pseudo-random number generator, to make the results replicable.
-The user can also specify the seed directly via the `seed` argument to `cv()`.
-
-To perform LOO CV, we can set the `k` argument to `cv()` to the number of cases in the data, here `k=392`, or, more conveniently, to `k="loo"` or `k="n"`:
+Because the division of the data into 10 folds is random, \code{cv()} explicitly (randomly) generates and saves a seed for \proglang{R}'s pseudo-random number generator, to make the results replicable.
+The user can also specify the seed directly via the \code{seed} argument to \code{cv()}.
+
+To perform LOO CV, we can set the \code{k} argument to \code{cv()} to the number of cases in the data, here \code{k=392}, or, more conveniently, to \code{k="loo"} or \code{k="n"}:
 
@@
 ```
-For LOO CV of a linear model, `cv()` by default uses the hatvalues from the model fit to the full data for the LOO updates, and reports only the CV estimate of MSE.
+For LOO CV of a linear model, \code{cv()} by default uses the hatvalues from the model fit to the full data for the LOO updates, and reports only the CV estimate of MSE.
 Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted.
@@
 ```
-The `"naive"` and `"Woodbury"` methods also return the bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an issue for LOO CV.
+The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an issue for LOO CV.
 
 This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed.
-In this comparison, we include the `cv.glm()` function from the **boot** package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM.
-We use the `microbenchmark()` function from the package of the same name for the timings [@Mersmann:2023]:
+In this comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM.
+We use the \code{microbenchmark()} function from the package of the same name for the timings [@Mersmann:2023]:
 ```{r cv.lm.timings, cache=TRUE}
@@
 
-[^microbenchmark]: Out of impatience, we asked `microbenchmark()` to execute each command only 10 times rather than the default 100.
+[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100.
 With the exception of the last columns, the output is self-explanatory.
 The last column shows which methods have average timings that are statistically distinguishable.
-Because of the small number of repetitions (i.e., 10), the `"hatvalues"` and `"Woodbury"` methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default `times=100` repetitions.
+Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default \code{times=100} repetitions.
 
@@
 
-The `cv()` function also has a method that can be applied to a list of regression models for the same data, composed using the `models()` function.
+The \code{cv()} function also has a method that can be applied to a list of regression models for the same data, composed using the \code{models()} function.
 For $k$-fold CV, the same folds are used for the competing models, which reduces random error in their comparison.
-This result can also be obtained by specifying a common seed for R's random-number generator while applying `cv()` separately to each model, but employing a list of models is more convenient for both $k$-fold and LOO CV (where there is no random component to the composition of the $n$ folds).
-
-We illustrate with the polynomial regression models of varying degree for the `Auto` data (discussed previously), beginning by fitting and saving the 10 models:
+This result can also be obtained by specifying a common seed for \proglang{R}'s random-number generator while applying \code{cv()} separately to each model, but employing a list of models is more convenient for both $k$-fold and LOO CV (where there is no random component to the composition of the $n$ folds).
+
+We illustrate with the polynomial regression models of varying degree for the \code{Auto} data (discussed previously), beginning by fitting and saving the 10 models:
 ```{r polyomial-models}
@@
 ```
-We then apply `cv()` to the list of 10 models (the `data` argument is required):
+We then apply \code{cv()} to the list of 10 models (the \code{data} argument is required):
 ```{r polynomial-regression-CV}
@@
 ```
-Because we didn't supply names for the models in the calls to the `models()` function, the names `model.1`, `model.2`, etc., are generated by the function.
+Because we didn't supply names for the models in the calls to the \code{models()} function, the names \code{model.1}, \code{model.2}, etc., are generated by the function.
 
@@
 
-Alternatively, we can use the `plot()` method for `"cvModList"` objects to compare the models, though with separate graphs for 10-fold and LOO CV:
+Alternatively, we can use the \code{plot()} method for \code{"cvModList"} objects to compare the models, though with separate graphs for 10-fold and LOO CV:
 
@@
 
-The `Mroz` data set from the **carData** package [associated with @FoxWeisberg:2019] has been used by several authors to illustrate binary logistic regression; see, in particular @FoxWeisberg:2019.
+The \code{Mroz} data set from the \pkg{carData} package [associated with @FoxWeisberg:2019] has been used by several authors to illustrate binary logistic regression; see, in particular @FoxWeisberg:2019.
 The data were originally drawn from the U.S.
@@
 ```
-The response variable in the logistic regression is `lfp`, labor-force participation, a factor coded `"yes"` or `"no"`. The remaining variables are predictors:
-
-* `k5`, number of children 5 years old of younger in the woman's household;
-* `k618`, number of children between 6 and 18 years old;
-* `age`, in years;
-* `wc`, wife's college attendance, `"yes"` or `"no"`;
-* `hc`, husband's college attendance;
-* `lwg`, the woman's log wage rate if she is employed, or her *imputed* wage rate, if she is not [a variable that @FoxWeisberg:2019 show is problematically defined]; and
-* `inc`, family income, in $1000s, exclusive of wife's income.
-
-We use the `glm()` function to fit a binary logistic regression to the `Mroz` data:
+The response variable in the logistic regression is \code{lfp}, labor-force participation, a factor coded \code{"yes"} or \code{"no"}. The remaining variables are predictors:
+
+* \code{k5}, number of children 5 years old of younger in the woman's household;
+* \code{k618}, number of children between 6 and 18 years old;
+* \code{age}, in years;
+* \code{wc}, wife's college attendance, \code{"yes"} or \code{"no"};
+* \code{hc}, husband's college attendance;
+* \code{lwg}, the woman's log wage rate if she is employed, or her *imputed* wage rate, if she is not [a variable that @FoxWeisberg:2019 show is problematically defined]; and
+* \code{inc}, family income, in $1000s, exclusive of wife's income.
+
+We use the \code{glm()} function to fit a binary logistic regression to the \code{Mroz} data:
 ```{r Mroz-logistic-regresion}
@@
 ```
-In addition to the usually summary output for a GLM, we show the result of applying the `BayesRule()` function from the **cv** package to predictions derived from the fitted model.
-Bayes rule, which predicts a "success" in a binary regression model when the fitted probability of success [i.e., $\phi = \Pr(y = 1)$] is $\widehat{\phi} \ge .5$ and a "failure" if $\widehat{\phi} < .5$.[^BayesRule] The first argument to `BayesRule()` is the binary {0, 1} response, and the second argument is the predicted probability of success.
-`BayesRule()` returns the proportion of predictions that are *in error*, as appropriate for a "cost" function.
-
-[^BayesRule]: `BayesRule()` does some error checking; `BayesRule2()` is similar, but omits the error checking, and so can be faster for large problems.
-
-The value returned by `BayesRule()` is associated with an "attribute" named `"casewise loss"` and set to `"y != round(yhat)"`, signifying that the Bayes rule CV criterion is computed as the mean of casewise values, here 0 if the prediction for a case matches the observed value and 1 if it does not (signifying a prediction error).
-The `mse()` function for numeric responses is also calculated as a casewise average.
-Some other criteria, such as the median absolute error, computed by the `medAbsErr()` function in the **cv** package, aren't averages of casewise components.
+In addition to the usually summary output for a GLM, we show the result of applying the \code{BayesRule()} function from the \pkg{cv} package to predictions derived from the fitted model.
+Bayes rule, which predicts a "success" in a binary regression model when the fitted probability of success [i.e., $\phi = \Pr(y = 1)$] is $\widehat{\phi} \ge .5$ and a "failure" if $\widehat{\phi} < .5$.[^BayesRule] The first argument to \code{BayesRule()} is the binary {0, 1} response, and the second argument is the predicted probability of success.
+\code{BayesRule()} returns the proportion of predictions that are *in error*, as appropriate for a "cost" function.
+
+[^BayesRule]: \code{BayesRule()} does some error checking; \code{BayesRule2()} is similar, but omits the error checking, and so can be faster for large problems.
+
+The value returned by \code{BayesRule()} is associated with an "attribute" named \code{"casewise loss"} and set to \code{"y != round(yhat)"}, signifying that the Bayes rule CV criterion is computed as the mean of casewise values, here 0 if the prediction for a case matches the observed value and 1 if it does not (signifying a prediction error).
+The \code{mse()} function for numeric responses is also calculated as a casewise average.
+Some other criteria, such as the median absolute error, computed by the \code{medAbsErr()} function in the \pkg{cv} package, aren't averages of casewise components.
 The distinction is important because, to our knowledge, the statistical theory of cross-validation, for example, in @DavisonHinkley:1997, @BatesHastieTibshirani:2023, and @ArlotCelisse:2010, is developed for CV criteria like MSE that are means of casewise components.
@@
 
-The `"glm"` method for `cv()` is largely similar to the `"lm"` method, although the default algorithm, selected explicitly by `method="exact"`, refits the model with each fold removed (and is thus equivalent to `method="naive"` for `"lm"` models).
-For generalized linear models, `method="Woodbury"` or (for LOO CV) `method="hatvalues"` provide approximate results (see the last section of the vignette for details):
+The \code{"glm"} method for \code{cv()} is largely similar to the \code{"lm"} method, although the default algorithm, selected explicitly by \code{method="exact"}, refits the model with each fold removed (and is thus equivalent to \code{method="naive"} for \code{"lm"} models).
+For generalized linear models, \code{method="Woodbury"} or (for LOO CV) \code{method="hatvalues"} provide approximate results (see the last section of the vignette for details):
 
@@
 ```
-To ensure that the two methods use the same 10 folds, we specify the seed for R's random-number generator explicitly; here, and as is common in our experience, the `"exact"` and `"Woodbury"` algorithms produce nearly identical results.
+To ensure that the two methods use the same 10 folds, we specify the seed for \proglang{R}'s random-number generator explicitly; here, and as is common in our experience, the \code{"exact"} and \code{"Woodbury"} algorithms produce nearly identical results.
 The CV estimates of prediction error are slightly higher than the estimate based on all of the cases.
@@
 The printed output includes a 95% confidence interval for the bias-adjusted Bayes rule CV criterion.
-@BatesHastieTibshirani:2023 show that these confidence intervals are unreliable for models fit to small samples, and by default `cv()` computes them only when the sample size is 400 or larger and when the CV criterion employed is an average of casewise components, as is the case for Bayes rule.
+@BatesHastieTibshirani:2023 show that these confidence intervals are unreliable for models fit to small samples, and by default \code{cv()} computes them only when the sample size is 400 or larger and when the CV criterion employed is an average of casewise components, as is the case for Bayes rule.
 See the final section of the vignette for details of the computation of confidence intervals for bias-adjusted CV criteria.
@@
 
-As for linear models, we report some timings for the various `cv()` methods of computation in LOO CV as well as for the `cv.glm()` function from the **boot** package (which, recall, refits the model with each case removed, and thus is comparable to `cv()` with `method="exact"`):
+As for linear models, we report some timings for the various \code{cv()` methods of computation in LOO CV as well as for the \code{cv.glm()} function from the \pkg{boot} package (which, recall, refits the model with each case removed, and thus is comparable to \code{cv()} with \code{method="exact"}):
 ```{r glm.timings, cache=TRUE}
@@
 In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed.
-The `lmer()` function in the **lme4** package is capable of modeling both nested and crossed random effects, and the `cv()` methods for mixed models in the **cv** package pertain to both nested and crossed random effects.
+The \code{lmer()} function in the \pkg{lme4} package is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects.
 We present an example of the latter later in the vignette.
@@
 
-The HSB data are included in the `MathAchieve` and `MathAchSchool` data sets in the **nlme** package  [@PinheiroBates:2000].
-`MathAchieve` includes individual-level data on 7185  students in 160 high schools, and `MathAchSchool` includes school-level data:
+The HSB data are included in the \code{MathAchieve} and \code{MathAchSchool} data sets in the \pkg{nlme} package  [@PinheiroBates:2000].
+\code{MathAchieve} includes individual-level data on 7185  students in 160 high schools, and \code{MathAchSchool} includes school-level data:
 ```{r HSB-data}
@@
 
-We'll use only the `School`, `SES` (students' socioeconomic status), and `MathAch` (their score on a standardized math-achievement test) variables in the `MathAchieve` data set, and `Sector` (`"Catholic"` or `"Public"`) in the `MathAchSchool` data set.
-
-Some data-management is required before fitting a mixed-effects model to the HSB data, for which we use the **dplyr** package [@WickhamEtAl:2023]:
+We'll use only the \code{School}, \code{SES} (students' socioeconomic status), and \code{MathAch} (their score on a standardized math-achievement test) variables in the \code{MathAchieve} data set, and \code{Sector} (\code{"Catholic"} or \code{"Public"}) in the \code{MathAchSchool} data set.
+
+Some data-management is required before fitting a mixed-effects model to the HSB data, for which we use the \pkg{dplyr} package [@WickhamEtAl:2023]:
 ```{r HSB-data-management, cache=TRUE}
@@
 ```
-In the process, we created two new school-level variables: `meanses`, which is the average SES for students in each school; and `cses`, which is school-average SES centered at its mean.
+In the process, we created two new school-level variables: \code{meanses}, which is the average SES for students in each school; and \code{cses}, which is school-average SES centered at its mean.
 For details, see @FoxWeisberg:2019 [Sec.
@@
 
-Still following Fox and Weisberg, we proceed to use the `lmer()` function in the **lme4** package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
+Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
 ```{r HSB-lmer, cache=TRUE}
@@
 ```
-For cluster-level CV, the `clusterVariables` argument tells `cv()` how the clusters are defined.
-Were there more than one clustering variable, say classes within schools, these would be provided as a character vector of variable names: `clusterVariables = c("school", "class")`.
-For cluster-level CV, the default is `k = "loo"`, that is, leave one cluster out at a time; we instead specify `k = 10` folds of clusters, each fold therefore comprising $160/10 = 16$ schools.
-
-If the `clusterVariables` argument is omitted, then case-level CV is employed, with `k = 10` folds as the default, here each with $7185/10 \approx 719$ students.
+For cluster-level CV, the \code{clusterVariables} argument tells \code{cv()} how the clusters are defined.
+Were there more than one clustering variable, say classes within schools, these would be provided as a character vector of variable names: \code{clusterVariables = c("school", "class")}.
+For cluster-level CV, the default is \code{k = "loo"}, that is, leave one cluster out at a time; we instead specify \code{k = 10} folds of clusters, each fold therefore comprising $160/10 = 16$ schools.
+
+If the \code{clusterVariables} argument is omitted, then case-level CV is employed, with \code{k = 10} folds as the default, here each with $7185/10 \approx 719$ students.
 Notice that one of the 10 models refit with a fold removed failed to converge.
@@
 
-There is also a `cv()` method for linear mixed models fit by the `lme()` function in the **nlme** package, and the arguments for `cv()` in this case are the same as for a model fit by `lmer()` or `glmer()`.
+There is also a \code{cv()} method for linear mixed models fit by the \code{lme()} function in the \pkg{nlme} package, and the arguments for \code{cv()} in this case are the same as for a model fit by \code{lmer()} or \code{glmer()}.
 We illustrate with the mixed model fit to the HSB data:
@@
 ```
-We used the same random-number generator seeds as in the previous example cross-validating the model fit by `lmer()`, and so the same folds are employed in both cases.[^optimizer] The estimated covariance components and fixed effects in the summary output differ slightly between the `lmer()` and `lme()` solutions, although both functions seek to maximize the REML criterion.
+We used the same random-number generator seeds as in the previous example cross-validating the model fit by \code{lmer()}, and so the same folds are employed in both cases.[^optimizer] The estimated covariance components and fixed effects in the summary output differ slightly between the \code{lmer()} and \code{lme()} solutions, although both functions seek to maximize the REML criterion.
 This is, of course, to be expected when different algorithms are used for numerical optimization.
-To the precision reported, the cluster-level CV results for the `lmer()` and `lme()` models are identical, while the case-level CV results are very similar but not identical.
-
-[^optimizer]: The observant reader will notice that we set the argument `control=list(opt="optim")` in the call to `lme()`, changing the optimizer employed from the default `"nlminb"`.
-We did this because with the default optimizer, `lme()` encountered the same convergence issue as `lmer()`, but rather than issuing a warning, `lme()` failed, reporting an error.
-As it turns out, setting the optimizer to `"optim"` avoids this problem.
+To the precision reported, the cluster-level CV results for the \code{lmer()} and \code{lme()} models are identical, while the case-level CV results are very similar but not identical.
+
+[^optimizer]: The observant reader will notice that we set the argument \code{control=list(opt="optim")} in the call to \code{lme()}, changing the optimizer employed from the default \code{"nlminb"}.
+We did this because with the default optimizer, \code{lme()} encountered the same convergence issue as \code{lmer()}, but rather than issuing a warning, \code{lme()} failed, reporting an error.
+As it turns out, setting the optimizer to \code{"optim"} avoids this problem.
 
@@
 
-[^2]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.
+[^2]: We find it convenient to use the \pkg{lattice} [@Sarkar:2008] and \pkg{latticeExtra} [@SarkarAndrews:2022] packages for this and other graphs in this section.
 
@@
 
-We proceed to obtain predictions from the random-intercept model (`mod.0`) and the other models (`mod.1`, `mod.2`, and `mod.3`) based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students):
+We proceed to obtain predictions from the random-intercept model (\code{mod.0}) and the other models (\code{mod.1}, \code{mod.2}, and \code{mod.3}) based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students):
 
@@
 
-Predictions based on the random-intercept model `mod.0` for the first 10 students are shown in the following graph:
+Predictions based on the random-intercept model \code{mod.0} for the first 10 students are shown in the following graph:
 
@@
 
-Our next model, `mod.1`, includes a fixed intercept and fixed effect of `x` along with a random intercept:
+Our next model, \code{mod.1}, includes a fixed intercept and fixed effect of \code{x} along with a random intercept:
 
@@
 
-The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, `mod.0`.
-We therefore anticipate (and show later in this section) that case-based cross-validation will prefer `mod1` to `mod0`, but that cluster-based cross-validation will prefer `mod0` to `mod1`.
-
-Our third model, `mod.2`, includes the contextual effect of $x$---that is, the cluster mean `xm`---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:
+The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, \code{mod.0}.
+We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod1} to \code{mod0}, but that cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.
+
+Our third model, \code{mod.2}, includes the contextual effect of $x$---that is, the cluster mean \code{xm]---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:
 
@@
 
-This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | group)`, which is the model that generated the data once the coefficient of the contextual predictor `xm` is set to 0 (as it is in `mod.3`, discussed below).
-
-Predictions from model `mod.2` appear in the following graph:
+This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that generated the data once the coefficient of the contextual predictor \code{xm} is set to 0 (as it is in \code{mod.3], discussed below).
+
+Predictions from model \code{mod.2} appear in the following graph:
 ```{r plot-fits-mod2}
@@
 
-Depending on the estimated variance parameters of the model, a mixed model like `mod.2` will apply varying degrees of shrinkage to the random-intercept BLUPs that correspond to variation in the heights of the parallel fitted lines for the individual students.
-In our contrived data, the `mod.2` applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student.
-The fit of the mixed model `mod.2` is consequently similar to that of a fixed-effects model with age and a categorical predictor for individual students (i.e., treating students as a factor, and not shown here).
-
-The  mixed model `mod.2` therefore fits individual observations well, and we anticipate a favorable assessment using individual-based cross-validation.
-In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for model `mod.2` compared to the smaller model `mod.0`, which includes only fixed and random intercepts.
+Depending on the estimated variance parameters of the model, a mixed model like \code{mod.2} will apply varying degrees of shrinkage to the random-intercept BLUPs that correspond to variation in the heights of the parallel fitted lines for the individual students.
+In our contrived data, the \code{mod.2} applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student.
+The fit of the mixed model \code{mod.2} is consequently similar to that of a fixed-effects model with age and a categorical predictor for individual students (i.e., treating students as a factor, and not shown here).
+
+The  mixed model \code{mod.2} therefore fits individual observations well, and we anticipate a favorable assessment using individual-based cross-validation.
+In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for model \code{mod.2} compared to the smaller model \code{mod.0}, which includes only fixed and random intercepts.
 
@@
 
-Finally, `mod.3` directly estimates the model used to generate the data.
-As mentioned, it is a constrained version of `mod.2`, with the coefficient of `xm` set to 0, and with `x` expressed as a deviation from the cluster mean `xm`:
+Finally, \code{mod.3} directly estimates the model used to generate the data.
+As mentioned, it is a constrained version of \code{mod.2}, with the coefficient of \code{xm} set to 0, and with \code{x} expressed as a deviation from the cluster mean \code{xm}:
 
@@
 
-The predictions from `mod.3` are therefore similar to those from `mod.2`:
+The predictions from \code{mod.3} are therefore similar to those from \code{mod.2}:
 
@@
 We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.
-In order to reduce between-model random variability in comparisons of models, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model:
+In order to reduce between-model random variability in comparisons of models, we apply \code{cv()} to the list of models created by the \code{models()} function (introduced previously), performing cross-validation with the same folds for each model:
 
@@
 
-In summary, model `mod.1`, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
-Model `mod.2`, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV.
-The data-generating model, `mod.3`, which includes the fixed effect of `x - xm` in place of separate terms in `x` and `xm`, isn't distinguishable from model `mod.2`, which includes `x` and `xm` separately, even though `mod.2` has an unnecessary parameter (recall that the population coefficient of `xm` is 0 when `x` is expressed as deviations from the contextual mean).
+In summary, model \code{mod.1}, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
+Model \code{mod.2}, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV.
+The data-generating model, \code{mod.3}, which includes the fixed effect of \code{x - xm} in place of separate terms in \code{x} and \code{xm}, isn't distinguishable from model \code{mod.2}, which includes \code{x} and \code{xm} separately, even though \code{mod.2} has an unnecessary parameter (recall that the population coefficient of \code{xm} is 0 when \code{x} is expressed as deviations from the contextual mean).
 These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
@@
 
-We'll illustrate with data on pig growth, introduced by @DiggleLiangZeger:1994 [Table 3.1]. The data are in the `Pigs` data frame in the **cv** package:
+We'll illustrate with data on pig growth, introduced by @DiggleLiangZeger:1994 [Table 3.1]. The data are in the \code{Pigs} data frame in the \pkg{cv} package:
 ```{r pigs}
@@
 Each of 48 pigs is observed weekly over a period of 9 weeks, with the weight of the pig recorded in kg.
-The data are in "long" format, as is appropriate for use with the `lmer()` function in the **lme4** package.
+The data are in "long" format, as is appropriate for use with the \code{lmer()} function in the \pkg{lme4} package.
 The data are very regular, with no missing cases.
@@
 
-The **Stata** mixed-effects models manual proposes a model with crossed random effects for the `Pigs` data [@Stata:2023 page 37]:
+The \proglang{Stata} mixed-effects models manual proposes a model with crossed random effects for the \code{Pigs} data [@Stata:2023 page 37]:
 
@@
 [^pig-alternative]: These are repeated-measures data, which would be more conventionally modeled with autocorrelated errors within pigs.
-The `lme()` function in the **nlme** package, for example, is capable of fitting a mixed-model of this form.
-
-The **Stata** manual fits the mixed model by maximum likelihood (rather than REML), and we duplicate the results reported there using `lmer()`:
+The \code{lme()} function in the \pkg{nlme} package, for example, is capable of fitting a mixed-model of this form.
+
+The \proglang{Stata} manual fits the mixed model by maximum likelihood (rather than REML), and we duplicate the results reported there using \code{lmer()}:
 ```{r pigs-lmer}
@@
 ```
-We opt for the non-default `"bobyqa"` optimizer because it provides more numerically stable results for subsequent cross-validation in this example.
-
-We can then cross-validate the model by omitting folds composed of pigs, folds composed of weeks, or folds composed of pig-weeks (which in the `Pigs` data set correspond to individual cases, using only the fixed effects):
+We opt for the non-default \code{"bobyqa"} optimizer because it provides more numerically stable results for subsequent cross-validation in this example.
+
+We can then cross-validate the model by omitting folds composed of pigs, folds composed of weeks, or folds composed of pig-weeks (which in the \code{Pigs} data set correspond to individual cases, using only the fixed effects):
 ```{r pigs-cv}
@@
 [^crossed-vs-nested]: Even though there is only one observation per combination of pigs and weeks, we can use the BLUP for the omitted case because of the crossed structure of the random effects; that is each pig-week has a pig random effect and a week random effect.
-Although it probably isn't sensible, we can imagine a mixed model for the pig data that employs nested random effects, which would be specified by `lmer(weight ~ week + (1 | id/week), data=Pigs)`---that is, a random intercept that varies by combinations of `id` (pig) and `week`.
-This model can't be fit, however: With only one case per combination of `id` and `week`, the nested random-effect variance is indistinguishable from the case-level variance.
+Although it probably isn't sensible, we can imagine a mixed model for the pig data that employs nested random effects, which would be specified by \code{lmer(weight ~ week + (1 | id/week), data=Pigs)}---that is, a random intercept that varies by combinations of \code{id} (pig) and \code{week}.
+This model can't be fit, however: With only one case per combination of \code{id} and \code{week}, the nested random-effect variance is indistinguishable from the case-level variance.
 
@@
 (An exception is LOO cross-validation, which is not at all random.) To get a sense of the magnitude of the sampling error, we can repeat the CV procedure with different randomly selected partitions into folds.
-All of the CV functions in the **cv** package are capable of repeated cross-validation, with the number of repetitions controlled by the `reps` argument, which defaults to `1`.
+All of the CV functions in the \pkg{cv} package are capable of repeated cross-validation, with the number of repetitions controlled by the \code{reps} argument, which defaults to \code{1}.
 
@@
 ```
-When `reps` > `1`, the result returned by `cv()` is an object of class `"cvList"`---literally a list of `"cv"` objects.
+When \code{reps} > \code{1}, the result returned by \code{cv()} is an object of class \code{"cvList"}---literally a list of \code{"cv"} objects.
 The results are reported for each repetition and then averaged across repetitions, with the standard deviations of the CV criterion and the biased-adjusted CV criterion given in parentheses.
@@
 
-Notice that the seed that's set in the `cv()` command pertains to the first repetition and the seeds for the remaining repetitions are then selected pseudo-randomly.[^replicates] Setting the first seed, however, makes the entire process easily replicable, and the seed for each repetition is stored in the corresponding element of the `"cvList"` object (which isn't, however, saved in the example).
-
-[^replicates]: Because of the manner in which the computation is performed, the order of the replicates in the `"cvList"` object returned by `cv()` isn't the same as the order in which the replicates are computed.
-Each element of the result, however, is a `"cv"` object with the correct random-number seed saved, and so this technical detail can be safely ignored.
-The individual `"cv"` objects are printed in the order in which they are stored rather than the order in which they are computed.
-
-It's also possible to replicate CV when comparing competing models via the `cv()` method for `"modList"` objects.
-Recall our comparison of polynomial regressions of varying degree fit to the `Auto` data; we performed 10-fold CV for each of 10 models.
+Notice that the seed that's set in the \code{cv()} command pertains to the first repetition and the seeds for the remaining repetitions are then selected pseudo-randomly.[^replicates] Setting the first seed, however, makes the entire process easily replicable, and the seed for each repetition is stored in the corresponding element of the \code{"cvList"} object (which isn't, however, saved in the example).
+
+[^replicates]: Because of the manner in which the computation is performed, the order of the replicates in the \code{"cvList"} object returned by \code{cv()} isn't the same as the order in which the replicates are computed.
+Each element of the result, however, is a \code{"cv"} object with the correct random-number seed saved, and so this technical detail can be safely ignored.
+The individual \code{"cv"} objects are printed in the order in which they are stored rather than the order in which they are computed.
+
+It's also possible to replicate CV when comparing competing models via the \code{cv()} method for \code{"modList"} objects.
+Recall our comparison of polynomial regressions of varying degree fit to the \code{Auto} data; we performed 10-fold CV for each of 10 models.
 Here, we replicate that process 5 times for each model and graph the results:
@@
 ```
-Next, using the `stepAIC()` function in the **MASS** package [@VenablesRipley:2002], let us perform a forward stepwise regression to select a "best" model, starting with the null model, and using AIC as the model-selection criterion (see the help page for `stepAIC()` for details):[^selection-order]
+Next, using the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], let us perform a forward stepwise regression to select a "best" model, starting with the null model, and using AIC as the model-selection criterion (see the help page for \code{stepAIC()} for details):[^selection-order]
 
@@
 
-The `cvSelect()` function in the **cv** package allows us to cross-validate the whole model-selection procedure.
-The first argument to `cvSelect()` is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion.
-The `selectStepAIC()` function, also in **cv** and based on `stepAIC()`, is suitable for use with `cvSelect()`:
+The \code{cvSelect()} function in the \pkg{cv} package allows us to cross-validate the whole model-selection procedure.
+The first argument to \code{cvSelect()} is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion.
+The \code{selectStepAIC()} function, also in \pkg{cv} and based on \code{stepAIC()}, is suitable for use with \code{cvSelect()}:
 ```{r cvSelect-artificial-data, cache=TRUE}
@@
 
-The other arguments to `cvSelect()` are:
-
-* `data`, the data set to which the model is fit;
-* `seed`, an optional seed for R's pseudo-random-number generator; as for `cv()`, if the seed isn't supplied by the user, a seed is randomly selected and saved;
-* additional arguments required by the model-selection function, here the starting `model` argument, the `direction` of model selection, and the `scope` of models considered (from the model with only a regression constant to the model with all 100 predictors).
-
-By default, `cvSelect()` performs 10-fold CV, and produces an estimate of MSE for the model-selection procedure even *larger* than the true error variance, $\sigma^2 = 1$.
-
-Also by default, when the number of folds is 10 or fewer, `cvSelect()` saves the coefficients of the selected models.
-In this example, the `compareFolds()` function reveals that the variables retained by the model-selection process in the several folds are quite different:
+The other arguments to \code{cvSelect()} are:
+
+* \code{data}, the data set to which the model is fit;
+* \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, a seed is randomly selected and saved;
+* additional arguments required by the model-selection function, here the starting \code{model} argument, the \code{direction} of model selection, and the \code{scope} of models considered (from the model with only a regression constant to the model with all 100 predictors).
+
+By default, \code{cvSelect()} performs 10-fold CV, and produces an estimate of MSE for the model-selection procedure even *larger* than the true error variance, $\sigma^2 = 1$.
+
+Also by default, when the number of folds is 10 or fewer, \code{cvSelect()} saves the coefficients of the selected models.
+In this example, the \code{compareFolds()} function reveals that the variables retained by the model-selection process in the several folds are quite different:
 ```{r compare-selected-models}
@@
 For a contrasting example we apply model selection to Mroz's logistic regression for married women's labor-force participation.
-First, recall the logistic regression model that we fit to the `Mroz` data:
+First, recall the logistic regression model that we fit to the \code{Mroz} data:
 ```{r recall-Mroz-regression}
@@
 
-Applying stepwise model selection Mroz's logistic regression, using BIC as the model-selection criterion (via the argument `k=log(nrow(Mroz))` to `stepAIC()`) selects 5 of the 7 original predictors:
+Applying stepwise model selection Mroz's logistic regression, using BIC as the model-selection criterion (via the argument \code{k=log(nrow(Mroz))} to \code{stepAIC()}) selects 5 of the 7 original predictors:
 ```{r mroz-selection}
@@
 ```
-Bayes rule applied to the selected model misclassifies 32% of the cases in the `Mroz` data.
+Bayes rule applied to the selected model misclassifies 32% of the cases in the \code{Mroz} data.
 
@@
 ```
-Setting `AIC=FALSE` in the call to `cvSelect()` uses the BIC rather than the AIC as the model-selection criterion.
-As it turns out, exactly the same predictors are selected when each of the 10 folds are omitted, and the several coefficient estimates are very similar, as we show using `compareFolds()`:
+Setting \code{AIC=FALSE} in the call to \code{cvSelect()} uses the BIC rather than the AIC as the model-selection criterion.
+As it turns out, exactly the same predictors are selected when each of the 10 folds are omitted, and the several coefficient estimates are very similar, as we show using \code{compareFolds()}:
 ```{r compare-selected-models-mroz}
@@
 
-The **cv** package also provides a `cvSelect()` procedure, `selectTrans()`, for choosing transformations of the predictors and the response in regression.
+The \pkg{cv} package also provides a \code{cvSelect()} procedure, \code{selectTrans()}, for choosing transformations of the predictors and the response in regression.
 
@@
 To illustrate, we adapt an example appearing in several places in @FoxWeisberg:2019 (for example in Chapter 3 on transforming data), using data on the prestige and other characteristics of 102 Canadian occupations circa 1970.
-The data are in the `Prestige` data frame in the **carData** package:
+The data are in the \code{Prestige} data frame in the \pkg{carData} package:
 ```{r Prestige-data}
@@
 ```
-The variables in the `Prestige` data set are:
-
-* `education`: average years of education for incumbents in the occupation, from the 1971 Canadian Census.
-* `income`: average dollars of annual income for the occupation, from the Census.
-* `women`: percentage of occupational incumbents who were women, also from the Census.
-* `prestige`: the average prestige rating of the occupation on a 0--100 "thermometer" scale, in a Canadian social survey conducted around the same time.
-* `type`, type of occupation, and `census`, the Census occupational code, which are not used in our example.
-
-The object of a regression analysis for the `Prestige` data (and their original purpose) is to predict occupational prestige from the other variables in the data set.
-
-A scatterplot matrix (using the `scatterplotMatrix()` function in the **car** package) of the numeric variables in the data reveals that the distributions of `income` and `women` are positively skewed, and that some of the relationships among the three predictors, and between the predictors and the response (i.e., `prestige`), are nonlinear:
+The variables in the \code{Prestige} data set are:
+
+* \code{education}: average years of education for incumbents in the occupation, from the 1971 Canadian Census.
+* \code{income}: average dollars of annual income for the occupation, from the Census.
+* \code{women}: percentage of occupational incumbents who were women, also from the Census.
+* \code{prestige}: the average prestige rating of the occupation on a 0--100 "thermometer" scale, in a Canadian social survey conducted around the same time.
+* \code{type}, type of occupation, and \code{census}, the Census occupational code, which are not used in our example.
+
+The object of a regression analysis for the \code{Prestige} data (and their original purpose) is to predict occupational prestige from the other variables in the data set.
+
+A scatterplot matrix (using the \code{scatterplotMatrix()} function in the \pkg{car} package) of the numeric variables in the data reveals that the distributions of \code{income} and \code{women} are positively skewed, and that some of the relationships among the three predictors, and between the predictors and the response (i.e., \code{prestige}), are nonlinear:
 ```{r scatterplot-matrix}
@@
 ```
-The `powerTransform()` function in the **car** package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964].
+The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964].
 Several "families" of power transformations can be used, including the original Box-Cox family, simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap.
@@
 3] for details.
-Because `women` has some zero values, we use the Yeo-Johnson family:
+Because \code{women} has some 0 values, we use the Yeo-Johnson family:
 ```{r power-transform-Prestige}
@@
 ```
-We thus have evidence of the desirability of transforming `income` (by the $1/3$ power) and `women` (by the $0.16$ power---which is close to the "0" power, i.e., the log transformation), but not `education`.
+We thus have evidence of the desirability of transforming \code{income} (by the $1/3$ power) and \code{women} (by the $0.16$ power---which is close to the "0" power, i.e., the log transformation), but not \code{education}.
 Applying the "rounded" power transformations makes the predictors better-behaved:
@@
 ```
-Similarly, component+residual plots for the two regressions, produced by the `crPlots()` function in the **car** package, suggest that the partial relationship of `prestige` to `income` is more nearly linear in the transformed data, but the transformation of `women` fails to capture what appears to be a slight quadratic partial relationship; the partial relationship of `prestige` to `education` is close to linear in both regressions:
+Similarly, component+residual plots for the two regressions, produced by the \code{crPlots()} function in the \pkg{car} package, suggest that the partial relationship of \code{prestige} to \code{income} is more nearly linear in the transformed data, but the transformation of \code{women] fails to capture what appears to be a slight quadratic partial relationship; the partial relationship of \code{prestige} to \code{education} is close to linear in both regressions:
 ```{r CR-plots-untransformed}
@@
 
-Having transformed the predictors towards multinormality, we now consider whether there's evidence for transforming the response (using `powerTransform()` for Box and Cox's original method), and we discover that there's not:
+Having transformed the predictors towards multinormality, we now consider whether there's evidence for transforming the response (using \code{powerTransform()} for Box and Cox's original method), and we discover that there's not:
 ```{r transform-response}
@@
 
-The `selectTrans()` function in the **cv** package automates the process of selecting predictor and response transformations.
-The function takes a `data` set and "working" `model` as arguments, along with the candidate `predictors` and `response` for transformation, and the transformation `family` to employ.
-If the `predictors` argument is missing then only the response is transformed, and if the `response` argument is missing, only the supplied predictors are transformed.
-The default `family` for transforming the predictors is `"bcPower"`---the original Box-Cox family---as is the default `family.y` for transforming the response; here we specify `family="yjPower` because of the zeros in `women`.
-`selectTrans()` returns the result of applying a lack-of-fit criterion to the model after the selected transformation is applied, with the default `criterion=mse`:
+The \code{selectTrans()} function in the \pkg{cv} package automates the process of selecting predictor and response transformations.
+The function takes a \code{data} set and "working" \code{model} as arguments, along with the candidate \code{predictors} and \code{response} for transformation, and the transformation \code{family} to employ.
+If the \code{predictors} argument is missing then only the response is transformed, and if the \code{response} argument is missing, only the supplied predictors are transformed.
+The default \code{family} for transforming the predictors is \code{"bcPower"}---the original Box-Cox family---as is the default \code{family.y} for transforming the response; here we specify \code{family="yjPower} because of the 0s in \code{women}.
+\code{selectTrans()} returns the result of applying a lack-of-fit criterion to the model after the selected transformation is applied, with the default \code{criterion=mse}:
 ```{r selectTrans}
@@
 ```
-`selectTrans()` also takes an optional `indices` argument, making it suitable for doing computations on a subset of the data (i.e., a CV fold), and hence for use with `cvSelect()` (see `?selectTrans` for details):
+\code{selectTrans()} also takes an optional \code{indices} argument, making it suitable for doing computations on a subset of the data (i.e., a CV fold), and hence for use with \code{cvSelect()} (see \code{?selectTrans} for details):
 
@@
 The results suggest that the predictive power of the transformed regression is reliably greater than that of the untransformed regression (though in both case, the cross-validated MSE is considerably higher than the MSE computed for the whole data).
-Examining the selected transformations for each fold reveals that the predictor `education` and the response `prestige` are never transformed; that the $1/3$ power is selected for `income` in all of the folds; and that the transformation selected for `women` varies narrowly across the folds between the $0$th power (i.e., log) and the $1/3$ power.
+Examining the selected transformations for each fold reveals that the predictor \code{education} and the response \code{prestige} are never transformed; that the $1/3$ power is selected for \code{income} in all of the folds; and that the transformation selected for \code{women} varies narrowly across the folds between the $0$th power (i.e., log) and the $1/3$ power.
 
@@
 In this section, we consider the case where we both select variable transformations and then proceed to select predictors.
-It's insufficient to apply these steps sequentially, first, for example, using `cvSelect()` with `selectTrans()` and then with `selectStepAIC()`; rather we should apply the whole model-selection procedure with each fold omitted.
-The `selectTransAndStepAIC()` function, also supplied by the **cv** package, does exactly that.
-
-To illustrate this process, we return to the `Auto` data set:
+It's insufficient to apply these steps sequentially, first, for example, using \code{cvSelect()} with \code{selectTrans()} and then with \code{selectStepAIC()}; rather we should apply the whole model-selection procedure with each fold omitted.
+The \code{selectTransAndStepAIC()} function, also supplied by the \pkg{cv} package, does exactly that.
+
+To illustrate this process, we return to the \code{Auto} data set:
 ```{r Auto-redux}
@@
 ```
-We previously used the `Auto` here in a preliminary example where we employed CV to inform the selection of the order of a polynomial regression of `mpg` on `horsepower`.
-Here, we consider more generally the problem of predicting `mpg` from the other variables in the `Auto` data.
+We previously used the \code{Auto} here in a preliminary example where we employed CV to inform the selection of the order of a polynomial regression of \code{mpg} on \code{horsepower}.
+Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data.
 We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set:
@@
 ```
-A comment before we proceed: `origin` is clearly categorical and so converting it to a factor is natural, but we could imagine treating `cylinders` and `year` as numeric predictors.
-There are, however, only 5 distinct values of `cylinders` (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare.
+A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors.
+There are, however, only 5 distinct values of \code{cylinders} (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare.
 and none of the cars has 7 cylinders.
-There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between `mpg` and `year` is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.
-
-[^year]: Of course, making the decision to treat `year` as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process.
+There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.
+
+[^year]: Of course, making the decision to treat \code{year} as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process.
 It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation.
@@
 ```
-The component+residual plots, created with the `crPlots()` function in the previously loaded **car** package, clearly reveal the inadequacy of the model.
+The component+residual plots, created with the \code{crPlots()} function in the previously loaded \pkg{car} package, clearly reveal the inadequacy of the model.
 
@@
 
-Having transformed both the numeric predictors and the response, we proceed to use the `stepAIC()` function in the **MASS** package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log(n)$):
+Having transformed both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log(n)$):
 ```{r}
@@
 ```
-The selected model includes three of the numeric predictors, `horsepower`, `weight`, and `acceleration`, along with the factors `year` and `origin`.
+The selected model includes three of the numeric predictors, \code{horsepower}, \code{weight}, and \code{acceleration}, along with the factors \code{year} and \code{origin}.
 We can calculate the MSE for this model, but we expect that the result will be optimistic because we used the whole data to help specify the model
@@
 ```
-A perhaps subtle point is that we compute the MSE for the selected model on the original `mpg` response scale rather than the log scale, so as to make the selected model comparable to the working model.
-That's slightly uncomfortable given the skewed distribution of `mpg`.
-An alternative is to use the median absolute error instead of the mean-squared error, employing the `medAbsErr()` function from the **cv** package:
+A perhaps subtle point is that we compute the MSE for the selected model on the original \code{mpg} response scale rather than the log scale, so as to make the selected model comparable to the working model.
+That's slightly uncomfortable given the skewed distribution of \code{mpg}.
+An alternative is to use the median absolute error instead of the mean-squared error, employing the \code{medAbsErr()} function from the \pkg{cv} package:
 ```{r Auto-median-absolute-error}
@@
 
-Now let's use `cvSelect()` with `selectTransAndStepAIC()` to automate and cross-validate the whole model-specification process:
+Now let's use \code{cvSelect()} with \code{selectTransAndStepAIC()} to automate and cross-validate the whole model-specification process:
 ```{r Auto-transform-and-select}
@@
 ```
-Here, as for `selectTrans()`, the `predictors` and `response` arguments specify candidate variables for transformation,
-and `AIC=FALSE` uses the BIC for model selection.
-The starting model, `m.auto`, is the working model fit to the `Auto` data.
+Here, as for \code{selectTrans()}, the \code{predictors} and \code{response} arguments specify candidate variables for transformation,
+and \code{AIC=FALSE} uses the BIC for model selection.
+The starting model, \code{m.auto}, is the working model fit to the \code{Auto} data.
 The CV criterion isn't bias-adjusted because median absolute error isn't a mean of casewise error components.
@@
 
-* `selectTransStepAIC()` automatically computes CV cost criteria, here the median absolute error, on the untransformed response scale.
-* The estimate of the median absolute error that we obtain by cross-validating the whole model-specification process is a little larger than the median absolute error computed for the model we fit to the `Auto` data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
-* When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of `compareFolds()`), we see that there is little uncertainty in choosing variable transformations (the `lam.*`s for the $x$s and `lambda` for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: `horsepower`, `weight`, and `year` are always included among the selected predictors; `acceleration` and `displacement` are each included respectively in 4 and 3 of 10 selected models; and `cylinders` and `origin` are each included in only 1 of 10 models.
-Recall that when we selected predictors for the full data, we obtained a model with `horsepower`, `weight`, `acceleration`, `year`, and `origin`.
+* \code{selectTransStepAIC()} automatically computes CV cost criteria, here the median absolute error, on the untransformed response scale.
+* The estimate of the median absolute error that we obtain by cross-validating the whole model-specification process is a little larger than the median absolute error computed for the model we fit to the \code{Auto} data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
+* When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of \code{compareFolds()}), we see that there is little uncertainty in choosing variable transformations (the \code{lam.*}s for the $x$s and \code{lambda} for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: \code{horsepower}, \code{weight}, and \code{year} are always included among the selected predictors; \code{acceleration} and \code{displacement} are each included respectively in 4 and 3 of 10 selected models; and \code{cylinders} and \code{origin} are each included in only 1 of 10 models.
+Recall that when we selected predictors for the full data, we obtained a model with \code{horsepower}, \code{weight}, \code{acceleration}, \code{year}, and \code{origin}.
 
@@
 
-The CV functions in the **cv** package are all capable of performing parallel computations by setting the `ncores` argument (specifying the number of computer cores to be used) to a number > `1` (which is the default).
+The CV functions in the \pkg{cv} package are all capable of performing parallel computations by setting the \code{ncores} argument (specifying the number of computer cores to be used) to a number > \code{1} (which is the default).
 Parallel computation can be advantageous for large problems, reducing the execution time of the program.
@@
 
-The **cv** package is designed to be extensible in several directions.
-In this vignette, we discuss three kinds of extensions, ordered by increasing general complexity: (1) adding a cross-validation cost criterion; (2) adding a model class that's not directly accommodated by the `cv()` default method or by another directly inherited method, with separate consideration of mixed-effects models; and (3) adding a new model-selection procedure suitable for use with `selectModel()`.
+The \pkg{cv} package is designed to be extensible in several directions.
+In this vignette, we discuss three kinds of extensions, ordered by increasing general complexity: (1) adding a cross-validation cost criterion; (2) adding a model class that's not directly accommodated by the \code{cv()} default method or by another directly inherited method, with separate consideration of mixed-effects models; and (3) adding a new model-selection procedure suitable for use with \code{selectModel()}.
 
@@
 
-A cost criterion suitable for use with `cv()` or `cvSelect()` should take two arguments, `y` (the observed response vector) and `yhat` (a vector of fitted or predicted response values), and return a numeric index of lack of fit.
-The **cv** package supplies several such criteria: `mse(y, yhat)`, which returns the mean-squared prediction error for a numeric response; `rmse(y, yhat)`, which returns the (square-)root mean-squared error; `medAbsErr(y, yhat)`, which returns the median absolute error; and `BayesRule(y, yhat)` (and its non-error-checking version, `BayesRule2(y, yhat))`, suitable for use with a binary regression model, where `y` is the binary response coded `0` for a "failure" or `1` for a "success"; where `yhat` is the predicted probability of success; and where the proportion of *incorrectly* classified cases is returned.
+A cost criterion suitable for use with \code{cv()} or \code{cvSelect()} should take two arguments, \code{y} (the observed response vector) and \code{yhat} (a vector of fitted or predicted response values), and return a numeric index of lack of fit.
+The \pkg{cv} package supplies several such criteria: \code{mse(y, yhat)}, which returns the mean-squared prediction error for a numeric response; \code{rmse(y, yhat)}, which returns the (square-)root mean-squared error; \code{medAbsErr(y, yhat)}, which returns the median absolute error; and \code{BayesRule(y, yhat)} (and its non-error-checking version, \code{BayesRule2(y, yhat))}, suitable for use with a binary regression model, where \code{y` is the binary response coded \code{0} for a "failure" or \code{1} for a "success"; where \code{yhat} is the predicted probability of success; and where the proportion of *incorrectly* classified cases is returned.
 
@@
 The ROC curve is a graphical representation of the classification power of a binary regression model, and the area under the ROC curve ("AUC"), which varies from 0 to 1, is a common summary measure based on the ROC [see @Wikipedia-ROC:2023].
-The **Metrics** package [@HamnerFrasco:2018] includes a variety of measures useful for model selection, including an `auc()` function.
+The \pkg{Metrics} package [@HamnerFrasco:2018] includes a variety of measures useful for model selection, including an \code{auc()} function.
 We convert the AUC into a cost measure by taking its complement:
@@
 
-We then apply `AUCcomp()` to the the Mroz logistic regression discussed in the main **cv** package vignette, which we reproduce here, using the `Mroz` data frame from the **carData** package [@FoxWeisberg:2019]:
+We then apply \code{AUCcomp()} to the the Mroz logistic regression discussed in the main \pkg{cv} package vignette, which we reproduce here, using the \code{Mroz} data frame from the \pkg{carData} package [@FoxWeisberg:2019]:
 ```{r Mroz-regression}
@@
 
-As we explain in the vignette "Cross-validation of regression models," the `cv()` function differentiates between CV criteria that are averages of casewise components and criteria that are not.
+As we explain in the vignette "Cross-validation of regression models," the \code{cv()} function differentiates between CV criteria that are averages of casewise components and criteria that are not.
 Computation of bias corrections and confidence intervals is limited to the former.
@@
 
-`cv()` looks for a `"casewise loss"` attribute of the value returned by a CV criterion function.
-If this attribute exists, then the criterion is treated as the mean of casewise components, and `cv()` uses the unexported function `getLossFn()` to construct a function that returns the casewise components of the criterion.
-
-We illustrate with the `mse()`:
+\code{cv()} looks for a \code{"casewise loss"} attribute of the value returned by a CV criterion function.
+If this attribute exists, then the criterion is treated as the mean of casewise components, and \code{cv()} uses the unexported function \code{getLossFn()} to construct a function that returns the casewise components of the criterion.
+
+We illustrate with the \code{mse()}:
 ```{r}
@@
 ```
-For this scheme to work, the "casewise loss" attribute must be a character string (or vector of character strings), here `"(y - yhat)^2"`, that evaluates to an expression that is a function of `y` and `yhat`, and that computes the vector of casewise components of the CV criterion.
+For this scheme to work, the "casewise loss" attribute must be a character string (or vector of character strings), here \code{"(y - yhat)\^2"}, that evaluates to an expression that is a function of \code{y} and \code{yhat}, and that computes the vector of casewise components of the CV criterion.
 
@@
 
-Suppose that we want to cross-validate a multinomial logistic regression model fit by the `multinom()` function in the **nnet** package [@VenablesRipley:2002].
+Suppose that we want to cross-validate a multinomial logistic regression model fit by the \code{multinom()} function in the \pkg{nnet} package [@VenablesRipley:2002].
 We borrow an example from @Fox:2016 [Sec.
 14.2.1], with data from the British Election Panel Study on vote choice in the 2001 British election.
-Data for the example are in the `BEPS` data frame in the **carData** package:
+Data for the example are in the \code{BEPS} data frame in the \pkg{carData} package:
 ```{r BEPS-data}
@@
 ```
-The polytomous (multi-category) response variable is `vote`, a factor with levels `"Conservative"`, `"Labour"`, and `"Liberal Democrat"`. The predictors of `vote` are:
-
-* `age`, in years;
-* `econ.cond.national` and `econ.cond.household`, the respondent's ratings of the state of the economy, on 1 to 5 scales.
-* `Blair`, `Hague`, and `Kennedy`, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
-* `Europe`, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism."
-* `political.knowledge`, knowledge of the parties' positions on European integration, with scores from 0 to 3.
-* `gender`, `"female"` or `"male"`.
-
-The model fit to the data includes an interaction between `Europe` and `political.knowledge`; the other predictors enter the model additively:
+The polytomous (multi-category) response variable is \code{vote}, a factor with levels \code{"Conservative"}, \code{"Labour"}, and \code{"Liberal Democrat"}. The predictors of \code{vote} are:
+
+* \code{age}, in years;
+* \code{econ.cond.national} and \code{econ.cond.household}, the respondent's ratings of the state of the economy, on 1 to 5 scales.
+* \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
+* \code{Europe}, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism."
+* \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3.
+* \code{gender}, \code{"female"} or \code{"male"}.
+
+The model fit to the data includes an interaction between \code{Europe} and \code{political.knowledge}; the other predictors enter the model additively:
 ```{r BEPS-model}
@@
 ```
-Most of the predictors, including the `Europe` $\times$ `political.knowledge` interaction, are associated with very small $p$-values; the `Anova()` function is from the **car** package [@FoxWeisberg:2019].
-
-Here's an "effect plot", using the the **effects** package [@FoxWeisberg:2019] to visualize the `Europe` $\times$ `political.knowledge` interaction in a "stacked-area" graph:
+Most of the predictors, including the \code{Europe} $\times$ \code{political.knowledge} interaction, are associated with very small $p$-values; the \code{Anova()} function is from the \pkg{car} package [@FoxWeisberg:2019].
+
+Here's an "effect plot", using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the \code{Europe} $\times$ \code{political.knowledge} interaction in a "stacked-area" graph:
 ```{r BEPS-plot, fig.width=9, fig.height=5}
@@
 To cross-validate this multinomial-logit model we need an appropriate cost criterion.
-None of the criteria supplied by the **cv** package---for example, neither `mse()`, which is appropriate for a numeric response, nor `BayesRule()`, which is appropriate for a binary response---will do.
+None of the criteria supplied by the \pkg{cv} package---for example, neither \code{mse()}, which is appropriate for a numeric response, nor \code{BayesRule()}, which is appropriate for a binary response---will do.
 One possibility is to adapt Bayes rule to a polytomous response:
@@
 ```
-The `predict()` method for `"multinom"` models called with argument `type="class"` reports the Bayes-rule prediction for each case---that is, the response category with the highest predicted probability.
-Our `BayesRuleMulti()` function calculates the proportion of misclassified cases.
-Because this value is the mean of casewise components, we attach a `"casewise loss"` attribute to the result (as explained in the preceding section).
+The \code{predict()} method for \code{"multinom"} models called with argument \code{type="class"} reports the Bayes-rule prediction for each case---that is, the response category with the highest predicted probability.
+Our \code{BayesRuleMulti()} function calculates the proportion of misclassified cases.
+Because this value is the mean of casewise components, we attach a \code{"casewise loss"} attribute to the result (as explained in the preceding section).
 
@@
 
-We check first whether the default `cv()` method works "out-of-the-box" for the `"multinom"` model:
+We check first whether the default \code{cv()} method works "out-of-the-box" for the \code{"multinom"} model:
 ```{r BEPS-test-default, error=TRUE}
@@
 ```
-The default method of `GetResponse()` (a function supplied by the **cv** package---see `?GetResponse`) fails for a `"multinom"` object.
-A straightforward solution is to supply a `GetResponse.multinom()` method that returns the factor response [using the `get_response()` function from the **insight** package, @LudeckeWaggonerMakowski:2019],
+The default method of \code{GetResponse()} (a function supplied by the \pkg{cv} package---see \code{?GetResponse}) fails for a \code{"multinom"} object.
+A straightforward solution is to supply a \code{GetResponse.multinom()} method that returns the factor response [using the `get_response()` function from the \pkg{insight} package, @LudeckeWaggonerMakowski:2019],
 ```{r GetResponse.multinom}
@@
 ```
-A `traceback()` (not shown) reveals that the problem is that the default method of `cv()` calls the `"multinom"` method for `predict()` with the argument `type="response"`, when the correct argument should be `type="class"`.
-We therefore must write a "`multinom`" method for `cv()`, but that proves to be very simple:
+A \code{traceback()} (not shown) reveals that the problem is that the default method of \code{cv()} calls the \code{"multinom"} method for \code{predict()} with the argument \code{type="response"}, when the correct argument should be \code{type="class"}.
+We therefore must write a \code{"multinom"} method for \code{cv()}, but that proves to be very simple:
 ```{r cv.nultinom}
@@
 ```
-That is, we simply call the default `cv()` method with the `type` argument properly set.
-In addition to supplying the correct `type` argument, our method sets the default `criterion` for the `cv.multinom()` method to `BayesRuleMulti`.
+That is, we simply call the default \code{cv()} method with the \code{type} argument properly set.
+In addition to supplying the correct \code{type} argument, our method sets the default \code{criterion} for the \code{cv.multinom()} method to \code{BayesRuleMulti}.
 
@@
 ```
-Prior to invoking `cv()`, we called `update()` with `trace=FALSE` to suppress the iteration history reported by default by `multinom()`---it would be tedious to see the iteration history for each fold.
+Prior to invoking \code{cv()}, we called \code{update()} with \code{trace=FALSE} to suppress the iteration history reported by default by \code{multinom()}---it would be tedious to see the iteration history for each fold.
 The cross-validated polytomous Bayes-rule criterion confirms that the fitted model does substantially better than the marginal Bayes-rule prediction that everyone votes for Labour.
@@
 
-Adding a `cv()` method for a mixed-model class is somewhat more complicated.
-We provide the `cvMixed()` function to facilitate this process, and to see how that works, consider the `"lme"` method from the **cv** package:
+Adding a \code{cv()} method for a mixed-model class is somewhat more complicated.
+We provide the \code{cvMixed()} function to facilitate this process, and to see how that works, consider the \code{"lme"} method from the \pkg{cv} package:
 ```{r cv.lme}
@@
 ```
-Notice that `cv.lme()` sets up a call to `cvMixed()`, which does the computational work.
-
-Most of the arguments of `cvMixed()` are familiar:
-
-* `model` is the mixed-model object, here of class `"lme"`.
-
-* `package` is the name of the package in which the mixed-modeling function used to fit the model, here  `lme()`, resides---i.e., `"nlme"`; `cvMixed()` uses this argument to retrieve the package namespace.
-
-* `data` is the data set to which the model is fit, by default extracted by the `get_data()` function in the **insight** package.
-
-* `criterion` is the CV criterion, defaulting to the `mse()` function.
-
-* `k` is the number of CV folds, defaulting to `"loo"` for CV by clusters and `10` for CV by cases.
-
-* `reps` is the number of times the CV process is repeated, defaulting to `1`.
-
-* `seed` is the seed for R's random-number generator, defaulting to a randomly selected (and saved) value.
-
-* `ncores` is the number of cores to use for parallel computation; if `1`, the default, then the computation isn't parallelized.
-
-* `clusterVariables` is a character vector of the names of variables defining clusters; if missing, then CV is based on cases rather than clusters.
+Notice that \code{cv.lme()} sets up a call to \code{cvMixed()}, which does the computational work.
+
+Most of the arguments of \code{cvMixed()} are familiar:
+
+* \code{model} is the mixed-model object, here of class \code{"lme"}.
+
+* \code{package} is the name of the package in which the mixed-modeling function used to fit the model, here  \code{lme()}, resides---i.e., \code{"nlme"}; \code{cvMixed()} uses this argument to retrieve the package namespace.
+
+* \code{data} is the data set to which the model is fit, by default extracted by the \code{get_data()} function in the \pkg{insight} package.
+
+* \code{criterion} is the CV criterion, defaulting to the \code{mse()} function.
+
+* \code{k} is the number of CV folds, defaulting to \code{"loo"} for CV by clusters and \code{10} for CV by cases.
+
+* \code{reps} is the number of times the CV process is repeated, defaulting to \code{1}.
+
+* \code{seed} is the seed for \proglang{R}'s random-number generator, defaulting to a randomly selected (and saved) value.
+
+* \code{ncores} is the number of cores to use for parallel computation; if \code{1}, the default, then the computation isn't parallelized.
+
+* \code{clusterVariables} is a character vector of the names of variables defining clusters; if missing, then CV is based on cases rather than clusters.
 
@@
 
-* `predict.clusters.args` is a named list of arguments to be passed to the `predict()` function to obtain predictions for the full data set from a model fit to a subset of the data for cluster-based CV.
-The first two arguments should be `object` and `newdata`.
-It is typically necessary to tell `cvMixed()` how to base predictions only on fixed effects; in the case of `"lme"` models, this is done by setting `level = 0`.
-
-* Similarly, `predict.cases.args` is a named list of arguments to be passed to `predict()` for case-based CV. Setting `level = 1` includes random effects in the predictions.
-
-Finally, any additional arguments, absorbed by `...`, are passed to `update()` when the model is refit with each fold omitted. `cvMixed()` returns an object of class `"cv"`.
+* \code{predict.clusters.args} is a named list of arguments to be passed to the \code{predict()} function to obtain predictions for the full data set from a model fit to a subset of the data for cluster-based CV.
+The first two arguments should be \code{object} and \code{newdata}.
+It is typically necessary to tell \code{cvMixed()} how to base predictions only on fixed effects; in the case of \code{"lme"} models, this is done by setting \code{level = 0}.
+
+* Similarly, \code{predict.cases.args} is a named list of arguments to be passed to \code{predict()} for case-based CV. Setting \code{level = 1} includes random effects in the predictions.
+
+Finally, any additional arguments, absorbed by \code{...}, are passed to \code{update()} when the model is refit with each fold omitted. \code{cvMixed()} returns an object of class \code{"cv"}.
 
 Now imagine that we want to support a new class of mixed-effects models.
-To be concrete, we illustrate with the `glmmPQL()` function in the **MASS** package [@VenablesRipley:2002], which fits generalized-linear mixed-effects models by penalized quasi-likelihood.[^glmmPQL] Not coincidentally, the arguments of `glmmPQL()` are similar to those of `lme()` (with an additional `family` argument), because the former iteratively invokes the latter; so `cv.glmmPQL()` should resemble `cv.lme()`.
-
-[^glmmPQL]: This example is somewhat artificial in that `glmmPQL()` has largely been superseded by computationally superior functions, such the `glmer()` function in the **lme4** package.
-There is, however, one situation in which `glmmPQL()` might prove useful: to specify serial dependency in case-level errors within clusters for longitudinal data, which is not currently supported by `glmer()`.
-
-As it turns out, neither the default method for `GetResponse()` nor `insight::get_data()` work for `"glmmPQL"` objects.
-These objects include a `"data"` element, however, and so we can simply extract this element as the default for the `data` argument of our `cv.glmmPQL()` method.
+To be concrete, we illustrate with the \code{glmmPQL()} function in the \pkg{MASS} package [@VenablesRipley:2002], which fits generalized-linear mixed-effects models by penalized quasi-likelihood.[^glmmPQL] Not coincidentally, the arguments of \code{glmmPQL()} are similar to those of \code{lme()} (with an additional \code{family} argument), because the former iteratively invokes the latter; so \code{cv.glmmPQL()} should resemble \code{cv.lme()}.
+
+[^glmmPQL]: This example is somewhat artificial in that \code{glmmPQL()} has largely been superseded by computationally superior functions, such the \code{glmer()} function in the \pkg{lme4} package.
+There is, however, one situation in which \code{glmmPQL()} might prove useful: to specify serial dependency in case-level errors within clusters for longitudinal data, which is not currently supported by \code{glmer()}.
+
+As it turns out, neither the default method for \code{GetResponse()} nor \code{insight::get_data()} work for \code{"glmmPQL"} objects.
+These objects include a \code{"data"} element, however, and so we can simply extract this element as the default for the \code{data} argument of our \code{cv.glmmPQL()} method.
 
@@
 
-Writing the `cv()` method is then straightforward:
+Writing the \code{cv()} method is then straightforward:
 ```{r cv.glmmPQL}
@@
 ```
-We set the argument `verbose=FALSE` to suppress `glmmPQL()`'s iteration counter when `cvMixed()` calls `update()`.
-
-Let's apply our newly minted method to a logistic regression with a random intercept in an example that appears in `?glmmPQL`:
+We set the argument \code{verbose=FALSE} to suppress \code{glmmPQL()}'s iteration counter when \code{cvMixed()} calls \code{update()}.
+
+Let's apply our newly minted method to a logistic regression with a random intercept in an example that appears in \code{?glmmPQL}:
 ```{r glmmPQL-example}
@@
 ```
-We compare this result to that obtained from `glmer()` in the **lme4** package:
+We compare this result to that obtained from \code{glmer()} in the \pkg{lme4} package:
 ```{r compare-to-lme4}
@@
 
-Finally, we try out our `cv.glmmPQL()` method, cross-validating both by clusters and by cases,
+Finally, we try out our \code{cv.glmmPQL()} method, cross-validating both by clusters and by cases,
 ```{r cv-example-glmmPQL, cache=TRUE}
@@
 ```
-and again compare to `glmer()`:
+and again compare to \code{glmer()}:
 ```{r cv-example-glmer, cache=TRUE}
@@
 
-The `selectStepAIC()` function supplied by the **cv** package, which is based on the `stepAIC()` function from the **nnet** package [@VenablesRipley:2002] for stepwise model selection, is suitable for the `procedure` argument of `cvSelect()`.
-The use of `selectStepAIC()` is illustrated in the principal vignette for the package.
-
-We'll employ `selectStepAIC()` as a "template" for writing a CV model-selection procedure.
-To see the code for this function, type `cv::selectStepAIC` at the R command prompt, or examine the sources for the **cv** package at https://github.com/gmonette/cv (the code for `selectStepAIC()` is in https://github.com/gmonette/cv/blob/main/R/cvSelect.R).
+The \code{selectStepAIC()} function supplied by the \pkg{cv} package, which is based on the \code{stepAIC()} function from the \pkg{nnet} package [@VenablesRipley:2002] for stepwise model selection, is suitable for the \code{procedure} argument of \code{cvSelect()}.
+The use of \code{selectStepAIC()} is illustrated in the principal vignette for the package.
+
+We'll employ \code{selectStepAIC()} as a "template" for writing a CV model-selection procedure.
+To see the code for this function, type \code{cv::selectStepAIC} at the \proglang{R} command prompt, or examine the sources for the \pkg{cv} package at https://github.com/gmonette/cv (the code for \code{selectStepAIC()} is in https://github.com/gmonette/cv/blob/main/R/cvSelect.R).
 
 Another approach to model selection is all-subsets regression.
-The `regsubsets()` function in the **leaps** package [@LumleyMiller:2020] implements an efficient algorithm for selecting the best-fitting linear least-squares regressions for subsets of predictors of all sizes, from 1 through the maximum number of candidate predictors.[^1] To illustrate the use of `regsubsets()`, we employ the `swiss` data frame supplied by the **leaps** package:
-
-[^1]: The `regsubsets()` function computes several measures of model predictive performance, including the $R^2$ and $R^2$ adjusted for degrees of freedom, the residual sums of squares, Mallows's $C_p$, and the BIC.
+The \code{regsubsets()} function in the \pkg{leaps} package [@LumleyMiller:2020] implements an efficient algorithm for selecting the best-fitting linear least-squares regressions for subsets of predictors of all sizes, from 1 through the maximum number of candidate predictors.[^1] To illustrate the use of \code{regsubsets()}, we employ the \code{swiss} data frame supplied by the \pkg{leaps} package:
+
+[^1]: The \code{regsubsets()} function computes several measures of model predictive performance, including the $R^2$ and $R^2$ adjusted for degrees of freedom, the residual sums of squares, Mallows's $C_p$, and the BIC.
 Several of these are suitable for comparing models with differing numbers of coefficients---we use the BIC below---but all necessarily agree when comparing models with the *same* number of coefficients.
@@
 
-* `Fertility`: A standardized fertility measure.
-* `Agriculture`: The percentage of the male population engaged in agriculture.
-* `Examination`: The percentage of draftees into the Swiss army receiving the highest grade on an examination.
-* `Education`: The percentage of draftees with more than a primary-school education.
-* `Catholic`: The percentage of the population who were Catholic.
-* `Infant.Mortality`: The infant-mortality rate, expressed as the percentage of live births surviving less than a year.
-
-Following @LumleyMiller:2020, we treat `Fertility` as the response and the other variables as predictors in a linear least-squares regression:
+* \code{Fertility}: A standardized fertility measure.
+* \code{Agriculture}: The percentage of the male population engaged in agriculture.
+* \code{Examination}: The percentage of draftees into the Swiss army receiving the highest grade on an examination.
+* \code{Education}: The percentage of draftees with more than a primary-school education.
+* \code{Catholic}: The percentage of the population who were Catholic.
+* \code{Infant.Mortality}: The infant-mortality rate, expressed as the percentage of live births surviving less than a year.
+
+Following @LumleyMiller:2020, we treat \code{Fertility} as the response and the other variables as predictors in a linear least-squares regression:
 ```{r swiss-lm}
@@
 ```
-The graph, produced by the `subsets()` function in the **car** package, shows that the model with the smallest BIC is the best model with 4 predictors, including `Agriculture`, `Education`, `Catholic`, and `Infant.Mortality`, but not `Examination`:
+The graph, produced by the \code{subsets()} function in the \pkg{car} package, shows that the model with the smallest BIC is the best model with 4 predictors, including \code{Agriculture}, \code{Education}, \code{Catholic}, and \code{Infant.Mortality}, but not \code{Examination}:
 ```{r best-model}
@@
 
-Here's a function named `selectSubsets()`, meant to be used with `cvSelect()`, suitable for cross-validating the model-selection process:
+Here's a function named \code{selectSubsets()}, meant to be used with \code{cvSelect()}, suitable for cross-validating the model-selection process:
 ```{r selectSubsets}
@@
 ```
-A slightly tricky point is that because of scoping issues, `predict()` doesn't work with the model fit omitting the $i$th fold, and so the fitted values for all cases are computed directly as $\widehat{\mathbf{y}}_{-i} = \mathbf{X} \mathbf{b}_{-i}$, where $\mathbf{X}$ is the model-matrix for all of the cases, and $\mathbf{b}_{-i}$ is the vector of least-squares coefficients for the selected model with the $i$th fold omitted.
-
-Additionally, the command `lm(y[-indices] ~ X[-indices, x.names.i] - 1)`, which is the selected model with the $i$th fold deleted, produces awkward coefficient names like `"X[-indices, x.names.i]Infant.Mortality"`.
-Purely for aesthetic reasons, the command `sub("X\\[-indices, x.names.i\\]", "", names(coefs))` fixes these awkward names, removing the extraneous text, `"X[-indices, x.names.i]"`.
-
-Applying `selectSubsets()` to the full data produces the full-data cross-validated RMSE (which we obtained previously):
+A slightly tricky point is that because of scoping issues, \code{predict()} doesn't work with the model fit omitting the $i$th fold, and so the fitted values for all cases are computed directly as $\widehat{\mathbf{y}}_{-i} = \mathbf{X} \mathbf{b}_{-i}$, where $\mathbf{X}$ is the model-matrix for all of the cases, and $\mathbf{b}_{-i}$ is the vector of least-squares coefficients for the selected model with the $i$th fold omitted.
+
+Additionally, the command \code{lm(y[-indices] ~ X[-indices, x.names.i] - 1)}, which is the selected model with the $i$th fold deleted, produces awkward coefficient names like \code{"X[-indices, x.names.i]Infant.Mortality"}.
+Purely for aesthetic reasons, the command `sub("X\\[-indices, x.names.i\\]", "", names(coefs))` fixes these awkward names, removing the extraneous text, \code{"X[-indices, x.names.i]"}.
+
+Applying \code{selectSubsets()} to the full data produces the full-data cross-validated RMSE (which we obtained previously):
 %%```{r test-selectSubsets}
@@
 
-Then, using `selectSubsets()` in cross-validation, we get:
+Then, using \code{selectSubsets()} in cross-validation, we get:
 ```{r cvSelect-selectSubsets}
@@
 Cross-validation shows that model selection exacts a penalty in RMSE.
-Examining the models selected for the 10 folds reveals that there is some uncertainty in identifying the predictors in the "best" model, with `Agriculture` sometimes appearing and sometimes not:
+Examining the models selected for the 10 folds reveals that there is some uncertainty in identifying the predictors in the "best" model, with \code{Agriculture} sometimes appearing and sometimes not:
 ```{r best-models-by-folds}
@@
 
-The most straightforward way to implement cross-validation in R for statistical modeling functions that are written in the canonical manner is to use `update()` to refit the model with each fold removed.
-This is the approach taken in the default method for `cv()`, and it is appropriate if the cases are independently sampled.
+The most straightforward way to implement cross-validation in \proglang{R} for statistical modeling functions that are written in the canonical manner is to use \code{update()} to refit the model with each fold removed.
+This is the approach taken in the default method for \code{cv()}, and it is appropriate if the cases are independently sampled.
 Refitting the model in this manner for each fold is generally feasible when the number of folds in modest, but can be prohibitively costly for leave-one-out cross-validation when the number of cases is large.
 
-The `"lm"` and `"glm"` methods for `cv()` take advantage of computational efficiencies by avoiding refitting the model with each fold removed.
+The \code{"lm"} and \code{"glm"} methods for \code{cv()} take advantage of computational efficiencies by avoiding refitting the model with each fold removed.
 Consider, in particular, the weighted linear model $\mathbf{y}_{n \times 1} = \mathbf{X}_{n \times p}\boldsymbol{\beta}_{p \times 1} + \boldsymbol{\varepsilon}_{n \times 1}$, where $\boldsymbol{\varepsilon} \sim \mathbf{N}_n \left(\mathbf{0}, \sigma^2 \mathbf{W}^{-1}_{n \times n}\right)$.
@@
 [^WLS]: This is a definitional formula, which assumes that the model matrix $\mathbf{X}$ is of full column rank, and which can be subject to numerical instability when $\mathbf{X}$ is ill-conditioned.
-`lm()` uses the singular-value decomposition of the model matrix to obtain computationally more stable results.
+\code{lm()} uses the singular-value decomposition of the model matrix to obtain computationally more stable results.
 
@@
 A disadvantage of the Woodbury identity, however, is that it entails explicit matrix inversion and thus may be numerically unstable.
-The inverse of $\mathbf{A} = \mathbf{X}^T \mathbf{W} \mathbf{X}$ is available directly in the `"lm"` object, but the second term on the right-hand side of the Woodbury identity requires a matrix inversion with each fold deleted.
+The inverse of $\mathbf{A} = \mathbf{X}^T \mathbf{W} \mathbf{X}$ is available directly in the \code{"lm"} object, but the second term on the right-hand side of the Woodbury identity requires a matrix inversion with each fold deleted.
 (In contrast, the inverse of each $\mathbf{C} = \mathbf{W}_\mathbf{j}$ is straightforward because $\mathbf{W}$ is diagonal.)
@@
 
-`glm()` computes the maximum-likelihood estimates for a generalized linear model by iterated weighted least squares [see, e.g., @FoxWeisberg:2019, Sec.
+\code{glm()} computes the maximum-likelihood estimates for a generalized linear model by iterated weighted least squares [see, e.g., @FoxWeisberg:2019, Sec.
 6.12].
-The last iteration is therefore just a WLS fit of the "working response" on the model matrix using "working weights." Both the working weights and the working response at convergence are available from the information in the object returned by `glm()`.
+The last iteration is therefore just a WLS fit of the "working response" on the model matrix using "working weights." Both the working weights and the working response at convergence are available from the information in the object returned by \code{glm()}.
 
@@
 The resulting fitted values for the deleted fold aren't exact---that is, except for the Gaussian family, the result isn't identical to what we would obtain by literally refitting the model---but in our (limited) experience, the approximation is very good, especially for LOO CV, which is when we would be most tempted to use it.
-Nevertheless, because these results are approximate, the default for the `"glm"` `cv()` method is to perform the exact computation, which entails refitting the model with each fold omitted.
+Nevertheless, because these results are approximate, the default for the \code{"glm"} \code{cv()} method is to perform the exact computation, which entails refitting the model with each fold omitted.
 
@@
 @BatesHastieTibshirani:2023 show that the coverage of this confidence interval is poor for small samples, and they suggest a much more computationally intensive procedure, called *nested cross-validation*, to compute better estimates of error and confidence intervals with better coverage for small samples.
-We may implement Bates et al.'s approach in a later release of the **cv** package.
+We may implement Bates et al.'s approach in a later release of the \pkg{cv} package.
 At present we use the confidence interval above for sufficiently large $n$, which, based on Bates et al.'s results, we take by default to be $n \ge 400$.
@@
 
-The following R function computes AUC from $\mathbf{\widehat{y}}_v$ and $\mathbf{y}_v$, accommodating the cases where $\mathbf{y}_v$ is all 0s or all 1s:
+The following \proglang{R} function computes AUC from $\mathbf{\widehat{y}}_v$ and $\mathbf{y}_v$, accommodating the cases where $\mathbf{y}_v$ is all 0s or all 1s:
 

### 2024-01-02 scratch/notes_john/JSS-article/JSS-article.tex
use JSS markup \pkg() etc.

@@
 }
-\title{\pkg{cv}: An R Package for Cross-Validation of Regression Models}
+\title{\pkg{cv}: An \proglang{R} Package for Cross-Validation of
+Regression Models}
 
@@
 
-The data for this example are drawn from the \textbf{ISLR2} package for
-R, associated with \citet{JamesEtAl:2021}. The presentation here is
-close (though not identical) to that in the original source \citep[
-Secs. 5.1, 5.3]{JamesEtAl:2021}, and it demonstrates the use of the
-\texttt{cv()} function in the \textbf{cv} package.\footnote{\citet{JamesEtAl:2021}
-  use the \texttt{cv.glm()} function in the \textbf{boot} package
+The data for this example are drawn from the \pkg{ISLR2} package for
+\proglang{R}, associated with \citet{JamesEtAl:2021}. The presentation
+here is close (though not identical) to that in the original source
+\citep[ Secs. 5.1, 5.3]{JamesEtAl:2021}, and it demonstrates the use of
+the \texttt{cv()} function in the \pkg{cv} package.\footnote{\citet{JamesEtAl:2021}
+  use the \texttt{cv.glm()} function in the \pkg{boot} package
   \citep{CantyRipley2022, DavisonHinkley:1997}. Despite its name,
@@
 The code for this graph uses the \texttt{mse()} function from the
-\textbf{cv} package to compute the MSE for each fit.
+\pkg{cv} package to compute the MSE for each fit.
 
@@
 \begin{CodeOutput}
-R RNG seed set to 534539
+R RNG seed set to 44768
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.22944
-bias-adjusted cross-validation criterion = 19.21651
+cross-validation criterion = 19.52622
+bias-adjusted cross-validation criterion = 19.49747
 full-sample criterion = 18.98477 
@@
 
-The \texttt{"lm"} method by default uses \texttt{mse()} as the CV
-criterion and the Woodbury matrix identity to update the regression with
-each fold deleted without having literally to refit the model.
-Computational details are discussed in the final section of this
-vignette. The function reports the CV estimate of MSE, a biased-adjusted
-estimate of the MSE (the bias adjustment is explained in the final
-section), and the MSE is also computed for the original, full-sample
-regression. Because the division of the data into 10 folds is random,
-\texttt{cv()} explicitly (randomly) generates and saves a seed for R's
-pseudo-random number generator, to make the results replicable. The user
-can also specify the seed directly via the \texttt{seed} argument to
-\texttt{cv()}.
-
-To perform LOO CV, we can set the \texttt{k} argument to \texttt{cv()}
-to the number of cases in the data, here \texttt{k=392}, or, more
-conveniently, to \texttt{k="loo"} or \texttt{k="n"}:
+The \code{"lm"} method by default uses \code{mse()} as the CV criterion
+and the Woodbury matrix identity to update the regression with each fold
+deleted without having literally to refit the model. Computational
+details are discussed in the final section of this vignette. The
+function reports the CV estimate of MSE, a biased-adjusted estimate of
+the MSE (the bias adjustment is explained in the final section), and the
+MSE is also computed for the original, full-sample regression. Because
+the division of the data into 10 folds is random, \code{cv()} explicitly
+(randomly) generates and saves a seed for \proglang{R}'s pseudo-random
+number generator, to make the results replicable. The user can also
+specify the seed directly via the \code{seed} argument to \code{cv()}.
+
+To perform LOO CV, we can set the \code{k} argument to \code{cv()} to
+the number of cases in the data, here \code{k=392}, or, more
+conveniently, to \code{k="loo"} or \code{k="n"}:
 
@@
 
-For LOO CV of a linear model, \texttt{cv()} by default uses the
-hatvalues from the model fit to the full data for the LOO updates, and
-reports only the CV estimate of MSE. Alternative methods are to use the
-Woodbury matrix identity or the ``naive'' approach of literally
-refitting the model with each case omitted. All three methods produce
-exact results for a linear model (within the precision of floating-point
+For LOO CV of a linear model, \code{cv()} by default uses the hatvalues
+from the model fit to the full data for the LOO updates, and reports
+only the CV estimate of MSE. Alternative methods are to use the Woodbury
+matrix identity or the ``naive'' approach of literally refitting the
+model with each case omitted. All three methods produce exact results
+for a linear model (within the precision of floating-point
 computations):
@@
 
-The \texttt{"naive"} and \texttt{"Woodbury"} methods also return the
+The \code{"naive"} and \code{"Woodbury"} methods also return the
 bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an
@@
 investigate their relative speed. In this comparison, we include the
-\texttt{cv.glm()} function from the \textbf{boot} package, which takes
-the naive approach, and for which we have to fit the linear model as an
-equivalent Gaussian GLM. We use the \texttt{microbenchmark()} function
+\code{cv.glm()} function from the \pkg{boot} package, which takes the
+naive approach, and for which we have to fit the linear model as an
+equivalent Gaussian GLM. We use the \code{microbenchmark()} function
 from the package of the same name for the timings \citep{Mersmann:2023}:
@@
 of magnitude faster than refitting the model.\footnote{Out of
-  impatience, we asked \texttt{microbenchmark()} to execute each command
+  impatience, we asked \code{microbenchmark()} to execute each command
   only 10 times rather than the default 100. With the exception of the
@@
   distinguishable. Because of the small number of repetitions (i.e.,
-  10), the \texttt{"hatvalues"} and \texttt{"Woodbury"} methods aren't
+  10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't
   distinguishable, but the difference between these methods persists
   when we perform more repetitions---we invite the reader to redo this
-  computation with the default \texttt{times=100} repetitions.}
+  computation with the default \code{times=100} repetitions.}
 
@@
 
-The \texttt{cv()} function also has a method that can be applied to a
-list of regression models for the same data, composed using the
-\texttt{models()} function. For \(k\)-fold CV, the same folds are used
-for the competing models, which reduces random error in their
-comparison. This result can also be obtained by specifying a common seed
-for R's random-number generator while applying \texttt{cv()} separately
-to each model, but employing a list of models is more convenient for
-both \(k\)-fold and LOO CV (where there is no random component to the
-composition of the \(n\) folds).
+The \code{cv()} function also has a method that can be applied to a list
+of regression models for the same data, composed using the
+\code{models()} function. For \(k\)-fold CV, the same folds are used for
+the competing models, which reduces random error in their comparison.
+This result can also be obtained by specifying a common seed for
+\proglang{R}'s random-number generator while applying \code{cv()}
+separately to each model, but employing a list of models is more
+convenient for both \(k\)-fold and LOO CV (where there is no random
+component to the composition of the \(n\) folds).
 
 We illustrate with the polynomial regression models of varying degree
-for the \texttt{Auto} data (discussed previously), beginning by fitting
+for the \code{Auto} data (discussed previously), beginning by fitting
 and saving the 10 models:
@@
 
-We then apply \texttt{cv()} to the list of 10 models (the \texttt{data}
+We then apply \code{cv()} to the list of 10 models (the \code{data}
 argument is required):
@@
 Because we didn't supply names for the models in the calls to the
-\texttt{models()} function, the names \texttt{model.1},
-\texttt{model.2}, etc., are generated by the function.
+\code{models()} function, the names \code{model.1}, \code{model.2},
+etc., are generated by the function.
 
@@
 
-Alternatively, we can use the \texttt{plot()} method for
-\texttt{"cvModList"} objects to compare the models, though with separate
+Alternatively, we can use the \code{plot()} method for
+\code{"cvModList"} objects to compare the models, though with separate
 graphs for 10-fold and LOO CV:
@@
 
-The \texttt{Mroz} data set from the \textbf{carData} package
+The \code{Mroz} data set from the \pkg{carData} package
 \citep[associated with][]{FoxWeisberg:2019} has been used by several
@@
 
-The response variable in the logistic regression is \texttt{lfp},
-labor-force participation, a factor coded \texttt{"yes"} or
-\texttt{"no"}. The remaining variables are predictors:
+The response variable in the logistic regression is \code{lfp},
+labor-force participation, a factor coded \code{"yes"} or \code{"no"}.
+The remaining variables are predictors:
 
@@
 \item
-  \texttt{k5}, number of children 5 years old of younger in the woman's
+  \code{k5}, number of children 5 years old of younger in the woman's
   household;
 \item
-  \texttt{k618}, number of children between 6 and 18 years old;
+  \code{k618}, number of children between 6 and 18 years old;
 \item
-  \texttt{age}, in years;
+  \code{age}, in years;
 \item
-  \texttt{wc}, wife's college attendance, \texttt{"yes"} or
-  \texttt{"no"};
+  \code{wc}, wife's college attendance, \code{"yes"} or \code{"no"};
 \item
-  \texttt{hc}, husband's college attendance;
+  \code{hc}, husband's college attendance;
 \item
-  \texttt{lwg}, the woman's log wage rate if she is employed, or her
+  \code{lwg}, the woman's log wage rate if she is employed, or her
   \emph{imputed} wage rate, if she is not \citep[a variable that][ show
@@
 \item
-  \texttt{inc}, family income, in \$1000s, exclusive of wife's income.
+  \code{inc}, family income, in \$1000s, exclusive of wife's income.
 \end{itemize}
 
-We use the \texttt{glm()} function to fit a binary logistic regression
-to the \texttt{Mroz} data:
+We use the \code{glm()} function to fit a binary logistic regression to
+the \code{Mroz} data:
 
@@
 In addition to the usually summary output for a GLM, we show the result
-of applying the \texttt{BayesRule()} function from the \textbf{cv}
-package to predictions derived from the fitted model. Bayes rule, which
-predicts a ``success'' in a binary regression model when the fitted
-probability of success {[}i.e., \(\phi = \Pr(y = 1)\){]} is
-\(\widehat{\phi} \ge .5\) and a ``failure'' if
-\(\widehat{\phi} < .5\).\footnote{\texttt{BayesRule()} does some error
-  checking; \texttt{BayesRule2()} is similar, but omits the error
-  checking, and so can be faster for large problems.} The first argument
-to \texttt{BayesRule()} is the binary \{0, 1\} response, and the second
-argument is the predicted probability of success. \texttt{BayesRule()}
-returns the proportion of predictions that are \emph{in error}, as
-appropriate for a ``cost'' function.
-
-The value returned by \texttt{BayesRule()} is associated with an
-``attribute'' named \texttt{"casewise\ loss"} and set to
-\texttt{"y\ !=\ round(yhat)"}, signifying that the Bayes rule CV
-criterion is computed as the mean of casewise values, here 0 if the
-prediction for a case matches the observed value and 1 if it does not
-(signifying a prediction error). The \texttt{mse()} function for numeric
-responses is also calculated as a casewise average. Some other criteria,
-such as the median absolute error, computed by the \texttt{medAbsErr()}
-function in the \textbf{cv} package, aren't averages of casewise
-components. The distinction is important because, to our knowledge, the
-statistical theory of cross-validation, for example, in
-\citet{DavisonHinkley:1997}, \citet{BatesHastieTibshirani:2023}, and
-\citet{ArlotCelisse:2010}, is developed for CV criteria like MSE that
-are means of casewise components. As a consequence, we limit computation
-of bias adjustment and confidence intervals (see below) to criteria that
-are casewise averages.
+of applying the \code{BayesRule()} function from the \pkg{cv} package to
+predictions derived from the fitted model. Bayes rule, which predicts a
+``success'' in a binary regression model when the fitted probability of
+success {[}i.e., \(\phi = \Pr(y = 1)\){]} is \(\widehat{\phi} \ge .5\)
+and a ``failure'' if \(\widehat{\phi} < .5\).\footnote{\code{BayesRule()}
+  does some error checking; \code{BayesRule2()} is similar, but omits
+  the error checking, and so can be faster for large problems.} The
+first argument to \code{BayesRule()} is the binary \{0, 1\} response,
+and the second argument is the predicted probability of success.
+\code{BayesRule()} returns the proportion of predictions that are
+\emph{in error}, as appropriate for a ``cost'' function.
+
+The value returned by \code{BayesRule()} is associated with an
+``attribute'' named \code{"casewise loss"} and set to
+\code{"y != round(yhat)"}, signifying that the Bayes rule CV criterion
+is computed as the mean of casewise values, here 0 if the prediction for
+a case matches the observed value and 1 if it does not (signifying a
+prediction error). The \code{mse()} function for numeric responses is
+also calculated as a casewise average. Some other criteria, such as the
+median absolute error, computed by the \code{medAbsErr()} function in
+the \pkg{cv} package, aren't averages of casewise components. The
+distinction is important because, to our knowledge, the statistical
+theory of cross-validation, for example, in \citet{DavisonHinkley:1997},
+\citet{BatesHastieTibshirani:2023}, and \citet{ArlotCelisse:2010}, is
+developed for CV criteria like MSE that are means of casewise
+components. As a consequence, we limit computation of bias adjustment
+and confidence intervals (see below) to criteria that are casewise
+averages.
 
@@
 
-The \texttt{"glm"} method for \texttt{cv()} is largely similar to the
-\texttt{"lm"} method, although the default algorithm, selected
-explicitly by \texttt{method="exact"}, refits the model with each fold
-removed (and is thus equivalent to \texttt{method="naive"} for
-\texttt{"lm"} models). For generalized linear models,
-\texttt{method="Woodbury"} or (for LOO CV) \texttt{method="hatvalues"}
-provide approximate results (see the last section of the vignette for
-details):
+The \code{"glm"} method for \code{cv()} is largely similar to the
+\code{"lm"} method, although the default algorithm, selected explicitly
+by \code{method="exact"}, refits the model with each fold removed (and
+is thus equivalent to \code{method="naive"} for \code{"lm"} models). For
+generalized linear models, \code{method="Woodbury"} or (for LOO CV)
+\code{method="hatvalues"} provide approximate results (see the last
+section of the vignette for details):
 
@@
 To ensure that the two methods use the same 10 folds, we specify the
-seed for R's random-number generator explicitly; here, and as is common
-in our experience, the \texttt{"exact"} and \texttt{"Woodbury"}
+seed for \proglang{R}'s random-number generator explicitly; here, and as
+is common in our experience, the \code{"exact"} and \code{"Woodbury"}
 algorithms produce nearly identical results. The CV estimates of
@@
 are unreliable for models fit to small samples, and by default
-\texttt{cv()} computes them only when the sample size is 400 or larger
-and when the CV criterion employed is an average of casewise components,
-as is the case for Bayes rule. See the final section of the vignette for
+\code{cv()} computes them only when the sample size is 400 or larger and
+when the CV criterion employed is an average of casewise components, as
+is the case for Bayes rule. See the final section of the vignette for
 details of the computation of confidence intervals for bias-adjusted CV
@@
 As for linear models, we report some timings for the various
-\texttt{cv()} methods of computation in LOO CV as well as for the
-\texttt{cv.glm()} function from the \textbf{boot} package (which,
+\textbackslash code\{cv()` methods of computation in LOO CV as well as
+for the \code{cv.glm()} function from the \pkg{boot} package (which,
 recall, refits the model with each case removed, and thus is comparable
-to \texttt{cv()} with \texttt{method="exact"}):
+to \code{cv()} with \code{method="exact"}):
 
@@
   random effects based on classes within schools are crossed. The
-  \texttt{lmer()} function in the \textbf{lme4} package is capable of
-  modeling both nested and crossed random effects, and the \texttt{cv()}
-  methods for mixed models in the \textbf{cv} package pertain to both
+  \code{lmer()} function in the \pkg{lme4} package is capable of
+  modeling both nested and crossed random effects, and the \code{cv()}
+  methods for mixed models in the \pkg{cv} package pertain to both
   nested and crossed random effects. We present an example of the latter
@@
 
-The HSB data are included in the \texttt{MathAchieve} and
-\texttt{MathAchSchool} data sets in the \textbf{nlme} package
-\citep{PinheiroBates:2000}. \texttt{MathAchieve} includes
-individual-level data on 7185 students in 160 high schools, and
-\texttt{MathAchSchool} includes school-level data:
+The HSB data are included in the \code{MathAchieve} and
+\code{MathAchSchool} data sets in the \pkg{nlme} package
+\citep{PinheiroBates:2000}. \code{MathAchieve} includes individual-level
+data on 7185 students in 160 high schools, and \code{MathAchSchool}
+includes school-level data:
 
@@
 
-We'll use only the \texttt{School}, \texttt{SES} (students'
-socioeconomic status), and \texttt{MathAch} (their score on a
-standardized math-achievement test) variables in the
-\texttt{MathAchieve} data set, and \texttt{Sector} (\texttt{"Catholic"}
-or \texttt{"Public"}) in the \texttt{MathAchSchool} data set.
+We'll use only the \code{School}, \code{SES} (students' socioeconomic
+status), and \code{MathAch} (their score on a standardized
+math-achievement test) variables in the \code{MathAchieve} data set, and
+\code{Sector} (\code{"Catholic"} or \code{"Public"}) in the
+\code{MathAchSchool} data set.
 
 Some data-management is required before fitting a mixed-effects model to
-the HSB data, for which we use the \textbf{dplyr} package
+the HSB data, for which we use the \pkg{dplyr} package
 \citep{WickhamEtAl:2023}:
@@
 In the process, we created two new school-level variables:
-\texttt{meanses}, which is the average SES for students in each school;
-and \texttt{cses}, which is school-average SES centered at its mean. For
+\code{meanses}, which is the average SES for students in each school;
+and \code{cses}, which is school-average SES centered at its mean. For
 details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.
 
-Still following Fox and Weisberg, we proceed to use the \texttt{lmer()}
-function in the \textbf{lme4} package \citep{BatesEtAl:2015} to fit a
-mixed model for math achievement to the HSB data:
+Still following Fox and Weisberg, we proceed to use the \code{lmer()}
+function in the \pkg{lme4} package \citep{BatesEtAl:2015} to fit a mixed
+model for math achievement to the HSB data:
 
@@
 
-For cluster-level CV, the \texttt{clusterVariables} argument tells
-\texttt{cv()} how the clusters are defined. Were there more than one
+For cluster-level CV, the \code{clusterVariables} argument tells
+\code{cv()} how the clusters are defined. Were there more than one
 clustering variable, say classes within schools, these would be provided
 as a character vector of variable names:
-\texttt{clusterVariables\ =\ c("school",\ "class")}. For cluster-level
-CV, the default is \texttt{k\ =\ "loo"}, that is, leave one cluster out
-at a time; we instead specify \texttt{k\ =\ 10} folds of clusters, each
-fold therefore comprising \(160/10 = 16\) schools.
-
-If the \texttt{clusterVariables} argument is omitted, then case-level CV
-is employed, with \texttt{k\ =\ 10} folds as the default, here each with
+\code{clusterVariables = c("school", "class")}. For cluster-level CV,
+the default is \code{k = "loo"}, that is, leave one cluster out at a
+time; we instead specify \code{k = 10} folds of clusters, each fold
+therefore comprising \(160/10 = 16\) schools.
+
+If the \code{clusterVariables} argument is omitted, then case-level CV
+is employed, with \code{k = 10} folds as the default, here each with
 \(7185/10 \approx 719\) students. Notice that one of the 10 models refit
@@
 
-There is also a \texttt{cv()} method for linear mixed models fit by the
-\texttt{lme()} function in the \textbf{nlme} package, and the arguments
-for \texttt{cv()} in this case are the same as for a model fit by
-\texttt{lmer()} or \texttt{glmer()}. We illustrate with the mixed model
-fit to the HSB data:
+There is also a \code{cv()} method for linear mixed models fit by the
+\code{lme()} function in the \pkg{nlme} package, and the arguments for
+\code{cv()} in this case are the same as for a model fit by
+\code{lmer()} or \code{glmer()}. We illustrate with the mixed model fit
+to the HSB data:
 
@@
 We used the same random-number generator seeds as in the previous
-example cross-validating the model fit by \texttt{lmer()}, and so the
-same folds are employed in both cases.\footnote{The observant reader
-  will notice that we set the argument
-  \texttt{control=list(opt="optim")} in the call to \texttt{lme()},
-  changing the optimizer employed from the default \texttt{"nlminb"}. We
-  did this because with the default optimizer, \texttt{lme()}
-  encountered the same convergence issue as \texttt{lmer()}, but rather
-  than issuing a warning, \texttt{lme()} failed, reporting an error. As
-  it turns out, setting the optimizer to \texttt{"optim"} avoids this
-  problem.} The estimated covariance components and fixed effects in the
-summary output differ slightly between the \texttt{lmer()} and
-\texttt{lme()} solutions, although both functions seek to maximize the
-REML criterion. This is, of course, to be expected when different
-algorithms are used for numerical optimization. To the precision
-reported, the cluster-level CV results for the \texttt{lmer()} and
-\texttt{lme()} models are identical, while the case-level CV results are
-very similar but not identical.
+example cross-validating the model fit by \code{lmer()}, and so the same
+folds are employed in both cases.\footnote{The observant reader will
+  notice that we set the argument \code{control=list(opt="optim")} in
+  the call to \code{lme()}, changing the optimizer employed from the
+  default \code{"nlminb"}. We did this because with the default
+  optimizer, \code{lme()} encountered the same convergence issue as
+  \code{lmer()}, but rather than issuing a warning, \code{lme()} failed,
+  reporting an error. As it turns out, setting the optimizer to
+  \code{"optim"} avoids this problem.} The estimated covariance
+components and fixed effects in the summary output differ slightly
+between the \code{lmer()} and \code{lme()} solutions, although both
+functions seek to maximize the REML criterion. This is, of course, to be
+expected when different algorithms are used for numerical optimization.
+To the precision reported, the cluster-level CV results for the
+\code{lmer()} and \code{lme()} models are identical, while the
+case-level CV results are very similar but not identical.
 
@@
 description above and meant to highlight the issues that arise in
-cross-validating mixed-effects models:\footnote{The
-  \texttt{regsubsets()} function computes several measures of model
-  predictive performance, including the \(R^2\) and \(R^2\) adjusted for
-  degrees of freedom, the residual sums of squares, Mallows's \(C_p\),
-  and the BIC. Several of these are suitable for comparing models with
-  differing numbers of coefficients---we use the BIC below---but all
-  necessarily agree when comparing models with the \emph{same} number of
-  coefficients.}
+cross-validating mixed-effects models:\footnote{The \code{regsubsets()}
+  function computes several measures of model predictive performance,
+  including the \(R^2\) and \(R^2\) adjusted for degrees of freedom, the
+  residual sums of squares, Mallows's \(C_p\), and the BIC. Several of
+  these are suitable for comparing models with differing numbers of
+  coefficients---we use the BIC below---but all necessarily agree when
+  comparing models with the \emph{same} number of coefficients.}
 
@@
 95\% concentration ellipse for each cluster:\footnote{We find it
-  convenient to use the \textbf{lattice} \citep{Sarkar:2008} and
-  \textbf{latticeExtra} \citep{SarkarAndrews:2022} packages for this and
+  convenient to use the \pkg{lattice} \citep{Sarkar:2008} and
+  \pkg{latticeExtra} \citep{SarkarAndrews:2022} packages for this and
   other graphs in this section.}
@@
 We proceed to obtain predictions from the random-intercept model
-(\texttt{mod.0}) and the other models (\texttt{mod.1}, \texttt{mod.2},
-and \texttt{mod.3}) based on fixed effects alone, as would be used for
+(\code{mod.0}) and the other models (\code{mod.1}, \code{mod.2}, and
+\code{mod.3}) based on fixed effects alone, as would be used for
 cross-validation based on clusters (i.e., students), and for fixed and
@@
 
-Predictions based on the random-intercept model \texttt{mod.0} for the
+Predictions based on the random-intercept model \code{mod.0} for the
 first 10 students are shown in the following graph:
@@
 
-Our next model, \texttt{mod.1}, includes a fixed intercept and fixed
-effect of \texttt{x} along with a random intercept:
+Our next model, \code{mod.1}, includes a fixed intercept and fixed
+effect of \code{x} along with a random intercept:
 
@@
 clusters, are very poor---indeed, much worse than the fixed-effects-only
-predictions based on the simpler random-intercept model, \texttt{mod.0}.
+predictions based on the simpler random-intercept model, \code{mod.0}.
 We therefore anticipate (and show later in this section) that case-based
-cross-validation will prefer \texttt{mod1} to \texttt{mod0}, but that
-cluster-based cross-validation will prefer \texttt{mod0} to
-\texttt{mod1}.
-
-Our third model, \texttt{mod.2}, includes the contextual effect of
-\(x\)---that is, the cluster mean \texttt{xm}---along with \(x\) and the
-intercept in the fixed-effect part of the model, and a random intercept:
+cross-validation will prefer \code{mod1} to \code{mod0}, but that
+cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.
+
+Our third model, \code{mod.2}, includes the contextual effect of
+\(x\)---that is, the cluster mean \textbackslash code\{xm{]}---along
+with \(x\) and the intercept in the fixed-effect part of the model, and
+a random intercept:
 
@@
 This model is equivalent to fitting
-\texttt{y\ \textasciitilde{}\ I(x\ -\ xm)\ +\ xm\ +\ (1\ \textbar{}\ group)},
-which is the model that generated the data once the coefficient of the
-contextual predictor \texttt{xm} is set to 0 (as it is in
-\texttt{mod.3}, discussed below).
-
-Predictions from model \texttt{mod.2} appear in the following graph:
+\code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that
+generated the data once the coefficient of the contextual predictor
+\code{xm} is set to 0 (as it is in \textbackslash code\{mod.3{]},
+discussed below).
+
+Predictions from model \code{mod.2} appear in the following graph:
 
@@
 Depending on the estimated variance parameters of the model, a mixed
-model like \texttt{mod.2} will apply varying degrees of shrinkage to the
+model like \code{mod.2} will apply varying degrees of shrinkage to the
 random-intercept BLUPs that correspond to variation in the heights of
 the parallel fitted lines for the individual students. In our contrived
-data, the \texttt{mod.2} applies little shrinkage, allowing substantial
+data, the \code{mod.2} applies little shrinkage, allowing substantial
 variability in the heights of the fitted lines, which closely approach
 the observed values for each student. The fit of the mixed model
-\texttt{mod.2} is consequently similar to that of a fixed-effects model
+\code{mod.2} is consequently similar to that of a fixed-effects model
 with age and a categorical predictor for individual students (i.e.,
@@
 
-The mixed model \texttt{mod.2} therefore fits individual observations
+The mixed model \code{mod.2} therefore fits individual observations
 well, and we anticipate a favorable assessment using individual-based
@@
 alone, and so we expect that cluster-based cross-validation won't show
-an advantage for model \texttt{mod.2} compared to the smaller model
-\texttt{mod.0}, which includes only fixed and random intercepts.
+an advantage for model \code{mod.2} compared to the smaller model
+\code{mod.0}, which includes only fixed and random intercepts.
 
@@
 
-Finally, \texttt{mod.3} directly estimates the model used to generate
-the data. As mentioned, it is a constrained version of \texttt{mod.2},
-with the coefficient of \texttt{xm} set to 0, and with \texttt{x}
-expressed as a deviation from the cluster mean \texttt{xm}:
+Finally, \code{mod.3} directly estimates the model used to generate the
+data. As mentioned, it is a constrained version of \code{mod.2}, with
+the coefficient of \code{xm} set to 0, and with \code{x} expressed as a
+deviation from the cluster mean \code{xm}:
 
@@
 
-The predictions from \texttt{mod.3} are therefore similar to those from
-\texttt{mod.2}:
+The predictions from \code{mod.3} are therefore similar to those from
+\code{mod.2}:
 
@@
 effects only. In order to reduce between-model random variability in
-comparisons of models, we apply \texttt{cv()} to the list of models
-created by the \texttt{models()} function (introduced previously),
+comparisons of models, we apply \code{cv()} to the list of models
+created by the \code{models()} function (introduced previously),
 performing cross-validation with the same folds for each model:
@@
 
-In summary, model \texttt{mod.1}, with \(x\) alone and without the
+In summary, model \code{mod.1}, with \(x\) alone and without the
 contextual mean of \(x\), is assessed as fitting very poorly by
 cluster-based CV, but relatively much better by case-based CV. Model
-\texttt{mod.2}, which includes both \(x\) and its contextual mean,
+\code{mod.2}, which includes both \(x\) and its contextual mean,
 produces better results using both cluster-based and case-based CV. The
-data-generating model, \texttt{mod.3}, which includes the fixed effect
-of \texttt{x\ -\ xm} in place of separate terms in \texttt{x} and
-\texttt{xm}, isn't distinguishable from model \texttt{mod.2}, which
-includes \texttt{x} and \texttt{xm} separately, even though
-\texttt{mod.2} has an unnecessary parameter (recall that the population
-coefficient of \texttt{xm} is 0 when \texttt{x} is expressed as
-deviations from the contextual mean). These conclusions are consistent
-with our observations based on graphing predictions from the various
-models, and they illustrate the desirability of assessing mixed-effect
-models at different hierarchical levels.
+data-generating model, \code{mod.3}, which includes the fixed effect of
+\code{x - xm} in place of separate terms in \code{x} and \code{xm},
+isn't distinguishable from model \code{mod.2}, which includes \code{x}
+and \code{xm} separately, even though \code{mod.2} has an unnecessary
+parameter (recall that the population coefficient of \code{xm} is 0 when
+\code{x} is expressed as deviations from the contextual mean). These
+conclusions are consistent with our observations based on graphing
+predictions from the various models, and they illustrate the
+desirability of assessing mixed-effect models at different hierarchical
+levels.
 
@@
 We'll illustrate with data on pig growth, introduced by \citet[Table
-3.1]{DiggleLiangZeger:1994}. The data are in the \texttt{Pigs} data
-frame in the \textbf{cv} package:
+3.1]{DiggleLiangZeger:1994}. The data are in the \code{Pigs} data frame
+in the \pkg{cv} package:
 
@@
 weight of the pig recorded in kg. The data are in ``long'' format, as is
-appropriate for use with the \texttt{lmer()} function in the
-\textbf{lme4} package. The data are very regular, with no missing cases.
+appropriate for use with the \code{lmer()} function in the \pkg{lme4}
+package. The data are very regular, with no missing cases.
 
@@
 
-The \textbf{Stata} mixed-effects models manual proposes a model with
-crossed random effects for the \texttt{Pigs} data \citep[ page
+The \proglang{Stata} mixed-effects models manual proposes a model with
+crossed random effects for the \code{Pigs} data \citep[ page
 37]{Stata:2023}:
@@
   repeated-measures data, which would be more conventionally modeled
-  with autocorrelated errors within pigs. The \texttt{lme()} function in
-  the \textbf{nlme} package, for example, is capable of fitting a
+  with autocorrelated errors within pigs. The \code{lme()} function in
+  the \pkg{nlme} package, for example, is capable of fitting a
   mixed-model of this form.} we think that this is a reasonable
@@
 
-The \textbf{Stata} manual fits the mixed model by maximum likelihood
+The \proglang{Stata} manual fits the mixed model by maximum likelihood
 (rather than REML), and we duplicate the results reported there using
-\texttt{lmer()}:
+\code{lmer()}:
 
@@
 
-We opt for the non-default \texttt{"bobyqa"} optimizer because it
-provides more numerically stable results for subsequent cross-validation
-in this example.
+We opt for the non-default \code{"bobyqa"} optimizer because it provides
+more numerically stable results for subsequent cross-validation in this
+example.
 
@@
 folds composed of weeks, or folds composed of pig-weeks (which in the
-\texttt{Pigs} data set correspond to individual cases, using only the
+\code{Pigs} data set correspond to individual cases, using only the
 fixed effects):
@@
   effects, which would be specified by
-  \texttt{lmer(weight\ \textasciitilde{}\ week\ +\ (1\ \textbar{}\ id/week),\ data=Pigs)}---that
-  is, a random intercept that varies by combinations of \texttt{id}
-  (pig) and \texttt{week}. This model can't be fit, however: With only
-  one case per combination of \texttt{id} and \texttt{week}, the nested
-  random-effect variance is indistinguishable from the case-level
-  variance.} As well, the difference between the MSE computed for the
-model fit to the full data and the CV estimates of the MSE is greater
-here than for cluster-based predictions.
+  \code{lmer(weight ~ week + (1 | id/week), data=Pigs)}---that is, a
+  random intercept that varies by combinations of \code{id} (pig) and
+  \code{week}. This model can't be fit, however: With only one case per
+  combination of \code{id} and \code{week}, the nested random-effect
+  variance is indistinguishable from the case-level variance.} As well,
+the difference between the MSE computed for the model fit to the full
+data and the CV estimates of the MSE is greater here than for
+cluster-based predictions.
 
@@
 procedure with different randomly selected partitions into folds. All of
-the CV functions in the \textbf{cv} package are capable of repeated
+the CV functions in the \pkg{cv} package are capable of repeated
 cross-validation, with the number of repetitions controlled by the
-\texttt{reps} argument, which defaults to \texttt{1}.
+\code{reps} argument, which defaults to \code{1}.
 
@@
 
-When \texttt{reps} \textgreater{} \texttt{1}, the result returned by
-\texttt{cv()} is an object of class \texttt{"cvList"}---literally a list
-of \texttt{"cv"} objects. The results are reported for each repetition
-and then averaged across repetitions, with the standard deviations of
-the CV criterion and the biased-adjusted CV criterion given in
-parentheses. In this example, there is therefore little variation across
-repetitions, increasing our confidence in the reliability of the
-results.
-
-Notice that the seed that's set in the \texttt{cv()} command pertains to
+When \code{reps} \textgreater{} \code{1}, the result returned by
+\code{cv()} is an object of class \code{"cvList"}---literally a list of
+\code{"cv"} objects. The results are reported for each repetition and
+then averaged across repetitions, with the standard deviations of the CV
+criterion and the biased-adjusted CV criterion given in parentheses. In
+this example, there is therefore little variation across repetitions,
+increasing our confidence in the reliability of the results.
+
+Notice that the seed that's set in the \code{cv()} command pertains to
 the first repetition and the seeds for the remaining repetitions are
@@
   the computation is performed, the order of the replicates in the
-  \texttt{"cvList"} object returned by \texttt{cv()} isn't the same as
-  the order in which the replicates are computed. Each element of the
-  result, however, is a \texttt{"cv"} object with the correct
+  \code{"cvList"} object returned by \code{cv()} isn't the same as the
+  order in which the replicates are computed. Each element of the
+  result, however, is a \code{"cv"} object with the correct
   random-number seed saved, and so this technical detail can be safely
-  ignored. The individual \texttt{"cv"} objects are printed in the order
+  ignored. The individual \code{"cv"} objects are printed in the order
   in which they are stored rather than the order in which they are
@@
 easily replicable, and the seed for each repetition is stored in the
-corresponding element of the \texttt{"cvList"} object (which isn't,
+corresponding element of the \code{"cvList"} object (which isn't,
 however, saved in the example).
@@
 It's also possible to replicate CV when comparing competing models via
-the \texttt{cv()} method for \texttt{"modList"} objects. Recall our
+the \code{cv()} method for \code{"modList"} objects. Recall our
 comparison of polynomial regressions of varying degree fit to the
-\texttt{Auto} data; we performed 10-fold CV for each of 10 models. Here,
+\code{Auto} data; we performed 10-fold CV for each of 10 models. Here,
 we replicate that process 5 times for each model and graph the results:
@@
 
-Next, using the \texttt{stepAIC()} function in the \textbf{MASS} package
+Next, using the \code{stepAIC()} function in the \pkg{MASS} package
 \citep{VenablesRipley:2002}, let us perform a forward stepwise
@@
 using AIC as the model-selection criterion (see the help page for
-\texttt{stepAIC()} for details):\footnote{It's generally advantageous to
+\code{stepAIC()} for details):\footnote{It's generally advantageous to
   start with the largest model, here the one with 100 predictors, and
@@
 
-The \texttt{cvSelect()} function in the \textbf{cv} package allows us to
+The \code{cvSelect()} function in the \pkg{cv} package allows us to
 cross-validate the whole model-selection procedure. The first argument
-to \texttt{cvSelect()} is a model-selection function capable of
-refitting the model with a fold omitted and returning a CV criterion.
-The \texttt{selectStepAIC()} function, also in \textbf{cv} and based on
-\texttt{stepAIC()}, is suitable for use with \texttt{cvSelect()}:
+to \code{cvSelect()} is a model-selection function capable of refitting
+the model with a fold omitted and returning a CV criterion. The
+\code{selectStepAIC()} function, also in \pkg{cv} and based on
+\code{stepAIC()}, is suitable for use with \code{cvSelect()}:
 
@@
 
-The other arguments to \texttt{cvSelect()} are:
+The other arguments to \code{cvSelect()} are:
 
@@
 \item
-  \texttt{data}, the data set to which the model is fit;
+  \code{data}, the data set to which the model is fit;
 \item
-  \texttt{seed}, an optional seed for R's pseudo-random-number
-  generator; as for \texttt{cv()}, if the seed isn't supplied by the
-  user, a seed is randomly selected and saved;
+  \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number
+  generator; as for \code{cv()}, if the seed isn't supplied by the user,
+  a seed is randomly selected and saved;
 \item
   additional arguments required by the model-selection function, here
-  the starting \texttt{model} argument, the \texttt{direction} of model
-  selection, and the \texttt{scope} of models considered (from the model
+  the starting \code{model} argument, the \code{direction} of model
+  selection, and the \code{scope} of models considered (from the model
   with only a regression constant to the model with all 100 predictors).
@@
 
-By default, \texttt{cvSelect()} performs 10-fold CV, and produces an
+By default, \code{cvSelect()} performs 10-fold CV, and produces an
 estimate of MSE for the model-selection procedure even \emph{larger}
@@
 Also by default, when the number of folds is 10 or fewer,
-\texttt{cvSelect()} saves the coefficients of the selected models. In
-this example, the \texttt{compareFolds()} function reveals that the
-variables retained by the model-selection process in the several folds
-are quite different:
+\code{cvSelect()} saves the coefficients of the selected models. In this
+example, the \code{compareFolds()} function reveals that the variables
+retained by the model-selection process in the several folds are quite
+different:
 
@@
 regression for married women's labor-force participation. First, recall
-the logistic regression model that we fit to the \texttt{Mroz} data:
+the logistic regression model that we fit to the \code{Mroz} data:
 
@@
 as the model-selection criterion (via the argument
-\texttt{k=log(nrow(Mroz))} to \texttt{stepAIC()}) selects 5 of the 7
+\code{k=log(nrow(Mroz))} to \code{stepAIC()}) selects 5 of the 7
 original predictors:
@@
 Bayes rule applied to the selected model misclassifies 32\% of the cases
-in the \texttt{Mroz} data.
+in the \code{Mroz} data.
 
@@
 
-Setting \texttt{AIC=FALSE} in the call to \texttt{cvSelect()} uses the
-BIC rather than the AIC as the model-selection criterion. As it turns
-out, exactly the same predictors are selected when each of the 10 folds
-are omitted, and the several coefficient estimates are very similar, as
-we show using \texttt{compareFolds()}:
+Setting \code{AIC=FALSE} in the call to \code{cvSelect()} uses the BIC
+rather than the AIC as the model-selection criterion. As it turns out,
+exactly the same predictors are selected when each of the 10 folds are
+omitted, and the several coefficient estimates are very similar, as we
+show using \code{compareFolds()}:
 
@@
 
-The \textbf{cv} package also provides a \texttt{cvSelect()} procedure,
-\texttt{selectTrans()}, for choosing transformations of the predictors
-and the response in regression.
+The \pkg{cv} package also provides a \code{cvSelect()} procedure,
+\code{selectTrans()}, for choosing transformations of the predictors and
+the response in regression.
 
@@
 data), using data on the prestige and other characteristics of 102
-Canadian occupations circa 1970. The data are in the \texttt{Prestige}
-data frame in the \textbf{carData} package:
+Canadian occupations circa 1970. The data are in the \code{Prestige}
+data frame in the \pkg{carData} package:
 
@@
 
-The variables in the \texttt{Prestige} data set are:
+The variables in the \code{Prestige} data set are:
 
@@
 \item
-  \texttt{education}: average years of education for incumbents in the
+  \code{education}: average years of education for incumbents in the
   occupation, from the 1971 Canadian Census.
 \item
-  \texttt{income}: average dollars of annual income for the occupation,
+  \code{income}: average dollars of annual income for the occupation,
   from the Census.
 \item
-  \texttt{women}: percentage of occupational incumbents who were women,
+  \code{women}: percentage of occupational incumbents who were women,
   also from the Census.
 \item
-  \texttt{prestige}: the average prestige rating of the occupation on a
+  \code{prestige}: the average prestige rating of the occupation on a
   0--100 ``thermometer'' scale, in a Canadian social survey conducted
@@
 \item
-  \texttt{type}, type of occupation, and \texttt{census}, the Census
+  \code{type}, type of occupation, and \code{census}, the Census
   occupational code, which are not used in our example.
@@
 
-The object of a regression analysis for the \texttt{Prestige} data (and
+The object of a regression analysis for the \code{Prestige} data (and
 their original purpose) is to predict occupational prestige from the
@@
 
-A scatterplot matrix (using the \texttt{scatterplotMatrix()} function in
-the \textbf{car} package) of the numeric variables in the data reveals
-that the distributions of \texttt{income} and \texttt{women} are
-positively skewed, and that some of the relationships among the three
-predictors, and between the predictors and the response (i.e.,
-\texttt{prestige}), are nonlinear:
+A scatterplot matrix (using the \code{scatterplotMatrix()} function in
+the \pkg{car} package) of the numeric variables in the data reveals that
+the distributions of \code{income} and \code{women} are positively
+skewed, and that some of the relationships among the three predictors,
+and between the predictors and the response (i.e., \code{prestige}), are
+nonlinear:
 
@@
 
-The \texttt{powerTransform()} function in the \textbf{car} package
-transforms variables towards multivariate normality by a generalization
-of Box and Cox's maximum-likelihood-like approach \citep{BoxCox:1964}.
-Several ``families'' of power transformations can be used, including the
+The \code{powerTransform()} function in the \pkg{car} package transforms
+variables towards multivariate normality by a generalization of Box and
+Cox's maximum-likelihood-like approach \citep{BoxCox:1964}. Several
+``families'' of power transformations can be used, including the
 original Box-Cox family, simple powers (and roots), and two adaptations
@@
 \citet[Chap. 8]{Weisberg:2014}, and \citet[Chap. 3]{FoxWeisberg:2019}
-for details. Because \texttt{women} has some zero values, we use the
+for details. Because \code{women} has some 0 values, we use the
 Yeo-Johnson family:
@@
 
-We thus have evidence of the desirability of transforming
-\texttt{income} (by the \(1/3\) power) and \texttt{women} (by the
-\(0.16\) power---which is close to the ``0'' power, i.e., the log
-transformation), but not \texttt{education}. Applying the ``rounded''
-power transformations makes the predictors better-behaved:
+We thus have evidence of the desirability of transforming \code{income}
+(by the \(1/3\) power) and \code{women} (by the \(0.16\) power---which
+is close to the ``0'' power, i.e., the log transformation), but not
+\code{education}. Applying the ``rounded'' power transformations makes
+the predictors better-behaved:
 
@@
 Similarly, component+residual plots for the two regressions, produced by
-the \texttt{crPlots()} function in the \textbf{car} package, suggest
-that the partial relationship of \texttt{prestige} to \texttt{income} is
-more nearly linear in the transformed data, but the transformation of
-\texttt{women} fails to capture what appears to be a slight quadratic
-partial relationship; the partial relationship of \texttt{prestige} to
-\texttt{education} is close to linear in both regressions:
+the \code{crPlots()} function in the \pkg{car} package, suggest that the
+partial relationship of \code{prestige} to \code{income} is more nearly
+linear in the transformed data, but the transformation of
+\textbackslash code\{women{]} fails to capture what appears to be a
+slight quadratic partial relationship; the partial relationship of
+\code{prestige} to \code{education} is close to linear in both
+regressions:
 
@@
 consider whether there's evidence for transforming the response (using
-\texttt{powerTransform()} for Box and Cox's original method), and we
+\code{powerTransform()} for Box and Cox's original method), and we
 discover that there's not:
@@
 
-The \texttt{selectTrans()} function in the \textbf{cv} package automates
-the process of selecting predictor and response transformations. The
-function takes a \texttt{data} set and ``working'' \texttt{model} as
-arguments, along with the candidate \texttt{predictors} and
-\texttt{response} for transformation, and the transformation
-\texttt{family} to employ. If the \texttt{predictors} argument is
-missing then only the response is transformed, and if the
-\texttt{response} argument is missing, only the supplied predictors are
-transformed. The default \texttt{family} for transforming the predictors
-is \texttt{"bcPower"}---the original Box-Cox family---as is the default
-\texttt{family.y} for transforming the response; here we specify
-\texttt{family="yjPower} because of the zeros in \texttt{women}.
-\texttt{selectTrans()} returns the result of applying a lack-of-fit
-criterion to the model after the selected transformation is applied,
-with the default \texttt{criterion=mse}:
+The \code{selectTrans()} function in the \pkg{cv} package automates the
+process of selecting predictor and response transformations. The
+function takes a \code{data} set and ``working'' \code{model} as
+arguments, along with the candidate \code{predictors} and
+\code{response} for transformation, and the transformation \code{family}
+to employ. If the \code{predictors} argument is missing then only the
+response is transformed, and if the \code{response} argument is missing,
+only the supplied predictors are transformed. The default \code{family}
+for transforming the predictors is \code{"bcPower"}---the original
+Box-Cox family---as is the default \code{family.y} for transforming the
+response; here we specify \code{family="yjPower} because of the 0s in
+\code{women}. \code{selectTrans()} returns the result of applying a
+lack-of-fit criterion to the model after the selected transformation is
+applied, with the default \code{criterion=mse}:
 
@@
 
-\texttt{selectTrans()} also takes an optional \texttt{indices} argument,
+\code{selectTrans()} also takes an optional \code{indices} argument,
 making it suitable for doing computations on a subset of the data (i.e.,
-a CV fold), and hence for use with \texttt{cvSelect()} (see
-\texttt{?selectTrans} for details):
+a CV fold), and hence for use with \code{cvSelect()} (see
+\code{?selectTrans} for details):
 
@@
 transformations for each fold reveals that the predictor
-\texttt{education} and the response \texttt{prestige} are never
-transformed; that the \(1/3\) power is selected for \texttt{income} in
-all of the folds; and that the transformation selected for
-\texttt{women} varies narrowly across the folds between the \(0\)th
-power (i.e., log) and the \(1/3\) power.
+\code{education} and the response \code{prestige} are never transformed;
+that the \(1/3\) power is selected for \code{income} in all of the
+folds; and that the transformation selected for \code{women} varies
+narrowly across the folds between the \(0\)th power (i.e., log) and the
+\(1/3\) power.
 
@@
 proceed to select predictors. It's insufficient to apply these steps
-sequentially, first, for example, using \texttt{cvSelect()} with
-\texttt{selectTrans()} and then with \texttt{selectStepAIC()}; rather we
+sequentially, first, for example, using \code{cvSelect()} with
+\code{selectTrans()} and then with \code{selectStepAIC()}; rather we
 should apply the whole model-selection procedure with each fold omitted.
-The \texttt{selectTransAndStepAIC()} function, also supplied by the
-\textbf{cv} package, does exactly that.
-
-To illustrate this process, we return to the \texttt{Auto} data set:
+The \code{selectTransAndStepAIC()} function, also supplied by the
+\pkg{cv} package, does exactly that.
+
+To illustrate this process, we return to the \code{Auto} data set:
 
@@
 
-We previously used the \texttt{Auto} here in a preliminary example where
+We previously used the \code{Auto} here in a preliminary example where
 we employed CV to inform the selection of the order of a polynomial
-regression of \texttt{mpg} on \texttt{horsepower}. Here, we consider
-more generally the problem of predicting \texttt{mpg} from the other
-variables in the \texttt{Auto} data. We begin with a bit of data
-management, and then examine the pairwise relationships among the
-numeric variables in the data set:
+regression of \code{mpg} on \code{horsepower}. Here, we consider more
+generally the problem of predicting \code{mpg} from the other variables
+in the \code{Auto} data. We begin with a bit of data management, and
+then examine the pairwise relationships among the numeric variables in
+the data set:
 
@@
 
-A comment before we proceed: \texttt{origin} is clearly categorical and
-so converting it to a factor is natural, but we could imagine treating
-\texttt{cylinders} and \texttt{year} as numeric predictors. There are,
-however, only 5 distinct values of \texttt{cylinders} (ranging from 3 to
+A comment before we proceed: \code{origin} is clearly categorical and so
+converting it to a factor is natural, but we could imagine treating
+\code{cylinders} and \code{year} as numeric predictors. There are,
+however, only 5 distinct values of \code{cylinders} (ranging from 3 to
 8), but cars with 3 or 5 cylinders are rare. and none of the cars has 7
 cylinders. There are similarly only 13 distinct years between 1970 and
-1982 in the data, and the relationship between \texttt{mpg} and
-\texttt{year} is difficult to characterize.\footnote{Of course, making
-  the decision to treat \texttt{year} as a factor on this basis could be
+1982 in the data, and the relationship between \code{mpg} and
+\code{year} is difficult to characterize.\footnote{Of course, making the
+  decision to treat \code{year} as a factor on this basis could be
   construed as cheating in the current context, which illustrates the
@@
 
-The component+residual plots, created with the \texttt{crPlots()}
-function in the previously loaded \textbf{car} package, clearly reveal
-the inadequacy of the model.
+The component+residual plots, created with the \code{crPlots()} function
+in the previously loaded \pkg{car} package, clearly reveal the
+inadequacy of the model.
 
@@
 Having transformed both the numeric predictors and the response, we
-proceed to use the \texttt{stepAIC()} function in the \textbf{MASS}
-package to perform predictor selection, employing the BIC
-model-selection criterion (by setting the \texttt{k} argument of
-\texttt{stepAIC()} to \(\log(n)\)):
+proceed to use the \code{stepAIC()} function in the \pkg{MASS} package
+to perform predictor selection, employing the BIC model-selection
+criterion (by setting the \texttt{k} argument of \texttt{stepAIC()} to
+\(\log(n)\)):
 
@@
 The selected model includes three of the numeric predictors,
-\texttt{horsepower}, \texttt{weight}, and \texttt{acceleration}, along
-with the factors \texttt{year} and \texttt{origin}. We can calculate the
-MSE for this model, but we expect that the result will be optimistic
-because we used the whole data to help specify the model
+\code{horsepower}, \code{weight}, and \code{acceleration}, along with
+the factors \code{year} and \code{origin}. We can calculate the MSE for
+this model, but we expect that the result will be optimistic because we
+used the whole data to help specify the model
 
@@
 A perhaps subtle point is that we compute the MSE for the selected model
-on the original \texttt{mpg} response scale rather than the log scale,
-so as to make the selected model comparable to the working model. That's
-slightly uncomfortable given the skewed distribution of \texttt{mpg}. An
+on the original \code{mpg} response scale rather than the log scale, so
+as to make the selected model comparable to the working model. That's
+slightly uncomfortable given the skewed distribution of \code{mpg}. An
 alternative is to use the median absolute error instead of the
-mean-squared error, employing the \texttt{medAbsErr()} function from the
-\textbf{cv} package:
+mean-squared error, employing the \code{medAbsErr()} function from the
+\pkg{cv} package:
 
@@
 
-Now let's use \texttt{cvSelect()} with \texttt{selectTransAndStepAIC()}
-to automate and cross-validate the whole model-specification process:
+Now let's use \code{cvSelect()} with \code{selectTransAndStepAIC()} to
+automate and cross-validate the whole model-specification process:
 
@@
 
-Here, as for \texttt{selectTrans()}, the \texttt{predictors} and
-\texttt{response} arguments specify candidate variables for
-transformation, and \texttt{AIC=FALSE} uses the BIC for model selection.
-The starting model, \texttt{m.auto}, is the working model fit to the
-\texttt{Auto} data. The CV criterion isn't bias-adjusted because median
+Here, as for \code{selectTrans()}, the \code{predictors} and
+\code{response} arguments specify candidate variables for
+transformation, and \code{AIC=FALSE} uses the BIC for model selection.
+The starting model, \code{m.auto}, is the working model fit to the
+\code{Auto} data. The CV criterion isn't bias-adjusted because median
 absolute error isn't a mean of casewise error components.
@@
 \item
-  \texttt{selectTransStepAIC()} automatically computes CV cost criteria,
+  \code{selectTransStepAIC()} automatically computes CV cost criteria,
   here the median absolute error, on the untransformed response scale.
@@
   larger than the median absolute error computed for the model we fit to
-  the \texttt{Auto} data separately selecting transformations of the
+  the \code{Auto} data separately selecting transformations of the
   predictors and the response and then selecting predictors for the
@@
   When we look at the transformations and predictors selected with each
-  of the 10 folds omitted (i.e., the output of \texttt{compareFolds()}),
+  of the 10 folds omitted (i.e., the output of \code{compareFolds()}),
   we see that there is little uncertainty in choosing variable
-  transformations (the \texttt{lam.*}s for the \(x\)s and
-  \texttt{lambda} for \(y\) in the output), but considerably more
-  uncertainty in subsequently selecting predictors: \texttt{horsepower},
-  \texttt{weight}, and \texttt{year} are always included among the
-  selected predictors; \texttt{acceleration} and \texttt{displacement}
-  are each included respectively in 4 and 3 of 10 selected models; and
-  \texttt{cylinders} and \texttt{origin} are each included in only 1 of
-  10 models. Recall that when we selected predictors for the full data,
-  we obtained a model with \texttt{horsepower}, \texttt{weight},
-  \texttt{acceleration}, \texttt{year}, and \texttt{origin}.
+  transformations (the \code{lam.*}s for the \(x\)s and \code{lambda}
+  for \(y\) in the output), but considerably more uncertainty in
+  subsequently selecting predictors: \code{horsepower}, \code{weight},
+  and \code{year} are always included among the selected predictors;
+  \code{acceleration} and \code{displacement} are each included
+  respectively in 4 and 3 of 10 selected models; and \code{cylinders}
+  and \code{origin} are each included in only 1 of 10 models. Recall
+  that when we selected predictors for the full data, we obtained a
+  model with \code{horsepower}, \code{weight}, \code{acceleration},
+  \code{year}, and \code{origin}.
 \end{itemize}
@@
 
-The CV functions in the \textbf{cv} package are all capable of
-performing parallel computations by setting the \texttt{ncores} argument
-(specifying the number of computer cores to be used) to a number
-\textgreater{} \texttt{1} (which is the default). Parallel computation
-can be advantageous for large problems, reducing the execution time of
-the program.
+The CV functions in the \pkg{cv} package are all capable of performing
+parallel computations by setting the \code{ncores} argument (specifying
+the number of computer cores to be used) to a number \textgreater{}
+\code{1} (which is the default). Parallel computation can be
+advantageous for large problems, reducing the execution time of the
+program.
 
@@
    user  system elapsed 
-  0.249   0.004   0.253 
+  0.241   0.010   0.253 
 \end{CodeOutput}
@@
    user  system elapsed 
-  0.035   0.004   0.949 
+  0.035   0.004   0.894 
 \end{CodeOutput}
@@
 
-The \textbf{cv} package is designed to be extensible in several
-directions. In this vignette, we discuss three kinds of extensions,
-ordered by increasing general complexity: (1) adding a cross-validation
-cost criterion; (2) adding a model class that's not directly
-accommodated by the \texttt{cv()} default method or by another directly
-inherited method, with separate consideration of mixed-effects models;
-and (3) adding a new model-selection procedure suitable for use with
-\texttt{selectModel()}.
+The \pkg{cv} package is designed to be extensible in several directions.
+In this vignette, we discuss three kinds of extensions, ordered by
+increasing general complexity: (1) adding a cross-validation cost
+criterion; (2) adding a model class that's not directly accommodated by
+the \code{cv()} default method or by another directly inherited method,
+with separate consideration of mixed-effects models; and (3) adding a
+new model-selection procedure suitable for use with
+\code{selectModel()}.
 
@@
 
-A cost criterion suitable for use with \texttt{cv()} or
-\texttt{cvSelect()} should take two arguments, \texttt{y} (the observed
-response vector) and \texttt{yhat} (a vector of fitted or predicted
-response values), and return a numeric index of lack of fit. The
-\textbf{cv} package supplies several such criteria:
-\texttt{mse(y,\ yhat)}, which returns the mean-squared prediction error
-for a numeric response; \texttt{rmse(y,\ yhat)}, which returns the
-(square-)root mean-squared error; \texttt{medAbsErr(y,\ yhat)}, which
-returns the median absolute error; and \texttt{BayesRule(y,\ yhat)} (and
-its non-error-checking version, \texttt{BayesRule2(y,\ yhat))}, suitable
-for use with a binary regression model, where \texttt{y} is the binary
-response coded \texttt{0} for a ``failure'' or \texttt{1} for a
-``success''; where \texttt{yhat} is the predicted probability of
-success; and where the proportion of \emph{incorrectly} classified cases
-is returned.
+A cost criterion suitable for use with \code{cv()} or \code{cvSelect()}
+should take two arguments, \code{y} (the observed response vector) and
+\code{yhat} (a vector of fitted or predicted response values), and
+return a numeric index of lack of fit. The \pkg{cv} package supplies
+several such criteria: \code{mse(y, yhat)}, which returns the
+mean-squared prediction error for a numeric response;
+\code{rmse(y, yhat)}, which returns the (square-)root mean-squared
+error; \code{medAbsErr(y, yhat)}, which returns the median absolute
+error; and \code{BayesRule(y, yhat)} (and its non-error-checking
+version, \code{BayesRule2(y, yhat))}, suitable for use with a binary
+regression model, where \textbackslash code\{y` is the binary response
+coded \code{0} for a ``failure'' or \code{1} for a ``success''; where
+\code{yhat} is the predicted probability of success; and where the
+proportion of \emph{incorrectly} classified cases is returned.
 
@@
 a common summary measure based on the ROC
-\citep[see][]{Wikipedia-ROC:2023}. The \textbf{Metrics} package
+\citep[see][]{Wikipedia-ROC:2023}. The \pkg{Metrics} package
 \citep{HamnerFrasco:2018} includes a variety of measures useful for
-model selection, including an \texttt{auc()} function. We convert the
-AUC into a cost measure by taking its complement:
+model selection, including an \code{auc()} function. We convert the AUC
+into a cost measure by taking its complement:
 
@@
 
-We then apply \texttt{AUCcomp()} to the the Mroz logistic regression
-discussed in the main \textbf{cv} package vignette, which we reproduce
-here, using the \texttt{Mroz} data frame from the \textbf{carData}
-package \citep{FoxWeisberg:2019}:
+We then apply \code{AUCcomp()} to the the Mroz logistic regression
+discussed in the main \pkg{cv} package vignette, which we reproduce
+here, using the \code{Mroz} data frame from the \pkg{carData} package
+\citep{FoxWeisberg:2019}:
 
@@
 As we explain in the vignette ``Cross-validation of regression models,''
-the \texttt{cv()} function differentiates between CV criteria that are
+the \code{cv()} function differentiates between CV criteria that are
 averages of casewise components and criteria that are not. Computation
@@
 
-\texttt{cv()} looks for a \texttt{"casewise\ loss"} attribute of the
-value returned by a CV criterion function. If this attribute exists,
-then the criterion is treated as the mean of casewise components, and
-\texttt{cv()} uses the unexported function \texttt{getLossFn()} to
-construct a function that returns the casewise components of the
-criterion.
-
-We illustrate with the \texttt{mse()}:
+\code{cv()} looks for a \code{"casewise loss"} attribute of the value
+returned by a CV criterion function. If this attribute exists, then the
+criterion is treated as the mean of casewise components, and \code{cv()}
+uses the unexported function \code{getLossFn()} to construct a function
+that returns the casewise components of the criterion.
+
+We illustrate with the \code{mse()}:
 
@@
 }
-<environment: 0x12bdf3978>
+<environment: 0x12f694d48>
 \end{CodeOutput}
@@
 character string (or vector of character strings), here
-\texttt{"(y\ -\ yhat)\^{}2"}, that evaluates to an expression that is a
-function of \texttt{y} and \texttt{yhat}, and that computes the vector
-of casewise components of the CV criterion.
+\code{"(y - yhat)\^2"}, that evaluates to an expression that is a
+function of \code{y} and \code{yhat}, and that computes the vector of
+casewise components of the CV criterion.
 
@@
 Suppose that we want to cross-validate a multinomial logistic regression
-model fit by the \texttt{multinom()} function in the \textbf{nnet}
-package \citep{VenablesRipley:2002}. We borrow an example from
-\citet[Sec. 14.2.1]{Fox:2016}, with data from the British Election Panel
-Study on vote choice in the 2001 British election. Data for the example
-are in the \texttt{BEPS} data frame in the \textbf{carData} package:
+model fit by the \code{multinom()} function in the \pkg{nnet} package
+\citep{VenablesRipley:2002}. We borrow an example from \citet[Sec.
+14.2.1]{Fox:2016}, with data from the British Election Panel Study on
+vote choice in the 2001 British election. Data for the example are in
+the \code{BEPS} data frame in the \pkg{carData} package:
 
@@
 
-The polytomous (multi-category) response variable is \texttt{vote}, a
-factor with levels \texttt{"Conservative"}, \texttt{"Labour"}, and
-\texttt{"Liberal\ Democrat"}. The predictors of \texttt{vote} are:
+The polytomous (multi-category) response variable is \code{vote}, a
+factor with levels \code{"Conservative"}, \code{"Labour"}, and
+\code{"Liberal Democrat"}. The predictors of \code{vote} are:
 
@@
 \item
-  \texttt{age}, in years;
+  \code{age}, in years;
 \item
-  \texttt{econ.cond.national} and \texttt{econ.cond.household}, the
+  \code{econ.cond.national} and \code{econ.cond.household}, the
   respondent's ratings of the state of the economy, on 1 to 5 scales.
 \item
-  \texttt{Blair}, \texttt{Hague}, and \texttt{Kennedy}, ratings of the
-  leaders of the Labour, Conservative, and Liberal Democratic parties,
-  on 1 to 5 scales.
+  \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders
+  of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5
+  scales.
 \item
-  \texttt{Europe}, an 11-point scale on attitude towards European
+  \code{Europe}, an 11-point scale on attitude towards European
   integration, with high scores representing ``Euro-skepticism.''
 \item
-  \texttt{political.knowledge}, knowledge of the parties' positions on
+  \code{political.knowledge}, knowledge of the parties' positions on
   European integration, with scores from 0 to 3.
 \item
-  \texttt{gender}, \texttt{"female"} or \texttt{"male"}.
+  \code{gender}, \code{"female"} or \code{"male"}.
 \end{itemize}
 
-The model fit to the data includes an interaction between
-\texttt{Europe} and \texttt{political.knowledge}; the other predictors
-enter the model additively:
+The model fit to the data includes an interaction between \code{Europe}
+and \code{political.knowledge}; the other predictors enter the model
+additively:
 
@@
 
-Most of the predictors, including the \texttt{Europe} \(\times\)
-\texttt{political.knowledge} interaction, are associated with very small
-\(p\)-values; the \texttt{Anova()} function is from the \textbf{car}
-package \citep{FoxWeisberg:2019}.
-
-Here's an ``effect plot'', using the the \textbf{effects} package
-\citep{FoxWeisberg:2019} to visualize the \texttt{Europe} \(\times\)
-\texttt{political.knowledge} interaction in a ``stacked-area'' graph:
+Most of the predictors, including the \code{Europe} \(\times\)
+\code{political.knowledge} interaction, are associated with very small
+\(p\)-values; the \code{Anova()} function is from the \pkg{car} package
+\citep{FoxWeisberg:2019}.
+
+Here's an ``effect plot'', using the the \pkg{effects} package
+\citep{FoxWeisberg:2019} to visualize the \code{Europe} \(\times\)
+\code{political.knowledge} interaction in a ``stacked-area'' graph:
 
@@
 To cross-validate this multinomial-logit model we need an appropriate
-cost criterion. None of the criteria supplied by the \textbf{cv}
-package---for example, neither \texttt{mse()}, which is appropriate for
-a numeric response, nor \texttt{BayesRule()}, which is appropriate for a
+cost criterion. None of the criteria supplied by the \pkg{cv}
+package---for example, neither \code{mse()}, which is appropriate for a
+numeric response, nor \code{BayesRule()}, which is appropriate for a
 binary response---will do. One possibility is to adapt Bayes rule to a
@@
 
-The \texttt{predict()} method for \texttt{"multinom"} models called with
-argument \texttt{type="class"} reports the Bayes-rule prediction for
-each case---that is, the response category with the highest predicted
-probability. Our \texttt{BayesRuleMulti()} function calculates the
+The \code{predict()} method for \code{"multinom"} models called with
+argument \code{type="class"} reports the Bayes-rule prediction for each
+case---that is, the response category with the highest predicted
+probability. Our \code{BayesRuleMulti()} function calculates the
 proportion of misclassified cases. Because this value is the mean of
-casewise components, we attach a \texttt{"casewise\ loss"} attribute to
-the result (as explained in the preceding section).
+casewise components, we attach a \code{"casewise loss"} attribute to the
+result (as explained in the preceding section).
 
@@
 
-We check first whether the default \texttt{cv()} method works
-``out-of-the-box'' for the \texttt{"multinom"} model:
+We check first whether the default \code{cv()} method works
+``out-of-the-box'' for the \code{"multinom"} model:
 
@@
 
-The default method of \texttt{GetResponse()} (a function supplied by the
-\textbf{cv} package---see \texttt{?GetResponse}) fails for a
-\texttt{"multinom"} object. A straightforward solution is to supply a
-\texttt{GetResponse.multinom()} method that returns the factor response
+The default method of \code{GetResponse()} (a function supplied by the
+\pkg{cv} package---see \code{?GetResponse}) fails for a
+\code{"multinom"} object. A straightforward solution is to supply a
+\code{GetResponse.multinom()} method that returns the factor response
 \citep[using the \texttt{get\_response()} function from the
-\textbf{insight} package,][]{LudeckeWaggonerMakowski:2019},
+\pkg{insight} package,][]{LudeckeWaggonerMakowski:2019},
 
@@
 
-A \texttt{traceback()} (not shown) reveals that the problem is that the
-default method of \texttt{cv()} calls the \texttt{"multinom"} method for
-\texttt{predict()} with the argument \texttt{type="response"}, when the
-correct argument should be \texttt{type="class"}. We therefore must
-write a ``\texttt{multinom}'' method for \texttt{cv()}, but that proves
-to be very simple:
+A \code{traceback()} (not shown) reveals that the problem is that the
+default method of \code{cv()} calls the \code{"multinom"} method for
+\code{predict()} with the argument \code{type="response"}, when the
+correct argument should be \code{type="class"}. We therefore must write
+a \code{"multinom"} method for \code{cv()}, but that proves to be very
+simple:
 
@@
 
-That is, we simply call the default \texttt{cv()} method with the
-\texttt{type} argument properly set. In addition to supplying the
-correct \texttt{type} argument, our method sets the default
-\texttt{criterion} for the \texttt{cv.multinom()} method to
-\texttt{BayesRuleMulti}.
+That is, we simply call the default \code{cv()} method with the
+\code{type} argument properly set. In addition to supplying the correct
+\code{type} argument, our method sets the default \code{criterion} for
+the \code{cv.multinom()} method to \code{BayesRuleMulti}.
 
@@
 
-Prior to invoking \texttt{cv()}, we called \texttt{update()} with
-\texttt{trace=FALSE} to suppress the iteration history reported by
-default by \texttt{multinom()}---it would be tedious to see the
-iteration history for each fold. The cross-validated polytomous
-Bayes-rule criterion confirms that the fitted model does substantially
-better than the marginal Bayes-rule prediction that everyone votes for
-Labour.
+Prior to invoking \code{cv()}, we called \code{update()} with
+\code{trace=FALSE} to suppress the iteration history reported by default
+by \code{multinom()}---it would be tedious to see the iteration history
+for each fold. The cross-validated polytomous Bayes-rule criterion
+confirms that the fitted model does substantially better than the
+marginal Bayes-rule prediction that everyone votes for Labour.
 
@@
 
-Adding a \texttt{cv()} method for a mixed-model class is somewhat more
-complicated. We provide the \texttt{cvMixed()} function to facilitate
-this process, and to see how that works, consider the \texttt{"lme"}
-method from the \textbf{cv} package:
+Adding a \code{cv()} method for a mixed-model class is somewhat more
+complicated. We provide the \code{cvMixed()} function to facilitate this
+process, and to see how that works, consider the \code{"lme"} method
+from the \pkg{cv} package:
 
@@
 }
-<bytecode: 0x117501408>
+<bytecode: 0x148948190>
 <environment: namespace:cv>
@@
 
-Notice that \texttt{cv.lme()} sets up a call to \texttt{cvMixed()},
-which does the computational work.
-
-Most of the arguments of \texttt{cvMixed()} are familiar:
+Notice that \code{cv.lme()} sets up a call to \code{cvMixed()}, which
+does the computational work.
+
+Most of the arguments of \code{cvMixed()} are familiar:
 
@@
 \item
-  \texttt{model} is the mixed-model object, here of class
-  \texttt{"lme"}.
+  \code{model} is the mixed-model object, here of class \code{"lme"}.
 \item
-  \texttt{package} is the name of the package in which the
-  mixed-modeling function used to fit the model, here \texttt{lme()},
-  resides---i.e., \texttt{"nlme"}; \texttt{cvMixed()} uses this argument
-  to retrieve the package namespace.
+  \code{package} is the name of the package in which the mixed-modeling
+  function used to fit the model, here \code{lme()}, resides---i.e.,
+  \code{"nlme"}; \code{cvMixed()} uses this argument to retrieve the
+  package namespace.
 \item
-  \texttt{data} is the data set to which the model is fit, by default
-  extracted by the \texttt{get\_data()} function in the \textbf{insight}
+  \code{data} is the data set to which the model is fit, by default
+  extracted by the \code{get_data()} function in the \pkg{insight}
   package.
 \item
-  \texttt{criterion} is the CV criterion, defaulting to the
-  \texttt{mse()} function.
+  \code{criterion} is the CV criterion, defaulting to the \code{mse()}
+  function.
 \item
-  \texttt{k} is the number of CV folds, defaulting to \texttt{"loo"} for
-  CV by clusters and \texttt{10} for CV by cases.
+  \code{k} is the number of CV folds, defaulting to \code{"loo"} for CV
+  by clusters and \code{10} for CV by cases.
 \item
-  \texttt{reps} is the number of times the CV process is repeated,
-  defaulting to \texttt{1}.
+  \code{reps} is the number of times the CV process is repeated,
+  defaulting to \code{1}.
 \item
-  \texttt{seed} is the seed for R's random-number generator, defaulting
-  to a randomly selected (and saved) value.
+  \code{seed} is the seed for \proglang{R}'s random-number generator,
+  defaulting to a randomly selected (and saved) value.
 \item
-  \texttt{ncores} is the number of cores to use for parallel
-  computation; if \texttt{1}, the default, then the computation isn't
-  parallelized.
+  \code{ncores} is the number of cores to use for parallel computation;
+  if \code{1}, the default, then the computation isn't parallelized.
 \item
-  \texttt{clusterVariables} is a character vector of the names of
+  \code{clusterVariables} is a character vector of the names of
   variables defining clusters; if missing, then CV is based on cases
@@
 \item
-  \texttt{predict.clusters.args} is a named list of arguments to be
-  passed to the \texttt{predict()} function to obtain predictions for
-  the full data set from a model fit to a subset of the data for
-  cluster-based CV. The first two arguments should be \texttt{object}
-  and \texttt{newdata}. It is typically necessary to tell
-  \texttt{cvMixed()} how to base predictions only on fixed effects; in
-  the case of \texttt{"lme"} models, this is done by setting
-  \texttt{level\ =\ 0}.
+  \code{predict.clusters.args} is a named list of arguments to be passed
+  to the \code{predict()} function to obtain predictions for the full
+  data set from a model fit to a subset of the data for cluster-based
+  CV. The first two arguments should be \code{object} and
+  \code{newdata}. It is typically necessary to tell \code{cvMixed()} how
+  to base predictions only on fixed effects; in the case of \code{"lme"}
+  models, this is done by setting \code{level = 0}.
 \item
-  Similarly, \texttt{predict.cases.args} is a named list of arguments to
-  be passed to \texttt{predict()} for case-based CV. Setting
-  \texttt{level\ =\ 1} includes random effects in the predictions.
+  Similarly, \code{predict.cases.args} is a named list of arguments to
+  be passed to \code{predict()} for case-based CV. Setting
+  \code{level = 1} includes random effects in the predictions.
 \end{itemize}
 
-Finally, any additional arguments, absorbed by \texttt{...}, are passed
-to \texttt{update()} when the model is refit with each fold omitted.
-\texttt{cvMixed()} returns an object of class \texttt{"cv"}.
+Finally, any additional arguments, absorbed by \code{...}, are passed to
+\code{update()} when the model is refit with each fold omitted.
+\code{cvMixed()} returns an object of class \code{"cv"}.
 
 Now imagine that we want to support a new class of mixed-effects models.
-To be concrete, we illustrate with the \texttt{glmmPQL()} function in
-the \textbf{MASS} package \citep{VenablesRipley:2002}, which fits
+To be concrete, we illustrate with the \code{glmmPQL()} function in the
+\pkg{MASS} package \citep{VenablesRipley:2002}, which fits
 generalized-linear mixed-effects models by penalized
 quasi-likelihood.\footnote{This example is somewhat artificial in that
-  \texttt{glmmPQL()} has largely been superseded by computationally
-  superior functions, such the \texttt{glmer()} function in the
-  \textbf{lme4} package. There is, however, one situation in which
-  \texttt{glmmPQL()} might prove useful: to specify serial dependency in
-  case-level errors within clusters for longitudinal data, which is not
-  currently supported by \texttt{glmer()}.} Not coincidentally, the
-arguments of \texttt{glmmPQL()} are similar to those of \texttt{lme()}
-(with an additional \texttt{family} argument), because the former
-iteratively invokes the latter; so \texttt{cv.glmmPQL()} should resemble
-\texttt{cv.lme()}.
-
-As it turns out, neither the default method for \texttt{GetResponse()}
-nor \texttt{insight::get\_data()} work for \texttt{"glmmPQL"} objects.
-These objects include a \texttt{"data"} element, however, and so we can
-simply extract this element as the default for the \texttt{data}
-argument of our \texttt{cv.glmmPQL()} method.
+  \code{glmmPQL()} has largely been superseded by computationally
+  superior functions, such the \code{glmer()} function in the \pkg{lme4}
+  package. There is, however, one situation in which \code{glmmPQL()}
+  might prove useful: to specify serial dependency in case-level errors
+  within clusters for longitudinal data, which is not currently
+  supported by \code{glmer()}.} Not coincidentally, the arguments of
+\code{glmmPQL()} are similar to those of \code{lme()} (with an
+additional \code{family} argument), because the former iteratively
+invokes the latter; so \code{cv.glmmPQL()} should resemble
+\code{cv.lme()}.
+
+As it turns out, neither the default method for \code{GetResponse()} nor
+\code{insight::get_data()} work for \code{"glmmPQL"} objects. These
+objects include a \code{"data"} element, however, and so we can simply
+extract this element as the default for the \code{data} argument of our
+\code{cv.glmmPQL()} method.
 
@@
 
-Writing the \texttt{cv()} method is then straightforward:
+Writing the \code{cv()} method is then straightforward:
 
@@
 
-We set the argument \texttt{verbose=FALSE} to suppress
-\texttt{glmmPQL()}'s iteration counter when \texttt{cvMixed()} calls
-\texttt{update()}.
+We set the argument \code{verbose=FALSE} to suppress \code{glmmPQL()}'s
+iteration counter when \code{cvMixed()} calls \code{update()}.
 
 Let's apply our newly minted method to a logistic regression with a
-random intercept in an example that appears in \texttt{?glmmPQL}:
+random intercept in an example that appears in \code{?glmmPQL}:
 
@@
 
-We compare this result to that obtained from \texttt{glmer()} in the
-\textbf{lme4} package:
+We compare this result to that obtained from \code{glmer()} in the
+\pkg{lme4} package:
 
@@
 
-Finally, we try out our \texttt{cv.glmmPQL()} method, cross-validating
+Finally, we try out our \code{cv.glmmPQL()} method, cross-validating
 both by clusters and by cases,
@@
 
-and again compare to \texttt{glmer()}:
+and again compare to \code{glmer()}:
 
@@
 
-The \texttt{selectStepAIC()} function supplied by the \textbf{cv}
-package, which is based on the \texttt{stepAIC()} function from the
-\textbf{nnet} package \citep{VenablesRipley:2002} for stepwise model
-selection, is suitable for the \texttt{procedure} argument of
-\texttt{cvSelect()}. The use of \texttt{selectStepAIC()} is illustrated
-in the principal vignette for the package.
-
-We'll employ \texttt{selectStepAIC()} as a ``template'' for writing a CV
+The \code{selectStepAIC()} function supplied by the \pkg{cv} package,
+which is based on the \code{stepAIC()} function from the \pkg{nnet}
+package \citep{VenablesRipley:2002} for stepwise model selection, is
+suitable for the \code{procedure} argument of \code{cvSelect()}. The use
+of \code{selectStepAIC()} is illustrated in the principal vignette for
+the package.
+
+We'll employ \code{selectStepAIC()} as a ``template'' for writing a CV
 model-selection procedure. To see the code for this function, type
-\texttt{cv::selectStepAIC} at the R command prompt, or examine the
-sources for the \textbf{cv} package at
+\code{cv::selectStepAIC} at the \proglang{R} command prompt, or examine
+the sources for the \pkg{cv} package at
 \url{https://github.com/gmonette/cv} (the code for
-\texttt{selectStepAIC()} is in
+\code{selectStepAIC()} is in
 \url{https://github.com/gmonette/cv/blob/main/R/cvSelect.R}).
@@
 Another approach to model selection is all-subsets regression. The
-\texttt{regsubsets()} function in the \textbf{leaps} package
+\code{regsubsets()} function in the \pkg{leaps} package
 \citep{LumleyMiller:2020} implements an efficient algorithm for
@@
 of predictors of all sizes, from 1 through the maximum number of
-candidate predictors.\footnote{The \texttt{regsubsets()} function
-  computes several measures of model predictive performance, including
-  the \(R^2\) and \(R^2\) adjusted for degrees of freedom, the residual
-  sums of squares, Mallows's \(C_p\), and the BIC. Several of these are
+candidate predictors.\footnote{The \code{regsubsets()} function computes
+  several measures of model predictive performance, including the
+  \(R^2\) and \(R^2\) adjusted for degrees of freedom, the residual sums
+  of squares, Mallows's \(C_p\), and the BIC. Several of these are
   suitable for comparing models with differing numbers of
@@
   comparing models with the \emph{same} number of coefficients.} To
-illustrate the use of \texttt{regsubsets()}, we employ the
-\texttt{swiss} data frame supplied by the \textbf{leaps} package:
+illustrate the use of \code{regsubsets()}, we employ the \code{swiss}
+data frame supplied by the \pkg{leaps} package:
 
@@
 \item
-  \texttt{Fertility}: A standardized fertility measure.
+  \code{Fertility}: A standardized fertility measure.
 \item
-  \texttt{Agriculture}: The percentage of the male population engaged in
+  \code{Agriculture}: The percentage of the male population engaged in
   agriculture.
 \item
-  \texttt{Examination}: The percentage of draftees into the Swiss army
+  \code{Examination}: The percentage of draftees into the Swiss army
   receiving the highest grade on an examination.
 \item
-  \texttt{Education}: The percentage of draftees with more than a
+  \code{Education}: The percentage of draftees with more than a
   primary-school education.
 \item
-  \texttt{Catholic}: The percentage of the population who were Catholic.
+  \code{Catholic}: The percentage of the population who were Catholic.
 \item
-  \texttt{Infant.Mortality}: The infant-mortality rate, expressed as the
+  \code{Infant.Mortality}: The infant-mortality rate, expressed as the
   percentage of live births surviving less than a year.
@@
 
-Following \citet{LumleyMiller:2020}, we treat \texttt{Fertility} as the
+Following \citet{LumleyMiller:2020}, we treat \code{Fertility} as the
 response and the other variables as predictors in a linear least-squares
@@
 
-The graph, produced by the \texttt{subsets()} function in the
-\textbf{car} package, shows that the model with the smallest BIC is the
-best model with 4 predictors, including \texttt{Agriculture},
-\texttt{Education}, \texttt{Catholic}, and \texttt{Infant.Mortality},
-but not \texttt{Examination}:
+The graph, produced by the \code{subsets()} function in the \pkg{car}
+package, shows that the model with the smallest BIC is the best model
+with 4 predictors, including \code{Agriculture}, \code{Education},
+\code{Catholic}, and \code{Infant.Mortality}, but not
+\code{Examination}:
 
@@
 
-Here's a function named \texttt{selectSubsets()}, meant to be used with
-\texttt{cvSelect()}, suitable for cross-validating the model-selection
+Here's a function named \code{selectSubsets()}, meant to be used with
+\code{cvSelect()}, suitable for cross-validating the model-selection
 process:
@@
 A slightly tricky point is that because of scoping issues,
-\texttt{predict()} doesn't work with the model fit omitting the \(i\)th
+\code{predict()} doesn't work with the model fit omitting the \(i\)th
 fold, and so the fitted values for all cases are computed directly as
@@
 Additionally, the command
-\texttt{lm(y{[}-indices{]}\ \textasciitilde{}\ X{[}-indices,\ x.names.i{]}\ -\ 1)},
-which is the selected model with the \(i\)th fold deleted, produces
-awkward coefficient names like
-\texttt{"X{[}-indices,\ x.names.i{]}Infant.Mortality"}. Purely for
-aesthetic reasons, the command
+\code{lm(y[-indices] ~ X[-indices, x.names.i] - 1)}, which is the
+selected model with the \(i\)th fold deleted, produces awkward
+coefficient names like \code{"X[-indices, x.names.i]Infant.Mortality"}.
+Purely for aesthetic reasons, the command
 \texttt{sub("X\textbackslash{}\textbackslash{}{[}-indices,\ x.names.i\textbackslash{}\textbackslash{}{]}",\ "",\ names(coefs))}
 fixes these awkward names, removing the extraneous text,
-\texttt{"X{[}-indices,\ x.names.i{]}"}.
-
-Applying \texttt{selectSubsets()} to the full data produces the
-full-data cross-validated RMSE (which we obtained previously):
+\code{"X[-indices, x.names.i]"}.
+
+Applying \code{selectSubsets()} to the full data produces the full-data
+cross-validated RMSE (which we obtained previously):
 \%\%\texttt{\{r\ test-selectSubsets\}\ \%\%selectSubsets(model=m.swiss)\ \%\%}
@@
 
-Then, using \texttt{selectSubsets()} in cross-validation, we get:
+Then, using \code{selectSubsets()} in cross-validation, we get:
 
@@
 some uncertainty in identifying the predictors in the ``best'' model,
-with \texttt{Agriculture} sometimes appearing and sometimes not:
+with \code{Agriculture} sometimes appearing and sometimes not:
 
@@
 
-The most straightforward way to implement cross-validation in R for
-statistical modeling functions that are written in the canonical manner
-is to use \texttt{update()} to refit the model with each fold removed.
-This is the approach taken in the default method for \texttt{cv()}, and
-it is appropriate if the cases are independently sampled. Refitting the
-model in this manner for each fold is generally feasible when the number
-of folds in modest, but can be prohibitively costly for leave-one-out
-cross-validation when the number of cases is large.
-
-The \texttt{"lm"} and \texttt{"glm"} methods for \texttt{cv()} take
-advantage of computational efficiencies by avoiding refitting the model
-with each fold removed. Consider, in particular, the weighted linear
-model
+The most straightforward way to implement cross-validation in
+\proglang{R} for statistical modeling functions that are written in the
+canonical manner is to use \code{update()} to refit the model with each
+fold removed. This is the approach taken in the default method for
+\code{cv()}, and it is appropriate if the cases are independently
+sampled. Refitting the model in this manner for each fold is generally
+feasible when the number of folds in modest, but can be prohibitively
+costly for leave-one-out cross-validation when the number of cases is
+large.
+
+The \code{"lm"} and \code{"glm"} methods for \code{cv()} take advantage
+of computational efficiencies by avoiding refitting the model with each
+fold removed. Consider, in particular, the weighted linear model
 \(\mathbf{y}_{n \times 1} = \mathbf{X}_{n \times p}\boldsymbol{\beta}_{p \times 1} + \boldsymbol{\varepsilon}_{n \times 1}\),
@@
   numerical instability when \(\mathbf{X}\) is ill-conditioned.
-  \texttt{lm()} uses the singular-value decomposition of the model
-  matrix to obtain computationally more stable results.} \[
+  \code{lm()} uses the singular-value decomposition of the model matrix
+  to obtain computationally more stable results.} \[
 \mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} 
@@
 inverse of \(\mathbf{A} = \mathbf{X}^T \mathbf{W} \mathbf{X}\) is
-available directly in the \texttt{"lm"} object, but the second term on
-the right-hand side of the Woodbury identity requires a matrix inversion
+available directly in the \code{"lm"} object, but the second term on the
+right-hand side of the Woodbury identity requires a matrix inversion
 with each fold deleted. (In contrast, the inverse of each
@@
 
-\texttt{glm()} computes the maximum-likelihood estimates for a
-generalized linear model by iterated weighted least squares \citep[see,
-e.g.,][Sec. 6.12]{FoxWeisberg:2019}. The last iteration is therefore
-just a WLS fit of the ``working response'' on the model matrix using
-``working weights.'' Both the working weights and the working response
-at convergence are available from the information in the object returned
-by \texttt{glm()}.
+\code{glm()} computes the maximum-likelihood estimates for a generalized
+linear model by iterated weighted least squares \citep[see, e.g.,][Sec.
+6.12]{FoxWeisberg:2019}. The last iteration is therefore just a WLS fit
+of the ``working response'' on the model matrix using ``working
+weights.'' Both the working weights and the working response at
+convergence are available from the information in the object returned by
+\code{glm()}.
 
@@
 is when we would be most tempted to use it. Nevertheless, because these
-results are approximate, the default for the \texttt{"glm"}
-\texttt{cv()} method is to perform the exact computation, which entails
-refitting the model with each fold omitted.
+results are approximate, the default for the \code{"glm"} \code{cv()}
+method is to perform the exact computation, which entails refitting the
+model with each fold omitted.
 
@@
 intervals with better coverage for small samples. We may implement Bates
-et al.'s approach in a later release of the \textbf{cv} package. At
-present we use the confidence interval above for sufficiently large
-\(n\), which, based on Bates et al.'s results, we take by default to be
+et al.'s approach in a later release of the \pkg{cv} package. At present
+we use the confidence interval above for sufficiently large \(n\),
+which, based on Bates et al.'s results, we take by default to be
 \(n \ge 400\).
@@
 
-The following R function computes AUC from \(\mathbf{\widehat{y}}_v\)
-and \(\mathbf{y}_v\), accommodating the cases where \(\mathbf{y}_v\) is
-all 0s or all 1s:
+The following \proglang{R} function computes AUC from
+\(\mathbf{\widehat{y}}_v\) and \(\mathbf{y}_v\), accommodating the cases
+where \(\mathbf{y}_v\) is all 0s or all 1s:
 

### 2024-01-03 scratch/notes_john/JSS-article/JSS-article-reduced.Rmd
added JSS-article-reduced.Rmd (preliminary attempt)

--- 2024-01-02 scratch/notes_john/JSS-article/JSS-article.Rmd
@@
 
-## Logistic regression for the Mroz data
-
-The \code{Mroz} data set from the \pkg{carData} package [associated with @FoxWeisberg:2019] has been used by several authors to illustrate binary logistic regression; see, in particular @FoxWeisberg:2019.
-The data were originally drawn from the U.S.
-Panel Study of Income Dynamics and pertain to married women.
-Here are a few cases in the data set:
-```{r Mroz-data}
-data("Mroz", package="carData")
-head(Mroz, 3)
-tail(Mroz, 3)
-```
-The response variable in the logistic regression is \code{lfp}, labor-force participation, a factor coded \code{"yes"} or \code{"no"}. The remaining variables are predictors:
-
-* \code{k5}, number of children 5 years old of younger in the woman's household;
-* \code{k618}, number of children between 6 and 18 years old;
-* \code{age}, in years;
-* \code{wc}, wife's college attendance, \code{"yes"} or \code{"no"};
-* \code{hc}, husband's college attendance;
-* \code{lwg}, the woman's log wage rate if she is employed, or her *imputed* wage rate, if she is not [a variable that @FoxWeisberg:2019 show is problematically defined]; and
-* \code{inc}, family income, in $1000s, exclusive of wife's income.
-
-We use the \code{glm()} function to fit a binary logistic regression to the \code{Mroz} data:
-```{r Mroz-logistic-regresion}
-m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
-summary(m.mroz)
-
-BayesRule(ifelse(Mroz$lfp == "yes", 1, 0), 
-          fitted(m.mroz, type="response"))
-```
-In addition to the usually summary output for a GLM, we show the result of applying the \code{BayesRule()} function from the \pkg{cv} package to predictions derived from the fitted model.
-Bayes rule, which predicts a "success" in a binary regression model when the fitted probability of success [i.e., $\phi = \Pr(y = 1)$] is $\widehat{\phi} \ge .5$ and a "failure" if $\widehat{\phi} < .5$.[^BayesRule] The first argument to \code{BayesRule()} is the binary {0, 1} response, and the second argument is the predicted probability of success.
-\code{BayesRule()} returns the proportion of predictions that are *in error*, as appropriate for a "cost" function.
-
-[^BayesRule]: \code{BayesRule()} does some error checking; \code{BayesRule2()} is similar, but omits the error checking, and so can be faster for large problems.
-
-The value returned by \code{BayesRule()} is associated with an "attribute" named \code{"casewise loss"} and set to \code{"y != round(yhat)"}, signifying that the Bayes rule CV criterion is computed as the mean of casewise values, here 0 if the prediction for a case matches the observed value and 1 if it does not (signifying a prediction error).
-The \code{mse()} function for numeric responses is also calculated as a casewise average.
-Some other criteria, such as the median absolute error, computed by the \code{medAbsErr()} function in the \pkg{cv} package, aren't averages of casewise components.
-The distinction is important because, to our knowledge, the statistical theory of cross-validation, for example, in @DavisonHinkley:1997, @BatesHastieTibshirani:2023, and @ArlotCelisse:2010, is developed for CV criteria like MSE that are means of casewise components.
-As a consequence, we limit computation of bias adjustment and confidence intervals (see below) to criteria that are casewise averages.
-
-In this example, the fitted logistic regression incorrectly predicts 31% of the responses; we expect this estimate to be optimistic given that the model is used to "predict" the data to which it is fit.
-
-The \code{"glm"} method for \code{cv()} is largely similar to the \code{"lm"} method, although the default algorithm, selected explicitly by \code{method="exact"}, refits the model with each fold removed (and is thus equivalent to \code{method="naive"} for \code{"lm"} models).
-For generalized linear models, \code{method="Woodbury"} or (for LOO CV) \code{method="hatvalues"} provide approximate results (see the last section of the vignette for details):
-
-```{r cv-Mroz-10-fold}
-cv(m.mroz, criterion=BayesRule, seed=248)
-
-cv(m.mroz, criterion=BayesRule, seed=248, method="Woodbury")
-```
-To ensure that the two methods use the same 10 folds, we specify the seed for \proglang{R}'s random-number generator explicitly; here, and as is common in our experience, the \code{"exact"} and \code{"Woodbury"} algorithms produce nearly identical results.
-The CV estimates of prediction error are slightly higher than the estimate based on all of the cases.
-
-The printed output includes a 95% confidence interval for the bias-adjusted Bayes rule CV criterion.
-@BatesHastieTibshirani:2023 show that these confidence intervals are unreliable for models fit to small samples, and by default \code{cv()} computes them only when the sample size is 400 or larger and when the CV criterion employed is an average of casewise components, as is the case for Bayes rule.
-See the final section of the vignette for details of the computation of confidence intervals for bias-adjusted CV criteria.
-
-Here are results of applying LOO CV to the Mroz model, using both the exact and the approximate methods:
-```{r cv-Mroz-LOO}
-cv(m.mroz, k="loo", criterion=BayesRule)
-
-cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury")
-
-cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues")
-```
-To the number of decimal digits shown, the three methods produce identical results for this example.
-
-As for linear models, we report some timings for the various \code{cv()` methods of computation in LOO CV as well as for the \code{cv.glm()} function from the \pkg{boot} package (which, recall, refits the model with each case removed, and thus is comparable to \code{cv()} with \code{method="exact"}):
-```{r glm.timings, cache=TRUE}
-microbenchmark::microbenchmark(
-  hatvalues=cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues"),
-  Woodbury=cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury"),
-  exact=cv(m.mroz, k="loo", criterion=BayesRule),
-  cv.glm=boot::cv.glm(Mroz, m.mroz,
-               cost=BayesRule),
-  times=10)
-```
-There is a substantial time penalty associated with exact computations.
-
 # Cross-validating mixed-effects models
@@
 
-## Example: Crossed random effects
-
-Crossed random effects arise when the structure of the data aren't strictly hierarchical.
-Nevertheless, crossed and nested random effects can be handled in much the same manner, by refitting the mixed-effects model to the data with a fold of clusters or cases removed and using the refitted model to predict the response in the removed fold.
-
-We'll illustrate with data on pig growth, introduced by @DiggleLiangZeger:1994 [Table 3.1]. The data are in the \code{Pigs} data frame in the \pkg{cv} package:
-```{r pigs}
-head(Pigs, 9)
-head(xtabs(~ id + week, data=Pigs), 3)
-tail(xtabs(~ id + week, data=Pigs), 3)
-```
-Each of 48 pigs is observed weekly over a period of 9 weeks, with the weight of the pig recorded in kg.
-The data are in "long" format, as is appropriate for use with the \code{lmer()} function in the \pkg{lme4} package.
-The data are very regular, with no missing cases.
-
-The following graph, showing the growth trajectories of the pigs, is similar to Figure 3.1 in @DiggleLiangZeger:1994; we add an overall least-squares line and a loess smooth, which are nearly indistinguishable:
-```{r pigs-graph}
-#| out.width = "60%",
-#| fig.height = 6,
-#| fig.cap = "Growth trajectories for 48 pigs, with overall least-squares line (sold blue) and loess line (broken magenta)."
-plot(weight ~ week, data=Pigs, type="n")
-for (i in unique(Pigs$id)){
-  with(Pigs, lines(x=1:9, y=Pigs[id == i, "weight"],
-                   col="gray"))
-}
-abline(lm(weight ~ week, data=Pigs), col="blue", lwd=2)
-lines(with(Pigs, loess.smooth(week, weight, span=0.5)),
-      col="magenta", lty=2, lwd=2)
-```
-The individual "growth curves" and the overall trend are generally linear, with some tendency for variability of pig weight to increase over weeks (a feature of the data that we ignore in the mixed model that we fit to the data below).
-
-The \proglang{Stata} mixed-effects models manual proposes a model with crossed random effects for the \code{Pigs} data [@Stata:2023 page 37]:
-
-> [S]uppose that we wish to fit
-$$
-\mathrm{weight}_{ij} = \beta_0 + \beta_1 \mathrm{week}_{ij} + u_i + v_j + \varepsilon_{ij} 
-$$
-for the $i = 1, \ldots, 9$ weeks and $j = 1, \dots, 48$ pigs and
-$$
-u_i \sim N(0, \sigma^2_u); v_j \sim N(0, \sigma^2_v ); \varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon)
-$$
-all independently.
-That is, we assume an overall population-average growth curve $\beta_0 + \beta_1 \mathrm{week}$ and a random pig-specific shift.
-In other words, the effect due to week, $u_i$, is systematic to that week and  common to all pigs.
-The rationale behind [this model] could be that, assuming that the pigs were measured contemporaneously, we might be concerned that week-specific random factors such as weather and feeding patterns had significant systematic effects on all pigs.
-
-Although we might prefer an alternative model,[^pig-alternative] we think that this is a reasonable specification.
-
-[^pig-alternative]: These are repeated-measures data, which would be more conventionally modeled with autocorrelated errors within pigs.
-The \code{lme()} function in the \pkg{nlme} package, for example, is capable of fitting a mixed-model of this form.
-
-The \proglang{Stata} manual fits the mixed model by maximum likelihood (rather than REML), and we duplicate the results reported there using \code{lmer()}:
-```{r pigs-lmer}
-m.p <- lmer(weight ~ week + (1 | id) + (1 | week),
-            data=Pigs, REML=FALSE, # i.e., ML
-            control=lmerControl(optimizer="bobyqa"))
-summary(m.p)
-```
-We opt for the non-default \code{"bobyqa"} optimizer because it provides more numerically stable results for subsequent cross-validation in this example.
-
-We can then cross-validate the model by omitting folds composed of pigs, folds composed of weeks, or folds composed of pig-weeks (which in the \code{Pigs} data set correspond to individual cases, using only the fixed effects):
-```{r pigs-cv}
-cv(m.p, clusterVariables="id")
-
-cv(m.p, clusterVariables="week")
-
-cv(m.p, clusterVariables=c("id", "week"), k=10, seed=8469)
-```
-We can also cross-validate the individual cases taking account of the random effects (employing the same 10 folds):
-```{r pigs-cv-cases}
-cv(m.p, k=10, seed=8469)
-```
-Because these predictions are based on BLUPs, they are more accurate than the predictions based only on fixed effects.[^crossed-vs-nested] As well, the difference between the MSE computed for the model fit to the full data and the CV estimates of the MSE is greater here than for cluster-based predictions.
-
-[^crossed-vs-nested]: Even though there is only one observation per combination of pigs and weeks, we can use the BLUP for the omitted case because of the crossed structure of the random effects; that is each pig-week has a pig random effect and a week random effect.
-Although it probably isn't sensible, we can imagine a mixed model for the pig data that employs nested random effects, which would be specified by \code{lmer(weight ~ week + (1 | id/week), data=Pigs)}---that is, a random intercept that varies by combinations of \code{id} (pig) and \code{week}.
-This model can't be fit, however: With only one case per combination of \code{id} and \code{week}, the nested random-effect variance is indistinguishable from the case-level variance.
-
-# Replicating cross-validation
-
-Assuming that the number of cases $n$ is a multiple of the number of folds $k$---a slightly simplifying assumption---the number of possible partitions of cases into folds is $\frac{n!}{[(n/k)!]^k}$, a number that grows very large very quickly.
-For example, for $n = 10$ and $k = 5$, so that the folds are each of size $n/k = 2$, there are $113,400$ possible partitions; for $n=100$ and $k=5$, where $n/k = 20$, still a small problem, the number of possible partitions is truly astronomical, $1.09\times 10^{66}$.
-
-Because the partition into folds that's employed is selected randomly, the resulting CV criterion estimates are subject to sampling error.
-(An exception is LOO cross-validation, which is not at all random.) To get a sense of the magnitude of the sampling error, we can repeat the CV procedure with different randomly selected partitions into folds.
-All of the CV functions in the \pkg{cv} package are capable of repeated cross-validation, with the number of repetitions controlled by the \code{reps} argument, which defaults to \code{1}.
-
-Here, for example, is 10-fold CV for the Mroz logistic regression, repeated 5 times:
-
-```{r mroz-reps}
-cv(m.mroz, criterion=BayesRule, seed=248, reps=5, 
-   method="Woodbury")
-```
-When \code{reps} > \code{1}, the result returned by \code{cv()} is an object of class \code{"cvList"}---literally a list of \code{"cv"} objects.
-The results are reported for each repetition and then averaged across repetitions, with the standard deviations of the CV criterion and the biased-adjusted CV criterion given in parentheses.
-In this example, there is therefore little variation across repetitions, increasing our confidence in the reliability of the results.
-
-Notice that the seed that's set in the \code{cv()} command pertains to the first repetition and the seeds for the remaining repetitions are then selected pseudo-randomly.[^replicates] Setting the first seed, however, makes the entire process easily replicable, and the seed for each repetition is stored in the corresponding element of the \code{"cvList"} object (which isn't, however, saved in the example).
-
-[^replicates]: Because of the manner in which the computation is performed, the order of the replicates in the \code{"cvList"} object returned by \code{cv()} isn't the same as the order in which the replicates are computed.
-Each element of the result, however, is a \code{"cv"} object with the correct random-number seed saved, and so this technical detail can be safely ignored.
-The individual \code{"cv"} objects are printed in the order in which they are stored rather than the order in which they are computed.
-
-It's also possible to replicate CV when comparing competing models via the \code{cv()} method for \code{"modList"} objects.
-Recall our comparison of polynomial regressions of varying degree fit to the \code{Auto} data; we performed 10-fold CV for each of 10 models.
-Here, we replicate that process 5 times for each model and graph the results:
-```{r model-comparison-with-reps}
-#| out.width = "100%",
-#| fig.height = 5,
-#| fig.cap = " Replicated cross-validated 10-fold CV as a function of polynomial degree, $p$"
-cv.auto.reps <- cv(models(m.1, m.2, m.3, m.4, m.5,
-                        m.6, m.7, m.8, m.9, m.10),
-                 data=Auto, seed=8004, reps=5)
-plot(cv.auto.reps)
-```
-The graph shows both the average CV criterion and its range for each of the competing models.
-
-# Cross-validating model selection
-
 ## A preliminary example
@@
 ```
-## Mroz's logistic regression revisited
-
-For a contrasting example we apply model selection to Mroz's logistic regression for married women's labor-force participation.
-First, recall the logistic regression model that we fit to the \code{Mroz} data:
-```{r recall-Mroz-regression}
-summary(m.mroz)
-```
-
-Applying stepwise model selection Mroz's logistic regression, using BIC as the model-selection criterion (via the argument \code{k=log(nrow(Mroz))} to \code{stepAIC()}) selects 5 of the 7 original predictors:
-```{r mroz-selection}
-m.mroz.sel <- stepAIC(m.mroz, k=log(nrow(Mroz)),
-                      trace=FALSE)
-summary(m.mroz.sel)
-BayesRule(Mroz$lfp == "yes",
-          predict(m.mroz.sel, type="response"))
-```
-Bayes rule applied to the selected model misclassifies 32% of the cases in the \code{Mroz} data.
-
-Cross-validating the selected model produces a similar, slightly larger, estimate of misclassification, about 33%:
-```{r cv-mroz-regression}
-cv(m.mroz.sel, criterion=BayesRule, seed=345266)
-```
-Is this estimate of predictive performance optimistic?
-
-We proceed to apply the model-selection procedure by cross-validation, producing more or less the same result:
-```{r cv-mroz-selection}
-m.mroz.sel.cv <- cvSelect(selectStepAIC, Mroz, 
-                          seed=6681,
-                          criterion=BayesRule,
-                          model=m.mroz,
-                          AIC=FALSE)
-m.mroz.sel.cv
-```
-Setting \code{AIC=FALSE} in the call to \code{cvSelect()} uses the BIC rather than the AIC as the model-selection criterion.
-As it turns out, exactly the same predictors are selected when each of the 10 folds are omitted, and the several coefficient estimates are very similar, as we show using \code{compareFolds()}:
-```{r compare-selected-models-mroz}
-compareFolds(m.mroz.sel.cv)
-```
-In this example, therefore, we appear to obtain a realistic estimate of model performance directly from the selected model, because there is little added uncertainty induced by model selection.
 
@@
 
-# Parallel computations
-
-The CV functions in the \pkg{cv} package are all capable of performing parallel computations by setting the \code{ncores} argument (specifying the number of computer cores to be used) to a number > \code{1} (which is the default).
-Parallel computation can be advantageous for large problems, reducing the execution time of the program.
-
-To illustrate, let's time model selection in Mroz's logistic regression, repeating the computation as performed previously and then doing it in parallel using 2 cores:
-```{r parallel-computation}
-system.time(m.mroz.sel.cv <- cvSelect(selectStepAIC, Mroz,
-                          seed=6681,
-                          criterion=BayesRule,
-                          model=m.mroz,
-                          AIC=FALSE))
-
-system.time(m.mroz.sel.cv.p <- cvSelect(selectStepAIC, Mroz,
-                          seed=6681,
-                          criterion=BayesRule,
-                          model=m.mroz,
-                          AIC=FALSE,
-                          ncores=2))
-all.equal(m.mroz.sel.cv, m.mroz.sel.cv.p)
-```
-
-In this small problem, the parallel computation is actually *slower*, because there is an overhead cost to parallelization, but we can see that it produces the same result as before.
-
 # Extending the cv package
@@
 In this vignette, we discuss three kinds of extensions, ordered by increasing general complexity: (1) adding a cross-validation cost criterion; (2) adding a model class that's not directly accommodated by the \code{cv()} default method or by another directly inherited method, with separate consideration of mixed-effects models; and (3) adding a new model-selection procedure suitable for use with \code{selectModel()}.
-
-## Adding a cost criterion
-
-A cost criterion suitable for use with \code{cv()} or \code{cvSelect()} should take two arguments, \code{y} (the observed response vector) and \code{yhat} (a vector of fitted or predicted response values), and return a numeric index of lack of fit.
-The \pkg{cv} package supplies several such criteria: \code{mse(y, yhat)}, which returns the mean-squared prediction error for a numeric response; \code{rmse(y, yhat)}, which returns the (square-)root mean-squared error; \code{medAbsErr(y, yhat)}, which returns the median absolute error; and \code{BayesRule(y, yhat)} (and its non-error-checking version, \code{BayesRule2(y, yhat))}, suitable for use with a binary regression model, where \code{y` is the binary response coded \code{0} for a "failure" or \code{1} for a "success"; where \code{yhat} is the predicted probability of success; and where the proportion of *incorrectly* classified cases is returned.
-
-To illustrate using a different prediction cost criterion, we'll base a cost criterion on the area under the receiver operating characteristic ("ROC") curve for a logistic regression.
-The ROC curve is a graphical representation of the classification power of a binary regression model, and the area under the ROC curve ("AUC"), which varies from 0 to 1, is a common summary measure based on the ROC [see @Wikipedia-ROC:2023].
-The \pkg{Metrics} package [@HamnerFrasco:2018] includes a variety of measures useful for model selection, including an \code{auc()} function.
-We convert the AUC into a cost measure by taking its complement:
-
-```{r AUCcomp}
-AUCcomp <- function(y, yhat) 1 - Metrics::auc(y, yhat)
-```
-
-We then apply \code{AUCcomp()} to the the Mroz logistic regression discussed in the main \pkg{cv} package vignette, which we reproduce here, using the \code{Mroz} data frame from the \pkg{carData} package [@FoxWeisberg:2019]:
-```{r Mroz-regression}
-data("Mroz", package="carData")
-m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
-summary(m.mroz)
-
-AUCcomp(with(Mroz, as.numeric(lfp == "yes")), fitted(m.mroz))
-```
-Cross-validating this cost measure is straightforward:
-```{r Mroz-CV-ROC}
-library("cv")
-cv(m.mroz, criterion=AUCcomp, seed=3639)
-```
-As expected, the cross-validated complement to the AUC is somewhat less optimistic than the criterion computed from the model fit to the whole data set.
-
-As we explain in the vignette "Cross-validation of regression models," the \code{cv()} function differentiates between CV criteria that are averages of casewise components and criteria that are not.
-Computation of bias corrections and confidence intervals is limited to the former.
-We show in the appendix to this vignette that the AUC, and hence its complement, cannot be expressed as averages of casewise components.
-
-\code{cv()} looks for a \code{"casewise loss"} attribute of the value returned by a CV criterion function.
-If this attribute exists, then the criterion is treated as the mean of casewise components, and \code{cv()} uses the unexported function \code{getLossFn()} to construct a function that returns the casewise components of the criterion.
-
-We illustrate with the \code{mse()}:
-```{r}
-mse
-
-cv:::getLossFn(mse(rnorm(100), rnorm(100)))
-```
-For this scheme to work, the "casewise loss" attribute must be a character string (or vector of character strings), here \code{"(y - yhat)\^2"}, that evaluates to an expression that is a function of \code{y} and \code{yhat}, and that computes the vector of casewise components of the CV criterion.
 
@@
 The cross-validated polytomous Bayes-rule criterion confirms that the fitted model does substantially better than the marginal Bayes-rule prediction that everyone votes for Labour.
-
-### Mixed-effects models
-
-Adding a \code{cv()} method for a mixed-model class is somewhat more complicated.
-We provide the \code{cvMixed()} function to facilitate this process, and to see how that works, consider the \code{"lme"} method from the \pkg{cv} package:
-```{r cv.lme}
-cv:::cv.lme
-```
-Notice that \code{cv.lme()} sets up a call to \code{cvMixed()}, which does the computational work.
-
-Most of the arguments of \code{cvMixed()} are familiar:
-
-* \code{model} is the mixed-model object, here of class \code{"lme"}.
-
-* \code{package} is the name of the package in which the mixed-modeling function used to fit the model, here  \code{lme()}, resides---i.e., \code{"nlme"}; \code{cvMixed()} uses this argument to retrieve the package namespace.
-
-* \code{data} is the data set to which the model is fit, by default extracted by the \code{get_data()} function in the \pkg{insight} package.
-
-* \code{criterion} is the CV criterion, defaulting to the \code{mse()} function.
-
-* \code{k} is the number of CV folds, defaulting to \code{"loo"} for CV by clusters and \code{10} for CV by cases.
-
-* \code{reps} is the number of times the CV process is repeated, defaulting to \code{1}.
-
-* \code{seed} is the seed for \proglang{R}'s random-number generator, defaulting to a randomly selected (and saved) value.
-
-* \code{ncores} is the number of cores to use for parallel computation; if \code{1}, the default, then the computation isn't parallelized.
-
-* \code{clusterVariables} is a character vector of the names of variables defining clusters; if missing, then CV is based on cases rather than clusters.
-
-The remaining two arguments are unfamiliar:
-
-* \code{predict.clusters.args} is a named list of arguments to be passed to the \code{predict()} function to obtain predictions for the full data set from a model fit to a subset of the data for cluster-based CV.
-The first two arguments should be \code{object} and \code{newdata}.
-It is typically necessary to tell \code{cvMixed()} how to base predictions only on fixed effects; in the case of \code{"lme"} models, this is done by setting \code{level = 0}.
-
-* Similarly, \code{predict.cases.args} is a named list of arguments to be passed to \code{predict()} for case-based CV. Setting \code{level = 1} includes random effects in the predictions.
-
-Finally, any additional arguments, absorbed by \code{...}, are passed to \code{update()} when the model is refit with each fold omitted. \code{cvMixed()} returns an object of class \code{"cv"}.
-
-Now imagine that we want to support a new class of mixed-effects models.
-To be concrete, we illustrate with the \code{glmmPQL()} function in the \pkg{MASS} package [@VenablesRipley:2002], which fits generalized-linear mixed-effects models by penalized quasi-likelihood.[^glmmPQL] Not coincidentally, the arguments of \code{glmmPQL()} are similar to those of \code{lme()} (with an additional \code{family} argument), because the former iteratively invokes the latter; so \code{cv.glmmPQL()} should resemble \code{cv.lme()}.
-
-[^glmmPQL]: This example is somewhat artificial in that \code{glmmPQL()} has largely been superseded by computationally superior functions, such the \code{glmer()} function in the \pkg{lme4} package.
-There is, however, one situation in which \code{glmmPQL()} might prove useful: to specify serial dependency in case-level errors within clusters for longitudinal data, which is not currently supported by \code{glmer()}.
-
-As it turns out, neither the default method for \code{GetResponse()} nor \code{insight::get_data()} work for \code{"glmmPQL"} objects.
-These objects include a \code{"data"} element, however, and so we can simply extract this element as the default for the \code{data} argument of our \code{cv.glmmPQL()} method.
-
-To get the response variable is more complicated: We refit the fixed part of the model as a GLM with only the regression constant on the right-hand side, and extract the response from that; because all we need is the response variable, we limit the number of GLM iterations to 1 and suppress warning messages about non-convergence:
-
-```{r GetResponse.glmmPQL}
-GetResponse.glmmPQL <- function(model, ...){
-  f <- formula(model)
-  f[[3]] <- 1 # regression constant only on RHS
-  model <- suppressWarnings(glm(f, data=model$data, family=model$family,
-                                control=list(maxit=1)))
-  cv::GetResponse(model)
-}
-```
-
-Writing the \code{cv()} method is then straightforward:
-```{r cv.glmmPQL}
-cv.glmmPQL <- function(model, data = model$data, criterion = mse,
-                     k, reps = 1, seed, ncores = 1, clusterVariables, ...){
-  cvMixed(
-    model,
-    package="MASS",
-    data=data,
-    criterion=criterion,
-    k=k,
-    reps=reps,
-    seed=seed,
-    ncores=ncores,
-    clusterVariables=clusterVariables,
-    predict.clusters.args=list(object=model,
-                               newdata=data,
-                               level=0,
-                               type="response"),
-    predict.cases.args=list(object=model,
-                            newdata=data,
-                            level=1,
-                            type="response"),
-    verbose=FALSE,
-    ...)
-}
-```
-We set the argument \code{verbose=FALSE} to suppress \code{glmmPQL()}'s iteration counter when \code{cvMixed()} calls \code{update()}.
-
-Let's apply our newly minted method to a logistic regression with a random intercept in an example that appears in \code{?glmmPQL}:
-```{r glmmPQL-example}
-library("MASS")
-m.pql <- glmmPQL(y ~ trt + I(week > 2), random = ~ 1 | ID,
-             family = binomial, data = bacteria)
-summary(m.pql)
-```
-We compare this result to that obtained from \code{glmer()} in the \pkg{lme4} package:
-```{r compare-to-lme4}
-library("lme4")
-m.glmer <- glmer(y ~ trt + I(week > 2) + (1 | ID),
-               family = binomial, data = bacteria)
-summary(m.glmer)
-
-  # comparison of fixed effects:
-car::compareCoefs(m.pql, m.glmer) 
-```
-The two sets of estimates are similar, but not identical
-
-Finally, we try out our \code{cv.glmmPQL()} method, cross-validating both by clusters and by cases,
-```{r cv-example-glmmPQL, cache=TRUE}
-cv(m.pql, clusterVariables="ID", criterion=BayesRule)
-
-cv(m.pql, data=bacteria, criterion=BayesRule, seed=1490)
-```
-and again compare to \code{glmer()}:
-```{r cv-example-glmer, cache=TRUE}
-cv(m.glmer, clusterVariables="ID", criterion=BayesRule)
-
-cv(m.glmer, data=bacteria, criterion=BayesRule, seed=1490)
-```
-
-## Adding a model-selection procedure
-
-The \code{selectStepAIC()} function supplied by the \pkg{cv} package, which is based on the \code{stepAIC()} function from the \pkg{nnet} package [@VenablesRipley:2002] for stepwise model selection, is suitable for the \code{procedure} argument of \code{cvSelect()}.
-The use of \code{selectStepAIC()} is illustrated in the principal vignette for the package.
-
-We'll employ \code{selectStepAIC()} as a "template" for writing a CV model-selection procedure.
-To see the code for this function, type \code{cv::selectStepAIC} at the \proglang{R} command prompt, or examine the sources for the \pkg{cv} package at https://github.com/gmonette/cv (the code for \code{selectStepAIC()} is in https://github.com/gmonette/cv/blob/main/R/cvSelect.R).
-
-Another approach to model selection is all-subsets regression.
-The \code{regsubsets()} function in the \pkg{leaps} package [@LumleyMiller:2020] implements an efficient algorithm for selecting the best-fitting linear least-squares regressions for subsets of predictors of all sizes, from 1 through the maximum number of candidate predictors.[^1] To illustrate the use of \code{regsubsets()}, we employ the \code{swiss} data frame supplied by the \pkg{leaps} package:
-
-[^1]: The \code{regsubsets()} function computes several measures of model predictive performance, including the $R^2$ and $R^2$ adjusted for degrees of freedom, the residual sums of squares, Mallows's $C_p$, and the BIC.
-Several of these are suitable for comparing models with differing numbers of coefficients---we use the BIC below---but all necessarily agree when comparing models with the *same* number of coefficients.
-
-```{r swiss}
-library("leaps")
-head(swiss)
-nrow(swiss)
-```
-The data set includes the following variables, for each of 47 French-speaking Swiss provinces circa 1888:
-
-* \code{Fertility}: A standardized fertility measure.
-* \code{Agriculture}: The percentage of the male population engaged in agriculture.
-* \code{Examination}: The percentage of draftees into the Swiss army receiving the highest grade on an examination.
-* \code{Education}: The percentage of draftees with more than a primary-school education.
-* \code{Catholic}: The percentage of the population who were Catholic.
-* \code{Infant.Mortality}: The infant-mortality rate, expressed as the percentage of live births surviving less than a year.
-
-Following @LumleyMiller:2020, we treat \code{Fertility} as the response and the other variables as predictors in a linear least-squares regression:
-```{r swiss-lm}
-m.swiss <- lm(Fertility ~ ., data=swiss)
-summary(m.swiss)
-
-cv(m.swiss, seed=8433)
-```
-Thus, the RMSE for the model fit to the complete data is considerably smaller than the CV estimate of the RMSE.
-Can we do better by selecting a subset of the predictors, taking account of the additional uncertainty induced by model selection?
-
-First, let's apply best-subset selection to the complete data set:
-```{r subset-selection}
-#| fig.cap = "Selecting the best model of each size."
-swiss.sub <- regsubsets(Fertility ~ ., data=swiss)
-summary(swiss.sub)
-
-(bics <- summary(swiss.sub)$bic)
-which.min(bics)
-
-car::subsets(swiss.sub, legend="topright")
-```
-The graph, produced by the \code{subsets()} function in the \pkg{car} package, shows that the model with the smallest BIC is the best model with 4 predictors, including \code{Agriculture}, \code{Education}, \code{Catholic}, and \code{Infant.Mortality}, but not \code{Examination}:
-```{r best-model}
-m.best <- update(m.swiss, . ~ . - Examination)
-summary(m.best)
-
-cv(m.best, seed=8433) # use same folds as before
-```
-The RMSE for the selected model is (of course) slightly higher than for the full model fit previously, but the cross-validated RMSE is a bit lower; as we explain in the main vignette, however, it isn't kosher to select and cross-validate a model on the same data.
-
-Here's a function named \code{selectSubsets()}, meant to be used with \code{cvSelect()}, suitable for cross-validating the model-selection process:
-```{r selectSubsets}
-selectSubsets <- function(data=insight::get_data(model), 
-                          model,
-                          indices,
-                          criterion=mse,
-                          save.coef=TRUE, ...){
-
-  if (inherits(model, "lm", which=TRUE) != 1)
-    stop("selectSubsets is appropriate only for 'lm' models")
-
-  y <- GetResponse(model)
-  formula <- formula(model)
-  X <- model.matrix(model)
-
-  if (missing(indices)) {
-    # select the best model from the full data by BIC
-    sel <- leaps::regsubsets(formula, data=data, ...)
-    bics <- summary(sel)$bic
-    best <- coef(sel, 1:length(bics))[[which.min(bics)]]
-    x.names <- names(best)
-    # fit the best model; intercept is already in X, hence - 1:
-    m.best <- lm(y ~ X[, x.names] - 1) 
-    fit.all <- predict(m.best, newdata=data)
-    return(criterion(y, fit.all)) # return the CV criterion
-  }
-
-  # select the best model omitting the i-th fold (given by indices)
-  sel.i <- leaps::regsubsets(formula, data[-indices, ], ...)
-  bics.i <- summary(sel.i)$bic
-  best.i <- coef(sel.i, 1:length(bics.i))[[which.min(bics.i)]]
-  x.names.i <- names(best.i)
-  m.best.i <- lm(y[-indices] ~ X[-indices, x.names.i] - 1)
-              # predict() doesn't work here:
-  fit.all.i <- as.vector(X[, x.names.i] %*% coef(m.best.i))
-  fit.i <- fit.all.i[indices]
-  # return the fitted values for i-th fold, CV criterion for all cases, 
-  #   and the regression coefficients
-  list(fit.i=fit.i, # fitted values for i-th fold
-       crit.all.i=criterion(y, fit.all.i), # CV crit for all cases
-       coefficients = if (save.coef){ # regression coefficients
-         coefs <- coef(m.best.i)
-
-         # fix coefficient names
-         names(coefs) <- sub("X\\[-indices, x.names.i\\]", "",
-                             names(coefs))
-
-         coefs
-       }  else {
-         NULL
-       }
-  )
-}
-```
-A slightly tricky point is that because of scoping issues, \code{predict()} doesn't work with the model fit omitting the $i$th fold, and so the fitted values for all cases are computed directly as $\widehat{\mathbf{y}}_{-i} = \mathbf{X} \mathbf{b}_{-i}$, where $\mathbf{X}$ is the model-matrix for all of the cases, and $\mathbf{b}_{-i}$ is the vector of least-squares coefficients for the selected model with the $i$th fold omitted.
-
-Additionally, the command \code{lm(y[-indices] ~ X[-indices, x.names.i] - 1)}, which is the selected model with the $i$th fold deleted, produces awkward coefficient names like \code{"X[-indices, x.names.i]Infant.Mortality"}.
-Purely for aesthetic reasons, the command `sub("X\\[-indices, x.names.i\\]", "", names(coefs))` fixes these awkward names, removing the extraneous text, \code{"X[-indices, x.names.i]"}.
-
-Applying \code{selectSubsets()} to the full data produces the full-data cross-validated RMSE (which we obtained previously):
-%%```{r test-selectSubsets}
-%%selectSubsets(model=m.swiss)
-%%```
-Similarly, applying the function to an imaginary "fold" of 5 cases returns the RMSE for the cases in the fold, based on the model selected and fit to the cases omitting the fold; the RMSE for all of the cases, based on the same model; and the coefficients of the selected model, which includes 4 or the 5 predictors (and the intercept):
-%%```{r test-selectSubsets-fold}
-%%selectSubsets(model=m.swiss, indices=seq(5, 45, by=10))
-%%```
-
-Then, using \code{selectSubsets()} in cross-validation, we get:
-```{r cvSelect-selectSubsets}
-(cv.swiss <- cvSelect(selectSubsets, model=m.swiss,
-                      data=swiss, seed=8433)) # use same folds
-```
-Cross-validation shows that model selection exacts a penalty in RMSE.
-Examining the models selected for the 10 folds reveals that there is some uncertainty in identifying the predictors in the "best" model, with \code{Agriculture} sometimes appearing and sometimes not:
-```{r best-models-by-folds}
-compareFolds(cv.swiss)
-```
 
@@
 At present we use the confidence interval above for sufficiently large $n$, which, based on Bates et al.'s results, we take by default to be $n \ge 400$.
-
-# Why the complement of AUC isn't a casewise CV criterion
-
-Consider calculating AUC for folds in which a validation set contains $n_v$ observations.
-To calculate AUC in the validation set, we need the vector of prediction criteria, $\widehat{\mathbf{y}}_{v_{(n_v \times 1)}} = (\widehat{y}_1, ..., \widehat{y}_{n_v})^T$, and the vector of observed responses in the validation set, $\mathbf{y}_{v_{(n_v \times 1)}} = (y_1, \ldots, y_{n_v})^T$ with $y_i \in \{0,1\}, \; i = 1, \ldots, n_v$.
-
-To construct the ROC curve, only the ordering of the values in 
-$\mathbf{\widehat{y}}_v$ is relevant. Thus, assuming that there are no ties, and reordering observations if necessary, we can set 
-$\mathbf{\widehat{y}}_v = (1, 2, \ldots, n_v)^T$.
-
-If the AUC can be expressed as the casewise mean or sum of a function $\mathrm{cv}(\widehat{y}_i,y_i)$, where $\mathrm{cv}: \{1,2,...,n_v\}\times\{0,1\} \rightarrow [0,1]$, then
-\begin{equation}
-\label{eq:cw}
-\tag{1}
-\sum_{i=1}^{n_v} \mathrm{cv}(\widehat{y}_i,y_i) = \mathrm{AUC}(\mathbf{\widehat{y}}_v,\mathbf{y}_v)
-\end{equation}
-must hold for all $2^{n_v}$ possible values of $\mathbf{y}_v = (y_1,...,y_{n_v})^T$.
-If all $y\mathrm{s}$ have the same value, either 1 or 0, then the definition of AUC is ambiguous.
-AUC could be considered undefined, or it could be set to 0 if all $y$s are 0 and to 1 if all $y$s are 1.
-If AUC is considered to be undefined in these cases, we have $2^{n_v} - 2$ admissible values for $\mathbf{y}_v$.
-
-Thus, equation (\ref{eq:cw}) produces either $2^{n_v}$ or $2^{n_v}-2$ constraints.
-Although there are only $2n_v$ possible values for the $\mathrm{cv(\cdot)}$ function, equation (\ref{eq:cw}) could, nevertheless, have consistent solutions.
-We therefore need to determine whether there is a value of $n_v$ for which (\ref{eq:cw}) has no consistent solution for all admissible values of $\mathbf{y}_v$.
-In that eventuality, we will have shown that AUC cannot, in general, be expressed through a casewise sum.
-
-If $n_v=3$, we show below that (\ref{eq:cw}) has no consistent solution if we include all possibilities for $\mathbf{y}_v$, but does if we exclude cases where all $y$s have the same value.
-If $n_v=4$, we show that there are no consistent solutions in either case.
-
-The following \proglang{R} function computes AUC from $\mathbf{\widehat{y}}_v$ and $\mathbf{y}_v$, accommodating the cases where $\mathbf{y}_v$ is all 0s or all 1s:
-
-```{r}
-AUC <- function(y, yhat = seq_along(y)) {
-  s <- sum(y)
-  if (s == 0) return(0)
-  if (s == length(y)) return(1)
-  Metrics::auc(y, yhat)
-}
-```
-
-We then define a function to generate all possible $\mathbf{y}_v$s of length $n_v$ as rows of the matrix $\mathbf{Y}_{(2^{n_v} \times n_v)}$:
-
-```{r}
-Ymat <- function(n_v, exclude_identical = FALSE) {
-  stopifnot(n_v > 0 && round(n_v) == n_v)    # n_v must be a positive integer
-  ret <- sapply(0:(2^n_v - 1),
-                function(x) as.integer(intToBits(x)) )[1:n_v, ]
-  ret <- if (is.matrix(ret)) t(ret) else matrix(ret)
-  colnames(ret) <- paste0("y", 1:ncol(ret))
-  if (exclude_identical) ret[-c(1, nrow(ret)), ] else ret
-}
-```
-
-For $n_v=3$,
-
-```{r}
-Ymat(3)
-```
-
-If we exclude $\mathbf{y}_v$s with identical values, then
-
-```{r}
-Ymat(3, exclude_identical = TRUE)
-```
-
-Here is $\mathbf{Y}$ with corresponding values of AUC:
-
-```{r}
-cbind(Ymat(3), AUC = apply(Ymat(3), 1, AUC))
-```
-
-The values of $\mathrm{cv}(\widehat{y}_i, y_i)$ that express AUC as a sum of casewise values are solutions of equation (\ref{eq:cw}), which can be written as solutions of the following system of $2^{n_v}$ linear simultaneous equations in $2n_v$ unknowns:
-\begin{equation}
-\label{eq:lin}
-\tag{2}
-(\mathbf{U} -\mathbf{Y}) \mathbf{c}_0 + \mathbf{Y} \mathbf{c}_1
-=
-[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]
-\begin{bmatrix}
-\mathbf{c}_0 \\ \mathbf{c}_1
-\end{bmatrix}
-= \mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})
-\end{equation}
-where $\mathbf{U}_{(2^{n_v} \times n_v)}$ is a matrix of 1s conformable with $\mathbf{Y}$; $\mathbf{c}_0 = [\mathrm{cv}(1,0), c(2,0), ..., \mathrm{cv}(n_v,0)]^T$; $\mathbf{c}_1 = [\mathrm{cv}(1,1), c(2,1), ..., \mathrm{cv}(n_v,1)]^T$; $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]_{(2^{n_v} \times 2n_v)}$ and
-$\begin{bmatrix}\begin{aligned}
-\mathbf{c}_0 \\ \mathbf{c}_1
-\end{aligned}
-\end{bmatrix}_{(2n_v \times 1)}$ 
-are partitioned matrices; and $\mathbf{\widehat{Y}}_{(2^{n_v} \times n_v)}$ is a matrix each of whose rows consists of the integers 1 to $n_v$.
-
-We can test whether equation (\ref{eq:lin}) has a solution for any given $n_v$ by trying to solve it as a least-squares problem, considering whether the residuals of the associated linear model are all 0, using the "design matrix" $[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]$ to predict the "outcome"
-$\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})_{(2^{n_v} \times 1)}$:
-
-```{r}
-resids <- function(n_v, exclude_identical = FALSE, 
-                   tol = sqrt(.Machine$double.eps)) {
-  Y <- Ymat(n_v, exclude_identical = exclude_identical)
-  AUC <- apply(Y, 1, AUC)
-  X <- cbind(1-Y, Y)
-  opts <- options(warn = -1)
-  on.exit(options(opts))
-  fit <- lsfit(X, AUC, intercept = FALSE)
-  ret <- max(abs(residuals(fit)))
-  if(ret < tol){
-    ret <- 0
-    solution <- coef(fit)
-    names(solution) <- paste0("c(", c(1:n_v, 1:n_v), ",", 
-                              rep(0:1, each = n_v), ")")
-    attr(ret, "solution") <- zapsmall(solution)
-  }
-  ret
-}
-```
-
-The case $n_v=3$, excluding identical $y$s, has a solution:
-
-```{r}
-resids(3, exclude_identical = TRUE)
-```
-
-But, if identical $y$s are included, the equation is not consistent:
-
-```{r}
-resids(3, exclude_identical = FALSE)
-```
-
-For $n_v=4$, there are no solutions in either case:
-
-```{r}
-resids(4, exclude_identical = TRUE)
-resids(4, exclude_identical = FALSE)
-```
-
-Consequently, the widely employed AUC measure of fit for binary regression cannot in general be used for a casewise cross-validation criterion.
-
-<!-- # Introduction -->
-
-<!-- This template demonstrates some of the basic LaTeX that you need to know to create a JSS article. -->
-
-<!-- ## Code formatting -->
-
-<!-- In general, don't use Markdown, but use the more precise LaTeX commands instead: -->
-
-<!-- * \proglang{Java} -->
-<!-- * \pkg{plyr} -->
-
-<!-- One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax). -->
-
-<!-- If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute.
-You can also provide a custom identifier if necessary.
-See the header of Section \ref{r-code} for example.
--->
-
-<!-- # \proglang{R} code {short-title="R code" #r-code} -->
-
-<!-- Can be inserted in regular R markdown blocks. -->
-
-<!-- ```{r} -->
-<!-- x <- 1:10 -->
-<!-- x -->
-<!-- ``` -->
-
-<!-- ## Features specific to \pkg{rticles} {short-title="Features specific to rticles"} -->
-
-<!-- * Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter).
-This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
--->
-<!-- * Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format. -->

### 2024-01-03 scratch/notes_john/JSS-article/JSS-article-reduced.tex
added JSS-article-reduced.Rmd (preliminary attempt)

--- 2024-01-02 scratch/notes_john/JSS-article/JSS-article.tex
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/mpg-horsepower-scatterplot-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-scatterplot-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/mpg-horsepower-MSE-se-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-MSE-se-1}
 
@@
 \begin{CodeOutput}
-R RNG seed set to 44768
+R RNG seed set to 423836
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.52622
-bias-adjusted cross-validation criterion = 19.49747
+cross-validation criterion = 19.26863
+bias-adjusted cross-validation criterion = 19.25351
 full-sample criterion = 18.98477 
@@
       expr        min         lq       mean     median         uq        max
- hatvalues    977.973    993.43   1143.433   1173.113   1202.981   1296.256
-  Woodbury   9932.004  10178.21  11461.993  10363.201  10865.943  17762.266
-     naive 215384.152 217809.02 224546.766 219116.505 221670.231 275518.278
-    cv.glm 384357.616 388159.59 409612.546 390652.100 441741.913 443226.605
+ hatvalues    984.287   1153.412   1160.394   1189.902   1199.414   1285.104
+  Woodbury  10145.122  10213.592  10525.532  10463.385  10657.581  11476.351
+     naive 216360.403 217763.218 223882.308 218184.226 219846.551 273572.951
+    cv.glm 380361.674 382182.689 400581.980 386284.739 436401.540 439866.368
  neval cld
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/polynomial-regression-CV-graph-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-1}
 
@@
 
-{\centering \includegraphics[width=0.45\linewidth]{JSS-article_files/figure-latex/polynomial-regression-CV-graph-2-1} \includegraphics[width=0.45\linewidth]{JSS-article_files/figure-latex/polynomial-regression-CV-graph-2-2}
+{\centering \includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-2-1} \includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-2-2}
 
@@
 clearly suggest over-fitting).
-
-\hypertarget{logistic-regression-for-the-mroz-data}{%
-\subsection{Logistic regression for the Mroz
-data}\label{logistic-regression-for-the-mroz-data}}
-
-The \code{Mroz} data set from the \pkg{carData} package
-\citep[associated with][]{FoxWeisberg:2019} has been used by several
-authors to illustrate binary logistic regression; see, in particular
-\citet{FoxWeisberg:2019}. The data were originally drawn from the U.S.
-Panel Study of Income Dynamics and pertain to married women. Here are a
-few cases in the data set:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> data("Mroz", package="carData")
-R> head(Mroz, 3)
-\end{CodeInput}
-\begin{CodeOutput}
-  lfp k5 k618 age wc hc       lwg   inc
-1 yes  1    0  32 no no 1.2101647 10.91
-2 yes  0    2  30 no no 0.3285041 19.50
-3 yes  1    3  35 no no 1.5141279 12.04
-\end{CodeOutput}
-\begin{CodeInput}
-R> tail(Mroz, 3)
-\end{CodeInput}
-\begin{CodeOutput}
-    lfp k5 k618 age wc hc       lwg    inc
-751  no  0    0  43 no no 0.8881401  9.952
-752  no  0    0  60 no no 1.2249736 24.984
-753  no  0    3  39 no no 0.8532125 28.363
-\end{CodeOutput}
-\end{CodeChunk}
-
-The response variable in the logistic regression is \code{lfp},
-labor-force participation, a factor coded \code{"yes"} or \code{"no"}.
-The remaining variables are predictors:
-
-\begin{itemize}
-\tightlist
-\item
-  \code{k5}, number of children 5 years old of younger in the woman's
-  household;
-\item
-  \code{k618}, number of children between 6 and 18 years old;
-\item
-  \code{age}, in years;
-\item
-  \code{wc}, wife's college attendance, \code{"yes"} or \code{"no"};
-\item
-  \code{hc}, husband's college attendance;
-\item
-  \code{lwg}, the woman's log wage rate if she is employed, or her
-  \emph{imputed} wage rate, if she is not \citep[a variable that][ show
-  is problematically defined]{FoxWeisberg:2019}; and
-\item
-  \code{inc}, family income, in \$1000s, exclusive of wife's income.
-\end{itemize}
-
-We use the \code{glm()} function to fit a binary logistic regression to
-the \code{Mroz} data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
-R> summary(m.mroz)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-glm(formula = lfp ~ ., family = binomial, data = Mroz)
-
-Coefficients:
-             Estimate Std. Error z value Pr(>|z|)
-(Intercept)  3.182140   0.644375   4.938 7.88e-07
-k5          -1.462913   0.197001  -7.426 1.12e-13
-k618        -0.064571   0.068001  -0.950 0.342337
-age         -0.062871   0.012783  -4.918 8.73e-07
-wcyes        0.807274   0.229980   3.510 0.000448
-hcyes        0.111734   0.206040   0.542 0.587618
-lwg          0.604693   0.150818   4.009 6.09e-05
-inc         -0.034446   0.008208  -4.196 2.71e-05
-
-(Dispersion parameter for binomial family taken to be 1)
-
-    Null deviance: 1029.75  on 752  degrees of freedom
-Residual deviance:  905.27  on 745  degrees of freedom
-AIC: 921.27
-
-Number of Fisher Scoring iterations: 4
-\end{CodeOutput}
-\begin{CodeInput}
-R> BayesRule(ifelse(Mroz$lfp == "yes", 1, 0), 
-+           fitted(m.mroz, type="response"))
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0.3067729
-attr(,"casewise loss")
-[1] "y != round(yhat)"
-\end{CodeOutput}
-\end{CodeChunk}
-
-In addition to the usually summary output for a GLM, we show the result
-of applying the \code{BayesRule()} function from the \pkg{cv} package to
-predictions derived from the fitted model. Bayes rule, which predicts a
-``success'' in a binary regression model when the fitted probability of
-success {[}i.e., \(\phi = \Pr(y = 1)\){]} is \(\widehat{\phi} \ge .5\)
-and a ``failure'' if \(\widehat{\phi} < .5\).\footnote{\code{BayesRule()}
-  does some error checking; \code{BayesRule2()} is similar, but omits
-  the error checking, and so can be faster for large problems.} The
-first argument to \code{BayesRule()} is the binary \{0, 1\} response,
-and the second argument is the predicted probability of success.
-\code{BayesRule()} returns the proportion of predictions that are
-\emph{in error}, as appropriate for a ``cost'' function.
-
-The value returned by \code{BayesRule()} is associated with an
-``attribute'' named \code{"casewise loss"} and set to
-\code{"y != round(yhat)"}, signifying that the Bayes rule CV criterion
-is computed as the mean of casewise values, here 0 if the prediction for
-a case matches the observed value and 1 if it does not (signifying a
-prediction error). The \code{mse()} function for numeric responses is
-also calculated as a casewise average. Some other criteria, such as the
-median absolute error, computed by the \code{medAbsErr()} function in
-the \pkg{cv} package, aren't averages of casewise components. The
-distinction is important because, to our knowledge, the statistical
-theory of cross-validation, for example, in \citet{DavisonHinkley:1997},
-\citet{BatesHastieTibshirani:2023}, and \citet{ArlotCelisse:2010}, is
-developed for CV criteria like MSE that are means of casewise
-components. As a consequence, we limit computation of bias adjustment
-and confidence intervals (see below) to criteria that are casewise
-averages.
-
-In this example, the fitted logistic regression incorrectly predicts
-31\% of the responses; we expect this estimate to be optimistic given
-that the model is used to ``predict'' the data to which it is fit.
-
-The \code{"glm"} method for \code{cv()} is largely similar to the
-\code{"lm"} method, although the default algorithm, selected explicitly
-by \code{method="exact"}, refits the model with each fold removed (and
-is thus equivalent to \code{method="naive"} for \code{"lm"} models). For
-generalized linear models, \code{method="Woodbury"} or (for LOO CV)
-\code{method="hatvalues"} provide approximate results (see the last
-section of the vignette for details):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(m.mroz, criterion=BayesRule, seed=248)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 248
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-method: exact
-criterion: BayesRule
-cross-validation criterion = 0.3240372
-bias-adjusted cross-validation criterion = 0.3195223
-95% CI for bias-adjusted CV criterion = (0.2860721, 0.3529724)
-full-sample criterion = 0.3067729 
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.mroz, criterion=BayesRule, seed=248, method="Woodbury")
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 248
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-method: Woodbury
-criterion: BayesRule
-cross-validation criterion = 0.3240372
-bias-adjusted cross-validation criterion = 0.3192577
-95% CI for bias-adjusted CV criterion = (0.2858076, 0.3527079)
-full-sample criterion = 0.3067729 
-\end{CodeOutput}
-\end{CodeChunk}
-
-To ensure that the two methods use the same 10 folds, we specify the
-seed for \proglang{R}'s random-number generator explicitly; here, and as
-is common in our experience, the \code{"exact"} and \code{"Woodbury"}
-algorithms produce nearly identical results. The CV estimates of
-prediction error are slightly higher than the estimate based on all of
-the cases.
-
-The printed output includes a 95\% confidence interval for the
-bias-adjusted Bayes rule CV criterion.
-\citet{BatesHastieTibshirani:2023} show that these confidence intervals
-are unreliable for models fit to small samples, and by default
-\code{cv()} computes them only when the sample size is 400 or larger and
-when the CV criterion employed is an average of casewise components, as
-is the case for Bayes rule. See the final section of the vignette for
-details of the computation of confidence intervals for bias-adjusted CV
-criteria.
-
-Here are results of applying LOO CV to the Mroz model, using both the
-exact and the approximate methods:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(m.mroz, k="loo", criterion=BayesRule)
-\end{CodeInput}
-\begin{CodeOutput}
-n-Fold Cross Validation
-method: exact
-criterion: BayesRule
-cross-validation criterion = 0.3200531
-bias-adjusted cross-validation criterion = 0.3183001
-95% CI for bias-adjusted CV criterion = (0.2849584, 0.3516418)
-full-sample criterion = 0.3067729 
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury")
-\end{CodeInput}
-\begin{CodeOutput}
-n-Fold Cross Validation
-method: Woodbury
-criterion: BayesRule
-cross-validation criterion = 0.3200531
-bias-adjusted cross-validation criterion = 0.3183001
-95% CI for bias-adjusted CV criterion = (0.2849584, 0.3516418)
-full-sample criterion = 0.3067729 
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues")
-\end{CodeInput}
-\begin{CodeOutput}
-n-Fold Cross Validation
-method: hatvalues
-criterion: BayesRule
-cross-validation criterion = 0.3200531
-\end{CodeOutput}
-\end{CodeChunk}
-
-To the number of decimal digits shown, the three methods produce
-identical results for this example.
-
-As for linear models, we report some timings for the various
-\textbackslash code\{cv()` methods of computation in LOO CV as well as
-for the \code{cv.glm()} function from the \pkg{boot} package (which,
-recall, refits the model with each case removed, and thus is comparable
-to \code{cv()} with \code{method="exact"}):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> microbenchmark::microbenchmark(
-+   hatvalues=cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues"),
-+   Woodbury=cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury"),
-+   exact=cv(m.mroz, k="loo", criterion=BayesRule),
-+   cv.glm=boot::cv.glm(Mroz, m.mroz,
-+                cost=BayesRule),
-+   times=10)
-\end{CodeInput}
-\begin{CodeOutput}
-Unit: milliseconds
-      expr         min          lq        mean      median          uq
- hatvalues    1.316141    1.326842    1.654063    1.407571    1.442462
-  Woodbury   37.127427   38.641639   39.368192   39.356904   40.562202
-     exact 1757.625925 1827.861385 1905.072191 1911.421128 1951.049944
-    cv.glm 2062.693354 2122.649130 2149.761134 2140.079440 2184.788525
-         max neval cld
-    4.022428    10 a  
-   41.060188    10 a  
- 2128.307540    10  b 
- 2299.544573    10   c
-\end{CodeOutput}
-\end{CodeChunk}
-
-There is a substantial time penalty associated with exact computations.
 
@@
 R> library(nlme)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Attaching package: 'nlme'
+\end{CodeOutput}
+\begin{CodeOutput}
+The following object is masked from 'package:lme4':
+
+    lmList
+\end{CodeOutput}
+\begin{CodeOutput}
+The following object is masked from 'package:dplyr':
+
+    collapse
+\end{CodeOutput}
+\begin{CodeInput}
 R> hsb.lme <- lme(mathach ~ mean.ses*cses + sector*cses,
@@
 description above and meant to highlight the issues that arise in
-cross-validating mixed-effects models:\footnote{The \code{regsubsets()}
-  function computes several measures of model predictive performance,
-  including the \(R^2\) and \(R^2\) adjusted for degrees of freedom, the
-  residual sums of squares, Mallows's \(C_p\), and the BIC. Several of
-  these are suitable for comparing models with differing numbers of
-  coefficients---we use the BIC below---but all necessarily agree when
-  comparing models with the \emph{same} number of coefficients.}
+cross-validating mixed-effects models:\footnote{We invite the interested
+  reader to experiment with varying the parameters of our example.}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot1-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot1-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod0-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod0-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod1-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod1-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod2-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod2-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/plot-fits-mod3-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod3-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/cross-validation-clusters-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/cross-validation-clusters-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/cross-validation-cases-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/cross-validation-cases-1}
 
@@
 levels.
-
-\hypertarget{example-crossed-random-effects}{%
-\subsection{Example: Crossed random
-effects}\label{example-crossed-random-effects}}
-
-Crossed random effects arise when the structure of the data aren't
-strictly hierarchical. Nevertheless, crossed and nested random effects
-can be handled in much the same manner, by refitting the mixed-effects
-model to the data with a fold of clusters or cases removed and using the
-refitted model to predict the response in the removed fold.
-
-We'll illustrate with data on pig growth, introduced by \citet[Table
-3.1]{DiggleLiangZeger:1994}. The data are in the \code{Pigs} data frame
-in the \pkg{cv} package:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> head(Pigs, 9)
-\end{CodeInput}
-\begin{CodeOutput}
-  id week weight
-1  1    1   24.0
-2  1    2   32.0
-3  1    3   39.0
-4  1    4   42.5
-5  1    5   48.0
-6  1    6   54.5
-7  1    7   61.0
-8  1    8   65.0
-9  1    9   72.0
-\end{CodeOutput}
-\begin{CodeInput}
-R> head(xtabs(~ id + week, data=Pigs), 3)
-\end{CodeInput}
-\begin{CodeOutput}
-   week
-id  1 2 3 4 5 6 7 8 9
-  1 1 1 1 1 1 1 1 1 1
-  2 1 1 1 1 1 1 1 1 1
-  3 1 1 1 1 1 1 1 1 1
-\end{CodeOutput}
-\begin{CodeInput}
-R> tail(xtabs(~ id + week, data=Pigs), 3)
-\end{CodeInput}
-\begin{CodeOutput}
-    week
-id   1 2 3 4 5 6 7 8 9
-  46 1 1 1 1 1 1 1 1 1
-  47 1 1 1 1 1 1 1 1 1
-  48 1 1 1 1 1 1 1 1 1
-\end{CodeOutput}
-\end{CodeChunk}
-
-Each of 48 pigs is observed weekly over a period of 9 weeks, with the
-weight of the pig recorded in kg. The data are in ``long'' format, as is
-appropriate for use with the \code{lmer()} function in the \pkg{lme4}
-package. The data are very regular, with no missing cases.
-
-The following graph, showing the growth trajectories of the pigs, is
-similar to Figure 3.1 in \citet{DiggleLiangZeger:1994}; we add an
-overall least-squares line and a loess smooth, which are nearly
-indistinguishable:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot(weight ~ week, data=Pigs, type="n")
-R> for (i in unique(Pigs$id)){
-+   with(Pigs, lines(x=1:9, y=Pigs[id == i, "weight"],
-+                    col="gray"))
-+ }
-R> abline(lm(weight ~ week, data=Pigs), col="blue", lwd=2)
-R> lines(with(Pigs, loess.smooth(week, weight, span=0.5)),
-+       col="magenta", lty=2, lwd=2)
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article_files/figure-latex/pigs-graph-1}
-
-}
-
-\caption[Growth trajectories for 48 pigs, with overall least-squares line (sold blue) and loess line (broken magenta)]{Growth trajectories for 48 pigs, with overall least-squares line (sold blue) and loess line (broken magenta).}\label{fig:pigs-graph}
-\end{figure}
-\end{CodeChunk}
-
-The individual ``growth curves'' and the overall trend are generally
-linear, with some tendency for variability of pig weight to increase
-over weeks (a feature of the data that we ignore in the mixed model that
-we fit to the data below).
-
-The \proglang{Stata} mixed-effects models manual proposes a model with
-crossed random effects for the \code{Pigs} data \citep[ page
-37]{Stata:2023}:
-
-\begin{quote}
-{[}S{]}uppose that we wish to fit \[
-\mathrm{weight}_{ij} = \beta_0 + \beta_1 \mathrm{week}_{ij} + u_i + v_j + \varepsilon_{ij} 
-\] for the \(i = 1, \ldots, 9\) weeks and \(j = 1, \dots, 48\) pigs and
-\[
-u_i \sim N(0, \sigma^2_u); v_j \sim N(0, \sigma^2_v ); \varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon)
-\] all independently. That is, we assume an overall population-average
-growth curve \(\beta_0 + \beta_1 \mathrm{week}\) and a random
-pig-specific shift. In other words, the effect due to week, \(u_i\), is
-systematic to that week and common to all pigs. The rationale behind
-{[}this model{]} could be that, assuming that the pigs were measured
-contemporaneously, we might be concerned that week-specific random
-factors such as weather and feeding patterns had significant systematic
-effects on all pigs.
-\end{quote}
-
-Although we might prefer an alternative model,\footnote{These are
-  repeated-measures data, which would be more conventionally modeled
-  with autocorrelated errors within pigs. The \code{lme()} function in
-  the \pkg{nlme} package, for example, is capable of fitting a
-  mixed-model of this form.} we think that this is a reasonable
-specification.
-
-The \proglang{Stata} manual fits the mixed model by maximum likelihood
-(rather than REML), and we duplicate the results reported there using
-\code{lmer()}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> m.p <- lmer(weight ~ week + (1 | id) + (1 | week),
-+             data=Pigs, REML=FALSE, # i.e., ML
-+             control=lmerControl(optimizer="bobyqa"))
-R> summary(m.p)
-\end{CodeInput}
-\begin{CodeOutput}
-Linear mixed model fit by maximum likelihood  ['lmerMod']
-Formula: weight ~ week + (1 | id) + (1 | week)
-   Data: Pigs
-Control: lmerControl(optimizer = "bobyqa")
-
-     AIC      BIC   logLik deviance df.resid 
-  2037.6   2058.0  -1013.8   2027.6      427
-
-Scaled residuals: 
-    Min      1Q  Median      3Q     Max 
--3.7750 -0.5418  0.0054  0.4762  3.9816
-
-Random effects:
- Groups   Name        Variance Std.Dev.
- id       (Intercept) 14.83622 3.8518  
- week     (Intercept)  0.08499 0.2915  
- Residual              4.29733 2.0730  
-Number of obs: 432, groups:  id, 48; week, 9
-
-Fixed effects:
-            Estimate Std. Error t value
-(Intercept) 19.35561    0.63340   30.56
-week         6.20990    0.05393  115.14
-
-Correlation of Fixed Effects:
-     (Intr)
-week -0.426
-\end{CodeOutput}
-\end{CodeChunk}
-
-We opt for the non-default \code{"bobyqa"} optimizer because it provides
-more numerically stable results for subsequent cross-validation in this
-example.
-
-We can then cross-validate the model by omitting folds composed of pigs,
-folds composed of weeks, or folds composed of pig-weeks (which in the
-\code{Pigs} data set correspond to individual cases, using only the
-fixed effects):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(m.p, clusterVariables="id")
-\end{CodeInput}
-\begin{CodeOutput}
-n-Fold Cross Validation based on 48 {id} clusters
-cross-validation criterion = 19.97309
-bias-adjusted cross-validation criterion = 19.96496
-95% CI for bias-adjusted CV criterion = (17.1249, 22.80502)
-full-sample criterion = 19.20076 
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.p, clusterVariables="week")
-\end{CodeInput}
-\begin{CodeOutput}
-boundary (singular) fit: see help('isSingular')
-\end{CodeOutput}
-\begin{CodeOutput}
-n-Fold Cross Validation based on 9 {week} clusters
-cross-validation criterion = 19.31219
-bias-adjusted cross-validation criterion = 19.30499
-95% CI for bias-adjusted CV criterion = (16.56624, 22.04373)
-full-sample criterion = 19.20076 
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.p, clusterVariables=c("id", "week"), k=10, seed=8469)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 8469
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation based on 432 {id, week} clusters
-cross-validation criterion = 19.2352
-bias-adjusted cross-validation criterion = 19.23294
-95% CI for bias-adjusted CV criterion = (16.49263, 21.97325)
-full-sample criterion = 19.20076 
-\end{CodeOutput}
-\end{CodeChunk}
-
-We can also cross-validate the individual cases taking account of the
-random effects (employing the same 10 folds):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(m.p, k=10, seed=8469)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 8469
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-cross-validation criterion = 5.15835
-bias-adjusted cross-validation criterion = 5.072933
-95% CI for bias-adjusted CV criterion = (4.122952, 6.022913)
-full-sample criterion = 3.795972 
-\end{CodeOutput}
-\end{CodeChunk}
-
-Because these predictions are based on BLUPs, they are more accurate
-than the predictions based only on fixed effects.\footnote{Even though
-  there is only one observation per combination of pigs and weeks, we
-  can use the BLUP for the omitted case because of the crossed structure
-  of the random effects; that is each pig-week has a pig random effect
-  and a week random effect. Although it probably isn't sensible, we can
-  imagine a mixed model for the pig data that employs nested random
-  effects, which would be specified by
-  \code{lmer(weight ~ week + (1 | id/week), data=Pigs)}---that is, a
-  random intercept that varies by combinations of \code{id} (pig) and
-  \code{week}. This model can't be fit, however: With only one case per
-  combination of \code{id} and \code{week}, the nested random-effect
-  variance is indistinguishable from the case-level variance.} As well,
-the difference between the MSE computed for the model fit to the full
-data and the CV estimates of the MSE is greater here than for
-cluster-based predictions.
-
-\hypertarget{replicating-cross-validation}{%
-\section{Replicating
-cross-validation}\label{replicating-cross-validation}}
-
-Assuming that the number of cases \(n\) is a multiple of the number of
-folds \(k\)---a slightly simplifying assumption---the number of possible
-partitions of cases into folds is \(\frac{n!}{[(n/k)!]^k}\), a number
-that grows very large very quickly. For example, for \(n = 10\) and
-\(k = 5\), so that the folds are each of size \(n/k = 2\), there are
-\(113,400\) possible partitions; for \(n=100\) and \(k=5\), where
-\(n/k = 20\), still a small problem, the number of possible partitions
-is truly astronomical, \(1.09\times 10^{66}\).
-
-Because the partition into folds that's employed is selected randomly,
-the resulting CV criterion estimates are subject to sampling error. (An
-exception is LOO cross-validation, which is not at all random.) To get a
-sense of the magnitude of the sampling error, we can repeat the CV
-procedure with different randomly selected partitions into folds. All of
-the CV functions in the \pkg{cv} package are capable of repeated
-cross-validation, with the number of repetitions controlled by the
-\code{reps} argument, which defaults to \code{1}.
-
-Here, for example, is 10-fold CV for the Mroz logistic regression,
-repeated 5 times:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(m.mroz, criterion=BayesRule, seed=248, reps=5, 
-+    method="Woodbury")
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 248
-\end{CodeOutput}
-\begin{CodeOutput}
-R RNG seed set to 68134
-\end{CodeOutput}
-\begin{CodeOutput}
-R RNG seed set to 767359
-\end{CodeOutput}
-\begin{CodeOutput}
-R RNG seed set to 556270
-\end{CodeOutput}
-\begin{CodeOutput}
-R RNG seed set to 882966
-\end{CodeOutput}
-\begin{CodeOutput}
-
-Replicate 1:
-10-Fold Cross Validation
-method: Woodbury
-criterion: BayesRule
-cross-validation criterion = 0.3200531
-bias-adjusted cross-validation criterion = 0.3130109
-95% CI for bias-adjusted CV criterion = (0.2796692, 0.3463526)
-full-sample criterion = 0.3067729
-
-Replicate 2:
-10-Fold Cross Validation
-method: Woodbury
-criterion: BayesRule
-cross-validation criterion = 0.3160691
-bias-adjusted cross-validation criterion = 0.311697
-95% CI for bias-adjusted CV criterion = (0.2784665, 0.3449274)
-full-sample criterion = 0.3067729
-
-Replicate 3:
-10-Fold Cross Validation
-method: Woodbury
-criterion: BayesRule
-cross-validation criterion = 0.314741
-bias-adjusted cross-validation criterion = 0.3086212
-95% CI for bias-adjusted CV criterion = (0.2754285, 0.341814)
-full-sample criterion = 0.3067729
-
-Replicate 4:
-10-Fold Cross Validation
-method: Woodbury
-criterion: BayesRule
-cross-validation criterion = 0.3240372
-bias-adjusted cross-validation criterion = 0.318069
-95% CI for bias-adjusted CV criterion = (0.2846189, 0.3515192)
-full-sample criterion = 0.3067729
-
-Replicate 5:
-10-Fold Cross Validation
-method: Woodbury
-criterion: BayesRule
-cross-validation criterion = 0.3240372
-bias-adjusted cross-validation criterion = 0.3192577
-95% CI for bias-adjusted CV criterion = (0.2858076, 0.3527079)
-full-sample criterion = 0.3067729
-
-Average:
-10-Fold Cross Validation
-method: Woodbury
-criterion: BayesRule
-cross-validation criterion = 0.3198318 (0.003886968)
-bias-adjusted cross-validation criterion = 0.3139445 (0.004009339)
-full-sample criterion = 0.3067729 
-\end{CodeOutput}
-\end{CodeChunk}
-
-When \code{reps} \textgreater{} \code{1}, the result returned by
-\code{cv()} is an object of class \code{"cvList"}---literally a list of
-\code{"cv"} objects. The results are reported for each repetition and
-then averaged across repetitions, with the standard deviations of the CV
-criterion and the biased-adjusted CV criterion given in parentheses. In
-this example, there is therefore little variation across repetitions,
-increasing our confidence in the reliability of the results.
-
-Notice that the seed that's set in the \code{cv()} command pertains to
-the first repetition and the seeds for the remaining repetitions are
-then selected pseudo-randomly.\footnote{Because of the manner in which
-  the computation is performed, the order of the replicates in the
-  \code{"cvList"} object returned by \code{cv()} isn't the same as the
-  order in which the replicates are computed. Each element of the
-  result, however, is a \code{"cv"} object with the correct
-  random-number seed saved, and so this technical detail can be safely
-  ignored. The individual \code{"cv"} objects are printed in the order
-  in which they are stored rather than the order in which they are
-  computed.} Setting the first seed, however, makes the entire process
-easily replicable, and the seed for each repetition is stored in the
-corresponding element of the \code{"cvList"} object (which isn't,
-however, saved in the example).
-
-It's also possible to replicate CV when comparing competing models via
-the \code{cv()} method for \code{"modList"} objects. Recall our
-comparison of polynomial regressions of varying degree fit to the
-\code{Auto} data; we performed 10-fold CV for each of 10 models. Here,
-we replicate that process 5 times for each model and graph the results:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv.auto.reps <- cv(models(m.1, m.2, m.3, m.4, m.5,
-+                         m.6, m.7, m.8, m.9, m.10),
-+                  data=Auto, seed=8004, reps=5)
-R> plot(cv.auto.reps)
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/model-comparison-with-reps-1}
-
-}
-
-\caption[ Replicated cross-validated 10-fold CV as a function of polynomial degree, $p$]{ Replicated cross-validated 10-fold CV as a function of polynomial degree, $p$}\label{fig:model-comparison-with-reps}
-\end{figure}
-\end{CodeChunk}
-
-The graph shows both the average CV criterion and its range for each of
-the competing models.
-
-\hypertarget{cross-validating-model-selection}{%
-\section{Cross-validating model
-selection}\label{cross-validating-model-selection}}
 
@@
 
-\hypertarget{mrozs-logistic-regression-revisited}{%
-\subsection{Mroz's logistic regression
-revisited}\label{mrozs-logistic-regression-revisited}}
-
-For a contrasting example we apply model selection to Mroz's logistic
-regression for married women's labor-force participation. First, recall
-the logistic regression model that we fit to the \code{Mroz} data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> summary(m.mroz)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-glm(formula = lfp ~ ., family = binomial, data = Mroz)
-
-Coefficients:
-             Estimate Std. Error z value Pr(>|z|)
-(Intercept)  3.182140   0.644375   4.938 7.88e-07
-k5          -1.462913   0.197001  -7.426 1.12e-13
-k618        -0.064571   0.068001  -0.950 0.342337
-age         -0.062871   0.012783  -4.918 8.73e-07
-wcyes        0.807274   0.229980   3.510 0.000448
-hcyes        0.111734   0.206040   0.542 0.587618
-lwg          0.604693   0.150818   4.009 6.09e-05
-inc         -0.034446   0.008208  -4.196 2.71e-05
-
-(Dispersion parameter for binomial family taken to be 1)
-
-    Null deviance: 1029.75  on 752  degrees of freedom
-Residual deviance:  905.27  on 745  degrees of freedom
-AIC: 921.27
-
-Number of Fisher Scoring iterations: 4
-\end{CodeOutput}
-\end{CodeChunk}
-
-Applying stepwise model selection Mroz's logistic regression, using BIC
-as the model-selection criterion (via the argument
-\code{k=log(nrow(Mroz))} to \code{stepAIC()}) selects 5 of the 7
-original predictors:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> m.mroz.sel <- stepAIC(m.mroz, k=log(nrow(Mroz)),
-+                       trace=FALSE)
-R> summary(m.mroz.sel)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-glm(formula = lfp ~ k5 + age + wc + lwg + inc, family = binomial, 
-    data = Mroz)
-
-Coefficients:
-            Estimate Std. Error z value Pr(>|z|)
-(Intercept)  2.90193    0.54290   5.345 9.03e-08
-k5          -1.43180    0.19320  -7.411 1.25e-13
-age         -0.05853    0.01142  -5.127 2.94e-07
-wcyes        0.87237    0.20639   4.227 2.37e-05
-lwg          0.61568    0.15014   4.101 4.12e-05
-inc         -0.03367    0.00780  -4.317 1.58e-05
-
-(Dispersion parameter for binomial family taken to be 1)
-
-    Null deviance: 1029.75  on 752  degrees of freedom
-Residual deviance:  906.46  on 747  degrees of freedom
-AIC: 918.46
-
-Number of Fisher Scoring iterations: 3
-\end{CodeOutput}
-\begin{CodeInput}
-R> BayesRule(Mroz$lfp == "yes",
-+           predict(m.mroz.sel, type="response"))
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0.3187251
-attr(,"casewise loss")
-[1] "y != round(yhat)"
-\end{CodeOutput}
-\end{CodeChunk}
-
-Bayes rule applied to the selected model misclassifies 32\% of the cases
-in the \code{Mroz} data.
-
-Cross-validating the selected model produces a similar, slightly larger,
-estimate of misclassification, about 33\%:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(m.mroz.sel, criterion=BayesRule, seed=345266)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 345266
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-method: exact
-criterion: BayesRule
-cross-validation criterion = 0.3306773
-bias-adjusted cross-validation criterion = 0.3333245
-95% CI for bias-adjusted CV criterion = (0.2996998, 0.3669493)
-full-sample criterion = 0.3187251 
-\end{CodeOutput}
-\end{CodeChunk}
-
-Is this estimate of predictive performance optimistic?
-
-We proceed to apply the model-selection procedure by cross-validation,
-producing more or less the same result:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> m.mroz.sel.cv <- cvSelect(selectStepAIC, Mroz, 
-+                           seed=6681,
-+                           criterion=BayesRule,
-+                           model=m.mroz,
-+                           AIC=FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 6681
-\end{CodeOutput}
-\begin{CodeInput}
-R> m.mroz.sel.cv
-\end{CodeInput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-cross-validation criterion = 0.3306773
-bias-adjusted cross-validation criterion = 0.3345203
-95% CI for bias-adjusted CV criterion = (0.3008955, 0.368145)
-full-sample criterion = 0.3187251 
-\end{CodeOutput}
-\end{CodeChunk}
-
-Setting \code{AIC=FALSE} in the call to \code{cvSelect()} uses the BIC
-rather than the AIC as the model-selection criterion. As it turns out,
-exactly the same predictors are selected when each of the 10 folds are
-omitted, and the several coefficient estimates are very similar, as we
-show using \code{compareFolds()}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> compareFolds(m.mroz.sel.cv)
-\end{CodeInput}
-\begin{CodeOutput}
-        (Intercept)     age     inc      k5     lwg wcyes
-Fold 1       2.5014 -0.0454 -0.0388 -1.3613  0.5653  0.85
-Fold 2       3.0789 -0.0659 -0.0306 -1.5335  0.6923  0.79
-Fold 3       3.0141 -0.0595 -0.0305 -1.3994  0.5428  0.86
-Fold 4       2.7251 -0.0543 -0.0354 -1.4474  0.6298  1.09
-Fold 5       2.7617 -0.0566 -0.0320 -1.4752  0.6324  0.74
-Fold 6       3.0234 -0.0621 -0.0348 -1.4537  0.6618  0.94
-Fold 7       2.9615 -0.0600 -0.0351 -1.4127  0.5835  0.97
-Fold 8       2.9598 -0.0603 -0.0329 -1.3865  0.6210  0.69
-Fold 9       3.2481 -0.0650 -0.0381 -1.4138  0.6093  0.94
-Fold 10      2.7724 -0.0569 -0.0295 -1.4503  0.6347  0.85
-\end{CodeOutput}
-\end{CodeChunk}
-
-In this example, therefore, we appear to obtain a realistic estimate of
-model performance directly from the selected model, because there is
-little added uncertainty induced by model selection.
-
 \hypertarget{cross-validating-choice-of-transformations-in-regression}{%
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/scatterplot-matrix-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/scatterplot-matrix-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/transformed-predictors-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/transformed-predictors-1}
 
@@
 
-{\centering \includegraphics{JSS-article_files/figure-latex/CR-plots-untransformed-1}
+{\centering \includegraphics{JSS-article-reduced_files/figure-latex/CR-plots-untransformed-1}
 
@@
 
-{\centering \includegraphics{JSS-article_files/figure-latex/CR-plots-transformed-1}
+{\centering \includegraphics{JSS-article-reduced_files/figure-latex/CR-plots-transformed-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/Auto-explore-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/Auto-explore-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/Auto-working-model-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/Auto-working-model-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/Auto-transformed-scatterplot-matrix-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/Auto-transformed-scatterplot-matrix-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article_files/figure-latex/Auto-CR-plots-transformed-1}
+{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/Auto-CR-plots-transformed-1}
 
@@
 
-\hypertarget{parallel-computations}{%
-\section{Parallel computations}\label{parallel-computations}}
-
-The CV functions in the \pkg{cv} package are all capable of performing
-parallel computations by setting the \code{ncores} argument (specifying
-the number of computer cores to be used) to a number \textgreater{}
-\code{1} (which is the default). Parallel computation can be
-advantageous for large problems, reducing the execution time of the
-program.
-
-To illustrate, let's time model selection in Mroz's logistic regression,
-repeating the computation as performed previously and then doing it in
-parallel using 2 cores:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> system.time(m.mroz.sel.cv <- cvSelect(selectStepAIC, Mroz,
-+                           seed=6681,
-+                           criterion=BayesRule,
-+                           model=m.mroz,
-+                           AIC=FALSE))
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 6681
-\end{CodeOutput}
-\begin{CodeOutput}
-   user  system elapsed 
-  0.241   0.010   0.253 
-\end{CodeOutput}
-\begin{CodeInput}
-R> system.time(m.mroz.sel.cv.p <- cvSelect(selectStepAIC, Mroz,
-+                           seed=6681,
-+                           criterion=BayesRule,
-+                           model=m.mroz,
-+                           AIC=FALSE,
-+                           ncores=2))
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 6681
-\end{CodeOutput}
-\begin{CodeOutput}
-   user  system elapsed 
-  0.035   0.004   0.894 
-\end{CodeOutput}
-\begin{CodeInput}
-R> all.equal(m.mroz.sel.cv, m.mroz.sel.cv.p)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] TRUE
-\end{CodeOutput}
-\end{CodeChunk}
-
-In this small problem, the parallel computation is actually
-\emph{slower}, because there is an overhead cost to parallelization, but
-we can see that it produces the same result as before.
-
 \hypertarget{extending-the-cv-package}{%
@@
 \code{selectModel()}.
-
-\hypertarget{adding-a-cost-criterion}{%
-\subsection{Adding a cost criterion}\label{adding-a-cost-criterion}}
-
-A cost criterion suitable for use with \code{cv()} or \code{cvSelect()}
-should take two arguments, \code{y} (the observed response vector) and
-\code{yhat} (a vector of fitted or predicted response values), and
-return a numeric index of lack of fit. The \pkg{cv} package supplies
-several such criteria: \code{mse(y, yhat)}, which returns the
-mean-squared prediction error for a numeric response;
-\code{rmse(y, yhat)}, which returns the (square-)root mean-squared
-error; \code{medAbsErr(y, yhat)}, which returns the median absolute
-error; and \code{BayesRule(y, yhat)} (and its non-error-checking
-version, \code{BayesRule2(y, yhat))}, suitable for use with a binary
-regression model, where \textbackslash code\{y` is the binary response
-coded \code{0} for a ``failure'' or \code{1} for a ``success''; where
-\code{yhat} is the predicted probability of success; and where the
-proportion of \emph{incorrectly} classified cases is returned.
-
-To illustrate using a different prediction cost criterion, we'll base a
-cost criterion on the area under the receiver operating characteristic
-(``ROC'') curve for a logistic regression. The ROC curve is a graphical
-representation of the classification power of a binary regression model,
-and the area under the ROC curve (``AUC''), which varies from 0 to 1, is
-a common summary measure based on the ROC
-\citep[see][]{Wikipedia-ROC:2023}. The \pkg{Metrics} package
-\citep{HamnerFrasco:2018} includes a variety of measures useful for
-model selection, including an \code{auc()} function. We convert the AUC
-into a cost measure by taking its complement:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> AUCcomp <- function(y, yhat) 1 - Metrics::auc(y, yhat)
-\end{CodeInput}
-\end{CodeChunk}
-
-We then apply \code{AUCcomp()} to the the Mroz logistic regression
-discussed in the main \pkg{cv} package vignette, which we reproduce
-here, using the \code{Mroz} data frame from the \pkg{carData} package
-\citep{FoxWeisberg:2019}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> data("Mroz", package="carData")
-R> m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
-R> summary(m.mroz)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-glm(formula = lfp ~ ., family = binomial, data = Mroz)
-
-Coefficients:
-             Estimate Std. Error z value Pr(>|z|)
-(Intercept)  3.182140   0.644375   4.938 7.88e-07
-k5          -1.462913   0.197001  -7.426 1.12e-13
-k618        -0.064571   0.068001  -0.950 0.342337
-age         -0.062871   0.012783  -4.918 8.73e-07
-wcyes        0.807274   0.229980   3.510 0.000448
-hcyes        0.111734   0.206040   0.542 0.587618
-lwg          0.604693   0.150818   4.009 6.09e-05
-inc         -0.034446   0.008208  -4.196 2.71e-05
-
-(Dispersion parameter for binomial family taken to be 1)
-
-    Null deviance: 1029.75  on 752  degrees of freedom
-Residual deviance:  905.27  on 745  degrees of freedom
-AIC: 921.27
-
-Number of Fisher Scoring iterations: 4
-\end{CodeOutput}
-\begin{CodeInput}
-R> AUCcomp(with(Mroz, as.numeric(lfp == "yes")), fitted(m.mroz))
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0.2636161
-\end{CodeOutput}
-\end{CodeChunk}
-
-Cross-validating this cost measure is straightforward:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("cv")
-R> cv(m.mroz, criterion=AUCcomp, seed=3639)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 3639
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-method: exact
-criterion: AUCcomp
-cross-validation criterion = 0.2747088
-full-sample criterion = 0.2636161 
-\end{CodeOutput}
-\end{CodeChunk}
-
-As expected, the cross-validated complement to the AUC is somewhat less
-optimistic than the criterion computed from the model fit to the whole
-data set.
-
-As we explain in the vignette ``Cross-validation of regression models,''
-the \code{cv()} function differentiates between CV criteria that are
-averages of casewise components and criteria that are not. Computation
-of bias corrections and confidence intervals is limited to the former.
-We show in the appendix to this vignette that the AUC, and hence its
-complement, cannot be expressed as averages of casewise components.
-
-\code{cv()} looks for a \code{"casewise loss"} attribute of the value
-returned by a CV criterion function. If this attribute exists, then the
-criterion is treated as the mean of casewise components, and \code{cv()}
-uses the unexported function \code{getLossFn()} to construct a function
-that returns the casewise components of the criterion.
-
-We illustrate with the \code{mse()}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> mse
-\end{CodeInput}
-\begin{CodeOutput}
- [1] 23.94366 18.98477 18.94499 18.87633 18.42697 18.24065 18.07817 18.06613
- [9] 18.02697 18.00953
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv:::getLossFn(mse(rnorm(100), rnorm(100)))
-\end{CodeInput}
-\begin{CodeOutput}
-function (y, yhat) 
-{
-    (y - yhat)^2
-}
-<environment: 0x12f694d48>
-\end{CodeOutput}
-\end{CodeChunk}
-
-For this scheme to work, the ``casewise loss'' attribute must be a
-character string (or vector of character strings), here
-\code{"(y - yhat)\^2"}, that evaluates to an expression that is a
-function of \code{y} and \code{yhat}, and that computes the vector of
-casewise components of the CV criterion.
 
@@
 
-\begin{center}\includegraphics{JSS-article_files/figure-latex/BEPS-plot-1} \end{center}
+\begin{center}\includegraphics{JSS-article-reduced_files/figure-latex/BEPS-plot-1} \end{center}
 
@@
 marginal Bayes-rule prediction that everyone votes for Labour.
-
-\hypertarget{mixed-effects-models}{%
-\subsubsection{Mixed-effects models}\label{mixed-effects-models}}
-
-Adding a \code{cv()} method for a mixed-model class is somewhat more
-complicated. We provide the \code{cvMixed()} function to facilitate this
-process, and to see how that works, consider the \code{"lme"} method
-from the \pkg{cv} package:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv:::cv.lme
-\end{CodeInput}
-\begin{CodeOutput}
-function(model, data = insight::get_data(model), criterion = mse,
-                   k, reps = 1, seed, ncores = 1, clusterVariables, ...){
-  cvMixed(
-    model,
-    package="nlme",
-    data=data,
-    criterion=criterion,
-    k=k,
-    reps=reps,
-    seed=seed,
-    ncores=ncores,
-    clusterVariables=clusterVariables,
-    predict.clusters.args=list(object=model,
-                               newdata=data,
-                               level=0),
-    predict.cases.args=list(object=model,
-                            newdata=data,
-                            level=1),
-    ...)
-}
-<bytecode: 0x148948190>
-<environment: namespace:cv>
-\end{CodeOutput}
-\end{CodeChunk}
-
-Notice that \code{cv.lme()} sets up a call to \code{cvMixed()}, which
-does the computational work.
-
-Most of the arguments of \code{cvMixed()} are familiar:
-
-\begin{itemize}
-\item
-  \code{model} is the mixed-model object, here of class \code{"lme"}.
-\item
-  \code{package} is the name of the package in which the mixed-modeling
-  function used to fit the model, here \code{lme()}, resides---i.e.,
-  \code{"nlme"}; \code{cvMixed()} uses this argument to retrieve the
-  package namespace.
-\item
-  \code{data} is the data set to which the model is fit, by default
-  extracted by the \code{get_data()} function in the \pkg{insight}
-  package.
-\item
-  \code{criterion} is the CV criterion, defaulting to the \code{mse()}
-  function.
-\item
-  \code{k} is the number of CV folds, defaulting to \code{"loo"} for CV
-  by clusters and \code{10} for CV by cases.
-\item
-  \code{reps} is the number of times the CV process is repeated,
-  defaulting to \code{1}.
-\item
-  \code{seed} is the seed for \proglang{R}'s random-number generator,
-  defaulting to a randomly selected (and saved) value.
-\item
-  \code{ncores} is the number of cores to use for parallel computation;
-  if \code{1}, the default, then the computation isn't parallelized.
-\item
-  \code{clusterVariables} is a character vector of the names of
-  variables defining clusters; if missing, then CV is based on cases
-  rather than clusters.
-\end{itemize}
-
-The remaining two arguments are unfamiliar:
-
-\begin{itemize}
-\item
-  \code{predict.clusters.args} is a named list of arguments to be passed
-  to the \code{predict()} function to obtain predictions for the full
-  data set from a model fit to a subset of the data for cluster-based
-  CV. The first two arguments should be \code{object} and
-  \code{newdata}. It is typically necessary to tell \code{cvMixed()} how
-  to base predictions only on fixed effects; in the case of \code{"lme"}
-  models, this is done by setting \code{level = 0}.
-\item
-  Similarly, \code{predict.cases.args} is a named list of arguments to
-  be passed to \code{predict()} for case-based CV. Setting
-  \code{level = 1} includes random effects in the predictions.
-\end{itemize}
-
-Finally, any additional arguments, absorbed by \code{...}, are passed to
-\code{update()} when the model is refit with each fold omitted.
-\code{cvMixed()} returns an object of class \code{"cv"}.
-
-Now imagine that we want to support a new class of mixed-effects models.
-To be concrete, we illustrate with the \code{glmmPQL()} function in the
-\pkg{MASS} package \citep{VenablesRipley:2002}, which fits
-generalized-linear mixed-effects models by penalized
-quasi-likelihood.\footnote{This example is somewhat artificial in that
-  \code{glmmPQL()} has largely been superseded by computationally
-  superior functions, such the \code{glmer()} function in the \pkg{lme4}
-  package. There is, however, one situation in which \code{glmmPQL()}
-  might prove useful: to specify serial dependency in case-level errors
-  within clusters for longitudinal data, which is not currently
-  supported by \code{glmer()}.} Not coincidentally, the arguments of
-\code{glmmPQL()} are similar to those of \code{lme()} (with an
-additional \code{family} argument), because the former iteratively
-invokes the latter; so \code{cv.glmmPQL()} should resemble
-\code{cv.lme()}.
-
-As it turns out, neither the default method for \code{GetResponse()} nor
-\code{insight::get_data()} work for \code{"glmmPQL"} objects. These
-objects include a \code{"data"} element, however, and so we can simply
-extract this element as the default for the \code{data} argument of our
-\code{cv.glmmPQL()} method.
-
-To get the response variable is more complicated: We refit the fixed
-part of the model as a GLM with only the regression constant on the
-right-hand side, and extract the response from that; because all we need
-is the response variable, we limit the number of GLM iterations to 1 and
-suppress warning messages about non-convergence:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> GetResponse.glmmPQL <- function(model, ...){
-+   f <- formula(model)
-+   f[[3]] <- 1 # regression constant only on RHS
-+   model <- suppressWarnings(glm(f, data=model$data, family=model$family,
-+                                 control=list(maxit=1)))
-+   cv::GetResponse(model)
-+ }
-\end{CodeInput}
-\end{CodeChunk}
-
-Writing the \code{cv()} method is then straightforward:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv.glmmPQL <- function(model, data = model$data, criterion = mse,
-+                      k, reps = 1, seed, ncores = 1, clusterVariables, ...){
-+   cvMixed(
-+     model,
-+     package="MASS",
-+     data=data,
-+     criterion=criterion,
-+     k=k,
-+     reps=reps,
-+     seed=seed,
-+     ncores=ncores,
-+     clusterVariables=clusterVariables,
-+     predict.clusters.args=list(object=model,
-+                                newdata=data,
-+                                level=0,
-+                                type="response"),
-+     predict.cases.args=list(object=model,
-+                             newdata=data,
-+                             level=1,
-+                             type="response"),
-+     verbose=FALSE,
-+     ...)
-+ }
-\end{CodeInput}
-\end{CodeChunk}
-
-We set the argument \code{verbose=FALSE} to suppress \code{glmmPQL()}'s
-iteration counter when \code{cvMixed()} calls \code{update()}.
-
-Let's apply our newly minted method to a logistic regression with a
-random intercept in an example that appears in \code{?glmmPQL}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("MASS")
-R> m.pql <- glmmPQL(y ~ trt + I(week > 2), random = ~ 1 | ID,
-+              family = binomial, data = bacteria)
-\end{CodeInput}
-\begin{CodeOutput}
-iteration 1
-\end{CodeOutput}
-\begin{CodeOutput}
-iteration 2
-\end{CodeOutput}
-\begin{CodeOutput}
-iteration 3
-\end{CodeOutput}
-\begin{CodeOutput}
-iteration 4
-\end{CodeOutput}
-\begin{CodeOutput}
-iteration 5
-\end{CodeOutput}
-\begin{CodeOutput}
-iteration 6
-\end{CodeOutput}
-\begin{CodeInput}
-R> summary(m.pql)
-\end{CodeInput}
-\begin{CodeOutput}
-Linear mixed-effects model fit by maximum likelihood
-  Data: bacteria 
-  AIC BIC logLik
-   NA  NA     NA
-
-Random effects:
- Formula: ~1 | ID
-        (Intercept)  Residual
-StdDev:    1.410637 0.7800511
-
-Variance function:
- Structure: fixed weights
- Formula: ~invwt 
-Fixed effects:  y ~ trt + I(week > 2) 
-                    Value Std.Error  DF   t-value p-value
-(Intercept)      3.412014 0.5185033 169  6.580506  0.0000
-trtdrug         -1.247355 0.6440635  47 -1.936696  0.0588
-trtdrug+        -0.754327 0.6453978  47 -1.168779  0.2484
-I(week > 2)TRUE -1.607257 0.3583379 169 -4.485311  0.0000
- Correlation: 
-                (Intr) trtdrg trtdr+
-trtdrug         -0.598              
-trtdrug+        -0.571  0.460       
-I(week > 2)TRUE -0.537  0.047 -0.001
-
-Standardized Within-Group Residuals:
-       Min         Q1        Med         Q3        Max 
--5.1985361  0.1572336  0.3513075  0.4949482  1.7448845
-
-Number of Observations: 220
-Number of Groups: 50 
-\end{CodeOutput}
-\end{CodeChunk}
-
-We compare this result to that obtained from \code{glmer()} in the
-\pkg{lme4} package:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("lme4")
-R> m.glmer <- glmer(y ~ trt + I(week > 2) + (1 | ID),
-+                family = binomial, data = bacteria)
-R> summary(m.glmer)
-\end{CodeInput}
-\begin{CodeOutput}
-Generalized linear mixed model fit by maximum likelihood (Laplace
-  Approximation) [glmerMod]
- Family: binomial  ( logit )
-Formula: y ~ trt + I(week > 2) + (1 | ID)
-   Data: bacteria
-
-     AIC      BIC   logLik deviance df.resid 
-   202.3    219.2    -96.1    192.3      215
-
-Scaled residuals: 
-    Min      1Q  Median      3Q     Max 
--4.5615  0.1359  0.3022  0.4217  1.1276
-
-Random effects:
- Groups Name        Variance Std.Dev.
- ID     (Intercept) 1.543    1.242   
-Number of obs: 220, groups:  ID, 50
-
-Fixed effects:
-                Estimate Std. Error z value Pr(>|z|)
-(Intercept)       3.5479     0.6958   5.099 3.41e-07
-trtdrug          -1.3667     0.6770  -2.019 0.043517
-trtdrug+         -0.7826     0.6831  -1.146 0.251925
-I(week > 2)TRUE  -1.5985     0.4759  -3.359 0.000783
-
-Correlation of Fixed Effects:
-            (Intr) trtdrg trtdr+
-trtdrug     -0.593              
-trtdrug+    -0.537  0.487       
-I(wk>2)TRUE -0.656  0.126  0.064
-\end{CodeOutput}
-\begin{CodeInput}
-R>   # comparison of fixed effects:
-R> car::compareCoefs(m.pql, m.glmer) 
-\end{CodeInput}
-\begin{CodeOutput}
-Warning in car::compareCoefs(m.pql, m.glmer): models to be compared are of
-different classes
-\end{CodeOutput}
-\begin{CodeOutput}
-Calls:
-1: glmmPQL(fixed = y ~ trt + I(week > 2), random = ~1 | ID, family = 
-  binomial, data = bacteria)
-2: glmer(formula = y ~ trt + I(week > 2) + (1 | ID), data = bacteria, 
-  family = binomial)
-
-                Model 1 Model 2
-(Intercept)       3.412   3.548
-SE                0.514   0.696
-
-trtdrug          -1.247  -1.367
-SE                0.638   0.677
-
-trtdrug+         -0.754  -0.783
-SE                0.640   0.683
-
-I(week > 2)TRUE  -1.607  -1.598
-SE                0.355   0.476
-
-\end{CodeOutput}
-\end{CodeChunk}
-
-The two sets of estimates are similar, but not identical
-
-Finally, we try out our \code{cv.glmmPQL()} method, cross-validating
-both by clusters and by cases,
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(m.pql, clusterVariables="ID", criterion=BayesRule)
-\end{CodeInput}
-\begin{CodeOutput}
-n-Fold Cross Validation based on 50 {ID} clusters
-cross-validation criterion = 0.1954545
-bias-adjusted cross-validation criterion = 0.1954545
-full-sample criterion = 0.1954545 
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.pql, data=bacteria, criterion=BayesRule, seed=1490)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 1490
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-cross-validation criterion = 0.2090909
-bias-adjusted cross-validation criterion = 0.2072727
-full-sample criterion = 0.1454545 
-\end{CodeOutput}
-\end{CodeChunk}
-
-and again compare to \code{glmer()}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(m.glmer, clusterVariables="ID", criterion=BayesRule)
-\end{CodeInput}
-\begin{CodeOutput}
-n-Fold Cross Validation based on 50 {ID} clusters
-cross-validation criterion = 0.1954545
-bias-adjusted cross-validation criterion = 0.1954545
-full-sample criterion = 0.1954545 
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.glmer, data=bacteria, criterion=BayesRule, seed=1490)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 1490
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-cross-validation criterion = 0.1954545
-bias-adjusted cross-validation criterion = 0.1936364
-full-sample criterion = 0.15 
-\end{CodeOutput}
-\end{CodeChunk}
-
-\hypertarget{adding-a-model-selection-procedure}{%
-\subsection{Adding a model-selection
-procedure}\label{adding-a-model-selection-procedure}}
-
-The \code{selectStepAIC()} function supplied by the \pkg{cv} package,
-which is based on the \code{stepAIC()} function from the \pkg{nnet}
-package \citep{VenablesRipley:2002} for stepwise model selection, is
-suitable for the \code{procedure} argument of \code{cvSelect()}. The use
-of \code{selectStepAIC()} is illustrated in the principal vignette for
-the package.
-
-We'll employ \code{selectStepAIC()} as a ``template'' for writing a CV
-model-selection procedure. To see the code for this function, type
-\code{cv::selectStepAIC} at the \proglang{R} command prompt, or examine
-the sources for the \pkg{cv} package at
-\url{https://github.com/gmonette/cv} (the code for
-\code{selectStepAIC()} is in
-\url{https://github.com/gmonette/cv/blob/main/R/cvSelect.R}).
-
-Another approach to model selection is all-subsets regression. The
-\code{regsubsets()} function in the \pkg{leaps} package
-\citep{LumleyMiller:2020} implements an efficient algorithm for
-selecting the best-fitting linear least-squares regressions for subsets
-of predictors of all sizes, from 1 through the maximum number of
-candidate predictors.\footnote{The \code{regsubsets()} function computes
-  several measures of model predictive performance, including the
-  \(R^2\) and \(R^2\) adjusted for degrees of freedom, the residual sums
-  of squares, Mallows's \(C_p\), and the BIC. Several of these are
-  suitable for comparing models with differing numbers of
-  coefficients---we use the BIC below---but all necessarily agree when
-  comparing models with the \emph{same} number of coefficients.} To
-illustrate the use of \code{regsubsets()}, we employ the \code{swiss}
-data frame supplied by the \pkg{leaps} package:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("leaps")
-R> head(swiss)
-\end{CodeInput}
-\begin{CodeOutput}
-             Fertility Agriculture Examination Education Catholic
-Courtelary        80.2        17.0          15        12     9.96
-Delemont          83.1        45.1           6         9    84.84
-Franches-Mnt      92.5        39.7           5         5    93.40
-Moutier           85.8        36.5          12         7    33.77
-Neuveville        76.9        43.5          17        15     5.16
-Porrentruy        76.1        35.3           9         7    90.57
-             Infant.Mortality
-Courtelary               22.2
-Delemont                 22.2
-Franches-Mnt             20.2
-Moutier                  20.3
-Neuveville               20.6
-Porrentruy               26.6
-\end{CodeOutput}
-\begin{CodeInput}
-R> nrow(swiss)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 47
-\end{CodeOutput}
-\end{CodeChunk}
-
-The data set includes the following variables, for each of 47
-French-speaking Swiss provinces circa 1888:
-
-\begin{itemize}
-\tightlist
-\item
-  \code{Fertility}: A standardized fertility measure.
-\item
-  \code{Agriculture}: The percentage of the male population engaged in
-  agriculture.
-\item
-  \code{Examination}: The percentage of draftees into the Swiss army
-  receiving the highest grade on an examination.
-\item
-  \code{Education}: The percentage of draftees with more than a
-  primary-school education.
-\item
-  \code{Catholic}: The percentage of the population who were Catholic.
-\item
-  \code{Infant.Mortality}: The infant-mortality rate, expressed as the
-  percentage of live births surviving less than a year.
-\end{itemize}
-
-Following \citet{LumleyMiller:2020}, we treat \code{Fertility} as the
-response and the other variables as predictors in a linear least-squares
-regression:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> m.swiss <- lm(Fertility ~ ., data=swiss)
-R> summary(m.swiss)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-lm(formula = Fertility ~ ., data = swiss)
-
-Residuals:
-     Min       1Q   Median       3Q      Max 
--15.2743  -5.2617   0.5032   4.1198  15.3213
-
-Coefficients:
-                 Estimate Std. Error t value Pr(>|t|)
-(Intercept)      66.91518   10.70604   6.250 1.91e-07
-Agriculture      -0.17211    0.07030  -2.448  0.01873
-Examination      -0.25801    0.25388  -1.016  0.31546
-Education        -0.87094    0.18303  -4.758 2.43e-05
-Catholic          0.10412    0.03526   2.953  0.00519
-Infant.Mortality  1.07705    0.38172   2.822  0.00734
-
-Residual standard error: 7.165 on 41 degrees of freedom
-Multiple R-squared:  0.7067,    Adjusted R-squared:  0.671 
-F-statistic: 19.76 on 5 and 41 DF,  p-value: 5.594e-10
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.swiss, seed=8433)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 8433
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-method: Woodbury
-criterion: mse
-cross-validation criterion = 59.68315
-bias-adjusted cross-validation criterion = 58.84614
-full-sample criterion = 44.78815 
-\end{CodeOutput}
-\end{CodeChunk}
-
-Thus, the RMSE for the model fit to the complete data is considerably
-smaller than the CV estimate of the RMSE. Can we do better by selecting
-a subset of the predictors, taking account of the additional uncertainty
-induced by model selection?
-
-First, let's apply best-subset selection to the complete data set:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> swiss.sub <- regsubsets(Fertility ~ ., data=swiss)
-R> summary(swiss.sub)
-\end{CodeInput}
-\begin{CodeOutput}
-Subset selection object
-Call: regsubsets.formula(Fertility ~ ., data = swiss)
-5 Variables  (and intercept)
-                 Forced in Forced out
-Agriculture          FALSE      FALSE
-Examination          FALSE      FALSE
-Education            FALSE      FALSE
-Catholic             FALSE      FALSE
-Infant.Mortality     FALSE      FALSE
-1 subsets of each size up to 5
-Selection Algorithm: exhaustive
-         Agriculture Examination Education Catholic Infant.Mortality
-1  ( 1 ) " "         " "         "*"       " "      " "             
-2  ( 1 ) " "         " "         "*"       "*"      " "             
-3  ( 1 ) " "         " "         "*"       "*"      "*"             
-4  ( 1 ) "*"         " "         "*"       "*"      "*"             
-5  ( 1 ) "*"         "*"         "*"       "*"      "*"             
-\end{CodeOutput}
-\begin{CodeInput}
-R> (bics <- summary(swiss.sub)$bic)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] -19.60287 -28.61139 -35.65643 -37.23388 -34.55301
-\end{CodeOutput}
-\begin{CodeInput}
-R> which.min(bics)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 4
-\end{CodeOutput}
-\begin{CodeInput}
-R> car::subsets(swiss.sub, legend="topright")
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics{JSS-article_files/figure-latex/subset-selection-1}
-
-}
-
-\caption[Selecting the best model of each size]{Selecting the best model of each size.}\label{fig:subset-selection}
-\end{figure}
-\end{CodeChunk}
-
-The graph, produced by the \code{subsets()} function in the \pkg{car}
-package, shows that the model with the smallest BIC is the best model
-with 4 predictors, including \code{Agriculture}, \code{Education},
-\code{Catholic}, and \code{Infant.Mortality}, but not
-\code{Examination}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> m.best <- update(m.swiss, . ~ . - Examination)
-R> summary(m.best)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-lm(formula = Fertility ~ Agriculture + Education + Catholic + 
-    Infant.Mortality, data = swiss)
-
-Residuals:
-     Min       1Q   Median       3Q      Max 
--14.6765  -6.0522   0.7514   3.1664  16.1422
-
-Coefficients:
-                 Estimate Std. Error t value Pr(>|t|)
-(Intercept)      62.10131    9.60489   6.466 8.49e-08
-Agriculture      -0.15462    0.06819  -2.267  0.02857
-Education        -0.98026    0.14814  -6.617 5.14e-08
-Catholic          0.12467    0.02889   4.315 9.50e-05
-Infant.Mortality  1.07844    0.38187   2.824  0.00722
-
-Residual standard error: 7.168 on 42 degrees of freedom
-Multiple R-squared:  0.6993,    Adjusted R-squared:  0.6707 
-F-statistic: 24.42 on 4 and 42 DF,  p-value: 1.717e-10
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.best, seed=8433) # use same folds as before
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 8433
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-method: Woodbury
-criterion: mse
-cross-validation criterion = 58.46664
-bias-adjusted cross-validation criterion = 57.7779
-full-sample criterion = 45.91637 
-\end{CodeOutput}
-\end{CodeChunk}
-
-The RMSE for the selected model is (of course) slightly higher than for
-the full model fit previously, but the cross-validated RMSE is a bit
-lower; as we explain in the main vignette, however, it isn't kosher to
-select and cross-validate a model on the same data.
-
-Here's a function named \code{selectSubsets()}, meant to be used with
-\code{cvSelect()}, suitable for cross-validating the model-selection
-process:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> selectSubsets <- function(data=insight::get_data(model), 
-+                           model,
-+                           indices,
-+                           criterion=mse,
-+                           save.coef=TRUE, ...){
-+   
-+   if (inherits(model, "lm", which=TRUE) != 1)
-+     stop("selectSubsets is appropriate only for 'lm' models")
-+   
-+   y <- GetResponse(model)
-+   formula <- formula(model)
-+   X <- model.matrix(model)
-+ 
-+   if (missing(indices)) {
-+     # select the best model from the full data by BIC
-+     sel <- leaps::regsubsets(formula, data=data, ...)
-+     bics <- summary(sel)$bic
-+     best <- coef(sel, 1:length(bics))[[which.min(bics)]]
-+     x.names <- names(best)
-+     # fit the best model; intercept is already in X, hence - 1:
-+     m.best <- lm(y ~ X[, x.names] - 1) 
-+     fit.all <- predict(m.best, newdata=data)
-+     return(criterion(y, fit.all)) # return the CV criterion
-+   }
-+ 
-+   # select the best model omitting the i-th fold (given by indices)
-+   sel.i <- leaps::regsubsets(formula, data[-indices, ], ...)
-+   bics.i <- summary(sel.i)$bic
-+   best.i <- coef(sel.i, 1:length(bics.i))[[which.min(bics.i)]]
-+   x.names.i <- names(best.i)
-+   m.best.i <- lm(y[-indices] ~ X[-indices, x.names.i] - 1)
-+               # predict() doesn't work here:
-+   fit.all.i <- as.vector(X[, x.names.i] %*% coef(m.best.i))
-+   fit.i <- fit.all.i[indices]
-+   # return the fitted values for i-th fold, CV criterion for all cases, 
-+   #   and the regression coefficients
-+   list(fit.i=fit.i, # fitted values for i-th fold
-+        crit.all.i=criterion(y, fit.all.i), # CV crit for all cases
-+        coefficients = if (save.coef){ # regression coefficients
-+          coefs <- coef(m.best.i)
-+          
-+          # fix coefficient names
-+          names(coefs) <- sub("X\\[-indices, x.names.i\\]", "",
-+                              names(coefs))
-+          
-+          coefs
-+        }  else {
-+          NULL
-+        }
-+   )
-+ }
-\end{CodeInput}
-\end{CodeChunk}
-
-A slightly tricky point is that because of scoping issues,
-\code{predict()} doesn't work with the model fit omitting the \(i\)th
-fold, and so the fitted values for all cases are computed directly as
-\(\widehat{\mathbf{y}}_{-i} = \mathbf{X} \mathbf{b}_{-i}\), where
-\(\mathbf{X}\) is the model-matrix for all of the cases, and
-\(\mathbf{b}_{-i}\) is the vector of least-squares coefficients for the
-selected model with the \(i\)th fold omitted.
-
-Additionally, the command
-\code{lm(y[-indices] ~ X[-indices, x.names.i] - 1)}, which is the
-selected model with the \(i\)th fold deleted, produces awkward
-coefficient names like \code{"X[-indices, x.names.i]Infant.Mortality"}.
-Purely for aesthetic reasons, the command
-\texttt{sub("X\textbackslash{}\textbackslash{}{[}-indices,\ x.names.i\textbackslash{}\textbackslash{}{]}",\ "",\ names(coefs))}
-fixes these awkward names, removing the extraneous text,
-\code{"X[-indices, x.names.i]"}.
-
-Applying \code{selectSubsets()} to the full data produces the full-data
-cross-validated RMSE (which we obtained previously):
-\%\%\texttt{\{r\ test-selectSubsets\}\ \%\%selectSubsets(model=m.swiss)\ \%\%}
-Similarly, applying the function to an imaginary ``fold'' of 5 cases
-returns the RMSE for the cases in the fold, based on the model selected
-and fit to the cases omitting the fold; the RMSE for all of the cases,
-based on the same model; and the coefficients of the selected model,
-which includes 4 or the 5 predictors (and the intercept):
-\%\%\texttt{\{r\ test-selectSubsets-fold\}\ \%\%selectSubsets(model=m.swiss,\ indices=seq(5,\ 45,\ by=10))\ \%\%}
-
-Then, using \code{selectSubsets()} in cross-validation, we get:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> (cv.swiss <- cvSelect(selectSubsets, model=m.swiss,
-+                       data=swiss, seed=8433)) # use same folds
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 8433
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-cross-validation criterion = 65.83495
-bias-adjusted cross-validation criterion = 63.64431
-full-sample criterion = 45.91637 
-\end{CodeOutput}
-\end{CodeChunk}
-
-Cross-validation shows that model selection exacts a penalty in RMSE.
-Examining the models selected for the 10 folds reveals that there is
-some uncertainty in identifying the predictors in the ``best'' model,
-with \code{Agriculture} sometimes appearing and sometimes not:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> compareFolds(cv.swiss)
-\end{CodeInput}
-\begin{CodeOutput}
-        (Intercept) Catholic Education Infant.Mortality Agriculture
-Fold 1      59.0852   0.1397   -1.0203           1.2985       -0.17
-Fold 2      67.0335   0.1367   -1.0499           0.9413       -0.20
-Fold 3      55.0453   0.1221   -0.8757           1.3541       -0.15
-Fold 4      62.5543   0.1236   -0.9719           1.0679       -0.16
-Fold 5      50.4643   0.1057   -0.7863           1.2144            
-Fold 6      68.0289   0.1195   -1.0073           0.8294       -0.17
-Fold 7      66.5219   0.1357   -1.0827           0.9523       -0.19
-Fold 8      46.3507   0.0776   -0.7637           1.4463            
-Fold 9      62.2632   0.1230   -1.0067           1.1000       -0.17
-Fold 10     52.5112   0.1005   -0.7232           1.0809            
-\end{CodeOutput}
-\end{CodeChunk}
 
@@
 
-\hypertarget{why-the-complement-of-auc-isnt-a-casewise-cv-criterion}{%
-\section{Why the complement of AUC isn't a casewise CV
-criterion}\label{why-the-complement-of-auc-isnt-a-casewise-cv-criterion}}
-
-Consider calculating AUC for folds in which a validation set contains
-\(n_v\) observations. To calculate AUC in the validation set, we need
-the vector of prediction criteria,
-\(\widehat{\mathbf{y}}_{v_{(n_v \times 1)}} = (\widehat{y}_1, ..., \widehat{y}_{n_v})^T\),
-and the vector of observed responses in the validation set,
-\(\mathbf{y}_{v_{(n_v \times 1)}} = (y_1, \ldots, y_{n_v})^T\) with
-\(y_i \in \{0,1\}, \; i = 1, \ldots, n_v\).
-
-To construct the ROC curve, only the ordering of the values in
-\(\mathbf{\widehat{y}}_v\) is relevant. Thus, assuming that there are no
-ties, and reordering observations if necessary, we can set
-\(\mathbf{\widehat{y}}_v = (1, 2, \ldots, n_v)^T\).
-
-If the AUC can be expressed as the casewise mean or sum of a function
-\(\mathrm{cv}(\widehat{y}_i,y_i)\), where
-\(\mathrm{cv}: \{1,2,...,n_v\}\times\{0,1\} \rightarrow [0,1]\), then
-\begin{equation}
-\label{eq:cw}
-\tag{1}
-\sum_{i=1}^{n_v} \mathrm{cv}(\widehat{y}_i,y_i) = \mathrm{AUC}(\mathbf{\widehat{y}}_v,\mathbf{y}_v)
-\end{equation} must hold for all \(2^{n_v}\) possible values of
-\(\mathbf{y}_v = (y_1,...,y_{n_v})^T\). If all \(y\mathrm{s}\) have the
-same value, either 1 or 0, then the definition of AUC is ambiguous. AUC
-could be considered undefined, or it could be set to 0 if all \(y\)s are
-0 and to 1 if all \(y\)s are 1. If AUC is considered to be undefined in
-these cases, we have \(2^{n_v} - 2\) admissible values for
-\(\mathbf{y}_v\).
-
-Thus, equation (\ref{eq:cw}) produces either \(2^{n_v}\) or
-\(2^{n_v}-2\) constraints. Although there are only \(2n_v\) possible
-values for the \(\mathrm{cv(\cdot)}\) function, equation (\ref{eq:cw})
-could, nevertheless, have consistent solutions. We therefore need to
-determine whether there is a value of \(n_v\) for which (\ref{eq:cw})
-has no consistent solution for all admissible values of
-\(\mathbf{y}_v\). In that eventuality, we will have shown that AUC
-cannot, in general, be expressed through a casewise sum.
-
-If \(n_v=3\), we show below that (\ref{eq:cw}) has no consistent
-solution if we include all possibilities for \(\mathbf{y}_v\), but does
-if we exclude cases where all \(y\)s have the same value. If \(n_v=4\),
-we show that there are no consistent solutions in either case.
-
-The following \proglang{R} function computes AUC from
-\(\mathbf{\widehat{y}}_v\) and \(\mathbf{y}_v\), accommodating the cases
-where \(\mathbf{y}_v\) is all 0s or all 1s:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> AUC <- function(y, yhat = seq_along(y)) {
-+   s <- sum(y)
-+   if (s == 0) return(0)
-+   if (s == length(y)) return(1)
-+   Metrics::auc(y, yhat)
-+ }
-\end{CodeInput}
-\end{CodeChunk}
-
-We then define a function to generate all possible \(\mathbf{y}_v\)s of
-length \(n_v\) as rows of the matrix
-\(\mathbf{Y}_{(2^{n_v} \times n_v)}\):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> Ymat <- function(n_v, exclude_identical = FALSE) {
-+   stopifnot(n_v > 0 && round(n_v) == n_v)    # n_v must be a positive integer
-+   ret <- sapply(0:(2^n_v - 1),
-+                 function(x) as.integer(intToBits(x)) )[1:n_v, ]
-+   ret <- if (is.matrix(ret)) t(ret) else matrix(ret)
-+   colnames(ret) <- paste0("y", 1:ncol(ret))
-+   if (exclude_identical) ret[-c(1, nrow(ret)), ] else ret
-+ }
-\end{CodeInput}
-\end{CodeChunk}
-
-For \(n_v=3\),
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> Ymat(3)
-\end{CodeInput}
-\begin{CodeOutput}
-     y1 y2 y3
-[1,]  0  0  0
-[2,]  1  0  0
-[3,]  0  1  0
-[4,]  1  1  0
-[5,]  0  0  1
-[6,]  1  0  1
-[7,]  0  1  1
-[8,]  1  1  1
-\end{CodeOutput}
-\end{CodeChunk}
-
-If we exclude \(\mathbf{y}_v\)s with identical values, then
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> Ymat(3, exclude_identical = TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-     y1 y2 y3
-[1,]  1  0  0
-[2,]  0  1  0
-[3,]  1  1  0
-[4,]  0  0  1
-[5,]  1  0  1
-[6,]  0  1  1
-\end{CodeOutput}
-\end{CodeChunk}
-
-Here is \(\mathbf{Y}\) with corresponding values of AUC:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cbind(Ymat(3), AUC = apply(Ymat(3), 1, AUC))
-\end{CodeInput}
-\begin{CodeOutput}
-     y1 y2 y3 AUC
-[1,]  0  0  0 0.0
-[2,]  1  0  0 0.0
-[3,]  0  1  0 0.5
-[4,]  1  1  0 0.0
-[5,]  0  0  1 1.0
-[6,]  1  0  1 0.5
-[7,]  0  1  1 1.0
-[8,]  1  1  1 1.0
-\end{CodeOutput}
-\end{CodeChunk}
-
-The values of \(\mathrm{cv}(\widehat{y}_i, y_i)\) that express AUC as a
-sum of casewise values are solutions of equation (\ref{eq:cw}), which
-can be written as solutions of the following system of \(2^{n_v}\)
-linear simultaneous equations in \(2n_v\) unknowns: \begin{equation}
-\label{eq:lin}
-\tag{2}
-(\mathbf{U} -\mathbf{Y}) \mathbf{c}_0 + \mathbf{Y} \mathbf{c}_1
-=
-[\mathbf{U} -\mathbf{Y}, \mathbf{Y}]
-\begin{bmatrix}
-\mathbf{c}_0 \\ \mathbf{c}_1
-\end{bmatrix}
-= \mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})
-\end{equation} where \(\mathbf{U}_{(2^{n_v} \times n_v)}\) is a matrix
-of 1s conformable with \(\mathbf{Y}\);
-\(\mathbf{c}_0 = [\mathrm{cv}(1,0), c(2,0), ..., \mathrm{cv}(n_v,0)]^T\);
-\(\mathbf{c}_1 = [\mathrm{cv}(1,1), c(2,1), ..., \mathrm{cv}(n_v,1)]^T\);
-\([\mathbf{U} -\mathbf{Y}, \mathbf{Y}]_{(2^{n_v} \times 2n_v)}\) and
-\(\begin{bmatrix}\begin{aligned} \mathbf{c}_0 \\ \mathbf{c}_1 \end{aligned} \end{bmatrix}_{(2n_v \times 1)}\)
-are partitioned matrices; and
-\(\mathbf{\widehat{Y}}_{(2^{n_v} \times n_v)}\) is a matrix each of
-whose rows consists of the integers 1 to \(n_v\).
-
-We can test whether equation (\ref{eq:lin}) has a solution for any given
-\(n_v\) by trying to solve it as a least-squares problem, considering
-whether the residuals of the associated linear model are all 0, using
-the ``design matrix'' \([\mathbf{U} -\mathbf{Y}, \mathbf{Y}]\) to
-predict the ``outcome''
-\(\mathrm{AUC}(\mathbf{\widehat{Y}},\mathbf{Y})_{(2^{n_v} \times 1)}\):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> resids <- function(n_v, exclude_identical = FALSE, 
-+                    tol = sqrt(.Machine$double.eps)) {
-+   Y <- Ymat(n_v, exclude_identical = exclude_identical)
-+   AUC <- apply(Y, 1, AUC)
-+   X <- cbind(1-Y, Y)
-+   opts <- options(warn = -1)
-+   on.exit(options(opts))
-+   fit <- lsfit(X, AUC, intercept = FALSE)
-+   ret <- max(abs(residuals(fit)))
-+   if(ret < tol){
-+     ret <- 0
-+     solution <- coef(fit)
-+     names(solution) <- paste0("c(", c(1:n_v, 1:n_v), ",", 
-+                               rep(0:1, each = n_v), ")")
-+     attr(ret, "solution") <- zapsmall(solution)
-+   }
-+   ret
-+ }
-\end{CodeInput}
-\end{CodeChunk}
-
-The case \(n_v=3\), excluding identical \(y\)s, has a solution:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> resids(3, exclude_identical = TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0
-attr(,"solution")
-c(1,0) c(2,0) c(3,0) c(1,1) c(2,1) c(3,1) 
-   1.0    0.0   -0.5    0.5    0.0    0.0 
-\end{CodeOutput}
-\end{CodeChunk}
-
-But, if identical \(y\)s are included, the equation is not consistent:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> resids(3, exclude_identical = FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0.125
-\end{CodeOutput}
-\end{CodeChunk}
-
-For \(n_v=4\), there are no solutions in either case:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> resids(4, exclude_identical = TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0.08333333
-\end{CodeOutput}
-\begin{CodeInput}
-R> resids(4, exclude_identical = FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0.25
-\end{CodeOutput}
-\end{CodeChunk}
-
-Consequently, the widely employed AUC measure of fit for binary
-regression cannot in general be used for a casewise cross-validation
-criterion.
-
 \bibliography{cv.bib}

### 2024-01-04 scratch/notes_john/JSS-article/JSS-article-reduced.Rmd
try multicolumn figure

@@
   \usepackage{amsmath}
-output: rticles::jss_article
+output: 
+  rticles::jss_article:
+    extra_dependencies: "subfig"
 bibliography: ["cv.bib"]
@@
 ```{r, polynomial-regression-CV-graph-2, fig.show="hold"}
-#| out.width = "45%",
-#| fig.height = 6,
-#| fig.cap = "Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$"
+#| fig.height = 5,
+#| fig.cap = "Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$",
+#| fig.subcap=c('', ''),
+#| fig.ncol = 2, out.width = "50%", fig.align = "center"
 plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",

### 2024-01-04 scratch/notes_john/JSS-article/JSS-article-reduced.tex
try multicolumn figure

@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{subfig}
 
@@
 \begin{CodeOutput}
-R RNG seed set to 423836
+R RNG seed set to 886597
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.26863
-bias-adjusted cross-validation criterion = 19.25351
+cross-validation criterion = 19.41669
+bias-adjusted cross-validation criterion = 19.39385
 full-sample criterion = 18.98477 
@@
 
-{\centering \includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-2-1} \includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-2-2}
+{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-2-1} }\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-2-2} }
 
@@
 
-\caption[Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$]{Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$}\label{fig:polynomial-regression-CV-graph-2}
+\caption[Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$]{Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$}\label{fig:polynomial-regression-CV-graph-2}
 \end{figure}

### 2024-01-04 scratch/notes_john/JSS-article/JSS-article-reduced.Rmd
put plots for contrived mixed-models data in single figure; text needs adjustment

@@
 
-```{r plot1}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Hierarchical data set, showing the first 10 of 100 students."
+```{r plot1, echo=FALSE, results=FALSE}
+#| fig.height = 4,
+#| fig.cap = "(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data",
+#| fig.subcap=rep("", 5),
+#| fig.ncol = 2, out.width = "45%", fig.align = "center"
 library("lattice")
 library("latticeExtra")
+library("lme4", quietly=TRUE)
 plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
@@
 plot # display graph
-```
-
-The between-student effect of age is 0 but the within-student effect is 1.
-Due to the large variation in ages between students, the least-squares regression of yodeling skill on age (for the 500 observations among all 100 students) produces an estimated slope close to 0 (though with a small $p$-value), because the slope is heavily weighted toward the between-student effect:
-
-```{r}
-summary(lm(y ~ x, data=Data))
-```
-
-The initial mixed-effects model that we fit to the data is a simple random-intercepts model:
-```{r include=FALSE, echo=FALSE}
-library(lme4) # necessary for some reason to knit vignette in RStudio, harmless otherwise
-```
-```{r}
-# random intercept only:
+
 mod.0 <- lmer(y ~ 1 + (1 | group), Data)
-summary(mod.0)
-```
-
-We will shortly consider three other, more complex, mixed models; because of data-management considerations, it is convenient to fit them now, but we defer discussion of these models:
-
-```{r}
-# effect of x and random intercept:
 mod.1 <- lmer(y ~ x + (1 | group), Data)
-
-# effect of x, contextual (student) mean of x, and random intercept:
 mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
-        # equivalent to y ~ I(x - xm) + xm + (1 | group)
-
-# model generating the data (where Bb = 0)
 mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
-```
-
-We proceed to obtain predictions from the random-intercept model (\code{mod.0}) and the other models (\code{mod.1}, \code{mod.2}, and \code{mod.3}) based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students):
-
-```{r}
+
 Data <- within(Data, {
@@
 })
-```
-
-We then prepare the data for plotting:
-
-```{r}
+
 Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
@@
 )
-```
-
-Predictions based on the random-intercept model \code{mod.0} for the first 10 students are shown in the following graph:
-
-```{r plot-fits-mod0}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Predictions from the random intercept model."
+
 (plot +
@@
     lines=list(lty=c(1, 3))))
-```
-
-The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean.
-Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.
-
-Our next model, \code{mod.1}, includes a fixed intercept and fixed effect of \code{x} along with a random intercept:
-
-```{r}
-summary(mod.1)
-```
-
-Predictions from this model appear in the following graph:
-
-```{r plot-fits-mod1}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Predictions from the model with random intercepts and $x$ as a fixed-effect predictor."
+
 (plot +
@@
     lines=list(lty=c(1, 3))))
-```
-
-The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, \code{mod.0}.
-We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod1} to \code{mod0}, but that cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.
-
-Our third model, \code{mod.2}, includes the contextual effect of $x$---that is, the cluster mean \code{xm]---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:
-
-```{r}
-summary(mod.2)
-```
-
-This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that generated the data once the coefficient of the contextual predictor \code{xm} is set to 0 (as it is in \code{mod.3], discussed below).
-
-Predictions from model \code{mod.2} appear in the following graph:
-```{r plot-fits-mod2}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Predictors from the model with random intercepts, $x$, and the group (student) mean of $x$ as predictors."
+
 (plot +
@@
     lines=list(lty=c(1, 3))))
-```
-
-Depending on the estimated variance parameters of the model, a mixed model like \code{mod.2} will apply varying degrees of shrinkage to the random-intercept BLUPs that correspond to variation in the heights of the parallel fitted lines for the individual students.
-In our contrived data, the \code{mod.2} applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student.
-The fit of the mixed model \code{mod.2} is consequently similar to that of a fixed-effects model with age and a categorical predictor for individual students (i.e., treating students as a factor, and not shown here).
-
-The  mixed model \code{mod.2} therefore fits individual observations well, and we anticipate a favorable assessment using individual-based cross-validation.
-In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for model \code{mod.2} compared to the smaller model \code{mod.0}, which includes only fixed and random intercepts.
-
-Had the mixed model applied considerable shrinkage, then neither cluster-based nor case-based cross-validation would show much improvement over the random-intercept-only model.
-In our experience, the degree of shrinkage does not vary smoothly as parameters are changed but tends to be "all or nothing," and near the tipping point, the behavior of estimates can be affected considerably by the choice of algorithm used to fit the model.
-
-Finally, \code{mod.3} directly estimates the model used to generate the data.
-As mentioned, it is a constrained version of \code{mod.2}, with the coefficient of \code{xm} set to 0, and with \code{x} expressed as a deviation from the cluster mean \code{xm}:
-
-```{r}
-summary(mod.3)
-```
-
-The predictions from \code{mod.3} are therefore similar to those from \code{mod.2}:
-
-```{r plot-fits-mod3}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Predictions from the estimated model generating the data."
+
 (plot +
@@
     lines=list(lty=c(1, 3))))
-```
+
+```
+
+The between-student effect of age is 0 but the within-student effect is 1.
+Due to the large variation in ages between students, the least-squares regression of yodeling skill on age (for the 500 observations among all 100 students) produces an estimated slope close to 0 (though with a small $p$-value), because the slope is heavily weighted toward the between-student effect:
+
+```{r}
+summary(lm(y ~ x, data=Data))
+```
+
+The initial mixed-effects model that we fit to the data is a simple random-intercepts model:
+```{r include=FALSE, echo=FALSE}
+library(lme4) # necessary for some reason to knit vignette in RStudio, harmless otherwise
+```
+```{r}
+# random intercept only:
+mod.0 <- lmer(y ~ 1 + (1 | group), Data)
+summary(mod.0)
+```
+
+We will shortly consider three other, more complex, mixed models; because of data-management considerations, it is convenient to fit them now, but we defer discussion of these models:
+
+```{r}
+# effect of x and random intercept:
+mod.1 <- lmer(y ~ x + (1 | group), Data)
+
+# effect of x, contextual (student) mean of x, and random intercept:
+mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
+        # equivalent to y ~ I(x - xm) + xm + (1 | group)
+
+# model generating the data (where Bb = 0)
+mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
+```
+
+We proceed to obtain predictions from the random-intercept model (\code{mod.0}) and the other models (\code{mod.1}, \code{mod.2}, and \code{mod.3}) based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students):
+
+We then prepare the data for plotting:
+
+Predictions based on the random-intercept model \code{mod.0} for the first 10 students are shown in the following graph:
+
+The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean.
+Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.
+
+Our next model, \code{mod.1}, includes a fixed intercept and fixed effect of \code{x} along with a random intercept:
+
+```{r}
+summary(mod.1)
+```
+
+Predictions from this model appear in the following graph:
+
+The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, \code{mod.0}.
+We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod1} to \code{mod0}, but that cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.
+
+Our third model, \code{mod.2}, includes the contextual effect of $x$---that is, the cluster mean \code{xm]---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:
+
+```{r}
+summary(mod.2)
+```
+
+This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that generated the data once the coefficient of the contextual predictor \code{xm} is set to 0 (as it is in \code{mod.3], discussed below).
+
+Predictions from model \code{mod.2} appear in the following graph:
+
+Depending on the estimated variance parameters of the model, a mixed model like \code{mod.2} will apply varying degrees of shrinkage to the random-intercept BLUPs that correspond to variation in the heights of the parallel fitted lines for the individual students.
+In our contrived data, the \code{mod.2} applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student.
+The fit of the mixed model \code{mod.2} is consequently similar to that of a fixed-effects model with age and a categorical predictor for individual students (i.e., treating students as a factor, and not shown here).
+
+The  mixed model \code{mod.2} therefore fits individual observations well, and we anticipate a favorable assessment using individual-based cross-validation.
+In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for model \code{mod.2} compared to the smaller model \code{mod.0}, which includes only fixed and random intercepts.
+
+Had the mixed model applied considerable shrinkage, then neither cluster-based nor case-based cross-validation would show much improvement over the random-intercept-only model.
+In our experience, the degree of shrinkage does not vary smoothly as parameters are changed but tends to be "all or nothing," and near the tipping point, the behavior of estimates can be affected considerably by the choice of algorithm used to fit the model.
+
+Finally, \code{mod.3} directly estimates the model used to generate the data.
+As mentioned, it is a constrained version of \code{mod.2}, with the coefficient of \code{xm} set to 0, and with \code{x} expressed as a deviation from the cluster mean \code{xm}:
+
+```{r}
+summary(mod.3)
+```
+
+The predictions from \code{mod.3} are therefore similar to those from \code{mod.2}:
 

### 2024-01-04 scratch/notes_john/JSS-article/JSS-article-reduced.tex
put plots for contrived mixed-models data in single figure; text needs adjustment

@@
 \begin{CodeOutput}
-R RNG seed set to 886597
+R RNG seed set to 838816
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.41669
-bias-adjusted cross-validation criterion = 19.39385
+cross-validation criterion = 19.36368
+bias-adjusted cross-validation criterion = 19.34363
 full-sample criterion = 18.98477 
@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> library("lattice")
-R> library("latticeExtra")
-R> plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
-+                ylim=c(4, 16),
-+                par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
-+     layer(panel.ellipse(..., center.cex=0))
-R> plot # display graph
-\end{CodeInput}
 \begin{figure}
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot1-1}
+{\centering \subfloat[\label{fig:plot1-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-1} }\subfloat[\label{fig:plot1-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-2} }\newline\subfloat[\label{fig:plot1-3}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-3} }\subfloat[\label{fig:plot1-4}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-4} }\newline\subfloat[\label{fig:plot1-5}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-5} }
 
@@
 
-\caption[Hierarchical data set, showing the first 10 of 100 students]{Hierarchical data set, showing the first 10 of 100 students.}\label{fig:plot1}
+\caption[(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data]{(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data}\label{fig:plot1}
 \end{figure}
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> Data <- within(Data, {
-+   fit_mod0.fe <- predict(mod.0, re.form = ~ 0) # fixed effects only
-+   fit_mod0.re <- predict(mod.0) # fixed and random effects (BLUPs)
-+   fit_mod1.fe <- predict(mod.1, re.form = ~ 0)
-+   fit_mod1.re <- predict(mod.1)
-+   fit_mod2.fe <- predict(mod.2, re.form = ~ 0)
-+   fit_mod2.re <- predict(mod.2)
-+   fit_mod3.fe <- predict(mod.3, re.form = ~ 0)
-+   fit_mod3.re <- predict(mod.3)
-+ })
-\end{CodeInput}
-\end{CodeChunk}
-
 We then prepare the data for plotting:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
-+               timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
-R> Data_long$id <- 1:nrow(Data_long)
-R> Data_long <- reshape(Data_long, direction = "long", sep = "_", 
-+               timevar = "modelcode",  varying = grep("_", names(Data_long)))
-R> Data_long$model <- factor(
-+   c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
-+   [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
-+ )
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 first 10 students are shown in the following graph:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> (plot +
-+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
-+          groups=group, type="l", lwd=2) +
-+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
-+          groups=group, type="l", lwd=2, lty=3)
-+ ) |> update(
-+   main="Model: y ~ 1 + (1 | group)",
-+   key=list(
-+     corner=c(0.05, 0.05),
-+     text=list(c("fixed effects only","fixed and random")),
-+     lines=list(lty=c(1, 3))))
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod0-1}
-
-}
-
-\caption[Predictions from the random intercept model]{Predictions from the random intercept model.}\label{fig:plot-fits-mod0}
-\end{figure}
-\end{CodeChunk}
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> (plot +
-+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
-+          groups=group, type="l", lwd=2) +
-+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
-+          groups=group, type="l", lwd=2, lty=3)
-+ ) |> update(
-+   main="Model: y ~ 1 + x + (1 | group)",
-+   ylim=c(-15, 35),
-+   key=list(
-+     corner=c(0.95, 0.05),
-+     text=list(c("fixed effects only","fixed and random")),
-+     lines=list(lty=c(1, 3))))
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod1-1}
-
-}
-
-\caption[Predictions from the model with random intercepts and $x$ as a fixed-effect predictor]{Predictions from the model with random intercepts and $x$ as a fixed-effect predictor.}\label{fig:plot-fits-mod1}
-\end{figure}
-\end{CodeChunk}
-
 The BLUPs fit the observed data very closely, but predictions based on
@@
 Predictions from model \code{mod.2} appear in the following graph:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> (plot +
-+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
-+          groups=group, type="l", lwd=2) +
-+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
-+          groups=group, type="l", lwd=2, lty=3)
-+ ) |> update(
-+   main="Model: y ~ 1 + x + xm + (1 | group)",
-+   ylim=c(4, 16),
-+   key=list(
-+     corner=c(0.05, 0.05),
-+     text=list(c("fixed effects only","fixed and random")),
-+     lines=list(lty=c(1, 3))))
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod2-1}
-
-}
-
-\caption[Predictors from the model with random intercepts, $x$, and the group (student) mean of $x$ as predictors]{Predictors from the model with random intercepts, $x$, and the group (student) mean of $x$ as predictors.}\label{fig:plot-fits-mod2}
-\end{figure}
-\end{CodeChunk}
 
@@
 \code{mod.2}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> (plot +
-+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
-+          groups=group, type="l", lwd=2) +
-+   xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
-+          groups=group, type="l", lwd=2, lty=3)
-+ ) |> update(
-+   main="Model: y ~ 1 + I(x - xm) + (1 | group)",
-+   ylim=c(4, 16),
-+   key=list(
-+     corner=c(0.05, 0.05),
-+     text=list(c("fixed effects only","fixed and random")),
-+     lines=list(lty=c(1, 3))))
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/plot-fits-mod3-1}
-
-}
-
-\caption[Predictions from the estimated model generating the data]{Predictions from the estimated model generating the data.}\label{fig:plot-fits-mod3}
-\end{figure}
-\end{CodeChunk}
 

### 2024-01-04 scratch/notes_john/JSS-article/JSS-article-reduced.Rmd
tweak loading of lme4

@@
 The initial mixed-effects model that we fit to the data is a simple random-intercepts model:
-```{r include=FALSE, echo=FALSE}
-library(lme4) # necessary for some reason to knit vignette in RStudio, harmless otherwise
-```
+
 ```{r}

### 2024-01-04 scratch/notes_john/JSS-article/JSS-article-reduced.tex
tweak loading of lme4

@@
 \begin{CodeOutput}
-R RNG seed set to 838816
+R RNG seed set to 866709
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.36368
-bias-adjusted cross-validation criterion = 19.34363
+cross-validation criterion = 19.27395
+bias-adjusted cross-validation criterion = 19.25863
 full-sample criterion = 18.98477 

### 2024-01-04 scratch/notes_john/JSS-article/JSS-article-reduced.Rmd
combine, delete, and resize graphs

@@
 
-```{r mpg-horsepower-scatterplot}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "`mpg` vs `horsepower` for the `Auto` data"
-plot(mpg ~ horsepower, data=Auto)
-```
 The relationship between the two variables is monotone, decreasing, and nonlinear.
@@
 ```{r mpg-horsepower-scatterplot-polynomials}
-#| out.width = "100%",
+#| out.width = "50%",
 #| fig.height = 5,
@@
 ```{r mpg-horsepower-MSE-se}
-#| out.width = "100%",
+#| out.width = "50%",
 #| fig.height = 5,
@@
 Because we didn't supply names for the models in the calls to the \code{models()} function, the names \code{model.1}, \code{model.2}, etc., are generated by the function.
-
-Finally, we extract and graph the adjusted MSEs for $10$-fold CV and the MSEs for LOO CV:
-```{r polynomial-regression-CV-graph}
-#| out.width = "100%",
-#| fig.height = 5,
-#| fig.cap = "Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$"
-cv.mse.10 <- sapply(cv.auto.10, function(x) x[["adj CV crit"]])
-cv.mse.loo <- sapply(cv.auto.loo, function(x) x[["CV crit"]])
-plot(c(1, 10), range(cv.mse.10, cv.mse.loo), type="n",
-     xlab="Degree of polynomial, p",
-     ylab="Cross-Validated MSE")
-lines(1:10, cv.mse.10, lwd=2, lty=1, col=2, pch=16, type="b")
-lines(1:10, cv.mse.loo, lwd=2, lty=2, col=3, pch=17, type="b")
-legend("topright", inset=0.02,
-       legend=c("10-Fold CV", "LOO CV"),
-       lwd=2, lty=2:1, col=3:2, pch=17:16)
-```
 
@@
 ```{r cross-validation-clusters}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept models with varying fixed effects.
+#| fig.height = 5,
+#| fig.subcap=rep("", 2),
+#| fig.ncol = 2, out.width = "45%", fig.align = "center",
+#| fig.cap = "10-fold (a) cluster-based and (b) case-based cross-validation comparing random intercept models with varying fixed effects.
 The error bars show the 95% confidence interval around the CV estimate of the MSE for each model."
@@
                   "~ 1 + x + xm"=mod.2, "~ 1 + I(x - xm)"=mod.3)
+
 cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
 plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")
-```
-
-```{r cross-validation-cases}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "10-fold case-based cross-validation comparing random intercept models with varying fixed effects."
+
 cvs_cases <- cv(modlist, data=Data, seed=9693)
@@
 These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
+
+# Cross-validating model selection
 
@@
 ```{r scatterplot-matrix}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Scatterplot matrix for the `Prestige` data."
+#| fig.height = 5,
+#| fig.cap = "Scatterplot matrix for the `Prestige` data, (a) untransformed and (b) transformed",
+#| fig.subcap=rep("", 2),
+#| fig.ncol = 2, out.width = "50%", fig.align = "center"
 library("car")
@@
                   data=Prestige, smooth=list(spread=FALSE))
-```
-The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964].
-Several "families" of power transformations can be used, including the original Box-Cox family, simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap.
-8], and @FoxWeisberg:2019 [Chap.
-3] for details.
-Because \code{women} has some 0 values, we use the Yeo-Johnson family:
-```{r power-transform-Prestige}
+
 trans <- powerTransform( cbind(income, education, women) ~ 1,
@@
 summary(trans)
-```
-We thus have evidence of the desirability of transforming \code{income} (by the $1/3$ power) and \code{women} (by the $0.16$ power---which is close to the "0" power, i.e., the log transformation), but not \code{education}.
-Applying the "rounded" power transformations makes the predictors better-behaved:
-```{r transformed-predictors}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Scatterplot matrix for the `Prestige` data with the predictors transformed."
+
 P <- Prestige[, c("prestige", "income", "education", "women")]
@@
 }
-summary(P)
-
 scatterplotMatrix(~ prestige + income + education + women,
                   data=P, smooth=list(spread=FALSE))
-```
+
+```
+The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964].
+Several "families" of power transformations can be used, including the original Box-Cox family, simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap.
+8], and @FoxWeisberg:2019 [Chap.
+3] for details.
+Because \code{women} has some 0 values, we use the Yeo-Johnson family:
+
+We thus have evidence of the desirability of transforming \code{income} (by the $1/3$ power) and \code{women} (by the $0.16$ power---which is close to the "0" power, i.e., the log transformation), but not \code{education}.
+Applying the "rounded" power transformations makes the predictors better-behaved:
+
 Comparing the MSE for the regressions with the original and transformed predictors shows a advantage to the latter:
@@
 Similarly, component+residual plots for the two regressions, produced by the \code{crPlots()} function in the \pkg{car} package, suggest that the partial relationship of \code{prestige} to \code{income} is more nearly linear in the transformed data, but the transformation of \code{women] fails to capture what appears to be a slight quadratic partial relationship; the partial relationship of \code{prestige} to \code{education} is close to linear in both regressions:
-```{r CR-plots-untransformed}
-#| fig.cap = "Component+residual plots for the `Prestige` regression with the original predictors."
+
+```{r CR-plots}
+#| fig.height = 5,
+#| fig.cap = "Component+residual plots for the `Prestige` regression with (a) the original predictors and (b) transformed predictors",
+#| fig.subcap=rep("", 2),
+#| fig.ncol = 1, out.width = "60%", fig.align = "center"
 crPlots(m.pres)
-```
-```{r CR-plots-transformed}
-#| fig.cap = "Component+residual plots for the `Prestige` regression with transformed predictors."
+
 crPlots(m.pres.trans)
@@
 ```{r Auto-explore}
-#| out.width = "100%",
-#| fig.height = 7,
+#| out.width = "60%",
+#| fig.height = 5,
 #| fig.cap = "Scatterplot matrix for the numeric variables in the `Auto` data"
@@
 ```{r Auto-working-model}
-#| out.width = "100%",
-#| fig.height = 7,
+#| out.width = "60%",
+#| fig.height = 5,
 #| fig.cap = "Component+residual plots for the working model fit to the `Auto` data"
@@
 ```{r Auto-transformed-scatterplot-matrix}
-#| out.width = "100%",
-#| fig.height = 7,
+#| out.width = "60%",
+#| fig.height = 5,
 #| fig.cap = "Scatterplot matrix for the transformed numeric variables in the `Auto` data"
@@
 ```{r Auto-CR-plots-transformed}
-#| out.width = "100%",
-#| fig.height = 7,
+#| out.width = "60%",
+#| fig.height = 5,
 #| fig.cap = "Component+residual plots for the model fit to the transformed `Auto` data"

### 2024-01-04 scratch/notes_john/JSS-article/JSS-article-reduced.tex
combine, delete, and resize graphs

@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot(mpg ~ horsepower, data=Auto)
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-scatterplot-1}
-
-}
-
-\caption[`mpg` vs `horsepower` for the `Auto` data]{`mpg` vs `horsepower` for the `Auto` data}\label{fig:mpg-horsepower-scatterplot}
-\end{figure}
-\end{CodeChunk}
-
 The relationship between the two variables is monotone, decreasing, and
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
+{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-MSE-se-1}
+{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-MSE-se-1}
 
@@
 \begin{CodeOutput}
-R RNG seed set to 866709
+R RNG seed set to 577350
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.27395
-bias-adjusted cross-validation criterion = 19.25863
+cross-validation criterion = 19.19107
+bias-adjusted cross-validation criterion = 19.18016
 full-sample criterion = 18.98477 
@@
 etc., are generated by the function.
-
-Finally, we extract and graph the adjusted MSEs for \(10\)-fold CV and
-the MSEs for LOO CV:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv.mse.10 <- sapply(cv.auto.10, function(x) x[["adj CV crit"]])
-R> cv.mse.loo <- sapply(cv.auto.loo, function(x) x[["CV crit"]])
-R> plot(c(1, 10), range(cv.mse.10, cv.mse.loo), type="n",
-+      xlab="Degree of polynomial, p",
-+      ylab="Cross-Validated MSE")
-R> lines(1:10, cv.mse.10, lwd=2, lty=1, col=2, pch=16, type="b")
-R> lines(1:10, cv.mse.loo, lwd=2, lty=2, col=3, pch=17, type="b")
-R> legend("topright", inset=0.02,
-+        legend=c("10-Fold CV", "LOO CV"),
-+        lwd=2, lty=2:1, col=3:2, pch=17:16)
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-1}
-
-}
-
-\caption[Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$]{Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$}\label{fig:polynomial-regression-CV-graph}
-\end{figure}
-\end{CodeChunk}
 
@@
 +                   "~ 1 + x + xm"=mod.2, "~ 1 + I(x - xm)"=mod.3)
+R> 
 R> cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
 R> plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/cross-validation-clusters-1}
-
-}
-
-\caption[10-fold cluster-based cross-validation comparing random intercept models with varying fixed effects]{10-fold cluster-based cross-validation comparing random intercept models with varying fixed effects.
-The error bars show the 95\% confidence interval around the CV estimate of the MSE for each model.}\label{fig:cross-validation-clusters}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
+R> 
 R> cvs_cases <- cv(modlist, data=Data, seed=9693)
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/cross-validation-cases-1}
+{\centering \subfloat[\label{fig:cross-validation-clusters-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/cross-validation-clusters-1} }\subfloat[\label{fig:cross-validation-clusters-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/cross-validation-clusters-2} }
 
@@
 
-\caption[10-fold case-based cross-validation comparing random intercept models with varying fixed effects]{10-fold case-based cross-validation comparing random intercept models with varying fixed effects.}\label{fig:cross-validation-cases}
+\caption[10-fold (a) cluster-based and (b) case-based cross-validation comparing random intercept models with varying fixed effects]{10-fold (a) cluster-based and (b) case-based cross-validation comparing random intercept models with varying fixed effects.
+The error bars show the 95\% confidence interval around the CV estimate of the MSE for each model.}\label{fig:cross-validation-clusters}
 \end{figure}
@@
 levels.
+
+\hypertarget{cross-validating-model-selection}{%
+\section{Cross-validating model
+selection}\label{cross-validating-model-selection}}
 
@@
 +                   data=Prestige, smooth=list(spread=FALSE))
+R> 
+R> trans <- powerTransform( cbind(income, education, women) ~ 1,
++                          data=Prestige, family="yjPower")
+R> summary(trans)
+\end{CodeInput}
+\begin{CodeOutput}
+yjPower Transformations to Multinormality 
+          Est Power Rounded Pwr Wald Lwr Bnd Wald Upr Bnd
+income       0.2678        0.33       0.1051       0.4304
+education    0.5162        1.00      -0.2822       1.3145
+women        0.1630        0.16       0.0112       0.3149
+
+ Likelihood ratio test that all transformation parameters are equal to 0
+                               LRT df      pval
+LR test, lambda = (0 0 0) 15.73879  3 0.0012827
+\end{CodeOutput}
+\begin{CodeInput}
+R> P <- Prestige[, c("prestige", "income", "education", "women")]
+R> (lambdas <- trans$roundlam)
+\end{CodeInput}
+\begin{CodeOutput}
+   income education     women 
+0.3300000 1.0000000 0.1630182 
+\end{CodeOutput}
+\begin{CodeInput}
+R> names(lambdas) <- c("income", "education", "women")
+R> for (var in c("income", "education", "women")){
++   P[, var] <- yjPower(P[, var], lambda=lambdas[var])
++ }
+R> scatterplotMatrix(~ prestige + income + education + women,
++                   data=P, smooth=list(spread=FALSE))
 \end{CodeInput}
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/scatterplot-matrix-1}
+{\centering \subfloat[\label{fig:scatterplot-matrix-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/scatterplot-matrix-1} }\subfloat[\label{fig:scatterplot-matrix-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/scatterplot-matrix-2} }
 
@@
 
-\caption[Scatterplot matrix for the `Prestige` data]{Scatterplot matrix for the `Prestige` data.}\label{fig:scatterplot-matrix}
+\caption[Scatterplot matrix for the `Prestige` data, (a) untransformed and (b) transformed]{Scatterplot matrix for the `Prestige` data, (a) untransformed and (b) transformed}\label{fig:scatterplot-matrix}
 \end{figure}
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> trans <- powerTransform( cbind(income, education, women) ~ 1,
-+                          data=Prestige, family="yjPower")
-R> summary(trans)
-\end{CodeInput}
-\begin{CodeOutput}
-yjPower Transformations to Multinormality 
-          Est Power Rounded Pwr Wald Lwr Bnd Wald Upr Bnd
-income       0.2678        0.33       0.1051       0.4304
-education    0.5162        1.00      -0.2822       1.3145
-women        0.1630        0.16       0.0112       0.3149
-
- Likelihood ratio test that all transformation parameters are equal to 0
-                               LRT df      pval
-LR test, lambda = (0 0 0) 15.73879  3 0.0012827
-\end{CodeOutput}
-\end{CodeChunk}
-
 We thus have evidence of the desirability of transforming \code{income}
@@
 the predictors better-behaved:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> P <- Prestige[, c("prestige", "income", "education", "women")]
-R> (lambdas <- trans$roundlam)
-\end{CodeInput}
-\begin{CodeOutput}
-   income education     women 
-0.3300000 1.0000000 0.1630182 
-\end{CodeOutput}
-\begin{CodeInput}
-R> names(lambdas) <- c("income", "education", "women")
-R> for (var in c("income", "education", "women")){
-+   P[, var] <- yjPower(P[, var], lambda=lambdas[var])
-+ }
-R> summary(P)
-\end{CodeInput}
-\begin{CodeOutput}
-    prestige         income        education          women      
- Min.   :14.80   Min.   :22.15   Min.   : 6.380   Min.   :0.000  
- 1st Qu.:35.23   1st Qu.:44.17   1st Qu.: 8.445   1st Qu.:1.731  
- Median :43.60   Median :50.26   Median :10.540   Median :3.362  
- Mean   :46.83   Mean   :50.76   Mean   :10.738   Mean   :3.502  
- 3rd Qu.:59.27   3rd Qu.:56.24   3rd Qu.:12.648   3rd Qu.:5.591  
- Max.   :87.20   Max.   :83.62   Max.   :15.970   Max.   :6.830  
-\end{CodeOutput}
-\begin{CodeInput}
-R> scatterplotMatrix(~ prestige + income + education + women,
-+                   data=P, smooth=list(spread=FALSE))
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/transformed-predictors-1}
-
-}
-
-\caption[Scatterplot matrix for the `Prestige` data with the predictors transformed]{Scatterplot matrix for the `Prestige` data with the predictors transformed.}\label{fig:transformed-predictors}
-\end{figure}
-\end{CodeChunk}
 
@@
 R> crPlots(m.pres)
+R> 
+R> crPlots(m.pres.trans)
 \end{CodeInput}
@@
 
-{\centering \includegraphics{JSS-article-reduced_files/figure-latex/CR-plots-untransformed-1}
+{\centering \subfloat[\label{fig:CR-plots-1}]{\includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/CR-plots-1} }\newline\subfloat[\label{fig:CR-plots-2}]{\includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/CR-plots-2} }\newline
 
@@
 
-\caption[Component+residual plots for the `Prestige` regression with the original predictors]{Component+residual plots for the `Prestige` regression with the original predictors.}\label{fig:CR-plots-untransformed}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> crPlots(m.pres.trans)
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics{JSS-article-reduced_files/figure-latex/CR-plots-transformed-1}
-
-}
-
-\caption[Component+residual plots for the `Prestige` regression with transformed predictors]{Component+residual plots for the `Prestige` regression with transformed predictors.}\label{fig:CR-plots-transformed}
+\caption[Component+residual plots for the `Prestige` regression with (a) the original predictors and (b) transformed predictors]{Component+residual plots for the `Prestige` regression with (a) the original predictors and (b) transformed predictors}\label{fig:CR-plots}
 \end{figure}
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/Auto-explore-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/Auto-explore-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/Auto-working-model-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/Auto-working-model-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/Auto-transformed-scatterplot-matrix-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/Auto-transformed-scatterplot-matrix-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{JSS-article-reduced_files/figure-latex/Auto-CR-plots-transformed-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/Auto-CR-plots-transformed-1}
 

### 2024-01-10 scratch/notes_john/JSS-article/JSS-article-reduced.Rmd
edit JSS paper

@@
   # For running headers, if needed
-  short:     "\\pkg{cv}: Cross-Validation"
+  short:     "\\pkg{cv}: Cross-Validation of Regression Models"
 abstract: >
@@
 options(prompt = 'R> ', continue = '+ ')
-```
-
-# Cross-validation
+palette(car::carPalette())
+```
+
+# Introduction
 
@@
 Emulating this process in a single study by data-division is less common: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data.
-Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, particularly in smaller samples, the results can vary substantially based on the random division of the data: See @Harrell:2015 [Sec.
+Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples: See @Harrell:2015 [Sec.
 5.3] for this and other remarks about data-division and cross-validation.
@@
 Each fitted model is then used to predict the response variable for the cases in the omitted fold.
-A CV criterion or "cost" measure, such as the mean-squared error ("MSE") of prediction, is then computed using these predicted values.
+A CV criterion (also termed a "cost" or "loss" measure), such as the mean-squared error ("MSE") of prediction, is then computed using these predicted values.
 In the extreme $k = n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."
@@
 In contrast, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) are somewhat biased but have smaller variance.
-It is also possible to correct $k$-fold CV for bias (see below).
-
-# Examples
-
-## Polynomial regression for the Auto data
-
-The data for this example are drawn from the \pkg{ISLR2} package for \proglang{R}, associated with @JamesEtAl:2021.
+It is also possible to correct $k$-fold CV for bias (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+
+The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical model objects.
+The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
+```{r cv-methods}
+library("cv", quietly=TRUE)
+methods("cv")
+```
+The \code{"modList"} method for \code{cv()} cross-validates several competing models, not necessarily of the same class, using the same division of the data into folds.
+The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.
+
+Cross-validating mixed-effects models involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.
+
+The \code{cvSelect()} function, discussed in Section \ref{cross-validating-model-selection}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors.
+
+The \code{"default"} \code{cv()} method works (perhaps with a bit of coaxing) with many other existing regression-model classes for which there is an \code{update()} method that accepts a \code{data} argument.
+More generally, the \pkg{cv} package is designed to be extensible, as discussed in Section \ref{extending-the-cv-package}.
+
+In the interest of brevity, we won't describe all of the features of the \pkg{cv} package here, concentrating on the aspects of the package that are relatively novel.
+For example, the \code{cv()} and \code{cvSelect()} functions can perform computations in parallel and can independently replicate a cross-validation procedure several times.
+These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provides greater detail on some of topics that we do describe, such as extensions to the package.
+
+# Preliminary Example: Polynomial regression
+
+The data for the example in this section are drawn from the \pkg{ISLR2} package for \proglang{R}, associated with @JamesEtAl:2021.
 The presentation here is close (though not identical) to that in the original source [@JamesEtAl:2021 Secs.
-5.1, 5.3], and it demonstrates the use of the `cv()` function in the \pkg{cv} package.[^boot]
+5.1, 5.3], and it demonstrates the use of the `cv()` function.[^boot]
 
@@
 
-The `Auto` dataset contains information about 392 cars:
+The \code{Auto} dataset contains information about 392 cars:
 
@@
 data("Auto", package="ISLR2")
-head(Auto)
-dim(Auto)
+summary(Auto)
 ```
@@
 
-We'll focus here on the relationship of `mpg` (miles per gallon) to `horsepower`, as displayed in the following scatterplot:
-
-The relationship between the two variables is monotone, decreasing, and nonlinear.
-Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in the following figure:
-
-[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is probably not the best choice here.
-Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader).
-
-```{r mpg-horsepower-scatterplot-polynomials}
+```{r mpg-horsepower-scatterplot-polynomials, echo=FALSE}
 #| out.width = "50%",
 #| fig.height = 5,
-#| fig.cap = "`mpg` vs `horsepower` for the `Auto` data"
+#| fig.cap = "mpg vs horsepower for the Auto data, showing fitted polynomials of degree 1 through 5."
 plot(mpg ~ horsepower, data=Auto)
@@
 ```
+
+We'll focus here on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}.
+The relationship between the two variables is monotone, decreasing, and nonlinear.
+Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}.
 The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).
 
-The following graph shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$.
+[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here.
+Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader).
+We revisit the \code{Auto} data in Section \ref{cross-validating-model-selection}.
+
+Figure \ref{fig:mpg-horsepower-MSE-se} shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$.
 The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.
 
-```{r mpg-horsepower-MSE-se}
+```{r mpg-horsepower-MSE-se, echo=FALSE}
 #| out.width = "50%",
@@
 #| fig.cap = "Estimated squared error as a function of polynomial degree, $p$"
-library("cv") # for mse() and other functions
 
@@
 ```
-The code for this graph uses the `mse()` function from the \pkg{cv} package to compute the MSE for each fit.
-
-### Using cv()
-
-The generic `cv()` function has an `"lm"` method, which by default performs $k = 10$-fold CV:
+
+The generic `cv()` function has an `"lm"` method,
+```{r cv-lm-method}
+args(cv:::cv.lm)
+```
+which takes the following arguments:
+
+* \code{model}, an \code{"lm"} object, the only required argument.
+
+* \code{data}, which can usually be inferred from the \code{model} object.
+
+* \code{criterion}, a function to compute the CV criterion (defaulting to \code{mse}).
+
+* \code{k}, the number of folds to employ (defaulting to \code{10}); the character value \code{"n"} or \code{"loo"} may be supplied to specify leave-one-out cross-validation.
+
+* \code{reps}, the number of times to repeat the CV procedure (defaulting to 1).
+
+* \code{seed}, the seed for \proglang{R}'s pseudo-random number generator; if not specified a value is randomly selected, reported, and saved, so that the CV procedure is replicable.
+
+* \code{confint}, whether or not to compute a confidence interval for the CV criterion, defaulting to \code{TRUE} if there are at least 400 cases; a confidence interval is computed only if the CV criterion can be expressed as the average of casewise components (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals} for details).
+
+* \code{level}, the level for the confidence interval (defaulting to \code{0.95}).
+
+* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computation-for-linear-and-generalized-linear-models}).
+
+* \code{ncores}, the number of cores to employ for parallel computation; if \code{cores = 1} (the default), the computations are not parallelized.
+
+To illustrate, we perform 10-fold CV for a quadratic polynomial fit to the \code{Auto} data:
 ```{r cv-lm-1}
@@
 
-cv(m.auto)
-```
-The \code{"lm"} method by default uses \code{mse()} as the CV criterion and the Woodbury matrix identity to update the regression with each fold deleted without having literally to refit the model.
-Computational details are discussed in the final section of this vignette.
-The function reports the CV estimate of MSE, a biased-adjusted estimate of the MSE (the bias adjustment is explained in the final section), and the MSE is also computed for the original, full-sample regression.
-Because the division of the data into 10 folds is random, \code{cv()} explicitly (randomly) generates and saves a seed for \proglang{R}'s pseudo-random number generator, to make the results replicable.
-The user can also specify the seed directly via the \code{seed} argument to \code{cv()}.
-
-To perform LOO CV, we can set the \code{k} argument to \code{cv()} to the number of cases in the data, here \code{k=392}, or, more conveniently, to \code{k="loo"} or \code{k="n"}:
+cv(m.auto, confint=TRUE)
+```
+The function reports the CV estimate of MSE, a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression.
+Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint=TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.
+
+To perform LOO CV:
 
@@
 ```
-For LOO CV of a linear model, \code{cv()} by default uses the hatvalues from the model fit to the full data for the LOO updates, and reports only the CV estimate of MSE.
+The \code{"hatvalues"} method reports only the CV estimate of MSE.
 Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted.
@@
 ```{r cv.lm-3}
-cv(m.auto, k="loo", method="naive")
-
-cv(m.auto, k="loo", method="Woodbury")
-```
-The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an issue for LOO CV.
+cv(m.auto, k="loo", method="naive", confint=TRUE)
+
+cv(m.auto, k="loo", method="Woodbury", confint=TRUE)
+```
+The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE (and a confidence interval around it) along with the full-sample MSE, but bias isn't an issue for LOO CV.
 
 This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed.
-In this comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM.
-We use the \code{microbenchmark()} function from the package of the same name for the timings [@Mersmann:2023]:
+In the following comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM.
+We use the \code{microbenchmark()} function from the package of the same name  [@Mersmann:2023] for the timings:
 ```{r cv.lm.timings, cache=TRUE}
@@
 [^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100.
-With the exception of the last columns, the output is self-explanatory.
+With the exception of the last column, the output is self-explanatory.
 The last column shows which methods have average timings that are statistically distinguishable.
@@
 
-### Comparing competing models
+## Comparing competing models
 
@@
 
-We illustrate with the polynomial regression models of varying degree for the \code{Auto} data (discussed previously), beginning by fitting and saving the 10 models:
+We illustrate with the polynomial regression models of varying degree for the \code{Auto} data, beginning by fitting and saving the 10 models:
 ```{r polyomial-models}
-for (p in 1:10){
-  assign(paste0("m.", p),
-         lm(mpg ~ poly(horsepower, p), data=Auto))
-}
-objects(pattern="m\\.[0-9]")
-summary(m.2) # for example, the quadratic fit
+mlist <- vector(10, mode="list")
+for (p in 1:10) mlist[[p]] <- lm(mpg ~ poly(horsepower, p), data = Auto)
+names(mlist) <- paste0("m.", 1:10)
+mlist[2] # e.g., the quadratic fit
 ```
@@
 # 10-fold CV
-cv.auto.10 <- cv(models(m.1, m.2, m.3, m.4, m.5,
-                     m.6, m.7, m.8, m.9, m.10),
-              data=Auto, seed=2120)
-cv.auto.10[1:2] # for the linear and quadratic models
+mlist <- do.call(models, mlist) # create "modList" object
+
+cv.auto.10 <- cv(mlist, data=Auto, seed=2120)
+cv.auto.10[2] # e.g., for quadratic model
 
 # LOO CV
-cv.auto.loo <- cv(models(m.1, m.2, m.3, m.4, m.5,
-                        m.6, m.7, m.8, m.9, m.10),
-                 data=Auto, k="loo")
-cv.auto.loo[1:2] # linear and quadratic models
-```
-Because we didn't supply names for the models in the calls to the \code{models()} function, the names \code{model.1}, \code{model.2}, etc., are generated by the function.
-
-Alternatively, we can use the \code{plot()} method for \code{"cvModList"} objects to compare the models, though with separate graphs for 10-fold and LOO CV:
+cv.auto.loo <- cv(mlist, data=Auto, k="loo")
+cv.auto.loo[2] # e.g., for quadratic model
+```
+The \code{models()} function takes an arbitrary number of regression models as its arguments, which are optionally named, to create a \code{"modList"} object.
+Because we generated the polynomial regression models in a named list, we conveniently employ \code{do.call()} to supply the models as arguments to \code{models()}.
+Note that the names created for the list (e.g., \code{"m.2"}) are then used for the models.
+We can also invoke the \code{plot()} method for \code{"cvModList"} objects to compare the models (see Figure \ref{fig:polynomial-regression-CV-graph-2}):
 
@@
 #| fig.height = 5,
-#| fig.cap = "Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$",
+#| fig.cap = "Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$.",
 #| fig.subcap=c('', ''),
@@
 
-In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model, reported above (except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting).
+In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model (cf., Figure \ref{fig:mpg-horsepower-MSE-se} on page \pageref{fig:mpg-horsepower-MSE-se}), except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting.
 
@@
 That is, predicting the response in each fold from the model fit to data in the other folds is like using the model fit to all of the data to predict the response for new cases from the values of the predictors for those new cases.
-As we explained, the application of this idea to independently sampled cases is straightforward---simply partition the data into random folds of equal size and leave each fold out in turn, or, in the case of LOO CV, simply omit each case in turn.
-
-In contrast, mixed-effects models are fit to *dependent* data, in which cases as clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases repeated observations on the individuals over time.[^crossed-effects]
+As we explained, the application of this idea to independently sampled cases is straightforward.
+
+In contrast, mixed-effects models are fit to *dependent* data, in which cases are clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases are repeated observations on the individuals over time.[^crossed-effects]
 
@@
 The \code{lmer()} function in the \pkg{lme4} package is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects.
-We present an example of the latter later in the vignette.
+We present an example of the latter in a vignette for the \pkg{cv} package.
 
@@
 Treat CV as analogous to predicting the response for a newly observed case in an *existing cluster*.
-In this instance, the folds comprise one or more individual cases, and the predictions can use both the fixed and random effects.
+In this instance, the folds comprise one or more individual cases, and the predictions can use both the fixed and random effects---so-called "best-linear-unbiased predictors" or "BLUPs."
 
@@
 The HSB data are included in the \code{MathAchieve} and \code{MathAchSchool} data sets in the \pkg{nlme} package  [@PinheiroBates:2000].
-\code{MathAchieve} includes individual-level data on 7185  students in 160 high schools, and \code{MathAchSchool} includes school-level data:
+\code{MathAchieve} comprises individual-level data on 7185  students in 160 high schools, and \code{MathAchSchool} contains school-level data:
 ```{r HSB-data}
@@
 
-Some data-management is required before fitting a mixed-effects model to the HSB data, for which we use the \pkg{dplyr} package [@WickhamEtAl:2023]:
+Some data-management is required before fitting a mixed-effects model to the HSB data:
 ```{r HSB-data-management, cache=TRUE}
-library("dplyr")
-MathAchieve %>% group_by(School) %>%
-  summarize(mean.ses = mean(SES)) -> Temp
-Temp <- merge(MathAchSchool, Temp, by="School")
-HSB <- merge(Temp[, c("School", "Sector", "mean.ses")],
-             MathAchieve[, c("School", "SES", "MathAch")], by="School")
+HSB <- MathAchieve
+HSB <- merge(MathAchSchool[, c("School", "Sector")],
+             HSB[, c("School", "SES", "MathAch")], by="School")
 names(HSB) <- tolower(names(HSB))
-
-HSB$cses <- with(HSB, ses - mean.ses)
-```
-In the process, we created two new school-level variables: \code{meanses}, which is the average SES for students in each school; and \code{cses}, which is school-average SES centered at its mean.
+HSB <- within(HSB, {
+  mean.ses <- ave(ses, school)
+  cses <- ses - mean.ses
+})
+```
+In the process, we merged variables from the school-level and student-level data sets and created two new school-level variables: \code{mean.ses}, which is the average SES for students in each school; and \code{cses}, which is each students' SES centered at their school means.
 For details, see @FoxWeisberg:2019 [Sec.
@@
 ```{r HSB-lmer, cache=TRUE}
-library("lme4")
+library("lme4", quietly=TRUE)
 hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
@@
 Convergence problems are common in mixed-effects modeling.
-The apparent issue here is that an estimated variance component is close to or equal to 0, which is at a boundary of the parameter space.
+The issue here is that an estimated variance component is close to or equal to 0, which is at a boundary of the parameter space.
 That shouldn't disqualify the fitted model for the kind of prediction required for cross-validation.
 
-There is also a \code{cv()} method for linear mixed models fit by the \code{lme()} function in the \pkg{nlme} package, and the arguments for \code{cv()} in this case are the same as for a model fit by \code{lmer()} or \code{glmer()}.
-We illustrate with the mixed model fit to the HSB data:
-```{r hsb-lme, cache=TRUE}
-library(nlme)
-hsb.lme <- lme(mathach ~ mean.ses*cses + sector*cses,
-                 random = ~ cses | school, data=HSB,
-               control=list(opt="optim"))
-summary(hsb.lme)
-
-cv(hsb.lme, k=10, clusterVariables="school", seed=5240)
-
-cv(hsb.lme, seed=1575)
-```
-We used the same random-number generator seeds as in the previous example cross-validating the model fit by \code{lmer()}, and so the same folds are employed in both cases.[^optimizer] The estimated covariance components and fixed effects in the summary output differ slightly between the \code{lmer()} and \code{lme()} solutions, although both functions seek to maximize the REML criterion.
-This is, of course, to be expected when different algorithms are used for numerical optimization.
-To the precision reported, the cluster-level CV results for the \code{lmer()} and \code{lme()} models are identical, while the case-level CV results are very similar but not identical.
-
-[^optimizer]: The observant reader will notice that we set the argument \code{control=list(opt="optim")} in the call to \code{lme()}, changing the optimizer employed from the default \code{"nlminb"}.
-We did this because with the default optimizer, \code{lme()} encountered the same convergence issue as \code{lmer()}, but rather than issuing a warning, \code{lme()} failed, reporting an error.
-As it turns out, setting the optimizer to \code{"optim"} avoids this problem.
+\code{cv()} also has methods for mixed models fit by the \code{glmer()} function in the \pkg{lme4} package, the \code{lme()} function in the \pkg{nlme} package [@PinheiroBates:2000], and the \code{glmmTMB()} function in the \pkg{glmmTMB} package [@BrooksEtAl], along with a simple procedure for extending \code{cv()} to other classes of mixed-effects models.
+See the vignettes in the \pkg{cv} package for details.
 
@@
 
-We introduce an artificial data set that exemplifies aspects of cross-validation particular to hierarchical models.
+In this section, we introduce an artificial data set that exemplifies aspects of cross-validation particular to hierarchical models.
 Using this data set, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model.
@@
 
-Consider a researcher studying improvement in a skill, yodeling, for example, among students enrolled in a four-year yodeling program.
-The plan is to measure each student's skill level at the beginning of the program and every year thereafter until the end of the program, resulting in five annual measurements for each student.
-It turns out that yodeling appeals to students of all ages, and students enrolling in the program range in age from 20 to 70.
-Moreover, participants' untrained yodeling skill is similar at all ages, as is their rate of progress with training.
+Consider a researcher studying improvement in a skill, singing, for example, among students enrolled in a four-year voice program at a music conservatory.
+The plan is to measure each student's skill level at the beginning of the program and every year thereafter until the end of the program, resulting in 5 annual measurements for each student.
+It turns out that singing appeals to students of all ages, and students enrolling in the program range in age from 20 to 70.
+Moreover, participants' untrained singing skill is similar at all ages, as is their rate of progress with training.
 All students complete the four-year program.
 
-The researcher, who has more expertise in yodeling than in modeling, decides to model the response, $y$, yodeling skill, as a function of age, $x$, reasoning that students get older during their stay in the program, and (incorrectly) that age can serve as a proxy for elapsed time.
+The researcher, who has more expertise in singing than in modeling, decides to model the response, $y$, singing skill, as a function of age, $x$, reasoning that students get older during their stay in the program, and (incorrectly) that age can serve as a proxy for elapsed time.
 The researcher knows that a mixed model should be used to account for clustering due to the expected similarity of measurements taken from each student.
@@
   x = Ax + rep(1:Nx, length.out = Nw*Nb)
-) |>
-  within(
-    {
+) |> within ({
   xm  <- ave(x, group, FUN = mean) # within-group mean
@@
     rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
-    }
-  )
-```
-
-Here is a scatterplot of the data for a representative group of 10 (without loss of generality, the first 10) of 100 students, showing the 95% concentration ellipse for each cluster:[^2]
-
-[^2]: We find it convenient to use the \pkg{lattice} [@Sarkar:2008] and \pkg{latticeExtra} [@SarkarAndrews:2022] packages for this and other graphs in this section.
+})
+```
+
+Figure \ref{fig:plot1} (a) shows a scatterplot the data for a representative group of 10 (without loss of generality, the first 10) of the 100 students, displaying the 95% concentration ellipse for each cluster:
 
@@
 The between-student effect of age is 0 but the within-student effect is 1.
-Due to the large variation in ages between students, the least-squares regression of yodeling skill on age (for the 500 observations among all 100 students) produces an estimated slope close to 0 (though with a small $p$-value), because the slope is heavily weighted toward the between-student effect:
+Due to the large variation in ages between students, the least-squares regression of singing skill on age (for the 500 observations among all 100 students) produces an estimated slope close to 0 (though with a small $p$-value), because the slope is heavily weighted toward the between-student effect:
 
@@
 
-We will shortly consider three other, more complex, mixed models; because of data-management considerations, it is convenient to fit them now, but we defer discussion of these models:
+We will shortly consider three other, more complex, mixed models.
+
+We proceed to obtain predictions from the random-intercept model (\code{mod.0}) for fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---the BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students).
+Predictions for the random-intercept model, \code{mod.0}, for the first 10 students are shown in Figure \ref{fig:plot1} (b).
+The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean.
+Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.
+
+Our next model, \code{mod.1}, includes a fixed intercept and fixed effect of \code{x} along with a random intercept:
 
@@
 mod.1 <- lmer(y ~ x + (1 | group), Data)
-
+summary(mod.1)
+```
+
+Predictions from this model appear in Figure \ref{fig:plot1} (c).
+The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, \code{mod.0}.
+We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod1} to \code{mod0}, but that cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.
+
+Our third model, \code{mod.2}, includes the "contextual effect" of $x$---that is, the cluster mean \code{xm}---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:
+
+```{r}
 # effect of x, contextual (student) mean of x, and random intercept:
@@
         # equivalent to y ~ I(x - xm) + xm + (1 | group)
-
+summary(mod.2)
+```
+
+This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that generated the data once the coefficient of the contextual predictor \code{xm} is set to 0 (as it is in \code{mod.3}, discussed below).
+
+Predictions from \code{mod.2} appear in Figure \ref{fig:plot1} (d).
+Depending on the estimated variance parameters of the model, a mixed model like \code{mod.2} will apply varying degrees of shrinkage to the random-intercept BLUPs that correspond to variation in the heights of the parallel fitted lines for the individual students.
+In our contrived data, \code{mod.2} applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student.
+The fit of the mixed model \code{mod.2} is consequently similar to that of a fixed-effects model with age and a categorical predictor for individual students (i.e., treating students as a factor, and not shown here).
+
+The  mixed model \code{mod.2} therefore fits the individual observations well, and we anticipate a favorable assessment using individual-based cross-validation.
+In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for  \code{mod.2} compared to the smaller \code{mod.0}, which includes only fixed and random intercepts.
+
+Had the mixed model applied considerable shrinkage, then neither cluster-based nor case-based cross-validation would show much improvement over the random-intercept-only model.
+In our experience, the degree of shrinkage does not vary smoothly as parameters are changed but tends to be "all or nothing," and near the tipping point, the behavior of estimates can be affected considerably by the choice of algorithm used to fit the model.
+
+Finally, \code{mod.3} directly estimates the model used to generate the data.
+As mentioned, it is a constrained version of \code{mod.2}, with the coefficient of \code{xm} set to 0, and with \code{x} expressed as a deviation from the cluster mean \code{xm}:
+
+```{r}
 # model generating the data (where Bb = 0)
 mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
-```
-
-We proceed to obtain predictions from the random-intercept model (\code{mod.0}) and the other models (\code{mod.1}, \code{mod.2}, and \code{mod.3}) based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students):
-
-We then prepare the data for plotting:
-
-Predictions based on the random-intercept model \code{mod.0} for the first 10 students are shown in the following graph:
-
-The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean.
-Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.
-
-Our next model, \code{mod.1}, includes a fixed intercept and fixed effect of \code{x} along with a random intercept:
-
-```{r}
-summary(mod.1)
-```
-
-Predictions from this model appear in the following graph:
-
-The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, \code{mod.0}.
-We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod1} to \code{mod0}, but that cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.
-
-Our third model, \code{mod.2}, includes the contextual effect of $x$---that is, the cluster mean \code{xm]---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:
-
-```{r}
-summary(mod.2)
-```
-
-This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that generated the data once the coefficient of the contextual predictor \code{xm} is set to 0 (as it is in \code{mod.3], discussed below).
-
-Predictions from model \code{mod.2} appear in the following graph:
-
-Depending on the estimated variance parameters of the model, a mixed model like \code{mod.2} will apply varying degrees of shrinkage to the random-intercept BLUPs that correspond to variation in the heights of the parallel fitted lines for the individual students.
-In our contrived data, the \code{mod.2} applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student.
-The fit of the mixed model \code{mod.2} is consequently similar to that of a fixed-effects model with age and a categorical predictor for individual students (i.e., treating students as a factor, and not shown here).
-
-The  mixed model \code{mod.2} therefore fits individual observations well, and we anticipate a favorable assessment using individual-based cross-validation.
-In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for model \code{mod.2} compared to the smaller model \code{mod.0}, which includes only fixed and random intercepts.
-
-Had the mixed model applied considerable shrinkage, then neither cluster-based nor case-based cross-validation would show much improvement over the random-intercept-only model.
-In our experience, the degree of shrinkage does not vary smoothly as parameters are changed but tends to be "all or nothing," and near the tipping point, the behavior of estimates can be affected considerably by the choice of algorithm used to fit the model.
-
-Finally, \code{mod.3} directly estimates the model used to generate the data.
-As mentioned, it is a constrained version of \code{mod.2}, with the coefficient of \code{xm} set to 0, and with \code{x} expressed as a deviation from the cluster mean \code{xm}:
-
-```{r}
 summary(mod.3)
@@
 
-The predictions from \code{mod.3} are therefore similar to those from \code{mod.2}:
-
-We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.
-In order to reduce between-model random variability in comparisons of models, we apply \code{cv()} to the list of models created by the \code{models()} function (introduced previously), performing cross-validation with the same folds for each model:
+The predictions from \code{mod.3}, shown in Figure \ref{fig:plot1} (e), are therefore similar to those from \code{mod.2}.
+
+We next carry out case-based cross-validation, which, as we have explained, includes both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which includes fixed effects only.
+In order to reduce between-model random variability in comparisons of models, we apply \code{cv()} to the list of models created by the \code{models()} function (introduced previously), performing cross-validation with the same folds for each model (see Figure \ref{fig:cross-validation-clusters}):
 
@@
 
-In summary, model \code{mod.1}, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
-Model \code{mod.2}, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV.
-The data-generating model, \code{mod.3}, which includes the fixed effect of \code{x - xm} in place of separate terms in \code{x} and \code{xm}, isn't distinguishable from model \code{mod.2}, which includes \code{x} and \code{xm} separately, even though \code{mod.2} has an unnecessary parameter (recall that the population coefficient of \code{xm} is 0 when \code{x} is expressed as deviations from the contextual mean).
-These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
+In summary, \code{mod.1}, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
+\code{mod.2}, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV.
+The data-generating model, \code{mod.3}, which includes the fixed effect of \code{x - xm} in place of separate terms in \code{x} and \code{xm}, isn't distinguishable from \code{mod.2}, which includes \code{x} and \code{xm} separately, even though \code{mod.2} has an unnecessary parameter (recall that the population coefficient of \code{xm} is 0 when \code{x} is expressed as deviations from the contextual mean).
+These conclusions are consistent with our observations based on graphing predictions from the various models in Figure \ref{fig:plot1} (on page \pageref{fig:plot1}), and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
 
@@
 
+As @HastieTibshiraniFriedman:2009 [Sec.
+7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain, if the whole data are used to select or fine-tune a statistical model, then subsequent cross-validation of the model is intrinsically misleading, because the model is selected to fit the whole data, including the part of the data that remains when each fold is removed.
+
 ## A preliminary example
 
-As @HastieTibshiraniFriedman:2009 [Sec.
-7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain, if the whole data are used to select or fine-tune a statistical model, subsequent cross-validation of the model is intrinsically misleading, because the model is selected to fit the whole data, including the part of the data that remains when each fold is removed.
-
 The following example is similar in spirit to one employed by @HastieTibshiraniFriedman:2009.
-Suppose that we randomly generate $n = 1000$ independent observations for a response variable variable $y \sim N(\mu = 10, \sigma^2 = 0)$, and independently sample $1000$ observations for $p = 100$ "predictors," $x_1, \ldots, x_{100}$, each from $x_j \sim N(0, 1)$.
-The response has nothing to do with the predictors and so the population linear-regression model $y_i = \alpha + \beta_1 x_{i1} + \cdots + \beta_{100} x_{i,100} + \varepsilon_i$ has $\alpha = 10$ and all $\beta_j = 0$.
+Suppose that we randomly generate $n = 1000$ independent observations for a response variable variable $y \sim N(\mu = 10, \sigma^2 = 1)$, and independently sample $1000$ observations for $p = 100$ "predictors," $x_1, \ldots, x_{100}$, each from $x_j \sim N(0, 1)$.
+The response has nothing to do with the predictors and so the population linear-regression model $y_i = \alpha + \beta_1 x_{i1} + \cdots + \beta_{100} x_{i,100} + \varepsilon_i$ has $\alpha = 10$, all $\beta_j = 0$, and $\sigma_\varepsilon^2 = 1$.
 
@@
 )
-head(D[, 1:6])
 ```
@@
 m.null <- lm(y ~ 1, data=D)
+summary(m.null)
 anova(m.null, m.full)
-
-summary(m.null)
-```
-Next, using the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], let us perform a forward stepwise regression to select a "best" model, starting with the null model, and using AIC as the model-selection criterion (see the help page for \code{stepAIC()} for details):[^selection-order]
+```
+Next, using the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], let us perform a forward stepwise regression to select a "best" model, starting with the null model, and using AIC as the model-selection criterion (see \code{?stepAIC()} for details):[^selection-order]
 
@@
 The resulting model has 15 predictors, a very modest $R^2 = .044$, but a small $p$-value for its omnibus $F$-test (which, of course, is entirely spurious because the same data were used to select and test the model).
-The MSE for the selected model is smaller than the true error variance $\sigma^2 = 1$, as is the estimated error variance for the selected model, $\widehat{\sigma}^2 = 0.973^2 = 0.947$.
+The MSE for the selected model is smaller than the true error variance $\sigma^2 = 1$, as is the estimated error variance for the selected model, $\widehat{\sigma}^2 = 0.973^2 = 0.947$.[^casewise-loss]
+
+[^casewise-loss]: The \code{"casewise loss"} attribute attached to the value returned by the \code{mse()} function signals to \code{cv()} that MSE is an average of casewise components (and how to calculate the casewise components), which has implications for computing bias correction and confidences intervals; see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}.
 
@@
 
-The \code{cvSelect()} function in the \pkg{cv} package allows us to cross-validate the whole model-selection procedure.
+The \code{cvSelect()} function in the \pkg{cv} package, in contrast, allows us to cross-validate the whole model-selection procedure.
 The first argument to \code{cvSelect()} is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion.
@@
 
-* \code{data}, the data set to which the model is fit;
-* \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, a seed is randomly selected and saved;
+* \code{data}, the data set to which the model is fit.
+* \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, a seed is randomly selected and saved.
 * additional arguments required by the model-selection function, here the starting \code{model} argument, the \code{direction} of model selection, and the \code{scope} of models considered (from the model with only a regression constant to the model with all 100 predictors).
 
-By default, \code{cvSelect()} performs 10-fold CV, and produces an estimate of MSE for the model-selection procedure even *larger* than the true error variance, $\sigma^2 = 1$.
+By default, \code{cvSelect()} performs 10-fold CV, and it produces an estimate of MSE for the model-selection procedure even *larger* than the true error variance, $\sigma^2 = 1$.
 
@@
 If the predictors *aren't* linearly related, then the relationships between them can often be straightened by power transformations.
-Transformations can be selected after graphical examination of the data, or by analytic methods.
+Transformations can be selected after graphical examination of the data, or by analytic methods, such as transforming the predictors towards multivariate normality, which implies linearity.
 Once the relationships between the predictors are linearized, it can be advantageous similarly to transform the response variable towards normality.
@@
 data("Prestige", package="carData")
-head(Prestige)
 summary(Prestige)
@@
 * \code{women}: percentage of occupational incumbents who were women, also from the Census.
-* \code{prestige}: the average prestige rating of the occupation on a 0--100 "thermometer" scale, in a Canadian social survey conducted around the same time.
+* \code{prestige}: the average prestige rating of the occupation on a 0--100 "thermometer" scale, from a Canadian social survey conducted around the same time.
 * \code{type}, type of occupation, and \code{census}, the Census occupational code, which are not used in our example.
@@
 
-A scatterplot matrix (using the \code{scatterplotMatrix()} function in the \pkg{car} package) of the numeric variables in the data reveals that the distributions of \code{income} and \code{women} are positively skewed, and that some of the relationships among the three predictors, and between the predictors and the response (i.e., \code{prestige}), are nonlinear:
-```{r scatterplot-matrix}
+The scatterplot matrix in Figure \ref{fig:scatterplot-matrix} (a) (produced by the \code{scatterplotMatrix()} function in the \pkg{car} package) for the numeric variables in the data reveals that the distributions of \code{income} and \code{women} are positively skewed, and that some of the relationships among the three predictors, and between the predictors and the response (i.e., \code{prestige}), are nonlinear:
+```{r scatterplot-matrix, echo=FALSE, results=FALSE}
 #| fig.height = 5,
-#| fig.cap = "Scatterplot matrix for the `Prestige` data, (a) untransformed and (b) transformed",
+#| fig.cap = "Scatterplot matrix for the Prestige data, (a) untransformed and (b) with the predictors transformed towards multivariate normality.",
 #| fig.subcap=rep("", 2),
 #| fig.ncol = 2, out.width = "50%", fig.align = "center"
-library("car")
+library("car", quietly=TRUE)
 scatterplotMatrix(~ prestige + income + education + women,
                   data=Prestige, smooth=list(spread=FALSE))
-
 trans <- powerTransform( cbind(income, education, women) ~ 1,
                          data=Prestige, family="yjPower")
-summary(trans)
-
 P <- Prestige[, c("prestige", "income", "education", "women")]
@@
                   data=P, smooth=list(spread=FALSE))
-
-```
+```
+
+```{r eval=FALSE}
+library("car", quietly=TRUE)
+scatterplotMatrix(~ prestige + income + education + women,
+                  data=Prestige, smooth=list(spread=FALSE))
+```
+
 The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964].
 Several "families" of power transformations can be used, including the original Box-Cox family, simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap.
-8], and @FoxWeisberg:2019 [Chap.
+8] and @FoxWeisberg:2019 [Chap.
 3] for details.
@@
 
+```{r transform-predictors}
+trans <- powerTransform( cbind(income, education, women) ~ 1,
+                         data=Prestige, family="yjPower")
+summary(trans)
+```
+
 We thus have evidence of the desirability of transforming \code{income} (by the $1/3$ power) and \code{women} (by the $0.16$ power---which is close to the "0" power, i.e., the log transformation), but not \code{education}.
-Applying the "rounded" power transformations makes the predictors better-behaved:
-
-Comparing the MSE for the regressions with the original and transformed predictors shows a advantage to the latter:
+Applying the "rounded" power transformations makes the predictors better-behaved (cf., Figures \ref{fig:scatterplot-matrix} (a) and (b)):
+
+```{r eval=FALSE}
+P <- Prestige[, c("prestige", "income", "education", "women")]
+(lambdas <- trans$roundlam)
+names(lambdas) <- c("income", "education", "women")
+for (var in c("income", "education", "women")){
+  P[, var] <- yjPower(P[, var], lambda=lambdas[var])
+}
+scatterplotMatrix(~ prestige + income + education + women,
+                  data=P, smooth=list(spread=FALSE))
+```
+
+Comparing the MSE for the regressions with the original and transformed predictors shows an advantage to the latter:
 ```{r prestige-regressions}
@@
 ```
-Similarly, component+residual plots for the two regressions, produced by the \code{crPlots()} function in the \pkg{car} package, suggest that the partial relationship of \code{prestige} to \code{income} is more nearly linear in the transformed data, but the transformation of \code{women] fails to capture what appears to be a slight quadratic partial relationship; the partial relationship of \code{prestige} to \code{education} is close to linear in both regressions:
+Similarly, component+residual plots for the two regressions in Figure \ref{fig:CR-plots}, produced by the \code{crPlots()} function in the \pkg{car} package, suggest that the partial relationship of \code{prestige} to \code{income} is more nearly linear in the transformed data, but the transformation of \code{women} fails to capture what appears to be a slight quadratic partial relationship; the partial relationship of \code{prestige} to \code{education} is close to linear in both regressions:
 
@@
 #| fig.height = 5,
-#| fig.cap = "Component+residual plots for the `Prestige` regression with (a) the original predictors and (b) transformed predictors",
+#| fig.cap = "Component+residual plots for the Prestige regression with (a) the original predictors and (b) transformed predictors",
 #| fig.subcap=rep("", 2),
@@
 crPlots(m.pres)
-
 crPlots(m.pres.trans)
@@
 
-Having transformed the predictors towards multinormality, we now consider whether there's evidence for transforming the response (using \code{powerTransform()} for Box and Cox's original method), and we discover that there's not:
+Having transformed the predictors towards multivariate normality, we now consider whether there's evidence for transforming the response (using \code{powerTransform()} for Box and Cox's original method), and we discover that there's not:
 ```{r transform-response}
@@
 The function takes a \code{data} set and "working" \code{model} as arguments, along with the candidate \code{predictors} and \code{response} for transformation, and the transformation \code{family} to employ.
-If the \code{predictors} argument is missing then only the response is transformed, and if the \code{response} argument is missing, only the supplied predictors are transformed.
-The default \code{family} for transforming the predictors is \code{"bcPower"}---the original Box-Cox family---as is the default \code{family.y} for transforming the response; here we specify \code{family="yjPower} because of the 0s in \code{women}.
+If the \code{predictors} argument is missing then only the \code{response} is transformed, and if the \code{response} argument is missing, only the supplied \code{predictors} are transformed.
+The default \code{family} for transforming the predictors is \code{"bcPower"}---the original Box-Cox family---as is the default \code{family.y} for transforming the response; here we specify \code{family="yjPower"} because of the 0s in \code{women}.
 \code{selectTrans()} returns the result of applying a lack-of-fit criterion to the model after the selected transformation is applied, with the default \code{criterion=mse}:
@@
 ```
-\code{selectTrans()} also takes an optional \code{indices} argument, making it suitable for doing computations on a subset of the data (i.e., a CV fold), and hence for use with \code{cvSelect()} (see \code{?selectTrans} for details):
+\code{selectTrans()} also takes an optional \code{indices} argument, making it suitable for doing computations on a subset of the data (i.e., a CV fold), and hence for use with \code{cvSelect()} (see \code{?selectTrans} and the vignette on extending the \pkg{cv} package for details):
 
@@
 That's what we did in the preceding two sections, first for predictor selection and then for selection of transformations in regression.
-In this section, we consider the case where we both select variable transformations and then proceed to select predictors.
-It's insufficient to apply these steps sequentially, first, for example, using \code{cvSelect()} with \code{selectTrans()} and then with \code{selectStepAIC()}; rather we should apply the whole model-selection procedure with each fold omitted.
+In this section, we consider the case where we both choose variable transformations and then proceed to select predictors.
+It's insufficient to apply these steps sequentially, first, for example, using \code{cvSelect()} with \code{selectTrans()} and then with \code{selectStepAIC()}; rather, we should apply the whole model-selection procedure with each fold omitted.
 The \code{selectTransAndStepAIC()} function, also supplied by the \pkg{cv} package, does exactly that.
@@
 ```{r Auto-redux}
-summary(Auto)
+names(Auto)
 xtabs(~ year, data=Auto)
@@
 ```
-We previously used the \code{Auto} here in a preliminary example where we employed CV to inform the selection of the order of a polynomial regression of \code{mpg} on \code{horsepower}.
+The \code{Auto} appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}.
 Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data.
-We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set:
+We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore}):
 ```{r Auto-explore}
@@
 #| fig.height = 5,
-#| fig.cap = "Scatterplot matrix for the numeric variables in the `Auto` data"
+#| fig.cap = "Scatterplot matrix for the numeric variables in the Auto data"
 Auto$cylinders <- factor(Auto$cylinders,
@@
 scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
-                  smooth=list(spread=FALSE), data=Auto)
+                  smooth=list(spread=FALSE), data=Auto, pch=".")
 ```
 A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors.
-There are, however, only 5 distinct values of \code{cylinders} (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare.
-and none of the cars has 7 cylinders.
+There are, however, only 5 distinct values of \code{cylinders} (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders.
 There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.
@@
 #| fig.height = 5,
-#| fig.cap = "Component+residual plots for the working model fit to the `Auto` data"
+#| fig.cap = "Component+residual plots for the working model fit to the Auto data"
 m.auto <- lm(mpg ~ ., data = Auto)
-summary(m.auto)
-
-Anova(m.auto)
-
 crPlots(m.auto)
 ```
-The component+residual plots, created with the \code{crPlots()} function in the previously loaded \pkg{car} package, clearly reveal the inadequacy of the model.
+The component+residual plots in Figure \ref{fig:Auto-working-model} clearly reveal the inadequacy of the model.
 
@@
 ```{r Auto-transform}
-num.predictors <- c("displacement", "horsepower", "weight", "acceleration")
+num.predictors <- c("displacement", "horsepower", "weight", 
+                    "acceleration")
 tr.x <- powerTransform(Auto[, num.predictors])
@@
 }
-head(A)
-
 m <- update(m.auto, data=A)
@@
 summary(powerTransform(m))
-
 m <- update(m, log(mpg) ~ .)
-summary(m)
-
-Anova(m)
-```
-
-The transformed numeric variables are much better-behaved:
+```
+
+The transformed numeric variables are much better-behaved (cf., Figures \ref{fig:Auto-explore} on page \pageref{fig:Auto-explore} and \ref{fig:Auto-transformed-scatterplot-matrix}:
 ```{r Auto-transformed-scatterplot-matrix}
@@
 #| fig.height = 5,
-#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the `Auto` data"
+#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the Auto data"
 scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                   + acceleration, 
-                  smooth=list(spread=FALSE), data=A)
-```
-And the partial relationships in the model fit to the transformed data are much more nearly linear:
+                  smooth=list(spread=FALSE), data=A, pch=".")
+```
+And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figures \ref{fig:Auto-working-model} on page \pageref{fig:Auto-working-model}  and \ref{fig:Auto-CR-plots-transformed}:
 ```{r Auto-CR-plots-transformed}
@@
 #| fig.height = 5,
-#| fig.cap = "Component+residual plots for the model fit to the transformed `Auto` data"
+#| fig.cap = "Component+residual plots for the model fit to the transformed Auto data"
 crPlots(m)
@@
 summary(m.step)
-
-Anova(m.step)
 ```
 The selected model includes three of the numeric predictors, \code{horsepower}, \code{weight}, and \code{acceleration}, along with the factors \code{year} and \code{origin}.
-We can calculate the MSE for this model, but we expect that the result will be optimistic because we used the whole data to help specify the model
+We can calculate the MSE for this model, but we expect that the result will be optimistic because we used the whole data to help specify the model:
 ```{r MSE-whole-selected-model}
@@
 ```
-A perhaps subtle point is that we compute the MSE for the selected model on the original \code{mpg} response scale rather than the log scale, so as to make the selected model comparable to the working model.
-That's slightly uncomfortable given the skewed distribution of \code{mpg}.
-An alternative is to use the median absolute error instead of the mean-squared error, employing the \code{medAbsErr()} function from the \pkg{cv} package:
-```{r Auto-median-absolute-error}
-medAbsErr(Auto$mpg, exp(fitted(m.step)))
-medAbsErr(Auto$mpg, fitted(m.auto))
-```
+A perhaps subtle point is that we compute the MSE for the selected model on the original \code{mpg} response scale rather than the log scale, so as to make the selected model comparable to the working model.[^medAbsErr]
+
+[^medAbsErr]: That's slightly uncomfortable given the skewed distribution of \code{mpg}.
+An alternative is to use a robust measure of model lack-of-fit, such as the median absolute error instead of the mean-squared error, employing the \code{medAbsErr()} function from the \pkg{cv} package.
+The median absolute error, however, cannot be expressed as a casewise average (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
 
@@
                 predictors=num.predictors,
-                response="mpg", AIC=FALSE, criterion=medAbsErr)
+                response="mpg", AIC=FALSE)
 cvs
@@
 ```
-Here, as for \code{selectTrans()}, the \code{predictors} and \code{response} arguments specify candidate variables for transformation,
-and \code{AIC=FALSE} uses the BIC for model selection.
+Here, as for \code{selectTrans()}, the \code{predictors} and \code{response} arguments specify candidate variables for transformation, and \code{AIC=FALSE} uses the BIC for model selection.
 The starting model, \code{m.auto}, is the working model fit to the \code{Auto} data.
-The CV criterion isn't bias-adjusted because median absolute error isn't a mean of casewise error components.
 
@@
 
-* \code{selectTransStepAIC()} automatically computes CV cost criteria, here the median absolute error, on the untransformed response scale.
-* The estimate of the median absolute error that we obtain by cross-validating the whole model-specification process is a little larger than the median absolute error computed for the model we fit to the \code{Auto} data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
-* When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of \code{compareFolds()}), we see that there is little uncertainty in choosing variable transformations (the \code{lam.*}s for the $x$s and \code{lambda} for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: \code{horsepower}, \code{weight}, and \code{year} are always included among the selected predictors; \code{acceleration} and \code{displacement} are each included respectively in 4 and 3 of 10 selected models; and \code{cylinders} and \code{origin} are each included in only 1 of 10 models.
+* \code{selectTransStepAIC()} automatically computes CV cost criteria, here the MSE, on the \emph{untransformed} response scale.
+* The estimate of the MSE that we obtain by cross-validating the whole model-specification process is larger than the MSE computed for the model we fit to the \code{Auto} data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
+* When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of \code{compareFolds()}), we see that there is little uncertainty in choosing variable transformations (the \code{lam.*}s for the $x$s and \code{lambda} for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: \code{horsepower}, \code{weight}, and \code{year} are always included among the selected predictors; \code{acceleration} and \code{displacement} are included respectively in 4 and 3 of 10 selected models; and \code{cylinders} and \code{origin} are each included in only 1 of 10 models.
 Recall that when we selected predictors for the full data, we obtained a model with \code{horsepower}, \code{weight}, \code{acceleration}, \code{year}, and \code{origin}.
@@
 
-The \pkg{cv} package is designed to be extensible in several directions.
-In this vignette, we discuss three kinds of extensions, ordered by increasing general complexity: (1) adding a cross-validation cost criterion; (2) adding a model class that's not directly accommodated by the \code{cv()} default method or by another directly inherited method, with separate consideration of mixed-effects models; and (3) adding a new model-selection procedure suitable for use with \code{selectModel()}.
-
-## Adding a model class not covered by the default cv() method
-
-### Independently sampled cases
+The \pkg{cv} package is designed to be extensible in several directions; in order of increasing general complexity, we can add: (1) a cross-validation cost criterion; (2) a model class that's not directly accommodated by the \code{cv()} default method or by another directly inherited method; and (3) a new model-selection procedure suitable for use with \code{selectModel()}.
+In this section, we illustrate (1) and (2); more extensive examples may be found in the vignette on extending the \pkg{cv} package.
 
@@
 14.2.1], with data from the British Election Panel Study on vote choice in the 2001 British election.
-Data for the example are in the \code{BEPS} data frame in the \pkg{carData} package:
+Data for the example are in the \code{BEPS} data set in the \pkg{carData} package:
 ```{r BEPS-data}
 data("BEPS", package="carData")
-head(BEPS)
+summary(BEPS)
 ```
@@
 
-The model fit to the data includes an interaction between \code{Europe} and \code{political.knowledge}; the other predictors enter the model additively:
+The model fit to the data includes an interaction between \code{Europe} and \code{political.knowledge}, which was the focus of the original research on which this example is based [@AndersenHeathSinnott:2002]; the other predictors enter the model additively:
 ```{r BEPS-model}
-library("nnet")
+library("nnet", quietly=TRUE)
 m.beps <- multinom(vote ~ age + gender + economic.cond.national +
                        economic.cond.household + Blair + Hague + Kennedy +
-                       Europe*political.knowledge, data=BEPS)
-
-car::Anova(m.beps)
-```
-Most of the predictors, including the \code{Europe} $\times$ \code{political.knowledge} interaction, are associated with very small $p$-values; the \code{Anova()} function is from the \pkg{car} package [@FoxWeisberg:2019].
-
-Here's an "effect plot", using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the \code{Europe} $\times$ \code{political.knowledge} interaction in a "stacked-area" graph:
+                       Europe*political.knowledge, data=BEPS, trace=FALSE)
+
+```
+We set the argument \code{trace=FALSE} in the call to \code{multinom()} to suppress reporting the iteration history, which will be particularly appreciated when we repeatedly refit the model in cross-validation.
+
+The \code{Europe} $\times$ \code{political.knowledge} interaction is associated with a very small $p$-value.
+Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the interaction in a "stacked-area" graph:
 ```{r BEPS-plot, fig.width=9, fig.height=5}
+#| fig.cap = "Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study."
 plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
@@
 To cross-validate this multinomial-logit model we need an appropriate cost criterion.
-None of the criteria supplied by the \pkg{cv} package---for example, neither \code{mse()}, which is appropriate for a numeric response, nor \code{BayesRule()}, which is appropriate for a binary response---will do.
-One possibility is to adapt Bayes rule to a polytomous response:
+None of the criteria in the \pkg{cv} package will do---for example, \code{mse()} is appropriate only for a numeric response.
+The \code{BayesRule()} criterion, also supplied by \pkg{cv}, which is for a binary response, comes close:
+```{r BayesRule}
+BayesRule
+```
+After doing some error checking, \code{BayesRule()} rounds the predicted proabability of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction and then reports the proportion of incorrect predictions.
+Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, making possible the computation of bias correction and confidence intervals (as discussion in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+
+It is straightforward to adapt Bayes's rule to a polytomous response:
 ```{r BayesRuleMulti}
@@
 ```
-The \code{predict()} method for \code{"multinom"} models called with argument \code{type="class"} reports the Bayes-rule prediction for each case---that is, the response category with the highest predicted probability.
+The \code{predict()} method for \code{"multinom"} models called with argument \code{type="class"} reports the Bayes's rule prediction for each case---that is, the response category with the highest predicted probability.
 Our \code{BayesRuleMulti()} function calculates the proportion of misclassified cases.
-Because this value is the mean of casewise components, we attach a \code{"casewise loss"} attribute to the result (as explained in the preceding section).
+Because this value is also the mean of casewise components, we attach a \code{"casewise loss"} attribute to the result.
 
@@
 ```
-and so the marginal Bayes-rule prediction, that everyone will vote Labour, produces an error rate of $1 - 0.47213 = 0.52787$.
+and so the marginal Bayes's rule prediction, that everyone will vote Labour, produces an error rate of $1 - 0.47213 = 0.52787$.
 The multinomial-logit model appears to do substantially better than that, but does its performance hold up to cross-validation?
@@
 ```
-That is, we simply call the default \code{cv()} method with the \code{type} argument properly set.
+That is, we simply invoke the default \code{cv()} method via \code{NextMethod()}, with the \code{type} argument properly set.
 In addition to supplying the correct \code{type} argument, our method sets the default \code{criterion} for the \code{cv.multinom()} method to \code{BayesRuleMulti}.
@@
 ```{r BEPS-cv}
-m.beps <- update(m.beps, trace=FALSE)
 cv(m.beps, seed=3465)
 ```
-Prior to invoking \code{cv()}, we called \code{update()} with \code{trace=FALSE} to suppress the iteration history reported by default by \code{multinom()}---it would be tedious to see the iteration history for each fold.
-The cross-validated polytomous Bayes-rule criterion confirms that the fitted model does substantially better than the marginal Bayes-rule prediction that everyone votes for Labour.
+The cross-validated polytomous Bayes's rule criterion confirms that the fitted model does substantially better than the marginal Bayes's rule prediction that everyone votes for Labour.
 
@@
 This approach can break down when one or more hatvalues are equal to 1, in which case the formula for $\widehat{y}_{-i}$ requires division by 0.
-
-To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identify [@Wikipedia-Woodbury:2023],
+In this case, the "training" set omitting the observation with hatvalue = 1 is rank-deficient and the predictors for the left-out case are outside the linear span of the predictors in the training set.
+
+To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identify [@Hager:1989],
 $$

### 2024-01-10 scratch/notes_john/JSS-article/JSS-article-reduced.tex
edit JSS paper

@@
 \Plaintitle{cv: An R Package for Cross-Validation of Regression Models}
-\Shorttitle{\pkg{cv}: Cross-Validation}
+\Shorttitle{\pkg{cv}: Cross-Validation of Regression Models}
 
@@
 
-\hypertarget{cross-validation}{%
-\section{Cross-validation}\label{cross-validation}}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 sample size and thus increases sampling error; and (2), even more
-disconcertingly, particularly in smaller samples, the results can vary
-substantially based on the random division of the data: See \citet[Sec.
+disconcertingly, the results can vary substantially based on the random
+division of the data, particularly in smaller samples: See \citet[Sec.
 5.3]{Harrell:2015} for this and other remarks about data-division and
@@
 fold out in turn. Each fitted model is then used to predict the response
-variable for the cases in the omitted fold. A CV criterion or ``cost''
-measure, such as the mean-squared error (``MSE'') of prediction, is then
-computed using these predicted values. In the extreme \(k = n\), the
-number of cases in the data, thus omitting individual cases and
-refitting the model \(n\) times---a procedure termed ``leave-one-out
-(LOO) cross-validation.''
+variable for the cases in the omitted fold. A CV criterion (also termed
+a ``cost'' or ``loss'' measure), such as the mean-squared error
+(``MSE'') of prediction, is then computed using these predicted values.
+In the extreme \(k = n\), the number of cases in the data, thus omitting
+individual cases and refitting the model \(n\) times---a procedure
+termed ``leave-one-out (LOO) cross-validation.''
 
@@
 employed choices) are somewhat biased but have smaller variance. It is
-also possible to correct \(k\)-fold CV for bias (see below).
-
-\hypertarget{examples}{%
-\section{Examples}\label{examples}}
-
-\hypertarget{polynomial-regression-for-the-auto-data}{%
-\subsection{Polynomial regression for the Auto
-data}\label{polynomial-regression-for-the-auto-data}}
-
-The data for this example are drawn from the \pkg{ISLR2} package for
-\proglang{R}, associated with \citet{JamesEtAl:2021}. The presentation
-here is close (though not identical) to that in the original source
-\citep[ Secs. 5.1, 5.3]{JamesEtAl:2021}, and it demonstrates the use of
-the \texttt{cv()} function in the \pkg{cv} package.\footnote{\citet{JamesEtAl:2021}
+also possible to correct \(k\)-fold CV for bias (see Section
+\ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+
+The \pkg{cv} package for \proglang{R} automates the process of
+cross-validation for standard \proglang{R} statistical model objects.
+The principal function in the packages, also named \code{cv()}, has
+methods for objects produced by a number of commonly employed
+regression-modeling functions, including those for mixed-effects models:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library("cv", quietly=TRUE)
+R> methods("cv")
+\end{CodeInput}
+\begin{CodeOutput}
+[1] cv.default* cv.glm*     cv.glmmTMB* cv.lm*      cv.lme*     cv.merMod* 
+[7] cv.modList* cv.rlm*    
+see '?methods' for accessing help and source code
+\end{CodeOutput}
+\end{CodeChunk}
+
+The \code{"modList"} method for \code{cv()} cross-validates several
+competing models, not necessarily of the same class, using the same
+division of the data into folds. The \code{cv()} function is introduced
+in the context of a preliminary example in Section
+\ref{preliminary-example-polynomial-regression} of the paper.
+
+Cross-validating mixed-effects models involves special considerations
+that we take up in Section \ref{cross-validating-mixed-effects-models}.
+
+The \code{cvSelect()} function, discussed in Section
+\ref{cross-validating-model-selection}, cross-validates a complex
+model-specification process that may, for example, involve choice of
+data transformations and predictors.
+
+The \code{"default"} \code{cv()} method works (perhaps with a bit of
+coaxing) with many other existing regression-model classes for which
+there is an \code{update()} method that accepts a \code{data} argument.
+More generally, the \pkg{cv} package is designed to be extensible, as
+discussed in Section \ref{extending-the-cv-package}.
+
+In the interest of brevity, we won't describe all of the features of the
+\pkg{cv} package here, concentrating on the aspects of the package that
+are relatively novel. For example, the \code{cv()} and \code{cvSelect()}
+functions can perform computations in parallel and can independently
+replicate a cross-validation procedure several times. These and other
+features not discussed in this paper are taken up in the vignettes
+distributed with the package, which also provides greater detail on some
+of topics that we do describe, such as extensions to the package.
+
+\hypertarget{preliminary-example-polynomial-regression}{%
+\section{Preliminary Example: Polynomial
+regression}\label{preliminary-example-polynomial-regression}}
+
+The data for the example in this section are drawn from the \pkg{ISLR2}
+package for \proglang{R}, associated with \citet{JamesEtAl:2021}. The
+presentation here is close (though not identical) to that in the
+original source \citep[ Secs. 5.1, 5.3]{JamesEtAl:2021}, and it
+demonstrates the use of the \texttt{cv()} function.\footnote{\citet{JamesEtAl:2021}
   use the \texttt{cv.glm()} function in the \pkg{boot} package
@@
 
-The \texttt{Auto} dataset contains information about 392 cars:
+The \code{Auto} dataset contains information about 392 cars:
 
@@
 R> data("Auto", package="ISLR2")
-R> head(Auto)
-\end{CodeInput}
-\begin{CodeOutput}
-  mpg cylinders displacement horsepower weight acceleration year origin
-1  18         8          307        130   3504         12.0   70      1
-2  15         8          350        165   3693         11.5   70      1
-3  18         8          318        150   3436         11.0   70      1
-4  16         8          304        150   3433         12.0   70      1
-5  17         8          302        140   3449         10.5   70      1
-6  15         8          429        198   4341         10.0   70      1
-                       name
-1 chevrolet chevelle malibu
-2         buick skylark 320
-3        plymouth satellite
-4             amc rebel sst
-5               ford torino
-6          ford galaxie 500
-\end{CodeOutput}
-\begin{CodeInput}
-R> dim(Auto)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 392   9
+R> summary(Auto)
+\end{CodeInput}
+\begin{CodeOutput}
+      mpg          cylinders      displacement     horsepower        weight    
+ Min.   : 9.00   Min.   :3.000   Min.   : 68.0   Min.   : 46.0   Min.   :1613  
+ 1st Qu.:17.00   1st Qu.:4.000   1st Qu.:105.0   1st Qu.: 75.0   1st Qu.:2225  
+ Median :22.75   Median :4.000   Median :151.0   Median : 93.5   Median :2804  
+ Mean   :23.45   Mean   :5.472   Mean   :194.4   Mean   :104.5   Mean   :2978  
+ 3rd Qu.:29.00   3rd Qu.:8.000   3rd Qu.:275.8   3rd Qu.:126.0   3rd Qu.:3615  
+ Max.   :46.60   Max.   :8.000   Max.   :455.0   Max.   :230.0   Max.   :5140
+
+  acceleration        year           origin                      name    
+ Min.   : 8.00   Min.   :70.00   Min.   :1.000   amc matador       :  5  
+ 1st Qu.:13.78   1st Qu.:73.00   1st Qu.:1.000   ford pinto        :  5  
+ Median :15.50   Median :76.00   Median :1.000   toyota corolla    :  5  
+ Mean   :15.54   Mean   :75.98   Mean   :1.577   amc gremlin       :  4  
+ 3rd Qu.:17.02   3rd Qu.:79.00   3rd Qu.:2.000   amc hornet        :  4  
+ Max.   :24.80   Max.   :82.00   Max.   :3.000   chevrolet chevette:  4  
+                                                 (Other)           :365  
 \end{CodeOutput}
@@
 
-We'll focus here on the relationship of \texttt{mpg} (miles per gallon)
-to \texttt{horsepower}, as displayed in the following scatterplot:
-
-The relationship between the two variables is monotone, decreasing, and
-nonlinear. Following \citet{JamesEtAl:2021}, we'll consider
-approximating the relationship by a polynomial regression, with the
-degree of the polynomial \(p\) ranging from 1 (a linear regression) to
-10.\footnote{Although it serves to illustrate the use of CV, a
-  polynomial is probably not the best choice here. Consider, for example
-  the scatterplot for log-transformed \texttt{mpg} and
-  \texttt{horsepower}, produced by
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
+
+}
+
+\caption[mpg vs horsepower for the Auto data, showing fitted polynomials of degree 1 through 5]{mpg vs horsepower for the Auto data, showing fitted polynomials of degree 1 through 5.}\label{fig:mpg-horsepower-scatterplot-polynomials}
+\end{figure}
+\end{CodeChunk}
+
+We'll focus here on the relationship of \code{mpg} (miles per gallon) to
+\code{horsepower}, as displayed in Figure
+\ref{fig:mpg-horsepower-scatterplot-polynomials}. The relationship
+between the two variables is monotone, decreasing, and nonlinear.
+Following \citet{JamesEtAl:2021}, we'll consider approximating the
+relationship by a polynomial regression, with the degree of the
+polynomial \(p\) ranging from 1 (a linear regression) to 10.\footnote{Although
+  it serves to illustrate the use of CV, a polynomial is not the best
+  choice here. Consider, for example the scatterplot for log-transformed
+  \texttt{mpg} and \texttt{horsepower}, produced by
   \texttt{plot(mpg\ \textasciitilde{}\ horsepower,\ data=Auto,\ log="xy")}
-  (execution of which is left to the reader).} Polynomial fits for
-\(p = 1\) to \(5\) are shown in the following figure:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot(mpg ~ horsepower, data=Auto)
-R> horsepower <- with(Auto, 
-+                    seq(min(horsepower), max(horsepower), 
-+                        length=1000))
-R> for (p in 1:5){
-+   m <- lm(mpg ~ poly(horsepower,p), data=Auto)
-+   mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
-+   lines(horsepower, mpg, col=p + 1, lty=p, lwd=2)
-+ }
-R> legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=2,
-+        title="Degree", inset=0.02)
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
-
-}
-
-\caption[`mpg` vs `horsepower` for the `Auto` data]{`mpg` vs `horsepower` for the `Auto` data}\label{fig:mpg-horsepower-scatterplot-polynomials}
-\end{figure}
-\end{CodeChunk}
-
-The linear fit is clearly inappropriate; the fits for \(p = 2\)
-(quadratic) through \(4\) are very similar; and the fit for \(p = 5\)
-may over-fit the data by chasing one or two relatively high \texttt{mpg}
-values at the right (but see the CV results reported below).
-
-The following graph shows two measures of estimated (squared) error as a
-function of polynomial-regression degree: The mean-squared error
-(``MSE''), defined as
+  (execution of which is left to the reader). We revisit the \code{Auto}
+  data in Section \ref{cross-validating-model-selection}.} Polynomial
+fits for \(p = 1\) to \(5\) are shown in Figure
+\ref{fig:mpg-horsepower-scatterplot-polynomials}. The linear fit is
+clearly inappropriate; the fits for \(p = 2\) (quadratic) through \(4\)
+are very similar; and the fit for \(p = 5\) may over-fit the data by
+chasing one or two relatively high \texttt{mpg} values at the right (but
+see the CV results reported below).
+
+Figure \ref{fig:mpg-horsepower-MSE-se} shows two measures of estimated
+(squared) error as a function of polynomial-regression degree: The
+mean-squared error (``MSE''), defined as
 \(\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\), and
@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> library("cv") # for mse() and other functions
-\end{CodeInput}
-\begin{CodeOutput}
-Loading required package: doParallel
-\end{CodeOutput}
-\begin{CodeOutput}
-Loading required package: foreach
-\end{CodeOutput}
-\begin{CodeOutput}
-Loading required package: iterators
-\end{CodeOutput}
-\begin{CodeOutput}
-Loading required package: parallel
-\end{CodeOutput}
-\begin{CodeInput}
-R> var <- mse <- numeric(10)
-R> for (p in 1:10){
-+   m <- lm(mpg ~ poly(horsepower, p), data=Auto)
-+   mse[p] <- mse(Auto$mpg, fitted(m))
-+   var[p] <- summary(m)$sigma^2
-+ }
-R> 
-R> plot(c(1, 10), range(mse, var), type="n",
-+      xlab="Degree of polynomial, p",
-+      ylab="Estimated Squared Error")
-R> lines(1:10, mse, lwd=2, lty=1, col=2, pch=16, type="b")
-R> lines(1:10, var, lwd=2, lty=2, col=3, pch=17, type="b")
-R> legend("topright", inset=0.02,
-+        legend=c(expression(hat(sigma)^2), "MSE"),
-+        lwd=2, lty=2:1, col=3:2, pch=17:16)
-\end{CodeInput}
 \begin{figure}
@@
 
-The code for this graph uses the \texttt{mse()} function from the
-\pkg{cv} package to compute the MSE for each fit.
-
-\hypertarget{using-cv}{%
-\subsubsection{Using cv()}\label{using-cv}}
-
-The generic \texttt{cv()} function has an \texttt{"lm"} method, which by
-default performs \(k = 10\)-fold CV:
+The generic \texttt{cv()} function has an \texttt{"lm"} method,
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> args(cv:::cv.lm)
+\end{CodeInput}
+\begin{CodeOutput}
+function (model, data = insight::get_data(model), criterion = mse, 
+    k = 10, reps = 1, seed, confint = n >= 400, level = 0.95, 
+    method = c("auto", "hatvalues", "Woodbury", "naive"), ncores = 1, 
+    ...) 
+NULL
+\end{CodeOutput}
+\end{CodeChunk}
+
+which takes the following arguments:
+
+\begin{itemize}
+\item
+  \code{model}, an \code{"lm"} object, the only required argument.
+\item
+  \code{data}, which can usually be inferred from the \code{model}
+  object.
+\item
+  \code{criterion}, a function to compute the CV criterion (defaulting
+  to \code{mse}).
+\item
+  \code{k}, the number of folds to employ (defaulting to \code{10}); the
+  character value \code{"n"} or \code{"loo"} may be supplied to specify
+  leave-one-out cross-validation.
+\item
+  \code{reps}, the number of times to repeat the CV procedure
+  (defaulting to 1).
+\item
+  \code{seed}, the seed for \proglang{R}'s pseudo-random number
+  generator; if not specified a value is randomly selected, reported,
+  and saved, so that the CV procedure is replicable.
+\item
+  \code{confint}, whether or not to compute a confidence interval for
+  the CV criterion, defaulting to \code{TRUE} if there are at least 400
+  cases; a confidence interval is computed only if the CV criterion can
+  be expressed as the average of casewise components (see Section
+  \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}
+  for details).
+\item
+  \code{level}, the level for the confidence interval (defaulting to
+  \code{0.95}).
+\item
+  \code{method}, the computational method to employ: \code{"hatvalues"}
+  is relevant only for LOO CV and bases computation on the hatvalues for
+  the linear model; \code{"Woodbury"} employs the Woodbury matrix to
+  compute the CV criterion with each fold deleted; \code{"naive"}
+  updates the model using the \code{update()} function; and
+  \code{"auto"} selects \code{"hatvalues"} for LOO CV and
+  \code{"Woodbury"} for \(k\)-fold CV, both of which are much more
+  efficient than literally updating the least-squares fit (see below and
+  Section
+  \ref{efficient-computation-for-linear-and-generalized-linear-models}).
+\item
+  \code{ncores}, the number of cores to employ for parallel computation;
+  if \code{cores = 1} (the default), the computations are not
+  parallelized.
+\end{itemize}
+
+To illustrate, we perform 10-fold CV for a quadratic polynomial fit to
+the \code{Auto} data:
 
@@
                       Estimate Std. Error t value Pr(>|t|)    
-(Intercept)            23.4459     0.2209  106.13   <2e-16
-poly(horsepower, 2)1 -120.1377     4.3739  -27.47   <2e-16
-poly(horsepower, 2)2   44.0895     4.3739   10.08   <2e-16
+(Intercept)            23.4459     0.2209  106.13   <2e-16 ***
+poly(horsepower, 2)1 -120.1377     4.3739  -27.47   <2e-16 ***
+poly(horsepower, 2)2   44.0895     4.3739   10.08   <2e-16 ***
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
 
@@
 \begin{CodeInput}
-R> cv(m.auto)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 577350
+R> cv(m.auto, confint=TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+R RNG seed set to 542508
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.19107
-bias-adjusted cross-validation criterion = 19.18016
+cross-validation criterion = 19.57393
+bias-adjusted cross-validation criterion = 19.54281
+95% CI for bias-adjusted CV criterion = (16.03528, 23.05034)
 full-sample criterion = 18.98477 
@@
 
-The \code{"lm"} method by default uses \code{mse()} as the CV criterion
-and the Woodbury matrix identity to update the regression with each fold
-deleted without having literally to refit the model. Computational
-details are discussed in the final section of this vignette. The
-function reports the CV estimate of MSE, a biased-adjusted estimate of
-the MSE (the bias adjustment is explained in the final section), and the
-MSE is also computed for the original, full-sample regression. Because
-the division of the data into 10 folds is random, \code{cv()} explicitly
-(randomly) generates and saves a seed for \proglang{R}'s pseudo-random
-number generator, to make the results replicable. The user can also
-specify the seed directly via the \code{seed} argument to \code{cv()}.
-
-To perform LOO CV, we can set the \code{k} argument to \code{cv()} to
-the number of cases in the data, here \code{k=392}, or, more
-conveniently, to \code{k="loo"} or \code{k="n"}:
+The function reports the CV estimate of MSE, a bias-adjusted estimate of
+the MSE (the bias adjustment is explained in Section
+\ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}),
+and the MSE is also computed for the original, full-sample regression.
+Because the number of cases \(n = 392 < 400\) for the \code{Auto} data,
+we set the argument \code{confint=TRUE} to obtain a confidence interval
+for the MSE, which proves to be quite wide.
+
+To perform LOO CV:
 
@@
 
-For LOO CV of a linear model, \code{cv()} by default uses the hatvalues
-from the model fit to the full data for the LOO updates, and reports
-only the CV estimate of MSE. Alternative methods are to use the Woodbury
-matrix identity or the ``naive'' approach of literally refitting the
-model with each case omitted. All three methods produce exact results
-for a linear model (within the precision of floating-point
-computations):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(m.auto, k="loo", method="naive")
+The \code{"hatvalues"} method reports only the CV estimate of MSE.
+Alternative methods are to use the Woodbury matrix identity or the
+``naive'' approach of literally refitting the model with each case
+omitted. All three methods produce exact results for a linear model
+(within the precision of floating-point computations):
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> cv(m.auto, k="loo", method="naive", confint=TRUE)
 \end{CodeInput}
@@
 n-Fold Cross Validation
-method: naive
 criterion: mse
@@
 bias-adjusted cross-validation criterion = 19.24787
+95% CI for bias-adjusted CV criterion = (15.77884, 22.71691)
 full-sample criterion = 18.98477 
@@
 \begin{CodeInput}
-R> cv(m.auto, k="loo", method="Woodbury")
+R> cv(m.auto, k="loo", method="Woodbury", confint=TRUE)
 \end{CodeInput}
@@
 The \code{"naive"} and \code{"Woodbury"} methods also return the
-bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an
-issue for LOO CV.
+bias-adjusted estimate of MSE (and a confidence interval around it)
+along with the full-sample MSE, but bias isn't an issue for LOO CV.
 
@@
 approaches are essentially instantaneous, but it is still of interest to
-investigate their relative speed. In this comparison, we include the
-\code{cv.glm()} function from the \pkg{boot} package, which takes the
-naive approach, and for which we have to fit the linear model as an
-equivalent Gaussian GLM. We use the \code{microbenchmark()} function
-from the package of the same name for the timings \citep{Mersmann:2023}:
+investigate their relative speed. In the following comparison, we
+include the \code{cv.glm()} function from the \pkg{boot} package, which
+takes the naive approach, and for which we have to fit the linear model
+as an equivalent Gaussian GLM. We use the \code{microbenchmark()}
+function from the package of the same name \citep{Mersmann:2023} for the
+timings:
 
@@
   only 10 times rather than the default 100. With the exception of the
-  last columns, the output is self-explanatory. The last column shows
+  last column, the output is self-explanatory. The last column shows
   which methods have average timings that are statistically
@@
 \hypertarget{comparing-competing-models}{%
-\subsubsection{Comparing competing
+\subsection{Comparing competing
 models}\label{comparing-competing-models}}
@@
 We illustrate with the polynomial regression models of varying degree
-for the \code{Auto} data (discussed previously), beginning by fitting
-and saving the 10 models:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> for (p in 1:10){
-+   assign(paste0("m.", p),
-+          lm(mpg ~ poly(horsepower, p), data=Auto))
-+ }
-R> objects(pattern="m\\.[0-9]")
-\end{CodeInput}
-\begin{CodeOutput}
- [1] "m.1"  "m.10" "m.2"  "m.3"  "m.4"  "m.5"  "m.6"  "m.7"  "m.8"  "m.9" 
-\end{CodeOutput}
-\begin{CodeInput}
-R> summary(m.2) # for example, the quadratic fit
-\end{CodeInput}
-\begin{CodeOutput}
+for the \code{Auto} data, beginning by fitting and saving the 10 models:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> mlist <- vector(10, mode="list")
+R> for (p in 1:10) mlist[[p]] <- lm(mpg ~ poly(horsepower, p), data = Auto)
+R> names(mlist) <- paste0("m.", 1:10)
+R> mlist[2] # e.g., the quadratic fit
+\end{CodeInput}
+\begin{CodeOutput}
+$m.2
 
@@
 
-Residuals:
-     Min       1Q   Median       3Q      Max 
--14.7135  -2.5943  -0.0859   2.2868  15.8961
-
 Coefficients:
-                      Estimate Std. Error t value Pr(>|t|)
-(Intercept)            23.4459     0.2209  106.13   <2e-16
-poly(horsepower, p)1 -120.1377     4.3739  -27.47   <2e-16
-poly(horsepower, p)2   44.0895     4.3739   10.08   <2e-16
-
-Residual standard error: 4.374 on 389 degrees of freedom
-Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
-F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
+         (Intercept)  poly(horsepower, p)1  poly(horsepower, p)2  
+               23.45               -120.14                 44.09  
 \end{CodeOutput}
@@
 R> # 10-fold CV
-R> cv.auto.10 <- cv(models(m.1, m.2, m.3, m.4, m.5,
-+                      m.6, m.7, m.8, m.9, m.10),
-+               data=Auto, seed=2120)
-R> cv.auto.10[1:2] # for the linear and quadratic models
-\end{CodeInput}
-\begin{CodeOutput}
-
-Model model.1:
-10-Fold Cross Validation
-method: Woodbury
-cross-validation criterion = 24.24642
-bias-adjusted cross-validation criterion = 24.23039
-full-sample criterion = 23.94366
-
-Model model.2:
+R> mlist <- do.call(models, mlist) # create "modList" object
+R> 
+R> cv.auto.10 <- cv(mlist, data=Auto, seed=2120)
+R> cv.auto.10[2] # e.g., for quadratic model
+\end{CodeInput}
+\begin{CodeOutput}
+
+Model m.2:
 10-Fold Cross Validation
@@
 R> # LOO CV
-R> cv.auto.loo <- cv(models(m.1, m.2, m.3, m.4, m.5,
-+                         m.6, m.7, m.8, m.9, m.10),
-+                  data=Auto, k="loo")
-R> cv.auto.loo[1:2] # linear and quadratic models
-\end{CodeInput}
-\begin{CodeOutput}
-
-Model model.1:
-n-Fold Cross Validation
-method: hatvalues
-cross-validation criterion = 24.23151
-Model model.2:
+R> cv.auto.loo <- cv(mlist, data=Auto, k="loo")
+R> cv.auto.loo[2] # e.g., for quadratic model
+\end{CodeInput}
+\begin{CodeOutput}
+
+Model m.2:
 n-Fold Cross Validation
@@
 
-Because we didn't supply names for the models in the calls to the
-\code{models()} function, the names \code{model.1}, \code{model.2},
-etc., are generated by the function.
-
-Alternatively, we can use the \code{plot()} method for
-\code{"cvModList"} objects to compare the models, though with separate
-graphs for 10-fold and LOO CV:
+The \code{models()} function takes an arbitrary number of regression
+models as its arguments, which are optionally named, to create a
+\code{"modList"} object. Because we generated the polynomial regression
+models in a named list, we conveniently employ \code{do.call()} to
+supply the models as arguments to \code{models()}. Note that the names
+created for the list (e.g., \code{"m.2"}) are then used for the models.
+We can also invoke the \code{plot()} method for \code{"cvModList"}
+objects to compare the models (see Figure
+\ref{fig:polynomial-regression-CV-graph-2}):
 
@@
 
-\caption[Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$]{Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$}\label{fig:polynomial-regression-CV-graph-2}
+\caption[Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$]{Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$.}\label{fig:polynomial-regression-CV-graph-2}
 \end{figure}
@@
 and also results that are similar to those produced by the estimated
-error variance \(\widehat{\sigma}^2\) for each model, reported above
-(except for the highest-degree polynomials, where the CV results more
-clearly suggest over-fitting).
+error variance \(\widehat{\sigma}^2\) for each model (cf., Figure
+\ref{fig:mpg-horsepower-MSE-se} on page
+\pageref{fig:mpg-horsepower-MSE-se}), except for the highest-degree
+polynomials, where the CV results more clearly suggest over-fitting.
 
@@
 predictors for those new cases. As we explained, the application of this
-idea to independently sampled cases is straightforward---simply
-partition the data into random folds of equal size and leave each fold
-out in turn, or, in the case of LOO CV, simply omit each case in turn.
+idea to independently sampled cases is straightforward.
 
 In contrast, mixed-effects models are fit to \emph{dependent} data, in
-which cases as clustered, such as hierarchical data, where the clusters
+which cases are clustered, such as hierarchical data, where the clusters
 comprise higher-level units (e.g., students clustered in schools), or
-longitudinal data, where the clusters are individuals and the cases
+longitudinal data, where the clusters are individuals and the cases are
 repeated observations on the individuals over time.\footnote{There are,
@@
   nested and crossed random effects. We present an example of the latter
-  later in the vignette.}
+  in a vignette for the \pkg{cv} package.}
 
@@
   comprise one or more individual cases, and the predictions can use
-  both the fixed and random effects.
+  both the fixed and random effects---so-called ``best-linear-unbiased
+  predictors'' or ``BLUPs.''
 \end{enumerate}
@@
 \code{MathAchSchool} data sets in the \pkg{nlme} package
-\citep{PinheiroBates:2000}. \code{MathAchieve} includes individual-level
-data on 7185 students in 160 high schools, and \code{MathAchSchool}
-includes school-level data:
+\citep{PinheiroBates:2000}. \code{MathAchieve} comprises
+individual-level data on 7185 students in 160 high schools, and
+\code{MathAchSchool} contains school-level data:
 
@@
 Some data-management is required before fitting a mixed-effects model to
-the HSB data, for which we use the \pkg{dplyr} package
-\citep{WickhamEtAl:2023}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("dplyr")
-\end{CodeInput}
-\begin{CodeOutput}
-
-Attaching package: 'dplyr'
-\end{CodeOutput}
-\begin{CodeOutput}
-The following objects are masked from 'package:stats':
-
-    filter, lag
-\end{CodeOutput}
-\begin{CodeOutput}
-The following objects are masked from 'package:base':
-
-    intersect, setdiff, setequal, union
-\end{CodeOutput}
-\begin{CodeInput}
-R> MathAchieve %>% group_by(School) %>%
-+   summarize(mean.ses = mean(SES)) -> Temp
-R> Temp <- merge(MathAchSchool, Temp, by="School")
-R> HSB <- merge(Temp[, c("School", "Sector", "mean.ses")],
-+              MathAchieve[, c("School", "SES", "MathAch")], by="School")
+the HSB data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> HSB <- MathAchieve
+R> HSB <- merge(MathAchSchool[, c("School", "Sector")],
++              HSB[, c("School", "SES", "MathAch")], by="School")
 R> names(HSB) <- tolower(names(HSB))
-R> 
-R> HSB$cses <- with(HSB, ses - mean.ses)
-\end{CodeInput}
-\end{CodeChunk}
-
-In the process, we created two new school-level variables:
-\code{meanses}, which is the average SES for students in each school;
-and \code{cses}, which is school-average SES centered at its mean. For
-details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.
+R> HSB <- within(HSB, {
++   mean.ses <- ave(ses, school)
++   cses <- ses - mean.ses
++ })
+\end{CodeInput}
+\end{CodeChunk}
+
+In the process, we merged variables from the school-level and
+student-level data sets and created two new school-level variables:
+\code{mean.ses}, which is the average SES for students in each school;
+and \code{cses}, which is each students' SES centered at their school
+means. For details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.
 
@@
 \begin{CodeInput}
-R> library("lme4")
-\end{CodeInput}
-\begin{CodeOutput}
-Loading required package: Matrix
-\end{CodeOutput}
-\begin{CodeInput}
+R> library("lme4", quietly=TRUE)
 R> hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
@@
 with a fold removed failed to converge. Convergence problems are common
-in mixed-effects modeling. The apparent issue here is that an estimated
-variance component is close to or equal to 0, which is at a boundary of
-the parameter space. That shouldn't disqualify the fitted model for the
-kind of prediction required for cross-validation.
-
-There is also a \code{cv()} method for linear mixed models fit by the
-\code{lme()} function in the \pkg{nlme} package, and the arguments for
-\code{cv()} in this case are the same as for a model fit by
-\code{lmer()} or \code{glmer()}. We illustrate with the mixed model fit
-to the HSB data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library(nlme)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Attaching package: 'nlme'
-\end{CodeOutput}
-\begin{CodeOutput}
-The following object is masked from 'package:lme4':
-
-    lmList
-\end{CodeOutput}
-\begin{CodeOutput}
-The following object is masked from 'package:dplyr':
-
-    collapse
-\end{CodeOutput}
-\begin{CodeInput}
-R> hsb.lme <- lme(mathach ~ mean.ses*cses + sector*cses,
-+                  random = ~ cses | school, data=HSB,
-+                control=list(opt="optim"))
-R> summary(hsb.lme)
-\end{CodeInput}
-\begin{CodeOutput}
-Linear mixed-effects model fit by REML
-  Data: HSB 
-       AIC      BIC    logLik
-  46524.78 46593.57 -23252.39
-
-Random effects:
- Formula: ~cses | school
- Structure: General positive-definite, Log-Cholesky parametrization
-            StdDev     Corr  
-(Intercept) 1.54117685 (Intr)
-cses        0.01817364 0.006 
-Residual    6.06349216
-
-Fixed effects:  mathach ~ mean.ses * cses + sector * cses 
-                        Value Std.Error   DF  t-value p-value
-(Intercept)         12.128207 0.1991964 7022 60.88567   0e+00
-mean.ses             5.336665 0.3689784  157 14.46335   0e+00
-cses                 2.942145 0.1512240 7022 19.45554   0e+00
-sectorCatholic       1.224531 0.3061139  157  4.00025   1e-04
-mean.ses:cses        1.044406 0.2910747 7022  3.58810   3e-04
-cses:sectorCatholic -1.642148 0.2331162 7022 -7.04433   0e+00
- Correlation: 
-                    (Intr) men.ss cses   sctrCt mn.ss:
-mean.ses             0.256                            
-cses                 0.000  0.000                     
-sectorCatholic      -0.699 -0.356  0.000              
-mean.ses:cses        0.000  0.000  0.295  0.000       
-cses:sectorCatholic  0.000  0.000 -0.696  0.000 -0.351
-
-Standardized Within-Group Residuals:
-        Min          Q1         Med          Q3         Max 
--3.17010624 -0.72487654  0.01489162  0.75426269  2.96549829
-
-Number of Observations: 7185
-Number of Groups: 160 
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(hsb.lme, k=10, clusterVariables="school", seed=5240)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 5240
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation based on 160 {school} clusters
-cross-validation criterion = 39.1569
-bias-adjusted cross-validation criterion = 39.14881
-95% CI for bias-adjusted CV criterion = (38.06591, 40.23171)
-full-sample criterion = 39.0062 
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(hsb.lme, seed=1575)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 1575
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-cross-validation criterion = 37.44163
-bias-adjusted cross-validation criterion = 37.40222
-95% CI for bias-adjusted CV criterion = (36.35136, 38.45309)
-full-sample criterion = 36.14707 
-\end{CodeOutput}
-\end{CodeChunk}
-
-We used the same random-number generator seeds as in the previous
-example cross-validating the model fit by \code{lmer()}, and so the same
-folds are employed in both cases.\footnote{The observant reader will
-  notice that we set the argument \code{control=list(opt="optim")} in
-  the call to \code{lme()}, changing the optimizer employed from the
-  default \code{"nlminb"}. We did this because with the default
-  optimizer, \code{lme()} encountered the same convergence issue as
-  \code{lmer()}, but rather than issuing a warning, \code{lme()} failed,
-  reporting an error. As it turns out, setting the optimizer to
-  \code{"optim"} avoids this problem.} The estimated covariance
-components and fixed effects in the summary output differ slightly
-between the \code{lmer()} and \code{lme()} solutions, although both
-functions seek to maximize the REML criterion. This is, of course, to be
-expected when different algorithms are used for numerical optimization.
-To the precision reported, the cluster-level CV results for the
-\code{lmer()} and \code{lme()} models are identical, while the
-case-level CV results are very similar but not identical.
+in mixed-effects modeling. The issue here is that an estimated variance
+component is close to or equal to 0, which is at a boundary of the
+parameter space. That shouldn't disqualify the fitted model for the kind
+of prediction required for cross-validation.
+
+\code{cv()} also has methods for mixed models fit by the \code{glmer()}
+function in the \pkg{lme4} package, the \code{lme()} function in the
+\pkg{nlme} package \citep{PinheiroBates:2000}, and the \code{glmmTMB()}
+function in the \pkg{glmmTMB} package \citep{BrooksEtAl}, along with a
+simple procedure for extending \code{cv()} to other classes of
+mixed-effects models. See the vignettes in the \pkg{cv} package for
+details.
 
@@
 
-We introduce an artificial data set that exemplifies aspects of
-cross-validation particular to hierarchical models. Using this data set,
-we show that model comparisons employing cluster-based and those
-employing case-based cross-validation may not agree on a ``best'' model.
-Furthermore, commonly used measures of fit, such as mean-squared error,
-do not necessarily become smaller as models become larger, even when the
-models are nested, and even when the measure of fit is computed for the
-whole data set.
-
-Consider a researcher studying improvement in a skill, yodeling, for
-example, among students enrolled in a four-year yodeling program. The
-plan is to measure each student's skill level at the beginning of the
-program and every year thereafter until the end of the program,
-resulting in five annual measurements for each student. It turns out
-that yodeling appeals to students of all ages, and students enrolling in
-the program range in age from 20 to 70. Moreover, participants'
-untrained yodeling skill is similar at all ages, as is their rate of
+In this section, we introduce an artificial data set that exemplifies
+aspects of cross-validation particular to hierarchical models. Using
+this data set, we show that model comparisons employing cluster-based
+and those employing case-based cross-validation may not agree on a
+``best'' model. Furthermore, commonly used measures of fit, such as
+mean-squared error, do not necessarily become smaller as models become
+larger, even when the models are nested, and even when the measure of
+fit is computed for the whole data set.
+
+Consider a researcher studying improvement in a skill, singing, for
+example, among students enrolled in a four-year voice program at a music
+conservatory. The plan is to measure each student's skill level at the
+beginning of the program and every year thereafter until the end of the
+program, resulting in 5 annual measurements for each student. It turns
+out that singing appeals to students of all ages, and students enrolling
+in the program range in age from 20 to 70. Moreover, participants'
+untrained singing skill is similar at all ages, as is their rate of
 progress with training. All students complete the four-year program.
 
-The researcher, who has more expertise in yodeling than in modeling,
-decides to model the response, \(y\), yodeling skill, as a function of
+The researcher, who has more expertise in singing than in modeling,
+decides to model the response, \(y\), singing skill, as a function of
 age, \(x\), reasoning that students get older during their stay in the
@@
 +   x = Ax + rep(1:Nx, length.out = Nw*Nb)
-+ ) |>
-+   within(
-+     {
++ ) |> within ({
 +   xm  <- ave(x, group, FUN = mean) # within-group mean
@@
 +     rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
-+     }
-+   )
-\end{CodeInput}
-\end{CodeChunk}
-
-Here is a scatterplot of the data for a representative group of 10
-(without loss of generality, the first 10) of 100 students, showing the
-95\% concentration ellipse for each cluster:\footnote{We find it
-  convenient to use the \pkg{lattice} \citep{Sarkar:2008} and
-  \pkg{latticeExtra} \citep{SarkarAndrews:2022} packages for this and
-  other graphs in this section.}
++ })
+\end{CodeInput}
+\end{CodeChunk}
+
+Figure \ref{fig:plot1} (a) shows a scatterplot the data for a
+representative group of 10 (without loss of generality, the first 10) of
+the 100 students, displaying the 95\% concentration ellipse for each
+cluster:
 
@@
 1. Due to the large variation in ages between students, the
-least-squares regression of yodeling skill on age (for the 500
+least-squares regression of singing skill on age (for the 500
 observations among all 100 students) produces an estimated slope close
@@
             Estimate Std. Error t value Pr(>|t|)    
-(Intercept) 9.050430   0.347189  26.068  < 2e-16
-x           0.020908   0.007273   2.875  0.00422
+(Intercept) 9.050430   0.347189  26.068  < 2e-16 ***
+x           0.020908   0.007273   2.875  0.00422 ** 
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
 
@@
 
-We will shortly consider three other, more complex, mixed models;
-because of data-management considerations, it is convenient to fit them
-now, but we defer discussion of these models:
+We will shortly consider three other, more complex, mixed models.
+
+We proceed to obtain predictions from the random-intercept model
+(\code{mod.0}) for fixed effects alone, as would be used for
+cross-validation based on clusters (i.e., students), and for fixed and
+random effects---the BLUPs---as would be used for cross-validation based
+on cases (i.e., occasions within students). Predictions for the
+random-intercept model, \code{mod.0}, for the first 10 students are
+shown in Figure \ref{fig:plot1} (b). The fixed-effect predictions for
+the various individuals are identical---the estimated fixed-effects
+intercept or estimated general mean of \(y\)---while the BLUPs are the
+sums of the fixed-effects intercept and the random intercepts, and are
+only slightly shrunken towards the general mean. Because in our
+artificial data there is no population relationship between age and
+skill, the fixed-effect-only predictions and the BLUPs are not very
+different.
+
+Our next model, \code{mod.1}, includes a fixed intercept and fixed
+effect of \code{x} along with a random intercept:
 
@@
 R> mod.1 <- lmer(y ~ x + (1 | group), Data)
-R> 
-R> # effect of x, contextual (student) mean of x, and random intercept:
-R> mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
-R>         # equivalent to y ~ I(x - xm) + xm + (1 | group)
-R> 
-R> # model generating the data (where Bb = 0)
-R> mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
-\end{CodeInput}
-\end{CodeChunk}
-
-We proceed to obtain predictions from the random-intercept model
-(\code{mod.0}) and the other models (\code{mod.1}, \code{mod.2}, and
-\code{mod.3}) based on fixed effects alone, as would be used for
-cross-validation based on clusters (i.e., students), and for fixed and
-random effects---so-called best linear unbiased predictions or
-BLUPs---as would be used for cross-validation based on cases (i.e.,
-occasions within students):
-
-We then prepare the data for plotting:
-
-Predictions based on the random-intercept model \code{mod.0} for the
-first 10 students are shown in the following graph:
-
-The fixed-effect predictions for the various individuals are
-identical---the estimated fixed-effects intercept or estimated general
-mean of \(y\)---while the BLUPs are the sums of the fixed-effects
-intercept and the random intercepts, and are only slightly shrunken
-towards the general mean. Because in our artificial data there is no
-population relationship between age and skill, the fixed-effect-only
-predictions and the BLUPs are not very different.
-
-Our next model, \code{mod.1}, includes a fixed intercept and fixed
-effect of \code{x} along with a random intercept:
-
-\begin{CodeChunk}
-\begin{CodeInput}
 R> summary(mod.1)
@@
 
-Predictions from this model appear in the following graph:
-
-The BLUPs fit the observed data very closely, but predictions based on
-the fixed effects alone, with a common intercept and slope for all
-clusters, are very poor---indeed, much worse than the fixed-effects-only
+Predictions from this model appear in Figure \ref{fig:plot1} (c). The
+BLUPs fit the observed data very closely, but predictions based on the
+fixed effects alone, with a common intercept and slope for all clusters,
+are very poor---indeed, much worse than the fixed-effects-only
 predictions based on the simpler random-intercept model, \code{mod.0}.
@@
 
-Our third model, \code{mod.2}, includes the contextual effect of
-\(x\)---that is, the cluster mean \textbackslash code\{xm{]}---along
-with \(x\) and the intercept in the fixed-effect part of the model, and
-a random intercept:
-
-\begin{CodeChunk}
-\begin{CodeInput}
+Our third model, \code{mod.2}, includes the ``contextual effect'' of
+\(x\)---that is, the cluster mean \code{xm}---along with \(x\) and the
+intercept in the fixed-effect part of the model, and a random intercept:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # effect of x, contextual (student) mean of x, and random intercept:
+R> mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
+R>         # equivalent to y ~ I(x - xm) + xm + (1 | group)
 R> summary(mod.2)
@@
 generated the data once the coefficient of the contextual predictor
-\code{xm} is set to 0 (as it is in \textbackslash code\{mod.3{]},
-discussed below).
-
-Predictions from model \code{mod.2} appear in the following graph:
-
+\code{xm} is set to 0 (as it is in \code{mod.3}, discussed below).
+
+Predictions from \code{mod.2} appear in Figure \ref{fig:plot1} (d).
 Depending on the estimated variance parameters of the model, a mixed
@@
 the parallel fitted lines for the individual students. In our contrived
-data, the \code{mod.2} applies little shrinkage, allowing substantial
+data, \code{mod.2} applies little shrinkage, allowing substantial
 variability in the heights of the fitted lines, which closely approach
@@
 
-The mixed model \code{mod.2} therefore fits individual observations
+The mixed model \code{mod.2} therefore fits the individual observations
 well, and we anticipate a favorable assessment using individual-based
@@
 alone, and so we expect that cluster-based cross-validation won't show
-an advantage for model \code{mod.2} compared to the smaller model
-\code{mod.0}, which includes only fixed and random intercepts.
+an advantage for \code{mod.2} compared to the smaller \code{mod.0},
+which includes only fixed and random intercepts.
 
@@
 \begin{CodeInput}
+R> # model generating the data (where Bb = 0)
+R> mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
 R> summary(mod.3)
@@
 
-The predictions from \code{mod.3} are therefore similar to those from
-\code{mod.2}:
+The predictions from \code{mod.3}, shown in Figure \ref{fig:plot1} (e),
+are therefore similar to those from \code{mod.2}.
 
 We next carry out case-based cross-validation, which, as we have
-explained, is based on both fixed and predicted random effects (i.e.,
-BLUPs), and cluster-based cross-validation, which is based on fixed
-effects only. In order to reduce between-model random variability in
-comparisons of models, we apply \code{cv()} to the list of models
-created by the \code{models()} function (introduced previously),
-performing cross-validation with the same folds for each model:
+explained, includes both fixed and predicted random effects (i.e.,
+BLUPs), and cluster-based cross-validation, which includes fixed effects
+only. In order to reduce between-model random variability in comparisons
+of models, we apply \code{cv()} to the list of models created by the
+\code{models()} function (introduced previously), performing
+cross-validation with the same folds for each model (see Figure
+\ref{fig:cross-validation-clusters}):
 
@@
 
-In summary, model \code{mod.1}, with \(x\) alone and without the
-contextual mean of \(x\), is assessed as fitting very poorly by
-cluster-based CV, but relatively much better by case-based CV. Model
-\code{mod.2}, which includes both \(x\) and its contextual mean,
-produces better results using both cluster-based and case-based CV. The
-data-generating model, \code{mod.3}, which includes the fixed effect of
-\code{x - xm} in place of separate terms in \code{x} and \code{xm},
-isn't distinguishable from model \code{mod.2}, which includes \code{x}
-and \code{xm} separately, even though \code{mod.2} has an unnecessary
-parameter (recall that the population coefficient of \code{xm} is 0 when
-\code{x} is expressed as deviations from the contextual mean). These
-conclusions are consistent with our observations based on graphing
-predictions from the various models, and they illustrate the
-desirability of assessing mixed-effect models at different hierarchical
-levels.
+In summary, \code{mod.1}, with \(x\) alone and without the contextual
+mean of \(x\), is assessed as fitting very poorly by cluster-based CV,
+but relatively much better by case-based CV. \code{mod.2}, which
+includes both \(x\) and its contextual mean, produces better results
+using both cluster-based and case-based CV. The data-generating model,
+\code{mod.3}, which includes the fixed effect of \code{x - xm} in place
+of separate terms in \code{x} and \code{xm}, isn't distinguishable from
+\code{mod.2}, which includes \code{x} and \code{xm} separately, even
+though \code{mod.2} has an unnecessary parameter (recall that the
+population coefficient of \code{xm} is 0 when \code{x} is expressed as
+deviations from the contextual mean). These conclusions are consistent
+with our observations based on graphing predictions from the various
+models in Figure \ref{fig:plot1} (on page \pageref{fig:plot1}), and they
+illustrate the desirability of assessing mixed-effect models at
+different hierarchical levels.
 
@@
 
+As \citet[Sec. 7.10.2: ``The Wrong and Right Way to Do
+Cross-validation'']{HastieTibshiraniFriedman:2009} explain, if the whole
+data are used to select or fine-tune a statistical model, then
+subsequent cross-validation of the model is intrinsically misleading,
+because the model is selected to fit the whole data, including the part
+of the data that remains when each fold is removed.
+
 \hypertarget{a-preliminary-example}{%
 \subsection{A preliminary example}\label{a-preliminary-example}}
-
-As \citet[Sec. 7.10.2: ``The Wrong and Right Way to Do
-Cross-validation'']{HastieTibshiraniFriedman:2009} explain, if the whole
-data are used to select or fine-tune a statistical model, subsequent
-cross-validation of the model is intrinsically misleading, because the
-model is selected to fit the whole data, including the part of the data
-that remains when each fold is removed.
 
@@
 \(n = 1000\) independent observations for a response variable variable
-\(y \sim N(\mu = 10, \sigma^2 = 0)\), and independently sample \(1000\)
+\(y \sim N(\mu = 10, \sigma^2 = 1)\), and independently sample \(1000\)
 observations for \(p = 100\) ``predictors,'' \(x_1, \ldots, x_{100}\),
@@
 \(y_i = \alpha + \beta_1 x_{i1} + \cdots + \beta_{100} x_{i,100} + \varepsilon_i\)
-has \(\alpha = 10\) and all \(\beta_j = 0\).
+has \(\alpha = 10\), all \(\beta_j = 0\), and
+\(\sigma_\varepsilon^2 = 1\).
 
@@
 + )
-R> head(D[, 1:6])
-\end{CodeInput}
-\begin{CodeOutput}
-          y        X.1        X.2         X.3         X.4         X.5
-1 10.031647 -1.2388628 -0.2648705 -0.03539048 -2.57697337  0.81104761
-2  9.664989  0.1228689 -0.1774440  0.37290421 -0.93513788  0.62867324
-3 10.023249 -0.9505172 -0.7348667 -1.05978180  0.88294443  0.02391808
-4  8.990969  1.1357103  0.3241085  0.11036901  1.37630285 -0.42211426
-5  9.071249  1.4947403  1.8753802  0.10574793  0.29213991 -0.18456833
-6 11.349283 -0.1845331 -0.7803709 -1.23803778 -0.01094861  0.69103395
-\end{CodeOutput}
+\end{CodeInput}
 \end{CodeChunk}
@@
 R> m.null <- lm(y ~ 1, data=D)
+R> summary(m.null)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Call:
+lm(formula = y ~ 1, data = D)
+
+Residuals:
+    Min      1Q  Median      3Q     Max 
+-3.4585 -0.6809  0.0190  0.6365  2.9346
+
+Coefficients:
+            Estimate Std. Error t value Pr(>|t|)    
+(Intercept)  9.93704    0.03122   318.3   <2e-16 ***
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+Residual standard error: 0.9872 on 999 degrees of freedom
+\end{CodeOutput}
+\begin{CodeInput}
 R> anova(m.null, m.full)
@@
 \end{CodeOutput}
-\begin{CodeInput}
-R> summary(m.null)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-lm(formula = y ~ 1, data = D)
-
-Residuals:
-    Min      1Q  Median      3Q     Max 
--3.4585 -0.6809  0.0190  0.6365  2.9346
-
-Coefficients:
-            Estimate Std. Error t value Pr(>|t|)
-(Intercept)  9.93704    0.03122   318.3   <2e-16
-
-Residual standard error: 0.9872 on 999 degrees of freedom
-\end{CodeOutput}
 \end{CodeChunk}
@@
 regression to select a ``best'' model, starting with the null model, and
-using AIC as the model-selection criterion (see the help page for
-\code{stepAIC()} for details):\footnote{It's generally advantageous to
-  start with the largest model, here the one with 100 predictors, and
-  proceed by backward elimination. In this demonstration, however, where
-  all of the \(\beta\)s are really 0, the selected model will be small,
-  and so we proceed by forward selection from the null model to save
-  computing time.}
+using AIC as the model-selection criterion (see \code{?stepAIC()} for
+details):\footnote{It's generally advantageous to start with the largest
+  model, here the one with 100 predictors, and proceed by backward
+  elimination. In this demonstration, however, where all of the
+  \(\beta\)s are really 0, the selected model will be small, and so we
+  proceed by forward selection from the null model to save computing
+  time.}
 
@@
             Estimate Std. Error t value Pr(>|t|)    
-(Intercept)  9.93716    0.03098 320.803  < 2e-16
-X.99        -0.09103    0.03082  -2.953  0.00322
-X.90        -0.08205    0.03135  -2.617  0.00901
-X.87        -0.06942    0.03105  -2.235  0.02561
+(Intercept)  9.93716    0.03098 320.803  < 2e-16 ***
+X.99        -0.09103    0.03082  -2.953  0.00322 ** 
+X.90        -0.08205    0.03135  -2.617  0.00901 ** 
+X.87        -0.06942    0.03105  -2.235  0.02561 *  
 X.40        -0.04759    0.03076  -1.547  0.12211    
-X.65        -0.05523    0.03147  -1.755  0.07952
-X.91         0.05245    0.03084   1.700  0.08937
+X.65        -0.05523    0.03147  -1.755  0.07952 .  
+X.91         0.05245    0.03084   1.700  0.08937 .  
 X.53        -0.04921    0.03048  -1.615  0.10672    
-X.45         0.05543    0.03182   1.742  0.08183
+X.45         0.05543    0.03182   1.742  0.08183 .  
 X.31         0.04525    0.03108   1.456  0.14570    
-X.56         0.05433    0.03273   1.660  0.09723
+X.56         0.05433    0.03273   1.660  0.09723 .  
 X.61        -0.05085    0.03170  -1.604  0.10908    
@@
 X.92         0.04430    0.03100   1.429  0.15329    
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
 
@@
 variance \(\sigma^2 = 1\), as is the estimated error variance for the
-selected model, \(\widehat{\sigma}^2 = 0.973^2 = 0.947\).
+selected model, \(\widehat{\sigma}^2 = 0.973^2 = 0.947\).\footnote{The
+  \code{"casewise loss"} attribute attached to the value returned by the
+  \code{mse()} function signals to \code{cv()} that MSE is an average of
+  casewise components (and how to calculate the casewise components),
+  which has implications for computing bias correction and confidences
+  intervals; see Section
+  \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}.}
 
@@
 
-The \code{cvSelect()} function in the \pkg{cv} package allows us to
-cross-validate the whole model-selection procedure. The first argument
-to \code{cvSelect()} is a model-selection function capable of refitting
-the model with a fold omitted and returning a CV criterion. The
-\code{selectStepAIC()} function, also in \pkg{cv} and based on
-\code{stepAIC()}, is suitable for use with \code{cvSelect()}:
+The \code{cvSelect()} function in the \pkg{cv} package, in contrast,
+allows us to cross-validate the whole model-selection procedure. The
+first argument to \code{cvSelect()} is a model-selection function
+capable of refitting the model with a fold omitted and returning a CV
+criterion. The \code{selectStepAIC()} function, also in \pkg{cv} and
+based on \code{stepAIC()}, is suitable for use with \code{cvSelect()}:
 
@@
 \item
-  \code{data}, the data set to which the model is fit;
+  \code{data}, the data set to which the model is fit.
 \item
@@
   generator; as for \code{cv()}, if the seed isn't supplied by the user,
-  a seed is randomly selected and saved;
+  a seed is randomly selected and saved.
 \item
@@
 
-By default, \code{cvSelect()} performs 10-fold CV, and produces an
+By default, \code{cvSelect()} performs 10-fold CV, and it produces an
 estimate of MSE for the model-selection procedure even \emph{larger}
@@
 selected after graphical examination of the data, or by analytic
-methods. Once the relationships between the predictors are linearized,
-it can be advantageous similarly to transform the response variable
-towards normality.
+methods, such as transforming the predictors towards multivariate
+normality, which implies linearity. Once the relationships between the
+predictors are linearized, it can be advantageous similarly to transform
+the response variable towards normality.
 
@@
 R> data("Prestige", package="carData")
-R> head(Prestige)
-\end{CodeInput}
-\begin{CodeOutput}
-                    education income women prestige census type
-gov.administrators      13.11  12351 11.16     68.8   1113 prof
-general.managers        12.26  25879  4.02     69.1   1130 prof
-accountants             12.77   9271 15.70     63.4   1171 prof
-purchasing.officers     11.42   8865  9.11     56.8   1175 prof
-chemists                14.62   8403 11.68     73.5   2111 prof
-physicists              15.64  11030  5.13     77.6   2113 prof
-\end{CodeOutput}
-\begin{CodeInput}
 R> summary(Prestige)
@@
   \code{prestige}: the average prestige rating of the occupation on a
-  0--100 ``thermometer'' scale, in a Canadian social survey conducted
+  0--100 ``thermometer'' scale, from a Canadian social survey conducted
   around the same time.
@@
 
-A scatterplot matrix (using the \code{scatterplotMatrix()} function in
-the \pkg{car} package) of the numeric variables in the data reveals that
-the distributions of \code{income} and \code{women} are positively
-skewed, and that some of the relationships among the three predictors,
-and between the predictors and the response (i.e., \code{prestige}), are
+The scatterplot matrix in Figure \ref{fig:scatterplot-matrix} (a)
+(produced by the \code{scatterplotMatrix()} function in the \pkg{car}
+package) for the numeric variables in the data reveals that the
+distributions of \code{income} and \code{women} are positively skewed,
+and that some of the relationships among the three predictors, and
+between the predictors and the response (i.e., \code{prestige}), are
 nonlinear:
@@
 \begin{CodeChunk}
-\begin{CodeInput}
-R> library("car")
-\end{CodeInput}
-\begin{CodeOutput}
-Loading required package: carData
-\end{CodeOutput}
-\begin{CodeInput}
+\begin{figure}
+
+{\centering \subfloat[\label{fig:scatterplot-matrix-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/scatterplot-matrix-1} }\subfloat[\label{fig:scatterplot-matrix-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/scatterplot-matrix-2} }
+
+}
+
+\caption[Scatterplot matrix for the Prestige data, (a) untransformed and (b) with the predictors transformed towards multivariate normality]{Scatterplot matrix for the Prestige data, (a) untransformed and (b) with the predictors transformed towards multivariate normality.}\label{fig:scatterplot-matrix}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library("car", quietly=TRUE)
 R> scatterplotMatrix(~ prestige + income + education + women,
 +                   data=Prestige, smooth=list(spread=FALSE))
-R> 
-R> trans <- powerTransform( cbind(income, education, women) ~ 1,
-+                          data=Prestige, family="yjPower")
-R> summary(trans)
-\end{CodeInput}
-\begin{CodeOutput}
-yjPower Transformations to Multinormality 
-          Est Power Rounded Pwr Wald Lwr Bnd Wald Upr Bnd
-income       0.2678        0.33       0.1051       0.4304
-education    0.5162        1.00      -0.2822       1.3145
-women        0.1630        0.16       0.0112       0.3149
-
- Likelihood ratio test that all transformation parameters are equal to 0
-                               LRT df      pval
-LR test, lambda = (0 0 0) 15.73879  3 0.0012827
-\end{CodeOutput}
-\begin{CodeInput}
-R> P <- Prestige[, c("prestige", "income", "education", "women")]
-R> (lambdas <- trans$roundlam)
-\end{CodeInput}
-\begin{CodeOutput}
-   income education     women 
-0.3300000 1.0000000 0.1630182 
-\end{CodeOutput}
-\begin{CodeInput}
-R> names(lambdas) <- c("income", "education", "women")
-R> for (var in c("income", "education", "women")){
-+   P[, var] <- yjPower(P[, var], lambda=lambdas[var])
-+ }
-R> scatterplotMatrix(~ prestige + income + education + women,
-+                   data=P, smooth=list(spread=FALSE))
-\end{CodeInput}
-\begin{figure}
-
-{\centering \subfloat[\label{fig:scatterplot-matrix-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/scatterplot-matrix-1} }\subfloat[\label{fig:scatterplot-matrix-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/scatterplot-matrix-2} }
-
-}
-
-\caption[Scatterplot matrix for the `Prestige` data, (a) untransformed and (b) transformed]{Scatterplot matrix for the `Prestige` data, (a) untransformed and (b) transformed}\label{fig:scatterplot-matrix}
-\end{figure}
+\end{CodeInput}
 \end{CodeChunk}
@@
 zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see
-\citet[Chap. 8]{Weisberg:2014}, and \citet[Chap. 3]{FoxWeisberg:2019}
-for details. Because \code{women} has some 0 values, we use the
-Yeo-Johnson family:
+\citet[Chap. 8]{Weisberg:2014} and \citet[Chap. 3]{FoxWeisberg:2019} for
+details. Because \code{women} has some 0 values, we use the Yeo-Johnson
+family:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> trans <- powerTransform( cbind(income, education, women) ~ 1,
++                          data=Prestige, family="yjPower")
+R> summary(trans)
+\end{CodeInput}
+\begin{CodeOutput}
+yjPower Transformations to Multinormality 
+          Est Power Rounded Pwr Wald Lwr Bnd Wald Upr Bnd
+income       0.2678        0.33       0.1051       0.4304
+education    0.5162        1.00      -0.2822       1.3145
+women        0.1630        0.16       0.0112       0.3149
+
+ Likelihood ratio test that all transformation parameters are equal to 0
+                               LRT df      pval
+LR test, lambda = (0 0 0) 15.73879  3 0.0012827
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 \code{education}. Applying the ``rounded'' power transformations makes
-the predictors better-behaved:
+the predictors better-behaved (cf., Figures \ref{fig:scatterplot-matrix}
+(a) and (b)):
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> P <- Prestige[, c("prestige", "income", "education", "women")]
+R> (lambdas <- trans$roundlam)
+R> names(lambdas) <- c("income", "education", "women")
+R> for (var in c("income", "education", "women")){
++   P[, var] <- yjPower(P[, var], lambda=lambdas[var])
++ }
+R> scatterplotMatrix(~ prestige + income + education + women,
++                   data=P, smooth=list(spread=FALSE))
+\end{CodeInput}
+\end{CodeChunk}
 
 Comparing the MSE for the regressions with the original and transformed
-predictors shows a advantage to the latter:
+predictors shows an advantage to the latter:
 
@@
 
-Similarly, component+residual plots for the two regressions, produced by
-the \code{crPlots()} function in the \pkg{car} package, suggest that the
-partial relationship of \code{prestige} to \code{income} is more nearly
-linear in the transformed data, but the transformation of
-\textbackslash code\{women{]} fails to capture what appears to be a
-slight quadratic partial relationship; the partial relationship of
-\code{prestige} to \code{education} is close to linear in both
-regressions:
+Similarly, component+residual plots for the two regressions in Figure
+\ref{fig:CR-plots}, produced by the \code{crPlots()} function in the
+\pkg{car} package, suggest that the partial relationship of
+\code{prestige} to \code{income} is more nearly linear in the
+transformed data, but the transformation of \code{women} fails to
+capture what appears to be a slight quadratic partial relationship; the
+partial relationship of \code{prestige} to \code{education} is close to
+linear in both regressions:
 
@@
 R> crPlots(m.pres)
-R> 
 R> crPlots(m.pres.trans)
@@
 
-\caption[Component+residual plots for the `Prestige` regression with (a) the original predictors and (b) transformed predictors]{Component+residual plots for the `Prestige` regression with (a) the original predictors and (b) transformed predictors}\label{fig:CR-plots}
+\caption[Component+residual plots for the Prestige regression with (a) the original predictors and (b) transformed predictors]{Component+residual plots for the Prestige regression with (a) the original predictors and (b) transformed predictors}\label{fig:CR-plots}
 \end{figure}
@@
 
-Having transformed the predictors towards multinormality, we now
+Having transformed the predictors towards multivariate normality, we now
 consider whether there's evidence for transforming the response (using
@@
 to employ. If the \code{predictors} argument is missing then only the
-response is transformed, and if the \code{response} argument is missing,
-only the supplied predictors are transformed. The default \code{family}
-for transforming the predictors is \code{"bcPower"}---the original
-Box-Cox family---as is the default \code{family.y} for transforming the
-response; here we specify \code{family="yjPower} because of the 0s in
-\code{women}. \code{selectTrans()} returns the result of applying a
-lack-of-fit criterion to the model after the selected transformation is
-applied, with the default \code{criterion=mse}:
+\code{response} is transformed, and if the \code{response} argument is
+missing, only the supplied \code{predictors} are transformed. The
+default \code{family} for transforming the predictors is
+\code{"bcPower"}---the original Box-Cox family---as is the default
+\code{family.y} for transforming the response; here we specify
+\code{family="yjPower"} because of the 0s in \code{women}.
+\code{selectTrans()} returns the result of applying a lack-of-fit
+criterion to the model after the selected transformation is applied,
+with the default \code{criterion=mse}:
 
@@
 a CV fold), and hence for use with \code{cvSelect()} (see
-\code{?selectTrans} for details):
+\code{?selectTrans} and the vignette on extending the \pkg{cv} package
+for details):
 
@@
 for selection of transformations in regression. In this section, we
-consider the case where we both select variable transformations and then
+consider the case where we both choose variable transformations and then
 proceed to select predictors. It's insufficient to apply these steps
 sequentially, first, for example, using \code{cvSelect()} with
-\code{selectTrans()} and then with \code{selectStepAIC()}; rather we
+\code{selectTrans()} and then with \code{selectStepAIC()}; rather, we
 should apply the whole model-selection procedure with each fold omitted.
@@
 \begin{CodeInput}
-R> summary(Auto)
-\end{CodeInput}
-\begin{CodeOutput}
-      mpg          cylinders      displacement     horsepower        weight    
- Min.   : 9.00   Min.   :3.000   Min.   : 68.0   Min.   : 46.0   Min.   :1613  
- 1st Qu.:17.00   1st Qu.:4.000   1st Qu.:105.0   1st Qu.: 75.0   1st Qu.:2225  
- Median :22.75   Median :4.000   Median :151.0   Median : 93.5   Median :2804  
- Mean   :23.45   Mean   :5.472   Mean   :194.4   Mean   :104.5   Mean   :2978  
- 3rd Qu.:29.00   3rd Qu.:8.000   3rd Qu.:275.8   3rd Qu.:126.0   3rd Qu.:3615  
- Max.   :46.60   Max.   :8.000   Max.   :455.0   Max.   :230.0   Max.   :5140
-
-  acceleration        year           origin                      name    
- Min.   : 8.00   Min.   :70.00   Min.   :1.000   amc matador       :  5  
- 1st Qu.:13.78   1st Qu.:73.00   1st Qu.:1.000   ford pinto        :  5  
- Median :15.50   Median :76.00   Median :1.000   toyota corolla    :  5  
- Mean   :15.54   Mean   :75.98   Mean   :1.577   amc gremlin       :  4  
- 3rd Qu.:17.02   3rd Qu.:79.00   3rd Qu.:2.000   amc hornet        :  4  
- Max.   :24.80   Max.   :82.00   Max.   :3.000   chevrolet chevette:  4  
-                                                 (Other)           :365  
+R> names(Auto)
+\end{CodeInput}
+\begin{CodeOutput}
+[1] "mpg"          "cylinders"    "displacement" "horsepower"   "weight"      
+[6] "acceleration" "year"         "origin"       "name"        
 \end{CodeOutput}
@@
 
-We previously used the \code{Auto} here in a preliminary example where
-we employed CV to inform the selection of the order of a polynomial
-regression of \code{mpg} on \code{horsepower}. Here, we consider more
-generally the problem of predicting \code{mpg} from the other variables
-in the \code{Auto} data. We begin with a bit of data management, and
-then examine the pairwise relationships among the numeric variables in
-the data set:
+The \code{Auto} appeared in a preliminary example in Section
+\ref{preliminary-example-polynomial-regression}, where we employed CV to
+inform the selection of the degree of a polynomial regression of
+\code{mpg} on \code{horsepower}. Here, we consider more generally the
+problem of predicting \code{mpg} from the other variables in the
+\code{Auto} data. We begin with a bit of data management, and then
+examine the pairwise relationships among the numeric variables in the
+data set (Figure \ref{fig:Auto-explore}):
 
@@
 R> scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
-+                   smooth=list(spread=FALSE), data=Auto)
++                   smooth=list(spread=FALSE), data=Auto, pch=".")
 \end{CodeInput}
@@
 
-\caption[Scatterplot matrix for the numeric variables in the `Auto` data]{Scatterplot matrix for the numeric variables in the `Auto` data}\label{fig:Auto-explore}
+\caption[Scatterplot matrix for the numeric variables in the Auto data]{Scatterplot matrix for the numeric variables in the Auto data}\label{fig:Auto-explore}
 \end{figure}
@@
 however, only 5 distinct values of \code{cylinders} (ranging from 3 to
-8), but cars with 3 or 5 cylinders are rare. and none of the cars has 7
+8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7
 cylinders. There are similarly only 13 distinct years between 1970 and
@@
 R> m.auto <- lm(mpg ~ ., data = Auto)
-R> summary(m.auto)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-lm(formula = mpg ~ ., data = Auto)
-
-Residuals:
-    Min      1Q  Median      3Q     Max 
--9.0064 -1.7450 -0.0917  1.5251 10.9504
-
-Coefficients:
-               Estimate Std. Error t value Pr(>|t|)
-(Intercept)  37.0341323  1.9693933  18.805  < 2e-16
-cylinders5.6 -2.6029412  0.6552000  -3.973 8.54e-05
-cylinders8   -0.5824578  1.1714516  -0.497 0.619335
-displacement  0.0174253  0.0067340   2.588 0.010043
-horsepower   -0.0413534  0.0133786  -3.091 0.002145
-weight       -0.0055479  0.0006323  -8.774  < 2e-16
-acceleration  0.0615272  0.0883132   0.697 0.486431
-year71        0.9680584  0.8373899   1.156 0.248408
-year72       -0.6014345  0.8251155  -0.729 0.466517
-year73       -0.6876890  0.7402723  -0.929 0.353510
-year74        1.3755758  0.8765000   1.569 0.117408
-year75        0.9299288  0.8590716   1.082 0.279742
-year76        1.5598929  0.8225051   1.897 0.058669
-year77        2.9094161  0.8417285   3.456 0.000611
-year78        3.1751976  0.7989396   3.974 8.48e-05
-year79        5.0192987  0.8457587   5.935 6.76e-09
-year80        9.0997634  0.8972933  10.141  < 2e-16
-year81        6.6886597  0.8852181   7.556 3.28e-13
-year82        8.0711248  0.8706683   9.270  < 2e-16
-originEurope  2.0466642  0.5171236   3.958 9.07e-05
-originJapan   2.1448874  0.5077169   4.225 3.02e-05
-
-Residual standard error: 2.924 on 371 degrees of freedom
-Multiple R-squared:  0.8668,    Adjusted R-squared:  0.8596 
-F-statistic: 120.7 on 20 and 371 DF,  p-value: < 2.2e-16
-\end{CodeOutput}
-\begin{CodeInput}
-R> Anova(m.auto)
-\end{CodeInput}
-\begin{CodeOutput}
-Anova Table (Type II tests)
-
-Response: mpg
-             Sum Sq  Df F value    Pr(>F)
-cylinders     292.3   2 17.0915 7.935e-08
-displacement   57.3   1  6.6959  0.010043
-horsepower     81.7   1  9.5544  0.002145
-weight        658.3   1 76.9801 < 2.2e-16
-acceleration    4.2   1  0.4854  0.486431
-year         3016.8  12 29.3987 < 2.2e-16
-origin        190.3   2 11.1287 2.024e-05
-Residuals    3172.5 371                  
-\end{CodeOutput}
-\begin{CodeInput}
 R> crPlots(m.auto)
@@
 
-\caption[Component+residual plots for the working model fit to the `Auto` data]{Component+residual plots for the working model fit to the `Auto` data}\label{fig:Auto-working-model}
+\caption[Component+residual plots for the working model fit to the Auto data]{Component+residual plots for the working model fit to the Auto data}\label{fig:Auto-working-model}
 \end{figure}
@@
 
-The component+residual plots, created with the \code{crPlots()} function
-in the previously loaded \pkg{car} package, clearly reveal the
-inadequacy of the model.
+The component+residual plots in Figure \ref{fig:Auto-working-model}
+clearly reveal the inadequacy of the model.
 
@@
 \begin{CodeInput}
-R> num.predictors <- c("displacement", "horsepower", "weight", "acceleration")
+R> num.predictors <- c("displacement", "horsepower", "weight", 
++                     "acceleration")
 R> tr.x <- powerTransform(Auto[, num.predictors])
@@
 + }
-R> head(A)
-\end{CodeInput}
-\begin{CodeOutput}
-                          mpg cylinders displacement horsepower   weight
-chevrolet.chevelle.malibu  18         8     5.726848   4.867534 8.161660
-buick.skylark.320          15         8     5.857933   5.105945 8.214194
-plymouth.satellite         18         8     5.762051   5.010635 8.142063
-amc.rebel.sst              16         8     5.717028   5.010635 8.141190
-ford.torino                17         8     5.710427   4.941642 8.145840
-ford.galaxie.500           15         8     6.061457   5.288267 8.375860
-                          acceleration year  origin
-chevrolet.chevelle.malibu     2.484907   70 America
-buick.skylark.320             2.442347   70 America
-plymouth.satellite            2.397895   70 America
-amc.rebel.sst                 2.484907   70 America
-ford.torino                   2.351375   70 America
-ford.galaxie.500              2.302585   70 America
-\end{CodeOutput}
-\begin{CodeInput}
 R> m <- update(m.auto, data=A)
@@
 R> m <- update(m, log(mpg) ~ .)
-R> summary(m)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-lm(formula = log(mpg) ~ cylinders + displacement + horsepower + 
-    weight + acceleration + year + origin, data = A)
-
-Residuals:
-     Min       1Q   Median       3Q      Max 
--0.33412 -0.05774  0.00410  0.06072  0.38081
-
-Coefficients:
-             Estimate Std. Error t value Pr(>|t|)
-(Intercept)   8.89652    0.35822  24.835  < 2e-16
-cylinders5.6 -0.06355    0.02574  -2.469   0.0140
-cylinders8   -0.07691    0.03900  -1.972   0.0493
-displacement  0.02799    0.05146   0.544   0.5868
-horsepower   -0.29010    0.05631  -5.152 4.20e-07
-weight       -0.54274    0.08193  -6.624 1.23e-10
-acceleration -0.14214    0.05630  -2.525   0.0120
-year71        0.02505    0.02891   0.866   0.3869
-year72       -0.01680    0.02894  -0.580   0.5620
-year73       -0.04257    0.02602  -1.636   0.1026
-year74        0.04932    0.03041   1.622   0.1056
-year75        0.04715    0.02959   1.594   0.1118
-year76        0.07087    0.02845   2.491   0.0132
-year77        0.13241    0.02927   4.523 8.21e-06
-year78        0.14472    0.02777   5.211 3.13e-07
-year79        0.23354    0.02921   7.994 1.67e-14
-year80        0.32381    0.03170  10.216  < 2e-16
-year81        0.25655    0.03094   8.291 2.10e-15
-year82        0.30756    0.03036  10.131  < 2e-16
-originEurope  0.04921    0.01955   2.518   0.0122
-originJapan   0.04409    0.01947   2.265   0.0241
-
-Residual standard error: 0.1043 on 371 degrees of freedom
-Multiple R-squared:  0.9108,    Adjusted R-squared:  0.906 
-F-statistic: 189.4 on 20 and 371 DF,  p-value: < 2.2e-16
-\end{CodeOutput}
-\begin{CodeInput}
-R> Anova(m)
-\end{CodeInput}
-\begin{CodeOutput}
-Anova Table (Type II tests)
-
-Response: log(mpg)
-             Sum Sq  Df F value    Pr(>F)
-cylinders    0.0663   2  3.0521   0.04845
-displacement 0.0032   1  0.2959   0.58679
-horsepower   0.2885   1 26.5420 4.198e-07
-weight       0.4769   1 43.8805 1.229e-10
-acceleration 0.0693   1  6.3745   0.01199
-year         4.4521  12 34.1339 < 2.2e-16
-origin       0.0807   2  3.7128   0.02532
-Residuals    4.0325 371                  
-\end{CodeOutput}
-\end{CodeChunk}
-
-The transformed numeric variables are much better-behaved:
+\end{CodeInput}
+\end{CodeChunk}
+
+The transformed numeric variables are much better-behaved (cf., Figures
+\ref{fig:Auto-explore} on page \pageref{fig:Auto-explore} and
+\ref{fig:Auto-transformed-scatterplot-matrix}:
 
@@
 +                   + acceleration, 
-+                   smooth=list(spread=FALSE), data=A)
++                   smooth=list(spread=FALSE), data=A, pch=".")
 \end{CodeInput}
@@
 
-\caption[Scatterplot matrix for the transformed numeric variables in the `Auto` data]{Scatterplot matrix for the transformed numeric variables in the `Auto` data}\label{fig:Auto-transformed-scatterplot-matrix}
+\caption[Scatterplot matrix for the transformed numeric variables in the Auto data]{Scatterplot matrix for the transformed numeric variables in the Auto data}\label{fig:Auto-transformed-scatterplot-matrix}
 \end{figure}
@@
 And the partial relationships in the model fit to the transformed data
-are much more nearly linear:
+are much more nearly linear (cf., Figures \ref{fig:Auto-working-model}
+on page \pageref{fig:Auto-working-model} and
+\ref{fig:Auto-CR-plots-transformed}:
 
@@
 
-\caption[Component+residual plots for the model fit to the transformed `Auto` data]{Component+residual plots for the model fit to the transformed `Auto` data}\label{fig:Auto-CR-plots-transformed}
+\caption[Component+residual plots for the model fit to the transformed Auto data]{Component+residual plots for the model fit to the transformed Auto data}\label{fig:Auto-CR-plots-transformed}
 \end{figure}
@@
               Estimate Std. Error t value Pr(>|t|)    
-(Intercept)   9.434594   0.261529  36.075  < 2e-16
-horsepower   -0.276254   0.056143  -4.921 1.30e-06
-weight       -0.609071   0.056003 -10.876  < 2e-16
-acceleration -0.131380   0.053195  -2.470  0.01397
+(Intercept)   9.434594   0.261529  36.075  < 2e-16 ***
+horsepower   -0.276254   0.056143  -4.921 1.30e-06 ***
+weight       -0.609071   0.056003 -10.876  < 2e-16 ***
+acceleration -0.131380   0.053195  -2.470  0.01397 *  
 year71        0.027984   0.028936   0.967  0.33412    
@@
 year73       -0.039529   0.026014  -1.520  0.12947    
-year74        0.052752   0.029986   1.759  0.07936
-year75        0.053199   0.029280   1.817  0.07004
-year76        0.074317   0.028212   2.634  0.00878
-year77        0.137931   0.028875   4.777 2.56e-06
-year78        0.145876   0.027529   5.299 1.99e-07
-year79        0.236036   0.029080   8.117 6.99e-15
-year80        0.335274   0.031148  10.764  < 2e-16
-year81        0.262872   0.030555   8.603  < 2e-16
-year82        0.323391   0.029608  10.922  < 2e-16
-originEurope  0.055818   0.016785   3.326  0.00097
-originJapan   0.043554   0.017479   2.492  0.01314
+year74        0.052752   0.029986   1.759  0.07936 .  
+year75        0.053199   0.029280   1.817  0.07004 .  
+year76        0.074317   0.028212   2.634  0.00878 ** 
+year77        0.137931   0.028875   4.777 2.56e-06 ***
+year78        0.145876   0.027529   5.299 1.99e-07 ***
+year79        0.236036   0.029080   8.117 6.99e-15 ***
+year80        0.335274   0.031148  10.764  < 2e-16 ***
+year81        0.262872   0.030555   8.603  < 2e-16 ***
+year82        0.323391   0.029608  10.922  < 2e-16 ***
+originEurope  0.055818   0.016785   3.326  0.00097 ***
+originJapan   0.043554   0.017479   2.492  0.01314 *  
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
 
@@
 F-statistic: 219.8 on 17 and 374 DF,  p-value: < 2.2e-16
-\end{CodeOutput}
-\begin{CodeInput}
-R> Anova(m.step)
-\end{CodeInput}
-\begin{CodeOutput}
-Anova Table (Type II tests)
-
-Response: log(mpg)
-             Sum Sq  Df  F value    Pr(>F)
-horsepower   0.2663   1  24.2120 1.296e-06
-weight       1.3010   1 118.2805 < 2.2e-16
-acceleration 0.0671   1   6.0998  0.013965
-year         4.7589  12  36.0549 < 2.2e-16
-origin       0.1366   2   6.2090  0.002225
-Residuals    4.1137 374                   
 \end{CodeOutput}
@@
 this model, but we expect that the result will be optimistic because we
-used the whole data to help specify the model
+used the whole data to help specify the model:
 
@@
 on the original \code{mpg} response scale rather than the log scale, so
-as to make the selected model comparable to the working model. That's
-slightly uncomfortable given the skewed distribution of \code{mpg}. An
-alternative is to use the median absolute error instead of the
+as to make the selected model comparable to the working
+model.\footnote{That's slightly uncomfortable given the skewed
+  distribution of \code{mpg}. An alternative is to use a robust measure
+  of model lack-of-fit, such as the median absolute error instead of the
   mean-squared error, employing the \code{medAbsErr()} function from the
-\pkg{cv} package:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> medAbsErr(Auto$mpg, exp(fitted(m.step)))
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 1.339604
-\end{CodeOutput}
-\begin{CodeInput}
-R> medAbsErr(Auto$mpg, fitted(m.auto))
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 1.666121
-\end{CodeOutput}
-\end{CodeChunk}
+  \pkg{cv} package. The median absolute error, however, cannot be
+  expressed as a casewise average (see Section
+  \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).}
 
@@
 +                 predictors=num.predictors,
-+                 response="mpg", AIC=FALSE, criterion=medAbsErr)
++                 response="mpg", AIC=FALSE)
 \end{CodeInput}
@@
 10-Fold Cross Validation
-cross-validation criterion = 1.495075
-full-sample criterion = 1.339604 
+cross-validation criterion = 7.485557
+bias-adjusted cross-validation criterion = 7.343535
+full-sample criterion = 6.512144 
 \end{CodeOutput}
@@
 The starting model, \code{m.auto}, is the working model fit to the
-\code{Auto} data. The CV criterion isn't bias-adjusted because median
-absolute error isn't a mean of casewise error components.
+\code{Auto} data.
 
@@
   \code{selectTransStepAIC()} automatically computes CV cost criteria,
-  here the median absolute error, on the untransformed response scale.
-\item
-  The estimate of the median absolute error that we obtain by
-  cross-validating the whole model-specification process is a little
-  larger than the median absolute error computed for the model we fit to
-  the \code{Auto} data separately selecting transformations of the
-  predictors and the response and then selecting predictors for the
-  whole data set.
+  here the MSE, on the \emph{untransformed} response scale.
+\item
+  The estimate of the MSE that we obtain by cross-validating the whole
+  model-specification process is larger than the MSE computed for the
+  model we fit to the \code{Auto} data separately selecting
+  transformations of the predictors and the response and then selecting
+  predictors for the whole data set.
 \item
@@
   and \code{year} are always included among the selected predictors;
-  \code{acceleration} and \code{displacement} are each included
-  respectively in 4 and 3 of 10 selected models; and \code{cylinders}
-  and \code{origin} are each included in only 1 of 10 models. Recall
-  that when we selected predictors for the full data, we obtained a
-  model with \code{horsepower}, \code{weight}, \code{acceleration},
+  \code{acceleration} and \code{displacement} are included respectively
+  in 4 and 3 of 10 selected models; and \code{cylinders} and
+  \code{origin} are each included in only 1 of 10 models. Recall that
+  when we selected predictors for the full data, we obtained a model
+  with \code{horsepower}, \code{weight}, \code{acceleration},
   \code{year}, and \code{origin}.
@@
 
-The \pkg{cv} package is designed to be extensible in several directions.
-In this vignette, we discuss three kinds of extensions, ordered by
-increasing general complexity: (1) adding a cross-validation cost
-criterion; (2) adding a model class that's not directly accommodated by
-the \code{cv()} default method or by another directly inherited method,
-with separate consideration of mixed-effects models; and (3) adding a
-new model-selection procedure suitable for use with
-\code{selectModel()}.
-
-\hypertarget{adding-a-model-class-not-covered-by-the-default-cv-method}{%
-\subsection{Adding a model class not covered by the default cv()
-method}\label{adding-a-model-class-not-covered-by-the-default-cv-method}}
-
-\hypertarget{independently-sampled-cases}{%
-\subsubsection{Independently sampled
-cases}\label{independently-sampled-cases}}
+The \pkg{cv} package is designed to be extensible in several directions;
+in order of increasing general complexity, we can add: (1) a
+cross-validation cost criterion; (2) a model class that's not directly
+accommodated by the \code{cv()} default method or by another directly
+inherited method; and (3) a new model-selection procedure suitable for
+use with \code{selectModel()}. In this section, we illustrate (1) and
+(2); more extensive examples may be found in the vignette on extending
+the \pkg{cv} package.
 
@@
 vote choice in the 2001 British election. Data for the example are in
-the \code{BEPS} data frame in the \pkg{carData} package:
+the \code{BEPS} data set in the \pkg{carData} package:
 
@@
 R> data("BEPS", package="carData")
-R> head(BEPS)
-\end{CodeInput}
-\begin{CodeOutput}
-              vote age economic.cond.national economic.cond.household Blair
-1 Liberal Democrat  43                      3                       3     4
-2           Labour  36                      4                       4     4
-3           Labour  35                      4                       4     5
-4           Labour  24                      4                       2     2
-5           Labour  41                      2                       2     1
-6           Labour  47                      3                       4     4
-  Hague Kennedy Europe political.knowledge gender
-1     1       4      2                   2 female
-2     4       4      5                   2   male
-3     2       3      3                   2   male
-4     1       3      4                   0 female
-5     1       4      6                   2   male
-6     4       2      4                   2   male
+R> summary(BEPS)
+\end{CodeInput}
+\begin{CodeOutput}
+               vote          age        economic.cond.national
+ Conservative    :462   Min.   :24.00   Min.   :1.000         
+ Labour          :720   1st Qu.:41.00   1st Qu.:3.000         
+ Liberal Democrat:343   Median :53.00   Median :3.000         
+                        Mean   :54.18   Mean   :3.246         
+                        3rd Qu.:67.00   3rd Qu.:4.000         
+                        Max.   :93.00   Max.   :5.000         
+ economic.cond.household     Blair           Hague          Kennedy     
+ Min.   :1.00            Min.   :1.000   Min.   :1.000   Min.   :1.000  
+ 1st Qu.:3.00            1st Qu.:2.000   1st Qu.:2.000   1st Qu.:2.000  
+ Median :3.00            Median :4.000   Median :2.000   Median :3.000  
+ Mean   :3.14            Mean   :3.334   Mean   :2.747   Mean   :3.135  
+ 3rd Qu.:4.00            3rd Qu.:4.000   3rd Qu.:4.000   3rd Qu.:4.000  
+ Max.   :5.00            Max.   :5.000   Max.   :5.000   Max.   :5.000  
+     Europe       political.knowledge    gender   
+ Min.   : 1.000   Min.   :0.000       female:812  
+ 1st Qu.: 4.000   1st Qu.:0.000       male  :713  
+ Median : 6.000   Median :2.000                   
+ Mean   : 6.729   Mean   :1.542                   
+ 3rd Qu.:10.000   3rd Qu.:2.000                   
+ Max.   :11.000   Max.   :3.000                   
 \end{CodeOutput}
@@
 The model fit to the data includes an interaction between \code{Europe}
-and \code{political.knowledge}; the other predictors enter the model
+and \code{political.knowledge}, which was the focus of the original
+research on which this example is based
+\citep{AndersenHeathSinnott:2002}; the other predictors enter the model
 additively:
@@
 \begin{CodeInput}
-R> library("nnet")
+R> library("nnet", quietly=TRUE)
 R> m.beps <- multinom(vote ~ age + gender + economic.cond.national +
 +                        economic.cond.household + Blair + Hague + Kennedy +
-+                        Europe*political.knowledge, data=BEPS)
-\end{CodeInput}
-\begin{CodeOutput}
-# weights:  36 (22 variable)
-initial  value 1675.383740 
-iter  10 value 1240.047788
-iter  20 value 1163.199642
-iter  30 value 1116.519687
-final  value 1116.519666 
-converged
-\end{CodeOutput}
-\begin{CodeInput}
-R> car::Anova(m.beps)
-\end{CodeInput}
-\begin{CodeOutput}
-Analysis of Deviance Table (Type II tests)
-
-Response: vote
-                           LR Chisq Df Pr(>Chisq)
-age                          13.872  2  0.0009721
-gender                        0.453  2  0.7972568
-economic.cond.national       30.604  2  2.262e-07
-economic.cond.household       5.652  2  0.0592570
-Blair                       135.369  2  < 2.2e-16
-Hague                       166.770  2  < 2.2e-16
-Kennedy                      68.878  2  1.105e-15
-Europe                       78.033  2  < 2.2e-16
-political.knowledge          55.568  2  8.582e-13
-Europe:political.knowledge   50.804  2  9.291e-12
-\end{CodeOutput}
-\end{CodeChunk}
-
-Most of the predictors, including the \code{Europe} \(\times\)
-\code{political.knowledge} interaction, are associated with very small
-\(p\)-values; the \code{Anova()} function is from the \pkg{car} package
-\citep{FoxWeisberg:2019}.
-
-Here's an ``effect plot'', using the the \pkg{effects} package
-\citep{FoxWeisberg:2019} to visualize the \code{Europe} \(\times\)
-\code{political.knowledge} interaction in a ``stacked-area'' graph:
++                        Europe*political.knowledge, data=BEPS, trace=FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+We set the argument \code{trace=FALSE} in the call to \code{multinom()}
+to suppress reporting the iteration history, which will be particularly
+appreciated when we repeatedly refit the model in cross-validation.
+
+The \code{Europe} \(\times\) \code{political.knowledge} interaction is
+associated with a very small \(p\)-value. Figure \ref{fig:BEPS-plot}
+shows an ``effect plot,'' using the the \pkg{effects} package
+\citep{FoxWeisberg:2019} to visualize the interaction in a
+``stacked-area'' graph:
 
@@
 \end{CodeInput}
-
-\begin{center}\includegraphics{JSS-article-reduced_files/figure-latex/BEPS-plot-1} \end{center}
-
+\begin{figure}
+
+{\centering \includegraphics{JSS-article-reduced_files/figure-latex/BEPS-plot-1}
+
+}
+
+\caption[Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study]{Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study.}\label{fig:BEPS-plot}
+\end{figure}
 \end{CodeChunk}
@@
 To cross-validate this multinomial-logit model we need an appropriate
-cost criterion. None of the criteria supplied by the \pkg{cv}
-package---for example, neither \code{mse()}, which is appropriate for a
-numeric response, nor \code{BayesRule()}, which is appropriate for a
-binary response---will do. One possibility is to adapt Bayes rule to a
-polytomous response:
+cost criterion. None of the criteria in the \pkg{cv} package will
+do---for example, \code{mse()} is appropriate only for a numeric
+response. The \code{BayesRule()} criterion, also supplied by \pkg{cv},
+which is for a binary response, comes close:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> BayesRule
+\end{CodeInput}
+\begin{CodeOutput}
+function(y, yhat){
+  if (!all(y %in% c(0, 1))) stop("response values not all 0 or 1")
+  if (any(yhat < 0) || any(yhat > 1)) stop("fitted values outside of interval [0, 1]")
+  yhat <- round(yhat)
+  result <- mean(y != yhat) # proportion in error
+  attr(result, "casewise loss") <- "y != round(yhat)"
+  result
+}
+<bytecode: 0x129d04270>
+<environment: namespace:cv>
+\end{CodeOutput}
+\end{CodeChunk}
+
+After doing some error checking, \code{BayesRule()} rounds the predicted
+proabability of a 1 (``success'') response in a binary regression model
+to 0 or 1 to obtain a categorical prediction and then reports the
+proportion of incorrect predictions. Because the Bayes's rule criterion
+is an average of casewise components (as, e.g., is the MSE), a
+\code{"casewise loss"} attribute is attached to the result, making
+possible the computation of bias correction and confidence intervals (as
+discussion in Section
+\ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+
+It is straightforward to adapt Bayes's rule to a polytomous response:
 
@@
 The \code{predict()} method for \code{"multinom"} models called with
-argument \code{type="class"} reports the Bayes-rule prediction for each
-case---that is, the response category with the highest predicted
+argument \code{type="class"} reports the Bayes's rule prediction for
+each case---that is, the response category with the highest predicted
 probability. Our \code{BayesRuleMulti()} function calculates the
-proportion of misclassified cases. Because this value is the mean of
-casewise components, we attach a \code{"casewise loss"} attribute to the
-result (as explained in the preceding section).
+proportion of misclassified cases. Because this value is also the mean
+of casewise components, we attach a \code{"casewise loss"} attribute to
+the result.
 
@@
 
-and so the marginal Bayes-rule prediction, that everyone will vote
+and so the marginal Bayes's rule prediction, that everyone will vote
 Labour, produces an error rate of \(1 - 0.47213 = 0.52787\). The
@@
 \begin{CodeOutput}
-# weights:  36 (22 variable)
-initial  value 1507.296060 
-iter  10 value 1134.575036
-iter  20 value 1037.413231
-iter  30 value 1007.705242
-iter  30 value 1007.705235
-iter  30 value 1007.705235
-final  value 1007.705235 
-converged
-\end{CodeOutput}
-\begin{CodeOutput}
 Error in match.arg(type): 'arg' should be one of "class", "probs"
@@
 
-That is, we simply call the default \code{cv()} method with the
-\code{type} argument properly set. In addition to supplying the correct
-\code{type} argument, our method sets the default \code{criterion} for
-the \code{cv.multinom()} method to \code{BayesRuleMulti}.
+That is, we simply invoke the default \code{cv()} method via
+\code{NextMethod()}, with the \code{type} argument properly set. In
+addition to supplying the correct \code{type} argument, our method sets
+the default \code{criterion} for the \code{cv.multinom()} method to
+\code{BayesRuleMulti}.
 
@@
 \begin{CodeInput}
-R> m.beps <- update(m.beps, trace=FALSE)
 R> cv(m.beps, seed=3465)
@@
 
-Prior to invoking \code{cv()}, we called \code{update()} with
-\code{trace=FALSE} to suppress the iteration history reported by default
-by \code{multinom()}---it would be tedious to see the iteration history
-for each fold. The cross-validated polytomous Bayes-rule criterion
-confirms that the fitted model does substantially better than the
-marginal Bayes-rule prediction that everyone votes for Labour.
+The cross-validated polytomous Bayes's rule criterion confirms that the
+fitted model does substantially better than the marginal Bayes's rule
+prediction that everyone votes for Labour.
 
@@
 hatvalues are equal to 1, in which case the formula for
-\(\widehat{y}_{-i}\) requires division by 0.
+\(\widehat{y}_{-i}\) requires division by 0. In this case, the
+``training'' set omitting the observation with hatvalue = 1 is
+rank-deficient and the predictors for the left-out case are outside the
+linear span of the predictors in the training set.
 
@@
 than one case, we make use of the Woodbury matrix identify
-\citep{Wikipedia-Woodbury:2023}, \[
+\citep{Hager:1989}, \[
 \left(\mathbf{A}_{m \times m} + \mathbf{U}_{m \times k} 

### 2024-01-11 scratch/notes_john/JSS-article/JSS-article-reduced.Rmd
small edits to JSS paper

@@
   mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
-  lines(horsepower, mpg, col=p + 1, lty=p, lwd=2)
+  lines(horsepower, mpg, col=p + 1, lty=p, lwd=4)
 }
-legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=2,
+legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=4,
        title="Degree", inset=0.02)
@@
 Because we generated the polynomial regression models in a named list, we conveniently employ \code{do.call()} to supply the models as arguments to \code{models()}.
-Note that the names created for the list (e.g., \code{"m.2"}) are then used for the models.
+The names created for the list (e.g., \code{"m.2"}) are then used for the models.
 We can also invoke the \code{plot()} method for \code{"cvModList"} objects to compare the models (see Figure \ref{fig:polynomial-regression-CV-graph-2}):
@@
 ```
-In the process, we merged variables from the school-level and student-level data sets and created two new school-level variables: \code{mean.ses}, which is the average SES for students in each school; and \code{cses}, which is each students' SES centered at their school means.
+In the process, we merged variables from the school-level and student-level data sets and created two new school-level variables: \code{mean.ses}, which is the average SES for students in each school; and \code{cses}, which is the individual students' SES centered at their school means.
 For details, see @FoxWeisberg:2019 [Sec.

### 2024-01-11 scratch/notes_john/JSS-article/JSS-article-reduced.tex
small edits to JSS paper

@@
                       Estimate Std. Error t value Pr(>|t|)
-(Intercept)            23.4459     0.2209  106.13   <2e-16 ***
-poly(horsepower, 2)1 -120.1377     4.3739  -27.47   <2e-16 ***
-poly(horsepower, 2)2   44.0895     4.3739   10.08   <2e-16 ***
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+(Intercept)            23.4459     0.2209  106.13   <2e-16
+poly(horsepower, 2)1 -120.1377     4.3739  -27.47   <2e-16
+poly(horsepower, 2)2   44.0895     4.3739   10.08   <2e-16
 
@@
 \begin{CodeOutput}
-R RNG seed set to 542508
+R RNG seed set to 32925
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.57393
-bias-adjusted cross-validation criterion = 19.54281
-95% CI for bias-adjusted CV criterion = (16.03528, 23.05034)
+cross-validation criterion = 19.41501
+bias-adjusted cross-validation criterion = 19.3924
+95% CI for bias-adjusted CV criterion = (15.88674, 22.89806)
 full-sample criterion = 18.98477 
@@
 models in a named list, we conveniently employ \code{do.call()} to
-supply the models as arguments to \code{models()}. Note that the names
-created for the list (e.g., \code{"m.2"}) are then used for the models.
-We can also invoke the \code{plot()} method for \code{"cvModList"}
-objects to compare the models (see Figure
+supply the models as arguments to \code{models()}. The names created for
+the list (e.g., \code{"m.2"}) are then used for the models. We can also
+invoke the \code{plot()} method for \code{"cvModList"} objects to
+compare the models (see Figure
 \ref{fig:polynomial-regression-CV-graph-2}):
@@
 \code{mean.ses}, which is the average SES for students in each school;
-and \code{cses}, which is each students' SES centered at their school
-means. For details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.
+and \code{cses}, which is the individual students' SES centered at their
+school means. For details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.
 
@@
             Estimate Std. Error t value Pr(>|t|)
-(Intercept) 9.050430   0.347189  26.068  < 2e-16 ***
-x           0.020908   0.007273   2.875  0.00422 ** 
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+(Intercept) 9.050430   0.347189  26.068  < 2e-16
+x           0.020908   0.007273   2.875  0.00422
 
@@
             Estimate Std. Error t value Pr(>|t|)
-(Intercept)  9.93704    0.03122   318.3   <2e-16 ***
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+(Intercept)  9.93704    0.03122   318.3   <2e-16
 
@@
             Estimate Std. Error t value Pr(>|t|)
-(Intercept)  9.93716    0.03098 320.803  < 2e-16 ***
-X.99        -0.09103    0.03082  -2.953  0.00322 ** 
-X.90        -0.08205    0.03135  -2.617  0.00901 ** 
-X.87        -0.06942    0.03105  -2.235  0.02561 *  
+(Intercept)  9.93716    0.03098 320.803  < 2e-16
+X.99        -0.09103    0.03082  -2.953  0.00322
+X.90        -0.08205    0.03135  -2.617  0.00901
+X.87        -0.06942    0.03105  -2.235  0.02561
 X.40        -0.04759    0.03076  -1.547  0.12211
-X.65        -0.05523    0.03147  -1.755  0.07952 .  
-X.91         0.05245    0.03084   1.700  0.08937 .  
+X.65        -0.05523    0.03147  -1.755  0.07952
+X.91         0.05245    0.03084   1.700  0.08937
 X.53        -0.04921    0.03048  -1.615  0.10672
-X.45         0.05543    0.03182   1.742  0.08183 .  
+X.45         0.05543    0.03182   1.742  0.08183
 X.31         0.04525    0.03108   1.456  0.14570
-X.56         0.05433    0.03273   1.660  0.09723 .  
+X.56         0.05433    0.03273   1.660  0.09723
 X.61        -0.05085    0.03170  -1.604  0.10908
@@
 X.92         0.04430    0.03100   1.429  0.15329
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
 
@@
               Estimate Std. Error t value Pr(>|t|)
-(Intercept)   9.434594   0.261529  36.075  < 2e-16 ***
-horsepower   -0.276254   0.056143  -4.921 1.30e-06 ***
-weight       -0.609071   0.056003 -10.876  < 2e-16 ***
-acceleration -0.131380   0.053195  -2.470  0.01397 *  
+(Intercept)   9.434594   0.261529  36.075  < 2e-16
+horsepower   -0.276254   0.056143  -4.921 1.30e-06
+weight       -0.609071   0.056003 -10.876  < 2e-16
+acceleration -0.131380   0.053195  -2.470  0.01397
 year71        0.027984   0.028936   0.967  0.33412
@@
 year73       -0.039529   0.026014  -1.520  0.12947
-year74        0.052752   0.029986   1.759  0.07936 .  
-year75        0.053199   0.029280   1.817  0.07004 .  
-year76        0.074317   0.028212   2.634  0.00878 ** 
-year77        0.137931   0.028875   4.777 2.56e-06 ***
-year78        0.145876   0.027529   5.299 1.99e-07 ***
-year79        0.236036   0.029080   8.117 6.99e-15 ***
-year80        0.335274   0.031148  10.764  < 2e-16 ***
-year81        0.262872   0.030555   8.603  < 2e-16 ***
-year82        0.323391   0.029608  10.922  < 2e-16 ***
-originEurope  0.055818   0.016785   3.326  0.00097 ***
-originJapan   0.043554   0.017479   2.492  0.01314 *  
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+year74        0.052752   0.029986   1.759  0.07936
+year75        0.053199   0.029280   1.817  0.07004
+year76        0.074317   0.028212   2.634  0.00878
+year77        0.137931   0.028875   4.777 2.56e-06
+year78        0.145876   0.027529   5.299 1.99e-07
+year79        0.236036   0.029080   8.117 6.99e-15
+year80        0.335274   0.031148  10.764  < 2e-16
+year81        0.262872   0.030555   8.603  < 2e-16
+year82        0.323391   0.029608  10.922  < 2e-16
+originEurope  0.055818   0.016785   3.326  0.00097
+originJapan   0.043554   0.017479   2.492  0.01314
 
@@
 }
-<bytecode: 0x129d04270>
+<bytecode: 0x13e7ed098>
 <environment: namespace:cv>

### 2024-01-12 scratch/notes_john/JSS-article/JSS-article-reduced.Rmd
fix typo in JSS paper

@@
 
-* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computation-for-linear-and-generalized-linear-models}).
+* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).
 

### 2024-01-12 scratch/notes_john/JSS-article/JSS-article-reduced.tex
fix typo in JSS paper

@@
   Section
-  \ref{efficient-computation-for-linear-and-generalized-linear-models}).
+  \ref{efficient-computations-for-linear-and-generalized-linear-models}).
 \item
@@
 \begin{CodeOutput}
-R RNG seed set to 32925
+R RNG seed set to 735559
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.41501
-bias-adjusted cross-validation criterion = 19.3924
-95% CI for bias-adjusted CV criterion = (15.88674, 22.89806)
+cross-validation criterion = 19.44017
+bias-adjusted cross-validation criterion = 19.41622
+95% CI for bias-adjusted CV criterion = (15.87764, 22.95479)
 full-sample criterion = 18.98477 
@@
 }
-<bytecode: 0x13e7ed098>
+<bytecode: 0x13db9d388>
 <environment: namespace:cv>

### 2024-01-12 scratch/notes_john/JSS-article/JSS-article-reduced-2.Rmd
add JSS-article-reduced-2

--- 2024-01-12 scratch/notes_john/JSS-article/JSS-article-reduced.Rmd
@@
 
-The \code{cvSelect()} function, discussed in Section \ref{cross-validating-model-selection}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors.
+The \code{cvSelect()} function, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors.
 
@@
 
-# Cross-validating model selection
+# Cross-validating model specification
 
 As @HastieTibshiraniFriedman:2009 [Sec.
-7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain, if the whole data are used to select or fine-tune a statistical model, then subsequent cross-validation of the model is intrinsically misleading, because the model is selected to fit the whole data, including the part of the data that remains when each fold is removed.
-
-## A preliminary example
-
-The following example is similar in spirit to one employed by @HastieTibshiraniFriedman:2009.
-Suppose that we randomly generate $n = 1000$ independent observations for a response variable variable $y \sim N(\mu = 10, \sigma^2 = 1)$, and independently sample $1000$ observations for $p = 100$ "predictors," $x_1, \ldots, x_{100}$, each from $x_j \sim N(0, 1)$.
-The response has nothing to do with the predictors and so the population linear-regression model $y_i = \alpha + \beta_1 x_{i1} + \cdots + \beta_{100} x_{i,100} + \varepsilon_i$ has $\alpha = 10$, all $\beta_j = 0$, and $\sigma_\varepsilon^2 = 1$.
-
-```{r generate-selection-data}
-set.seed(24361) # for reproducibility
-D <- data.frame(
-  y = rnorm(1000, mean=10),
-  X = matrix(rnorm(1000*100), 1000, 100)
-)
-```
-Least-squares provides accurate estimates of the regression constant $\alpha = 10$ and the error variance $\sigma^2 = 1$ for the "null model" including only the regression constant; moreover, the omnibus $F$-test of the correct null hypothesis that all of the $\beta$s are 0 for the "full model" with all 100 $x$s is associated with a large $p$-value:
-```{r omnibus-F}
-m.full <- lm(y ~ ., data=D)
-m.null <- lm(y ~ 1, data=D)
-summary(m.null)
-anova(m.null, m.full)
-```
-Next, using the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], let us perform a forward stepwise regression to select a "best" model, starting with the null model, and using AIC as the model-selection criterion (see \code{?stepAIC()} for details):[^selection-order]
-
-[^selection-order]: It's generally advantageous to start with the largest model, here the one with 100 predictors, and proceed by backward elimination.
-In this demonstration, however, where all of the $\beta$s are really 0, the selected model will be small, and so we proceed by forward selection from the null model to save computing time.
-
-```{r forward-selection}
-library("MASS")  # for stepAIC()
-m.select <- stepAIC(m.null,
-                    direction="forward", trace=FALSE,
-                    scope=list(lower=~1, upper=formula(m.full)))
-summary(m.select)
-mse(D$y, fitted(m.select))
-```
-The resulting model has 15 predictors, a very modest $R^2 = .044$, but a small $p$-value for its omnibus $F$-test (which, of course, is entirely spurious because the same data were used to select and test the model).
-The MSE for the selected model is smaller than the true error variance $\sigma^2 = 1$, as is the estimated error variance for the selected model, $\widehat{\sigma}^2 = 0.973^2 = 0.947$.[^casewise-loss]
-
-[^casewise-loss]: The \code{"casewise loss"} attribute attached to the value returned by the \code{mse()} function signals to \code{cv()} that MSE is an average of casewise components (and how to calculate the casewise components), which has implications for computing bias correction and confidences intervals; see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}.
-
-If we cross-validate the selected model, we also obtain an optimistic estimate of its predictive power (although the confidence interval for the bias-adjusted MSE includes 1):
-```{r cv-selectedModel}
-cv(m.select, seed=2529)
-```
-
-The \code{cvSelect()} function in the \pkg{cv} package, in contrast, allows us to cross-validate the whole model-selection procedure.
-The first argument to \code{cvSelect()} is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion.
-The \code{selectStepAIC()} function, also in \pkg{cv} and based on \code{stepAIC()}, is suitable for use with \code{cvSelect()}:
-```{r cvSelect-artificial-data, cache=TRUE}
-cv.select <- cvSelect(selectStepAIC, data=D, seed=3791,
-                      model=m.null, direction="forward",
-                      scope=list(lower=~1, 
-                                 upper=formula(m.full)))
-cv.select
-```
-
-The other arguments to \code{cvSelect()} are:
-
-* \code{data}, the data set to which the model is fit.
-* \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, a seed is randomly selected and saved.
-* additional arguments required by the model-selection function, here the starting \code{model} argument, the \code{direction} of model selection, and the \code{scope} of models considered (from the model with only a regression constant to the model with all 100 predictors).
-
-By default, \code{cvSelect()} performs 10-fold CV, and it produces an estimate of MSE for the model-selection procedure even *larger* than the true error variance, $\sigma^2 = 1$.
-
-Also by default, when the number of folds is 10 or fewer, \code{cvSelect()} saves the coefficients of the selected models.
-In this example, the \code{compareFolds()} function reveals that the variables retained by the model-selection process in the several folds are quite different:
-```{r compare-selected-models}
-compareFolds(cv.select)
-```
-
-## Cross-validating choice of transformations in regression
-
-The \pkg{cv} package also provides a \code{cvSelect()} procedure, \code{selectTrans()}, for choosing transformations of the predictors and the response in regression.
-
-Some background: As @Weisberg:2014 [Sec.
-8.2] explains, there are technical advantages to having (numeric) predictors in linear regression analysis that are themselves linearly related.
-If the predictors *aren't* linearly related, then the relationships between them can often be straightened by power transformations.
-Transformations can be selected after graphical examination of the data, or by analytic methods, such as transforming the predictors towards multivariate normality, which implies linearity.
-Once the relationships between the predictors are linearized, it can be advantageous similarly to transform the response variable towards normality.
-
-Selecting transformations analytically raises the possibility of automating the process, as would be required for cross-validation.
-One could, in principle, apply graphical methods to select transformations for each fold, but because a data analyst couldn't forget the choices made for previous folds, the process wouldn't really be applied independently to the folds.
-
-To illustrate, we adapt an example appearing in several places in @FoxWeisberg:2019 (for example in Chapter 3 on transforming data), using data on the prestige and other characteristics of 102 Canadian occupations circa 1970.
-The data are in the \code{Prestige} data frame in the \pkg{carData} package:
-```{r Prestige-data}
-data("Prestige", package="carData")
-summary(Prestige)
-```
-The variables in the \code{Prestige} data set are:
-
-* \code{education}: average years of education for incumbents in the occupation, from the 1971 Canadian Census.
-* \code{income}: average dollars of annual income for the occupation, from the Census.
-* \code{women}: percentage of occupational incumbents who were women, also from the Census.
-* \code{prestige}: the average prestige rating of the occupation on a 0--100 "thermometer" scale, from a Canadian social survey conducted around the same time.
-* \code{type}, type of occupation, and \code{census}, the Census occupational code, which are not used in our example.
-
-The object of a regression analysis for the \code{Prestige} data (and their original purpose) is to predict occupational prestige from the other variables in the data set.
-
-The scatterplot matrix in Figure \ref{fig:scatterplot-matrix} (a) (produced by the \code{scatterplotMatrix()} function in the \pkg{car} package) for the numeric variables in the data reveals that the distributions of \code{income} and \code{women} are positively skewed, and that some of the relationships among the three predictors, and between the predictors and the response (i.e., \code{prestige}), are nonlinear:
-```{r scatterplot-matrix, echo=FALSE, results=FALSE}
-#| fig.height = 5,
-#| fig.cap = "Scatterplot matrix for the Prestige data, (a) untransformed and (b) with the predictors transformed towards multivariate normality.",
-#| fig.subcap=rep("", 2),
-#| fig.ncol = 2, out.width = "50%", fig.align = "center"
-library("car", quietly=TRUE)
-scatterplotMatrix(~ prestige + income + education + women,
-                  data=Prestige, smooth=list(spread=FALSE))
-trans <- powerTransform( cbind(income, education, women) ~ 1,
-                         data=Prestige, family="yjPower")
-P <- Prestige[, c("prestige", "income", "education", "women")]
-(lambdas <- trans$roundlam)
-names(lambdas) <- c("income", "education", "women")
-for (var in c("income", "education", "women")){
-  P[, var] <- yjPower(P[, var], lambda=lambdas[var])
-}
-scatterplotMatrix(~ prestige + income + education + women,
-                  data=P, smooth=list(spread=FALSE))
-```
-
-```{r eval=FALSE}
-library("car", quietly=TRUE)
-scatterplotMatrix(~ prestige + income + education + women,
-                  data=Prestige, smooth=list(spread=FALSE))
-```
-
-The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964].
-Several "families" of power transformations can be used, including the original Box-Cox family, simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap.
-8] and @FoxWeisberg:2019 [Chap.
-3] for details.
-Because \code{women} has some 0 values, we use the Yeo-Johnson family:
-
-```{r transform-predictors}
-trans <- powerTransform( cbind(income, education, women) ~ 1,
-                         data=Prestige, family="yjPower")
-summary(trans)
-```
-
-We thus have evidence of the desirability of transforming \code{income} (by the $1/3$ power) and \code{women} (by the $0.16$ power---which is close to the "0" power, i.e., the log transformation), but not \code{education}.
-Applying the "rounded" power transformations makes the predictors better-behaved (cf., Figures \ref{fig:scatterplot-matrix} (a) and (b)):
-
-```{r eval=FALSE}
-P <- Prestige[, c("prestige", "income", "education", "women")]
-(lambdas <- trans$roundlam)
-names(lambdas) <- c("income", "education", "women")
-for (var in c("income", "education", "women")){
-  P[, var] <- yjPower(P[, var], lambda=lambdas[var])
-}
-scatterplotMatrix(~ prestige + income + education + women,
-                  data=P, smooth=list(spread=FALSE))
-```
-
-Comparing the MSE for the regressions with the original and transformed predictors shows an advantage to the latter:
-```{r prestige-regressions}
-m.pres <- lm(prestige ~ income + education + women, data=Prestige)
-m.pres.trans <- lm(prestige ~ income + education + women, data=P)
-mse(Prestige$prestige, fitted(m.pres))
-mse(P$prestige, fitted(m.pres.trans))
-```
-Similarly, component+residual plots for the two regressions in Figure \ref{fig:CR-plots}, produced by the \code{crPlots()} function in the \pkg{car} package, suggest that the partial relationship of \code{prestige} to \code{income} is more nearly linear in the transformed data, but the transformation of \code{women} fails to capture what appears to be a slight quadratic partial relationship; the partial relationship of \code{prestige} to \code{education} is close to linear in both regressions:
-
-```{r CR-plots}
-#| fig.height = 5,
-#| fig.cap = "Component+residual plots for the Prestige regression with (a) the original predictors and (b) transformed predictors",
-#| fig.subcap=rep("", 2),
-#| fig.ncol = 1, out.width = "60%", fig.align = "center"
-crPlots(m.pres)
-crPlots(m.pres.trans)
-```
-
-Having transformed the predictors towards multivariate normality, we now consider whether there's evidence for transforming the response (using \code{powerTransform()} for Box and Cox's original method), and we discover that there's not:
-```{r transform-response}
-summary(powerTransform(m.pres.trans))
-```
-
-The \code{selectTrans()} function in the \pkg{cv} package automates the process of selecting predictor and response transformations.
-The function takes a \code{data} set and "working" \code{model} as arguments, along with the candidate \code{predictors} and \code{response} for transformation, and the transformation \code{family} to employ.
-If the \code{predictors} argument is missing then only the \code{response} is transformed, and if the \code{response} argument is missing, only the supplied \code{predictors} are transformed.
-The default \code{family} for transforming the predictors is \code{"bcPower"}---the original Box-Cox family---as is the default \code{family.y} for transforming the response; here we specify \code{family="yjPower"} because of the 0s in \code{women}.
-\code{selectTrans()} returns the result of applying a lack-of-fit criterion to the model after the selected transformation is applied, with the default \code{criterion=mse}:
-```{r selectTrans}
-selectTrans(data=Prestige, model=m.pres,
-            predictors=c("income", "education", "women"),
-            response="prestige", family="yjPower")
-```
-\code{selectTrans()} also takes an optional \code{indices} argument, making it suitable for doing computations on a subset of the data (i.e., a CV fold), and hence for use with \code{cvSelect()} (see \code{?selectTrans} and the vignette on extending the \pkg{cv} package for details):
-
-```{r cv-select-transformations}
-cvs <- cvSelect(selectTrans, data=Prestige, model=m.pres, seed=1463,
-                predictors=c("income", "education", "women"),
-                response="prestige",
-                family="yjPower")
-cvs
-
-cv(m.pres, seed=1463) # untransformed model with same folds
-
-compareFolds(cvs)
-```
-The results suggest that the predictive power of the transformed regression is reliably greater than that of the untransformed regression (though in both case, the cross-validated MSE is considerably higher than the MSE computed for the whole data).
-Examining the selected transformations for each fold reveals that the predictor \code{education} and the response \code{prestige} are never transformed; that the $1/3$ power is selected for \code{income} in all of the folds; and that the transformation selected for \code{women} varies narrowly across the folds between the $0$th power (i.e., log) and the $1/3$ power.
-
-## Selecting both transformations and predictors[^Venables]
-
-[^Venables]: The presentation in the section benefits from an email conversation with Bill Venables, who of course isn't responsible for the use to which we've put his insightful remarks.
-
-As we mentioned, @HastieTibshiraniFriedman:2009 [Sec.
-7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain that honest cross-validation has to take account of model specification and selection.
-Statistical modeling is at least partly a craft, and one could imagine applying that craft to successive partial data sets, each with a fold removed.
+7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain, if the whole data are used to specify or fine-tune a statistical model, then subsequent cross-validation of the model is intrinsically misleading, because the model is selected to fit the whole data, including the part of the data that remains when each fold is removed.
+Statistical modeling is partly a craft, and one could imagine applying that craft to successive partial data sets, each with a fold removed.
 The resulting procedure would be tedious, though possibly worth the effort, but it would also be difficult to realize in practice: After all, we can hardly erase our memory of statistical modeling choices between analyzing partial data sets.
-
-Alternatively, if we're able to automate the process of model selection, then we can more realistically apply CV mechanically.
-That's what we did in the preceding two sections, first for predictor selection and then for selection of transformations in regression.
-In this section, we consider the case where we both choose variable transformations and then proceed to select predictors.
-It's insufficient to apply these steps sequentially, first, for example, using \code{cvSelect()} with \code{selectTrans()} and then with \code{selectStepAIC()}; rather, we should apply the whole model-selection procedure with each fold omitted.
-The \code{selectTransAndStepAIC()} function, also supplied by the \pkg{cv} package, does exactly that.
-
-To illustrate this process, we return to the \code{Auto} data set:
+Alternatively, if we're able to automate the process of model specification, then we can more realistically apply CV mechanically.
+
+The \code{cvSelect()} function in the \pkg{cv} package cross-validates a model-specification process in a general manner.
+Functions for three such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package [@FoxWeisberg:2019], transforms predictors and the response in a regression model towards normality; and \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially.
+In the vignette on extending the \pkg{cv} package, we explain how to add model-selection procedures.
+
+To illustrate cross-validation of model specification, we return to the \code{Auto} data set:[^Venables]
 ```{r Auto-redux}
@@
 ```
+
+[^Venables]: This example benefits from an email conversation with Bill Venables, who of course isn't responsible for the use to which we've put his insightful remarks.
+
 The \code{Auto} appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}.
 Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data.
-We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore}):
+We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
 ```{r Auto-explore}
@@
 
+library("car", quietly=TRUE)
 scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
@@
 
-We begin with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
+We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
 ```{r Auto-working-model}
@@
 m.auto <- lm(mpg ~ ., data = Auto)
-crPlots(m.auto)
+crPlots(m.auto, pch=".")
 ```
@@
 
-We proceed to transform the numeric predictors towards multi-normality: 
+Some background: As @Weisberg:2014 [Sec.
+8.2] explains, there are technical advantages to having (numeric) predictors in linear regression analysis that are themselves linearly related.
+If the predictors *aren't* linearly related, then the relationships between them can often be straightened by power transformations.
+Transformations can be selected after graphical examination of the data, or by analytic methods, such as transforming the predictors towards multivariate normality, which implies linearity.
+Once the relationships between the predictors are linearized, it can be advantageous similarly to transform the conditional distribution of the response variable towards normality.
+Selecting transformations analytically raises the possibility of automating the process, as would be required for cross-validation.
+
+The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964].
+Several "families" of power transformations can be used, including the original Box-Cox family (which is the default), simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and 0s: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap.
+8] and @FoxWeisberg:2019 [Chap.
+3] for details.
+We proceed to transform the numeric predictors in the \code{Auto} regression towards multivariate normality:
 ```{r Auto-transform}
@@
 ```
-Finally, we perform Box-Cox regression to transform the response (also obtaining a log transformation):
+
+Having transformed the predictors towards multivariate normality, we now consider whether there's evidence for transforming the response (using \code{powerTransform()} for Box and Cox's original method), also obtaining a log transformation:
 ```{r Auto-Box-Cox}
@@
 
-The transformed numeric variables are much better-behaved (cf., Figures \ref{fig:Auto-explore} on page \pageref{fig:Auto-explore} and \ref{fig:Auto-transformed-scatterplot-matrix}:
+The transformed numeric variables are much better-behaved (cf., Figures \ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and \ref{fig:Auto-transformed-scatterplot-matrix}):
 ```{r Auto-transformed-scatterplot-matrix}
@@
 ```
-And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figures \ref{fig:Auto-working-model} on page \pageref{fig:Auto-working-model}  and \ref{fig:Auto-CR-plots-transformed}:
+And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figures \ref{fig:Auto-working-model}, on page \pageref{fig:Auto-working-model}, and \ref{fig:Auto-CR-plots-transformed}):
 ```{r Auto-CR-plots-transformed}
@@
 #| fig.cap = "Component+residual plots for the model fit to the transformed Auto data"
-crPlots(m)
-```
-
-Having transformed both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log(n)$):
+crPlots(m, pch=".")
+```
+
+After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log(n)$):
 ```{r}
+library("MASS")
 m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
@@
 
-Now let's use \code{cvSelect()} with \code{selectTransAndStepAIC()} to automate and cross-validate the whole model-specification process:
+The \code{cvSelect()} function in the \pkg{cv} package allows us to cross-validate the whole model-selection procedure.
+The first argument to \code{cvSelect()}---here \code{selectTransAndStepAIC}---is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion:
 ```{r Auto-transform-and-select}
@@
 ```
-Here, as for \code{selectTrans()}, the \code{predictors} and \code{response} arguments specify candidate variables for transformation, and \code{AIC=FALSE} uses the BIC for model selection.
+The other arguments to \code{cvSelect()} are (see \code{?cvSelect} for additional optional arguments and details):
+
+* \code{data}, the data set to which the model is fit.
+* \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, then a seed is randomly selected and saved.
+* additional arguments required by the model-selection function, here the starting \code{model} for predictor selection; the names of the variables, \code{predictors} and \code{response}, that are candidates for transformation; and \code{AIC=FALSE}, which specifies use of the BIC for model selection.
 The starting model, \code{m.auto}, is the working model fit to the \code{Auto} data.

### 2024-01-12 scratch/notes_john/JSS-article/JSS-article-reduced-2.tex
add JSS-article-reduced-2

--- 2024-01-12 scratch/notes_john/JSS-article/JSS-article-reduced.tex
@@
 The \code{cvSelect()} function, discussed in Section
-\ref{cross-validating-model-selection}, cross-validates a complex
+\ref{cross-validating-model-specification}, cross-validates a complex
 model-specification process that may, for example, involve choice of
@@
 
-{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
+{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
 
@@
 
-{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/mpg-horsepower-MSE-se-1}
+{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/mpg-horsepower-MSE-se-1}
 
@@
 \begin{CodeOutput}
-R RNG seed set to 735559
+R RNG seed set to 392419
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.44017
-bias-adjusted cross-validation criterion = 19.41622
-95% CI for bias-adjusted CV criterion = (15.87764, 22.95479)
+cross-validation criterion = 19.16505
+bias-adjusted cross-validation criterion = 19.15554
+95% CI for bias-adjusted CV criterion = (15.71577, 22.59531)
 full-sample criterion = 18.98477 
@@
       expr        min         lq       mean     median         uq        max
- hatvalues    984.287   1153.412   1160.394   1189.902   1199.414   1285.104
-  Woodbury  10145.122  10213.592  10525.532  10463.385  10657.581  11476.351
-     naive 216360.403 217763.218 223882.308 218184.226 219846.551 273572.951
-    cv.glm 380361.674 382182.689 400581.980 386284.739 436401.540 439866.368
+ hatvalues    981.745   1172.477   1163.211   1189.943   1203.063   1325.817
+  Woodbury   9958.777  10178.496  10309.593  10232.370  10356.067  10998.127
+     naive 214083.345 217623.654 247100.940 220301.795 277217.933 331491.314
+    cv.glm 381699.053 383011.832 401940.298 387351.887 439080.439 442256.709
  neval cld
@@
 
-{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-2-1} }\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/polynomial-regression-CV-graph-2-2} }
+{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/polynomial-regression-CV-graph-2-1} }\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/polynomial-regression-CV-graph-2-2} }
 
@@
 
-{\centering \subfloat[\label{fig:plot1-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-1} }\subfloat[\label{fig:plot1-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-2} }\newline\subfloat[\label{fig:plot1-3}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-3} }\subfloat[\label{fig:plot1-4}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-4} }\newline\subfloat[\label{fig:plot1-5}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/plot1-5} }
+{\centering \subfloat[\label{fig:plot1-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-1} }\subfloat[\label{fig:plot1-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-2} }\newline\subfloat[\label{fig:plot1-3}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-3} }\subfloat[\label{fig:plot1-4}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-4} }\newline\subfloat[\label{fig:plot1-5}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-5} }
 
@@
 
-{\centering \subfloat[\label{fig:cross-validation-clusters-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/cross-validation-clusters-1} }\subfloat[\label{fig:cross-validation-clusters-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced_files/figure-latex/cross-validation-clusters-2} }
+{\centering \subfloat[\label{fig:cross-validation-clusters-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/cross-validation-clusters-1} }\subfloat[\label{fig:cross-validation-clusters-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/cross-validation-clusters-2} }
 
@@
 
-\hypertarget{cross-validating-model-selection}{%
+\hypertarget{cross-validating-model-specification}{%
 \section{Cross-validating model
-selection}\label{cross-validating-model-selection}}
+specification}\label{cross-validating-model-specification}}
 
@@
 Cross-validation'']{HastieTibshiraniFriedman:2009} explain, if the whole
-data are used to select or fine-tune a statistical model, then
+data are used to specify or fine-tune a statistical model, then
 subsequent cross-validation of the model is intrinsically misleading,
 because the model is selected to fit the whole data, including the part
-of the data that remains when each fold is removed.
-
-\hypertarget{a-preliminary-example}{%
-\subsection{A preliminary example}\label{a-preliminary-example}}
-
-The following example is similar in spirit to one employed by
-\citet{HastieTibshiraniFriedman:2009}. Suppose that we randomly generate
-\(n = 1000\) independent observations for a response variable variable
-\(y \sim N(\mu = 10, \sigma^2 = 1)\), and independently sample \(1000\)
-observations for \(p = 100\) ``predictors,'' \(x_1, \ldots, x_{100}\),
-each from \(x_j \sim N(0, 1)\). The response has nothing to do with the
-predictors and so the population linear-regression model
-\(y_i = \alpha + \beta_1 x_{i1} + \cdots + \beta_{100} x_{i,100} + \varepsilon_i\)
-has \(\alpha = 10\), all \(\beta_j = 0\), and
-\(\sigma_\varepsilon^2 = 1\).
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> set.seed(24361) # for reproducibility
-R> D <- data.frame(
-+   y = rnorm(1000, mean=10),
-+   X = matrix(rnorm(1000*100), 1000, 100)
-+ )
-\end{CodeInput}
-\end{CodeChunk}
-
-Least-squares provides accurate estimates of the regression constant
-\(\alpha = 10\) and the error variance \(\sigma^2 = 1\) for the ``null
-model'' including only the regression constant; moreover, the omnibus
-\(F\)-test of the correct null hypothesis that all of the \(\beta\)s are
-0 for the ``full model'' with all 100 \(x\)s is associated with a large
-\(p\)-value:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> m.full <- lm(y ~ ., data=D)
-R> m.null <- lm(y ~ 1, data=D)
-R> summary(m.null)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-lm(formula = y ~ 1, data = D)
-
-Residuals:
-    Min      1Q  Median      3Q     Max 
--3.4585 -0.6809  0.0190  0.6365  2.9346
-
-Coefficients:
-            Estimate Std. Error t value Pr(>|t|)
-(Intercept)  9.93704    0.03122   318.3   <2e-16
-
-Residual standard error: 0.9872 on 999 degrees of freedom
-\end{CodeOutput}
-\begin{CodeInput}
-R> anova(m.null, m.full)
-\end{CodeInput}
-\begin{CodeOutput}
-Analysis of Variance Table
-
-Model 1: y ~ 1
-Model 2: y ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7 + X.8 + X.9 + X.10 + 
-    X.11 + X.12 + X.13 + X.14 + X.15 + X.16 + X.17 + X.18 + X.19 + 
-    X.20 + X.21 + X.22 + X.23 + X.24 + X.25 + X.26 + X.27 + X.28 + 
-    X.29 + X.30 + X.31 + X.32 + X.33 + X.34 + X.35 + X.36 + X.37 + 
-    X.38 + X.39 + X.40 + X.41 + X.42 + X.43 + X.44 + X.45 + X.46 + 
-    X.47 + X.48 + X.49 + X.50 + X.51 + X.52 + X.53 + X.54 + X.55 + 
-    X.56 + X.57 + X.58 + X.59 + X.60 + X.61 + X.62 + X.63 + X.64 + 
-    X.65 + X.66 + X.67 + X.68 + X.69 + X.70 + X.71 + X.72 + X.73 + 
-    X.74 + X.75 + X.76 + X.77 + X.78 + X.79 + X.80 + X.81 + X.82 + 
-    X.83 + X.84 + X.85 + X.86 + X.87 + X.88 + X.89 + X.90 + X.91 + 
-    X.92 + X.93 + X.94 + X.95 + X.96 + X.97 + X.98 + X.99 + X.100
-  Res.Df    RSS  Df Sum of Sq      F Pr(>F)
-1    999 973.65                            
-2    899 888.44 100    85.208 0.8622  0.825
-\end{CodeOutput}
-\end{CodeChunk}
-
-Next, using the \code{stepAIC()} function in the \pkg{MASS} package
-\citep{VenablesRipley:2002}, let us perform a forward stepwise
-regression to select a ``best'' model, starting with the null model, and
-using AIC as the model-selection criterion (see \code{?stepAIC()} for
-details):\footnote{It's generally advantageous to start with the largest
-  model, here the one with 100 predictors, and proceed by backward
-  elimination. In this demonstration, however, where all of the
-  \(\beta\)s are really 0, the selected model will be small, and so we
-  proceed by forward selection from the null model to save computing
-  time.}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("MASS")  # for stepAIC()
-R> m.select <- stepAIC(m.null,
-+                     direction="forward", trace=FALSE,
-+                     scope=list(lower=~1, upper=formula(m.full)))
-R> summary(m.select)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-lm(formula = y ~ X.99 + X.90 + X.87 + X.40 + X.65 + X.91 + X.53 + 
-    X.45 + X.31 + X.56 + X.61 + X.60 + X.46 + X.35 + X.92, data = D)
-
-Residuals:
-    Min      1Q  Median      3Q     Max 
--3.2620 -0.6446  0.0236  0.6406  3.1180
-
-Coefficients:
-            Estimate Std. Error t value Pr(>|t|)
-(Intercept)  9.93716    0.03098 320.803  < 2e-16
-X.99        -0.09103    0.03082  -2.953  0.00322
-X.90        -0.08205    0.03135  -2.617  0.00901
-X.87        -0.06942    0.03105  -2.235  0.02561
-X.40        -0.04759    0.03076  -1.547  0.12211
-X.65        -0.05523    0.03147  -1.755  0.07952
-X.91         0.05245    0.03084   1.700  0.08937
-X.53        -0.04921    0.03048  -1.615  0.10672
-X.45         0.05543    0.03182   1.742  0.08183
-X.31         0.04525    0.03108   1.456  0.14570
-X.56         0.05433    0.03273   1.660  0.09723
-X.61        -0.05085    0.03170  -1.604  0.10908
-X.60        -0.05133    0.03194  -1.607  0.10832
-X.46         0.05158    0.03272   1.576  0.11529
-X.35         0.04696    0.03146   1.493  0.13584
-X.92         0.04430    0.03100   1.429  0.15329
-
-Residual standard error: 0.9725 on 984 degrees of freedom
-Multiple R-squared:  0.04419,   Adjusted R-squared:  0.02962 
-F-statistic: 3.033 on 15 and 984 DF,  p-value: 8.338e-05
-\end{CodeOutput}
-\begin{CodeInput}
-R> mse(D$y, fitted(m.select))
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0.9306254
-attr(,"casewise loss")
-[1] "(y - yhat)^2"
-\end{CodeOutput}
-\end{CodeChunk}
-
-The resulting model has 15 predictors, a very modest \(R^2 = .044\), but
-a small \(p\)-value for its omnibus \(F\)-test (which, of course, is
-entirely spurious because the same data were used to select and test the
-model). The MSE for the selected model is smaller than the true error
-variance \(\sigma^2 = 1\), as is the estimated error variance for the
-selected model, \(\widehat{\sigma}^2 = 0.973^2 = 0.947\).\footnote{The
-  \code{"casewise loss"} attribute attached to the value returned by the
-  \code{mse()} function signals to \code{cv()} that MSE is an average of
-  casewise components (and how to calculate the casewise components),
-  which has implications for computing bias correction and confidences
-  intervals; see Section
-  \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}.}
-
-If we cross-validate the selected model, we also obtain an optimistic
-estimate of its predictive power (although the confidence interval for
-the bias-adjusted MSE includes 1):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(m.select, seed=2529)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 2529
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-method: Woodbury
-criterion: mse
-cross-validation criterion = 0.9593695
-bias-adjusted cross-validation criterion = 0.9578478
-95% CI for bias-adjusted CV criterion = (0.8766138, 1.039082)
-full-sample criterion = 0.9306254 
-\end{CodeOutput}
-\end{CodeChunk}
-
-The \code{cvSelect()} function in the \pkg{cv} package, in contrast,
-allows us to cross-validate the whole model-selection procedure. The
-first argument to \code{cvSelect()} is a model-selection function
-capable of refitting the model with a fold omitted and returning a CV
-criterion. The \code{selectStepAIC()} function, also in \pkg{cv} and
-based on \code{stepAIC()}, is suitable for use with \code{cvSelect()}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv.select <- cvSelect(selectStepAIC, data=D, seed=3791,
-+                       model=m.null, direction="forward",
-+                       scope=list(lower=~1, 
-+                                  upper=formula(m.full)))
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 3791
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv.select
-\end{CodeInput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-cross-validation criterion = 1.06873
-bias-adjusted cross-validation criterion = 1.061183
-95% CI for bias-adjusted CV criterion = (0.9717229, 1.150642)
-full-sample criterion = 0.9306254 
-\end{CodeOutput}
-\end{CodeChunk}
-
-The other arguments to \code{cvSelect()} are:
-
-\begin{itemize}
-\tightlist
-\item
-  \code{data}, the data set to which the model is fit.
-\item
-  \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number
-  generator; as for \code{cv()}, if the seed isn't supplied by the user,
-  a seed is randomly selected and saved.
-\item
-  additional arguments required by the model-selection function, here
-  the starting \code{model} argument, the \code{direction} of model
-  selection, and the \code{scope} of models considered (from the model
-  with only a regression constant to the model with all 100 predictors).
-\end{itemize}
-
-By default, \code{cvSelect()} performs 10-fold CV, and it produces an
-estimate of MSE for the model-selection procedure even \emph{larger}
-than the true error variance, \(\sigma^2 = 1\).
-
-Also by default, when the number of folds is 10 or fewer,
-\code{cvSelect()} saves the coefficients of the selected models. In this
-example, the \code{compareFolds()} function reveals that the variables
-retained by the model-selection process in the several folds are quite
-different:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> compareFolds(cv.select)
-\end{CodeInput}
-\begin{CodeOutput}
-        (Intercept)    X.87    X.90    X.99    X.91    X.54    X.53    X.56
-Fold 1       9.9187 -0.0615 -0.0994 -0.0942  0.0512  0.0516                
-Fold 2       9.9451 -0.0745 -0.0899 -0.0614          0.0587          0.0673
-Fold 3       9.9423 -0.0783 -0.0718 -0.0987  0.0601                  0.0512
-Fold 4       9.9410 -0.0860 -0.0831 -0.0867  0.0570         -0.0508        
-Fold 5       9.9421 -0.0659 -0.0849 -0.1004  0.0701  0.0511 -0.0487  0.0537
-Fold 6       9.9633 -0.0733 -0.0874 -0.0960  0.0555  0.0629 -0.0478        
-Fold 7       9.9279 -0.0618 -0.0960 -0.0838  0.0533         -0.0464        
-Fold 8       9.9453 -0.0610 -0.0811 -0.0818          0.0497 -0.0612  0.0560
-Fold 9       9.9173 -0.0663 -0.0894 -0.1100  0.0504  0.0524          0.0747
-Fold 10      9.9449 -0.0745 -0.0906 -0.0891  0.0535  0.0482 -0.0583  0.0642
-           X.40    X.45    X.65    X.68    X.92    X.15    X.26    X.46    X.60
-Fold 1                  -0.0590                 -0.0456  0.0658  0.0608        
-Fold 2                                   0.0607          0.0487                
-Fold 3  -0.0496         -0.0664          0.0494                                
-Fold 4  -0.0597  0.0579 -0.0531          0.0519 -0.0566                 -0.0519
-Fold 5                           0.0587                          0.0527 -0.0603
-Fold 6  -0.0596  0.0552          0.0474                                        
-Fold 7           0.0572          0.0595                                        
-Fold 8           0.0547 -0.0617  0.0453  0.0493 -0.0613  0.0591  0.0703 -0.0588
-Fold 9  -0.0552  0.0573 -0.0635  0.0492         -0.0513  0.0484         -0.0507
-Fold 10 -0.0558                          0.0529                  0.0710        
-           X.61     X.8    X.28    X.29    X.31    X.35    X.70    X.89    X.17
-Fold 1  -0.0490          0.0616 -0.0537                  0.0638                
-Fold 2           0.0671                  0.0568                  0.0523        
-Fold 3  -0.0631          0.0616                                                
-Fold 4           0.0659         -0.0549          0.0527                  0.0527
-Fold 5           0.0425                  0.0672  0.0613          0.0493        
-Fold 6           0.0559         -0.0629  0.0498          0.0487                
-Fold 7                                                           0.0611  0.0472
-Fold 8  -0.0719                                          0.0586                
-Fold 9                   0.0525                                                
-Fold 10 -0.0580                                  0.0603                        
-           X.25     X.4    X.64    X.81    X.97    X.11     X.2    X.33    X.47
-Fold 1                                   0.0604          0.0575                
-Fold 2   0.0478          0.0532  0.0518                                        
-Fold 3                           0.0574                          0.0473        
-Fold 4                   0.0628                                                
-Fold 5   0.0518                                                                
-Fold 6                                           0.0521                        
-Fold 7           0.0550                                                        
-Fold 8                                                                         
-Fold 9                                   0.0556                          0.0447
-Fold 10          0.0516                                                        
-            X.6    X.72    X.73    X.77    X.79 X.88
-Fold 1   0.0476                                     
-Fold 2                   0.0514                     
-Fold 3                                              
-Fold 4                                  -0.0473     
-Fold 5           0.0586                         0.07
-Fold 6                          -0.0489             
-Fold 7                                              
-Fold 8                                              
-Fold 9                                              
-Fold 10                                             
-\end{CodeOutput}
-\end{CodeChunk}
-
-\hypertarget{cross-validating-choice-of-transformations-in-regression}{%
-\subsection{Cross-validating choice of transformations in
-regression}\label{cross-validating-choice-of-transformations-in-regression}}
-
-The \pkg{cv} package also provides a \code{cvSelect()} procedure,
-\code{selectTrans()}, for choosing transformations of the predictors and
-the response in regression.
-
-Some background: As \citet[Sec. 8.2]{Weisberg:2014} explains, there are
-technical advantages to having (numeric) predictors in linear regression
-analysis that are themselves linearly related. If the predictors
-\emph{aren't} linearly related, then the relationships between them can
-often be straightened by power transformations. Transformations can be
-selected after graphical examination of the data, or by analytic
-methods, such as transforming the predictors towards multivariate
-normality, which implies linearity. Once the relationships between the
-predictors are linearized, it can be advantageous similarly to transform
-the response variable towards normality.
-
-Selecting transformations analytically raises the possibility of
-automating the process, as would be required for cross-validation. One
-could, in principle, apply graphical methods to select transformations
-for each fold, but because a data analyst couldn't forget the choices
-made for previous folds, the process wouldn't really be applied
-independently to the folds.
-
-To illustrate, we adapt an example appearing in several places in
-\citet{FoxWeisberg:2019} (for example in Chapter 3 on transforming
-data), using data on the prestige and other characteristics of 102
-Canadian occupations circa 1970. The data are in the \code{Prestige}
-data frame in the \pkg{carData} package:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> data("Prestige", package="carData")
-R> summary(Prestige)
-\end{CodeInput}
-\begin{CodeOutput}
-   education          income          women           prestige    
- Min.   : 6.380   Min.   :  611   Min.   : 0.000   Min.   :14.80  
- 1st Qu.: 8.445   1st Qu.: 4106   1st Qu.: 3.592   1st Qu.:35.23  
- Median :10.540   Median : 5930   Median :13.600   Median :43.60  
- Mean   :10.738   Mean   : 6798   Mean   :28.979   Mean   :46.83  
- 3rd Qu.:12.648   3rd Qu.: 8187   3rd Qu.:52.203   3rd Qu.:59.27  
- Max.   :15.970   Max.   :25879   Max.   :97.510   Max.   :87.20  
-     census       type   
- Min.   :1113   bc  :44  
- 1st Qu.:3120   prof:31  
- Median :5135   wc  :23  
- Mean   :5402   NA's: 4  
- 3rd Qu.:8312            
- Max.   :9517            
-\end{CodeOutput}
-\end{CodeChunk}
-
-The variables in the \code{Prestige} data set are:
-
-\begin{itemize}
-\tightlist
-\item
-  \code{education}: average years of education for incumbents in the
-  occupation, from the 1971 Canadian Census.
-\item
-  \code{income}: average dollars of annual income for the occupation,
-  from the Census.
-\item
-  \code{women}: percentage of occupational incumbents who were women,
-  also from the Census.
-\item
-  \code{prestige}: the average prestige rating of the occupation on a
-  0--100 ``thermometer'' scale, from a Canadian social survey conducted
-  around the same time.
-\item
-  \code{type}, type of occupation, and \code{census}, the Census
-  occupational code, which are not used in our example.
-\end{itemize}
-
-The object of a regression analysis for the \code{Prestige} data (and
-their original purpose) is to predict occupational prestige from the
-other variables in the data set.
-
-The scatterplot matrix in Figure \ref{fig:scatterplot-matrix} (a)
-(produced by the \code{scatterplotMatrix()} function in the \pkg{car}
-package) for the numeric variables in the data reveals that the
-distributions of \code{income} and \code{women} are positively skewed,
-and that some of the relationships among the three predictors, and
-between the predictors and the response (i.e., \code{prestige}), are
-nonlinear:
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \subfloat[\label{fig:scatterplot-matrix-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/scatterplot-matrix-1} }\subfloat[\label{fig:scatterplot-matrix-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced_files/figure-latex/scatterplot-matrix-2} }
-
-}
-
-\caption[Scatterplot matrix for the Prestige data, (a) untransformed and (b) with the predictors transformed towards multivariate normality]{Scatterplot matrix for the Prestige data, (a) untransformed and (b) with the predictors transformed towards multivariate normality.}\label{fig:scatterplot-matrix}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("car", quietly=TRUE)
-R> scatterplotMatrix(~ prestige + income + education + women,
-+                   data=Prestige, smooth=list(spread=FALSE))
-\end{CodeInput}
-\end{CodeChunk}
-
-The \code{powerTransform()} function in the \pkg{car} package transforms
-variables towards multivariate normality by a generalization of Box and
-Cox's maximum-likelihood-like approach \citep{BoxCox:1964}. Several
-``families'' of power transformations can be used, including the
-original Box-Cox family, simple powers (and roots), and two adaptations
-of the Box-Cox family to data that may include negative values and
-zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see
-\citet[Chap. 8]{Weisberg:2014} and \citet[Chap. 3]{FoxWeisberg:2019} for
-details. Because \code{women} has some 0 values, we use the Yeo-Johnson
-family:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> trans <- powerTransform( cbind(income, education, women) ~ 1,
-+                          data=Prestige, family="yjPower")
-R> summary(trans)
-\end{CodeInput}
-\begin{CodeOutput}
-yjPower Transformations to Multinormality 
-          Est Power Rounded Pwr Wald Lwr Bnd Wald Upr Bnd
-income       0.2678        0.33       0.1051       0.4304
-education    0.5162        1.00      -0.2822       1.3145
-women        0.1630        0.16       0.0112       0.3149
-
- Likelihood ratio test that all transformation parameters are equal to 0
-                               LRT df      pval
-LR test, lambda = (0 0 0) 15.73879  3 0.0012827
-\end{CodeOutput}
-\end{CodeChunk}
-
-We thus have evidence of the desirability of transforming \code{income}
-(by the \(1/3\) power) and \code{women} (by the \(0.16\) power---which
-is close to the ``0'' power, i.e., the log transformation), but not
-\code{education}. Applying the ``rounded'' power transformations makes
-the predictors better-behaved (cf., Figures \ref{fig:scatterplot-matrix}
-(a) and (b)):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> P <- Prestige[, c("prestige", "income", "education", "women")]
-R> (lambdas <- trans$roundlam)
-R> names(lambdas) <- c("income", "education", "women")
-R> for (var in c("income", "education", "women")){
-+   P[, var] <- yjPower(P[, var], lambda=lambdas[var])
-+ }
-R> scatterplotMatrix(~ prestige + income + education + women,
-+                   data=P, smooth=list(spread=FALSE))
-\end{CodeInput}
-\end{CodeChunk}
-
-Comparing the MSE for the regressions with the original and transformed
-predictors shows an advantage to the latter:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> m.pres <- lm(prestige ~ income + education + women, data=Prestige)
-R> m.pres.trans <- lm(prestige ~ income + education + women, data=P)
-R> mse(Prestige$prestige, fitted(m.pres))
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 59.15265
-attr(,"casewise loss")
-[1] "(y - yhat)^2"
-\end{CodeOutput}
-\begin{CodeInput}
-R> mse(P$prestige, fitted(m.pres.trans))
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 50.60016
-attr(,"casewise loss")
-[1] "(y - yhat)^2"
-\end{CodeOutput}
-\end{CodeChunk}
-
-Similarly, component+residual plots for the two regressions in Figure
-\ref{fig:CR-plots}, produced by the \code{crPlots()} function in the
-\pkg{car} package, suggest that the partial relationship of
-\code{prestige} to \code{income} is more nearly linear in the
-transformed data, but the transformation of \code{women} fails to
-capture what appears to be a slight quadratic partial relationship; the
-partial relationship of \code{prestige} to \code{education} is close to
-linear in both regressions:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> crPlots(m.pres)
-R> crPlots(m.pres.trans)
-\end{CodeInput}
-\begin{figure}
-
-{\centering \subfloat[\label{fig:CR-plots-1}]{\includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/CR-plots-1} }\newline\subfloat[\label{fig:CR-plots-2}]{\includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/CR-plots-2} }\newline
-
-}
-
-\caption[Component+residual plots for the Prestige regression with (a) the original predictors and (b) transformed predictors]{Component+residual plots for the Prestige regression with (a) the original predictors and (b) transformed predictors}\label{fig:CR-plots}
-\end{figure}
-\end{CodeChunk}
-
-Having transformed the predictors towards multivariate normality, we now
-consider whether there's evidence for transforming the response (using
-\code{powerTransform()} for Box and Cox's original method), and we
-discover that there's not:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> summary(powerTransform(m.pres.trans))
-\end{CodeInput}
-\begin{CodeOutput}
-bcPower Transformation to Normality 
-   Est Power Rounded Pwr Wald Lwr Bnd Wald Upr Bnd
-Y1    1.0194           1       0.6773       1.3615
-
-Likelihood ratio test that transformation parameter is equal to 0
- (log transformation)
-                          LRT df       pval
-LR test, lambda = (0) 32.2174  1 1.3785e-08
-
-Likelihood ratio test that no transformation is needed
-                             LRT df    pval
-LR test, lambda = (1) 0.01238421  1 0.91139
-\end{CodeOutput}
-\end{CodeChunk}
-
-The \code{selectTrans()} function in the \pkg{cv} package automates the
-process of selecting predictor and response transformations. The
-function takes a \code{data} set and ``working'' \code{model} as
-arguments, along with the candidate \code{predictors} and
-\code{response} for transformation, and the transformation \code{family}
-to employ. If the \code{predictors} argument is missing then only the
-\code{response} is transformed, and if the \code{response} argument is
-missing, only the supplied \code{predictors} are transformed. The
-default \code{family} for transforming the predictors is
-\code{"bcPower"}---the original Box-Cox family---as is the default
-\code{family.y} for transforming the response; here we specify
-\code{family="yjPower"} because of the 0s in \code{women}.
-\code{selectTrans()} returns the result of applying a lack-of-fit
-criterion to the model after the selected transformation is applied,
-with the default \code{criterion=mse}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> selectTrans(data=Prestige, model=m.pres,
-+             predictors=c("income", "education", "women"),
-+             response="prestige", family="yjPower")
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 50.60016
-attr(,"casewise loss")
-[1] "(y - yhat)^2"
-\end{CodeOutput}
-\end{CodeChunk}
-
-\code{selectTrans()} also takes an optional \code{indices} argument,
-making it suitable for doing computations on a subset of the data (i.e.,
-a CV fold), and hence for use with \code{cvSelect()} (see
-\code{?selectTrans} and the vignette on extending the \pkg{cv} package
-for details):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cvs <- cvSelect(selectTrans, data=Prestige, model=m.pres, seed=1463,
-+                 predictors=c("income", "education", "women"),
-+                 response="prestige",
-+                 family="yjPower")
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 1463
-\end{CodeOutput}
-\begin{CodeInput}
-R> cvs
-\end{CodeInput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-cross-validation criterion = 54.4871
-bias-adjusted cross-validation criterion = 54.30824
-full-sample criterion = 50.60016 
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.pres, seed=1463) # untransformed model with same folds
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 1463
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-method: Woodbury
-criterion: mse
-cross-validation criterion = 63.2926
-bias-adjusted cross-validation criterion = 63.07251
-full-sample criterion = 59.15265 
-\end{CodeOutput}
-\begin{CodeInput}
-R> compareFolds(cvs)
-\end{CodeInput}
-\begin{CodeOutput}
-        lam.education lam.income lam.women lambda
-Fold 1          1.000      0.330     0.330      1
-Fold 2          1.000      0.330     0.169      1
-Fold 3          1.000      0.330     0.330      1
-Fold 4          1.000      0.330     0.330      1
-Fold 5          1.000      0.330     0.000      1
-Fold 6          1.000      0.330     0.330      1
-Fold 7          1.000      0.330     0.330      1
-Fold 8          1.000      0.330     0.000      1
-Fold 9          1.000      0.330     0.000      1
-Fold 10         1.000      0.330     0.000      1
-\end{CodeOutput}
-\end{CodeChunk}
-
-The results suggest that the predictive power of the transformed
-regression is reliably greater than that of the untransformed regression
-(though in both case, the cross-validated MSE is considerably higher
-than the MSE computed for the whole data). Examining the selected
-transformations for each fold reveals that the predictor
-\code{education} and the response \code{prestige} are never transformed;
-that the \(1/3\) power is selected for \code{income} in all of the
-folds; and that the transformation selected for \code{women} varies
-narrowly across the folds between the \(0\)th power (i.e., log) and the
-\(1/3\) power.
-
-\hypertarget{selecting-both-transformations-and-predictorsvenables}{%
-\subsection[Selecting both transformations and
-predictors]{\texorpdfstring{Selecting both transformations and
-predictors\footnote{The presentation in the section benefits from an
-  email conversation with Bill Venables, who of course isn't responsible
-  for the use to which we've put his insightful remarks.}}{Selecting both transformations and predictors}}\label{selecting-both-transformations-and-predictorsvenables}}
-
-As we mentioned, \citet[Sec. 7.10.2: ``The Wrong and Right Way to Do
-Cross-validation'']{HastieTibshiraniFriedman:2009} explain that honest
-cross-validation has to take account of model specification and
-selection. Statistical modeling is at least partly a craft, and one
-could imagine applying that craft to successive partial data sets, each
-with a fold removed. The resulting procedure would be tedious, though
-possibly worth the effort, but it would also be difficult to realize in
-practice: After all, we can hardly erase our memory of statistical
-modeling choices between analyzing partial data sets.
-
-Alternatively, if we're able to automate the process of model selection,
-then we can more realistically apply CV mechanically. That's what we did
-in the preceding two sections, first for predictor selection and then
-for selection of transformations in regression. In this section, we
-consider the case where we both choose variable transformations and then
-proceed to select predictors. It's insufficient to apply these steps
-sequentially, first, for example, using \code{cvSelect()} with
-\code{selectTrans()} and then with \code{selectStepAIC()}; rather, we
-should apply the whole model-selection procedure with each fold omitted.
-The \code{selectTransAndStepAIC()} function, also supplied by the
-\pkg{cv} package, does exactly that.
-
-To illustrate this process, we return to the \code{Auto} data set:
+of the data that remains when each fold is removed. Statistical modeling
+is partly a craft, and one could imagine applying that craft to
+successive partial data sets, each with a fold removed. The resulting
+procedure would be tedious, though possibly worth the effort, but it
+would also be difficult to realize in practice: After all, we can hardly
+erase our memory of statistical modeling choices between analyzing
+partial data sets. Alternatively, if we're able to automate the process
+of model specification, then we can more realistically apply CV
+mechanically.
+
+The \code{cvSelect()} function in the \pkg{cv} package cross-validates a
+model-specification process in a general manner. Functions for three
+such model-specification processes are included in the package:
+\code{selectStepAIC()}, based on the \code{stepAIC()} function in the
+\pkg{MASS} package \citep{VenablesRipley:2002}, performs stepwise
+predictor selection for regression models; \code{selectTrans()}, based
+on the \code{powerTransform()} function in the \pkg{car} package
+\citep{FoxWeisberg:2019}, transforms predictors and the response in a
+regression model towards normality; and
+\code{selectTransAndStepAIC()}---the use of which we illustrate in the
+current section---performs both of these procedures sequentially. In the
+vignette on extending the \pkg{cv} package, we explain how to add
+model-selection procedures.
+
+To illustrate cross-validation of model specification, we return to the
+\code{Auto} data set:\footnote{This example benefits from an email
+  conversation with Bill Venables, who of course isn't responsible for
+  the use to which we've put his insightful remarks.}
 
@@
 examine the pairwise relationships among the numeric variables in the
-data set (Figure \ref{fig:Auto-explore}):
+data set (Figure \ref{fig:Auto-explore}, produced by the
+\code{scatterplotMatrix()} function in the \pkg{car} package):
 
@@
 R> 
+R> library("car", quietly=TRUE)
 R> scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/Auto-explore-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-explore-1}
 
@@
 
-We begin with a ``working model'' that specifies linear partial
+We start with a ``working model'' that specifies linear partial
 relationships of the response to the numeric predictors:
@@
 R> m.auto <- lm(mpg ~ ., data = Auto)
-R> crPlots(m.auto)
+R> crPlots(m.auto, pch=".")
 \end{CodeInput}
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/Auto-working-model-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-working-model-1}
 
@@
 
-We proceed to transform the numeric predictors towards multi-normality:
+Some background: As \citet[Sec. 8.2]{Weisberg:2014} explains, there are
+technical advantages to having (numeric) predictors in linear regression
+analysis that are themselves linearly related. If the predictors
+\emph{aren't} linearly related, then the relationships between them can
+often be straightened by power transformations. Transformations can be
+selected after graphical examination of the data, or by analytic
+methods, such as transforming the predictors towards multivariate
+normality, which implies linearity. Once the relationships between the
+predictors are linearized, it can be advantageous similarly to transform
+the conditional distribution of the response variable towards normality.
+Selecting transformations analytically raises the possibility of
+automating the process, as would be required for cross-validation.
+
+The \code{powerTransform()} function in the \pkg{car} package transforms
+variables towards multivariate normality by a generalization of Box and
+Cox's maximum-likelihood-like approach \citep{BoxCox:1964}. Several
+``families'' of power transformations can be used, including the
+original Box-Cox family (which is the default), simple powers (and
+roots), and two adaptations of the Box-Cox family to data that may
+include negative values and 0s: the Box-Cox-with-negatives family and
+the Yeo-Johnson family; see \citet[Chap. 8]{Weisberg:2014} and
+\citet[Chap. 3]{FoxWeisberg:2019} for details. We proceed to transform
+the numeric predictors in the \code{Auto} regression towards
+multivariate normality:
 
@@
 
-Finally, we perform Box-Cox regression to transform the response (also
-obtaining a log transformation):
+Having transformed the predictors towards multivariate normality, we now
+consider whether there's evidence for transforming the response (using
+\code{powerTransform()} for Box and Cox's original method), also
+obtaining a log transformation:
 
@@
 The transformed numeric variables are much better-behaved (cf., Figures
-\ref{fig:Auto-explore} on page \pageref{fig:Auto-explore} and
-\ref{fig:Auto-transformed-scatterplot-matrix}:
+\ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and
+\ref{fig:Auto-transformed-scatterplot-matrix}):
 
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/Auto-transformed-scatterplot-matrix-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-transformed-scatterplot-matrix-1}
 
@@
 And the partial relationships in the model fit to the transformed data
-are much more nearly linear (cf., Figures \ref{fig:Auto-working-model}
-on page \pageref{fig:Auto-working-model} and
-\ref{fig:Auto-CR-plots-transformed}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> crPlots(m)
+are much more nearly linear (cf., Figures \ref{fig:Auto-working-model},
+on page \pageref{fig:Auto-working-model}, and
+\ref{fig:Auto-CR-plots-transformed}):
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> crPlots(m, pch=".")
 \end{CodeInput}
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced_files/figure-latex/Auto-CR-plots-transformed-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-CR-plots-transformed-1}
 
@@
 
-Having transformed both the numeric predictors and the response, we
+After transforming both the numeric predictors and the response, we
 proceed to use the \code{stepAIC()} function in the \pkg{MASS} package
@@
 \begin{CodeInput}
+R> library("MASS")
 R> m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
@@
 
-Now let's use \code{cvSelect()} with \code{selectTransAndStepAIC()} to
-automate and cross-validate the whole model-specification process:
+The \code{cvSelect()} function in the \pkg{cv} package allows us to
+cross-validate the whole model-selection procedure. The first argument
+to \code{cvSelect()}---here \code{selectTransAndStepAIC}---is a
+model-selection function capable of refitting the model with a fold
+omitted and returning a CV criterion:
 
@@
 
-Here, as for \code{selectTrans()}, the \code{predictors} and
-\code{response} arguments specify candidate variables for
-transformation, and \code{AIC=FALSE} uses the BIC for model selection.
-The starting model, \code{m.auto}, is the working model fit to the
-\code{Auto} data.
+The other arguments to \code{cvSelect()} are (see \code{?cvSelect} for
+additional optional arguments and details):
+
+\begin{itemize}
+\tightlist
+\item
+  \code{data}, the data set to which the model is fit.
+\item
+  \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number
+  generator; as for \code{cv()}, if the seed isn't supplied by the user,
+  then a seed is randomly selected and saved.
+\item
+  additional arguments required by the model-selection function, here
+  the starting \code{model} for predictor selection; the names of the
+  variables, \code{predictors} and \code{response}, that are candidates
+  for transformation; and \code{AIC=FALSE}, which specifies use of the
+  BIC for model selection. The starting model, \code{m.auto}, is the
+  working model fit to the \code{Auto} data.
+\end{itemize}
 
@@
 
-{\centering \includegraphics{JSS-article-reduced_files/figure-latex/BEPS-plot-1}
+{\centering \includegraphics{JSS-article-reduced-2_files/figure-latex/BEPS-plot-1}
 
@@
 }
-<bytecode: 0x13db9d388>
+<bytecode: 0x13ca97218>
 <environment: namespace:cv>

### 2024-01-13 scratch/notes_john/JSS-article/JSS-article-reduced-2.Rmd
further edits to JSS draft

@@
 
-The \code{Auto} appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}.
+The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}.
 Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data.
@@
 
-[^year]: Of course, making the decision to treat \code{year} as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process.
+[^year]: Making the decision to treat \code{year} as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process.
 It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation.
@@
 * \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, then a seed is randomly selected and saved.
-* additional arguments required by the model-selection function, here the starting \code{model} for predictor selection; the names of the variables, \code{predictors} and \code{response}, that are candidates for transformation; and \code{AIC=FALSE}, which specifies use of the BIC for model selection.
-The starting model, \code{m.auto}, is the working model fit to the \code{Auto} data.
+* Arguments required by the model-selection function: the starting \code{model} (\code{m.auto}) for predictor selection, the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation, and \code{AIC=FALSE}, which specifies use of the BIC for model selection.
 

### 2024-01-13 scratch/notes_john/JSS-article/JSS-article-reduced-2.tex
further edits to JSS draft

@@
 \begin{CodeOutput}
-R RNG seed set to 392419
+R RNG seed set to 352413
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.16505
-bias-adjusted cross-validation criterion = 19.15554
-95% CI for bias-adjusted CV criterion = (15.71577, 22.59531)
+cross-validation criterion = 19.42674
+bias-adjusted cross-validation criterion = 19.40347
+95% CI for bias-adjusted CV criterion = (15.89518, 22.91177)
 full-sample criterion = 18.98477 
@@
 
-The \code{Auto} appeared in a preliminary example in Section
+The \code{Auto} data appeared in a preliminary example in Section
 \ref{preliminary-example-polynomial-regression}, where we employed CV to
@@
 1982 in the data, and the relationship between \code{mpg} and
-\code{year} is difficult to characterize.\footnote{Of course, making the
-  decision to treat \code{year} as a factor on this basis could be
-  construed as cheating in the current context, which illustrates the
-  difficulty of automating the whole model-selection process. It's
-  rarely desirable, in our opinion, to forgo exploration of the data to
-  ensure the purity of model validation. We believe, however, that it's
-  still useful to automate as much of the process as we can to obtain a
-  more realistic, if still biased, estimate of the predictive power of a
-  model.} It's apparent that most these variables are positively skewed
-and that many of the pairwise relationships among them are nonlinear.
+\code{year} is difficult to characterize.\footnote{Making the decision
+  to treat \code{year} as a factor on this basis could be construed as
+  cheating in the current context, which illustrates the difficulty of
+  automating the whole model-selection process. It's rarely desirable,
+  in our opinion, to forgo exploration of the data to ensure the purity
+  of model validation. We believe, however, that it's still useful to
+  automate as much of the process as we can to obtain a more realistic,
+  if still biased, estimate of the predictive power of a model.} It's
+apparent that most these variables are positively skewed and that many
+of the pairwise relationships among them are nonlinear.
 
@@
 \item
-  additional arguments required by the model-selection function, here
-  the starting \code{model} for predictor selection; the names of the
-  variables, \code{predictors} and \code{response}, that are candidates
-  for transformation; and \code{AIC=FALSE}, which specifies use of the
-  BIC for model selection. The starting model, \code{m.auto}, is the
-  working model fit to the \code{Auto} data.
+  Arguments required by the model-selection function: the starting
+  \code{model} (\code{m.auto}) for predictor selection, the names of the
+  variables---\code{predictors} and \code{response}---that are
+  candidates for transformation, and \code{AIC=FALSE}, which specifies
+  use of the BIC for model selection.
 \end{itemize}
@@
 }
-<bytecode: 0x13ca97218>
+<bytecode: 0x128c4cd38>
 <environment: namespace:cv>

### 2024-01-14 scratch/notes_john/JSS-article/JSS-article-reduced-2.Rmd
updates NEWS file; add Georges's orcid

@@
   - name: Georges Monette
-    orcid: 0000-0000-0000-0000
+    orcid: 0000-0003-0076-5532
     affiliation: 'York University'
@@
 
-## Example: Contrived hierarchical data
+## Example: CV of clusters versus individuals using contrived data
 

### 2024-01-14 scratch/notes_john/JSS-article/JSS-article-reduced-2.tex
updates NEWS file; add Georges's orcid

@@
 John Fox~\orcidlink{0000-0002-1196-8012}\\McMaster
-University \And Georges Monette~\orcidlink{0000-0000-0000-0000}\\York
+University \And Georges Monette~\orcidlink{0000-0003-0076-5532}\\York
 University
@@
 \begin{CodeOutput}
-R RNG seed set to 352413
+R RNG seed set to 801552
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.42674
-bias-adjusted cross-validation criterion = 19.40347
-95% CI for bias-adjusted CV criterion = (15.89518, 22.91177)
+cross-validation criterion = 19.32751
+bias-adjusted cross-validation criterion = 19.30935
+95% CI for bias-adjusted CV criterion = (15.83674, 22.78196)
 full-sample criterion = 18.98477 
@@
 
-\hypertarget{example-contrived-hierarchical-data}{%
-\subsection{Example: Contrived hierarchical
-data}\label{example-contrived-hierarchical-data}}
+\hypertarget{example-cv-of-clusters-versus-individuals-using-contrived-data}{%
+\subsection{Example: CV of clusters versus individuals using contrived
+data}\label{example-cv-of-clusters-versus-individuals-using-contrived-data}}
 
@@
 }
-<bytecode: 0x128c4cd38>
+<bytecode: 0x123a02cc0>
 <environment: namespace:cv>

### 2024-01-15 scratch/notes_john/JSS-article/JSS-article-reduced-2.Rmd
compress JSS paper section on contrived data

@@
 
-## Example: CV of clusters versus individuals using contrived data
+## Example: Contrasting cluster-based and case-based CV
 
@@
 
-Figure \ref{fig:plot1} (a) shows a scatterplot the data for a representative group of 10 (without loss of generality, the first 10) of the 100 students, displaying the 95% concentration ellipse for each cluster:
+Figure \ref{fig:plot1} (a) shows a scatterplot the data for a representative group of 10 (without loss of generality, the first 10) of the 100 students, displaying the 95% concentration ellipse for each cluster.
 
@@
 #| fig.height = 4,
-#| fig.cap = "(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data",
+#| fig.cap = "(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) predictions from several mixed models fit to the data",
 #| fig.subcap=rep("", 5),
@@
   xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
-         groups=group, type="l", lwd=2, lty=3)
+         groups=group, type="l", lwd=2, lty=2)
 ) |> update(
@@
     text=list(c("fixed effects only","fixed and random")),
-    lines=list(lty=c(1, 3))))
+    lines=list(lty=c(1, 2), lwd=2)))
 
@@
   xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
-         groups=group, type="l", lwd=2, lty=3)
+         groups=group, type="l", lwd=2, lty=2)
 ) |> update(
   main="Model: y ~ 1 + x + (1 | group)",
-  ylim=c(-15, 35),
-  key=list(
-    corner=c(0.95, 0.05),
-    text=list(c("fixed effects only","fixed and random")),
-    lines=list(lty=c(1, 3))))
+  ylim=c(-15, 35) #,
+  # key=list(
+  #   corner=c(0.95, 0.05),
+  #   text=list(c("fixed effects only","fixed and random")),
+  #   lines=list(lty=c(1, 3)))
+  )
 
@@
   xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
-         groups=group, type="l", lwd=2, lty=3)
+         groups=group, type="l", lwd=2, lty=2)
 ) |> update(
   main="Model: y ~ 1 + x + xm + (1 | group)",
-  ylim=c(4, 16),
-  key=list(
-    corner=c(0.05, 0.05),
-    text=list(c("fixed effects only","fixed and random")),
-    lines=list(lty=c(1, 3))))
+  ylim=c(4, 16) #,
+  # key=list(
+  #   corner=c(0.05, 0.05),
+  #   text=list(c("fixed effects only","fixed and random")),
+  #   lines=list(lty=c(1, 3)))
+  )
 
@@
   xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
-         groups=group, type="l", lwd=2, lty=3)
+         groups=group, type="l", lwd=2, lty=2)
 ) |> update(
   main="Model: y ~ 1 + I(x - xm) + (1 | group)",
-  ylim=c(4, 16),
-  key=list(
-    corner=c(0.05, 0.05),
-    text=list(c("fixed effects only","fixed and random")),
-    lines=list(lty=c(1, 3))))
+  ylim=c(4, 16) #,
+  # key=list(
+  #   corner=c(0.05, 0.05),
+  #   text=list(c("fixed effects only","fixed and random")),
+  #   lines=list(lty=c(1, 3)))
+  )
 
@@
 
-The initial mixed-effects model that we fit to the data is a simple random-intercepts model:
-
-```{r}
-# random intercept only:
+We proceed to fit several mixed-effects models to the data, using the `compareCoefs()` function from the \pkg{car} package [@FoxWeisberg:2019] to display the fixed-effect estimates for these models.
+We discuss each of these models below.
+```{r fitted-mixed-models}
 mod.0 <- lmer(y ~ 1 + (1 | group), Data)
-summary(mod.0)
+mod.1 <- lmer(y ~ x + (1 | group), Data)
+mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
+mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
+library("car", quietly=TRUE)
+compareCoefs(mod.0, mod.1, mod.2, mod.3)
 ```
 
-We will shortly consider three other, more complex, mixed models.
-
-We proceed to obtain predictions from the random-intercept model (\code{mod.0}) for fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---the BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students).
-Predictions for the random-intercept model, \code{mod.0}, for the first 10 students are shown in Figure \ref{fig:plot1} (b).
+The initial mixed-effects model, \code{mod.0}, is a simple random-intercepts model.
+We obtain predictions from this model for the fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---the BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students).
+Predictions from \code{mod.0} for the first 10 students are shown in Figure \ref{fig:plot1} (b).
 The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean.
@@
 
-Our next model, \code{mod.1}, includes a fixed intercept and fixed effect of \code{x} along with a random intercept:
-
-```{r}
-# effect of x and random intercept:
-mod.1 <- lmer(y ~ x + (1 | group), Data)
-summary(mod.1)
-```
-
+Our next model, \code{mod.1}, includes a fixed intercept and the fixed effect of \code{x}, along with a random intercept.
 Predictions from this model appear in Figure \ref{fig:plot1} (c).
 The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, \code{mod.0}.
-We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod1} to \code{mod0}, but that cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.
+We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod.1} to \code{mod.0}, but that cluster-based cross-validation will prefer \code{mod.0} to \code{mod.1}.
 
-Our third model, \code{mod.2}, includes the "contextual effect" of $x$---that is, the cluster mean \code{xm}---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:
-
-```{r}
-# effect of x, contextual (student) mean of x, and random intercept:
-mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
-        # equivalent to y ~ I(x - xm) + xm + (1 | group)
-summary(mod.2)
-```
-
+Our third model, \code{mod.2}, includes the "contextual effect" of $x$---that is, the cluster mean \code{xm}---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept.
 This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that generated the data once the coefficient of the contextual predictor \code{xm} is set to 0 (as it is in \code{mod.3}, discussed below).
@@
 Finally, \code{mod.3} directly estimates the model used to generate the data.
-As mentioned, it is a constrained version of \code{mod.2}, with the coefficient of \code{xm} set to 0, and with \code{x} expressed as a deviation from the cluster mean \code{xm}:
-
-```{r}
-# model generating the data (where Bb = 0)
-mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
-summary(mod.3)
-```
-
+As mentioned, it is a constrained version of \code{mod.2}, with the coefficient of \code{xm} set to 0, and with \code{x} expressed as a deviation from the cluster mean \code{xm}.
 The predictions from \code{mod.3}, shown in Figure \ref{fig:plot1} (e), are therefore similar to those from \code{mod.2}.
@@
 
-library("car", quietly=TRUE)
 scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 

### 2024-01-15 scratch/notes_john/JSS-article/JSS-article-reduced-2.tex
compress JSS paper section on contrived data

@@
 \begin{CodeOutput}
-R RNG seed set to 801552
+R RNG seed set to 610656
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.32751
-bias-adjusted cross-validation criterion = 19.30935
-95% CI for bias-adjusted CV criterion = (15.83674, 22.78196)
+cross-validation criterion = 19.39106
+bias-adjusted cross-validation criterion = 19.36966
+95% CI for bias-adjusted CV criterion = (15.86196, 22.87735)
 full-sample criterion = 18.98477 
@@
 
-\hypertarget{example-cv-of-clusters-versus-individuals-using-contrived-data}{%
-\subsection{Example: CV of clusters versus individuals using contrived
-data}\label{example-cv-of-clusters-versus-individuals-using-contrived-data}}
+\hypertarget{example-contrasting-cluster-based-and-case-based-cv}{%
+\subsection{Example: Contrasting cluster-based and case-based
+CV}\label{example-contrasting-cluster-based-and-case-based-cv}}
 
@@
 the 100 students, displaying the 95\% concentration ellipse for each
-cluster:
+cluster.
 
@@
 
-\caption[(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data]{(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) several mixed models fit to the data}\label{fig:plot1}
+\caption[(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) predictions from several mixed models fit to the data]{(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) predictions from several mixed models fit to the data}\label{fig:plot1}
 \end{figure}
@@
 
-The initial mixed-effects model that we fit to the data is a simple
-random-intercepts model:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # random intercept only:
+We proceed to fit several mixed-effects models to the data, using the
+\texttt{compareCoefs()} function from the \pkg{car} package
+\citep{FoxWeisberg:2019} to display the fixed-effect estimates for these
+models. We discuss each of these models below.
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> mod.0 <- lmer(y ~ 1 + (1 | group), Data)
-R> summary(mod.0)
-\end{CodeInput}
-\begin{CodeOutput}
-Linear mixed model fit by REML ['lmerMod']
-Formula: y ~ 1 + (1 | group)
-   Data: Data
-
-REML criterion at convergence: 2103.1
-
-Scaled residuals: 
-     Min       1Q   Median       3Q      Max 
--2.03514 -0.72645 -0.01169  0.78477  2.04377
-
-Random effects:
- Groups   Name        Variance Std.Dev.
- group    (Intercept) 2.900    1.703   
- Residual             2.712    1.647   
-Number of obs: 500, groups:  group, 100
-
-Fixed effects:
-            Estimate Std. Error t value
-(Intercept)  10.0018     0.1855   53.91
-\end{CodeOutput}
-\end{CodeChunk}
-
-We will shortly consider three other, more complex, mixed models.
-
-We proceed to obtain predictions from the random-intercept model
-(\code{mod.0}) for fixed effects alone, as would be used for
-cross-validation based on clusters (i.e., students), and for fixed and
-random effects---the BLUPs---as would be used for cross-validation based
-on cases (i.e., occasions within students). Predictions for the
-random-intercept model, \code{mod.0}, for the first 10 students are
-shown in Figure \ref{fig:plot1} (b). The fixed-effect predictions for
-the various individuals are identical---the estimated fixed-effects
-intercept or estimated general mean of \(y\)---while the BLUPs are the
-sums of the fixed-effects intercept and the random intercepts, and are
-only slightly shrunken towards the general mean. Because in our
-artificial data there is no population relationship between age and
-skill, the fixed-effect-only predictions and the BLUPs are not very
-different.
-
-Our next model, \code{mod.1}, includes a fixed intercept and fixed
-effect of \code{x} along with a random intercept:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # effect of x and random intercept:
 R> mod.1 <- lmer(y ~ x + (1 | group), Data)
-R> summary(mod.1)
-\end{CodeInput}
-\begin{CodeOutput}
-Linear mixed model fit by REML ['lmerMod']
-Formula: y ~ x + (1 | group)
-   Data: Data
-
-REML criterion at convergence: 1564.5
-
-Scaled residuals: 
-     Min       1Q   Median       3Q      Max 
--2.90160 -0.63501  0.01879  0.55407  2.82932
-
-Random effects:
- Groups   Name        Variance Std.Dev.
- group    (Intercept) 192.9406 13.8903 
- Residual               0.2569  0.5068 
-Number of obs: 500, groups:  group, 100
-
-Fixed effects:
-             Estimate Std. Error t value
-(Intercept) -33.91892    1.56446  -21.68
-x             0.96529    0.01581   61.05
-
-Correlation of Fixed Effects:
-  (Intr)
-x -0.460
-\end{CodeOutput}
-\end{CodeChunk}
-
-Predictions from this model appear in Figure \ref{fig:plot1} (c). The
-BLUPs fit the observed data very closely, but predictions based on the
-fixed effects alone, with a common intercept and slope for all clusters,
-are very poor---indeed, much worse than the fixed-effects-only
-predictions based on the simpler random-intercept model, \code{mod.0}.
-We therefore anticipate (and show later in this section) that case-based
-cross-validation will prefer \code{mod1} to \code{mod0}, but that
-cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.
+R> mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
+R> mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
+R> library("car", quietly=TRUE)
+R> compareCoefs(mod.0, mod.1, mod.2, mod.3)
+\end{CodeInput}
+\begin{CodeOutput}
+Calls:
+1: lmer(formula = y ~ 1 + (1 | group), data = Data)
+2: lmer(formula = y ~ x + (1 | group), data = Data)
+3: lmer(formula = y ~ x + xm + (1 | group), data = Data)
+4: lmer(formula = y ~ I(x - xm) + (1 | group), data = Data)
+
+            Model 1 Model 2 Model 3 Model 4
+(Intercept)  10.002 -33.919   9.479  10.002
+SE            0.186   1.564   0.617   0.186
+
+x                    0.9653  0.9915        
+SE                   0.0158  0.0160
+
+xm                          -0.9800        
+SE                           0.0206
+
+I(x - xm)                             0.992
+SE                                    0.016
+
+\end{CodeOutput}
+\end{CodeChunk}
+
+The initial mixed-effects model, \code{mod.0}, is a simple
+random-intercepts model. We obtain predictions from this model for the
+fixed effects alone, as would be used for cross-validation based on
+clusters (i.e., students), and for fixed and random effects---the
+BLUPs---as would be used for cross-validation based on cases (i.e.,
+occasions within students). Predictions from \code{mod.0} for the first
+10 students are shown in Figure \ref{fig:plot1} (b). The fixed-effect
+predictions for the various individuals are identical---the estimated
+fixed-effects intercept or estimated general mean of \(y\)---while the
+BLUPs are the sums of the fixed-effects intercept and the random
+intercepts, and are only slightly shrunken towards the general mean.
+Because in our artificial data there is no population relationship
+between age and skill, the fixed-effect-only predictions and the BLUPs
+are not very different.
+
+Our next model, \code{mod.1}, includes a fixed intercept and the fixed
+effect of \code{x}, along with a random intercept. Predictions from this
+model appear in Figure \ref{fig:plot1} (c). The BLUPs fit the observed
+data very closely, but predictions based on the fixed effects alone,
+with a common intercept and slope for all clusters, are very
+poor---indeed, much worse than the fixed-effects-only predictions based
+on the simpler random-intercept model, \code{mod.0}. We therefore
+anticipate (and show later in this section) that case-based
+cross-validation will prefer \code{mod.1} to \code{mod.0}, but that
+cluster-based cross-validation will prefer \code{mod.0} to \code{mod.1}.
 
@@
 \(x\)---that is, the cluster mean \code{xm}---along with \(x\) and the
-intercept in the fixed-effect part of the model, and a random intercept:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # effect of x, contextual (student) mean of x, and random intercept:
-R> mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
-R>         # equivalent to y ~ I(x - xm) + xm + (1 | group)
-R> summary(mod.2)
-\end{CodeInput}
-\begin{CodeOutput}
-Linear mixed model fit by REML ['lmerMod']
-Formula: y ~ x + xm + (1 | group)
-   Data: Data
-
-REML criterion at convergence: 1169.2
-
-Scaled residuals: 
-     Min       1Q   Median       3Q      Max 
--2.98466 -0.63750  0.00191  0.55682  2.73246
-
-Random effects:
- Groups   Name        Variance Std.Dev.
- group    (Intercept) 3.3986   1.8435  
- Residual             0.2552   0.5052  
-Number of obs: 500, groups:  group, 100
-
-Fixed effects:
-            Estimate Std. Error t value
-(Intercept)  9.47866    0.61705   15.36
-x            0.99147    0.01597   62.07
-xm          -0.97998    0.02055  -47.68
-
-Correlation of Fixed Effects:
-   (Intr) x     
-x   0.000       
-xm -0.600 -0.777
-\end{CodeOutput}
-\end{CodeChunk}
-
+intercept in the fixed-effect part of the model, and a random intercept.
 This model is equivalent to fitting
@@
 the coefficient of \code{xm} set to 0, and with \code{x} expressed as a
-deviation from the cluster mean \code{xm}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # model generating the data (where Bb = 0)
-R> mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
-R> summary(mod.3)
-\end{CodeInput}
-\begin{CodeOutput}
-Linear mixed model fit by REML ['lmerMod']
-Formula: y ~ I(x - xm) + (1 | group)
-   Data: Data
-
-REML criterion at convergence: 1163.2
-
-Scaled residuals: 
-     Min       1Q   Median       3Q      Max 
--2.97703 -0.63204  0.00627  0.56032  2.72489
-
-Random effects:
- Groups   Name        Variance Std.Dev.
- group    (Intercept) 3.3913   1.8415  
- Residual             0.2552   0.5052  
-Number of obs: 500, groups:  group, 100
-
-Fixed effects:
-            Estimate Std. Error t value
-(Intercept) 10.00176    0.18553   53.91
-I(x - xm)    0.99147    0.01597   62.07
-
-Correlation of Fixed Effects:
-          (Intr)
-I(x - xm) 0.000 
-\end{CodeOutput}
-\end{CodeChunk}
-
-The predictions from \code{mod.3}, shown in Figure \ref{fig:plot1} (e),
-are therefore similar to those from \code{mod.2}.
+deviation from the cluster mean \code{xm}. The predictions from
+\code{mod.3}, shown in Figure \ref{fig:plot1} (e), are therefore similar
+to those from \code{mod.2}.
 
@@
 R> 
-R> library("car", quietly=TRUE)
 R> scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
@@
 }
-<bytecode: 0x123a02cc0>
+<bytecode: 0x139564668>
 <environment: namespace:cv>

### 2024-02-03 R/cv-mixed-models.R
add details arg to cv() and methods

@@ -30,6 +30,10 @@
 #' @param seed for R's random number generator; optional, if not
 #' supplied a random seed will be selected and saved; not needed
 #' for n-fold cross-validation
+#' @param details if \code{TRUE} (the default if the number of
+#' folds \code{k <= 10}), save detailed information about the value of the
+#' CV criterion for the cases in each fold and coefficients
+#' with that fold deleted.
 #' @param ncores number of cores to use for parallel computations
 #'        (default is \code{1}, i.e., computations aren't done in parallel)
 #' @param clusterVariables a character vector of names of the variables
@@ -46,7 +50,10 @@
 #' the first two elements should be
 #' \code{model} and \code{newdata}; see the "Extending the cv package" vignette
 #' (\code{vignette("cv-extend", package="cv")}).
-#'
+#' @param blups a function to be used to compute BLUPs for
+#' case-based CV when \code{details = TRUE}.
+#' @param fixed.effects a function to be used to compute fixed-effect
+#' coefficients for cluster-based CV when \code{details = TRUE}.
 #' @param ... for \code{cv()} methods, to match generic,
 #' and for \code{cvMixed()}, arguments to be passed to \code{update()}.
 #'
@@ -91,10 +98,12 @@ cvMixed <- function(model,
                     reps=1,
                     confint, level=0.95,
                     seed,
+                    details = k <= 10,
                     ncores=1,
                     clusterVariables,
                     predict.clusters.args=list(object=model, newdata=data),
                     predict.cases.args=list(object=model, newdata=data),
+                    blups, fixed.effects,
                     ...){
 
   pkg.env <- getNamespace(package)
@@ -109,7 +118,8 @@ cvMixed <- function(model,
     fit.all.i <- do.call(predict, predict.clusters.args)
     fit.i <- fit.all.i[index <- !index]
     list(fit.i=fit.i, crit.all.i=criterion(y, fit.all.i),
-         indices.i=index)
+         indices.i=index,
+         coef.i=fixed.effects(predict.clusters.args$object))
   }
 
   f.cases <- function(i, predict.clusters.args, predict.cases.args, ...){
@@ -121,14 +131,15 @@ cvMixed <- function(model,
     fit.all.i <- do.call(predict, predict.cases.args)
     fit.i <- fit.all.i[indices.i]
[…]

### 2024-02-03 R/cv.R

@@ -14,16 +14,20 @@
 #' @param k perform k-fold cross-validation (default is \code{10}); \code{k}
 #' may be a number or \code{"loo"} or \code{"n"} for n-fold (leave-one-out)
 #' cross-validation.
-#' @param reps number of times to replicate k-fold CV (default is \code{1})
+#' @param reps number of times to replicate k-fold CV (default is \code{1}).
 #' @param confint if \code{TRUE} (the default if the number of cases is 400
 #' or greater), compute a confidence interval for the bias-corrected CV
 #' criterion, if the criterion is the average of casewise components.
 #' @param level confidence level (default \code{0.95}).
 #' @param seed for R's random number generator; optional, if not
 #' supplied a random seed will be selected and saved; not needed
-#' for n-fold cross-validation
+#' for n-fold cross-validation.
+#' @param details if \code{TRUE} (the default if the number of
+#' folds \code{k <= 10}), save detailed information about the value of the
+#' CV criterion for the cases in each fold and the regression coefficients
+#' with that fold deleted.
 #' @param ncores number of cores to use for parallel computations
-#'        (default is \code{1}, i.e., computations aren't done in parallel)
+#'        (default is \code{1}, i.e., computations aren't done in parallel).
 #' @param method computational method to apply to a linear (i.e. \code{"lm"}) model
 #' or to a generalized linear (i.e., \code{"glm"}) model. See Details for an explanation
 #' of the available options.
@@ -122,6 +126,7 @@ cv <- function(model, data, criterion, k, reps=1, seed, ...){
 #' @export
 cv.default <- function(model, data=insight::get_data(model),
                        criterion=mse, k=10, reps=1, seed,
+                       details = k <= 10,
                        confint = n >= 400, level=0.95,
                        ncores=1,
                        type="response", ...){
@@ -132,7 +137,8 @@ cv.default <- function(model, data=insight::get_data(model),
     model.i <- update(model, data=data[ - indices.i, ])
     fit.all.i <- predict(model.i, newdata=data, type=type, ...)
     fit.i <- fit.all.i[indices.i]
[…]

### 2024-02-03 scratch/notes_john/cv-details-examples.R

@@ -1,0 +1,41 @@
+data("Auto", package="ISLR2")
+m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
+cv.lm.loo <- cv(m.auto,  k="loo")
+cv.lm.loo$details
+cv.lm.wood <- cv(m.auto, seed=1234)
+cv.lm.wood$details
+cv.lm.naive <- cv(m.auto, seed=1234, method="naive")
+cv.lm.naive$details
+all.equal(cv.lm.wood$details, cv.lm.naive$details)
+max(mapply(cv.lm.wood$details$coefficients,
+       cv.lm.naive$details$coefficients,
+       FUN=function(x, y) max(abs(x - y)/x)))
+cbind(unlist(cv.lm.wood$details$coefficients),
+      unlist(cv.lm.naive$details$coefficients))
+
+data("Mroz", package="carData")
+m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
+cv.glm.naive <- cv(m.mroz, criterion=BayesRule, seed=123)
+cv.glm.naive$details
+cv.glm.wood <- cv(m.mroz, criterion=BayesRule, seed=123,
+                  method="Woodbury")
+cv.glm.wood$details
+all.equal(cv.glm.wood$details, cv.glm.naive$details)
+all.equal(lapply(cv.glm.wood$details$coefficients, round),
+          lapply(cv.glm.naive$details$coefficients, round))
+max(abs(mapply("-", cv.glm.wood$details$coefficients,
+       cv.glm.naive$details$coefficients)))
+
+library("lme4")
+# from ?lmer:
+(fm1 <- lmer(Reaction ~ Days + (Days | Subject), sleepstudy))
+cvloo <- cv(fm1, clusterVariables="Subject") # LOO CV of clusters
+cvcase <- cv(fm1, seed=447) # 10-fold CV of cases
+cvclus <- cv(fm1, clusterVariables="Subject", k=5,
+   seed=834) # 5-fold CV of clusters
+cvcase$details
+cvclus$details
+lapply(cvcase$details$coefficients,
+       function(x) rownames(x$Subject))
+
+

### 2024-02-05 R/cv.R
implement details arg for cv.formula()

@@ -97,7 +97,8 @@
 #' data("Auto", package="ISLR2")
 #' m.auto <- lm(mpg ~ horsepower, data=Auto)
 #' cv(m.auto,  k="loo")
-#' cv(m.auto, seed=1234)
+#' (cv.auto <- cv(m.auto, seed=1234))
+#' compareFolds(cv.auto)
 #' cv(m.auto, seed=1234, reps=3)
 #'
 #' data("Mroz", package="carData")

### 2024-02-05 scratch/notes_john/cv-details-examples.R

@@ -38,4 +38,7 @@ cvclus$details
 lapply(cvcase$details$coefficients,
        function(x) rownames(x$Subject))
 
+data("Auto", package="ISLR2")
+m.auto <- lm(mpg ~ . - name - origin, data=Auto)
+cv.sel <- cv(selectStepAIC, Auto, seed=123, working.model=m.auto)
 

### 2024-02-05 vignettes/cv-extend.R

@@ -193,7 +193,7 @@ selectSubsets <- function(data=insight::get_data(model),
                           model,
                           indices,
                           criterion=mse,
-                          save.coef=TRUE, ...){
+                          details=TRUE, ...){
   
   if (inherits(model, "lm", which=TRUE) != 1)
     stop("selectSubsets is appropriate only for 'lm' models")
@@ -227,7 +227,7 @@ selectSubsets <- function(data=insight::get_data(model),
   #   and the regression coefficients
   list(fit.i=fit.i, # fitted values for i-th fold
        crit.all.i=criterion(y, fit.all.i), # CV crit for all cases
-       coefficients = if (save.coef){ # regression coefficients
+       coefficients = if (details){ # regression coefficients
          coefs <- coef(m.best.i)
          
          # fix coefficient names

### 2024-02-15 R/cv.R
add selectModelList(), with small changes to cvSelect() and methods.

@@ -35,6 +35,9 @@
 #' \code{type} argument of \code{predict()};
 #' the default is `type="response"`, which is appropriate, e.g., for a `"glm"` model
 #' and may be recognized or ignored by \code{predict()} methods for other model classes.
+#' @param start if \code{TRUE} (the default is \code{FALSE}), the \code{start} argument
+#' set to the vector of regression coefficients for the model fit to the full data is passed
+#' to \code{update()} possibly making the CV updates faster, e.g. for a GLM.
 #' @param ... to match generic; passed to \code{predict()} for the default method.
 #'
 #' @returns The \code{cv()} methods return an object of class \code{"cv"}, with the CV criterion
@@ -134,12 +137,17 @@ cv.default <- function(model, data=insight::get_data(model),
                        details = k <= 10,
                        confint = n >= 400, level=0.95,
                        ncores=1,
-                       type="response", ...){
+                       type="response",
+                       start=FALSE, ...){
 
   f <- function(i){
     # helper function to compute cv criterion for each fold
     indices.i <- indices[starts[i]:ends[i]]
-    model.i <- update(model, data=data[ - indices.i, ])
+    model.i <- if (start) {
+      update(model, data=data[ - indices.i, ], start=b)
+    } else {
+      update(model, data=data[ - indices.i, ])
+      }
     fit.all.i <- predict(model.i, newdata=data, type=type, ...)
     fit.i <- fit.all.i[indices.i]
     list(fit.i=fit.i, crit.all.i=criterion(y, fit.all.i),
@@ -152,7 +160,12 @@ cv.default <- function(model, data=insight::get_data(model),
     indices.i <- indices[starts[i]:ends[i]]
     # the following deals with a scoping issue that can
     #   occur with args passed via ... (which is saved in dots)
-    predict.args <- c(list(object=update(model, data=data[ - indices.i, ]),
+    predict.args <- c(list(
+      object= if (start) {
+          update(model, data=data[ - indices.i, ], start=b)
+        } else {
+          update(model, data=data[ - indices.i, ])
+        },
       newdata=data, type=type), dots)
     fit.all.i <- do.call(predict, predict.args)
     fit.i <- fit.all.i[indices.i]
@@ -161,6 +174,7 @@ cv.default <- function(model, data=insight::get_data(model),
   }
 
   y <- GetResponse(model)
+  b <- coef(model)
   n <- nrow(data)
   if (is.character(k)){
     if (k == "n" || k == "loo") {
[…]

### 2024-02-15 R/getResponse.R

@@ -65,3 +65,7 @@ GetResponse.lme <- function(model, ...) insight::get_response(model)
 #' @describeIn GetResponse \code{glmmTMB} method
 #' @export
 GetResponse.glmmTMB <- GetResponse.merMod
+
+#' @describeIn GetResponse \code{modList} method
+#' @export
+GetResponse.modList <- function(model, ...) GetResponse(model[[1]])

### 2024-02-15 vignettes/cv-extend.R

@@ -193,7 +193,10 @@ selectSubsets <- function(data=insight::get_data(model),
                           model,
                           indices,
                           criterion=mse,
-                          details=TRUE, ...){
+                          details=TRUE, 
+                          seed,
+                          save.model=FALSE, 
+                          ...){
   
   if (inherits(model, "lm", which=TRUE) != 1)
     stop("selectSubsets is appropriate only for 'lm' models")
@@ -203,6 +206,7 @@ selectSubsets <- function(data=insight::get_data(model),
   X <- model.matrix(model)
 
   if (missing(indices)) {
+    if (missing(seed) || is.null(seed)) seed <- sample(1e6, 1L)
     # select the best model from the full data by BIC
     sel <- leaps::regsubsets(formula, data=data, ...)
     bics <- summary(sel)$bic
@@ -211,7 +215,10 @@ selectSubsets <- function(data=insight::get_data(model),
     # fit the best model; intercept is already in X, hence - 1:
     m.best <- lm(y ~ X[, x.names] - 1) 
     fit.all <- predict(m.best, newdata=data)
-    return(criterion(y, fit.all)) # return the CV criterion
+    return(list(
+      criterion = criterion(y, fit.all),
+      model = if (save.model) m.best else NULL
+      ))
   }
 
   # select the best model omitting the i-th fold (given by indices)

### 2024-02-15 vignettes/cv-extend.R
update vignettes for nested CV

@@ -217,7 +217,7 @@ selectSubsets <- function(data=insight::get_data(model),
     fit.all <- predict(m.best, newdata=data)
     return(list(
       criterion = criterion(y, fit.all),
-      model = if (save.model) m.best else NULL
+      model = if (save.model) m.best else NULL # return best model
       ))
   }
 

### 2024-02-18 vignettes/cv.R
add recursive arg to cv.modList()

@@ -396,15 +396,24 @@ legend("topright", inset=0.02,
        legend=c("10-Fold CV", "LOO CV"),
        lwd=2, lty=2:1, col=3:2, pch=17:16)
 
-## ----nested-CV-polynomials----------------------------------------------------
-nestedCV.auto <- cv(selectModelList, Auto,
-                    working.model=models(m.1, m.2, m.3, m.4, m.5,
-                                        m.6, m.7, m.8, m.9, m.10),
-                    save.model=TRUE,
-                    seed=2120)
-nestedCV.auto
-nestedCV.auto$selected.model
+## ----recursive-CV-polynomials-------------------------------------------------
+recursiveCV.auto <- cv(selectModelList, Auto,
+                       working.model=models(m.1, m.2, m.3, m.4, m.5,
+                                            m.6, m.7, m.8, m.9, m.10),
+                       save.model=TRUE,
+                       seed=2120)
+recursiveCV.auto
+recursiveCV.auto$selected.model
 cv(m.7, seed=2120) # same seed for same folds
+
+## ----recursive-cv-alt---------------------------------------------------------
+recursiveCV.auto.alt <- cv(models(m.1, m.2, m.3, m.4, m.5,
+                                  m.6, m.7, m.8, m.9, m.10),
+                           data=Auto,
+                           seed=2120,
+                           recursive=TRUE,
+                           save.model=TRUE)
+all.equal(recursiveCV.auto, recursiveCV.auto.alt)
 
 ## ----recall-Mroz-regression---------------------------------------------------
 summary(m.mroz)

### 2024-03-06 R/cv.R
further improvement to cv.default()

@@ -42,9 +42,10 @@
 #' to the full data, possibly making the CV updates faster, e.g., for a GLM.
 #' @param ... to match generic; passed to \code{predict()} for the default method.
 #' @param model.function a regression function, typically for a new \code{cv()} method that
-#' that calls \code{cv(default())} via \code{NextMethod()},
+#' that calls \code{cv.default()} via \code{NextMethod()},
 #' residing in a package that's not a declared dependency of the \pkg{cv} package,
-#' e.g., \code{nnet::multinom}.
+#' e.g., \code{nnet::multinom}. It's usually not necessary to specify
+#' \code{model.function} to make \code{cv.default()} work.
 #'
 #' @returns The \code{cv()} methods return an object of class \code{"cv"}, with the CV criterion
 #' (\code{"CV crit"}), the bias-adjusted CV criterion (\code{"adj CV crit"}),
@@ -128,7 +129,8 @@ cv <- function(model, data, criterion, k, reps=1, seed, ...){
 }
 
 #' @describeIn cv \code{"default"} method.
-#' @importFrom stats coef family fitted lm.wfit lsfit model.frame
+#' @importFrom stats coef family fitted getCall
+#' lm.wfit lsfit model.frame
 #' model.matrix model.response predict qnorm
 #' update weighted.mean weights
 #' residuals hatvalues printCoefmat sd
@@ -148,7 +150,7 @@ cv.default <- function(model, data=insight::get_data(model),
                        ncores=1,
                        type="response",
                        start=FALSE,
-                       model.function=NULL,
+                       model.function,
                        ...){
 
   f <- function(i, ...){
@@ -185,8 +187,14 @@ cv.default <- function(model, data=insight::get_data(model),
          coef.i=coef(predict.args$object))
   }
 
+  if (missing(model.function)){
+    model.function <- getCall(model)[[1]]
+    model.function.name <- as.character(model.function)
+    model.function <- eval(model.function)
+  } else {
     model.function.name <- sub("^.*\\:\\:", "",
-                       deparse(substitute(model.function)))
+                               deparse(substitute(model.function)))
+  }
 
   n <- nrow(data)
 
@@ -263,7 +271,7 @@ cv.lm <- function(model, data=insight::get_data(model),
   if (method == "auto"){
     method <- if (k == n) "hatvalues" else "Woodbury"
   }
-  if (method == "naive") return(NextMethod())
[…]

### 2024-03-06 scratch/notes_john/cv.multinom.R

@@ -2,8 +2,7 @@
 cv.multinom <- function (model, data, criterion=BayesRuleMulti, k, reps,
                          seed, ...){
   NextMethod(type="class", criterion=criterion,
-             criterion.name=deparse(substitute(criterion)),
-             model.function=nnet::multinom)
+             criterion.name=deparse(substitute(criterion)))
 }
 
 # new version

### 2024-03-06 vignettes/cv-extend.R
small revision to cv-extend vignette

@@ -100,12 +100,12 @@ cv(m.beps, seed=3465, criterion=BayesRuleMulti)
 ## ----cv.nultinom--------------------------------------------------------------
 cv.multinom <- function (model, data, criterion=BayesRuleMulti, k, reps,
                          seed, ...){
+  model <- update(model, trace=FALSE)
   NextMethod(type="class", criterion=criterion,
              criterion.name=deparse(substitute(criterion)))
 }
 
 ## ----BEPS-cv------------------------------------------------------------------
-m.beps <- update(m.beps, trace=FALSE)
 cv(m.beps, seed=3465)
 
 ## ----cv.lme-------------------------------------------------------------------

### 2024-03-12 vignettes/cv.R
reformat code in vignettes

@@ -38,84 +38,153 @@ dim(Auto)
 plot(mpg ~ horsepower, data=Auto)
 
 ## ----mpg-horsepower-scatterplot-polynomials-----------------------------------
-plot(mpg ~ horsepower, data=Auto)
-horsepower <- with(Auto, 
-                   seq(min(horsepower), max(horsepower), 
-                       length=1000))
-for (p in 1:5){
-  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
-  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
-  lines(horsepower, mpg, col=p + 1, lty=p, lwd=2)
+plot(mpg ~ horsepower, data = Auto)
+horsepower <- with(Auto,
+                   seq(min(horsepower), max(horsepower),
+                       length = 1000))
+for (p in 1:5) {
+  m <- lm(mpg ~ poly(horsepower, p), data = Auto)
+  mpg <- predict(m, newdata = data.frame(horsepower = horsepower))
+  lines(horsepower,
+        mpg,
+        col = p + 1,
+        lty = p,
+        lwd = 2)
 }
-legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=2,
-       title="Degree", inset=0.02)
+legend(
+  "topright",
+  legend = 1:5,
+  col = 2:6,
+  lty = 1:5,
+  lwd = 2,
+  title = "Degree",
+  inset = 0.02
+)
 
 ## ----mpg-horsepower-MSE-se----------------------------------------------------
 library("cv") # for mse() and other functions
 
 var <- mse <- numeric(10)
-for (p in 1:10){
-  m <- lm(mpg ~ poly(horsepower, p), data=Auto)
+for (p in 1:10) {
+  m <- lm(mpg ~ poly(horsepower, p), data = Auto)
   mse[p] <- mse(Auto$mpg, fitted(m))
-  var[p] <- summary(m)$sigma^2
+  var[p] <- summary(m)$sigma ^ 2
 }
 
-plot(c(1, 10), range(mse, var), type="n",
-     xlab="Degree of polynomial, p",
-     ylab="Estimated Squared Error")
-lines(1:10, mse, lwd=2, lty=1, col=2, pch=16, type="b")
-lines(1:10, var, lwd=2, lty=2, col=3, pch=17, type="b")
-legend("topright", inset=0.02,
[…]

### 2024-04-03 vignettes/cv-extend.R
reorganize vignettes

@@ -395,74 +395,3 @@ selectSubsets(model=m.swiss, indices=seq(5, 45, by=10))
 ## ----best-models-by-folds-----------------------------------------------------
 compareFolds(cv.swiss)
 
-## -----------------------------------------------------------------------------
-AUC <- function(y, yhat = seq_along(y)) {
-  s <- sum(y)
-  if (s == 0)
-    return(0)
-  if (s == length(y))
-    return(1)
-  Metrics::auc(y, yhat)
-}
-
-## -----------------------------------------------------------------------------
-Ymat <- function(n_v, exclude_identical = FALSE) {
-  stopifnot(n_v > 0 &&
-              round(n_v) == n_v)    # n_v must be a positive integer
-  ret <- sapply(0:(2 ^ n_v - 1),
-                function(x)
-                  as.integer(intToBits(x)))[1:n_v,]
-  ret <- if (is.matrix(ret))
-    t(ret)
-  else
-    matrix(ret)
-  colnames(ret) <- paste0("y", 1:ncol(ret))
-  if (exclude_identical)
-    ret[-c(1, nrow(ret)),]
-  else
-    ret
-}
-
-## -----------------------------------------------------------------------------
-Ymat(3)
-
-## -----------------------------------------------------------------------------
-Ymat(3, exclude_identical = TRUE)
-
-## -----------------------------------------------------------------------------
-cbind(Ymat(3), AUC = apply(Ymat(3), 1, AUC))
-
-## -----------------------------------------------------------------------------
-resids <- function(n_v,
-                   exclude_identical = FALSE,
-                   tol = sqrt(.Machine$double.eps)) {
-  Y <- Ymat(n_v, exclude_identical = exclude_identical)
-  AUC <- apply(Y, 1, AUC)
-  X <- cbind(1 - Y, Y)
-  opts <- options(warn = -1)
-  on.exit(options(opts))
-  fit <- lsfit(X, AUC, intercept = FALSE)
-  ret <- max(abs(residuals(fit)))
-  if (ret < tol) {
-    ret <- 0
-    solution <- coef(fit)
-    names(solution) <- paste0("c(", c(1:n_v, 1:n_v), ",",
-                              rep(0:1, each = n_v), ")")
-    attr(ret, "solution") <- zapsmall(solution)
-  }
-  ret
-}
-
-## -----------------------------------------------------------------------------
-resids(3, exclude_identical = TRUE)
-
-## -----------------------------------------------------------------------------
[…]

### 2024-04-03 vignettes/cv-selection.R

@@ -1,0 +1,355 @@
+## ----setup, include = FALSE---------------------------------------------------
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  message = TRUE,
+  warning = TRUE,
+  fig.align = "center",
+  fig.height = 6,
+  fig.width = 7,
+  fig.path = "fig/",
+  dev = "png",
+  comment = "#>" #,
+)
+
+# save some typing
+knitr::set_alias(w = "fig.width",
+                 h = "fig.height",
+                 cap = "fig.cap")
+
+# colorize text: use inline as `r colorize(text, color)`
+colorize <- function(x, color) {
+  if (knitr::is_latex_output()) {
+    sprintf("\\textcolor{%s}{%s}", color, x)
+  } else if (knitr::is_html_output()) {
+    sprintf("<span style='color: %s;'>%s</span>", color,
+      x)
+  } else x
+}
+
+
+.opts <- options(digits = 5)
+
+## ----generate-selection-data--------------------------------------------------
+set.seed(24361) # for reproducibility
+D <- data.frame(y = rnorm(1000, mean = 10),
+                X = matrix(rnorm(1000 * 100), 1000, 100))
+head(D[, 1:6])
+
+## ----omnibus-F----------------------------------------------------------------
+m.full <- lm(y ~ ., data = D)
+m.null <- lm(y ~ 1, data = D)
+anova(m.null, m.full)
+
+summary(m.null)
+
+## ----forward-selection--------------------------------------------------------
+library("MASS")  # for stepAIC()
+m.select <- stepAIC(
+  m.null,
+  direction = "forward",
+  trace = FALSE,
+  scope = list(lower =  ~ 1, upper = formula(m.full))
+)
+summary(m.select)
+
+library("cv")
+mse(D$y, fitted(m.select))
+
+## ----cv-selectedModel---------------------------------------------------------
+library("cv")
+
+cv(m.select, seed = 2529)
+
+## ----compare-selected-models--------------------------------------------------
+compareFolds(cv.select)
+
+## ----polynomial-regression-CV-graph-duplicated, echo=FALSE--------------------
+data("Auto", package="ISLR2")
+for (p in 1:10) {
+  command <- paste0("m.", p, "<- lm(mpg ~ poly(horsepower, ", p,
+                    "), data=Auto)")
+  eval(parse(text = command))
+}
+# 10-fold CV
+cv.auto.10 <- cv(
+  models(m.1, m.2, m.3, m.4, m.5,
[…]

### 2024-04-05 vignettes/cv-extend.R
tweak model-selection example in cv-extend vignette

@@ -385,9 +385,9 @@ selectSubsets(model=m.swiss)
 selectSubsets(model=m.swiss, indices=seq(5, 45, by=10))
 
 ## ----cvSelect-selectSubsets---------------------------------------------------
-(cv.swiss <- cvSelect(
+(cv.swiss <- cv(
   selectSubsets,
-  model = m.swiss,
+  working.model = m.swiss,
   data = swiss,
   seed = 8433 # use same folds
 ))

### 2024-04-14 R/cv.R
change "cross-validation of" to "cross-validating"

@@ -105,7 +105,7 @@
 #' inheriting the \code{"lm"} method for which the default \code{"auto"}
 #' computational method would be inappropriate).
 #'
-#' For additional details, see the "Cross-validation of regression models"
+#' For additional details, see the "Cross-validating regression models"
 #' vignette (\code{vignette("cv", package="cv")}).
 #'
 #' \code{cv()} is designed to be extensible to other classes of regression

### 2024-05-21 scratch/notes_john/JSS-article/JSS-article-3.Rmd
add and begin to work on JSS-article-3

--- 2024-01-15 scratch/notes_john/JSS-article/JSS-article-reduced-2.Rmd
@@
 title:
-  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validation of Regression Models"
+  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validating Regression Models"
   # If you use tex in the formatted title, also supply version without
-  plain:     "cv: An R Package for Cross-Validation of Regression Models"
+  plain:     "cv: An R Package for Cross-Validating Regression Models"
   # For running headers, if needed
-  short:     "\\pkg{cv}: Cross-Validation of Regression Models"
+  short:     "\\pkg{cv}: Cross-Validating Regression Models"
 abstract: >
@@
 Emulating this process in a single study by data-division is less common: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data.
-Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples: See @Harrell:2015 [Sec.
+Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples.
+See @Harrell:2015 [Sec.
 5.3] for this and other remarks about data-division and cross-validation.
@@
 
-Cross-validating mixed-effects models involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.
+Cross-validating mixed-effects models, using the \code{"glmmTMB"}, \code{"lme"}, and  \code{"merMod"} methods for \code{cv()}, involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.
 
-The \code{cvSelect()} function, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors.
+The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.
 
@@
 
+A number of existing \proglang{R} packages include functions for cross-validating regression models.
+We briefly situate the \pkg{cv} package relative to other \proglang{R} software for cross-validation in Section \ref{comparing-cv-to-other-r-software-for-cross-validation}.
+
 In the interest of brevity, we won't describe all of the features of the \pkg{cv} package here, concentrating on the aspects of the package that are relatively novel.
-For example, the \code{cv()} and \code{cvSelect()} functions can perform computations in parallel and can independently replicate a cross-validation procedure several times.
+For example, the \code{cv()} function can perform computations in parallel and can independently replicate a cross-validation procedure several times.
 These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provides greater detail on some of topics that we do describe, such as extensions to the package.
@@
 #| fig.height = 5,
-#| fig.cap = "mpg vs horsepower for the Auto data, showing fitted polynomials of degree 1 through 5."
+#| fig.cap = "\\code{mpg} vs. \\code{horsepower} for the \\code{Auto} data, showing fitted polynomials of degree 1 through 5."
 plot(mpg ~ horsepower, data=Auto)
@@
 Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader).
-We revisit the \code{Auto} data in Section \ref{cross-validating-model-selection}.
+We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.
 
@@
 
-* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).
+* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).
 
@@
 ```
-On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.[^microbenchmark]
+On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.
 
-[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100.
+<!-- [^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100.
 With the exception of the last column, the output is self-explanatory.
@@
 Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default \code{times=100} repetitions.
+-->
 
@@
 In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed.
-The \code{lmer()} function in the \pkg{lme4} package is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects.
+The \code{lmer()} function in the \pkg{lme4} package, for example, is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects.
 We present an example of the latter in a vignette for the \pkg{cv} package.
@@
 
-The \code{cvSelect()} function in the \pkg{cv} package cross-validates a model-specification process in a general manner.
-Functions for three such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package [@FoxWeisberg:2019], transforms predictors and the response in a regression model towards normality; and \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially.
-In the vignette on extending the \pkg{cv} package, we explain how to add model-selection procedures.
+The \code{"function"} method for \code{cv()} cross-validates a model-specification process in a general manner.
+Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package [@FoxWeisberg:2019], transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model.
+In a vignette on extending the \pkg{cv} package, we explain how to add model-selection procedures.
+
+## Example: Data transformation and predictor selection for the Auto data
 
@@
 #| fig.height = 5,
-#| fig.cap = "Scatterplot matrix for the numeric variables in the Auto data"
+#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data"
 Auto$cylinders <- factor(Auto$cylinders,
-                         labels=c("3.4", "3.4", "5.6", "5.6", "8"))
+                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
 Auto$year <- as.factor(Auto$year)
@@
 It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation.
-We believe, however, that it's still useful to automate as much of the process as we can to obtain a more realistic, if still biased, estimate of the predictive power of a model.
+We believe, however, that it's still useful to automate as much of the process as we can, as illustrated here, to obtain a more realistic, if still biased, estimate of the predictive power of a model.
 
@@
 #| fig.height = 5,
-#| fig.cap = "Component+residual plots for the working model fit to the Auto data"
+#| fig.cap = "Component+residual plots for the working model fit to the \\code{Auto} data"
 m.auto <- lm(mpg ~ ., data = Auto)
@@
 Once the relationships between the predictors are linearized, it can be advantageous similarly to transform the conditional distribution of the response variable towards normality.
-Selecting transformations analytically raises the possibility of automating the process, as would be required for cross-validation.
+Selecting transformations analytically raises the possibility of automating the process, as required for cross-validation.
 
@@
 #| fig.height = 5,
-#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the Auto data"
+#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data"
 scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
@@
 ```
-And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figures \ref{fig:Auto-working-model}, on page \pageref{fig:Auto-working-model}, and \ref{fig:Auto-CR-plots-transformed}):
+And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-working-model}, on page \pageref{fig:Auto-working-model}, and Figure \ref{fig:Auto-CR-plots-transformed}):
 ```{r Auto-CR-plots-transformed}
@@
 #| fig.height = 5,
-#| fig.cap = "Component+residual plots for the model fit to the transformed Auto data"
+#| fig.cap = "Component+residual plots for the model fit to the transformed \\code{Auto} data"
 crPlots(m, pch=".")
@@
 
-After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log(n)$):
+After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log n$):
 ```{r}
@@
 
-The \code{cvSelect()} function in the \pkg{cv} package allows us to cross-validate the whole model-selection procedure.
-The first argument to \code{cvSelect()}---here \code{selectTransAndStepAIC}---is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion:
+The \code{"function"} method for \code{cv()} allows us to cross-validate the whole model-selection procedure.
+The first argument to \code{cv()}---here \code{selectTransAndStepAIC}---is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion:
 ```{r Auto-transform-and-select}
 num.predictors
-cvs <- cvSelect(selectTransStepAIC, data=Auto, seed=76692, model=m.auto,
+cvs <- cv(selectTransStepAIC, data=Auto, seed=76692, 
+          working.model=m.auto,
           predictors=num.predictors,
@@
 ```
-The other arguments to \code{cvSelect()} are (see \code{?cvSelect} for additional optional arguments and details):
+The other arguments to \code{cv()} are (see \code{?cv:::cv.function} for additional optional arguments and details):
 
@@
 * \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, then a seed is randomly selected and saved.
-* Arguments required by the model-selection function: the starting \code{model} (\code{m.auto}) for predictor selection, the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation, and \code{AIC=FALSE}, which specifies use of the BIC for model selection.
+* Arguments required by the model-selection function: the starting \code{working.model} (\code{m.auto}) for transformation and predictor selection, the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation, and \code{AIC=FALSE}, which specifies use of the BIC for model selection.
 
@@
 * \code{selectTransStepAIC()} automatically computes CV cost criteria, here the MSE, on the \emph{untransformed} response scale.
-* The estimate of the MSE that we obtain by cross-validating the whole model-specification process is larger than the MSE computed for the model we fit to the \code{Auto} data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
+* As we anticipated, the estimate of the MSE that we obtain by cross-validating the whole model-specification process is larger than the MSE computed for the model we fit to the \code{Auto} data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
 * When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of \code{compareFolds()}), we see that there is little uncertainty in choosing variable transformations (the \code{lam.*}s for the $x$s and \code{lambda} for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: \code{horsepower}, \code{weight}, and \code{year} are always included among the selected predictors; \code{acceleration} and \code{displacement} are included respectively in 4 and 3 of 10 selected models; and \code{cylinders} and \code{origin} are each included in only 1 of 10 models.
@@
 
+## Example: Applying recursive CV to polynomial regression for the Auto data
+
 # Extending the cv package
 
-The \pkg{cv} package is designed to be extensible in several directions; in order of increasing general complexity, we can add: (1) a cross-validation cost criterion; (2) a model class that's not directly accommodated by the \code{cv()} default method or by another directly inherited method; and (3) a new model-selection procedure suitable for use with \code{selectModel()}.
+The \pkg{cv} package is designed to be extensible in several directions; in order of increasing general complexity, we can add: (1) a cross-validation cost criterion; (2) a model class that's not directly accommodated by the \code{cv()} default method or by another directly inherited method; and (3) a new model-selection procedure suitable for use with the \code{"function"} method for \code{cv()}.
 In this section, we illustrate (1) and (2); more extensive examples may be found in the vignette on extending the \pkg{cv} package.
@@
                        economic.cond.household + Blair + Hague + Kennedy +
-                       Europe*political.knowledge, data=BEPS, trace=FALSE)
+                       Europe*political.knowledge, data=BEPS)
 
 ```
-We set the argument \code{trace=FALSE} in the call to \code{multinom()} to suppress reporting the iteration history, which will be particularly appreciated when we repeatedly refit the model in cross-validation.
+<!-- We set the argument \code{trace=FALSE} in the call to \code{multinom()} to suppress reporting the iteration history, which will be particularly appreciated when we repeatedly refit the model in cross-validation.
+-->
 
@@
 ```
+As political knowledge increases, voters tend to align their votes more closely with the party positions on European integration: The Conservative Party was "Eurosceptic," while Labour and the Liberal Democrats were supportive of the UK's participation in the EU.
 
@@
 After doing some error checking, \code{BayesRule()} rounds the predicted proabability of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction and then reports the proportion of incorrect predictions.
-Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, making possible the computation of bias correction and confidence intervals (as discussion in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, enabling the computation of bias correction and confidence intervals (as discussion in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
 
@@
 ```{r cv.nultinom}
-cv.multinom <- function (model, data, criterion=BayesRuleMulti, k, reps,
-                         seed, ...){
-  NextMethod(type="class", criterion=criterion)
+cv.multinom <- function (model, data, criterion = BayesRuleMulti, 
+                         k, reps, seed, ...) {
+    model <- update(model, trace = FALSE)
+    NextMethod(
+      type = "class",
+      criterion = criterion,
+      criterion.name = deparse(substitute(criterion))
+    )
   }
 ```
-That is, we simply invoke the default \code{cv()} method via \code{NextMethod()}, with the \code{type} argument properly set.
-In addition to supplying the correct \code{type} argument, our method sets the default \code{criterion} for the \code{cv.multinom()} method to \code{BayesRuleMulti}.
+That is, we simply call the default `cv()` method with the `type` argument properly set.
+In addition to supplying the correct `type` argument, our method sets the default `criterion` for the `cv.multinom()` method to `BayesRuleMulti`.
+Adding the argument `criterion.name=` `deparse(substitute(criterion))` is inessential, but it insures that printed output will include the name of the criterion function that's employed, whether it's the default `BayesRuleMulti` or something else.
+Prior to invoking `NextMethod()`, we called `update()` with `trace=FALSE` to suppress the iteration history reported by default by `multinom()`---it would be tedious to see the iteration history for each fold.
 
@@
 The cross-validated polytomous Bayes's rule criterion confirms that the fitted model does substantially better than the marginal Bayes's rule prediction that everyone votes for Labour.
+
+# Comparing cv to other R software for cross-validation
 

### 2024-05-21 scratch/notes_john/JSS-article/JSS-article-3.tex
add and begin to work on JSS-article-3

--- 2024-01-15 scratch/notes_john/JSS-article/JSS-article-reduced-2.tex
@@
 }
-\title{\pkg{cv}: An \proglang{R} Package for Cross-Validation of
-Regression Models}
+\title{\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression
+Models}
 
 \Plainauthor{John Fox, Georges Monette}
-\Plaintitle{cv: An R Package for Cross-Validation of Regression Models}
-\Shorttitle{\pkg{cv}: Cross-Validation of Regression Models}
+\Plaintitle{cv: An R Package for Cross-Validating Regression Models}
+\Shorttitle{\pkg{cv}: Cross-Validating Regression Models}
 
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 disconcertingly, the results can vary substantially based on the random
-division of the data, particularly in smaller samples: See \citet[Sec.
+division of the data, particularly in smaller samples. See \citet[Sec.
 5.3]{Harrell:2015} for this and other remarks about data-division and
@@
 \begin{CodeOutput}
-[1] cv.default* cv.glm*     cv.glmmTMB* cv.lm*      cv.lme*     cv.merMod* 
-[7] cv.modList* cv.rlm*    
+[1] cv.default*  cv.function* cv.glm*      cv.glmmTMB*  cv.lm*      
+[6] cv.lme*      cv.merMod*   cv.modList*  cv.rlm*     
 see '?methods' for accessing help and source code
@@
 
-Cross-validating mixed-effects models involves special considerations
-that we take up in Section \ref{cross-validating-mixed-effects-models}.
-
-The \code{cvSelect()} function, discussed in Section
+Cross-validating mixed-effects models, using the \code{"glmmTMB"},
+\code{"lme"}, and \code{"merMod"} methods for \code{cv()}, involves
+special considerations that we take up in Section
+\ref{cross-validating-mixed-effects-models}.
+
+The \code{"function"} method for \code{cv()}, discussed in Section
 \ref{cross-validating-model-specification}, cross-validates a complex
 model-specification process that may, for example, involve choice of
-data transformations and predictors.
+data transformations and predictors, or model selection via CV itself.
 
@@
 
+A number of existing \proglang{R} packages include functions for
+cross-validating regression models. We briefly situate the \pkg{cv}
+package relative to other \proglang{R} software for cross-validation in
+Section \ref{comparing-cv-to-other-r-software-for-cross-validation}.
+
 In the interest of brevity, we won't describe all of the features of the
 \pkg{cv} package here, concentrating on the aspects of the package that
-are relatively novel. For example, the \code{cv()} and \code{cvSelect()}
-functions can perform computations in parallel and can independently
-replicate a cross-validation procedure several times. These and other
-features not discussed in this paper are taken up in the vignettes
-distributed with the package, which also provides greater detail on some
-of topics that we do describe, such as extensions to the package.
-
-\hypertarget{preliminary-example-polynomial-regression}{%
+are relatively novel. For example, the \code{cv()} function can perform
+computations in parallel and can independently replicate a
+cross-validation procedure several times. These and other features not
+discussed in this paper are taken up in the vignettes distributed with
+the package, which also provides greater detail on some of topics that
+we do describe, such as extensions to the package.
+
 \section{Preliminary Example: Polynomial
-regression}\label{preliminary-example-polynomial-regression}}
+regression}\label{preliminary-example-polynomial-regression}
 
@@
 
-{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
+{\centering \includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
 
@@
 
-\caption[mpg vs horsepower for the Auto data, showing fitted polynomials of degree 1 through 5]{mpg vs horsepower for the Auto data, showing fitted polynomials of degree 1 through 5.}\label{fig:mpg-horsepower-scatterplot-polynomials}
+\caption{\code{mpg} vs. \code{horsepower} for the \code{Auto} data, showing fitted polynomials of degree 1 through 5.}\label{fig:mpg-horsepower-scatterplot-polynomials}
 \end{figure}
@@
   (execution of which is left to the reader). We revisit the \code{Auto}
-  data in Section \ref{cross-validating-model-selection}.} Polynomial
-fits for \(p = 1\) to \(5\) are shown in Figure
+  data in Section \ref{cross-validating-model-specification}.}
+Polynomial fits for \(p = 1\) to \(5\) are shown in Figure
 \ref{fig:mpg-horsepower-scatterplot-polynomials}. The linear fit is
@@
 
-{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/mpg-horsepower-MSE-se-1}
+{\centering \includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-MSE-se-1}
 
@@
 function (model, data = insight::get_data(model), criterion = mse, 
-    k = 10, reps = 1, seed, confint = n >= 400, level = 0.95, 
-    method = c("auto", "hatvalues", "Woodbury", "naive"), ncores = 1, 
-    ...) 
+    k = 10L, reps = 1L, seed = NULL, details = k <= 10L, confint = n >= 
+        400L, level = 0.95, method = c("auto", "hatvalues", "Woodbury", 
+        "naive"), ncores = 1L, ...) 
 NULL
@@
   is relevant only for LOO CV and bases computation on the hatvalues for
-  the linear model; \code{"Woodbury"} employs the Woodbury matrix to
-  compute the CV criterion with each fold deleted; \code{"naive"}
-  updates the model using the \code{update()} function; and
-  \code{"auto"} selects \code{"hatvalues"} for LOO CV and
+  the linear model; \code{"Woodbury"} employs the Woodbury matrix
+  identity to compute the CV criterion with each fold deleted;
+  \code{"naive"} updates the model using the \code{update()} function;
+  and \code{"auto"} selects \code{"hatvalues"} for LOO CV and
   \code{"Woodbury"} for \(k\)-fold CV, both of which are much more
@@
 \begin{CodeOutput}
-R RNG seed set to 610656
+R RNG seed set to 32663
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.39106
-bias-adjusted cross-validation criterion = 19.36966
-95% CI for bias-adjusted CV criterion = (15.86196, 22.87735)
+cross-validation criterion = 19.24054
+bias-adjusted cross-validation criterion = 19.22691
+95% CI for bias-adjusted CV criterion = (15.76529, 22.68854)
 full-sample criterion = 18.98477 
@@
 n-Fold Cross Validation
+method: naive
 criterion: mse
@@
 \begin{CodeOutput}
-Unit: microseconds
+Unit: milliseconds
       expr        min         lq       mean     median         uq        max
- hatvalues    981.745   1172.477   1163.211   1189.943   1203.063   1325.817
-  Woodbury   9958.777  10178.496  10309.593  10232.370  10356.067  10998.127
-     naive 214083.345 217623.654 247100.940 220301.795 277217.933 331491.314
-    cv.glm 381699.053 383011.832 401940.298 387351.887 439080.439 442256.709
- neval cld
-    10 a  
-    10 a  
-    10  b 
-    10   c
+ hatvalues   1.015119   1.201546   1.191919   1.206446   1.222333   1.258987
+  Woodbury  12.108407  12.390569  19.157808  12.634949  15.289884  73.319275
+     naive 219.469269 220.993977 223.441857 223.066916 226.155180 228.677377
+    cv.glm 379.740606 383.517075 390.172306 384.790925 386.413397 442.643421
+ neval
+    10
+    10
+    10
+    10
 \end{CodeOutput}
@@
 faster than employing Woodbury matrix updates, and more than two orders
-of magnitude faster than refitting the model.\footnote{Out of
-  impatience, we asked \code{microbenchmark()} to execute each command
-  only 10 times rather than the default 100. With the exception of the
-  last column, the output is self-explanatory. The last column shows
-  which methods have average timings that are statistically
-  distinguishable. Because of the small number of repetitions (i.e.,
-  10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't
-  distinguishable, but the difference between these methods persists
-  when we perform more repetitions---we invite the reader to redo this
-  computation with the default \code{times=100} repetitions.}
-
-\hypertarget{comparing-competing-models}{%
+of magnitude faster than refitting the model.
+
 \subsection{Comparing competing
-models}\label{comparing-competing-models}}
+models}\label{comparing-competing-models}
 
@@
 
-{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/polynomial-regression-CV-graph-2-1} }\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/polynomial-regression-CV-graph-2-2} }
+{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/polynomial-regression-CV-graph-2-1} }\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/polynomial-regression-CV-graph-2-2} }
 
@@
 
-\hypertarget{cross-validating-mixed-effects-models}{%
 \section{Cross-validating mixed-effects
-models}\label{cross-validating-mixed-effects-models}}
+models}\label{cross-validating-mixed-effects-models}
 
@@
   random effects based on classes within schools are crossed. The
-  \code{lmer()} function in the \pkg{lme4} package is capable of
-  modeling both nested and crossed random effects, and the \code{cv()}
-  methods for mixed models in the \pkg{cv} package pertain to both
-  nested and crossed random effects. We present an example of the latter
-  in a vignette for the \pkg{cv} package.}
+  \code{lmer()} function in the \pkg{lme4} package, for example, is
+  capable of modeling both nested and crossed random effects, and the
+  \code{cv()} methods for mixed models in the \pkg{cv} package pertain
+  to both nested and crossed random effects. We present an example of
+  the latter in a vignette for the \pkg{cv} package.}
 
@@
 
-\hypertarget{example-the-high-school-and-beyond-data}{%
 \subsection{Example: The High-School and Beyond
-data}\label{example-the-high-school-and-beyond-data}}
+data}\label{example-the-high-school-and-beyond-data}
 
@@
 10-Fold Cross Validation based on 160 {school} clusters
+criterion: mse
 cross-validation criterion = 39.15662
@@
 Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv, :
-Model failed to converge with max|grad| = 0.00587228 (tol = 0.002, component 1)
+Model failed to converge with max|grad| = 0.00587207 (tol = 0.002, component 1)
 \end{CodeOutput}
@@
 10-Fold Cross Validation
+criterion: mse
 cross-validation criterion = 37.44473
@@
 
-\hypertarget{example-contrasting-cluster-based-and-case-based-cv}{%
 \subsection{Example: Contrasting cluster-based and case-based
-CV}\label{example-contrasting-cluster-based-and-case-based-cv}}
+CV}\label{example-contrasting-cluster-based-and-case-based-cv}
 
@@
 
-{\centering \subfloat[\label{fig:plot1-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-1} }\subfloat[\label{fig:plot1-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-2} }\newline\subfloat[\label{fig:plot1-3}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-3} }\subfloat[\label{fig:plot1-4}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-4} }\newline\subfloat[\label{fig:plot1-5}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-5} }
+{\centering \subfloat[\label{fig:plot1-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-1} }\subfloat[\label{fig:plot1-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-2} }\newline\subfloat[\label{fig:plot1-3}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-3} }\subfloat[\label{fig:plot1-4}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-4} }\newline\subfloat[\label{fig:plot1-5}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-5} }
 
@@
 
-{\centering \subfloat[\label{fig:cross-validation-clusters-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/cross-validation-clusters-1} }\subfloat[\label{fig:cross-validation-clusters-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/cross-validation-clusters-2} }
+{\centering \subfloat[\label{fig:cross-validation-clusters-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/cross-validation-clusters-1} }\subfloat[\label{fig:cross-validation-clusters-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/cross-validation-clusters-2} }
 
@@
 
-\hypertarget{cross-validating-model-specification}{%
 \section{Cross-validating model
-specification}\label{cross-validating-model-specification}}
+specification}\label{cross-validating-model-specification}
 
@@
 
-The \code{cvSelect()} function in the \pkg{cv} package cross-validates a
-model-specification process in a general manner. Functions for three
-such model-specification processes are included in the package:
+The \code{"function"} method for \code{cv()} cross-validates a
+model-specification process in a general manner. Functions for four such
+model-specification processes are included in the package:
 \code{selectStepAIC()}, based on the \code{stepAIC()} function in the
@@
 \citep{FoxWeisberg:2019}, transforms predictors and the response in a
-regression model towards normality; and
-\code{selectTransAndStepAIC()}---the use of which we illustrate in the
-current section---performs both of these procedures sequentially. In the
-vignette on extending the \pkg{cv} package, we explain how to add
-model-selection procedures.
+regression model towards normality; \code{selectTransAndStepAIC()}---the
+use of which we illustrate in the current section---performs both of
+these procedures sequentially; and \code{selectModelList()}---also
+illustrated in the current section---uses CV both to select one of
+several competing models, and then, recursively, to estimate prediction
+error for the selected model. In a vignette on extending the \pkg{cv}
+package, we explain how to add model-selection procedures.
+
+\subsection{Example: Data transformation and predictor selection for the
+Auto
+data}\label{example-data-transformation-and-predictor-selection-for-the-auto-data}
 
@@
 R> Auto$cylinders <- factor(Auto$cylinders,
-+                          labels=c("3.4", "3.4", "5.6", "5.6", "8"))
++                          labels=c("3-4", "3-4", "5-6", "5-6", "8"))
 R> Auto$year <- as.factor(Auto$year)
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-explore-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-explore-1}
 
@@
 
-\caption[Scatterplot matrix for the numeric variables in the Auto data]{Scatterplot matrix for the numeric variables in the Auto data}\label{fig:Auto-explore}
+\caption[Scatterplot matrix for the numeric variables in the \code{Auto} data]{Scatterplot matrix for the numeric variables in the \code{Auto} data}\label{fig:Auto-explore}
 \end{figure}
@@
   of model validation. We believe, however, that it's still useful to
-  automate as much of the process as we can to obtain a more realistic,
-  if still biased, estimate of the predictive power of a model.} It's
-apparent that most these variables are positively skewed and that many
-of the pairwise relationships among them are nonlinear.
+  automate as much of the process as we can, as illustrated here, to
+  obtain a more realistic, if still biased, estimate of the predictive
+  power of a model.} It's apparent that most these variables are
+positively skewed and that many of the pairwise relationships among them
+are nonlinear.
 
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-working-model-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-working-model-1}
 
@@
 
-\caption[Component+residual plots for the working model fit to the Auto data]{Component+residual plots for the working model fit to the Auto data}\label{fig:Auto-working-model}
+\caption[Component+residual plots for the working model fit to the \code{Auto} data]{Component+residual plots for the working model fit to the \code{Auto} data}\label{fig:Auto-working-model}
 \end{figure}
@@
 Selecting transformations analytically raises the possibility of
-automating the process, as would be required for cross-validation.
+automating the process, as required for cross-validation.
 
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-transformed-scatterplot-matrix-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-transformed-scatterplot-matrix-1}
 
@@
 
-\caption[Scatterplot matrix for the transformed numeric variables in the Auto data]{Scatterplot matrix for the transformed numeric variables in the Auto data}\label{fig:Auto-transformed-scatterplot-matrix}
+\caption[Scatterplot matrix for the transformed numeric variables in the \code{Auto} data]{Scatterplot matrix for the transformed numeric variables in the \code{Auto} data}\label{fig:Auto-transformed-scatterplot-matrix}
 \end{figure}
@@
 And the partial relationships in the model fit to the transformed data
-are much more nearly linear (cf., Figures \ref{fig:Auto-working-model},
-on page \pageref{fig:Auto-working-model}, and
+are much more nearly linear (cf., Figure \ref{fig:Auto-working-model},
+on page \pageref{fig:Auto-working-model}, and Figure
 \ref{fig:Auto-CR-plots-transformed}):
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-CR-plots-transformed-1}
+{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-CR-plots-transformed-1}
 
@@
 
-\caption[Component+residual plots for the model fit to the transformed Auto data]{Component+residual plots for the model fit to the transformed Auto data}\label{fig:Auto-CR-plots-transformed}
+\caption[Component+residual plots for the model fit to the transformed \code{Auto} data]{Component+residual plots for the model fit to the transformed \code{Auto} data}\label{fig:Auto-CR-plots-transformed}
 \end{figure}
@@
 criterion (by setting the \texttt{k} argument of \texttt{stepAIC()} to
-\(\log(n)\)):
+\(\log n\)):
 
@@
 
-The \code{cvSelect()} function in the \pkg{cv} package allows us to
-cross-validate the whole model-selection procedure. The first argument
-to \code{cvSelect()}---here \code{selectTransAndStepAIC}---is a
-model-selection function capable of refitting the model with a fold
-omitted and returning a CV criterion:
+The \code{"function"} method for \code{cv()} allows us to cross-validate
+the whole model-selection procedure. The first argument to
+\code{cv()}---here \code{selectTransAndStepAIC}---is a model-selection
+function capable of refitting the model with a fold omitted and
+returning a CV criterion:
 
@@
 \begin{CodeInput}
-R> cvs <- cvSelect(selectTransStepAIC, data=Auto, seed=76692, model=m.auto,
+R> cvs <- cv(selectTransStepAIC, data=Auto, seed=76692, 
++           working.model=m.auto,
 +           predictors=num.predictors,
@@
 10-Fold Cross Validation
+criterion: mse
 cross-validation criterion = 7.485557
@@
 \begin{CodeOutput}
+CV criterion by folds:
+   fold.1    fold.2    fold.3    fold.4    fold.5    fold.6    fold.7    fold.8 
+ 6.000589 10.509334  7.444420 10.617864  5.049238  7.183160  4.077682 10.501909 
+   fold.9   fold.10 
+ 9.181184  4.250736
+
+Coefficients by folds:
         (Intercept) horsepower lam.acceleration lam.displacement lam.horsepower
@@
 Fold 10  0.04881  0.07862  0.14101  0.14313  0.23258  0.35649  0.26214  0.32421
-        acceleration displacement cylinders5.6 cylinders8 originEurope
+        acceleration displacement cylinders5-6 cylinders8 originEurope
 Fold 1                                                                
@@
 
-The other arguments to \code{cvSelect()} are (see \code{?cvSelect} for
+The other arguments to \code{cv()} are (see \code{?cv:::cv.function} for
 additional optional arguments and details):
@@
   Arguments required by the model-selection function: the starting
-  \code{model} (\code{m.auto}) for predictor selection, the names of the
-  variables---\code{predictors} and \code{response}---that are
-  candidates for transformation, and \code{AIC=FALSE}, which specifies
-  use of the BIC for model selection.
+  \code{working.model} (\code{m.auto}) for transformation and predictor
+  selection, the names of the variables---\code{predictors} and
+  \code{response}---that are candidates for transformation, and
+  \code{AIC=FALSE}, which specifies use of the BIC for model selection.
 \end{itemize}
@@
 \item
-  The estimate of the MSE that we obtain by cross-validating the whole
-  model-specification process is larger than the MSE computed for the
-  model we fit to the \code{Auto} data separately selecting
-  transformations of the predictors and the response and then selecting
-  predictors for the whole data set.
+  As we anticipated, the estimate of the MSE that we obtain by
+  cross-validating the whole model-specification process is larger than
+  the MSE computed for the model we fit to the \code{Auto} data
+  separately selecting transformations of the predictors and the
+  response and then selecting predictors for the whole data set.
 \item
@@
 
-\hypertarget{extending-the-cv-package}{%
-\section{Extending the cv package}\label{extending-the-cv-package}}
+\subsection{Example: Applying recursive CV to polynomial regression for
+the Auto
+data}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-data}
+
+\section{Extending the cv package}\label{extending-the-cv-package}
 
@@
 inherited method; and (3) a new model-selection procedure suitable for
-use with \code{selectModel()}. In this section, we illustrate (1) and
-(2); more extensive examples may be found in the vignette on extending
-the \pkg{cv} package.
+use with the \code{"function"} method for \code{cv()}. In this section,
+we illustrate (1) and (2); more extensive examples may be found in the
+vignette on extending the \pkg{cv} package.
 
@@
 +                        economic.cond.household + Blair + Hague + Kennedy +
-+                        Europe*political.knowledge, data=BEPS, trace=FALSE)
-\end{CodeInput}
-\end{CodeChunk}
-
-We set the argument \code{trace=FALSE} in the call to \code{multinom()}
-to suppress reporting the iteration history, which will be particularly
-appreciated when we repeatedly refit the model in cross-validation.
++                        Europe*political.knowledge, data=BEPS)
+\end{CodeInput}
+\begin{CodeOutput}
+# weights:  36 (22 variable)
+initial  value 1675.383740 
+iter  10 value 1240.047788
+iter  20 value 1163.199642
+iter  30 value 1116.519687
+final  value 1116.519666 
+converged
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-{\centering \includegraphics{JSS-article-reduced-2_files/figure-latex/BEPS-plot-1}
+{\centering \includegraphics{JSS-article-3_files/figure-latex/BEPS-plot-1}
 
@@
 \end{CodeChunk}
+
+As political knowledge increases, voters tend to align their votes more
+closely with the party positions on European integration: The
+Conservative Party was ``Eurosceptic,'' while Labour and the Liberal
+Democrats were supportive of the UK's participation in the EU.
 
@@
 \begin{CodeOutput}
-function(y, yhat){
-  if (!all(y %in% c(0, 1))) stop("response values not all 0 or 1")
-  if (any(yhat < 0) || any(yhat > 1)) stop("fitted values outside of interval [0, 1]")
+function (y, yhat) 
+{
+    if (!all(y %in% c(0, 1))) 
+        stop("response values not all 0 or 1")
+    if (any(yhat < 0) || any(yhat > 1)) 
+        stop("fitted values outside of interval [0, 1]")
     yhat <- round(yhat)
-  result <- mean(y != yhat) # proportion in error
+    result <- mean(y != yhat)
     attr(result, "casewise loss") <- "y != round(yhat)"
@@
 }
-<bytecode: 0x139564668>
+<bytecode: 0x12000eac8>
 <environment: namespace:cv>
@@
 is an average of casewise components (as, e.g., is the MSE), a
-\code{"casewise loss"} attribute is attached to the result, making
-possible the computation of bias correction and confidence intervals (as
-discussion in Section
+\code{"casewise loss"} attribute is attached to the result, enabling the
+computation of bias correction and confidence intervals (as discussion
+in Section
 \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
@@
 \begin{CodeOutput}
+# weights:  36 (22 variable)
+initial  value 1507.296060 
+iter  10 value 1134.575036
+iter  20 value 1037.413231
+iter  30 value 1007.705242
+iter  30 value 1007.705235
+iter  30 value 1007.705235
+final  value 1007.705235 
+converged
+\end{CodeOutput}
+\begin{CodeOutput}
 Error in match.arg(type): 'arg' should be one of "class", "probs"
@@
 \begin{CodeInput}
-R> cv.multinom <- function (model, data, criterion=BayesRuleMulti, k, reps,
-+                          seed, ...){
-+   NextMethod(type="class", criterion=criterion)
+R> cv.multinom <- function (model, data, criterion = BayesRuleMulti, 
++                          k, reps, seed, ...) {
++     model <- update(model, trace = FALSE)
++     NextMethod(
++       type = "class",
++       criterion = criterion,
++       criterion.name = deparse(substitute(criterion))
++     )
 +   }
@@
 
-That is, we simply invoke the default \code{cv()} method via
-\code{NextMethod()}, with the \code{type} argument properly set. In
-addition to supplying the correct \code{type} argument, our method sets
-the default \code{criterion} for the \code{cv.multinom()} method to
-\code{BayesRuleMulti}.
+That is, we simply call the default \texttt{cv()} method with the
+\texttt{type} argument properly set. In addition to supplying the
+correct \texttt{type} argument, our method sets the default
+\texttt{criterion} for the \texttt{cv.multinom()} method to
+\texttt{BayesRuleMulti}. Adding the argument \texttt{criterion.name=}
+\texttt{deparse(substitute(criterion))} is inessential, but it insures
+that printed output will include the name of the criterion function
+that's employed, whether it's the default \texttt{BayesRuleMulti} or
+something else. Prior to invoking \texttt{NextMethod()}, we called
+\texttt{update()} with \texttt{trace=FALSE} to suppress the iteration
+history reported by default by \texttt{multinom()}---it would be tedious
+to see the iteration history for each fold.
 
@@
 10-Fold Cross Validation
+criterion: BayesRuleMulti
 cross-validation criterion = 0.3245902
@@
 
-\hypertarget{computational-notes}{%
-\section{Computational notes}\label{computational-notes}}
-
-\hypertarget{efficient-computations-for-linear-and-generalized-linear-models}{%
+\section{Comparing cv to other R software for
+cross-validation}\label{comparing-cv-to-other-r-software-for-cross-validation}
+
+\section{Computational notes}\label{computational-notes}
+
 \subsection{Efficient computations for linear and generalized linear
-models}\label{efficient-computations-for-linear-and-generalized-linear-models}}
+models}\label{efficient-computations-for-linear-and-generalized-linear-models}
 
@@
 
-\hypertarget{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}{%
 \subsection{Computation of the bias-corrected CV criterion and
 confidence
-intervals}\label{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}}
+intervals}\label{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}
 

### 2024-05-22 scratch/notes_john/JSS-article/JSS-article-3.Rmd
add section on recursive CV to JSS paper

@@
 ```
-The \code{"modList"} method for \code{cv()} cross-validates several competing models, not necessarily of the same class, using the same division of the data into folds.
+* The \code{"lm"} and \code{"glm"} methods are for linear and generalized-linear models fit respectively by the standard \proglang{R} \code{lm()} and \code{glm()} functions.
+
+* The \code{"modList"} method for \code{cv()} cross-validates several competing models, not necessarily of the same class, using the same division of the data into folds.
 The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.
 
-Cross-validating mixed-effects models, using the \code{"glmmTMB"}, \code{"lme"}, and  \code{"merMod"} methods for \code{cv()}, involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.
+* Cross-validating mixed-effects models, using the \code{"glmmTMB"}, \code{"lme"}, and  \code{"merMod"} methods for \code{cv()}, involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.
 
-The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.
+* The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.
 
-The \code{"default"} \code{cv()} method works (perhaps with a bit of coaxing) with many other existing regression-model classes for which there is an \code{update()} method that accepts a \code{data} argument.
+* The \code{"default"} \code{cv()} method works (perhaps with a bit of coaxing) with many other existing regression-model classes for which there is an \code{update()} method that accepts a \code{data} argument.
 More generally, the \pkg{cv} package is designed to be extensible, as discussed in Section \ref{extending-the-cv-package}.
@@
 
+The final section of the paper (Section \ref{computational-notes}) describes some computational details, including efficient CV computations for linear and generalized-linear models and computation of bias-corrected CV criteria.
+
 In the interest of brevity, we won't describe all of the features of the \pkg{cv} package here, concentrating on the aspects of the package that are relatively novel.
 For example, the \code{cv()} function can perform computations in parallel and can independently replicate a cross-validation procedure several times.
+There are also data management facilities in the package, such as coercing the objects produced by the \code{cv()} function into data frames for further analysis.
 These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provides greater detail on some of topics that we do describe, such as extensions to the package.
@@
 Despite its name, `cv.glm()` is an independent function and not a method of a `cv()` generic function.
+The \pkg{boot} package is part of the standard \proglang{R} distribution.
 
@@
 m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
-boot::cv.glm(Auto, m.auto.glm)$delta
+boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE
 
@@
 In this instance, the folds comprise one or more whole clusters; we refit the model with all of the cases in clusters in the current fold removed; and then we predict the response for the cases in clusters in the current fold.
-These predictions are based only on fixed effects because the random effects for the omitted clusters are presumably unknown, as they would be for data on cases in newly observed clusters.
+These predictions are based only on fixed effects because the random effects for the omitted clusters are construed to be unknown, as they would be for data on cases in newly observed clusters.
 
@@
 ```
-In the process, we merged variables from the school-level and student-level data sets and created two new school-level variables: \code{mean.ses}, which is the average SES for students in each school; and \code{cses}, which is the individual students' SES centered at their school means.
+In the process, we merge variables from the school-level and student-level data sets, and create two new school-level variables: \code{mean.ses}, which is the average SES for students in each school; and \code{cses}, which is the individual students' SES centered at their school means.
 For details, see @FoxWeisberg:2019 [Sec.
@@
 Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package [@FoxWeisberg:2019], transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model.
-In a vignette on extending the \pkg{cv} package, we explain how to add model-selection procedures.
+In a vignette on extending the \pkg{cv} package, we explain how to specify additional model-selection procedures.
 
@@
 #| fig.height = 5,
-#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data"
+#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data.
+In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression."
 Auto$cylinders <- factor(Auto$cylinders,
@@
 #| fig.height = 5,
-#| fig.cap = "Component+residual plots for the working model fit to the \\code{Auto} data"
+#| fig.cap = "Component+residual plots for the working model fit to the \\code{Auto} data.
+The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths."
 m.auto <- lm(mpg ~ ., data = Auto)
@@
 ```
-We then apply the (rounded) transformations---all, as it turns out, logs---to the data and re-estimate the model:
+We then apply the (rounded) transformations---all, as it turns out, logs (i.e., "Oth" powers)---to the data and re-estimate the model:
 ```{r Auto-with-transformed-predictors}
@@
 
-The transformed numeric variables are much better-behaved (cf., Figures \ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and \ref{fig:Auto-transformed-scatterplot-matrix}):
+The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed-scatterplot-matrix}):
 ```{r Auto-transformed-scatterplot-matrix}
@@
 * \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, then a seed is randomly selected and saved.
-* Arguments required by the model-selection function: the starting \code{working.model} (\code{m.auto}) for transformation and predictor selection, the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation, and \code{AIC=FALSE}, which specifies use of the BIC for model selection.
+* Arguments required by the model-selection function: the starting \code{working.model} (here, \code{m.auto}) for transformation and predictor selection; the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation; and \code{AIC=FALSE}, which specifies use of the BIC for model selection.
 
@@
 
+In Section \ref{comparing-competing-models}, following @JamesEtAl:2021[Secs.
+5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of `mpg` to `horsepower` for the `Auto` data, saving the 10 models, named `m.1` through `m.10`, in `mlist`.
+We then used `cv()` to compare the cross-validated MSE for the 10 models, discovering that the 7th degree polynomial had the smallest MSE (by a small margin).
+
+If we then select the 7th degree polynomial model, intending to use it for prediction, the CV estimate of the MSE for this model will be optimistic.
+One solution is to cross-validate the process of using CV to select the "best" model---that is, to apply CV to CV recursively.
+The function `selectModelList()`, which is suitable for use with `cv()`, implements this idea.
+
+Applying `selectModelList()` to the `Auto` polynomial-regression models, and using 10-fold CV, we obtain:
+```{r recursive-CV-polynomials}
+recursiveCV.auto <- cv(selectModelList, data=Auto, 
+                       working.model=mlist,
+                       save.model=TRUE, seed=2120)
+recursiveCV.auto
+recursiveCV.auto$selected.model
+cv(mlist[[7]], seed=2120) # same seed for same folds
+```
+As expected, recursive CV produces a larger estimate of MSE for the selected 7th degree polynomial model than CV applied directly to this model.
+
+We can equivalently call `cv()` with the list of models as its first argument and set the argument `recursive=TRUE`:
+```{r recursive-cv-alt}
+cv(mlist, data=Auto, seed=2120, recursive=TRUE, save.model=TRUE)
+```
+
 # Extending the cv package
 
-The \pkg{cv} package is designed to be extensible in several directions; in order of increasing general complexity, we can add: (1) a cross-validation cost criterion; (2) a model class that's not directly accommodated by the \code{cv()} default method or by another directly inherited method; and (3) a new model-selection procedure suitable for use with the \code{"function"} method for \code{cv()}.
-In this section, we illustrate (1) and (2); more extensive examples may be found in the vignette on extending the \pkg{cv} package.
+The \pkg{cv} package is designed to be extensible in several directions.
+In order of increasing general complexity, we can add: (1) a cross-validation cost criterion; (2) a model class that's not directly accommodated by the \code{cv()} default method or by another directly inherited method; and (3) a new model-selection procedure suitable for use with the \code{"function"} method for \code{cv()}.
+In this section, we illustrate (1) and (2); more diverse and extensive examples may be found in the vignette on extending the \pkg{cv} package.
 
@@
             fixed.predictors=list(given.values=c(gendermale=0.5))),
-     lines=list(col=c("blue", "red", "orange")),
+     lines=list(col=c("blue", "red", "orange")), # party colors
      axes=list(x=list(rug=FALSE), y=list(style="stacked")))
 ```
-As political knowledge increases, voters tend to align their votes more closely with the party positions on European integration: The Conservative Party was "Eurosceptic," while Labour and the Liberal Democrats were supportive of the UK's participation in the EU.
+As political knowledge increases, voters tend to align their votes more closely with the party positions on European integration: The Conservative Party was relatively Euro-skeptic, while Labour and the Liberal Democrats were more supportive of the UK's participation in the EU.
 
@@
 ```
-After doing some error checking, \code{BayesRule()} rounds the predicted proabability of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction and then reports the proportion of incorrect predictions.
-Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, enabling the computation of bias correction and confidence intervals (as discussion in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+After doing some error checking, \code{BayesRule()} rounds the predicted proabability of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction, and then reports the proportion of incorrect predictions.
+Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, enabling the computation of bias correction and confidence intervals (as discussed in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
 
@@
 ```
-That is, we simply call the default `cv()` method with the `type` argument properly set.
+That is, we simply call the default `cv()` method (via \code{NextMethod()}) with the `type` argument properly set.
 In addition to supplying the correct `type` argument, our method sets the default `criterion` for the `cv.multinom()` method to `BayesRuleMulti`.

### 2024-05-22 scratch/notes_john/JSS-article/JSS-article-3.tex
add section on recursive CV to JSS paper

@@
 
+\begin{itemize}
+\item
+  The \code{"lm"} and \code{"glm"} methods are for linear and
+  generalized-linear models fit respectively by the standard
+  \proglang{R} \code{lm()} and \code{glm()} functions.
+\item
   The \code{"modList"} method for \code{cv()} cross-validates several
   competing models, not necessarily of the same class, using the same
-division of the data into folds. The \code{cv()} function is introduced
-in the context of a preliminary example in Section
+  division of the data into folds. The \code{cv()} function is
+  introduced in the context of a preliminary example in Section
   \ref{preliminary-example-polynomial-regression} of the paper.
-
+\item
   Cross-validating mixed-effects models, using the \code{"glmmTMB"},
@@
   \ref{cross-validating-mixed-effects-models}.
-
+\item
   The \code{"function"} method for \code{cv()}, discussed in Section
@@
   data transformations and predictors, or model selection via CV itself.
-
+\item
   The \code{"default"} \code{cv()} method works (perhaps with a bit of
   coaxing) with many other existing regression-model classes for which
-there is an \code{update()} method that accepts a \code{data} argument.
-More generally, the \pkg{cv} package is designed to be extensible, as
-discussed in Section \ref{extending-the-cv-package}.
+  there is an \code{update()} method that accepts a \code{data}
+  argument. More generally, the \pkg{cv} package is designed to be
+  extensible, as discussed in Section \ref{extending-the-cv-package}.
+\end{itemize}
 
@@
 
+The final section of the paper (Section \ref{computational-notes})
+describes some computational details, including efficient CV
+computations for linear and generalized-linear models and computation of
+bias-corrected CV criteria.
+
 In the interest of brevity, we won't describe all of the features of the
@@
 computations in parallel and can independently replicate a
-cross-validation procedure several times. These and other features not
-discussed in this paper are taken up in the vignettes distributed with
-the package, which also provides greater detail on some of topics that
-we do describe, such as extensions to the package.
+cross-validation procedure several times. There are also data management
+facilities in the package, such as coercing the objects produced by the
+\code{cv()} function into data frames for further analysis. These and
+other features not discussed in this paper are taken up in the vignettes
+distributed with the package, which also provides greater detail on some
+of topics that we do describe, such as extensions to the package.
 
@@
   \texttt{cv.glm()} is an independent function and not a method of a
-  \texttt{cv()} generic function.}
+  \texttt{cv()} generic function. The \pkg{boot} package is part of the
+  standard \proglang{R} distribution.}
 
@@
 \begin{CodeOutput}
-R RNG seed set to 32663
+R RNG seed set to 423102
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.24054
-bias-adjusted cross-validation criterion = 19.22691
-95% CI for bias-adjusted CV criterion = (15.76529, 22.68854)
+cross-validation criterion = 19.27853
+bias-adjusted cross-validation criterion = 19.26291
+95% CI for bias-adjusted CV criterion = (15.80408, 22.72175)
 full-sample criterion = 18.98477 
@@
 R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
-R> boot::cv.glm(Auto, m.auto.glm)$delta
+R> boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE
 \end{CodeInput}
@@
       expr        min         lq       mean     median         uq        max
- hatvalues   1.015119   1.201546   1.191919   1.206446   1.222333   1.258987
-  Woodbury  12.108407  12.390569  19.157808  12.634949  15.289884  73.319275
-     naive 219.469269 220.993977 223.441857 223.066916 226.155180 228.677377
-    cv.glm 379.740606 383.517075 390.172306 384.790925 386.413397 442.643421
+ hatvalues   1.139062   1.184244   1.202678   1.194248   1.205769   1.297117
+  Woodbury  12.106521  12.310701  13.115158  12.337515  12.808810  16.149900
+     naive 217.941773 219.491204 225.861284 220.053909 222.015410 278.078974
+    cv.glm 373.900074 379.259963 397.907632 382.333036 437.228141 441.112358
  neval
@@
   predictions are based only on fixed effects because the random effects
-  for the omitted clusters are presumably unknown, as they would be for
-  data on cases in newly observed clusters.
+  for the omitted clusters are construed to be unknown, as they would be
+  for data on cases in newly observed clusters.
 \item
@@
 
-In the process, we merged variables from the school-level and
-student-level data sets and created two new school-level variables:
+In the process, we merge variables from the school-level and
+student-level data sets, and create two new school-level variables:
 \code{mean.ses}, which is the average SES for students in each school;
@@
 error for the selected model. In a vignette on extending the \pkg{cv}
-package, we explain how to add model-selection procedures.
+package, we explain how to specify additional model-selection
+procedures.
 
@@
 
-\caption[Scatterplot matrix for the numeric variables in the \code{Auto} data]{Scatterplot matrix for the numeric variables in the \code{Auto} data}\label{fig:Auto-explore}
+\caption[Scatterplot matrix for the numeric variables in the \code{Auto} data]{Scatterplot matrix for the numeric variables in the \code{Auto} data.
+In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.}\label{fig:Auto-explore}
 \end{figure}
@@
 
-\caption[Component+residual plots for the working model fit to the \code{Auto} data]{Component+residual plots for the working model fit to the \code{Auto} data}\label{fig:Auto-working-model}
+\caption[Component+residual plots for the working model fit to the \code{Auto} data]{Component+residual plots for the working model fit to the \code{Auto} data.
+The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.}\label{fig:Auto-working-model}
 \end{figure}
@@
 
-We then apply the (rounded) transformations---all, as it turns out,
-logs---to the data and re-estimate the model:
+We then apply the (rounded) transformations---all, as it turns out, logs
+(i.e., ``Oth'' powers)---to the data and re-estimate the model:
 
@@
 
-The transformed numeric variables are much better-behaved (cf., Figures
-\ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and
+The transformed numeric variables are much better-behaved (cf., Figure
+\ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and Figure
 \ref{fig:Auto-transformed-scatterplot-matrix}):
@@
   Arguments required by the model-selection function: the starting
-  \code{working.model} (\code{m.auto}) for transformation and predictor
-  selection, the names of the variables---\code{predictors} and
-  \code{response}---that are candidates for transformation, and
+  \code{working.model} (here, \code{m.auto}) for transformation and
+  predictor selection; the names of the variables---\code{predictors}
+  and \code{response}---that are candidates for transformation; and
   \code{AIC=FALSE}, which specifies use of the BIC for model selection.
@@
 
+In Section \ref{comparing-competing-models}, following \citet[Secs. 5.1,
+5.3]{JamesEtAl:2021}, we fit polynomial regressions up to degree 10 to
+the relationship of \texttt{mpg} to \texttt{horsepower} for the
+\texttt{Auto} data, saving the 10 models, named \texttt{m.1} through
+\texttt{m.10}, in \texttt{mlist}. We then used \texttt{cv()} to compare
+the cross-validated MSE for the 10 models, discovering that the 7th
+degree polynomial had the smallest MSE (by a small margin).
+
+If we then select the 7th degree polynomial model, intending to use it
+for prediction, the CV estimate of the MSE for this model will be
+optimistic. One solution is to cross-validate the process of using CV to
+select the ``best'' model---that is, to apply CV to CV recursively. The
+function \texttt{selectModelList()}, which is suitable for use with
+\texttt{cv()}, implements this idea.
+
+Applying \texttt{selectModelList()} to the \texttt{Auto}
+polynomial-regression models, and using 10-fold CV, we obtain:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> recursiveCV.auto <- cv(selectModelList, data=Auto, 
++                        working.model=mlist,
++                        save.model=TRUE, seed=2120)
+\end{CodeInput}
+\begin{CodeOutput}
+R RNG seed set to 2120
+\end{CodeOutput}
+\begin{CodeInput}
+R> recursiveCV.auto
+\end{CodeInput}
+\begin{CodeOutput}
+10-Fold Cross Validation
+criterion: mse
+cross-validation criterion = 20.01249
+bias-adjusted cross-validation criterion = 20.61866
+full-sample criterion = 18.74615 
+\end{CodeOutput}
+\begin{CodeInput}
+R> recursiveCV.auto$selected.model
+\end{CodeInput}
+\begin{CodeOutput}
+
+Call:
+lm(formula = mpg ~ poly(horsepower, p), data = Auto)
+
+Coefficients:
+         (Intercept)  poly(horsepower, p)1  poly(horsepower, p)2  
+              23.446              -120.138                44.090  
+poly(horsepower, p)3  poly(horsepower, p)4  poly(horsepower, p)5  
+              -3.949                -5.188                13.272  
+poly(horsepower, p)6  poly(horsepower, p)7  
+              -8.546                 7.981  
+\end{CodeOutput}
+\begin{CodeInput}
+R> cv(mlist[[7]], seed=2120) # same seed for same folds
+\end{CodeInput}
+\begin{CodeOutput}
+R RNG seed set to 2120
+\end{CodeOutput}
+\begin{CodeOutput}
+10-Fold Cross Validation
+method: Woodbury
+criterion: mse
+cross-validation criterion = 18.89822
+bias-adjusted cross-validation criterion = 18.85429
+full-sample criterion = 18.07817 
+\end{CodeOutput}
+\end{CodeChunk}
+
+As expected, recursive CV produces a larger estimate of MSE for the
+selected 7th degree polynomial model than CV applied directly to this
+model.
+
+We can equivalently call \texttt{cv()} with the list of models as its
+first argument and set the argument \texttt{recursive=TRUE}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> cv(mlist, data=Auto, seed=2120, recursive=TRUE, save.model=TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+R RNG seed set to 2120
+\end{CodeOutput}
+\begin{CodeOutput}
+10-Fold Cross Validation
+cross-validation criterion = 20.01249
+bias-adjusted cross-validation criterion = 20.61866
+full-sample criterion = 18.74615 
+\end{CodeOutput}
+\end{CodeChunk}
+
 \section{Extending the cv package}\label{extending-the-cv-package}
 
-The \pkg{cv} package is designed to be extensible in several directions;
-in order of increasing general complexity, we can add: (1) a
+The \pkg{cv} package is designed to be extensible in several directions.
+In order of increasing general complexity, we can add: (1) a
 cross-validation cost criterion; (2) a model class that's not directly
@@
 use with the \code{"function"} method for \code{cv()}. In this section,
-we illustrate (1) and (2); more extensive examples may be found in the
-vignette on extending the \pkg{cv} package.
+we illustrate (1) and (2); more diverse and extensive examples may be
+found in the vignette on extending the \pkg{cv} package.
 
@@
 +             fixed.predictors=list(given.values=c(gendermale=0.5))),
-+      lines=list(col=c("blue", "red", "orange")),
++      lines=list(col=c("blue", "red", "orange")), # party colors
 +      axes=list(x=list(rug=FALSE), y=list(style="stacked")))
@@
 closely with the party positions on European integration: The
-Conservative Party was ``Eurosceptic,'' while Labour and the Liberal
-Democrats were supportive of the UK's participation in the EU.
+Conservative Party was relatively Euro-skeptic, while Labour and the
+Liberal Democrats were more supportive of the UK's participation in the
+EU.
 
@@
 }
-<bytecode: 0x12000eac8>
+<bytecode: 0x12e288c00>
 <environment: namespace:cv>
@@
 proabability of a 1 (``success'') response in a binary regression model
-to 0 or 1 to obtain a categorical prediction and then reports the
+to 0 or 1 to obtain a categorical prediction, and then reports the
 proportion of incorrect predictions. Because the Bayes's rule criterion
@@
 \code{"casewise loss"} attribute is attached to the result, enabling the
-computation of bias correction and confidence intervals (as discussion
-in Section
+computation of bias correction and confidence intervals (as discussed in
+Section
 \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
@@
 
-That is, we simply call the default \texttt{cv()} method with the
-\texttt{type} argument properly set. In addition to supplying the
-correct \texttt{type} argument, our method sets the default
-\texttt{criterion} for the \texttt{cv.multinom()} method to
-\texttt{BayesRuleMulti}. Adding the argument \texttt{criterion.name=}
-\texttt{deparse(substitute(criterion))} is inessential, but it insures
-that printed output will include the name of the criterion function
-that's employed, whether it's the default \texttt{BayesRuleMulti} or
-something else. Prior to invoking \texttt{NextMethod()}, we called
-\texttt{update()} with \texttt{trace=FALSE} to suppress the iteration
-history reported by default by \texttt{multinom()}---it would be tedious
-to see the iteration history for each fold.
+That is, we simply call the default \texttt{cv()} method (via
+\code{NextMethod()}) with the \texttt{type} argument properly set. In
+addition to supplying the correct \texttt{type} argument, our method
+sets the default \texttt{criterion} for the \texttt{cv.multinom()}
+method to \texttt{BayesRuleMulti}. Adding the argument
+\texttt{criterion.name=} \texttt{deparse(substitute(criterion))} is
+inessential, but it insures that printed output will include the name of
+the criterion function that's employed, whether it's the default
+\texttt{BayesRuleMulti} or something else. Prior to invoking
+\texttt{NextMethod()}, we called \texttt{update()} with
+\texttt{trace=FALSE} to suppress the iteration history reported by
+default by \texttt{multinom()}---it would be tedious to see the
+iteration history for each fold.
 

### 2024-05-23 scratch/notes_john/JSS-article/JSS-article-3.tex
version -> 2.0.1; fix typo in notes vignette

@@
 \begin{CodeOutput}
-R RNG seed set to 423102
+R RNG seed set to 486347
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.27853
-bias-adjusted cross-validation criterion = 19.26291
-95% CI for bias-adjusted CV criterion = (15.80408, 22.72175)
+cross-validation criterion = 19.52274
+bias-adjusted cross-validation criterion = 19.49438
+95% CI for bias-adjusted CV criterion = (15.96188, 23.02687)
 full-sample criterion = 18.98477 
@@
 \begin{CodeOutput}
-function (y, yhat) 
-{
+function(y, yhat) {
   if (!all(y %in% c(0, 1)))
     stop("response values not all 0 or 1")
-    if (any(yhat < 0) || any(yhat > 1)) 
+  if (any(yhat < 0) ||
+      any(yhat > 1))
     stop("fitted values outside of interval [0, 1]")
   yhat <- round(yhat)
-    result <- mean(y != yhat)
+  result <- mean(y != yhat) # proportion in error
   attr(result, "casewise loss") <- "y != round(yhat)"
@@
 }
-<bytecode: 0x12e288c00>
+<bytecode: 0x111a6b780>
 <environment: namespace:cv>

### 2024-05-23 scratch/notes_john/JSS-article/JSS-article-3.Rmd
add section on other R CV packages to JSS paper

@@
 
+The \pkg{cv} package is far from unique in implementing cross-validation of regression models in \proglang{R}.
+We've already mentioned the \code{cv.glm()} function in the \pkg{boot} package.
+A general review of CV facilities in \proglang{R} is beyond the scope of the current paper, but in this section we remark selectively on other \proglang{R} packages that support CV.
+
+Our goal in implementing CV is to make it conveniently available to \proglang{R} users with limited programming skills,[^cv-by-programming] and hence to encourage its use.
+Towards this end, the \pkg{cv} package provides a simple, consistent interface through the \code{cv()} generic function, with specific methods for different classes of standard \proglang{R} statistical models.
+Moreover, we have tried to make writing methods for additional classes of statistical models as simple and straightforward as possible.
+Because CV is computationally intensive, we also aim for efficiency, by enabling parallel computations generally, and by exploiting properties of specific classes of statistical models (in particular, computations based on hatvalues and the Woodbury matrix identity for linear and generalized linear model).
+Finally, the package includes some unusual features, such as general support for mixed-effects models and for cross-validating complex model-specification procedures.
+
+[^cv-by-programming]: \proglang{R} users with sophisticated programming skills would generally find it unchallenging to implement CV directly for specific applications.
+Even in this case, however, there's an argument for using a simple pre-programmed interface to CV, to minimize programming effort and to avoid mistakes.
+
+Some \proglang{R} packages for statistical learning include facilities for cross-validation.
+Notable examples are the \pkg{caret} package [@Kuhn:2008] and the \pkg{tidyfit} package [@Pfitzinger:2024]; the latter employs \pkg{rsample} [@FrickEtAl:2024] for CV, which we will discuss presently This contrasts with the approach taken in the \pkg{cv} package, which, as we have explained, directly supports classes of standard \proglang{R} statistical models.
+
+Other packages, including \pkg{cvms} [@OlsenZachariae:2024], \pkg{mlexperiments} [@Kapsner:2024], \pkg{origami} [@CoyleEtAl:2022], and \pkg{rsample} (which also supports bootstrapping), modularize the CV process to a greater or lesser extent.
+Advantages of this approach include flexibility and generality, but a disadvantage is that use of these packages entails nontrivial programming effort and skill.
+
+We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package.
+Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]
+```{r rsample-example, cache=TRUE}
+library(rsample)
+
+data("attrition", package = "modeldata")
+nrow(attrition)
+
+# formula for model to be fit; Attrition is binary:
+mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome
+
+rs_obj <- loo_cv(attrition) # original uses vfold_cv()
+
+# user-supplied function:
+holdout_results <- function(splits, ...) {
+  mod <- glm(..., data=analysis(splits), family=binomial)
+  holdout <- assessment(splits)
+  res <- broom::augment(mod, newdata=holdout)
+  lvls <- levels(holdout$Attrition)
+  predictions <- factor(ifelse(res$.fitted > 0, lvls[2], lvls[1]),
+                        levels=lvls)
+  res$correct <- predictions == holdout$Attrition
+  res
+}
+
+library(purrr, quietly=TRUE)
+# apply CV:
+rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
+# summarize results:
+rs_obj$accuracy <- map_dbl(rs_obj$results,
+                           function(x) mean(x$correct))
+# accuracy is the complement of the Bayes-rule CV criterion:
+1 - summary(rs_obj$accuracy) 
+```
+[^k-fold-used-in-rsample]: The example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise the code from the vignette is only trivially modified here.
+
+And here's how we can perform the same computation using the \code{cv()} function in the \pkg{cv} package:
+```{r rsample-example-using-cv, cache=TRUE}
+mod.attrition <- glm(mod_form, data=attrition, family=binomial)
+(cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
+```
+
+Let's compare the computational efficiency of the two implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
+```{r compare-rsample-cv, cache=TRUE}
+all.equal(cv.attrition$`CV crit`,
+          cv(mod.attrition, k="loo", criterion=BayesRule,
+             method="Woodbury")$`CV crit`,
+          )
+all.equal(cv.attrition$`CV crit`,
+          cv(mod.attrition, k="loo", criterion=BayesRule,
+             method="hatvalues")$`CV crit`
+          )
+
+microbenchmark::microbenchmark(
+  rsample={
+    rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
+    rs_obj$accuracy <- map_dbl(rs_obj$results,
+                               function(x) mean(x$correct))
+  },
+  cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
+  cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
+             method="Woodbury"),
+  cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
+                  method="hatvalues"),
+  times=10
+)
+```
+Computing time for \pkg{rsample} for this problem is similar to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.
+
 # Computational notes
@@
 Let $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ represent a cross-validation cost criterion, such as mean-squared error, computed for all of the $n$ values of the response $\mathbf{y}$ based on fitted values $\widehat{\mathbf{y}}$ from the model fit to all of the data.
-We require that $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ is the mean of casewise components, that is, $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n\mathrm{cv}(y_i, \widehat{y}_i)$.[^contrast-function] For example, $\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$.
+We require that $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ is the mean of casewise components, that is, $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n\mathrm{cv}(y_i, \widehat{y}_i)$.[^contrast-function] For example, $\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$.[^casewise-loss]
+
+[^casewise-loss]: Some commonly employed CV criteria---such as the root-mean-squared error (RMSE), median absolute error, and, for binary-regression models, the complement of the area under the receiver operating characteristic (ROC) curve---are not means of casewise components.
+That the RMSE and median absolute error aren't means of casewise components is obvious; for the complement of the area under the ROC curve, see the vignette on extending the \pkg{cv} package.
 

### 2024-05-23 scratch/notes_john/JSS-article/JSS-article-3.tex
add section on other R CV packages to JSS paper

--- 2024-05-22 scratch/notes_john/JSS-article/JSS-article-3.tex
@@
 \begin{CodeOutput}
-R RNG seed set to 423102
+R RNG seed set to 226376
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.27853
-bias-adjusted cross-validation criterion = 19.26291
-95% CI for bias-adjusted CV criterion = (15.80408, 22.72175)
+cross-validation criterion = 19.33326
+bias-adjusted cross-validation criterion = 19.31477
+95% CI for bias-adjusted CV criterion = (15.83621, 22.79333)
 full-sample criterion = 18.98477 
@@
 }
-<bytecode: 0x12e288c00>
+<bytecode: 0x14013a7f0>
 <environment: namespace:cv>
@@
 cross-validation}\label{comparing-cv-to-other-r-software-for-cross-validation}
+
+The \pkg{cv} package is far from unique in implementing cross-validation
+of regression models in \proglang{R}. We've already mentioned the
+\code{cv.glm()} function in the \pkg{boot} package. A general review of
+CV facilities in \proglang{R} is beyond the scope of the current paper,
+but in this section we remark selectively on other \proglang{R} packages
+that support CV.
+
+Our goal in implementing CV is to make it conveniently available to
+\proglang{R} users with limited programming skills,\footnote{\proglang{R}
+  users with sophisticated programming skills would generally find it
+  unchallenging to implement CV directly for specific applications. Even
+  in this case, however, there's an argument for using a simple
+  pre-programmed interface to CV, to minimize programming effort and to
+  avoid mistakes.} and hence to encourage its use. Towards this end, the
+\pkg{cv} package provides a simple, consistent interface through the
+\code{cv()} generic function, with specific methods for different
+classes of standard \proglang{R} statistical models. Moreover, we have
+tried to make writing methods for additional classes of statistical
+models as simple and straightforward as possible. Because CV is
+computationally intensive, we also aim for efficiency, by enabling
+parallel computations generally, and by exploiting properties of
+specific classes of statistical models (in particular, computations
+based on hatvalues and the Woodbury matrix identity for linear and
+generalized linear model). Finally, the package includes some unusual
+features, such as general support for mixed-effects models and for
+cross-validating complex model-specification procedures.
+
+Some \proglang{R} packages for statistical learning include facilities
+for cross-validation. Notable examples are the \pkg{caret} package
+\citep{Kuhn:2008} and the \pkg{tidyfit} package \citep{Pfitzinger:2024};
+the latter employs \pkg{rsample} \citep{FrickEtAl:2024} for CV, which we
+will discuss presently This contrasts with the approach taken in the
+\pkg{cv} package, which, as we have explained, directly supports classes
+of standard \proglang{R} statistical models.
+
+Other packages, including \pkg{cvms} \citep{OlsenZachariae:2024},
+\pkg{mlexperiments} \citep{Kapsner:2024}, \pkg{origami}
+\citep{CoyleEtAl:2022}, and \pkg{rsample} (which also supports
+bootstrapping), modularize the CV process to a greater or lesser extent.
+Advantages of this approach include flexibility and generality, but a
+disadvantage is that use of these packages entails nontrivial
+programming effort and skill.
+
+We illustrate with an example adapted from a vignette in the
+\pkg{rsample} package, which uses the \code{attrition} data set from the
+\pkg{modeldata} package. Here, with minimal explanation, is how one
+would perform LOO CV for the example using \pkg{rsample}:\footnote{The
+  example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV,
+  but otherwise the code from the vignette is only trivially modified
+  here.}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library(rsample)
+R> 
+R> data("attrition", package = "modeldata")
+R> nrow(attrition)
+\end{CodeInput}
+\begin{CodeOutput}
+[1] 1470
+\end{CodeOutput}
+\begin{CodeInput}
+R> # formula for model to be fit; Attrition is binary:
+R> mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome
+R> 
+R> rs_obj <- loo_cv(attrition) # original uses vfold_cv()
+R> 
+R> # user-supplied function:
+R> holdout_results <- function(splits, ...) {
++   mod <- glm(..., data=analysis(splits), family=binomial)
++   holdout <- assessment(splits)
++   res <- broom::augment(mod, newdata=holdout)
++   lvls <- levels(holdout$Attrition)
++   predictions <- factor(ifelse(res$.fitted > 0, lvls[2], lvls[1]),
++                         levels=lvls)
++   res$correct <- predictions == holdout$Attrition
++   res
++ }
+R> 
+R> library(purrr, quietly=TRUE)
+R> # apply CV:
+R> rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
+R> # summarize results:
+R> rs_obj$accuracy <- map_dbl(rs_obj$results,
++                            function(x) mean(x$correct))
+R> # accuracy is the complement of the Bayes-rule CV criterion:
+R> 1 - summary(rs_obj$accuracy) 
+\end{CodeInput}
+\begin{CodeOutput}
+   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+ 1.0000  0.0000  0.0000  0.1612  0.0000  0.0000 
+\end{CodeOutput}
+\end{CodeChunk}
+
+And here's how we can perform the same computation using the \code{cv()}
+function in the \pkg{cv} package:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> mod.attrition <- glm(mod_form, data=attrition, family=binomial)
+R> (cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
+\end{CodeInput}
+\begin{CodeOutput}
+n-Fold Cross Validation
+method: exact
+criterion: BayesRule
+cross-validation criterion = 0.1612245
+bias-adjusted cross-validation criterion = 0.1612245
+95% CI for bias-adjusted CV criterion = (0.1424194, 0.1800296)
+full-sample criterion = 0.1612245 
+\end{CodeOutput}
+\end{CodeChunk}
+
+Let's compare the computational efficiency of the two implementations,
+also showing the use of \code{method="Woodbury"} and
+\code{method="hatvalues"} for \code{cv()}, which, in this example,
+produce numerically identical results to the default
+\code{method="exact"}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> all.equal(cv.attrition$`CV crit`,
++           cv(mod.attrition, k="loo", criterion=BayesRule,
++              method="Woodbury")$`CV crit`,
++           )
+\end{CodeInput}
+\begin{CodeOutput}
+[1] TRUE
+\end{CodeOutput}
+\begin{CodeInput}
+R> all.equal(cv.attrition$`CV crit`,
++           cv(mod.attrition, k="loo", criterion=BayesRule,
++              method="hatvalues")$`CV crit`
++           )
+\end{CodeInput}
+\begin{CodeOutput}
+[1] TRUE
+\end{CodeOutput}
+\begin{CodeInput}
+R> microbenchmark::microbenchmark(
++   rsample={
++     rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
++     rs_obj$accuracy <- map_dbl(rs_obj$results,
++                                function(x) mean(x$correct))
++   },
++   cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
++   cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
++              method="Woodbury"),
++   cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
++                   method="hatvalues"),
++   times=10
++ )
+\end{CodeInput}
+\begin{CodeOutput}
+Warning in microbenchmark::microbenchmark(rsample = {: less accurate nanosecond
+times to avoid potential integer overflows
+\end{CodeOutput}
+\begin{CodeOutput}
+Unit: milliseconds
+         expr         min          lq        mean      median          uq
+      rsample 5814.989369 5864.285227 5952.773502 5897.248509 6023.438043
+     cv.exact 4739.827837 4827.064193 4902.546681 4910.249278 4952.683150
+      cv.wood  103.663047  104.760125  110.628578  111.201532  116.027007
+ cv.hatvalues    1.217905    1.222251    1.409719    1.479649    1.543937
+         max neval
+ 6207.205988    10
+ 5090.904318    10
+  118.610540    10
+    1.591538    10
+\end{CodeOutput}
+\end{CodeChunk}
+
+Computing time for \pkg{rsample} for this problem is similar to
+\code{cv()} with \code{method="exact"}, but more than an order of
+magitude slower than \code{cv()} with \code{method="Woodbury"}, and more
+than three orders of magnitude slower then \code{cv()} with
+\code{method="hatvalues"}.
 
@@
   ``contrast function.''} For example,
-\(\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\).
+\(\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\).\footnote{Some
+  commonly employed CV criteria---such as the root-mean-squared error
+  (RMSE), median absolute error, and, for binary-regression models, the
+  complement of the area under the receiver operating characteristic
+  (ROC) curve---are not means of casewise components. That the RMSE and
+  median absolute error aren't means of casewise components is obvious;
+  for the complement of the area under the ROC curve, see the vignette
+  on extending the \pkg{cv} package.}
 

### 2024-05-24 scratch/notes_john/JSS-article/JSS-article-3.Rmd
add abstract to JSS paper + minor edits

@@
 abstract: >
-  The abstract of the article.
+  We describe the \pkg{cv} package, which implements cross-validation for standard \proglang{R} regression models through a uniform and simple to use generic function.
+Methods for the \code{cv()} function are provided for many commonly employed classes of statistical models, including mixed-effects models, and it is straightforward to extend the \pkg{cv} package by writing methods for additional classes of models.
+The \code{cv()} function also can cross-validate complex model-selection procedures, such as those that include variable transformations, predictor selection, and selection among competing models.
+The \code{cv()} function generally supports parallel computations, and the supplied methods for linear and generalized linear models take advantage of computational efficiencies.
 keywords:
@@
 
-## Example: Applying recursive CV to polynomial regression for the Auto data
+## Example: Applying recursive CV to polynomial regression for the Auto data[^recursive-cv]
+
+[^recursive-cv]: What we call here "recursive CV" has also been termed "nested CV" [e.g., in the \pkg{cvms} package by @OlsenZachariae:2024].
+We prefer, however, to reserve the term "nested CV" for the procedure described by [@BatesHastieTibshirani:2023].
 
@@
 
-We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package.
+We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024].
 Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]

### 2024-05-24 scratch/notes_john/JSS-article/JSS-article-3.tex
add abstract to JSS paper + minor edits

@@
 \Abstract{
-The abstract of the article.
+We describe the \pkg{cv} package, which implements cross-validation for
+standard \proglang{R} regression models through a uniform and simple to
+use generic function. Methods for the \code{cv()} function are provided
+for many commonly employed classes of statistical models, including
+mixed-effects models, and it is straightforward to extend the \pkg{cv}
+package by writing methods for additional classes of models. The
+\code{cv()} function also can cross-validate complex model-selection
+procedures, such as those that include variable transformations,
+predictor selection, and selection among competing models. The
+\code{cv()} function generally supports parallel computations, and the
+supplied methods for linear and generalized linear models take advantage
+of computational efficiencies.
 }
@@
 \begin{CodeOutput}
-R RNG seed set to 226376
+R RNG seed set to 891621
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.33326
-bias-adjusted cross-validation criterion = 19.31477
-95% CI for bias-adjusted CV criterion = (15.83621, 22.79333)
+cross-validation criterion = 19.30934
+bias-adjusted cross-validation criterion = 19.29207
+95% CI for bias-adjusted CV criterion = (15.80161, 22.78253)
 full-sample criterion = 18.98477 
@@
 \begin{CodeOutput}
-Unit: milliseconds
+Unit: microseconds
       expr        min         lq       mean     median         uq        max
- hatvalues   1.139062   1.184244   1.202678   1.194248   1.205769   1.297117
-  Woodbury  12.106521  12.310701  13.115158  12.337515  12.808810  16.149900
-     naive 217.941773 219.491204 225.861284 220.053909 222.015410 278.078974
-    cv.glm 373.900074 379.259963 397.907632 382.333036 437.228141 441.112358
+ hatvalues    992.282   1149.681   1181.427   1199.188   1223.235   1356.485
+  Woodbury  12297.663  12375.317  14678.447  12908.952  16017.675  22227.535
+     naive 217204.347 222086.545 249652.690 250411.416 277112.440 279141.530
+    cv.glm 379237.823 382052.063 401064.952 388823.726 435052.394 440563.573
  neval
@@
 
-\subsection{Example: Applying recursive CV to polynomial regression for
-the Auto
-data}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-data}
+\subsection[Example: Applying recursive CV to polynomial regression for
+the Auto data]{\texorpdfstring{Example: Applying recursive CV to
+polynomial regression for the Auto
+data\footnote{What we call here ``recursive CV'' has also been termed
+  ``nested CV'' \citep[e.g., in the \pkg{cvms} package
+  by][]{OlsenZachariae:2024}. We prefer, however, to reserve the term
+  ``nested CV'' for the procedure described by
+  \citep{BatesHastieTibshirani:2023}.}}{Example: Applying recursive CV to polynomial regression for the Auto data}}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-datarecursive-cv}
 
@@
 }
-<bytecode: 0x14013a7f0>
+<bytecode: 0x1293c2ea8>
 <environment: namespace:cv>
@@
 \pkg{rsample} package, which uses the \code{attrition} data set from the
-\pkg{modeldata} package. Here, with minimal explanation, is how one
-would perform LOO CV for the example using \pkg{rsample}:\footnote{The
-  example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV,
-  but otherwise the code from the vignette is only trivially modified
-  here.}
+\pkg{modeldata} package \citep{Kuhn:2024}. Here, with minimal
+explanation, is how one would perform LOO CV for this example using
+\pkg{rsample}:\footnote{The example in the \pkg{rsample} vignette uses
+  10-fold rather than LOO CV, but otherwise the code from the vignette
+  is only trivially modified here.}
 
@@
 R> library(purrr, quietly=TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Attaching package: 'purrr'
+\end{CodeOutput}
+\begin{CodeOutput}
+The following object is masked from 'package:car':
+
+    some
+\end{CodeOutput}
+\begin{CodeOutput}
+The following objects are masked from 'package:foreach':
+
+    accumulate, when
+\end{CodeOutput}
+\begin{CodeInput}
 R> # apply CV:
@@
 \begin{CodeOutput}
-Warning in microbenchmark::microbenchmark(rsample = {: less accurate nanosecond
-times to avoid potential integer overflows
-\end{CodeOutput}
-\begin{CodeOutput}
 Unit: milliseconds
          expr         min          lq        mean      median          uq
-      rsample 5814.989369 5864.285227 5952.773502 5897.248509 6023.438043
-     cv.exact 4739.827837 4827.064193 4902.546681 4910.249278 4952.683150
-      cv.wood  103.663047  104.760125  110.628578  111.201532  116.027007
- cv.hatvalues    1.217905    1.222251    1.409719    1.479649    1.543937
+      rsample 5858.845429 5918.511745 6182.166132 6032.719008 6513.721496
+     cv.exact 4823.821011 4879.622175 4986.028516 4973.975331 5051.786505
+      cv.wood  103.815690  107.823973  112.389204  112.138341  117.148398
+ cv.hatvalues    1.248655    1.299823    1.540665    1.587007    1.654432
          max neval
- 6207.205988    10
- 5090.904318    10
-  118.610540    10
-    1.591538    10
+ 6707.589832    10
+ 5305.047728    10
+  121.483287    10
+    1.916832    10
 \end{CodeOutput}
@@
 \code{cv()} with \code{method="exact"}, but more than an order of
-magitude slower than \code{cv()} with \code{method="Woodbury"}, and more
-than three orders of magnitude slower then \code{cv()} with
+magnitude slower than \code{cv()} with \code{method="Woodbury"}, and
+more than three orders of magnitude slower then \code{cv()} with
 \code{method="hatvalues"}.

### 2024-05-25 scratch/notes_john/JSS-article/JSS-article-3.Rmd
tweak draft JSS paper

@@
 
-We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024].
+We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024] (and which is by a wide margin the most downloaded of the CV packages mentioned in the preceding paragraph).
 Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]

### 2024-05-25 scratch/notes_john/JSS-article/JSS-article-3.tex
tweak draft JSS paper

@@
 \begin{CodeOutput}
-R RNG seed set to 891621
+R RNG seed set to 508110
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.30934
-bias-adjusted cross-validation criterion = 19.29207
-95% CI for bias-adjusted CV criterion = (15.80161, 22.78253)
+cross-validation criterion = 19.23279
+bias-adjusted cross-validation criterion = 19.21953
+95% CI for bias-adjusted CV criterion = (15.76747, 22.67159)
 full-sample criterion = 18.98477 
@@
 }
-<bytecode: 0x1293c2ea8>
+<bytecode: 0x14aa65750>
 <environment: namespace:cv>
@@
 \pkg{rsample} package, which uses the \code{attrition} data set from the
-\pkg{modeldata} package \citep{Kuhn:2024}. Here, with minimal
-explanation, is how one would perform LOO CV for this example using
-\pkg{rsample}:\footnote{The example in the \pkg{rsample} vignette uses
-  10-fold rather than LOO CV, but otherwise the code from the vignette
-  is only trivially modified here.}
+\pkg{modeldata} package \citep{Kuhn:2024} (and which is by a wide margin
+the most downloaded of the CV packages mentioned in the preceding
+paragraph). Here, with minimal explanation, is how one would perform LOO
+CV for this example using \pkg{rsample}:\footnote{The example in the
+  \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise
+  the code from the vignette is only trivially modified here.}
 

### 2024-05-25 scratch/notes_john/JSS-article/JSS-article-3.Rmd
further tweak JSS draft paper

@@
 (cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
+all.equal(cv.attrition$`CV crit`, 1 - mean(rs_obj$accuracy), 
+          check.attributes=FALSE)
 ```

### 2024-05-25 scratch/notes_john/JSS-article/JSS-article-3.tex
further tweak JSS draft paper

@@
 \begin{CodeOutput}
-R RNG seed set to 508110
+R RNG seed set to 866542
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.23279
-bias-adjusted cross-validation criterion = 19.21953
-95% CI for bias-adjusted CV criterion = (15.76747, 22.67159)
+cross-validation criterion = 19.09701
+bias-adjusted cross-validation criterion = 19.09108
+95% CI for bias-adjusted CV criterion = (15.65339, 22.52877)
 full-sample criterion = 18.98477 
@@
 }
-<bytecode: 0x14aa65750>
+<bytecode: 0x13b740948>
 <environment: namespace:cv>
@@
 \end{CodeOutput}
+\begin{CodeInput}
+R> all.equal(cv.attrition$`CV crit`, 1 - mean(rs_obj$accuracy), 
++           check.attributes=FALSE)
+\end{CodeInput}
+\begin{CodeOutput}
+[1] TRUE
+\end{CodeOutput}
 \end{CodeChunk}

### 2024-05-26 scratch/notes_john/JSS-article/JSS-article-3.Rmd
further edits to draft JSS paper

@@
 [^recursive-cv]: What we call here "recursive CV" has also been termed "nested CV" [e.g., in the \pkg{cvms} package by @OlsenZachariae:2024].
-We prefer, however, to reserve the term "nested CV" for the procedure described by [@BatesHastieTibshirani:2023].
+We prefer, however, to reserve the term "nested CV" for the procedure described by @BatesHastieTibshirani:2023.
 
 In Section \ref{comparing-competing-models}, following @JamesEtAl:2021[Secs.
-5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of `mpg` to `horsepower` for the `Auto` data, saving the 10 models, named `m.1` through `m.10`, in `mlist`.
+5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of `mpg` to `horsepower` for the `Auto` data, saving the 10 models, named `m.1` through `m.10`, in the list `mlist`.
 We then used `cv()` to compare the cross-validated MSE for the 10 models, discovering that the 7th degree polynomial had the smallest MSE (by a small margin).
 
-If we then select the 7th degree polynomial model, intending to use it for prediction, the CV estimate of the MSE for this model will be optimistic.
+If we select the 7th degree polynomial model, intending to use it for prediction, the CV estimate of the MSE for this model will be optimistic.
 One solution is to cross-validate the process of using CV to select the "best" model---that is, to apply CV to CV recursively.
@@
 Some \proglang{R} packages for statistical learning include facilities for cross-validation.
-Notable examples are the \pkg{caret} package [@Kuhn:2008] and the \pkg{tidyfit} package [@Pfitzinger:2024]; the latter employs \pkg{rsample} [@FrickEtAl:2024] for CV, which we will discuss presently This contrasts with the approach taken in the \pkg{cv} package, which, as we have explained, directly supports classes of standard \proglang{R} statistical models.
+Notable examples are the \pkg{caret} package [@Kuhn:2008] and the \pkg{tidyfit} package [@Pfitzinger:2024]; the latter employs \pkg{rsample} [@FrickEtAl:2024] for CV, which we will discuss presently.
+This contrasts with the approach taken in the \pkg{cv} package, which, as we have explained, directly supports classes of standard \proglang{R} statistical models.
 
@@
 
-library(purrr, quietly=TRUE)
+library(purrr, warn.conflicts=FALSE)
 # apply CV:

### 2024-05-26 scratch/notes_john/JSS-article/JSS-article-3.tex
further edits to draft JSS paper

@@
 \begin{CodeOutput}
-R RNG seed set to 866542
+R RNG seed set to 968919
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.09701
-bias-adjusted cross-validation criterion = 19.09108
-95% CI for bias-adjusted CV criterion = (15.65339, 22.52877)
+cross-validation criterion = 19.22048
+bias-adjusted cross-validation criterion = 19.20805
+95% CI for bias-adjusted CV criterion = (15.74739, 22.66871)
 full-sample criterion = 18.98477 
@@
   ``nested CV'' for the procedure described by
-  \citep{BatesHastieTibshirani:2023}.}}{Example: Applying recursive CV to polynomial regression for the Auto data}}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-datarecursive-cv}
+  \citet{BatesHastieTibshirani:2023}.}}{Example: Applying recursive CV to polynomial regression for the Auto data}}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-datarecursive-cv}
 
@@
 \texttt{Auto} data, saving the 10 models, named \texttt{m.1} through
-\texttt{m.10}, in \texttt{mlist}. We then used \texttt{cv()} to compare
-the cross-validated MSE for the 10 models, discovering that the 7th
-degree polynomial had the smallest MSE (by a small margin).
-
-If we then select the 7th degree polynomial model, intending to use it
-for prediction, the CV estimate of the MSE for this model will be
+\texttt{m.10}, in the list \texttt{mlist}. We then used \texttt{cv()} to
+compare the cross-validated MSE for the 10 models, discovering that the
+7th degree polynomial had the smallest MSE (by a small margin).
+
+If we select the 7th degree polynomial model, intending to use it for
+prediction, the CV estimate of the MSE for this model will be
 optimistic. One solution is to cross-validate the process of using CV to
@@
 }
-<bytecode: 0x13b740948>
+<bytecode: 0x142fa0f70>
 <environment: namespace:cv>
@@
 the latter employs \pkg{rsample} \citep{FrickEtAl:2024} for CV, which we
-will discuss presently This contrasts with the approach taken in the
+will discuss presently. This contrasts with the approach taken in the
 \pkg{cv} package, which, as we have explained, directly supports classes
@@
 R> 
-R> library(purrr, quietly=TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Attaching package: 'purrr'
-\end{CodeOutput}
-\begin{CodeOutput}
-The following object is masked from 'package:car':
-
-    some
-\end{CodeOutput}
-\begin{CodeOutput}
-The following objects are masked from 'package:foreach':
-
-    accumulate, when
-\end{CodeOutput}
-\begin{CodeInput}
+R> library(purrr, warn.conflicts=FALSE)
 R> # apply CV:

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.Rmd
add direct CV computation to speed comparisons in JSS draft paper

@@
 
-[^cv-by-programming]: \proglang{R} users with sophisticated programming skills would generally find it unchallenging to implement CV directly for specific applications.
+[^cv-by-programming]: \proglang{R} users with sophisticated programming skills would generally find it unchallenging to implement CV directly for specific applications (as we illustrate below).
 Even in this case, however, there's an argument for using a simple pre-programmed interface to CV, to minimize programming effort and to avoid mistakes.
@@
 
-We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024] (and which is by a wide margin the most downloaded of the CV packages mentioned in the preceding paragraph).
+We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024] (and which is by a wide margin the most downloaded from the RStudio CRAN mirror of the CV packages mentioned in the preceding paragraph).
 Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]
@@
 
+In this case, it's simple (and, in our opinion, simpler) to program LOO CV directly. For example
+```{r direct-loo-cv, cache=TRUE}
+mod.attrition <- glm(mod_form, data=attrition, family=binomial)
+n <- nrow(attrition)
+yhat <- numeric(n)
+for (i in 1:n){
+  m <- update(mod.attrition, data=attrition[-i, ])
+  yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
+}
+mean(((attrition$Attrition == "Yes") - round(yhat))^2)
+```
+
 And here's how we can perform the same computation using the \code{cv()} function in the \pkg{cv} package:
 ```{r rsample-example-using-cv, cache=TRUE}
-mod.attrition <- glm(mod_form, data=attrition, family=binomial)
 (cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
@@
 
-Let's compare the computational efficiency of the two implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
+Let's compare the computational efficiency of the three implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
 ```{r compare-rsample-cv, cache=TRUE}
@@
   },
+  direct={
+    n <- nrow(attrition)
+    yhat <- numeric(n)
+    for (i in 1:n){
+      m <- update(mod.attrition, data=attrition[-i, ])
+      yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
+    }
+    mean(((attrition$Attrition == "Yes") - round(yhat))^2)
+  },
   cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
@@
 ```
-Computing time for \pkg{rsample} for this problem is similar to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.
+Computing time for \pkg{rsample} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.
 

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.tex
add direct CV computation to speed comparisons in JSS draft paper

@@
 \begin{CodeOutput}
-R RNG seed set to 968919
+R RNG seed set to 716237
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.22048
-bias-adjusted cross-validation criterion = 19.20805
-95% CI for bias-adjusted CV criterion = (15.74739, 22.66871)
+cross-validation criterion = 19.28027
+bias-adjusted cross-validation criterion = 19.2646
+95% CI for bias-adjusted CV criterion = (15.81196, 22.71725)
 full-sample criterion = 18.98477 
@@
       expr        min         lq       mean     median         uq        max
- hatvalues    992.282   1149.681   1181.427   1199.188   1223.235   1356.485
-  Woodbury  12297.663  12375.317  14678.447  12908.952  16017.675  22227.535
-     naive 217204.347 222086.545 249652.690 250411.416 277112.440 279141.530
-    cv.glm 379237.823 382052.063 401064.952 388823.726 435052.394 440563.573
+ hatvalues    984.369   1182.604   1188.963   1202.345   1220.242   1293.468
+  Woodbury  12366.543  12508.690  13408.099  12769.347  12815.124  16566.050
+     naive 216467.126 217748.581 225187.170 219534.685 222515.077 276207.898
+    cv.glm 373527.917 380937.683 393430.301 382341.584 389266.341 443234.190
  neval
@@
 }
-<bytecode: 0x142fa0f70>
+<bytecode: 0x126d8b808>
 <environment: namespace:cv>
@@
   users with sophisticated programming skills would generally find it
-  unchallenging to implement CV directly for specific applications. Even
-  in this case, however, there's an argument for using a simple
-  pre-programmed interface to CV, to minimize programming effort and to
-  avoid mistakes.} and hence to encourage its use. Towards this end, the
-\pkg{cv} package provides a simple, consistent interface through the
-\code{cv()} generic function, with specific methods for different
-classes of standard \proglang{R} statistical models. Moreover, we have
-tried to make writing methods for additional classes of statistical
-models as simple and straightforward as possible. Because CV is
-computationally intensive, we also aim for efficiency, by enabling
-parallel computations generally, and by exploiting properties of
-specific classes of statistical models (in particular, computations
-based on hatvalues and the Woodbury matrix identity for linear and
-generalized linear model). Finally, the package includes some unusual
-features, such as general support for mixed-effects models and for
-cross-validating complex model-specification procedures.
+  unchallenging to implement CV directly for specific applications (as
+  we illustrate below). Even in this case, however, there's an argument
+  for using a simple pre-programmed interface to CV, to minimize
+  programming effort and to avoid mistakes.} and hence to encourage its
+use. Towards this end, the \pkg{cv} package provides a simple,
+consistent interface through the \code{cv()} generic function, with
+specific methods for different classes of standard \proglang{R}
+statistical models. Moreover, we have tried to make writing methods for
+additional classes of statistical models as simple and straightforward
+as possible. Because CV is computationally intensive, we also aim for
+efficiency, by enabling parallel computations generally, and by
+exploiting properties of specific classes of statistical models (in
+particular, computations based on hatvalues and the Woodbury matrix
+identity for linear and generalized linear model). Finally, the package
+includes some unusual features, such as general support for
+mixed-effects models and for cross-validating complex
+model-specification procedures.
 
@@
 \pkg{modeldata} package \citep{Kuhn:2024} (and which is by a wide margin
-the most downloaded of the CV packages mentioned in the preceding
-paragraph). Here, with minimal explanation, is how one would perform LOO
-CV for this example using \pkg{rsample}:\footnote{The example in the
-  \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise
-  the code from the vignette is only trivially modified here.}
+the most downloaded from the RStudio CRAN mirror of the CV packages
+mentioned in the preceding paragraph). Here, with minimal explanation,
+is how one would perform LOO CV for this example using
+\pkg{rsample}:\footnote{The example in the \pkg{rsample} vignette uses
+  10-fold rather than LOO CV, but otherwise the code from the vignette
+  is only trivially modified here.}
 
@@
 
+In this case, it's simple (and, in our opinion, simpler) to program LOO
+CV directly. For example
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> mod.attrition <- glm(mod_form, data=attrition, family=binomial)
+R> n <- nrow(attrition)
+R> yhat <- numeric(n)
+R> for (i in 1:n){
++   m <- update(mod.attrition, data=attrition[-i, ])
++   yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
++ }
+R> mean(((attrition$Attrition == "Yes") - round(yhat))^2)
+\end{CodeInput}
+\begin{CodeOutput}
+[1] 0.1612245
+\end{CodeOutput}
+\end{CodeChunk}
+
 And here's how we can perform the same computation using the \code{cv()}
@@
 \begin{CodeInput}
-R> mod.attrition <- glm(mod_form, data=attrition, family=binomial)
 R> (cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
@@
 
-Let's compare the computational efficiency of the two implementations,
+Let's compare the computational efficiency of the three implementations,
 also showing the use of \code{method="Woodbury"} and
@@
 +                                function(x) mean(x$correct))
++   },
++   direct={
++     n <- nrow(attrition)
++     yhat <- numeric(n)
++     for (i in 1:n){
++       m <- update(mod.attrition, data=attrition[-i, ])
++       yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
++     }
++     mean(((attrition$Attrition == "Yes") - round(yhat))^2)
 +   },
@@
 \begin{CodeOutput}
+Warning in microbenchmark::microbenchmark(rsample = {: less accurate nanosecond
+times to avoid potential integer overflows
+\end{CodeOutput}
+\begin{CodeOutput}
 Unit: milliseconds
          expr         min          lq        mean      median         uq
-      rsample 5858.845429 5918.511745 6182.166132 6032.719008 6513.721496
-     cv.exact 4823.821011 4879.622175 4986.028516 4973.975331 5051.786505
-      cv.wood  103.815690  107.823973  112.389204  112.138341  117.148398
- cv.hatvalues    1.248655    1.299823    1.540665    1.587007    1.654432
+      rsample 5707.408772 5711.011032 5810.464433 5828.739437 5853.64979
+       direct 5187.215696 5206.152940 5281.503347 5230.005408 5348.24496
+     cv.exact 4659.560948 4691.022749 4735.789317 4737.251540 4790.76759
+      cv.wood   99.232628  104.541882  106.917914  106.051051  108.49436
+ cv.hatvalues    1.187401    1.224875    2.751965    1.356813    1.56374
         max neval
- 6707.589832    10
- 5305.047728    10
-  121.483287    10
-    1.916832    10
-\end{CodeOutput}
-\end{CodeChunk}
-
-Computing time for \pkg{rsample} for this problem is similar to
-\code{cv()} with \code{method="exact"}, but more than an order of
-magnitude slower than \code{cv()} with \code{method="Woodbury"}, and
-more than three orders of magnitude slower then \code{cv()} with
-\code{method="hatvalues"}.
+ 5999.09487    10
+ 5500.60457    10
+ 4811.15176    10
+  119.72115    10
+   15.35548    10
+\end{CodeOutput}
+\end{CodeChunk}
+
+Computing time for \pkg{rsample} for this problem is similar to direct
+computation of LOO CV and to \code{cv()} with \code{method="exact"}, but
+more than an order of magnitude slower than \code{cv()} with
+\code{method="Woodbury"}, and more than three orders of magnitude slower
+then \code{cv()} with \code{method="hatvalues"}.
 

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.Rmd
add caret to package comparisons in draft JSS paper

@@
 
-We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024] (and which is by a wide margin the most downloaded from the RStudio CRAN mirror of the CV packages mentioned in the preceding paragraph).
+<!-- We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024] (and which is by a wide margin the most downloaded from the RStudio CRAN mirror of the CV packages mentioned in the preceding paragraph).
+-->
+
+We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024], comparing LOO CV computed using the \code{cv()} function in the \pkg{cv} package to LOO CV computed using the \pkg{caret} and \pkg{rsample} packages.
+At more than 15 million cumulative downloads, \pkg{caret} is by a very wide margin the most downloaded from the RStudio CRAN mirror of the \proglang{R} packages mentioned above.
+Among the packages that focus more on cross-validation, \pkg{rsample} is the most downloaded, at nearly 4 million downloads.
+
 Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]
 ```{r rsample-example, cache=TRUE}
-library(rsample)
+library("rsample")
 
@@
 
-library(purrr, warn.conflicts=FALSE)
+library("purrr", warn.conflicts=FALSE)
 # apply CV:
@@
 
+The \pkg{caret} package doesn't directly support generalized linear models fit by \code{glm()}, but it can be made to do so by defining a custom "method" for the binary logistic regression fit to the \code{attrition} data:
+```{r rsample-example-using-caret, cache=TRUE}
+library("caret", quietly=TRUE, warn.conflicts=FALSE)
+
+glm_mod <- getModelInfo("glm", regex = FALSE)[[1]]
+glm_mod$fit <- function(x, y, wts, param, lev, last, classProbs, ...) {
+  x$Attrition <- y
+  glm(Attrition ~ JobSatisfaction + Gender + MonthlyIncome,
+      family = binomial, data = x)
+}
+
+mod <- train(x = attrition[, c("JobSatisfaction", "Gender",
+                               "MonthlyIncome")],
+             y = attrition$Attrition,
+             method = glm_mod,
+             trControl = trainControl(method="LOOCV"))
+mod # reports complement of Bayes-rule criterion
+```
+
 And here's how we can perform the same computation using the \code{cv()} function in the \pkg{cv} package:
@@
 
-Let's compare the computational efficiency of the three implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
+Let's compare the computational efficiency of the various implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
 ```{r compare-rsample-cv, cache=TRUE}
@@
   },
+  caret=train(x = attrition[, c("JobSatisfaction", "Gender",
+                                "MonthlyIncome")],
+              y = attrition$Attrition,
+              method = glm_mod,
+              trControl = trainControl(method="LOOCV")),
   direct={
@@
 ```
-Computing time for \pkg{rsample} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.
+Computing time for \pkg{rsample} and \pkg{caret} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.
 

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.tex
add caret to package comparisons in draft JSS paper

@@
 \begin{CodeOutput}
-R RNG seed set to 716237
+R RNG seed set to 799417
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.28027
-bias-adjusted cross-validation criterion = 19.2646
-95% CI for bias-adjusted CV criterion = (15.81196, 22.71725)
+cross-validation criterion = 19.20609
+bias-adjusted cross-validation criterion = 19.19442
+95% CI for bias-adjusted CV criterion = (15.73743, 22.6514)
 full-sample criterion = 18.98477 
@@
 }
-<bytecode: 0x126d8b808>
+<bytecode: 0x1118adc78>
 <environment: namespace:cv>
@@
 \pkg{rsample} package, which uses the \code{attrition} data set from the
-\pkg{modeldata} package \citep{Kuhn:2024} (and which is by a wide margin
-the most downloaded from the RStudio CRAN mirror of the CV packages
-mentioned in the preceding paragraph). Here, with minimal explanation,
-is how one would perform LOO CV for this example using
-\pkg{rsample}:\footnote{The example in the \pkg{rsample} vignette uses
-  10-fold rather than LOO CV, but otherwise the code from the vignette
-  is only trivially modified here.}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library(rsample)
+\pkg{modeldata} package \citep{Kuhn:2024}, comparing LOO CV computed
+using the \code{cv()} function in the \pkg{cv} package to LOO CV
+computed using the \pkg{caret} and \pkg{rsample} packages. At more than
+15 million cumulative downloads, \pkg{caret} is by a very wide margin
+the most downloaded from the RStudio CRAN mirror of the \proglang{R}
+packages mentioned above. Among the packages that focus more on
+cross-validation, \pkg{rsample} is the most downloaded, at nearly 4
+million downloads.
+
+Here, with minimal explanation, is how one would perform LOO CV for this
+example using \pkg{rsample}:\footnote{The example in the \pkg{rsample}
+  vignette uses 10-fold rather than LOO CV, but otherwise the code from
+  the vignette is only trivially modified here.}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library("rsample")
 R> 
@@
 R> 
-R> library(purrr, warn.conflicts=FALSE)
+R> library("purrr", warn.conflicts=FALSE)
 R> # apply CV:
@@
 
+The \pkg{caret} package doesn't directly support generalized linear
+models fit by \code{glm()}, but it can be made to do so by defining a
+custom ``method'' for the binary logistic regression fit to the
+\code{attrition} data:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library("caret", quietly=TRUE, warn.conflicts=FALSE)
+R> 
+R> glm_mod <- getModelInfo("glm", regex = FALSE)[[1]]
+R> glm_mod$fit <- function(x, y, wts, param, lev, last, classProbs, ...) {
++   x$Attrition <- y
++   glm(Attrition ~ JobSatisfaction + Gender + MonthlyIncome,
++       family = binomial, data = x)
++ }
+R> 
+R> mod <- train(x = attrition[, c("JobSatisfaction", "Gender",
++                                "MonthlyIncome")],
++              y = attrition$Attrition,
++              method = glm_mod,
++              trControl = trainControl(method="LOOCV"))
+R> mod # reports complement of Bayes-rule criterion
+\end{CodeInput}
+\begin{CodeOutput}
+Generalized Linear Model
+
+1470 samples
+   3 predictor
+   2 classes: 'No', 'Yes'
+
+No pre-processing
+Resampling: Leave-One-Out Cross-Validation 
+Summary of sample sizes: 1469, 1469, 1469, 1469, 1469, 1469, ... 
+Resampling results:
+
+  Accuracy   Kappa
+  0.8387755  0    
+\end{CodeOutput}
+\end{CodeChunk}
+
 And here's how we can perform the same computation using the \code{cv()}
@@
 
-Let's compare the computational efficiency of the three implementations,
-also showing the use of \code{method="Woodbury"} and
+Let's compare the computational efficiency of the various
+implementations, also showing the use of \code{method="Woodbury"} and
 \code{method="hatvalues"} for \code{cv()}, which, in this example,
@@
 +   },
++   caret=train(x = attrition[, c("JobSatisfaction", "Gender",
++                                 "MonthlyIncome")],
++               y = attrition$Attrition,
++               method = glm_mod,
++               trControl = trainControl(method="LOOCV")),
 +   direct={
@@
          expr         min          lq        mean      median          uq
-      rsample 5707.408772 5711.011032 5810.464433 5828.739437 5853.64979
-       direct 5187.215696 5206.152940 5281.503347 5230.005408 5348.24496
-     cv.exact 4659.560948 4691.022749 4735.789317 4737.251540 4790.76759
-      cv.wood   99.232628  104.541882  106.917914  106.051051  108.49436
- cv.hatvalues    1.187401    1.224875    2.751965    1.356813    1.56374
+      rsample 5853.035606 5948.291603 6163.101903 6121.042951 6144.223100
+        caret 6804.763276 6951.219335 7097.717763 7061.439656 7255.566640
+       direct 5310.127259 5350.848336 5554.513023 5484.144809 5690.480487
+     cv.exact 4751.498733 4797.980556 4879.481352 4870.303039 4927.941413
+      cv.wood  101.606159  103.891294  108.285899  108.577430  112.075263
+ cv.hatvalues    1.206999    1.471613    1.514548    1.554679    1.614416
          max neval
- 5999.09487    10
- 5500.60457    10
- 4811.15176    10
-  119.72115    10
-   15.35548    10
-\end{CodeOutput}
-\end{CodeChunk}
-
-Computing time for \pkg{rsample} for this problem is similar to direct
-computation of LOO CV and to \code{cv()} with \code{method="exact"}, but
-more than an order of magnitude slower than \code{cv()} with
-\code{method="Woodbury"}, and more than three orders of magnitude slower
-then \code{cv()} with \code{method="hatvalues"}.
+ 6981.899881    10
+ 7541.571242    10
+ 6081.412412    10
+ 5118.834994    10
+  116.574603    10
+    1.759433    10
+\end{CodeOutput}
+\end{CodeChunk}
+
+Computing time for \pkg{rsample} and \pkg{caret} for this problem is
+similar to direct computation of LOO CV and to \code{cv()} with
+\code{method="exact"}, but more than an order of magnitude slower than
+\code{cv()} with \code{method="Woodbury"}, and more than three orders of
+magnitude slower then \code{cv()} with \code{method="hatvalues"}.
 

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.Rmd
tweak JSS draft paper

@@
 ```
+In fairness, \pkg{caret} really isn't intended for this kind of simple CV computation.
 

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.tex
tweak JSS draft paper

@@
 \begin{CodeOutput}
-R RNG seed set to 799417
+R RNG seed set to 507470
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.20609
-bias-adjusted cross-validation criterion = 19.19442
-95% CI for bias-adjusted CV criterion = (15.73743, 22.6514)
+cross-validation criterion = 19.41715
+bias-adjusted cross-validation criterion = 19.39447
+95% CI for bias-adjusted CV criterion = (15.86881, 22.92014)
 full-sample criterion = 18.98477 
@@
 }
-<bytecode: 0x1118adc78>
+<bytecode: 0x14ad5f380>
 <environment: namespace:cv>
@@
 \end{CodeChunk}
+
+In fairness, \pkg{caret} really isn't intended for this kind of simple
+CV computation.
 

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.Rmd
simply use of caret in example in JSS draft paper

--- 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.Rmd
@@
 
-The \pkg{caret} package doesn't directly support generalized linear models fit by \code{glm()}, but it can be made to do so by defining a custom "method" for the binary logistic regression fit to the \code{attrition} data:
+The \pkg{caret} package supports generalized linear models fit by \code{glm()} and can straightforwardly perform LOO CV:
 ```{r rsample-example-using-caret, cache=TRUE}
 library("caret", quietly=TRUE, warn.conflicts=FALSE)
-
-glm_mod <- getModelInfo("glm", regex = FALSE)[[1]]
-glm_mod$fit <- function(x, y, wts, param, lev, last, classProbs, ...) {
-  x$Attrition <- y
-  glm(Attrition ~ JobSatisfaction + Gender + MonthlyIncome,
-      family = binomial, data = x)
-}
-
-mod <- train(x = attrition[, c("JobSatisfaction", "Gender",
+# reports complement of Bayes-rule criterion:
+train(x = attrition[, c("JobSatisfaction", "Gender",
                                "MonthlyIncome")],
              y = attrition$Attrition,
-             method = glm_mod,
+             method = "glm",
              trControl = trainControl(method="LOOCV"))
-mod # reports complement of Bayes-rule criterion
 ```
@@
               y = attrition$Attrition,
-              method = glm_mod,
+              method = "glm",
               trControl = trainControl(method="LOOCV")),

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.tex
simply use of caret in example in JSS draft paper

--- 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.tex
@@
 \begin{CodeOutput}
-R RNG seed set to 799417
+R RNG seed set to 45649
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.20609
-bias-adjusted cross-validation criterion = 19.19442
-95% CI for bias-adjusted CV criterion = (15.73743, 22.6514)
+cross-validation criterion = 19.13181
+bias-adjusted cross-validation criterion = 19.12404
+95% CI for bias-adjusted CV criterion = (15.69034, 22.55774)
 full-sample criterion = 18.98477 
@@
 }
-<bytecode: 0x1118adc78>
+<bytecode: 0x145d6f3a0>
 <environment: namespace:cv>
@@
 
-The \pkg{caret} package doesn't directly support generalized linear
-models fit by \code{glm()}, but it can be made to do so by defining a
-custom ``method'' for the binary logistic regression fit to the
-\code{attrition} data:
+The \pkg{caret} package supports generalized linear models fit by
+\code{glm()} and can straightforwardly perform LOO CV:
 
@@
 R> library("caret", quietly=TRUE, warn.conflicts=FALSE)
-R> 
-R> glm_mod <- getModelInfo("glm", regex = FALSE)[[1]]
-R> glm_mod$fit <- function(x, y, wts, param, lev, last, classProbs, ...) {
-+   x$Attrition <- y
-+   glm(Attrition ~ JobSatisfaction + Gender + MonthlyIncome,
-+       family = binomial, data = x)
-+ }
-R> 
-R> mod <- train(x = attrition[, c("JobSatisfaction", "Gender",
+R> # reports complement of Bayes-rule criterion:
+R> train(x = attrition[, c("JobSatisfaction", "Gender",
 +                                "MonthlyIncome")],
 +              y = attrition$Attrition,
-+              method = glm_mod,
++              method = "glm",
 +              trControl = trainControl(method="LOOCV"))
-R> mod # reports complement of Bayes-rule criterion
 \end{CodeInput}
@@
 +               y = attrition$Attrition,
-+               method = glm_mod,
++               method = "glm",
 +               trControl = trainControl(method="LOOCV")),
@@
          expr         min          lq        mean      median          uq
-      rsample 5853.035606 5948.291603 6163.101903 6121.042951 6144.223100
-        caret 6804.763276 6951.219335 7097.717763 7061.439656 7255.566640
-       direct 5310.127259 5350.848336 5554.513023 5484.144809 5690.480487
-     cv.exact 4751.498733 4797.980556 4879.481352 4870.303039 4927.941413
-      cv.wood  101.606159  103.891294  108.285899  108.577430  112.075263
- cv.hatvalues    1.206999    1.471613    1.514548    1.554679    1.614416
+      rsample 5741.137299 5867.248666 6155.761099 6095.937708 6218.031915
+        caret 6805.241869 6931.865408 7135.256404 7052.951569 7377.336681
+       direct 5262.786773 5312.612884 5455.994681 5388.428711 5475.565416
+     cv.exact 4740.642384 4771.993936 4870.961647 4847.941434 4981.140799
+      cv.wood  104.161361  106.036578  112.172855  113.107500  115.602534
+ cv.hatvalues    1.262554    1.462716    1.497295    1.520198    1.569685
          max neval
- 6981.899881    10
- 7541.571242    10
- 6081.412412    10
- 5118.834994    10
-  116.574603    10
-    1.759433    10
+ 7162.266671    10
+ 7648.302729    10
+ 6045.941344    10
+ 5054.087753    10
+  120.361527    10
+    1.666896    10
 \end{CodeOutput}

### 2024-06-17 scratch/notes_john/JSS-article/JSS-article-3.Rmd
update draft JSS paper for new models()

@@
 # 10-fold CV
-mlist <- do.call(models, mlist) # create "modList" object
+mlist <- models(mlist) # create "modList" object
 
@@
 The \code{models()} function takes an arbitrary number of regression models as its arguments, which are optionally named, to create a \code{"modList"} object.
-Because we generated the polynomial regression models in a named list, we conveniently employ \code{do.call()} to supply the models as arguments to \code{models()}.
-The names created for the list (e.g., \code{"m.2"}) are then used for the models.
-We can also invoke the \code{plot()} method for \code{"cvModList"} objects to compare the models (see Figure \ref{fig:polynomial-regression-CV-graph-2}):
+Alternatively, as here, we can supply a pre-constructed, optionally named, list of models as the single argument to \code{models()}.
+To visualize the results, comparing the models, we invoke the \code{plot()} method for the \code{"cvModList"} object returned by \code{cv()} (see Figure \ref{fig:polynomial-regression-CV-graph-2}):
 

### 2024-06-17 scratch/notes_john/JSS-article/JSS-article-3.tex
update draft JSS paper for new models()

@@
 \begin{CodeOutput}
-R RNG seed set to 45649
+R RNG seed set to 969978
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.13181
-bias-adjusted cross-validation criterion = 19.12404
-95% CI for bias-adjusted CV criterion = (15.69034, 22.55774)
+cross-validation criterion = 19.18252
+bias-adjusted cross-validation criterion = 19.17193
+95% CI for bias-adjusted CV criterion = (15.72359, 22.62027)
 full-sample criterion = 18.98477 
@@
       expr        min         lq       mean     median         uq        max
- hatvalues    984.369   1182.604   1188.963   1202.345   1220.242   1293.468
-  Woodbury  12366.543  12508.690  13408.099  12769.347  12815.124  16566.050
-     naive 216467.126 217748.581 225187.170 219534.685 222515.077 276207.898
-    cv.glm 373527.917 380937.683 393430.301 382341.584 389266.341 443234.190
+ hatvalues    996.956   1181.292   1496.971   1216.777   1233.157   4243.664
+  Woodbury  12189.054  12270.726  13187.437  12580.645  12664.121  16369.455
+     naive 216220.101 218010.571 231464.303 220980.816 223550.942 280828.393
+    cv.glm 381634.027 383670.456 404475.689 390533.979 441471.559 442930.011
  neval
@@
 R> # 10-fold CV
-R> mlist <- do.call(models, mlist) # create "modList" object
+R> mlist <- models(mlist) # create "modList" object
 R> 
@@
 models as its arguments, which are optionally named, to create a
-\code{"modList"} object. Because we generated the polynomial regression
-models in a named list, we conveniently employ \code{do.call()} to
-supply the models as arguments to \code{models()}. The names created for
-the list (e.g., \code{"m.2"}) are then used for the models. We can also
-invoke the \code{plot()} method for \code{"cvModList"} objects to
-compare the models (see Figure
+\code{"modList"} object. Alternatively, as here, we can supply a
+pre-constructed, optionally named, list of models as the single argument
+to \code{models()}. To visualize the results, comparing the models, we
+invoke the \code{plot()} method for the \code{"cvModList"} object
+returned by \code{cv()} (see Figure
 \ref{fig:polynomial-regression-CV-graph-2}):
@@
 Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv, :
-Model failed to converge with max|grad| = 0.00587207 (tol = 0.002, component 1)
+Model failed to converge with max|grad| = 0.00587131 (tol = 0.002, component 1)
 \end{CodeOutput}
@@
 }
-<bytecode: 0x145d6f3a0>
+<bytecode: 0x13b56c108>
 <environment: namespace:cv>
@@
 R> library("caret", quietly=TRUE, warn.conflicts=FALSE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Attaching package: 'ggplot2'
+\end{CodeOutput}
+\begin{CodeOutput}
+The following object is masked _by_ '.GlobalEnv':
+
+    mpg
+\end{CodeOutput}
+\begin{CodeOutput}
+The following object is masked from 'package:latticeExtra':
+
+    layer
+\end{CodeOutput}
+\begin{CodeInput}
 R> # reports complement of Bayes-rule criterion:
@@
 \begin{CodeOutput}
-Warning in microbenchmark::microbenchmark(rsample = {: less accurate nanosecond
-times to avoid potential integer overflows
-\end{CodeOutput}
-\begin{CodeOutput}
 Unit: milliseconds
          expr         min          lq        mean     median          uq
-      rsample 5741.137299 5867.248666 6155.761099 6095.937708 6218.031915
-        caret 6805.241869 6931.865408 7135.256404 7052.951569 7377.336681
-       direct 5262.786773 5312.612884 5455.994681 5388.428711 5475.565416
-     cv.exact 4740.642384 4771.993936 4870.961647 4847.941434 4981.140799
-      cv.wood  104.161361  106.036578  112.172855  113.107500  115.602534
- cv.hatvalues    1.262554    1.462716    1.497295    1.520198    1.569685
+      rsample 5892.991705 5997.040644 6197.075663 6180.82761 6298.898798
+        caret 6863.187210 6952.637853 7008.026442 7024.60940 7072.257280
+       direct 5483.414538 5488.340073 5603.901283 5564.68712 5655.604370
+     cv.exact 4829.717139 4885.941915 5049.808394 5021.14425 5155.600842
+      cv.wood  104.694812  110.656458  114.844686  117.06384  119.173552
+ cv.hatvalues    1.281824    1.513761    1.656785    1.59406    1.668454
          max neval
- 7162.266671    10
- 7648.302729    10
- 6045.941344    10
- 5054.087753    10
-  120.361527    10
-    1.666896    10
+ 6871.248999    10
+ 7095.635316    10
+ 5972.080828    10
+ 5486.199381    10
+  121.229702    10
+    2.391448    10
 \end{CodeOutput}

### 2024-06-28 scratch/notes_georges/vignettes_drafts/cv-mixed_modified.Rmd
draft cv_mixed

---
title: "Cross-validating mixed-effects models"
author: "John Fox and Georges Monette"
date: "`r Sys.Date()`"
package: cv
output: 
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["cv.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Cross-validating mixed-effects models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.align = "center",
  fig.height = 6,
  fig.width = 7,
  fig.path = "fig/",
  dev = "png",
  comment = "#>" #,
)
library(cv)
library(lme4)
# save some typing
knitr::set_alias(w = "fig.width",
                 h = "fig.height",
                 cap = "fig.cap")

# colorize text: use inline as `r colorize(text, color)`
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}

  # temporary fix

  GetResponse.glmmTMB <- function (model, ...)
  {
    y <- insight::get_response(model)
    if(is.matrix(y) && dim(y)[2] == 1) y <- as.vector(y)
    if (is.factor(y)) {
      levels <- levels(y)
      failure <- levels[1L]
      if (length(levels) > 2L) {
        message("Note: the response has more than 2 levels.\n",
                " The first level ('", failure, "') denotes failure (0),\n",
                " the others success (1)")
      }
      y <- as.numeric(y != failure)
    }
    if (!is.vector(y))
      stop("non-vector response")
    if (!is.numeric(y))
      stop("non-numeric response")
    y
  }
  assignInNamespace("GetResponse.glmmTMB", GetResponse.glmmTMB, ns = 'cv')

.opts <- options(digits = 5)
```

The fundamental analogy for cross-validation is to the collection of new data. That is, predicting the response in each fold from the model fit to data in the other folds is like using the model fit to all of the data to predict the response for new cases from the values of the predictors for those new cases. As we explained in the introductory vignette on cross-validating regression models, the application of this idea to independently sampled cases is straightforward---simply partition the data into random folds of equal size and leave each fold out in turn, or, in the case of LOO CV, simply omit each case in turn.

In contrast, mixed-effects models are fit to *dependent* data, in which cases as clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases repeated observations on the individuals over time.[^crossed-effects]

[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects. For example, consider students within classes within schools. In primary schools, students typically are in a single class, and so classes are nested within schools. In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed. The `lmer()` function in the **lme4** package is capable of modeling both nested and crossed random effects, and the `cv()` methods for mixed models in the **cv** package pertain to both nested and crossed random effects. We present an example of the latter later in the vignette.

We can think of two approaches to applying cross-validation to clustered data:[^cv-faq]

[^cv-faq]: We subsequently discovered that @Vehtari:2023 [Section 8] makes similar points.

1. Treat CV as analogous to predicting the response for one or more cases in a *newly observed cluster*. In this instance, the folds comprise one or more whole clusters; we refit the model with all of the cases in clusters in the current fold removed; and then we predict the response for the cases in clusters in the current fold. These predictions are based only on fixed effects because the random effects for the omitted clusters are presumably unknown, as they would be for data on cases in newly observed clusters.

2. Treat CV as analogous to predicting the response for a newly observed case in an *existing cluster*. In this instance, the folds comprise one or more individual cases, and the predictions can use both the fixed and random effects.

## Example: The High-School and Beyond data

Following their use by @RaudenbushBryk:2002, data from the 1982 *High School and Beyond* (HSB) survey have become a staple of the literature on mixed-effects models. The HSB data are used by @FoxWeisberg:2019 [Sec. 7.2.2] to illustrate the application of linear mixed models to hierarchical data, and we'll closely follow their example here.

The HSB data are included in the `MathAchieve` and `MathAchSchool` data sets in the **nlme** package  [@PinheiroBates:2000]. `MathAchieve` includes individual-level data on 7185  students in 160 high schools, and `MathAchSchool` includes school-level data:
```{r HSB-data}
data("MathAchieve", package = "nlme")
dim(MathAchieve)
head(MathAchieve, 3)
tail(MathAchieve, 3)

data("MathAchSchool", package = "nlme")
dim(MathAchSchool)
head(MathAchSchool, 2)
tail(MathAchSchool, 2)
```
The first few students are in school number 1224 and the last few in school 9586.

We'll use only the `School`, `SES` (students' socioeconomic status), and `MathAch` (their score on a standardized math-achievement test) variables in the `MathAchieve` data set, and `Sector` (`"Catholic"` or `"Public"`) in the `MathAchSchool` data set.

Some data-management is required before fitting a mixed-effects model to the HSB data, for which we use the **dplyr** package [@WickhamEtAl:2023]:
```{r HSB-data-management, cache=TRUE}
library("dplyr")
MathAchieve %>% group_by(School) %>%
  summarize(mean.ses = mean(SES)) -> Temp
Temp <- merge(MathAchSchool, Temp, by = "School")
HSB <- merge(Temp[, c("School", "Sector", "mean.ses")],
             MathAchieve[, c("School", "SES", "MathAch")], by = "School")
names(HSB) <- tolower(names(HSB))

HSB$cses <- with(HSB, ses - mean.ses)
```
In the process, we created two new school-level variables: `meanses`, which is the average SES for students in each school; and `cses`, which is school-average SES centered at its mean. For details, see @FoxWeisberg:2019 [Sec. 7.2.2].

Still following Fox and Weisberg, we proceed to use the `lmer()` function in the **lme4** package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
```{r HSB-lmer, cache=TRUE}
library("lme4")
hsb.lmer <- lmer(mathach ~ mean.ses * cses + sector * cses
                 + (cses | school), data = HSB)
summary(hsb.lmer, correlation = FALSE)
```

We can then cross-validate at the cluster (i.e., school) level,
```{r HSB-lmer-CV-cluster, cache=TRUE}
library("cv")

cv(hsb.lmer,
   k = 10,
   clusterVariables = "school",
   seed = 5240)
```
or at the case (i.e., student) level,
```{r HSB-lmer-CV-case, cache=TRUE}
cv(hsb.lmer, seed = 1575)
```
For cluster-level CV, the `clusterVariables` argument tells `cv()` how the clusters are defined. Were there more than one clustering variable, say classes within schools, these would be provided as a character vector of variable names: `clusterVariables = c("school", "class")`. For cluster-level CV, the default is `k = "loo"`, that is, leave one cluster out at a time; we instead specify `k = 10` folds of clusters, each fold therefore comprising $160/10 = 16$ schools.

If the `clusterVariables` argument is omitted, then case-level CV is employed, with `k = 10` folds as the default, here each with $7185/10 \approx 719$ students. Notice that one of the 10 models refit with a fold removed failed to converge. Convergence problems are common in mixed-effects modeling. The apparent issue here is that an estimated variance component is close to or equal to 0, which is at a boundary of the parameter space. That shouldn't disqualify the fitted model for the kind of prediction required for cross-validation.

There is also a `cv()` method for linear mixed models fit by the `lme()` function in the **nlme** package, and the arguments for `cv()` in this case are the same as for a model fit by `lmer()` or `glmer()`. We illustrate with the mixed model fit to the HSB data:
```{r hsb-lme, cache=TRUE}
library("nlme")
hsb.lme <- lme(
  mathach ~ mean.ses * cses + sector * cses,
  random = ~ cses | school,
  data = HSB,
  control = list(opt = "optim")
)
summary(hsb.lme)

cv(hsb.lme,
   k = 10,
   clusterVariables = "school",
   seed = 5240)

cv(hsb.lme, seed = 1575)
```
We used the same random-number generator seeds as in the previous example cross-validating the model fit by `lmer()`, and so the same folds are employed in both cases.[^optimizer] The estimated covariance components and fixed effects in the summary output differ slightly between the `lmer()` and `lme()` solutions, although both functions seek to maximize the REML criterion. This is, of course, to be expected when different algorithms are used for numerical optimization.  To the precision reported, the cluster-level CV results for the `lmer()` and `lme()` models are identical, while the case-level CV results are very similar but not identical.

[^optimizer]: The observant reader will notice that we set the argument `control=list(opt="optim")` in the call to `lme()`, changing the optimizer employed from the default `"nlminb"`. We did this because with the default optimizer, `lme()` encountered the same convergence issue as `lmer()`, but rather than issuing a warning, `lme()` failed, reporting an error. As it turns out, setting the optimizer to `"optim"` avoids this problem.

## Example: Contrived hierarchical data

We introduce four artificial data sets that exemplify aspects of cross-validation particular to hierarchical models. Using these data sets, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model. Furthermore, commonly used measures of fit, such as mean-squared error, do not necessarily become smaller as models become larger, even when the models are nested, and even when the measure of fit is computed for the whole data set.

The four datasets differ in the relative magnitude of between-cluster variance compared with within-cluster variance. They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which determines the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.

For these analyses, we will use the `glmmTMB()` function in the **glmmTMB** package [@BrooksEtAl:2017] because, in our experience, it is more likely to converge than functions in the **nlme** and the **lme4** packages for models with low between-cluster variance.

Consider a researcher studying the effect of the dosage of a drug on a measure of the severity of symptoms for some hypothetical disease. The researcher has longitudinal data on 20 patients, each of whom was observed on five occasions in which patients received different dosages of the drug.  The data are observational with dosages prescribed by the patients' physicians so that patients who were more severely affected by the disease received higher dosages of the drug.

A figure below shows the four simulated contrived data sets illustrating possible results for data obtained in such a scenario. The relative configuration of dosages, 'x', and symptoms, 'y', are identical within patients in each of the four datasets. Within patients, higher dosages are generally associated with a reduction in symptoms.

Between patients, however, higher dosages are associated with higher levels of symptoms. A plausible mechanism would be a reversal of causality: within patients, higher dosages may alleviate symptoms, but, between patients, higher morbidity may cause the prescription of higher dosages.

The four data sets differ in the between-patient variance of patient centroids from a regression line. They exhibit a progression from low to high variance around the regression line.  We start by generating a data set to serve as a common template for the four sample data sets.  We then apply different multipliers to the between-patient variation using parameters consistent with the description above o highlight the issues that arise in cross-validating mixed-effects models:[^1]

[^1]: We invite the interested reader to experiment with varying the parameters of our example.

```{r include=FALSE, echo=FALSE}
library(glmmTMB) # necessary for some reason to knit vignette in RStudio, harmless otherwise
```
```{r}

# Parameters:

Nb <- 20     # number of patients
Nw <- 5      # number of occasions for each patient

Bb <- 1.0    # between-patient regression coefficient on patient means
Bw <- -0.5   # within-patient effect of x

SD_between <- c(0,5,6,8)                  # SD between patients
SD_within <- rep(2.5, length(SD_between)) # SD within patients

Nv <- length(SD_within)       # number of variance profiles
SD_ratio <- paste0(SD_between,' / ',SD_within)
SD_ratio <- factor(SD_ratio, levels = SD_ratio)

set.seed(833885)

Data_template <- expand.grid(patient = 1:Nb, obs = 1:Nw) |>
  within({
    xw <- seq(-2,2, length.out = Nw)[obs]
    x <- patient + xw
    xm  <- ave(x, patient)   # within-patient mean

    # Scaled random error within each SD_ratio_i group

    re_std <- scale(resid(lm(rnorm(Nb*Nw)~ x)))
    re_between <- ave(re_std, patient)
    re_within <- re_std - re_between
    re_between <- scale(re_between)/sqrt(Nw)
    re_within <- scale(re_within)
  })

Data <- do.call(
  rbind,
  lapply(
    1:Nv,
    function(i) {
      cbind(Data_template, SD_ratio_i = i)
    }
  )
)

Data <- within(
  Data,
  {
    SD_within_ <- SD_within[SD_ratio_i]
    SD_between_ <- SD_between[SD_ratio_i]
    SD_ratio <- SD_ratio[SD_ratio_i]
    y <- 10 +
      Bb * xm +                  # contextual effect
      Bw * (x - xm) +            # within-patient effect
      SD_within_ * re_within +   # within patient random effect
      SD_between_ * re_between   # adjustment to between patient random effect
  }
)
```
Here is a scatterplot of the data sets showing estimated 50% concentration ellipses for each cluster:[^2]

[^2]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.

```{r plot1}
#| out.width = "100%",
#| fig.height = 4.5,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the between-patient SD/within-patient SD."
library("lattice")
library("latticeExtra")
plot <- xyplot(y ~ x | SD_ratio, data=Data, group=patient,
           layout = c(Nv,1),
           par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
      layer(panel.ellipse(..., center.pch = 16, center.cex = 1.5,  level = 0.5))
plot # display graph
```
The population from which these datasets are generated has a between-patient effect of dosage of 1 and a within-patient effect of -0.5. The researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models have sequentially nested fixed effects models: 1) intercept only, `~ 1`, 2) intercept and effect of 'x', `~ 1 + x` and 3) intercept, effect of 'x' and a contextual variable, 'xm', consisting of the within-patient mean of 'x', `~ 1 + x + xm`. The fourth model uses an intercept and the 'centered-within-group' variable 'x - xm', `~ 1 + I(x = xm)`. We thus fit four models to four datasets for a total of sixteen models.

```{r}
model.formulas <- c(
  intercept =  y ~ 1 + (1 | patient),
  x         =  y ~ 1 + x + (1 | patient),
  'x + xm'  =  y ~ 1 + x + xm + (1 | patient),
  '(x-xm)'  =  y ~ 1 + I(x - xm) + (1 | patient)
)

fits <- lapply(split(Data, ~ SD_ratio),
               function(d) {
                 lapply(model.formulas, function(form) {
                   glmmTMB(form, data = d)
                 })
               })
```
We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within patients):
```{r}
pred.BLUPs <- lapply(fits, lapply, predict)                # predicted fixed and random effects
pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  # predicted fixed effects
```
We then prepare the data for plotting:
```{r}
Dataf <- lapply(split(Data, ~ SD_ratio),
                    function(d) {
                      lapply(names(model.formulas), function(form) cbind(d, formula = form))
                    }) |> 
             lapply(function(dlist) do.call(rbind, dlist) ) |> 
             do.call(rbind, args = _)

Dataf <- within(
  Dataf,
  {
    pred.fixed <- unlist(pred.fixed)
    pred.BLUPs <- unlist(pred.BLUPs)
    formula <- factor(formula, levels = names(model.formulas))
  }
)
```

The fixed-effects predictions from these models are shown in the following graph:
```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed-effect predictions using each model applied to data sets with varying between- and within-patient variance ratios."
xyplot(pred.fixed  ~ x |SD_ratio * formula, Dataf, type = 'l', groups = patient,
       ylab = 'fixed-effect predictions') |> 
  useOuterStrips() |> print()
```
The BLUPs from these models are shown in the following graph:
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient variance ratios."
xyplot(pred.BLUPs  ~ x |SD_ratio * formula, Dataf, type = 'l', groups = patient,
       ylab = 'fixed- and random-effect predictions (BLUPS)') |> 
  useOuterStrips() |> print()
```
Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between 'y' and 'x' in the model with an intercept and 'x' as fixed-effect predictors, `~ 1 + x`. The inclusion of a contextual variable in the model corrects this problem.

Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform with the within-patient slopes for the `~ 1 + x` model in the two datasets with the smallest between-patient variances.

For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for datasets with larger between-subject variance.  With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.

We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, `~ 1`  to the larger `~ 1 + x` model when the between-cluster variance is relatively small, but that cluster-based cross-validation will prefer the latter to the former.

Comparing the '5 / 2.5' dataset with the '8 / 2.5' dataset, we will note that, with the former dataset, case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model but cluster-based cross-validation prefers the latter to the former. The situation is entirely reversed with the '8 / 2.5' dataset.

The third model, `~ 1 + x + xm`, includes a contextual effect of $x$---that is, the cluster mean `xm`---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept. This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | patient)`, which is the model that generated the data.  The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects only model with 'x' and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).

<!-- The  mixed model `mod.2` therefore fits individual observations well, and we anticipate a favorable assessment using individual-based cross-validation. In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for model `mod.2` compared to the smaller model `mod.0`, which includes only fixed and random intercepts. -->

<!-- With very small between-patient variance and, consequently, considerable shrinkage, then neither cluster-based nor case-based cross-validation would show much improvement over the random-intercept-only model. In our experience, the degree of shrinkage does not vary smoothly as parameters are changed but tends to be "all or nothing," and near the tipping point, the behavior of estimates can be affected considerably by the choice of algorithm used to fit the model. NOTE: This paragraph no longer applies to the current example in which the between-cluster slope is not zero-->

We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only. In order to reduce between-model random variability in comparisons of models on the same dataset, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.

<!-- The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean. Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different. -->
```{r echo=FALSE,include=FALSE}
library(cv) # unclear why it's necessary to reload cv
.opts <- options(warn = -2) 
```
```{r cross-validation}

model_lists <- lapply(fits, function(fitlist) do.call(models, fitlist))

cvs_cases <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]], k = 10, data = split(Data, ~ SD_ratio)[[i]])
         })

cvs_clusters <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]],  k = 10, data = split(Data, ~SD_ratio)[[i]], clusterVariables = 'patient')
         })
```
For a given dataset, we can plot the results for a list of models using the plot method for 'cvList'. For example
```{r}
#| out.width = "60%",
#| fig.height = 3,
#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept mixed models with varying fixed effects."
plot(cvs_clusters[[2]])
```
We assemble the results for all datasets to show them in a common figure.
```{r}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))

dsummary$cases <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
  })

dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(cases + clusters ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', reverse.rows = T, columns = 2), type = 'b',
       ylab = 'CV criterion (MSE)',
       layout= c(Nv,1),
       par.settings =
         list(superpose.line=list(lty = c(2, 3), lwd = 3),
              superpose.symbol=list(pch = 15:16, cex = 1.5)),
       scales = list(y = list(log = TRUE), x = list(alternating = F, rot = 60))) |> print()
```
In summary, when between-cluster variance is relatively large, model `~ 1 + x`, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. In all our examples, model `~ 1 + x + xm`, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

<!-- The data-generating model, `mod.3`, which includes the fixed effect of `x - xm` in place of separate terms in `x` and `xm`, isn't distinguishable from model `mod.2`, which includes `x` and `xm` separately, even though `mod.2` has an unnecessary parameter (recall that the population coefficient of `xm` is 0 when `x` is expressed as deviations from the contextual mean). -->

## Example: Crossed random effects

Crossed random effects arise when the structure of the data aren't strictly hierarchical. Nevertheless, crossed and nested random effects can be handled in much the same manner, by refitting the mixed-effects model to the data with a fold of clusters or cases removed and using the refitted model to predict the response in the removed fold.

We'll illustrate with data on pig growth, introduced by @DiggleLiangZeger:1994 [Table 3.1]. The data are in the `Pigs` data frame in the **cv** package:
```{r pigs}
head(Pigs, 9)
head(xtabs( ~ id + week, data = Pigs), 3)
tail(xtabs( ~ id + week, data = Pigs), 3)
```
Each of 48 pigs is observed weekly over a period of 9 weeks, with the weight of the pig recorded in kg. The data are in "long" format, as is appropriate for use with the `lmer()` function in the **lme4** package. The data are very regular, with no missing cases.

The following graph, showing the growth trajectories of the pigs, is similar to Figure 3.1 in @DiggleLiangZeger:1994; we add an overall least-squares line and a loess smooth, which are nearly indistinguishable:
```{r pigs-graph}
#| out.width = "60%",
#| fig.height = 6,
#| fig.cap = "Growth trajectories for 48 pigs, with overall least-squares line (sold blue) and loess line (broken magenta)."
plot(weight ~ week, data = Pigs, type = "n")
for (i in unique(Pigs$id)) {
  with(Pigs, lines(
    x = 1:9,
    y = Pigs[id == i, "weight"],
    col = "gray"
  ))
}
abline(lm(weight ~ week, data = Pigs),
       col = "blue",
       lwd = 2)
lines(
  with(Pigs, loess.smooth(week, weight, span = 0.5)),
  col = "magenta",
  lty = 2,
  lwd = 2
)
```
The individual "growth curves" and the overall trend are generally linear, with some tendency for variability of pig weight to increase over weeks (a feature of the data that we ignore in the mixed model that we fit to the data below).

The **Stata** mixed-effects models manual proposes a model with crossed random effects for the `Pigs` data [@Stata:2023 page 37]:

> [S]uppose that we wish to fit
$$
\mathrm{weight}_{ij} = \beta_0 + \beta_1 \mathrm{week}_{ij} + u_i + v_j + \varepsilon_{ij} 
$$
for the $i = 1, \ldots, 9$ weeks and $j = 1, \dots, 48$ pigs and
$$
u_i \sim N(0, \sigma^2_u); v_j \sim N(0, \sigma^2_v ); \varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon)
$$
all independently. That is, we assume an overall population-average growth curve $\beta_0 + \beta_1 \mathrm{week}$ and a random pig-specific shift. In other words, the effect due to week, $u_i$, is systematic to that week and  common to all pigs. The rationale behind [this model] could be that, assuming that the pigs were measured contemporaneously, we might be concerned that week-specific random factors such as weather and feeding patterns had significant systematic effects on all pigs.

Although we might prefer an alternative model,[^pig-alternative] we think that this is a reasonable specification.

[^pig-alternative]: These are repeated-measures data, which would be more conventionally modeled with autocorrelated errors within pigs. The `lme()` function in the **nlme** package, for example, is capable of fitting a mixed-model of this form.

The **Stata** manual fits the mixed model by maximum likelihood (rather than REML), and we duplicate the results reported there using `lmer()`:
```{r pigs-lmer}
m.p <- lmer(
  weight ~ week + (1 | id) + (1 | week),
  data = Pigs,
  REML = FALSE, # i.e., ML
  control = lmerControl(optimizer = "bobyqa")
)
summary(m.p)
```
We opt for the non-default `"bobyqa"` optimizer because it provides more numerically stable results for subsequent cross-validation in this example.

We can then cross-validate the model by omitting folds composed of pigs, folds composed of weeks, or folds composed of pig-weeks (which in the `Pigs` data set correspond to individual cases, using only the fixed effects):
```{r pigs-cv}
cv(m.p, clusterVariables = "id")

cv(m.p, clusterVariables = "week")

cv(
  m.p,
  clusterVariables = c("id", "week"),
  k = 10,
  seed = 8469
)
```
We can also cross-validate the individual cases taking account of the random effects (employing the same 10 folds):
```{r pigs-cv-cases}
cv(m.p, k = 10, seed = 8469)
```
Because these predictions are based on BLUPs, they are more accurate than the predictions based only on fixed effects.[^crossed-vs-nested] As well, the difference between the MSE computed for the model fit to the full data and the CV estimates of the MSE is greater here than for cluster-based predictions.

[^crossed-vs-nested]: Even though there is only one observation per combination of pigs and weeks, we can use the BLUP for the omitted case because of the crossed structure of the random effects; that is each pig-week has a pig random effect and a week random effect. Although it probably isn't sensible, we can imagine a mixed model for the pig data that employs nested random effects, which would be specified by `lmer(weight ~ week + (1 | id/week), data=Pigs)`---that is, a random intercept that varies by combinations of `id` (pig) and `week`. This model can't be fit, however: With only one case per combination of `id` and `week`, the nested random-effect variance is indistinguishable from the case-level variance.

```{r coda, include = FALSE}
options(.opts)
```

## References

### 2024-06-30 scratch/notes_georges/vignettes_drafts/cv-mixed_modified-edited.Rmd
cv-mixed with John's 2024-06-29 edits

--- 2024-06-28 scratch/notes_georges/vignettes_drafts/cv-mixed_modified.Rmd
@@
 
-Consider a researcher studying the effect of the dosage of a drug on a measure of the severity of symptoms for some hypothetical disease.
+Consider a researcher studying the effect of the dosage of a drug on the severity of symptoms for a hypothetical disease.
 The researcher has longitudinal data on 20 patients, each of whom was observed on five occasions in which patients received different dosages of the drug.
-The data are observational with dosages prescribed by the patients' physicians so that patients who were more severely affected by the disease received higher dosages of the drug.
+The data are observational, with dosages prescribed by the patients' physicians, so that patients who were more severely affected by the disease received higher dosages of the drug.
 
-A figure below shows the four simulated contrived data sets illustrating possible results for data obtained in such a scenario.
-The relative configuration of dosages, 'x', and symptoms, 'y', are identical within patients in each of the four datasets.
+Our four contrived data sets (see below) illustrate possible results for data obtained in such a scenario.
+The relative configuration of dosages `x` and symptoms `y` are identical within patients in each of the four datasets.
 Within patients, higher dosages are generally associated with a reduction in symptoms.
@@
 Between patients, however, higher dosages are associated with higher levels of symptoms.
-A plausible mechanism would be a reversal of causality: within patients, higher dosages may alleviate symptoms, but, between patients, higher morbidity may cause the prescription of higher dosages.
+A plausible mechanism is a reversal of causality: Within patients, higher dosages alleviate symptoms, but between patients higher morbidity causes the prescription of higher dosages.
 
-The four data sets differ in the between-patient variance of patient centroids from a regression line.
-They exhibit a progression from low to high variance around the regression line.
+The four data sets differ in the between-patient variance of patient centroids from a common between-patients regression line.
+The data sets exhibit a progression from low to high variance around the common regression line.
+
 We start by generating a data set to serve as a common template for the four sample data sets.
-We then apply different multipliers to the between-patient variation using parameters consistent with the description above o highlight the issues that arise in cross-validating mixed-effects models:[^1]
+We then apply different multipliers to the between-patient variation using parameters consistent with the description above to highlight the issues that arise in cross-validating mixed-effects models:[^data-template]
 
-[^1]: We invite the interested reader to experiment with varying the parameters of our example.
+[^data-template]: We invite the interested reader to experiment with varying the parameters of our example.
 
 ```{r include=FALSE, echo=FALSE}
-library(glmmTMB) # necessary for some reason to knit vignette in RStudio, harmless otherwise
+library("glmmTMB") # necessary for some reason to knit vignette in RStudio, harmless otherwise
 ```
@@
 
-SD_between <- c(0,5,6,8)                  # SD between patients
+SD_between <- c(0, 5, 6, 8)               # SD between patients
 SD_within <- rep(2.5, length(SD_between)) # SD within patients
@@
 Nv <- length(SD_within)       # number of variance profiles
-SD_ratio <- paste0(SD_between,' / ',SD_within)
+SD_ratio <- paste0('SD ratio = ', SD_between,' / ',SD_within)
 SD_ratio <- factor(SD_ratio, levels = SD_ratio)
@@
   within({
-    xw <- seq(-2,2, length.out = Nw)[obs]
+    xw <- seq(-2, 2, length.out = Nw)[obs]
     x <- patient + xw
@@
 
-    re_std <- scale(resid(lm(rnorm(Nb*Nw)~ x)))
+    re_std <- scale(resid(lm(rnorm(Nb*Nw) ~ x)))
     re_between <- ave(re_std, patient)
@@
 ```
-Here is a scatterplot of the data sets showing estimated 50% concentration ellipses for each cluster:[^2]
+Here is a scatterplot of the data sets showing estimated 50% concentration ellipses for each cluster:[^lattice]
 
-[^2]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.
+[^lattice]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.
 
@@
 #| fig.height = 4.5,
-#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the between-patient SD/within-patient SD."
+#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance.
+The labels above each panel show the between-patient SD/within-patient SD.
+The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel."
 library("lattice")
@@
 plot <- xyplot(y ~ x | SD_ratio, data=Data, group=patient,
-           layout = c(Nv,1),
+           layout = c(Nv, 1),
            par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
-      layer(panel.ellipse(..., center.pch = 16, center.cex = 1.5,  level = 0.5))
+      layer(panel.ellipse(..., center.pch = 16, center.cex = 1.5,  
+                          level = 0.5),
+            panel.abline(a=10, b=1))
 plot # display graph
 ```
-The population from which these datasets are generated has a between-patient effect of dosage of 1 and a within-patient effect of -0.5.
+The population from which these datasets is generated has a between-patient effect of dosage of 1 and a within-patient effect of $-0.5$.
 The researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept.
 We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.
-The first three models have sequentially nested fixed effects models: 1) intercept only, `~ 1`, 2) intercept and effect of 'x', `~ 1 + x` and 3) intercept, effect of 'x' and a contextual variable, 'xm', consisting of the within-patient mean of 'x', `~ 1 + x + xm`.
-The fourth model uses an intercept and the 'centered-within-group' variable 'x - xm', `~ 1 + I(x = xm)`.
-We thus fit four models to four datasets for a total of sixteen models.
+The first three models have sequentially nested fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`.
+The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`.
+We thus fit four models to four datasets for a total of 16 models.
 
@@
 model.formulas <- c(
-  intercept =  y ~ 1 + (1 | patient),
-  x         =  y ~ 1 + x + (1 | patient),
-  'x + xm'  =  y ~ 1 + x + xm + (1 | patient),
-  '(x-xm)'  =  y ~ 1 + I(x - xm) + (1 | patient)
+  ' ~ 1'             =  y ~ 1 + (1 | patient),
+  '~ 1 + x'          =  y ~ 1 + x + (1 | patient),
+  '~ 1 + x + xm'     =  y ~ 1 + x + xm + (1 | patient),
+  '~ 1 + I(x - xm)'  =  y ~ 1 + I(x - xm) + (1 | patient)
 )
@@
 ```
-We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within patients):
+We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or "BLUPs"---as would be used for cross-validation based on cases (i.e., occasions within patients):
 ```{r}
-pred.BLUPs <- lapply(fits, lapply, predict)                # predicted fixed and random effects
-pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  # predicted fixed effects
+# predicted fixed and random effects:
+pred.BLUPs <- lapply(fits, lapply, predict)
+# predicted fixed effects:
+pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  
 ```
@@
                     function(d) {
-                      lapply(names(model.formulas), function(form) cbind(d, formula = form))
+                      lapply(names(model.formulas), 
+                             function(form) cbind(d, formula = form))
                     }) |> 
@@
 #| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient variance ratios."
-xyplot(pred.BLUPs  ~ x |SD_ratio * formula, Dataf, type = 'l', groups = patient,
+xyplot(pred.BLUPs  ~ x |SD_ratio * formula, 
+       Dataf, type = 'l', groups = patient,
        ylab = 'fixed- and random-effect predictions (BLUPS)') |> 
@@
 ```
-Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between 'y' and 'x' in the model with an intercept and 'x' as fixed-effect predictors, `~ 1 + x`.
+Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`.
 The inclusion of a contextual variable in the model corrects this problem.
@@
 
-Comparing the '5 / 2.5' dataset with the '8 / 2.5' dataset, we will note that, with the former dataset, case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model but cluster-based cross-validation prefers the latter to the former.
+We will discover that case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model for the '5 / 2.5' dataset, but cluster-based cross-validation prefers the latter model to the former.
 The situation is entirely reversed with the '8 / 2.5' dataset.
 
-The third model, `~ 1 + x + xm`, includes a contextual effect of $x$---that is, the cluster mean `xm`---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept.
+The third model, `~ 1 + x + xm`, includes a contextual effect of `x`---that is, the cluster mean `xm`---along with `x` and the intercept in the fixed-effect part of the model, along with a random intercept.
 This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | patient)`, which is the model that generated the data.
-The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects only model with 'x' and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).
+The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects only model with `x` and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).
 
@@
          function(i){
-           cv(model_lists[[i]], k = 10, data = split(Data, ~ SD_ratio)[[i]])
+           cv(model_lists[[i]], k = 10, 
+              data = split(Data, ~ SD_ratio)[[i]])
          })
@@
          function(i){
-           cv(model_lists[[i]],  k = 10, data = split(Data, ~SD_ratio)[[i]], clusterVariables = 'patient')
+           cv(model_lists[[i]],  k = 10, 
+              data = split(Data, ~SD_ratio)[[i]], 
+              clusterVariables = 'patient')
          })
 ```
-For a given dataset, we can plot the results for a list of models using the plot method for 'cvList'. For example
+For a given dataset, we can plot the results for a list of models using the plot method for `"cvList"` objects. For example
 ```{r}
@@
 ```
-In summary, when between-cluster variance is relatively large, model `~ 1 + x`, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
-In all our examples, model `~ 1 + x + xm`, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV.
+In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
+In all our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.
 These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

### 2024-06-30 scratch/notes_john/cv-mixed_modified-edited.Rmd
tweak some plots in mixed-models vignette

--- 2024-06-30 scratch/notes_georges/vignettes_drafts/cv-mixed_modified-edited.Rmd
@@
 #| out.width = "60%",
-#| fig.height = 3,
+#| fig.height = 6,
 #| fig.cap = "10-fold cluster-based cross-validation comparing random intercept mixed models with varying fixed effects."
-plot(cvs_clusters[[2]])
+plot(cvs_clusters[[2]], main="Comparison of Fixed Effects")
 ```
@@
        auto.key = list(space = 'top', reverse.rows = T, columns = 2), type = 'b',
+       xlab = "Fixed Effects",
        ylab = 'CV criterion (MSE)',

### 2024-07-01 scratch/notes_georges/vignettes_drafts/cv-mixed_modified-edited.Rmd
Change temporary fix to GetResponse.glmmTMB to 2024-06-30 version

@@
   # temporary fix 
-
-  GetResponse.glmmTMB <- function (model, ...)
-  {
+GetResponse.glmmTMB <- function (model, ...){
    y <- insight::get_response(model)
-    if(is.matrix(y) && dim(y)[2] == 1) y <- as.vector(y)
-    if (is.factor(y)) {
-      levels <- levels(y)
-      failure <- levels[1L]
-      if (length(levels) > 2L) {
-        message("Note: the response has more than 2 levels.\n",
-                " The first level ('", failure, "') denotes failure (0),\n",
-                " the others success (1)")
-      }
-      y <- as.numeric(y != failure)
-    }
+   if (length(dim(y)) == 2 && dim(y)[2] == 1) y <- y[, 1]
    if (!is.vector(y))

### 2024-07-01 scratch/notes_georges/AUC/AUC_not_casewise-edited-1.Rmd
clean up scratch

--- 2024-01-01 scratch/notes_john/JSS-article/JSS-article.Rmd
@@
 ---
-documentclass: jss
-author:
-  - name: John Fox
-    orcid: 0000-0002-1196-8012
-    affiliation: 'McMaster University'
-    address: |
-      | Hamilton, Ontario, Canada
-    email: \email{jfox@mcmaster.ca}
-    url: https://www.john-fox.ca/
-  - name: Georges Monette
-    orcid: 0000-0000-0000-0000
-    affiliation: 'York University'
-    # To add another line, use \AND at the end of the previous one as above
-    address: |
-      | Toronto, Ontario, Canada
-title:
-  formatted: "\\pkg{cv}: An R Package for Cross-Validation of Regression Models"
-  # If you use tex in the formatted title, also supply version without
-  plain:     "cv: An R Package for Cross-Validation of Regression Models"
-  # For running headers, if needed
-  short:     "\\pkg{cv}: Cross-Validation"
-abstract: >
-  The abstract of the article.
-keywords:
-  # at least one keyword must be supplied
-  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
-  plain:     [cross-validation, regression analysis, model selection, R]
-preamble: >
-  \usepackage{amsmath}
-output: rticles::jss_article
-bibliography: ["cv.bib"]
+title: "Appendix: Why the complement of AUC isn't a casewise CV criterion"
+output: html_document
 ---
-
-```{r, setup, include=FALSE}
-options(prompt = 'R> ', continue = '+ ')
-```
-
-# Cross-validation
-
-Cross-validation (CV) is an essentially simple and intuitively reasonable approach to estimating the predictive accuracy of regression models.
-CV is developed in many standard sources on regression modeling and "machine learning"---we particularly recommend @JamesEtAl:2021 [Secs.
-5.1, 5.3]---and so we will describe the method only briefly here before taking up computational issues and some examples.
-See @ArlotCelisse:2010 for a wide-ranging, if technical, survey of cross-validation and related methods that emphasizes the statistical properties of CV.
-
-Validating research by replication on independently collected data is a common scientific norm.
-Emulating this process in a single study by data-division is less common: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data.
-Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, particularly in smaller samples, the results can vary substantially based on the random division of the data: See @Harrell:2015 [Sec.
-5.3] for this and other remarks about data-division and cross-validation.
-
-Cross-validation speaks to both of these issues.
-In CV, the data are randomly divided as equally as possible into several, say $k$, parts, called "folds." The statistical model is fit $k$ times, leaving each fold out in turn.
-Each fitted model is then used to predict the response variable for the cases in the omitted fold.
-A CV criterion or "cost" measure, such as the mean-squared error ("MSE") of prediction, is then computed using these predicted values.
-In the extreme $k = n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."
-
-Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of prediction error.
-The $n$ regression models are highly statistical dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has relatively large variance.
-In contrast, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) are somewhat biased but have smaller variance.
-It is also possible to correct $k$-fold CV for bias (see below).
-
-# Examples
-
-## Polynomial regression for the Auto data
-
-The data for this example are drawn from the **ISLR2** package for R, associated with @JamesEtAl:2021.
-The presentation here is close (though not identical) to that in the original source [@JamesEtAl:2021 Secs.
-5.1, 5.3], and it demonstrates the use of the `cv()` function in the **cv** package.[^boot]
-
-[^boot]: @JamesEtAl:2021 use the `cv.glm()` function in the **boot** package [@CantyRipley2022; @DavisonHinkley:1997].
-Despite its name, `cv.glm()` is an independent function and not a method of a `cv()` generic function.
-
-The `Auto` dataset contains information about 392 cars:
-
-```{r Auto}
-data("Auto", package="ISLR2")
-head(Auto)
-dim(Auto)
-```
-With the exception of `origin` (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see `help("Auto", package="ISLR2")`.
-
-We'll focus here on the relationship of `mpg` (miles per gallon) to `horsepower`, as displayed in the following scatterplot:
-
-```{r mpg-horsepower-scatterplot}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "`mpg` vs `horsepower` for the `Auto` data"
-plot(mpg ~ horsepower, data=Auto)
-```
-The relationship between the two variables is monotone, decreasing, and nonlinear.
-Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in the following figure:
-
-[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is probably not the best choice here.
-Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader).
-
-```{r mpg-horsepower-scatterplot-polynomials}
-#| out.width = "100%",
-#| fig.height = 5,
-#| fig.cap = "`mpg` vs `horsepower` for the `Auto` data"
-plot(mpg ~ horsepower, data=Auto)
-horsepower <- with(Auto, 
-                   seq(min(horsepower), max(horsepower), 
-                       length=1000))
-for (p in 1:5){
-  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
-  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
-  lines(horsepower, mpg, col=p + 1, lty=p, lwd=2)
-}
-legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=2,
-       title="Degree", inset=0.02)
-```
-The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).
-
-The following graph shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$.
-The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.
-
-```{r mpg-horsepower-MSE-se}
-#| out.width = "100%",
-#| fig.height = 5,
-#| fig.cap = "Estimated squared error as a function of polynomial degree, $p$"
-library("cv") # for mse() and other functions
-
-var <- mse <- numeric(10)
-for (p in 1:10){
-  m <- lm(mpg ~ poly(horsepower, p), data=Auto)
-  mse[p] <- mse(Auto$mpg, fitted(m))
-  var[p] <- summary(m)$sigma^2
-}
-
-plot(c(1, 10), range(mse, var), type="n",
-     xlab="Degree of polynomial, p",
-     ylab="Estimated Squared Error")
-lines(1:10, mse, lwd=2, lty=1, col=2, pch=16, type="b")
-lines(1:10, var, lwd=2, lty=2, col=3, pch=17, type="b")
-legend("topright", inset=0.02,
-       legend=c(expression(hat(sigma)^2), "MSE"),
-       lwd=2, lty=2:1, col=3:2, pch=17:16)
-```
-The code for this graph uses the `mse()` function from the **cv** package to compute the MSE for each fit.
-
-### Using cv()
-
-The generic `cv()` function has an `"lm"` method, which by default performs $k = 10$-fold CV:
-```{r cv-lm-1}
-m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
-summary(m.auto)
-
-cv(m.auto)
-```
-The `"lm"` method by default uses `mse()` as the CV criterion and the Woodbury matrix identity to update the regression with each fold deleted without having literally to refit the model.
-Computational details are discussed in the final section of this vignette.
-The function reports the CV estimate of MSE, a biased-adjusted estimate of the MSE (the bias adjustment is explained in the final section), and the MSE is also computed for the original, full-sample regression.
-Because the division of the data into 10 folds is random, `cv()` explicitly (randomly) generates and saves a seed for R's pseudo-random number generator, to make the results replicable.
-The user can also specify the seed directly via the `seed` argument to `cv()`.
-
-To perform LOO CV, we can set the `k` argument to `cv()` to the number of cases in the data, here `k=392`, or, more conveniently, to `k="loo"` or `k="n"`:
-
-```{r cv.lm-2`}
-cv(m.auto, k="loo")
-```
-For LOO CV of a linear model, `cv()` by default uses the hatvalues from the model fit to the full data for the LOO updates, and reports only the CV estimate of MSE.
-Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted.
-All three methods produce exact results for a linear model (within the precision of floating-point computations):
-```{r cv.lm-3}
-cv(m.auto, k="loo", method="naive")
-
-cv(m.auto, k="loo", method="Woodbury")
-```
-The `"naive"` and `"Woodbury"` methods also return the bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an issue for LOO CV.
-
-This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed.
-In this comparison, we include the `cv.glm()` function from the **boot** package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM.
-We use the `microbenchmark()` function from the package of the same name for the timings [@Mersmann:2023]:
-```{r cv.lm.timings, cache=TRUE}
-m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
-boot::cv.glm(Auto, m.auto.glm)$delta
-
-microbenchmark::microbenchmark(
-  hatvalues = cv(m.auto, k="loo"),
-  Woodbury = cv(m.auto, k="loo", method="Woodbury"),
-  naive = cv(m.auto, k="loo", method="naive"),
-  cv.glm = boot::cv.glm(Auto, m.auto.glm),
-  times=10
-)
-```
-On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.[^microbenchmark]
-
-[^microbenchmark]: Out of impatience, we asked `microbenchmark()` to execute each command only 10 times rather than the default 100.
-With the exception of the last columns, the output is self-explanatory.
-The last column shows which methods have average timings that are statistically distinguishable.
-Because of the small number of repetitions (i.e., 10), the `"hatvalues"` and `"Woodbury"` methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default `times=100` repetitions.
-
-### Comparing competing models
-
-The `cv()` function also has a method that can be applied to a list of regression models for the same data, composed using the `models()` function.
-For $k$-fold CV, the same folds are used for the competing models, which reduces random error in their comparison.
-This result can also be obtained by specifying a common seed for R's random-number generator while applying `cv()` separately to each model, but employing a list of models is more convenient for both $k$-fold and LOO CV (where there is no random component to the composition of the $n$ folds).
-
-We illustrate with the polynomial regression models of varying degree for the `Auto` data (discussed previously), beginning by fitting and saving the 10 models:
-```{r polyomial-models}
-for (p in 1:10){
-  assign(paste0("m.", p),
-         lm(mpg ~ poly(horsepower, p), data=Auto))
-}
-objects(pattern="m\\.[0-9]")
-summary(m.2) # for example, the quadratic fit
-```
-We then apply `cv()` to the list of 10 models (the `data` argument is required):
-```{r polynomial-regression-CV}
-# 10-fold CV
-cv.auto.10 <- cv(models(m.1, m.2, m.3, m.4, m.5,
-                     m.6, m.7, m.8, m.9, m.10),
-              data=Auto, seed=2120)
-cv.auto.10[1:2] # for the linear and quadratic models
-
-# LOO CV
-cv.auto.loo <- cv(models(m.1, m.2, m.3, m.4, m.5,
-                        m.6, m.7, m.8, m.9, m.10),
-                 data=Auto, k="loo")
-cv.auto.loo[1:2] # linear and quadratic models
-```
-Because we didn't supply names for the models in the calls to the `models()` function, the names `model.1`, `model.2`, etc., are generated by the function.
-
-Finally, we extract and graph the adjusted MSEs for $10$-fold CV and the MSEs for LOO CV:
-```{r polynomial-regression-CV-graph}
-#| out.width = "100%",
-#| fig.height = 5,
-#| fig.cap = "Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$"
-cv.mse.10 <- sapply(cv.auto.10, function(x) x[["adj CV crit"]])
-cv.mse.loo <- sapply(cv.auto.loo, function(x) x[["CV crit"]])
-plot(c(1, 10), range(cv.mse.10, cv.mse.loo), type="n",
-     xlab="Degree of polynomial, p",
-     ylab="Cross-Validated MSE")
-lines(1:10, cv.mse.10, lwd=2, lty=1, col=2, pch=16, type="b")
-lines(1:10, cv.mse.loo, lwd=2, lty=2, col=3, pch=17, type="b")
-legend("topright", inset=0.02,
-       legend=c("10-Fold CV", "LOO CV"),
-       lwd=2, lty=2:1, col=3:2, pch=17:16)
-```
-
-Alternatively, we can use the `plot()` method for `"cvModList"` objects to compare the models, though with separate graphs for 10-fold and LOO CV:
-
-```{r, polynomial-regression-CV-graph-2, fig.show="hold"}
-#| out.width = "45%",
-#| fig.height = 6,
-#| fig.cap = "Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$"
-plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",
-     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
-plot(cv.auto.loo, main="Polynomial Regressions, LOO CV",
-     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
-```
-
-In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model, reported above (except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting).
-
-## Logistic regression for the Mroz data
-
-The `Mroz` data set from the **carData** package [associated with @FoxWeisberg:2019] has been used by several authors to illustrate binary logistic regression; see, in particular @FoxWeisberg:2019.
-The data were originally drawn from the U.S.
-Panel Study of Income Dynamics and pertain to married women.
-Here are a few cases in the data set:
-```{r Mroz-data}
-data("Mroz", package="carData")
-head(Mroz, 3)
-tail(Mroz, 3)
-```
-The response variable in the logistic regression is `lfp`, labor-force participation, a factor coded `"yes"` or `"no"`. The remaining variables are predictors:
-
-* `k5`, number of children 5 years old of younger in the woman's household;
-* `k618`, number of children between 6 and 18 years old;
-* `age`, in years;
-* `wc`, wife's college attendance, `"yes"` or `"no"`;
-* `hc`, husband's college attendance;
-* `lwg`, the woman's log wage rate if she is employed, or her *imputed* wage rate, if she is not [a variable that @FoxWeisberg:2019 show is problematically defined]; and
-* `inc`, family income, in $1000s, exclusive of wife's income.
-
-We use the `glm()` function to fit a binary logistic regression to the `Mroz` data:
-```{r Mroz-logistic-regresion}
-m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
-summary(m.mroz)
-
-BayesRule(ifelse(Mroz$lfp == "yes", 1, 0), 
-          fitted(m.mroz, type="response"))
-```
-In addition to the usually summary output for a GLM, we show the result of applying the `BayesRule()` function from the **cv** package to predictions derived from the fitted model.
-Bayes rule, which predicts a "success" in a binary regression model when the fitted probability of success [i.e., $\phi = \Pr(y = 1)$] is $\widehat{\phi} \ge .5$ and a "failure" if $\widehat{\phi} < .5$.[^BayesRule] The first argument to `BayesRule()` is the binary {0, 1} response, and the second argument is the predicted probability of success.
-`BayesRule()` returns the proportion of predictions that are *in error*, as appropriate for a "cost" function.
-
-[^BayesRule]: `BayesRule()` does some error checking; `BayesRule2()` is similar, but omits the error checking, and so can be faster for large problems.
-
-The value returned by `BayesRule()` is associated with an "attribute" named `"casewise loss"` and set to `"y != round(yhat)"`, signifying that the Bayes rule CV criterion is computed as the mean of casewise values, here 0 if the prediction for a case matches the observed value and 1 if it does not (signifying a prediction error).
-The `mse()` function for numeric responses is also calculated as a casewise average.
-Some other criteria, such as the median absolute error, computed by the `medAbsErr()` function in the **cv** package, aren't averages of casewise components.
-The distinction is important because, to our knowledge, the statistical theory of cross-validation, for example, in @DavisonHinkley:1997, @BatesHastieTibshirani:2023, and @ArlotCelisse:2010, is developed for CV criteria like MSE that are means of casewise components.
-As a consequence, we limit computation of bias adjustment and confidence intervals (see below) to criteria that are casewise averages.
-
-In this example, the fitted logistic regression incorrectly predicts 31% of the responses; we expect this estimate to be optimistic given that the model is used to "predict" the data to which it is fit.
-
-The `"glm"` method for `cv()` is largely similar to the `"lm"` method, although the default algorithm, selected explicitly by `method="exact"`, refits the model with each fold removed (and is thus equivalent to `method="naive"` for `"lm"` models).
-For generalized linear models, `method="Woodbury"` or (for LOO CV) `method="hatvalues"` provide approximate results (see the last section of the vignette for details):
-
-```{r cv-Mroz-10-fold}
-cv(m.mroz, criterion=BayesRule, seed=248)
-
-cv(m.mroz, criterion=BayesRule, seed=248, method="Woodbury")
-```
-To ensure that the two methods use the same 10 folds, we specify the seed for R's random-number generator explicitly; here, and as is common in our experience, the `"exact"` and `"Woodbury"` algorithms produce nearly identical results.
-The CV estimates of prediction error are slightly higher than the estimate based on all of the cases.
-
-The printed output includes a 95% confidence interval for the bias-adjusted Bayes rule CV criterion.
-@BatesHastieTibshirani:2023 show that these confidence intervals are unreliable for models fit to small samples, and by default `cv()` computes them only when the sample size is 400 or larger and when the CV criterion employed is an average of casewise components, as is the case for Bayes rule.
-See the final section of the vignette for details of the computation of confidence intervals for bias-adjusted CV criteria.
-
-Here are results of applying LOO CV to the Mroz model, using both the exact and the approximate methods:
-```{r cv-Mroz-LOO}
-cv(m.mroz, k="loo", criterion=BayesRule)
-
-cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury")
-
-cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues")
-```
-To the number of decimal digits shown, the three methods produce identical results for this example.
-
-As for linear models, we report some timings for the various `cv()` methods of computation in LOO CV as well as for the `cv.glm()` function from the **boot** package (which, recall, refits the model with each case removed, and thus is comparable to `cv()` with `method="exact"`):
-```{r glm.timings, cache=TRUE}
-microbenchmark::microbenchmark(
-  hatvalues=cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues"),
-  Woodbury=cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury"),
-  exact=cv(m.mroz, k="loo", criterion=BayesRule),
-  cv.glm=boot::cv.glm(Mroz, m.mroz,
-               cost=BayesRule),
-  times=10)
-```
-There is a substantial time penalty associated with exact computations.
-
-# Cross-validating mixed-effects models
-
-The fundamental analogy for cross-validation is to the collection of new data.
-That is, predicting the response in each fold from the model fit to data in the other folds is like using the model fit to all of the data to predict the response for new cases from the values of the predictors for those new cases.
-As we explained, the application of this idea to independently sampled cases is straightforward---simply partition the data into random folds of equal size and leave each fold out in turn, or, in the case of LOO CV, simply omit each case in turn.
-
-In contrast, mixed-effects models are fit to *dependent* data, in which cases as clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases repeated observations on the individuals over time.[^crossed-effects]
-
-[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects.
-For example, consider students within classes within schools.
-In primary schools, students typically are in a single class, and so classes are nested within schools.
-In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed.
-The `lmer()` function in the **lme4** package is capable of modeling both nested and crossed random effects, and the `cv()` methods for mixed models in the **cv** package pertain to both nested and crossed random effects.
-We present an example of the latter later in the vignette.
-
-We can think of two approaches to applying cross-validation to clustered data:[^cv-faq]
-
-[^cv-faq]: We subsequently discovered that @Vehtari:2023 [Section 8] makes similar points.
-
-1.
-Treat CV as analogous to predicting the response for one or more cases in a *newly observed cluster*.
-In this instance, the folds comprise one or more whole clusters; we refit the model with all of the cases in clusters in the current fold removed; and then we predict the response for the cases in clusters in the current fold.
-These predictions are based only on fixed effects because the random effects for the omitted clusters are presumably unknown, as they would be for data on cases in newly observed clusters.
-
-2.
-Treat CV as analogous to predicting the response for a newly observed case in an *existing cluster*.
-In this instance, the folds comprise one or more individual cases, and the predictions can use both the fixed and random effects.
-
-## Example: The High-School and Beyond data
-
-Following their use by @RaudenbushBryk:2002, data from the 1982 *High School and Beyond* (HSB) survey have become a staple of the literature on mixed-effects models.
-The HSB data are used by @FoxWeisberg:2019 [Sec.
-7.2.2] to illustrate the application of linear mixed models to hierarchical data, and we'll closely follow their example here.
-
-The HSB data are included in the `MathAchieve` and `MathAchSchool` data sets in the **nlme** package  [@PinheiroBates:2000].
-`MathAchieve` includes individual-level data on 7185  students in 160 high schools, and `MathAchSchool` includes school-level data:
-```{r HSB-data}
-data("MathAchieve", package="nlme")
-dim(MathAchieve)
-head(MathAchieve, 3)
-tail(MathAchieve, 3)
-
-data("MathAchSchool", package="nlme")
-dim(MathAchSchool)
-head(MathAchSchool, 2)
-tail(MathAchSchool, 2)
-```
-The first few students are in school number 1224 and the last few in school 9586.
-
-We'll use only the `School`, `SES` (students' socioeconomic status), and `MathAch` (their score on a standardized math-achievement test) variables in the `MathAchieve` data set, and `Sector` (`"Catholic"` or `"Public"`) in the `MathAchSchool` data set.
-
-Some data-management is required before fitting a mixed-effects model to the HSB data, for which we use the **dplyr** package [@WickhamEtAl:2023]:
-```{r HSB-data-management, cache=TRUE}
-library("dplyr")
-MathAchieve %>% group_by(School) %>%
-  summarize(mean.ses = mean(SES)) -> Temp
-Temp <- merge(MathAchSchool, Temp, by="School")
-HSB <- merge(Temp[, c("School", "Sector", "mean.ses")],
-             MathAchieve[, c("School", "SES", "MathAch")], by="School")
-names(HSB) <- tolower(names(HSB))
-
-HSB$cses <- with(HSB, ses - mean.ses)
-```
-In the process, we created two new school-level variables: `meanses`, which is the average SES for students in each school; and `cses`, which is school-average SES centered at its mean.
-For details, see @FoxWeisberg:2019 [Sec.
-7.2.2].
-
-Still following Fox and Weisberg, we proceed to use the `lmer()` function in the **lme4** package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
-```{r HSB-lmer, cache=TRUE}
-library("lme4")
-hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
-                   + (cses | school), data=HSB)
-summary(hsb.lmer, correlation=FALSE)
-```
-
-We can then cross-validate at the cluster (i.e., school) level,
-```{r HSB-lmer-CV-cluster, cache=TRUE}
-cv(hsb.lmer, k=10, clusterVariables="school", seed=5240)
-```
-or at the case (i.e., student) level,
-```{r HSB-lmer-CV-case, cache=TRUE}
-cv(hsb.lmer, seed=1575)
-```
-For cluster-level CV, the `clusterVariables` argument tells `cv()` how the clusters are defined.
-Were there more than one clustering variable, say classes within schools, these would be provided as a character vector of variable names: `clusterVariables = c("school", "class")`.
-For cluster-level CV, the default is `k = "loo"`, that is, leave one cluster out at a time; we instead specify `k = 10` folds of clusters, each fold therefore comprising $160/10 = 16$ schools.
-
-If the `clusterVariables` argument is omitted, then case-level CV is employed, with `k = 10` folds as the default, here each with $7185/10 \approx 719$ students.
-Notice that one of the 10 models refit with a fold removed failed to converge.
-Convergence problems are common in mixed-effects modeling.
-The apparent issue here is that an estimated variance component is close to or equal to 0, which is at a boundary of the parameter space.
-That shouldn't disqualify the fitted model for the kind of prediction required for cross-validation.
-
-There is also a `cv()` method for linear mixed models fit by the `lme()` function in the **nlme** package, and the arguments for `cv()` in this case are the same as for a model fit by `lmer()` or `glmer()`.
-We illustrate with the mixed model fit to the HSB data:
-```{r hsb-lme, cache=TRUE}
-library(nlme)
-hsb.lme <- lme(mathach ~ mean.ses*cses + sector*cses,
-                 random = ~ cses | school, data=HSB,
-               control=list(opt="optim"))
-summary(hsb.lme)
-
-cv(hsb.lme, k=10, clusterVariables="school", seed=5240)
-
-cv(hsb.lme, seed=1575)
-```
-We used the same random-number generator seeds as in the previous example cross-validating the model fit by `lmer()`, and so the same folds are employed in both cases.[^optimizer] The estimated covariance components and fixed effects in the summary output differ slightly between the `lmer()` and `lme()` solutions, although both functions seek to maximize the REML criterion.
-This is, of course, to be expected when different algorithms are used for numerical optimization.
-To the precision reported, the cluster-level CV results for the `lmer()` and `lme()` models are identical, while the case-level CV results are very similar but not identical.
-
-[^optimizer]: The observant reader will notice that we set the argument `control=list(opt="optim")` in the call to `lme()`, changing the optimizer employed from the default `"nlminb"`.
-We did this because with the default optimizer, `lme()` encountered the same convergence issue as `lmer()`, but rather than issuing a warning, `lme()` failed, reporting an error.
-As it turns out, setting the optimizer to `"optim"` avoids this problem.
-
-## Example: Contrived hierarchical data
-
-We introduce an artificial data set that exemplifies aspects of cross-validation particular to hierarchical models.
-Using this data set, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model.
-Furthermore, commonly used measures of fit, such as mean-squared error, do not necessarily become smaller as models become larger, even when the models are nested, and even when the measure of fit is computed for the whole data set.
-
-Consider a researcher studying improvement in a skill, yodeling, for example, among students enrolled in a four-year yodeling program.
-The plan is to measure each student's skill level at the beginning of the program and every year thereafter until the end of the program, resulting in five annual measurements for each student.
-It turns out that yodeling appeals to students of all ages, and students enrolling in the program range in age from 20 to 70.
-Moreover, participants' untrained yodeling skill is similar at all ages, as is their rate of progress with training.
-All students complete the four-year program.
-
-The researcher, who has more expertise in yodeling than in modeling, decides to model the response, $y$, yodeling skill, as a function of age, $x$, reasoning that students get older during their stay in the program, and (incorrectly) that age can serve as a proxy for elapsed time.
-The researcher knows that a mixed model should be used to account for clustering due to the expected similarity of measurements taken from each student.
-
-We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^1]
-
-[^1]: We invite the interested reader to experiment with varying the parameters of our example.
-
-```{r data}
-# Parameters:
-set.seed(9693) 
-Nb <- 100     # number of groups
-Nw <- 5       # number of individuals within groups
-Bb <- 0       # between-group regression coefficient on group mean
-SDre <- 2.0   # between-group SD of random level relative to group mean of x
-SDwithin <- 0.5  # within group SD
-Bw <- 1          # within group effect of x
-Ay <- 10         # intercept for response
-Ax <- 20         # starting level of x
-Nx <- Nw*10      # number of distinct x values
-
-Data <- data.frame(
-  group = factor(rep(1:Nb, each=Nw)),
-  x = Ax + rep(1:Nx, length.out = Nw*Nb)
-) |>
-  within(
-    {
-      xm  <- ave(x, group, FUN = mean) # within-group mean
-      y <- Ay +
-        Bb * xm +                    # contextual effect
-        Bw * (x - xm) +              # within-group effect
-        rnorm(Nb, sd=SDre)[group] +  # random level by group
-        rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
-    }
-  )
-```
-
-Here is a scatterplot of the data for a representative group of 10 (without loss of generality, the first 10) of 100 students, showing the 95% concentration ellipse for each cluster:[^2]
-
-[^2]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.
-
-```{r plot1}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Hierarchical data set, showing the first 10 of 100 students."
-library("lattice")
-library("latticeExtra")
-plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
-               ylim=c(4, 16),
-               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
-    layer(panel.ellipse(..., center.cex=0))
-plot # display graph
-```
-
-The between-student effect of age is 0 but the within-student effect is 1.
-Due to the large variation in ages between students, the least-squares regression of yodeling skill on age (for the 500 observations among all 100 students) produces an estimated slope close to 0 (though with a small $p$-value), because the slope is heavily weighted toward the between-student effect:
-
-```{r}
-summary(lm(y ~ x, data=Data))
-```
-
-The initial mixed-effects model that we fit to the data is a simple random-intercepts model:
-```{r include=FALSE, echo=FALSE}
-library(lme4) # necessary for some reason to knit vignette in RStudio, harmless otherwise
-```
-```{r}
-# random intercept only:
-mod.0 <- lmer(y ~ 1 + (1 | group), Data)
-summary(mod.0)
-```
-
-We will shortly consider three other, more complex, mixed models; because of data-management considerations, it is convenient to fit them now, but we defer discussion of these models:
-
-```{r}
-# effect of x and random intercept:
-mod.1 <- lmer(y ~ x + (1 | group), Data)
-
-# effect of x, contextual (student) mean of x, and random intercept:
-mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
-        # equivalent to y ~ I(x - xm) + xm + (1 | group)
-
-# model generating the data (where Bb = 0)
-mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
-```
-
-We proceed to obtain predictions from the random-intercept model (`mod.0`) and the other models (`mod.1`, `mod.2`, and `mod.3`) based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students):
-
-```{r}
-Data <- within(Data, {
-  fit_mod0.fe <- predict(mod.0, re.form = ~ 0) # fixed effects only
-  fit_mod0.re <- predict(mod.0) # fixed and random effects (BLUPs)
-  fit_mod1.fe <- predict(mod.1, re.form = ~ 0)
-  fit_mod1.re <- predict(mod.1)
-  fit_mod2.fe <- predict(mod.2, re.form = ~ 0)
-  fit_mod2.re <- predict(mod.2)
-  fit_mod3.fe <- predict(mod.3, re.form = ~ 0)
-  fit_mod3.re <- predict(mod.3)
-})
-```
-
-We then prepare the data for plotting:
-
-```{r}
-Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
-              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
-Data_long$id <- 1:nrow(Data_long)
-Data_long <- reshape(Data_long, direction = "long", sep = "_", 
-              timevar = "modelcode",  varying = grep("_", names(Data_long)))
-Data_long$model <- factor(
-  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
-  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
-)
-```
-
-Predictions based on the random-intercept model `mod.0` for the first 10 students are shown in the following graph:
-
-```{r plot-fits-mod0}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Predictions from the random intercept model."
-(plot +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
-         groups=group, type="l", lwd=2) +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
-         groups=group, type="l", lwd=2, lty=3)
-) |> update(
-  main="Model: y ~ 1 + (1 | group)",
-  key=list(
-    corner=c(0.05, 0.05),
-    text=list(c("fixed effects only","fixed and random")),
-    lines=list(lty=c(1, 3))))
-```
-
-The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean.
-Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.
-
-Our next model, `mod.1`, includes a fixed intercept and fixed effect of `x` along with a random intercept:
-
-```{r}
-summary(mod.1)
-```
-
-Predictions from this model appear in the following graph:
-
-```{r plot-fits-mod1}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Predictions from the model with random intercepts and $x$ as a fixed-effect predictor."
-(plot +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
-         groups=group, type="l", lwd=2) +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
-         groups=group, type="l", lwd=2, lty=3)
-) |> update(
-  main="Model: y ~ 1 + x + (1 | group)",
-  ylim=c(-15, 35),
-  key=list(
-    corner=c(0.95, 0.05),
-    text=list(c("fixed effects only","fixed and random")),
-    lines=list(lty=c(1, 3))))
-```
-
-The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, `mod.0`.
-We therefore anticipate (and show later in this section) that case-based cross-validation will prefer `mod1` to `mod0`, but that cluster-based cross-validation will prefer `mod0` to `mod1`.
-
-Our third model, `mod.2`, includes the contextual effect of $x$---that is, the cluster mean `xm`---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:
-
-```{r}
-summary(mod.2)
-```
-
-This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | group)`, which is the model that generated the data once the coefficient of the contextual predictor `xm` is set to 0 (as it is in `mod.3`, discussed below).
-
-Predictions from model `mod.2` appear in the following graph:
-```{r plot-fits-mod2}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Predictors from the model with random intercepts, $x$, and the group (student) mean of $x$ as predictors."
-(plot +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
-         groups=group, type="l", lwd=2) +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
-         groups=group, type="l", lwd=2, lty=3)
-) |> update(
-  main="Model: y ~ 1 + x + xm + (1 | group)",
-  ylim=c(4, 16),
-  key=list(
-    corner=c(0.05, 0.05),
-    text=list(c("fixed effects only","fixed and random")),
-    lines=list(lty=c(1, 3))))
-```
-
-Depending on the estimated variance parameters of the model, a mixed model like `mod.2` will apply varying degrees of shrinkage to the random-intercept BLUPs that correspond to variation in the heights of the parallel fitted lines for the individual students.
-In our contrived data, the `mod.2` applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student.
-The fit of the mixed model `mod.2` is consequently similar to that of a fixed-effects model with age and a categorical predictor for individual students (i.e., treating students as a factor, and not shown here).
-
-The  mixed model `mod.2` therefore fits individual observations well, and we anticipate a favorable assessment using individual-based cross-validation.
-In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for model `mod.2` compared to the smaller model `mod.0`, which includes only fixed and random intercepts.
-
-Had the mixed model applied considerable shrinkage, then neither cluster-based nor case-based cross-validation would show much improvement over the random-intercept-only model.
-In our experience, the degree of shrinkage does not vary smoothly as parameters are changed but tends to be "all or nothing," and near the tipping point, the behavior of estimates can be affected considerably by the choice of algorithm used to fit the model.
-
-Finally, `mod.3` directly estimates the model used to generate the data.
-As mentioned, it is a constrained version of `mod.2`, with the coefficient of `xm` set to 0, and with `x` expressed as a deviation from the cluster mean `xm`:
-
-```{r}
-summary(mod.3)
-```
-
-The predictions from `mod.3` are therefore similar to those from `mod.2`:
-
-```{r plot-fits-mod3}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Predictions from the estimated model generating the data."
-(plot +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
-         groups=group, type="l", lwd=2) +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
-         groups=group, type="l", lwd=2, lty=3)
-) |> update(
-  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
-  ylim=c(4, 16),
-  key=list(
-    corner=c(0.05, 0.05),
-    text=list(c("fixed effects only","fixed and random")),
-    lines=list(lty=c(1, 3))))
-```
-
-We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.
-In order to reduce between-model random variability in comparisons of models, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model:
-
-```{r cross-validation-clusters}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept models with varying fixed effects.
-The error bars show the 95% confidence interval around the CV estimate of the MSE for each model."
-modlist <- models("~ 1"=mod.0, "~ 1 + x"=mod.1, 
-                  "~ 1 + x + xm"=mod.2, "~ 1 + I(x - xm)"=mod.3)
-cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
-plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")
-```
-
-```{r cross-validation-cases}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "10-fold case-based cross-validation comparing random intercept models with varying fixed effects."
-cvs_cases <- cv(modlist, data=Data, seed=9693)
-plot(cvs_cases, main="Model Comparison, Case-Based CV")
-```
-
-In summary, model `mod.1`, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
-Model `mod.2`, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV.
-The data-generating model, `mod.3`, which includes the fixed effect of `x - xm` in place of separate terms in `x` and `xm`, isn't distinguishable from model `mod.2`, which includes `x` and `xm` separately, even though `mod.2` has an unnecessary parameter (recall that the population coefficient of `xm` is 0 when `x` is expressed as deviations from the contextual mean).
-These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
-
-## Example: Crossed random effects
-
-Crossed random effects arise when the structure of the data aren't strictly hierarchical.
-Nevertheless, crossed and nested random effects can be handled in much the same manner, by refitting the mixed-effects model to the data with a fold of clusters or cases removed and using the refitted model to predict the response in the removed fold.
-
-We'll illustrate with data on pig growth, introduced by @DiggleLiangZeger:1994 [Table 3.1]. The data are in the `Pigs` data frame in the **cv** package:
-```{r pigs}
-head(Pigs, 9)
-head(xtabs(~ id + week, data=Pigs), 3)
-tail(xtabs(~ id + week, data=Pigs), 3)
-```
-Each of 48 pigs is observed weekly over a period of 9 weeks, with the weight of the pig recorded in kg.
-The data are in "long" format, as is appropriate for use with the `lmer()` function in the **lme4** package.
-The data are very regular, with no missing cases.
-
-The following graph, showing the growth trajectories of the pigs, is similar to Figure 3.1 in @DiggleLiangZeger:1994; we add an overall least-squares line and a loess smooth, which are nearly indistinguishable:
-```{r pigs-graph}
-#| out.width = "60%",
-#| fig.height = 6,
-#| fig.cap = "Growth trajectories for 48 pigs, with overall least-squares line (sold blue) and loess line (broken magenta)."
-plot(weight ~ week, data=Pigs, type="n")
-for (i in unique(Pigs$id)){
-  with(Pigs, lines(x=1:9, y=Pigs[id == i, "weight"],
-                   col="gray"))
-}
-abline(lm(weight ~ week, data=Pigs), col="blue", lwd=2)
-lines(with(Pigs, loess.smooth(week, weight, span=0.5)),
-      col="magenta", lty=2, lwd=2)
-```
-The individual "growth curves" and the overall trend are generally linear, with some tendency for variability of pig weight to increase over weeks (a feature of the data that we ignore in the mixed model that we fit to the data below).
-
-The **Stata** mixed-effects models manual proposes a model with crossed random effects for the `Pigs` data [@Stata:2023 page 37]:
-
-> [S]uppose that we wish to fit
-$$
-\mathrm{weight}_{ij} = \beta_0 + \beta_1 \mathrm{week}_{ij} + u_i + v_j + \varepsilon_{ij} 
-$$
-for the $i = 1, \ldots, 9$ weeks and $j = 1, \dots, 48$ pigs and
-$$
-u_i \sim N(0, \sigma^2_u); v_j \sim N(0, \sigma^2_v ); \varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon)
-$$
-all independently.
-That is, we assume an overall population-average growth curve $\beta_0 + \beta_1 \mathrm{week}$ and a random pig-specific shift.
-In other words, the effect due to week, $u_i$, is systematic to that week and  common to all pigs.
-The rationale behind [this model] could be that, assuming that the pigs were measured contemporaneously, we might be concerned that week-specific random factors such as weather and feeding patterns had significant systematic effects on all pigs.
-
-Although we might prefer an alternative model,[^pig-alternative] we think that this is a reasonable specification.
-
-[^pig-alternative]: These are repeated-measures data, which would be more conventionally modeled with autocorrelated errors within pigs.
-The `lme()` function in the **nlme** package, for example, is capable of fitting a mixed-model of this form.
-
-The **Stata** manual fits the mixed model by maximum likelihood (rather than REML), and we duplicate the results reported there using `lmer()`:
-```{r pigs-lmer}
-m.p <- lmer(weight ~ week + (1 | id) + (1 | week),
-            data=Pigs, REML=FALSE, # i.e., ML
-            control=lmerControl(optimizer="bobyqa"))
-summary(m.p)
-```
-We opt for the non-default `"bobyqa"` optimizer because it provides more numerically stable results for subsequent cross-validation in this example.
-
-We can then cross-validate the model by omitting folds composed of pigs, folds composed of weeks, or folds composed of pig-weeks (which in the `Pigs` data set correspond to individual cases, using only the fixed effects):
-```{r pigs-cv}
-cv(m.p, clusterVariables="id")
-
-cv(m.p, clusterVariables="week")
-
-cv(m.p, clusterVariables=c("id", "week"), k=10, seed=8469)
-```
-We can also cross-validate the individual cases taking account of the random effects (employing the same 10 folds):
-```{r pigs-cv-cases}
-cv(m.p, k=10, seed=8469)
-```
-Because these predictions are based on BLUPs, they are more accurate than the predictions based only on fixed effects.[^crossed-vs-nested] As well, the difference between the MSE computed for the model fit to the full data and the CV estimates of the MSE is greater here than for cluster-based predictions.
-
-[^crossed-vs-nested]: Even though there is only one observation per combination of pigs and weeks, we can use the BLUP for the omitted case because of the crossed structure of the random effects; that is each pig-week has a pig random effect and a week random effect.
-Although it probably isn't sensible, we can imagine a mixed model for the pig data that employs nested random effects, which would be specified by `lmer(weight ~ week + (1 | id/week), data=Pigs)`---that is, a random intercept that varies by combinations of `id` (pig) and `week`.
-This model can't be fit, however: With only one case per combination of `id` and `week`, the nested random-effect variance is indistinguishable from the case-level variance.
-
-# Replicating cross-validation
-
-Assuming that the number of cases $n$ is a multiple of the number of folds $k$---a slightly simplifying assumption---the number of possible partitions of cases into folds is $\frac{n!}{[(n/k)!]^k}$, a number that grows very large very quickly.
-For example, for $n = 10$ and $k = 5$, so that the folds are each of size $n/k = 2$, there are $113,400$ possible partitions; for $n=100$ and $k=5$, where $n/k = 20$, still a small problem, the number of possible partitions is truly astronomical, $1.09\times 10^{66}$.
-
-Because the partition into folds that's employed is selected randomly, the resulting CV criterion estimates are subject to sampling error.
-(An exception is LOO cross-validation, which is not at all random.) To get a sense of the magnitude of the sampling error, we can repeat the CV procedure with different randomly selected partitions into folds.
-All of the CV functions in the **cv** package are capable of repeated cross-validation, with the number of repetitions controlled by the `reps` argument, which defaults to `1`.
-
-Here, for example, is 10-fold CV for the Mroz logistic regression, repeated 5 times:
-
-```{r mroz-reps}
-cv(m.mroz, criterion=BayesRule, seed=248, reps=5, 
-   method="Woodbury")
-```
-When `reps` > `1`, the result returned by `cv()` is an object of class `"cvList"`---literally a list of `"cv"` objects.
-The results are reported for each repetition and then averaged across repetitions, with the standard deviations of the CV criterion and the biased-adjusted CV criterion given in parentheses.
-In this example, there is therefore little variation across repetitions, increasing our confidence in the reliability of the results.
-
-Notice that the seed that's set in the `cv()` command pertains to the first repetition and the seeds for the remaining repetitions are then selected pseudo-randomly.[^replicates] Setting the first seed, however, makes the entire process easily replicable, and the seed for each repetition is stored in the corresponding element of the `"cvList"` object (which isn't, however, saved in the example).
-
-[^replicates]: Because of the manner in which the computation is performed, the order of the replicates in the `"cvList"` object returned by `cv()` isn't the same as the order in which the replicates are computed.
-Each element of the result, however, is a `"cv"` object with the correct random-number seed saved, and so this technical detail can be safely ignored.
-The individual `"cv"` objects are printed in the order in which they are stored rather than the order in which they are computed.
-
-It's also possible to replicate CV when comparing competing models via the `cv()` method for `"modList"` objects.
-Recall our comparison of polynomial regressions of varying degree fit to the `Auto` data; we performed 10-fold CV for each of 10 models.
-Here, we replicate that process 5 times for each model and graph the results:
-```{r model-comparison-with-reps}
-#| out.width = "100%",
-#| fig.height = 5,
-#| fig.cap = " Replicated cross-validated 10-fold CV as a function of polynomial degree, $p$"
-cv.auto.reps <- cv(models(m.1, m.2, m.3, m.4, m.5,
-                        m.6, m.7, m.8, m.9, m.10),
-                 data=Auto, seed=8004, reps=5)
-plot(cv.auto.reps)
-```
-The graph shows both the average CV criterion and its range for each of the competing models.
-
-# Cross-validating model selection
-
-## A preliminary example
-
-As @HastieTibshiraniFriedman:2009 [Sec.
-7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain, if the whole data are used to select or fine-tune a statistical model, subsequent cross-validation of the model is intrinsically misleading, because the model is selected to fit the whole data, including the part of the data that remains when each fold is removed.
-
-The following example is similar in spirit to one employed by @HastieTibshiraniFriedman:2009.
-Suppose that we randomly generate $n = 1000$ independent observations for a response variable variable $y \sim N(\mu = 10, \sigma^2 = 0)$, and independently sample $1000$ observations for $p = 100$ "predictors," $x_1, \ldots, x_{100}$, each from $x_j \sim N(0, 1)$.
-The response has nothing to do with the predictors and so the population linear-regression model $y_i = \alpha + \beta_1 x_{i1} + \cdots + \beta_{100} x_{i,100} + \varepsilon_i$ has $\alpha = 10$ and all $\beta_j = 0$.
-
-```{r generate-selection-data}
-set.seed(24361) # for reproducibility
-D <- data.frame(
-  y = rnorm(1000, mean=10),
-  X = matrix(rnorm(1000*100), 1000, 100)
-)
-head(D[, 1:6])
-```
-Least-squares provides accurate estimates of the regression constant $\alpha = 10$ and the error variance $\sigma^2 = 1$ for the "null model" including only the regression constant; moreover, the omnibus $F$-test of the correct null hypothesis that all of the $\beta$s are 0 for the "full model" with all 100 $x$s is associated with a large $p$-value:
-```{r omnibus-F}
-m.full <- lm(y ~ ., data=D)
-m.null <- lm(y ~ 1, data=D)
-anova(m.null, m.full)
-
-summary(m.null)
-```
-Next, using the `stepAIC()` function in the **MASS** package [@VenablesRipley:2002], let us perform a forward stepwise regression to select a "best" model, starting with the null model, and using AIC as the model-selection criterion (see the help page for `stepAIC()` for details):[^selection-order]
-
-[^selection-order]: It's generally advantageous to start with the largest model, here the one with 100 predictors, and proceed by backward elimination.
-In this demonstration, however, where all of the $\beta$s are really 0, the selected model will be small, and so we proceed by forward selection from the null model to save computing time.
-
-```{r forward-selection}
-library("MASS")  # for stepAIC()
-m.select <- stepAIC(m.null,
-                    direction="forward", trace=FALSE,
-                    scope=list(lower=~1, upper=formula(m.full)))
-summary(m.select)
-mse(D$y, fitted(m.select))
-```
-The resulting model has 15 predictors, a very modest $R^2 = .044$, but a small $p$-value for its omnibus $F$-test (which, of course, is entirely spurious because the same data were used to select and test the model).
-The MSE for the selected model is smaller than the true error variance $\sigma^2 = 1$, as is the estimated error variance for the selected model, $\widehat{\sigma}^2 = 0.973^2 = 0.947$.
-
-If we cross-validate the selected model, we also obtain an optimistic estimate of its predictive power (although the confidence interval for the bias-adjusted MSE includes 1):
-```{r cv-selectedModel}
-cv(m.select, seed=2529)
-```
-
-The `cvSelect()` function in the **cv** package allows us to cross-validate the whole model-selection procedure.
-The first argument to `cvSelect()` is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion.
-The `selectStepAIC()` function, also in **cv** and based on `stepAIC()`, is suitable for use with `cvSelect()`:
-```{r cvSelect-artificial-data, cache=TRUE}
-cv.select <- cvSelect(selectStepAIC, data=D, seed=3791,
-                      model=m.null, direction="forward",
-                      scope=list(lower=~1, 
-                                 upper=formula(m.full)))
-cv.select
-```
-
-The other arguments to `cvSelect()` are:
-
-* `data`, the data set to which the model is fit;
-* `seed`, an optional seed for R's pseudo-random-number generator; as for `cv()`, if the seed isn't supplied by the user, a seed is randomly selected and saved;
-* additional arguments required by the model-selection function, here the starting `model` argument, the `direction` of model selection, and the `scope` of models considered (from the model with only a regression constant to the model with all 100 predictors).
-
-By default, `cvSelect()` performs 10-fold CV, and produces an estimate of MSE for the model-selection procedure even *larger* than the true error variance, $\sigma^2 = 1$.
-
-Also by default, when the number of folds is 10 or fewer, `cvSelect()` saves the coefficients of the selected models.
-In this example, the `compareFolds()` function reveals that the variables retained by the model-selection process in the several folds are quite different:
-```{r compare-selected-models}
-compareFolds(cv.select)
-```
-## Mroz's logistic regression revisited
-
-For a contrasting example we apply model selection to Mroz's logistic regression for married women's labor-force participation.
-First, recall the logistic regression model that we fit to the `Mroz` data:
-```{r recall-Mroz-regression}
-summary(m.mroz)
-```
-
-Applying stepwise model selection Mroz's logistic regression, using BIC as the model-selection criterion (via the argument `k=log(nrow(Mroz))` to `stepAIC()`) selects 5 of the 7 original predictors:
-```{r mroz-selection}
-m.mroz.sel <- stepAIC(m.mroz, k=log(nrow(Mroz)),
-                      trace=FALSE)
-summary(m.mroz.sel)
-BayesRule(Mroz$lfp == "yes",
-          predict(m.mroz.sel, type="response"))
-```
-Bayes rule applied to the selected model misclassifies 32% of the cases in the `Mroz` data.
-
-Cross-validating the selected model produces a similar, slightly larger, estimate of misclassification, about 33%:
-```{r cv-mroz-regression}
-cv(m.mroz.sel, criterion=BayesRule, seed=345266)
-```
-Is this estimate of predictive performance optimistic?
-
-We proceed to apply the model-selection procedure by cross-validation, producing more or less the same result:
-```{r cv-mroz-selection}
-m.mroz.sel.cv <- cvSelect(selectStepAIC, Mroz, 
-                          seed=6681,
-                          criterion=BayesRule,
-                          model=m.mroz,
-                          AIC=FALSE)
-m.mroz.sel.cv
-```
-Setting `AIC=FALSE` in the call to `cvSelect()` uses the BIC rather than the AIC as the model-selection criterion.
-As it turns out, exactly the same predictors are selected when each of the 10 folds are omitted, and the several coefficient estimates are very similar, as we show using `compareFolds()`:
-```{r compare-selected-models-mroz}
-compareFolds(m.mroz.sel.cv)
-```
-In this example, therefore, we appear to obtain a realistic estimate of model performance directly from the selected model, because there is little added uncertainty induced by model selection.
-
-## Cross-validating choice of transformations in regression
-
-The **cv** package also provides a `cvSelect()` procedure, `selectTrans()`, for choosing transformations of the predictors and the response in regression.
-
-Some background: As @Weisberg:2014 [Sec.
-8.2] explains, there are technical advantages to having (numeric) predictors in linear regression analysis that are themselves linearly related.
-If the predictors *aren't* linearly related, then the relationships between them can often be straightened by power transformations.
-Transformations can be selected after graphical examination of the data, or by analytic methods.
-Once the relationships between the predictors are linearized, it can be advantageous similarly to transform the response variable towards normality.
-
-Selecting transformations analytically raises the possibility of automating the process, as would be required for cross-validation.
-One could, in principle, apply graphical methods to select transformations for each fold, but because a data analyst couldn't forget the choices made for previous folds, the process wouldn't really be applied independently to the folds.
-
-To illustrate, we adapt an example appearing in several places in @FoxWeisberg:2019 (for example in Chapter 3 on transforming data), using data on the prestige and other characteristics of 102 Canadian occupations circa 1970.
-The data are in the `Prestige` data frame in the **carData** package:
-```{r Prestige-data}
-data("Prestige", package="carData")
-head(Prestige)
-summary(Prestige)
-```
-The variables in the `Prestige` data set are:
-
-* `education`: average years of education for incumbents in the occupation, from the 1971 Canadian Census.
-* `income`: average dollars of annual income for the occupation, from the Census.
-* `women`: percentage of occupational incumbents who were women, also from the Census.
-* `prestige`: the average prestige rating of the occupation on a 0--100 "thermometer" scale, in a Canadian social survey conducted around the same time.
-* `type`, type of occupation, and `census`, the Census occupational code, which are not used in our example.
-
-The object of a regression analysis for the `Prestige` data (and their original purpose) is to predict occupational prestige from the other variables in the data set.
-
-A scatterplot matrix (using the `scatterplotMatrix()` function in the **car** package) of the numeric variables in the data reveals that the distributions of `income` and `women` are positively skewed, and that some of the relationships among the three predictors, and between the predictors and the response (i.e., `prestige`), are nonlinear:
-```{r scatterplot-matrix}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Scatterplot matrix for the `Prestige` data."
-library("car")
-scatterplotMatrix(~ prestige + income + education + women,
-                  data=Prestige, smooth=list(spread=FALSE))
-```
-The `powerTransform()` function in the **car** package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964].
-Several "families" of power transformations can be used, including the original Box-Cox family, simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap.
-8], and @FoxWeisberg:2019 [Chap.
-3] for details.
-Because `women` has some zero values, we use the Yeo-Johnson family:
-```{r power-transform-Prestige}
-trans <- powerTransform( cbind(income, education, women) ~ 1,
-                         data=Prestige, family="yjPower")
-summary(trans)
-```
-We thus have evidence of the desirability of transforming `income` (by the $1/3$ power) and `women` (by the $0.16$ power---which is close to the "0" power, i.e., the log transformation), but not `education`.
-Applying the "rounded" power transformations makes the predictors better-behaved:
-```{r transformed-predictors}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Scatterplot matrix for the `Prestige` data with the predictors transformed."
-P <- Prestige[, c("prestige", "income", "education", "women")]
-(lambdas <- trans$roundlam)
-names(lambdas) <- c("income", "education", "women")
-for (var in c("income", "education", "women")){
-  P[, var] <- yjPower(P[, var], lambda=lambdas[var])
-}
-summary(P)
-
-scatterplotMatrix(~ prestige + income + education + women,
-                  data=P, smooth=list(spread=FALSE))
-```
-Comparing the MSE for the regressions with the original and transformed predictors shows a advantage to the latter:
-```{r prestige-regressions}
-m.pres <- lm(prestige ~ income + education + women, data=Prestige)
-m.pres.trans <- lm(prestige ~ income + education + women, data=P)
-mse(Prestige$prestige, fitted(m.pres))
-mse(P$prestige, fitted(m.pres.trans))
-```
-Similarly, component+residual plots for the two regressions, produced by the `crPlots()` function in the **car** package, suggest that the partial relationship of `prestige` to `income` is more nearly linear in the transformed data, but the transformation of `women` fails to capture what appears to be a slight quadratic partial relationship; the partial relationship of `prestige` to `education` is close to linear in both regressions:
-```{r CR-plots-untransformed}
-#| fig.cap = "Component+residual plots for the `Prestige` regression with the original predictors."
-crPlots(m.pres)
-```
-```{r CR-plots-transformed}
-#| fig.cap = "Component+residual plots for the `Prestige` regression with transformed predictors."
-crPlots(m.pres.trans)
-```
-
-Having transformed the predictors towards multinormality, we now consider whether there's evidence for transforming the response (using `powerTransform()` for Box and Cox's original method), and we discover that there's not:
-```{r transform-response}
-summary(powerTransform(m.pres.trans))
-```
-
-The `selectTrans()` function in the **cv** package automates the process of selecting predictor and response transformations.
-The function takes a `data` set and "working" `model` as arguments, along with the candidate `predictors` and `response` for transformation, and the transformation `family` to employ.
-If the `predictors` argument is missing then only the response is transformed, and if the `response` argument is missing, only the supplied predictors are transformed.
-The default `family` for transforming the predictors is `"bcPower"`---the original Box-Cox family---as is the default `family.y` for transforming the response; here we specify `family="yjPower` because of the zeros in `women`.
-`selectTrans()` returns the result of applying a lack-of-fit criterion to the model after the selected transformation is applied, with the default `criterion=mse`:
-```{r selectTrans}
-selectTrans(data=Prestige, model=m.pres,
-            predictors=c("income", "education", "women"),
-            response="prestige", family="yjPower")
-```
-`selectTrans()` also takes an optional `indices` argument, making it suitable for doing computations on a subset of the data (i.e., a CV fold), and hence for use with `cvSelect()` (see `?selectTrans` for details):
-
-```{r cv-select-transformations}
-cvs <- cvSelect(selectTrans, data=Prestige, model=m.pres, seed=1463,
-                predictors=c("income", "education", "women"),
-                response="prestige",
-                family="yjPower")
-cvs
-
-cv(m.pres, seed=1463) # untransformed model with same folds
-
-compareFolds(cvs)
-```
-The results suggest that the predictive power of the transformed regression is reliably greater than that of the untransformed regression (though in both case, the cross-validated MSE is considerably higher than the MSE computed for the whole data).
-Examining the selected transformations for each fold reveals that the predictor `education` and the response `prestige` are never transformed; that the $1/3$ power is selected for `income` in all of the folds; and that the transformation selected for `women` varies narrowly across the folds between the $0$th power (i.e., log) and the $1/3$ power.
-
-## Selecting both transformations and predictors[^Venables]
-
-[^Venables]: The presentation in the section benefits from an email conversation with Bill Venables, who of course isn't responsible for the use to which we've put his insightful remarks.
-
-As we mentioned, @HastieTibshiraniFriedman:2009 [Sec.
-7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain that honest cross-validation has to take account of model specification and selection.
-Statistical modeling is at least partly a craft, and one could imagine applying that craft to successive partial data sets, each with a fold removed.
-The resulting procedure would be tedious, though possibly worth the effort, but it would also be difficult to realize in practice: After all, we can hardly erase our memory of statistical modeling choices between analyzing partial data sets.
-
-Alternatively, if we're able to automate the process of model selection, then we can more realistically apply CV mechanically.
-That's what we did in the preceding two sections, first for predictor selection and then for selection of transformations in regression.
-In this section, we consider the case where we both select variable transformations and then proceed to select predictors.
-It's insufficient to apply these steps sequentially, first, for example, using `cvSelect()` with `selectTrans()` and then with `selectStepAIC()`; rather we should apply the whole model-selection procedure with each fold omitted.
-The `selectTransAndStepAIC()` function, also supplied by the **cv** package, does exactly that.
-
-To illustrate this process, we return to the `Auto` data set:
-```{r Auto-redux}
-summary(Auto)
-xtabs(~ year, data=Auto)
-xtabs(~ origin, data=Auto)
-xtabs(~ cylinders, data=Auto)
-```
-We previously used the `Auto` here in a preliminary example where we employed CV to inform the selection of the order of a polynomial regression of `mpg` on `horsepower`.
-Here, we consider more generally the problem of predicting `mpg` from the other variables in the `Auto` data.
-We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set:
-```{r Auto-explore}
-#| out.width = "100%",
-#| fig.height = 7,
-#| fig.cap = "Scatterplot matrix for the numeric variables in the `Auto` data"
-Auto$cylinders <- factor(Auto$cylinders,
-                         labels=c("3.4", "3.4", "5.6", "5.6", "8"))
-Auto$year <- as.factor(Auto$year)
-Auto$origin <- factor(Auto$origin,
-                      labels=c("America", "Europe", "Japan"))
-rownames(Auto) <- make.names(Auto$name, unique=TRUE)
-Auto$name <- NULL
-
-scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
-                  smooth=list(spread=FALSE), data=Auto)
-```
-A comment before we proceed: `origin` is clearly categorical and so converting it to a factor is natural, but we could imagine treating `cylinders` and `year` as numeric predictors.
-There are, however, only 5 distinct values of `cylinders` (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare.
-and none of the cars has 7 cylinders.
-There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between `mpg` and `year` is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.
-
-[^year]: Of course, making the decision to treat `year` as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process.
-It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation.
-We believe, however, that it's still useful to automate as much of the process as we can to obtain a more realistic, if still biased, estimate of the predictive power of a model.
-
-We begin with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
-```{r Auto-working-model}
-#| out.width = "100%",
-#| fig.height = 7,
-#| fig.cap = "Component+residual plots for the working model fit to the `Auto` data"
-m.auto <- lm(mpg ~ ., data = Auto)
-summary(m.auto)
-
-Anova(m.auto)
-
-crPlots(m.auto)
-```
-The component+residual plots, created with the `crPlots()` function in the previously loaded **car** package, clearly reveal the inadequacy of the model.
-
-We proceed to transform the numeric predictors towards multi-normality: 
-```{r Auto-transform}
-num.predictors <- c("displacement", "horsepower", "weight", "acceleration")
-tr.x <- powerTransform(Auto[, num.predictors])
-summary(tr.x)
-```
-We then apply the (rounded) transformations---all, as it turns out, logs---to the data and re-estimate the model:
-```{r Auto-with-transformed-predictors}
-A <- Auto
-powers <- tr.x$roundlam
-for (pred in num.predictors){
-  A[, pred] <- bcPower(A[, pred], lambda=powers[pred])
-}
-head(A)
-
-m <- update(m.auto, data=A)
-```
-Finally, we perform Box-Cox regression to transform the response (also obtaining a log transformation):
-```{r Auto-Box-Cox}
-summary(powerTransform(m))
-
-m <- update(m, log(mpg) ~ .)
-summary(m)
-
-Anova(m)
-```
-
-The transformed numeric variables are much better-behaved:
-```{r Auto-transformed-scatterplot-matrix}
-#| out.width = "100%",
-#| fig.height = 7,
-#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the `Auto` data"
-scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
-                  + acceleration, 
-                  smooth=list(spread=FALSE), data=A)
-```
-And the partial relationships in the model fit to the transformed data are much more nearly linear:
-```{r Auto-CR-plots-transformed}
-#| out.width = "100%",
-#| fig.height = 7,
-#| fig.cap = "Component+residual plots for the model fit to the transformed `Auto` data"
-crPlots(m)
-```
-
-Having transformed both the numeric predictors and the response, we proceed to use the `stepAIC()` function in the **MASS** package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log(n)$):
-```{r}
-m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
-summary(m.step)
-
-Anova(m.step)
-```
-The selected model includes three of the numeric predictors, `horsepower`, `weight`, and `acceleration`, along with the factors `year` and `origin`.
-We can calculate the MSE for this model, but we expect that the result will be optimistic because we used the whole data to help specify the model
-```{r MSE-whole-selected-model}
-mse(Auto$mpg, exp(fitted(m.step)))
-```
-This is considerably smaller than the MSE for the original working model:
-```{r MSE-working-model}
-mse(Auto$mpg, fitted(m.auto))
-```
-A perhaps subtle point is that we compute the MSE for the selected model on the original `mpg` response scale rather than the log scale, so as to make the selected model comparable to the working model.
-That's slightly uncomfortable given the skewed distribution of `mpg`.
-An alternative is to use the median absolute error instead of the mean-squared error, employing the `medAbsErr()` function from the **cv** package:
-```{r Auto-median-absolute-error}
-medAbsErr(Auto$mpg, exp(fitted(m.step)))
-medAbsErr(Auto$mpg, fitted(m.auto))
-```
-
-Now let's use `cvSelect()` with `selectTransAndStepAIC()` to automate and cross-validate the whole model-specification process:
-```{r Auto-transform-and-select}
-num.predictors
-cvs <- cvSelect(selectTransStepAIC, data=Auto, seed=76692, model=m.auto,
-                predictors=num.predictors,
-                response="mpg", AIC=FALSE, criterion=medAbsErr)
-cvs
-
-compareFolds(cvs)
-```
-Here, as for `selectTrans()`, the `predictors` and `response` arguments specify candidate variables for transformation,
-and `AIC=FALSE` uses the BIC for model selection.
-The starting model, `m.auto`, is the working model fit to the `Auto` data.
-The CV criterion isn't bias-adjusted because median absolute error isn't a mean of casewise error components.
-
-Some noteworthy points:
-
-* `selectTransStepAIC()` automatically computes CV cost criteria, here the median absolute error, on the untransformed response scale.
-* The estimate of the median absolute error that we obtain by cross-validating the whole model-specification process is a little larger than the median absolute error computed for the model we fit to the `Auto` data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
-* When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of `compareFolds()`), we see that there is little uncertainty in choosing variable transformations (the `lam.*`s for the $x$s and `lambda` for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: `horsepower`, `weight`, and `year` are always included among the selected predictors; `acceleration` and `displacement` are each included respectively in 4 and 3 of 10 selected models; and `cylinders` and `origin` are each included in only 1 of 10 models.
-Recall that when we selected predictors for the full data, we obtained a model with `horsepower`, `weight`, `acceleration`, `year`, and `origin`.
-
-# Parallel computations
-
-The CV functions in the **cv** package are all capable of performing parallel computations by setting the `ncores` argument (specifying the number of computer cores to be used) to a number > `1` (which is the default).
-Parallel computation can be advantageous for large problems, reducing the execution time of the program.
-
-To illustrate, let's time model selection in Mroz's logistic regression, repeating the computation as performed previously and then doing it in parallel using 2 cores:
-```{r parallel-computation}
-system.time(m.mroz.sel.cv <- cvSelect(selectStepAIC, Mroz,
-                          seed=6681,
-                          criterion=BayesRule,
-                          model=m.mroz,
-                          AIC=FALSE))
-
-system.time(m.mroz.sel.cv.p <- cvSelect(selectStepAIC, Mroz,
-                          seed=6681,
-                          criterion=BayesRule,
-                          model=m.mroz,
-                          AIC=FALSE,
-                          ncores=2))
-all.equal(m.mroz.sel.cv, m.mroz.sel.cv.p)
-```
-
-In this small problem, the parallel computation is actually *slower*, because there is an overhead cost to parallelization, but we can see that it produces the same result as before.
-
-# Extending the cv package
-
-The **cv** package is designed to be extensible in several directions.
-In this vignette, we discuss three kinds of extensions, ordered by increasing general complexity: (1) adding a cross-validation cost criterion; (2) adding a model class that's not directly accommodated by the `cv()` default method or by another directly inherited method, with separate consideration of mixed-effects models; and (3) adding a new model-selection procedure suitable for use with `selectModel()`.
-
-## Adding a cost criterion
-
-A cost criterion suitable for use with `cv()` or `cvSelect()` should take two arguments, `y` (the observed response vector) and `yhat` (a vector of fitted or predicted response values), and return a numeric index of lack of fit.
-The **cv** package supplies several such criteria: `mse(y, yhat)`, which returns the mean-squared prediction error for a numeric response; `rmse(y, yhat)`, which returns the (square-)root mean-squared error; `medAbsErr(y, yhat)`, which returns the median absolute error; and `BayesRule(y, yhat)` (and its non-error-checking version, `BayesRule2(y, yhat))`, suitable for use with a binary regression model, where `y` is the binary response coded `0` for a "failure" or `1` for a "success"; where `yhat` is the predicted probability of success; and where the proportion of *incorrectly* classified cases is returned.
-
-To illustrate using a different prediction cost criterion, we'll base a cost criterion on the area under the receiver operating characteristic ("ROC") curve for a logistic regression.
-The ROC curve is a graphical representation of the classification power of a binary regression model, and the area under the ROC curve ("AUC"), which varies from 0 to 1, is a common summary measure based on the ROC [see @Wikipedia-ROC:2023].
-The **Metrics** package [@HamnerFrasco:2018] includes a variety of measures useful for model selection, including an `auc()` function.
-We convert the AUC into a cost measure by taking its complement:
-
-```{r AUCcomp}
-AUCcomp <- function(y, yhat) 1 - Metrics::auc(y, yhat)
-```
-
-We then apply `AUCcomp()` to the the Mroz logistic regression discussed in the main **cv** package vignette, which we reproduce here, using the `Mroz` data frame from the **carData** package [@FoxWeisberg:2019]:
-```{r Mroz-regression}
-data("Mroz", package="carData")
-m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
-summary(m.mroz)
-
-AUCcomp(with(Mroz, as.numeric(lfp == "yes")), fitted(m.mroz))
-```
-Cross-validating this cost measure is straightforward:
-```{r Mroz-CV-ROC}
-library("cv")
-cv(m.mroz, criterion=AUCcomp, seed=3639)
-```
-As expected, the cross-validated complement to the AUC is somewhat less optimistic than the criterion computed from the model fit to the whole data set.
-
-As we explain in the vignette "Cross-validation of regression models," the `cv()` function differentiates between CV criteria that are averages of casewise components and criteria that are not.
-Computation of bias corrections and confidence intervals is limited to the former.
-We show in the appendix to this vignette that the AUC, and hence its complement, cannot be expressed as averages of casewise components.
-
-`cv()` looks for a `"casewise loss"` attribute of the value returned by a CV criterion function.
-If this attribute exists, then the criterion is treated as the mean of casewise components, and `cv()` uses the unexported function `getLossFn()` to construct a function that returns the casewise components of the criterion.
-
-We illustrate with the `mse()`:
-```{r}
-mse
-
-cv:::getLossFn(mse(rnorm(100), rnorm(100)))
-```
-For this scheme to work, the "casewise loss" attribute must be a character string (or vector of character strings), here `"(y - yhat)^2"`, that evaluates to an expression that is a function of `y` and `yhat`, and that computes the vector of casewise components of the CV criterion.
-
-## Adding a model class not covered by the default cv() method
-
-### Independently sampled cases
-
-Suppose that we want to cross-validate a multinomial logistic regression model fit by the `multinom()` function in the **nnet** package [@VenablesRipley:2002].
-We borrow an example from @Fox:2016 [Sec.
-14.2.1], with data from the British Election Panel Study on vote choice in the 2001 British election.
-Data for the example are in the `BEPS` data frame in the **carData** package:
-```{r BEPS-data}
-data("BEPS", package="carData")
-head(BEPS)
-```
-The polytomous (multi-category) response variable is `vote`, a factor with levels `"Conservative"`, `"Labour"`, and `"Liberal Democrat"`. The predictors of `vote` are:
-
-* `age`, in years;
-* `econ.cond.national` and `econ.cond.household`, the respondent's ratings of the state of the economy, on 1 to 5 scales.
-* `Blair`, `Hague`, and `Kennedy`, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
-* `Europe`, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism."
-* `political.knowledge`, knowledge of the parties' positions on European integration, with scores from 0 to 3.
-* `gender`, `"female"` or `"male"`.
-
-The model fit to the data includes an interaction between `Europe` and `political.knowledge`; the other predictors enter the model additively:
-```{r BEPS-model}
-library("nnet")
-m.beps <- multinom(vote ~ age + gender + economic.cond.national +
-                       economic.cond.household + Blair + Hague + Kennedy +
-                       Europe*political.knowledge, data=BEPS)
-
-car::Anova(m.beps)
-```
-Most of the predictors, including the `Europe` $\times$ `political.knowledge` interaction, are associated with very small $p$-values; the `Anova()` function is from the **car** package [@FoxWeisberg:2019].
-
-Here's an "effect plot", using the the **effects** package [@FoxWeisberg:2019] to visualize the `Europe` $\times$ `political.knowledge` interaction in a "stacked-area" graph:
-```{r BEPS-plot, fig.width=9, fig.height=5}
-plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
-            xlevels=list(Europe=1:11, political.knowledge=0:3),
-            fixed.predictors=list(given.values=c(gendermale=0.5))),
-     lines=list(col=c("blue", "red", "orange")),
-     axes=list(x=list(rug=FALSE), y=list(style="stacked")))
-```
-
-To cross-validate this multinomial-logit model we need an appropriate cost criterion.
-None of the criteria supplied by the **cv** package---for example, neither `mse()`, which is appropriate for a numeric response, nor `BayesRule()`, which is appropriate for a binary response---will do.
-One possibility is to adapt Bayes rule to a polytomous response:
-```{r BayesRuleMulti}
-head(BEPS$vote)
-yhat <- predict(m.beps, type="class")
-head(yhat)
-
-BayesRuleMulti <- function(y, yhat){
-  result <- mean(y != yhat)
-  attr(result, "casewise loss") <- "y != yhat"
-  result
-}
-
-BayesRuleMulti(BEPS$vote, yhat)
-```
-The `predict()` method for `"multinom"` models called with argument `type="class"` reports the Bayes-rule prediction for each case---that is, the response category with the highest predicted probability.
-Our `BayesRuleMulti()` function calculates the proportion of misclassified cases.
-Because this value is the mean of casewise components, we attach a `"casewise loss"` attribute to the result (as explained in the preceding section).
-
-The marginal proportions for the response categories are
-```{r BEPS-response-distribution}
-xtabs(~ vote, data=BEPS)/nrow(BEPS)
-```
-and so the marginal Bayes-rule prediction, that everyone will vote Labour, produces an error rate of $1 - 0.47213 = 0.52787$.
-The multinomial-logit model appears to do substantially better than that, but does its performance hold up to cross-validation?
-
-We check first whether the default `cv()` method works "out-of-the-box" for the `"multinom"` model:
-```{r BEPS-test-default, error=TRUE}
-cv(m.beps, seed=3465, criterion=BayesRuleMulti)
-```
-The default method of `GetResponse()` (a function supplied by the **cv** package---see `?GetResponse`) fails for a `"multinom"` object.
-A straightforward solution is to supply a `GetResponse.multinom()` method that returns the factor response [using the `get_response()` function from the **insight** package, @LudeckeWaggonerMakowski:2019],
-```{r GetResponse.multinom}
-GetResponse.multinom <- function(model, ...) {
-  insight::get_response(model)
-}
-
-head(GetResponse(m.beps))
-```
-and to try again:
-```{r BEPS-test-default-2, error=TRUE}
-cv(m.beps, seed=3465, criterion=BayesRuleMulti)
-```
-A `traceback()` (not shown) reveals that the problem is that the default method of `cv()` calls the `"multinom"` method for `predict()` with the argument `type="response"`, when the correct argument should be `type="class"`.
-We therefore must write a "`multinom`" method for `cv()`, but that proves to be very simple:
-```{r cv.nultinom}
-cv.multinom <- function (model, data, criterion=BayesRuleMulti, k, reps,
-                         seed, ...){
-  NextMethod(type="class", criterion=criterion)
-}
-```
-That is, we simply call the default `cv()` method with the `type` argument properly set.
-In addition to supplying the correct `type` argument, our method sets the default `criterion` for the `cv.multinom()` method to `BayesRuleMulti`.
-
-Then:
-```{r BEPS-cv}
-m.beps <- update(m.beps, trace=FALSE)
-cv(m.beps, seed=3465)
-```
-Prior to invoking `cv()`, we called `update()` with `trace=FALSE` to suppress the iteration history reported by default by `multinom()`---it would be tedious to see the iteration history for each fold.
-The cross-validated polytomous Bayes-rule criterion confirms that the fitted model does substantially better than the marginal Bayes-rule prediction that everyone votes for Labour.
-
-### Mixed-effects models
-
-Adding a `cv()` method for a mixed-model class is somewhat more complicated.
-We provide the `cvMixed()` function to facilitate this process, and to see how that works, consider the `"lme"` method from the **cv** package:
-```{r cv.lme}
-cv:::cv.lme
-```
-Notice that `cv.lme()` sets up a call to `cvMixed()`, which does the computational work.
-
-Most of the arguments of `cvMixed()` are familiar:
-
-* `model` is the mixed-model object, here of class `"lme"`.
-
-* `package` is the name of the package in which the mixed-modeling function used to fit the model, here  `lme()`, resides---i.e., `"nlme"`; `cvMixed()` uses this argument to retrieve the package namespace.
-
-* `data` is the data set to which the model is fit, by default extracted by the `get_data()` function in the **insight** package.
-
-* `criterion` is the CV criterion, defaulting to the `mse()` function.
-
-* `k` is the number of CV folds, defaulting to `"loo"` for CV by clusters and `10` for CV by cases.
-
-* `reps` is the number of times the CV process is repeated, defaulting to `1`.
-
-* `seed` is the seed for R's random-number generator, defaulting to a randomly selected (and saved) value.
-
-* `ncores` is the number of cores to use for parallel computation; if `1`, the default, then the computation isn't parallelized.
-
-* `clusterVariables` is a character vector of the names of variables defining clusters; if missing, then CV is based on cases rather than clusters.
-
-The remaining two arguments are unfamiliar:
-
-* `predict.clusters.args` is a named list of arguments to be passed to the `predict()` function to obtain predictions for the full data set from a model fit to a subset of the data for cluster-based CV.
-The first two arguments should be `object` and `newdata`.
-It is typically necessary to tell `cvMixed()` how to base predictions only on fixed effects; in the case of `"lme"` models, this is done by setting `level = 0`.
-
-* Similarly, `predict.cases.args` is a named list of arguments to be passed to `predict()` for case-based CV. Setting `level = 1` includes random effects in the predictions.
-
-Finally, any additional arguments, absorbed by `...`, are passed to `update()` when the model is refit with each fold omitted. `cvMixed()` returns an object of class `"cv"`.
-
-Now imagine that we want to support a new class of mixed-effects models.
-To be concrete, we illustrate with the `glmmPQL()` function in the **MASS** package [@VenablesRipley:2002], which fits generalized-linear mixed-effects models by penalized quasi-likelihood.[^glmmPQL] Not coincidentally, the arguments of `glmmPQL()` are similar to those of `lme()` (with an additional `family` argument), because the former iteratively invokes the latter; so `cv.glmmPQL()` should resemble `cv.lme()`.
-
-[^glmmPQL]: This example is somewhat artificial in that `glmmPQL()` has largely been superseded by computationally superior functions, such the `glmer()` function in the **lme4** package.
-There is, however, one situation in which `glmmPQL()` might prove useful: to specify serial dependency in case-level errors within clusters for longitudinal data, which is not currently supported by `glmer()`.
-
-As it turns out, neither the default method for `GetResponse()` nor `insight::get_data()` work for `"glmmPQL"` objects.
-These objects include a `"data"` element, however, and so we can simply extract this element as the default for the `data` argument of our `cv.glmmPQL()` method.
-
-To get the response variable is more complicated: We refit the fixed part of the model as a GLM with only the regression constant on the right-hand side, and extract the response from that; because all we need is the response variable, we limit the number of GLM iterations to 1 and suppress warning messages about non-convergence:
-
-```{r GetResponse.glmmPQL}
-GetResponse.glmmPQL <- function(model, ...){
-  f <- formula(model)
-  f[[3]] <- 1 # regression constant only on RHS
-  model <- suppressWarnings(glm(f, data=model$data, family=model$family,
-                                control=list(maxit=1)))
-  cv::GetResponse(model)
-}
-```
-
-Writing the `cv()` method is then straightforward:
-```{r cv.glmmPQL}
-cv.glmmPQL <- function(model, data = model$data, criterion = mse,
-                     k, reps = 1, seed, ncores = 1, clusterVariables, ...){
-  cvMixed(
-    model,
-    package="MASS",
-    data=data,
-    criterion=criterion,
-    k=k,
-    reps=reps,
-    seed=seed,
-    ncores=ncores,
-    clusterVariables=clusterVariables,
-    predict.clusters.args=list(object=model,
-                               newdata=data,
-                               level=0,
-                               type="response"),
-    predict.cases.args=list(object=model,
-                            newdata=data,
-                            level=1,
-                            type="response"),
-    verbose=FALSE,
-    ...)
-}
-```
-We set the argument `verbose=FALSE` to suppress `glmmPQL()`'s iteration counter when `cvMixed()` calls `update()`.
-
-Let's apply our newly minted method to a logistic regression with a random intercept in an example that appears in `?glmmPQL`:
-```{r glmmPQL-example}
-library("MASS")
-m.pql <- glmmPQL(y ~ trt + I(week > 2), random = ~ 1 | ID,
-             family = binomial, data = bacteria)
-summary(m.pql)
-```
-We compare this result to that obtained from `glmer()` in the **lme4** package:
-```{r compare-to-lme4}
-library("lme4")
-m.glmer <- glmer(y ~ trt + I(week > 2) + (1 | ID),
-               family = binomial, data = bacteria)
-summary(m.glmer)
-
-  # comparison of fixed effects:
-car::compareCoefs(m.pql, m.glmer) 
-```
-The two sets of estimates are similar, but not identical
-
-Finally, we try out our `cv.glmmPQL()` method, cross-validating both by clusters and by cases,
-```{r cv-example-glmmPQL, cache=TRUE}
-cv(m.pql, clusterVariables="ID", criterion=BayesRule)
-
-cv(m.pql, data=bacteria, criterion=BayesRule, seed=1490)
-```
-and again compare to `glmer()`:
-```{r cv-example-glmer, cache=TRUE}
-cv(m.glmer, clusterVariables="ID", criterion=BayesRule)
-
-cv(m.glmer, data=bacteria, criterion=BayesRule, seed=1490)
-```
-
-## Adding a model-selection procedure
-
-The `selectStepAIC()` function supplied by the **cv** package, which is based on the `stepAIC()` function from the **nnet** package [@VenablesRipley:2002] for stepwise model selection, is suitable for the `procedure` argument of `cvSelect()`.
-The use of `selectStepAIC()` is illustrated in the principal vignette for the package.
-
-We'll employ `selectStepAIC()` as a "template" for writing a CV model-selection procedure.
-To see the code for this function, type `cv::selectStepAIC` at the R command prompt, or examine the sources for the **cv** package at https://github.com/gmonette/cv (the code for `selectStepAIC()` is in https://github.com/gmonette/cv/blob/main/R/cvSelect.R).
-
-Another approach to model selection is all-subsets regression.
-The `regsubsets()` function in the **leaps** package [@LumleyMiller:2020] implements an efficient algorithm for selecting the best-fitting linear least-squares regressions for subsets of predictors of all sizes, from 1 through the maximum number of candidate predictors.[^1] To illustrate the use of `regsubsets()`, we employ the `swiss` data frame supplied by the **leaps** package:
-
-[^1]: The `regsubsets()` function computes several measures of model predictive performance, including the $R^2$ and $R^2$ adjusted for degrees of freedom, the residual sums of squares, Mallows's $C_p$, and the BIC.
-Several of these are suitable for comparing models with differing numbers of coefficients---we use the BIC below---but all necessarily agree when comparing models with the *same* number of coefficients.
-
-```{r swiss}
-library("leaps")
-head(swiss)
-nrow(swiss)
-```
-The data set includes the following variables, for each of 47 French-speaking Swiss provinces circa 1888:
-
-* `Fertility`: A standardized fertility measure.
-* `Agriculture`: The percentage of the male population engaged in agriculture.
-* `Examination`: The percentage of draftees into the Swiss army receiving the highest grade on an examination.
-* `Education`: The percentage of draftees with more than a primary-school education.
-* `Catholic`: The percentage of the population who were Catholic.
-* `Infant.Mortality`: The infant-mortality rate, expressed as the percentage of live births surviving less than a year.
-
-Following @LumleyMiller:2020, we treat `Fertility` as the response and the other variables as predictors in a linear least-squares regression:
-```{r swiss-lm}
-m.swiss <- lm(Fertility ~ ., data=swiss)
-summary(m.swiss)
-
-cv(m.swiss, seed=8433)
-```
-Thus, the RMSE for the model fit to the complete data is considerably smaller than the CV estimate of the RMSE.
-Can we do better by selecting a subset of the predictors, taking account of the additional uncertainty induced by model selection?
-
-First, let's apply best-subset selection to the complete data set:
-```{r subset-selection}
-#| fig.cap = "Selecting the best model of each size."
-swiss.sub <- regsubsets(Fertility ~ ., data=swiss)
-summary(swiss.sub)
-
-(bics <- summary(swiss.sub)$bic)
-which.min(bics)
-
-car::subsets(swiss.sub, legend="topright")
-```
-The graph, produced by the `subsets()` function in the **car** package, shows that the model with the smallest BIC is the best model with 4 predictors, including `Agriculture`, `Education`, `Catholic`, and `Infant.Mortality`, but not `Examination`:
-```{r best-model}
-m.best <- update(m.swiss, . ~ . - Examination)
-summary(m.best)
-
-cv(m.best, seed=8433) # use same folds as before
-```
-The RMSE for the selected model is (of course) slightly higher than for the full model fit previously, but the cross-validated RMSE is a bit lower; as we explain in the main vignette, however, it isn't kosher to select and cross-validate a model on the same data.
-
-Here's a function named `selectSubsets()`, meant to be used with `cvSelect()`, suitable for cross-validating the model-selection process:
-```{r selectSubsets}
-selectSubsets <- function(data=insight::get_data(model), 
-                          model,
-                          indices,
-                          criterion=mse,
-                          save.coef=TRUE, ...){
-
-  if (inherits(model, "lm", which=TRUE) != 1)
-    stop("selectSubsets is appropriate only for 'lm' models")
-
-  y <- GetResponse(model)
-  formula <- formula(model)
-  X <- model.matrix(model)
-
-  if (missing(indices)) {
-    # select the best model from the full data by BIC
-    sel <- leaps::regsubsets(formula, data=data, ...)
-    bics <- summary(sel)$bic
-    best <- coef(sel, 1:length(bics))[[which.min(bics)]]
-    x.names <- names(best)
-    # fit the best model; intercept is already in X, hence - 1:
-    m.best <- lm(y ~ X[, x.names] - 1) 
-    fit.all <- predict(m.best, newdata=data)
-    return(criterion(y, fit.all)) # return the CV criterion
-  }
-
-  # select the best model omitting the i-th fold (given by indices)
-  sel.i <- leaps::regsubsets(formula, data[-indices, ], ...)
-  bics.i <- summary(sel.i)$bic
-  best.i <- coef(sel.i, 1:length(bics.i))[[which.min(bics.i)]]
-  x.names.i <- names(best.i)
-  m.best.i <- lm(y[-indices] ~ X[-indices, x.names.i] - 1)
-              # predict() doesn't work here:
-  fit.all.i <- as.vector(X[, x.names.i] %*% coef(m.best.i))
-  fit.i <- fit.all.i[indices]
-  # return the fitted values for i-th fold, CV criterion for all cases, 
-  #   and the regression coefficients
-  list(fit.i=fit.i, # fitted values for i-th fold
-       crit.all.i=criterion(y, fit.all.i), # CV crit for all cases
-       coefficients = if (save.coef){ # regression coefficients
-         coefs <- coef(m.best.i)
-
-         # fix coefficient names
-         names(coefs) <- sub("X\\[-indices, x.names.i\\]", "",
-                             names(coefs))
-
-         coefs
-       }  else {
-         NULL
-       }
-  )
-}
-```
-A slightly tricky point is that because of scoping issues, `predict()` doesn't work with the model fit omitting the $i$th fold, and so the fitted values for all cases are computed directly as $\widehat{\mathbf{y}}_{-i} = \mathbf{X} \mathbf{b}_{-i}$, where $\mathbf{X}$ is the model-matrix for all of the cases, and $\mathbf{b}_{-i}$ is the vector of least-squares coefficients for the selected model with the $i$th fold omitted.
-
-Additionally, the command `lm(y[-indices] ~ X[-indices, x.names.i] - 1)`, which is the selected model with the $i$th fold deleted, produces awkward coefficient names like `"X[-indices, x.names.i]Infant.Mortality"`.
-Purely for aesthetic reasons, the command `sub("X\\[-indices, x.names.i\\]", "", names(coefs))` fixes these awkward names, removing the extraneous text, `"X[-indices, x.names.i]"`.
-
-Applying `selectSubsets()` to the full data produces the full-data cross-validated RMSE (which we obtained previously):
-%%```{r test-selectSubsets}
-%%selectSubsets(model=m.swiss)
-%%```
-Similarly, applying the function to an imaginary "fold" of 5 cases returns the RMSE for the cases in the fold, based on the model selected and fit to the cases omitting the fold; the RMSE for all of the cases, based on the same model; and the coefficients of the selected model, which includes 4 or the 5 predictors (and the intercept):
-%%```{r test-selectSubsets-fold}
-%%selectSubsets(model=m.swiss, indices=seq(5, 45, by=10))
-%%```
-
-Then, using `selectSubsets()` in cross-validation, we get:
-```{r cvSelect-selectSubsets}
-(cv.swiss <- cvSelect(selectSubsets, model=m.swiss,
-                      data=swiss, seed=8433)) # use same folds
-```
-Cross-validation shows that model selection exacts a penalty in RMSE.
-Examining the models selected for the 10 folds reveals that there is some uncertainty in identifying the predictors in the "best" model, with `Agriculture` sometimes appearing and sometimes not:
-```{r best-models-by-folds}
-compareFolds(cv.swiss)
-```
-
-# Computational notes
-
-## Efficient computations for linear and generalized linear models
-
-The most straightforward way to implement cross-validation in R for statistical modeling functions that are written in the canonical manner is to use `update()` to refit the model with each fold removed.
-This is the approach taken in the default method for `cv()`, and it is appropriate if the cases are independently sampled.
-Refitting the model in this manner for each fold is generally feasible when the number of folds in modest, but can be prohibitively costly for leave-one-out cross-validation when the number of cases is large.
-
-The `"lm"` and `"glm"` methods for `cv()` take advantage of computational efficiencies by avoiding refitting the model with each fold removed.
-Consider, in particular, the weighted linear model $\mathbf{y}_{n \times 1} = \mathbf{X}_{n \times p}\boldsymbol{\beta}_{p \times 1} + \boldsymbol{\varepsilon}_{n \times 1}$, where $\boldsymbol{\varepsilon} \sim \mathbf{N}_n \left(\mathbf{0}, \sigma^2 \mathbf{W}^{-1}_{n \times n}\right)$.
-Here, $\mathbf{y}$ is the response vector,  $\mathbf{X}$ the model matrix, and $\boldsymbol{\varepsilon}$ the error vector, each for $n$ cases, and $\boldsymbol{\beta}$ is the vector of $p$ population regression coefficients.
-The errors are assumed to be multivariately normally distributed with 0 means and covariance matrix $\sigma^2 \mathbf{W}^{-1}$, where $\mathbf{W} = \mathrm{diag}(w_i)$ is a diagonal matrix of inverse-variance weights.
-For the linear model with constant error variance, the weight matrix is taken to be $\mathbf{W} = \mathbf{I}_n$, the order-$n$ identity matrix.
-
-The weighted-least-squares (WLS) estimator of $\boldsymbol{\beta}$ is [see, e.g., @Fox:2016, Sec. 12.2.2] [^WLS]
-$$
-\mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} 
-  \mathbf{X}^T \mathbf{W} \mathbf{y}
-$$
-
-[^WLS]: This is a definitional formula, which assumes that the model matrix $\mathbf{X}$ is of full column rank, and which can be subject to numerical instability when $\mathbf{X}$ is ill-conditioned.
-`lm()` uses the singular-value decomposition of the model matrix to obtain computationally more stable results.
-
-Fitted values are then $\widehat{\mathbf{y}} = \mathbf{X}\mathbf{b}_{\mathrm{WLS}}$.
-
-The LOO fitted value for the $i$th case can be efficiently computed by $\widehat{y}_{-i} = y_i - e_i/(1 - h_i)$ where $h_i = \mathbf{x}^T_i \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} \mathbf{x}_i$ (the so-called "hatvalue").
-Here, $\mathbf{x}^T_i$ is the $i$th row of $\mathbf{X}$, and $\mathbf{x}_i$ is the $i$th row written as a column vector.
-This approach can break down when one or more hatvalues are equal to 1, in which case the formula for $\widehat{y}_{-i}$ requires division by 0.
-
-To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identify [@Wikipedia-Woodbury:2023],
-$$
-\left(\mathbf{A}_{m \times m} + \mathbf{U}_{m \times k} 
-\mathbf{C}_{k \times k} \mathbf{V}_{k \times m} \right)^{-1} = \mathbf{A}^{-1} - \mathbf{A}^{-1}\mathbf{U} \left(\mathbf{C}^{-1} + 
-\mathbf{VA}^{-1}\mathbf{U} \right)^{-1} \mathbf{VA}^{-1}
-$$
-where $\mathbf{A}$ is a nonsingular order-$n$ matrix. We apply this result by letting
-\begin{align*}
-	\mathbf{A} &= \mathbf{X}^T \mathbf{W} \mathbf{X} \\
-	\mathbf{U} &= \mathbf{X}_\mathbf{j}^T \\
-	\mathbf{V} &= - \mathbf{X}_\mathbf{j} \\
-	\mathbf{C} &= \mathbf{W}_\mathbf{j} \\
-\end{align*}
-where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jm})^T$ represents the vector of indices for the cases in the $j$th fold, $j = 1, \ldots, k$.
-The negative sign in $\mathbf{V} = - \mathbf{X}_\mathbf{j}$ reflects the *removal*, rather than addition, of the cases in $\mathbf{j}$.
-
-Applying the Woodbury identity isn't quite as fast as using the hatvalues, but it is generally much faster than refitting the model.
-A disadvantage of the Woodbury identity, however, is that it entails explicit matrix inversion and thus may be numerically unstable.
-The inverse of $\mathbf{A} = \mathbf{X}^T \mathbf{W} \mathbf{X}$ is available directly in the `"lm"` object, but the second term on the right-hand side of the Woodbury identity requires a matrix inversion with each fold deleted.
-(In contrast, the inverse of each $\mathbf{C} = \mathbf{W}_\mathbf{j}$ is straightforward because $\mathbf{W}$ is diagonal.)
-
-The Woodbury identity also requires that the model matrix be of full rank.
-We impose that restriction in our code by removing redundant regressors from the model matrix for all of the cases, but that doesn't preclude rank deficiency from surfacing when a fold is removed.
-Rank deficiency of $\mathbf{X}$ doesn't disqualify cross-validation because all we need are fitted values under the estimated model.
-
-`glm()` computes the maximum-likelihood estimates for a generalized linear model by iterated weighted least squares [see, e.g., @FoxWeisberg:2019, Sec.
-6.12].
-The last iteration is therefore just a WLS fit of the "working response" on the model matrix using "working weights." Both the working weights and the working response at convergence are available from the information in the object returned by `glm()`.
-
-We then treat re-estimation of the model with a case or cases deleted as a WLS problem, using the hatvalues or the Woodbury matrix identity.
-The resulting fitted values for the deleted fold aren't exact---that is, except for the Gaussian family, the result isn't identical to what we would obtain by literally refitting the model---but in our (limited) experience, the approximation is very good, especially for LOO CV, which is when we would be most tempted to use it.
-Nevertheless, because these results are approximate, the default for the `"glm"` `cv()` method is to perform the exact computation, which entails refitting the model with each fold omitted.
-
-## Computation of the bias-corrected CV criterion and confidence intervals
-
-Let $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ represent a cross-validation cost criterion, such as mean-squared error, computed for all of the $n$ values of the response $\mathbf{y}$ based on fitted values $\widehat{\mathbf{y}}$ from the model fit to all of the data.
-We require that $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ is the mean of casewise components, that is, $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n\mathrm{cv}(y_i, \widehat{y}_i)$.[^contrast-function] For example, $\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$.
-
-[^contrast-function]: @ArlotCelisse:2010 term the casewise loss, $\mathrm{cv}(y_i, \widehat{y}_i)$, the "contrast function."
-
-We divide the $n$ cases into $k$ folds of approximately $n_j \approx n/k$ cases each, where $n = \sum n_j$. As above, let $\mathbf{j}$ denote the indices of the cases in the $j$th fold.
-
-Now define $\mathrm{CV}_j = \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}^{(j)})$.
-The superscript $(j)$ on $\widehat{\mathbf{y}}^{(j)}$ represents fitted values computed  for all of the cases from the model with fold $j$ omitted.
-Let $\widehat{\mathbf{y}}^{(-i)}$ represent the vector of fitted values for all $n$ cases where the fitted value for the $i$th case is computed from the model fit with the fold including the $i$th case omitted (i.e., fold $j$ for which $i \in \mathbf{j}$).
-
-Then the cross-validation criterion is just $\mathrm{CV} = \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}^{(-i)})$.
-Following @DavisonHinkley:1997[pp. 293--295], the bias-adjusted cross-validation criterion is
-$$
-\mathrm{CV}_{\mathrm{adj}} = \mathrm{CV} + \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) - \frac{1}{n} \sum_{j=1}^{k} n_j \mathrm{CV}_j
-$$
-
-We compute the standard error of CV as 
-$$
-\mathrm{SE}(\mathrm{CV}) = \frac{1}{\sqrt n} \sqrt{ \frac{\sum_{i=1}^n \left[ \mathrm{cv}(y_i, \widehat{y}_i^{(-i)} ) - \mathrm{CV} \right]^2 }{n - 1} }
-$$
-that is, as the standard deviation of the casewise components of CV divided by the square-root of the number of cases.
-
-We then use $\mathrm{SE}(\mathrm{CV})$ to construct a $100 \times (1 - \alpha)$% confidence interval around the *adjusted* CV estimate of error:
-$$
-\left[ \mathrm{CV}_{\mathrm{adj}} - z_{1 - \alpha/2}\mathrm{SE}(\mathrm{CV}), \mathrm{CV}_{\mathrm{adj}} + z_{1 - \alpha/2}\mathrm{SE}(\mathrm{CV})  \right]
-$$
-where $z_{1 - \alpha/2}$ is the $1 - \alpha/2$ quantile of the standard-normal distribution (e.g, $z \approx 1.96$ for a 95% confidence interval, for which $1 - \alpha/2 = .975$).
-
-@BatesHastieTibshirani:2023 show that the coverage of this confidence interval is poor for small samples, and they suggest a much more computationally intensive procedure, called *nested cross-validation*, to compute better estimates of error and confidence intervals with better coverage for small samples.
-We may implement Bates et al.'s approach in a later release of the **cv** package.
-At present we use the confidence interval above for sufficiently large $n$, which, based on Bates et al.'s results, we take by default to be $n \ge 400$.
-
-# Why the complement of AUC isn't a casewise CV criterion
 
@@
 Consequently, the widely employed AUC measure of fit for binary regression cannot in general be used for a casewise cross-validation criterion.
-
-<!-- # Introduction -->
-
-<!-- This template demonstrates some of the basic LaTeX that you need to know to create a JSS article. -->
-
-<!-- ## Code formatting -->
-
-<!-- In general, don't use Markdown, but use the more precise LaTeX commands instead: -->
-
-<!-- * \proglang{Java} -->
-<!-- * \pkg{plyr} -->
-
-<!-- One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax). -->
-
-<!-- If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute.
-You can also provide a custom identifier if necessary.
-See the header of Section \ref{r-code} for example.
--->
-
-<!-- # \proglang{R} code {short-title="R code" #r-code} -->
-
-<!-- Can be inserted in regular R markdown blocks. -->
-
-<!-- ```{r} -->
-<!-- x <- 1:10 -->
-<!-- x -->
-<!-- ``` -->
-
-<!-- ## Features specific to \pkg{rticles} {short-title="Features specific to rticles"} -->
-
-<!-- * Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter).
-This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
--->
-<!-- * Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format. -->

### 2024-07-01 scratch/notes_georges/AUC/AUC_not_casewise.Rmd
clean up scratch

---
title: 'Can AUC be expressed using a case-wise criterion? No'
output: html_document
---

Consider calculating AUC for folds in which a
validation set contains $n$ observations.
To calculate AUC in the validation set, we need
the vector of prediction criteria,
$\mathbf{\hat{y}} = (\hat{y}_1, ..., \hat{y}_1)'$,
and the vector of observed responses in the validation set,
$\mathbf{y} = (y_1, ..., y_n)'$,
with 
$y_i \in \{0,1\}, \; i = 1,...,n$.

To construct the ROC curve, only the ordering of
the values in 
$\mathbf{\hat{y}}$ is needed.
Thus, assuming that there are no ties and reordering
observations if necessary, we can set 
$\mathbf{\hat{y}} = (1,2,...,n)'$.

If the AUC can be expressed as the case-wise mean or sum of a function 
$cv(\hat{y}_i,y_i)$
where $cv: \{1,2,...,n\}\times\{0,1\} \rightarrow [0,1]$, then
\begin{equation}
\label{eq:cw}
\tag{1}
\sum_{i=1}^n cv(\hat{y}_i,y_i) = AUC(\mathbf{\hat{y}},\mathbf{y})
\end{equation}
must hold for all $2^n$ possibilities for $\mathbf{y} = (y_1,...,y_n)$.
If all $y$s have the same value, either 1s or 0s, the definition
of AUC is ambiguous. It could be considered to be undefined, or
it could be set to 0 if all $y$s are 0, and to 1 if all $y$s are 1.
If AUC is considered to be undefined in these cases, we have
$2^n - 2$
admissible values for $\mathbf{y}$.

Thus, equation (\ref{eq:cw}) produces either $2^n$ or $2^n-2$ constraints.
There are, however, only $2n$ possible values for the $cv$ function.
Equation (\ref{eq:cw})
could, nevertheless, have consistent solutions.  We therefore need to verify
whether there is a value of $n$ for which (\ref{eq:cw}) has no consistent solution
for all admissible values of $\mathbf{y}$.
In that eventuality, we will have shown that AUC cannot, in general,
be expressed through a case-wise criterion.

If $n=3$, we show below that (\ref{eq:cw}) has no consistent solution
if we include all possibilities for $\mathbf{y}$, but does
if we exclude cases when all $y$s have the same value.
If $n=4$, we show that there no consistent solutions
in either case.

The following function computes AUC as function of $\mathbf{\hat{y}}$
and $\mathbf{y}$

```{r}
AUC <- function(y, yhat = seq_along(y)) {
  # AUC:
  # y is a vector of 0s and 1s
  # yhat is a predictor of y, only used for its order
  #    i.e. the result is invariant under monotonic transformations of yhat
  #    If there are ties in yhat, this function splits the ties
  #
  s <- sum(y)
  n <- length(y)
  if(s == 0) return(0)
  if(s == n) return(1)
  a <- sum(outer(yhat[y==0],yhat[y==1], `<`)) + sum(outer(yhat[y==0],yhat[y==1], `==`))/2
  a / (s*(n-s))
}
```

We generate all possible $\mathbf{y}$s of length $n$ as rows of the matrix
$\mathbf{Y}$ with the following function:

```{r}
Ymat <- function(n, exclude_identical = FALSE) {
  stopifnot(n > 0 && round(n) == n)    # n must be a positive integer
  ret <- sapply(0:(2^n - 1),
                function(x) as.integer(intToBits(x)) )[1:n, ]
  ret <- if (is.matrix(ret)) t(ret) else matrix(ret)
  colnames(ret) <- paste0('y', 1:ncol(ret))
  if (exclude_identical) ret[-c(1, nrow(ret)), ] else ret
}
```

For $n=3$:

```{r}
Ymat(3)
```

If we exclude $\mathbf{y}$s with identical values:

```{r}
Ymat(3, TRUE)
```

The following shows $\mathbf{Y}$ with corresponding values of AUC:

```{r}
cbind(Ymat(3), AUC = apply(Ymat(3), 1, AUC))
```

The values of $cv(\hat{y}_i, y_i)$ that express
AUC as a sum of case-wise values are solutions of equation (\ref{eq:cw}),
which can be expressed as solutions of the following linear equation:
\begin{equation}
\label{eq:lin}
\tag{2}
(\mathbf{U} -\mathbf{Y}) \mathbf{c}_0 + \mathbf{Y} \mathbf{c}_1
=
[\mathbf{U} -\mathbf{Y} \quad \mathbf{Y}]
\begin{bmatrix}
\mathbf{c}_0 \\ \mathbf{c}_1
\end{bmatrix}
= AUC(\mathbf{\hat{Y}},\mathbf{Y})
\end{equation}

where $\mathbf{U}$ is a matrix of 1's conformable with $\mathbf{Y}$,
$\mathbf{c}_0 = (cv(1,0), c(2,0), ..., cv(n,0))'$,
$\mathbf{c}_1 = (cv(1,1), c(2,1), ..., cv(n,1))'$,
$[\mathbf{U} -\mathbf{Y} \quad \mathbf{Y}]$ and
$\begin{bmatrix}\begin{aligned}
\mathbf{c}_0 \\ \mathbf{c}_1
\end{aligned}
\end{bmatrix}$ are partitioned matrices, and
$\mathbf{\hat{Y}}$ is a matrix each of whose rows consists of the integers
1 to $n$.

We can test whether equation (\ref{eq:lin}) has a solution for any given $n$
by considering whether the residuals of the associated linear model,
using the design matrix
$[\mathbf{U} -\mathbf{Y} \quad \mathbf{Y}]$ to
predict the outcome
$AUC(\mathbf{\hat{Y}},\mathbf{Y})$, are zero.

```{r}
resids <- function(n, exclude_identical = FALSE, tol = 1e-8) {
  # ss
  Y <- Ymat(n, exclude_identical = exclude_identical)
  AUC <- apply(Y, 1, AUC)
  X <- cbind(1-Y, Y)
  opts <- options(warn = -1)
  on.exit(options(opts))
  fit <- lsfit(X, AUC, intercept = FALSE)
  ret <- sum(abs(fit$residuals))
  if(ret < tol){
    solution <- fit$coef
    names(solution) <- paste0('c(', c(1:n,1:n), ',', rep(0:1, each = n),')')
    attr(ret, 'solution') <- zapsmall(solution)
  }
  ret
}
```

The case $n=3$, excluding identical $y$s, has a solution:

```{r}
resids(3, exclude_identical = TRUE)
```

But, if identical $y$s are included, the equation is not consistent:

```{r}
resids(3, exclude_identical = FALSE)
```

For $n=4$, there are no solutions in either case:

```{r}
resids(4, exclude_identical = TRUE)
resids(4, exclude_identical = FALSE)
```

We have demonstated that a widely used criterion, AUC, cannot, in general, be cross-validated
using a case-wise approach.

### 2024-07-01 scratch/notes_georges/Bias/bias.R

@@ -1,0 +1,410 @@
+#'
+#' Experimenting with CV bias and variance
+#'
+
+{
+  # Initialize
+
+  library(glmnet)
+  library(cv)
+  library(insight)
+  library(latticeExtra)
+  library(spida2)
+  library(car)
+
+  getResponse <- insight::get_response
+  spida2::setwd_here()
+}
+
+
+{
+  #
+  # Supernova data from Efron and Hastie (2013) Computer Age Statistical Inference
+  #
+
+  if(!file.exists('supernova.csv')){
+    # import data
+    supernova.in <- read.table("http://hastie.su.domains/CASI_files/DATA/supernova.txt", header=TRUE)
+    write.csv(supernova.in, file = 'supernova.csv')
+  }
+  supernova <- read.csv('supernova.csv', row.names = 1)
+}
+
+fit1 <- lm(Magnitude ~ E1, supernova)
+fit2 <- lm(Magnitude ~ E1 + E2, supernova)
+full <- lm(Magnitude ~ ., supernova)    # E1 to E10
+
+anova(fit1, fit2, full)
+
+summary(fit1)
+summary(fit2)
+summary(full)
+
+xyplot(Magnitude~predict(full), supernova, pch = 16) +
+  layer(panel.dell(...)) +
+  layer(panel.lmline(..., lty = 1)) +
+  layer(panel.abline(a=0,b=1, lty = 2, lwd = 3, col = 'red'))
+
+cvk <- cv(models(fit1, fit2, full), data = supernova, k = 10)
+cvloo <- cv(models(fit1, fit2, full), data = supernova, k = 'loo')
+cvkreps <- cv(models(fit1, fit2, full), data = supernova, k = 10, reps = 100)
+plot(cvk)
+plot(cvloo)
+plot(cvkreps)
+
+
+{
+
+  #
+  # Simulate N samples using the supernova data
+  #
+  N <- 1000
+  sims <- lapply(1:N, \(i) supernova[sample(1:39, replace = T),])
+
+  mods <- lapply(
+    sims,
+    function(d) {
+      list(
+        fit1 = lm(Magnitude ~ E1, d),
+        fit2 = lm(Magnitude ~ E1 + E2, d),
+        full = lm(Magnitude ~ ., d)
+      )
+    }
+  )
+
+  kfold.l <- lapply(mods, lapply, cv)
+
[…]

### 2024-07-01 scratch/notes_georges/Bias/trinity.R

@@ -1,0 +1,75 @@
+#
+#
+# Using numDeriv to obtain numerical Rao(score), Wald and Wilks(LRT) tests
+# from the likelihood function for the binomial
+# with two parametrizations: probability and log odds
+#
+#
+
+
+logLikp <- function(n,Y) {
+  #
+  # creates the log likelihood for a binomial with n observations and Y successes
+  # as a function of probability of a success
+  #
+  ret <- function(p) {
+    Y*log(p) + (n-Y)*log(1-p)
+  }
+  phat <- Y/n
+  attr(ret, 'mle') <- phat
+  ret
+}
+logLiklo <- function(n,Y) {
+  #
+  # creates the log likelihood for a binomial with n observations and Y successes
+  # as a function of the log odds of success
+  #
+  ret <- function(lo) {
+    Y*log(1/(1+exp(-lo))) + (n-Y)*log(1/(1+exp(lo)))
+  }
+  phat <- Y/n
+  attr(ret, 'mle') <- log(phat/(1-phat))
+  ret
+}
+
+tests <- function(llik, h0) {
+  mle <- attr(llik,'mle')
+  c(
+    wald = -(mle - h0)^2 * hessian(llik, mle),
+    wald2 = -(mle - h0)^2 * hessian(llik, h0),
+    rao = -(grad(llik, h0)^2) / hessian(llik, h0),
+    wilks = 2 * (llik(mle) - llik(h0))
+  )
+}
+
+
+
+llp <- logLikp(20,8)
+lllo <- logLiklo(20,8)
+
+tests(llp, .4)                   # should be 0
+tests(lllo, log(.4/((1 -.4))))   # should be 0
+
+# Rao and wilks should be the same except for numerical errors in differentiation
+tests(llp, .5)                   # should be 0
+tests(lllo, log(.5/((1 -.5))))   # should be 0
+
+tests(llp, .3)                   # should be 0
+tests(lllo, log(.3/((1 -.3))))   # should be 0
+
+tests(llp, .5)
+tests(llp, .3)
+tests(lllo, log(.4/(1-.4)))
+tests(lllo, log(.5/(1-.5)))
+tests(lllo, log(.3/(1-.3)))
+
+library(latticeExtra)
+plot(seq(0,1,.01), sapply(seq(0,1,.01), function(x) llp(x)))
+
+h0s <- seq(.001,.999,.001)
+
+ts <- lapply(h0s, function(h0) tests(llp, h0))
+
+
+
+help(p=numDeriv)

### 2024-07-03 scratch/notes_john/JSS-article/JSS-article-3.Rmd
fix discussion of bias and variance in JSS paper draft

@@
 Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of prediction error.
-The $n$ regression models are highly statistical dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has relatively large variance.
-In contrast, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) are somewhat biased but have smaller variance.
-It is also possible to correct $k$-fold CV for bias (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+The $n$ regression models are highly statistically dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has larger variance than if the predictions from the models fit to the $n$ data sets were independent.
+
+Because predictions are based on smaller data sets, each of size approximately $n - n/k$, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) is more biased than estimated prediction error for LOO CV.
+It is possible, however, to correct $k$-fold CV for bias (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+
+The relative *variance* of prediction error for LOO CV and $k$-fold CV (with $k < n$) is more complicated: Because the overlap between the data sets with each fold omitted is smaller for $k$-fold CV than for LOO CV, the dependencies among the predictions are smaller for the former than for the latter, tending to produce smaller variance in prediction error for $k$-fold CV.
+In contrast, there are factors that tend to inflate the variance of prediction error in $k$-fold CV, including the reduced size of the data sets with each fold omitted and the randomness induced by the selection of folds---in LOO CV the folds are not random.
 

### 2024-07-03 scratch/notes_john/JSS-article/JSS-article-3.tex
fix discussion of bias and variance in JSS paper draft

@@
 produces a nearly unbiased estimate of prediction error. The \(n\)
-regression models are highly statistical dependent, however, based as
+regression models are highly statistically dependent, however, based as
 they are on nearly the same data, and so the resulting estimate of
-prediction error has relatively large variance. In contrast, estimated
-prediction error for \(k\)-fold CV with \(k = 5\) or \(10\) (commonly
-employed choices) are somewhat biased but have smaller variance. It is
-also possible to correct \(k\)-fold CV for bias (see Section
+prediction error has larger variance than if the predictions from the
+models fit to the \(n\) data sets were independent.
+
+Because predictions are based on smaller data sets, each of size
+approximately \(n - n/k\), estimated prediction error for \(k\)-fold CV
+with \(k = 5\) or \(10\) (commonly employed choices) is more biased than
+estimated prediction error for LOO CV. It is possible, however, to
+correct \(k\)-fold CV for bias (see Section
 \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+
+The relative \emph{variance} of prediction error for LOO CV and
+\(k\)-fold CV (with \(k < n\)) is more complicated: Because the overlap
+between the data sets with each fold omitted is smaller for \(k\)-fold
+CV than for LOO CV, the dependencies among the predictions are smaller
+for the former than for the latter, tending to produce smaller variance
+in prediction error for \(k\)-fold CV. In contrast, there are factors
+that tend to inflate the variance of prediction error in \(k\)-fold CV,
+including the reduced size of the data sets with each fold omitted and
+the randomness induced by the selection of folds---in LOO CV the folds
+are not random.
 
@@
 \begin{CodeOutput}
-R RNG seed set to 969978
+R RNG seed set to 431467
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.18252
-bias-adjusted cross-validation criterion = 19.17193
-95% CI for bias-adjusted CV criterion = (15.72359, 22.62027)
+cross-validation criterion = 19.192
+bias-adjusted cross-validation criterion = 19.18107
+95% CI for bias-adjusted CV criterion = (15.71858, 22.64356)
 full-sample criterion = 18.98477 
@@
 }
-<bytecode: 0x13b56c108>
+<bytecode: 0x13fb26ea8>
 <environment: namespace:cv>

### 2024-07-04 scratch/notes_georges/JSS-article/JSS-article-3-GM.Rmd
incorporate new text in the JSS article for the section on mixed models from the revised vignette.

--- 2024-06-17 scratch/notes_john/JSS-article/JSS-article-3.Rmd
@@
 palette(car::carPalette())
+library(car)     # included this to avoid scatterplotMatrix() error
 ```
@@
 
-In this section, we introduce an artificial data set that exemplifies aspects of cross-validation particular to hierarchical models.
-Using this data set, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model.
+We introduce four artificial data sets that exemplify aspects of cross-validation particular to hierarchical models.
+Using these data sets, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model.
 Furthermore, commonly used measures of fit, such as mean-squared error, do not necessarily become smaller as models become larger, even when the models are nested, and even when the measure of fit is computed for the whole data set.
 
-Consider a researcher studying improvement in a skill, singing, for example, among students enrolled in a four-year voice program at a music conservatory.
-The plan is to measure each student's skill level at the beginning of the program and every year thereafter until the end of the program, resulting in 5 annual measurements for each student.
-It turns out that singing appeals to students of all ages, and students enrolling in the program range in age from 20 to 70.
-Moreover, participants' untrained singing skill is similar at all ages, as is their rate of progress with training.
-All students complete the four-year program.
-
-The researcher, who has more expertise in singing than in modeling, decides to model the response, $y$, singing skill, as a function of age, $x$, reasoning that students get older during their stay in the program, and (incorrectly) that age can serve as a proxy for elapsed time.
-The researcher knows that a mixed model should be used to account for clustering due to the expected similarity of measurements taken from each student.
-
-We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^1]
-
-[^1]: We invite the interested reader to experiment with varying the parameters of our example.
-
-```{r data}
+The four datasets differ in the relative magnitude of between-cluster variance compared with within-cluster variance.
+They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.
+
+For these analyses, we use the `glmmTMB()` function in the **glmmTMB** package [@BrooksEtAl] because, in our experience, it is more likely to converge than functions in the **nlme** and the **lme4** packages for models with low between-cluster variance.
+
+Consider a researcher studying the effect of the dosage of a drug on the severity of symptoms for a hypothetical disease.
+The researcher has longitudinal data on 20 patients, each of whom was observed on five occasions in which patients received different dosages of the drug.
+The data are observational, with dosages prescribed by the patients' physicians, so that patients who were more severely affected by the disease received higher dosages of the drug.
+
+Our four contrived data sets (see below) illustrate possible results for data obtained in such a scenario.
+The relative configuration of dosages `x` and symptoms `y` are identical within patients in each of the four datasets.
+Within patients, higher dosages are generally associated with a reduction in symptoms.
+
+Between patients, however, higher dosages are associated with higher levels of symptoms.
+A plausible mechanism is a reversal of causality: Within patients, higher dosages alleviate symptoms, but, between patients, higher morbidity causes the prescription of higher dosages.
+
+The four data sets differ in the between-patient variance of patient centroids from a common between-patients regression line.
+The data sets exhibit a progression from low to high variance around the common regression line.
+
+We start by generating a data set to serve as a common template for the four sample data sets.
+We then apply different multipliers to the between-patient variation using parameters consistent with the description above to highlight the issues that arise in cross-validating mixed-effects models:[^data-template]
+
+[^data-template]: We invite the interested reader to experiment with varying the parameters of our example.
+
+```{r include=FALSE, echo=FALSE}
+library("glmmTMB") # necessary for some reason to knit vignette in RStudio, harmless otherwise
+```
+```{r parameters}
 # Parameters:
-set.seed(9693) 
-Nb <- 100     # number of groups
-Nw <- 5       # number of individuals within groups
-Bb <- 0       # between-group regression coefficient on group mean
-SDre <- 2.0   # between-group SD of random level relative to group mean of x
-SDwithin <- 0.5  # within group SD
-Bw <- 1          # within group effect of x
-Ay <- 10         # intercept for response
-Ax <- 20         # starting level of x
-Nx <- Nw*10      # number of distinct x values
-
-Data <- data.frame(
-  group = factor(rep(1:Nb, each=Nw)),
-  x = Ax + rep(1:Nx, length.out = Nw*Nb)
-) |> within ({
-  xm  <- ave(x, group, FUN = mean) # within-group mean
-  y <- Ay +
+Nb <- 20     # number of patients
+Nw <- 5      # number of occasions for each patient
+Bb <- 1.0    # between-patient regression coefficient on patient means
+Bw <- -0.5   # within-patient effect of x
+
+SD_between <- c(0, 5, 6, 8)               # SD between patients
+SD_within <- rep(2.5, length(SD_between)) # SD within patients
+
+Nv <- length(SD_within)       # number of variance profiles
+SD_ratio <- paste0('SD ratio = ', SD_between,' / ',SD_within)
+SD_ratio <- factor(SD_ratio, levels = SD_ratio)
+
+set.seed(833885)
+
+Data_template <- expand.grid(patient = 1:Nb, obs = 1:Nw) |>
+  within({
+    xw <- seq(-2, 2, length.out = Nw)[obs]
+    x <- patient + xw
+    xm  <- ave(x, patient)   # within-patient mean
+    # Scaled random error within each SD_ratio_i group
+    re_std <- scale(resid(lm(rnorm(Nb*Nw) ~ x)))
+    re_between <- ave(re_std, patient)
+    re_within <- re_std - re_between
+    re_between <- scale(re_between)/sqrt(Nw)
+    re_within <- scale(re_within)
+  })
+Data <- do.call(
+  rbind,
+  lapply(
+    1:Nv,
+    function(i) {
+      cbind(Data_template, SD_ratio_i = i)
+    }
+  )
+)
+Data <- within(
+  Data,
+  {
+    SD_within_ <- SD_within[SD_ratio_i]
+    SD_between_ <- SD_between[SD_ratio_i]
+    SD_ratio <- SD_ratio[SD_ratio_i]
+    y <- 10 +
       Bb * xm +                  # contextual effect
-    Bw * (x - xm) +              # within-group effect
-    rnorm(Nb, sd=SDre)[group] +  # random level by group
-    rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
-})
-```
-
-Figure \ref{fig:plot1} (a) shows a scatterplot the data for a representative group of 10 (without loss of generality, the first 10) of the 100 students, displaying the 95% concentration ellipse for each cluster.
-
-```{r plot1, echo=FALSE, results=FALSE}
-#| fig.height = 4,
-#| fig.cap = "(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) predictions from several mixed models fit to the data",
-#| fig.subcap=rep("", 5),
-#| fig.ncol = 2, out.width = "45%", fig.align = "center"
+      Bw * (x - xm) +            # within-patient effect
+      SD_within_ * re_within +   # within patient random effect
+      SD_between_ * re_between   # adjustment to between patient random effect
+  }
+)
+```
+Figure \ref{fig:plot1} display scatterplots of each data set showing estimated 50% concentration ellipses for each cluster.[^lattice]
+
+[^lattice]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.
+
+```{r plot1, echo=FALSE}
+#| out.width = "100%",
+#| fig.height = 3.5,
+#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance.
+The labels above each panel show the between-patient SD/within-patient SD.
+The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel."
 library("lattice")
 library("latticeExtra")
-library("lme4", quietly=TRUE)
-plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
-               ylim=c(4, 16),
-               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
-    layer(panel.ellipse(..., center.cex=0))
-plot # display graph
-
-mod.0 <- lmer(y ~ 1 + (1 | group), Data)
-mod.1 <- lmer(y ~ x + (1 | group), Data)
-mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
-mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
-
-Data <- within(Data, {
-  fit_mod0.fe <- predict(mod.0, re.form = ~ 0) # fixed effects only
-  fit_mod0.re <- predict(mod.0) # fixed and random effects (BLUPs)
-  fit_mod1.fe <- predict(mod.1, re.form = ~ 0)
-  fit_mod1.re <- predict(mod.1)
-  fit_mod2.fe <- predict(mod.2, re.form = ~ 0)
-  fit_mod2.re <- predict(mod.2)
-  fit_mod3.fe <- predict(mod.3, re.form = ~ 0)
-  fit_mod3.re <- predict(mod.3)
+layer <- latticeExtra::layer # to avert conflicts with ggplot2::layer
+xyplot(y ~ x | SD_ratio, data = Data, group = patient,
+       par.strip.text = list(cex = 0.7),
+       scales = list(x = list(alternating = FALSE)),
+       layout = c(Nv, 1),
+       par.settings = list(superpose.symbol = list(pch = 1, cex = 0.7))) +
+  layer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
+                      level = 0.5),
+        panel.abline(a = 10, b = 1))
+```
+The population from which these datasets are generated has a between-patient effect of dosage of 1 and a within-patient effect of $-0.5$.
+The researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept.
+We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.
+The first three models have sequentially nested fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`.
+The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`.
+We thus fit four models to four datasets for a total of 16 models.
+
+```{r model-fits}
+model.formulas <- c(
+  ' ~ 1'             =  y ~ 1 + (1 | patient),
+  '~ 1 + x'          =  y ~ 1 + x + (1 | patient),
+  '~ 1 + x + xm'     =  y ~ 1 + x + xm + (1 | patient),
+  '~ 1 + I(x - xm)'  =  y ~ 1 + I(x - xm) + (1 | patient)
+)
+fits <- lapply(split(Data, ~ SD_ratio),
+               function(d) {
+                 lapply(model.formulas, function(form) {
+                   glmmTMB(form, data = d)
                  })
-
-Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
-              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
-Data_long$id <- 1:nrow(Data_long)
-Data_long <- reshape(Data_long, direction = "long", sep = "_", 
-              timevar = "modelcode",  varying = grep("_", names(Data_long)))
-Data_long$model <- factor(
-  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
-  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
+               })
+```
+We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or "BLUPs"---as would be used for cross-validation based on cases (i.e., occasions within patients):
+```{r predict}
+pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  
+pred.BLUPs <- lapply(fits, lapply, predict)
+```
+<!-- We then prepare the data for plotting: -->
+
+```{r data-predictions,include=FALSE}
+Dataf <- lapply(split(Data, ~ SD_ratio),
+                function(d) {
+                  lapply(names(model.formulas), 
+                         function(form) cbind(d, formula = form))
+                }) |> 
+  lapply(function(dlist) do.call(rbind, dlist)) |> 
+  do.call(rbind, args = _) |> 
+  within(
+    {
+      pred.fixed <- unlist(pred.fixed)
+      pred.BLUPs <- unlist(pred.BLUPs)
+      panel <- factor(formula, levels = c(names(model.formulas), 'data'))
+    }
   )
-
-(plot +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
-         groups=group, type="l", lwd=2) +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
-         groups=group, type="l", lwd=2, lty=2)
-) |> update(
-  main="Model: y ~ 1 + (1 | group)",
-  key=list(
-    corner=c(0.05, 0.05),
-    text=list(c("fixed effects only","fixed and random")),
-    lines=list(lty=c(1, 2), lwd=2)))
-
-(plot +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
-         groups=group, type="l", lwd=2) +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
-         groups=group, type="l", lwd=2, lty=2)
-) |> update(
-  main="Model: y ~ 1 + x + (1 | group)",
-  ylim=c(-15, 35) #,
-  # key=list(
-  #   corner=c(0.95, 0.05),
-  #   text=list(c("fixed effects only","fixed and random")),
-  #   lines=list(lty=c(1, 3)))
-  )
-
-(plot +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "fe"),
-         groups=group, type="l", lwd=2) +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod2" & effect == "re"),
-         groups=group, type="l", lwd=2, lty=2)
-) |> update(
-  main="Model: y ~ 1 + x + xm + (1 | group)",
-  ylim=c(4, 16) #,
-  # key=list(
-  #   corner=c(0.05, 0.05),
-  #   text=list(c("fixed effects only","fixed and random")),
-  #   lines=list(lty=c(1, 3)))
-  )
-
-(plot +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "fe"),
-         groups=group, type="l", lwd=2) +
-  xyplot(fit ~ x, subset(Data_long, modelcode == "mod3" & effect == "re"),
-         groups=group, type="l", lwd=2, lty=2)
-) |> update(
-  main="Model: y ~ 1 + I(x - xm) + (1 | group)",
-  ylim=c(4, 16) #,
-  # key=list(
-  #   corner=c(0.05, 0.05),
-  #   text=list(c("fixed effects only","fixed and random")),
-  #   lines=list(lty=c(1, 3)))
-  )
-
-```
-
-The between-student effect of age is 0 but the within-student effect is 1.
-Due to the large variation in ages between students, the least-squares regression of singing skill on age (for the 500 observations among all 100 students) produces an estimated slope close to 0 (though with a small $p$-value), because the slope is heavily weighted toward the between-student effect:
-
-```{r}
-summary(lm(y ~ x, data=Data))
-```
-
-We proceed to fit several mixed-effects models to the data, using the `compareCoefs()` function from the \pkg{car} package [@FoxWeisberg:2019] to display the fixed-effect estimates for these models.
-We discuss each of these models below.
-```{r fitted-mixed-models}
-mod.0 <- lmer(y ~ 1 + (1 | group), Data)
-mod.1 <- lmer(y ~ x + (1 | group), Data)
-mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
-mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
-library("car", quietly=TRUE)
-compareCoefs(mod.0, mod.1, mod.2, mod.3)
-```
-
-The initial mixed-effects model, \code{mod.0}, is a simple random-intercepts model.
-We obtain predictions from this model for the fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---the BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students).
-Predictions from \code{mod.0} for the first 10 students are shown in Figure \ref{fig:plot1} (b).
-The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean.
-Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.
-
-Our next model, \code{mod.1}, includes a fixed intercept and the fixed effect of \code{x}, along with a random intercept.
-Predictions from this model appear in Figure \ref{fig:plot1} (c).
-The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, \code{mod.0}.
-We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod.1} to \code{mod.0}, but that cluster-based cross-validation will prefer \code{mod.0} to \code{mod.1}.
-
-Our third model, \code{mod.2}, includes the "contextual effect" of $x$---that is, the cluster mean \code{xm}---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept.
-This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that generated the data once the coefficient of the contextual predictor \code{xm} is set to 0 (as it is in \code{mod.3}, discussed below).
-
-Predictions from \code{mod.2} appear in Figure \ref{fig:plot1} (d).
-Depending on the estimated variance parameters of the model, a mixed model like \code{mod.2} will apply varying degrees of shrinkage to the random-intercept BLUPs that correspond to variation in the heights of the parallel fitted lines for the individual students.
-In our contrived data, \code{mod.2} applies little shrinkage, allowing substantial variability in the heights of the fitted lines, which closely approach the observed values for each student.
-The fit of the mixed model \code{mod.2} is consequently similar to that of a fixed-effects model with age and a categorical predictor for individual students (i.e., treating students as a factor, and not shown here).
-
-The  mixed model \code{mod.2} therefore fits the individual observations well, and we anticipate a favorable assessment using individual-based cross-validation.
-In contrast, the large variability in the BLUPs results in larger residuals for predictions based on fixed effects alone, and so we expect that cluster-based cross-validation won't show an advantage for  \code{mod.2} compared to the smaller \code{mod.0}, which includes only fixed and random intercepts.
-
-Had the mixed model applied considerable shrinkage, then neither cluster-based nor case-based cross-validation would show much improvement over the random-intercept-only model.
-In our experience, the degree of shrinkage does not vary smoothly as parameters are changed but tends to be "all or nothing," and near the tipping point, the behavior of estimates can be affected considerably by the choice of algorithm used to fit the model.
-
-Finally, \code{mod.3} directly estimates the model used to generate the data.
-As mentioned, it is a constrained version of \code{mod.2}, with the coefficient of \code{xm} set to 0, and with \code{x} expressed as a deviation from the cluster mean \code{xm}.
-The predictions from \code{mod.3}, shown in Figure \ref{fig:plot1} (e), are therefore similar to those from \code{mod.2}.
-
-We next carry out case-based cross-validation, which, as we have explained, includes both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which includes fixed effects only.
-In order to reduce between-model random variability in comparisons of models, we apply \code{cv()} to the list of models created by the \code{models()} function (introduced previously), performing cross-validation with the same folds for each model (see Figure \ref{fig:cross-validation-clusters}):
-
-```{r cross-validation-clusters}
-#| fig.height = 5,
-#| fig.subcap=rep("", 2),
-#| fig.ncol = 2, out.width = "45%", fig.align = "center",
-#| fig.cap = "10-fold (a) cluster-based and (b) case-based cross-validation comparing random intercept models with varying fixed effects.
-The error bars show the 95% confidence interval around the CV estimate of the MSE for each model."
-modlist <- models("~ 1"=mod.0, "~ 1 + x"=mod.1, 
-                  "~ 1 + x + xm"=mod.2, "~ 1 + I(x - xm)"=mod.3)
-
-cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
-plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")
-
-cvs_cases <- cv(modlist, data=Data, seed=9693)
-plot(cvs_cases, main="Model Comparison, Case-Based CV")
-```
-
-In summary, \code{mod.1}, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
-\code{mod.2}, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV.
-The data-generating model, \code{mod.3}, which includes the fixed effect of \code{x - xm} in place of separate terms in \code{x} and \code{xm}, isn't distinguishable from \code{mod.2}, which includes \code{x} and \code{xm} separately, even though \code{mod.2} has an unnecessary parameter (recall that the population coefficient of \code{xm} is 0 when \code{x} is expressed as deviations from the contextual mean).
-These conclusions are consistent with our observations based on graphing predictions from the various models in Figure \ref{fig:plot1} (on page \pageref{fig:plot1}), and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
+Data$panel <- factor('data', levels = c(names(model.formulas), 'data'))
+```
+The fixed-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-fixed} along with a summary of the data using 50% concentration ellipses.
+Figure \ref{fig:plot-fits-blups} similarly displays predicted BLUPs.
+```{r plot-fits-fixed,echo=FALSE}
+#| out.width = "100%",
+#| fig.height = 6,
+#| fig.cap = "Fixed-effect predictions using each model applied to data sets with varying between- and within-patient variance ratios.
+The top row shows summaries of the within-patient data using estimated 50% concentration ellipses."
+{
+  xyplot(y ~ x |SD_ratio * panel, Data,
+         groups = patient, type = 'n',    
+         par.strip.text = list(cex = 0.6),
+         scales = list(x = list(alternating = FALSE)),
+         drop.unused.levels = FALSE,
+         ylab = list('fixed-effect predictions',cex=.7)) + 
+    glayer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
+                         level = 0.5),
+           panel.abline(a = 10, b = 1)) +
+    xyplot(pred.fixed  ~ x |SD_ratio * panel, Dataf, type = 'l', 
+           groups = patient,
+           drop.unused.levels = F)
+}|> useOuterStrips()
+```
+<!-- The BLUPs from these models, shown in Figure \ref{fig:plot-fits-blups}, are produced as follows: -->
+```{r plot-fits-blups,echo=FALSE}
+#| out.width = "100%",
+#| fig.height = 6,
+#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient variance ratios.
+The top row shows summaries of the within-patient data using estimated 50% concentration ellipses."
+{   xyplot(y ~ x | SD_ratio * panel, Data,
+         groups = patient, type = 'n',
+         drop.unused.levels = F, 
+         par.strip.text = list(cex = 0.6),
+         scales = list(x = list(alternating = FALSE)),
+         ylab = list('fixed- and random-effect predictions (BLUPS)', cex = 0.7)) +
+    glayer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
+                         level = 0.5),
+           panel.abline(a = 10, b = 1)) +
+    xyplot(pred.BLUPs  ~ x | SD_ratio * panel, Dataf, type = 'l', 
+           groups = patient,
+           drop.unused.levels = F) 
+}|> useOuterStrips()
+```
+Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`.
+The inclusion of a contextual variable in the model corrects this problem.
+
+Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform with the within-patient slopes for the `~ 1 + x` model in the two datasets with the smallest between-patient variances.
+
+For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for datasets with larger between-subject variance.
+With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.
+
+We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, `~ 1`  to the larger `~ 1 + x` model when the between-cluster variance is relatively small, but that cluster-based cross-validation will prefer the latter to the former.
+
+We will discover that case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model for the '5 / 2.5' dataset, but cluster-based cross-validation prefers the latter model to the former.
+The situation is entirely reversed with the '8 / 2.5' dataset.
+
+The third model, `~ 1 + x + xm`, includes a contextual effect of `x`---that is, the cluster mean `xm`---along with `x` and the intercept in the fixed-effect part of the model, along with a random intercept.
+This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | patient)`, which is the model that generated the data.
+The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects only model with `x` and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).
+
+We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.
+In order to reduce between-model random variability in comparisons of models on the same dataset, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.
+
+```{r echo=FALSE,include=FALSE}
+library(cv) # unclear why it's necessary to reload cv
+.opts <- options(warn = -2) 
+```
+```{r cross-validation,cache=TRUE,message=FALSE}
+model_lists <- lapply(fits, function(fitlist) do.call(models, fitlist))
+cvs_cases <-
+  lapply(1:Nv,
+         function(i){
+           cv(model_lists[[i]], k = 10, 
+              data = split(Data, ~ SD_ratio)[[i]])
+         })
+cvs_clusters <-
+  lapply(1:Nv,
+         function(i){
+           cv(model_lists[[i]],  k = 10, 
+              data = split(Data, ~SD_ratio)[[i]], 
+              clusterVariables = 'patient')
+         })
+```
+For a given dataset, we can plot the results for a list of models using the plot method for `"cvList"` objects. For example, this code produces Figure \ref{fig:plot-cv-example}.
+```{r plot-cv-example}
+#| out.width = "60%",
+#| fig.height = 3.5,
+#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept mixed models with varying fixed effects."
+plot(cvs_clusters[[2]], main="Comparison of Fixed Effects",)
+```
+We assemble the results for all datasets to show them in a common figure.
+```{r cross-validation-data}
+names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
+dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))
+dsummary$cases <-
+  sapply(1:nrow(dsummary), function(i){
+    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
+  })
+dsummary$clusters <-
+  sapply(1:nrow(dsummary), function(i){
+    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
+  })
+```
+```{r cross-validation-data-plot}
+#| out.width = "100%",
+#| fig.height = 4,
+#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
+xyplot(cases + clusters ~ model|SD_ratio_i, dsummary,
+       auto.key = list(space = 'top', reverse.rows = T, columns = 2), 
+       type = 'b', ylab = 'CV criterion (MSE)',
+       par.strip.text = list(cex = 0.7),
+       layout= c(Nv, 1),
+       par.settings =
+         list(superpose.line=list(lty = c(1, 2), lwd = 2),
+              superpose.symbol=list(pch = 15:16, cex = 1)),
+       scales = list(y = list(log = TRUE), 
+                     x = list(alternating = F, rot = 45))) 
+```
+In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
+With all our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.
+These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
 

### 2024-07-04 scratch/notes_georges/JSS-article/JSS-article-3-GM.Rmd
reduce example 3.2 to 7 3/4 pages

@@
 ```
-Figure \ref{fig:plot1} display scatterplots of each data set showing estimated 50% concentration ellipses for each cluster.[^lattice]
+Figure \ref{fig:plot1} displays scatterplots of each data set showing estimated 50% concentration ellipses for each cluster.[^lattice]
 
@@
 The labels above each panel show the between-patient SD/within-patient SD.
+The ellipses are estimated 50% concentration ellipses for each patient.
 The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel."
@@
 ```
-We assemble the results for all datasets to show them in a common figure.
-```{r cross-validation-data}
+The results for all datasets are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.
+```{r cross-validation-data,include=FALSE}
 names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
@@
 ```
-```{r cross-validation-data-plot}
+```{r cross-validation-data-plot,echo=FALSE}
 #| out.width = "100%",
-#| fig.height = 4,
+#| fig.height = 3.5,
 #| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."

### 2024-07-07 scratch/notes_georges/JSS-article/JSS-article-3-GM.Rmd
case- and cluster-based section compressed to 5 3/4 pages

@@
 
-We start by generating a data set to serve as a common template for the four sample data sets.
-We then apply different multipliers to the between-patient variation using parameters consistent with the description above to highlight the issues that arise in cross-validating mixed-effects models:[^data-template]
-
-[^data-template]: We invite the interested reader to experiment with varying the parameters of our example.
+<!-- We start by generating a data set to serve as a common template for the four sample data sets.
+We then apply different multipliers to the between-patient variation using parameters consistent with the description above to highlight the issues that arise in cross-validating mixed-effects models:[^data-template] -->
+
+<!-- [^data-template]: We invite the interested reader to experiment with varying the parameters of our example.  -->
 
@@
 ```
-```{r parameters}
+```{r parameters, include=FALSE}
 # Parameters:
@@
 
-```{r model-fits}
+```{r model-fits,include=FALSE}
 model.formulas <- c(
@@
 We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or "BLUPs"---as would be used for cross-validation based on cases (i.e., occasions within patients):
-```{r predict}
+```{r predict,include=FALSE}
 pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  
@@
 ```
-```{r cross-validation,cache=TRUE,message=FALSE}
+```{r cross-validation,cache=TRUE,message=FALSE,include=FALSE}
 model_lists <- lapply(fits, function(fitlist) do.call(models, fitlist))
@@
 ```
-For a given dataset, we can plot the results for a list of models using the plot method for `"cvList"` objects. For example, this code produces Figure \ref{fig:plot-cv-example}.
-```{r plot-cv-example}
+For a given dataset, we can plot the results for a list of models using the plot method for `"cvList"` objects. 
+For example, a comparison cluster-based cross-validation using fixed effects for the models fit to the third dataset (with "SD ratio = 6 / 2.5") is shown in Figure \ref{fig:plot-cv-example}.
+```{r plot-cv-example,echo=FALSE}
 #| out.width = "60%",
-#| fig.height = 3.5,
+#| fig.height = 3.8,
 #| fig.cap = "10-fold cluster-based cross-validation comparing random intercept mixed models with varying fixed effects."
-plot(cvs_clusters[[2]], main="Comparison of Fixed Effects",)
-```
-The results for all datasets are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.
+plot(cvs_clusters[[3]], main="Comparison of Fixed Effects",)
+```
+The results for all datasets, using both cluster-based and case-based cross-validation, are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.
 ```{r cross-validation-data,include=FALSE}
@@
 #| out.width = "100%",
-#| fig.height = 3.5,
+#| fig.height = 3.3,
 #| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
-xyplot(cases + clusters ~ model|SD_ratio_i, dsummary,
-       auto.key = list(space = 'top', reverse.rows = T, columns = 2), 
-       type = 'b', ylab = 'CV criterion (MSE)',
+xyplot(clusters + cases ~ model|SD_ratio_i, dsummary,
+       auto.key = list(space = 'top', reverse.rows = T, columns = 2, cex = 0.7), 
+       type = 'b', ylab = list('CV criterion (MSE)', cex = .7),
+       xlab = '',
        par.strip.text = list(cex = 0.7),
@@
        par.settings =
-         list(superpose.line=list(lty = c(1, 2), lwd = 2),
-              superpose.symbol=list(pch = 15:16, cex = 1)),
-       scales = list(y = list(log = TRUE), 
-                     x = list(alternating = F, rot = 45))) 
+         list(superpose.line=list(lty = c(1, 2), lwd = 1),
+              superpose.symbol=list(cex = 0.7)),
+       scales = list(y = list(log = TRUE, cex = 0.7), 
+                     x = list(alternating = F, rot = 45, cex = 0.7))) 
 ```

### 2024-07-08 scratch/notes_georges/JSS-article/JSS-article-3-GM.Rmd
Contrasting cluster-based and case-based CV down to 4 pages.

@@
 
-The \code{Auto} dataset contains information about 392 cars:
+The \code{Auto} data set contains information about 392 cars:
 
@@
 
-The four datasets differ in the relative magnitude of between-cluster variance compared with within-cluster variance.
+The four data sets differ in the relative magnitude of between-cluster variance compared with within-cluster variance.
 They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.
@@
 
-Our four contrived data sets (see below) illustrate possible results for data obtained in such a scenario.
-The relative configuration of dosages `x` and symptoms `y` are identical within patients in each of the four datasets.
 Within patients, higher dosages are generally associated with a reduction in symptoms.
-
 Between patients, however, higher dosages are associated with higher levels of symptoms.
@@
 
-The four data sets differ in the between-patient variance of patient centroids from a common between-patients regression line.
-The data sets exhibit a progression from low to high variance around the common regression line.
-
-<!-- We start by generating a data set to serve as a common template for the four sample data sets.
-We then apply different multipliers to the between-patient variation using parameters consistent with the description above to highlight the issues that arise in cross-validating mixed-effects models:[^data-template] -->
-
-<!-- [^data-template]: We invite the interested reader to experiment with varying the parameters of our example.  -->
+To construct the four data sets, we first construct a template data set from a population with a between patient effect of dosage of $1.0$ and a within-patient effect of $-0.5$.
+We then apply four different multipliers to the between-patient variation from the regression line of patient centroids.
+The resulting data sets, shown in Figure \ref{fig:plot1}, differ in the between-patient variance of patient centroids from a common between-patient regression line, exhibiting a progression from low to high variance around the common regression line.[^lattice]
+
+Note that each patient has the same relative configuration of dosages and symptoms in each of the four data sets.
+To help visualize the structure of the data, estimated 50% concentration ellipses are shown for each patient.
+
+Using such data, the researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept.
+We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.
+The first three models are sequentially nested with fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`.
+The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`.
+We thus fit four models to four data sets for a total of 16 models.
 
@@
 ```
-Figure \ref{fig:plot1} displays scatterplots of each data set showing estimated 50% concentration ellipses for each cluster.[^lattice]
 
@@
 #| out.width = "100%",
-#| fig.height = 3.5,
+#| fig.height = 3.0,
 #| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance.
@@
 xyplot(y ~ x | SD_ratio, data = Data, group = patient,
+       ylab = list('symptoms', cex = 0.7),
+       xlab = list('dosage', cex = 0.7),
        par.strip.text = list(cex = 0.7),
-       scales = list(x = list(alternating = FALSE)),
+       scales = list(cex = 0.7, x = list(alternating = FALSE)),
        layout = c(Nv, 1),
@@
 ```
-The population from which these datasets are generated has a between-patient effect of dosage of 1 and a within-patient effect of $-0.5$.
-The researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept.
-We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.
-The first three models have sequentially nested fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`.
-The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`.
-We thus fit four models to four datasets for a total of 16 models.
 
@@
 ```
-<!-- We then prepare the data for plotting: -->
-
 ```{r data-predictions,include=FALSE}
@@
   )
+Datap <- reshape(Dataf, direction = "long", sep = '.',
+                 varying = c('pred.fixed', 'pred.BLUPs'),
+                 timevar = 'prediction_type')
+Datap$panel2 <- with(Datap, paste0(formula,":",prediction_type))
+flevels <- c(unique(Datap$panel2), 'data')
+Datap$panel2 <- factor(Datap$panel2, levels = flevels)
 Data$panel <- factor('data', levels = c(names(model.formulas), 'data'))
-```
-The fixed-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-fixed} along with a summary of the data using 50% concentration ellipses.
-Figure \ref{fig:plot-fits-blups} similarly displays predicted BLUPs.
-```{r plot-fits-fixed,echo=FALSE}
-#| out.width = "100%",
-#| fig.height = 6,
-#| fig.cap = "Fixed-effect predictions using each model applied to data sets with varying between- and within-patient variance ratios.
-The top row shows summaries of the within-patient data using estimated 50% concentration ellipses."
-{
-  xyplot(y ~ x |SD_ratio * panel, Data,
-         groups = patient, type = 'n',    
-         par.strip.text = list(cex = 0.6),
-         scales = list(x = list(alternating = FALSE)),
-         drop.unused.levels = FALSE,
-         ylab = list('fixed-effect predictions',cex=.7)) + 
-    glayer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
-                         level = 0.5),
-           panel.abline(a = 10, b = 1)) +
-    xyplot(pred.fixed  ~ x |SD_ratio * panel, Dataf, type = 'l', 
-           groups = patient,
-           drop.unused.levels = F)
-}|> useOuterStrips()
-```
-<!-- The BLUPs from these models, shown in Figure \ref{fig:plot-fits-blups}, are produced as follows: -->
-```{r plot-fits-blups,echo=FALSE}
+Data$panel2 <- factor('data', levels = flevels)
+```
+The fixed- and random-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-combined} along with a summary of the data using 50% concentration ellipses.
+
+```{r plot-fits-combined,echo=FALSE}
 #| out.width = "100%",
@@
 The top row shows summaries of the within-patient data using estimated 50% concentration ellipses."
-{   xyplot(y ~ x | SD_ratio * panel, Data,
+levs3 <- levels(Data$panel2)
+levs3 <- sub(':fixed','', levs3)
+levs3 <- sub(':BLUPs',' ', levs3)
+Data$panel3 <- factor(sub(':fixed','', Data$panel2,), levels = levs3)
+Datap$panel3 <- factor(sub(':BLUPs',' ', Data$panel2,), levels = levs3)
+
+{   xyplot(y ~ x | SD_ratio * panel3, Data,
          groups = patient, type = 'n',
          drop.unused.levels = F,
-         par.strip.text = list(cex = 0.6),
-         scales = list(x = list(alternating = FALSE)),
-         ylab = list('fixed- and random-effect predictions (BLUPS)', cex = 0.7)) +
+         par.strip.text = list(cex = 0.4),
+         scales = list(x = list(alternating = FALSE, cex = 0.7),
+                       y = list(alternating =2, cex = 0.5)),
+         xlab = list('dosage', cex = 0.7),
+         between =
+             list(y = c(0,0,0,0.25,0,0,0,.25)),
+         ylab = list(
+           paste0(
+             paste0(rep(' ',18),  collapse = ''),
+                  'fixed-effects',
+             paste0(rep(' ',52), collapse = ''),
+             'BLUPs',
+             paste0(rep(' ',32), collapse = ''),
+             'data'),
+           cex = 0.7)) +
     glayer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,
@@
            panel.abline(a = 10, b = 1)) +
-    xyplot(pred.BLUPs  ~ x | SD_ratio * panel, Dataf, type = 'l', 
+    xyplot(pred  ~ x | SD_ratio * panel2, Datap, type = 'l',
            groups = patient,
            drop.unused.levels = F)
-}|> useOuterStrips()
-```
+} -> re.plot
+useOuterStrips(re.plot)
+
+```
+
 Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`.
@@
 
-Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform with the within-patient slopes for the `~ 1 + x` model in the two datasets with the smallest between-patient variances.
-
-For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for datasets with larger between-subject variance.
+Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform with the within-patient slopes for the `~ 1 + x` model in the two data sets with the smallest between-patient variances.
+
+For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variance.
 With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.
@@
 
-We will discover that case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model for the '5 / 2.5' dataset, but cluster-based cross-validation prefers the latter model to the former.
-The situation is entirely reversed with the '8 / 2.5' dataset.
+We will discover that case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model for the '5 / 2.5' data set, but cluster-based cross-validation prefers the latter model to the former.
+The situation is entirely reversed with the '8 / 2.5' data set.
 
@@
 We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.
-In order to reduce between-model random variability in comparisons of models on the same dataset, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.
+In order to reduce between-model random variability in comparisons of models on the same data set, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.
 
@@
 ```
-For a given dataset, we can plot the results for a list of models using the plot method for `"cvList"` objects. 
-For example, a comparison cluster-based cross-validation using fixed effects for the models fit to the third dataset (with "SD ratio = 6 / 2.5") is shown in Figure \ref{fig:plot-cv-example}.
-```{r plot-cv-example,echo=FALSE}
-#| out.width = "60%",
-#| fig.height = 3.8,
-#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept mixed models with varying fixed effects."
-plot(cvs_clusters[[3]], main="Comparison of Fixed Effects",)
-```
-The results for all datasets, using both cluster-based and case-based cross-validation, are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.
+The results for all data sets, using both cluster-based and case-based cross-validation, are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.
 ```{r cross-validation-data,include=FALSE}
@@
 xyplot(clusters + cases ~ model|SD_ratio_i, dsummary,
-       auto.key = list(space = 'top', reverse.rows = T, columns = 2, cex = 0.7), 
+       auto.key = list(space = 'top', columns = 2, cex = 0.7), 
        type = 'b', ylab = list('CV criterion (MSE)', cex = .7),
@@
 ```
+
 And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-working-model}, on page \pageref{fig:Auto-working-model}, and Figure \ref{fig:Auto-CR-plots-transformed}):
@@
 ```
+
 The selected model includes three of the numeric predictors, \code{horsepower}, \code{weight}, and \code{acceleration}, along with the factors \code{year} and \code{origin}.
@@
 ```
+
 This is considerably smaller than the MSE for the original working model:
@@
 ```
+
 A perhaps subtle point is that we compute the MSE for the selected model on the original \code{mpg} response scale rather than the log scale, so as to make the selected model comparable to the working model.[^medAbsErr]
@@
 ```
+
 The other arguments to \code{cv()} are (see \code{?cv:::cv.function} for additional optional arguments and details):

### 2024-07-08 scratch/notes_john/JSS-article/JSS-article-3.Rmd
some changes to John's JSS draft

@@
 The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
+
 ```{r cv-methods}
@@
 ```
-* The \code{"lm"} and \code{"glm"} methods are for linear and generalized-linear models fit respectively by the standard \proglang{R} \code{lm()} and \code{glm()} functions.
-
-* The \code{"modList"} method for \code{cv()} cross-validates several competing models, not necessarily of the same class, using the same division of the data into folds.
+
+-   The \code{"lm"} and \code{"glm"} methods are for linear and generalized-linear models fit respectively by the standard \proglang{R} \code{lm()} and \code{glm()} functions.
+
+-   The \code{"modList"} method for \code{cv()} cross-validates several competing models, not necessarily of the same class, using the same division of the data into folds.
 The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.
 
-* Cross-validating mixed-effects models, using the \code{"glmmTMB"}, \code{"lme"}, and  \code{"merMod"} methods for \code{cv()}, involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.
-
-* The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.
-
-* The \code{"default"} \code{cv()} method works (perhaps with a bit of coaxing) with many other existing regression-model classes for which there is an \code{update()} method that accepts a \code{data} argument.
+-   Cross-validating mixed-effects models, using the \code{"glmmTMB"}, \code{"lme"}, and \code{"merMod"} methods for \code{cv()}, involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.
+
+-   The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.
+
+-   The \code{"default"} \code{cv()} method works (perhaps with a bit of coaxing) with many other existing regression-model classes for which there is an \code{update()} method that accepts a \code{data} argument.
 More generally, the \pkg{cv} package is designed to be extensible, as discussed in Section \ref{extending-the-cv-package}.
@@
 The presentation here is close (though not identical) to that in the original source [@JamesEtAl:2021 Secs.
-5.1, 5.3], and it demonstrates the use of the `cv()` function.[^boot]
-
-[^boot]: @JamesEtAl:2021 use the `cv.glm()` function in the \pkg{boot} package [@CantyRipley2022; @DavisonHinkley:1997].
+5.1, 5.3], and it demonstrates the use of the `cv()` function.[^1]
+
+[^1]: @JamesEtAl:2021 use the `cv.glm()` function in the \pkg{boot} package [@CantyRipley2022; @DavisonHinkley:1997].
 Despite its name, `cv.glm()` is an independent function and not a method of a `cv()` generic function.
@@
 ```
+
 With the exception of `origin` (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see `help("Auto", package="ISLR2")`.
 
-```{r mpg-horsepower-scatterplot-polynomials, echo=FALSE}
-#| out.width = "50%",
+```{r mpg-horsepower-polynomials, fig.show="hold", echo=FALSE}
+#| out.width = "45%",
 #| fig.height = 5,
-#| fig.cap = "\\code{mpg} vs. \\code{horsepower} for the \\code{Auto} data, showing fitted polynomials of degree 1 through 5."
+#| fig.cap = "(a) \\code{mpg} vs. \\code{horsepower} for the \\code{Auto} data, showing fitted polynomials of degree 1 through 5. (b) Estimated squared error as a function of polynomial degree, $p$.",
+#| fig.subcap=rep("", 2)
 plot(mpg ~ horsepower, data=Auto)
@@
        title="Degree", inset=0.02)
-```
-
-We'll focus here on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}.
-The relationship between the two variables is monotone, decreasing, and nonlinear.
-Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}.
-The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).
-
-[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here.
-Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader).
-We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.
-
-Figure \ref{fig:mpg-horsepower-MSE-se} shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$.
-The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.
-
-```{r mpg-horsepower-MSE-se, echo=FALSE}
-#| out.width = "50%",
-#| fig.height = 5,
-#| fig.cap = "Estimated squared error as a function of polynomial degree, $p$"
 
@@
 
+We'll focus here on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}.
+The relationship between the two variables is monotone, decreasing, and nonlinear.
+Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^2] Polynomial fits for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}.
+The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).
+
+[^2]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here.
+Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader).
+We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.
+
+Figure \ref{fig:mpg-horsepower-MSE-se} shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$.
+The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.
+
 The generic `cv()` function has an `"lm"` method,
+
 ```{r cv-lm-method}
@@
 ```
+
 which takes the following arguments:
 
-* \code{model}, an \code{"lm"} object, the only required argument.
-
-* \code{data}, which can usually be inferred from the \code{model} object.
-
-* \code{criterion}, a function to compute the CV criterion (defaulting to \code{mse}).
-
-* \code{k}, the number of folds to employ (defaulting to \code{10}); the character value \code{"n"} or \code{"loo"} may be supplied to specify leave-one-out cross-validation.
-
-* \code{reps}, the number of times to repeat the CV procedure (defaulting to 1).
-
-* \code{seed}, the seed for \proglang{R}'s pseudo-random number generator; if not specified a value is randomly selected, reported, and saved, so that the CV procedure is replicable.
-
-* \code{confint}, whether or not to compute a confidence interval for the CV criterion, defaulting to \code{TRUE} if there are at least 400 cases; a confidence interval is computed only if the CV criterion can be expressed as the average of casewise components (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals} for details).
-
-* \code{level}, the level for the confidence interval (defaulting to \code{0.95}).
-
-* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).
-
-* \code{ncores}, the number of cores to employ for parallel computation; if \code{cores = 1} (the default), the computations are not parallelized.
+-   \code{model}, an \code{"lm"} object, the only required argument.
+
+-   \code{data}, which can usually be inferred from the \code{model} object.
+
+-   \code{criterion}, a function to compute the CV criterion (defaulting to \code{mse}).
+
+-   \code{k}, the number of folds to employ (defaulting to \code{10}); the character value \code{"n"} or \code{"loo"} may be supplied to specify leave-one-out cross-validation.
+
+-   \code{reps}, the number of times to repeat the CV procedure (defaulting to 1).
+
+-   \code{seed}, the seed for \proglang{R}'s pseudo-random number generator; if not specified a value is randomly selected, reported, and saved, so that the CV procedure is replicable.
+
+-   \code{confint}, whether or not to compute a confidence interval for the CV criterion, defaulting to \code{TRUE} if there are at least 400 cases; a confidence interval is computed only if the CV criterion can be expressed as the average of casewise components (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals} for details).
+
+-   \code{level}, the level for the confidence interval (defaulting to \code{0.95}).
+
+-   \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).
+
+-   \code{ncores}, the number of cores to employ for parallel computation; if \code{cores = 1} (the default), the computations are not parallelized.
 
 To illustrate, we perform 10-fold CV for a quadratic polynomial fit to the \code{Auto} data:
+
 ```{r cv-lm-1}
@@
 ```
+
 The function reports the CV estimate of MSE, a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression.
@@
 ```
+
 The \code{"hatvalues"} method reports only the CV estimate of MSE.
@@
 All three methods produce exact results for a linear model (within the precision of floating-point computations):
+
 ```{r cv.lm-3}
@@
 ```
+
 The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE (and a confidence interval around it) along with the full-sample MSE, but bias isn't an issue for LOO CV.
@@
 We use the \code{microbenchmark()} function from the package of the same name [@Mersmann:2023] for the timings:
+
 ```{r cv.lm.timings, cache=TRUE}
@@
 ```
+
 On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.
@@
 We illustrate with the polynomial regression models of varying degree for the \code{Auto} data, beginning by fitting and saving the 10 models:
+
 ```{r polyomial-models}
@@
 ```
+
 We then apply \code{cv()} to the list of 10 models (the \code{data} argument is required):
+
 ```{r polynomial-regression-CV}
@@
 ```
+
 The \code{models()} function takes an arbitrary number of regression models as its arguments, which are optionally named, to create a \code{"modList"} object.
@@
 
-In contrast, mixed-effects models are fit to *dependent* data, in which cases are clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases are repeated observations on the individuals over time.[^crossed-effects]
-
-[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects.
+In contrast, mixed-effects models are fit to *dependent* data, in which cases are clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases are repeated observations on the individuals over time.[^3]
+
+[^3]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects.
 For example, consider students within classes within schools.
@@
 
-We can think of two approaches to applying cross-validation to clustered data:[^cv-faq]
-
-[^cv-faq]: We subsequently discovered that @Vehtari:2023 [Section 8] makes similar points.
+We can think of two approaches to applying cross-validation to clustered data:[^4]
+
+[^4]: We subsequently discovered that @Vehtari:2023 [Section 8] makes similar points.
 
@@
 \code{MathAchieve} comprises individual-level data on 7185 students in 160 high schools, and \code{MathAchSchool} contains school-level data:
+
 ```{r HSB-data}
@@
 ```
+
 The first few students are in school number 1224 and the last few in school 9586.
@@
 Some data-management is required before fitting a mixed-effects model to the HSB data:
+
 ```{r HSB-data-management, cache=TRUE}
@@
 ```
+
 In the process, we merge variables from the school-level and student-level data sets, and create two new school-level variables: \code{mean.ses}, which is the average SES for students in each school; and \code{cses}, which is the individual students' SES centered at their school means.
@@
 Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
+
 ```{r HSB-lmer, cache=TRUE}
@@
 We can then cross-validate at the cluster (i.e., school) level,
+
 ```{r HSB-lmer-CV-cluster, cache=TRUE}
@@
 ```
+
 or at the case (i.e., student) level,
+
 ```{r HSB-lmer-CV-case, cache=TRUE}
@@
 ```
+
 For cluster-level CV, the \code{clusterVariables} argument tells \code{cv()} how the clusters are defined.
@@
 
-We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^1]
-
-[^1]: We invite the interested reader to experiment with varying the parameters of our example.
+We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^5]
+
+[^5]: We invite the interested reader to experiment with varying the parameters of our example.
 
@@
 We discuss each of these models below.
+
 ```{r fitted-mixed-models}
@@
 
-To illustrate cross-validation of model specification, we return to the \code{Auto} data set:[^Venables]
+To illustrate cross-validation of model specification, we return to the \code{Auto} data set:[^6]
+
+[^6]: This example benefits from an email conversation with Bill Venables, who of course isn't responsible for the use to which we've put his insightful remarks.
+
 ```{r Auto-redux}
@@
 
-[^Venables]: This example benefits from an email conversation with Bill Venables, who of course isn't responsible for the use to which we've put his insightful remarks.
-
 The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}.
@@
 We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
-```{r Auto-explore}
-#| out.width = "60%",
+
+```{r Auto-explore, fig.show="hold"}
+#| out.width = "45%",
 #| fig.height = 5,
-#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data.
-In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression."
+#| fig.cap = "(a) Scatterplot matrix for the numeric variables in the \\code{Auto} data.
+In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.
+(b) Component+residual plots for the working model fit to the \\code{Auto} data.
+The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.",
+#| fig.subcap=rep("", 2)
 Auto$cylinders <- factor(Auto$cylinders,
@@
                   smooth=list(spread=FALSE), data=Auto, pch=".")
-```
+
+m.auto <- lm(mpg ~ ., data = Auto)
+crPlots(m.auto, pch=".")
+```
+
+```{r Auto-explore-show, eval=FALSE}
+Auto$cylinders <- factor(Auto$cylinders,
+                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
+Auto$year <- as.factor(Auto$year)
+Auto$origin <- factor(Auto$origin,
+                      labels=c("America", "Europe", "Japan"))
+rownames(Auto) <- make.names(Auto$name, unique=TRUE)
+Auto$name <- NULL
+
+scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
+                  smooth=list(spread=FALSE), data=Auto, pch=".")
+```
+
 A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors.
 There are, however, only 5 distinct values of \code{cylinders} (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders.
-There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.
-
-[^year]: Making the decision to treat \code{year} as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process.
+There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^7] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.
+
+[^7]: Making the decision to treat \code{year} as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process.
 It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation.
@@
 We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
-```{r Auto-working-model}
-#| out.width = "60%",
-#| fig.height = 5,
-#| fig.cap = "Component+residual plots for the working model fit to the \\code{Auto} data.
-The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths."
+
+```{r Auto-working-model-show, eval=FALSE}
 m.auto <- lm(mpg ~ ., data = Auto)
@@
 ```
+
 The component+residual plots in Figure \ref{fig:Auto-working-model} clearly reveal the inadequacy of the model.
@@
 We proceed to transform the numeric predictors in the \code{Auto} regression towards multivariate normality:
+
 ```{r Auto-transform}
@@
 ```
+
 We then apply the (rounded) transformations---all, as it turns out, logs (i.e., "Oth" powers)---to the data and re-estimate the model:
+
 ```{r Auto-with-transformed-predictors}
@@
 Having transformed the predictors towards multivariate normality, we now consider whether there's evidence for transforming the response (using \code{powerTransform()} for Box and Cox's original method), also obtaining a log transformation:
+
 ```{r Auto-Box-Cox}
@@
 The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed-scatterplot-matrix}):
-```{r Auto-transformed-scatterplot-matrix}
-#| out.width = "60%",
+
+```{r Auto-transformed-scatterplot-matrix, fig.show="hold", echo=FALSE}
+#| out.width = "45%",
 #| fig.height = 5,
-#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data"
+#| fig.cap = "(a) Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data. (b) Component+residual plots for the working model fit to the transformed \\code{Auto} data.",
+#| fig.subcap=rep("", 2)
 scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
@@
                   smooth=list(spread=FALSE), data=A, pch=".")
+crPlots(m, pch=".")
+```
+
+```{r Auto-transformed-scatterplot-matrix-show, eval=FALSE}
+scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
+                  + acceleration, 
+                  smooth=list(spread=FALSE), data=A, pch=".")
 ```
 And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-working-model}, on page \pageref{fig:Auto-working-model}, and Figure \ref{fig:Auto-CR-plots-transformed}):
-```{r Auto-CR-plots-transformed}
-#| out.width = "60%",
-#| fig.height = 5,
-#| fig.cap = "Component+residual plots for the model fit to the transformed \\code{Auto} data"
+
+```{r Auto-CR-plots-transformed-show, eval=FALSE}
 crPlots(m, pch=".")
@@
 After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log n$):
+
 ```{r}
@@
 ```
+
 The selected model includes three of the numeric predictors, \code{horsepower}, \code{weight}, and \code{acceleration}, along with the factors \code{year} and \code{origin}.
 We can calculate the MSE for this model, but we expect that the result will be optimistic because we used the whole data to help specify the model:
+
 ```{r MSE-whole-selected-model}
@@
 ```
+
 This is considerably smaller than the MSE for the original working model:
+
 ```{r MSE-working-model}
@@
 ```
-A perhaps subtle point is that we compute the MSE for the selected model on the original \code{mpg} response scale rather than the log scale, so as to make the selected model comparable to the working model.[^medAbsErr]
-
-[^medAbsErr]: That's slightly uncomfortable given the skewed distribution of \code{mpg}.
+
+A perhaps subtle point is that we compute the MSE for the selected model on the original \code{mpg} response scale rather than the log scale, so as to make the selected model comparable to the working model.[^8]
+
+[^8]: That's slightly uncomfortable given the skewed distribution of \code{mpg}.
 An alternative is to use a robust measure of model lack-of-fit, such as the median absolute error instead of the mean-squared error, employing the \code{medAbsErr()} function from the \pkg{cv} package.
@@
 The first argument to \code{cv()}---here \code{selectTransAndStepAIC}---is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion:
+
 ```{r Auto-transform-and-select}
@@
 ```
+
 The other arguments to \code{cv()} are (see \code{?cv:::cv.function} for additional optional arguments and details):
 
-* \code{data}, the data set to which the model is fit.
-* \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, then a seed is randomly selected and saved.
-* Arguments required by the model-selection function: the starting \code{working.model} (here, \code{m.auto}) for transformation and predictor selection; the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation; and \code{AIC=FALSE}, which specifies use of the BIC for model selection.
+-   \code{data}, the data set to which the model is fit.
+-   \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, then a seed is randomly selected and saved.
+-   Arguments required by the model-selection function: the starting \code{working.model} (here, \code{m.auto}) for transformation and predictor selection; the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation; and \code{AIC=FALSE}, which specifies use of the BIC for model selection.
 
@@
 
-* \code{selectTransStepAIC()} automatically computes CV cost criteria, here the MSE, on the \emph{untransformed} response scale.
-* As we anticipated, the estimate of the MSE that we obtain by cross-validating the whole model-specification process is larger than the MSE computed for the model we fit to the \code{Auto} data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
-* When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of \code{compareFolds()}), we see that there is little uncertainty in choosing variable transformations (the \code{lam.*}s for the $x$s and \code{lambda} for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: \code{horsepower}, \code{weight}, and \code{year} are always included among the selected predictors; \code{acceleration} and \code{displacement} are included respectively in 4 and 3 of 10 selected models; and \code{cylinders} and \code{origin} are each included in only 1 of 10 models.
+-   \code{selectTransStepAIC()} automatically computes CV cost criteria, here the MSE, on the \emph{untransformed} response scale.
+-   As we anticipated, the estimate of the MSE that we obtain by cross-validating the whole model-specification process is larger than the MSE computed for the model we fit to the \code{Auto} data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
+-   When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of \code{compareFolds()}), we see that there is little uncertainty in choosing variable transformations (the \code{lam.*}s for the $x$s and \code{lambda} for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: \code{horsepower}, \code{weight}, and \code{year} are always included among the selected predictors; \code{acceleration} and \code{displacement} are included respectively in 4 and 3 of 10 selected models; and \code{cylinders} and \code{origin} are each included in only 1 of 10 models.
 Recall that when we selected predictors for the full data, we obtained a model with \code{horsepower}, \code{weight}, \code{acceleration}, \code{year}, and \code{origin}.
 
-## Example: Applying recursive CV to polynomial regression for the Auto data[^recursive-cv]
-
-[^recursive-cv]: What we call here "recursive CV" has also been termed "nested CV" [e.g., in the \pkg{cvms} package by @OlsenZachariae:2024].
+## Example: Applying recursive CV to polynomial regression for the Auto data[^9]
+
+[^9]: What we call here "recursive CV" has also been termed "nested CV" [e.g., in the \pkg{cvms} package by @OlsenZachariae:2024].
 We prefer, however, to reserve the term "nested CV" for the procedure described by @BatesHastieTibshirani:2023.
 
-In Section \ref{comparing-competing-models}, following @JamesEtAl:2021[Secs.
+In Section \ref{comparing-competing-models}, following @JamesEtAl:2021 [Secs.
 5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of `mpg` to `horsepower` for the `Auto` data, saving the 10 models, named `m.1` through `m.10`, in the list `mlist`.
@@
 Applying `selectModelList()` to the `Auto` polynomial-regression models, and using 10-fold CV, we obtain:
+
 ```{r recursive-CV-polynomials}
@@
 ```
+
 As expected, recursive CV produces a larger estimate of MSE for the selected 7th degree polynomial model than CV applied directly to this model.
@@
 We can equivalently call `cv()` with the list of models as its first argument and set the argument `recursive=TRUE`:
+
 ```{r recursive-cv-alt}
@@
 Data for the example are in the \code{BEPS} data set in the \pkg{carData} package:
+
 ```{r BEPS-data}
@@
 ```
+
 The polytomous (multi-category) response variable is \code{vote}, a factor with levels \code{"Conservative"}, \code{"Labour"}, and \code{"Liberal Democrat"}. The predictors of \code{vote} are:
 
-* \code{age}, in years;
-* \code{econ.cond.national} and \code{econ.cond.household}, the respondent's ratings of the state of the economy, on 1 to 5 scales.
-* \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
-* \code{Europe}, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism."
-* \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3.
-* \code{gender}, \code{"female"} or \code{"male"}.
+-   \code{age}, in years;
+-   \code{econ.cond.national} and \code{econ.cond.household}, the respondent's ratings of the state of the economy, on 1 to 5 scales.
+-   \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
+-   \code{Europe}, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism."
+-   \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3.
+-   \code{gender}, \code{"female"} or \code{"male"}.
 
 The model fit to the data includes an interaction between \code{Europe} and \code{political.knowledge}, which was the focus of the original research on which this example is based [@AndersenHeathSinnott:2002]; the other predictors enter the model additively:
+
 ```{r BEPS-model}
@@
 ```
+
 <!-- We set the argument \code{trace=FALSE} in the call to \code{multinom()} to suppress reporting the iteration history, which will be particularly appreciated when we repeatedly refit the model in cross-validation.
@@
 Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the interaction in a "stacked-area" graph:
+
 ```{r BEPS-plot, fig.width=9, fig.height=5}
@@
 ```
+
 As political knowledge increases, voters tend to align their votes more closely with the party positions on European integration: The Conservative Party was relatively Euro-skeptic, while Labour and the Liberal Democrats were more supportive of the UK's participation in the EU.
@@
 The \code{BayesRule()} criterion, also supplied by \pkg{cv}, which is for a binary response, comes close:
+
 ```{r BayesRule}
@@
 ```
+
 After doing some error checking, \code{BayesRule()} rounds the predicted proabability of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction, and then reports the proportion of incorrect predictions.
@@
 It is straightforward to adapt Bayes's rule to a polytomous response:
+
 ```{r BayesRuleMulti}
@@
 ```
+
 The \code{predict()} method for \code{"multinom"} models called with argument \code{type="class"} reports the Bayes's rule prediction for each case---that is, the response category with the highest predicted probability.
@@
 The marginal proportions for the response categories are
+
 ```{r BEPS-response-distribution}
@@
 ```
+
 and so the marginal Bayes's rule prediction, that everyone will vote Labour, produces an error rate of $1 - 0.47213 = 0.52787$.
@@
 We check first whether the default \code{cv()} method works "out-of-the-box" for the \code{"multinom"} model:
+
 ```{r BEPS-test-default, error=TRUE}
@@
 ```
+
 The default method of \code{GetResponse()} (a function supplied by the \pkg{cv} package---see \code{?GetResponse}) fails for a \code{"multinom"} object.
 A straightforward solution is to supply a \code{GetResponse.multinom()} method that returns the factor response [using the `get_response()` function from the \pkg{insight} package, @LudeckeWaggonerMakowski:2019],
+
 ```{r GetResponse.multinom}
@@
 ```
+
 and to try again:
+
 ```{r BEPS-test-default-2, error=TRUE}
@@
 ```
+
 A \code{traceback()} (not shown) reveals that the problem is that the default method of \code{cv()} calls the \code{"multinom"} method for \code{predict()} with the argument \code{type="response"}, when the correct argument should be \code{type="class"}.
 We therefore must write a \code{"multinom"} method for \code{cv()}, but that proves to be very simple:
+
 ```{r cv.nultinom}
@@
 ```
+
 That is, we simply call the default `cv()` method (via \code{NextMethod()}) with the `type` argument properly set.
@@
 Then:
+
 ```{r BEPS-cv}
@@
 ```
+
 The cross-validated polytomous Bayes's rule criterion confirms that the fitted model does substantially better than the marginal Bayes's rule prediction that everyone votes for Labour.
@@
 
-Our goal in implementing CV is to make it conveniently available to \proglang{R} users with limited programming skills,[^cv-by-programming] and hence to encourage its use.
+Our goal in implementing CV is to make it conveniently available to \proglang{R} users with limited programming skills,[^10] and hence to encourage its use.
 Towards this end, the \pkg{cv} package provides a simple, consistent interface through the \code{cv()} generic function, with specific methods for different classes of standard \proglang{R} statistical models.
@@
 
-[^cv-by-programming]: \proglang{R} users with sophisticated programming skills would generally find it unchallenging to implement CV directly for specific applications (as we illustrate below).
+[^10]: \proglang{R} users with sophisticated programming skills would generally find it unchallenging to implement CV directly for specific applications (as we illustrate below).
 Even in this case, however, there's an argument for using a simple pre-programmed interface to CV, to minimize programming effort and to avoid mistakes.
@@
 
-Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]
+Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^11]
+
+[^11]: The example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise the code from the vignette is only trivially modified here.
+
 ```{r rsample-example, cache=TRUE}
@@
 ```
-[^k-fold-used-in-rsample]: The example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise the code from the vignette is only trivially modified here.
 
 In this case, it's simple (and, in our opinion, simpler) to program LOO CV directly. For example
+
 ```{r direct-loo-cv, cache=TRUE}
@@
 The \pkg{caret} package supports generalized linear models fit by \code{glm()} and can straightforwardly perform LOO CV:
+
 ```{r rsample-example-using-caret, cache=TRUE}
@@
 And here's how we can perform the same computation using the \code{cv()} function in the \pkg{cv} package:
+
 ```{r rsample-example-using-cv, cache=TRUE}
@@
 Let's compare the computational efficiency of the various implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
+
 ```{r compare-rsample-cv, cache=TRUE}
@@
 ```
+
 Computing time for \pkg{rsample} and \pkg{caret} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.
@@
 
-The weighted-least-squares (WLS) estimator of $\boldsymbol{\beta}$ is [see, e.g., @Fox:2016, Sec. 12.2.2] [^WLS]
-$$
+The weighted-least-squares (WLS) estimator of $\boldsymbol{\beta}$ is [see, e.g., @Fox:2016, Sec. 12.2.2] [^12] $$
 \mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} 
@@
 
-[^WLS]: This is a definitional formula, which assumes that the model matrix $\mathbf{X}$ is of full column rank, and which can be subject to numerical instability when $\mathbf{X}$ is ill-conditioned.
+[^12]: This is a definitional formula, which assumes that the model matrix $\mathbf{X}$ is of full column rank, and which can be subject to numerical instability when $\mathbf{X}$ is ill-conditioned.
 \code{lm()} uses the singular-value decomposition of the model matrix to obtain computationally more stable results.
@@
 
-To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identify [@Hager:1989],
-$$
+To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identify [@Hager:1989], $$
 \left(\mathbf{A}_{m \times m} + \mathbf{U}_{m \times k} 
@@
 \mathbf{VA}^{-1}\mathbf{U} \right)^{-1} \mathbf{VA}^{-1}
-$$
-where $\mathbf{A}$ is a nonsingular order-$n$ matrix. We apply this result by letting
-\begin{align*}
+$$ where $\mathbf{A}$ is a nonsingular order-$n$ matrix. We apply this result by letting \begin{align*}
     \mathbf{A} &= \mathbf{X}^T \mathbf{W} \mathbf{X} \\
@@
     \mathbf{C} &= \mathbf{W}_\mathbf{j} \\
-\end{align*}
-where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jm})^T$ represents the vector of indices for the cases in the $j$th fold, $j = 1, \ldots, k$.
+\end{align*} where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jm})^T$ represents the vector of indices for the cases in the $j$th fold, $j = 1, \ldots, k$.
 The negative sign in $\mathbf{V} = - \mathbf{X}_\mathbf{j}$ reflects the *removal*, rather than addition, of the cases in $\mathbf{j}$.
@@
 Let $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ represent a cross-validation cost criterion, such as mean-squared error, computed for all of the $n$ values of the response $\mathbf{y}$ based on fitted values $\widehat{\mathbf{y}}$ from the model fit to all of the data.
-We require that $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ is the mean of casewise components, that is, $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n\mathrm{cv}(y_i, \widehat{y}_i)$.[^contrast-function] For example, $\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$.[^casewise-loss]
-
-[^casewise-loss]: Some commonly employed CV criteria---such as the root-mean-squared error (RMSE), median absolute error, and, for binary-regression models, the complement of the area under the receiver operating characteristic (ROC) curve---are not means of casewise components.
+We require that $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}})$ is the mean of casewise components, that is, $\mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n\mathrm{cv}(y_i, \widehat{y}_i)$.[^13] For example, $\mathrm{MSE}(\mathbf{y}, \widehat{\mathbf{y}}) = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$.[^14]
+
+[^13]: @ArlotCelisse:2010 term the casewise loss, $\mathrm{cv}(y_i, \widehat{y}_i)$, the "contrast function."
+
+[^14]: Some commonly employed CV criteria---such as the root-mean-squared error (RMSE), median absolute error, and, for binary-regression models, the complement of the area under the receiver operating characteristic (ROC) curve---are not means of casewise components.
 That the RMSE and median absolute error aren't means of casewise components is obvious; for the complement of the area under the ROC curve, see the vignette on extending the \pkg{cv} package.
-
-[^contrast-function]: @ArlotCelisse:2010 term the casewise loss, $\mathrm{cv}(y_i, \widehat{y}_i)$, the "contrast function."
 
@@
 Then the cross-validation criterion is just $\mathrm{CV} = \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}^{(-i)})$.
-Following @DavisonHinkley:1997[pp. 293--295], the bias-adjusted cross-validation criterion is
-$$
+Following @DavisonHinkley:1997 [pp.
+293--295], the bias-adjusted cross-validation criterion is $$
 \mathrm{CV}_{\mathrm{adj}} = \mathrm{CV} + \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}) - \frac{1}{n} \sum_{j=1}^{k} n_j \mathrm{CV}_j
@@
 
-We compute the standard error of CV as 
-$$
+We compute the standard error of CV as $$
 \mathrm{SE}(\mathrm{CV}) = \frac{1}{\sqrt n} \sqrt{ \frac{\sum_{i=1}^n \left[ \mathrm{cv}(y_i, \widehat{y}_i^{(-i)} ) - \mathrm{CV} \right]^2 }{n - 1} }
-$$
-that is, as the standard deviation of the casewise components of CV divided by the square-root of the number of cases.
-
-We then use $\mathrm{SE}(\mathrm{CV})$ to construct a $100 \times (1 - \alpha)$% confidence interval around the *adjusted* CV estimate of error:
-$$
+$$ that is, as the standard deviation of the casewise components of CV divided by the square-root of the number of cases.
+
+We then use $\mathrm{SE}(\mathrm{CV})$ to construct a $100 \times (1 - \alpha)$% confidence interval around the *adjusted* CV estimate of error: $$
 \left[ \mathrm{CV}_{\mathrm{adj}} - z_{1 - \alpha/2}\mathrm{SE}(\mathrm{CV}), \mathrm{CV}_{\mathrm{adj}} + z_{1 - \alpha/2}\mathrm{SE}(\mathrm{CV})  \right]
-$$
-where $z_{1 - \alpha/2}$ is the $1 - \alpha/2$ quantile of the standard-normal distribution (e.g, $z \approx 1.96$ for a 95% confidence interval, for which $1 - \alpha/2 = .975$).
+$$ where $z_{1 - \alpha/2}$ is the $1 - \alpha/2$ quantile of the standard-normal distribution (e.g, $z \approx 1.96$ for a 95% confidence interval, for which $1 - \alpha/2 = .975$).
 

### 2024-07-08 scratch/notes_john/JSS-article/JSS-article-3.tex
some changes to John's JSS draft

@@
 
-{\centering \includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
+{\centering \subfloat[\label{fig:mpg-horsepower-polynomials-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-polynomials-1} }\subfloat[\label{fig:mpg-horsepower-polynomials-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-polynomials-2} }
 
@@
 
-\caption{\code{mpg} vs. \code{horsepower} for the \code{Auto} data, showing fitted polynomials of degree 1 through 5.}\label{fig:mpg-horsepower-scatterplot-polynomials}
+\caption{(a) \code{mpg} vs.
+\code{horsepower} for the \code{Auto} data, showing fitted polynomials of degree 1 through 5.
+(b) Estimated squared error as a function of polynomial degree, $p$.}\label{fig:mpg-horsepower-polynomials}
 \end{figure}
@@
 
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-MSE-se-1}
-
-}
-
-\caption[Estimated squared error as a function of polynomial degree, $p$]{Estimated squared error as a function of polynomial degree, $p$}\label{fig:mpg-horsepower-MSE-se}
-\end{figure}
-\end{CodeChunk}
-
 The generic \texttt{cv()} function has an \texttt{"lm"} method,
@@
 \begin{CodeOutput}
-R RNG seed set to 431467
+R RNG seed set to 180983
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.192
-bias-adjusted cross-validation criterion = 19.18107
-95% CI for bias-adjusted CV criterion = (15.71858, 22.64356)
+cross-validation criterion = 19.37365
+bias-adjusted cross-validation criterion = 19.35319
+95% CI for bias-adjusted CV criterion = (15.87805, 22.82834)
 full-sample criterion = 18.98477 
@@
 +                   smooth=list(spread=FALSE), data=Auto, pch=".")
+R> 
+R> m.auto <- lm(mpg ~ ., data = Auto)
+R> crPlots(m.auto, pch=".")
 \end{CodeInput}
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-explore-1}
+{\centering \subfloat[\label{fig:Auto-explore-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/Auto-explore-1} }\subfloat[\label{fig:Auto-explore-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/Auto-explore-2} }
 
@@
 
-\caption[Scatterplot matrix for the numeric variables in the \code{Auto} data]{Scatterplot matrix for the numeric variables in the \code{Auto} data.
-In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.}\label{fig:Auto-explore}
+\caption{(a) Scatterplot matrix for the numeric variables in the \code{Auto} data.
+In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.
+(b) Component+residual plots for the working model fit to the \code{Auto} data.
+The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.}\label{fig:Auto-explore}
 \end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> Auto$cylinders <- factor(Auto$cylinders,
++                          labels=c("3-4", "3-4", "5-6", "5-6", "8"))
+R> Auto$year <- as.factor(Auto$year)
+R> Auto$origin <- factor(Auto$origin,
++                       labels=c("America", "Europe", "Japan"))
+R> rownames(Auto) <- make.names(Auto$name, unique=TRUE)
+R> Auto$name <- NULL
+R> 
+R> scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
++                   smooth=list(spread=FALSE), data=Auto, pch=".")
+\end{CodeInput}
 \end{CodeChunk}
@@
 \end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-working-model-1}
-
-}
-
-\caption[Component+residual plots for the working model fit to the \code{Auto} data]{Component+residual plots for the working model fit to the \code{Auto} data.
-The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.}\label{fig:Auto-working-model}
-\end{figure}
 \end{CodeChunk}
@@
 \begin{CodeChunk}
+\begin{figure}
+
+{\centering \subfloat[\label{fig:Auto-transformed-scatterplot-matrix-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/Auto-transformed-scatterplot-matrix-1} }\subfloat[\label{fig:Auto-transformed-scatterplot-matrix-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/Auto-transformed-scatterplot-matrix-2} }
+
+}
+
+\caption{(a) Scatterplot matrix for the transformed numeric variables in the \code{Auto} data.
+(b) Component+residual plots for the working model fit to the transformed \code{Auto} data.}\label{fig:Auto-transformed-scatterplot-matrix}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
 \begin{CodeInput}
@@
 \end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-transformed-scatterplot-matrix-1}
-
-}
-
-\caption[Scatterplot matrix for the transformed numeric variables in the \code{Auto} data]{Scatterplot matrix for the transformed numeric variables in the \code{Auto} data}\label{fig:Auto-transformed-scatterplot-matrix}
-\end{figure}
 \end{CodeChunk}
@@
 \end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-CR-plots-transformed-1}
-
-}
-
-\caption[Component+residual plots for the model fit to the transformed \code{Auto} data]{Component+residual plots for the model fit to the transformed \code{Auto} data}\label{fig:Auto-CR-plots-transformed}
-\end{figure}
 \end{CodeChunk}
@@
   ``nested CV'' for the procedure described by
-  \citet{BatesHastieTibshirani:2023}.}}{Example: Applying recursive CV to polynomial regression for the Auto data}}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-datarecursive-cv}
+  \citet{BatesHastieTibshirani:2023}.}}{Example: Applying recursive CV to polynomial regression for the Auto data}}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-data9}
 
@@
 }
-<bytecode: 0x13fb26ea8>
+<bytecode: 0x10f0f2c50>
 <environment: namespace:cv>

### 2024-07-08 scratch/notes_john/JSS-article/JSS-article-4.Rmd
added JSS-article-4.Rmd and .pdf under scratch\notes_john. Edited Secs. 1-3.

--- 2024-07-08 scratch/notes_georges/JSS-article/JSS-article-3-GM.Rmd
@@
 options(prompt = 'R> ', continue = '+ ')
+options(digits=5)
 palette(car::carPalette())
@@
 Validating research by replication on independently collected data is a common scientific norm.
-Emulating this process in a single study by data-division is less common: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data.
+Emulating this process in a single study by data-division is less common [see, e.g., @Barnard:1974]: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data.
 Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples.
@@
 
-* The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.
+* The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a potentially complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.
 
@@
 
-```{r mpg-horsepower-scatterplot-polynomials, echo=FALSE}
-#| out.width = "50%",
+```{r mpg-horsepower-polynomials, fig.show="hold", echo=FALSE}
+#| out.width = "45%",
 #| fig.height = 5,
-#| fig.cap = "\\code{mpg} vs. \\code{horsepower} for the \\code{Auto} data, showing fitted polynomials of degree 1 through 5."
+#| fig.cap = "(a) \\code{mpg} vs.
+\\code{horsepower} for the \\code{Auto} data, showing fitted polynomials of degree 1 through 5.
+(b) Estimated squared error as a function of polynomial degree, $p = 1, \\ldots, 10$.",
+#| fig.subcap=rep("", 2)
 plot(mpg ~ horsepower, data=Auto)
@@
 
+We'll focus on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-polynomials} (a).
+The relationship between the two variables is monotone, decreasing, and nonlinear.
+Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-polynomials} (a).
+The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).
+
+[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here.
+Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader).
+We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.
+
+Figure \ref{fig:mpg-horsepower-polynomials} (b) shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$.
+The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.
+
 The generic `cv()` function has an `"lm"` method,
@@
 
-* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).
-
-* \code{ncores}, the number of cores to employ for parallel computation; if \code{cores = 1} (the default), the computations are not parallelized.
-
-To illustrate, we perform 10-fold CV for a quadratic polynomial fit to the \code{Auto} data:
+* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} (the default) selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).
+
+* \code{ncores}, the number of cores to employ for parallel computation; if \code{ncores = 1} (the default), the computations are not parallelized.
+
+To illustrate, we perform 10-fold CV for a quadratic polynomial fit to the \code{Auto} data:[^car-package]
+
+[^car-package]: We load the \pkg{car} package [@FoxWeisberg:2019] here for the \code{brief()} function and to use below.
+
 ```{r cv-lm-1}
+library("car", quietly=TRUE)
+
 m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
-summary(m.auto)
+brief(m.auto)
 
@@
 Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted.
-All three methods produce exact results for a linear model (within the precision of floating-point computations):
+All three methods produce exact results for a linear model (within the precision of floating-point computations); for example:
 ```{r cv.lm-3}
 cv(m.auto, k="loo", method="naive", confint=TRUE)
-
-cv(m.auto, k="loo", method="Woodbury", confint=TRUE)
 ```
@@
 #| fig.height = 5,
-#| fig.cap = "Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$.",
+#| fig.cap = "Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p = 1, 2, \\ldots, 10$.",
 #| fig.subcap=c('', ''),
@@
 
-In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model (cf., Figure \ref{fig:mpg-horsepower-MSE-se} on page \pageref{fig:mpg-horsepower-MSE-se}), except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting.
+In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model (cf., Figure \ref{fig:mpg-horsepower-polynomials} (b) on page \pageref{fig:mpg-horsepower-polynomials}), except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting.
 
@@
 The \code{lmer()} function in the \pkg{lme4} package, for example, is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects.
-We present an example of the latter in a vignette for the \pkg{cv} package.
+We present an example of crossed random effects in a vignette for the \pkg{cv} package.
 
@@
 
-Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
+Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data (summarizing the model with the \code{S()} function from the \pkg{car} package):
 ```{r HSB-lmer, cache=TRUE}
@@
                    + (cses | school), data=HSB)
-summary(hsb.lmer, correlation=FALSE)
+S(hsb.lmer, brief=TRUE)
 ```
@@
 
-The four data sets differ in the relative magnitude of between-cluster variance compared with within-cluster variance.
-They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.
-
-For these analyses, we use the `glmmTMB()` function in the **glmmTMB** package [@BrooksEtAl] because, in our experience, it is more likely to converge than functions in the **nlme** and the **lme4** packages for models with low between-cluster variance.
+The four data sets differ in the magnitude of between-cluster variance compared to within-cluster variance.
+They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.[^code-in-complements]
+
+[^code-in-complements]: The purpose of this section is to explain some of the complexities of cross-validating mixed-effects models, which motivates the approach that we take for mixed models in the \pkg{cv} package.
+The examples in this section introduce no features of the \pkg{cv} package that we haven't already encountered, and the generation of data for these examples involves unenlightening data management.
+Consequently, we don't show the code for data generation, model fitting, and graphs in this section; the code is in the complementary materials for the paper, however.
+For these analyses, we used the `glmmTMB()` function in the \pkg{glmmTMB} package [@BrooksEtAl] because, in our experience, it is more likely to converge than functions in the \pkg{nlme} and the \pkg{lme4} packages for models with low between-cluster variance.
+We used the \pkg{lattice} [@Sarkar:2008] and \pkg{latticeExtra} [@SarkarAndrews:2022] packages for the graphs in this section.
 
 Consider a researcher studying the effect of the dosage of a drug on the severity of symptoms for a hypothetical disease.
-The researcher has longitudinal data on 20 patients, each of whom was observed on five occasions in which patients received different dosages of the drug.
+The researcher has longitudinal data on 20 patients, each of whom was observed on 5 occasions in which patients received different dosages of the drug.
 The data are observational, with dosages prescribed by the patients' physicians, so that patients who were more severely affected by the disease received higher dosages of the drug.
@@
 
-To construct the four data sets, we first construct a template data set from a population with a between patient effect of dosage of $1.0$ and a within-patient effect of $-0.5$.
-We then apply four different multipliers to the between-patient variation from the regression line of patient centroids.
-The resulting data sets, shown in Figure \ref{fig:plot1}, differ in the between-patient variance of patient centroids from a common between-patient regression line, exhibiting a progression from low to high variance around the common regression line.[^lattice]
-
-Note that each patient has the same relative configuration of dosages and symptoms in each of the four data sets.
-To help visualize the structure of the data, estimated 50% concentration ellipses are shown for each patient.
-
-Using such data, the researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept.
-We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.
-The first three models are sequentially nested with fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`.
-The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`.
-We thus fit four models to four data sets for a total of 16 models.
+To generate the four data sets, we first constructed a template data set from a population with a between-patient effect of dosage of $1.0$ and a within-patient effect of $-0.5$.
+We then applied four different multipliers to the between-patient variation from the regression line of patient centroids.
+The resulting data sets, shown in Figure \ref{fig:plot1}, differ in the between-patient variance of patient centroids from a common between-patient regression line, exhibiting a progression from low to high variance around the common regression line.
+Each patient has the same relative configuration of dosages and symptoms in each of the four data sets.
+To help visualize the structure of the data, estimated 50% Gaussian concentration ellipses are shown for each patient.
 
@@
 ```
-```{r parameters, include=FALSE}
+```{r parameters, include=FALSE, cache=TRUE}
 # Parameters:
@@
 
-[^lattice]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.
-
-```{r plot1, echo=FALSE}
-#| out.width = "100%",
-#| fig.height = 3.0,
+```{r plot1, echo=FALSE, cache=TRUE}
+#| out.width = "75%",
+#| fig.height = 2.5,
 #| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance.
@@
 The ellipses are estimated 50% concentration ellipses for each patient.
-The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel."
+The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel, where $y$ is symptoms and $x$ is dosage."
 library("lattice")
@@
 xyplot(y ~ x | SD_ratio, data = Data, group = patient,
-       ylab = list('symptoms', cex = 0.7),
-       xlab = list('dosage', cex = 0.7),
+       ylab = list('symptoms (y)', cex = 0.7),
+       xlab = list('dosage (x)', cex = 0.7),
        par.strip.text = list(cex = 0.7),
@@
 
-```{r model-fits,include=FALSE}
+Using data like these, a researcher may attempt to obtain an estimate of the within-patient effect of dosage by fitting mixed models with a random intercept.
+We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.
+The first three models are sequentially nested in their fixed effects: (1) `~ 1`, intercept only; (2) `~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`.
+The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`.
+We thus fit four models to four data sets for a total of 16 models.
+
+```{r model-fits,include=FALSE, cache=TRUE}
 model.formulas <- c(
@@
 ```
-We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or "BLUPs"---as would be used for cross-validation based on cases (i.e., occasions within patients):
+
+We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---that is, the BLUPs---as would be used for cross-validation based on cases (i.e., occasions within patients):
 ```{r predict,include=FALSE}
@@
 ```
-```{r data-predictions,include=FALSE}
+```{r data-predictions,include=FALSE, cache=TRUE}
 Dataf <- lapply(split(Data, ~ SD_ratio),
@@
 ```
-The fixed- and random-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-combined} along with a summary of the data using 50% concentration ellipses.
+The fixed- and random-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-combined} along with a summary of the data using 50% concentration ellipses (see the top row of the figure).
 
@@
 
-{   xyplot(y ~ x | SD_ratio * panel3, Data,
+{ library("lattice")
+  library("latticeExtra")
+  xyplot(y ~ x | SD_ratio * panel3, Data,
          groups = patient, type = 'n',
@@
 
-Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`.
+Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions, and also of BLUPs, towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`.
 The inclusion of a contextual variable in the model corrects this problem.
 
-Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform with the within-patient slopes for the `~ 1 + x` model in the two data sets with the smallest between-patient variances.
+Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform to the within-patient slopes for the `~ 1 + x` model in the two data sets with the smallest between-patient variances.
 
 For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variance.
-With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.
+With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.
 
@@
 
-The third model, `~ 1 + x + xm`, includes a contextual effect of `x`---that is, the cluster mean `xm`---along with `x` and the intercept in the fixed-effect part of the model, along with a random intercept.
+The third model, `~ 1 + x + xm`, includes a contextual effect of `x`---that is, the cluster mean `xm`---along with `x` and the intercept in the fixed-effect part of the model, and with a random intercept.
 This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | patient)`, which is the model that generated the data.
-The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects only model with `x` and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).
-
-We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.
-In order to reduce between-model random variability in comparisons of models on the same data set, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.
-
-```{r echo=FALSE,include=FALSE}
+The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects-only model with `x` and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).
+
+We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.[^cv-list] The results for all data sets, using both cluster-based and case-based cross-validation, are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.
+
+[^cv-list]: In order to reduce between-model random variability in comparisons of models on the same data set, we applied `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.
+
+```{r echo=FALSE,include=FALSE,cache=TRUE}
 library(cv) # unclear why it's necessary to reload cv
@@
 ```
-The results for all data sets, using both cluster-based and case-based cross-validation, are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.
-```{r cross-validation-data,include=FALSE}
+
+```{r cross-validation-data,include=FALSE,cache=TRUE}
 names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
@@
 ```{r cross-validation-data-plot,echo=FALSE}
-#| out.width = "100%",
-#| fig.height = 3.3,
+#| out.width = "75%",
+#| fig.height = 3,
 #| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
@@
 The \code{"function"} method for \code{cv()} cross-validates a model-specification process in a general manner.
-Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package [@FoxWeisberg:2019], transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model.
+Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package, transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model.
 In a vignette on extending the \pkg{cv} package, we explain how to specify additional model-selection procedures.

### 2024-07-09 scratch/notes_john/JSS-article/JSS-article-4.Rmd
further edits to JSS-article-4

@@
 
-microbenchmark::microbenchmark(
+print(microbenchmark::microbenchmark(
   hatvalues = cv(m.auto, k="loo"),
@@
   cv.glm = boot::cv.glm(Auto, m.auto.glm),
-  times=10
-)
-```
-On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.
-
-<!-- [^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100.
+  times=10,
+  unit = "relative"
+  ), signif = 3)
+```
+On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.[^microbenchmark]
+
+[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100.
 With the exception of the last column, the output is self-explanatory.
 The last column shows which methods have average timings that are statistically distinguishable.
-Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default \code{times=100} repetitions.
--->
+Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the patient reader to redo this computation with the default \code{times=100} repetitions.
 
@@
 
-[^cv-list]: In order to reduce between-model random variability in comparisons of models on the same data set, we applied `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.
+[^cv-list]: In order to reduce between-model random variability in comparisons of models on the same data set, we applied `cv()` to a list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.
 
@@
 The \code{"function"} method for \code{cv()} cross-validates a model-specification process in a general manner.
-Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package, transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model.
-In a vignette on extending the \pkg{cv} package, we explain how to specify additional model-selection procedures.
+Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package, transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model.[^adding-model-selection]
+
+[^adding-model-selection]: In a vignette on extending the \pkg{cv} package, we explain how to specify additional model-selection procedures.
 
@@
 Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data.
-We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
-```{r Auto-explore}
-#| out.width = "60%",
+We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore} (a), produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
+```{r Auto-explore, fig.show="hold", echo=FALSE}
+#| out.width = "45%",
 #| fig.height = 5,
-#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data.
-In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression."
+#| fig.cap = "(a) Scatterplot matrix for the numeric variables in the \\code{Auto} data.
+In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.
+(b) Component+residual plots for the working model fit to the \\code{Auto} data.
+The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.",
+#| fig.subcap=rep("", 2)
 Auto$cylinders <- factor(Auto$cylinders,
@@
                   smooth=list(spread=FALSE), data=Auto, pch=".")
-```
+
+m.auto <- lm(mpg ~ ., data = Auto)
+crPlots(m.auto, pch=".")
+```
+
+```{r Auto-explore-show, eval=FALSE}
+Auto$cylinders <- factor(Auto$cylinders,
+                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
+Auto$year <- as.factor(Auto$year)
+Auto$origin <- factor(Auto$origin,
+                      labels=c("America", "Europe", "Japan"))
+rownames(Auto) <- make.names(Auto$name, unique=TRUE)
+Auto$name <- NULL
+
+scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
+                  smooth=list(spread=FALSE), data=Auto, pch=".")
+```
+
 A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors.
@@
 We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
-```{r Auto-working-model}
-#| out.width = "60%",
-#| fig.height = 5,
-#| fig.cap = "Component+residual plots for the working model fit to the \\code{Auto} data.
-The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths."
+```{r Auto-working-model, eval=FALSE}
 m.auto <- lm(mpg ~ ., data = Auto)
@@
 ```
-The component+residual plots in Figure \ref{fig:Auto-working-model} clearly reveal the inadequacy of the model.
+The component+residual plots in Figure \ref{fig:Auto-explore} (b), produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.
 
@@
 ```
-We then apply the (rounded) transformations---all, as it turns out, logs (i.e., "Oth" powers)---to the data and re-estimate the model:
+We then apply the (rounded) transformations---all, as it turns out, logs (i.e., "zeroth" powers)---to the data and re-estimate the model:
 ```{r Auto-with-transformed-predictors}
@@
 
-The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed-scatterplot-matrix}):
-```{r Auto-transformed-scatterplot-matrix}
-#| out.width = "60%",
+The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore} (a), on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed} (a)):
+```{r Auto-transformed, fig.show="hold", echo=FALSE}
+#| out.width = "45%",
 #| fig.height = 5,
-#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data"
+#| fig.cap = "(a) Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data. (b) Component+residual plots for the working model fit to the transformed \\code{Auto} data.",
+#| fig.subcap=rep("", 2)
 scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
@@
                   smooth=list(spread=FALSE), data=A, pch=".")
-```
-
-And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-working-model}, on page \pageref{fig:Auto-working-model}, and Figure \ref{fig:Auto-CR-plots-transformed}):
-```{r Auto-CR-plots-transformed}
-#| out.width = "60%",
-#| fig.height = 5,
-#| fig.cap = "Component+residual plots for the model fit to the transformed \\code{Auto} data"
 crPlots(m, pch=".")
@@
 
+```{r Auto-transformed-scatterplot-matrix-show, eval=FALSE}
+scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
+                  + acceleration, 
+                  smooth=list(spread=FALSE), data=A, pch=".")
+```
+
+And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore} (b), on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed} (b)):
+```{r Auto-CR-plots-transformed, eval=FALSE}
+crPlots(m, pch=".")
+```
+
 After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log n$):
-```{r}
+```{r stepAIC}
 library("MASS")
 m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
-summary(m.step)
+brief(m.step)
 ```
@@
 
-The other arguments to \code{cv()} are (see \code{?cv:::cv.function} for additional optional arguments and details):
+The other arguments to \code{cv()} are (see \code{?cv.function} for additional optional arguments and details):
 
@@
 
-microbenchmark::microbenchmark(
+print(microbenchmark::microbenchmark(
+  caret=train(x = attrition[, c("JobSatisfaction", "Gender",
+                                "MonthlyIncome")],
+              y = attrition$Attrition,
+              method = "glm",
+              trControl = trainControl(method="LOOCV")),
   rsample={
@@
   },
-  caret=train(x = attrition[, c("JobSatisfaction", "Gender",
-                                "MonthlyIncome")],
-              y = attrition$Attrition,
-              method = "glm",
-              trControl = trainControl(method="LOOCV")),
   direct={
@@
                   method="hatvalues"),
-  times=10
-)
-```
-Computing time for \pkg{rsample} and \pkg{caret} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.
+  times=10,
+  unit = "relative"
+  ), signif=3)
+```
+Computing time for \pkg{rsample} and \pkg{caret} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but almost two orders of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.
 

### 2024-07-09 scratch/notes_john/JSS-article/JSS-article-4.Rmd
finish first edit of JSS-article-4

@@
 In order of increasing general complexity, we can add: (1) a cross-validation cost criterion; (2) a model class that's not directly accommodated by the \code{cv()} default method or by another directly inherited method; and (3) a new model-selection procedure suitable for use with the \code{"function"} method for \code{cv()}.
-In this section, we illustrate (1) and (2); more diverse and extensive examples may be found in the vignette on extending the \pkg{cv} package.
+In this section, we illustrate (1) and (2); more diverse and extensive examples (including of (3)) may be found in the vignette on extending the \pkg{cv} package.
 
@@
 
-<!-- We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024] (and which is by a wide margin the most downloaded from the RStudio CRAN mirror of the CV packages mentioned in the preceding paragraph).
--->
-
 We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024], comparing LOO CV computed using the \code{cv()} function in the \pkg{cv} package to LOO CV computed using the \pkg{caret} and \pkg{rsample} packages.
@@
 ```{r rsample-example-using-caret, cache=TRUE}
-library("caret", quietly=TRUE, warn.conflicts=FALSE)
+library("ggplot2", warn.conflicts=FALSE) # used by caret
+library("caret", quietly=TRUE)
 # reports complement of Bayes-rule criterion:
@@
 ```
+where "accuracy" is the complement of the misclassification rate.
 
@@
 
-Let's compare the computational efficiency of the various implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
+Let's compare the computational efficiency of the various implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:[^times-10]
+
+[^times-10]: As before, we ask \code{microbenchmark()} to report relative timings, and also as before, we set \code{times = 10} rather than the default \code{times = 100}, with the consequence that we fail to detect a statistically reliable difference between using the Woodbury matrix identity and using hatvalues for the \code{cv()} computations.
+
 ```{r compare-rsample-cv, cache=TRUE}
-all.equal(cv.attrition$`CV crit`,
-          cv(mod.attrition, k="loo", criterion=BayesRule,
-             method="Woodbury")$`CV crit`,
-          )
-all.equal(cv.attrition$`CV crit`,
-          cv(mod.attrition, k="loo", criterion=BayesRule,
-             method="hatvalues")$`CV crit`
-          )
-
 print(microbenchmark::microbenchmark(

### 2024-07-10 scratch/notes_john/JSS-article/JSS-article-4.Rmd
further edits to JSS-article-4

@@
 See @Harrell:2015 [Sec.
-5.3] for this and other remarks about data-division and cross-validation.
+5.3] for cogent remarks about data-division and cross-validation.
 
@@
 A CV criterion (also termed a "cost" or "loss" measure), such as the mean-squared error ("MSE") of prediction, is then computed using these predicted values.
-In the extreme $k = n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."
+In the extreme, $k = n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."
 
 Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of prediction error.
-The $n$ regression models are highly statistical dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has relatively large variance.
-In contrast, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) are somewhat biased but have smaller variance.
-It is also possible to correct $k$-fold CV for bias (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+The $n$ regression models are highly statistically dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has larger variance than if the predictions from the models fit to the $n$ data sets were independent.
+
+Because predictions are based on smaller data sets, each of size approximately $n - n/k$, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) is more biased than estimated prediction error for LOO CV.
+It is possible, however, to correct $k$-fold CV for bias (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+
+The relative *variance* of prediction error for LOO CV and $k$-fold CV (with $k < n$) is more complicated: Because the overlap between the data sets with each fold omitted is smaller for $k$-fold CV than for LOO CV, the dependencies among the predictions are smaller for the former than for the latter, tending to produce smaller variance in prediction error for $k$-fold CV.
+In contrast, there are factors that tend to inflate the variance of prediction error in $k$-fold CV, including the reduced size of the data sets with each fold omitted and the randomness induced by the selection of folds---in LOO CV the folds are not random.
 
@@
 ```
+ The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.
+
 * The \code{"lm"} and \code{"glm"} methods are for linear and generalized-linear models fit respectively by the standard \proglang{R} \code{lm()} and \code{glm()} functions.
@@
 * The \code{"modList"} method for \code{cv()} cross-validates several competing models, not necessarily of the same class, using the same division of the data into folds.
-The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.
 
@@
 
-The final section of the paper (Section \ref{computational-notes}) describes some computational details, including efficient CV computations for linear and generalized-linear models and computation of bias-corrected CV criteria.
+The final section of the paper (Section \ref{computational-notes}) describes some computational details, including efficient CV computations for linear and generalized-linear models.
+and computation of bias-corrected CV criteria.
 
@@
 
+set.seed(19412) # for reproducibility
 print(microbenchmark::microbenchmark(
@@
 The \code{lmer()} function in the \pkg{lme4} package, for example, is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects.
-We present an example of crossed random effects in a vignette for the \pkg{cv} package.
+We present an example of crossed random effects in the \pkg{cv} package vignette on mixed models.
 
@@
 
-We introduce four artificial data sets that exemplify aspects of cross-validation particular to hierarchical models.
+In this section, we introduce four artificial data sets that exemplify aspects of cross-validation particular to hierarchical models.
 Using these data sets, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model.
@@
 The four data sets differ in the magnitude of between-cluster variance compared to within-cluster variance.
-They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.[^code-in-complements]
-
-[^code-in-complements]: The purpose of this section is to explain some of the complexities of cross-validating mixed-effects models, which motivates the approach that we take for mixed models in the \pkg{cv} package.
+They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.
+
+The purpose of this section is to explain some of the complexities of cross-validating mixed-effects models, which motivates the approach that we take for mixed models in the \pkg{cv} package.
 The examples in this section introduce no features of the \pkg{cv} package that we haven't already encountered, and the generation of data for these examples involves unenlightening data management.
-Consequently, we don't show the code for data generation, model fitting, and graphs in this section; the code is in the complementary materials for the paper, however.
-For these analyses, we used the `glmmTMB()` function in the \pkg{glmmTMB} package [@BrooksEtAl] because, in our experience, it is more likely to converge than functions in the \pkg{nlme} and the \pkg{lme4} packages for models with low between-cluster variance.
-We used the \pkg{lattice} [@Sarkar:2008] and \pkg{latticeExtra} [@SarkarAndrews:2022] packages for the graphs in this section.
-
-Consider a researcher studying the effect of the dosage of a drug on the severity of symptoms for a hypothetical disease.
+Consequently, we don't show the code for data generation, model fitting, and graphs in this section.[^code-in-complements]
+
+[^code-in-complements]: The code for this section is in the complementary materials for the paper.
+We used the `glmmTMB()` function in the \pkg{glmmTMB} package [@BrooksEtAl] for the models fit here because, in our experience, it is more likely to converge than functions in the \pkg{nlme} and the \pkg{lme4} packages for models with low between-cluster variance.
+We used the \pkg{lattice} [@Sarkar:2008] and \pkg{latticeExtra} [@SarkarAndrews:2022] packages for the graphs.
+
+Consider a researcher studying the effect of the dosage $x$ of a drug on the severity of symptoms $y$ for a hypothetical disease.
 The researcher has longitudinal data on 20 patients, each of whom was observed on 5 occasions in which patients received different dosages of the drug.
@@
 #| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance.
-The labels above each panel show the between-patient SD/within-patient SD.
-The ellipses are estimated 50% concentration ellipses for each patient.
+The labels above each panel show the ratio of between-patient SD to within-patient SD.
+The ellipses are estimated 50% Gaussian concentration ellipses for each patient.
 The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel, where $y$ is symptoms and $x$ is dosage."
@@
 The first three models are sequentially nested in their fixed effects: (1) `~ 1`, intercept only; (2) `~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`.
-The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`.
+The final model, (4) `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`.
 We thus fit four models to four data sets for a total of 16 models.
@@
 
-We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---that is, the BLUPs---as would be used for cross-validation based on cases (i.e., occasions within patients):
+We proceed to obtain predictions from each model (1) using only the fixed effects, as for cross-validation based on clusters (i.e., patients), and (2) using both fixed and random effects---that is, the BLUPs---as for cross-validation based on cases (i.e., occasions within patients).
+The fixed- and random-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-combined} along with a summary of the data using 50% concentration ellipses (see the top row of the figure).
 ```{r predict,include=FALSE}
@@
 ```
-The fixed- and random-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-combined} along with a summary of the data using 50% concentration ellipses (see the top row of the figure).
 
@@
 #| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient variance ratios.
-The top row shows summaries of the within-patient data using estimated 50% concentration ellipses."
+The top row shows summaries of the within-patient data using estimated 50% concentration ellipses, along with the patient centroids."
 levs3 <- levels(Data$panel2)
@@
 Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions, and also of BLUPs, towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`.
-The inclusion of a contextual variable in the model corrects this problem.
+The inclusion of the contextual variable `xm` in the model corrects this problem.
 
 Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform to the within-patient slopes for the `~ 1 + x` model in the two data sets with the smallest between-patient variances.
-
 For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variance.
@@
 We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, `~ 1`  to the larger `~ 1 + x` model when the between-cluster variance is relatively small, but that cluster-based cross-validation will prefer the latter to the former.
-
 We will discover that case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model for the '5 / 2.5' data set, but cluster-based cross-validation prefers the latter model to the former.
@@
 ```
-In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
-With all our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.
+In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean `xm` of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
+In all of our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.
 These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

### 2024-07-10 scratch/notes_john/JSS-article/JSS-article-4.Rmd
finished edit of JSS-article-4

@@
 A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors.
-There are, however, only 5 distinct values of \code{cylinders} (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders.
+There are, however, only 5 distinct values of \code{cylinders}, ranging from 3 to 8, but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders.
 There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.
@@
 #| fig.height = 5,
-#| fig.cap = "(a) Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data. (b) Component+residual plots for the working model fit to the transformed \\code{Auto} data.",
+#| fig.cap = "(a) Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data. (b) Component+residual plots for the regression model fit to the transformed \\code{Auto} data.",
 #| fig.subcap=rep("", 2)
@@
 cvs
-
-compareFolds(cvs)
-```
-
-The other arguments to \code{cv()} are (see \code{?cv.function} for additional optional arguments and details):
+```
+
+The other arguments to \code{cv()} are:[^cv.function-args]
+
+[^cv.function-args]: See \code{?cv.function} for additional optional arguments and details.
 
@@
 * As we anticipated, the estimate of the MSE that we obtain by cross-validating the whole model-specification process is larger than the MSE computed for the model we fit to the \code{Auto} data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
-* When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of \code{compareFolds()}), we see that there is little uncertainty in choosing variable transformations (the \code{lam.*}s for the $x$s and \code{lambda} for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: \code{horsepower}, \code{weight}, and \code{year} are always included among the selected predictors; \code{acceleration} and \code{displacement} are included respectively in 4 and 3 of 10 selected models; and \code{cylinders} and \code{origin} are each included in only 1 of 10 models.
+* When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of \code{compareFolds(cvs)}, which isn't shown), we see that there is little uncertainty in choosing variable transformations, but considerably more uncertainty in subsequently selecting predictors: \code{horsepower}, \code{weight}, and \code{year} are always included among the selected predictors; \code{acceleration} and \code{displacement} are included respectively in 4 and 3 of 10 selected models; and \code{cylinders} and \code{origin} are each included in only 1 of 10 models.
 Recall that when we selected predictors for the full data, we obtained a model with \code{horsepower}, \code{weight}, \code{acceleration}, \code{year}, and \code{origin}.
@@
 recursiveCV.auto
-recursiveCV.auto$selected.model
+brief(recursiveCV.auto$selected.model)
 cv(mlist[[7]], seed=2120) # same seed for same folds
@@
 
-We can equivalently call `cv()` with the list of models as its first argument and set the argument `recursive=TRUE`:
-```{r recursive-cv-alt}
+We can equivalently call `cv()` with the list of models as its first argument and set the argument `recursive=TRUE` (identical output not shown):
+```{r recursive-cv-alt, eval=FALSE}
 cv(mlist, data=Auto, seed=2120, recursive=TRUE, save.model=TRUE)
@@
                        economic.cond.household + Blair + Hague + Kennedy +
-                       Europe*political.knowledge, data=BEPS)
-
-```
-<!-- We set the argument \code{trace=FALSE} in the call to \code{multinom()} to suppress reporting the iteration history, which will be particularly appreciated when we repeatedly refit the model in cross-validation.
--->
-
-The \code{Europe} $\times$ \code{political.knowledge} interaction is associated with a very small $p$-value.
-Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the interaction in a "stacked-area" graph:
+                       Europe*political.knowledge, data=BEPS, trace=FALSE)
+
+```
+
+The  interaction is associated with a very small $p$-value.
+Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the \code{Europe} $\times$ \code{political.knowledge} interaction in a "stacked-area" graph:
 ```{r BEPS-plot, fig.width=9, fig.height=5}
@@
 
-Other packages, including \pkg{cvms} [@OlsenZachariae:2024], \pkg{mlexperiments} [@Kapsner:2024], \pkg{origami} [@CoyleEtAl:2022], and \pkg{rsample} (which also supports bootstrapping), modularize the CV process to a greater or lesser extent.
+Other packages, including \pkg{cvms} [@OlsenZachariae:2024], \pkg{mlexperiments} [@Kapsner:2024], \pkg{origami} [@CoyleEtAl:2022], and \pkg{rsample} (which also implements bootstrapping), modularize the CV process to a greater or lesser extent.
 Advantages of this approach include flexibility and generality, but a disadvantage is that use of these packages entails nontrivial programming effort and skill.
@@
 ```{r compare-rsample-cv, cache=TRUE}
+set.seed(53437)
 print(microbenchmark::microbenchmark(
-  caret=train(x = attrition[, c("JobSatisfaction", "Gender",
-                                "MonthlyIncome")],
-              y = attrition$Attrition,
-              method = "glm",
-              trControl = trainControl(method="LOOCV")),
-  rsample={
-    rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
-    rs_obj$accuracy <- map_dbl(rs_obj$results,
-                               function(x) mean(x$correct))
-  },
+  cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
+                  method="hatvalues"),
+  cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
+             method="Woodbury"),
+  cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
   direct={
@@
   },
-  cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
-  cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
-             method="Woodbury"),
-  cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
-                  method="hatvalues"),
+  rsample={
+    rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
+    rs_obj$accuracy <- map_dbl(rs_obj$results,
+                               function(x) mean(x$correct))
+  },
+  caret=train(x = attrition[, c("JobSatisfaction", "Gender",
+                                "MonthlyIncome")],
+              y = attrition$Attrition,
+              method = "glm",
+              trControl = trainControl(method="LOOCV")),
   times=10,

### 2024-07-10 scratch/notes_john/JSS-article/JSS-article-4.Rmd
apply JSS style to JSS-article-4 (first pass)

@@
 ```{r cv-methods}
-library("cv", quietly=TRUE)
+library("cv", quietly = TRUE)
 methods("cv")
@@
 ```{r Auto}
-data("Auto", package="ISLR2")
+data("Auto", package = "ISLR2")
 summary(Auto)
@@
 ```{r cv-lm-1}
-library("car", quietly=TRUE)
-
-m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
+library("car", quietly = TRUE)
+
+m.auto <- lm(mpg ~ poly(horsepower, 2), data = Auto)
 brief(m.auto)
 
-cv(m.auto, confint=TRUE)
+cv(m.auto, confint = TRUE)
 ```
@@
 ```{r cv.lm-2`}
-cv(m.auto, k="loo")
+cv(m.auto, k = "loo")
 ```
@@
 ```{r cv.lm-3}
-cv(m.auto, k="loo", method="naive", confint=TRUE)
+cv(m.auto, k = "loo", method = "naive", confint = TRUE)
 ```
@@
 This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed.
-In the following comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM.
+In the following comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM; the output shown is for the CV estimate of MSE and the bias-corrected CV estimate.
 We use the \code{microbenchmark()} function from the package of the same name  [@Mersmann:2023] for the timings:
 ```{r cv.lm.timings, cache=TRUE}
-m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
-boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE
-
-set.seed(19412) # for reproducibility
+m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data = Auto)
+boot::cv.glm(Auto, m.auto.glm)$delta
+
+set.seed(19412)
 print(microbenchmark::microbenchmark(
-  hatvalues = cv(m.auto, k="loo"),
-  Woodbury = cv(m.auto, k="loo", method="Woodbury"),
-  naive = cv(m.auto, k="loo", method="naive"),
+  hatvalues = cv(m.auto, k = "loo"),
+  Woodbury = cv(m.auto, k = "loo", method = "Woodbury"),
+  naive = cv(m.auto, k = "loo", method = "naive"),
   cv.glm = boot::cv.glm(Auto, m.auto.glm),
-  times=10,
-  unit = "relative"
-  ), signif = 3)
+  times = 10, unit = "relative"), signif = 3)
 ```
@@
 ```{r polyomial-models}
-mlist <- vector(10, mode="list")
+mlist <- vector(10, mode = "list")
 for (p in 1:10) mlist[[p]] <- lm(mpg ~ poly(horsepower, p), data = Auto)
 names(mlist) <- paste0("m.", 1:10)
-mlist[2] # e.g., the quadratic fit
-```
-We then apply \code{cv()} to the list of 10 models (the \code{data} argument is required):
+mlist[2]
+```
+where, for example \code{mlist[2]} is the quadratic fit.
+
+We then apply \code{cv()} to the list of 10 models for both 10-fold and LOO CV (the \code{data} argument to \code{cv()} is required):
 ```{r polynomial-regression-CV}
-# 10-fold CV
-mlist <- models(mlist) # create "modList" object
-
-cv.auto.10 <- cv(mlist, data=Auto, seed=2120)
-cv.auto.10[2] # e.g., for quadratic model
-
-# LOO CV
-cv.auto.loo <- cv(mlist, data=Auto, k="loo")
-cv.auto.loo[2] # e.g., for quadratic model
+mlist <- models(mlist)
+
+cv.auto.10 <- cv(mlist, data = Auto, seed = 2120)
+cv.auto.10[2]
+
+cv.auto.loo <- cv(mlist, data = Auto, k = "loo")
+cv.auto.loo[2]
 ```
@@
 #| fig.ncol = 2, out.width = "50%", fig.align = "center"
-plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",
-     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
-plot(cv.auto.loo, main="Polynomial Regressions, LOO CV",
-     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
+plot(cv.auto.10, main = "Polynomial Regressions, 10-Fold CV",
+     axis.args = list(labels = 1:10), xlab = "Degree of Polynomial, p")
+plot(cv.auto.loo, main = "Polynomial Regressions, LOO CV",
+     axis.args = list(labels = 1:10), xlab = "Degree of Polynomial, p")
 ```
@@
 ```{r HSB-data}
-data("MathAchieve", package="nlme")
+data("MathAchieve", package = "nlme")
 dim(MathAchieve)
@@
 
-data("MathAchSchool", package="nlme")
+data("MathAchSchool", package = "nlme")
 dim(MathAchSchool)
@@
 HSB <- merge(MathAchSchool[, c("School", "Sector")],
-             HSB[, c("School", "SES", "MathAch")], by="School")
+             HSB[, c("School", "SES", "MathAch")], by = "School")
 names(HSB) <- tolower(names(HSB))
@@
 ```{r HSB-lmer, cache=TRUE}
-library("lme4", quietly=TRUE)
+library("lme4", quietly = TRUE)
 hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
-                   + (cses | school), data=HSB)
-S(hsb.lmer, brief=TRUE)
+                   + (cses | school), data = HSB)
+S(hsb.lmer, brief = TRUE)
 ```
@@
 ```{r HSB-lmer-CV-cluster, cache=TRUE}
-cv(hsb.lmer, k=10, clusterVariables="school", seed=5240)
+cv(hsb.lmer, k = 10, clusterVariables = "school", seed = 5240)
 ```
@@
 ```{r HSB-lmer-CV-case, cache=TRUE}
-cv(hsb.lmer, seed=1575)
+cv(hsb.lmer, seed = 1575)
 ```
@@
 names(Auto)
-xtabs(~ year, data=Auto)
-xtabs(~ origin, data=Auto)
-xtabs(~ cylinders, data=Auto)
+xtabs(~ year, data = Auto)
+xtabs(~ origin, data = Auto)
+xtabs(~ cylinders, data = Auto)
 ```
@@
 In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.
-(b) Component+residual plots for the working model fit to the \\code{Auto} data.
+(b) Component + residual plots for the working model fit to the \\code{Auto} data.
 The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.",
@@
 Auto$cylinders <- factor(Auto$cylinders,
-                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
+                         labels = c("3-4", "3-4", "5-6", "5-6", "8"))
 Auto$year <- as.factor(Auto$year)
 Auto$origin <- factor(Auto$origin,
-                      labels=c("America", "Europe", "Japan"))
-rownames(Auto) <- make.names(Auto$name, unique=TRUE)
+                      labels = c("America", "Europe", "Japan"))
+rownames(Auto) <- make.names(Auto$name, unique = TRUE)
 Auto$name <- NULL
 
-scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
-                  smooth=list(spread=FALSE), data=Auto, pch=".")
+scatterplotMatrix(~ mpg + displacement + horsepower + weight 
+                  + acceleration,
+                  smooth = list(spread = FALSE), data = Auto, pch= ".")
 
 m.auto <- lm(mpg ~ ., data = Auto)
-crPlots(m.auto, pch=".")
+crPlots(m.auto, pch = ".")
 ```
@@
 Auto$cylinders <- factor(Auto$cylinders,
-                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
+                         labels = c("3-4", "3-4", "5-6", "5-6", "8"))
 Auto$year <- as.factor(Auto$year)
 Auto$origin <- factor(Auto$origin,
-                      labels=c("America", "Europe", "Japan"))
-rownames(Auto) <- make.names(Auto$name, unique=TRUE)
+                      labels = c("America", "Europe", "Japan"))
+rownames(Auto) <- make.names(Auto$name, unique = TRUE)
 Auto$name <- NULL
 
-scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
-                  smooth=list(spread=FALSE), data=Auto, pch=".")
+scatterplotMatrix(~ mpg + displacement + horsepower + weight 
+                  + acceleration, 
+                  smooth = list(spread = FALSE), data = Auto, pch = ".")
 ```
@@
 m.auto <- lm(mpg ~ ., data = Auto)
-crPlots(m.auto, pch=".")
-```
-The component+residual plots in Figure \ref{fig:Auto-explore} (b), produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.
+crPlots(m.auto, pch = ".")
+```
+The component + residual plots in Figure \ref{fig:Auto-explore} (b), produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.
 
@@
 for (pred in num.predictors){
-  A[, pred] <- bcPower(A[, pred], lambda=powers[pred])
+  A[, pred] <- bcPower(A[, pred], lambda = powers[pred])
 }
-m <- update(m.auto, data=A)
+m <- update(m.auto, data = A)
 ```
@@
 #| fig.height = 5,
-#| fig.cap = "(a) Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data. (b) Component+residual plots for the regression model fit to the transformed \\code{Auto} data.",
+#| fig.cap = "(a) Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data. (b) Component + residual plots for the regression model fit to the transformed \\code{Auto} data.",
 #| fig.subcap=rep("", 2)
@@
                   + acceleration, 
-                  smooth=list(spread=FALSE), data=A, pch=".")
-crPlots(m, pch=".")
+                  smooth=list(spread = FALSE), data = A, pch = ".")
+crPlots(m, pch = ".")
 ```
@@
                   + acceleration, 
-                  smooth=list(spread=FALSE), data=A, pch=".")
+                  smooth = list(spread = FALSE), data = A, pch = ".")
 ```
@@
 ```{r Auto-CR-plots-transformed, eval=FALSE}
-crPlots(m, pch=".")
+crPlots(m, pch = ".")
 ```
@@
 library("MASS")
-m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
+m.step <- stepAIC(m, k = log(nrow(A)), trace = FALSE)
 brief(m.step)
@@
 num.predictors
-cvs <- cv(selectTransStepAIC, data=Auto, seed=76692, 
-          working.model=m.auto,
-          predictors=num.predictors,
-          response="mpg", AIC=FALSE)
+cvs <- cv(selectTransStepAIC, data = Auto, seed = 76692, 
+          working.model = m.auto, predictors = num.predictors,
+          response = "mpg", AIC = FALSE)
 cvs
@@
 ```{r recursive-CV-polynomials}
-recursiveCV.auto <- cv(selectModelList, data=Auto, 
-                       working.model=mlist,
-                       save.model=TRUE, seed=2120)
+recursiveCV.auto <- cv(selectModelList, data = Auto, 
+                       working.model = mlist, save.model = TRUE, 
+                       seed = 2120)
 recursiveCV.auto
 brief(recursiveCV.auto$selected.model)
-cv(mlist[[7]], seed=2120) # same seed for same folds
-```
-As expected, recursive CV produces a larger estimate of MSE for the selected 7th degree polynomial model than CV applied directly to this model.
+cv(mlist[[7]], seed = 2120)
+```
+As expected, recursive CV produces a larger estimate of MSE for the selected 7th degree polynomial model than CV applied directly to this model (using the same seed for \proglang{R}'s random-number generator to obtain the same folds).
 
@@
 ```{r recursive-cv-alt, eval=FALSE}
-cv(mlist, data=Auto, seed=2120, recursive=TRUE, save.model=TRUE)
+cv(mlist, data = Auto, seed = 2120, recursive = TRUE, save.model = TRUE)
 ```
@@
 ```{r BEPS-data}
-data("BEPS", package="carData")
+data("BEPS", package = "carData")
 summary(BEPS)
@@
 ```{r BEPS-model}
-library("nnet", quietly=TRUE)
-m.beps <- multinom(vote ~ age + gender + economic.cond.national +
-                       economic.cond.household + Blair + Hague + Kennedy +
-                       Europe*political.knowledge, data=BEPS, trace=FALSE)
+library("nnet", quietly = TRUE)
+m.beps <- multinom(vote ~ age + gender + economic.cond.national
+                        + economic.cond.household + Blair + Hague 
+                        + Kennedy + Europe * political.knowledge, 
+                   data = BEPS, trace = FALSE)
 
@@
 ```{r BEPS-plot, fig.width=9, fig.height=5}
-#| fig.cap = "Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study."
+#| fig.cap = "Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study, using party colors."
 plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
-            xlevels=list(Europe=1:11, political.knowledge=0:3),
-            fixed.predictors=list(given.values=c(gendermale=0.5))),
-     lines=list(col=c("blue", "red", "orange")), # party colors
-     axes=list(x=list(rug=FALSE), y=list(style="stacked")))
+            xlevels = list(Europe = 1:11, political.knowledge = 0:3),
+            fixed.predictors = list(given.values = c(gendermale = 0.5))),
+     lines = list(col = c("blue", "red", "orange")),
+     axes = list(x = list(rug = FALSE), y = list(style = "stacked")))
 ```
@@
 head(BEPS$vote)
-yhat <- predict(m.beps, type="class")
+yhat <- predict(m.beps, type = "class")
 head(yhat)
@@
 ```{r BEPS-response-distribution}
-xtabs(~ vote, data=BEPS)/nrow(BEPS)
+xtabs(~ vote, data = BEPS) / nrow(BEPS)
 ```
@@
 ```{r BEPS-test-default, error=TRUE}
-cv(m.beps, seed=3465, criterion=BayesRuleMulti)
+cv(m.beps, seed = 3465, criterion = BayesRuleMulti)
 ```
@@
 ```{r BEPS-test-default-2, error=TRUE}
-cv(m.beps, seed=3465, criterion=BayesRuleMulti)
+cv(m.beps, seed = 3465, criterion = BayesRuleMulti)
 ```
@@
     NextMethod(
-      type = "class",
-      criterion = criterion,
+      type = "class", criterion = criterion,
       criterion.name = deparse(substitute(criterion))
@@
 ```{r BEPS-cv}
-cv(m.beps, seed=3465)
+cv(m.beps, seed = 3465)
 ```
@@
 
-# formula for model to be fit; Attrition is binary:
 mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome
 
-rs_obj <- loo_cv(attrition) # original uses vfold_cv()
-
-# user-supplied function:
+rs_obj <- loo_cv(attrition)
+
 holdout_results <- function(splits, ...) {
-  mod <- glm(..., data=analysis(splits), family=binomial)
+  mod <- glm(..., data = analysis(splits), family = binomial)
   holdout <- assessment(splits)
-  res <- broom::augment(mod, newdata=holdout)
+  res <- broom::augment(mod, newdata = holdout)
   lvls <- levels(holdout$Attrition)
@@
 library("purrr", warn.conflicts=FALSE)
-# apply CV:
+
 rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
-# summarize results:
-rs_obj$accuracy <- map_dbl(rs_obj$results,
-                           function(x) mean(x$correct))
-# accuracy is the complement of the Bayes-rule CV criterion:
+rs_obj$accuracy <- map_dbl(rs_obj$results, function(x) mean(x$correct))
 1 - summary(rs_obj$accuracy) 
 ```
+The response variable in this example, \code{Attrition}, is binary.
+The function \code{holdout_results()} must be supplied by the user, and "accuracy" is the complement of the Bayes-rule classification error rate.
+
 [^k-fold-used-in-rsample]: The example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise the code from the vignette is only trivially modified here.
 
-In this case, it's simple (and, in our opinion, simpler) to program LOO CV directly. For example
+In this case, it's simple (and, in our opinion, simpler) to program LOO CV directly. For example,
 ```{r direct-loo-cv, cache=TRUE}
-mod.attrition <- glm(mod_form, data=attrition, family=binomial)
+mod.attrition <- glm(mod_form, data = attrition, family = binomial)
 n <- nrow(attrition)
@@
 for (i in 1:n){
-  m <- update(mod.attrition, data=attrition[-i, ])
-  yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
+  m <- update(mod.attrition, data = attrition[-i, ])
+  yhat[i] <- predict(m, newdata = attrition[i, ], type = "response")
 }
-mean(((attrition$Attrition == "Yes") - round(yhat))^2)
+mean(((attrition$Attrition == "Yes") - round(yhat)) ^ 2)
 ```
@@
 ```{r rsample-example-using-caret, cache=TRUE}
-library("ggplot2", warn.conflicts=FALSE) # used by caret
-library("caret", quietly=TRUE)
-# reports complement of Bayes-rule criterion:
-train(x = attrition[, c("JobSatisfaction", "Gender",
-                               "MonthlyIncome")],
-             y = attrition$Attrition,
-             method = "glm",
-             trControl = trainControl(method="LOOCV"))
-```
-where "accuracy" is the complement of the misclassification rate.
+library("ggplot2", warn.conflicts = FALSE)
+library("caret", quietly = TRUE)
+train(x = attrition[, c("JobSatisfaction", "Gender", "MonthlyIncome")],
+      y = attrition$Attrition, method = "glm",
+      trControl = trainControl(method = "LOOCV"))
+```
+where "accuracy", again, is the complement of the misclassification rate.
 
@@
 ```{r rsample-example-using-cv, cache=TRUE}
-(cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
+(cv.attrition <- cv(mod.attrition, k = "loo", criterion = BayesRule))
 all.equal(cv.attrition$`CV crit`, 1 - mean(rs_obj$accuracy), 
-          check.attributes=FALSE)
+          check.attributes = FALSE)
 ```
@@
 print(microbenchmark::microbenchmark(
-  cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
-                  method="hatvalues"),
-  cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
-             method="Woodbury"),
-  cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
-  direct={
+  cv.hatvalues = cv(mod.attrition, k = "loo", criterion = BayesRule,
+                    method = "hatvalues"),
+  cv.wood = cv(mod.attrition, k = "loo", criterion = BayesRule,
+               method = "Woodbury"),
+  cv.exact = cv(mod.attrition, k = "loo", criterion = BayesRule),
+  direct = {
     n <- nrow(attrition)
@@
     for (i in 1:n){
-      m <- update(mod.attrition, data=attrition[-i, ])
-      yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
+      m <- update(mod.attrition, data = attrition[-i, ])
+      yhat[i] <- predict(m, newdata = attrition[i, ], type = "response")
     }
-    mean(((attrition$Attrition == "Yes") - round(yhat))^2)
+    mean(((attrition$Attrition == "Yes") - round(yhat)) ^ 2)
   },
-  rsample={
+  rsample = {
     rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
@@
   },
-  caret=train(x = attrition[, c("JobSatisfaction", "Gender",
+  caret = train(x = attrition[, c("JobSatisfaction", "Gender",
                                   "MonthlyIncome")],
@@
               method = "glm", 
-              trControl = trainControl(method="LOOCV")),
-  times=10,
-  unit = "relative"
-  ), signif=3)
+              trControl = trainControl(method = "LOOCV")),
+  times = 10, unit = "relative"), signif = 3)
 ```
@@
 $$
-\mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} 
-  \mathbf{X}^T \mathbf{W} \mathbf{y}
+\mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^\top \mathbf{W} \mathbf{X} \right)^{-1} 
+  \mathbf{X}^\top \mathbf{W} \mathbf{y}
 $$
@@
 
-The LOO fitted value for the $i$th case can be efficiently computed by $\widehat{y}_{-i} = y_i - e_i/(1 - h_i)$ where $h_i = \mathbf{x}^T_i \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} \mathbf{x}_i$ (the so-called "hatvalue").
-Here, $\mathbf{x}^T_i$ is the $i$th row of $\mathbf{X}$, and $\mathbf{x}_i$ is the $i$th row written as a column vector.
+The LOO fitted value for the $i$th case can be efficiently computed by $\widehat{y}_{-i} = y_i - e_i/(1 - h_i)$ where $h_i = \mathbf{x}^\top_i \left( \mathbf{X}^\top \mathbf{W} \mathbf{X} \right)^{-1} \mathbf{x}_i$ (the so-called "hatvalue").
+Here, $\mathbf{x}^\top_i$ is the $i$th row of $\mathbf{X}$, and $\mathbf{x}_i$ is the $i$th row written as a column vector.
 This approach can break down when one or more hatvalues are equal to 1, in which case the formula for $\widehat{y}_{-i}$ requires division by 0.
@@
 \begin{align*}
-	\mathbf{A} &= \mathbf{X}^T \mathbf{W} \mathbf{X} \\
-	\mathbf{U} &= \mathbf{X}_\mathbf{j}^T \\
+	\mathbf{A} &= \mathbf{X}^\top \mathbf{W} \mathbf{X} \\
+	\mathbf{U} &= \mathbf{X}_\mathbf{j}^\top \\
 	\mathbf{V} &= - \mathbf{X}_\mathbf{j} \\
@@
 \end{align*}
-where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jm})^T$ represents the vector of indices for the cases in the $j$th fold, $j = 1, \ldots, k$.
+where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jm})^\top$ represents the vector of indices for the cases in the $j$th fold, $j = 1, \ldots, k$.
 The negative sign in $\mathbf{V} = - \mathbf{X}_\mathbf{j}$ reflects the *removal*, rather than addition, of the cases in $\mathbf{j}$.
@@
 A disadvantage of the Woodbury identity, however, is that it entails explicit matrix inversion and thus may be numerically unstable.
-The inverse of $\mathbf{A} = \mathbf{X}^T \mathbf{W} \mathbf{X}$ is available directly in the \code{"lm"} object, but the second term on the right-hand side of the Woodbury identity requires a matrix inversion with each fold deleted.
+The inverse of $\mathbf{A} = \mathbf{X}^\top \mathbf{W} \mathbf{X}$ is available directly in the \code{"lm"} object, but the second term on the right-hand side of the Woodbury identity requires a matrix inversion with each fold deleted.
 (In contrast, the inverse of each $\mathbf{C} = \mathbf{W}_\mathbf{j}$ is straightforward because $\mathbf{W}$ is diagonal.)

### 2024-07-11 scratch/notes_john/JSS-article/JSS-article-4.Rmd
make Figs. 6 & 7 larger in JSS-article-4

@@
 Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data.
-We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore} (a), produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
+We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore} (a), produced by the \code{scatterplotMatrix()} function in the \pkg{car} package); in each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.:
 ```{r Auto-explore, fig.show="hold", echo=FALSE}
-#| out.width = "45%",
+#| out.width = "65%",
+#| fig.align = 'center',
 #| fig.height = 5,
-#| fig.cap = "(a) Scatterplot matrix for the numeric variables in the \\code{Auto} data.
-In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.
-(b) Component + residual plots for the working model fit to the \\code{Auto} data.
-The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.",
+#| fig.ncol = 1,
+#| fig.cap = "(a) Scatterplot matrix for the numeric variables in the \\code{Auto} data.  (b) Component + residual plots for the working model fit to the \\code{Auto} data.",
 #| fig.subcap=rep("", 2)
@@
 ```
-The component + residual plots in Figure \ref{fig:Auto-explore} (b), produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.
+The component + residual (C+R) plots in Figure \ref{fig:Auto-explore} (b), produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.
+The broken blue lines in the C+R plots show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.
 
@@
 ```{r Auto-transformed, fig.show="hold", echo=FALSE}
-#| out.width = "45%",
+#| out.width = "65%",
+#| fig.align = 'center',
 #| fig.height = 5,
+#| fig.ncol = 1,
 #| fig.cap = "(a) Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data. (b) Component + residual plots for the regression model fit to the transformed \\code{Auto} data.",

### 2024-07-11 scratch/notes_john/JSS-article/JSS-article-4.Rmd
adjust size of Figs 5 & 6 in JSS-article-4

@@
 (b) Estimated squared error as a function of polynomial degree, $p = 1, \\ldots, 10$.",
-#| fig.subcap=rep("", 2)
+#| fig.subcap=rep("", 2),
+#| fig.sep = "~~~"
 plot(mpg ~ horsepower, data=Auto)
@@
 #| fig.subcap=c('', ''),
-#| fig.ncol = 2, out.width = "50%", fig.align = "center"
+#| fig.ncol = 2, out.width = "50%", fig.align = "center",
+#| fig.sep = "~~~"
 plot(cv.auto.10, main = "Polynomial Regressions, 10-Fold CV",
@@
 #| fig.align = 'center',
-#| fig.height = 5,
+#| fig.height = 4.5,
 #| fig.ncol = 1,

### 2024-07-11 scratch/notes_john/JSS-article/JSS-article-4.Rmd
improve C+R plots in JSS-article-4

@@
 m.auto <- lm(mpg ~ ., data = Auto)
-crPlots(m.auto, pch = ".")
+crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
 ```
@@
                   smooth=list(spread = FALSE), data = A, pch = ".")
-crPlots(m, pch = ".")
+crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
 ```
@@
 ```{r Auto-CR-plots-transformed, eval=FALSE}
-crPlots(m, pch = ".")
+crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
 ```

### 2024-07-11 scratch/notes_john/JSS-article/JSS-article-4.Rmd
tweak Figs 6 & 7 in JSS-article-4

@@
 Auto$origin <- factor(Auto$origin,
-                      labels = c("America", "Europe", "Japan"))
+                      labels = c("AMER", "EUR", "JAP"))
 rownames(Auto) <- make.names(Auto$name, unique = TRUE)

### 2024-07-14 scratch/notes_john/JSS-article/JSS-article-4.Rmd
change ` ` to \code{ } markup in JSS-article-4

@@
 The presentation here is close (though not identical) to that in the original source [@JamesEtAl:2021 Secs.
-5.1, 5.3], and it demonstrates the use of the `cv()` function.[^boot]
-
-[^boot]: @JamesEtAl:2021 use the `cv.glm()` function in the \pkg{boot} package [@CantyRipley2022; @DavisonHinkley:1997].
-Despite its name, `cv.glm()` is an independent function and not a method of a `cv()` generic function.
+5.1, 5.3], and it demonstrates the use of the \code{cv()} function.[^boot]
+
+[^boot]: @JamesEtAl:2021 use the \code{cv.glm()} function in the \pkg{boot} package [@CantyRipley2022; @DavisonHinkley:1997].
+Despite its name, \code{cv.glm()} is an independent function and not a method of a \code{cv()} generic function.
 The \pkg{boot} package is part of the standard \proglang{R} distribution.
@@
 ```
-With the exception of `origin` (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see `help("Auto", package="ISLR2")`.
+With the exception of \code{origin} (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see \code{help("Auto", package="ISLR2")}.
 
@@
 Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-polynomials} (a).
-The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).
+The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high \code{mpg} values at the right (but see the CV results reported below).
 
 [^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here.
-Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader).
+Consider, for example the scatterplot for log-transformed \code{mpg} and \code{horsepower}, produced by \code{plot(mpg ~ horsepower, data=Auto, log="xy")} (execution of which is left to the reader).
 We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.
@@
 
-The generic `cv()` function has an `"lm"` method,
+The generic \code{cv()} function has an \code{"lm"} method,
 ```{r cv-lm-method}
@@
 
-```{r cv.lm-2`}
+```{r cv.lm-2}
 cv(m.auto, k = "loo")
@@
 [^code-in-complements]: The code for this section is in the complementary materials for the paper.
-We used the `glmmTMB()` function in the \pkg{glmmTMB} package [@BrooksEtAl] for the models fit here because, in our experience, it is more likely to converge than functions in the \pkg{nlme} and the \pkg{lme4} packages for models with low between-cluster variance.
+We used the \code{glmmTMB()} function in the \pkg{glmmTMB} package [@BrooksEtAl] for the models fit here because, in our experience, it is more likely to converge than functions in the \pkg{nlme} and the \pkg{lme4} packages for models with low between-cluster variance.
 We used the \pkg{lattice} [@Sarkar:2008] and \pkg{latticeExtra} [@SarkarAndrews:2022] packages for the graphs.
@@
 The ellipses are estimated 50% Gaussian concentration ellipses for each patient.
-The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel, where $y$ is symptoms and $x$ is dosage."
+The population between-patient regression line, $\\E(y) = 10 + x$, is shown in each panel, where $y$ is symptoms and $x$ is dosage."
 library("lattice")
@@
 We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.
-The first three models are sequentially nested in their fixed effects: (1) `~ 1`, intercept only; (2) `~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`.
-The final model, (4) `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`.
+The first three models are sequentially nested in their fixed effects: (1) \code{~ 1}, intercept only; (2) \code{~ 1 + x}, intercept and effect of \code{x}; and (3) \code{~ 1 + x + xm}, intercept, effect of \code{x}, and a contextual variable, \code{xm}, consisting of the within-patient mean of \code{x}.
+The final model, (4) \code{~ 1 + I(x - xm)}, uses an intercept and the centered-within-group variable \code{x - xm}.
 We thus fit four models to four data sets for a total of 16 models.
@@
 
-Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions, and also of BLUPs, towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`.
-The inclusion of the contextual variable `xm` in the model corrects this problem.
-
-Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform to the within-patient slopes for the `~ 1 + x` model in the two data sets with the smallest between-patient variances.
-For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variance.
+Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions, and also of BLUPs, towards the between-patient relationship between \code{y} and \code{x} in the model with an intercept and \code{x} as fixed-effect predictors, \code{~ 1 + x}.
+The inclusion of the contextual variable \code{xm} in the model corrects this problem.
+
+Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform to the within-patient slopes for the \code{~ 1 + x} model in the two data sets with the smallest between-patient variances.
+For data with a small between-patient variance, fixed-effects predictions for the \code{~ 1 + x} model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variance.
 With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.
 
-We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, `~ 1`  to the larger `~ 1 + x` model when the between-cluster variance is relatively small, but that cluster-based cross-validation will prefer the latter to the former.
-We will discover that case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model for the '5 / 2.5' data set, but cluster-based cross-validation prefers the latter model to the former.
+We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, \code{~ 1}  to the larger \code{~ 1 + x} model when the between-cluster variance is relatively small, but that cluster-based cross-validation will prefer the latter to the former.
+We will discover that case-based cross-validation prefers the \code{~ 1 + x} model to the \code{~ 1} model for the '5 / 2.5' data set, but cluster-based cross-validation prefers the latter model to the former.
 The situation is entirely reversed with the '8 / 2.5' data set.
 
-The third model, `~ 1 + x + xm`, includes a contextual effect of `x`---that is, the cluster mean `xm`---along with `x` and the intercept in the fixed-effect part of the model, and with a random intercept.
-This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | patient)`, which is the model that generated the data.
-The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects-only model with `x` and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).
+The third model, \code{~ 1 + x + xm}, includes a contextual effect of \code{x}---that is, the cluster mean \code{xm}---along with \code{x} and the intercept in the fixed-effect part of the model, and with a random intercept.
+This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | patient)}, which is the model that generated the data.
+The fit of the mixed model \code{~ 1 + x + xm} is consequently similar to that of a fixed-effects-only model with \code{x} and a categorical predictor for individual patients (i.e., \code{y ~ factor(patient) + x}, treating patients as a factor, and not shown here).
 
@@
 
-[^cv-list]: In order to reduce between-model random variability in comparisons of models on the same data set, we applied `cv()` to a list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.
+[^cv-list]: In order to reduce between-model random variability in comparisons of models on the same data set, we applied \code{cv()} to a list of models created by the \code{models()} function (introduced previously), performing cross-validation with the same folds for each model.
 
@@
 ```
-In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean `xm` of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
-In all of our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.
+In summary, when between-cluster variance is relatively large, the model \code{~ 1 + x}, with \code{x} alone and without the contextual mean \code{xm} of \code{x}, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
+In all of our examples, the model \code{~ 1 + x + xm}, which includes both \code{x} and its contextual mean, produces better results using both cluster-based and case-based CV.
 These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
@@
 
-After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log n$):
+After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the \code{k} argument of \code{stepAIC()} to $\log n$):
 ```{r stepAIC}
@@
 In Section \ref{comparing-competing-models}, following @JamesEtAl:2021[Secs.
-5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of `mpg` to `horsepower` for the `Auto` data, saving the 10 models, named `m.1` through `m.10`, in the list `mlist`.
-We then used `cv()` to compare the cross-validated MSE for the 10 models, discovering that the 7th degree polynomial had the smallest MSE (by a small margin).
+5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of \code{mpg} to \code{horsepower} for the \code{Auto} data, saving the 10 models, named \code{m.1} through \code{m.10}, in the list \code{mlist}.
+We then used \code{cv()} to compare the cross-validated MSE for the 10 models, discovering that the 7th degree polynomial had the smallest MSE (by a small margin).
 
@@
 One solution is to cross-validate the process of using CV to select the "best" model---that is, to apply CV to CV recursively.
-The function `selectModelList()`, which is suitable for use with `cv()`, implements this idea.
-
-Applying `selectModelList()` to the `Auto` polynomial-regression models, and using 10-fold CV, we obtain:
+The function \code{selectModelList()}, which is suitable for use with \code{cv()}, implements this idea.
+
+Applying \code{selectModelList()} to the \code{Auto} polynomial-regression models, and using 10-fold CV, we obtain:
 ```{r recursive-CV-polynomials}
@@
 
-We can equivalently call `cv()` with the list of models as its first argument and set the argument `recursive=TRUE` (identical output not shown):
+We can equivalently call \code{cv()} with the list of models as its first argument and set the argument \code{recursive = TRUE} (identical output not shown):
 ```{r recursive-cv-alt, eval=FALSE}
@@
 ```
-That is, we simply call the default `cv()` method (via \code{NextMethod()}) with the `type` argument properly set.
-In addition to supplying the correct `type` argument, our method sets the default `criterion` for the `cv.multinom()` method to `BayesRuleMulti`.
-Adding the argument `criterion.name=` `deparse(substitute(criterion))` is inessential, but it insures that printed output will include the name of the criterion function that's employed, whether it's the default `BayesRuleMulti` or something else.
-Prior to invoking `NextMethod()`, we called `update()` with `trace=FALSE` to suppress the iteration history reported by default by `multinom()`---it would be tedious to see the iteration history for each fold.
+That is, we simply call the default \code{cv()} method (via \code{NextMethod()}) with the \code{type} argument properly set.
+In addition to supplying the correct \code{type} argument, our method sets the default \code{criterion} for the \code{cv.multinom()} method to \code{BayesRuleMulti}.
+Adding the argument \code{criterion.name = deparse(substitute(criterion))} is inessential, but it insures that printed output will include the name of the criterion function that's employed, whether it's the default \code{BayesRuleMulti} or something else.
+Prior to invoking \code{NextMethod()}, we called \code{update()} with \code{trace = FALSE} to suppress the iteration history reported by default by \code{multinom()}---it would be tedious to see the iteration history for each fold.
 

### 2024-07-14 scratch/notes_john/JSS-article/JSS-article-4.Rmd
finish replacing ` ` with \code{ } in JSS-article-4

@@
 The default method of \code{GetResponse()} (a function supplied by the \pkg{cv} package---see \code{?GetResponse}) fails for a \code{"multinom"} object.
-A straightforward solution is to supply a \code{GetResponse.multinom()} method that returns the factor response [using the `get_response()` function from the \pkg{insight} package, @LudeckeWaggonerMakowski:2019],
+A straightforward solution is to supply a \code{GetResponse.multinom()} method that returns the factor response [using the \code{get\_response()} function from the \pkg{insight} package, @LudeckeWaggonerMakowski:2019],
 ```{r GetResponse.multinom}

### 2024-07-14 scratch/notes_john/JSS-article/JSS-article-4.Rmd
add Georges's email address to JSS-article-4

@@
     affiliation: 'York University'
-    # To add another line, use \AND at the end of the previous one as above
     address: |
       | Toronto, Ontario, Canada
+    email: \email{georges@yorku.ca}
 title:

### 2024-07-20 scratch/notes_georges/JSS-article/JSS-article-4_corrected_typos_GM.Rmd
JSS-article-4 with modifications

--- 2024-07-14 scratch/notes_john/JSS-article/JSS-article-4.Rmd
@@
       | Toronto, Ontario, Canada
-    email: \email{georges@yorku.ca}
+    email: \email
+    url: http://blackwell.math.yorku.ca/gmonette
 title:
@@
 There are also data management facilities in the package, such as coercing the objects produced by the \code{cv()} function into data frames for further analysis.
-These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provides greater detail on some of topics that we do describe, such as extensions to the package.
+These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provide greater detail on some topics that we do describe, such as extensions to the package.
 
@@
 This is the approach taken in the default method for \code{cv()}, and it is appropriate if the cases are independently sampled.
-Refitting the model in this manner for each fold is generally feasible when the number of folds in modest, but can be prohibitively costly for leave-one-out cross-validation when the number of cases is large.
+Refitting the model in this manner for each fold is generally feasible when the number of folds is modest, but can be prohibitively costly for leave-one-out cross-validation when the number of cases is large.
 
@@
 [^WLS]: This is a definitional formula, which assumes that the model matrix $\mathbf{X}$ is of full column rank, and which can be subject to numerical instability when $\mathbf{X}$ is ill-conditioned.
-\code{lm()} uses the singular-value decomposition of the model matrix to obtain computationally more stable results.
+\code{lm()} uses the QR decomposition of the model matrix with pivoting to obtain computationally more stable results.
 
@@
 
-To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identify [@Hager:1989],
+To compute cross-validated fitted values when the folds contain more than one case, we make use of the Woodbury matrix identity [@Hager:1989],
 $$

### 2024-07-21 scratch/notes_john/JSS-article/JSS-submission/Fox-Monette-cv-package.Rmd
add scratch/ ... /JSS-submission/*

--- 2024-07-20 scratch/notes_georges/JSS-article/JSS-article-4_corrected_typos_GM.Rmd
@@
 
-# Preliminary Example: Polynomial regression
+# Preliminary example: Polynomial regression
 
@@
 Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data.
-We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore} (a), produced by the \code{scatterplotMatrix()} function in the \pkg{car} package); in each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.:
-```{r Auto-explore, fig.show="hold", echo=FALSE}
-#| out.width = "65%",
+We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore-scatterplot-matrix}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package); in each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.:
+
+```{r Auto-explore-scatterplot-matrix, echo=FALSE}
+#| out.width = "95%",
+#| fig.height = 5.5,
 #| fig.align = 'center',
-#| fig.height = 4.5,
-#| fig.ncol = 1,
-#| fig.cap = "(a) Scatterplot matrix for the numeric variables in the \\code{Auto} data.  (b) Component + residual plots for the working model fit to the \\code{Auto} data.",
-#| fig.subcap=rep("", 2)
+#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data."
 Auto$cylinders <- factor(Auto$cylinders,
@@
                   smooth = list(spread = FALSE), data = Auto, pch= ".")
-
+```
+
+```{r Auto-explore-cr-plots, echo=FALSE}
+#| out.width = "95%",
+#| fig.height = 5.5,
+#| fig.align = 'center',
+#| fig.cap = "Component + residual plots for the working model fit to the \\code{Auto} data."
 m.auto <- lm(mpg ~ ., data = Auto)
@@
 ```
-The component + residual (C+R) plots in Figure \ref{fig:Auto-explore} (b), produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.
+The component + residual (C+R) plots in Figure \ref{fig:Auto-explore-cr-plots}, produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.
 The broken blue lines in the C+R plots show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.
@@
 
-The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore} (a), on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed} (a)):
-```{r Auto-transformed, fig.show="hold", echo=FALSE}
-#| out.width = "65%",
+The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore-scatterplot-matrix}, on page \pageref{fig:Auto-explore-scatterplot-matrix}, and Figure \ref{fig:Auto-transformed-scatterplot-matrix}):
+```{r Auto-transformed-scatterplot-matrix, echo=FALSE}
+#| out.width = "95%",
 #| fig.align = 'center',
-#| fig.height = 4.5,
-#| fig.ncol = 1,
-#| fig.cap = "(a) Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data. (b) Component + residual plots for the regression model fit to the transformed \\code{Auto} data.",
-#| fig.subcap=rep("", 2)
+#| fig.height = 5.5,
+#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data."
 scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
@@
                   smooth=list(spread = FALSE), data = A, pch = ".")
+```
+
+```{r Auto-transformed-cr-plots, echo=FALSE}
+#| out.width = "95%",
+#| fig.align = 'center',
+#| fig.height = 5.5,
+#| fig.cap = "Component + residual plots for the regression model fit to the transformed \\code{Auto} data."
 crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
@@
 
-And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore} (b), on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed} (b)):
+And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore-cr-plots}, on page \pageref{fig:Auto-explore-cr-plots}, and Figure \ref{fig:Auto-transformed-cr-plots}:
 ```{r Auto-CR-plots-transformed, eval=FALSE}

### 2024-07-24 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
edit JSS paper

--- 2024-07-21 scratch/notes_john/JSS-article/JSS-submission/Fox-Monette-cv-package.Rmd
@@
 
-The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical model objects.
+The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical models.
 The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
@@
 For example, the \code{cv()} function can perform computations in parallel and can independently replicate a cross-validation procedure several times.
-There are also data management facilities in the package, such as coercing the objects produced by the \code{cv()} function into data frames for further analysis.
+There are also data-management facilities in the package, such as coercing the objects produced by the \code{cv()} function into data frames for further analysis.
 These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provide greater detail on some topics that we do describe, such as extensions to the package.
@@
 ```
-With the exception of \code{origin} (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see \code{help("Auto", package="ISLR2")}.
+With the exception of \code{origin} (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see \code{help("Auto",} \code{package = "ISLR2")}.
 
@@
 Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-polynomials} (a).
-The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high \code{mpg} values at the right (but see the CV results reported below).
+The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the degree-5 polynomial may over-fit the data by chasing one or two relatively high \code{mpg} values at the right (but see the CV results reported below).
 
@@
 The function reports the CV estimate of MSE, a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression.
-Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint=TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.
+Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint = TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.
 
@@
 The last column shows which methods have average timings that are statistically distinguishable.
-Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the patient reader to redo this computation with the default \code{times=100} repetitions.
+Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the patient reader to redo this computation with the default \code{times = 100} repetitions.
 
@@
 Alternatively, as here, we can supply a pre-constructed, optionally named, list of models as the single argument to \code{models()}.
-To visualize the results, comparing the models, we invoke the \code{plot()} method for the \code{"cvModList"} object returned by \code{cv()} (see Figure \ref{fig:polynomial-regression-CV-graph-2}):
+To visualize the results, comparing the models, we invoke the \code{plot()} method for the \code{"cvModList"} objects returned by \code{cv()} (see Figure \ref{fig:polynomial-regression-CV-graph-2}):
 
@@
 
-Following their use by @RaudenbushBryk:2002, data from the 1982 *High School and Beyond* (HSB) survey have become a staple of the literature on mixed-effects models.
+Following their use by @RaudenbushBryk:2002, data from the 1982 *High School and Beyond* ("HSB") survey have become a staple of the literature on mixed-effects models.
 The HSB data are used by @FoxWeisberg:2019 [Sec.
@@
 
-The four data sets differ in the magnitude of between-cluster variance compared to within-cluster variance.
-They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.
+The four data sets differ in the magnitude of between-cluster variation compared to within-cluster variation.
+They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variation, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.
 
@@
 [^code-in-complements]: The code for this section is in the complementary materials for the paper.
-We used the \code{glmmTMB()} function in the \pkg{glmmTMB} package [@BrooksEtAl] for the models fit here because, in our experience, it is more likely to converge than functions in the \pkg{nlme} and the \pkg{lme4} packages for models with low between-cluster variance.
+We used the \code{glmmTMB()} function in the \pkg{glmmTMB} package [@BrooksEtAl] for the models fit here because, in our experience, it is more likely to converge than functions in the \pkg{nlme} and the \pkg{lme4} packages for models with low between-cluster variation.
 We used the \pkg{lattice} [@Sarkar:2008] and \pkg{latticeExtra} [@SarkarAndrews:2022] packages for the graphs.
@@
 #| fig.height = 2.5,
-#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance.
+#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variation.
 The labels above each panel show the ratio of between-patient SD to within-patient SD.
@@
 Using data like these, a researcher may attempt to obtain an estimate of the within-patient effect of dosage by fitting mixed models with a random intercept.
-We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.
+We will illustrate how this approach results in estimates that are highly sensitive to the relative variation at the two levels of the mixed model by considering four models, all of which include a random intercept.
 The first three models are sequentially nested in their fixed effects: (1) \code{~ 1}, intercept only; (2) \code{~ 1 + x}, intercept and effect of \code{x}; and (3) \code{~ 1 + x + xm}, intercept, effect of \code{x}, and a contextual variable, \code{xm}, consisting of the within-patient mean of \code{x}.
@@
 #| fig.height = 6,
-#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient variance ratios.
+#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient standard deviation ratios.
 The top row shows summaries of the within-patient data using estimated 50% concentration ellipses, along with the patient centroids."
@@
 
-Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions, and also of BLUPs, towards the between-patient relationship between \code{y} and \code{x} in the model with an intercept and \code{x} as fixed-effect predictors, \code{~ 1 + x}.
+Data sets with relatively low between-patient variation result in strong shrinkage of fixed-effects predictions, and also of BLUPs, towards the between-patient relationship between \code{y} and \code{x} in the model with an intercept and \code{x} as fixed-effect predictors, \code{~ 1 + x}.
 The inclusion of the contextual variable \code{xm} in the model corrects this problem.
 
-Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform to the within-patient slopes for the \code{~ 1 + x} model in the two data sets with the smallest between-patient variances.
-For data with a small between-patient variance, fixed-effects predictions for the \code{~ 1 + x} model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variance.
-With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.
-
-We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, \code{~ 1}  to the larger \code{~ 1 + x} model when the between-cluster variance is relatively small, but that cluster-based cross-validation will prefer the latter to the former.
+Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform to the within-patient slopes for the \code{~ 1 + x} model in the two data sets with the smallest between-patient variation.
+For data with a small between-patient variation, fixed-effects predictions for the \code{~ 1 + x} model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variation.
+With data whose between-patient variation is relatively large, predictions based on the model with a common intercept and slope for all clusters are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.
+
+We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, \code{~ 1}  to the larger \code{~ 1 + x} model when the between-cluster variation is relatively small, but that cluster-based cross-validation will prefer the latter to the former.
 We will discover that case-based cross-validation prefers the \code{~ 1 + x} model to the \code{~ 1} model for the '5 / 2.5' data set, but cluster-based cross-validation prefers the latter model to the former.
@@
 
-The third model, \code{~ 1 + x + xm}, includes a contextual effect of \code{x}---that is, the cluster mean \code{xm}---along with \code{x} and the intercept in the fixed-effect part of the model, and with a random intercept.
+The third model, \code{~ 1 + x + xm}, includes a contextual effect of \code{x}---that is, the cluster mean \code{xm}---along with \code{x}, the intercept in the fixed-effect part of the model, and a random intercept.
 This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | patient)}, which is the model that generated the data.
@@
 
-We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.[^cv-list] The results for all data sets, using both cluster-based and case-based cross-validation, are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.
+We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.[^cv-list] The results for all data sets, using both cluster-based and case-based cross-validation, are assembled in Figure \ref{fig:cross-validation-data-plot}.
 
@@
 ```
-In summary, when between-cluster variance is relatively large, the model \code{~ 1 + x}, with \code{x} alone and without the contextual mean \code{xm} of \code{x}, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
+In summary, when between-cluster variation is relatively large, the model \code{~ 1 + x}, with \code{x} alone and without the contextual mean \code{xm} of \code{x}, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV.
 In all of our examples, the model \code{~ 1 + x + xm}, which includes both \code{x} and its contextual mean, produces better results using both cluster-based and case-based CV.
-These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
+These conclusions are consistent with our observations based on graphing predictions from the various models (in Figure \ref{fig:plot-fits-combined} on page \pageref{fig:plot-fits-combined}), and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
 
@@
 7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain, if the whole data are used to specify or fine-tune a statistical model, then subsequent cross-validation of the model is intrinsically misleading, because the model is selected to fit the whole data, including the part of the data that remains when each fold is removed.
-Statistical modeling is partly a craft, and one could imagine applying that craft to successive partial data sets, each with a fold removed.
+Statistical modeling is partly a craft, and one could imagine applying that craft independently to successive partial data sets, each with a fold removed.
 The resulting procedure would be tedious, though possibly worth the effort, but it would also be difficult to realize in practice: After all, we can hardly erase our memory of statistical modeling choices between analyzing partial data sets.
@@
 Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data.
-We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore-scatterplot-matrix}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package); in each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.:
-
-```{r Auto-explore-scatterplot-matrix, echo=FALSE}
+We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore-scatterplot-matrix}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package); in each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression:
+
+```{r Auto-explore-scatterplot-matrix}
 #| out.width = "95%",
@@
 
-```{r Auto-explore-cr-plots, echo=FALSE}
+A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors.
+There are, however, only 5 distinct values of \code{cylinders}, ranging from 3 to 8, but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders.
+There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most of the numeric variables are positively skewed and that many of the pairwise relationships among them are nonlinear.
+
+[^year]: Making the decision to treat \code{year} as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process.
+It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation.
+We believe, however, that it's still useful to automate as much of the process as we can, as illustrated here, to obtain a more realistic, if still biased, estimate of the predictive power of a model.
+
+We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
+
+```{r Auto-explore-cr-plots}
 #| out.width = "95%",
@@
 
-```{r Auto-explore-show, eval=FALSE}
-Auto$cylinders <- factor(Auto$cylinders,
-                         labels = c("3-4", "3-4", "5-6", "5-6", "8"))
-Auto$year <- as.factor(Auto$year)
-Auto$origin <- factor(Auto$origin,
-                      labels = c("AMER", "EUR", "JAP"))
-rownames(Auto) <- make.names(Auto$name, unique = TRUE)
-Auto$name <- NULL
-
-scatterplotMatrix(~ mpg + displacement + horsepower + weight 
-                  + acceleration, 
-                  smooth = list(spread = FALSE), data = Auto, pch = ".")
-```
-
-A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors.
-There are, however, only 5 distinct values of \code{cylinders}, ranging from 3 to 8, but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders.
-There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.
-
-[^year]: Making the decision to treat \code{year} as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process.
-It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation.
-We believe, however, that it's still useful to automate as much of the process as we can, as illustrated here, to obtain a more realistic, if still biased, estimate of the predictive power of a model.
-
-We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
-```{r Auto-working-model, eval=FALSE}
-m.auto <- lm(mpg ~ ., data = Auto)
-crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
-```
-The component + residual (C+R) plots in Figure \ref{fig:Auto-explore-cr-plots}, produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.
+The component + residual ("C+R") plots in Figure \ref{fig:Auto-explore-cr-plots}, produced by the \code{crPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.
 The broken blue lines in the C+R plots show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.
@@
 The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964].
-Several "families" of power transformations can be used, including the original Box-Cox family (which is the default), simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and 0s: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap.
+Several "families" of power transformations can be used, including the original Box-Cox family (which is the default), simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap.
 8] and @FoxWeisberg:2019 [Chap.
@@
 The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore-scatterplot-matrix}, on page \pageref{fig:Auto-explore-scatterplot-matrix}, and Figure \ref{fig:Auto-transformed-scatterplot-matrix}):
-```{r Auto-transformed-scatterplot-matrix, echo=FALSE}
+```{r Auto-transformed-scatterplot-matrix}
 #| out.width = "95%",
@@
 
-```{r Auto-transformed-cr-plots, echo=FALSE}
+And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore-cr-plots}, on page \pageref{fig:Auto-explore-cr-plots}, and Figure \ref{fig:Auto-transformed-cr-plots}:
+
+```{r Auto-transformed-cr-plots}
 #| out.width = "95%",
@@
 #| fig.cap = "Component + residual plots for the regression model fit to the transformed \\code{Auto} data."
-crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
-```
-
-```{r Auto-transformed-scatterplot-matrix-show, eval=FALSE}
-scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
-                  + acceleration, 
-                  smooth = list(spread = FALSE), data = A, pch = ".")
-```
-
-And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore-cr-plots}, on page \pageref{fig:Auto-explore-cr-plots}, and Figure \ref{fig:Auto-transformed-cr-plots}:
-```{r Auto-CR-plots-transformed, eval=FALSE}
-crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
+crPlots(m, pch = ".", ylab = "C+R", las = 2)
 ```
@@
 * \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, then a seed is randomly selected and saved.
-* Arguments required by the model-selection function: the starting \code{working.model} (here, \code{m.auto}) for transformation and predictor selection; the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation; and \code{AIC=FALSE}, which specifies use of the BIC for model selection.
+* Arguments required by the model-selection function: the starting \code{working.model} (here, \code{m.auto}) for transformation and predictor selection; the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation; and \code{AIC = FALSE}, which specifies use of the BIC for model selection.
 
@@
 We borrow an example from @Fox:2016 [Sec.
-14.2.1], with data from the British Election Panel Study on vote choice in the 2001 British election.
+14.2.1], with data from the British Election Panel Study on vote choice in the 2001 UK election.
 Data for the example are in the \code{BEPS} data set in the \pkg{carData} package:
@@
 * \code{age}, in years;
-* \code{econ.cond.national} and \code{econ.cond.household}, the respondent's ratings of the state of the economy, on 1 to 5 scales.
-* \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
-* \code{Europe}, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism."
-* \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3.
+* \code{econ.cond.national} and \code{econ.cond.household}, the respondent's ratings of the state of the economy, on 1 to 5 scales;
+* \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales;
+* \code{Europe}, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism";
+* \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3;
 * \code{gender}, \code{"female"} or \code{"male"}.
@@
 
-The  interaction is associated with a very small $p$-value.
 Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the \code{Europe} $\times$ \code{political.knowledge} interaction in a "stacked-area" graph:
@@
 ```
-After doing some error checking, \code{BayesRule()} rounds the predicted proabability of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction, and then reports the proportion of incorrect predictions.
+After doing some error checking, \code{BayesRule()} rounds the predicted proabability \code{yhat} of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction, and then reports the proportion of incorrect predictions.
 Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, enabling the computation of bias correction and confidence intervals (as discussed in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
@@
 ```
-The \code{predict()} method for \code{"multinom"} models called with argument \code{type="class"} reports the Bayes's rule prediction for each case---that is, the response category with the highest predicted probability.
-Our \code{BayesRuleMulti()} function calculates the proportion of misclassified cases.
+The \code{predict()} method for \code{"multinom"} models called with argument \code{type = "class"} reports the Bayes's rule prediction for each case---that is, the response category with the highest predicted probability.
+The argument \code{yhat} to our \code{BayesRuleMulti()} function is the vector of Bayes's rule categorical predictions, while \code{y} is the vector of observed categorical responses.
+The function calculates and returns the proportion of misclassified cases.
 Because this value is also the mean of casewise components, we attach a \code{"casewise loss"} attribute to the result.
@@
 Adding the argument \code{criterion.name = deparse(substitute(criterion))} is inessential, but it insures that printed output will include the name of the criterion function that's employed, whether it's the default \code{BayesRuleMulti} or something else.
-Prior to invoking \code{NextMethod()}, we called \code{update()} with \code{trace = FALSE} to suppress the iteration history reported by default by \code{multinom()}---it would be tedious to see the iteration history for each fold.
+Prior to invoking \code{NextMethod()}, we call \code{update()} with \code{trace = FALSE} to suppress the iteration history reported by default by \code{multinom()}---it would be tedious to see the iteration history for each fold.
 
@@
 The response variable in this example, \code{Attrition}, is binary.
-The function \code{holdout_results()} must be supplied by the user, and "accuracy" is the complement of the Bayes-rule classification error rate.
+The function \code{holdout_results()} must be supplied by the user, and "\code{accuracy}" is the complement of the Bayes-rule classification error rate.
 
@@
 library("ggplot2", warn.conflicts = FALSE)
-library("caret", quietly = TRUE)
+library("caret", quietly = TRUE, warn.conflicts = FALSE)
 train(x = attrition[, c("JobSatisfaction", "Gender", "MonthlyIncome")],
@@
 ```
-where "accuracy", again, is the complement of the misclassification rate.
+where "\code{Accuracy}", again, is the complement of the misclassification rate.
 
@@
 
-Let's compare the computational efficiency of the various implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:[^times-10]
+Let's compare the computational efficiency of the various implementations, also showing the use of \code{method = "Woodbury"} and \code{method =" hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method = "exact"}:[^times-10]
 
@@
 This approach can break down when one or more hatvalues are equal to 1, in which case the formula for $\widehat{y}_{-i}$ requires division by 0.
-In this case, the "training" set omitting the observation with hatvalue = 1 is rank-deficient and the predictors for the left-out case are outside the linear span of the predictors in the training set.
+Then the "training" set omitting the observation with hatvalue = 1 is rank-deficient and the predictors for the left-out case are outside the linear span of the predictors in the training set.
 
@@
 
-[^casewise-loss]: Some commonly employed CV criteria---such as the root-mean-squared error (RMSE), median absolute error, and, for binary-regression models, the complement of the area under the receiver operating characteristic (ROC) curve---are not means of casewise components.
+[^casewise-loss]: Some commonly employed CV criteria---such as the root-mean-squared error ("RMSE"), median absolute error, and, for binary-regression models, the complement of the area under the receiver operating characteristic ("ROC") curve---are not means of casewise components.
 That the RMSE and median absolute error aren't means of casewise components is obvious; for the complement of the area under the ROC curve, see the vignette on extending the \pkg{cv} package.
@@
 
-We divide the $n$ cases into $k$ folds of approximately $n_j \approx n/k$ cases each, where $n = \sum n_j$. As above, let $\mathbf{j}$ denote the indices of the cases in the $j$th fold.
+We divide the $n$ cases into $k$ folds of approximately $n_j \approx n/k$ cases each, $j = 1, \ldots, k$, where $n = \sum n_j$.
+As above, let $\mathbf{j}$ denote the indices of the cases in the $j$th fold.
 

### 2024-07-24 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
restrict C+R plots to numeric predictors

@@
 ```{r Auto-explore-cr-plots}
-#| out.width = "95%",
-#| fig.height = 5.5,
+#| out.width = "75%",
+#| fig.height = 4.5,
 #| fig.align = 'center',
-#| fig.cap = "Component + residual plots for the working model fit to the \\code{Auto} data."
+#| fig.cap = "Component + residual plots for the numeric predictors in the working model fit to the \\code{Auto} data."
 m.auto <- lm(mpg ~ ., data = Auto)
-crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
+crPlots(m.auto, 
+        terms = ~ displacement + horsepower + weight + acceleration, 
+        pch = ".", ylab = "C+R", las = 2)
 ```
@@
 ```{r Auto-transformed-cr-plots}
-#| out.width = "95%",
+#| out.width = "75%",
 #| fig.align = 'center',
-#| fig.height = 5.5,
+#| fig.height = 4.5,
 #| fig.cap = "Component + residual plots for the regression model fit to the transformed \\code{Auto} data."
-crPlots(m, pch = ".", ylab = "C+R", las = 2)
+crPlots(m, terms = ~ displacement + horsepower + weight + acceleration, 
+        pch = ".", ylab = "C+R", las = 2)
 ```

### 2024-07-24 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
adjust size of C+R plots to save 1 page

--- 2024-07-24 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
@@
 ```{r Auto-explore-cr-plots}
-#| out.width = "95%",
-#| fig.height = 5.5,
+#| out.width = "70%",
+#| fig.height = 4.25,
 #| fig.align = 'center',
-#| fig.cap = "Component + residual plots for the working model fit to the \\code{Auto} data."
+#| fig.cap = "Component + residual plots for the numeric predictors in the working model fit to the \\code{Auto} data."
 m.auto <- lm(mpg ~ ., data = Auto)
-crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
+crPlots(m.auto, 
+        terms = ~ displacement + horsepower + weight + acceleration, 
+        pch = ".", ylab = "C+R", las = 2)
 ```
@@
 ```{r Auto-transformed-cr-plots}
-#| out.width = "95%",
+#| out.width = "70%",
 #| fig.align = 'center',
-#| fig.height = 5.5,
+#| fig.height = 4.25,
 #| fig.cap = "Component + residual plots for the regression model fit to the transformed \\code{Auto} data."
-crPlots(m, pch = ".", ylab = "C+R", las = 2)
+crPlots(m, terms = ~ displacement + horsepower + weight + acceleration, 
+        pch = ".", ylab = "C+R", las = 2)
 ```

### 2024-07-26 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
further edit to the JSS paper

@@
 Validating research by replication on independently collected data is a common scientific norm.
-Emulating this process in a single study by data-division is less common [see, e.g., @Barnard:1974]: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data.
+Emulating this process in a single study by data-division is less common [see, e.g., @Barnard:1974]: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit (in the language of machine leaning, "train") a statistical model; and then the second part is used to assess the adequacy of ("validate") the model fit to the first part of the data.
 Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples.
@@
 The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical models.
-The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
+The principal function in the package, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
 ```{r cv-methods}
@@
 ```
-With the exception of \code{origin} (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see \code{help("Auto",} \code{package = "ISLR2")}.
+With the exception of \code{origin} (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: For details see \code{help("Auto",} \code{package = "ISLR2")}.
 
@@
 
-* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} (the default) selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).
+* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} (the default) selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than directly recomputing the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).
 
@@
 With the exception of the last column, the output is self-explanatory.
-The last column shows which methods have average timings that are statistically distinguishable.
+The last column shows which methods have average timings that are statistically distinguishable (as indicated by different letters).
 Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the patient reader to redo this computation with the default \code{times = 100} repetitions.
@@
 To generate the four data sets, we first constructed a template data set from a population with a between-patient effect of dosage of $1.0$ and a within-patient effect of $-0.5$.
-We then applied four different multipliers to the between-patient variation from the regression line of patient centroids.
-The resulting data sets, shown in Figure \ref{fig:plot1}, differ in the between-patient variance of patient centroids from a common between-patient regression line, exhibiting a progression from low to high variance around the common regression line.
+We then applied four different multipliers to the between-patient standard deviation from the regression line of patient centroids.
+The resulting data sets, shown in Figure \ref{fig:plot1}, differ in the between-patient variation of patient centroids from a common between-patient regression line, exhibiting a progression from low to high variation around the common regression line.
 Each patient has the same relative configuration of dosages and symptoms in each of the four data sets.
@@
 #| fig.height = 6,
-#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient standard deviation ratios.
+#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient standard-deviation ratios.
 The top row shows summaries of the within-patient data using estimated 50% concentration ellipses, along with the patient centroids."
@@
 We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, \code{~ 1}  to the larger \code{~ 1 + x} model when the between-cluster variation is relatively small, but that cluster-based cross-validation will prefer the latter to the former.
-We will discover that case-based cross-validation prefers the \code{~ 1 + x} model to the \code{~ 1} model for the '5 / 2.5' data set, but cluster-based cross-validation prefers the latter model to the former.
-The situation is entirely reversed with the '8 / 2.5' data set.
+We will discover that case-based cross-validation prefers the \code{~ 1 + x} model to the \code{~ 1} model for the "5 / 2.5" data set, but cluster-based cross-validation prefers the latter model to the former.
+The situation is entirely reversed with the "8 / 2.5" data set.
 
@@
 Auto$origin <- factor(Auto$origin,
-                      labels = c("AMER", "EUR", "JAP"))
+                      labels = c("America", "Europe", "Japan"))
 rownames(Auto) <- make.names(Auto$name, unique = TRUE)
@@
 
-* \code{age}, in years;
-* \code{econ.cond.national} and \code{econ.cond.household}, the respondent's ratings of the state of the economy, on 1 to 5 scales;
-* \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales;
-* \code{Europe}, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism";
-* \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3;
+* \code{age}, in years.
+* \code{econ.cond.national} and \code{econ.cond.household}, the respondent's ratings of the state of the economy, on 1 to 5 scales.
+* \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
+* \code{Europe}, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism."
+* \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3.
 * \code{gender}, \code{"female"} or \code{"male"}.
@@
 The response variable in this example, \code{Attrition}, is binary.
-The function \code{holdout_results()} must be supplied by the user, and "\code{accuracy}" is the complement of the Bayes-rule classification error rate.
+The function \code{holdout_results()} must be supplied by the user, and mean "\code{accuracy}" is the complement of the Bayes-rule classification error rate.
 
@@
 ```
-where "\code{Accuracy}", again, is the complement of the misclassification rate.
+where "\code{Accuracy}," again, is the complement of the misclassification rate.
 
@@
 
-The weighted-least-squares (WLS) estimator of $\boldsymbol{\beta}$ is [see, e.g., @Fox:2016, Sec. 12.2.2] [^WLS]
+The weighted-least-squares ("WLS") estimator of $\boldsymbol{\beta}$ is [see, e.g., @Fox:2016, Sec. 12.2.2] [^WLS]
 $$
@@
 $$
-where $\mathbf{A}$ is a nonsingular order-$n$ matrix. We apply this result by letting
+where $\mathbf{A}$ is a nonsingular order-$m$ matrix. We apply this result by letting
 \begin{align*}
@@
 \end{align*}
-where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jm})^\top$ represents the vector of indices for the cases in the $j$th fold, $j = 1, \ldots, k$.
+where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jn_j})^\top$ represents the vector of indices for the $n_j \approx n/k$ cases in the $j$th fold, $j = 1, \ldots, k$.
 The negative sign in $\mathbf{V} = - \mathbf{X}_\mathbf{j}$ reflects the *removal*, rather than addition, of the cases in $\mathbf{j}$.

### 2024-07-29 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
tweak edit to JSS paper

@@
 \end{align*}
-where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jn_j})^\top$ represents the vector of indices for the $n_j \approx n/k$ cases in the $j$th fold, $j = 1, \ldots, k$.
+where the subscript $\mathbf{j} = (i_{j1}, \ldots, i_{jm})^\top$ represents the vector of indices for the $m = n_j \approx n/k$ cases in the $j$th fold.
 The negative sign in $\mathbf{V} = - \mathbf{X}_\mathbf{j}$ reflects the *removal*, rather than addition, of the cases in $\mathbf{j}$.

### 2024-08-04 vignettes/cv-mixed.R
updated 2.0.2 tarball with vignette/fig folder deleted

@@ -28,6 +28,8 @@ colorize <- function(x, color) {
 }
 
 .opts <- options(digits = 5)
+CACHE <- FALSE
+
 
 ## ----HSB-data-----------------------------------------------------------------
 data("MathAchieve", package = "nlme")
@@ -39,6 +41,51 @@ data("MathAchSchool", package = "nlme")
 dim(MathAchSchool)
 head(MathAchSchool, 2)
 tail(MathAchSchool, 2)
+
+## ----HSB-data-management, cache=CACHE-----------------------------------------
+library("dplyr")
+MathAchieve %>% group_by(School) %>%
+  summarize(mean.ses = mean(SES)) -> Temp
+Temp <- merge(MathAchSchool, Temp, by = "School")
+HSB <- merge(Temp[, c("School", "Sector", "mean.ses")],
+             MathAchieve[, c("School", "SES", "MathAch")], by = "School")
+names(HSB) <- tolower(names(HSB))
+
+HSB$cses <- with(HSB, ses - mean.ses)
+
+## ----HSB-lmer, cache=CACHE----------------------------------------------------
+library("lme4")
+hsb.lmer <- lmer(mathach ~ mean.ses * cses + sector * cses
+                 + (cses | school), data = HSB)
+summary(hsb.lmer, correlation = FALSE)
+
+## ----HSB-lmer-CV-cluster, cache=CACHE-----------------------------------------
+library("cv")
+
+cv(hsb.lmer,
+   k = 10,
+   clusterVariables = "school",
+   seed = 5240)
+
+## ----HSB-lmer-CV-case, cache=CACHE--------------------------------------------
+cv(hsb.lmer, seed = 1575)
+
+## ----hsb-lme, cache=CACHE-----------------------------------------------------
+library("nlme")
+hsb.lme <- lme(
+  mathach ~ mean.ses * cses + sector * cses,
+  random = ~ cses | school,
+  data = HSB,
+  control = list(opt = "optim")
+)
+summary(hsb.lme)
+
+cv(hsb.lme,
+   k = 10,
+   clusterVariables = "school",
+   seed = 5240)
+
+cv(hsb.lme, seed = 1575)
 
 ## ----include=FALSE, echo=FALSE------------------------------------------------
 library("glmmTMB") # necessary for some reason to knit vignette in RStudio, harmless otherwise
@@ -189,7 +236,7 @@ Data$panel <- factor('data', levels = c(names(model.formulas), 'data'))
 library(cv) # unclear why it's necessary to reload cv
[…]

### 2024-08-14 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
small copy edits and one minor code change in code to reduce white space under Figure 5.

@@
 Methods for the \code{cv()} function are provided for many commonly employed classes of statistical models, including mixed-effects models, and it is straightforward to extend the \pkg{cv} package by writing methods for additional classes of models.
-The \code{cv()} function also can cross-validate complex model-selection procedures, such as those that include variable transformations, predictor selection, and selection among competing models.
+The \code{cv()} function can also cross-validate complex model-selection procedures, such as those that include variable transformations, predictor selection, and selection among competing models.
 The \code{cv()} function generally supports parallel computations, and the supplied methods for linear and generalized linear models take advantage of computational efficiencies.
@@
 A CV criterion (also termed a "cost" or "loss" measure), such as the mean-squared error ("MSE") of prediction, is then computed using these predicted values.
-In the extreme, $k = n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."
-
-Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of prediction error.
+In the extreme, $k$ equals $n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."
+
+Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of the prediction error with a model fit to $n$ cases.
 The $n$ regression models are highly statistically dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has larger variance than if the predictions from the models fit to the $n$ data sets were independent.
@@
 
-The final section of the paper (Section \ref{computational-notes}) describes some computational details, including efficient CV computations for linear and generalized-linear models.
-and computation of bias-corrected CV criteria.
+The final section of the paper (Section \ref{computational-notes}) describes some computational details, including efficient CV computations for linear and generalized-linear models, and computation of bias-corrected CV criteria.
 
@@
 
-We'll focus on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-polynomials} (a).
+We'll focus on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure\ \ref{fig:mpg-horsepower-polynomials} (a).
 The relationship between the two variables is monotone, decreasing, and nonlinear.
@@
 [^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here.
-Consider, for example the scatterplot for log-transformed \code{mpg} and \code{horsepower}, produced by \code{plot(mpg ~ horsepower, data=Auto, log="xy")} (execution of which is left to the reader).
+Consider, for example, the scatterplot for log-transformed \code{mpg} and \code{horsepower}, produced by \code{plot(mpg $\sim$ horsepower, data=Auto, log="xy")} (execution of which is left to the reader).
 We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.
@@
 ```
-where, for example \code{mlist[2]} is the quadratic fit.
+where, for example, \code{mlist[2]} is the quadratic fit.
 
@@
 In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model (cf., Figure \ref{fig:mpg-horsepower-polynomials} (b) on page \pageref{fig:mpg-horsepower-polynomials}), except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting.
+\newpage
 
@@
        par.settings =
-         list(superpose.line=list(lty = c(1, 2), lwd = 1),
-              superpose.symbol=list(cex = 0.7)),
+         list(layout.heights = list(bottom.padding = 0),
+              superpose.line = list(lty = c(1, 2), lwd = 1),
+              superpose.symbol = list(cex = 0.7)),
        scales = list(y = list(log = TRUE, cex = 0.7), 
@@
 
-And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore-cr-plots}, on page \pageref{fig:Auto-explore-cr-plots}, and Figure \ref{fig:Auto-transformed-cr-plots}:
+And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore-cr-plots}, on page \pageref{fig:Auto-explore-cr-plots}, and Figure \ref{fig:Auto-transformed-cr-plots}):
 
@@
 ```
-After doing some error checking, \code{BayesRule()} rounds the predicted proabability \code{yhat} of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction, and then reports the proportion of incorrect predictions.
+After doing some error checking, \code{BayesRule()} rounds the predicted probability \code{yhat} of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction, and then reports the proportion of incorrect predictions.
 Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, enabling the computation of bias correction and confidence intervals (as discussed in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).

### 2024-08-18 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
added 'compared' with a model fit to $n$ cases.

@@
 
-Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of the prediction error with a model fit to $n$ cases.
+Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of the prediction error compared with a model fit to $n$ cases.
 The $n$ regression models are highly statistically dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has larger variance than if the predictions from the models fit to the $n$ data sets were independent.

### 2024-09-02 vignettes/cv-mixed.R
update vignettes to use summary()

@@ -62,13 +62,13 @@ summary(hsb.lmer, correlation = FALSE)
 ## ----HSB-lmer-CV-cluster, cache=CACHE-----------------------------------------
 library("cv")
 
-cv(hsb.lmer,
+summary(cv(hsb.lmer,
    k = 10,
    clusterVariables = "school",
-   seed = 5240)
+   seed = 5240))
 
 ## ----HSB-lmer-CV-case, cache=CACHE--------------------------------------------
-cv(hsb.lmer, seed = 1575)
+summary(cv(hsb.lmer, seed = 1575))
 
 ## ----hsb-lme, cache=CACHE-----------------------------------------------------
 library("nlme")
@@ -80,12 +80,12 @@ hsb.lme <- lme(
 )
 summary(hsb.lme)
 
-cv(hsb.lme,
+summary(cv(hsb.lme,
    k = 10,
    clusterVariables = "school",
-   seed = 5240)
+   seed = 5240))
 
-cv(hsb.lme, seed = 1575)
+summary(cv(hsb.lme, seed = 1575))
 
 ## ----include=FALSE, echo=FALSE------------------------------------------------
 library("glmmTMB") # necessary for some reason to knit vignette in RStudio, harmless otherwise
@@ -317,19 +317,18 @@ m.p <- lmer(
 summary(m.p)
 
 ## ----pigs-cv------------------------------------------------------------------
-cv(m.p, clusterVariables = "id")
+summary(cv(m.p, clusterVariables = "id"))
 
-cv(m.p, clusterVariables = "week")
+summary(cv(m.p, clusterVariables = "week"))
 
-cv(
+summary(cv(
   m.p,
   clusterVariables = c("id", "week"),
   k = 10,
-  seed = 8469
-)
+  seed = 8469))
 
 ## ----pigs-cv-cases------------------------------------------------------------
-cv(m.p, k = 10, seed = 8469)
+summary(cv(m.p, k = 10, seed = 8469))
 
 ## ----coda, include = FALSE----------------------------------------------------
 options(.opts)

### 2024-09-02 vignettes/cv-notes.R

@@ -35,26 +35,26 @@ m.auto <- lm(mpg ~ poly(horsepower, 2), data = Auto)
 summary(m.auto)
 
 library("cv")
-cv(m.auto, k = "loo")  # default method = "hatvalues"
-cv(m.auto, k = "loo", method = "naive")
-cv(m.auto, k = "loo", method = "Woodbury")
+summary(cv(m.auto, k = "loo") ) # default method = "hatvalues"
+summary(cv(m.auto, k = "loo", method = "naive"))
+summary(cv(m.auto, k = "loo", method = "Woodbury"))
 
 ## ----Mroz-logistic-regression-------------------------------------------------
 data("Mroz", package="carData")
 m.mroz <- glm(lfp ~ ., data = Mroz, family = binomial)
 summary(m.mroz)
 
-cv(m.mroz, # default method = "exact"
+summary(cv(m.mroz, # default method = "exact"
    k = "loo", 
-   criterion = BayesRule)
-cv(m.mroz,
+   criterion = BayesRule))
+summary(cv(m.mroz,
    k = "loo",
    criterion = BayesRule,
-   method = "Woodbury")
-cv(m.mroz,
+   method = "Woodbury"))
+summary(cv(m.mroz,
    k = "loo",
    criterion = BayesRule,
-   method = "hatvalues")
+   method = "hatvalues"))
 
 ## -----------------------------------------------------------------------------
 AUC <- function(y, yhat = seq_along(y)) {

### 2024-09-04 vignettes/cv.R
update cv.Rmd vignette for new plot() methods

@@ -113,7 +113,13 @@ summary(m.auto)
 cv(m.auto)
 
 ## ----cv-summary---------------------------------------------------------------
-summary(cv(m.auto))
+summary(cv.m.auto <- cv(m.auto))
+
+## ----plot.cv.crit-------------------------------------------------------------
+plot(cv.m.auto) # CV criterion
+
+## ----plot.cv.coefs------------------------------------------------------------
+plot(cv.m.auto, what="coefficients") # coefficient estimates
 
 ## ----cv.lm-2`-----------------------------------------------------------------
 summary(cv(m.auto, k = "loo"))
@@ -235,13 +241,16 @@ summary(cv(m.mroz,
    method = "hatvalues"))
 
 ## ----mroz-reps----------------------------------------------------------------
-summary(cv(
+summary(cv.mroz.reps <- cv(
   m.mroz,
   criterion = BayesRule,
   seed = 248,
   reps = 5,
   method = "Woodbury"
 ))
+
+## ----cv.mroz.reps-------------------------------------------------------------
+plot(cv.mroz.reps)
 
 ## ----model-comparison-with-reps-----------------------------------------------
 cv.auto.reps <- cv(

### 2024-09-05 vignettes/cv-extend.R
get rid of fixed.effects arg to cv() mixed-models methods; retain arg in cvMixed() utility

@@ -234,7 +234,6 @@ cv.glmmPQL <- function(model,
                        seed,
                        ncores = 1,
                        clusterVariables,
-                       fixed.effects = nlme::fixef,
                        ...) {
   cvMixed(
     model,
@@ -258,7 +257,7 @@ cv.glmmPQL <- function(model,
       level = 1,
       type = "response"
     ),
-    fixed.effects = fixed.effects,
+    fixed.effects = nlme::fixef, 
     verbose = FALSE,
     ...
   )

### 2024-09-12 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package.Rmd
add scratch/JSS-submission/Revsion-1/*

--- 2024-08-18 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
@@
 
-cv(m.auto, confint = TRUE)
-```
-The function reports the CV estimate of MSE, a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression.
+(cv.auto <- cv(m.auto, confint = TRUE))
+summary(cv.auto)
+```
+The `print()` method for `"cv"` objects reports the CV estimate of MSE; the `summary()` method, which we'll use extensively in the sequel, in additions reports a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression.
 Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint = TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.
@@
 ```{r cv.lm-2}
-cv(m.auto, k = "loo")
-```
-The \code{"hatvalues"} method reports only the CV estimate of MSE.
+summary(cv(m.auto, k = "loo"))
+```
+The \code{"hatvalues"} method computes only the CV estimate of MSE.
 Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted.
@@
 ```{r cv.lm-3}
-cv(m.auto, k = "loo", method = "naive", confint = TRUE)
+summary(cv(m.auto, k = "loo", method = "naive", confint = TRUE))
 ```
@@
 ```{r HSB-lmer-CV-cluster, cache=TRUE}
-cv(hsb.lmer, k = 10, clusterVariables = "school", seed = 5240)
+summary(cv(hsb.lmer, k = 10, clusterVariables = "school", seed = 5240))
 ```
@@
 ```{r HSB-lmer-CV-case, cache=TRUE}
-cv(hsb.lmer, seed = 1575)
+summary(cv(hsb.lmer, seed = 1575))
 ```
@@
           response = "mpg", AIC = FALSE)
-cvs
+summary(cvs)
 ```
@@
                        seed = 2120)
-recursiveCV.auto
-brief(recursiveCV.auto$selected.model)
-cv(mlist[[7]], seed = 2120)
+summary(recursiveCV.auto)
+brief(m.sel <- selectedModel(recursiveCV.auto))
+cv(m.sel, seed = 2120)
 ```
@@
 ```{r recursive-cv-alt, eval=FALSE}
-cv(mlist, data = Auto, seed = 2120, recursive = TRUE, save.model = TRUE)
+summary(cv(mlist, data = Auto, seed = 2120, recursive = TRUE, 
+           save.model = TRUE))
 ```
@@
 ```{r BEPS-cv}
-cv(m.beps, seed = 3465)
+summary(cv(m.beps, seed = 3465))
 ```
@@
 ```{r rsample-example-using-cv, cache=TRUE}
-(cv.attrition <- cv(mod.attrition, k = "loo", criterion = BayesRule))
-all.equal(cv.attrition$`CV crit`, 1 - mean(rs_obj$accuracy), 
-          check.attributes = FALSE)
+cv(mod.attrition, k = "loo", criterion = BayesRule)
+1 - mean(rs_obj$accuracy)
 ```

### 2024-09-13 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package.Rmd
replace selectedModel() with cvInfo.cvSelect() method

@@
 summary(recursiveCV.auto)
-brief(m.sel <- selectedModel(recursiveCV.auto))
+brief(m.sel <- cvInfo(recursiveCV.auto, "selected model"))
 cv(m.sel, seed = 2120)

### 2024-09-13 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package-script.R

@@ -458,10 +458,10 @@ compareFolds(cvs)
 recursiveCV.auto <- cv(selectModelList, data = Auto,
                        working.model = mlist, save.model = TRUE,
                        seed = 2120)
-recursiveCV.auto
-selectedModel(recursiveCV.auto)
+summary(recursiveCV.auto)
+brief(m.sel <- cvInfo(recursiveCV.auto, "selected model"))
 # CV for selected model
-summary(cv(selectedModel(recursiveCV.auto), seed = 2120))
+cv(m.sel, seed = 2120)
 
   # equivalent, using recursive = TRUE
 

### 2024-09-13 vignettes/cv-selection.R

@@ -137,8 +137,8 @@ recursiveCV.auto <- cv(
   seed = 2120
 )
 summary(recursiveCV.auto)
-recursiveCV.auto$selected.model
-cv(m.7, seed = 2120) # same seed for same folds
+(m.sel <- cvInfo(recursiveCV.auto, "selected model"))
+cv(m.sel, seed = 2120) # same seed for same folds
 
 ## ----recursive-cv-alt---------------------------------------------------------
 recursiveCV.auto.alt <- cv(

### 2024-09-14 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package.Rmd
fix a typo in JSS paper

@@
 ```
-The `print()` method for `"cv"` objects reports the CV estimate of MSE; the `summary()` method, which we'll use extensively in the sequel, in additions reports a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression.
+The `print()` method for `"cv"` objects reports the CV estimate of MSE; the `summary()` method, which we'll use extensively in the sequel, in addition reports a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression.
 Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint = TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.

### 2024-09-14 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package.Rmd
Divide section in JSS paper on other software into R and SAS/Stata/Python

@@
 
+In this section with briefly compare the \code{cv} package to other packages in \proglang{R} that support cross-validation, and then even more briefly consider facilities for cross-validation in other commonly used statistical software, \proglang{SAS}, \proglang{Stata}, and \proglang{Python}.
+
+## Other R software
+
 The \pkg{cv} package is far from unique in implementing cross-validation of regression models in \proglang{R}.
@@
 
+## Other statistical software: SAS, Stata, and Python
+
 # Computational notes

### 2024-09-14 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package.Rmd
modify intro to JSS paper

@@
 A number of existing \proglang{R} packages include functions for cross-validating regression models.
-We briefly situate the \pkg{cv} package relative to other \proglang{R} software for cross-validation in Section \ref{comparing-cv-to-other-r-software-for-cross-validation}.
+We briefly situate the \pkg{cv} package relative to other \proglang{R} software for cross-validation, and to other statistical software (in particular, \proglang{SAS}, \proglang{Stata}, and \proglang{Python}), in Section \ref{comparing-cv-to-other-software-for-cross-validation}.
 
@@
 
-# Comparing cv to other R software for cross-validation
+# Comparing cv to other software for cross-validation
 

### 2024-09-15 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package.Rmd
change recursive -> meta

@@
 The \code{"function"} method for \code{cv()} cross-validates a model-specification process in a general manner.
-Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package, transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model.[^adding-model-selection]
+Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package, transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model, a process that we term "meta cross-validation."[^adding-model-selection]
 
@@
 
-## Example: Applying recursive CV to polynomial regression for the Auto data[^recursive-cv]
-
-[^recursive-cv]: What we call here "recursive CV" has also been termed "nested CV" [e.g., in the \pkg{cvms} package by @OlsenZachariae:2024].
+## Example: Applying meta CV to polynomial regression for the Auto data[^meta-cv]
+
+[^meta-cv]: What we call here "meta CV" has also been termed "nested CV" [e.g., in the \pkg{cvms} package by @OlsenZachariae:2024].
 We prefer, however, to reserve the term "nested CV" for the procedure described by @BatesHastieTibshirani:2023.
@@
 If we select the 7th degree polynomial model, intending to use it for prediction, the CV estimate of the MSE for this model will be optimistic.
-One solution is to cross-validate the process of using CV to select the "best" model---that is, to apply CV to CV recursively.
+One solution is to cross-validate the process of using CV to select the "best" model---that is, to apply meta-CV.
 The function \code{selectModelList()}, which is suitable for use with \code{cv()}, implements this idea.
@@
 Applying \code{selectModelList()} to the \code{Auto} polynomial-regression models, and using 10-fold CV, we obtain:
-```{r recursive-CV-polynomials}
-recursiveCV.auto <- cv(selectModelList, data = Auto, 
+```{r meta-CV-polynomials}
+metaCV.auto <- cv(selectModelList, data = Auto, 
                   working.model = mlist, save.model = TRUE, 
                        seed = 2120)
-summary(recursiveCV.auto)
-brief(m.sel <- cvInfo(recursiveCV.auto, "selected model"))
+summary(metaCV.auto)
+brief(m.sel <- cvInfo(metaCV.auto, "selected model"))
 cv(m.sel, seed = 2120)
 ```
-As expected, recursive CV produces a larger estimate of MSE for the selected 7th degree polynomial model than CV applied directly to this model (using the same seed for \proglang{R}'s random-number generator to obtain the same folds).
-
-We can equivalently call \code{cv()} with the list of models as its first argument and set the argument \code{recursive = TRUE} (identical output not shown):
-```{r recursive-cv-alt, eval=FALSE}
-summary(cv(mlist, data = Auto, seed = 2120, recursive = TRUE, 
+As expected, meta CV produces a larger estimate of MSE for the selected 7th degree polynomial model than CV applied directly to this model (using the same seed for \proglang{R}'s random-number generator to obtain the same folds).
+
+We can equivalently call \code{cv()} with the list of models as its first argument and set the argument \code{meta = TRUE} (identical output not shown):
+```{r meta-cv-alt, eval=FALSE}
+summary(cv(mlist, data = Auto, seed = 2120, meta = TRUE, 
            save.model = TRUE))

### 2024-09-15 R/cv-select.R

@@ -10,7 +10,7 @@
 #' \pkg{car} package; \code{selectTransAndStepAIC()} combines predictor and response
 #' transformations with predictor selection; and \code{selectModelList()}
 #' uses cross-validation to select a model from a list of models created by
-#' \code{\link{models}()} and employs (recursive) cross-validation to assess the predictive
+#' \code{\link{models}()} and employs (meta) cross-validation to assess the predictive
 #' accuracy of this procedure.
 #'
 #' @param data full data frame for model selection.
@@ -648,11 +648,11 @@ selectTransStepAIC <- function(data,
   )
 }
 
-#' @describeIn cv.function select a model using (recursive) CV.
+#' @describeIn cv.function select a model using (meta) CV.
 #' @param quietly if \code{TRUE} (the default), simple messages (for example about the
 #' value to which the random-number generator seed is set), but not warnings or
 #' errors, are suppressed.
-#' @param k.recurse the number of folds for recursive CV; defaults
+#' @param k.meta the number of folds for meta CV; defaults
 #' to the value of \code{k}; may be specified as \code{"loo"} or
 #' \code{"n"} as well as an integer.
 #' @export
@@ -662,7 +662,7 @@ selectModelList <-
            model,
            criterion = mse,
            k = 10L,
-           k.recurse = k,
+           k.meta = k,
            details =  k <= 10L,
            save.model = FALSE,
            seed = FALSE,
@@ -679,7 +679,7 @@ selectModelList <-
           model,
           data,
           criterion = criterion,
-          k = k.recurse,
+          k = k.meta,
           details = FALSE,
           quietly = quietly,
           seed = FALSE,
@@ -703,7 +703,7 @@ selectModelList <-
         model,
         data[-indices,],
         criterion = criterion,
-        k = k.recurse,
+        k = k.meta,
         details = details,
         quietly = quietly,
         seed = FALSE,
@@ -732,7 +732,7 @@ selectModelList <-
 #' m3 <- lm(prestige ~ (income + education)*type, data=Duncan)
 #' summary(cv.sel <- cv(selectModelList, data=Duncan, seed=5963,
 #'                      working.model=models(m1, m2, m3),
-#'                      save.model=TRUE)) # recursive CV
+#'                      save.model=TRUE)) # meta CV
 #' cvInfo(cv.sel, "selected model")
 #'
 #' @describeIn cv.function print the coefficients from the selected models

### 2024-09-15 R/cv.modList.R

@@ -3,7 +3,7 @@
 #' A \code{\link{cv}()} method for an object of class  \code{"modlist"},
 #' created by the \code{models()} function. This \code{cv()} method simplifies
 #' the process of cross-validating several models on the same set of CV folds
-#' and may also be used for recursive CV, where CV is used to select one from among
+#' and may also be used for meta CV, where CV is used to select one from among
 #' several models. \code{models()} performs some
 #' "sanity" checks, warning if the models are of different classes, and
 #' reporting an error if they are fit to apparently different data sets or
@@ -21,7 +21,7 @@
 #' to be used to create the same set of CV folds for all of the models;
 #' if omitted, a seed will be randomly generated and saved. Not used for
 #' leave-one-out CV.
-#' @param recursive if \code{TRUE} (the default is \code{FALSE}), cross-validation
+#' @param meta if \code{TRUE} (the default is \code{FALSE}), cross-validation
 #' is performed recursively to select a "best" model deleting each fold in turn
 #' by calculating the CV estimate of the criterion for the remaining folds;
 #' this is equivalent to employing the \code{\link{selectModelList}()} model-selection
@@ -68,7 +68,7 @@
 #' @param object an object to summarize.
 #' @return \code{models()} returns a \code{"modList"} object, the
 #' \code{cv()} method for which returns a \code{"cvModList"} object,
-#' or, when \code{recursive=TRUE}, an object of class \code{c("cvSelect", "cv")}.
+#' or, when \code{meta=TRUE}, an object of class \code{c("cvSelect", "cv")}.
 #' @seealso \code{\link{cv}}, \code{\link{cv.merMod}},
 #' \code{\link{selectModelList}}.
 #' @examples
@@ -88,10 +88,10 @@
 #'                     data=Duncan, seed=5963, confint=TRUE, level=0.50))
 #'                  # nb: n too small for accurate CIs
 #' plot(cv.models.ci)
-#' (cv.models.recursive <- cv(models(m1=m1, m2=m2, m3=m3),
-#'                            data=Duncan, seed=5963,
-#'                            recursive=TRUE, save.model=TRUE))
-#' cv.models.recursive$selected.model
+#' (cv.models.meta <- cv(models(m1=m1, m2=m2, m3=m3),
+#'                       data=Duncan, seed=5963,
+#'                       meta=TRUE, save.model=TRUE))
+#' cvInfo(cv.models.meta, "selected model")
 #' })
[…]

### 2024-09-15 tests/testthat/test-cvSelect.R

@@ -118,7 +118,7 @@ test_that("cvSelect parallel selectTransStepAIC", {
   )
 })
 
-# test recursive CV
+# test meta CV
 
 data("Auto", package="ISLR2")
 for (p in 1:10){
@@ -127,19 +127,19 @@ for (p in 1:10){
   eval(parse(text=command))
 }
 
-cv1 <- cv(selectModelList, Auto, k=10, k.recurse="loo",
+cv1 <- cv(selectModelList, Auto, k=10, k.meta="loo",
           working.model=models(m.1, m.2, m.3, m.4, m.5,
                                m.6, m.7, m.8, m.9, m.10),
           save.model=TRUE,
           seed=123)
 cv2 <- cv(models(m.1, m.2, m.3, m.4, m.5,
                  m.6, m.7, m.8, m.9, m.10),
-          recursive=TRUE,
-          k=10, k.recurse="loo",
+          meta=TRUE,
+          k=10, k.meta="loo",
           save.model=TRUE,
           seed=123)
 cv1$criterion <- cv2$criterion <- NULL
-test_that("cv(SelectModelList()) and cv.modList() produce same recursive CV/k-LOO", {
+test_that("cv(SelectModelList()) and cv.modList() produce same meta CV/k-LOO", {
   expect_equal(cv1, cv2)
 })
 
@@ -150,15 +150,15 @@ cv1 <- cv(selectModelList, Auto,
           seed=123)
 cv2 <- cv(models(m.1, m.2, m.3, m.4, m.5,
                  m.6, m.7, m.8, m.9, m.10),
-          recursive=TRUE,
+          meta=TRUE,
           save.model=TRUE,
           seed=123)
 cv1$criterion <- cv2$criterion <- NULL
-test_that("cv(SelectModelList()) and cv.modList() produce same recursive CV/k-fold", {
+test_that("cv(SelectModelList()) and cv.modList() produce same meta CV/k-fold", {
   expect_equal(cv1, cv2)
 })
 
-test_that("recursive SelectModelList() works with parallel computation loo", {
+test_that("meta SelectModelList() works with parallel computation loo", {
   expect_equal(
     cv(selectModelList, Auto, k="loo",
        working.model=models(m.1, m.2, m.3, m.4, m.5,
@@ -172,7 +172,7 @@ test_that("recursive SelectModelList() works with parallel computation loo", {
   )
 })
 
-test_that("recursive SelectModelList() works with parallel computation k-fold", {
[…]

### 2024-09-16 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package.Rmd
add footnote on cvInfo() and as.data.frame() to JSS paper

@@
 
-Applying \code{selectModelList()} to the \code{Auto} polynomial-regression models, and using 10-fold CV, we obtain:
+Applying \code{selectModelList()} to the \code{Auto} polynomial-regression models, and using 10-fold CV, we obtain:[^cvInfo]
 ```{r meta-CV-polynomials}
@@
 ```
+
+[^cvInfo]: Two functions are employed to access information in objects produced by functions in the \pkg{cv} package: \code{cvInfo()} and \code{as.data.frame()} are generic functions (the latter a standard-\proglang{R} generic) with methods for classes \code{"cv"}, \code{"cvList"}, \code{"cvModlist"}, and \code{"cvSelect"} (in the case of \code{cvInfo()}) , and \code{"cv"}, \code{"cvList"}, and \code{"cvModlist"} (in the case of \code{as.data.frame()}).
+Here we use \code{cvInfo(metaCV.auto, "selected model")} to retrieve the model selected by \code{cv()} via \code{selectModelList()}.
+See \code{?cv} and \code{?cvInfo.cvSelect} for details.
+
 As expected, meta CV produces a larger estimate of MSE for the selected 7th degree polynomial model than CV applied directly to this model (using the same seed for \proglang{R}'s random-number generator to obtain the same folds).

### 2024-09-22 scratch/JSS-submission/Revision-1/article-add.Rmd
additions re Python

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0003-0076-5532
    affiliation: 'York University'
    address: |
      | Toronto, Ontario, Canada
    email: \email
    url: http://blackwell.math.yorku.ca/gmonette
title: DRAFT Addition re Other Packages
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    extra_dependencies: "subfig"
bibliography: ["cv-add.bib"]
---

# Comparing cv to other software for cross-validation

We have examined provisions for cross-validation
in \proglang{SAS}, \proglang{Stata}, and \proglang{Python} although
we do not have recent experience with languages other than \proglang{R}.

In \proglang{Python} [@pedregosaScikitlearnMachineLearning:2011], 
statistical modelling is supported through extensive libraries
such as \pkg{scikit-learn} [@ScikitLearnDevelopers:2024]
and \pkg{Tensorflow} [@tensorflow:2024].  
Functions for cross-validation are provided in the \pkg{Scikit-learn}
library that includes functions
for k-fold, repeated k-fold, leave-one-out, leave-p-out cross-validation.
These functions can be applied to data sets consisting of independent observations,
hierarchical groups, or time series data. Documentation
for the \pkg{scikit-learn} provides
a sample program [@ScikitLearnDevelopers:2024] to show
users how to implement nested cross-validation.

### 2024-09-24 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package.Rmd
add section on other software to JSS paper

@@
   \usepackage{amsmath}
+  \usepackage{amssymb}
 output: 
@@
 
-## Other statistical software: SAS, Stata, and Python
+## Comparing cv to other software for cross-validation: SAS, Stata, and Python
+
+Although we do not have recent experience with statistical software for cross-validation other than in \proglang{R},[^other-software] we have examined provisions for CV in \proglang{SAS}, \proglang{Stata}, and \proglang{Python}, primarily by locating and reading documentation.
+
+[^other-software]: It is our impression that these three statistical systems, in addition to \proglang{R}, are most likely to be used for applications of cross-validation.
+We were once frequent users of \proglang{SAS}, though not recently, and have only cursory experience with \proglang{Stata} and \proglang{Python}, so we are not ideally situated to evaluate CV software for any of the three.
+
+### SAS
+
+Cross-validation in \proglang{SAS} is associated with particular procedures ("\code{PROC}s"), and is distributed across several products, including, for example, \code{PROC LOGISTIC} and \code{PROC PLS} in \proglang{SAS/STAT} [@SAS-Stat:2020], in \proglang{SAS Enterprise Miner} [@SAS-Enterprise-Miner:2018], and in \proglang{SAS Visual Data Mining and Machine Learning} [@SAS-Visual-Data-Mining:2020].
+These appear to be straightforward implementations of $k$-fold CV.
+
+In addition, \proglang{SAS} has programming capabilities, including via its macro language, and we're aware of several independent implementations of CV in \proglang{SAS} macros.
+These tend to be limited in scope.
+For example @Slaets-et-al:2014 employ and provide access to a \proglang{SAS} macro for $k$-fold CV of mixed-effects models, which repeatedly calls \code{PROC MIXED} in \proglang{SAS/STAT}, and which performs cluster-based resampling (one of the options that we provide for mixed-effects models in the \pkg{cv} package).
+
+### Stata
+
+As far as we can tell, \proglang{Stata} has no standard implementation of cross-validation.
+Like \proglang{R}, however, \proglang{Stata} is extensible through a programming language that allows users to define new commands in so-called "\code{ado}" files.
+
+Several such contributed \proglang{Stata} CV commands, varying in generality, are shared in an archive maintained at the Boston College Department of Economics.
+For example, the \code{CV\_REGRESS} command [@Rios-Avila:2018] computes leave-one-out CV for least-squares regression using the hatvalues, and thus is comparable to \code{cv()} for linear models with \code{method = "hatvalues"}.
+In contrast, the \code{CROSSVALIDATE} command [@Schonlau:2020] performs $k$-fold CV for a wide range of regression models fit in \proglang{Stata}, and is roughly comparable to the default method of \code{cv()}, although \code{CROSSVALIDATE} returns cross-validated fitted values, from which cost criteria can be computed, rather than (as \code{cv()}) cost-criteria directly.
+
+### Python
+
+The programming language \proglang{Python} is often the software of choice for "machine learning," in which cross-validation can play a prominent role.
+Machine learning and statistical modeling more generally are supported through extensive \proglang{Python} libraries, such as \pkg{scikit-learn} [@Pedregosa-et-al:2011; @ScikitLearnDevelopers:2024] and \pkg{TensorFlow} [@TensorFlow:2015, which isn't limited to use within \proglang{Python}].
+
+Functions for cross-validation are provided in \pkg{scikit-learn}, which supports $k$-fold, repeated $k$-fold, and leave-one-out CV.
+These functions can be applied to data sets consisting of independent observations, hierarchical groups, or time-series data.
+A sample program in the \pkg{scikit-learn} documentation [@ScikitLearnDevelopers:2024] demonstrates how to implement meta cross-validation (termed "nested cross-validation" in the documentation).
+
+The approach taken by \pkg{scikit-Learn} for independent observations is similar to that in the default \code{cv()} function in the \pkg{cv} package.
+The approach for hierarchical data is similar, but not identical, to that in \pkg{cv} (as described in Section \ref{cross-validating-mixed-effects-models}).
+Our \code{cv()} package doesn't currently provide for cross-validating time-series models.
 

### 2024-10-03 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package.Rmd
fix typos in JSS paper

@@
 The \code{"function"} method for \code{cv()} cross-validates a model-specification process in a general manner.
-Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package, transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model, a process that we term "meta cross-validation."[^adding-model-selection]
+Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package, transforms predictors and the response in a regression model towards normality; \code{selectTransStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model, a process that we term "meta cross-validation."[^adding-model-selection]
 
@@
 The \code{"function"} method for \code{cv()} allows us to cross-validate the whole model-selection procedure.
-The first argument to \code{cv()}---here \code{selectTransAndStepAIC}---is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion:
+The first argument to \code{cv()}---here \code{selectTransStepAIC}---is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion:
 ```{r Auto-transform-and-select}
@@
 ```
-Computing time for \pkg{rsample} and \pkg{caret} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but almost two orders of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.
+Computing time for \pkg{rsample} and \pkg{caret} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but almost two orders of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower than \code{cv()} with \code{method="hatvalues"}.
 

### 2025-03-04 scratch/JSS-submission/Resubmission/response-to-reviewer.Rmd
draft response to reviewer

---
title: "Response to the Reviewer"
author: "John Fox and Georges Monette"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1.  "I was surprised to see references to alternative packages only at the end of the paper rather than in the introduction. It might be beneficial to include a sentence in the introduction to clarify this."

Actually, we already had the following two sentences in the introductory section: "A number of existing R packages include functions for cross-validating regression models. We briefly situate the **cv** package relative to other R software for cross-validation, and to other statistical software (in particular, SAS, Stata, and Python), in Section 6." We elaborate slightly in the resubmission, indicating *which* R packages we take up in Section 6. We believe that the discussion itself should remain at the end of the paper, *after* we have discussed the facilities for cross-validation in the **cv** package.

2.  "What is the purpose of using insight::get_data on page 5? The insight package is also mentioned on page 31. I believe it would be helpful to include a brief explanation of why it is being used."

As suggested, we added a brief explanation of the advantage of using the **insight** package, which provides a robust way to access model components for many classes of statistical models.

3.  "Does the package allow for the addition of an uncertainty interval to the plots in Figure 2? I think that including a measure of uncertainty would be useful."

Yes, it does. As we explained on p. 5 of the original paper in discussing the `confint` argument to `cv()`, "`confint`, whether or not to compute a confidence interval for the CV criterion, defaulting to `TRUE` if there are at least 400 cases." The `Auto` data used in the preliminary example has fewer than 400 cases. When there are confidence limits in the `"cv"` objects, the `plot()` method used for Fig. 2 draws confidence intervals. We've added a footnote explaining this behaviour and have also explained why we adopted the $n \ge 400$ rule for computing confidence intervals by default.

4.  "In Section 4.1, six pages of the text are devoted to transformations before the actual CV content is addressed. Would it be possible/appropriate to shorten the discussion on transformations?"

We've compressed this material, primarily by removing the scatterplot matrix and C+R plots for the transformed data. We'd prefer not to abbreviate the section further: The purpose of the section is to show how to automate a complex model-selection method, so we think that it's important to explain how the method is applied "manually" before automating it.

5.  "On page 21, when you state “We proceed to transform the numeric predictors in the Auto regression towards multivariate normality”, I assume that the transformations being applied are marginal rather than joint, as joint transformations would cause the covariates to lose their interpretation. If so, then perhaps it would be worth clarifying that powerTransform adjusts the margins to achieve multivariate Gaussianity."

Yes, the predictors are transformed individually with the object of making their joint distribution as close to multinormal as possible. We have clarified this point in the resubmssion.

6.  "On page 29, when computing mse(Auto\$mpg, exp(fitted(m.step))), I think that you could use the relationship between the normal and log-normal distributions to correct for bias."

Yes, one could compute the MSE on the log-scale for both models (which we think is what the reviewer is getting at here). This is a good point, and has a more general implication: It's necessary to compare models on the same response scale, but that need not be the original response scale. We've added to what was footnote 12 on p. 25 of the original submission to explain this option (crediting the reviewer).

7.  "Finally, the computational section in Section 7 might be better placed in an appendix. It seems somewhat odd to conclude the paper with these details."

We agree with this point, and what was Section 7 is now an appendix.

### 2025-03-04 scratch/JSS-submission/Resubmission/response-to-reviewer.Rmd
preliminary edits to JSS resubmission

@@
 Actually, we already had the following two sentences in the introductory section: "A number of existing R packages include functions for cross-validating regression models.
-We briefly situate the **cv** package relative to other R software for cross-validation, and to other statistical software (in particular, SAS, Stata, and Python), in Section 6." We elaborate slightly in the resubmission, indicating *which* R packages we take up in Section 6.
-We believe that the discussion itself should remain at the end of the paper, *after* we have discussed the facilities for cross-validation in the **cv** package.
+We briefly situate the **cv** package relative to other R software for cross-validation, and to other statistical software (in particular, SAS, Stata, and Python), in Section 6." We elaborate slightly in the resubmission, indicating in the introduction *which* R packages we take up in Section 6.
+We believe that the discussion itself should remain at the end of the paper, *after* we have described the facilities for cross-validation in the **cv** package.
 
@@
 2 draws confidence intervals.
-We've added a footnote explaining this behaviour and have also explained why we adopted the $n \ge 400$ rule for computing confidence intervals by default.
+We've added a footnote explaining this behaviour and have also pointed to the appendix of the paper for an explanation of the $n \ge 400$ rule for computing confidence intervals by default.
 
@@
 
-6.  "On page 29, when computing mse(Auto\$mpg, exp(fitted(m.step))), I think that you could use the relationship between the normal and log-normal distributions to correct for bias."
+6.  "On page 29, when computing `mse(Auto\$mpg, exp(fitted(m.step)))`, I think that you could use the relationship between the normal and log-normal distributions to correct for bias."
 
@@
 We've added to what was footnote 12 on p.
-25 of the original submission to explain this option (crediting the reviewer).
+25 of the original submission (now footnote xx on p.
+xx) to explain this option (crediting the reviewer).
 

### 2025-03-04 scratch/JSS-submission/Resubmission/Fox-Monette-cv-package.Rmd
preliminary edits to JSS resubmission

--- 2024-10-03 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package.Rmd
@@
 Because predictions are based on smaller data sets, each of size approximately $n - n/k$, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) is more biased than estimated prediction error for LOO CV.
-It is possible, however, to correct $k$-fold CV for bias (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+It is possible, however, to correct $k$-fold CV for bias (see Appendix A.2).
 
@@
 A number of existing \proglang{R} packages include functions for cross-validating regression models.
-We briefly situate the \pkg{cv} package relative to other \proglang{R} software for cross-validation, and to other statistical software (in particular, \proglang{SAS}, \proglang{Stata}, and \proglang{Python}), in Section \ref{comparing-cv-to-other-software-for-cross-validation}.
-
-The final section of the paper (Section \ref{computational-notes}) describes some computational details, including efficient CV computations for linear and generalized-linear models, and computation of bias-corrected CV criteria.
+In Section \ref{comparing-cv-to-other-software-for-cross-validation}, we briefly situate the \pkg{cv} package relative to other \proglang{R} software for cross-validation (directly comparing \pkg{cv} to the \pkg{rsample} and \pkg{caret} packages), and to other statistical software (in particular, \proglang{SAS}, \proglang{Stata}, and \proglang{Python}).
+
+An appendix to the paper describes some computational details, including efficient CV computations for linear and generalized-linear models, and computation of bias-corrected CV criteria.
 
@@
 
-* \code{data}, which can usually be inferred from the \code{model} object.
+* \code{data}, which can usually be inferred from the \code{model} object, using the \code{get_data()` function from the \pkg{insight} package [@LudeckeWaggonerMakowski:2019].
+The \pkg{insight} package provides a robust way to access model components for many classes of statistical models.
 
@@
 
-* \code{confint}, whether or not to compute a confidence interval for the CV criterion, defaulting to \code{TRUE} if there are at least 400 cases; a confidence interval is computed only if the CV criterion can be expressed as the average of casewise components (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals} for details).
+* \code{confint}, whether or not to compute a confidence interval for the CV criterion, defaulting to \code{TRUE} if there are at least 400 cases; a confidence interval is computed only if the CV criterion can be expressed as the average of casewise components (see Appendix A.2 for details).
 
@@
 
+```{r include=FALSE, echo=FALSE}
+# to make results below reproducible
+set.seed(74595)
+```
 ```{r cv-lm-1}
@@
 ```
-The `print()` method for `"cv"` objects reports the CV estimate of MSE; the `summary()` method, which we'll use extensively in the sequel, in addition reports a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression.
+The `print()` method for `"cv"` objects reports the CV estimate of MSE; the `summary()` method, which we'll use extensively in the sequel, in addition reports a bias-adjusted estimate of the MSE (the bias adjustment is explained in Appendix A.2), and the MSE is also computed for the original, full-sample regression.
 Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint = TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.
@@
 An alternative is to use a robust measure of model lack-of-fit, such as the median absolute error instead of the mean-squared error, employing the \code{medAbsErr()} function from the \pkg{cv} package.
-The median absolute error, however, cannot be expressed as a casewise average (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+The median absolute error, however, cannot be expressed as a casewise average (see Appendix A.2).
 
@@
 After doing some error checking, \code{BayesRule()} rounds the predicted probability \code{yhat} of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction, and then reports the proportion of incorrect predictions.
-Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, enabling the computation of bias correction and confidence intervals (as discussed in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, enabling the computation of bias correction and confidence intervals (as discussed in Appendix A.2).
 
@@
 The default method of \code{GetResponse()} (a function supplied by the \pkg{cv} package---see \code{?GetResponse}) fails for a \code{"multinom"} object.
-A straightforward solution is to supply a \code{GetResponse.multinom()} method that returns the factor response [using the \code{get\_response()} function from the \pkg{insight} package, @LudeckeWaggonerMakowski:2019],
+A straightforward solution is to supply a \code{GetResponse.multinom()} method that returns the factor response [using the \code{get\_response()} function from the \pkg{insight} package,
 ```{r GetResponse.multinom}
@@
 
-The approach taken by \pkg{scikit-Learn} for independent observations is similar to that in the default \code{cv()} function in the \pkg{cv} package.
+The approach taken by \pkg{scikit-learn} for independent observations is similar to that in the default \code{cv()} method in the \pkg{cv} package.
 The approach for hierarchical data is similar, but not identical, to that in \pkg{cv} (as described in Section \ref{cross-validating-mixed-effects-models}).
-Our \code{cv()} package doesn't currently provide for cross-validating time-series models.
-
-# Computational notes
-
-## Efficient computations for linear and generalized linear models
+Our \pkg{cv} package doesn't currently provide for cross-validating time-series models, but we anticipate that the next version of the package will.
+
+# Appendix: Computational notes {-}
+
+## A.1 Efficient computations for linear and generalized linear models {-}
 
@@
 
-## Computation of the bias-corrected CV criterion and confidence intervals
+## A.2 Computation of the bias-corrected CV criterion and confidence intervals {-}
 

### 2025-03-04 scratch/JSS-submission/Resubmission/response-to-reviewer.Rmd
complete draft JSS resubmission and response to reviewer

@@
 ---
-title: "Response to the Reviewer"
+title: "Point-by-Point Response to the Reviewer"
 author: "John Fox and Georges Monette"
@@
 2 draws confidence intervals.
-We've added a footnote explaining this behaviour and have also pointed to the appendix of the paper for an explanation of the $n \ge 400$ rule for computing confidence intervals by default.
+We've added a footnote (fn.
+5) explaining this behaviour and have also in describing the `confint` argument pointed to the appendix of the paper for an explanation of the $n \ge 400$ rule for computing confidence intervals by default.
 
@@
 
-We've compressed this material, primarily by removing the scatterplot matrix and C+R plots for the transformed data.
+We've compressed this material by removing the code and graphs for the scatterplot matrix and C+R plots for the transformed data, shortening the paper by 2 pages.
 We'd prefer not to abbreviate the section further: The purpose of the section is to show how to automate a complex model-selection method, so we think that it's important to explain how the method is applied "manually" before automating it.
@@
 
-Yes, the predictors are transformed individually with the object of making their joint distribution as close to multinormal as possible. We have clarified this point in the resubmssion.
+Yes, the predictors are transformed individually with the object of making their joint distribution as close to multinormal as possible. We have clarified this point in the resubmission.
 
@@
 We've added to what was footnote 12 on p.
-25 of the original submission (now footnote xx on p.
-xx) to explain this option (crediting the reviewer).
+25 of the original submission (now footnote 13 on p.
+23) to explain this option (crediting the reviewer).
 

### 2025-03-04 scratch/JSS-submission/Resubmission/Fox-Monette-cv-package.Rmd
complete draft JSS resubmission and response to reviewer

@@
 Alternatively, as here, we can supply a pre-constructed, optionally named, list of models as the single argument to \code{models()}.
-To visualize the results, comparing the models, we invoke the \code{plot()} method for the \code{"cvModList"} objects returned by \code{cv()} (see Figure \ref{fig:polynomial-regression-CV-graph-2}):
+To visualize the results, comparing the models, we invoke the \code{plot()} method for the \code{"cvModList"} objects returned by \code{cv()} (see Figure \ref{fig:polynomial-regression-CV-graph-2}):[^plot-confint]
 
@@
 ```
+
+[^plot-confint]: The objects \code{cv.auto.10} and \code{cv.auto.loo} do not include confidence intervals for the CV MSE estimates because, as we explained, there are fewer than 400 cases in the \code{Auto} data.
+Consequently, confidence intervals are not shown in the graphs in Figure \ref{fig:polynomial-regression-CV-graph-2}.
+We can force `cv()` to compute confidence intervals by adding the argument `confint = TRUE`; we invite the reader to do this and to verify that the resulting confidence intervals are very broad, as was the case for the quadratic model fit to the \code{Auto} data in the preceding section.
 
@@
 The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964].
+Although the multinormal likelihood is maximized, the variables are transformed individually.
 Several "families" of power transformations can be used, including the original Box-Cox family (which is the default), simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap.
@@
 ```
-
-The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore-scatterplot-matrix}, on page \pageref{fig:Auto-explore-scatterplot-matrix}, and Figure \ref{fig:Auto-transformed-scatterplot-matrix}):
-```{r Auto-transformed-scatterplot-matrix}
-#| out.width = "95%",
-#| fig.align = 'center',
-#| fig.height = 5.5,
-#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data."
-scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
-                  + acceleration, 
-                  smooth=list(spread = FALSE), data = A, pch = ".")
-```
-
-And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore-cr-plots}, on page \pageref{fig:Auto-explore-cr-plots}, and Figure \ref{fig:Auto-transformed-cr-plots}):
-
-```{r Auto-transformed-cr-plots}
-#| out.width = "70%",
-#| fig.align = 'center',
-#| fig.height = 4.25,
-#| fig.cap = "Component + residual plots for the regression model fit to the transformed \\code{Auto} data."
-crPlots(m, terms = ~ displacement + horsepower + weight + acceleration, 
-        pch = ".", ylab = "C+R", las = 2)
-```
+The transformed numeric variables are much better-behaved and the partial relationships in the model fit to the transformed data are much more nearly linear (as we invite the reader to verify by redrawing the scatterplot matrix for the transformed variables and the C+R plots for the regression fit to the transformed data).
 
@@
 The median absolute error, however, cannot be expressed as a casewise average (see Appendix A.2).
+Yet another possibility, suggested to us by a reviewer of an earlier version of this paper, is to compute the MSE for both models on the log-response scale rather than on the original scale of the response.
+This suggestion is quite general, in that the essential point is to use the \emph{same} response scale for all models to be compared, not necessarilty the \emph{original} scale of the response.
 
@@
 
-[^times-10]: As before, we ask \code{microbenchmark()} to report relative timings, and also as before, we set \code{times = 10} rather than the default \code{times = 100}, with the consequence that we fail to detect a statistically reliable difference between using the Woodbury matrix identity and using hatvalues for the \code{cv()} computations.
+[^times-10]: As before, we ask \code{microbenchmark()} to report relative timings, and also as before, we set \code{times = 10} rather than the default \code{times = 100}.
 

### 2025-03-07 scratch/JSS-submission/Resubmission/response-to-reviewer.Rmd
tweak response to reviewer
tweak respose to reviewer

@@
 
-2.  "What is the purpose of using insight::get_data on page 5? The insight package is also mentioned on page 31. I believe it would be helpful to include a brief explanation of why it is being used."
+2.
+"What is the purpose of using `insight::get_data` on page 5?
+The **insight** package is also mentioned on page 31.
+I believe it would be helpful to include a brief explanation of why it is being used."
 
@@
 We've added a footnote (fn.
-5) explaining this behaviour and have also in describing the `confint` argument pointed to the appendix of the paper for an explanation of the $n \ge 400$ rule for computing confidence intervals by default.
+5) explaining this behaviour, and have also, in describing the `confint` argument, pointed to the appendix of the paper for an explanation of the $n \ge 400$ rule for computing confidence intervals by default.
 

### 2025-03-07 scratch/JSS-submission/Resubmission/response-to-reviewer.Rmd
further tweaks to response to reviewer

@@
 ```
+
+We thank the reviewer for his or her close reading of the paper.
 
@@
 5.
-"On page 21, when you state “We proceed to transform the numeric predictors in the Auto regression towards multivariate normality”, I assume that the transformations being applied are marginal rather than joint, as joint transformations would cause the covariates to lose their interpretation.
-If so, then perhaps it would be worth clarifying that powerTransform adjusts the margins to achieve multivariate Gaussianity."
+"On page 21, when you state “We proceed to transform the numeric predictors in the `Auto` regression towards multivariate normality”, I assume that the transformations being applied are marginal rather than joint, as joint transformations would cause the covariates to lose their interpretation.
+If so, then perhaps it would be worth clarifying that `powerTransform` adjusts the margins to achieve multivariate Gaussianity."
 
@@
 
-6.  "On page 29, when computing `mse(Auto\$mpg, exp(fitted(m.step)))`, I think that you could use the relationship between the normal and log-normal distributions to correct for bias."
+6.  "On page 29, when computing `mse(Auto$mpg, exp(fitted(m.step)))`, I think that you could use the relationship between the normal and log-normal distributions to correct for bias."
 

### 2025-03-08 scratch/JSS-submission/Resubmission/response-to-reviewer.Rmd
more tweaks to response to reviewer

@@
 
-Actually, we already had the following two sentences in the introductory section: "A number of existing R packages include functions for cross-validating regression models.
-We briefly situate the **cv** package relative to other R software for cross-validation, and to other statistical software (in particular, SAS, Stata, and Python), in Section 6." We elaborate slightly in the resubmission, indicating in the introduction *which* R packages we take up in Section 6.
+Actually, we already had the following two sentences in the introductory section: "A number of existing \textsf{R} packages include functions for cross-validating regression models.
+We briefly situate the **cv** package relative to other \textsf{R} software for cross-validation, and to other statistical software (in particular, \textsf{SAS}, \textsf{Stata}, and \textsf{Python}), in Section 6." We elaborate slightly in the resubmission, indicating in the introduction *which* \textsf{R} packages we take up in Section 6.
 We believe that the discussion itself should remain at the end of the paper, *after* we have described the facilities for cross-validation in the **cv** package.

### 2025-03-08 scratch/JSS-submission/Resubmission/Fox-Monette-cv-package.Rmd
fix typo etc in JSS paper

@@
 
-* \code{data}, which can usually be inferred from the \code{model} object, using the \code{get_data()` function from the \pkg{insight} package [@LudeckeWaggonerMakowski:2019].
+* \code{data}, the data to which the model was fit, and which can usually be inferred from the \code{model} object, using the \code{get_data()} function from the \pkg{insight} package [@LudeckeWaggonerMakowski:2019].
 The \pkg{insight} package provides a robust way to access model components for many classes of statistical models.

### 2025-05-05 scratch/JSS-submission/Final_submission/jss5535.Rmd
create Final_submission/add files

--- 2025-03-08 scratch/JSS-submission/Resubmission/Fox-Monette-cv-package.Rmd
@@
     extra_dependencies: "subfig"
-bibliography: ["cv.bib"]
+bibliography: ["jss5535.bib"]
 ---
@@
 
-* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} (the default) selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than directly recomputing the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).
+* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} (the default) selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than directly recomputing the least-squares fit (see below and Appendix A.1).
 

### 2025-05-05 scratch/JSS-submission/Final_submission/jss5535.tex
create Final_submission/add files

--- 2024-07-03 scratch/notes_john/JSS-article/JSS-article-3.tex
@@
 package by writing methods for additional classes of models. The
-\code{cv()} function also can cross-validate complex model-selection
+\code{cv()} function can also cross-validate complex model-selection
 procedures, such as those that include variable transformations,
@@
     Toronto, Ontario, Canada\\
-
+  E-mail: \email\\
+  URL: \url{http://blackwell.math.yorku.ca/gmonette}\\~\\
   }
@@
 
-\usepackage{amsmath}
+\usepackage{amsmath} \usepackage{amssymb}
 
@@
 common scientific norm. Emulating this process in a single study by
-data-division is less common: The data are randomly divided into two,
-possibly equal-size, parts; the first part is used to develop and fit a
-statistical model; and then the second part is used to assess the
-adequacy of the model fit to the first part of the data. Data-division,
-however, suffers from two problems: (1) Dividing the data decreases the
-sample size and thus increases sampling error; and (2), even more
-disconcertingly, the results can vary substantially based on the random
-division of the data, particularly in smaller samples. See \citet[Sec.
-5.3]{Harrell:2015} for this and other remarks about data-division and
-cross-validation.
+data-division is less common \citep[see, e.g.,][]{Barnard:1974}: The
+data are randomly divided into two, possibly equal-size, parts; the
+first part is used to develop and fit (in the language of machine
+leaning, ``train'') a statistical model; and then the second part is
+used to assess the adequacy of (``validate'') the model fit to the first
+part of the data. Data-division, however, suffers from two problems: (1)
+Dividing the data decreases the sample size and thus increases sampling
+error; and (2), even more disconcertingly, the results can vary
+substantially based on the random division of the data, particularly in
+smaller samples. See \citet[Sec. 5.3]{Harrell:2015} for cogent remarks
+about data-division and cross-validation.
 
@@
 (``MSE'') of prediction, is then computed using these predicted values.
-In the extreme \(k = n\), the number of cases in the data, thus omitting
-individual cases and refitting the model \(n\) times---a procedure
-termed ``leave-one-out (LOO) cross-validation.''
+In the extreme, \(k\) equals \(n\), the number of cases in the data,
+thus omitting individual cases and refitting the model \(n\) times---a
+procedure termed ``leave-one-out (LOO) cross-validation.''
 
 Because the \(n\) models are each fit to \(n - 1\) cases, LOO CV
-produces a nearly unbiased estimate of prediction error. The \(n\)
-regression models are highly statistically dependent, however, based as
-they are on nearly the same data, and so the resulting estimate of
-prediction error has larger variance than if the predictions from the
-models fit to the \(n\) data sets were independent.
+produces a nearly unbiased estimate of the prediction error compared
+with a model fit to \(n\) cases. The \(n\) regression models are highly
+statistically dependent, however, based as they are on nearly the same
+data, and so the resulting estimate of prediction error has larger
+variance than if the predictions from the models fit to the \(n\) data
+sets were independent.
 
@@
 estimated prediction error for LOO CV. It is possible, however, to
-correct \(k\)-fold CV for bias (see Section
-\ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+correct \(k\)-fold CV for bias (see Appendix A.2).
 
@@
 The \pkg{cv} package for \proglang{R} automates the process of
-cross-validation for standard \proglang{R} statistical model objects.
-The principal function in the packages, also named \code{cv()}, has
-methods for objects produced by a number of commonly employed
+cross-validation for standard \proglang{R} statistical models. The
+principal function in the package, also named \code{cv()}, has methods
+for objects produced by a number of commonly employed
 regression-modeling functions, including those for mixed-effects models:
@@
 \begin{CodeInput}
-R> library("cv", quietly=TRUE)
+R> library("cv", quietly = TRUE)
 R> methods("cv")
@@
 
+The \code{cv()} function is introduced in the context of a preliminary
+example in Section \ref{preliminary-example-polynomial-regression} of
+the paper.
+
 \begin{itemize}
@@
   competing models, not necessarily of the same class, using the same
-  division of the data into folds. The \code{cv()} function is
-  introduced in the context of a preliminary example in Section
-  \ref{preliminary-example-polynomial-regression} of the paper.
+  division of the data into folds.
 \item
@@
   The \code{"function"} method for \code{cv()}, discussed in Section
-  \ref{cross-validating-model-specification}, cross-validates a complex
-  model-specification process that may, for example, involve choice of
-  data transformations and predictors, or model selection via CV itself.
+  \ref{cross-validating-model-specification}, cross-validates a
+  potentially complex model-specification process that may, for example,
+  involve choice of data transformations and predictors, or model
+  selection via CV itself.
 \item
@@
 A number of existing \proglang{R} packages include functions for
-cross-validating regression models. We briefly situate the \pkg{cv}
-package relative to other \proglang{R} software for cross-validation in
-Section \ref{comparing-cv-to-other-r-software-for-cross-validation}.
-
-The final section of the paper (Section \ref{computational-notes})
-describes some computational details, including efficient CV
-computations for linear and generalized-linear models and computation of
-bias-corrected CV criteria.
+cross-validating regression models. In Section
+\ref{comparing-cv-to-other-software-for-cross-validation}, we briefly
+situate the \pkg{cv} package relative to other \proglang{R} software for
+cross-validation (directly comparing \pkg{cv} to the \pkg{rsample} and
+\pkg{caret} packages), and to other statistical software (in particular,
+\proglang{SAS}, \proglang{Stata}, and \proglang{Python}).
+
+An appendix to the paper describes some computational details, including
+efficient CV computations for linear and generalized-linear models, and
+computation of bias-corrected CV criteria.
 
@@
 computations in parallel and can independently replicate a
-cross-validation procedure several times. There are also data management
+cross-validation procedure several times. There are also data-management
 facilities in the package, such as coercing the objects produced by the
@@
 other features not discussed in this paper are taken up in the vignettes
-distributed with the package, which also provides greater detail on some
-of topics that we do describe, such as extensions to the package.
-
-\section{Preliminary Example: Polynomial
+distributed with the package, which also provide greater detail on some
+topics that we do describe, such as extensions to the package.
+
+\section{Preliminary example: Polynomial
 regression}\label{preliminary-example-polynomial-regression}
@@
 original source \citep[ Secs. 5.1, 5.3]{JamesEtAl:2021}, and it
-demonstrates the use of the \texttt{cv()} function.\footnote{\citet{JamesEtAl:2021}
-  use the \texttt{cv.glm()} function in the \pkg{boot} package
+demonstrates the use of the \code{cv()} function.\footnote{\citet{JamesEtAl:2021}
+  use the \code{cv.glm()} function in the \pkg{boot} package
   \citep{CantyRipley2022, DavisonHinkley:1997}. Despite its name,
-  \texttt{cv.glm()} is an independent function and not a method of a
-  \texttt{cv()} generic function. The \pkg{boot} package is part of the
+  \code{cv.glm()} is an independent function and not a method of a
+  \code{cv()} generic function. The \pkg{boot} package is part of the
   standard \proglang{R} distribution.}
 
-The \code{Auto} dataset contains information about 392 cars:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> data("Auto", package="ISLR2")
+The \code{Auto} data set contains information about 392 cars:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> data("Auto", package = "ISLR2")
 R> summary(Auto)
@@
       mpg         cylinders     displacement   horsepower        weight
- Min.   : 9.00   Min.   :3.000   Min.   : 68.0   Min.   : 46.0   Min.   :1613  
- 1st Qu.:17.00   1st Qu.:4.000   1st Qu.:105.0   1st Qu.: 75.0   1st Qu.:2225  
- Median :22.75   Median :4.000   Median :151.0   Median : 93.5   Median :2804  
- Mean   :23.45   Mean   :5.472   Mean   :194.4   Mean   :104.5   Mean   :2978  
- 3rd Qu.:29.00   3rd Qu.:8.000   3rd Qu.:275.8   3rd Qu.:126.0   3rd Qu.:3615  
- Max.   :46.60   Max.   :8.000   Max.   :455.0   Max.   :230.0   Max.   :5140
+ Min.   : 9.0   Min.   :3.00   Min.   : 68   Min.   : 46.0   Min.   :1613
+ 1st Qu.:17.0   1st Qu.:4.00   1st Qu.:105   1st Qu.: 75.0   1st Qu.:2225
+ Median :22.8   Median :4.00   Median :151   Median : 93.5   Median :2804
+ Mean   :23.4   Mean   :5.47   Mean   :194   Mean   :104.5   Mean   :2978
+ 3rd Qu.:29.0   3rd Qu.:8.00   3rd Qu.:276   3rd Qu.:126.0   3rd Qu.:3615
+ Max.   :46.6   Max.   :8.00   Max.   :455   Max.   :230.0   Max.   :5140
 
   acceleration       year        origin                     name
- Min.   : 8.00   Min.   :70.00   Min.   :1.000   amc matador       :  5  
- 1st Qu.:13.78   1st Qu.:73.00   1st Qu.:1.000   ford pinto        :  5  
- Median :15.50   Median :76.00   Median :1.000   toyota corolla    :  5  
- Mean   :15.54   Mean   :75.98   Mean   :1.577   amc gremlin       :  4  
- 3rd Qu.:17.02   3rd Qu.:79.00   3rd Qu.:2.000   amc hornet        :  4  
- Max.   :24.80   Max.   :82.00   Max.   :3.000   chevrolet chevette:  4  
+ Min.   : 8.0   Min.   :70   Min.   :1.00   amc matador       :  5
+ 1st Qu.:13.8   1st Qu.:73   1st Qu.:1.00   ford pinto        :  5
+ Median :15.5   Median :76   Median :1.00   toyota corolla    :  5
+ Mean   :15.5   Mean   :76   Mean   :1.58   amc gremlin       :  4
+ 3rd Qu.:17.0   3rd Qu.:79   3rd Qu.:2.00   amc hornet        :  4
+ Max.   :24.8   Max.   :82   Max.   :3.00   chevrolet chevette:  4
                                             (Other)           :365
@@
 
-With the exception of \texttt{origin} (which we don't use here), these
+With the exception of \code{origin} (which we don't use here), these
 variables are largely self-explanatory, except possibly for units of
-measurement: for details see \texttt{help("Auto",\ package="ISLR2")}.
+measurement: For details see \code{help("Auto",}
+\code{package = "ISLR2")}.
 
@@
 
-{\centering \includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}
+{\centering \subfloat[\label{fig:mpg-horsepower-polynomials-1}]{\includegraphics[width=0.45\linewidth]{Figures/mpg-horsepower-polynomials-1} }~~~\subfloat[\label{fig:mpg-horsepower-polynomials-2}]{\includegraphics[width=0.45\linewidth]{Figures/mpg-horsepower-polynomials-2} }
 
@@
 
-\caption{\code{mpg} vs. \code{horsepower} for the \code{Auto} data, showing fitted polynomials of degree 1 through 5.}\label{fig:mpg-horsepower-scatterplot-polynomials}
+\caption{(a) \code{mpg} vs.
+\code{horsepower} for the \code{Auto} data, showing fitted polynomials of degree 1 through 5.
+(b) Estimated squared error as a function of polynomial degree, $p = 1, \ldots, 10$.}\label{fig:mpg-horsepower-polynomials}
 \end{figure}
@@
 
-We'll focus here on the relationship of \code{mpg} (miles per gallon) to
-\code{horsepower}, as displayed in Figure
-\ref{fig:mpg-horsepower-scatterplot-polynomials}. The relationship
+We'll focus on the relationship of \code{mpg} (miles per gallon) to
+\code{horsepower}, as displayed in
+Figure~\ref{fig:mpg-horsepower-polynomials} (a). The relationship
 between the two variables is monotone, decreasing, and nonlinear.
@@
   it serves to illustrate the use of CV, a polynomial is not the best
-  choice here. Consider, for example the scatterplot for log-transformed
-  \texttt{mpg} and \texttt{horsepower}, produced by
-  \texttt{plot(mpg\ \textasciitilde{}\ horsepower,\ data=Auto,\ log="xy")}
-  (execution of which is left to the reader). We revisit the \code{Auto}
-  data in Section \ref{cross-validating-model-specification}.}
-Polynomial fits for \(p = 1\) to \(5\) are shown in Figure
-\ref{fig:mpg-horsepower-scatterplot-polynomials}. The linear fit is
-clearly inappropriate; the fits for \(p = 2\) (quadratic) through \(4\)
-are very similar; and the fit for \(p = 5\) may over-fit the data by
-chasing one or two relatively high \texttt{mpg} values at the right (but
-see the CV results reported below).
-
-Figure \ref{fig:mpg-horsepower-MSE-se} shows two measures of estimated
-(squared) error as a function of polynomial-regression degree: The
-mean-squared error (``MSE''), defined as
+  choice here. Consider, for example, the scatterplot for
+  log-transformed \code{mpg} and \code{horsepower}, produced by
+  \code{plot(mpg $\sim$ horsepower, data=Auto, log="xy")} (execution of
+  which is left to the reader). We revisit the \code{Auto} data in
+  Section \ref{cross-validating-model-specification}.} Polynomial fits
+for \(p = 1\) to \(5\) are shown in Figure
+\ref{fig:mpg-horsepower-polynomials} (a). The linear fit is clearly
+inappropriate; the fits for \(p = 2\) (quadratic) through \(4\) are very
+similar; and the degree-5 polynomial may over-fit the data by chasing
+one or two relatively high \code{mpg} values at the right (but see the
+CV results reported below).
+
+Figure \ref{fig:mpg-horsepower-polynomials} (b) shows two measures of
+estimated (squared) error as a function of polynomial-regression degree:
+The mean-squared error (``MSE''), defined as
 \(\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\), and
@@
 
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-MSE-se-1}
-
-}
-
-\caption[Estimated squared error as a function of polynomial degree, $p$]{Estimated squared error as a function of polynomial degree, $p$}\label{fig:mpg-horsepower-MSE-se}
-\end{figure}
-\end{CodeChunk}
-
-The generic \texttt{cv()} function has an \texttt{"lm"} method,
+The generic \code{cv()} function has an \code{"lm"} method,
 
@@
 \item
-  \code{data}, which can usually be inferred from the \code{model}
-  object.
+  \code{data}, the data to which the model was fit, and which can
+  usually be inferred from the \code{model} object, using the
+  \code{get_data()} function from the \pkg{insight} package
+  \citep{LudeckeWaggonerMakowski:2019}. The \pkg{insight} package
+  provides a robust way to access model components for many classes of
+  statistical models.
 \item
@@
   cases; a confidence interval is computed only if the CV criterion can
-  be expressed as the average of casewise components (see Section
-  \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}
+  be expressed as the average of casewise components (see Appendix A.2
   for details).
@@
   \code{"naive"} updates the model using the \code{update()} function;
-  and \code{"auto"} selects \code{"hatvalues"} for LOO CV and
-  \code{"Woodbury"} for \(k\)-fold CV, both of which are much more
-  efficient than literally updating the least-squares fit (see below and
-  Section
-  \ref{efficient-computations-for-linear-and-generalized-linear-models}).
+  and \code{"auto"} (the default) selects \code{"hatvalues"} for LOO CV
+  and \code{"Woodbury"} for \(k\)-fold CV, both of which are much more
+  efficient than directly recomputing the least-squares fit (see below
+  and Appendix A.1).
 \item
   \code{ncores}, the number of cores to employ for parallel computation;
-  if \code{cores = 1} (the default), the computations are not
+  if \code{ncores = 1} (the default), the computations are not
   parallelized.
@@
 To illustrate, we perform 10-fold CV for a quadratic polynomial fit to
-the \code{Auto} data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
-R> summary(m.auto)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-lm(formula = mpg ~ poly(horsepower, 2), data = Auto)
-
-Residuals:
-     Min       1Q   Median       3Q      Max 
--14.7135  -2.5943  -0.0859   2.2868  15.8961
-
-Coefficients:
-                      Estimate Std. Error t value Pr(>|t|)
-(Intercept)            23.4459     0.2209  106.13   <2e-16
-poly(horsepower, 2)1 -120.1377     4.3739  -27.47   <2e-16
-poly(horsepower, 2)2   44.0895     4.3739   10.08   <2e-16
-
-Residual standard error: 4.374 on 389 degrees of freedom
-Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
-F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.auto, confint=TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 431467
-\end{CodeOutput}
+the \code{Auto} data:\footnote{We load the \pkg{car} package
+  \citep{FoxWeisberg:2019} here for the \code{brief()} function and to
+  use below.}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library("car", quietly = TRUE)
+R>
+R> m.auto <- lm(mpg ~ poly(horsepower, 2), data = Auto)
+R> brief(m.auto)
+\end{CodeInput}
+\begin{CodeOutput}
+           (Intercept) poly(horsepower, 2)1 poly(horsepower, 2)2
+Estimate        23.446              -120.14                44.09
+Std. Error       0.221                 4.37                 4.37
+
+ Residual SD = 4.37 on 389 df, R-squared = 0.688
+\end{CodeOutput}
+\begin{CodeInput}
+R> (cv.auto <- cv(m.auto, confint = TRUE))
+\end{CodeInput}
+\begin{CodeOutput}
+R RNG seed set to 623225
+\end{CodeOutput}
+\begin{CodeOutput}
+cross-validation criterion (mse) = 19.409
+\end{CodeOutput}
+\begin{CodeInput}
+R> summary(cv.auto)
+\end{CodeInput}
 \begin{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 19.192
-bias-adjusted cross-validation criterion = 19.18107
-95% CI for bias-adjusted CV criterion = (15.71858, 22.64356)
-full-sample criterion = 18.98477 
-\end{CodeOutput}
-\end{CodeChunk}
-
-The function reports the CV estimate of MSE, a bias-adjusted estimate of
-the MSE (the bias adjustment is explained in Section
-\ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}),
-and the MSE is also computed for the original, full-sample regression.
-Because the number of cases \(n = 392 < 400\) for the \code{Auto} data,
-we set the argument \code{confint=TRUE} to obtain a confidence interval
-for the MSE, which proves to be quite wide.
+cross-validation criterion = 19.409
+bias-adjusted cross-validation criterion = 19.386
+95% CI for bias-adjusted CV criterion = (15.885, 22.887)
+full-sample criterion = 18.985
+\end{CodeOutput}
+\end{CodeChunk}
+
+The \texttt{print()} method for \texttt{"cv"} objects reports the CV
+estimate of MSE; the \texttt{summary()} method, which we'll use
+extensively in the sequel, in addition reports a bias-adjusted estimate
+of the MSE (the bias adjustment is explained in Appendix A.2), and the
+MSE is also computed for the original, full-sample regression. Because
+the number of cases \(n = 392 < 400\) for the \code{Auto} data, we set
+the argument \code{confint = TRUE} to obtain a confidence interval for
+the MSE, which proves to be quite wide.
 
@@
 \begin{CodeInput}
-R> cv(m.auto, k="loo")
+R> summary(cv(m.auto, k = "loo"))
 \end{CodeInput}
@@
 criterion: mse
-cross-validation criterion = 19.24821
-\end{CodeOutput}
-\end{CodeChunk}
-
-The \code{"hatvalues"} method reports only the CV estimate of MSE.
+cross-validation criterion = 19.248
+\end{CodeOutput}
+\end{CodeChunk}
+
+The \code{"hatvalues"} method computes only the CV estimate of MSE.
 Alternative methods are to use the Woodbury matrix identity or the
@@
 omitted. All three methods produce exact results for a linear model
-(within the precision of floating-point computations):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(m.auto, k="loo", method="naive", confint=TRUE)
+(within the precision of floating-point computations); for example:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> summary(cv(m.auto, k = "loo", method = "naive", confint = TRUE))
 \end{CodeInput}
@@
 criterion: mse
-cross-validation criterion = 19.24821
-bias-adjusted cross-validation criterion = 19.24787
-95% CI for bias-adjusted CV criterion = (15.77884, 22.71691)
-full-sample criterion = 18.98477 
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(m.auto, k="loo", method="Woodbury", confint=TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-n-Fold Cross Validation
-method: Woodbury
-criterion: mse
-cross-validation criterion = 19.24821
-bias-adjusted cross-validation criterion = 19.24787
-95% CI for bias-adjusted CV criterion = (15.77884, 22.71691)
-full-sample criterion = 18.98477 
+cross-validation criterion = 19.248
+bias-adjusted cross-validation criterion = 19.248
+95% CI for bias-adjusted CV criterion = (15.779, 22.717)
+full-sample criterion = 18.985
 \end{CodeOutput}
@@
 takes the naive approach, and for which we have to fit the linear model
-as an equivalent Gaussian GLM. We use the \code{microbenchmark()}
-function from the package of the same name \citep{Mersmann:2023} for the
-timings:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
-R> boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 19.24821 19.24787
-\end{CodeOutput}
-\begin{CodeInput}
-R> microbenchmark::microbenchmark(
-+   hatvalues = cv(m.auto, k="loo"),
-+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
-+   naive = cv(m.auto, k="loo", method="naive"),
+as an equivalent Gaussian GLM; the output shown is for the CV estimate
+of MSE and the bias-corrected CV estimate. We use the
+\code{microbenchmark()} function from the package of the same name
+\citep{Mersmann:2023} for the timings:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data = Auto)
+R> boot::cv.glm(Auto, m.auto.glm)$delta
+\end{CodeInput}
+\begin{CodeOutput}
+[1] 19.248 19.248
+\end{CodeOutput}
+\begin{CodeInput}
+R> set.seed(19412)
+R> print(microbenchmark::microbenchmark(
++   hatvalues = cv(m.auto, k = "loo"),
++   Woodbury = cv(m.auto, k = "loo", method = "Woodbury"),
++   naive = cv(m.auto, k = "loo", method = "naive"),
 +   cv.glm = boot::cv.glm(Auto, m.auto.glm),
-+   times=10
-+ )
++   times = 10, unit = "relative"), signif = 3)
 \end{CodeInput}
@@
 \begin{CodeOutput}
-Unit: microseconds
-      expr        min         lq       mean     median         uq        max
- hatvalues    996.956   1181.292   1496.971   1216.777   1233.157   4243.664
-  Woodbury  12189.054  12270.726  13187.437  12580.645  12664.121  16369.455
-     naive 216220.101 218010.571 231464.303 220980.816 223550.942 280828.393
-    cv.glm 381634.027 383670.456 404475.689 390533.979 441471.559 442930.011
- neval
-    10
-    10
-    10
-    10
+Unit: relative
+      expr    min     lq   mean median     uq   max neval cld
+ hatvalues   1.00   1.00   1.00   1.00   1.00  1.00    10 a
+  Woodbury   7.47   7.15   5.77   6.74   6.62  2.35    10 a
+     naive 124.00 119.00  98.20 112.00 111.00 47.50    10  b
+    cv.glm 217.00 207.00 172.00 197.00 198.00 75.20    10   c
 \end{CodeOutput}
@@
 faster than employing Woodbury matrix updates, and more than two orders
-of magnitude faster than refitting the model.
+of magnitude faster than refitting the model.\footnote{Out of
+  impatience, we asked \code{microbenchmark()} to execute each command
+  only 10 times rather than the default 100. With the exception of the
+  last column, the output is self-explanatory. The last column shows
+  which methods have average timings that are statistically
+  distinguishable (as indicated by different letters). Because of the
+  small number of repetitions (i.e., 10), the \code{"hatvalues"} and
+  \code{"Woodbury"} methods aren't distinguishable, but the difference
+  between these methods persists when we perform more repetitions---we
+  invite the patient reader to redo this computation with the default
+  \code{times = 100} repetitions.}
 
@@
 \begin{CodeInput}
-R> mlist <- vector(10, mode="list")
+R> mlist <- vector(10, mode = "list")
 R> for (p in 1:10) mlist[[p]] <- lm(mpg ~ poly(horsepower, p), data = Auto)
 R> names(mlist) <- paste0("m.", 1:10)
-R> mlist[2] # e.g., the quadratic fit
+R> mlist[2]
 \end{CodeInput}
@@
          (Intercept)  poly(horsepower, p)1  poly(horsepower, p)2
-               23.45               -120.14                 44.09  
-\end{CodeOutput}
-\end{CodeChunk}
-
-We then apply \code{cv()} to the list of 10 models (the \code{data}
-argument is required):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # 10-fold CV
-R> mlist <- models(mlist) # create "modList" object
+                23.4                -120.1                  44.1
+\end{CodeOutput}
+\end{CodeChunk}
+
+where, for example, \code{mlist[2]} is the quadratic fit.
+
+We then apply \code{cv()} to the list of 10 models for both 10-fold and
+LOO CV (the \code{data} argument to \code{cv()} is required):
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> mlist <- models(mlist)
 R>
-R> cv.auto.10 <- cv(mlist, data=Auto, seed=2120)
-R> cv.auto.10[2] # e.g., for quadratic model
-\end{CodeInput}
-\begin{CodeOutput}
-
+R> cv.auto.10 <- cv(mlist, data = Auto, seed = 2120)
+R> cv.auto.10[2]
+\end{CodeInput}
+\begin{CodeOutput}
 Model m.2:
-10-Fold Cross Validation
-method: Woodbury
-cross-validation criterion = 19.34601
-bias-adjusted cross-validation criterion = 19.32699
-full-sample criterion = 18.98477 
-\end{CodeOutput}
-\begin{CodeInput}
-R> # LOO CV
-R> cv.auto.loo <- cv(mlist, data=Auto, k="loo")
-R> cv.auto.loo[2] # e.g., for quadratic model
-\end{CodeInput}
-\begin{CodeOutput}
-
+cross-validation criterion = 19.346
+\end{CodeOutput}
+\begin{CodeInput}
+R> cv.auto.loo <- cv(mlist, data = Auto, k = "loo")
+R> cv.auto.loo[2]
+\end{CodeInput}
+\begin{CodeOutput}
 Model m.2:
-n-Fold Cross Validation
-method: hatvalues
-cross-validation criterion = 19.24821
+cross-validation criterion = 19.248
 \end{CodeOutput}
@@
 to \code{models()}. To visualize the results, comparing the models, we
-invoke the \code{plot()} method for the \code{"cvModList"} object
+invoke the \code{plot()} method for the \code{"cvModList"} objects
 returned by \code{cv()} (see Figure
-\ref{fig:polynomial-regression-CV-graph-2}):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",
-+      axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
-R> plot(cv.auto.loo, main="Polynomial Regressions, LOO CV",
-+      axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
+\ref{fig:polynomial-regression-CV-graph-2}):\footnote{The objects
+  \code{cv.auto.10} and \code{cv.auto.loo} do not include confidence
+  intervals for the CV MSE estimates because, as we explained, there are
+  fewer than 400 cases in the \code{Auto} data. Consequently, confidence
+  intervals are not shown in the graphs in Figure
+  \ref{fig:polynomial-regression-CV-graph-2}. We can force \texttt{cv()}
+  to compute confidence intervals by adding the argument
+  \texttt{confint\ =\ TRUE}; we invite the reader to do this and to
+  verify that the resulting confidence intervals are very broad, as was
+  the case for the quadratic model fit to the \code{Auto} data in the
+  preceding section.}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot(cv.auto.10, main = "Polynomial Regressions, 10-Fold CV",
++      axis.args = list(labels = 1:10), xlab = "Degree of Polynomial, p")
+R> plot(cv.auto.loo, main = "Polynomial Regressions, LOO CV",
++      axis.args = list(labels = 1:10), xlab = "Degree of Polynomial, p")
 \end{CodeInput}
@@
 
-{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/polynomial-regression-CV-graph-2-1} }\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/polynomial-regression-CV-graph-2-2} }
+{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{Figures/polynomial-regression-CV-graph-2-1} }~~~\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{Figures/polynomial-regression-CV-graph-2-2} }
 
@@
 
-\caption[Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$]{Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p$.}\label{fig:polynomial-regression-CV-graph-2}
+\caption[Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p = 1, 2, \ldots, 10$]{Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p = 1, 2, \ldots, 10$.}\label{fig:polynomial-regression-CV-graph-2}
 \end{figure}
@@
 error variance \(\widehat{\sigma}^2\) for each model (cf., Figure
-\ref{fig:mpg-horsepower-MSE-se} on page
-\pageref{fig:mpg-horsepower-MSE-se}), except for the highest-degree
+\ref{fig:mpg-horsepower-polynomials} (b) on page
+\pageref{fig:mpg-horsepower-polynomials}), except for the highest-degree
 polynomials, where the CV results more clearly suggest over-fitting.
+\newpage
 
@@
   to both nested and crossed random effects. We present an example of
-  the latter in a vignette for the \pkg{cv} package.}
+  crossed random effects in the \pkg{cv} package vignette on mixed
+  models.}
 
@@
 Following their use by \citet{RaudenbushBryk:2002}, data from the 1982
-\emph{High School and Beyond} (HSB) survey have become a staple of the
-literature on mixed-effects models. The HSB data are used by \citet[Sec.
-7.2.2]{FoxWeisberg:2019} to illustrate the application of linear mixed
-models to hierarchical data, and we'll closely follow their example
-here.
+\emph{High School and Beyond} (``HSB'') survey have become a staple of
+the literature on mixed-effects models. The HSB data are used by
+\citet[Sec. 7.2.2]{FoxWeisberg:2019} to illustrate the application of
+linear mixed models to hierarchical data, and we'll closely follow their
+example here.
 
@@
 \begin{CodeInput}
-R> data("MathAchieve", package="nlme")
+R> data("MathAchieve", package = "nlme")
 R> dim(MathAchieve)
@@
 \begin{CodeInput}
-R> data("MathAchSchool", package="nlme")
+R> data("MathAchSchool", package = "nlme")
 R> dim(MathAchSchool)
@@
 R> HSB <- merge(MathAchSchool[, c("School", "Sector")],
-+              HSB[, c("School", "SES", "MathAch")], by="School")
++              HSB[, c("School", "SES", "MathAch")], by = "School")
 R> names(HSB) <- tolower(names(HSB))
@@
 function in the \pkg{lme4} package \citep{BatesEtAl:2015} to fit a mixed
-model for math achievement to the HSB data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("lme4", quietly=TRUE)
+model for math achievement to the HSB data (summarizing the model with
+the \code{S()} function from the \pkg{car} package):
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library("lme4", quietly = TRUE)
 R> hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
-+                    + (cses | school), data=HSB)
-R> summary(hsb.lmer, correlation=FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
-Linear mixed model fit by REML ['lmerMod']
-Formula: mathach ~ mean.ses * cses + sector * cses + (cses | school)
-   Data: HSB
-
-REML criterion at convergence: 46503.7
-
-Scaled residuals: 
-     Min       1Q   Median       3Q      Max 
--3.15926 -0.72319  0.01704  0.75444  2.95822
-
-Random effects:
- Groups   Name        Variance Std.Dev. Corr
- school   (Intercept)  2.380   1.5426       
-          cses         0.101   0.3179   0.39
- Residual             36.721   6.0598       
++                    + (cses | school), data = HSB)
+R> S(hsb.lmer, brief = TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Estimates of Fixed Effects:
+                    Estimate Std. Error z value Pr(>|z|)
+(Intercept)           12.128      0.199   60.86  < 2e-16
+mean.ses               5.333      0.369   14.45  < 2e-16
+cses                   2.945      0.156   18.93  < 2e-16
+sectorCatholic         1.227      0.306    4.00  6.2e-05
+mean.ses:cses          1.039      0.299    3.48  0.00051
+cses:sectorCatholic   -1.643      0.240   -6.85  7.3e-12
+
+Estimates of Random Effects (Covariance Components):
+ Groups   Name        Std.Dev. Corr
+ school   (Intercept) 1.543
+          cses        0.318    0.39
+ Residual             6.060
+
 Number of obs: 7185, groups:  school, 160
 
-Fixed effects:
-                    Estimate Std. Error t value
-(Intercept)          12.1279     0.1993  60.856
-mean.ses              5.3329     0.3692  14.446
-cses                  2.9450     0.1556  18.928
-sectorCatholic        1.2266     0.3063   4.005
-mean.ses:cses         1.0393     0.2989   3.477
-cses:sectorCatholic  -1.6427     0.2398  -6.851
+logLik     df    AIC    BIC
+-23252     10  46524  46592
 \end{CodeOutput}
@@
 \begin{CodeInput}
-R> cv(hsb.lmer, k=10, clusterVariables="school", seed=5240)
+R> summary(cv(hsb.lmer, k = 10, clusterVariables = "school", seed = 5240))
 \end{CodeInput}
@@
 criterion: mse
-cross-validation criterion = 39.15662
-bias-adjusted cross-validation criterion = 39.14844
-95% CI for bias-adjusted CV criterion = (38.06554, 40.23135)
-full-sample criterion = 39.00599 
+cross-validation criterion = 39.157
+bias-adjusted cross-validation criterion = 39.148
+95% CI for bias-adjusted CV criterion = (38.066, 40.231)
+full-sample criterion = 39.006
 \end{CodeOutput}
@@
 \begin{CodeInput}
-R> cv(hsb.lmer, seed=1575)
+R> summary(cv(hsb.lmer, seed = 1575))
 \end{CodeInput}
@@
 Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv, :
-Model failed to converge with max|grad| = 0.00587131 (tol = 0.002, component 1)
+Model failed to converge with max|grad| = 0.00587207 (tol = 0.002, component 1)
 \end{CodeOutput}
@@
 criterion: mse
-cross-validation criterion = 37.44473
-bias-adjusted cross-validation criterion = 37.33801
-95% CI for bias-adjusted CV criterion = (36.28761, 38.38841)
-full-sample criterion = 36.06767 
+cross-validation criterion = 37.445
+bias-adjusted cross-validation criterion = 37.338
+95% CI for bias-adjusted CV criterion = (36.288, 38.388)
+full-sample criterion = 36.068
 \end{CodeOutput}
@@
 
-In this section, we introduce an artificial data set that exemplifies
+In this section, we introduce four artificial data sets that exemplify
 aspects of cross-validation particular to hierarchical models. Using
-this data set, we show that model comparisons employing cluster-based
+these data sets, we show that model comparisons employing cluster-based
 and those employing case-based cross-validation may not agree on a
@@
 
-Consider a researcher studying improvement in a skill, singing, for
-example, among students enrolled in a four-year voice program at a music
-conservatory. The plan is to measure each student's skill level at the
-beginning of the program and every year thereafter until the end of the
-program, resulting in 5 annual measurements for each student. It turns
-out that singing appeals to students of all ages, and students enrolling
-in the program range in age from 20 to 70. Moreover, participants'
-untrained singing skill is similar at all ages, as is their rate of
-progress with training. All students complete the four-year program.
-
-The researcher, who has more expertise in singing than in modeling,
-decides to model the response, \(y\), singing skill, as a function of
-age, \(x\), reasoning that students get older during their stay in the
-program, and (incorrectly) that age can serve as a proxy for elapsed
-time. The researcher knows that a mixed model should be used to account
-for clustering due to the expected similarity of measurements taken from
-each student.
-
-We start by generating the data, using parameters consistent with the
-description above and meant to highlight the issues that arise in
-cross-validating mixed-effects models:\footnote{We invite the interested
-  reader to experiment with varying the parameters of our example.}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> # Parameters:
-R> set.seed(9693) 
-R> Nb <- 100     # number of groups
-R> Nw <- 5       # number of individuals within groups
-R> Bb <- 0       # between-group regression coefficient on group mean
-R> SDre <- 2.0   # between-group SD of random level relative to group mean of x
-R> SDwithin <- 0.5  # within group SD
-R> Bw <- 1          # within group effect of x
-R> Ay <- 10         # intercept for response
-R> Ax <- 20         # starting level of x
-R> Nx <- Nw*10      # number of distinct x values
-R> 
-R> Data <- data.frame(
-+   group = factor(rep(1:Nb, each=Nw)),
-+   x = Ax + rep(1:Nx, length.out = Nw*Nb)
-+ ) |> within ({
-+   xm  <- ave(x, group, FUN = mean) # within-group mean
-+   y <- Ay +
-+     Bb * xm +                    # contextual effect
-+     Bw * (x - xm) +              # within-group effect
-+     rnorm(Nb, sd=SDre)[group] +  # random level by group
-+     rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
-+ })
-\end{CodeInput}
-\end{CodeChunk}
-
-Figure \ref{fig:plot1} (a) shows a scatterplot the data for a
-representative group of 10 (without loss of generality, the first 10) of
-the 100 students, displaying the 95\% concentration ellipse for each
-cluster.
+The four data sets differ in the magnitude of between-cluster variation
+compared to within-cluster variation. They serve to illustrate how
+fitting mixed models, and, consequently, the cross-validation of mixed
+models, is sensitive to relative variation, which affects the degree of
+shrinkage of within-cluster estimates of effects towards between-cluster
+estimates.
+
+The purpose of this section is to explain some of the complexities of
+cross-validating mixed-effects models, which motivates the approach that
+we take for mixed models in the \pkg{cv} package. The examples in this
+section introduce no features of the \pkg{cv} package that we haven't
+already encountered, and the generation of data for these examples
+involves unenlightening data management. Consequently, we don't show the
+code for data generation, model fitting, and graphs in this
+section.\footnote{The code for this section is in the complementary
+  materials for the paper. We used the \code{glmmTMB()} function in the
+  \pkg{glmmTMB} package \citep{BrooksEtAl} for the models fit here
+  because, in our experience, it is more likely to converge than
+  functions in the \pkg{nlme} and the \pkg{lme4} packages for models
+  with low between-cluster variation. We used the \pkg{lattice}
+  \citep{Sarkar:2008} and \pkg{latticeExtra} \citep{SarkarAndrews:2022}
+  packages for the graphs.}
+
+Consider a researcher studying the effect of the dosage \(x\) of a drug
+on the severity of symptoms \(y\) for a hypothetical disease. The
+researcher has longitudinal data on 20 patients, each of whom was
+observed on 5 occasions in which patients received different dosages of
+the drug. The data are observational, with dosages prescribed by the
+patients' physicians, so that patients who were more severely affected
+by the disease received higher dosages of the drug.
+
+Within patients, higher dosages are generally associated with a
+reduction in symptoms. Between patients, however, higher dosages are
+associated with higher levels of symptoms. A plausible mechanism is a
+reversal of causality: Within patients, higher dosages alleviate
+symptoms, but, between patients, higher morbidity causes the
+prescription of higher dosages.
+
+To generate the four data sets, we first constructed a template data set
+from a population with a between-patient effect of dosage of \(1.0\) and
+a within-patient effect of \(-0.5\). We then applied four different
+multipliers to the between-patient standard deviation from the
+regression line of patient centroids. The resulting data sets, shown in
+Figure \ref{fig:plot1}, differ in the between-patient variation of
+patient centroids from a common between-patient regression line,
+exhibiting a progression from low to high variation around the common
+regression line. Each patient has the same relative configuration of
+dosages and symptoms in each of the four data sets. To help visualize
+the structure of the data, estimated 50\% Gaussian concentration
+ellipses are shown for each patient.
 
@@
 
-{\centering \subfloat[\label{fig:plot1-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-1} }\subfloat[\label{fig:plot1-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-2} }\newline\subfloat[\label{fig:plot1-3}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-3} }\subfloat[\label{fig:plot1-4}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-4} }\newline\subfloat[\label{fig:plot1-5}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-5} }
+{\centering \includegraphics[width=0.75\linewidth]{Figures/plot1-1}
 
@@
 
-\caption[(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) predictions from several mixed models fit to the data]{(a) Hierarchical data set, showing the first 10 of 100 students, and (b)--(e) predictions from several mixed models fit to the data}\label{fig:plot1}
+\caption[Data sets showing identical within-patient configurations with increasing between-patient variation]{Data sets showing identical within-patient configurations with increasing between-patient variation.
+The labels above each panel show the ratio of between-patient SD to within-patient SD.
+The ellipses are estimated 50\% Gaussian concentration ellipses for each patient.
+The population between-patient regression line, $\E(y) = 10 + x$, is shown in each panel, where $y$ is symptoms and $x$ is dosage.}\label{fig:plot1}
 \end{figure}
@@
 
-The between-student effect of age is 0 but the within-student effect is
-1. Due to the large variation in ages between students, the
-least-squares regression of singing skill on age (for the 500
-observations among all 100 students) produces an estimated slope close
-to 0 (though with a small \(p\)-value), because the slope is heavily
-weighted toward the between-student effect:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> summary(lm(y ~ x, data=Data))
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-lm(formula = y ~ x, data = Data)
-
-Residuals:
-    Min      1Q  Median      3Q     Max 
--5.7713 -1.6583 -0.0894  1.5520  7.6240
-
-Coefficients:
-            Estimate Std. Error t value Pr(>|t|)
-(Intercept) 9.050430   0.347189  26.068  < 2e-16
-x           0.020908   0.007273   2.875  0.00422
-
-Residual standard error: 2.347 on 498 degrees of freedom
-Multiple R-squared:  0.01632,   Adjusted R-squared:  0.01435 
-F-statistic: 8.263 on 1 and 498 DF,  p-value: 0.004219
-\end{CodeOutput}
-\end{CodeChunk}
-
-We proceed to fit several mixed-effects models to the data, using the
-\texttt{compareCoefs()} function from the \pkg{car} package
-\citep{FoxWeisberg:2019} to display the fixed-effect estimates for these
-models. We discuss each of these models below.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> mod.0 <- lmer(y ~ 1 + (1 | group), Data)
-R> mod.1 <- lmer(y ~ x + (1 | group), Data)
-R> mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
-R> mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
-R> library("car", quietly=TRUE)
-R> compareCoefs(mod.0, mod.1, mod.2, mod.3)
-\end{CodeInput}
-\begin{CodeOutput}
-Calls:
-1: lmer(formula = y ~ 1 + (1 | group), data = Data)
-2: lmer(formula = y ~ x + (1 | group), data = Data)
-3: lmer(formula = y ~ x + xm + (1 | group), data = Data)
-4: lmer(formula = y ~ I(x - xm) + (1 | group), data = Data)
-
-            Model 1 Model 2 Model 3 Model 4
-(Intercept)  10.002 -33.919   9.479  10.002
-SE            0.186   1.564   0.617   0.186
-
-x                    0.9653  0.9915        
-SE                   0.0158  0.0160
-
-xm                          -0.9800        
-SE                           0.0206
-
-I(x - xm)                             0.992
-SE                                    0.016
-
-\end{CodeOutput}
-\end{CodeChunk}
-
-The initial mixed-effects model, \code{mod.0}, is a simple
-random-intercepts model. We obtain predictions from this model for the
-fixed effects alone, as would be used for cross-validation based on
-clusters (i.e., students), and for fixed and random effects---the
-BLUPs---as would be used for cross-validation based on cases (i.e.,
-occasions within students). Predictions from \code{mod.0} for the first
-10 students are shown in Figure \ref{fig:plot1} (b). The fixed-effect
-predictions for the various individuals are identical---the estimated
-fixed-effects intercept or estimated general mean of \(y\)---while the
-BLUPs are the sums of the fixed-effects intercept and the random
-intercepts, and are only slightly shrunken towards the general mean.
-Because in our artificial data there is no population relationship
-between age and skill, the fixed-effect-only predictions and the BLUPs
-are not very different.
-
-Our next model, \code{mod.1}, includes a fixed intercept and the fixed
-effect of \code{x}, along with a random intercept. Predictions from this
-model appear in Figure \ref{fig:plot1} (c). The BLUPs fit the observed
-data very closely, but predictions based on the fixed effects alone,
-with a common intercept and slope for all clusters, are very
-poor---indeed, much worse than the fixed-effects-only predictions based
-on the simpler random-intercept model, \code{mod.0}. We therefore
-anticipate (and show later in this section) that case-based
-cross-validation will prefer \code{mod.1} to \code{mod.0}, but that
-cluster-based cross-validation will prefer \code{mod.0} to \code{mod.1}.
-
-Our third model, \code{mod.2}, includes the ``contextual effect'' of
-\(x\)---that is, the cluster mean \code{xm}---along with \(x\) and the
-intercept in the fixed-effect part of the model, and a random intercept.
-This model is equivalent to fitting
-\code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that
-generated the data once the coefficient of the contextual predictor
-\code{xm} is set to 0 (as it is in \code{mod.3}, discussed below).
-
-Predictions from \code{mod.2} appear in Figure \ref{fig:plot1} (d).
-Depending on the estimated variance parameters of the model, a mixed
-model like \code{mod.2} will apply varying degrees of shrinkage to the
-random-intercept BLUPs that correspond to variation in the heights of
-the parallel fitted lines for the individual students. In our contrived
-data, \code{mod.2} applies little shrinkage, allowing substantial
-variability in the heights of the fitted lines, which closely approach
-the observed values for each student. The fit of the mixed model
-\code{mod.2} is consequently similar to that of a fixed-effects model
-with age and a categorical predictor for individual students (i.e.,
-treating students as a factor, and not shown here).
-
-The mixed model \code{mod.2} therefore fits the individual observations
-well, and we anticipate a favorable assessment using individual-based
-cross-validation. In contrast, the large variability in the BLUPs
-results in larger residuals for predictions based on fixed effects
-alone, and so we expect that cluster-based cross-validation won't show
-an advantage for \code{mod.2} compared to the smaller \code{mod.0},
-which includes only fixed and random intercepts.
-
-Had the mixed model applied considerable shrinkage, then neither
-cluster-based nor case-based cross-validation would show much
-improvement over the random-intercept-only model. In our experience, the
-degree of shrinkage does not vary smoothly as parameters are changed but
-tends to be ``all or nothing,'' and near the tipping point, the behavior
-of estimates can be affected considerably by the choice of algorithm
-used to fit the model.
-
-Finally, \code{mod.3} directly estimates the model used to generate the
-data. As mentioned, it is a constrained version of \code{mod.2}, with
-the coefficient of \code{xm} set to 0, and with \code{x} expressed as a
-deviation from the cluster mean \code{xm}. The predictions from
-\code{mod.3}, shown in Figure \ref{fig:plot1} (e), are therefore similar
-to those from \code{mod.2}.
+Using data like these, a researcher may attempt to obtain an estimate of
+the within-patient effect of dosage by fitting mixed models with a
+random intercept. We will illustrate how this approach results in
+estimates that are highly sensitive to the relative variation at the two
+levels of the mixed model by considering four models, all of which
+include a random intercept. The first three models are sequentially
+nested in their fixed effects: (1) \code{~ 1}, intercept only; (2)
+\code{~ 1 + x}, intercept and effect of \code{x}; and (3)
+\code{~ 1 + x + xm}, intercept, effect of \code{x}, and a contextual
+variable, \code{xm}, consisting of the within-patient mean of \code{x}.
+The final model, (4) \code{~ 1 + I(x - xm)}, uses an intercept and the
+centered-within-group variable \code{x - xm}. We thus fit four models to
+four data sets for a total of 16 models.
+
+We proceed to obtain predictions from each model (1) using only the
+fixed effects, as for cross-validation based on clusters (i.e.,
+patients), and (2) using both fixed and random effects---that is, the
+BLUPs---as for cross-validation based on cases (i.e., occasions within
+patients). The fixed- and random-effects predictions from these models
+are displayed in Figure \ref{fig:plot-fits-combined} along with a
+summary of the data using 50\% concentration ellipses (see the top row
+of the figure).
+
+\begin{CodeChunk}
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{Figures/plot-fits-combined-1}
+
+}
+
+\caption[Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient standard-deviation ratios]{Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient standard-deviation ratios.
+The top row shows summaries of the within-patient data using estimated 50\% concentration ellipses, along with the patient centroids.}\label{fig:plot-fits-combined}
+\end{figure}
+\end{CodeChunk}
+
+Data sets with relatively low between-patient variation result in strong
+shrinkage of fixed-effects predictions, and also of BLUPs, towards the
+between-patient relationship between \code{y} and \code{x} in the model
+with an intercept and \code{x} as fixed-effect predictors,
+\code{~ 1 + x}. The inclusion of the contextual variable \code{xm} in
+the model corrects this problem.
+
+Although the BLUPs fit the observed data more closely than predictions
+based on fixed effects alone, the slopes of within-patient BLUPs do not
+conform to the within-patient slopes for the \code{~ 1 + x} model in the
+two data sets with the smallest between-patient variation. For data with
+a small between-patient variation, fixed-effects predictions for the
+\code{~ 1 + x} model have a slope that is close to the between-patient
+slope but provide better overall predictions than the fixed-effect
+predictions for data sets with larger between-subject variation. With
+data whose between-patient variation is relatively large, predictions
+based on the model with a common intercept and slope for all clusters
+are very poor---indeed, much worse than the fixed-effects-only
+predictions based on the simpler random-intercept model.
+
+We therefore anticipate (and show later in this section) that case-based
+cross-validation may prefer the intercept-only model, \code{~ 1} to the
+larger \code{~ 1 + x} model when the between-cluster variation is
+relatively small, but that cluster-based cross-validation will prefer
+the latter to the former. We will discover that case-based
+cross-validation prefers the \code{~ 1 + x} model to the \code{~ 1}
+model for the ``5 / 2.5'' data set, but cluster-based cross-validation
+prefers the latter model to the former. The situation is entirely
+reversed with the ``8 / 2.5'' data set.
+
+The third model, \code{~ 1 + x + xm}, includes a contextual effect of
+\code{x}---that is, the cluster mean \code{xm}---along with \code{x},
+the intercept in the fixed-effect part of the model, and a random
+intercept. This model is equivalent to fitting
+\code{y ~ I(x - xm) + xm + (1 | patient)}, which is the model that
+generated the data. The fit of the mixed model \code{~ 1 + x + xm} is
+consequently similar to that of a fixed-effects-only model with \code{x}
+and a categorical predictor for individual patients (i.e.,
+\code{y ~ factor(patient) + x}, treating patients as a factor, and not
+shown here).
 
 We next carry out case-based cross-validation, which, as we have
-explained, includes both fixed and predicted random effects (i.e.,
-BLUPs), and cluster-based cross-validation, which includes fixed effects
-only. In order to reduce between-model random variability in comparisons
-of models, we apply \code{cv()} to the list of models created by the
-\code{models()} function (introduced previously), performing
-cross-validation with the same folds for each model (see Figure
-\ref{fig:cross-validation-clusters}):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> modlist <- models("~ 1"=mod.0, "~ 1 + x"=mod.1, 
-+                   "~ 1 + x + xm"=mod.2, "~ 1 + I(x - xm)"=mod.3)
-R> 
-R> cvs_clusters <- cv(modlist, data=Data, cluster="group", k=10, seed=6449)
-R> plot(cvs_clusters, main="Model Comparison, Cluster-Based CV")
-R> 
-R> cvs_cases <- cv(modlist, data=Data, seed=9693)
-R> plot(cvs_cases, main="Model Comparison, Case-Based CV")
-\end{CodeInput}
+explained, is based on both fixed and predicted random effects (i.e.,
+BLUPs), and cluster-based cross-validation, which is based on fixed
+effects only.\footnote{In order to reduce between-model random
+  variability in comparisons of models on the same data set, we applied
+  \code{cv()} to a list of models created by the \code{models()}
+  function (introduced previously), performing cross-validation with the
+  same folds for each model.} The results for all data sets, using both
+cluster-based and case-based cross-validation, are assembled in Figure
+\ref{fig:cross-validation-data-plot}.
+
+\begin{CodeChunk}
 \begin{figure}
 
-{\centering \subfloat[\label{fig:cross-validation-clusters-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/cross-validation-clusters-1} }\subfloat[\label{fig:cross-validation-clusters-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/cross-validation-clusters-2} }
+{\centering \includegraphics[width=0.75\linewidth]{Figures/cross-validation-data-plot-1}
 
@@
 
-\caption[10-fold (a) cluster-based and (b) case-based cross-validation comparing random intercept models with varying fixed effects]{10-fold (a) cluster-based and (b) case-based cross-validation comparing random intercept models with varying fixed effects.
-The error bars show the 95\% confidence interval around the CV estimate of the MSE for each model.}\label{fig:cross-validation-clusters}
+\caption[10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects]{10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects.}\label{fig:cross-validation-data-plot}
 \end{figure}
@@
 
-In summary, \code{mod.1}, with \(x\) alone and without the contextual
-mean of \(x\), is assessed as fitting very poorly by cluster-based CV,
-but relatively much better by case-based CV. \code{mod.2}, which
-includes both \(x\) and its contextual mean, produces better results
-using both cluster-based and case-based CV. The data-generating model,
-\code{mod.3}, which includes the fixed effect of \code{x - xm} in place
-of separate terms in \code{x} and \code{xm}, isn't distinguishable from
-\code{mod.2}, which includes \code{x} and \code{xm} separately, even
-though \code{mod.2} has an unnecessary parameter (recall that the
-population coefficient of \code{xm} is 0 when \code{x} is expressed as
-deviations from the contextual mean). These conclusions are consistent
-with our observations based on graphing predictions from the various
-models in Figure \ref{fig:plot1} (on page \pageref{fig:plot1}), and they
-illustrate the desirability of assessing mixed-effect models at
-different hierarchical levels.
+In summary, when between-cluster variation is relatively large, the
+model \code{~ 1 + x}, with \code{x} alone and without the contextual
+mean \code{xm} of \code{x}, is assessed as fitting very poorly by
+cluster-based CV, but relatively much better by case-based CV. In all of
+our examples, the model \code{~ 1 + x + xm}, which includes both
+\code{x} and its contextual mean, produces better results using both
+cluster-based and case-based CV. These conclusions are consistent with
+our observations based on graphing predictions from the various models
+(in Figure \ref{fig:plot-fits-combined} on page
+\pageref{fig:plot-fits-combined}), and they illustrate the desirability
+of assessing mixed-effect models at different hierarchical levels.
 
@@
 of the data that remains when each fold is removed. Statistical modeling
-is partly a craft, and one could imagine applying that craft to
-successive partial data sets, each with a fold removed. The resulting
-procedure would be tedious, though possibly worth the effort, but it
-would also be difficult to realize in practice: After all, we can hardly
-erase our memory of statistical modeling choices between analyzing
-partial data sets. Alternatively, if we're able to automate the process
-of model specification, then we can more realistically apply CV
-mechanically.
+is partly a craft, and one could imagine applying that craft
+independently to successive partial data sets, each with a fold removed.
+The resulting procedure would be tedious, though possibly worth the
+effort, but it would also be difficult to realize in practice: After
+all, we can hardly erase our memory of statistical modeling choices
+between analyzing partial data sets. Alternatively, if we're able to
+automate the process of model specification, then we can more
+realistically apply CV mechanically.
 
@@
 predictor selection for regression models; \code{selectTrans()}, based
-on the \code{powerTransform()} function in the \pkg{car} package
-\citep{FoxWeisberg:2019}, transforms predictors and the response in a
-regression model towards normality; \code{selectTransAndStepAIC()}---the
-use of which we illustrate in the current section---performs both of
-these procedures sequentially; and \code{selectModelList()}---also
-illustrated in the current section---uses CV both to select one of
-several competing models, and then, recursively, to estimate prediction
-error for the selected model. In a vignette on extending the \pkg{cv}
-package, we explain how to specify additional model-selection
-procedures.
+on the \code{powerTransform()} function in the \pkg{car} package,
+transforms predictors and the response in a regression model towards
+normality; \code{selectTransStepAIC()}---the use of which we illustrate
+in the current section---performs both of these procedures sequentially;
+and \code{selectModelList()}---also illustrated in the current
+section---uses CV both to select one of several competing models, and
+then, recursively, to estimate prediction error for the selected model,
+a process that we term ``meta cross-validation.''\footnote{In a vignette
+  on extending the \pkg{cv} package, we explain how to specify
+  additional model-selection procedures.}
 
@@
 \begin{CodeInput}
-R> xtabs(~ year, data=Auto)
+R> xtabs(~ year, data = Auto)
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> xtabs(~ origin, data=Auto)
+R> xtabs(~ origin, data = Auto)
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> xtabs(~ cylinders, data=Auto)
+R> xtabs(~ cylinders, data = Auto)
 \end{CodeInput}
@@
 examine the pairwise relationships among the numeric variables in the
-data set (Figure \ref{fig:Auto-explore}, produced by the
-\code{scatterplotMatrix()} function in the \pkg{car} package):
+data set (Figure \ref{fig:Auto-explore-scatterplot-matrix}, produced by
+the \code{scatterplotMatrix()} function in the \pkg{car} package); in
+each panel, the solid line shows the linear least-squares fit and the
+broken line is for a nonparametric regression:
 
@@
 R> Auto$cylinders <- factor(Auto$cylinders,
-+                          labels=c("3-4", "3-4", "5-6", "5-6", "8"))
++                          labels = c("3-4", "3-4", "5-6", "5-6", "8"))
 R> Auto$year <- as.factor(Auto$year)
 R> Auto$origin <- factor(Auto$origin,
-+                       labels=c("America", "Europe", "Japan"))
-R> rownames(Auto) <- make.names(Auto$name, unique=TRUE)
++                       labels = c("America", "Europe", "Japan"))
+R> rownames(Auto) <- make.names(Auto$name, unique = TRUE)
 R> Auto$name <- NULL
 R>
-R> scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
-+                   smooth=list(spread=FALSE), data=Auto, pch=".")
+R> scatterplotMatrix(~ mpg + displacement + horsepower + weight
++                   + acceleration,
++                   smooth = list(spread = FALSE), data = Auto, pch= ".")
 \end{CodeInput}
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-explore-1}
+{\centering \includegraphics[width=0.95\linewidth]{Figures/Auto-explore-scatterplot-matrix-1}
 
@@
 
-\caption[Scatterplot matrix for the numeric variables in the \code{Auto} data]{Scatterplot matrix for the numeric variables in the \code{Auto} data.
-In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.}\label{fig:Auto-explore}
+\caption[Scatterplot matrix for the numeric variables in the \code{Auto} data]{Scatterplot matrix for the numeric variables in the \code{Auto} data.}\label{fig:Auto-explore-scatterplot-matrix}
 \end{figure}
@@
 \code{cylinders} and \code{year} as numeric predictors. There are,
-however, only 5 distinct values of \code{cylinders} (ranging from 3 to
-8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7
+however, only 5 distinct values of \code{cylinders}, ranging from 3 to
+8, but cars with 3 or 5 cylinders are rare, and none of the cars has 7
 cylinders. There are similarly only 13 distinct years between 1970 and
@@
   obtain a more realistic, if still biased, estimate of the predictive
-  power of a model.} It's apparent that most these variables are
-positively skewed and that many of the pairwise relationships among them
-are nonlinear.
+  power of a model.} It's apparent that most of the numeric variables
+are positively skewed and that many of the pairwise relationships among
+them are nonlinear.
 
@@
 R> m.auto <- lm(mpg ~ ., data = Auto)
-R> crPlots(m.auto, pch=".")
+R> crPlots(m.auto,
++         terms = ~ displacement + horsepower + weight + acceleration,
++         pch = ".", ylab = "C+R", las = 2)
 \end{CodeInput}
@@
 
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-working-model-1}
+{\centering \includegraphics[width=0.7\linewidth]{Figures/Auto-explore-cr-plots-1}
 
@@
 
-\caption[Component+residual plots for the working model fit to the \code{Auto} data]{Component+residual plots for the working model fit to the \code{Auto} data.
-The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.}\label{fig:Auto-working-model}
+\caption[Component + residual plots for the numeric predictors in the working model fit to the \code{Auto} data]{Component + residual plots for the numeric predictors in the working model fit to the \code{Auto} data.}\label{fig:Auto-explore-cr-plots}
 \end{figure}
@@
 
-The component+residual plots in Figure \ref{fig:Auto-working-model}
-clearly reveal the inadequacy of the model.
+The component + residual (``C+R'') plots in Figure
+\ref{fig:Auto-explore-cr-plots}, produced by the \code{crPlots()}
+function in the \pkg{car} package, clearly reveal the inadequacy of the
+model. The broken blue lines in the C+R plots show linear least-squares
+fits, while the solid magenta lines are nonparametric-regression
+smooths.
 
@@
 variables towards multivariate normality by a generalization of Box and
-Cox's maximum-likelihood-like approach \citep{BoxCox:1964}. Several
-``families'' of power transformations can be used, including the
-original Box-Cox family (which is the default), simple powers (and
-roots), and two adaptations of the Box-Cox family to data that may
-include negative values and 0s: the Box-Cox-with-negatives family and
-the Yeo-Johnson family; see \citet[Chap. 8]{Weisberg:2014} and
-\citet[Chap. 3]{FoxWeisberg:2019} for details. We proceed to transform
-the numeric predictors in the \code{Auto} regression towards
+Cox's maximum-likelihood-like approach \citep{BoxCox:1964}. Although the
+multinormal likelihood is maximized, the variables are transformed
+individually. Several ``families'' of power transformations can be used,
+including the original Box-Cox family (which is the default), simple
+powers (and roots), and two adaptations of the Box-Cox family to data
+that may include negative values and zeros: the Box-Cox-with-negatives
+family and the Yeo-Johnson family; see \citet[Chap. 8]{Weisberg:2014}
+and \citet[Chap. 3]{FoxWeisberg:2019} for details. We proceed to
+transform the numeric predictors in the \code{Auto} regression towards
 multivariate normality:
@@
                                LRT df  pval
-LR test, lambda = (0 0 0 0) 4.872911  4 0.30059
+LR test, lambda = (0 0 0 0) 4.8729  4 0.301
 
@@
                                LRT df   pval
-LR test, lambda = (1 1 1 1) 390.0777  4 < 2.22e-16
+LR test, lambda = (1 1 1 1) 390.08  4 <2e-16
 \end{CodeOutput}
@@
 We then apply the (rounded) transformations---all, as it turns out, logs
-(i.e., ``Oth'' powers)---to the data and re-estimate the model:
+(i.e., ``zeroth'' powers)---to the data and re-estimate the model:
 
@@
 R> for (pred in num.predictors){
-+   A[, pred] <- bcPower(A[, pred], lambda=powers[pred])
++   A[, pred] <- bcPower(A[, pred], lambda = powers[pred])
 + }
-R> m <- update(m.auto, data=A)
+R> m <- update(m.auto, data = A)
 \end{CodeInput}
@@
                              LRT df  pval
-LR test, lambda = (0) 0.0008015428  1 0.97741
+LR test, lambda = (0) 0.00080154  1 0.977
 
@@
                          LRT df   pval
-LR test, lambda = (1) 124.1307  1 < 2.22e-16
+LR test, lambda = (1) 124.13  1 <2e-16
 \end{CodeOutput}
@@
 
-The transformed numeric variables are much better-behaved (cf., Figure
-\ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and Figure
-\ref{fig:Auto-transformed-scatterplot-matrix}):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
-+                   + acceleration, 
-+                   smooth=list(spread=FALSE), data=A, pch=".")
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-transformed-scatterplot-matrix-1}
-
-}
-
-\caption[Scatterplot matrix for the transformed numeric variables in the \code{Auto} data]{Scatterplot matrix for the transformed numeric variables in the \code{Auto} data}\label{fig:Auto-transformed-scatterplot-matrix}
-\end{figure}
-\end{CodeChunk}
-
-And the partial relationships in the model fit to the transformed data
-are much more nearly linear (cf., Figure \ref{fig:Auto-working-model},
-on page \pageref{fig:Auto-working-model}, and Figure
-\ref{fig:Auto-CR-plots-transformed}):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> crPlots(m, pch=".")
-\end{CodeInput}
-\begin{figure}
-
-{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-CR-plots-transformed-1}
-
-}
-
-\caption[Component+residual plots for the model fit to the transformed \code{Auto} data]{Component+residual plots for the model fit to the transformed \code{Auto} data}\label{fig:Auto-CR-plots-transformed}
-\end{figure}
-\end{CodeChunk}
+The transformed numeric variables are much better-behaved and the
+partial relationships in the model fit to the transformed data are much
+more nearly linear (as we invite the reader to verify by redrawing the
+scatterplot matrix for the transformed variables and the C+R plots for
+the regression fit to the transformed data).
 
@@
 to perform predictor selection, employing the BIC model-selection
-criterion (by setting the \texttt{k} argument of \texttt{stepAIC()} to
+criterion (by setting the \code{k} argument of \code{stepAIC()} to
 \(\log n\)):
@@
 R> library("MASS")
-R> m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
-R> summary(m.step)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-lm(formula = log(mpg) ~ horsepower + weight + acceleration + 
-    year + origin, data = A)
-
-Residuals:
-     Min       1Q   Median       3Q      Max 
--0.35230 -0.05682  0.00677  0.06741  0.35861
-
-Coefficients:
-              Estimate Std. Error t value Pr(>|t|)
-(Intercept)   9.434594   0.261529  36.075  < 2e-16
-horsepower   -0.276254   0.056143  -4.921 1.30e-06
-weight       -0.609071   0.056003 -10.876  < 2e-16
-acceleration -0.131380   0.053195  -2.470  0.01397
-year71        0.027984   0.028936   0.967  0.33412
-year72       -0.007111   0.028446  -0.250  0.80274
-year73       -0.039529   0.026014  -1.520  0.12947
-year74        0.052752   0.029986   1.759  0.07936
-year75        0.053199   0.029280   1.817  0.07004
-year76        0.074317   0.028212   2.634  0.00878
-year77        0.137931   0.028875   4.777 2.56e-06
-year78        0.145876   0.027529   5.299 1.99e-07
-year79        0.236036   0.029080   8.117 6.99e-15
-year80        0.335274   0.031148  10.764  < 2e-16
-year81        0.262872   0.030555   8.603  < 2e-16
-year82        0.323391   0.029608  10.922  < 2e-16
-originEurope  0.055818   0.016785   3.326  0.00097
-originJapan   0.043554   0.017479   2.492  0.01314
-
-Residual standard error: 0.1049 on 374 degrees of freedom
-Multiple R-squared:  0.909, Adjusted R-squared:  0.9049 
-F-statistic: 219.8 on 17 and 374 DF,  p-value: < 2.2e-16
+R> m.step <- stepAIC(m, k = log(nrow(A)), trace = FALSE)
+R> brief(m.step)
+\end{CodeInput}
+\begin{CodeOutput}
+           (Intercept) horsepower weight acceleration year71   year72  year73
+Estimate         9.435    -0.2763 -0.609      -0.1314 0.0280 -0.00711 -0.0395
+Std. Error       0.262     0.0561  0.056       0.0532 0.0289  0.02845  0.0260
+           year74 year75 year76 year77 year78 year79 year80 year81 year82
+Estimate   0.0528 0.0532 0.0743 0.1379 0.1459 0.2360 0.3353 0.2629 0.3234
+Std. Error 0.0300 0.0293 0.0282 0.0289 0.0275 0.0291 0.0311 0.0306 0.0296
+           originEurope originJapan
+Estimate         0.0558      0.0436
+Std. Error       0.0168      0.0175
+
+ Residual SD = 0.105 on 374 df, R-squared = 0.909
 \end{CodeOutput}
@@
 \begin{CodeOutput}
-[1] 6.512144
+[1] 6.5121
 attr(,"casewise loss")
@@
 \begin{CodeOutput}
-[1] 8.093171
+[1] 8.0932
 attr(,"casewise loss")
@@
   \pkg{cv} package. The median absolute error, however, cannot be
-  expressed as a casewise average (see Section
-  \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).}
+  expressed as a casewise average (see Appendix A.2). Yet another
+  possibility, suggested to us by a reviewer of an earlier version of
+  this paper, is to compute the MSE for both models on the log-response
+  scale rather than on the original scale of the response. This
+  suggestion is quite general, in that the essential point is to use the
+  \emph{same} response scale for all models to be compared, not
+  necessarilty the \emph{original} scale of the response.}
 
@@
 the whole model-selection procedure. The first argument to
-\code{cv()}---here \code{selectTransAndStepAIC}---is a model-selection
+\code{cv()}---here \code{selectTransStepAIC}---is a model-selection
 function capable of refitting the model with a fold omitted and
@@
 \begin{CodeInput}
-R> cvs <- cv(selectTransStepAIC, data=Auto, seed=76692, 
-+           working.model=m.auto,
-+           predictors=num.predictors,
-+           response="mpg", AIC=FALSE)
+R> cvs <- cv(selectTransStepAIC, data = Auto, seed = 76692,
++           working.model = m.auto, predictors = num.predictors,
++           response = "mpg", AIC = FALSE)
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> cvs
+R> summary(cvs)
 \end{CodeInput}
@@
 criterion: mse
-cross-validation criterion = 7.485557
-bias-adjusted cross-validation criterion = 7.343535
-full-sample criterion = 6.512144 
-\end{CodeOutput}
-\begin{CodeInput}
-R> compareFolds(cvs)
-\end{CodeInput}
-\begin{CodeOutput}
-CV criterion by folds:
-   fold.1    fold.2    fold.3    fold.4    fold.5    fold.6    fold.7    fold.8 
- 6.000589 10.509334  7.444420 10.617864  5.049238  7.183160  4.077682 10.501909 
-   fold.9   fold.10 
- 9.181184  4.250736
-
-Coefficients by folds:
-        (Intercept) horsepower lam.acceleration lam.displacement lam.horsepower
-Fold 1      9.71384   -0.17408          0.50000          0.00000        0.00000
-Fold 2      9.21713   -0.31480          0.00000          0.00000        0.00000
-Fold 3      9.61824   -0.19248          0.00000          0.00000        0.00000
-Fold 4      8.69910   -0.25523          0.50000          0.00000        0.00000
-Fold 5      9.14403   -0.14934          0.00000          0.00000        0.00000
-Fold 6      9.63481   -0.16739          0.50000          0.00000        0.00000
-Fold 7      9.98933   -0.36847          0.00000          0.00000       -0.15447
-Fold 8      9.06301   -0.29721          0.00000          0.00000        0.00000
-Fold 9      8.88315   -0.22684          0.00000          0.00000        0.00000
-Fold 10     9.61727   -0.17086          0.00000          0.00000        0.00000
-        lam.weight   lambda   weight   year71   year72   year73   year74
-Fold 1     0.00000  0.00000 -0.74636  0.03764 -0.00327 -0.02477  0.05606
-Fold 2     0.00000  0.00000 -0.47728  0.02173 -0.01488 -0.03770  0.04312
-Fold 3     0.00000  0.00000 -0.72085  0.01128 -0.02569 -0.03872  0.05187
-Fold 4     0.00000  0.00000 -0.53846  0.02153 -0.02922 -0.05181  0.04136
-Fold 5     0.00000  0.00000 -0.69081  0.02531 -0.01062 -0.04625  0.05039
-Fold 6     0.00000  0.00000 -0.74049  0.02456  0.00759 -0.03412  0.06266
-Fold 7     0.00000  0.00000 -0.72843  0.02532 -0.01271 -0.04144  0.04568
-Fold 8     0.00000  0.00000 -0.46392  0.02702 -0.02041 -0.05605  0.04437
-Fold 9     0.00000  0.00000 -0.47136  0.00860 -0.03620 -0.04835  0.01906
-Fold 10    0.00000  0.00000 -0.73550  0.02937 -0.00899 -0.03814  0.05408
-          year75   year76   year77   year78   year79   year80   year81   year82
-Fold 1   0.07080  0.07250  0.14420  0.14281  0.23266  0.35127  0.25635  0.30546
-Fold 2   0.04031  0.06718  0.13094  0.14917  0.21871  0.33192  0.26196  0.30943
-Fold 3   0.03837  0.06399  0.11593  0.12601  0.20499  0.32821  0.24478  0.29204
-Fold 4   0.04072  0.05537  0.12292  0.14083  0.22878  0.32947  0.25140  0.27248
-Fold 5   0.05596  0.07044  0.13356  0.14724  0.24675  0.33331  0.26938  0.32594
-Fold 6   0.06940  0.07769  0.14211  0.14647  0.23532  0.34761  0.26737  0.33062
-Fold 7   0.03614  0.07385  0.12976  0.14040  0.23976  0.33998  0.27652  0.30659
-Fold 8   0.06573  0.08135  0.13158  0.13987  0.23011  0.32880  0.25886  0.30538
-Fold 9   0.03018  0.05846  0.10536  0.11722  0.20665  0.31533  0.23352  0.29375
-Fold 10  0.04881  0.07862  0.14101  0.14313  0.23258  0.35649  0.26214  0.32421
-        acceleration displacement cylinders5-6 cylinders8 originEurope
-Fold 1                                                                
-Fold 2      -0.18909     -0.09197                                     
-Fold 3                                                                
-Fold 4      -0.03484                  -0.09080   -0.10909             
-Fold 5                                                         0.06261
-Fold 6                                                                
-Fold 7                                                                
-Fold 8      -0.17676     -0.10542                                     
-Fold 9      -0.14514     -0.13452                                     
-Fold 10                                                               
-        originJapan
-Fold 1             
-Fold 2             
-Fold 3             
-Fold 4             
-Fold 5         0.04
-Fold 6             
-Fold 7             
-Fold 8             
-Fold 9             
-Fold 10            
-\end{CodeOutput}
-\end{CodeChunk}
-
-The other arguments to \code{cv()} are (see \code{?cv:::cv.function} for
-additional optional arguments and details):
+cross-validation criterion = 7.4856
+bias-adjusted cross-validation criterion = 7.3435
+full-sample criterion = 6.5121
+\end{CodeOutput}
+\end{CodeChunk}
+
+The other arguments to \code{cv()} are:\footnote{See \code{?cv.function}
+  for additional optional arguments and details.}
 
@@
   and \code{response}---that are candidates for transformation; and
-  \code{AIC=FALSE}, which specifies use of the BIC for model selection.
+  \code{AIC = FALSE}, which specifies use of the BIC for model
+  selection.
 \end{itemize}
@@
   When we look at the transformations and predictors selected with each
-  of the 10 folds omitted (i.e., the output of \code{compareFolds()}),
-  we see that there is little uncertainty in choosing variable
-  transformations (the \code{lam.*}s for the \(x\)s and \code{lambda}
-  for \(y\) in the output), but considerably more uncertainty in
-  subsequently selecting predictors: \code{horsepower}, \code{weight},
-  and \code{year} are always included among the selected predictors;
-  \code{acceleration} and \code{displacement} are included respectively
-  in 4 and 3 of 10 selected models; and \code{cylinders} and
-  \code{origin} are each included in only 1 of 10 models. Recall that
-  when we selected predictors for the full data, we obtained a model
-  with \code{horsepower}, \code{weight}, \code{acceleration},
+  of the 10 folds omitted (i.e., the output of \code{compareFolds(cvs)},
+  which isn't shown), we see that there is little uncertainty in
+  choosing variable transformations, but considerably more uncertainty
+  in subsequently selecting predictors: \code{horsepower},
+  \code{weight}, and \code{year} are always included among the selected
+  predictors; \code{acceleration} and \code{displacement} are included
+  respectively in 4 and 3 of 10 selected models; and \code{cylinders}
+  and \code{origin} are each included in only 1 of 10 models. Recall
+  that when we selected predictors for the full data, we obtained a
+  model with \code{horsepower}, \code{weight}, \code{acceleration},
   \code{year}, and \code{origin}.
@@
 
-\subsection[Example: Applying recursive CV to polynomial regression for
-the Auto data]{\texorpdfstring{Example: Applying recursive CV to
-polynomial regression for the Auto
-data\footnote{What we call here ``recursive CV'' has also been termed
+\subsection[Example: Applying meta CV to polynomial regression for the
+Auto data]{\texorpdfstring{Example: Applying meta CV to polynomial
+regression for the Auto
+data\footnote{What we call here ``meta CV'' has also been termed
   ``nested CV'' \citep[e.g., in the \pkg{cvms} package
@@
   ``nested CV'' for the procedure described by
-  \citet{BatesHastieTibshirani:2023}.}}{Example: Applying recursive CV to polynomial regression for the Auto data}}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-datarecursive-cv}
+  \citet{BatesHastieTibshirani:2023}.}}{Example: Applying meta CV to polynomial regression for the Auto data}}\label{example-applying-meta-cv-to-polynomial-regression-for-the-auto-datameta-cv}
 
@@
 5.3]{JamesEtAl:2021}, we fit polynomial regressions up to degree 10 to
-the relationship of \texttt{mpg} to \texttt{horsepower} for the
-\texttt{Auto} data, saving the 10 models, named \texttt{m.1} through
-\texttt{m.10}, in the list \texttt{mlist}. We then used \texttt{cv()} to
-compare the cross-validated MSE for the 10 models, discovering that the
-7th degree polynomial had the smallest MSE (by a small margin).
+the relationship of \code{mpg} to \code{horsepower} for the \code{Auto}
+data, saving the 10 models, named \code{m.1} through \code{m.10}, in the
+list \code{mlist}. We then used \code{cv()} to compare the
+cross-validated MSE for the 10 models, discovering that the 7th degree
+polynomial had the smallest MSE (by a small margin).
 
@@
 optimistic. One solution is to cross-validate the process of using CV to
-select the ``best'' model---that is, to apply CV to CV recursively. The
-function \texttt{selectModelList()}, which is suitable for use with
-\texttt{cv()}, implements this idea.
-
-Applying \texttt{selectModelList()} to the \texttt{Auto}
-polynomial-regression models, and using 10-fold CV, we obtain:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> recursiveCV.auto <- cv(selectModelList, data=Auto, 
-+                        working.model=mlist,
-+                        save.model=TRUE, seed=2120)
+select the ``best'' model---that is, to apply meta-CV. The function
+\code{selectModelList()}, which is suitable for use with \code{cv()},
+implements this idea.
+
+Applying \code{selectModelList()} to the \code{Auto}
+polynomial-regression models, and using 10-fold CV, we
+obtain:\footnote{Two functions are employed to access information in
+  objects produced by functions in the \pkg{cv} package: \code{cvInfo()}
+  and \code{as.data.frame()} are generic functions (the latter a
+  standard-\proglang{R} generic) with methods for classes \code{"cv"},
+  \code{"cvList"}, \code{"cvModlist"}, and \code{"cvSelect"} (in the
+  case of \code{cvInfo()}) , and \code{"cv"}, \code{"cvList"}, and
+  \code{"cvModlist"} (in the case of \code{as.data.frame()}). Here we
+  use \code{cvInfo(metaCV.auto, "selected model")} to retrieve the model
+  selected by \code{cv()} via \code{selectModelList()}. See \code{?cv}
+  and \code{?cvInfo.cvSelect} for details.}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> metaCV.auto <- cv(selectModelList, data = Auto,
++                   working.model = mlist, save.model = TRUE,
++                        seed = 2120)
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> recursiveCV.auto
+R> summary(metaCV.auto)
 \end{CodeInput}
@@
 criterion: mse
-cross-validation criterion = 20.01249
-bias-adjusted cross-validation criterion = 20.61866
-full-sample criterion = 18.74615 
-\end{CodeOutput}
-\begin{CodeInput}
-R> recursiveCV.auto$selected.model
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-lm(formula = mpg ~ poly(horsepower, p), data = Auto)
-
-Coefficients:
+cross-validation criterion = 20.012
+bias-adjusted cross-validation criterion = 20.619
+full-sample criterion = 18.746
+\end{CodeOutput}
+\begin{CodeInput}
+R> brief(m.sel <- cvInfo(metaCV.auto, "selected model"))
+\end{CodeInput}
+\begin{CodeOutput}
            (Intercept) poly(horsepower, p)1 poly(horsepower, p)2
-              23.446              -120.138                44.090  
+Estimate        23.446               -120.1                 44.1
+Std. Error       0.217                  4.3                  4.3
            poly(horsepower, p)3 poly(horsepower, p)4 poly(horsepower, p)5
-              -3.949                -5.188                13.272  
+Estimate                  -3.95                -5.19                 13.3
+Std. Error                 4.30                 4.30                  4.3
            poly(horsepower, p)6 poly(horsepower, p)7
-              -8.546                 7.981  
-\end{CodeOutput}
-\begin{CodeInput}
-R> cv(mlist[[7]], seed=2120) # same seed for same folds
+Estimate                  -8.55                 7.98
+Std. Error                 4.30                 4.30
+
+ Residual SD = 4.3 on 384 df, R-squared = 0.702
+\end{CodeOutput}
+\begin{CodeInput}
+R> cv(m.sel, seed = 2120)
 \end{CodeInput}
@@
 \begin{CodeOutput}
-10-Fold Cross Validation
-method: Woodbury
-criterion: mse
-cross-validation criterion = 18.89822
-bias-adjusted cross-validation criterion = 18.85429
-full-sample criterion = 18.07817 
-\end{CodeOutput}
-\end{CodeChunk}
-
-As expected, recursive CV produces a larger estimate of MSE for the
-selected 7th degree polynomial model than CV applied directly to this
-model.
-
-We can equivalently call \texttt{cv()} with the list of models as its
-first argument and set the argument \texttt{recursive=TRUE}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> cv(mlist, data=Auto, seed=2120, recursive=TRUE, save.model=TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-R RNG seed set to 2120
-\end{CodeOutput}
-\begin{CodeOutput}
-10-Fold Cross Validation
-cross-validation criterion = 20.01249
-bias-adjusted cross-validation criterion = 20.61866
-full-sample criterion = 18.74615 
-\end{CodeOutput}
+cross-validation criterion (mse) = 18.898
+\end{CodeOutput}
+\end{CodeChunk}
+
+As expected, meta CV produces a larger estimate of MSE for the selected
+7th degree polynomial model than CV applied directly to this model
+(using the same seed for \proglang{R}'s random-number generator to
+obtain the same folds).
+
+We can equivalently call \code{cv()} with the list of models as its
+first argument and set the argument \code{meta = TRUE} (identical output
+not shown):
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> summary(cv(mlist, data = Auto, seed = 2120, meta = TRUE,
++            save.model = TRUE))
+\end{CodeInput}
 \end{CodeChunk}
@@
 use with the \code{"function"} method for \code{cv()}. In this section,
-we illustrate (1) and (2); more diverse and extensive examples may be
-found in the vignette on extending the \pkg{cv} package.
+we illustrate (1) and (2); more diverse and extensive examples
+(including of (3)) may be found in the vignette on extending the
+\pkg{cv} package.
 
@@
 14.2.1]{Fox:2016}, with data from the British Election Panel Study on
-vote choice in the 2001 British election. Data for the example are in
-the \code{BEPS} data set in the \pkg{carData} package:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> data("BEPS", package="carData")
+vote choice in the 2001 UK election. Data for the example are in the
+\code{BEPS} data set in the \pkg{carData} package:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> data("BEPS", package = "carData")
 R> summary(BEPS)
@@
                vote          age       economic.cond.national
- Conservative    :462   Min.   :24.00   Min.   :1.000         
- Labour          :720   1st Qu.:41.00   1st Qu.:3.000         
- Liberal Democrat:343   Median :53.00   Median :3.000         
-                        Mean   :54.18   Mean   :3.246         
-                        3rd Qu.:67.00   3rd Qu.:4.000         
-                        Max.   :93.00   Max.   :5.000         
+ Conservative    :462   Min.   :24.0   Min.   :1.00
+ Labour          :720   1st Qu.:41.0   1st Qu.:3.00
+ Liberal Democrat:343   Median :53.0   Median :3.00
+                        Mean   :54.2   Mean   :3.25
+                        3rd Qu.:67.0   3rd Qu.:4.00
+                        Max.   :93.0   Max.   :5.00
  economic.cond.household     Blair          Hague         Kennedy
- Min.   :1.00            Min.   :1.000   Min.   :1.000   Min.   :1.000  
- 1st Qu.:3.00            1st Qu.:2.000   1st Qu.:2.000   1st Qu.:2.000  
- Median :3.00            Median :4.000   Median :2.000   Median :3.000  
- Mean   :3.14            Mean   :3.334   Mean   :2.747   Mean   :3.135  
- 3rd Qu.:4.00            3rd Qu.:4.000   3rd Qu.:4.000   3rd Qu.:4.000  
- Max.   :5.00            Max.   :5.000   Max.   :5.000   Max.   :5.000  
+ Min.   :1.00            Min.   :1.00   Min.   :1.00   Min.   :1.00
+ 1st Qu.:3.00            1st Qu.:2.00   1st Qu.:2.00   1st Qu.:2.00
+ Median :3.00            Median :4.00   Median :2.00   Median :3.00
+ Mean   :3.14            Mean   :3.33   Mean   :2.75   Mean   :3.14
+ 3rd Qu.:4.00            3rd Qu.:4.00   3rd Qu.:4.00   3rd Qu.:4.00
+ Max.   :5.00            Max.   :5.00   Max.   :5.00   Max.   :5.00
      Europe      political.knowledge    gender
- Min.   : 1.000   Min.   :0.000       female:812  
- 1st Qu.: 4.000   1st Qu.:0.000       male  :713  
- Median : 6.000   Median :2.000                   
- Mean   : 6.729   Mean   :1.542                   
- 3rd Qu.:10.000   3rd Qu.:2.000                   
- Max.   :11.000   Max.   :3.000                   
+ Min.   : 1.00   Min.   :0.00        female:812
+ 1st Qu.: 4.00   1st Qu.:0.00        male  :713
+ Median : 6.00   Median :2.00
+ Mean   : 6.73   Mean   :1.54
+ 3rd Qu.:10.00   3rd Qu.:2.00
+ Max.   :11.00   Max.   :3.00
 \end{CodeOutput}
@@
 \item
-  \code{age}, in years;
+  \code{age}, in years.
 \item
@@
 \begin{CodeInput}
-R> library("nnet", quietly=TRUE)
-R> m.beps <- multinom(vote ~ age + gender + economic.cond.national +
-+                        economic.cond.household + Blair + Hague + Kennedy +
-+                        Europe*political.knowledge, data=BEPS)
-\end{CodeInput}
-\begin{CodeOutput}
-# weights:  36 (22 variable)
-initial  value 1675.383740 
-iter  10 value 1240.047788
-iter  20 value 1163.199642
-iter  30 value 1116.519687
-final  value 1116.519666 
-converged
-\end{CodeOutput}
-\end{CodeChunk}
-
-The \code{Europe} \(\times\) \code{political.knowledge} interaction is
-associated with a very small \(p\)-value. Figure \ref{fig:BEPS-plot}
-shows an ``effect plot,'' using the the \pkg{effects} package
-\citep{FoxWeisberg:2019} to visualize the interaction in a
+R> library("nnet", quietly = TRUE)
+R> m.beps <- multinom(vote ~ age + gender + economic.cond.national
++                         + economic.cond.household + Blair + Hague
++                         + Kennedy + Europe * political.knowledge,
++                    data = BEPS, trace = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+Figure \ref{fig:BEPS-plot} shows an ``effect plot,'' using the the
+\pkg{effects} package \citep{FoxWeisberg:2019} to visualize the
+\code{Europe} \(\times\) \code{political.knowledge} interaction in a
 ``stacked-area'' graph:
@@
 R> plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
-+             xlevels=list(Europe=1:11, political.knowledge=0:3),
-+             fixed.predictors=list(given.values=c(gendermale=0.5))),
-+      lines=list(col=c("blue", "red", "orange")), # party colors
-+      axes=list(x=list(rug=FALSE), y=list(style="stacked")))
++             xlevels = list(Europe = 1:11, political.knowledge = 0:3),
++             fixed.predictors = list(given.values = c(gendermale = 0.5))),
++      lines = list(col = c("blue", "red", "orange")),
++      axes = list(x = list(rug = FALSE), y = list(style = "stacked")))
 \end{CodeInput}
@@
 
-{\centering \includegraphics{JSS-article-3_files/figure-latex/BEPS-plot-1}
+{\centering \includegraphics{Figures/BEPS-plot-1}
 
@@
 
-\caption[Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study]{Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study.}\label{fig:BEPS-plot}
+\caption[Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study, using party colors]{Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study, using party colors.}\label{fig:BEPS-plot}
 \end{figure}
@@
 \begin{CodeOutput}
-function(y, yhat) {
+function (y, yhat)
+{
     if (!all(y %in% c(0, 1)))
         stop("response values not all 0 or 1")
-  if (any(yhat < 0) ||
-      any(yhat > 1))
+    if (any(yhat < 0) || any(yhat > 1))
         stop("fitted values outside of interval [0, 1]")
     yhat <- round(yhat)
-  result <- mean(y != yhat) # proportion in error
+    result <- mean(y != yhat)
     attr(result, "casewise loss") <- "y != round(yhat)"
@@
 }
-<bytecode: 0x13fb26ea8>
+<bytecode: 0x14fd342e8>
 <environment: namespace:cv>
@@
 After doing some error checking, \code{BayesRule()} rounds the predicted
-proabability of a 1 (``success'') response in a binary regression model
-to 0 or 1 to obtain a categorical prediction, and then reports the
-proportion of incorrect predictions. Because the Bayes's rule criterion
-is an average of casewise components (as, e.g., is the MSE), a
-\code{"casewise loss"} attribute is attached to the result, enabling the
-computation of bias correction and confidence intervals (as discussed in
-Section
-\ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
+probability \code{yhat} of a 1 (``success'') response in a binary
+regression model to 0 or 1 to obtain a categorical prediction, and then
+reports the proportion of incorrect predictions. Because the Bayes's
+rule criterion is an average of casewise components (as, e.g., is the
+MSE), a \code{"casewise loss"} attribute is attached to the result,
+enabling the computation of bias correction and confidence intervals (as
+discussed in Appendix A.2).
 
@@
 \begin{CodeInput}
-R> yhat <- predict(m.beps, type="class")
+R> yhat <- predict(m.beps, type = "class")
 R> head(yhat)
@@
 \begin{CodeOutput}
-[1] 0.3186885
+[1] 0.31869
 attr(,"casewise loss")
@@
 The \code{predict()} method for \code{"multinom"} models called with
-argument \code{type="class"} reports the Bayes's rule prediction for
+argument \code{type = "class"} reports the Bayes's rule prediction for
 each case---that is, the response category with the highest predicted
-probability. Our \code{BayesRuleMulti()} function calculates the
-proportion of misclassified cases. Because this value is also the mean
-of casewise components, we attach a \code{"casewise loss"} attribute to
-the result.
+probability. The argument \code{yhat} to our \code{BayesRuleMulti()}
+function is the vector of Bayes's rule categorical predictions, while
+\code{y} is the vector of observed categorical responses. The function
+calculates and returns the proportion of misclassified cases. Because
+this value is also the mean of casewise components, we attach a
+\code{"casewise loss"} attribute to the result.
 
@@
 \begin{CodeInput}
-R> xtabs(~ vote, data=BEPS)/nrow(BEPS)
+R> xtabs(~ vote, data = BEPS) / nrow(BEPS)
 \end{CodeInput}
@@
     Conservative           Labour Liberal Democrat
-       0.3029508        0.4721311        0.2249180 
+         0.30295          0.47213          0.22492
 \end{CodeOutput}
@@
 \begin{CodeInput}
-R> cv(m.beps, seed=3465, criterion=BayesRuleMulti)
+R> cv(m.beps, seed = 3465, criterion = BayesRuleMulti)
 \end{CodeInput}
@@
 \code{GetResponse.multinom()} method that returns the factor response
-\citep[using the \texttt{get\_response()} function from the
-\pkg{insight} package,][]{LudeckeWaggonerMakowski:2019},
+{[}using the \code{get\_response()} function from the \pkg{insight}
+package,
 
@@
 R RNG seed set to 3465
-\end{CodeOutput}
-\begin{CodeOutput}
-# weights:  36 (22 variable)
-initial  value 1507.296060 
-iter  10 value 1134.575036
-iter  20 value 1037.413231
-iter  30 value 1007.705242
-iter  30 value 1007.705235
-iter  30 value 1007.705235
-final  value 1007.705235 
-converged
 \end{CodeOutput}
@@
 +     NextMethod(
-+       type = "class",
-+       criterion = criterion,
++       type = "class", criterion = criterion,
 +       criterion.name = deparse(substitute(criterion))
@@
 
-That is, we simply call the default \texttt{cv()} method (via
-\code{NextMethod()}) with the \texttt{type} argument properly set. In
-addition to supplying the correct \texttt{type} argument, our method
-sets the default \texttt{criterion} for the \texttt{cv.multinom()}
-method to \texttt{BayesRuleMulti}. Adding the argument
-\texttt{criterion.name=} \texttt{deparse(substitute(criterion))} is
-inessential, but it insures that printed output will include the name of
-the criterion function that's employed, whether it's the default
-\texttt{BayesRuleMulti} or something else. Prior to invoking
-\texttt{NextMethod()}, we called \texttt{update()} with
-\texttt{trace=FALSE} to suppress the iteration history reported by
-default by \texttt{multinom()}---it would be tedious to see the
-iteration history for each fold.
+That is, we simply call the default \code{cv()} method (via
+\code{NextMethod()}) with the \code{type} argument properly set. In
+addition to supplying the correct \code{type} argument, our method sets
+the default \code{criterion} for the \code{cv.multinom()} method to
+\code{BayesRuleMulti}. Adding the argument
+\code{criterion.name = deparse(substitute(criterion))} is inessential,
+but it insures that printed output will include the name of the
+criterion function that's employed, whether it's the default
+\code{BayesRuleMulti} or something else. Prior to invoking
+\code{NextMethod()}, we call \code{update()} with \code{trace = FALSE}
+to suppress the iteration history reported by default by
+\code{multinom()}---it would be tedious to see the iteration history for
+each fold.
 
@@
 \begin{CodeInput}
-R> cv(m.beps, seed=3465)
+R> summary(cv(m.beps, seed = 3465))
 \end{CodeInput}
@@
 criterion: BayesRuleMulti
-cross-validation criterion = 0.3245902
-bias-adjusted cross-validation criterion = 0.3236756
-95% CI for bias-adjusted CV criterion = (0.300168, 0.3471831)
-full-sample criterion = 0.3186885 
+cross-validation criterion = 0.32459
+bias-adjusted cross-validation criterion = 0.32368
+95% CI for bias-adjusted CV criterion = (0.30017, 0.34718)
+full-sample criterion = 0.31869
 \end{CodeOutput}
@@
 
-\section{Comparing cv to other R software for
-cross-validation}\label{comparing-cv-to-other-r-software-for-cross-validation}
+\section{Comparing cv to other software for
+cross-validation}\label{comparing-cv-to-other-software-for-cross-validation}
+
+In this section with briefly compare the \code{cv} package to other
+packages in \proglang{R} that support cross-validation, and then even
+more briefly consider facilities for cross-validation in other commonly
+used statistical software, \proglang{SAS}, \proglang{Stata}, and
+\proglang{Python}.
+
+\subsection{Other R software}\label{other-r-software}
 
@@
 \pkg{mlexperiments} \citep{Kapsner:2024}, \pkg{origami}
-\citep{CoyleEtAl:2022}, and \pkg{rsample} (which also supports
+\citep{CoyleEtAl:2022}, and \pkg{rsample} (which also implements
 bootstrapping), modularize the CV process to a greater or lesser extent.
@@
 \begin{CodeInput}
-R> # formula for model to be fit; Attrition is binary:
 R> mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome
 R>
-R> rs_obj <- loo_cv(attrition) # original uses vfold_cv()
+R> rs_obj <- loo_cv(attrition)
 R>
-R> # user-supplied function:
 R> holdout_results <- function(splits, ...) {
-+   mod <- glm(..., data=analysis(splits), family=binomial)
++   mod <- glm(..., data = analysis(splits), family = binomial)
 +   holdout <- assessment(splits)
-+   res <- broom::augment(mod, newdata=holdout)
++   res <- broom::augment(mod, newdata = holdout)
 +   lvls <- levels(holdout$Attrition)
@@
 R> library("purrr", warn.conflicts=FALSE)
-R> # apply CV:
+R>
 R> rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
-R> # summarize results:
-R> rs_obj$accuracy <- map_dbl(rs_obj$results,
-+                            function(x) mean(x$correct))
-R> # accuracy is the complement of the Bayes-rule CV criterion:
+R> rs_obj$accuracy <- map_dbl(rs_obj$results, function(x) mean(x$correct))
 R> 1 - summary(rs_obj$accuracy)
@@
    Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
- 1.0000  0.0000  0.0000  0.1612  0.0000  0.0000 
-\end{CodeOutput}
-\end{CodeChunk}
+  1.000   0.000   0.000   0.161   0.000   0.000
+\end{CodeOutput}
+\end{CodeChunk}
+
+The response variable in this example, \code{Attrition}, is binary. The
+function \code{holdout_results()} must be supplied by the user, and mean
+``\code{accuracy}'' is the complement of the Bayes-rule classification
+error rate.
 
 In this case, it's simple (and, in our opinion, simpler) to program LOO
-CV directly. For example
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> mod.attrition <- glm(mod_form, data=attrition, family=binomial)
+CV directly. For example,
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> mod.attrition <- glm(mod_form, data = attrition, family = binomial)
 R> n <- nrow(attrition)
@@
 R> for (i in 1:n){
-+   m <- update(mod.attrition, data=attrition[-i, ])
-+   yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
++   m <- update(mod.attrition, data = attrition[-i, ])
++   yhat[i] <- predict(m, newdata = attrition[i, ], type = "response")
 + }
-R> mean(((attrition$Attrition == "Yes") - round(yhat))^2)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0.1612245
+R> mean(((attrition$Attrition == "Yes") - round(yhat)) ^ 2)
+\end{CodeInput}
+\begin{CodeOutput}
+[1] 0.16122
 \end{CodeOutput}
@@
 \begin{CodeInput}
-R> library("caret", quietly=TRUE, warn.conflicts=FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Attaching package: 'ggplot2'
-\end{CodeOutput}
-\begin{CodeOutput}
-The following object is masked _by_ '.GlobalEnv':
-
-    mpg
-\end{CodeOutput}
-\begin{CodeOutput}
-The following object is masked from 'package:latticeExtra':
-
-    layer
-\end{CodeOutput}
-\begin{CodeInput}
-R> # reports complement of Bayes-rule criterion:
-R> train(x = attrition[, c("JobSatisfaction", "Gender",
-+                                "MonthlyIncome")],
-+              y = attrition$Attrition,
-+              method = "glm",
-+              trControl = trainControl(method="LOOCV"))
+R> library("ggplot2", warn.conflicts = FALSE)
+R> library("caret", quietly = TRUE, warn.conflicts = FALSE)
+R> train(x = attrition[, c("JobSatisfaction", "Gender", "MonthlyIncome")],
++       y = attrition$Attrition, method = "glm",
++       trControl = trainControl(method = "LOOCV"))
 \end{CodeInput}
@@
   Accuracy  Kappa
-  0.8387755  0    
-\end{CodeOutput}
-\end{CodeChunk}
+  0.83878   0
+\end{CodeOutput}
+\end{CodeChunk}
+
+where ``\code{Accuracy},'' again, is the complement of the
+misclassification rate.
 
@@
 \begin{CodeInput}
-R> (cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
-\end{CodeInput}
-\begin{CodeOutput}
-n-Fold Cross Validation
-method: exact
-criterion: BayesRule
-cross-validation criterion = 0.1612245
-bias-adjusted cross-validation criterion = 0.1612245
-95% CI for bias-adjusted CV criterion = (0.1424194, 0.1800296)
-full-sample criterion = 0.1612245 
-\end{CodeOutput}
-\begin{CodeInput}
-R> all.equal(cv.attrition$`CV crit`, 1 - mean(rs_obj$accuracy), 
-+           check.attributes=FALSE)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] TRUE
+R> cv(mod.attrition, k = "loo", criterion = BayesRule)
+\end{CodeInput}
+\begin{CodeOutput}
+cross-validation criterion (BayesRule) = 0.16122
+\end{CodeOutput}
+\begin{CodeInput}
+R> 1 - mean(rs_obj$accuracy)
+\end{CodeInput}
+\begin{CodeOutput}
+[1] 0.16122
 \end{CodeOutput}
@@
 Let's compare the computational efficiency of the various
-implementations, also showing the use of \code{method="Woodbury"} and
-\code{method="hatvalues"} for \code{cv()}, which, in this example,
+implementations, also showing the use of \code{method = "Woodbury"} and
+\code{method =" hatvalues"} for \code{cv()}, which, in this example,
 produce numerically identical results to the default
-\code{method="exact"}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> all.equal(cv.attrition$`CV crit`,
-+           cv(mod.attrition, k="loo", criterion=BayesRule,
-+              method="Woodbury")$`CV crit`,
-+           )
-\end{CodeInput}
-\begin{CodeOutput}
-[1] TRUE
-\end{CodeOutput}
-\begin{CodeInput}
-R> all.equal(cv.attrition$`CV crit`,
-+           cv(mod.attrition, k="loo", criterion=BayesRule,
-+              method="hatvalues")$`CV crit`
-+           )
-\end{CodeInput}
-\begin{CodeOutput}
-[1] TRUE
-\end{CodeOutput}
-\begin{CodeInput}
-R> microbenchmark::microbenchmark(
-+   rsample={
+\code{method = "exact"}:\footnote{As before, we ask
+  \code{microbenchmark()} to report relative timings, and also as
+  before, we set \code{times = 10} rather than the default
+  \code{times = 100}.}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> set.seed(53437)
+R> print(microbenchmark::microbenchmark(
++   cv.hatvalues = cv(mod.attrition, k = "loo", criterion = BayesRule,
++                     method = "hatvalues"),
++   cv.wood = cv(mod.attrition, k = "loo", criterion = BayesRule,
++                method = "Woodbury"),
++   cv.exact = cv(mod.attrition, k = "loo", criterion = BayesRule),
++   direct = {
++     n <- nrow(attrition)
++     yhat <- numeric(n)
++     for (i in 1:n){
++       m <- update(mod.attrition, data = attrition[-i, ])
++       yhat[i] <- predict(m, newdata = attrition[i, ], type = "response")
++     }
++     mean(((attrition$Attrition == "Yes") - round(yhat)) ^ 2)
++   },
++   rsample = {
 +     rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
@@
 +   },
-+   caret=train(x = attrition[, c("JobSatisfaction", "Gender",
++   caret = train(x = attrition[, c("JobSatisfaction", "Gender",
 +                                   "MonthlyIncome")],
@@
 +               method = "glm",
-+               trControl = trainControl(method="LOOCV")),
-+   direct={
-+     n <- nrow(attrition)
-+     yhat <- numeric(n)
-+     for (i in 1:n){
-+       m <- update(mod.attrition, data=attrition[-i, ])
-+       yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
-+     }
-+     mean(((attrition$Attrition == "Yes") - round(yhat))^2)
-+   },
-+   cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
-+   cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
-+              method="Woodbury"),
-+   cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
-+                   method="hatvalues"),
-+   times=10
-+ )
-\end{CodeInput}
-\begin{CodeOutput}
-Unit: milliseconds
-         expr         min          lq        mean     median          uq
-      rsample 5892.991705 5997.040644 6197.075663 6180.82761 6298.898798
-        caret 6863.187210 6952.637853 7008.026442 7024.60940 7072.257280
-       direct 5483.414538 5488.340073 5603.901283 5564.68712 5655.604370
-     cv.exact 4829.717139 4885.941915 5049.808394 5021.14425 5155.600842
-      cv.wood  104.694812  110.656458  114.844686  117.06384  119.173552
- cv.hatvalues    1.281824    1.513761    1.656785    1.59406    1.668454
-         max neval
- 6871.248999    10
- 7095.635316    10
- 5972.080828    10
- 5486.199381    10
-  121.229702    10
-    2.391448    10
++               trControl = trainControl(method = "LOOCV")),
++   times = 10, unit = "relative"), signif = 3)
+\end{CodeInput}
+\begin{CodeOutput}
+Warning in microbenchmark::microbenchmark(cv.hatvalues = cv(mod.attrition, :
+less accurate nanosecond times to avoid potential integer overflows
+\end{CodeOutput}
+\begin{CodeOutput}
+Unit: relative
+         expr    min   lq   mean median     uq    max neval   cld
+ cv.hatvalues    1.0    1    1.0    1.0    1.0    1.0    10 a
+      cv.wood   50.4   46   47.5   48.5   46.1   47.2    10 a
+     cv.exact 2240.0 2010 2000.0 1980.0 1900.0 1960.0    10  b
+       direct 2500.0 2240 2200.0 2200.0 2100.0 2040.0    10   c
+      rsample 2770.0 2510 2560.0 2460.0 2330.0 3100.0    10    d
+        caret 3250.0 2930 2880.0 2870.0 2760.0 2710.0    10     e
 \end{CodeOutput}
@@
 similar to direct computation of LOO CV and to \code{cv()} with
-\code{method="exact"}, but more than an order of magnitude slower than
+\code{method="exact"}, but almost two orders of magnitude slower than
 \code{cv()} with \code{method="Woodbury"}, and more than three orders of
-magnitude slower then \code{cv()} with \code{method="hatvalues"}.
-
-\section{Computational notes}\label{computational-notes}
-
-\subsection{Efficient computations for linear and generalized linear
-models}\label{efficient-computations-for-linear-and-generalized-linear-models}
+magnitude slower than \code{cv()} with \code{method="hatvalues"}.
+
+\subsection{Comparing cv to other software for cross-validation: SAS,
+Stata, and
+Python}\label{comparing-cv-to-other-software-for-cross-validation-sas-stata-and-python}
+
+Although we do not have recent experience with statistical software for
+cross-validation other than in \proglang{R},\footnote{It is our
+  impression that these three statistical systems, in addition to
+  \proglang{R}, are most likely to be used for applications of
+  cross-validation. We were once frequent users of \proglang{SAS},
+  though not recently, and have only cursory experience with
+  \proglang{Stata} and \proglang{Python}, so we are not ideally situated
+  to evaluate CV software for any of the three.} we have examined
+provisions for CV in \proglang{SAS}, \proglang{Stata}, and
+\proglang{Python}, primarily by locating and reading documentation.
+
+\subsubsection{SAS}\label{sas}
+
+Cross-validation in \proglang{SAS} is associated with particular
+procedures (``\code{PROC}s''), and is distributed across several
+products, including, for example, \code{PROC LOGISTIC} and
+\code{PROC PLS} in \proglang{SAS/STAT} \citep{SAS-Stat:2020}, in
+\proglang{SAS Enterprise Miner} \citep{SAS-Enterprise-Miner:2018}, and
+in \proglang{SAS Visual Data Mining and Machine Learning}
+\citep{SAS-Visual-Data-Mining:2020}. These appear to be straightforward
+implementations of \(k\)-fold CV.
+
+In addition, \proglang{SAS} has programming capabilities, including via
+its macro language, and we're aware of several independent
+implementations of CV in \proglang{SAS} macros. These tend to be limited
+in scope. For example \citet{Slaets-et-al:2014} employ and provide
+access to a \proglang{SAS} macro for \(k\)-fold CV of mixed-effects
+models, which repeatedly calls \code{PROC MIXED} in \proglang{SAS/STAT},
+and which performs cluster-based resampling (one of the options that we
+provide for mixed-effects models in the \pkg{cv} package).
+
+\subsubsection{Stata}\label{stata}
+
+As far as we can tell, \proglang{Stata} has no standard implementation
+of cross-validation. Like \proglang{R}, however, \proglang{Stata} is
+extensible through a programming language that allows users to define
+new commands in so-called ``\code{ado}'' files.
+
+Several such contributed \proglang{Stata} CV commands, varying in
+generality, are shared in an archive maintained at the Boston College
+Department of Economics. For example, the \code{CV\_REGRESS} command
+\citep{Rios-Avila:2018} computes leave-one-out CV for least-squares
+regression using the hatvalues, and thus is comparable to \code{cv()}
+for linear models with \code{method = "hatvalues"}. In contrast, the
+\code{CROSSVALIDATE} command \citep{Schonlau:2020} performs \(k\)-fold
+CV for a wide range of regression models fit in \proglang{Stata}, and is
+roughly comparable to the default method of \code{cv()}, although
+\code{CROSSVALIDATE} returns cross-validated fitted values, from which
+cost criteria can be computed, rather than (as \code{cv()})
+cost-criteria directly.
+
+\subsubsection{Python}\label{python}
+
+The programming language \proglang{Python} is often the software of
+choice for ``machine learning,'' in which cross-validation can play a
+prominent role. Machine learning and statistical modeling more generally
+are supported through extensive \proglang{Python} libraries, such as
+\pkg{scikit-learn}
+\citep{Pedregosa-et-al:2011, ScikitLearnDevelopers:2024} and
+\pkg{TensorFlow} \citep[which isn't limited to use within
+\proglang{Python}]{TensorFlow:2015}.
+
+Functions for cross-validation are provided in \pkg{scikit-learn}, which
+supports \(k\)-fold, repeated \(k\)-fold, and leave-one-out CV. These
+functions can be applied to data sets consisting of independent
+observations, hierarchical groups, or time-series data. A sample program
+in the \pkg{scikit-learn} documentation
+\citep{ScikitLearnDevelopers:2024} demonstrates how to implement meta
+cross-validation (termed ``nested cross-validation'' in the
+documentation).
+
+The approach taken by \pkg{scikit-learn} for independent observations is
+similar to that in the default \code{cv()} method in the \pkg{cv}
+package. The approach for hierarchical data is similar, but not
+identical, to that in \pkg{cv} (as described in Section
+\ref{cross-validating-mixed-effects-models}). Our \pkg{cv} package
+doesn't currently provide for cross-validating time-series models, but
+we anticipate that the next version of the package will.
+
+\section*{Appendix: Computational
+notes}\label{appendix-computational-notes}
+\addcontentsline{toc}{section}{Appendix: Computational notes}
+
+\subsection*{A.1 Efficient computations for linear and generalized
+linear
+models}\label{a.1-efficient-computations-for-linear-and-generalized-linear-models}
+\addcontentsline{toc}{subsection}{A.1 Efficient computations for linear
+and generalized linear models}
 
@@
 sampled. Refitting the model in this manner for each fold is generally
-feasible when the number of folds in modest, but can be prohibitively
+feasible when the number of folds is modest, but can be prohibitively
 costly for leave-one-out cross-validation when the number of cases is
@@
 
-The weighted-least-squares (WLS) estimator of \(\boldsymbol{\beta}\) is
-\citep[see, e.g.,][Sec. 12.2.2]{Fox:2016} \footnote{This is a
+The weighted-least-squares (``WLS'') estimator of \(\boldsymbol{\beta}\)
+is \citep[see, e.g.,][Sec. 12.2.2]{Fox:2016} \footnote{This is a
   definitional formula, which assumes that the model matrix
@@
   numerical instability when \(\mathbf{X}\) is ill-conditioned.
-  \code{lm()} uses the singular-value decomposition of the model matrix
-  to obtain computationally more stable results.} \[
-\mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} 
-  \mathbf{X}^T \mathbf{W} \mathbf{y}
+  \code{lm()} uses the QR decomposition of the model matrix with
+  pivoting to obtain computationally more stable results.} \[
+\mathbf{b}_{\mathrm{WLS}} = \left( \mathbf{X}^\top \mathbf{W} \mathbf{X} \right)^{-1}
+  \mathbf{X}^\top \mathbf{W} \mathbf{y}
 \]
@@
 \(\widehat{y}_{-i} = y_i - e_i/(1 - h_i)\) where
-\(h_i = \mathbf{x}^T_i \left( \mathbf{X}^T \mathbf{W} \mathbf{X} \right)^{-1} \mathbf{x}_i\)
-(the so-called ``hatvalue''). Here, \(\mathbf{x}^T_i\) is the \(i\)th
+\(h_i = \mathbf{x}^\top_i \left( \mathbf{X}^\top \mathbf{W} \mathbf{X} \right)^{-1} \mathbf{x}_i\)
+(the so-called ``hatvalue''). Here, \(\mathbf{x}^\top_i\) is the \(i\)th
 row of \(\mathbf{X}\), and \(\mathbf{x}_i\) is the \(i\)th row written
@@
 hatvalues are equal to 1, in which case the formula for
-\(\widehat{y}_{-i}\) requires division by 0. In this case, the
-``training'' set omitting the observation with hatvalue = 1 is
-rank-deficient and the predictors for the left-out case are outside the
-linear span of the predictors in the training set.
+\(\widehat{y}_{-i}\) requires division by 0. Then the ``training'' set
+omitting the observation with hatvalue = 1 is rank-deficient and the
+predictors for the left-out case are outside the linear span of the
+predictors in the training set.
 
 To compute cross-validated fitted values when the folds contain more
-than one case, we make use of the Woodbury matrix identify
+than one case, we make use of the Woodbury matrix identity
 \citep{Hager:1989}, \[
@@
 \mathbf{VA}^{-1}\mathbf{U} \right)^{-1} \mathbf{VA}^{-1}
-\] where \(\mathbf{A}\) is a nonsingular order-\(n\) matrix. We apply
+\] where \(\mathbf{A}\) is a nonsingular order-\(m\) matrix. We apply
 this result by letting \begin{align*}
-    \mathbf{A} &= \mathbf{X}^T \mathbf{W} \mathbf{X} \\
-    \mathbf{U} &= \mathbf{X}_\mathbf{j}^T \\
+    \mathbf{A} &= \mathbf{X}^\top \mathbf{W} \mathbf{X} \\
+    \mathbf{U} &= \mathbf{X}_\mathbf{j}^\top \\
     \mathbf{V} &= - \mathbf{X}_\mathbf{j} \\
@@
 \end{align*} where the subscript
-\(\mathbf{j} = (i_{j1}, \ldots, i_{jm})^T\) represents the vector of
-indices for the cases in the \(j\)th fold, \(j = 1, \ldots, k\). The
+\(\mathbf{j} = (i_{j1}, \ldots, i_{jm})^\top\) represents the vector of
+indices for the \(m = n_j \approx n/k\) cases in the \(j\)th fold. The
 negative sign in \(\mathbf{V} = - \mathbf{X}_\mathbf{j}\) reflects the
@@
 explicit matrix inversion and thus may be numerically unstable. The
-inverse of \(\mathbf{A} = \mathbf{X}^T \mathbf{W} \mathbf{X}\) is
+inverse of \(\mathbf{A} = \mathbf{X}^\top \mathbf{W} \mathbf{X}\) is
 available directly in the \code{"lm"} object, but the second term on the
@@
 
-\subsection{Computation of the bias-corrected CV criterion and
+\subsection*{A.2 Computation of the bias-corrected CV criterion and
 confidence
-intervals}\label{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}
+intervals}\label{a.2-computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}
+\addcontentsline{toc}{subsection}{A.2 Computation of the bias-corrected
+CV criterion and confidence intervals}
 
@@
   commonly employed CV criteria---such as the root-mean-squared error
-  (RMSE), median absolute error, and, for binary-regression models, the
-  complement of the area under the receiver operating characteristic
-  (ROC) curve---are not means of casewise components. That the RMSE and
-  median absolute error aren't means of casewise components is obvious;
-  for the complement of the area under the ROC curve, see the vignette
-  on extending the \pkg{cv} package.}
+  (``RMSE''), median absolute error, and, for binary-regression models,
+  the complement of the area under the receiver operating characteristic
+  (``ROC'') curve---are not means of casewise components. That the RMSE
+  and median absolute error aren't means of casewise components is
+  obvious; for the complement of the area under the ROC curve, see the
+  vignette on extending the \pkg{cv} package.}
 
 We divide the \(n\) cases into \(k\) folds of approximately
-\(n_j \approx n/k\) cases each, where \(n = \sum n_j\). As above, let
-\(\mathbf{j}\) denote the indices of the cases in the \(j\)th fold.
+\(n_j \approx n/k\) cases each, \(j = 1, \ldots, k\), where
+\(n = \sum n_j\). As above, let \(\mathbf{j}\) denote the indices of the
+cases in the \(j\)th fold.
 
@@
 \(\mathrm{CV} = \mathrm{CV}(\mathbf{y}, \widehat{\mathbf{y}}^{(-i)})\).
-Following \citet[pp.~293--295]{DavisonHinkley:1997}, the bias-adjusted
+Following \citet[293--295]{DavisonHinkley:1997}, the bias-adjusted
 cross-validation criterion is \[
@@
 
-\bibliography{cv.bib}
+\bibliography{jss5535.bib}
 

### 2025-05-09 scratch/JSS-submission/Final_submission/jss5535.Rmd
edit .Rmd, .tex, and update .pdf files

@@
 [^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here.
-Consider, for example, the scatterplot for log-transformed \code{mpg} and \code{horsepower}, produced by \code{plot(mpg $\sim$ horsepower, data=Auto, log="xy")} (execution of which is left to the reader).
+Consider, for example, the scatterplot for log-transformed \code{mpg} and \code{horsepower}, produced by \code{plot(mpg $\sim$ horsepower, data = Auto, log = "xy")} (execution of which is left to the reader).
 We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.
@@
 Figure \ref{fig:mpg-horsepower-polynomials} (b) shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$.
+Here $y_i$ is the response and $\widehat{y}_i$ the fitted value for the $i$th case.
 The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.
@@
 The final model, (4) \code{~ 1 + I(x - xm)}, uses an intercept and the centered-within-group variable \code{x - xm}.
-We thus fit four models to four data sets for a total of 16 models.
+We thus fit four models to each of the four data sets for a total of 16 models.
 
@@
 In Section \ref{comparing-competing-models}, following @JamesEtAl:2021[Secs.
-5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of \code{mpg} to \code{horsepower} for the \code{Auto} data, saving the 10 models, named \code{m.1} through \code{m.10}, in the list \code{mlist}.
+5.1, 5.3], we fit polynomial regressions up to degree 10 to the relationship of \code{mpg} to \code{horsepower} for the \code{Auto} data, saving the 10 models, named \code{m.1} through \code{m.10}, in the object \code{mlist}.
 We then used \code{cv()} to compare the cross-validated MSE for the 10 models, discovering that the 7th degree polynomial had the smallest MSE (by a small margin).
@@
 * \code{Blair}, \code{Hague}, and \code{Kennedy}, ratings of the leaders of the Labour, Conservative, and Liberal Democratic parties, on 1 to 5 scales.
-* \code{Europe}, an 11-point scale on attitude towards European integration, with high scores representing "Euro-skepticism."
+* \code{Europe}, an 11-point scale for attitude towards European integration, with high scores representing "Euro-skepticism."
 * \code{political.knowledge}, knowledge of the parties' positions on European integration, with scores from 0 to 3.
@@
 ```
-As political knowledge increases, voters tend to align their votes more closely with the party positions on European integration: The Conservative Party was relatively Euro-skeptic, while Labour and the Liberal Democrats were more supportive of the UK's participation in the EU.
+As political knowledge increased, voters tended to align their votes more closely with the party positions on European integration: The Conservative Party was relatively Euro-skeptic, while Labour and the Liberal Democrats were more supportive of the UK's participation in the EU.
 
@@
 The approach for hierarchical data is similar, but not identical, to that in \pkg{cv} (as described in Section \ref{cross-validating-mixed-effects-models}).
-Our \pkg{cv} package doesn't currently provide for cross-validating time-series models, but we anticipate that the next version of the package will.
+Our \pkg{cv} package doesn't currently provide for cross-validating time-series models, but we anticipate that a future version of the package will.
 

### 2025-05-09 scratch/JSS-submission/Final_submission/jss5535.tex
edit .Rmd, .tex, and update .pdf files

@@
 substantially based on the random division of the data, particularly in
-smaller samples. See \citet[Sec. 5.3]{Harrell:2015} for cogent remarks
+smaller samples. See \citet[Sec.~5.3]{Harrell:2015} for cogent remarks
 about data-division and cross-validation.
@@
 The \code{cv()} function is introduced in the context of a preliminary
-example in Section \ref{preliminary-example-polynomial-regression} of
+example in Section~\ref{preliminary-example-polynomial-regression} of
 the paper.
@@
   argument. More generally, the \pkg{cv} package is designed to be
-  extensible, as discussed in Section \ref{extending-the-cv-package}.
+  extensible, as discussed in Section~\ref{extending-the-cv-package}.
 \end{itemize}
@@
 
-{\centering \subfloat[\label{fig:mpg-horsepower-polynomials-1}]{\includegraphics[width=0.45\linewidth]{Figures/mpg-horsepower-polynomials-1} }~~~\subfloat[\label{fig:mpg-horsepower-polynomials-2}]{\includegraphics[width=0.45\linewidth]{Figures/mpg-horsepower-polynomials-2} }
+{\centering \subfloat[\label{fig:mpg-horsepower-polynomials-1}]{\includegraphics[width=0.45\linewidth]{jss5535_files/figure-latex/mpg-horsepower-polynomials-1} }~~~\subfloat[\label{fig:mpg-horsepower-polynomials-2}]{\includegraphics[width=0.45\linewidth]{jss5535_files/figure-latex/mpg-horsepower-polynomials-2} }
 
@@
   log-transformed \code{mpg} and \code{horsepower}, produced by
-  \code{plot(mpg $\sim$ horsepower, data=Auto, log="xy")} (execution of
-  which is left to the reader). We revisit the \code{Auto} data in
-  Section \ref{cross-validating-model-specification}.} Polynomial fits
+  \code{plot(mpg $\sim$ horsepower, data = Auto, log = "xy")} (execution
+  of which is left to the reader). We revisit the \code{Auto} data in
+  Section~\ref{cross-validating-model-specification}.} Polynomial fits
 for \(p = 1\) to \(5\) are shown in Figure
@@
 
-Figure \ref{fig:mpg-horsepower-polynomials} (b) shows two measures of
+Figure~\ref{fig:mpg-horsepower-polynomials} (b) shows two measures of
 estimated (squared) error as a function of polynomial-regression degree:
@@
 \(\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2\).
-The former necessarily declines with \(p\) (or, more strictly, can't
-increase with \(p\)), while the latter gets slightly larger for the
-largest values of \(p\), with the ``best'' value, by a small margin, for
-\(p = 7\).
+Here \(y_i\) is the response and \(\widehat{y}_i\) the fitted value for
+the \(i\)th case. The former necessarily declines with \(p\) (or, more
+strictly, can't increase with \(p\)), while the latter gets slightly
+larger for the largest values of \(p\), with the ``best'' value, by a
+small margin, for \(p = 7\).
 
@@
 
-{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{Figures/polynomial-regression-CV-graph-2-1} }~~~\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{Figures/polynomial-regression-CV-graph-2-2} }
+{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{jss5535_files/figure-latex/polynomial-regression-CV-graph-2-1} }~~~\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{jss5535_files/figure-latex/polynomial-regression-CV-graph-2-2} }
 
@@
 the literature on mixed-effects models. The HSB data are used by
-\citet[Sec. 7.2.2]{FoxWeisberg:2019} to illustrate the application of
+\citet[Sec.~7.2.2]{FoxWeisberg:2019} to illustrate the application of
 linear mixed models to hierarchical data, and we'll closely follow their
@@
 and \code{cses}, which is the individual students' SES centered at their
-school means. For details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.
+school means. For details, see \citet[Sec.~7.2.2]{FoxWeisberg:2019}.
 
@@
 regression line of patient centroids. The resulting data sets, shown in
-Figure \ref{fig:plot1}, differ in the between-patient variation of
+Figure~\ref{fig:plot1}, differ in the between-patient variation of
 patient centroids from a common between-patient regression line,
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{Figures/plot1-1}
+{\centering \includegraphics[width=0.75\linewidth]{jss5535_files/figure-latex/plot1-1}
 
@@
 centered-within-group variable \code{x - xm}. We thus fit four models to
-four data sets for a total of 16 models.
+each of the four data sets for a total of 16 models.
 
@@
 patients). The fixed- and random-effects predictions from these models
-are displayed in Figure \ref{fig:plot-fits-combined} along with a
+are displayed in Figure~\ref{fig:plot-fits-combined} along with a
 summary of the data using 50\% concentration ellipses (see the top row
@@
 
-{\centering \includegraphics[width=1\linewidth]{Figures/plot-fits-combined-1}
+{\centering \includegraphics[width=1\linewidth]{jss5535_files/figure-latex/plot-fits-combined-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{Figures/cross-validation-data-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{jss5535_files/figure-latex/cross-validation-data-plot-1}
 
@@
 our observations based on graphing predictions from the various models
-(in Figure \ref{fig:plot-fits-combined} on page
+(in Figure~\ref{fig:plot-fits-combined} on page
 \pageref{fig:plot-fits-combined}), and they illustrate the desirability
@@
 
-As \citet[Sec. 7.10.2: ``The Wrong and Right Way to Do
+As \citet[Sec.~7.10.2: ``The Wrong and Right Way to Do
 Cross-validation'']{HastieTibshiraniFriedman:2009} explain, if the whole
@@
 
-\subsection{Example: Data transformation and predictor selection for the
-Auto
+\subsection[Example: Data transformation and predictor selection for the Auto
+data]{Example: Data transformation and predictor selection for the
+\code{Auto}
 data}\label{example-data-transformation-and-predictor-selection-for-the-auto-data}
@@
 examine the pairwise relationships among the numeric variables in the
-data set (Figure \ref{fig:Auto-explore-scatterplot-matrix}, produced by
+data set (Figure~\ref{fig:Auto-explore-scatterplot-matrix}, produced by
 the \code{scatterplotMatrix()} function in the \pkg{car} package); in
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{Figures/Auto-explore-scatterplot-matrix-1}
+{\centering \includegraphics[width=0.95\linewidth]{jss5535_files/figure-latex/Auto-explore-scatterplot-matrix-1}
 
@@
 
-{\centering \includegraphics[width=0.7\linewidth]{Figures/Auto-explore-cr-plots-1}
+{\centering \includegraphics[width=0.7\linewidth]{jss5535_files/figure-latex/Auto-explore-cr-plots-1}
 
@@
 
-Some background: As \citet[Sec. 8.2]{Weisberg:2014} explains, there are
+Some background: As \citet[Sec.~8.2]{Weisberg:2014} explains, there are
 technical advantages to having (numeric) predictors in linear regression
@@
 that may include negative values and zeros: the Box-Cox-with-negatives
-family and the Yeo-Johnson family; see \citet[Chap. 8]{Weisberg:2014}
-and \citet[Chap. 3]{FoxWeisberg:2019} for details. We proceed to
+family and the Yeo-Johnson family; see \citet[Chap.~8]{Weisberg:2014}
+and \citet[Chap.~3]{FoxWeisberg:2019} for details. We proceed to
 transform the numeric predictors in the \code{Auto} regression towards
@@
 Auto data]{\texorpdfstring{Example: Applying meta CV to polynomial
-regression for the Auto
+regression for the \code{Auto}
 data\footnote{What we call here ``meta CV'' has also been termed
@@
 
-In Section \ref{comparing-competing-models}, following \citet[Secs. 5.1,
+In Section~\ref{comparing-competing-models}, following \citet[Secs. 5.1,
 5.3]{JamesEtAl:2021}, we fit polynomial regressions up to degree 10 to
@@
 data, saving the 10 models, named \code{m.1} through \code{m.10}, in the
-list \code{mlist}. We then used \code{cv()} to compare the
+object \code{mlist}. We then used \code{cv()} to compare the
 cross-validated MSE for the 10 models, discovering that the 7th degree
@@
 model fit by the \code{multinom()} function in the \pkg{nnet} package
-\citep{VenablesRipley:2002}. We borrow an example from \citet[Sec.
-14.2.1]{Fox:2016}, with data from the British Election Panel Study on
+\citep{VenablesRipley:2002}. We borrow an example from \citet[Sec.~14.2.1]{Fox:2016},
+with data from the British Election Panel Study on
 vote choice in the 2001 UK election. Data for the example are in the
@@
 \item
-  \code{Europe}, an 11-point scale on attitude towards European
+  \code{Europe}, an 11-point scale for attitude towards European
   integration, with high scores representing ``Euro-skepticism.''
@@
 
-Figure \ref{fig:BEPS-plot} shows an ``effect plot,'' using the the
+Figure~\ref{fig:BEPS-plot} shows an ``effect plot,'' using the the
 \pkg{effects} package \citep{FoxWeisberg:2019} to visualize the
@@
 
-{\centering \includegraphics{Figures/BEPS-plot-1}
+{\centering \includegraphics{jss5535_files/figure-latex/BEPS-plot-1}
 
@@
 
-As political knowledge increases, voters tend to align their votes more
-closely with the party positions on European integration: The
+As political knowledge increased, voters tended to align their votes
+more closely with the party positions on European integration: The
 Conservative Party was relatively Euro-skeptic, while Labour and the
@@
 }
-<bytecode: 0x14fd342e8>
+<bytecode: 0x11188dee8>
 <environment: namespace:cv>
@@
 
-\subsection{Other R software}\label{other-r-software}
+\subsection[Other R software]{Other \proglang{R} software}\label{other-r-software}
 
@@
 
-\subsection{Comparing cv to other software for cross-validation: SAS,
-Stata, and
-Python}\label{comparing-cv-to-other-software-for-cross-validation-sas-stata-and-python}
+\subsection[Comparing cv to other software for cross-validation: SAS,
+Stata, and Python]
+{Comparing cv to other software for cross-validation: \proglang{SAS},
+\proglang{Stata}, and \proglang{Python}}
+\label{comparing-cv-to-other-software-for-cross-validation-sas-stata-and-python}
 
@@
 
-\subsubsection{SAS}\label{sas}
+\subsubsection[SAS]{\proglang{SAS}}\label{sas}
 
@@
 
-\subsubsection{Stata}\label{stata}
+\subsubsection[Stata]{\proglang{Stata}}\label{stata}
 
@@
 
-\subsubsection{Python}\label{python}
+\subsubsection[Python]{\proglang{Python}}\label{python}
 
@@
 doesn't currently provide for cross-validating time-series models, but
-we anticipate that the next version of the package will.
+we anticipate that a future version of the package will.
 
@@
 The weighted-least-squares (``WLS'') estimator of \(\boldsymbol{\beta}\)
-is \citep[see, e.g.,][Sec. 12.2.2]{Fox:2016} \footnote{This is a
+is \citep[see, e.g.,][Sec.~12.2.2]{Fox:2016} \footnote{This is a
   definitional formula, which assumes that the model matrix
@@
 \code{glm()} computes the maximum-likelihood estimates for a generalized
-linear model by iterated weighted least squares \citep[see, e.g.,][Sec.
-6.12]{FoxWeisberg:2019}. The last iteration is therefore just a WLS fit
+linear model by iterated weighted least squares \citep[see, e.g.,][Sec.~6.12]{FoxWeisberg:2019}.
+The last iteration is therefore just a WLS fit
 of the ``working response'' on the model matrix using ``working

### 2025-05-09 scratch/JSS-submission/Final_submission/jss5535.tex
correct refs to /Figures dir in .tex file

@@
 
-{\centering \subfloat[\label{fig:mpg-horsepower-polynomials-1}]{\includegraphics[width=0.45\linewidth]{jss5535_files/figure-latex/mpg-horsepower-polynomials-1} }~~~\subfloat[\label{fig:mpg-horsepower-polynomials-2}]{\includegraphics[width=0.45\linewidth]{jss5535_files/figure-latex/mpg-horsepower-polynomials-2} }
+{\centering \subfloat[\label{fig:mpg-horsepower-polynomials-1}]{\includegraphics[width=0.45\linewidth]{Figures/mpg-horsepower-polynomials-1} }~~~\subfloat[\label{fig:mpg-horsepower-polynomials-2}]{\includegraphics[width=0.45\linewidth]{Figures/mpg-horsepower-polynomials-2} }
 
@@
 
-{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{jss5535_files/figure-latex/polynomial-regression-CV-graph-2-1} }~~~\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{jss5535_files/figure-latex/polynomial-regression-CV-graph-2-2} }
+{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{Figures/polynomial-regression-CV-graph-2-1} }~~~\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{Figures/polynomial-regression-CV-graph-2-2} }
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{jss5535_files/figure-latex/plot1-1}
+{\centering \includegraphics[width=0.75\linewidth]{Figures/plot1-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{jss5535_files/figure-latex/plot-fits-combined-1}
+{\centering \includegraphics[width=1\linewidth]{Figures/plot-fits-combined-1}
 
@@
 
-{\centering \includegraphics[width=0.75\linewidth]{jss5535_files/figure-latex/cross-validation-data-plot-1}
+{\centering \includegraphics[width=0.75\linewidth]{Figures/cross-validation-data-plot-1}
 
@@
 
-{\centering \includegraphics[width=0.95\linewidth]{jss5535_files/figure-latex/Auto-explore-scatterplot-matrix-1}
+{\centering \includegraphics[width=0.95\linewidth]{Figures/Auto-explore-scatterplot-matrix-1}
 
@@
 
-{\centering \includegraphics[width=0.7\linewidth]{jss5535_files/figure-latex/Auto-explore-cr-plots-1}
+{\centering \includegraphics[width=0.7\linewidth]{Figures/Auto-explore-cr-plots-1}
 
@@
 
-{\centering \includegraphics{jss5535_files/figure-latex/BEPS-plot-1}
+{\centering \includegraphics{Figures/BEPS-plot-1}
 

### 2025-05-11 scratch/JSS-submission/Final_submission/jss5535.tex
fix "the the" typo
add Georges's email

@@
     Toronto, Ontario, Canada\\
-  E-mail: \email\\
+  E-mail: \email{georges@yorku.ca}\\
   URL: \url{http://blackwell.math.yorku.ca/gmonette}\\~\\
@@
 
-Figure~\ref{fig:BEPS-plot} shows an ``effect plot,'' using the the
+Figure~\ref{fig:BEPS-plot} shows an ``effect plot,'' using the
 \pkg{effects} package \citep{FoxWeisberg:2019} to visualize the

### 2025-10-19 scratch/JSS-submission/Final_submission/jss5535.Rmd

@@
       | Toronto, Ontario, Canada
-    email: \email
+    email: \email{georges@yorku.ca}
     url: http://blackwell.math.yorku.ca/gmonette
@@
 
-In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model (cf., Figure \ref{fig:mpg-horsepower-polynomials} (b) on page \pageref{fig:mpg-horsepower-polynomials}), except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting.
+In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model (cf., Figure \ref{fig:mpg-horsepower-polynomials}~(b) on page \pageref{fig:mpg-horsepower-polynomials}), except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting.
 \newpage
@@
 
-Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the \code{Europe} $\times$ \code{political.knowledge} interaction in a "stacked-area" graph:
+Figure \ref{fig:BEPS-plot} shows an "effect plot," using the \pkg{effects} package [@FoxWeisberg:2019] to visualize the \code{Europe} $\times$ \code{political.knowledge} interaction in a "stacked-area" graph:
 ```{r BEPS-plot, fig.width=9, fig.height=5}

### 2025-10-19 scratch/JSS-submission/Revision-2/Fox-Monette-cv-package.Rmd

--- 2024-10-03 scratch/JSS-submission/Revision-1/Fox-Monette-cv-package.Rmd
@@
 
-The approach taken by \pkg{scikit-Learn} for independent observations is similar to that in the default \code{cv()} function in the \pkg{cv} package.
+The approach taken by \pkg{scikit-learn} for independent observations is similar to that in the default \code{cv()} function in the \pkg{cv} package.
 The approach for hierarchical data is similar, but not identical, to that in \pkg{cv} (as described in Section \ref{cross-validating-mixed-effects-models}).
